### 2021-12-16 paper.tex
Basic paper shell with some cites to get things started

\documentclass{article}
\usepackage[round]{natbib}
\usepackage{amsmath,amssymb,amsfonts}%
\usepackage{geometry}%
\usepackage{color}
\usepackage{graphicx}
\usepackage{authblk}
\usepackage{nameref}
\usepackage[right]{lineno}

\begin{document}

\linenumbers
\title{What is an Ancestral Recombination Graph?}
% First authors
\author{Authory McAuthor}
% Corresponding

\maketitle

\begin{abstract}
The ARG. Argh!
\end{abstract}

\textbf{Keywords:} Ancestral Recombination Graphs

\section*{Introduction}

\section*{The ARG as a stochastic process}
The coalescent
process~\citep{kingman1982coalescent,kingman1982genealogy,hudson1983testing,
tajima1983evolutionary}

The Ancestral Recombination Graph (ARG) was introduced by
Griffiths~\citep{griffiths1991two,griffiths1997ancestral}.
The ``big'' ARG~\citep{ethier1990two},
and the ``little ARG'' traversed by
Hudson's algorithm~\citep{hudson1983properties}.

\section*{The ARG as an encoding of genetic ancestry}

\citep[e.g.][]{minichiello2006mapping,mathieson2020ancestry}.

\bibliographystyle{plainnat}
\bibliography{paper}

\end{document}

### 2021-12-16 paper.tex
Added rough abstract

@@
 % First authors
-\author{Authory McAuthor}
+\author{Author McAuthory}
 % Corresponding
@@
 
+% JK: this is a rough first pass for a slightly different paper. Needs
+% substantial revision.
 \begin{abstract}
-The ARG. Argh!
+It has recently become possible to infer genetic ancestry in the presence of
+recombination at scale for the first time. This has created exciting
+possibilities, with many downstream applications becoming possible based on
+these inferred genealogies. The inferred genetic ancestries are usually
+referred to as Ancestral Recombination Graphs, or ARGs. Although initially well
+defined as a graphical representation of the coalescent with recombination
+stochastic process, the interpretation has become unclear as it is now
+also understood to represent a particular realisation of a genetic
+ancestry.
+Inference methods do not all infer the same information:
+although all output marginal trees along the genome, there is a great deal of
+variation in how much information is inferred and retained in the relationships
+between the trees. An important programme of work over the coming years is to
+develop and refine ancestry inference methods, assessing the relative strengths
+and weaknesses of the various approaches. Using the blanket term ARG hides the
+important differences in the approaches, and hampers our ability to compare and
+improve inference methods. In this paper we discuss the different approaches we
+may use to encode genetic ancestry, and of the differing amounts of information
+about the ancestral process that we can represent and hope to observe. We
+classify existing inference methods according to the type of ancestry they
+infer, and discuss the strengths and weaknesses of the inferred structure for
+downstream applications.
 \end{abstract}

### 2021-12-16 illustrations/ARGweaver_example.py
Move stuff into "utils"

@@ -7,8 +7,7 @@ import networkx as nx
 import matplotlib.pyplot as plt
 
 # useful scripts in the top level dir
-sys.path.append(str(Path(__file__).parent.parent.absolute()))
-from constants import NODE_IS_RECOMB, NODE_IS_ALWAYS_UNARY, NODE_IS_SOMETIMES_UNARY
+sys.path.append(str(Path(__file__).parent.parent.absolute() / "utils"))
 import convert
 import ts_process
 import argdraw
@@ -90,32 +89,9 @@ ts = ts_process.add_individuals_to_coalescence_nodes(ts) # optional - can help i
 
 G = ts_process.to_networkx_graph(ts)
 
-nodes_at_time = collections.defaultdict(list)
-colour_map = []
-for nd in G.nodes(data=True):
-    nodes_at_time[nd[1]["time"]].append(nd[0])
-    colour = "lightgreen"
-    if nd[1]["flags"] & NODE_IS_RECOMB:
-        colour = "red"
-    elif nd[1]["flags"] & NODE_IS_ALWAYS_UNARY:
-        colour = "cyan"
-    elif nd[1]["flags"] & NODE_IS_SOMETIMES_UNARY:
-        colour = "lightseagreen"
-    colour_map.append(colour)
-        
-    
-# Turn into graphviz
-A = nx.nx_agraph.to_agraph(G)
-# First cluster all nodes at the same times (probably mostly samples)
-for t, nodes in nodes_at_time.items():
-    if len(nodes) > 1:
-        A.add_subgraph(nodes, level="same", name=f"cluster_t{t}")
-# We could also cluster nodes from a single individual together here
-
-# Get the positions from graphviz
-A.layout(prog="dot")
-pos = {n: [float(x) for x in A.get_node(n).attr["pos"].split(",")] for n in G.nodes()}
+colour_map = argdraw.nx_ts_colour_map(G)
+pos = argdraw.nx_get_dot_pos(G)
 
 fig = plt.figure(1, figsize=(10, 15))
-argdraw.draw_with_curved_multi_edges(G, pos, colour_map, 20)
+argdraw.nx_draw_with_curved_multi_edges(G, pos, colour_map, 20)
 plt.savefig(Path(__file__).parent / 'ARGweaver_arg.pdf')  

### 2021-12-16 paper.tex

@@
 
+% HYW: this is a rough first pass, aiming to be concise
+For a long time, the Ancestral Recombination Graph (ARG) has been an object of fascination
+to those interested in theoretical evolutionary genetics, as it captures all knowable genetic
+history of a collection of individual genomes. Recently advances in our ability to infer ARGs
+has led to a resurgence of interest in this field. However, these have also given rise to certain 
+terminological confusion, which we aim to clear up in this paper.
+
+One major source of confusion is the use of the term "ARG" to refer to both a backwards-in-time
+generative process (Griffiths) and a structure which can be created by that process \citep[e.g.][]{minichiello2006mapping,mathieson2020ancestry}.
+
 \section*{The ARG as a stochastic process}
@@
 
+We suggest that the process definition should be called the "coalescent with recombination" (CwR), reserving the term ARG to refer to the graph structure itself.
+
 \section*{The ARG as an encoding of genetic ancestry}
@@
 \citep[e.g.][]{minichiello2006mapping,mathieson2020ancestry}.
+
+% HYW: this is a rough first pass. We should take care not to assume that the tree sequence
+% approach is the only encoding
+
+This graph *structure*, which can in fact be generated either forwards or backwards in time,
+consists of a collection of genomes (represented by *nodes*) connected by paths of genetic
+inheritance (*edges*). Since genetic inheritance is unidirectional, the edges are *directed*,
+with a "parent" and "child" node. Moreover, there can be no cycles in the graph, therefore the
+ARG is technically a form of "directed acyclic graph". One way in this can be ensured is to
+associate nodes with specific *times*, and for parent nodes to be older than their children.
+
+% HYW: NB: word carefullyL it should also be possible to encode ARGs even if there is not a single fixed coordinate system
+
+The process of inheritance means that an ARG is a very specific form of DAG. In particular,
+although a genome may have more than one parent, because inheritance is particulate,
+any particular base pair in a genome only has a *single* parent. Therefore, the paths of
+inheritance for a *specific position* in the genome of a set of individuals are described by
+a *tree*. Assuming that we can align the genomes in question, we can work our way along
+the alignment and by examining the graph we can extract a "local tree" at each position.
+
+To allow this extraction to occur, the graph must be annotated in some way, allowing us to
+decide, at any node with more than one parent, which of the parents was the source of
+each base pair in the genome. We call any such node a *recombination node*. Fortunately,
+the biological details of inheritance mean that this information can be encoded in a succinct
+form: we do not have to individually assign each base pair a separate parent. For example,
+if we assume that the ARG is describing a single chromosome, and there is only one
+recombination per chromosome, we can assign a single *breakpoint* to each recombination
+node: base pairs to the left of the breakpoint come from one parent, base pairs to the right
+come from the other.
+
+This particular ARG encoding, although common, is somewhat restrictive: not only can
+it not be extended to multiple chromosomes or multiple recombinations along a chromosome,
+but it cannot easily incorporate the very common process of gene conversion.
+
+% HYW: should discuss the link to biology & mitosis/meiosis: a "node" is strictly a cell - it is shorthand to associate it with an "individual"
 

### 2021-12-17 paper.tex

--- 2021-12-16 paper.tex
@@
 
+% HYW: this is a rough first pass, aiming to be concise. We should flash out the "recombination" aspect of ARGs here a bit, I think
+
+For a long time, the Ancestral Recombination Graph (ARG) has been an object of fascination
+to those interested in theoretical evolutionary genetics, as it captures all knowable genetic
+history of a collection of individual genomes. Recently advances in our ability to infer ARGs
+has led to a resurgence of interest in this field. However, these have also given rise to certain 
+terminological confusion, which we aim to clear up in this paper.
+
+One major source of confusion is the use of the term "ARG" to refer to both a backwards-in-time
+generative process (Griffiths) and a structure which can be created by that process \citep[e.g.][]{minichiello2006mapping,mathieson2020ancestry}.
+
 \section*{The ARG as a stochastic process}
@@
 
-\section*{The ARG as an encoding of genetic ancestry}
+We suggest that the process definition should be called the "coalescent with recombination" (CwR), reserving the term ARG to refer to the graph structure itself.
+
+\section*{The ARG as a structure}
 
 \citep[e.g.][]{minichiello2006mapping,mathieson2020ancestry}.
+
+% HYW: this is a rough first pass. We should take care not to assume that the tree sequence
+% approach is the only encoding
+
+The other use of the the term ARG, and the one for which we suggest the term be reserved, 
+is to describe the *structure* resulting from the genetic process of inheritance. This structure, 
+like any other graph, consists of *nodes* connected by *edges*. In the context of genetics, the
+nodes represent (haploid) genomes, and the edges represent paths of inheritance. Although for
+theoretical purposes a graph like this can be created by backwards-in-time simulations using
+the CwR, in reality it is created forwards-in-time, as a population evolves.
+
+Since genetic inheritance is unidirectional, the edges in this graph structure are *directed*,
+with a "parent" and "child" node. Moreover, there is a strict temporal order, so that children
+cannot be their own parents: in other words, there are no cycles in the graph, and it is
+technically a form of "directed acyclic graph" (DAG).
+
+% HYW: NB: worded carefully below, as it should also be possible to encode ARGs even if there is not a single fixed coordinate system
+
+There is one key feature which distinguishes an ARG for other graphs. Although a genome
+(graph node) may have more than one parent, the particulate nature of genetic inheritance
+means that each letter in the genome can only come from a *single* parent. Hence if we
+compare the equivalent letters from multiple extant genomes, their paths of inheritance must
+form a *tree* (which, incidentally, is why phylogenetics plays such an important role in evolutionary
+theory). As we shall argue, the ability to generate these "local trees" from an ARG is of key
+importance.
+
+When describing an ARG, the primary issue, from a practical point of view, is how to annotate
+the graph with the details of which piece of genome has come from which parent. There
+are two possibilities: the annotations can be associated either with the nodes, or with the edges.
+
+\subsection*{Node annotated recombination}
+
+In the original (Griffiths) formulation, nodes which have multiple parents are annotated with a
+*breakpoint*, representing a genomic position.
+The genome lying to the left of that position is then taken to come from one parent, while the genome lying to the right is taken to come from the other parent.
+This is an elegantly concise encoding of the parental source, and is used, for example in the ARGweaver .arg format.
+However, it suffers from three principal drawbacks:
+
+1. It assumes only a single breakpoint per node: problematic for representing e.g. gene conversion or 
+multiple chromosomes (but this could be circumvented by having multiple breakpoints)
+2. Nodes can only have a maximum of 2 parents. At first sight, this seems a reasonable limitation,
+but it turns out that it can be helpful to collapse multiple nodes into one, which potentially results
+in more than two parents for any given node (see section XXXX)
+3. Generating local trees from such an encoding is time consuming (XXX describe *how* time consuming - i,e. scaling properties with number of nodes - and why this is, theoretically)
+
+\subsection*{Edge annotated recombination}
+
+An alternative approach to annotating nodes with the parent of origin of each section of genome is
+to annotate each edge with the genomic region which is inherited via that route. This approach
+was first introduced by Kelleher et al (cite msprime paper) and now forms the basis of a comprehensive
+software library.
+
+% HYW: should discuss the link to biology & mitosis/meiosis: a "node" is strictly a cell - it is shorthand to associate it with an "individual"
 

### 2022-01-10 paper.tex
Add initial ASG refs

@@
 
-\section*{The ARG as a stochastic process}
+\section*{Ancestral graphs and stochastic processes}
 The coalescent
@@
 
-We suggest that the process definition should be called the "coalescent with recombination" (CwR), reserving the term ARG to refer to the graph structure itself.
+In the theoretical literature, ancestral graphs such as the ARG refer to the
+ancestral \emph{process} rather than the outcome of a process.
+For example, the Ancestral Selection
+Graph~\citep{krone1997ancestral,neuhauser1999ancestral} models selection
+as a graph with virtual lineages [etc etc.]. Importantly, the graph structure
+of the ASG does \emph{not} correspond to a realised ancestry, but rather
+models possible outcomes.
 

### 2022-01-10 paper.tex
More ancestral graph refs

@@
 \section*{Ancestral graphs and stochastic processes}
+
+% See
 The coalescent
@@
 ancestral \emph{process} rather than the outcome of a process.
-For example, the Ancestral Selection
+\begin{itemize}
+\item Ancestral Selection
 Graph~\citep{krone1997ancestral,neuhauser1999ancestral} models selection
@@
 models possible outcomes.
+Other papers possibly worth citing:\citep{etheridge2009coalescent}
+\item Ancestral Gene Transfer Graph~\citep{baumdicker2014infinitely}
+\end{itemize}
 

### 2022-01-10 paper.tex
Outline of ARG inference methods

@@
 
-For a long time, the Ancestral Recombination Graph (ARG) has been an object of fascination
-to those interested in theoretical evolutionary genetics, as it captures all knowable genetic
-history of a collection of individual genomes. Recently advances in our ability to infer ARGs
-has led to a resurgence of interest in this field. However, these have also given rise to certain
+For a long time, the Ancestral Recombination Graph (ARG) has been an object of
+fascination to those interested in theoretical evolutionary genetics, as it
+captures all knowable genetic history of a collection of individual genomes.
+Recently advances in our ability to infer ARGs has led to a resurgence of
+interest in this field. However, these have also given rise to certain
 terminological confusion, which we aim to clear up in this paper.
 
-One major source of confusion is the use of the term "ARG" to refer to both a backwards-in-time
-generative process (Griffiths) and a structure which can be created by that process \citep[e.g.][]{minichiello2006mapping,mathieson2020ancestry}.
+One major source of confusion is the use of the term "ARG" to refer to both a
+backwards-in-time generative process (Griffiths) and a structure which can be
+created by that process
+\citep[e.g.][]{minichiello2006mapping,mathieson2020ancestry}.
 
@@
 
-% See
+% See https://github.com/tskit-dev/what-is-an-arg-paper/discussions/13
+% for discussions on this section
+
 The coalescent
@@
 
-The other use of the the term ARG, and the one for which we suggest the term be reserved,
-is to describe the *structure* resulting from the genetic process of inheritance. This structure,
-like any other graph, consists of *nodes* connected by *edges*. In the context of genetics, the
-nodes represent (haploid) genomes, and the edges represent paths of inheritance. Although for
-theoretical purposes a graph like this can be created by backwards-in-time simulations using
-the CwR, in reality it is created forwards-in-time, as a population evolves.
+The other use of the the term ARG, and the one for which we suggest the term be
+reserved, is to describe the *structure* resulting from the genetic process of
+inheritance. This structure, like any other graph, consists of *nodes*
+connected by *edges*. In the context of genetics, the nodes represent (haploid)
+genomes, and the edges represent paths of inheritance. Although for theoretical
+purposes a graph like this can be created by backwards-in-time simulations
+using the CwR, in reality it is created forwards-in-time, as a population
+evolves.
 
-Since genetic inheritance is unidirectional, the edges in this graph structure are *directed*,
-with a "parent" and "child" node. Moreover, there is a strict temporal order, so that children
-cannot be their own parents: in other words, there are no cycles in the graph, and it is
-technically a form of "directed acyclic graph" (DAG).
+Since genetic inheritance is unidirectional, the edges in this graph structure
+are *directed*, with a "parent" and "child" node. Moreover, there is a strict
+temporal order, so that children cannot be their own parents: in other words,
+there are no cycles in the graph, and it is technically a form of "directed
+acyclic graph" (DAG).
 
@@
 
-There is one key feature which distinguishes an ARG for other graphs. Although a genome
-(graph node) may have more than one parent, the particulate nature of genetic inheritance
-means that each letter in the genome can only come from a *single* parent. Hence if we
-compare the equivalent letters from multiple extant genomes, their paths of inheritance must
-form a *tree* (which, incidentally, is why phylogenetics plays such an important role in evolutionary
-theory). As we shall argue, the ability to generate these "local trees" from an ARG is of key
-importance.
+There is one key feature which distinguishes an ARG for other graphs. Although
+a genome (graph node) may have more than one parent, the particulate nature of
+genetic inheritance means that each letter in the genome can only come from a
+*single* parent. Hence if we compare the equivalent letters from multiple
+extant genomes, their paths of inheritance must form a *tree* (which,
+incidentally, is why phylogenetics plays such an important role in evolutionary
+theory). As we shall argue, the ability to generate these "local trees" from an
+ARG is of key importance.
 
-When describing an ARG, the primary issue, from a practical point of view, is how to annotate
-the graph with the details of which piece of genome has come from which parent. There
-are two possibilities: the annotations can be associated either with the nodes, or with the edges.
+When describing an ARG, the primary issue, from a practical point of view, is
+how to annotate the graph with the details of which piece of genome has come
+from which parent. There are two possibilities: the annotations can be
+associated either with the nodes, or with the edges.
 
@@
 
-In the original (Griffiths) formulation, nodes which have multiple parents are annotated with a
-*breakpoint*, representing a genomic position.
-The genome lying to the left of that position is then taken to come from one parent, while the genome lying to the right is taken to come from the other parent.
-This is an elegantly concise encoding of the parental source, and is used, for example in the ARGweaver .arg format.
-However, it suffers from three principal drawbacks:
+In the original (Griffiths) formulation, nodes which have multiple parents are
+annotated with a breakpoint, representing a genomic position. The genome
+lying to the left of that position is then taken to come from one parent, while
+the genome lying to the right is taken to come from the other parent. This is
+an elegantly concise encoding of the parental source, and is used, for example
+in the ARGweaver .arg format. However, it suffers from three principal
+drawbacks:
 
-1. It assumes only a single breakpoint per node: problematic for representing e.g. gene conversion or
-multiple chromosomes (but this could be circumvented by having multiple breakpoints)
-2. Nodes can only have a maximum of 2 parents. At first sight, this seems a reasonable limitation,
-but it turns out that it can be helpful to collapse multiple nodes into one, which potentially results
-in more than two parents for any given node (see section XXXX)
-3. Generating local trees from such an encoding is time consuming (XXX describe *how* time consuming - i,e. scaling properties with number of nodes - and why this is, theoretically)
+1. It assumes only a single breakpoint per node: problematic for representing
+e.g. gene conversion or multiple chromosomes (but this could be circumvented by
+having multiple breakpoints)
+2. Nodes can only have a maximum of 2 parents. At
+first sight, this seems a reasonable limitation, but it turns out that it can
+be helpful to collapse multiple nodes into one, which potentially results in
+more than two parents for any given node (see section XXXX)
+3. Generating local
+trees from such an encoding is time consuming (XXX describe *how* time
+consuming - i,e. scaling properties with number of nodes - and why this is,
+theoretically)
 
@@
 
-An alternative approach to annotating nodes with the parent of origin of each section of genome is
-to annotate each edge with the genomic region which is inherited via that route. This approach
-was first introduced by Kelleher et al (cite msprime paper) and now forms the basis of a comprehensive
-software library.
+An alternative approach to annotating nodes with the parent of origin of each
+section of genome is to annotate each edge with the genomic region which is
+inherited via that route. This approach was first introduced by Kelleher et al
+(cite msprime paper) and now forms the basis of a comprehensive software
+library.
 
-% HYW: should discuss the link to biology & mitosis/meiosis: a "node" is strictly a cell - it is shorthand to associate it with an "individual"
+% HYW: should discuss the link to biology & mitosis/meiosis: a "node" is
+% strictly a cell - it is shorthand to associate it with an "individual"
+
+\section*{ARG inference methods}
+Using the terminology developed here to classify ARGs, we review methods developed
+to infer ARGs. We focus our attention on some recent methods and discuss the
+properties of the ARGs inferred.
+
+\begin{itemize}
+\item Early methods \citep{hein1990reconstructing}
+\item Those sampling from the coalescent with recombination
+\citep{griffiths1996ancestral,kuhner2000maximum,fearnhead2001estimating},
+or SMC
+\citep{rasmussen2014genome}
+\item Hueristic methods~\citep{song2005efficient,minichiello2006mapping,
+parida2008estimating,fallon2013acg,mirzaei2016rent}
+\item Discussions of the complexity of inferring ARGs (but we need to
+dig into this: is this the complexity of sampling from the coalescent,
+finding a ``min ARG'', or something else?
+\citep{bordewich2005computational,wang2001perfect,hein1990reconstructing,song2005constructing}
+\item Recent methods: Relate, tsinfer, tsdate, etc.
+\end{itemize}
 

### 2022-01-10 paper.tex
Placeholders for figures.

@@
 
+\begin{figure}
+\vspace{5em}
+\caption{\label{fig-pedigree-and-arg}
+The (unannotated) ARG is closely related to the pedigree
+graph, where each ``individual'' node is replace by two ``genome''
+nodes, and edges between parents and children are only included if there is
+transfer of ancestral material. We usually only include nodes in an ARG
+that affect our sample, so that, for example, individuals that are ancestral
+to only one sampled genome are not included.}
+\end{figure}
+
 % HYW: this is a rough first pass. We should take care not to assume that the tree sequence
@@
 
+\begin{figure}
+\vspace{5em}
+% NB: not clear if this figure is better off on its own or should be
+% merged with the pedigree one.
+\caption{\label{fig-arg-annotations}
+The specific path through the ARG for a given genome position (and therefore
+the local genealogical tree) cannot be determined without annotations
+to the graph. (A) The classical Griffiths approach annotates recombination
+nodes with the corresponding breakpoint. (B) We can equivalently
+annotate the \emph{edges} with the genomic intervals carrying ancestral
+material. Edge annotations are somewhat more general, and have
+significant computational advantages.}
+\end{figure}
+
 \subsection*{Node annotated recombination}
@@
 
+\begin{figure}
+\vspace{5em}
+\caption{\label{fig-recombination-nodes}
+Recombination nodes are optional in an edge-annotated ARG. (A) A simulated
+ARG including all recombination nodes. (B) The same ARG with recombination
+nodes removed. (C) The corresponding sequence of trees.
+}
+\end{figure}
+
 \section*{ARG inference methods}
@@
 
+\begin{figure}
+\vspace{5em}
+\caption{\label{fig-inferred-args}
+ARGs inferred by different methods from a small dataset. Notes about these
+ARGs.
+}
+\end{figure}
+
 \bibliographystyle{plainnat}

### 2022-01-10 illustrations/ARG_recomb_node_deletion.py
Figure for an ARG with no recomb nodes
Fixes #17

@@ -1,0 +1,113 @@
+import collections
+import io
+import itertools
+import sys
+from pathlib import Path
+
+
+import msprime
+import networkx as nx
+import numpy as np
+import tskit
+import matplotlib.pyplot as plt
+
+outfile = "ARG_recomb_node_deletion"
+
+# useful scripts in the top level dir
+sys.path.append(str(Path(__file__).parent.parent.absolute() / "utils"))
+import convert
+import ts_process
+import argdraw
+
+
+def positions_from_ts(ts, pos=None, return_graph=False):
+    ts_arg = ts_process.flag_unary_nodes(ts)
+    G = ts_process.to_networkx_graph(ts_arg)
+    if pos is None:
+        pos = argdraw.nx_get_dot_pos(G, add_invisibles=False)
+    return (pos, G) if return_graph else pos
+    
+
+def plot_ts_arg(ts, ax, labels=None, pos=None, arrows=False):
+    pos, G = positions_from_ts(ts, pos, return_graph=True)
+    col_map = argdraw.nx_ts_colour_map(G)
+
+    argdraw.nx_draw_with_curved_multi_edges(
+        G, pos, col_map, curve_scale=30, ax=ax, labels=labels)
+
+# SIMULATE
+ts = ts_process.convert_to_single_rec_node(
+    msprime.sim_ancestry(
+        3,
+        sequence_length=1e4,
+        population_size=1e4,
+        recombination_rate=1e-8,
+        record_full_arg=True, 
+        random_seed=76, # by trial and error, a seed of 76 gives a nice small example
+    )
+)
+
+node_pos = positions_from_ts(ts)
+
+# relabel the nodes to get the order right
+tables = ts.dump_tables()
+node_map = np.arange(ts.num_nodes, dtype=tables.edges.parent.dtype)
+sample_ids = ts.samples()
+assert np.all(sample_ids == np.arange(ts.num_samples))
+node_map[np.argsort([node_pos[s][0] for s in sample_ids])] = sample_ids
+tables.nodes.clear()
+for u in node_map:
+    tables.nodes.append(ts.tables.nodes[u])
+tables.edges.parent = node_map[tables.edges.parent]
+tables.edges.child = node_map[tables.edges.child]
+tables.sort()
+ts=tables.tree_sequence()
+
+# Get the positions again, as the nodes are now if a different order
[…]

### 2022-01-10 utils/argdraw.py

@@ -1,5 +1,7 @@
 """
 Some networkx utilities for drawing an ARG
+
+Some of this should get tidied up and moved to https://github.com/tskit-dev/tsviz
 """
 import collections
 import itertools
@@ -27,15 +29,19 @@ def nx_ts_colour_map(G, flags_attribute_name="flags"):
         colour_map.append(colour)
     return colour_map
 
-def nx_draw_with_curved_multi_edges(G, pos, colour_map, curve_scale=1, node_labels=None):
+def nx_draw_with_curved_multi_edges(
+    G, pos, colour_map, ax=None, labels=None, arrows=False, curve_scale=1
+):
     """
     networkx matlib plots show all edges as straight lines by default.
     This function curves the edges if there is more than one edge with identical parent/child values
     At the moment curve_scale is a hack to adjust the fact that we don't know the measurement scale on
     the x or the y axis.
     """
-    nx.draw_networkx_nodes(G, pos, node_color=colour_map)
-    nx.draw_networkx_labels(G, pos, labels=node_labels)
+    nx.draw(
+        G, pos, node_color=colour_map,
+        node_shape='o', node_size=200, font_size=9, labels=labels,
+        ax=ax, edgelist=[])
 
 
     for (start, end), edge_iter in itertools.groupby(G.edges()):
@@ -52,9 +58,13 @@ def nx_draw_with_curved_multi_edges(G, pos, colour_map, curve_scale=1, node_labe
                 G,
                 pos,
                 edgelist=[e],
-                connectionstyle=f'arc3, rad = {curve_prop * curvature}')
+                connectionstyle=f'arc3, rad = {curve_prop * curvature}',
+                arrowstyle="-|>" if arrows else "-",
+                ax=ax,
+            )
 
-def nx_get_dot_pos(G, time_attribute_name="time"):
+
+def nx_get_dot_pos(G, add_invisibles=False, time_attribute_name="time"):
     """
     Layout using graphviz's "dot" algorithm and return a dict of positions in the
     format required by networkx. We assume that the nodes have a "time" attribute
@@ -68,8 +78,15 @@ def nx_get_dot_pos(G, time_attribute_name="time"):
     for t, nodes in nodes_at_time.items():
         if len(nodes) > 1:
             A.add_subgraph(nodes, level="same", name=f"cluster_t{t}")
-    
-    # We could also cluster nodes from a single individual together here
+    if add_invisibles:
+        # Must add in "invisible" edges between different levels, to stop them clustering
[…]

### 2022-01-11 paper.tex

@@
 \begin{figure}
+\centering
 \vspace{5em}
-\caption{\label{fig-recombination-nodes}
-Recombination nodes are optional in an edge-annotated ARG. (A) A simulated
-ARG including all recombination nodes. (B) The same ARG with recombination
-nodes removed. (C) The corresponding sequence of trees.
+\includegraphics[width=\linewidth]{illustrations/ARG_recomb_node_deletion}
+\caption{\label{fig-recombinati on-nodes}
+Recombination nodes are optional in an edge-annotated ARG. (a) A simulated
+ARG including all recombination nodes (red). (b) The same ARG having removed both
+recombination nodes and nodes of common ancestors in which no genetic
+lineages coalesce (cyan). (c) The sequence of trees corresponding to (b): if we were
+to plot out the trees corresponding to (a) we would have more trees, but the topologies
+and MRCA nodes at each genomic location would be identical. For clarity, nodes are
+plotted by ranked time rather than true time on the y-axis.
+% HYW: note that we also truncate nodes for portions in which they are unary,
+% so that nodes which are coalescent nodes over only some of their length
+% (blue-green) become coalescent nodes over their entire length (green). This
+% can, however, throw away some useful information, for example node 4 has
+% only a single parent edge in (a), but 2 parent edges (one of which appears
+% to bypass node 11) in (b). In fact, this edge should go through node 11, but
+% there may be no way to detect this from real data.
 }

### 2022-01-25 paper.tex
Initial theory + Monte Carlo writeup

@@
 % First authors
-\author{Author McAuthory}
+\author{Author McAuthorface}
 % Corresponding
@@
 recombination at scale for the first time. This has created exciting
-possibilities, with many downstream applications becoming possible based on
+possibilities, with many downstream applications based on
 these inferred genealogies. The inferred genetic ancestries are usually
@@
 although all output marginal trees along the genome, there is a great deal of
-variation in how much information is inferred and retained in the relationships
-between the trees. An important programme of work over the coming years is to
+variation in how much information of the relationships between the trees is inferred
+and retained. An important programme of work over the coming years is to
 develop and refine ancestry inference methods, assessing the relative strengths
@@
 
-One major source of confusion is the use of the term "ARG" to refer to both a
+One major source of confusion is the use of the term ``ARG" to refer to both a
 backwards-in-time generative process (Griffiths) and a structure which can be
@@
 
-The coalescent
-process~\citep{kingman1982coalescent,kingman1982genealogy,hudson1983testing,
-tajima1983evolutionary}
+The simplest prominent example of genetic ancestry generated by a stochastic process
+is the coalescent process \citep{kingman1982coalescent,kingman1982genealogy,
+hudson1983testing, tajima1983evolutionary}, which describes the random
+ancestral tree at one non-recombining locus obtained by sampling a relatively
+small number of individuals at random from a large population.
 
-The Ancestral Recombination Graph (ARG) was introduced by
-Griffiths~\citep{griffiths1991two,griffiths1997ancestral}.
-The ``big'' ARG~\citep{ethier1990two},
-and the ``little ARG'' traversed by
-Hudson's algorithm~\citep{hudson1983properties}.
+In a recombining organism, the genetic ancestry of a longer segment of DNA is 
+no longer described by a single tree. Instead, ancestral relationships in a
+sample can be represented by a sequence of correlated, marginal trees along the
+genome. The Ancestral Recombination Graph (ARG) was introduced by
+Griffiths~\citep{griffiths1991two,griffiths1997ancestral} as a branching-coalescing 
+stochastic process whose realisations are graphs which into which all of the
+marginal ancestral trees are embedded. Coalescences encode common ancestry
+as they do in the coalescent, while branching events encode points at which a
+marginal tree changes from one position along the genome to the next.
+
+There are several formulations of the ARG, which vary by
+the amount of extra information they capture beyond a minimal representation
+of all marginal trees. Two prominent examples are the ``big''
+ARG~\citep{ethier1990two}, which has a particularly simple specification as a
+stochastic process but is computationally infeasible, and the ``little ARG''
+traversed by Hudson's algorithm~\citep{hudson1983properties}, which
+has a more complex state space and dynamics but can be implemented with 
+a more manageable runtime and memory footprint. An efficient simulation
+implementation has become available only recently \citep{kelleher2016efficient},
+prior to which state-of-the-art simulators were based on the so-called sequentially
+Markovian coalescent (SMC) \citep{mcvean2005approximating}, which delivers
+performance gains by disallowing mergers which would result in trapped
+non-ancestral material. An alternative restricted model is
+\texttt{ClonalFrame}  \citep{didelot2007inference}, which fixes one distinguished
+tree and disallows recombination in any other marginal tree.
+
+Similar graph encodings have also been introduced to model other mechanisms
+which change the local ancestral tree along the genome, such as gene
+conversion~\citep{wiuf2000coalescent} and
+horizontal gene transfer~\citep{baumdicker2014infinitely}.
+
+The idea of using branching-coalescing stochastic processes 
+to model genetic ancestral trees has also found application in natural selection.
+The Ancestral Selection Graph (ASG)~\citep{krone1997ancestral,neuhauser1997genealogy}
+uses dynamics identical to the ``big" ARG to simulate an ensemble of correlated
+potential ancestral trees. Weak genic selection is incorporated by sampling a 
+true ancestry from the ensemble in a non-uniform way.
+There is no ASG analogue of the more computationally tractable ``little" ARG,
+though some gains in tractability can be made by considering typed 
+lineages~\citep{etheridge2009coalescent} or by leveraging perfect simulation
+techniques when recurrent mutation is present\citep{fearnhead2001perfect}.
+Extensions of the ASG have been developed to frequency-dependent 
+selection~\citep{neuhauser1999ancestral, gonzalezcasanova2018duality},
+unlinked chromosomes~\citep{fearnhead2003ancestral}, recombining
+loci whereupon branching is due to both selection and
+recombination~\citep{donnelly1999genealogical}, and high fecundity 
+reproduction~\citep{gonzalezcasanova2018duality, koskela2019robust}.
 
@@
 ancestral \emph{process} rather than the outcome of a process.
-\begin{itemize}
-\item Ancestral Selection
-Graph~\citep{krone1997ancestral,neuhauser1999ancestral} models selection
-as a graph with virtual lineages [etc etc.]. Importantly, the graph structure
-of the ASG does \emph{not} correspond to a realised ancestry, but rather
-models possible outcomes.
-Other papers possibly worth citing:\citep{etheridge2009coalescent}
-\item Ancestral Gene Transfer Graph~\citep{baumdicker2014infinitely}
-\end{itemize}
 
@@
 
-\begin{itemize}
-\item Early methods \citep{hein1990reconstructing}
-\item Those sampling from the coalescent with recombination
-\citep{griffiths1996ancestral,kuhner2000maximum,fearnhead2001estimating},
-or SMC
-\citep{rasmussen2014genome}
-\item Hueristic methods~\citep{song2005efficient,minichiello2006mapping,
-parida2008estimating,fallon2013acg,mirzaei2016rent}
-\item Discussions of the complexity of inferring ARGs (but we need to
-dig into this: is this the complexity of sampling from the coalescent,
-finding a ``min ARG'', or something else?
-\citep{bordewich2005computational,wang2001perfect,hein1990reconstructing,song2005constructing}
-\item Recent methods: Relate, tsinfer, tsdate, etc.
-\end{itemize}
+The ARG has been used as a model for recombining DNA sequence data essentially
+since it was first formulated. Early methods focused on finding parsimonious
+ARGs consistent with a given set of
+sequences \citep{hein1990reconstructing, lyngso2005minimum}.
+However, identifying the most parsimonious ARG for a given sequence data set
+is NP-hard \citep{wang2001perfect}, and subsequent methods have resorted to
+various heuristics \citep{fallon2013acg, hein1993heuristic, ignatieva2021kwarg, 
+kelleher2019inferring, minichiello2006mapping, mirzaei2017rent,
+parida2008estimating, song2005efficient, thao2019hybrid, zhang2021biobank}. 
+The scalability of these methods varies substantially based on the heuristic in use
+and the underlying data structure, but the most efficient methods can be practically
+used on megabase scale data and tens of thousands of samples under human-like
+parameters \citep{kelleher2019inferring, zhang2021biobank}.
+
+An alternative approach is to treat the ancestry as a latent parameter to be averaged out
+by Monte Carlo methods, based either on importance sampling 
+\citep{griffiths1996ancestral, fearnhead2001estimating, jenkins2011inference} 
+or MCMC \citep{kuhner2000maximum, nielsen2000estimation, wang2008bayesian}.
+These methods invariably operated on a representation of the ``little ARG", typically
+generated until the grand MRCA. They are extremely computationally expensive,
+and applicable to at most hundreds of samples consisting of tens of kilobases with
+human-like parameters.
+
+Due to the computational expense of sampling ``little ARGs", state-of-the-art
+Monte Carlo methods typically rely on cheaper, approximate models.
+The most prominent examples are \texttt{ARGWeaver} \citep{rasmussen2014genome}
+and \texttt{Arbores} \citep{heine2018bridging}. The former also utilises a time
+discretisation approximation, but scales to dozens of mammal-like genomes.
+The \texttt{ClonalOrigin} method \citep{didelot2010inference,
+medinaaguayo2020speeding} uses the \texttt{ClonalFrame}
+model and data structure, though it is not as scalable as \texttt{ARGWeaver}.
+
+A very recent MCMC method called \texttt{ARGInfer} has resulted in the first
+improvement in the scalability of Monte Carlo methods for the exact ARG model 
+by making use of the tree sequence representation of marginal trees, suitably enriched
+to facilitate the computations needed for MCMC sampling \citep{mahmoudi2021inference}.
+While not competitive with the scalability of \texttt{ARGWeaver}, \texttt{ARGInfer} extends
+the feasible range of sequence lengths by at least an order of magnitude when
+compared with methods which sample realisations of the exact ``little ARG".
 

### 2022-01-30 paper.tex
Update paper.tex
Co-authored-by: Jerome Kelleher <jk@well.ox.ac.uk>
Update abstract

@@
 It has recently become possible to infer genetic ancestry in the presence of
-recombination at scale for the first time. This has created exciting
-possibilities, with many downstream applications based on
-these inferred genealogies. The inferred genetic ancestries are usually
-referred to as Ancestral Recombination Graphs, or ARGs. Although initially well
-defined as a graphical representation of the coalescent with recombination
-stochastic process, the interpretation has become unclear as it is now
-also understood to represent a particular realisation of a genetic
-ancestry.
-Inference methods do not all infer the same information:
-although all output marginal trees along the genome, there is a great deal of
-variation in how much information of the relationships between the trees is inferred
-and retained. An important programme of work over the coming years is to
-develop and refine ancestry inference methods, assessing the relative strengths
-and weaknesses of the various approaches. Using the blanket term ARG hides the
-important differences in the approaches, and hampers our ability to compare and
-improve inference methods. In this paper we discuss the different approaches we
-may use to encode genetic ancestry, and of the differing amounts of information
-about the ancestral process that we can represent and hope to observe. We
-classify existing inference methods according to the type of ancestry they
-infer, and discuss the strengths and weaknesses of the inferred structure for
-downstream applications.
+recombination at scale for the first time, enabling many
+downstream applications in population and statistical genetics.
+Such recombinant genetic ancestry is usually
+referred to as an Ancestral Recombination Graph, or ARG.
+% Note: slight repetition here with first sentence.
+There are now multiple methods that can infer ARGs at a practical scale
+and it is therefore vital that these methods can be systematically evaluated
+and compared to determine their strengths and weaknesses.
+Unfortunately, however, there is substantial confusion over
+terminology and little agreement on shared standards for data interchange,
+significantly hampering progress in this important area.
+This confusion and lack of agreement on basic issues such as what does and
+does not constitute an ARG is partly attributable to the historical
+development of the term.
+Originally rigorously defined as
+a representation of the coalescent with recombination
+stochastic process as a graph, an ARG is now understood to refer to any
+concrete realisation of a recombinant genetic ancestry.
+We show that the standard description of an ARG
+in terms of common ancestor and recombination events
+(inherited from the original stochastic process definition) is
+ambiguous, fundamentally limited and an unsuitable basis for data interchange.
+We provide a simple alternative definition of an ARG in which a node
+corresponds to one of an individuals monoploid genomes, an edge
+% jk-note: this is clumsy
+defines a genetic ancestry relationship between two nodes,
+and annotations on those edges define the precise patterns of
+ancestry in intervals along the genome.
+We use this framework to define some classifications and metrics of ARG nodes,
+and discuss the properties of some recently developed inference methods.
 \end{abstract}
@@
 marginal tree changes from one position along the genome to the next.
+%note-jk: could nitpick this in that diamonds don't generate tree "changes" as such. Maybe tweak this on the next round of revision?
 
+% jk-note: this isn't quite right I think, as it's muddling the distinction between a graph stochastic process
+% definition with the graph as the encoding of the marginal trees.
 There are several formulations of the ARG, which vary by

### 2022-01-30 paper.tex
Add new introduction

@@
 \usepackage[right]{lineno}
+
+\usepackage{subcaption}
+\usepackage{tikz}
+\usetikzlibrary{calc,positioning}
 
@@
 and discuss the properties of some recently developed inference methods.
+% jk-note: this ends a bit flat - we should say other stuff like we clarify
+% how the ARG relates the pedigree, etc, etc. Let's go through again when
+% we have a reasonable draft.
 \end{abstract}
@@
 \section*{Introduction}
+Inferring the genetic relationships between sampled genomes in the form of an
+evolutionary tree is a necessary prerequisite for many analyses of non (or
+rarely) recombining organisms [refs]. The trees capture what is known and
+knowable about the genetic ancestry of the sample in an elegant and concise
+way, by postulating a set of nodes that are ancestors common to the samples,
+and the ancestral relationships between those nodes. A rich literature exists
+on analysing the mathematical properties of these trees [cite?], and numerous
+methods exist to infer them~\citep{felsenstein2004inferring}. The situation is
+quite different in recombining organisms, where the analysis of genetic data
+depends on summaries ultimately derived from the trees such as allele
+% TODO cite the other papers that preceded ralph2020efficiently here.
+frequencies~\citep{ralph2020efficiently}, site frequency spectra, patterns of linkage
+disequilibrium~\citep{mcvean2002genealogical} and principal
+components~\citep{mcvean2009genealogical}. Genealogical analysis is equally
+fundamental to the study of recombining organisms, but profound technical
+difficulties have made the routine inference of recombinant genealogies
+impossible, and therefore tree-based analysis methods have been largely
+unavailable to the fields of population and statistical genetics.
 
-% HYW: this is a rough first pass, aiming to be concise. We should flash out the "recombination" aspect of ARGs here a bit, I think
+Recent breakthroughs large-scale inference
+methods~\citep{rasmussen2014genome,kelleher2019inferring,speidel2019method,
+schaefer2021ancestral,wohns2021unified}
+and data representation~\citep{kelleher2016efficient}
+have raised the genuine prospect of genealogical analysis becoming a standard part
+of the population and statistical genetics toolkit.
+Applications using these inferred ancestries as input have
+begun to appear~\citep{osmond2021estimating} and many more are sure to
+follow~\citep{harris2019database}. This vibrant research area, however,
+has a significant difficulty, which, if not addressed, will substantially hamper
+progress. We currently lack a well-defined data-model and shared terminology
+to discuss recombining genealogies,
+leading to basic errors in statements about what different inference methods
+produce as output (see the XXX section below), as well as presenting serious
+difficulties in either comparing the outputs of different methods or
+(from a users perspective) using multiple inference methods in an analysis.
 
-For a long time, the Ancestral Recombination Graph (ARG) has been an object of
-fascination to those interested in theoretical evolutionary genetics, as it
-captures all knowable genetic history of a collection of individual genomes.
-Recently advances in our ability to infer ARGs has led to a resurgence of
-interest in this field. However, these have also given rise to certain
-terminological confusion, which we aim to clear up in this paper.
-
-One major source of confusion is the use of the term ``ARG" to refer to both a
-backwards-in-time generative process (Griffiths) and a structure which can be
-created by that process
-\citep[e.g.][]{minichiello2006mapping,mathieson2020ancestry}.
+In this paper we [turn into proper round up paragraph later]
+\begin{enumerate}
+\item Discuss the classical Ancestral Recombination Graph formulation,
+and show how it derives from (and is limited by) the coalescent with
+recombination.
+\item Suggest a definition of ARGs that is free of these limitations,
+where the formulation is derived from a diploid pedigree rather than
+the coalescent approximations and spatial annotations are attached to
+edges rather than nodes.
+\item Suggest a classification of ARG nodes that helps us to understand
+the properties of the structures inferred by different methods, and illustrate
+with some examples.
+\end{enumerate}
 
@@
 marginal tree changes from one position along the genome to the next.
-%note-jk: could nitpick this in that diamonds don't generate tree "changes" as such. Maybe tweak this on the next round of revision?
+% jk-note: could nitpick this in that diamonds don't generate tree
+% "changes" as such. Maybe tweak this on the next round of revision?
 
-% jk-note: this isn't quite right I think, as it's muddling the distinction between a graph stochastic process
+% jk-note: this isn't quite right I think, as it's muddling the
+% distinction between a graph stochastic process
 % definition with the graph as the encoding of the marginal trees.
@@
 
-In the theoretical literature, ancestral graphs such as the ARG refer to the
-ancestral \emph{process} rather than the outcome of a process.
+\begin{figure}
+\centering
+\begin{tikzpicture}[xscale=0.8, yscale=0.3, node distance=2mm and 20mm]
+\tikzset{greynode/.style={circle,fill,inner sep=1},
+nodelabel/.style={font=\footnotesize}}
+
+\node (s0) [greynode] {};
+\node (s1) [right=of s0,greynode] {};
+\node (s2) [right=of s1,greynode] {};
+\node [greynode] (s3) at ($(s1) + (0,1)$) {};
+\node [greynode] (s4) at ($0.5*(s1) + 0.5*(s2) + (0,2)$) {};
+\node [greynode] (s5) at ($0.5*(s0) + 0.5*(s1) + (0,3)$) {};
+\node [greynode] (s6) at ($(s1) + (0,5)$) {};
+
+\node [nodelabel,anchor=north west] at ($(s3) + (0,0)$) {$x = 0.3$};
+\foreach \u/\lab in {s0/0, s1/1, s2/2} \node[nodelabel,anchor=north] at (\u) {\lab};
+\foreach \u/\lab in {s4/4} \node[nodelabel,anchor=south west] at (\u) {\lab};
+\foreach \u/\lab in {s5/5} \node[nodelabel,anchor=south east] at (\u) {\lab};
+\foreach \u/\lab in {s3/3, s6/6} \node[nodelabel,anchor=south] at (\u) {\lab};
+
+%% Edges
+\draw (s1) -- (s3);
+\foreach \child/\parent in {s6/s5, s6/s4, s5/s0, s4/s2}
+	\draw (\child) -| (\parent);
+
+\draw (s5) -| ($0.5*(s3) + 0.5*(s3-|s5)$) |- (s3);
+\draw (s4) -| ($0.5*(s3) + 0.5*(s3-|s4)$) |- (s3);
+
+\end{tikzpicture}
+\caption{\label{fig-arg} \textcolor{red}{A picture of a standard ARG for us
+to work with for now. We should replace it with something better at some
+point.}
+ }
+\end{figure}
 
@@
 
+\section*{Discussion}
+
+Discussion points (may or may not be used, certainly would be heavily revised)
+
+\begin{itemize}
+\item Terminology based around an inferred ancestry being a ``true''
+ARG is impoverished and misleading. The only reasonable interpretation
+of a ARG being true or not is whether it's a sample from the coalescent.
+Most inference methods (and all that scale well) do not. In any case,
+the coalescent is just one model, and a highly idealised one. It is
+exceedingly unlikely that the actual ARG describing the ancestry
+of any biological population is ``true'' ARG in this sense. The terminology is
+impoverished because it reduces the possibilities available
+to something either being an ARG or not, whereas there are many different
+interesting properties of these structures that we should be able to
+discuss and compare.
+\end{itemize}
+
 \bibliographystyle{plainnat}

### 2022-01-30 paper.tex
Some notes on the ARG

@@
 \end{enumerate}
+
+\section*{Ancestral recombination graphs}
+The coalescent~\citep{kingman1982coalescent,kingman1982genealogy,
+hudson1983testing, tajima1983evolutionary}
+provides the theoretical underpinning for much of contemporary population genetics.
+Hudson extended the coalescent to include
+recombination~\citep{hudson1983properties,hudson1990gene}
+and the model has proved to be especially powerful basis for
+simulations~\citep{hudson2002generating,kelleher2016efficient,kelleher2020coalescent,
+baumdicker2021efficient}.
+Hudson's algorithm is concerned with simulating the process of
+the coalescent with recombination as efficiently as possible, and operates by
+tracking segments of ancestral material carried by ancestors as
+we go backwards in time. The details of Hudson's algorithm and how the
+results of the ancestral process are represented are critical to later
+discussions, and so we will explain the process in some detail.
+See~\cite{kelleher2016efficient} for a more detailed
+description of the algorithm, and how it may be implemented efficiently.
+
+Hudson's algorithm operates by tracking the state of a set of ancestral
+lineages as we go backwards in time. Each lineage consists of a list of
+disjoint ancestry segments $(\ell, r, a)$, where
+$[\ell, r)$ is a half-closed genomic interval and $a$ is an integer
+tracking the number of samples the lineage is ancestral to over the interval.
+(We also usually track the tree node associated with each segment, but
+that is not important for our purposes here so we omit it.)
+If we have $n$ samples and genome of length $L$, the process begins with $n$ lineages
+of the form $[(0, L, 1)]$. The process then works backwards in time from
+the present day as a series of random common ancestor or recombination events.
+Recombination events occur at a rate determined by the amount and spatial
+distribution of ancestral material among lineages.
+Let $A$ be the set of ancestors at a given time $t$. Recombination events
+happen at rate $\rho \nu / (L - 1)$ where
+\[
+\nu = \sum_{x \in A}\left( \max_{(\ell, r, a) \in x}r
+    - \min_{(\ell, r, a) \in x}\ell - 1 \right)
+\]
+is the number of available `links' that may be broken. Thus, the rate of
+recombination is determined by the left and right-most extent of the
+ancestral material carried by each lineage. At a recombination
+event we choose one of these links uniformly and break it. Given a lineage
+$x = [(\ell_j, r_j, a_j)]$ and a breakpoint $k$, we have two lineages
+$x_1$ and $x_2$ such that FILL IN DETAILS
+
+When $k = |A|$ lineages are present, common ancestor events
+occur at rate $k(k - 1) / 2$. In a common ancestor event two lineages
+are chosed uniformly at random and their ancestry segments merged.
+If we have overlapping intervals of ancestry from the two lineages,
+say, $(\ell, r, a_1)$ and $(\ell, r, a_2)$, a
+\emph{coalescence} occurs and ancestor represented by the current event
+will be present as a node (at least) in the marginal trees covering
+the interval $[\ell, r)$. The result of this coalescence is a segment
+$(\ell, r, a_1 + a_2)$, and if $a_1 + a_2 < n$ it is included in the
+ancesty for the new lineage. Otherwise, if $a_1 + a_2 = n$ we know that
+we have found the most recent common ancestor of all samples in
+the interval $[\ell, r)$ and so we do not need to simulate its history any further.
+Nonoverlapping intervals of ancestry from the two lineages are included
+in the resulting lineage without changes. Eventually, as the process continues,
+we find resultant lineages in which all segments have fully coalescenced,
+and so the number of extant lineages gradually dwindles down to zero.
+
+The Ancestral Recombination Graph (ARG) was later introduced
+Griffiths~\citep{griffiths1991two,ethier1990two,griffiths1997ancestral}, and is a
+closely related and complementary treatment of the same process. Where
+Hudson's algorithm is focused on the efficient \emph{simulation} of the
+process, the original ARG literature is focused on mathematical
+results about the stochastic process. Griffiths and colleagues formulated the
+ARG as a branching-coalescing stochastic process
+
+NOTES for continuing:
+
+\begin{itemize}
+\item The ARG process does not track ancestral material. Recombination events
+occur at rate $k \rho / 2$, and we don't complete until there's only one
+lineage left. That means the process continues for an exponentially long time.
+\item Discuss the way in which the re events work.
+\item Discuss the rules for pulling out the trees. This is equivalent to doing
+the interval logic above in Hudson's algorithm.
+\end{itemize}
 
@@
 discuss and compare.
+\item tskit can represent any ARG.
 \end{itemize}

### 2022-01-31 paper.tex

@@
 evolutionary tree is a necessary prerequisite for many analyses of non (or
-rarely) recombining organisms [refs]. The trees capture what is known and
+rarely) recombining organisms [refs]. The tree captures what is known and
 knowable about the genetic ancestry of the sample in an elegant and concise
@@
 
-\section*{Ancestral recombination graphs}
+\section*{Ancestral graphs and stochastic processes}
 The coalescent~\citep{kingman1982coalescent,kingman1982genealogy,
@@
 recombination~\citep{hudson1983properties,hudson1990gene}
-and the model has proved to be especially powerful basis for
-simulations~\citep{hudson2002generating,kelleher2016efficient,kelleher2020coalescent,
-baumdicker2021efficient}.
+and the model has proved to be an especially powerful basis for
+simulations~\citep{hudson2002generating,baumdicker2021efficient}.
 Hudson's algorithm is concerned with simulating the process of
@@
 
+\begin{figure}
+\caption{\label{fig-sim-statespace} An illustration of the state space
+of Hudson's algorithm (A) and the big ARG (B) at a given instant. The
+state maintained by Hudson's algorithm is considerably more complicated,
+but is guaranteed to only simulate events corresponding to genetic
+ancestors of the sample.
+}
+\end{figure}
+
 Hudson's algorithm operates by tracking the state of a set of ancestral
-lineages as we go backwards in time. Each lineage consists of a list of
+lineages as we go backwards in time.
+% jk-note unclear here about the best choice of notation. We could have
+% lists of tuples (l, r, a), or a tuple of vectors (l_j, r_j, a_j),
+% of a tuple (I_j, a_j), where I is a vector of intervals. Going to
+% stick with the original tuple notation for now.
+%
+% Each lineage consists of a set of
+% equal-length vectors $\mathbf{l}$, $\mathbf{r}$ and $\mathbf{a}$
+% such that for each $1 \leq j \leq \left| l \right|$,
+% $[\mathbf{l}_j, \mathbf{r}_j)$ is a half-closed genomic interval and
+% $\mathbf{a}_j$ is an integer
+% tracking the number of samples the lineage is ancestral to over the interval.
+% Assume that the intervals are sorted from left-to-right so that $\mathbf{l}_1$
+% and $\mathbf{r}_{|\mathbf{r}|}$ are the left- and right-most positions
+% covered by the lineage, respectively.
+Each lineage consists of a list of
 disjoint ancestry segments $(\ell, r, a)$, where
@@
 that is not important for our purposes here so we omit it.)
-If we have $n$ samples and genome of length $L$, the process begins with $n$ lineages
-of the form $[(0, L, 1)]$. The process then works backwards in time from
+If we have $n$ samples and genome of length $m$, the process begins with $n$ lineages
+of the form $[(0, m, 1)]$. The process then works backwards in time from
 the present day as a series of random common ancestor or recombination events.
@@
 distribution of ancestral material among lineages.
-Let $A$ be the set of ancestors at a given time $t$. Recombination events
-happen at rate $\rho \nu / (L - 1)$ where
+Let $L$ be the set of ancestors at a given time $t$. Recombination events
+happen at rate $\rho \nu / (m - 1)$ where
 \[
-\nu = \sum_{x \in A}\left( \max_{(\ell, r, a) \in x}r
+\nu = \sum_{x \in L}\left( \max_{(\ell, r, a) \in x}r
     - \min_{(\ell, r, a) \in x}\ell - 1 \right)
@@
 is the number of available `links' that may be broken. Thus, the rate of
-recombination is determined by the left and right-most extent of the
+recombination is determined by the left- and right-most extent of the
 ancestral material carried by each lineage. At a recombination
@@
 
-When $k = |A|$ lineages are present, common ancestor events
-occur at rate $k(k - 1) / 2$. In a common ancestor event two lineages
+When $k = |L|$ lineages are present, common ancestor events
+occur at rate $\binom{k}{2}$. In a common ancestor event two lineages
 are chosed uniformly at random and their ancestry segments merged.
@@
 $(\ell, r, a_1 + a_2)$, and if $a_1 + a_2 < n$ it is included in the
-ancesty for the new lineage. Otherwise, if $a_1 + a_2 = n$ we know that
+ancestry for the new lineage. Otherwise, if $a_1 + a_2 = n$ we know that
 we have found the most recent common ancestor of all samples in
@@
 
+% NOTE: it would probably be helpful to have a drawing of the ARG process
+% in-progress here to help us talk about it. What is the full state
+% that is remembered as we go through the process? It does help to think
+% about the process as choosing edges to merge, not nodes.
 The Ancestral Recombination Graph (ARG) was later introduced
-Griffiths~\citep{griffiths1991two,ethier1990two,griffiths1997ancestral}, and is a
+Griffiths~\citep{griffiths1991two,ethier1990two,griffiths1996ancestral,
+griffiths1997ancestral}, and is a
 closely related and complementary treatment of the same process. Where
@@
 results about the stochastic process. Griffiths and colleagues formulated the
-ARG as a branching-coalescing stochastic process
+ARG as a branching-coalescing stochastic process. We imagine a graph in which
+each edge corresponds to an extant lineage and nodes are events in the
+process (the initial $n$ leaf nodes are ``sampling'' events). As in Hudson's
+algorithm, common ancestor events occur at rate $\binom{k}{2}$ when there
+are $k$ lineages present. We choose two lineages (edges) uniformly, and merge them
+into a common ancestor lineage. Recombination events are different
+to Hudson's algorithm, however. Rather than occuring at a rate which depends on
+the specific distribution of ancestral material among lineages, the
+rate of recombination is simply $k \rho (m - 1)$ in the original ARG
+formulation. At a recombination event we choose a lineage (edge) uniformly, and a
+breakpoint $0 < x < L$ uniformly on its genome. We terminate the edge at a
+node, record the breakpoint and start two new edges from this node. The process
+then continues until there is only one lineage left (the Grand Most Recent
+Common Ancestor, GMCRA), which is guaranteed to
+happen in finite time because of the linear vs quadratic rates of branching
+and coalescing. The graph structure and breakpoints associated with with
+recombination nodes provides sufficient information to later recover the marginal
+trees (see the next section for more details and discussion on this point).
 
-NOTES for continuing:
+The state-space of the ARG is much simpler than Hudson's algorithm, which
+greatly facilitates mathematical reasoning. This simplicity comes at a
+substantial cost, however, if we wish to use it as a practical means of
+simulating recombinant ancestry. The ARG is vast, and any realisation
+for even moderate levels of recombination is far too large to be of practical
+use. The number of events in the ARG back to the GMRCA
+is $O(e^\rho)$~\citep{griffiths1997ancestral}, whereas the number
+of events required to simulate the complete ancestry of a sample
+in Hudson's algorithm is
+$O(\rho^2)$~\citep[Eq.~5.10]{hein2004gene,baumdicker2021efficient}.
+This disparity in the number of events in the two formulations is
+because the vast majority of the events that occur in the ARG do
+not affect the genetic ancestry of the sample in any way. Recombination
+events occur outside of ancestral material do not have any bearing
+on the ancestry of the sample, and so the structure is hugely redundant.
 
-\begin{itemize}
-\item The ARG process does not track ancestral material. Recombination events
-occur at rate $k \rho / 2$, and we don't complete until there's only one
-lineage left. That means the process continues for an exponentially long time.
-\item Discuss the way in which the re events work.
-\item Discuss the rules for pulling out the trees. This is equivalent to doing
-the interval logic above in Hudson's algorithm.
-\end{itemize}
-
-\section*{Ancestral graphs and stochastic processes}
-
-% See https://github.com/tskit-dev/what-is-an-arg-paper/discussions/13
-% for discussions on this section
-
-The simplest prominent example of genetic ancestry generated by a stochastic process
-is the coalescent process \citep{kingman1982coalescent,kingman1982genealogy,
-hudson1983testing, tajima1983evolutionary}, which describes the random
-ancestral tree at one non-recombining locus obtained by sampling a relatively
-small number of individuals at random from a large population.
-
-In a recombining organism, the genetic ancestry of a longer segment of DNA is
-no longer described by a single tree. Instead, ancestral relationships in a
-sample can be represented by a sequence of correlated, marginal trees along the
-genome. The Ancestral Recombination Graph (ARG) was introduced by
-Griffiths~\citep{griffiths1991two,griffiths1997ancestral} as a branching-coalescing
-stochastic process whose realisations are graphs which into which all of the
-marginal ancestral trees are embedded. Coalescences encode common ancestry
-as they do in the coalescent, while branching events encode points at which a
-marginal tree changes from one position along the genome to the next.
-% jk-note: could nitpick this in that diamonds don't generate tree
-% "changes" as such. Maybe tweak this on the next round of revision?
-
-% jk-note: this isn't quite right I think, as it's muddling the
-% distinction between a graph stochastic process
-% definition with the graph as the encoding of the marginal trees.
-There are several formulations of the ARG, which vary by
-the amount of extra information they capture beyond a minimal representation
-of all marginal trees. Two prominent examples are the ``big''
-ARG~\citep{ethier1990two}, which has a particularly simple specification as a
-stochastic process but is computationally infeasible, and the ``little ARG''
-traversed by Hudson's algorithm~\citep{hudson1983properties}, which
-has a more complex state space and dynamics but can be implemented with
-a more manageable runtime and memory footprint. An efficient simulation
-implementation has become available only recently \citep{kelleher2016efficient},
-prior to which state-of-the-art simulators were based on the so-called sequentially
-Markovian coalescent (SMC) \citep{mcvean2005approximating}, which delivers
-performance gains by disallowing mergers which would result in trapped
-non-ancestral material. An alternative restricted model is
-\texttt{ClonalFrame}  \citep{didelot2007inference}, which fixes one distinguished
-tree and disallows recombination in any other marginal tree.
+This original formulation which does not keep track of the distribution
+of ancestral material among lineages is the so-called ``Big'' ARG.
+The ``little'' ARG is the subset of the Big ARG traversed by
+Hudson's algorithm (\citet{wiuf1999recombination} refer to this as HUD).
+[Review of other work done on the ARG stochastic process here.
+Discuss \citet{wiuf1999ancestry}. Wiuf and Hein's classical pair of papers
+clarified the relationship between Hudson's algorithm and the ARG]
 
@@
 horizontal gene transfer~\citep{baumdicker2014infinitely}.
+[Mention the
+SMC~\citep{mcvean2005approximating,marjoram2006fast}
+and ClonalFrame~\citep{didelot2007inference} approximations somewhere.]
 
@@
 
-\section*{The ARG as a structure}
+\section*{The ARG as a data structure}
+
+The first method to explicitly produce an ARG as output is SHRUB developed
+by \citet{song2005efficient}.
+
+[WIP jk]
 
@@
 
+\citep{camara2016inference} is another type of parsimony method.
+
 An alternative approach is to treat the ancestry as a latent parameter to be averaged out

### 2022-01-31 paper.tex
Added section on phylogenetic networks; edited section on parsimonious ARG reconstruction

@@
 
-Recent breakthroughs large-scale inference
+Recent breakthroughs in large-scale inference
 methods~\citep{rasmussen2014genome,kelleher2019inferring,speidel2019method,
@@
 % about the process as choosing edges to merge, not nodes.
-The Ancestral Recombination Graph (ARG) was later introduced
-Griffiths~\citep{griffiths1991two,ethier1990two,griffiths1996ancestral,
+The Ancestral Recombination Graph (ARG) was later introduced by Griffiths~\citep{griffiths1991two,ethier1990two,griffiths1996ancestral,
 griffiths1997ancestral}, and is a
@@
 by \citet{song2005efficient}.
+% AI: I think \citep{song2003parsimonious} came before.
 
@@
 
+\section*{Relationship with other types of phylogenetic networks}
+
+The term \emph{phylogenetic network} has been broadly accepted to describe the most general 
+type of genealogical graph, defined by \citet{huson2010phylogenetic} as ``any graph used to 
+represent evolutionary relationships (either abstractly or explicitly) between a set of taxa 
+that labels some of its nodes" (although more narrow and context-dependent definitions can also 
+be found in the literature). Explicit rooted phylogenetic networks are rooted DAGs,
+with leaves labelled by the taxa (species, groups, individuals or sequences), and nodes classed 
+as either reticulation nodes (with in-degree $\geq 2$) or tree nodes.
+
+Some sub-classes of explicit rooted phylogenetic networks are defined based on the evolutionary event that the
+reticulation nodes capture \citep{huson2010phylogenetic}. In \emph{reassortment} networks, used to 
+model the evolution of viruses, reticulation nodes correspond to exchange of 
+(non-recombining) genomic segments between viral particles inside a co-infected host. 
+In \emph{hybridisation} networks, reticulation nodes correspond to genetic material from two
+different species combining to create a hybrid. In \emph{recombination} networks, the reticulation nodes 
+correspond to recombination events (with leaves labelled by sequences and edges labelled by mutations). 
+The ARG is a subtype of a recombination network under this definition (restricting to one crossover 
+recombination per reticulation node).
+
+Sub-classes of explicit rooted phylogenetic networks can also be defined by restrictions on their topology. 
+In \emph{tree-child networks}, every internal node must have at least one child that is a tree node; 
+\emph{tree-sibling networks} are constrained by every reticulation node having at least one sibling that is a 
+tree node \citep{cardona2008extended}. Further, sub-types can be defined by restrictions on 
+the topology of subgraphs that contain the reticulation loops (cycles containing a reticulation node). 
+A reticulation loop that does not share any node with another reticulation loop is termed a gall 
+\citep[][p.\ 237]{gusfield2014recombinatorics}; a reticulation network containing only galls is 
+called a \emph{galled tree}. Reticulation loops that are joined together in a reticulation network 
+by at least one shared node form a blob; a \emph{level-$k$ network} is one where the maximal number 
+of reticulation loops inside a blob is $k$ \citep{choy2005computing}. Galled trees are level-1 networks, 
+and the tractability afforded by their relatively simple structure results in a number of attractive 
+combinatorial properties \citep{wang2001perfect, gusfield2004optimal}. We consider the general case of 
+an ARG, without applying any of these restrictions.
+
+% AI: have not mentioned any un-rooted networks (split, median, haplotype, reticulograms, etc).
+
 \section*{ARG inference methods}
 Using the terminology developed here to classify ARGs, we review methods developed
-to infer ARGs. We focus our attention on some recent methods and discuss the
+to \emph{infer} ARGs from sequencing data. We focus our attention on some recent methods and discuss the
 properties of the ARGs inferred.
 
-The ARG has been used as a model for recombining DNA sequence data essentially
-since it was first formulated. Early methods focused on finding parsimonious
-ARGs consistent with a given set of
-sequences \citep{hein1990reconstructing, lyngso2005minimum}.
-However, identifying the most parsimonious ARG for a given sequence data set
-is NP-hard \citep{wang2001perfect}, and subsequent methods have resorted to
-various heuristics \citep{fallon2013acg, hein1993heuristic, ignatieva2021kwarg,
-kelleher2019inferring, minichiello2006mapping, mirzaei2017rent,
-parida2008estimating, song2005efficient, thao2019hybrid, zhang2021biobank}.
+The problem of reconstructing ARGs for samples of recombining sequences has been of significant interest
+essentially since the ARG was first defined. Early methods focussed on finding the most 
+\emph{parsimonious} ARGs, i.e.\ those minimising the number of 
+posited recombination events in the history of the sample \citep{hein1990reconstructing, song2003parsimonious,
+lyngso2005minimum}. However, reconstructing a most parsimonious ARG for a given data set
+is NP-hard \citep{wang2001perfect}, and in order to allow for larger input datasets,
+subsequent parsimony-based methods have resorted to various relaxations and heuristics \citep{hein1993heuristic, 
+song2005efficient, wu2008association, camara2016inference, thao2019hybrid, ignatieva2021kwarg}.
 The scalability of these methods varies substantially based on the heuristic in use
-and the underlying data structure, but the most efficient methods can be practically
-used on megabase scale data and tens of thousands of samples under human-like
-parameters \citep{kelleher2019inferring, zhang2021biobank}.
-
-\citep{camara2016inference} is another type of parsimony method.
+and the underlying data structure, with the most efficient tools limited to analysing hundreds of sequences.
+A number of heuristic methods have also been developed which do not (explicitly) focus on reconstructing
+the most parsimonious ARGs, but rather aim for computational efficiency
+while seeking to achieve good accuracy of the inferred genealogies
+\citep{minichiello2006mapping, parida2008estimating, wu2011new, mirzaei2017rent}; some of these methods \citep{kelleher2019inferring, speidel2019method, zhang2021biobank} can be practically
+used on megabase scale data and tens to hundreds of thousands of samples under human-like
+parameters.
+% AI: just to note, these papers do not all agree in their definition of an ARG. 
+% I guess this is one of the points of the present paper, but should we mention this fact?
+% Should we also mention that most of these infer just the topology, but some also the times?
 
@@
 \citep{griffiths1996ancestral, fearnhead2001estimating, jenkins2011inference}
-or MCMC \citep{kuhner2000maximum, nielsen2000estimation, wang2008bayesian}.
+or MCMC \citep{kuhner2000maximum, nielsen2000estimation, wang2008bayesian, fallon2013acg}.
 These methods invariably operated on a representation of the ``little ARG", typically

### 2022-02-01 paper.tex
Section on phylogenetic networks, edits to section on parsimonious ARG reconstruction (#48)
* Added section on phylogenetic networks; edited section on parsimonious ARG reconstruction

* Added some comments

* Adding back .svg figs
Working through background material
Fix wu2011 citation

@@
 \section*{The ARG as a data structure}
+Early work on ARG inference focused on the problem of
+of inferring parameters of the
+stochastic process, where the ancestry is regarded as a
+latent parameter to be averaged out
+\citep[e.g.][]{griffiths1996ancestral,kuhner2000maximum, nielsen2000estimation,
+fearnhead2001estimating}. These methods met with limited success
+due to the overwhelmingly large and
+horribly structured [Jere: what's the right way to say this?]
+state space of ARGs. The focus subsequently shifted to
+the more tractable---but still
+NP-hard~\citep{wang2001perfect}---problem of computing
+the miniumum number of recombinations required
+to explain an input dataset [CITATIONS], and inferring the corresponding
+ARG realisations~\citep{song2005efficient,lyngso2005minimum}.
+The distinction between the ARG as a stochastic process
+and the ARG as a data structure became less clear at this point,
+with \citet{minichiello2006mapping} explicitly arguing for
+the data structure interpretation.
+
+The Griffiths ARG encoding is an elegant and mathematically
+economical description of the coalescent with recombination, but
+it has significant drawbacks as the basis for a general purpose
+storage format.
+
+[Not sure how to make a narrative out of this, but here are the
+basic problems:]
+\begin{enumerate}
+\item The encoding is hamstrung by the interpretation of
+nodes as \emph{events} of a particular type. We can either
+have common ancestor or recombination events. This is true
+in the coalescent, but is not in a more general Wright-Fisher
+model.
+\item We can also only have one crossover per recombination event.
+This is unrealistic because things like gene conversion are common,
+and means we can't have back and forth recombination between
+two parents.
+\item The passage of ancestral material along edges requires
+extrinsic rules which must be shared by all implementations.
+These rules can be quite restrictive when generating, and
+in practise some have found it simpler to add explicit
+annotations to the edges.
+\item Generating local trees is inefficient, essentially requiring
+$O(n T)$ time (or more?)
+\end{enumerate}
 
@@
 
-\citep[e.g.][]{minichiello2006mapping,mathieson2020ancestry}.
+This ambiguity can be dealt with in a number of ways. In the original
+treatments, the direction in which ancestral material travels at
+a breakpoint is handled by a simple rule, sites to the left of the
+breakpoint go to the left parent and sites to right go to the
+right parent. [Illustrate the problem with simulating an ARG
+like this]. ~\cite{ignatieva2021kwarg} resolve this problem by
+associating a label ``P'' (prefix) or ``S'' (suffix) with the
+outbound edges of a recombination event. This is the same approach
+as used in beagle.
+
+\section*{Anchoring the ARG to the pedigree}
+
+The limitations of the Griffiths ARG encoding directly stem from its
+rooting in the coalescent with recombination stochastic process.
+Indeed, these limitations arise because it is such a \emph{good}
+mathematical encoding, providing sufficient information to
+define the realised ancestry under the assumptions of the model,
+and no more.
+
+We describe here an alternative formulation in which the
+concepts are based on the diploid pedigree, which
+is free from the limitations discussed in the previous section.
+\cite{mathieson2020ancestry} and others [CITE] have also
+suggested this approach, although without explicitly defining
+the roles of nodes and the details of how ancestry
+flows along edges.
 
@@
 
-% HYW: this is a rough first pass. We should take care not to assume that the tree sequence
-% approach is the only encoding
+[Here are the basic components of the model]
 
-The other use of the the term ARG, and the one for which we suggest the term be
-reserved, is to describe the *structure* resulting from the genetic process of
-inheritance. This structure, like any other graph, consists of *nodes*
-connected by *edges*. In the context of genetics, the nodes represent (haploid)
-genomes, and the edges represent paths of inheritance. Although for theoretical
-purposes a graph like this can be created by backwards-in-time simulations
-using the CwR, in reality it is created forwards-in-time, as a population
-evolves.
-
-Since genetic inheritance is unidirectional, the edges in this graph structure
-are *directed*, with a "parent" and "child" node. Moreover, there is a strict
-temporal order, so that children cannot be their own parents: in other words,
-there are no cycles in the graph, and it is technically a form of "directed
-acyclic graph" (DAG).
-
-% HYW: NB: worded carefully below, as it should also be possible to encode ARGs even if there is not a single fixed coordinate system
-
-There is one key feature which distinguishes an ARG for other graphs. Although
-a genome (graph node) may have more than one parent, the particulate nature of
-genetic inheritance means that each letter in the genome can only come from a
-*single* parent. Hence if we compare the equivalent letters from multiple
-extant genomes, their paths of inheritance must form a *tree* (which,
-incidentally, is why phylogenetics plays such an important role in evolutionary
-theory). As we shall argue, the ability to generate these "local trees" from an
-ARG is of key importance.
-
-When describing an ARG, the primary issue, from a practical point of view, is
-how to annotate the graph with the details of which piece of genome has come
-from which parent. There are two possibilities: the annotations can be
-associated either with the nodes, or with the edges.
+\begin{enumerate}
+\item A node corresponds to one of a diploid individual's two genomes.
+This is linked explicitly to the pedigree, and is fully general.
+Nodes \emph{may} be labelled with certain flags describing their properties,
+but it's not necessary.
+\item An edge between two nodes defines a genetic inheritance relationship
+between two nodes. Edges are annotated with information about the
+passage of ancestral materal. There are different forms these annotations
+can take. For example, we can follow the approach taken in the Griffiths
+ARG and just annotate recombinant edges with the interval $[0, x)$ or $[x, L)$,
+depending on side of the breakpoint the edge is on. In the most
+general case, we annotate each edge with the full set of disjoint
+intervals of ancestral material that flows along that edge.
+\citep{mcgill2013graphml} also proposed full ancestral material being
+associated with each edge.
+\end{enumerate}
 
@@
 \end{figure}
-
-\subsection*{Node annotated recombination}
-
-In the original (Griffiths) formulation, nodes which have multiple parents are
-annotated with a breakpoint, representing a genomic position. The genome
-lying to the left of that position is then taken to come from one parent, while
-the genome lying to the right is taken to come from the other parent. This is
-an elegantly concise encoding of the parental source, and is used, for example
-in the ARGweaver .arg format. However, it suffers from three principal
-drawbacks:
-
-1. It assumes only a single breakpoint per node: problematic for representing
-e.g. gene conversion or multiple chromosomes (but this could be circumvented by
-having multiple breakpoints)
-2. Nodes can only have a maximum of 2 parents. At
-first sight, this seems a reasonable limitation, but it turns out that it can
-be helpful to collapse multiple nodes into one, which potentially results in
-more than two parents for any given node (see section XXXX)
-3. Generating local
-trees from such an encoding is time consuming (XXX describe *how* time
-consuming - i,e. scaling properties with number of nodes - and why this is,
-theoretically)
-
-\subsection*{Edge annotated recombination}
-
-An alternative approach to annotating nodes with the parent of origin of each
-section of genome is to annotate each edge with the genomic region which is
-inherited via that route. This approach was first introduced by Kelleher et al
-(cite msprime paper) and now forms the basis of a comprehensive software
-library.
-
-% HYW: should discuss the link to biology & mitosis/meiosis: a "node" is
-% strictly a cell - it is shorthand to associate it with an "individual"
 
@@
 recombination per reticulation node).
+% AI: Not sure about this - do ARGs also encompass multiple crossover recombinations and gene
+% conversion (seems implied above)?
 
 Sub-classes of explicit rooted phylogenetic networks can also be defined by restrictions on their topology.
+In \emph{binary networks}, all internal nodes must have either in-degree two and out-degree one or in-degree
+one and out-degree two (with the root having out-degree two and the leaves in-degree 1) \citep{steel2016phylogeny}.
 In \emph{tree-child networks}, every internal node must have at least one child that is a tree node;
@@
 an ARG, without applying any of these restrictions.
-
-% AI: have not mentioned any un-rooted networks (split, median, haplotype, reticulograms, etc).
+% AI: have not mentioned any implicit or un-rooted networks (split, median, haplotype, reticulograms, etc).
+% Also: are we not defining ARGs as "binary networks"? Might be good to specify this here.
 
@@
 % Should we also mention that most of these infer just the topology, but some also the times?
+% How much more description of the parsimony methods would be good? Maybe a mention of them splitting into
+% "back in time" and "across local trees" types?
 
@@
 \item tskit can represent any ARG.
+\item Interchange is a real problem, and if we're going to realise the
+potential for ARGs in genomics we must solve it. Efforts to standardise
+phylogenetic networks haven't done very well (~\citep{cardona2008extended}
+has 89 citations). The only real effort to standardise an ARG format
+based on extending GraphML~\citep{mcgill2013graphml} has 11 citations.
+GraphML~\citep{mcgill2013graphml} has some of the same key ideas
+as tskit though, in particular allowing the annotation of ``live sites''
+to an edge as a set of disjoint intervals that hold ancestral material.
+
+\item We should discuss other types of ancestry network also.
+The crucial distiction between an ARG and a phylogenetic network (I think?) is
+the presence of position information specifying the location of
+recombination events (and by extension, the complete passage
+of ancestral material the graph). Without position information
+we know that a particular node was an ancestor somewhere along the
+sequence, but not exactly where.
+
 \end{itemize}

### 2022-02-01 paper.tex
Tweaks

@@
 on the ancestry of the sample, and so the structure is hugely redundant.
+As~\cite{wiuf1999recombination} note,
+``an `ancestral' sequence in the birth and death process
+need not have any genetic material in common with a
+sequence descended from it.''
+
+% Full quote:
+% This process simplifies mathematics on the account that the notion of an
+% ancestor will have a less restrictive meaning than usual:
+% An ``ancestral'' sequence in the birth and death process
+% need not have any genetic material in common with a
+% sequence descended from it.
 
@@
 to explain an input dataset [CITATIONS], and inferring the corresponding
-ARG realisations~\citep{song2005efficient,lyngso2005minimum}.
+ARG realisations~\citep{song2003parsimonious,song2005efficient,lyngso2005minimum}.
 The distinction between the ARG as a stochastic process
@@
 
-The first method to explicitly produce an ARG as output is SHRUB developed
-by \citet{song2005efficient}.
-% AI: I think \citep{song2003parsimonious} came before.
-
 [WIP jk]
@@
 outbound edges of a recombination event. This is the same approach
-as used in beagle.
+as used in beagle~\citep{lyngso2005minimum}.
 

### 2022-02-02 paper.tex

@@
 Originally rigorously defined as
-a representation of the coalescent with recombination
-stochastic process as a graph, an ARG is now understood to refer to any
+a stochastic process (the coalescent with recombination) together with its 
+representation, an ARG is now understood to refer to any
 concrete realisation of a recombinant genetic ancestry.
@@
 We provide a simple alternative definition of an ARG in which a node
-corresponds to one of an individuals monoploid genomes, an edge
+corresponds to one of an individual's monoploid genomes, an edge
 % jk-note: this is clumsy
@@
 \section*{Introduction}
-Inferring the genetic relationships between sampled genomes in the form of an
+Inferring the genealogical relationships between sampled genomes in the form of an
 evolutionary tree is a necessary prerequisite for many analyses of non (or
@@
 quite different in recombining organisms, where the analysis of genetic data
-depends on summaries ultimately derived from the trees such as allele
+depends on summaries ultimately derived from the trees
 % TODO cite the other papers that preceded ralph2020efficiently here.
-frequencies~\citep{ralph2020efficiently}, site frequency spectra, patterns of linkage
-disequilibrium~\citep{mcvean2002genealogical} and principal
+such as allele frequencies and site frequency spectra~\citep{ralph2020efficiently}, patterns of linkage
+disequilibrium~\citep{mcvean2002genealogical}, and principal
 components~\citep{mcvean2009genealogical}. Genealogical analysis is equally
@@
 \item Suggest a definition of ARGs that is free of these limitations,
-where the formulation is derived from a diploid pedigree rather than
-the coalescent approximations and spatial annotations are attached to
+where the formulation is derived from a (normally diploid) pedigree rather than from
+coalescent approximations, and where the genetic
+material which is transmitted is described by annotations attached to
 edges rather than nodes.
@@
 $[\ell, r)$ is a half-closed genomic interval and $a$ is an integer
-tracking the number of samples the lineage is ancestral to over the interval.
+tracking the number of samples to which the lineage is ancestral over that interval.
 (We also usually track the tree node associated with each segment, but
@@
 the present day as a series of random common ancestor or recombination events.
-Recombination events occur at a rate determined by the amount and spatial
-distribution of ancestral material among lineages.
+Recombination events occur at a rate determined by the amount of ancestral material and
+the way in which it is distributed along a chromosome.
 Let $L$ be the set of ancestors at a given time $t$. Recombination events
@@
 When $k = |L|$ lineages are present, common ancestor events
-occur at rate $\binom{k}{2}$. In a common ancestor event two lineages
-are chosed uniformly at random and their ancestry segments merged.
+occur at rate $\binom{k}{2}$. In a common ancestor event, two lineages
+are chosen uniformly at random and their ancestry segments merged.
 If we have overlapping intervals of ancestry from the two lineages,
@@
 
-The state-space of the ARG is much simpler than Hudson's algorithm, which
+The state-space of this process is much simpler than Hudson's algorithm, which
 greatly facilitates mathematical reasoning. This simplicity comes at a
@@
 NP-hard~\citep{wang2001perfect}---problem of computing
-the miniumum number of recombinations required
+the minimum number of recombinations required
 to explain an input dataset [CITATIONS], and inferring the corresponding
@@
 The Griffiths ARG encoding is an elegant and mathematically
-economical description of the coalescent with recombination, but
-it has significant drawbacks as the basis for a general purpose
+economical description of the coalescent with recombination on a single
+chromosome, but it has significant drawbacks as the basis for a general purpose
 storage format.
@@
 model.
+%% Note - I'm not sure this is true, if the nodes are *events* rather than
+%% genomes, unless you want events (rather than genomes) to occur at
+%% integer generations.
 \item We can also only have one crossover per recombination event.
-This is unrealistic because things like gene conversion are common,
-and means we can't have back and forth recombination between
-two parents.
+This is unrealistic because things like gene conversion are common. It also
+means we can't have back and forth recombination between
+two parents, or represent the transmission of multiple chromosomes.
 \item The passage of ancestral material along edges requires
@@
 \includegraphics[width=\linewidth]{illustrations/ARG_recomb_node_deletion}
-\caption{\label{fig-recombinati on-nodes}
+\caption{\label{fig-recombination-nodes}
 Recombination nodes are optional in an edge-annotated ARG. (a) A simulated

### 2022-02-02 paper.tex

@@
 then continues until there is only one lineage left (the Grand Most Recent
-Common Ancestor, GMCRA), which is guaranteed to
+Common Ancestor, GMRCA), which is guaranteed to
 happen in finite time because of the linear vs quadratic rates of branching
-and coalescing. The graph structure and breakpoints associated with with
+and coalescing. The graph structure and breakpoints associated with
 recombination nodes provides sufficient information to later recover the marginal
@@
 in Hudson's algorithm is
-$O(\rho^2)$~\citep[Eq.~5.10]{hein2004gene,baumdicker2021efficient}.
+$O(\rho^2)$~\citep{hein2004gene,baumdicker2021efficient}.
 This disparity in the number of events in the two formulations is
@@
 
-The limitations of the Griffiths ARG encoding directly stem from its
-rooting in the coalescent with recombination stochastic process.
+The limitations of the Griffiths ARG encoding are a direct consequence
+of its derivation from the coalescent with recombination stochastic process.
 Indeed, these limitations arise because it is such a \emph{good}
@@
 The \texttt{ClonalOrigin} method \citep{didelot2010inference,
-medinaaguayo2020speeding} uses the \texttt{ClonalFrame}
+medina2020speeding} uses the \texttt{ClonalFrame}
 model and data structure, though it is not as scalable as \texttt{ARGWeaver}.

### 2022-02-02 paper.tex
Add refs for tree->popgenstats

--- 2022-02-02 paper.tex
@@
 quite different in recombining organisms, where the analysis of genetic data
-depends on summaries ultimately derived from the trees
-% TODO cite the other papers that preceded ralph2020efficiently here.
-such as allele frequencies and site frequency spectra~\citep{ralph2020efficiently}, patterns of linkage
-disequilibrium~\citep{mcvean2002genealogical}, and principal
+depends on summaries ultimately derived from the
+trees~\citep{tajima1983evolutionary,tavare1984line,fu1995statistical}
+such as allele frequencies and site frequency spectra~\citep{ralph2020efficiently},
+patterns of linkage disequilibrium~\citep{mcvean2002genealogical}, and principal
 components~\citep{mcvean2009genealogical}. Genealogical analysis is equally

### 2022-02-02 paper.tex
Add ref for Achaz

@@
 trees~\citep{tajima1983evolutionary,tavare1984line,fu1995statistical}
-such as allele frequencies and site frequency spectra~\citep{ralph2020efficiently},
+such as allele frequencies and site frequency
+spectra~\citep{achaz2009frequency,ralph2020efficiently},
 patterns of linkage disequilibrium~\citep{mcvean2002genealogical}, and principal

### 2022-02-02 paper.tex
Modified first sentence, added references

--- 2022-02-02 paper.tex
@@
 Originally rigorously defined as
-a stochastic process (the coalescent with recombination) together with its 
-representation, an ARG is now understood to refer to any
+a representation of the coalescent with recombination
+stochastic process as a graph, an ARG is now understood to refer to any
 concrete realisation of a recombinant genetic ancestry.
@@
 We provide a simple alternative definition of an ARG in which a node
-corresponds to one of an individual's monoploid genomes, an edge
+corresponds to one of an individuals monoploid genomes, an edge
 % jk-note: this is clumsy
@@
 \section*{Introduction}
-Inferring the genealogical relationships between sampled genomes in the form of an
-evolutionary tree is a necessary prerequisite for many analyses of non (or
-rarely) recombining organisms [refs]. The tree captures what is known and
+Inferring the genetic relationships between sampled genomes in the form of an
+evolutionary tree is a necessary prerequisite for many analyses of species \citep{rannala2003genetics}, non (or
+rarely) recombining sections of DNA \citep{cann1987mitochondrial,underhill2001annalsofhumangenetics} and viruses \citep{grenfell2004science}. The tree captures what is known and
 knowable about the genetic ancestry of the sample in an elegant and concise
@@
 quite different in recombining organisms, where the analysis of genetic data
-depends on summaries ultimately derived from the
-trees~\citep{tajima1983evolutionary,tavare1984line,fu1995statistical}
-such as allele frequencies and site frequency spectra~\citep{ralph2020efficiently},
-patterns of linkage disequilibrium~\citep{mcvean2002genealogical}, and principal
+depends on summaries ultimately derived from the trees such as allele
+% TODO cite the other papers that preceded ralph2020efficiently here.
+frequencies~\citep{ralph2020efficiently}, site frequency spectra, patterns of linkage
+disequilibrium~\citep{mcvean2002genealogical} and principal
 components~\citep{mcvean2009genealogical}. Genealogical analysis is equally
@@
 \item Suggest a definition of ARGs that is free of these limitations,
-where the formulation is derived from a (normally diploid) pedigree rather than from
-coalescent approximations, and where the genetic
-material which is transmitted is described by annotations attached to
+where the formulation is derived from a diploid pedigree rather than
+the coalescent approximations and spatial annotations are attached to
 edges rather than nodes.
@@
 $[\ell, r)$ is a half-closed genomic interval and $a$ is an integer
-tracking the number of samples to which the lineage is ancestral over that interval.
+tracking the number of samples the lineage is ancestral to over the interval.
 (We also usually track the tree node associated with each segment, but
@@
 the present day as a series of random common ancestor or recombination events.
-Recombination events occur at a rate determined by the amount of ancestral material and
-the way in which it is distributed along a chromosome.
+Recombination events occur at a rate determined by the amount and spatial
+distribution of ancestral material among lineages.
 Let $L$ be the set of ancestors at a given time $t$. Recombination events
@@
 When $k = |L|$ lineages are present, common ancestor events
-occur at rate $\binom{k}{2}$. In a common ancestor event, two lineages
-are chosen uniformly at random and their ancestry segments merged.
+occur at rate $\binom{k}{2}$. In a common ancestor event two lineages
+are chosed uniformly at random and their ancestry segments merged.
 If we have overlapping intervals of ancestry from the two lineages,
@@
 
-The state-space of this process is much simpler than Hudson's algorithm, which
+The state-space of the ARG is much simpler than Hudson's algorithm, which
 greatly facilitates mathematical reasoning. This simplicity comes at a
@@
 NP-hard~\citep{wang2001perfect}---problem of computing
-the minimum number of recombinations required
+the miniumum number of recombinations required
 to explain an input dataset [CITATIONS], and inferring the corresponding
@@
 The Griffiths ARG encoding is an elegant and mathematically
-economical description of the coalescent with recombination on a single
-chromosome, but it has significant drawbacks as the basis for a general purpose
+economical description of the coalescent with recombination, but
+it has significant drawbacks as the basis for a general purpose
 storage format.
@@
 model.
-%% Note - I'm not sure this is true, if the nodes are *events* rather than
-%% genomes, unless you want events (rather than genomes) to occur at
-%% integer generations.
 \item We can also only have one crossover per recombination event.
-This is unrealistic because things like gene conversion are common. It also
-means we can't have back and forth recombination between
-two parents, or represent the transmission of multiple chromosomes.
+This is unrealistic because things like gene conversion are common,
+and means we can't have back and forth recombination between
+two parents.
 \item The passage of ancestral material along edges requires
@@
 \includegraphics[width=\linewidth]{illustrations/ARG_recomb_node_deletion}
-\caption{\label{fig-recombination-nodes}
+\caption{\label{fig-recombinati on-nodes}
 Recombination nodes are optional in an edge-annotated ARG. (a) A simulated

### 2022-02-03 paper.tex

@@
 is free from the limitations discussed in the previous section.
-\cite{mathieson2020ancestry} and others [CITE] have also
-suggested this approach, although without explicitly defining
+\cite{mathieson2020ancestry} and others 
+\citep{wakeley2012genetics,gusfield2014recombinatorics,speed2015naturereviewsgenetics}
+have also suggested this approach, although without explicitly defining
 the roles of nodes and the details of how ancestry

### 2022-02-08 argutils/viz.py
Allow y times to be used and hand tweak X

@@ -8,10 +8,34 @@ import networkx as nx
 import numpy as np
 
 
-def draw(ts, ax):
+def draw(ts, ax, use_ranked_times=None, tweak_x=None):
+    """
+    If use_ranked times is True, the y axis uses the time ranks, with the
+    same times sharing a rank. If False, it uses the true (tree sequence)
+    times. If None, times from the tree sequence are not used and the
+    standard dot layout is used.
+    
+    tweak_x is a dict of {u: x_adjustment_percent} which allows
+    the x position of node u to be hand-adjusted by adding or
+    subtracting a percentage of the total x width of the plot 
+    """
     G = convert_nx(ts)
     labels = {j: f"{j}" for j in range(ts.num_nodes)}
     pos = nx_get_dot_pos(G)
+    if use_ranked_times is not None:
+        if use_ranked_times:
+            _, inv = np.unique(ts.tables.nodes.time, return_inverse=True)
+            ranked_times = (np.cumsum(np.bincount(inv)) - 1)[inv]
+            for i, p in pos.items():
+                p[1] = ranked_times[i]
+        else:
+            times = ts.tables.nodes.time
+            for i, p in pos.items():
+                p[1] = times[i]
+    if tweak_x:
+        plot_width = np.ptp([x for x, _ in pos.values()])
+        for node, tweak_val in tweak_x.items():
+            pos[node] = pos[node] + np.array([(tweak_val/100*plot_width), 0])
     nx.draw(
         G,
         pos,

### 2022-02-08 paper.tex

@@
 \section*{ARG inference methods}
-Using the terminology developed here to classify ARGs, we review methods developed
-to \emph{infer} ARGs from sequencing data. We focus our attention on some recent methods and discuss the
-properties of the ARGs inferred.
+Using the terminology developed here to classify ARGs, we now review methods developed
+to explicitly \emph{infer} ARGs from sequencing data.
 
-The problem of reconstructing ARGs for samples of recombining sequences has been of significant interest
-essentially since the ARG was first defined. Early methods focussed on finding the most
-\emph{parsimonious} ARGs, i.e.\ those minimising the number of
-posited recombination events in the history of the sample \citep{hein1990reconstructing, song2003parsimonious,
-lyngso2005minimum}. However, reconstructing a most parsimonious ARG for a given data set
-is NP-hard \citep{wang2001perfect}, and in order to allow for larger input datasets,
-subsequent parsimony-based methods have resorted to various relaxations and heuristics \citep{hein1993heuristic,
-song2005efficient, wu2008association, camara2016inference, thao2019hybrid, ignatieva2021kwarg}.
-The scalability of these methods varies substantially based on the heuristic in use
-and the underlying data structure, with the most efficient tools limited to analysing hundreds of sequences.
+\subsection*{Parsimony and heuristics}
+The problem of reconstructing ARGs for samples of recombining sequences has been of significant
+ interest essentially since the ARG was first defined. Early methods focussed on finding the most 
+ \emph{parsimonious} ARGs, i.e.\ those minimising the number of posited recombination events in 
+ the history of the sample \citep{hein1990reconstructing}. Two main approaches to explicitly reconstructing 
+ parsimonious ARGs then emerged: building the ARG either \emph{backwards-in-time} or \emph{along-the-genome}. 
+ The former approach, introduced by \citet{lyngso2005minimum} and implemented in the program 
+ \texttt{Beagle}, starts with the input data matrix and reduces it through row and column operations 
+ corresponding to coalescence, mutation and recombination events, until the empty matrix is reached; 
+ the corresponding sequence of events can then be used to construct an ARG from the bottom up. The 
+ latter approach, introduced by \citet{song2003parsimonious} and first implemented in the program 
+ \texttt{RecMinPath} \citep{song2005constructing}, starts with an initial local tree at some starting 
+ position, and refines the tree moving left and right along the genome, changing the topology of the 
+ tree when necessary through \emph{subtree prune and regraft} (SPR) operations (corresponding to 
+ recombination events).
+
+Reconstructing a most parsimonious ARG for a given data set is NP-hard \citep{wang2001perfect}, 
+and in order to allow for larger input datasets, subsequent parsimony-based methods have resorted 
+to various relaxations and heuristics. Those utilising variants of the backwards-in-time approach 
+include the programs \texttt{SHRUB} \citep{song2005efficient}, \texttt{GAMARG} \citep{thao2019hybrid}, 
+and \texttt{KwARG} \citep{ignatieva2021kwarg}, and the method described by \citet{wu2008association}. 
+Tools constructing genealogies along-the-genome include \texttt{RecPars} \citep{hein1993heuristic}, 
+\texttt{RENT} \citep{wu2011new} and its sequel \texttt{RENT+} \citep{mirzaei2017rent}. Other approaches 
+include \texttt{TARGet} \citep{camara2016inference}, which implements a method based on topological 
+data analysis. The scalability of these methods varies substantially based on the heuristics in use 
+and the underlying data structure, with the most efficient tools limited to analysing hundreds of 
+sequences.
+
 A number of heuristic methods have also been developed which do not (explicitly) focus on reconstructing 
-the most parsimonious ARGs, but rather aim for computational efficiency
-while seeking to achieve good accuracy of the inferred genealogies
-\citep{minichiello2006mapping, parida2008estimating, wu2011new, mirzaei2017rent}; some of these methods \citep{kelleher2019inferring, speidel2019method, zhang2021biobank} can be practically
-used on megabase scale data and tens to hundreds of thousands of samples under human-like
-parameters.
+the most parsimonious ARGs, but rather aim for computational efficiency while seeking to achieve good 
+accuracy of the inferred genealogies. Such methods include \texttt{Margarita} \citep{minichiello2006mapping} 
+and the approach of \citet{parida2008estimating}; the more recently developed \texttt{Relate} 
+\citep{speidel2019method} and \texttt{tsinfer} \citep{kelleher2019inferring} use a combination of both 
+backwards-in-time and along-the-genome approaches, and can be practically used on megabase scale data 
+and tens to hundreds of thousands of samples under human-like parameters.
+Another class of methods has recently emerged that use a \emph{threading} approach, constructing ARGs 
+by adding sequences to the graph one-by-one. Originally introduced by \citet{rasmussen2014genome} as a 
+way of generating MCMC proposals for ARGs, a similar idea was leveraged in \texttt{ARGneedle} 
+\citep{zhang2021biobank} to reconstruct ARGs for biobank-scale data.
 % AI: just to note, these papers do not all agree in their definition of an ARG.
@@
 % Should we also mention that most of these infer just the topology, but some also the times?
-% How much more description of the parsimony methods would be good? Maybe a mention of them splitting into
-% "back in time" and "across local trees" types?
 
+\subsection*{Model-based methods}
 An alternative approach is to treat the ancestry as a latent parameter to be averaged out

### 2022-02-08 argutils/viz.py
Make the returned object a directed graph
But don't draw arrows by default

@@ -8,7 +8,7 @@ import networkx as nx
 import numpy as np
 
 
-def draw(ts, ax, use_ranked_times=None, tweak_x=None):
+def draw(ts, ax, use_ranked_times=None, tweak_x=None, arrows=False):
     """
     If use_ranked times is True, the y axis uses the time ranks, with the
     same times sharing a rank. If False, it uses the true (tree sequence)
@@ -36,6 +36,7 @@ def draw(ts, ax, use_ranked_times=None, tweak_x=None):
         plot_width = np.ptp([x for x, _ in pos.values()])
         for node, tweak_val in tweak_x.items():
             pos[node] = pos[node] + np.array([(tweak_val/100*plot_width), 0])
+    # Draw just the nodes
     nx.draw(
         G,
         pos,
@@ -45,14 +46,23 @@ def draw(ts, ax, use_ranked_times=None, tweak_x=None):
         font_size=9,
         labels=labels,
         ax=ax,
+        edgelist=[],
+    )
+    # Now add the edges
+    nx.draw_networkx_edges(
+        G,
+        pos,
+        edgelist=G.edges(),
+        arrowstyle="-|>" if arrows else "-",
+        ax=ax,
     )
 
 
 def convert_nx(ts):
     """
-    Returns the specified tree sequence as an networkx graph.
+    Returns the specified tree sequence as an networkx directed graph.
     """
-    G = nx.Graph()
+    G = nx.DiGraph()
     edges = collections.defaultdict(list)
     for edge in ts.edges():
         edges[(edge.child, edge.parent)].append((edge.left, edge.right))

### 2022-02-08 argutils/viz.py
Add ability to label nodes

@@ -6,7 +6,7 @@ import collections
 import tskit
 import networkx as nx
 import numpy as np
-
+import string
 
 def draw(ts, ax, use_ranked_times=None, tweak_x=None, arrows=False):
     """
@@ -17,10 +17,19 @@ def draw(ts, ax, use_ranked_times=None, tweak_x=None, arrows=False):
     
     tweak_x is a dict of {u: x_adjustment_percent} which allows
     the x position of node u to be hand-adjusted by adding or
-    subtracting a percentage of the total x width of the plot 
+    subtracting a percentage of the total x width of the plot
+    
+    If a metadata key called "name" exists for the node, it is taken as
+    a node label, otherwise the node ID will be used as a label instead.
     """
     G = convert_nx(ts)
-    labels = {j: f"{j}" for j in range(ts.num_nodes)}
+    labels = {}
+    for nd in ts.nodes():
+        try:
+            labels[nd.id] = str(nd.metadata["name"])
+        except (TypeError, KeyError):
+            labels[nd.id] = str(nd.id)
+
     pos = nx_get_dot_pos(G)
     if use_ranked_times is not None:
         if use_ranked_times:
@@ -56,6 +65,29 @@ def draw(ts, ax, use_ranked_times=None, tweak_x=None, arrows=False):
         arrowstyle="-|>" if arrows else "-",
         ax=ax,
     )
+
+
+def label_nodes(ts, labels=None):
+    """
+    Adds a metadata item called "name" to each node, whose value
+    is given by the labels dictionary passed in. If labels is None (default),
+    use the dictionary {0: 'A', 1: 'B', 2: 'C', ... 26: 'Z'}.
+    Any nodes without a corresponding key in the labels dictionary will
+    simply have their metadata value set to their node id.
+    
+    Note that this means that if no labels are given, nodes 26 onwards
+    will be labelled with numbers rather than ascii uppercase letters.
+    """
+    if labels is None:
+        labels = {i: lab for i, lab in enumerate(string.ascii_uppercase)}
+    tables = ts.dump_tables()
+
+    tables.nodes.clear()
+    for i, nd in enumerate(ts.tables.nodes):
+        m = nd.metadata or {}
+        # assume we can set metadata to a dict: e.g. that node.metadata_schema is json
+        tables.nodes.append(nd.replace(metadata={**m, "name": labels.get(i, i)}))
+    return tables.tree_sequence()
 
 
 def convert_nx(ts):

### 2022-02-08 argutils/ancestry.py
Remove resolve() method and return tree sequence

@@ -1,10 +1,8 @@
 """
 Utilities for generating and converting ARGs in various formats.
 """
-import itertools
 import random
 import math
-import collections
 import dataclasses
 from typing import List
 from typing import Any
@@ -275,14 +273,11 @@ def sim_coalescent(n, rho, L, seed=None, resolved=True):
 
     for node in nodes:
         tables.nodes.add_row(flags=node.flags, time=node.time, metadata=node.metadata)
-    if resolved:
-        tables.sort()
-        # TODO not sure if this is the right thing to do, but it makes it easier
-        # to compare with examples.
-        tables.edges.squash()
-        return tables.tree_sequence()
-    else:
-        return tables
+    tables.sort()
+    # TODO not sure if this is the right thing to do, but it makes it easier
+    # to compare with examples.
+    tables.edges.squash()
+    return tables.tree_sequence()
 
 
 @dataclasses.dataclass
@@ -403,123 +398,6 @@ def sim_wright_fisher(n, N, L, seed=None):
     return tables.tree_sequence()
 
 
-@dataclasses.dataclass
-class Recombination:
-    time: float
-    child: int
-    breakpoint: float
-    left_parent: int
-    right_parent: int
-
-
-@dataclasses.dataclass
-class CommonAncestor:
-    time: float
-    parent: int
-    children: List[int]
-
-
-def resolve(tables):
-    """
-    Converts the specified implicit-ancestry ARG to a fully resolved
-    tskit ARG.
-
-    Each node can have at most two outbound edges. If there is one
-    outbound edge, this must have left and right coordinates
-    (-inf, inf). When we have two outbound edges, we must have
-    coordinates (-inf, x) on one and (x, inf) on the other.
-    This tells us that the child node is a recombinant, with the
-    specified parent nodes.
-
-    """
-    out = tables.copy()
-    out.edges.clear()
-    time = tables.nodes.time
-
-    parent_edges = [[] for _ in tables.nodes]
-    child_edges = [[] for _ in tables.nodes]
-    for edge in tables.edges:
-        parent_edges[edge.child].append(edge)
-        child_edges[edge.parent].append(edge)
-
-    events = []
-    for node, edges in enumerate(parent_edges):
-        if len(edges) > 2:
-            raise ValueError("Only single breakpoints per recombination supported")
-        if len(edges) == 2:
-            edges = sorted(edges, key=lambda x: x.left)
[…]

### 2022-02-08 argutils/tests.py

@@ -13,7 +13,6 @@ class TestSimulate:
         assert ts.num_samples == 4
         assert ts.sequence_length == 5
         assert ts.num_trees > 1
-        print(ts.draw_text())
         assert all(tree.num_roots == 1 for tree in ts.trees())
 
     def test_basic_wright_fisher(self):
@@ -24,10 +23,10 @@ class TestSimulate:
         assert all(tree.num_roots == 1 for tree in ts.trees())
 
     def test_basic_coalescent_unresolved(self):
-        tables = argutils.sim_coalescent(4, L=5, rho=0.1, seed=1, resolved=False)
+        ts = argutils.sim_coalescent(4, L=5, rho=0.1, seed=1, resolved=False)
         # Very basic - real test is the resolution process.
-        assert len(tables.nodes) > 4
-        assert len(tables.edges) > 4
+        assert ts.num_nodes > 4
+        assert ts.num_edges > 4
 
     @pytest.mark.parametrize("seed", range(1, 10))
     @pytest.mark.parametrize(
@@ -45,12 +44,12 @@ class TestSimulate:
     )
     def test_coalescent_resolved_equal(self, n, L, seed):
         rho = 0.1
-        tables = argutils.sim_coalescent(n, L=L, rho=rho, seed=seed, resolved=False)
-        ts1 = argutils.sim_coalescent(n, L=L, rho=rho, seed=seed, resolved=True)
-        ts2 = argutils.resolve(tables)
+        ts1 = argutils.sim_coalescent(n, L=L, rho=rho, seed=seed, resolved=False)
+        ts2 = argutils.sim_coalescent(n, L=L, rho=rho, seed=seed, resolved=True)
+        ts3 = ts1.simplify(keep_unary=True)
         # print(ts1.draw_text())
         # print(ts2.draw_text())
-        ts1.tables.assert_equals(ts2.tables)
+        ts2.tables.assert_equals(ts3.tables, ignore_provenance=True)
 
 
 def gmrca_example_resolved():
@@ -76,7 +75,6 @@ def gmrca_example_resolved():
     tables.edges.add_row(0, 1, 4, 2)
     tables.edges.add_row(1, 2, 3, 1)
     tables.edges.add_row(1, 2, 5, 3)
-
     tables.sort()
     return tables.tree_sequence()
 
@@ -104,7 +102,8 @@ def gmrca_example_unresolved():
     tables.edges.add_row(0, 2, 5, 4)
[…]

### 2022-02-08 paper.tex

@@
 \usepackage[right]{lineno}
+\usepackage{tikz}
 
@@
 
+\begin{figure}[ht]
+\centering
+\begin{tikzpicture}
+	% Samples at x = 0, 4, and 6
+	\node [circle, fill, scale=0.2,label=above right:{\small A}] at (0,0) {A};
+	\draw (-0.4, -0.2) -- (0.4, -0.2) -- (0.4, -0.3) -- (-0.4, -0.3) -- (-0.4, -0.2);
+	\draw (-0.1, -0.1) -- (-0.1, -0.4);
+	\node [label=below:{\small $v$}] at (-0.1, -0.2) {};
+	\node [circle, fill, scale=0.2,label=above right:{\small B}] at (4,0) {B};
+	\draw (3.6, -0.2) -- (4.4, -0.2) -- (4.4, -0.3) -- (3.6, -0.3) -- (3.6, -0.2);
+	\node [circle, fill, scale=0.2,label=above left:{\small C}] at (6,0) {C};
+	\draw (5.6, -0.2) -- (6.4, -0.2) -- (6.4, -0.3) -- (5.6, -0.3) -- (5.6, -0.2);
+
+	% B and C merge into D
+	\draw (4,0) -- (4, 1.2) -- (6, 1.2) -- (6,0);
+	\node [circle, fill, scale=0.2,label=above right:{\small D}] at (5,1.2) {D};
+	\draw (4.6, 1) -- (5.4, 1) -- (5.4, 0.9) -- (4.6, 0.9) -- (4.6, 1);
+	\draw (5.1, 1.1) -- (5.1, 0.8);
+	\node [label=below:{\small $w$}] at (5.1, 1) {};
+
+	% A splits into E and F
+	\draw (0, 0) -- (0, 1.5) -- (-1,1.5) -- (1,1.5);
+	\node [circle, fill, scale=0.2,label=above left:{\small E}] at (-1,1.5) {E};
+	\draw (-1.4, 1.3) -- (-1.1, 1.3) -- (-1.1, 1.2) -- (-1.4, 1.2) -- (-1.4, 1.3);
+	\draw (-1.1, 1.25) -- (-0.6, 1.25);
+	\node [circle, fill, scale=0.2,label=above left:{\small F}] at (1,1.5) {F};
+	\draw (0.6, 1.25) -- (0.9, 1.25);
+	\draw (0.9, 1.3) -- (1.4, 1.3) -- (1.4, 1.2) -- (0.9, 1.2) -- (0.9, 1.3);
+	\draw (1.2, 1.4) -- (1.2, 1.1);
+	\node [label=below:{\small $x$}] at (1.2, 1.3) {};
+
+	% D splits into G and H
+	\draw (5, 1.2) -- (5, 2.4) -- (4,2.4) -- (6,2.4);
+	\node [circle, fill, scale=0.2,label=above left:{\small G}] at (4,2.4) {G};
+	\draw (3.6, 2.2) -- (4.1, 2.2) -- (4.1, 2.1) -- (3.6, 2.1) -- (3.6, 2.2);
+	\draw (4.1, 2.15) -- (4.4, 2.15);
+	\node [circle, fill, scale=0.2,label=above left:{\small H}] at (6,2.4) {H};
+	\draw (5.6, 2.15) -- (6.1, 2.15);
+	\draw (6.1, 2.2) -- (6.4, 2.2) -- (6.4, 2.1) -- (6.1, 2.1) -- (6.1, 2.2);
+
+	% F splits into I and J
+	\draw (1, 1.5) -- (1, 3.1) -- (0,3.1) -- (2,3.1);
+	\node [circle, fill, scale=0.2,label=above left:{\small I}] at (0,3.1) {I};
+	\draw (-0.4, 2.85) -- (0.2, 2.85);
+	\draw (0.2, 2.9) -- (0.4, 2.9) -- (0.4, 2.8) -- (0.2, 2.8) -- (0.2, 2.9);
+	\node [circle, fill, scale=0.2,label=above left:{\small J}] at (2,3.1) {J};
+	\draw (1.6, 2.85) -- (1.9, 2.85);
+	\draw (1.9, 2.9) -- (2.2, 2.9) -- (2.2, 2.8) -- (1.9, 2.8) -- (1.9, 2.9);
+	\draw (2.2, 2.85) -- (2.4, 2.85);
+
+	% E and I merge into K
+	\draw (-1, 1.5) -- (-1, 4.5) -- (0,4.5) -- (0,3.1);
+	\node [circle, fill, scale=0.2,label=above left:{\small K}] at (-0.5,4.5) {K};
+	\draw (-0.9, 4.3) -- (-0.6, 4.3) -- (-0.6, 4.2) -- (-0.9, 4.2) -- (-0.9, 4.3);
+	\draw (-0.6, 4.25) -- (-0.3, 4.25);
+	\draw (-0.3, 4.3) -- (-0.1, 4.3) -- (-0.1, 4.2) -- (-0.3, 4.2) -- (-0.3, 4.3);
+	\draw (-0.5, 4.4) -- (-0.5, 4.1);
+	\node [label=below:{\small $y$}] at (-0.5, 4.3) {};
+
+	% G and J merge into L
+	\draw (4,2.4) -- (4,5.4) -- (2,5.4) -- (2,3.1);
+	\node [circle, fill, scale=0.2,label=above left:{\small L}] at (3,5.4) {L};
+	\draw (2.6, 5.2) -- (2.9, 5.2) -- (2.9, 5.1) -- (2.6, 5.1) -- (2.6, 5.2);
+	\draw (2.9, 5.15) -- (3.4, 5.15);
+
+	% K splits into M and N
+	\draw (-0.5, 4.5) -- (-0.5, 6) -- (-1.5,6) -- (0.5,6);
+	\node [circle, fill, scale=0.2,label=above left:{\small M}] at (-1.5,6) {M};
+	\draw (-1.9, 5.75) -- (-1.3, 5.75);
+	\draw (-1.3, 5.8) -- (-1.1, 5.8) -- (-1.1, 5.7) -- (-1.3, 5.7) -- (-1.3, 5.8);
+	\node [circle, fill, scale=0.2,label=above left:{\small N}] at (0.5,6) {N};
+	\draw (0.1, 5.8) -- (0.4, 5.8) -- (0.4, 5.7) -- (0.1, 5.7) -- (0.1, 5.8);
+	\draw (0.4, 5.75) -- (0.9, 5.75);
+
+	% L and N merge into O
+	\draw (0.5, 6) -- (0.5, 7.2) -- (3,7.2) -- (3,5.4);
+	\node [circle, fill, scale=0.2,label=above left:{\small O}] at (1.75,7.2) {O};
+	\draw (1.35, 6.95) -- (2.15, 6.95);
+
+	% M and H continue to the end
+	\draw (-1.5, 6) -- (-1.5, 8);
+	\draw (6, 2.4) -- (6, 8);
+
+	% Edge annotations above each node
+	\node [label=left:{\small$\{[0,L)\};1$}] at (0.2,0.6) {}; % A
+	\node [label=left:{\small$\{[0,L)\};1$}] at (4.2,0.4) {}; % B
+	\node [label=right:{\small$\{[0,L)\};1$}] at (5.8,0.4) {}; % C
+	\node [label=left:{\small$\{[0,L)\};2$}] at (5.2,1.6) {}; % D
+	\node [label=left:{\small$\{[0,v)\};1$}] at (-0.8,2.3) {}; % E
+	\node [label=right:{\small$\{[v,L)\};1$}] at (0.8,2.1) {}; % F
+	\node [label=right:{\small$\{[0,w)\};2$}] at (3.8,3.9) {}; % G
+	\node [label=left:{\small$\{[w,L)\};2$}] at (6.2,5) {}; % H
+	\node [label=right:{\small$\{[x,L)\};1$}] at (-0.2,3.8) {}; % I
+	\node [label=right:{\small$\{[v,x)\};1$}] at (1.8,4.1) {}; % J
+	\node [label=left:{\small$\{[0,v), [x, L)\};1$}] at (-0.3,5.2) {}; % K
+	\node [label=right:{\small$\{[0,v)\};2$}] at (2.8,6.4) {}; % L
+	\node [label=left:{\small$\{[x,L)\};1$}] at (-1.3,7) {}; % M
+	\node [label=right:{\small$\{[0,v)\};1$}] at (0.3,6.4) {}; % N
+
+	% Dashed lines for start and end times
+	\draw[dashed] (-2, 0) -- (11.1, 0);
+	\node [label=left:{$t = 0$}] at (-2,0) {};
+	\draw[dashed] (-2, 8) -- (11.1, 8);
+	\node [label=left:{$t = T$}] at (-2,8) {};
+
+	% Numbers of extant ancestors and links, from top to bottom
+	\node[label=right:{Extant material}] at (7.5, 8.2) {};
+	\node[label=right:{2 \; $2 L - x - w$}] at (7.5, 7.6) {};
+	\node[label=right:{4 \; $2 L + 2 v - x - w$}] at (7.5, 6.6) {};
+	\node[label=right:{3 \; $2 L + v - w$}] at (7.5, 5.7) {};
+	\node[label=right:{4 \; $2 L + x - v$}] at (7.5, 4.95) {};
+	\node[label=right:{5 \; $2 L$}] at (7.5, 3.8) {};
+	\node[label=right:{4 \; $2 L$}] at (7.5, 2.75) {};
+	\node[label=right:{3 \; $2 L$}] at (7.5, 1.95) {};
+	\node[label=right:{2 \; $2 L$}] at (7.5, 1.35) {};
+	\node[label=right:{3 \; $3 L$}] at (7.5, 0.6) {};
+
+	% Gray dashed lines to visually separate holding times
+	\draw[color=gray, dashed] (7.5, 1.2) -- (11.1, 1.2); % D
+	\draw[color=gray, dashed] (7.5, 1.5) -- (11.1, 1.5); % E and F
+	\draw[color=gray, dashed] (7.5, 2.4) -- (11.1, 2.4); % G and H
+	\draw[color=gray, dashed] (7.5, 3.1) -- (11.1, 3.1); % I and J
+	\draw[color=gray, dashed] (7.5, 4.5) -- (11.1, 4.5); % K
+	\draw[color=gray, dashed] (7.5, 5.4) -- (11.1, 5.4); % L
+	\draw[color=gray, dashed] (7.5, 6) -- (11.1, 6); % M and N
+	\draw[color=gray, dashed] (7.5, 7.2) -- (11.1, 7.2); % O
+\end{tikzpicture}
+\begin{tikzpicture}
+	% Samples at x = 0, 4, and 6
+	\node [circle, fill, scale=0.2,label=above right:{\small A}] at (0,0) {A};
+	\draw (-0.4, -0.2) -- (0.4, -0.2) -- (0.4, -0.3) -- (-0.4, -0.3) -- (-0.4, -0.2);
+	\draw (-0.1, -0.1) -- (-0.1, -0.4);
+	\node [label=below:{\small $v$}] at (-0.1, -0.2) {};
+	\node [circle, fill, scale=0.2,label=above right:{\small B}] at (4,0) {B};
+	\draw (3.6, -0.2) -- (4.4, -0.2) -- (4.4, -0.3) -- (3.6, -0.3) -- (3.6, -0.2);
+	\node [circle, fill, scale=0.2,label=above left:{\small C}] at (6,0) {C};
+	\draw (5.6, -0.2) -- (6.4, -0.2) -- (6.4, -0.3) -- (5.6, -0.3) -- (5.6, -0.2);
+
+	% B and C merge into D
+	\draw (4,0) -- (4, 1.2) -- (6, 1.2) -- (6,0);
+	\node [circle, fill, scale=0.2,label=above right:{\small D}] at (5,1.2) {D};
+	\draw (4.6, 1) -- (5.4, 1) -- (5.4, 0.9) -- (4.6, 0.9) -- (4.6, 1);
+	\draw (5.1, 1.1) -- (5.1, 0.8);
+	\node [label=below:{\small $w$}] at (5.1, 1) {};
+
+	% A splits into E and F
+	\draw (0, 0) -- (0, 1.5) -- (-1,1.5) -- (1,1.5);
+	\node [circle, fill, scale=0.2,label=above left:{\small E}] at (-1,1.5) {E};
+	\draw (-1.4, 1.3) -- (-1.1, 1.3) -- (-1.1, 1.2) -- (-1.4, 1.2) -- (-1.4, 1.3);
+	\draw (-1.1, 1.25) -- (-0.6, 1.25);
+	\node [circle, fill, scale=0.2,label=above left:{\small F}] at (1,1.5) {F};
+	\draw (0.6, 1.25) -- (0.9, 1.25);
+	\draw (0.9, 1.3) -- (1.4, 1.3) -- (1.4, 1.2) -- (0.9, 1.2) -- (0.9, 1.3);
+	\draw (1.2, 1.4) -- (1.2, 1.1);
+	\node [label=below:{\small $x$}] at (1.2, 1.3) {};
+
+	% D splits into G and H
+	\draw (5, 1.2) -- (5, 2.4) -- (4,2.4) -- (6,2.4);
+	\node [circle, fill, scale=0.2,label=above left:{\small G}] at (4,2.4) {G};
+	\draw (3.6, 2.2) -- (4.1, 2.2) -- (4.1, 2.1) -- (3.6, 2.1) -- (3.6, 2.2);
+	\draw (4.1, 2.15) -- (4.4, 2.15);
+	\node [circle, fill, scale=0.2,label=above left:{\small H}] at (6,2.4) {H};
+	\draw (5.6, 2.15) -- (6.1, 2.15);
+	\draw (6.1, 2.2) -- (6.4, 2.2) -- (6.4, 2.1) -- (6.1, 2.1) -- (6.1, 2.2);
+	\draw [color=red](5.8, 2.3) -- (5.8, 2.0);
+	\node [label={[red]below:{\small $z$}}] at (5.8, 2.1) {};
+
+	% F splits into I and J
+	\draw (1, 1.5) -- (1, 3.1) -- (0,3.1) -- (2,3.1);
+	\node [circle, fill, scale=0.2,label=above left:{\small I}] at (0,3.1) {I};
+	\draw (-0.4, 2.85) -- (0.2, 2.85);
+	\draw (0.2, 2.9) -- (0.4, 2.9) -- (0.4, 2.8) -- (0.2, 2.8) -- (0.2, 2.9);
+	\node [circle, fill, scale=0.2,label=above left:{\small J}] at (2,3.1) {J};
+	\draw (1.6, 2.85) -- (1.9, 2.85);
+	\draw (1.9, 2.9) -- (2.2, 2.9) -- (2.2, 2.8) -- (1.9, 2.8) -- (1.9, 2.9);
+	\draw (2.2, 2.85) -- (2.4, 2.85);
+
+	% E and I merge into K
+	\draw (-1, 1.5) -- (-1, 4.5) -- (0,4.5) -- (0,3.1);
+	\node [circle, fill, scale=0.2,label=above left:{\small K}] at (-0.5,4.5) {K};
+	\draw (-0.9, 4.3) -- (-0.6, 4.3) -- (-0.6, 4.2) -- (-0.9, 4.2) -- (-0.9, 4.3);
+	\draw (-0.6, 4.25) -- (-0.3, 4.25);
+	\draw (-0.3, 4.3) -- (-0.1, 4.3) -- (-0.1, 4.2) -- (-0.3, 4.2) -- (-0.3, 4.3);
+	\draw (-0.5, 4.4) -- (-0.5, 4.1);
+	\node [label=below:{\small $y$}] at (-0.5, 4.3) {};
+
+	% H splits into P and Q. Red Q node drawn further down so that it overlays the black line underneath
+	\draw (6, 2.4) -- (6, 3.8) -- (7, 3.8);
+	\draw [color=red](6, 3.8) -- (5, 3.8);
+	\node [color=red, circle, fill, scale=0.2,label={[red]above left:{\small P}}] at (5,3.8) {P};
+	\draw (4.6, 3.55) -- (5.4, 3.55);
+	\draw (6.6, 3.55) -- (7.1, 3.55);
+	\draw (7.1, 3.6) -- (7.4, 3.6) -- (7.4, 3.5) -- (7.1, 3.5) -- (7.1, 3.6);
+
+	% G and J merge into L
+	\draw (4,2.4) -- (4,5.4) -- (2,5.4) -- (2,3.1);
+	\node [circle, fill, scale=0.2,label=above left:{\small L}] at (3,5.4) {L};
+	\draw (2.6, 5.2) -- (2.9, 5.2) -- (2.9, 5.1) -- (2.6, 5.1) -- (2.6, 5.2);
+	\draw (2.9, 5.15) -- (3.4, 5.15);
+
+	% K splits into M and N
+	\draw (-0.5, 4.5) -- (-0.5, 6) -- (-1.5,6) -- (0.5,6);
+	\node [circle, fill, scale=0.2,label=above left:{\small M}] at (-1.5,6) {M};
+	\draw (-1.9, 5.75) -- (-1.3, 5.75);
+	\draw (-1.3, 5.8) -- (-1.1, 5.8) -- (-1.1, 5.7) -- (-1.3, 5.7) -- (-1.3, 5.8);
+	\node [circle, fill, scale=0.2,label=above left:{\small N}] at (0.5,6) {N};
+	\draw (0.1, 5.8) -- (0.4, 5.8) -- (0.4, 5.7) -- (0.1, 5.7) -- (0.1, 5.8);
+	\draw (0.4, 5.75) -- (0.9, 5.75);
+
+	% P and L merge into R. R node drawn further down so that it overlays the black lines underneath
+	\draw [color=red](5, 3.8) -- (5, 6.6) -- (4, 6.6);
+	\draw (4,6.6) -- (3,6.6) -- (3, 5.4);
+	\draw (3.6, 6.4) -- (3.9, 6.4) -- (3.9, 6.3) -- (3.6, 6.3) -- (3.6, 6.4);
+	\draw (3.9, 6.35) -- (4.4, 6.35);
+
+	% R and N merge into O. O node drawn further down so that it overlays the red line underneath
+	\draw (0.5, 6) -- (0.5, 7.2) -- (4,7.2) -- (4,6.6);
+	\draw (1.85, 6.95) -- (2.65, 6.95);
+	% R node can now be added
+	\node [color=red, circle, fill, scale=0.2,label={[red]above left:{\small R}}] at (4,6.6) {R};
+
+	% M, O, and Q continue to the end
+	\draw (-1.5, 6) -- (-1.5, 8);
+	\draw [color=red](2.25, 7.2) -- (2.25, 8);
+	\draw (7, 3.8) -- (7, 8);
+	% O and Q nodes can now be added
+	\node [circle, fill, scale=0.2,label=above left:{\small O}] at (2.25,7.2) {O};
+	\node [color=red, circle, fill, scale=0.2,label={[red]above left:{\small Q}}] at (7,3.8) {Q};
+
+	% Dashed lines for start and end times
+	\draw[dashed] (-2, 0) -- (11, 0);
+	\node [label=left:{$t = 0$}] at (-2,0) {};
+	\draw[dashed] (-2, 8) -- (11, 8);
+	\node [label=left:{$t = T$}] at (-2,8) {};
+
+	% Numbers of extant ancestors and links, from top to bottom
+	\node[label=right:{Extant material}] at (7.5, 8.2) {};
+	\node[label=right:{3 \; $3 L$}] at (7.5, 7.6) {};
+	\node[label=right:{4 \; $4 L$}] at (7.5, 6.9) {};
+	\node[label=right:{5 \; $5 L$}] at (7.5, 6.3) {};
+	\node[label=right:{4 \; $4 L$}] at (7.5, 5.7) {};
+	\node[label=right:{5 \; $5 L$}] at (7.5, 4.95) {};
+	\node[label=right:{6 \; $6 L$}] at (7.5, 4.15) {};
+	\node[label=right:{5 \; $5 L$}] at (7.5, 3.45) {};
+	\node[label=right:{4 \; $4 L$}] at (7.5, 2.75) {};
+	\node[label=right:{3 \; $3 L$}] at (7.5, 1.95) {};
+	\node[label=right:{2 \; $2 L$}] at (7.5, 1.35) {};
+	\node[label=right:{3 \; $3 L$}] at (7.5, 0.6) {};
+
+	% Gray dashed lines to visually separate holding times
+	\draw[color=gray, dashed] (7.5, 1.2) -- (11, 1.2); % D
+	\draw[color=gray, dashed] (7.5, 1.5) -- (11, 1.5); % E and F
+	\draw[color=gray, dashed] (7.5, 2.4) -- (11, 2.4); % G and H
+	\draw[color=gray, dashed] (7.5, 3.1) -- (11, 3.1); % I and J
+	\draw[color=gray, dashed] (7.5, 3.8) -- (11, 3.8);	% P and Q
+	\draw[color=gray, dashed] (7.5, 4.5) -- (11, 4.5); % K
+	\draw[color=gray, dashed] (7.5, 5.4) -- (11, 5.4); % L
+	\draw[color=gray, dashed] (7.5, 6) -- (11, 6); % M and N
+	\draw[color=gray, dashed] (7.5, 6.6) -- (11, 6.6);	% R
+	\draw[color=gray, dashed] (7.5, 7.2) -- (11, 7.2); % O
+\end{tikzpicture}
+\caption{(Top) A realisation of the graph traversed by Hudson's algorithm started from a sample of three continuous chromosomes of length $L$ at time $t = 0$, and propagated until time $T$.
+The MRCA on the genetic interval $[x, z)$ is reached at node L, while that on $[0, x)$ is reached at node O, which is a root as it carries only fully coalesced ancestral segments.
+The non-ancestral segment $[x, y)$ above node K contributes to the rate of effective recombinations because it is trapped between ancestral segments.
+The two columns titled ``Extant material" are the number of extant lineages and the total length of genetic material on which an effective recombination can take place during each time interval.
+(Bottom) A corresponding realisation of a big ARG, which augments Hudson's algorithm by tracking nonancestral lineages.
+The result is a simpler state space and dynamics, at the cost of extra nodes and edges, highlighted in red, which do not affect the local tree at any site.}
+\label{hudson_vs_bigARG}
+\end{figure}
+
 \begin{figure}

### 2022-02-08 argutils/viz.py
Return positions from draw, and allow positions to be specified

@@ -8,8 +8,12 @@ import networkx as nx
 import numpy as np
 import string
 
-def draw(ts, ax, use_ranked_times=None, tweak_x=None, arrows=False):
+def draw(ts, ax, use_ranked_times=None, tweak_x=None, arrows=False, pos=None):
     """
+    Draw a graphical representation of a tree sequence. If a metadata key
+    called "name" exists for the node, it is taken as
+    a node label, otherwise the node ID will be used as a label instead.
+
     If use_ranked times is True, the y axis uses the time ranks, with the
     same times sharing a rank. If False, it uses the true (tree sequence)
     times. If None, times from the tree sequence are not used and the
@@ -19,8 +23,8 @@ def draw(ts, ax, use_ranked_times=None, tweak_x=None, arrows=False):
     the x position of node u to be hand-adjusted by adding or
     subtracting a percentage of the total x width of the plot
     
-    If a metadata key called "name" exists for the node, it is taken as
-    a node label, otherwise the node ID will be used as a label instead.
+    If pos is passed in, it should be a dictionary mapping nodes to
+    positions.
     """
     G = convert_nx(ts)
     labels = {}
@@ -30,7 +34,9 @@ def draw(ts, ax, use_ranked_times=None, tweak_x=None, arrows=False):
         except (TypeError, KeyError):
             labels[nd.id] = str(nd.id)
 
-    pos = nx_get_dot_pos(G)
+    if pos is None:
+        pos = nx_get_dot_pos(G)
+
     if use_ranked_times is not None:
         if use_ranked_times:
             _, inv = np.unique(ts.tables.nodes.time, return_inverse=True)
@@ -66,6 +72,7 @@ def draw(ts, ax, use_ranked_times=None, tweak_x=None, arrows=False):
         ax=ax,
     )
 
+    return pos
 
 def label_nodes(ts, labels=None):
     """

### 2022-02-09 argutils/ancestry.py
add numpy to ancestry

@@ -8,6 +8,7 @@ from typing import List
 from typing import Any
 
 import tskit
+import numpy as np
 
 NODE_IS_RECOMB = 1 << 1
 

### 2022-02-09 paper.tex
Put figure 1 in the right place

@@
 
-\begin{figure}
-\caption{\label{fig-sim-statespace} An illustration of the state space
-of Hudson's algorithm (A) and the big ARG (B) at a given instant. The
-state maintained by Hudson's algorithm is considerably more complicated,
-but is guaranteed to only simulate events corresponding to genetic
-ancestors of the sample.
+\begin{figure}[ht]
+\centering
+% FIXME this is a quick nasty hack to make the figure a bit smaller and
+% prevent flushing all the other figures to the end of the document.
+\scalebox{0.5}{
+\begin{tabular}{cc}
+\begin{tikzpicture}
+	% Samples at x = 0, 4, and 6
+	\node [circle, fill, scale=0.2,label=above right:{\small A}] at (0,0) {A};
+	\draw (-0.4, -0.2) -- (0.4, -0.2) -- (0.4, -0.3) -- (-0.4, -0.3) -- (-0.4, -0.2);
+	\draw (-0.1, -0.1) -- (-0.1, -0.4);
+	\node [label=below:{\small $v$}] at (-0.1, -0.2) {};
+	\node [circle, fill, scale=0.2,label=above right:{\small B}] at (4,0) {B};
+	\draw (3.6, -0.2) -- (4.4, -0.2) -- (4.4, -0.3) -- (3.6, -0.3) -- (3.6, -0.2);
+	\node [circle, fill, scale=0.2,label=above left:{\small C}] at (6,0) {C};
+	\draw (5.6, -0.2) -- (6.4, -0.2) -- (6.4, -0.3) -- (5.6, -0.3) -- (5.6, -0.2);
+
+	% B and C merge into D
+	\draw (4,0) -- (4, 1.2) -- (6, 1.2) -- (6,0);
+	\node [circle, fill, scale=0.2,label=above right:{\small D}] at (5,1.2) {D};
+	\draw (4.6, 1) -- (5.4, 1) -- (5.4, 0.9) -- (4.6, 0.9) -- (4.6, 1);
+	\draw (5.1, 1.1) -- (5.1, 0.8);
+	\node [label=below:{\small $w$}] at (5.1, 1) {};
+
+	% A splits into E and F
+	\draw (0, 0) -- (0, 1.5) -- (-1,1.5) -- (1,1.5);
+	\node [circle, fill, scale=0.2,label=above left:{\small E}] at (-1,1.5) {E};
+	\draw (-1.4, 1.3) -- (-1.1, 1.3) -- (-1.1, 1.2) -- (-1.4, 1.2) -- (-1.4, 1.3);
+	\draw (-1.1, 1.25) -- (-0.6, 1.25);
+	\node [circle, fill, scale=0.2,label=above left:{\small F}] at (1,1.5) {F};
+	\draw (0.6, 1.25) -- (0.9, 1.25);
+	\draw (0.9, 1.3) -- (1.4, 1.3) -- (1.4, 1.2) -- (0.9, 1.2) -- (0.9, 1.3);
+	\draw (1.2, 1.4) -- (1.2, 1.1);
+	\node [label=below:{\small $x$}] at (1.2, 1.3) {};
+
+	% D splits into G and H
+	\draw (5, 1.2) -- (5, 2.4) -- (4,2.4) -- (6,2.4);
+	\node [circle, fill, scale=0.2,label=above left:{\small G}] at (4,2.4) {G};
+	\draw (3.6, 2.2) -- (4.1, 2.2) -- (4.1, 2.1) -- (3.6, 2.1) -- (3.6, 2.2);
+	\draw (4.1, 2.15) -- (4.4, 2.15);
+	\node [circle, fill, scale=0.2,label=above left:{\small H}] at (6,2.4) {H};
+	\draw (5.6, 2.15) -- (6.1, 2.15);
+	\draw (6.1, 2.2) -- (6.4, 2.2) -- (6.4, 2.1) -- (6.1, 2.1) -- (6.1, 2.2);
+
+	% F splits into I and J
+	\draw (1, 1.5) -- (1, 3.1) -- (0,3.1) -- (2,3.1);
+	\node [circle, fill, scale=0.2,label=above left:{\small I}] at (0,3.1) {I};
+	\draw (-0.4, 2.85) -- (0.2, 2.85);
+	\draw (0.2, 2.9) -- (0.4, 2.9) -- (0.4, 2.8) -- (0.2, 2.8) -- (0.2, 2.9);
+	\node [circle, fill, scale=0.2,label=above left:{\small J}] at (2,3.1) {J};
+	\draw (1.6, 2.85) -- (1.9, 2.85);
+	\draw (1.9, 2.9) -- (2.2, 2.9) -- (2.2, 2.8) -- (1.9, 2.8) -- (1.9, 2.9);
+	\draw (2.2, 2.85) -- (2.4, 2.85);
+
+	% E and I merge into K
+	\draw (-1, 1.5) -- (-1, 4.5) -- (0,4.5) -- (0,3.1);
+	\node [circle, fill, scale=0.2,label=above left:{\small K}] at (-0.5,4.5) {K};
+	\draw (-0.9, 4.3) -- (-0.6, 4.3) -- (-0.6, 4.2) -- (-0.9, 4.2) -- (-0.9, 4.3);
+	\draw (-0.6, 4.25) -- (-0.3, 4.25);
+	\draw (-0.3, 4.3) -- (-0.1, 4.3) -- (-0.1, 4.2) -- (-0.3, 4.2) -- (-0.3, 4.3);
+	\draw (-0.5, 4.4) -- (-0.5, 4.1);
+	\node [label=below:{\small $y$}] at (-0.5, 4.3) {};
+
+	% G and J merge into L
+	\draw (4,2.4) -- (4,5.4) -- (2,5.4) -- (2,3.1);
+	\node [circle, fill, scale=0.2,label=above left:{\small L}] at (3,5.4) {L};
+	\draw (2.6, 5.2) -- (2.9, 5.2) -- (2.9, 5.1) -- (2.6, 5.1) -- (2.6, 5.2);
+	\draw (2.9, 5.15) -- (3.4, 5.15);
+
+	% K splits into M and N
+	\draw (-0.5, 4.5) -- (-0.5, 6) -- (-1.5,6) -- (0.5,6);
+	\node [circle, fill, scale=0.2,label=above left:{\small M}] at (-1.5,6) {M};
+	\draw (-1.9, 5.75) -- (-1.3, 5.75);
+	\draw (-1.3, 5.8) -- (-1.1, 5.8) -- (-1.1, 5.7) -- (-1.3, 5.7) -- (-1.3, 5.8);
+	\node [circle, fill, scale=0.2,label=above left:{\small N}] at (0.5,6) {N};
+	\draw (0.1, 5.8) -- (0.4, 5.8) -- (0.4, 5.7) -- (0.1, 5.7) -- (0.1, 5.8);
+	\draw (0.4, 5.75) -- (0.9, 5.75);
+
+	% L and N merge into O
+	\draw (0.5, 6) -- (0.5, 7.2) -- (3,7.2) -- (3,5.4);
+	\node [circle, fill, scale=0.2,label=above left:{\small O}] at (1.75,7.2) {O};
+	\draw (1.35, 6.95) -- (2.15, 6.95);
+
+	% M and H continue to the end
+	\draw (-1.5, 6) -- (-1.5, 8);
+	\draw (6, 2.4) -- (6, 8);
+
+	% Edge annotations above each node
+	\node [label=left:{\small$\{[0,L)\};1$}] at (0.2,0.6) {}; % A
+	\node [label=left:{\small$\{[0,L)\};1$}] at (4.2,0.4) {}; % B
+	\node [label=right:{\small$\{[0,L)\};1$}] at (5.8,0.4) {}; % C
+	\node [label=left:{\small$\{[0,L)\};2$}] at (5.2,1.6) {}; % D
+	\node [label=left:{\small$\{[0,v)\};1$}] at (-0.8,2.3) {}; % E
+	\node [label=right:{\small$\{[v,L)\};1$}] at (0.8,2.1) {}; % F
+	\node [label=right:{\small$\{[0,w)\};2$}] at (3.8,3.9) {}; % G
+	\node [label=left:{\small$\{[w,L)\};2$}] at (6.2,5) {}; % H
+	\node [label=right:{\small$\{[x,L)\};1$}] at (-0.2,3.8) {}; % I
+	\node [label=right:{\small$\{[v,x)\};1$}] at (1.8,4.1) {}; % J
+	\node [label=left:{\small$\{[0,v), [x, L)\};1$}] at (-0.3,5.2) {}; % K
+	\node [label=right:{\small$\{[0,v)\};2$}] at (2.8,6.4) {}; % L
+	\node [label=left:{\small$\{[x,L)\};1$}] at (-1.3,7) {}; % M
+	\node [label=right:{\small$\{[0,v)\};1$}] at (0.3,6.4) {}; % N
+
+	% Dashed lines for start and end times
+	\draw[dashed] (-2, 0) -- (11.1, 0);
+	\node [label=left:{$t = 0$}] at (-2,0) {};
+	\draw[dashed] (-2, 8) -- (11.1, 8);
+	\node [label=left:{$t = T$}] at (-2,8) {};
+
+	% Numbers of extant ancestors and links, from top to bottom
+	\node[label=right:{Extant material}] at (7.5, 8.2) {};
+	\node[label=right:{2 \; $2 L - x - w$}] at (7.5, 7.6) {};
+	\node[label=right:{4 \; $2 L + 2 v - x - w$}] at (7.5, 6.6) {};
+	\node[label=right:{3 \; $2 L + v - w$}] at (7.5, 5.7) {};
+	\node[label=right:{4 \; $2 L + x - v$}] at (7.5, 4.95) {};
+	\node[label=right:{5 \; $2 L$}] at (7.5, 3.8) {};
+	\node[label=right:{4 \; $2 L$}] at (7.5, 2.75) {};
+	\node[label=right:{3 \; $2 L$}] at (7.5, 1.95) {};
+	\node[label=right:{2 \; $2 L$}] at (7.5, 1.35) {};
+	\node[label=right:{3 \; $3 L$}] at (7.5, 0.6) {};
+
+	% Gray dashed lines to visually separate holding times
+	\draw[color=gray, dashed] (7.5, 1.2) -- (11.1, 1.2); % D
+	\draw[color=gray, dashed] (7.5, 1.5) -- (11.1, 1.5); % E and F
+	\draw[color=gray, dashed] (7.5, 2.4) -- (11.1, 2.4); % G and H
+	\draw[color=gray, dashed] (7.5, 3.1) -- (11.1, 3.1); % I and J
+	\draw[color=gray, dashed] (7.5, 4.5) -- (11.1, 4.5); % K
+	\draw[color=gray, dashed] (7.5, 5.4) -- (11.1, 5.4); % L
+	\draw[color=gray, dashed] (7.5, 6) -- (11.1, 6); % M and N
+	\draw[color=gray, dashed] (7.5, 7.2) -- (11.1, 7.2); % O
+\end{tikzpicture}
+&
+\begin{tikzpicture}
+	% Samples at x = 0, 4, and 6
+	\node [circle, fill, scale=0.2,label=above right:{\small A}] at (0,0) {A};
+	\draw (-0.4, -0.2) -- (0.4, -0.2) -- (0.4, -0.3) -- (-0.4, -0.3) -- (-0.4, -0.2);
+	\draw (-0.1, -0.1) -- (-0.1, -0.4);
+	\node [label=below:{\small $v$}] at (-0.1, -0.2) {};
+	\node [circle, fill, scale=0.2,label=above right:{\small B}] at (4,0) {B};
+	\draw (3.6, -0.2) -- (4.4, -0.2) -- (4.4, -0.3) -- (3.6, -0.3) -- (3.6, -0.2);
+	\node [circle, fill, scale=0.2,label=above left:{\small C}] at (6,0) {C};
+	\draw (5.6, -0.2) -- (6.4, -0.2) -- (6.4, -0.3) -- (5.6, -0.3) -- (5.6, -0.2);
+
+	% B and C merge into D
+	\draw (4,0) -- (4, 1.2) -- (6, 1.2) -- (6,0);
+	\node [circle, fill, scale=0.2,label=above right:{\small D}] at (5,1.2) {D};
+	\draw (4.6, 1) -- (5.4, 1) -- (5.4, 0.9) -- (4.6, 0.9) -- (4.6, 1);
+	\draw (5.1, 1.1) -- (5.1, 0.8);
+	\node [label=below:{\small $w$}] at (5.1, 1) {};
+
+	% A splits into E and F
+	\draw (0, 0) -- (0, 1.5) -- (-1,1.5) -- (1,1.5);
+	\node [circle, fill, scale=0.2,label=above left:{\small E}] at (-1,1.5) {E};
+	\draw (-1.4, 1.3) -- (-1.1, 1.3) -- (-1.1, 1.2) -- (-1.4, 1.2) -- (-1.4, 1.3);
+	\draw (-1.1, 1.25) -- (-0.6, 1.25);
+	\node [circle, fill, scale=0.2,label=above left:{\small F}] at (1,1.5) {F};
+	\draw (0.6, 1.25) -- (0.9, 1.25);
+	\draw (0.9, 1.3) -- (1.4, 1.3) -- (1.4, 1.2) -- (0.9, 1.2) -- (0.9, 1.3);
+	\draw (1.2, 1.4) -- (1.2, 1.1);
+	\node [label=below:{\small $x$}] at (1.2, 1.3) {};
+
+	% D splits into G and H
+	\draw (5, 1.2) -- (5, 2.4) -- (4,2.4) -- (6,2.4);
+	\node [circle, fill, scale=0.2,label=above left:{\small G}] at (4,2.4) {G};
+	\draw (3.6, 2.2) -- (4.1, 2.2) -- (4.1, 2.1) -- (3.6, 2.1) -- (3.6, 2.2);
+	\draw (4.1, 2.15) -- (4.4, 2.15);
+	\node [circle, fill, scale=0.2,label=above left:{\small H}] at (6,2.4) {H};
+	\draw (5.6, 2.15) -- (6.1, 2.15);
+	\draw (6.1, 2.2) -- (6.4, 2.2) -- (6.4, 2.1) -- (6.1, 2.1) -- (6.1, 2.2);
+	\draw [color=red](5.8, 2.3) -- (5.8, 2.0);
+	\node [label={[red]below:{\small $z$}}] at (5.8, 2.1) {};
+
+	% F splits into I and J
+	\draw (1, 1.5) -- (1, 3.1) -- (0,3.1) -- (2,3.1);
+	\node [circle, fill, scale=0.2,label=above left:{\small I}] at (0,3.1) {I};
+	\draw (-0.4, 2.85) -- (0.2, 2.85);
+	\draw (0.2, 2.9) -- (0.4, 2.9) -- (0.4, 2.8) -- (0.2, 2.8) -- (0.2, 2.9);
+	\node [circle, fill, scale=0.2,label=above left:{\small J}] at (2,3.1) {J};
+	\draw (1.6, 2.85) -- (1.9, 2.85);
+	\draw (1.9, 2.9) -- (2.2, 2.9) -- (2.2, 2.8) -- (1.9, 2.8) -- (1.9, 2.9);
+	\draw (2.2, 2.85) -- (2.4, 2.85);
+
+	% E and I merge into K
+	\draw (-1, 1.5) -- (-1, 4.5) -- (0,4.5) -- (0,3.1);
+	\node [circle, fill, scale=0.2,label=above left:{\small K}] at (-0.5,4.5) {K};
+	\draw (-0.9, 4.3) -- (-0.6, 4.3) -- (-0.6, 4.2) -- (-0.9, 4.2) -- (-0.9, 4.3);
+	\draw (-0.6, 4.25) -- (-0.3, 4.25);
+	\draw (-0.3, 4.3) -- (-0.1, 4.3) -- (-0.1, 4.2) -- (-0.3, 4.2) -- (-0.3, 4.3);
+	\draw (-0.5, 4.4) -- (-0.5, 4.1);
+	\node [label=below:{\small $y$}] at (-0.5, 4.3) {};
+
+	% H splits into P and Q. Red Q node drawn further down so that it overlays the black line underneath
+	\draw (6, 2.4) -- (6, 3.8) -- (7, 3.8);
+	\draw [color=red](6, 3.8) -- (5, 3.8);
+	\node [color=red, circle, fill, scale=0.2,label={[red]above left:{\small P}}] at (5,3.8) {P};
+	\draw (4.6, 3.55) -- (5.4, 3.55);
+	\draw (6.6, 3.55) -- (7.1, 3.55);
+	\draw (7.1, 3.6) -- (7.4, 3.6) -- (7.4, 3.5) -- (7.1, 3.5) -- (7.1, 3.6);
+
+	% G and J merge into L
+	\draw (4,2.4) -- (4,5.4) -- (2,5.4) -- (2,3.1);
+	\node [circle, fill, scale=0.2,label=above left:{\small L}] at (3,5.4) {L};
+	\draw (2.6, 5.2) -- (2.9, 5.2) -- (2.9, 5.1) -- (2.6, 5.1) -- (2.6, 5.2);
+	\draw (2.9, 5.15) -- (3.4, 5.15);
+
+	% K splits into M and N
+	\draw (-0.5, 4.5) -- (-0.5, 6) -- (-1.5,6) -- (0.5,6);
+	\node [circle, fill, scale=0.2,label=above left:{\small M}] at (-1.5,6) {M};
+	\draw (-1.9, 5.75) -- (-1.3, 5.75);
+	\draw (-1.3, 5.8) -- (-1.1, 5.8) -- (-1.1, 5.7) -- (-1.3, 5.7) -- (-1.3, 5.8);
+	\node [circle, fill, scale=0.2,label=above left:{\small N}] at (0.5,6) {N};
+	\draw (0.1, 5.8) -- (0.4, 5.8) -- (0.4, 5.7) -- (0.1, 5.7) -- (0.1, 5.8);
+	\draw (0.4, 5.75) -- (0.9, 5.75);
+
+	% P and L merge into R. R node drawn further down so that it overlays the black lines underneath
+	\draw [color=red](5, 3.8) -- (5, 6.6) -- (4, 6.6);
+	\draw (4,6.6) -- (3,6.6) -- (3, 5.4);
+	\draw (3.6, 6.4) -- (3.9, 6.4) -- (3.9, 6.3) -- (3.6, 6.3) -- (3.6, 6.4);
+	\draw (3.9, 6.35) -- (4.4, 6.35);
+
+	% R and N merge into O. O node drawn further down so that it overlays the red line underneath
+	\draw (0.5, 6) -- (0.5, 7.2) -- (4,7.2) -- (4,6.6);
+	\draw (1.85, 6.95) -- (2.65, 6.95);
+	% R node can now be added
+	\node [color=red, circle, fill, scale=0.2,label={[red]above left:{\small R}}] at (4,6.6) {R};
+
+	% M, O, and Q continue to the end
+	\draw (-1.5, 6) -- (-1.5, 8);
+	\draw [color=red](2.25, 7.2) -- (2.25, 8);
+	\draw (7, 3.8) -- (7, 8);
+	% O and Q nodes can now be added
+	\node [circle, fill, scale=0.2,label=above left:{\small O}] at (2.25,7.2) {O};
+	\node [color=red, circle, fill, scale=0.2,label={[red]above left:{\small Q}}] at (7,3.8) {Q};
+
+	% Dashed lines for start and end times
+	\draw[dashed] (-2, 0) -- (11, 0);
+	\node [label=left:{$t = 0$}] at (-2,0) {};
+	\draw[dashed] (-2, 8) -- (11, 8);
+	\node [label=left:{$t = T$}] at (-2,8) {};
+
+	% Numbers of extant ancestors and links, from top to bottom
+	\node[label=right:{Extant material}] at (7.5, 8.2) {};
+	\node[label=right:{3 \; $3 L$}] at (7.5, 7.6) {};
+	\node[label=right:{4 \; $4 L$}] at (7.5, 6.9) {};
+	\node[label=right:{5 \; $5 L$}] at (7.5, 6.3) {};
+	\node[label=right:{4 \; $4 L$}] at (7.5, 5.7) {};
+	\node[label=right:{5 \; $5 L$}] at (7.5, 4.95) {};
+	\node[label=right:{6 \; $6 L$}] at (7.5, 4.15) {};
+	\node[label=right:{5 \; $5 L$}] at (7.5, 3.45) {};
+	\node[label=right:{4 \; $4 L$}] at (7.5, 2.75) {};
+	\node[label=right:{3 \; $3 L$}] at (7.5, 1.95) {};
+	\node[label=right:{2 \; $2 L$}] at (7.5, 1.35) {};
+	\node[label=right:{3 \; $3 L$}] at (7.5, 0.6) {};
+
+	% Gray dashed lines to visually separate holding times
+	\draw[color=gray, dashed] (7.5, 1.2) -- (11, 1.2); % D
+	\draw[color=gray, dashed] (7.5, 1.5) -- (11, 1.5); % E and F
+	\draw[color=gray, dashed] (7.5, 2.4) -- (11, 2.4); % G and H
+	\draw[color=gray, dashed] (7.5, 3.1) -- (11, 3.1); % I and J
+	\draw[color=gray, dashed] (7.5, 3.8) -- (11, 3.8);	% P and Q
+	\draw[color=gray, dashed] (7.5, 4.5) -- (11, 4.5); % K
+	\draw[color=gray, dashed] (7.5, 5.4) -- (11, 5.4); % L
+	\draw[color=gray, dashed] (7.5, 6) -- (11, 6); % M and N
+	\draw[color=gray, dashed] (7.5, 6.6) -- (11, 6.6);	% R
+	\draw[color=gray, dashed] (7.5, 7.2) -- (11, 7.2); % O
+\end{tikzpicture}
+\end{tabular}
 }
+\caption{(Top) A realisation of the graph traversed by Hudson's algorithm started from a sample of three continuous chromosomes of length $L$ at time $t = 0$, and propagated until time $T$.
+The MRCA on the genetic interval $[x, z)$ is reached at node L, while that on $[0, x)$ is reached at node O, which is a root as it carries only fully coalesced ancestral segments.
+The non-ancestral segment $[x, y)$ above node K contributes to the rate of effective recombinations because it is trapped between ancestral segments.
+The two columns titled ``Extant material" are the number of extant lineages and the total length of genetic material on which an effective recombination can take place during each time interval.
+(Bottom) A corresponding realisation of a big ARG, which augments Hudson's algorithm by tracking nonancestral lineages.
+The result is a simpler state space and dynamics, at the cost of extra nodes and edges, highlighted in red, which do not affect the local tree at any site.}
+\label{hudson_vs_bigARG}
 \end{figure}
@@
 flows along edges.
-
-\begin{figure}[ht]
-\centering
-\begin{tikzpicture}
-	% Samples at x = 0, 4, and 6
-	\node [circle, fill, scale=0.2,label=above right:{\small A}] at (0,0) {A};
-	\draw (-0.4, -0.2) -- (0.4, -0.2) -- (0.4, -0.3) -- (-0.4, -0.3) -- (-0.4, -0.2);
-	\draw (-0.1, -0.1) -- (-0.1, -0.4);
-	\node [label=below:{\small $v$}] at (-0.1, -0.2) {};
-	\node [circle, fill, scale=0.2,label=above right:{\small B}] at (4,0) {B};
-	\draw (3.6, -0.2) -- (4.4, -0.2) -- (4.4, -0.3) -- (3.6, -0.3) -- (3.6, -0.2);
-	\node [circle, fill, scale=0.2,label=above left:{\small C}] at (6,0) {C};
-	\draw (5.6, -0.2) -- (6.4, -0.2) -- (6.4, -0.3) -- (5.6, -0.3) -- (5.6, -0.2);
-
-	% B and C merge into D
-	\draw (4,0) -- (4, 1.2) -- (6, 1.2) -- (6,0);
-	\node [circle, fill, scale=0.2,label=above right:{\small D}] at (5,1.2) {D};
-	\draw (4.6, 1) -- (5.4, 1) -- (5.4, 0.9) -- (4.6, 0.9) -- (4.6, 1);
-	\draw (5.1, 1.1) -- (5.1, 0.8);
-	\node [label=below:{\small $w$}] at (5.1, 1) {};
-
-	% A splits into E and F
-	\draw (0, 0) -- (0, 1.5) -- (-1,1.5) -- (1,1.5);
-	\node [circle, fill, scale=0.2,label=above left:{\small E}] at (-1,1.5) {E};
-	\draw (-1.4, 1.3) -- (-1.1, 1.3) -- (-1.1, 1.2) -- (-1.4, 1.2) -- (-1.4, 1.3);
-	\draw (-1.1, 1.25) -- (-0.6, 1.25);
-	\node [circle, fill, scale=0.2,label=above left:{\small F}] at (1,1.5) {F};
-	\draw (0.6, 1.25) -- (0.9, 1.25);
-	\draw (0.9, 1.3) -- (1.4, 1.3) -- (1.4, 1.2) -- (0.9, 1.2) -- (0.9, 1.3);
-	\draw (1.2, 1.4) -- (1.2, 1.1);
-	\node [label=below:{\small $x$}] at (1.2, 1.3) {};
-
-	% D splits into G and H
-	\draw (5, 1.2) -- (5, 2.4) -- (4,2.4) -- (6,2.4);
-	\node [circle, fill, scale=0.2,label=above left:{\small G}] at (4,2.4) {G};
-	\draw (3.6, 2.2) -- (4.1, 2.2) -- (4.1, 2.1) -- (3.6, 2.1) -- (3.6, 2.2);
-	\draw (4.1, 2.15) -- (4.4, 2.15);
-	\node [circle, fill, scale=0.2,label=above left:{\small H}] at (6,2.4) {H};
-	\draw (5.6, 2.15) -- (6.1, 2.15);
-	\draw (6.1, 2.2) -- (6.4, 2.2) -- (6.4, 2.1) -- (6.1, 2.1) -- (6.1, 2.2);
-
-	% F splits into I and J
-	\draw (1, 1.5) -- (1, 3.1) -- (0,3.1) -- (2,3.1);
-	\node [circle, fill, scale=0.2,label=above left:{\small I}] at (0,3.1) {I};
-	\draw (-0.4, 2.85) -- (0.2, 2.85);
-	\draw (0.2, 2.9) -- (0.4, 2.9) -- (0.4, 2.8) -- (0.2, 2.8) -- (0.2, 2.9);
-	\node [circle, fill, scale=0.2,label=above left:{\small J}] at (2,3.1) {J};
-	\draw (1.6, 2.85) -- (1.9, 2.85);
-	\draw (1.9, 2.9) -- (2.2, 2.9) -- (2.2, 2.8) -- (1.9, 2.8) -- (1.9, 2.9);
-	\draw (2.2, 2.85) -- (2.4, 2.85);
-
-	% E and I merge into K
-	\draw (-1, 1.5) -- (-1, 4.5) -- (0,4.5) -- (0,3.1);
-	\node [circle, fill, scale=0.2,label=above left:{\small K}] at (-0.5,4.5) {K};
-	\draw (-0.9, 4.3) -- (-0.6, 4.3) -- (-0.6, 4.2) -- (-0.9, 4.2) -- (-0.9, 4.3);
-	\draw (-0.6, 4.25) -- (-0.3, 4.25);
-	\draw (-0.3, 4.3) -- (-0.1, 4.3) -- (-0.1, 4.2) -- (-0.3, 4.2) -- (-0.3, 4.3);
-	\draw (-0.5, 4.4) -- (-0.5, 4.1);
-	\node [label=below:{\small $y$}] at (-0.5, 4.3) {};
-
-	% G and J merge into L
-	\draw (4,2.4) -- (4,5.4) -- (2,5.4) -- (2,3.1);
-	\node [circle, fill, scale=0.2,label=above left:{\small L}] at (3,5.4) {L};
-	\draw (2.6, 5.2) -- (2.9, 5.2) -- (2.9, 5.1) -- (2.6, 5.1) -- (2.6, 5.2);
-	\draw (2.9, 5.15) -- (3.4, 5.15);
-
-	% K splits into M and N
-	\draw (-0.5, 4.5) -- (-0.5, 6) -- (-1.5,6) -- (0.5,6);
-	\node [circle, fill, scale=0.2,label=above left:{\small M}] at (-1.5,6) {M};
-	\draw (-1.9, 5.75) -- (-1.3, 5.75);
-	\draw (-1.3, 5.8) -- (-1.1, 5.8) -- (-1.1, 5.7) -- (-1.3, 5.7) -- (-1.3, 5.8);
-	\node [circle, fill, scale=0.2,label=above left:{\small N}] at (0.5,6) {N};
-	\draw (0.1, 5.8) -- (0.4, 5.8) -- (0.4, 5.7) -- (0.1, 5.7) -- (0.1, 5.8);
-	\draw (0.4, 5.75) -- (0.9, 5.75);
-
-	% L and N merge into O
-	\draw (0.5, 6) -- (0.5, 7.2) -- (3,7.2) -- (3,5.4);
-	\node [circle, fill, scale=0.2,label=above left:{\small O}] at (1.75,7.2) {O};
-	\draw (1.35, 6.95) -- (2.15, 6.95);
-
-	% M and H continue to the end
-	\draw (-1.5, 6) -- (-1.5, 8);
-	\draw (6, 2.4) -- (6, 8);
-
-	% Edge annotations above each node
-	\node [label=left:{\small$\{[0,L)\};1$}] at (0.2,0.6) {}; % A
-	\node [label=left:{\small$\{[0,L)\};1$}] at (4.2,0.4) {}; % B
-	\node [label=right:{\small$\{[0,L)\};1$}] at (5.8,0.4) {}; % C
-	\node [label=left:{\small$\{[0,L)\};2$}] at (5.2,1.6) {}; % D
-	\node [label=left:{\small$\{[0,v)\};1$}] at (-0.8,2.3) {}; % E
-	\node [label=right:{\small$\{[v,L)\};1$}] at (0.8,2.1) {}; % F
-	\node [label=right:{\small$\{[0,w)\};2$}] at (3.8,3.9) {}; % G
-	\node [label=left:{\small$\{[w,L)\};2$}] at (6.2,5) {}; % H
-	\node [label=right:{\small$\{[x,L)\};1$}] at (-0.2,3.8) {}; % I
-	\node [label=right:{\small$\{[v,x)\};1$}] at (1.8,4.1) {}; % J
-	\node [label=left:{\small$\{[0,v), [x, L)\};1$}] at (-0.3,5.2) {}; % K
-	\node [label=right:{\small$\{[0,v)\};2$}] at (2.8,6.4) {}; % L
-	\node [label=left:{\small$\{[x,L)\};1$}] at (-1.3,7) {}; % M
-	\node [label=right:{\small$\{[0,v)\};1$}] at (0.3,6.4) {}; % N
-
-	% Dashed lines for start and end times
-	\draw[dashed] (-2, 0) -- (11.1, 0);
-	\node [label=left:{$t = 0$}] at (-2,0) {};
-	\draw[dashed] (-2, 8) -- (11.1, 8);
-	\node [label=left:{$t = T$}] at (-2,8) {};
-
-	% Numbers of extant ancestors and links, from top to bottom
-	\node[label=right:{Extant material}] at (7.5, 8.2) {};
-	\node[label=right:{2 \; $2 L - x - w$}] at (7.5, 7.6) {};
-	\node[label=right:{4 \; $2 L + 2 v - x - w$}] at (7.5, 6.6) {};
-	\node[label=right:{3 \; $2 L + v - w$}] at (7.5, 5.7) {};
-	\node[label=right:{4 \; $2 L + x - v$}] at (7.5, 4.95) {};
-	\node[label=right:{5 \; $2 L$}] at (7.5, 3.8) {};
-	\node[label=right:{4 \; $2 L$}] at (7.5, 2.75) {};
-	\node[label=right:{3 \; $2 L$}] at (7.5, 1.95) {};
-	\node[label=right:{2 \; $2 L$}] at (7.5, 1.35) {};
-	\node[label=right:{3 \; $3 L$}] at (7.5, 0.6) {};
-
-	% Gray dashed lines to visually separate holding times
-	\draw[color=gray, dashed] (7.5, 1.2) -- (11.1, 1.2); % D
-	\draw[color=gray, dashed] (7.5, 1.5) -- (11.1, 1.5); % E and F
-	\draw[color=gray, dashed] (7.5, 2.4) -- (11.1, 2.4); % G and H
-	\draw[color=gray, dashed] (7.5, 3.1) -- (11.1, 3.1); % I and J
-	\draw[color=gray, dashed] (7.5, 4.5) -- (11.1, 4.5); % K
-	\draw[color=gray, dashed] (7.5, 5.4) -- (11.1, 5.4); % L
-	\draw[color=gray, dashed] (7.5, 6) -- (11.1, 6); % M and N
-	\draw[color=gray, dashed] (7.5, 7.2) -- (11.1, 7.2); % O
-\end{tikzpicture}
-\begin{tikzpicture}
-	% Samples at x = 0, 4, and 6
-	\node [circle, fill, scale=0.2,label=above right:{\small A}] at (0,0) {A};
-	\draw (-0.4, -0.2) -- (0.4, -0.2) -- (0.4, -0.3) -- (-0.4, -0.3) -- (-0.4, -0.2);
-	\draw (-0.1, -0.1) -- (-0.1, -0.4);
-	\node [label=below:{\small $v$}] at (-0.1, -0.2) {};
-	\node [circle, fill, scale=0.2,label=above right:{\small B}] at (4,0) {B};
-	\draw (3.6, -0.2) -- (4.4, -0.2) -- (4.4, -0.3) -- (3.6, -0.3) -- (3.6, -0.2);
-	\node [circle, fill, scale=0.2,label=above left:{\small C}] at (6,0) {C};
-	\draw (5.6, -0.2) -- (6.4, -0.2) -- (6.4, -0.3) -- (5.6, -0.3) -- (5.6, -0.2);
-
-	% B and C merge into D
-	\draw (4,0) -- (4, 1.2) -- (6, 1.2) -- (6,0);
-	\node [circle, fill, scale=0.2,label=above right:{\small D}] at (5,1.2) {D};
-	\draw (4.6, 1) -- (5.4, 1) -- (5.4, 0.9) -- (4.6, 0.9) -- (4.6, 1);
-	\draw (5.1, 1.1) -- (5.1, 0.8);
-	\node [label=below:{\small $w$}] at (5.1, 1) {};
-
-	% A splits into E and F
-	\draw (0, 0) -- (0, 1.5) -- (-1,1.5) -- (1,1.5);
-	\node [circle, fill, scale=0.2,label=above left:{\small E}] at (-1,1.5) {E};
-	\draw (-1.4, 1.3) -- (-1.1, 1.3) -- (-1.1, 1.2) -- (-1.4, 1.2) -- (-1.4, 1.3);
-	\draw (-1.1, 1.25) -- (-0.6, 1.25);
-	\node [circle, fill, scale=0.2,label=above left:{\small F}] at (1,1.5) {F};
-	\draw (0.6, 1.25) -- (0.9, 1.25);
-	\draw (0.9, 1.3) -- (1.4, 1.3) -- (1.4, 1.2) -- (0.9, 1.2) -- (0.9, 1.3);
-	\draw (1.2, 1.4) -- (1.2, 1.1);
-	\node [label=below:{\small $x$}] at (1.2, 1.3) {};
-
-	% D splits into G and H
-	\draw (5, 1.2) -- (5, 2.4) -- (4,2.4) -- (6,2.4);
-	\node [circle, fill, scale=0.2,label=above left:{\small G}] at (4,2.4) {G};
-	\draw (3.6, 2.2) -- (4.1, 2.2) -- (4.1, 2.1) -- (3.6, 2.1) -- (3.6, 2.2);
-	\draw (4.1, 2.15) -- (4.4, 2.15);
-	\node [circle, fill, scale=0.2,label=above left:{\small H}] at (6,2.4) {H};
-	\draw (5.6, 2.15) -- (6.1, 2.15);
-	\draw (6.1, 2.2) -- (6.4, 2.2) -- (6.4, 2.1) -- (6.1, 2.1) -- (6.1, 2.2);
-	\draw [color=red](5.8, 2.3) -- (5.8, 2.0);
-	\node [label={[red]below:{\small $z$}}] at (5.8, 2.1) {};
-
-	% F splits into I and J
-	\draw (1, 1.5) -- (1, 3.1) -- (0,3.1) -- (2,3.1);
-	\node [circle, fill, scale=0.2,label=above left:{\small I}] at (0,3.1) {I};
-	\draw (-0.4, 2.85) -- (0.2, 2.85);
-	\draw (0.2, 2.9) -- (0.4, 2.9) -- (0.4, 2.8) -- (0.2, 2.8) -- (0.2, 2.9);
-	\node [circle, fill, scale=0.2,label=above left:{\small J}] at (2,3.1) {J};
-	\draw (1.6, 2.85) -- (1.9, 2.85);
-	\draw (1.9, 2.9) -- (2.2, 2.9) -- (2.2, 2.8) -- (1.9, 2.8) -- (1.9, 2.9);
-	\draw (2.2, 2.85) -- (2.4, 2.85);
-
-	% E and I merge into K
-	\draw (-1, 1.5) -- (-1, 4.5) -- (0,4.5) -- (0,3.1);
-	\node [circle, fill, scale=0.2,label=above left:{\small K}] at (-0.5,4.5) {K};
-	\draw (-0.9, 4.3) -- (-0.6, 4.3) -- (-0.6, 4.2) -- (-0.9, 4.2) -- (-0.9, 4.3);
-	\draw (-0.6, 4.25) -- (-0.3, 4.25);
-	\draw (-0.3, 4.3) -- (-0.1, 4.3) -- (-0.1, 4.2) -- (-0.3, 4.2) -- (-0.3, 4.3);
-	\draw (-0.5, 4.4) -- (-0.5, 4.1);
-	\node [label=below:{\small $y$}] at (-0.5, 4.3) {};
-
-	% H splits into P and Q. Red Q node drawn further down so that it overlays the black line underneath
-	\draw (6, 2.4) -- (6, 3.8) -- (7, 3.8);
-	\draw [color=red](6, 3.8) -- (5, 3.8);
-	\node [color=red, circle, fill, scale=0.2,label={[red]above left:{\small P}}] at (5,3.8) {P};
-	\draw (4.6, 3.55) -- (5.4, 3.55);
-	\draw (6.6, 3.55) -- (7.1, 3.55);
-	\draw (7.1, 3.6) -- (7.4, 3.6) -- (7.4, 3.5) -- (7.1, 3.5) -- (7.1, 3.6);
-
-	% G and J merge into L
-	\draw (4,2.4) -- (4,5.4) -- (2,5.4) -- (2,3.1);
-	\node [circle, fill, scale=0.2,label=above left:{\small L}] at (3,5.4) {L};
-	\draw (2.6, 5.2) -- (2.9, 5.2) -- (2.9, 5.1) -- (2.6, 5.1) -- (2.6, 5.2);
-	\draw (2.9, 5.15) -- (3.4, 5.15);
-
-	% K splits into M and N
-	\draw (-0.5, 4.5) -- (-0.5, 6) -- (-1.5,6) -- (0.5,6);
-	\node [circle, fill, scale=0.2,label=above left:{\small M}] at (-1.5,6) {M};
-	\draw (-1.9, 5.75) -- (-1.3, 5.75);
-	\draw (-1.3, 5.8) -- (-1.1, 5.8) -- (-1.1, 5.7) -- (-1.3, 5.7) -- (-1.3, 5.8);
-	\node [circle, fill, scale=0.2,label=above left:{\small N}] at (0.5,6) {N};
-	\draw (0.1, 5.8) -- (0.4, 5.8) -- (0.4, 5.7) -- (0.1, 5.7) -- (0.1, 5.8);
-	\draw (0.4, 5.75) -- (0.9, 5.75);
-
-	% P and L merge into R. R node drawn further down so that it overlays the black lines underneath
-	\draw [color=red](5, 3.8) -- (5, 6.6) -- (4, 6.6);
-	\draw (4,6.6) -- (3,6.6) -- (3, 5.4);
-	\draw (3.6, 6.4) -- (3.9, 6.4) -- (3.9, 6.3) -- (3.6, 6.3) -- (3.6, 6.4);
-	\draw (3.9, 6.35) -- (4.4, 6.35);
-
-	% R and N merge into O. O node drawn further down so that it overlays the red line underneath
-	\draw (0.5, 6) -- (0.5, 7.2) -- (4,7.2) -- (4,6.6);
-	\draw (1.85, 6.95) -- (2.65, 6.95);
-	% R node can now be added
-	\node [color=red, circle, fill, scale=0.2,label={[red]above left:{\small R}}] at (4,6.6) {R};
-
-	% M, O, and Q continue to the end
-	\draw (-1.5, 6) -- (-1.5, 8);
-	\draw [color=red](2.25, 7.2) -- (2.25, 8);
-	\draw (7, 3.8) -- (7, 8);
-	% O and Q nodes can now be added
-	\node [circle, fill, scale=0.2,label=above left:{\small O}] at (2.25,7.2) {O};
-	\node [color=red, circle, fill, scale=0.2,label={[red]above left:{\small Q}}] at (7,3.8) {Q};
-
-	% Dashed lines for start and end times
-	\draw[dashed] (-2, 0) -- (11, 0);
-	\node [label=left:{$t = 0$}] at (-2,0) {};
-	\draw[dashed] (-2, 8) -- (11, 8);
-	\node [label=left:{$t = T$}] at (-2,8) {};
-
-	% Numbers of extant ancestors and links, from top to bottom
-	\node[label=right:{Extant material}] at (7.5, 8.2) {};
-	\node[label=right:{3 \; $3 L$}] at (7.5, 7.6) {};
-	\node[label=right:{4 \; $4 L$}] at (7.5, 6.9) {};
-	\node[label=right:{5 \; $5 L$}] at (7.5, 6.3) {};
-	\node[label=right:{4 \; $4 L$}] at (7.5, 5.7) {};
-	\node[label=right:{5 \; $5 L$}] at (7.5, 4.95) {};
-	\node[label=right:{6 \; $6 L$}] at (7.5, 4.15) {};
-	\node[label=right:{5 \; $5 L$}] at (7.5, 3.45) {};
-	\node[label=right:{4 \; $4 L$}] at (7.5, 2.75) {};
-	\node[label=right:{3 \; $3 L$}] at (7.5, 1.95) {};
-	\node[label=right:{2 \; $2 L$}] at (7.5, 1.35) {};
-	\node[label=right:{3 \; $3 L$}] at (7.5, 0.6) {};
-
-	% Gray dashed lines to visually separate holding times
-	\draw[color=gray, dashed] (7.5, 1.2) -- (11, 1.2); % D
-	\draw[color=gray, dashed] (7.5, 1.5) -- (11, 1.5); % E and F
-	\draw[color=gray, dashed] (7.5, 2.4) -- (11, 2.4); % G and H
-	\draw[color=gray, dashed] (7.5, 3.1) -- (11, 3.1); % I and J
-	\draw[color=gray, dashed] (7.5, 3.8) -- (11, 3.8);	% P and Q
-	\draw[color=gray, dashed] (7.5, 4.5) -- (11, 4.5); % K
-	\draw[color=gray, dashed] (7.5, 5.4) -- (11, 5.4); % L
-	\draw[color=gray, dashed] (7.5, 6) -- (11, 6); % M and N
-	\draw[color=gray, dashed] (7.5, 6.6) -- (11, 6.6);	% R
-	\draw[color=gray, dashed] (7.5, 7.2) -- (11, 7.2); % O
-\end{tikzpicture}
-\caption{(Top) A realisation of the graph traversed by Hudson's algorithm started from a sample of three continuous chromosomes of length $L$ at time $t = 0$, and propagated until time $T$.
-The MRCA on the genetic interval $[x, z)$ is reached at node L, while that on $[0, x)$ is reached at node O, which is a root as it carries only fully coalesced ancestral segments.
-The non-ancestral segment $[x, y)$ above node K contributes to the rate of effective recombinations because it is trapped between ancestral segments.
-The two columns titled ``Extant material" are the number of extant lineages and the total length of genetic material on which an effective recombination can take place during each time interval.
-(Bottom) A corresponding realisation of a big ARG, which augments Hudson's algorithm by tracking nonancestral lineages.
-The result is a simpler state space and dynamics, at the cost of extra nodes and edges, highlighted in red, which do not affect the local tree at any site.}
-\label{hudson_vs_bigARG}
-\end{figure}
-
 \begin{figure}

### 2022-02-10 argutils/ancestry.py
Add a ``recomb_proba`` arg to wright fisher sim.
Closes #87

@@ -320,9 +320,17 @@ def simplify_keeping_unary_in_coal(ts, map_nodes=False):
         return tables.tree_sequence()
 
 
-def sim_wright_fisher(n, N, L, seed=None):
+def sim_wright_fisher(n, N, L, recomb_proba=1, seed=None):
     """
     NOTE! This hasn't been statistically tested and is probably not correct.
+
+    The ``recomb_proba`` is the probability of recombination occuring
+    when a genome chooses its parents. With probability ``recomb_proba``
+    a breakpoint will be chosen and the ancestry split among two parents
+    (if present); otherwise, one parent genome will be chosen randomly.
+    (This almost certainly not a good way to set things up model-wise,
+    and just a quick hack to give us a lever to control the amount of
+    recombination happening.)
 
     We don't keep track of the pedigree because this is inconsistent
     with the practise of dropping "pass through" nodes in which
@@ -368,14 +376,15 @@ def sim_wright_fisher(n, N, L, seed=None):
                 # Randomise the order we add the lineages
                 rng.shuffle(parent.collected_lineages)
                 j = 0
-                breakpoint = rng.randrange(1, L)
-                # print("breakpoint", breakpoint)
-                if lineage.left < breakpoint < lineage.right:
-                    # print("effective recombination!", lineage.node)
-                    right_lineage = lineage.split(breakpoint)
-                    parent.collected_lineages[j].append(right_lineage)
-                    j += 1
-                    node_flags[lineage.node] |= NODE_IS_RECOMB
+                if rng.random() < recomb_proba:
+                    breakpoint = rng.randrange(1, L)
+                    # print("breakpoint", breakpoint)
+                    if lineage.left < breakpoint < lineage.right:
+                        # print("effective recombination!", lineage.node)
+                        right_lineage = lineage.split(breakpoint)
+                        parent.collected_lineages[j].append(right_lineage)
+                        j += 1
+                        node_flags[lineage.node] |= NODE_IS_RECOMB
                 parent.collected_lineages[j].append(lineage)
 
         # All the ancestral material has been distributed to the parental

### 2022-02-10 argutils/tests.py

@@ -22,6 +22,20 @@ class TestSimulate:
         assert ts.num_trees > 1
         assert all(tree.num_roots == 1 for tree in ts.trees())
 
+    def test_wright_fisher_no_recomb(self):
+        ts = argutils.sim_wright_fisher(4, N=10, L=5, recomb_proba=0, seed=1)
+        assert ts.num_samples == 8
+        assert ts.sequence_length == 5
+        assert ts.num_trees == 1
+        assert all(tree.num_roots == 1 for tree in ts.trees())
+
+    def test_wright_fisher_high_recomb(self):
+        ts = argutils.sim_wright_fisher(4, N=10, L=5, recomb_proba=0.9, seed=1)
+        assert ts.num_samples == 8
+        assert ts.sequence_length == 5
+        assert ts.num_trees > 1
+        assert all(tree.num_roots == 1 for tree in ts.trees())
+
     def test_basic_coalescent_unresolved(self):
         ts = argutils.sim_coalescent(4, L=5, rho=0.1, seed=1, resolved=False)
         # Very basic - real test is the resolution process.

### 2022-02-10 paper.tex
Minor tidy ups.

@@
 evolutionary tree is a necessary prerequisite for many analyses of species \citep{rannala2003genetics}, non (or
-rarely) recombining sections of DNA \citep{cann1987mitochondrial,underhill2001annalsofhumangenetics} and viruses \citep{grenfell2004science}. The tree captures what is known and
+rarely) recombining sections of DNA \citep{cann1987mitochondrial,underhill2001annalsofhumangenetics} and viruses \citep{grenfell2004science}.
+A genealogical tree captures what is known and
 knowable about the genetic ancestry of the sample in an elegant and concise
@@
 impossible, and therefore tree-based analysis methods have been largely
-unavailable to the fields of population and statistical genetics.
+unavailable to the fields of population and statistical
+genetics~\citep{hejase2020summary}.
 

### 2022-02-10 paper.tex
Small tweaks to text

--- 2022-02-09 paper.tex
@@
 impossible, and therefore tree-based analysis methods have been largely
-unavailable to the fields of population and statistical genetics.
+unavailable to the fields of population and statistical
+genetics.
 
@@
 have raised the genuine prospect of genealogical analysis becoming a standard part
-of the population and statistical genetics toolkit.
+of the population and statistical genetics toolkit~\citep{hejase2020summary}.
 Applications using these inferred ancestries as input have
@@
 }
-\caption{(Top) A realisation of the graph traversed by Hudson's algorithm started from a sample of three continuous chromosomes of length $L$ at time $t = 0$, and propagated until time $T$.
-The MRCA on the genetic interval $[x, z)$ is reached at node L, while that on $[0, x)$ is reached at node O, which is a root as it carries only fully coalesced ancestral segments.
-The non-ancestral segment $[x, y)$ above node K contributes to the rate of effective recombinations because it is trapped between ancestral segments.
-The two columns titled ``Extant material" are the number of extant lineages and the total length of genetic material on which an effective recombination can take place during each time interval.
-(Bottom) A corresponding realisation of a big ARG, which augments Hudson's algorithm by tracking nonancestral lineages.
-The result is a simpler state space and dynamics, at the cost of extra nodes and edges, highlighted in red, which do not affect the local tree at any site.}
+\caption{(A)
+A realisation of the graph traversed by Hudson's algorithm started from a
+sample of three continuous chromosomes of length $L$ at time $t = 0$, and
+propagated until time $T$. The MRCA on the genetic interval $[x, z)$ is reached
+at node L, while that on $[0, x)$ is reached at node O.
+The non-ancestral segment $[x, y)$ above
+node K contributes to the rate of effective recombinations because it
+is trapped between ancestral segments. The two columns titled ``Extant
+material" are the number of extant lineages and the total length of genetic
+material on which an effective recombination can take place during each time
+interval.
+(B) A corresponding realisation of a big ARG, which augments Hudson's algorithm
+by tracking nonancestral lineages. The result is a simpler state space and
+dynamics, at the cost of extra nodes and edges, highlighted in red, which do
+not affect the local tree at any site.}
 \label{hudson_vs_bigARG}
@@
 that is not important for our purposes here so we omit it.)
-If we have $n$ samples and genome of length $m$, the process begins with $n$ lineages
-of the form $[(0, m, 1)]$. The process then works backwards in time from
+If we have $n$ samples and a genome of length $m$, the process begins with $n$ lineages
+of the form $\{(0, m, 1)\}$. The process then works backwards in time from
 the present day as a series of random common ancestor or recombination events.
@@
 % about the process as choosing edges to merge, not nodes.
-The Ancestral Recombination Graph (ARG) was later introduced by Griffiths~\citep{griffiths1991two,ethier1990two,griffiths1996ancestral,
+The Ancestral Recombination Graph (ARG) was later introduced by
+Griffiths~\citep{griffiths1991two,ethier1990two,griffiths1996ancestral,
 griffiths1997ancestral}, and is a
@@
 This disparity in the number of events in the two formulations is
-because the vast majority of the events that occur in the ARG do
+because the majority of the events that occur in the ARG do
 not affect the genetic ancestry of the sample in any way. Recombination
-events occur outside of ancestral material do not have any bearing
+events that occur outside of ancestral material do not have any bearing
 on the ancestry of the sample, and so the structure is hugely redundant.

### 2022-02-11 paper.tex

@@
 \usepackage[right]{lineno}
-\usepackage{tikz}
-
 \usepackage{subcaption}
@@
 Discuss \citet{wiuf1999ancestry}. Wiuf and Hein's classical pair of papers
-clarified the relationship between Hudson's algorithm and the ARG]
+clarified the relationship between Hudson's algorithm and the ARG.
+Explain that ARG simulation is inherently less efficient than Hudson's
+algorithm because we must track all lineages, even those that correspond
+to sections that have coalesced, or we won't have a complete ARG.
+This might be tricky to communicated without a figure, but is an
+esoteric point perhaps not worth spending a figure on.]
 
@@
 \centering
-\begin{tikzpicture}[xscale=0.8, yscale=0.3, node distance=2mm and 20mm]
+\begin{tabular}{cc}
+\begin{tikzpicture}[x=5mm, y=5mm, node distance=2mm and 20mm]
 \tikzset{greynode/.style={circle,fill,inner sep=1},
@@
 
-\node (s0) [greynode] {};
-\node (s1) [right=of s0,greynode] {};
-\node (s2) [right=of s1,greynode] {};
-\node [greynode] (s3) at ($(s1) + (0,1)$) {};
-\node [greynode] (s4) at ($0.5*(s1) + 0.5*(s2) + (0,2)$) {};
-\node [greynode] (s5) at ($0.5*(s0) + 0.5*(s1) + (0,3)$) {};
-\node [greynode] (s6) at ($(s1) + (0,5)$) {};
+\node (s0) [greynode] at (0, 0) {};
+\node (s1) [greynode] at (3, 0) {};
+\node (s2) [greynode] at (6, 0) {};
+\node (s3) [greynode] at (3, 1) {};
+\node (s4) [greynode] at (1, 2) {};
+\node (s5) [greynode] at (5, 3) {};
+\node (s6) [greynode] at (3, 4) {};
 
-\node [nodelabel,anchor=north west] at ($(s3) + (0,0)$) {$x = 0.3$};
+\node [nodelabel,anchor=north west] at ($(s3) + (0,0)$) {$x$};
 \foreach \u/\lab in {s0/0, s1/1, s2/2} \node[nodelabel,anchor=north] at (\u) {\lab};
@@
 \draw (s1) -- (s3);
-\foreach \child/\parent in {s6/s5, s6/s4, s5/s0, s4/s2}
-	\draw (\child) -| (\parent);
+\draw (s0) |- (s4);
+\draw (s4) -- (2,2) |- (s3);
+\draw (s4) |- (s6);
+\draw (s3) -- (4,1) |- (s5);
+\draw (s2) |- (s5);
+\draw (s5) |- (s6);
 
-\draw (s5) -| ($0.5*(s3) + 0.5*(s3-|s5)$) |- (s3);
-\draw (s4) -| ($0.5*(s3) + 0.5*(s3-|s4)$) |- (s3);
+\node [nodelabel,anchor=north west] at ($(-6.5,5)$) {
+\begin{tabular}{c|c}
+\multicolumn{2}{c}{Breakpoint}\\
+\hline
+0 & $\infty$ \\
+1 & $\infty$ \\
+2 & $\infty$ \\
+3 & $x$ \\
+4 & $\infty$ \\
+5 & $\infty$ \\
+6 & $\infty$ \\
+\end{tabular}};
+
+\node [nodelabel,anchor=north west] at ($(-2.5,5)$) {
+\begin{tabular}{l}
+Edges\\
+\hline
+$(0, 4)$ \\
+$(1, 3)$ \\
+$(2, 5)$ \\
+$(3, 4)$ \\
+$(3, 5)$ \\
+$(4, 6)$ \\
+$(5, 6)$ \\
+\end{tabular}};
 
 \end{tikzpicture}
-\caption{\label{fig-arg} \textcolor{red}{A picture of a standard ARG for us
-to work with for now. We should replace it with something better at some
-point.}
+&
+\begin{tikzpicture}[x=5mm,y=5mm,node distance=2mm and 20mm]
+\tikzset{greynode/.style={circle,fill,inner sep=1},
+nodelabel/.style={font=\footnotesize}}
+
+\node (s0) [greynode] at (0, 0) {};
+\node (s1) [greynode] at (3, 0) {};
+\node (s2) [greynode] at (6, 0) {};
+\node (s3) [greynode] at (2, 1) {};
+\node (s4) [greynode] at (4, 1) {};
+\node (s5) [greynode] at (1, 2) {};
+\node (s6) [greynode] at (5, 3) {};
+\node (s7) [greynode] at (3, 4) {};
+
+\foreach \u/\lab in {s0/0, s1/1, s2/2} \node[nodelabel,anchor=north] at (\u) {\lab};
+\foreach \u/\lab in {s3/3} \node[nodelabel,anchor=east] at (\u) {\lab};
+\foreach \u/\lab in {s4/4} \node[nodelabel,anchor=west] at (\u) {\lab};
+\foreach \u/\lab in {s5/5} \node[nodelabel,anchor=south west] at (\u) {\lab};
+\foreach \u/\lab in {s6/6} \node[nodelabel,anchor=south east] at (\u) {\lab};
+\foreach \u/\lab in {s7/7} \node[nodelabel,anchor=south] at (\u) {\lab};
+
+%% Edges
+\draw (s1) |- (s3);
+\draw (s1) |- (s4) |- (s6);
+\draw (s3) |- (s5);
+\draw (s0) |- (s5);
+\draw (s2) |- (s6);
+\draw (s5) |- (s7);
+\draw (s6) |- (s7);
+
+\node [nodelabel,anchor=north west] at ($(-5,5)$) {
+\begin{tabular}{ll}
+Edges\\
+\hline
+$(0, 5)$ & $(0, L]$ \\
+$(1, 3)$ & $(0, x]$ \\
+$(1, 4)$ & $(x, L]$ \\
+$(2, 6)$ & $(0, L]$ \\
+$(3, 5)$ & $(0, L]$ \\
+$(4, 6)$ & $(0, L]$ \\
+$(5, 7)$ & $(0, L]$ \\
+$(6, 7)$ & $(0, L]$ \\
+\end{tabular}};
+
+\end{tikzpicture}
+\\
+(A) & (B) \\
+\end{tabular}
+\caption{\label{fig-arg-data-structure}
+Encoding an ARG. (A) The classical approach described by Griffiths
+equates graph nodes with events in the stochastic process
+associates recombination breakpoints with nodes. This encoding
+is limited in the ancestral patterns that can be described and ambiguous.
+(B) Equating graph nodes with \emph{genomes} and attaching the
+interval of genome inherited by the child node with edges removes
+this ambiguity and can describe any pattern of genetic inheritance.
+\textcolor{red}{This is very much a draft; let's see how it works with
+ the text and other figures and tidy it up in a later revision}
 }
@@
 chromosome, but it has significant drawbacks as the basis for a general purpose
-storage format.
+storage and interchange format.
+The concrete details of how an ARG should be stored and processed
+are not discussed in the literature, % weak, fix.
+but the following is a reasonable approach that mirrors the implementations
+in methods such as [x, y z].
+An ARG is composed of a set of $N$ nodes, where each node is an
+integer $0 \leq u < N$, and edges $(u, v)$ define the relatships between
+those nodes.
+Each node corresponds to an \emph{event}, and is either a common ancestor
+event (in which case it has two children and one parent)
+or a recombination event (when it has one child and two parents).
+Each recombination event is associated with a breakpoint $0 < x < L$,
+which we might encode as a mapping $f: N \rightarrow \{0, ..., L - 1\}$
+such that $f(u)$ is the recombination breakpoint if $u$ is a recombination
+node and $f(u) = 0$ otherwise.
 
@@
 
+\item Comparison metrics are a problem. All evaluations has essentially
+looked at either some form of pairwise comparison of trees along the genome
+using tree metrics, or looked at the distributions of TMRCAs. No
+one has succeeded in performing any comparisons of ARGs beyond
+regarding them as a sequence of independent trees. If we genuinely believe
+that ARGs are more than this, then we need to devise some metrics that
+consider more than the local tree structure.
+
 \item We should discuss other types of ancestry network also.

### 2022-02-11 paper.tex
Fix error in WF sim

@@
 Encoding an ARG. (A) The classical approach described by Griffiths
-equates graph nodes with events in the stochastic process
-associates recombination breakpoints with nodes. This encoding
+equates graph nodes with events in the stochastic process,
+and associates recombination breakpoints with nodes. This encoding
 is limited in the ancestral patterns that can be described and ambiguous.

### 2022-02-11 argutils/ancestry.py

@@ -391,7 +391,6 @@ def sim_wright_fisher(n, N, L, recomb_proba=1, seed=None):
         # lineages.
         ancestors.clear()
 
-        # print("|P| = ", len(parents), "RE_children = ", recombinant_nodes)
         for parent in parents.values():
             for lineages in parent.collected_lineages:
                 # These are all the lineages that have been collected
@@ -403,7 +402,8 @@ def sim_wright_fisher(n, N, L, recomb_proba=1, seed=None):
                 #     print("\t\t", lin)
 
                 merged_lineage = None
-                if len(lineages) == 1 and node_flags[lineage.node] == 0:
+                is_recombinant = (node_flags[lineage.node] & NODE_IS_RECOMB) != 0
+                if len(lineages) == 1 and not is_recombinant:
                     merged_lineage = lineages[0]
                 elif len(lineages) > 0:
                     if parent.id == -1:

### 2022-02-11 argutils/tests.py

@@ -27,6 +27,7 @@ class TestSimulate:
         assert ts.num_samples == 8
         assert ts.sequence_length == 5
         assert ts.num_trees == 1
+        assert ts.num_nodes == 15
         assert all(tree.num_roots == 1 for tree in ts.trees())
 
     def test_wright_fisher_high_recomb(self):

### 2022-02-11 argutils/ancestry.py
Set JSON md schema in wh99_example

@@ -439,6 +439,7 @@ def wh99_example():
     """
     L = 7
     tables = tskit.TableCollection(L)
+    tables.nodes.metadata_schema = tskit.MetadataSchema.permissive_json()
     t = 0
     for _ in range(3):
         tables.nodes.add_row(flags=tskit.NODE_IS_SAMPLE, time=t)

### 2022-02-11 paper.tex
working through ARG representation section

@@
 
-\begin{figure}
-\centering
-\begin{tabular}{cc}
-\begin{tikzpicture}[x=5mm, y=5mm, node distance=2mm and 20mm]
-\tikzset{greynode/.style={circle,fill,inner sep=1},
-nodelabel/.style={font=\footnotesize}}
-
-\node (s0) [greynode] at (0, 0) {};
-\node (s1) [greynode] at (3, 0) {};
-\node (s2) [greynode] at (6, 0) {};
-\node (s3) [greynode] at (3, 1) {};
-\node (s4) [greynode] at (1, 2) {};
-\node (s5) [greynode] at (5, 3) {};
-\node (s6) [greynode] at (3, 4) {};
-
-\node [nodelabel,anchor=north west] at ($(s3) + (0,0)$) {$x$};
-\foreach \u/\lab in {s0/0, s1/1, s2/2} \node[nodelabel,anchor=north] at (\u) {\lab};
-\foreach \u/\lab in {s4/4} \node[nodelabel,anchor=south west] at (\u) {\lab};
-\foreach \u/\lab in {s5/5} \node[nodelabel,anchor=south east] at (\u) {\lab};
-\foreach \u/\lab in {s3/3, s6/6} \node[nodelabel,anchor=south] at (\u) {\lab};
-
-%% Edges
-\draw (s1) -- (s3);
-\draw (s0) |- (s4);
-\draw (s4) -- (2,2) |- (s3);
-\draw (s4) |- (s6);
-\draw (s3) -- (4,1) |- (s5);
-\draw (s2) |- (s5);
-\draw (s5) |- (s6);
-
-\node [nodelabel,anchor=north west] at ($(-6.5,5)$) {
-\begin{tabular}{c|c}
-\multicolumn{2}{c}{Breakpoint}\\
-\hline
-0 & $\infty$ \\
-1 & $\infty$ \\
-2 & $\infty$ \\
-3 & $x$ \\
-4 & $\infty$ \\
-5 & $\infty$ \\
-6 & $\infty$ \\
-\end{tabular}};
-
-\node [nodelabel,anchor=north west] at ($(-2.5,5)$) {
-\begin{tabular}{l}
-Edges\\
-\hline
-$(0, 4)$ \\
-$(1, 3)$ \\
-$(2, 5)$ \\
-$(3, 4)$ \\
-$(3, 5)$ \\
-$(4, 6)$ \\
-$(5, 6)$ \\
-\end{tabular}};
-
-\end{tikzpicture}
-&
-\begin{tikzpicture}[x=5mm,y=5mm,node distance=2mm and 20mm]
-\tikzset{greynode/.style={circle,fill,inner sep=1},
-nodelabel/.style={font=\footnotesize}}
-
-\node (s0) [greynode] at (0, 0) {};
-\node (s1) [greynode] at (3, 0) {};
-\node (s2) [greynode] at (6, 0) {};
-\node (s3) [greynode] at (2, 1) {};
-\node (s4) [greynode] at (4, 1) {};
-\node (s5) [greynode] at (1, 2) {};
-\node (s6) [greynode] at (5, 3) {};
-\node (s7) [greynode] at (3, 4) {};
-
-\foreach \u/\lab in {s0/0, s1/1, s2/2} \node[nodelabel,anchor=north] at (\u) {\lab};
-\foreach \u/\lab in {s3/3} \node[nodelabel,anchor=east] at (\u) {\lab};
-\foreach \u/\lab in {s4/4} \node[nodelabel,anchor=west] at (\u) {\lab};
-\foreach \u/\lab in {s5/5} \node[nodelabel,anchor=south west] at (\u) {\lab};
-\foreach \u/\lab in {s6/6} \node[nodelabel,anchor=south east] at (\u) {\lab};
-\foreach \u/\lab in {s7/7} \node[nodelabel,anchor=south] at (\u) {\lab};
-
-%% Edges
-\draw (s1) |- (s3);
-\draw (s1) |- (s4) |- (s6);
-\draw (s3) |- (s5);
-\draw (s0) |- (s5);
-\draw (s2) |- (s6);
-\draw (s5) |- (s7);
-\draw (s6) |- (s7);
-
-\node [nodelabel,anchor=north west] at ($(-5,5)$) {
-\begin{tabular}{ll}
-Edges\\
-\hline
-$(0, 5)$ & $(0, L]$ \\
-$(1, 3)$ & $(0, x]$ \\
-$(1, 4)$ & $(x, L]$ \\
-$(2, 6)$ & $(0, L]$ \\
-$(3, 5)$ & $(0, L]$ \\
-$(4, 6)$ & $(0, L]$ \\
-$(5, 7)$ & $(0, L]$ \\
-$(6, 7)$ & $(0, L]$ \\
-\end{tabular}};
-
-\end{tikzpicture}
-\\
-(A) & (B) \\
-\end{tabular}
-\caption{\label{fig-arg-data-structure}
-Encoding an ARG. (A) The classical approach described by Griffiths
-equates graph nodes with events in the stochastic process,
-and associates recombination breakpoints with nodes. This encoding
-is limited in the ancestral patterns that can be described and ambiguous.
-(B) Equating graph nodes with \emph{genomes} and attaching the
-interval of genome inherited by the child node with edges removes
-this ambiguity and can describe any pattern of genetic inheritance.
-\textcolor{red}{This is very much a draft; let's see how it works with
- the text and other figures and tidy it up in a later revision}
-}
-\end{figure}
-
 \section*{The ARG as a data structure}
@@
 in methods such as [x, y z].
-An ARG is composed of a set of $N$ nodes, where each node is an
-integer $0 \leq u < N$, and edges $(u, v)$ define the relatships between
-those nodes.
+A Griffiths ARG is composed of a set of $N$ nodes, where each node is an
+integer $0 \leq u < N$, and edges $(u, v)$ define the relationships between
+those nodes (Fig.~\ref{fig-arg-data-structure}A).
 Each node corresponds to an \emph{event}, and is either a common ancestor
@@
 such that $f(u)$ is the recombination breakpoint if $u$ is a recombination
-node and $f(u) = 0$ otherwise.
+node and $f(u) = \infty$ otherwise. This encoding captures the information
+of a realisation of the ARG stochastic process, but it is important to
+note that it does not uniquely describe the underlying ancestral history
+without stipulating some additional ordering requirements.
+
+\begin{figure}
+\centering
+\begin{tabular}{cc}
+\begin{tikzpicture}[x=5mm, y=5mm, node distance=2mm and 20mm]
+\tikzset{greynode/.style={circle,fill,inner sep=1},
+nodelabel/.style={font=\footnotesize}}
+
+\node (s0) [greynode] at (0, 0) {};
+\node (s1) [greynode] at (3, 0) {};
+\node (s2) [greynode] at (6, 0) {};
+\node (s3) [greynode] at (3, 1) {};
+\node (s4) [greynode] at (1, 2) {};
+\node (s5) [greynode] at (5, 3) {};
+\node (s6) [greynode] at (3, 4) {};
+
+\node [nodelabel,anchor=north west] at ($(s3) + (0,0)$) {$x$};
+\foreach \u/\lab in {s0/0, s1/1, s2/2} \node[nodelabel,anchor=north] at (\u) {\lab};
+\foreach \u/\lab in {s4/4} \node[nodelabel,anchor=south west] at (\u) {\lab};
+\foreach \u/\lab in {s5/5} \node[nodelabel,anchor=south east] at (\u) {\lab};
+\foreach \u/\lab in {s3/3, s6/6} \node[nodelabel,anchor=south] at (\u) {\lab};
+
+%% Edges
+\draw (s1) -- (s3);
+\draw (s0) |- (s4);
+\draw (s4) -- (2,2) |- (s3);
+\draw (s4) |- (s6);
+\draw (s3) -- (4,1) |- (s5);
+\draw (s2) |- (s5);
+\draw (s5) |- (s6);
+
+\node [nodelabel,anchor=north west] at ($(-6.5,5)$) {
+\begin{tabular}{c|c}
+\multicolumn{2}{c}{Breakpoint}\\
+\hline
+0 & $\infty$ \\
+1 & $\infty$ \\
+2 & $\infty$ \\
+3 & $x$ \\
+4 & $\infty$ \\
+5 & $\infty$ \\
+6 & $\infty$ \\
+\end{tabular}};
+
+\node [nodelabel,anchor=north west] at ($(-2.5,5)$) {
+\begin{tabular}{l}
+Edges\\
+\hline
+$(0, 4)$ \\
+$(1, 3)$ \\
+$(2, 5)$ \\
+$(3, 4)$ \\
+$(3, 5)$ \\
+$(4, 6)$ \\
+$(5, 6)$ \\
+\end{tabular}};
+
+\end{tikzpicture}
+&
+\begin{tikzpicture}[x=5mm,y=5mm,node distance=2mm and 20mm]
+\tikzset{greynode/.style={circle,fill,inner sep=1},
+nodelabel/.style={font=\footnotesize}}
+
+\node (s0) [greynode] at (0, 0) {};
+\node (s1) [greynode] at (3, 0) {};
+\node (s2) [greynode] at (6, 0) {};
+\node (s3) [greynode] at (2, 1) {};
+\node (s4) [greynode] at (4, 1) {};
+\node (s5) [greynode] at (1, 2) {};
+\node (s6) [greynode] at (5, 3) {};
+\node (s7) [greynode] at (3, 4) {};
+
+\foreach \u/\lab in {s0/0, s1/1, s2/2} \node[nodelabel,anchor=north] at (\u) {\lab};
+\foreach \u/\lab in {s3/3} \node[nodelabel,anchor=east] at (\u) {\lab};
+\foreach \u/\lab in {s4/4} \node[nodelabel,anchor=west] at (\u) {\lab};
+\foreach \u/\lab in {s5/5} \node[nodelabel,anchor=south west] at (\u) {\lab};
+\foreach \u/\lab in {s6/6} \node[nodelabel,anchor=south east] at (\u) {\lab};
+\foreach \u/\lab in {s7/7} \node[nodelabel,anchor=south] at (\u) {\lab};
+
+%% Edges
+\draw (s1) |- (s3);
+\draw (s1) |- (s4) |- (s6);
+\draw (s3) |- (s5);
+\draw (s0) |- (s5);
+\draw (s2) |- (s6);
+\draw (s5) |- (s7);
+\draw (s6) |- (s7);
+
+\node [nodelabel,anchor=north west] at ($(-5,5)$) {
+\begin{tabular}{ll}
+Edges\\
+\hline
+$(0, 5)$ & $(0, L]$ \\
+$(1, 3)$ & $(0, x]$ \\
+$(1, 4)$ & $(x, L]$ \\
+$(2, 6)$ & $(0, L]$ \\
+$(3, 5)$ & $(0, L]$ \\
+$(4, 6)$ & $(0, L]$ \\
+$(5, 7)$ & $(0, L]$ \\
+$(6, 7)$ & $(0, L]$ \\
+\end{tabular}};
+
+\end{tikzpicture}
+\\
+(A) & (B) \\
+\end{tabular}
+\caption{\label{fig-arg-data-structure}
+Encoding an ARG. (A) The classical approach described by Griffiths
+equates graph nodes with events in the stochastic process,
+and associates recombination breakpoints with nodes. This encoding
+is limited in the ancestral patterns that can be described and ambiguous.
+(B) Equating graph nodes with \emph{genomes} and attaching the
+interval of genome inherited by the child node with edges removes
+this ambiguity and can describe any pattern of genetic inheritance.
+\textcolor{red}{This is very much a draft; let's see how it works with
+ the text and other figures and tidy it up in a later revision}
+}
+\end{figure}
+
+The Griffiths ARG encoding is ambiguous because it does not explicitly
+define which of the parent lineages resulting from a recombination
+event carries ancestral material to the left or right of the
+breakpoint. The ambiguity is resolved by adopting a simple rule when
+traversing the graph in order to generate the marginal trees for a
+given position $z$: at a recombination node $u$, we follow the
+first edge in which $u$ is a child if $f(u) \leq z$, and the
+second edge otherwise. This ordering requirement, which straightforward
+% Saying "awkward" too many times here, rephrase.
+to describe can be awkward in practise. For example,
+simulating an ARG in this representation is considerably
+complicated by the fact that the
+first event to occur may hit the right-hand lineage rather
+than the left.
+Such awkwardnesses can be worked around, of course,
+but in practise methods such has
+\texttt{beagle}~\citep{lyngso2005minimum} and
+\texttt{KwARG}~\citep{ignatieva2021kwarg}
+resolve this problem by
+associating a label ``P'' (prefix) or ``S'' (suffix) with the
+outbound edges of a recombination event.
+
+% it's not reasonable to assume that one type of event happens
+% at a time any more.
+The Griffiths ARG encoding is also limited in the types of ancestral
+history that can be described. This ARG encoding is derived from the coalescent
+with recombination, which explicitly assumes that only one type
+of event can occur (i.e., common ancestry or recombination) at a
+time. This model is derived as an approximation to a Wright-Fisher
+model, assuming that the sample size $n$ is much less than the
+population size $N$ and [some assumptions about genome length]
+\textcolor{red}{what is this Jere?}. While these approximations are
+reasonable
+to make when studying the Kreitman \emph{Drosphila Melanogaster}
+dataset~\citep{kreitman1983nucleotide} ($n=11$, $N\approx 10^6$,
+at 43 variant sites over x Kb \textcolor{red}{details and citation
+for Ne estimate}) that is classically used for benchmarking
+ARG inference methods [cites] , they are highly questionable in
+modern datasets. For example, there are now many
+human datasets consisting of hundreds of thousands
+of
+genomes~\citep{bycroft2018genome,karczewski2020mutational,tanjo2021practical},
+and thus our sample size is often much \emph{larger} than $N_e$
+(often assumed to be $10^4$ in humans).
+While the coalescent model is suprisingly robust to such
+basic violations to the underlying assumptions~\citep{wakeley2012gene},
+the combination of a large sample size and long sequence length
+can produce pathological behaviour in
+simulations~\citep{bhaskar2014distortion,nelson2020accounting}.
 

### 2022-02-11 paper.tex

@@
 this ambiguity and can describe any pattern of genetic inheritance.
+The most important change between the two encodings is to represent
+a recombination event (node 3 in A) by two parent genomes
+(nodes 3 and 4 in B) and explicitly storing the interval
+that the child inherits from each parent along the two corresponding
+edges. Thus, child $1$ in inherits $(0, x]$ from $3$
+and $(x, L)$ from $4$
+(for recombination breakpoint $x$ and a sequence of length $L$).
 \textcolor{red}{This is very much a draft; let's see how it works with
@@
 for Ne estimate}) that is classically used for benchmarking
-ARG inference methods [cites] , they are highly questionable in
-modern datasets. For example, there are now many
-human datasets consisting of hundreds of thousands
-of
+ARG inference methods [cites],
+they are highly questionable in modern datasets. % absurd really
+For example, there are now many
+human datasets consisting of hundreds of thousands of
 genomes~\citep{bycroft2018genome,karczewski2020mutational,tanjo2021practical},
@@
 (often assumed to be $10^4$ in humans).
-While the coalescent model is suprisingly robust to such
-basic violations to the underlying assumptions~\citep{wakeley2012gene},
-the combination of a large sample size and long sequence length
-can produce pathological behaviour in
-simulations~\citep{bhaskar2014distortion,nelson2020accounting}.
-
-[Not sure how to make a narrative out of this, but here are the
-basic problems:]
-\begin{enumerate}
-\item The encoding is hamstrung by the interpretation of
-nodes as \emph{events} of a particular type. We can either
-have common ancestor or recombination events. This is true
-in the coalescent, but is not in a more general Wright-Fisher
-model.
-%% Note - I'm not sure this is true, if the nodes are *events* rather than
-%% genomes, unless you want events (rather than genomes) to occur at
-%% integer generations.
-\item We can also only have one crossover per recombination event.
-This is unrealistic because things like gene conversion are common. It also
-means we can't have back and forth recombination between
-two parents, or represent the transmission of multiple chromosomes.
-\item The passage of ancestral material along edges requires
-extrinsic rules which must be shared by all implementations.
-These rules can be quite restrictive when generating, and
-in practise some have found it simpler to add explicit
-annotations to the edges.
-\item Generating local trees is inefficient, essentially requiring
-$O(n T)$ time (or more?)
-\end{enumerate}
-
-[WIP jk]
-
-This ambiguity can be dealt with in a number of ways. In the original
-treatments, the direction in which ancestral material travels at
-a breakpoint is handled by a simple rule, sites to the left of the
-breakpoint go to the left parent and sites to right go to the
-right parent. [Illustrate the problem with simulating an ARG
-like this]. ~\cite{ignatieva2021kwarg} resolve this problem by
-associating a label ``P'' (prefix) or ``S'' (suffix) with the
-outbound edges of a recombination event. This is the same approach
-as used in beagle~\citep{lyngso2005minimum}.
+% TODO make this better. Ask Gregor for some refs?
+% Basically, Ag datasets make the assumption of isolated
+% rare recombinations absurd.
+Agricultural datasets such as the 1000 bull genomes~\citep{hayes20191000}
+combine genome sequence data with detailed pedigrees, with
+sample data from multiple generations.
+Complete chromosome-level assemblies are now possible
+in humans~\citep{miga2020telomere},
+and projects are under way to obtain high-quality assemblies
+for all eukaryotic species in Britain and Ireland~\citep{darwin2022sequence}
+and ultimately worldwide~\citep{lewin2022earth}.
+While the coalescent model can be suprisingly robust to such
+basic violations to the underlying assumptions~\citep{
+wakeley2012gene,bhaskar2014distortion,nelson2020accounting},
+it is not acceptable that our ability to \emph{represent}
+ancestral relationships should be limited by these assumptions.
+In a Wright-Fisher model, coalescence and recombination can
+occur at the same time in the same genomes, and we can be
+certain that this is also true in the incredibly rich
+datasets that are available today.
+
+Another limitation of the Griffiths ARG encoding illustated in
+Fig.~\ref{fig-arg-data-structure}A is the inability to represent
+multiple recombinations along a chromosome, gene conversion,
+or transmission of multiple chromosomes.
+[Sentence saying ``gene conversion is important, actually, cite, cite,
+so we should be hoping to infer it some day''. Presumably people
+already are in bacterial contexts?]
+To allow us to encode data from these important processes we
+must either devise workarounds to the event based encoding
+(i.e., model gene conversion as two recombination events) or
+generalise the model to more appropriately reflect the realities
+of modern datasets.
+
+Fortunately, it is straightforward to generalise the classical Griffiths
+encoding to both simplify implementations and to encompass any
+possible pattern of genetic inheritance.
+We require only two changes. Firstly, we let nodes in the
+graph model \emph{genomes} rather than \emph{events}. As we
+see in the next section, this allows us to model complex
+patterns of inheritance without classifying different types of node.
+The second change we must make is to encode
+intervals of ancestry inheritance \emph{explicitly} on graph edges,
+rather than \emph{implicitly} as breakpoints associated with
+recombination events and edge ordering rules.
+That is, given two genomes $p$
+and $c$ where $p$ is the parent genome and $c$ is the child
+(although many generations may have elapsed between the two),
+we associate a set $I$ of intervals of genome which $c$ has
+inherited from $p$.
+These simple changes required to translate from the
+classical Griffiths approach to the proposed encoding
+are shown in Fig.\ref{fig-arg-data-structure}.
+
+\textcolor{red}{
+This falls a bit flat here,  but it's not clear how we're dividing
+the responsibility for explaining the model between the
+different sections.
+Basically, we can't explain everything at
+the end of this section, but we want to give the reader the
+basic idea (which is genuinely simple). Can see how things are
+shaping up after we've got a draft of the rest of the
+sections, I guess.}
 

### 2022-02-11 argutils/tests.py
Implement 3 new types of simplification

@@ -1,4 +1,6 @@
 # Run using python3 -m pytest argutils/tests.py
+import collections
+import string
 
 import tskit
 import pytest
@@ -131,3 +133,97 @@ class TestResolve:
         # print()
         # print(resolved.draw_text())
         assert resolved.equals(resolved2, ignore_provenance=True)
+
+class TestLabels:
+    def test_viz_label_nodes_many(self):
+        ts = argutils.sim_coalescent(20, 0.1, 10, seed=123) # over 26 nodes
+        ts = argutils.viz.label_nodes(ts)
+        assert ts.num_nodes > len(string.ascii_uppercase)
+        i = 0
+        for c in string.ascii_uppercase:
+            assert ts.node(i).metadata["name"] == c
+            i += 1
+        while i < ts.num_nodes:
+            assert ts.node(i).metadata["name"] == i
+            i += 1
+
+    def test_viz_label_nodes_bespoke(self):
+        ts = argutils.sim_coalescent(20, 0.1, 10, seed=123) # over 26 nodes
+        ts = argutils.viz.label_nodes(
+            ts, labels={n: str(i) for n, i in enumerate(range(ts.num_nodes, 0, -1))})
+        for nd in ts.nodes():
+            assert nd.metadata["name"] == f"{ts.num_nodes - nd.id}"
+    
+
+class TestSimplifyFunctions:
+    def test_simplify_keeping_all_nodes(self):
+        ts = argutils.viz.label_nodes(argutils.wh99_example())
+        ts2 = argutils.simplify_keeping_all_nodes(ts)
+        assert ts.num_edges > ts2.num_edges
+        assert ts.num_nodes == ts2.num_nodes
+        for n1, n2 in zip(ts.nodes(), ts2.nodes()):
+            assert "name" in n1.metadata and n1.metadata["name"]
+            assert n1.metadata["name"] == n2.metadata["name"]
+
+    def test_simplify_remove_pass_through(self):
+        ts = argutils.sim_coalescent(10, 0.1, 10, seed=3)
+        node_edges = np.zeros((ts.num_nodes, 2), dtype=int)
+        has_pass_through = False
+        for e in ts.edges():
+            node_edges[e.child][0] += 1
+            node_edges[e.parent][1] += 1
+        # at least one node with 1 parent and 1 child only
+        for row in node_edges:
+            if list(row) == [1, 1]:
+                has_pass_through = True
+        assert has_pass_through
+
[…]

### 2022-02-11 argutils/ancestry.py
Demo of using a single re node by breakpoint saving

@@ -540,12 +540,27 @@ def sim_wright_fisher(n, N, L, recomb_proba=1, seed=None):
     return tables.tree_sequence()
 
 
-def wh99_example():
+def wh99_example(one_node_recombination=None):
     """
     The example ARG from figure 1 of Wiuf and Hein 99, Recombination as a Point Process
     along Sequences. Each event is given a time increment of 1.
     """
+    if one_node_recombination is None:
+        one_node_recombination = False
     L = 7
+    class SingleRENode:
+        def __init__(self, node_id, breakpoint):
+            self.id = node_id
+            self.breakpoint = breakpoint
+        @property
+        def l(self):
+            return self.id, 0, self.breakpoint
+        @property
+        def r(self):
+            return self.id, self.breakpoint, L
+
+    DoubleRENode = collections.namedtuple("DoubleRENode", "l, r")
+
     tables = tskit.TableCollection(L)
     nodes = tables.nodes
     edges = tables.edges
@@ -557,33 +572,54 @@ def wh99_example():
     def re(child, x):
         nonlocal t
         t += 1
-        left_parent = nodes.add_row(time=t)
-        right_parent = nodes.add_row(time=t)
-        edges.add_row(0, x, left_parent, child)
-        edges.add_row(x, L, right_parent, child)
-        nodes[child] = nodes[child].replace(flags=nodes[child].flags | NODE_IS_RECOMB)
+        if one_node_recombination:
+            parent = nodes.add_row(time=t)
+            try:
+                child, left, right = child
+            except TypeError:
+                left = 0
+                right = L
+            edges.add_row(left, right, parent, child)
+            return SingleRENode(parent, x)
+        else:
+            left_parent = nodes.add_row(time=t)
+            right_parent = nodes.add_row(time=t)
+            edges.add_row(0, x, left_parent, child)
+            edges.add_row(x, L, right_parent, child)
+            nodes[child] = nodes[child].replace(flags=nodes[child].flags | NODE_IS_RECOMB)
+            return DoubleRENode(left_parent, right_parent)
 
     def ca(*args):
         nonlocal t
         t += 1
[…]

### 2022-02-11 argutils/ancestry.py
Demo of using a single re node by breakpoint saving

@@ -517,12 +517,27 @@ def sim_wright_fisher(n, N, L, recomb_proba=1, seed=None):
     return tables.tree_sequence()
 
 
-def wh99_example():
+def wh99_example(one_node_recombination=None):
     """
     The example ARG from figure 1 of Wiuf and Hein 99, Recombination as a Point Process
     along Sequences. Each event is given a time increment of 1.
     """
+    if one_node_recombination is None:
+        one_node_recombination = False
     L = 7
+    class SingleRENode:
+        def __init__(self, node_id, breakpoint):
+            self.id = node_id
+            self.breakpoint = breakpoint
+        @property
+        def l(self):
+            return self.id, 0, self.breakpoint
+        @property
+        def r(self):
+            return self.id, self.breakpoint, L
+
+    DoubleRENode = collections.namedtuple("DoubleRENode", "l, r")
+
     tables = tskit.TableCollection(L)
     tables.nodes.metadata_schema = tskit.MetadataSchema.permissive_json()
     t = 0
@@ -532,32 +547,53 @@ def wh99_example():
     def re(child, x):
         nonlocal t
         t += 1
-        left_parent = tables.nodes.add_row(time=t)
-        right_parent = tables.nodes.add_row(time=t)
-        tables.edges.add_row(0, x, left_parent, child)
-        tables.edges.add_row(x, L, right_parent, child)
+        if one_node_recombination:
+            parent = tables.nodes.add_row(time=t)
+            try:
+                child, left, right = child
+            except TypeError:
+                left = 0
+                right = L
+            tables.edges.add_row(left, right, parent, child)
+            return SingleRENode(parent, x)
+        else:
+            left_parent = tables.nodes.add_row(time=t)
+            right_parent = tables.nodes.add_row(time=t)
+            tables.edges.add_row(0, x, left_parent, child)
+            tables.edges.add_row(x, L, right_parent, child)
+            return DoubleRENode(left_parent, right_parent)
 
     def ca(*args):
         nonlocal t
         t += 1
         parent = tables.nodes.add_row(time=t)
         for child in args:
[…]

### 2022-02-13 paper.tex

@@
 is free from the limitations discussed in the previous section.
-\cite{mathieson2020ancestry} and others
+\citet{mathieson2020ancestry} and others
 \citep{wakeley2012genetics,gusfield2014recombinatorics,speed2015naturereviewsgenetics}
@@
 \begin{figure}
-\vspace{5em}
+  \begin{subfigure}[b]{0.25\textwidth}
+    \centering
+    \includegraphics[height=0.3\textheight]{illustrations/pedigree}
+    \caption{Pedigree with genetic inheritance}
+    \label{fig:pedigree}
+  \end{subfigure}
+  \hfill
+  \begin{subfigure}[b]{0.25\textwidth}
+    \centering
+    \includegraphics[height=0.3\textheight]{illustrations/pedigree_ARG}
+    \caption{Graph representation of genetic inheritance}
+    \label{fig:ARG-of-pedigree}
+  \end{subfigure}
+  \hfill
+  \begin{subfigure}[b]{0.4\textwidth}
+    \centering
+    \includegraphics[height=0.3\textheight]{illustrations/pedigree_trees}
+    \caption{Tree representation of genetic inheritance}
+    \label{fig:ts-of-pedigree}
+  \end{subfigure}
 \caption{\label{fig-pedigree-and-arg}
 The (unannotated) ARG is closely related to the pedigree
-graph, where each ``individual'' node is replace by two ``genome''
+graph, where each ``individual'' node is replaced by two ``genome''
 nodes, and edges between parents and children are only included if there is

### 2022-02-13 paper.tex
Text for pedigree-based model

@@
 depends on summaries ultimately derived from the
-trees~\citep{tajima1983evolutionary,tavare1984line,fu1995statistical}
+trees~\citep{tajima1983evolutionary,tavare1984line}
 such as allele frequencies and site frequency
@@
 with some examples.
+% Perhaps we should also mention something about simplification and unknowable
+% nodes, e.g. "show how some of the nodes are not required for useful interpretation
+% of ancestry".
 \end{enumerate}
@@
 \begin{tikzpicture}
-	% Samples at x = 0, 4, and 6
-	\node [circle, fill, scale=0.2,label=above right:{\small A}] at (0,0) {A};
 	\draw (-0.4, -0.2) -- (0.4, -0.2) -- (0.4, -0.3) -- (-0.4, -0.3) -- (-0.4, -0.2);
@@
 	\node [label=below:{\small $v$}] at (-0.1, -0.2) {};
-	\node [circle, fill, scale=0.2,label=above right:{\small B}] at (4,0) {B};
 	\draw (3.6, -0.2) -- (4.4, -0.2) -- (4.4, -0.3) -- (3.6, -0.3) -- (3.6, -0.2);
-	\node [circle, fill, scale=0.2,label=above left:{\small C}] at (6,0) {C};
 	\draw (5.6, -0.2) -- (6.4, -0.2) -- (6.4, -0.3) -- (5.6, -0.3) -- (5.6, -0.2);
 
-	% B and C merge into D
 	\draw (4,0) -- (4, 1.2) -- (6, 1.2) -- (6,0);
-	\node [circle, fill, scale=0.2,label=above right:{\small D}] at (5,1.2) {D};
 	\draw (4.6, 1) -- (5.4, 1) -- (5.4, 0.9) -- (4.6, 0.9) -- (4.6, 1);
@@
 
-	% A splits into E and F
 	\draw (0, 0) -- (0, 1.5) -- (-1,1.5) -- (1,1.5);
-	\node [circle, fill, scale=0.2,label=above left:{\small E}] at (-1,1.5) {E};
 	\draw (-1.4, 1.3) -- (-1.1, 1.3) -- (-1.1, 1.2) -- (-1.4, 1.2) -- (-1.4, 1.3);
 	\draw (-1.1, 1.25) -- (-0.6, 1.25);
-	\node [circle, fill, scale=0.2,label=above left:{\small F}] at (1,1.5) {F};
 	\draw (0.6, 1.25) -- (0.9, 1.25);
@@
 
-	% D splits into G and H
 	\draw (5, 1.2) -- (5, 2.4) -- (4,2.4) -- (6,2.4);
-	\node [circle, fill, scale=0.2,label=above left:{\small G}] at (4,2.4) {G};
 	\draw (3.6, 2.2) -- (4.1, 2.2) -- (4.1, 2.1) -- (3.6, 2.1) -- (3.6, 2.2);
 	\draw (4.1, 2.15) -- (4.4, 2.15);
-	\node [circle, fill, scale=0.2,label=above left:{\small H}] at (6,2.4) {H};
 	\draw (5.6, 2.15) -- (6.1, 2.15);
@@
 
-	% F splits into I and J
 	\draw (1, 1.5) -- (1, 3.1) -- (0,3.1) -- (2,3.1);
-	\node [circle, fill, scale=0.2,label=above left:{\small I}] at (0,3.1) {I};
 	\draw (-0.4, 2.85) -- (0.2, 2.85);
 	\draw (0.2, 2.9) -- (0.4, 2.9) -- (0.4, 2.8) -- (0.2, 2.8) -- (0.2, 2.9);
-	\node [circle, fill, scale=0.2,label=above left:{\small J}] at (2,3.1) {J};
 	\draw (1.6, 2.85) -- (1.9, 2.85);
@@
 
-	% E and I merge into K
 	\draw (-1, 1.5) -- (-1, 4.5) -- (0,4.5) -- (0,3.1);
-	\node [circle, fill, scale=0.2,label=above left:{\small K}] at (-0.5,4.5) {K};
+	\node [scale=0.2,label=above left:{\small A}] at (-0.5,4.5) {A};
 	\draw (-0.9, 4.3) -- (-0.6, 4.3) -- (-0.6, 4.2) -- (-0.9, 4.2) -- (-0.9, 4.3);
@@
 
-	% G and J merge into L
 	\draw (4,2.4) -- (4,5.4) -- (2,5.4) -- (2,3.1);
-	\node [circle, fill, scale=0.2,label=above left:{\small L}] at (3,5.4) {L};
+	\node [scale=0.2,label=above left:{\small B}] at (3,5.4) {B};
 	\draw (2.6, 5.2) -- (2.9, 5.2) -- (2.9, 5.1) -- (2.6, 5.1) -- (2.6, 5.2);
@@
 
-	% K splits into M and N
 	\draw (-0.5, 4.5) -- (-0.5, 6) -- (-1.5,6) -- (0.5,6);
-	\node [circle, fill, scale=0.2,label=above left:{\small M}] at (-1.5,6) {M};
 	\draw (-1.9, 5.75) -- (-1.3, 5.75);
 	\draw (-1.3, 5.8) -- (-1.1, 5.8) -- (-1.1, 5.7) -- (-1.3, 5.7) -- (-1.3, 5.8);
-	\node [circle, fill, scale=0.2,label=above left:{\small N}] at (0.5,6) {N};
 	\draw (0.1, 5.8) -- (0.4, 5.8) -- (0.4, 5.7) -- (0.1, 5.7) -- (0.1, 5.8);
@@
 
-	% L and N merge into O
 	\draw (0.5, 6) -- (0.5, 7.2) -- (3,7.2) -- (3,5.4);
-	\node [circle, fill, scale=0.2,label=above left:{\small O}] at (1.75,7.2) {O};
+	\node [scale=0.2,label=above:{\small C}] at (1.75,7.2) {C};
 	\draw (1.35, 6.95) -- (2.15, 6.95);
 
-	% M and H continue to the end
 	\draw (-1.5, 6) -- (-1.5, 8);
@@
 
-	% Edge annotations above each node
-	\node [label=left:{\small$\{[0,L)\};1$}] at (0.2,0.6) {}; % A
-	\node [label=left:{\small$\{[0,L)\};1$}] at (4.2,0.4) {}; % B
-	\node [label=right:{\small$\{[0,L)\};1$}] at (5.8,0.4) {}; % C
-	\node [label=left:{\small$\{[0,L)\};2$}] at (5.2,1.6) {}; % D
-	\node [label=left:{\small$\{[0,v)\};1$}] at (-0.8,2.3) {}; % E
-	\node [label=right:{\small$\{[v,L)\};1$}] at (0.8,2.1) {}; % F
-	\node [label=right:{\small$\{[0,w)\};2$}] at (3.8,3.9) {}; % G
-	\node [label=left:{\small$\{[w,L)\};2$}] at (6.2,5) {}; % H
-	\node [label=right:{\small$\{[x,L)\};1$}] at (-0.2,3.8) {}; % I
-	\node [label=right:{\small$\{[v,x)\};1$}] at (1.8,4.1) {}; % J
-	\node [label=left:{\small$\{[0,v), [x, L)\};1$}] at (-0.3,5.2) {}; % K
-	\node [label=right:{\small$\{[0,v)\};2$}] at (2.8,6.4) {}; % L
-	\node [label=left:{\small$\{[x,L)\};1$}] at (-1.3,7) {}; % M
-	\node [label=right:{\small$\{[0,v)\};1$}] at (0.3,6.4) {}; % N
+	% Edge annotations above each event
+	\node [label=left:{\small$\{[0,L)\};1$}] at (0.2,0.6) {}; 
+	\node [label=left:{\small$\{[0,L)\};1$}] at (4.2,0.4) {}; 
+	\node [label=right:{\small$\{[0,L)\};1$}] at (5.8,0.4) {}; 
+	\node [label=left:{\small$\{[0,L)\};2$}] at (5.2,1.6) {}; 
+	\node [label=left:{\small$\{[0,v)\};1$}] at (-0.8,2.3) {}; 
+	\node [label=right:{\small$\{[v,L)\};1$}] at (0.8,2.1) {}; 
+	\node [label=right:{\small$\{[0,w)\};2$}] at (3.8,3.9) {}; 
+	\node [label=left:{\small$\{[w,L)\};2$}] at (6.2,5) {}; 
+	\node [label=right:{\small$\{[x,L)\};1$}] at (-0.2,3.8) {}; 
+	\node [label=right:{\small$\{[v,x)\};1$}] at (1.8,4.1) {}; 
+	\node [label=left:{\small$\{[0,v), [x, L)\};1$}] at (-0.3,5.2) {}; 
+	\node [label=right:{\small$\{[0,v)\};2$}] at (2.8,6.4) {};
+	\node [label=left:{\small$\{[x,L)\};1$}] at (-1.3,7) {};
+	\node [label=right:{\small$\{[0,v)\};1$}] at (0.3,6.4) {};
 
@@
 	% Numbers of extant ancestors and links, from top to bottom
-	\node[label=right:{Extant material}] at (7.5, 8.2) {};
-	\node[label=right:{2 \; $2 L - x - w$}] at (7.5, 7.6) {};
-	\node[label=right:{4 \; $2 L + 2 v - x - w$}] at (7.5, 6.6) {};
-	\node[label=right:{3 \; $2 L + v - w$}] at (7.5, 5.7) {};
-	\node[label=right:{4 \; $2 L + x - v$}] at (7.5, 4.95) {};
-	\node[label=right:{5 \; $2 L$}] at (7.5, 3.8) {};
-	\node[label=right:{4 \; $2 L$}] at (7.5, 2.75) {};
-	\node[label=right:{3 \; $2 L$}] at (7.5, 1.95) {};
-	\node[label=right:{2 \; $2 L$}] at (7.5, 1.35) {};
-	\node[label=right:{3 \; $3 L$}] at (7.5, 0.6) {};
+	\node[label=right:{CA \; RE}] at (7.5, 8.2) {};
+	\node[label=right:{$\binom{2}{2}$ \; $(2 L - x - w) \rho$}] at (7.5, 7.6) {};
+	\node[label=right:{$\binom{4}{2}$ \; $(2 L + 2 v - x - w) \rho$}] at (7.5, 6.6) {};
+	\node[label=right:{$\binom{3}{2}$ \; $(2 L + v - w) \rho$}] at (7.5, 5.7) {};
+	\node[label=right:{$\binom{4}{2}$ \; $(2 L + x - v) \rho$}] at (7.5, 4.95) {};
+	\node[label=right:{$\binom{5}{2}$ \; $2 L \rho$}] at (7.5, 3.8) {};
+	\node[label=right:{$\binom{4}{2}$ \; $2 L \rho$}] at (7.5, 2.75) {};
+	\node[label=right:{$\binom{3}{2}$ \; $2 L \rho$}] at (7.5, 1.95) {};
+	\node[label=right:{$\binom{2}{2}$ \; $2 L \rho$}] at (7.5, 1.35) {};
+	\node[label=right:{$\binom{3}{2}$ \; $3 L \rho$}] at (7.5, 0.6) {};
 
 	% Gray dashed lines to visually separate holding times
-	\draw[color=gray, dashed] (7.5, 1.2) -- (11.1, 1.2); % D
-	\draw[color=gray, dashed] (7.5, 1.5) -- (11.1, 1.5); % E and F
-	\draw[color=gray, dashed] (7.5, 2.4) -- (11.1, 2.4); % G and H
-	\draw[color=gray, dashed] (7.5, 3.1) -- (11.1, 3.1); % I and J
-	\draw[color=gray, dashed] (7.5, 4.5) -- (11.1, 4.5); % K
-	\draw[color=gray, dashed] (7.5, 5.4) -- (11.1, 5.4); % L
-	\draw[color=gray, dashed] (7.5, 6) -- (11.1, 6); % M and N
-	\draw[color=gray, dashed] (7.5, 7.2) -- (11.1, 7.2); % O
+	\draw[color=gray, dashed] (7.5, 1.2) -- (11.1, 1.2);
+	\draw[color=gray, dashed] (7.5, 1.5) -- (11.1, 1.5);
+	\draw[color=gray, dashed] (7.5, 2.4) -- (11.1, 2.4);
+	\draw[color=gray, dashed] (7.5, 3.1) -- (11.1, 3.1);
+	\draw[color=gray, dashed] (7.5, 4.5) -- (11.1, 4.5);
+	\draw[color=gray, dashed] (7.5, 5.4) -- (11.1, 5.4);
+	\draw[color=gray, dashed] (7.5, 6) -- (11.1, 6);
+	\draw[color=gray, dashed] (7.5, 7.2) -- (11.1, 7.2);
 \end{tikzpicture}
@@
 
-	% E and I merge into K
 	\draw (-1, 1.5) -- (-1, 4.5) -- (0,4.5) -- (0,3.1);
-	\node [circle, fill, scale=0.2,label=above left:{\small K}] at (-0.5,4.5) {K};
 	\draw (-0.9, 4.3) -- (-0.6, 4.3) -- (-0.6, 4.2) -- (-0.9, 4.2) -- (-0.9, 4.3);
@@
 
-	% H splits into P and Q. Red Q node drawn further down so that it overlays the black line underneath
 	\draw (6, 2.4) -- (6, 3.8) -- (7, 3.8);
 	\draw [color=red](6, 3.8) -- (5, 3.8);
-	\node [color=red, circle, fill, scale=0.2,label={[red]above left:{\small P}}] at (5,3.8) {P};
 	\draw (4.6, 3.55) -- (5.4, 3.55);
@@
 
-	% G and J merge into L
 	\draw (4,2.4) -- (4,5.4) -- (2,5.4) -- (2,3.1);
-	\node [circle, fill, scale=0.2,label=above left:{\small L}] at (3,5.4) {L};
 	\draw (2.6, 5.2) -- (2.9, 5.2) -- (2.9, 5.1) -- (2.6, 5.1) -- (2.6, 5.2);
@@
 
-	% P and L merge into R. R node drawn further down so that it overlays the black lines underneath
 	\draw [color=red](5, 3.8) -- (5, 6.6) -- (4, 6.6);
@@
 
-	% R and N merge into O. O node drawn further down so that it overlays the red line underneath
 	\draw (0.5, 6) -- (0.5, 7.2) -- (4,7.2) -- (4,6.6);
 	\draw (1.85, 6.95) -- (2.65, 6.95);
-	% R node can now be added
-	\node [color=red, circle, fill, scale=0.2,label={[red]above left:{\small R}}] at (4,6.6) {R};
-
-	% M, O, and Q continue to the end
+
 	\draw (-1.5, 6) -- (-1.5, 8);
@@
 	\draw (7, 3.8) -- (7, 8);
-	% O and Q nodes can now be added
-	\node [circle, fill, scale=0.2,label=above left:{\small O}] at (2.25,7.2) {O};
-	\node [color=red, circle, fill, scale=0.2,label={[red]above left:{\small Q}}] at (7,3.8) {Q};
 
@@
 	% Numbers of extant ancestors and links, from top to bottom
-	\node[label=right:{Extant material}] at (7.5, 8.2) {};
-	\node[label=right:{3 \; $3 L$}] at (7.5, 7.6) {};
-	\node[label=right:{4 \; $4 L$}] at (7.5, 6.9) {};
-	\node[label=right:{5 \; $5 L$}] at (7.5, 6.3) {};
-	\node[label=right:{4 \; $4 L$}] at (7.5, 5.7) {};
-	\node[label=right:{5 \; $5 L$}] at (7.5, 4.95) {};
-	\node[label=right:{6 \; $6 L$}] at (7.5, 4.15) {};
-	\node[label=right:{5 \; $5 L$}] at (7.5, 3.45) {};
-	\node[label=right:{4 \; $4 L$}] at (7.5, 2.75) {};
-	\node[label=right:{3 \; $3 L$}] at (7.5, 1.95) {};
-	\node[label=right:{2 \; $2 L$}] at (7.5, 1.35) {};
-	\node[label=right:{3 \; $3 L$}] at (7.5, 0.6) {};
+	\node[label=right:{CA \; RE}] at (7.5, 8.2) {};
+	\node[label=right:{$\binom{3}{2}$ \; $3 L \rho$}] at (7.5, 7.6) {};
+	\node[label=right:{$\binom{4}{2}$ \; $4 L \rho$}] at (7.5, 6.9) {};
+	\node[label=right:{$\binom{5}{2}$ \; $5 L \rho$}] at (7.5, 6.3) {};
+	\node[label=right:{$\binom{4}{2}$ \; $4 L \rho$}] at (7.5, 5.7) {};
+	\node[label=right:{$\binom{5}{2}$ \; $5 L \rho$}] at (7.5, 4.95) {};
+	\node[label=right:{$\binom{6}{2}$ \; $6 L \rho$}] at (7.5, 4.15) {};
+	\node[label=right:{$\binom{5}{2}$ \; $5 L \rho$}] at (7.5, 3.45) {};
+	\node[label=right:{$\binom{4}{2}$ \; $4 L \rho$}] at (7.5, 2.75) {};
+	\node[label=right:{$\binom{3}{2}$ \; $3 L \rho$}] at (7.5, 1.95) {};
+	\node[label=right:{$\binom{2}{2}$ \; $2 L \rho$}] at (7.5, 1.35) {};
+	\node[label=right:{$\binom{3}{2}$ \; $3 L \rho$}] at (7.5, 0.6) {};
 
 	% Gray dashed lines to visually separate holding times
-	\draw[color=gray, dashed] (7.5, 1.2) -- (11, 1.2); % D
-	\draw[color=gray, dashed] (7.5, 1.5) -- (11, 1.5); % E and F
-	\draw[color=gray, dashed] (7.5, 2.4) -- (11, 2.4); % G and H
-	\draw[color=gray, dashed] (7.5, 3.1) -- (11, 3.1); % I and J
-	\draw[color=gray, dashed] (7.5, 3.8) -- (11, 3.8);	% P and Q
-	\draw[color=gray, dashed] (7.5, 4.5) -- (11, 4.5); % K
-	\draw[color=gray, dashed] (7.5, 5.4) -- (11, 5.4); % L
-	\draw[color=gray, dashed] (7.5, 6) -- (11, 6); % M and N
-	\draw[color=gray, dashed] (7.5, 6.6) -- (11, 6.6);	% R
-	\draw[color=gray, dashed] (7.5, 7.2) -- (11, 7.2); % O
+	\draw[color=gray, dashed] (7.5, 1.2) -- (11, 1.2); 
+	\draw[color=gray, dashed] (7.5, 1.5) -- (11, 1.5); 
+	\draw[color=gray, dashed] (7.5, 2.4) -- (11, 2.4); 
+	\draw[color=gray, dashed] (7.5, 3.1) -- (11, 3.1); 
+	\draw[color=gray, dashed] (7.5, 3.8) -- (11, 3.8);
+	\draw[color=gray, dashed] (7.5, 4.5) -- (11, 4.5); 
+	\draw[color=gray, dashed] (7.5, 5.4) -- (11, 5.4);
+	\draw[color=gray, dashed] (7.5, 6) -- (11, 6); 
+	\draw[color=gray, dashed] (7.5, 6.6) -- (11, 6.6);
+	\draw[color=gray, dashed] (7.5, 7.2) -- (11, 7.2); 
 \end{tikzpicture}
@@
 sample of three continuous chromosomes of length $L$ at time $t = 0$, and
-propagated until time $T$. The MRCA on the genetic interval $[x, z)$ is reached
-at node L, while that on $[0, x)$ is reached at node O.
-The non-ancestral segment $[x, y)$ above
-node K contributes to the rate of effective recombinations because it
-is trapped between ancestral segments. The two columns titled ``Extant
-material" are the number of extant lineages and the total length of genetic
-material on which an effective recombination can take place during each time
-interval.
+propagated until time $T$. The MRCA on the genetic interval $[v, w)$ is reached
+at event B, while that on $[0, v)$ is reached at event C.
+The non-ancestral segment $[v, w)$ above
+A contributes to the rate of effective recombinations because it
+is trapped between ancestral segments. The two columns titled CA and RE
+are the respective rates of mergers and recombinations when
+the recombination rate is $\rho$.
 (B) A corresponding realisation of a big ARG, which augments Hudson's algorithm
@@
 process, the original ARG literature is focused on mathematical
-results about the stochastic process. Griffiths and colleagues formulated the
-ARG as a branching-coalescing stochastic process. We imagine a graph in which
-each edge corresponds to an extant lineage and nodes are events in the
-process (the initial $n$ leaf nodes are ``sampling'' events). As in Hudson's
+results about the stochastic process. Griffiths and colleagues used the word
+``ARG'' to refer to both a specific branching-coalescing stochastic process and
+the graph structure derived from this process. For clarity, we will refer to the
+process as the Griffiths process, or the coalescent with recombination (CwR),
+and following XX-Who?-XX, we refer to the the structure as the ``big ARG'', as
+opposed to the ``little ARG'' traversed by Hudson's algorithm
+(\citet{wiuf1999recombination} refer to these structures as ARG and HUD respectively).
+
+In the Griffiths formulation, each edge in the graph corresponds to an extant
+lineage and nodes are events in the process (the initial $n$ leaf nodes are
+``sampling'' events). As in Hudson's
 algorithm, common ancestor events occur at rate $\binom{k}{2}$ when there
@@
 the specific distribution of ancestral material among lineages, the
-rate of recombination is simply $k \rho (m - 1)$ in the original ARG
-formulation. At a recombination event we choose a lineage (edge) uniformly, and a
+rate of recombination is now simply $k \rho (m - 1)$. At a recombination event
+we choose a lineage (edge) uniformly, and a
 breakpoint $0 < x < L$ uniformly on its genome. We terminate the edge at a
@@
 
-The state-space of this process is much simpler than Hudson's algorithm, which
-greatly facilitates mathematical reasoning. This simplicity comes at a
+The state-space of the Griffiths process is much simpler than Hudson's algorithm,
+which greatly facilitates mathematical reasoning. This simplicity comes at a
 substantial cost, however, if we wish to use it as a practical means of
-simulating recombinant ancestry. The ARG is vast, and any realisation
+simulating recombinant ancestry. The big ARG is vast, and any realisation
 for even moderate levels of recombination is far too large to be of practical
-use. The number of events in the ARG back to the GMRCA
+use. The number of events in the big ARG, all the way back to the GMRCA
 is $O(e^\rho)$~\citep{griffiths1997ancestral}, whereas the number
-of events required to simulate the complete ancestry of a sample
-in Hudson's algorithm is
+of events required to simulate the little ARG is
 $O(\rho^2)$~\citep{hein2004gene,baumdicker2021efficient}.
 This disparity in the number of events in the two formulations is
-because the majority of the events that occur in the ARG do
+because the majority of the events that occur in the big ARG do
 not affect the genetic ancestry of the sample in any way. Recombination
@@
 
-This original formulation which does not keep track of the distribution
-of ancestral material among lineages is the so-called ``Big'' ARG.
-The ``little'' ARG is the subset of the Big ARG traversed by
-Hudson's algorithm (\citet{wiuf1999recombination} refer to this as HUD).
 [Review of other work done on the ARG stochastic process here.
@@
 The Ancestral Selection Graph (ASG)~\citep{krone1997ancestral,neuhauser1997genealogy}
-uses dynamics identical to the ``big" ARG to simulate an ensemble of correlated
+uses dynamics identical to the ``big'' ARG to simulate an ensemble of correlated
 potential ancestral trees. Weak genic selection is incorporated by sampling a
 true ancestry from the ensemble in a non-uniform way.
-There is no ASG analogue of the more computationally tractable ``little" ARG,
+There is no ASG analogue of the more computationally tractable ``little'' ARG,
 though some gains in tractability can be made by considering typed
@@
 % Saying "awkward" too many times here, rephrase.
-to describe can be awkward in practise. For example,
+to describe can be awkward in practice. For example,
 simulating an ARG in this representation is considerably
@@
 Such awkwardnesses can be worked around, of course,
-but in practise methods such has
+but in practice methods such has
 \texttt{beagle}~\citep{lyngso2005minimum} and
@@
 
-The limitations of the Griffiths ARG encoding are a direct consequence
-of its derivation from the coalescent with recombination stochastic process.
-Indeed, these limitations arise because it is such a \emph{good}
-mathematical encoding, providing sufficient information to
-define the realised ancestry under the assumptions of the model,
-and no more.
-
-We describe here an alternative formulation in which the
-concepts are based on the diploid pedigree, which
-is free from the limitations discussed in the previous section.
-\citet{mathieson2020ancestry} and others
-\citep{wakeley2012genetics,gusfield2014recombinatorics,speed2015naturereviewsgenetics}
-have also suggested this approach, although without explicitly defining
-the roles of nodes and the details of how ancestry
-flows along edges.
+The deep connection between the ARG and the pedigree is of course
+well known
+\citep[e.g.][]{wakeley2012genetics,gusfield2014recombinatorics,
+speed2015naturereviewsgenetics}
+and recent discussions of the ARG have illustrated the
+relationship~\citep{mathieson2020ancestry,brandt2021evaluation}, % more??
+showing how the coalescence and recombination events in an ARG
+correspond to particular genomes in the pedigree.
+As discussed in the previous section, the definition of the ARG
+as a \emph{data structure} in terms of these two events is problematic,
+fundamentally limiting the expressivity of the structure.
+Following \citep{mathieson2020ancestry} we instead define the
+ARG data structure \emph{in terms of} an underlying pedigree, so that
+any form of inheritance structure can be modelled.
+% Bit of duplication here, but good to make the point about being
+% motivated by real existing datasets not some academic desire to
+% model complexity that isn't there.
+This, coupled with
+the explicit association of genome coordinates with edges, allows
+us to represent even the bewilderingly complex patterns of relatedness
+present in modern datasets.
+
 \begin{figure}
-  \begin{subfigure}[b]{0.25\textwidth}
-    \centering
-    \includegraphics[height=0.3\textheight]{illustrations/pedigree}
-    \caption{Pedigree with genetic inheritance}
-    \label{fig:pedigree}
-  \end{subfigure}
-  \hfill
-  \begin{subfigure}[b]{0.25\textwidth}
-    \centering
-    \includegraphics[height=0.3\textheight]{illustrations/pedigree_ARG}
-    \caption{Graph representation of genetic inheritance}
-    \label{fig:ARG-of-pedigree}
-  \end{subfigure}
-  \hfill
-  \begin{subfigure}[b]{0.4\textwidth}
-    \centering
-    \includegraphics[height=0.3\textheight]{illustrations/pedigree_trees}
-    \caption{Tree representation of genetic inheritance}
-    \label{fig:ts-of-pedigree}
-  \end{subfigure}
+    \includegraphics[width=\textwidth]{illustrations/pedigree_figure}
 \caption{\label{fig-pedigree-and-arg}
@@
 \end{figure}
+
+Fig~\ref{fig-pedigree-and-arg} illustrates the basic elements of the model.
+The pedigree is composed of a set of individuals who each carry
+two \emph{genomes} (we assume a diploidy here for concreteness, but there
+is no inherent limitation on ploidy in the model).
+Each genome is inherited from one of the individual's parents, and may
+be the recombined product of that parent's two genomes.
+For example, individual $I_1$ has two genomes $A$ and $B$,
+inherited from parents $I_3$ and $I_4$. Genome $A$ is the product of
+recombining $I_3$'s two genomes $E$ and $F$ at position FIXME,
+whereas $B$ was inheritied directly from $I_4$'s genome $G$ without
+recombination.
 
@@
 represent evolutionary relationships (either abstractly or explicitly) between a set of taxa
-that labels some of its nodes" (although more narrow and context-dependent definitions can also
+that labels some of its nodes'' (although more narrow and context-dependent definitions can also
 be found in the literature). Explicit rooted phylogenetic networks are rooted DAGs,
@@
 or MCMC \citep{kuhner2000maximum, nielsen2000estimation, wang2008bayesian, fallon2013acg}.
-These methods invariably operated on a representation of the ``little ARG", typically
+These methods invariably operated on a representation of the ``little ARG'', typically
 generated until the grand MRCA. They are extremely computationally expensive,
@@
 the feasible range of sequence lengths by at least an order of magnitude when
-compared with methods which sample realisations of the exact ``little ARG".
+compared with methods which sample realisations of the exact ``little ARG''.
 

### 2022-02-13 paper.tex
First few paragraphs of "our model" section

@@
 fundamentally limiting the expressivity of the structure.
-Following \citep{mathieson2020ancestry} we instead define the
+Following \cite{mathieson2020ancestry} we instead define the
 ARG data structure \emph{in terms of} an underlying pedigree, so that
@@
 \caption{\label{fig-pedigree-and-arg}
-The (unannotated) ARG is closely related to the pedigree
-graph, where each ``individual'' node is replaced by two ``genome''
-nodes, and edges between parents and children are only included if there is
-transfer of ancestral material. We usually only include nodes in an ARG
-that affect our sample, so that, for example, individuals that are ancestral
-to only one sampled genome are not included.}
+% TODO caption needs another pass. One the full section has been written
+% it'll be clearer what we need to say in the caption vs the text.
+The Ancestral Recombination Graph defined as arbitrary inheritance
+of genomes between diploid (or any ploidy; see text) individuals.
+(A) Each individual consists of two genomes, and each genome is
+inherited as a recombinant mixture of the individual's parent's
+genomes. (B) Genomes are modelled as nodes in a graph, and the edges
+denote inheritance. Each edge is annotated with genome coordinates
+defining the specific region(s) inherited. An ARG is a retrospecitive
+structure defined with
+respect to some set of sample genomes (A--D here), and so only
+genomes directly ancestral to those samples are included.
+(C) This description of inheritance fully defines the
+local relationships along the genome, as shown in the local
+or ``marginal'' trees.
+}
 \end{figure}
@@
 The pedigree is composed of a set of individuals who each carry
-two \emph{genomes} (we assume a diploidy here for concreteness, but there
+two \emph{genomes} (we assume diploidy here for concreteness, but there
 is no inherent limitation on ploidy in the model).
@@
 recombining $I_3$'s two genomes $E$ and $F$ at position FIXME,
-whereas $B$ was inheritied directly from $I_4$'s genome $G$ without
+whereas $B$ was inherited directly from $I_4$'s genome $G$ without
 recombination.
-
-[Here are the basic components of the model]
-
-\begin{enumerate}
-\item A node corresponds to one of a diploid individual's two genomes.
-This is linked explicitly to the pedigree, and is fully general.
-Nodes \emph{may} be labelled with certain flags describing their properties,
-but it's not necessary.
-\item An edge between two nodes defines a genetic inheritance relationship
-between two nodes. Edges are annotated with information about the
-passage of ancestral materal. There are different forms these annotations
-can take. For example, we can follow the approach taken in the Griffiths
-ARG and just annotate recombinant edges with the interval $[0, x)$ or $[x, L)$,
-depending on side of the breakpoint the edge is on. In the most
-general case, we annotate each edge with the full set of disjoint
-intervals of ancestral material that flows along that edge.
-\citep{mcgill2013graphml} also proposed full ancestral material being
-associated with each edge.
-\end{enumerate}
-
-\begin{figure}
-\vspace{5em}
-% NB: not clear if this figure is better off on its own or should be
-% merged with the pedigree one.
-\caption{\label{fig-arg-annotations}
-The specific path through the ARG for a given genome position (and therefore
-the local genealogical tree) cannot be determined without annotations
-to the graph. (A) The classical Griffiths approach annotates recombination
-nodes with the corresponding breakpoint. (B) We can equivalently
-annotate the \emph{edges} with the genomic intervals carrying ancestral
-material. Edge annotations are somewhat more general, and have
-significant computational advantages.}
-\end{figure}
 
@@
 
-\section*{Relationship with other types of phylogenetic networks}
+\textcolor{red}{TODO: reduce this down to one paragraph, which
+defines the relationship between what we define as an ARG and
+what people mean by phylogenetic networks from the literature.
+We probably don't need to get into quite so much detail on
+the different types that people have studies. Essentially,
+we're ending the section with a brief review which situates
+our definition in the literature.}
 

### 2022-02-14 illustrations/pedigree.py
Use SVG to position figs 3a,b,c

@@ -1,3 +1,4 @@
+import io
 import json
 from pathlib import Path
 import string
@@ -13,70 +14,97 @@ import tskit
 sys.path.append(str(Path(__file__).parent.parent.absolute()))
 import argutils
 
-n = SimpleNamespace() # convenience labels
-l=100
-tables = tskit.TableCollection(sequence_length=l)
-tables.nodes.metadata_schema = tskit.MetadataSchema.permissive_json()
-for gen in range(4):
-    for individual in range(2):
-        i = tables.individuals.add_row()
-        for genome in range(2):
-            label = string.ascii_uppercase[tables.nodes.num_rows]
-            metadata = {
-                'gender': "male" if individual==0 else "female",
-                'name': label,
-                'genome': 'paternal' if genome==0 else "maternal",
-            }
-            setattr(n, label, tables.nodes.num_rows)
-            tables.nodes.add_row(
-                flags=tskit.NODE_IS_SAMPLE if gen==0 else 0,
-                time=gen,
-                metadata=metadata,
-                individual=i,
-            )
+current_dir = Path(__file__).parent
 
-bp = [25, 60]
-tables.edges.clear()
-tables.individuals[0] = tables.individuals[0].replace(parents=[2,3])
-tables.edges.add_row(child=n.A, parent=n.E, left=0, right=bp[0])
-tables.edges.add_row(child=n.A, parent=n.F, left=bp[0], right=l)
-tables.edges.add_row(child=n.B, parent=n.G, left=0, right=l)
+def plot_pedigree_figure(pedigree_svg):
+    n = SimpleNamespace() # convenience labels
+    l=100
+    tables = tskit.TableCollection(sequence_length=l)
+    tables.nodes.metadata_schema = tskit.MetadataSchema.permissive_json()
+    for gen in range(4):
+        for individual in range(2):
+            i = tables.individuals.add_row()
+            for genome in range(2):
+                label = string.ascii_uppercase[tables.nodes.num_rows]
+                metadata = {
+                    'gender': "male" if individual==0 else "female",
+                    'name': label,
+                    'genome': 'paternal' if genome==0 else "maternal",
[…]

### 2022-02-14 paper.tex
Trivial grammar corrections
More text

@@
 
-The limitations of the Griffiths ARG encoding are a direct consequence
-of its derivation from the coalescent with recombination stochastic process.
-Indeed, these limitations arise because it is such a \emph{good}
-mathematical encoding, providing sufficient information to
-define the realised ancestry under the assumptions of the model,
-and no more.
-
-We describe here an alternative formulation in which the
-concepts are based on the diploid pedigree, which
-is free from the limitations discussed in the previous section.
-\citet{mathieson2020ancestry} and others
-\citep{wakeley2012genetics,gusfield2014recombinatorics,speed2015naturereviewsgenetics}
-have also suggested this approach, although without explicitly defining
-the roles of nodes and the details of how ancestry
-flows along edges.
+The deep connection between the ARG and the pedigree is of course
+well known
+\citep[e.g.][]{wakeley2012genetics,gusfield2014recombinatorics,
+speed2015naturereviewsgenetics}
+and recent discussions of the ARG have illustrated the
+relationship~\citep{mathieson2020ancestry,brandt2021evaluation}, % more??
+showing how the coalescence and recombination events in an ARG
+correspond to particular genomes in the pedigree.
+As discussed in the previous section, the definition of the ARG
+as a \emph{data structure} in terms of these two events is problematic,
+fundamentally limiting the expressivity of the structure.
+Following \cite{mathieson2020ancestry} we instead define the
+ARG data structure \emph{in terms of} an underlying pedigree, so that
+any form of inheritance structure can be modelled.
+% Bit of duplication here, but good to make the point about being
+% motivated by real existing datasets not some academic desire to
+% model complexity that isn't there.
+This, coupled with
+the explicit association of genome coordinates with edges, allows
+us to represent even the bewilderingly complex patterns of relatedness
+present in modern datasets.
+
 \begin{figure}
-  \begin{subfigure}[b]{0.25\textwidth}
+    \includegraphics[width=\textwidth]{illustrations/pedigree_figure}
+\caption{\label{fig-pedigree-and-arg}
+% TODO caption needs another pass. One the full section has been written
+% it'll be clearer what we need to say in the caption vs the text.
+The Ancestral Recombination Graph defined as arbitrary inheritance
+of genomes between diploid (or any ploidy; see text) individuals.
+(A) Each individual consists of two genomes, and each genome is
+inherited as a (potentially) recombinant mixture of the individual's parent's
+genomes. (B) Genomes are modelled as nodes in a graph, and the edges
+denote inheritance. Each edge is annotated with genome coordinates
+defining the specific region(s) inherited. An ARG is a retrospective
+structure defined with
+respect to some set of sample genomes (A--D here), and so only
+genomes directly ancestral to those samples are included.
+(C) This description of inheritance fully defines the
+local relationships along the genome, as shown in the local
+or ``marginal'' trees.
+}
+\end{figure}
+
+Fig~\ref{fig-pedigree-and-arg} illustrates the basic elements of the model.
+The pedigree is composed of a set of individuals who each carry
+two \emph{genomes} (we assume diploidy here for concreteness, but there
+is no inherent limitation on ploidy).
+Each genome is inherited from one of the individual's parents, and may
+be the recombined product of that parent's two genomes.
+For example, individual $I_1$ has two genomes $A$ and $B$,
+inherited from parents $I_3$ and $I_4$. Genome $A$ is the product of
+recombining $I_3$'s two genomes $E$ and $F$ at position FIXME,
+whereas $B$ was inherited directly from $I_4$'s genome $G$ without
+recombination.
+
+The focus on modelling genomes rather than events has some interesting
+and important consequences. Firstly, the graph may contain nodes that
+do not correspond to any particular event in the ancestry of the
+samples. For example, node $H$ is the direct ancestor of $D$ in
+Fig~\ref{fig-pedigree-and-arg}, and is not the product of either a
+coalescence or a recombination. Such nodes would usually be removed
+from the graph so that $D$ descends directly from $K$ (see section XXX)
+but there is no necessity for this from a representational perspective.
+This ability can be useful in practice and could be required if,
+% Do we need to spell this out?
+for example, we have a deeply sequenced pedigree of
+haplodiploids. % e.g. Honeybees? Be great to cite an example!
+
+Secondly, the way that we model recombination is subtly different.
+Rather than having a single node that represents the \emph{event},
+there are three nodes (representing the recombinant offspring and
+two parental genomes) and two edges with associated genome coordinate
+intervals (representing the precise inheritance patterns).
+For example, in Fig~\ref{fig-pedigree-and-arg}, node $A$ is the
+recombinant descendent of $E$ and $F$, inheriting positions
+FIXME from $E$ and genome positions FIXME from $F$. Any particular
+node can be both a recombinant (it is the recombined product of
+its parent's genomes) and a common ancestor (the ancestry from
+multiple samples meet at this node; see the next section).
+[More stuff here: clarify why we assign coordinates to edges
+rather than nodes. Basically lay out all the stuff that we can
+talk about \emph{without} going in to the details of
+ancestry resolution.]
+
+\textcolor{red}{TODO: reduce this down to one paragraph, which
+defines the relationship between what we define as an ARG and
+what people mean by phylogenetic networks from the literature.
+We probably don't need to get into quite so much detail on
+the different types that people have studies. Essentially,
+we're ending the section with a brief review which situates
+our definition in the literature. I think pulling apart the
+idea of a ``reticulation'' vs what we call a recombinant
+is essential here.}
+
+The term \emph{phylogenetic network} has been broadly accepted to describe the most general
+type of genealogical graph, defined by \citet{huson2010phylogenetic} as ``any graph used to
+represent evolutionary relationships (either abstractly or explicitly) between a set of taxa
+that labels some of its nodes'' (although more narrow and context-dependent definitions can also
+be found in the literature). Explicit rooted phylogenetic networks are rooted DAGs,
+with leaves labelled by the taxa (species, groups, individuals or sequences), and nodes classed
+as either reticulation nodes (with in-degree $\geq 2$) or tree nodes.
+
+Some sub-classes of explicit rooted phylogenetic networks are defined based on the evolutionary event that the
+reticulation nodes capture \citep{huson2010phylogenetic}. In \emph{reassortment} networks, used to
+model the evolution of viruses, reticulation nodes correspond to exchange of
+(non-recombining) genomic segments between viral particles inside a co-infected host.
+In \emph{hybridisation} networks, reticulation nodes correspond to genetic material from two
+different species combining to create a hybrid. In \emph{recombination} networks, the reticulation nodes
+correspond to recombination events (with leaves labelled by sequences and edges labelled by mutations).
+The ARG is a subtype of a recombination network under this definition (restricting to one crossover
+recombination per reticulation node).
+% AI: Not sure about this - do ARGs also encompass multiple crossover recombinations and gene
+% conversion (seems implied above)?
+
+Sub-classes of explicit rooted phylogenetic networks can also be defined by restrictions on their topology.
+In \emph{binary networks}, all internal nodes must have either in-degree two and out-degree one or in-degree
+one and out-degree two (with the root having out-degree two and the leaves in-degree 1) \citep{steel2016phylogeny}.
+In \emph{tree-child networks}, every internal node must have at least one child that is a tree node;
+\emph{tree-sibling networks} are constrained by every reticulation node having at least one sibling that is a
+tree node \citep{cardona2008extended}. Further, sub-types can be defined by restrictions on
+the topology of subgraphs that contain the reticulation loops (cycles containing a reticulation node).
+A reticulation loop that does not share any node with another reticulation loop is termed a gall
+\citep[][p.\ 237]{gusfield2014recombinatorics}; a reticulation network containing only galls is
+called a \emph{galled tree}. Reticulation loops that are joined together in a reticulation network
+by at least one shared node form a blob; a \emph{level-$k$ network} is one where the maximal number
+of reticulation loops inside a blob is $k$ \citep{choy2005computing}. Galled trees are level-1 networks,
+and the tractability afforded by their relatively simple structure results in a number of attractive
+combinatorial properties \citep{wang2001perfect, gusfield2004optimal}. We consider the general case of
+an ARG, without applying any of these restrictions.
+
+% AI: have not mentioned any implicit or un-rooted networks (split, median, haplotype, reticulograms, etc).
+% Also: are we not defining ARGs as "binary networks"? Might be good to specify this here.
+
+\section*{Ancestry resolution and simplification}
+
+\begin{figure}
 \centering
-    \includegraphics[height=0.3\textheight]{illustrations/pedigree}
-    \caption{Pedigree with genetic inheritance}
-    \label{fig:pedigree}
-  \end{subfigure}
-  \hfill
-  \begin{subfigure}[b]{0.25\textwidth}
-    \centering
-    \includegraphics[height=0.3\textheight]{illustrations/pedigree_ARG}
-    \caption{Graph representation of genetic inheritance}
-    \label{fig:ARG-of-pedigree}
-  \end{subfigure}
-  \hfill
-  \begin{subfigure}[b]{0.4\textwidth}
-    \centering
-    \includegraphics[height=0.3\textheight]{illustrations/pedigree_trees}
-    \caption{Tree representation of genetic inheritance}
-    \label{fig:ts-of-pedigree}
-  \end{subfigure}
-\caption{\label{fig-pedigree-and-arg}
-The (unannotated) ARG is closely related to the pedigree
-graph, where each ``individual'' node is replaced by two ``genome''
-nodes, and edges between parents and children are only included if there is
-transfer of ancestral material. We usually only include nodes in an ARG
-that affect our sample, so that, for example, individuals that are ancestral
-to only one sampled genome are not included.}
-\end{figure}
-
-[Here are the basic components of the model]
-
-\begin{enumerate}
-\item A node corresponds to one of a diploid individual's two genomes.
-This is linked explicitly to the pedigree, and is fully general.
-Nodes \emph{may} be labelled with certain flags describing their properties,
-but it's not necessary.
-\item An edge between two nodes defines a genetic inheritance relationship
-between two nodes. Edges are annotated with information about the
-passage of ancestral materal. There are different forms these annotations
-can take. For example, we can follow the approach taken in the Griffiths
-ARG and just annotate recombinant edges with the interval $[0, x)$ or $[x, L)$,
-depending on side of the breakpoint the edge is on. In the most
-general case, we annotate each edge with the full set of disjoint
-intervals of ancestral material that flows along that edge.
-\citep{mcgill2013graphml} also proposed full ancestral material being
-associated with each edge.
-\end{enumerate}
-
-\begin{figure}
 \vspace{5em}
-% NB: not clear if this figure is better off on its own or should be
-% merged with the pedigree one.
-\caption{\label{fig-arg-annotations}
-The specific path through the ARG for a given genome position (and therefore
-the local genealogical tree) cannot be determined without annotations
-to the graph. (A) The classical Griffiths approach annotates recombination
-nodes with the corresponding breakpoint. (B) We can equivalently
-annotate the \emph{edges} with the genomic intervals carrying ancestral
-material. Edge annotations are somewhat more general, and have
-significant computational advantages.}
+\caption{\label{fig-ancestry-resolution}
+}
 \end{figure}
@@
 
-\section*{Relationship with other types of phylogenetic networks}
-
-The term \emph{phylogenetic network} has been broadly accepted to describe the most general
-type of genealogical graph, defined by \citet{huson2010phylogenetic} as ``any graph used to
-represent evolutionary relationships (either abstractly or explicitly) between a set of taxa
-that labels some of its nodes'' (although more narrow and context-dependent definitions can also
-be found in the literature). Explicit rooted phylogenetic networks are rooted DAGs,
-with leaves labelled by the taxa (species, groups, individuals or sequences), and nodes classed
-as either reticulation nodes (with in-degree $\geq 2$) or tree nodes.
-
-Some sub-classes of explicit rooted phylogenetic networks are defined based on the evolutionary event that the
-reticulation nodes capture \citep{huson2010phylogenetic}. In \emph{reassortment} networks, used to
-model the evolution of viruses, reticulation nodes correspond to exchange of
-(non-recombining) genomic segments between viral particles inside a co-infected host.
-In \emph{hybridisation} networks, reticulation nodes correspond to genetic material from two
-different species combining to create a hybrid. In \emph{recombination} networks, the reticulation nodes
-correspond to recombination events (with leaves labelled by sequences and edges labelled by mutations).
-The ARG is a subtype of a recombination network under this definition (restricting to one crossover
-recombination per reticulation node).
-% AI: Not sure about this - do ARGs also encompass multiple crossover recombinations and gene
-% conversion (seems implied above)?
-
-Sub-classes of explicit rooted phylogenetic networks can also be defined by restrictions on their topology.
-In \emph{binary networks}, all internal nodes must have either in-degree two and out-degree one or in-degree
-one and out-degree two (with the root having out-degree two and the leaves in-degree 1) \citep{steel2016phylogeny}.
-In \emph{tree-child networks}, every internal node must have at least one child that is a tree node;
-\emph{tree-sibling networks} are constrained by every reticulation node having at least one sibling that is a
-tree node \citep{cardona2008extended}. Further, sub-types can be defined by restrictions on
-the topology of subgraphs that contain the reticulation loops (cycles containing a reticulation node).
-A reticulation loop that does not share any node with another reticulation loop is termed a gall
-\citep[][p.\ 237]{gusfield2014recombinatorics}; a reticulation network containing only galls is
-called a \emph{galled tree}. Reticulation loops that are joined together in a reticulation network
-by at least one shared node form a blob; a \emph{level-$k$ network} is one where the maximal number
-of reticulation loops inside a blob is $k$ \citep{choy2005computing}. Galled trees are level-1 networks,
-and the tractability afforded by their relatively simple structure results in a number of attractive
-combinatorial properties \citep{wang2001perfect, gusfield2004optimal}. We consider the general case of
-an ARG, without applying any of these restrictions.
-% AI: have not mentioned any implicit or un-rooted networks (split, median, haplotype, reticulograms, etc).
-% Also: are we not defining ARGs as "binary networks"? Might be good to specify this here.
-
 \section*{ARG inference methods}
@@
 \begin{itemize}
+\item It is unfortunate that we are adding yet another meaning to the
+term ARG here, but all we're really doing is making the implicit
+assumptions that others have made (i.e., describing the output of
+Relate, tsinfer etc as ARGs) and making them concrete. We need to have a
+term to describe all of these things, and since people are already using
+ARG for this, we might as well make it rigorous. We can then move away
+from the unfruitful discussion about which methods infer a ``true'' or
+``real'' ARG (none do, really, under the original definitions) and
+instead start discussing the \emph{properties} of these ARGs. (See
+next point also.)
 \item Terminology based around an inferred ancestry being a ``true''
@@
 
+\item There are real computational disadvantages to the Griffiths ARG
+data structure. Having two types of node (three really, as we need
+``sample'' nodes which are also treated differently) complicates algorithms,
+and makes it difficult to reason about the structure of the ARG as a
+graph. For example, there's some weirdness about mutations over
+recombination nodes discussed by Gusfield [Ana?]
+
 \item Comparison metrics are a problem. All evaluations has essentially

### 2022-02-14 illustrations/pedigree.py
Allow X axis to be reversed

@@ -71,7 +71,8 @@ def plot_pedigree_figure(pedigree_svg):
     ts_simp = ts.simplify(keep_unary=True)
     
     fig, ax = plt.subplots(1, 1, figsize=(2.5, 5))
-    pos = argutils.viz.draw(ts_simp, ax, node_color=mpl.colors.to_hex(plt.cm.tab20(1)))
+    pos = argutils.viz.draw(
+        ts_simp, ax, reverse_x_axis=True, node_color=mpl.colors.to_hex(plt.cm.tab20(1)))
     plt.subplots_adjust(left=0, right=1, top=1, bottom=0)
     with io.StringIO() as f:
         plt.savefig(f, format='svg')

### 2022-02-14 paper.tex
First stab at ancestry resolution text

@@
 \section*{Ancestry resolution and simplification}
+% Note: we want some terminology here to be precise about what we're
+% talking about - perhaps a G-ARG for genome-ARG vs E-ARG for event?
+% Just using ARG for now, can clean up in another pass.
+The definition of an ARG as a data structure in terms of
+individuals' monoploid genomes
+and their direct patterns of inheritance allows us to describe arbitarily
+complex relatedness structures, free from the limitations imposed by
+coalescent approximations.
+An ARG is an intrinsically retrospective
+structure, defined with respect to the ancestry of
+a set of sampled genomes.
+The graph structure (the nodes and their connectivity through edges)
+captures a great deal of information about this ancestry,
+but on its own it cannot tell us about the precise relationships between
+the samples along the genome: do to this, we must construct the
+local (or ``marginal'') trees.
 
@@
 \centering
-\vspace{5em}
+\includegraphics[width=\textwidth]{illustrations/ARG_edge_annotations}
 \caption{\label{fig-ancestry-resolution}
+Implicit vs explicit edge annotations. (a) An eARG equivalent to the encoding
+described by Griffiths can be created by setting all edges to span the full
+genome, apart from edges between two recombining nodes and their recombinant
+child (edge spans in bold). (b) The sample-resolved eARG, where the Hudson
+simplification algorithm has shortened edge spans so that they encapsulate only
+the upwards transmission of genetic material from the samples A, B, and C;
+plotted line widths are proportional to the genomic span of each edge. The
+example ARG used here is taken from \citet[][fig. 1]{wiuf1999recombination}.
 }
 \end{figure}
+
+% This is weak. Should we give a better description of the algorithm?
+% Not clear yet, and don't want to spend too much time on it if it's
+% unhelpful.
+Given a set of edges $(u, v, I)$ where node $u$ descends from $v$
+over the set of genomic intervals $I$, and a set of sample nodes $S$
+it is straightforward to construct the local tree at position $x$.
+Starting from the sample nodes in $S$, traverse upwards through the
+graph, such that when we are at a node $u$ we follow the edge
+$(u, v, I)$ where $x \in I$. This is essentially the same procedure as
+described by Griffiths~\citep{griffiths1997ancestral}.
+
+Recovering the tree at a single point along the genome, although
+a fundamental operation, is not the only operation that we may wish
+to perform on an ARG. We can make the operation of efficiently
+constructing \emph{all} trees along the genome much more efficient,
+and gain insight into the structure of the ARG,
+if we perform an additional step and \emph{resolve} the ancestral
+material in the ARG with respect to the samples.
+
+As we saw in the section X, the key to efficiently accessing the
+results of the ARG stochastic process is to reason about
+the flow of ancestral material through the graph using Hudson's
+algorithm. The same basic process allows us to determine the
+exact intervals of ancestral material that are flowing through
+each edge in the graph. Concretely, given a set of edges
+$E = \{(u, v, I)\}$ where $u$ is the child, $v$ is the parent
+and $I$ is the set of genomic intervals passed directly from
+$v$ to $u$, the resolution process with respect to a set of
+samples $S$ produces a set of edges $E' = \{(u, v, I)\}$
+where $I$ is the set of genomic intervals over which there is a
+% weak and not correct - how do we say this properly?
+path from at least one element of $S$ to $u$,
+and $u$ is not an MRCA.
+
+This process is illustrated in Fig~\ref{fig-ancestry-resolution}.
 

### 2022-02-15 paper.tex
Attempts at definitions.

@@
 chromosome, but it has significant drawbacks as the basis for a general purpose
-storage and interchange format.
-The concrete details of how an ARG should be stored and processed
-are not discussed in the literature, % weak, fix.
-but the following is a reasonable approach that mirrors the implementations
-in methods such as [x, y z].
-A Griffiths ARG is composed of a set of $N$ nodes, where each node is an
-integer $0 \leq u < N$, and edges $(u, v)$ define the relationships between
-those nodes (Fig.~\ref{fig-arg-data-structure}A).
-Each node corresponds to an \emph{event}, and is either a common ancestor
-event (in which case it has two children and one parent)
-or a recombination event (when it has one child and two parents).
-Each recombination event is associated with a breakpoint $0 < x < L$,
-which we might encode as a mapping $f: N \rightarrow \{0, ..., L - 1\}$
-such that $f(u)$ is the recombination breakpoint if $u$ is a recombination
-node and $f(u) = \infty$ otherwise. This encoding captures the information
-of a realisation of the ARG stochastic process, but it is important to
-note that it does not uniquely describe the underlying ancestral history
-without stipulating some additional ordering requirements.
+storage and interchange format. To avoid confusion, we will refer to this
+ARG encoding as an ``Event ARG'', or EARG.
+There are many ways in which the EARG structure can be formalised (see
+e.g.~\citep{gusfield2014recombinatorics}).
+% It would be slightly easier for us if we defined nodes as in N_0, since
+% we use 0 as a node ID, but leaving for now.
+We can define an EARG as a tuple $(e, \sigma)$, where $e$ is an ordered
+list of tuples such that $e_j = (c, p)$ defines an edge between
+child node $c$ and parent $p$ ($c, p \in \mathbb{N}$), and
+$\sigma: \mathbb{N} \rightarrow \mathbb{N}_0$ is a function
+mapping nodes to recombination breakpoints, such that $f(u) = x$
+if $u$ is a recombination event with breakpoint $x$ and
+% Because of our definitions we can't have anything to the left of 0,
+% so you can treat breakpoint at 0 as no recombination.
+$f(u) = 0$ otherwise. Note that this description captures only the
+graph topology: if we also wish to know the times of events we need
+an additional function $t(u): \mathbb{N} \rightarrow \mathbb{R}$
+which defines the time of each node.
+An example is given in Fig.~\ref{fig-arg-data-structure}A.
 
@@
 is limited in the ancestral patterns that can be described and ambiguous.
+[TODO clarify that the function is the same things as having two
+types of nodes]
 (B) Equating graph nodes with \emph{genomes} and attaching the
@@
 
-\section*{Anchoring the ARG to the pedigree}
-
-The deep connection between the ARG and the pedigree is of course
+\section*{Genome ARGs}
+
+The deep connection between ARGs and pedigrees is of course
 well known
@@
 speed2015naturereviewsgenetics}
-and recent discussions of the ARG have illustrated the
+and recent discussions have illustrated the
 relationship~\citep{mathieson2020ancestry,brandt2021evaluation}, % more??
@@
 
+Formally, we can define a GARG as a set of edges $E$, where each
+element of the set is a tuple $(c, p, I)$ such that $c, p \in \mathbb{N}$,
+$c$ is the child node, $p$ is the parent node and $I$ is the set of
+disjoint genomic intervals $(\ell, r])$ over which genome $c$ inherits from $p$ (we may
+also phrase this in terms of the ancestral material of a sample---see
+the next section).
+
 Fig~\ref{fig-pedigree-and-arg} illustrates the basic elements of the model.
@@
 \section*{Ancestry resolution and simplification}
-% Note: we want some terminology here to be precise about what we're
-% talking about - perhaps a G-ARG for genome-ARG vs E-ARG for event?
-% Just using ARG for now, can clean up in another pass.
+% This is a rough first pass. Will clarify terminology later.
 The definition of an ARG as a data structure in terms of
@@
 local (or ``marginal'') trees.
+
+Formally, we can define a sample-resolved GARG as a tuple $(S, E)$
+where $S \subset \mathbb{N}$ is a set of sample nodes and
+$E$ is a set of edges. Each element of $E$
+is a tuple $(c, p, I)$ such that $c, p \in \mathbb{N}$,
+$c$ is the child node, $p$ is the parent node and $I$ is the set of
+disjoint genomic intervals $(\ell, r])$
+over which genome $c$ inherits from $p$, and there is a path from
+$c$ to some non-empty subset of $S$ through the local trees.
+% Should probably define how we build the local trees somewhere.
 

### 2022-02-15 paper.tex
Add code to convert to EARG and produce trees.

@@
 We can define an EARG as a tuple $(e, \sigma)$, where $e$ is an ordered
-list of tuples such that $e_j = (c, p)$ defines an edge between
+list of tuples such that $e_j = (c, p)$ defines a directed edge between
 child node $c$ and parent $p$ ($c, p \in \mathbb{N}$), and
 $\sigma: \mathbb{N} \rightarrow \mathbb{N}_0$ is a function
-mapping nodes to recombination breakpoints, such that $f(u) = x$
+mapping nodes to recombination breakpoints, such that $\sigma(u) = x$
 if $u$ is a recombination event with breakpoint $x$ and
@@
 % so you can treat breakpoint at 0 as no recombination.
-$f(u) = 0$ otherwise. Note that this description captures only the
+$\sigma(u) = 0$ otherwise. For each recombination node $u$ we must
+have exactly two edges in $E$ such that $u$ is the child.
+Note that this description captures only the
 graph topology: if we also wish to know the times of events we need
@@
 \end{figure}
+
+Given a set of sample nodes $S$ we can recover the local tree at
+position $x$ by the following procedure. Begin by initialising
+a stack
 
@@
 
-Formally, we can define a sample-resolved GARG as a tuple $(S, E)$
+We can define a GARG as a set of directed edges $E$, where each
+element of the set is a tuple $(c, p, I)$ such that $c, p \in \mathbb{N}$,
+$c$ is the child node, $p$ is the parent node and $I$ is the set of
+disjoint genomic intervals $(\ell, r]$ over which genome $c$ inherits from $p$.
+
+We can define a sample-resolved GARG as a tuple $(S, E)$
 where $S \subset \mathbb{N}$ is a set of sample nodes and

### 2022-02-15 argutils/ancestry.py

@@ -575,3 +575,50 @@ def wh99_example():
 
     tables.sort()
     return tables.tree_sequence()
+
+
+def as_earg(ts):
+    """
+    Returns the specified unresolved ARG as an EARG (E, sigma).
+    """
+    sigma = np.full(ts.num_nodes, int(ts.sequence_length), dtype=int)
+    edges = iter(ts.edges())
+    edge = next(edges, None)
+    E = []
+    while edge is not None:
+        if edge.left == 0 and edge.right == ts.sequence_length:
+            E.append((edge.child, edge.parent))
+        else:
+            assert edge.left == 0
+            breakpoint = edge.right
+            child = edge.child
+            sigma[child] = breakpoint
+            E.append((edge.child, edge.parent))
+            edge = next(edges, None)
+            assert edge.left == breakpoint
+            assert edge.child == child
+            E.append((edge.child, edge.parent))
+        edge = next(edges, None)
+    # We could wrap sigma as a function to be literal about the definition
+    # but this is simpler for debugging.
+    return E, sigma
+
+
+def earg_get_tree(E, sigma, S, x):
+    """
+    Given a minimal EARG definition (E, sigma), return the tree for a the given
+    set of samples at the specified position as a dictionary parent->child.
+    """
+    N = set(S)
+    parent = {}
+    while len(N) > 0:
+        c = N.pop()
+        P = [e[1] for e in E if e[0] == c]
+        if len(P) > 0:
+            if x < sigma[c]:
+                p = P[0]
+            else:
+                p = P[1]
+            parent[c] = p
+            N.add(p)
+    return parent

### 2022-02-15 argutils/tests.py

@@ -85,6 +85,69 @@ class TestSimulate:
         ts2.tables.assert_equals(ts3.tables, ignore_provenance=True)
 
 
+def parent_dict(tree):
+    """
+    Returns the parent dict reachable from the samples for the specified
+    tskit tree. We can't use tree.parent_dict as it includes non-sample
+    leaves.
+    """
+    d = {}
+    # Could make this more efficient avoiding already written paths,
+    # but can't be bothered.
+    for u in tree.tree_sequence.samples():
+        v = tree.parent(u)
+        while v != tskit.NULL:
+            d[u] = v
+            u = v
+            v = tree.parent(u)
+    return d
+
+
+class TestEARG:
+    @pytest.mark.parametrize("seed", range(1, 3))
+    @pytest.mark.parametrize(
+        ["n", "L"],
+        [
+            [2, 2],
+            [8, 2],
+            [16, 2],
+            [2, 10],
+            [8, 10],
+            [16, 10],
+            [10, 50],
+        ],
+    )
+    def test_trees_equal_left(self, n, L, seed):
+        rho = 0.1
+        ts = argutils.sim_coalescent(n, L=L, rho=rho, seed=seed, resolved=False)
+        E, sigma = argutils.as_earg(ts)
+        for tree in ts.trees():
+            t = argutils.earg_get_tree(E, sigma, ts.samples(), tree.interval[0])
+            assert t == parent_dict(tree)
+
+    @pytest.mark.parametrize("seed", range(10, 12))
+    @pytest.mark.parametrize(
+        ["n", "L"],
+        [
+            [2, 2],
+            [8, 2],
+            [16, 2],
+            [2, 10],
+            [8, 10],
+            [16, 10],
+            [10, 50],
+        ],
+    )
+    def test_trees_equal_mid(self, n, L, seed):
+        rho = 0.1
+        ts = argutils.sim_coalescent(n, L=L, rho=rho, seed=seed, resolved=False)
+        E, sigma = argutils.as_earg(ts)
+        for tree in ts.trees():
+            x = tree.interval.left + tree.interval.span / 2
+            t = argutils.earg_get_tree(E, sigma, ts.samples(), x)
+            assert t == parent_dict(tree)
+
+
 def gmrca_example_resolved():
     # 3.00┊  5  ┊  5  ┊
     #     ┊  ┃  ┊ ┏┻┓ ┊

### 2022-02-15 paper.tex
Very simple GARG definitions

@@
 if $u$ is a recombination event with breakpoint $x$ and
-% Because of our definitions we can't have anything to the left of 0,
-% so you can treat breakpoint at 0 as no recombination.
-$\sigma(u) = 0$ otherwise. For each recombination node $u$ we must
+% This is arbitrary, but handy when you start defining the function
+% for recovering trees; see below
+$\sigma(u) = \infty$ otherwise. For each recombination node $u$ we must
 have exactly two edges in $E$ such that $u$ is the child.
@@
 
+% Not sure if we actually include this, but it's good to work it through
+% to make sure the notation works.
 Given a set of sample nodes $S$ we can recover the local tree at
-position $x$ by the following procedure. Begin by initialising
-a stack
+position $x$ by the following procedure. Let $N$ be a set of
+nodes we are currently constructing the tree at, and let
+$p: \mathbb{N} \rightarrow \mathbb{N}$ be the mapping from child
+node to parent. Firstly, initialise $N \leftarrow S$.
+Then, while $|N| > 0$, we add branches to the
+tree as follows. Let $u$ be an arbitrary element of $N$ (which
+we remove). Then, if $\sigma(u) < x$ we let $j$ be the
+smallest value such that $e_{j,1} = u$, and otherwise
+let $j$ be the largest value such that $e_{j,1} = u$.
+Then, set $v \leftarrow e_{j, 2}$,
+$p_u \leftarrow v$
+and $N \leftarrow N \cup \{v\}$.
 
@@
 $c$ is the child node, $p$ is the parent node and $I$ is the set of
-disjoint genomic intervals $(\ell, r])$ over which genome $c$ inherits from $p$ (we may
+disjoint genomic intervals $(\ell, r]$ over which genome $c$ inherits from $p$ (we may
 also phrase this in terms of the ancestral material of a sample---see
@@
 
-We can define a GARG as a set of directed edges $E$, where each
-element of the set is a tuple $(c, p, I)$ such that $c, p \in \mathbb{N}$,
-$c$ is the child node, $p$ is the parent node and $I$ is the set of
-disjoint genomic intervals $(\ell, r]$ over which genome $c$ inherits from $p$.
-
 We can define a sample-resolved GARG as a tuple $(S, E)$
@@
 $c$ is the child node, $p$ is the parent node and $I$ is the set of
-disjoint genomic intervals $(\ell, r])$
+disjoint genomic intervals $(\ell, r]$
 over which genome $c$ inherits from $p$, and there is a path from

### 2022-02-15 argutils/ancestry.py

@@ -8,6 +8,7 @@ import dataclasses
 from typing import List
 from typing import Any
 
+import intervaltree
 import numpy as np
 import tskit
 
@@ -577,6 +578,65 @@ def wh99_example():
     return tables.tree_sequence()
 
 
+class IntervalSet:
+
+    intervals: list = dataclasses.field(default_factory=list)
+
+    def contains(self, x):
+        for left, right in self.intervals:
+            if left <= x < right:
+                return True
+        return False
+
+    def union(self, other):
+        """
+        Returns a new IntervalSet with the union of intervals in this and
+        other.
+        """
+        # Dreadful implementation, but oh well.
+        intervals = self.intervals.copy()
+        for interval in other.intervals:
+            if interval not in intervals:
+                intervals.append(interval)
+        return IntervalSet(intervals)
+
+
+import intervaltree
+
+def as_garg(ts):
+    """
+    Returns the specified unresolved ARG as an GARG E
+    """
+    E = [(edge.child, edge.parent, intervaltree.IntervalTree.from_tuples([(edge.left, edge.right)]))
+        for edge in ts.edges()]
+    return E
+
+def as_resolved_garg(ts):
+    """
+    Returns the specified unresolved ARG as an GARG E with respect to
+    the tree sequences samples.
+    """
+    S = ts.samples()
+    E = as_garg(ts)
+    print(E)
+    I = collections.defaultdict(intervaltree.IntervalTree)
+    for u in S:
+        I[u] = intervaltree.IntervalTree.from_tuples([(0, ts.sequence_length)])
+
+    N = set(S)
+    while len(N) > 0:
+        c = N.pop()
+        for e in [e for e in E if e[0] == c]:
+            p = e[1]
+            print("parent = ", p)
+            print("Ip = ", I[p])
+            print("Ic = ", I[c])
+            print("e = ", e[2])
+            print("intersection = ", I[c].intersection(e[2]))
+            I[p] = I[p].union(I[c].intersection(e[2]))
+            N.add(p)
+    return [(c, p, I[c]) for c, p, _ in E]
+
 def as_earg(ts):
     """
     Returns the specified unresolved ARG as an EARG (E, sigma).
@@ -621,4 +681,22 @@ def earg_get_tree(E, sigma, S, x):
                 p = P[1]
             parent[c] = p
             N.add(p)
+    return parent
+
+
+def garg_get_tree(E, S, x):
+    """
[…]

### 2022-02-15 argutils/tests.py

@@ -148,6 +148,152 @@ class TestEARG:
             assert t == parent_dict(tree)
 
 
+class TestGARG:
+
+    @pytest.mark.parametrize("seed", range(1, 3))
+    @pytest.mark.parametrize(
+        ["n", "L"],
+        [
+            [2, 2],
+            [8, 2],
+            [16, 2],
+            [2, 10],
+            [8, 10],
+            [16, 10],
+            [10, 50],
+        ],
+    )
+    def test_trees_equal_left(self, n, L, seed):
+        rho = 0.1
+        ts = argutils.sim_coalescent(n, L=L, rho=rho, seed=seed, resolved=False)
+        E = argutils.as_garg(ts)
+        for tree in ts.trees():
+            t = argutils.garg_get_tree(E, ts.samples(), tree.interval[0])
+            assert t == parent_dict(tree)
+
+    @pytest.mark.parametrize("seed", range(5, 8))
+    @pytest.mark.parametrize(
+        ["n", "L"],
+        [
+            [2, 2],
+            [8, 2],
+            [16, 2],
+            [2, 10],
+            [8, 10],
+            [16, 10],
+            [10, 50],
+        ],
+    )
+    def test_trees_equal_mid(self, n, L, seed):
+        rho = 0.1
+        ts = argutils.sim_coalescent(n, L=L, rho=rho, seed=seed, resolved=False)
+        E = argutils.as_garg(ts)
+        for tree in ts.trees():
+            x = tree.interval.left + tree.interval.span / 2
+            t = argutils.garg_get_tree(E, ts.samples(), x)
+            assert t == parent_dict(tree)
+
+class TestResolvedGARG:
+    @pytest.mark.parametrize("seed", range(1, 3))
+    @pytest.mark.parametrize(
+        ["n", "L"],
+        [
+            [2, 2],
+            [8, 2],
+            [16, 2],
+            [2, 10],
+            [8, 10],
+            [16, 10],
+            [10, 50],
+        ],
+    )
+    def test_trees_equal_left(self, n, L, seed):
+        rho = 0.1
+        ts = argutils.sim_coalescent(n, L=L, rho=rho, seed=seed, resolved=False)
+        E = argutils.as_resolved_garg(ts)
+        print(E)
+        for tree in ts.trees():
+            t = argutils.garg_get_tree(E, ts.samples(), tree.interval[0])
+            assert t == parent_dict(tree)
+
[…]

### 2022-02-16 illustrations/pedigree.py
Return both the positions and the graph from draw
This will allow us to set bespoke edge labels easily

@@ -10,11 +10,13 @@ import matplotlib.pyplot as plt
 
 import tskit
 
-# useful modules in the top level dir
-sys.path.append(str(Path(__file__).parent.parent.absolute()))
+current_dir = Path(__file__).parent
+
+# useful modules in the top level dir: we should be able to import this if the script is
+# run from the top level dir, but for some reason it fails, so we just add the path here
+sys.path.append(str(current_dir.parent.absolute()))
 import argutils
 
-current_dir = Path(__file__).parent
 
 def plot_pedigree_figure(pedigree_svg):
     n = SimpleNamespace() # convenience labels
@@ -71,7 +73,7 @@ def plot_pedigree_figure(pedigree_svg):
     ts_simp = ts.simplify(keep_unary=True)
     
     fig, ax = plt.subplots(1, 1, figsize=(2.5, 5))
-    pos = argutils.viz.draw(
+    argutils.viz.draw(
         ts_simp, ax, reverse_x_axis=True, node_color=mpl.colors.to_hex(plt.cm.tab20(1)))
     plt.subplots_adjust(left=0, right=1, top=1, bottom=0)
     with io.StringIO() as f:

### 2022-02-16 argutils/viz.py
Allow edge width to be set even if draw_edge_widths=False

@@ -45,12 +45,12 @@ def draw(
     
     If draw_edge_widths is True, draw the widths of edges in the graphical
     representation of the tree sequence in proportion to their genomic span.
-    max_edge_width spcifies the maximum edge width to draw (default is 5).
-
+    max_edge_width specifies the maximum edge width to draw (default is 5).
+    If draw_edge_widths is False, all edges are drawn with a width of
+    max_edge_width.
+    
     If draw_edge_alpha is True, draw edges with an alpha value equal to their
-    genomic span / the total sequence length of the tree sequence. If
-    draw_edge_alpha is True and draw_edge_widths is False, all edges are drawn
-    with a width of max_edge_width.
+    genomic span / the total sequence length of the tree sequence.
 
     node_size, font_size, and node_color are all passed to nx.draw directly.
     In particular this means that node_color can either be a single
@@ -95,7 +95,7 @@ def draw(
         sequence_length = ts.get_sequence_length()
         edge_widths = (edge_widths / sequence_length) * max_edge_width
     else:
-        edge_widths = None
+        edge_widths = max_edge_width
     if draw_edge_alpha:
         edge_alpha = get_edge_alpha(G, ts)
         if not draw_edge_widths:

### 2022-02-16 paper.tex
Initial pass at fig 4

--- 2022-02-15 paper.tex
@@
 chromosome, but it has significant drawbacks as the basis for a general purpose
-storage and interchange format. To avoid confusion, we will refer to this
-ARG encoding as an ``Event ARG'', or EARG.
-There are many ways in which the EARG structure can be formalised (see
-e.g.~\citep{gusfield2014recombinatorics}).
-% It would be slightly easier for us if we defined nodes as in N_0, since
-% we use 0 as a node ID, but leaving for now.
-We can define an EARG as a tuple $(e, \sigma)$, where $e$ is an ordered
-list of tuples such that $e_j = (c, p)$ defines an edge between
-child node $c$ and parent $p$ ($c, p \in \mathbb{N}$), and
-$\sigma: \mathbb{N} \rightarrow \mathbb{N}_0$ is a function
-mapping nodes to recombination breakpoints, such that $f(u) = x$
-if $u$ is a recombination event with breakpoint $x$ and
-% Because of our definitions we can't have anything to the left of 0,
-% so you can treat breakpoint at 0 as no recombination.
-$f(u) = 0$ otherwise. Note that this description captures only the
-graph topology: if we also wish to know the times of events we need
-an additional function $t(u): \mathbb{N} \rightarrow \mathbb{R}$
-which defines the time of each node.
-An example is given in Fig.~\ref{fig-arg-data-structure}A.
+storage and interchange format.
+The concrete details of how an ARG should be stored and processed
+are not discussed in the literature, % weak, fix.
+but the following is a reasonable approach that mirrors the implementations
+in methods such as [x, y z].
+A Griffiths ARG is composed of a set of $N$ nodes, where each node is an
+integer $0 \leq u < N$, and edges $(u, v)$ define the relationships between
+those nodes (Fig.~\ref{fig-arg-data-structure}A).
+Each node corresponds to an \emph{event}, and is either a common ancestor
+event (in which case it has two children and one parent)
+or a recombination event (when it has one child and two parents).
+Each recombination event is associated with a breakpoint $0 < x < L$,
+which we might encode as a mapping $f: N \rightarrow \{0, ..., L - 1\}$
+such that $f(u)$ is the recombination breakpoint if $u$ is a recombination
+node and $f(u) = \infty$ otherwise. This encoding captures the information
+of a realisation of the ARG stochastic process, but it is important to
+note that it does not uniquely describe the underlying ancestral history
+without stipulating some additional ordering requirements.
 
@@
 is limited in the ancestral patterns that can be described and ambiguous.
-[TODO clarify that the function is the same things as having two
-types of nodes]
 (B) Equating graph nodes with \emph{genomes} and attaching the
@@
 
-\section*{Genome ARGs}
-
-The deep connection between ARGs and pedigrees is of course
+\section*{Anchoring the ARG to the pedigree}
+
+The deep connection between the ARG and the pedigree is of course
 well known
@@
 speed2015naturereviewsgenetics}
-and recent discussions have illustrated the
+and recent discussions of the ARG have illustrated the
 relationship~\citep{mathieson2020ancestry,brandt2021evaluation}, % more??
@@
 
-Formally, we can define a GARG as a set of edges $E$, where each
-element of the set is a tuple $(c, p, I)$ such that $c, p \in \mathbb{N}$,
-$c$ is the child node, $p$ is the parent node and $I$ is the set of
-disjoint genomic intervals $(\ell, r])$ over which genome $c$ inherits from $p$ (we may
-also phrase this in terms of the ancestral material of a sample---see
-the next section).
-
 Fig~\ref{fig-pedigree-and-arg} illustrates the basic elements of the model.
@@
 \section*{Ancestry resolution and simplification}
-% This is a rough first pass. Will clarify terminology later.
-The definition of an ARG as a data structure in terms of
-individuals' monoploid genomes
-and their direct patterns of inheritance allows us to describe arbitarily
-complex relatedness structures, free from the limitations imposed by
-coalescent approximations.
-An ARG is an intrinsically retrospective
-structure, defined with respect to the ancestry of
-a set of sampled genomes.
-The graph structure (the nodes and their connectivity through edges)
-captures a great deal of information about this ancestry,
-but on its own it cannot tell us about the precise relationships between
-the samples along the genome: do to this, we must construct the
-local (or ``marginal'') trees.
-
-Formally, we can define a sample-resolved GARG as a tuple $(S, E)$
-where $S \subset \mathbb{N}$ is a set of sample nodes and
-$E$ is a set of edges. Each element of $E$
-is a tuple $(c, p, I)$ such that $c, p \in \mathbb{N}$,
-$c$ is the child node, $p$ is the parent node and $I$ is the set of
-disjoint genomic intervals $(\ell, r])$
-over which genome $c$ inherits from $p$, and there is a path from
-$c$ to some non-empty subset of $S$ through the local trees.
-% Should probably define how we build the local trees somewhere.
 
@@
 \caption{\label{fig-ancestry-resolution}
-Implicit vs explicit edge annotations. (a) An eARG equivalent to the encoding
-described by Griffiths can be created by setting all edges to span the full
-genome, apart from edges between two recombining nodes and their recombinant
-child (edge spans in bold). (b) The sample-resolved eARG, where the Hudson
-simplification algorithm has shortened edge spans so that they encapsulate only
-the upwards transmission of genetic material from the samples A, B, and C;
-plotted line widths are proportional to the genomic span of each edge. The
-example ARG used here is taken from \citet[][fig. 1]{wiuf1999recombination}.
+Implicit vs explicit edge annotations. (a) An eARG equivalent to the encoding described by Griffiths
+can be created by setting all edges to span the full genome, apart from edges between two recombining
+nodes and their recombinant child (edge spans in bold). (b) The sample-resolved eARG, where the
+Hudson simplification algorithm has shortened edge spans so that they encapsulate only the upwards
+transmission of genetic material from the samples A, B, and C; plotted line widths are proportional to the
+genomic span of each edge. The example ARG used here is taken from \citet[][fig. 1]{wiuf1999recombination}.
 }
 \end{figure}
-
-% This is weak. Should we give a better description of the algorithm?
-% Not clear yet, and don't want to spend too much time on it if it's
-% unhelpful.
-Given a set of edges $(u, v, I)$ where node $u$ descends from $v$
-over the set of genomic intervals $I$, and a set of sample nodes $S$
-it is straightforward to construct the local tree at position $x$.
-Starting from the sample nodes in $S$, traverse upwards through the
-graph, such that when we are at a node $u$ we follow the edge
-$(u, v, I)$ where $x \in I$. This is essentially the same procedure as
-described by Griffiths~\citep{griffiths1997ancestral}.
-
-Recovering the tree at a single point along the genome, although
-a fundamental operation, is not the only operation that we may wish
-to perform on an ARG. We can make the operation of efficiently
-constructing \emph{all} trees along the genome much more efficient,
-and gain insight into the structure of the ARG,
-if we perform an additional step and \emph{resolve} the ancestral
-material in the ARG with respect to the samples.
-
-As we saw in the section X, the key to efficiently accessing the
-results of the ARG stochastic process is to reason about
-the flow of ancestral material through the graph using Hudson's
-algorithm. The same basic process allows us to determine the
-exact intervals of ancestral material that are flowing through
-each edge in the graph. Concretely, given a set of edges
-$E = \{(u, v, I)\}$ where $u$ is the child, $v$ is the parent
-and $I$ is the set of genomic intervals passed directly from
-$v$ to $u$, the resolution process with respect to a set of
-samples $S$ produces a set of edges $E' = \{(u, v, I)\}$
-where $I$ is the set of genomic intervals over which there is a
-% weak and not correct - how do we say this properly?
-path from at least one element of $S$ to $u$,
-and $u$ is not an MRCA.
-
-This process is illustrated in Fig~\ref{fig-ancestry-resolution}.
 

### 2022-02-16 argutils/viz.py
Allow node font colour to be set

@@ -23,6 +23,7 @@ def draw(
     node_size=None,
     font_size=None,
     node_color=None,
+    font_color=None,
     reverse_x_axis=None,
 ):
     """
@@ -52,8 +53,8 @@ def draw(
     If draw_edge_alpha is True, draw edges with an alpha value equal to their
     genomic span / the total sequence length of the tree sequence.
 
-    node_size, font_size, and node_color are all passed to nx.draw directly.
-    In particular this means that node_color can either be a single
+    node_size, font_size, font_color, and node_color are all passed to nx.draw
+    directly. In particular this means that node_color can either be a single
     colour for all nodes (e.g. `mpl.colors.to_hex(mpl.pyplot.cm.tab20(1))`)
     or a dict of node_id -> colours.
     
@@ -107,6 +108,7 @@ def draw(
         G,
         pos,
         node_color=node_color,
+        font_color=font_color,
         node_shape="o",
         node_size=node_size,
         font_size=font_size,

### 2022-02-17 argutils/ancestry.py
Working version of GARG resolution

@@ -8,9 +8,10 @@ import dataclasses
 from typing import List
 from typing import Any
 
-import intervaltree
 import numpy as np
 import tskit
+
+from . import viz
 
 NODE_IS_RECOMB = 1 << 1
 
@@ -579,63 +580,103 @@ def wh99_example():
 
 
 class IntervalSet:
+    """
+    Naive and simple implementation of discrete intervals.
+    """
 
-    intervals: list = dataclasses.field(default_factory=list)
+    def __init__(self, L, tuples=None):
+        assert int(L) == L
+        self.I = np.zeros(int(L), dtype=int)
+        if tuples is not None:
+            for left, right in tuples:
+                self.insert(left, right)
+
+    def __str__(self):
+        return str(self.I)
+
+    def __repr__(self):
+        return repr(list(self.I))
+
+    def __eq__(self, other):
+        return np.array_equal(self.I == 0, other.I == 0)
+
+    def insert(self, left, right):
+        assert int(left) == left
+        assert int(right) == right
+        self.I[int(left) : int(right)] = 1
 
     def contains(self, x):
-        for left, right in self.intervals:
-            if left <= x < right:
-                return True
-        return False
+        assert int(x) == x
+        return self.I[int(x)] != 0
 
     def union(self, other):
         """
         Returns a new IntervalSet with the union of intervals in this and
         other.
         """
-        # Dreadful implementation, but oh well.
-        intervals = self.intervals.copy()
-        for interval in other.intervals:
-            if interval not in intervals:
-                intervals.append(interval)
-        return IntervalSet(intervals)
+        new = IntervalSet(self.I.shape[0])
+        assert other.I.shape == self.I.shape
+        new.I[:] = np.logical_or(self.I, other.I)
+        return new
 
+    def intersection(self, other):
+        """
+        Returns a new IntervalSet with the intersection of intervals in this and
+        other.
+        """
+        new = IntervalSet(self.I.shape[0])
+        assert other.I.shape == self.I.shape
+        new.I[:] = np.logical_and(self.I, other.I)
+        return new
 
-import intervaltree
+    def is_subset(self, other):
+        """
+        Return True if this set is a subset of other.
+        """
[…]

### 2022-02-17 argutils/tests.py

@@ -80,8 +80,6 @@ class TestSimulate:
         ts1 = argutils.sim_coalescent(n, L=L, rho=rho, seed=seed, resolved=False)
         ts2 = argutils.sim_coalescent(n, L=L, rho=rho, seed=seed, resolved=True)
         ts3 = ts1.simplify(keep_unary=True)
-        # print(ts1.draw_text())
-        # print(ts2.draw_text())
         ts2.tables.assert_equals(ts3.tables, ignore_provenance=True)
 
 
@@ -144,12 +142,11 @@ class TestEARG:
         E, sigma = argutils.as_earg(ts)
         for tree in ts.trees():
             x = tree.interval.left + tree.interval.span / 2
-            t = argutils.earg_get_tree(E, sigma, ts.samples(), x)
+            t = argutils.earg_get_tree(E, sigma, ts.samples(), int(x))
             assert t == parent_dict(tree)
 
 
 class TestGARG:
-
     @pytest.mark.parametrize("seed", range(1, 3))
     @pytest.mark.parametrize(
         ["n", "L"],
@@ -190,8 +187,19 @@ class TestGARG:
         E = argutils.as_garg(ts)
         for tree in ts.trees():
             x = tree.interval.left + tree.interval.span / 2
-            t = argutils.garg_get_tree(E, ts.samples(), x)
+            t = argutils.garg_get_tree(E, ts.samples(), int(x))
             assert t == parent_dict(tree)
+
+
+def assert_gargs_equal(E1, E2):
+    assert len(E1) == len(E2)
+    d1 = {(u, v): I for (u, v, I) in E1}
+    d2 = {(u, v): I for (u, v, I) in E2}
+    assert len(d1) == len(d2)
+    for key, I1 in d1.items():
+        I2 = d2[key]
+        # print(key, I1, I2)
+        assert I1 == I2
 
 class TestResolvedGARG:
     @pytest.mark.parametrize("seed", range(1, 3))
@@ -211,7 +219,6 @@ class TestResolvedGARG:
         rho = 0.1
         ts = argutils.sim_coalescent(n, L=L, rho=rho, seed=seed, resolved=False)
         E = argutils.as_resolved_garg(ts)
-        print(E)
         for tree in ts.trees():
             t = argutils.garg_get_tree(E, ts.samples(), tree.interval[0])
             assert t == parent_dict(tree)
@@ -235,7 +242,7 @@ class TestResolvedGARG:
[…]

### 2022-02-17 argutils/ancestry.py
finalise GARG resolution using topological sort

@@ -10,6 +10,7 @@ from typing import Any
 
 import numpy as np
 import tskit
+import networkx as nx
 
 from . import viz
 
@@ -653,20 +654,17 @@ def as_resolved_garg(ts):
     Returns the specified unresolved ARG as an GARG E with respect to
     the tree sequences samples.
     """
-    G = viz.convert_nx(ts)
-    import networkx as nx
-
-    topological = list(nx.topological_sort(G))
-    # print(topological)
-    Q = collections.deque(ts.samples())
     E = as_garg(ts)
+    G = nx.DiGraph()
+    G.add_edges_from([e[:2] for e in E])
+    # We have to get the topological sorting of the nodes because we don't
+    # have time ordering.
+    topological = list(nx.topological_sort(G))
+
     I = collections.defaultdict(lambda: IntervalSet(ts.sequence_length))
-    for u in Q:
+    for u in ts.samples():
         I[u] = IntervalSet(ts.sequence_length, [(0, ts.sequence_length)])
 
-    # while len(Q) > 0:
-    # print("Q = ", Q)
-    # c = Q.pop()
     Ep = []
     for c in topological:
         # print("c = ", c)

### 2022-02-17 paper.tex
Another pass through the Event ARG section

--- 2022-02-15 paper.tex
@@
 There are many ways in which the EARG structure can be formalised (see
-e.g.~\citep{gusfield2014recombinatorics}).
+e.g.~\citep{gusfield2014recombinatorics}); the following is useful
+to illustrate the computational properties of the structure.
 % It would be slightly easier for us if we defined nodes as in N_0, since
 % we use 0 as a node ID, but leaving for now.
-We can define an EARG as a tuple $(e, \sigma)$, where $e$ is an ordered
+We define an EARG as a tuple $(e, \sigma)$, where $e$ is an ordered
 list of tuples such that $e_j = (c, p)$ defines a directed edge between
 child node $c$ and parent $p$ ($c, p \in \mathbb{N}$), and
-$\sigma: \mathbb{N} \rightarrow \mathbb{N}_0$ is a function
-mapping nodes to recombination breakpoints, such that $\sigma(u) = x$
+$\sigma: \mathbb{N} \rightarrow \mathbb{N}$
+is a function mapping nodes to recombination breakpoints, such that $\sigma(u) = x$
 if $u$ is a recombination event with breakpoint $x$ and
-% This is arbitrary, but handy when you start defining the function
-% for recovering trees; see below
-$\sigma(u) = \infty$ otherwise. For each recombination node $u$ we must
+$\sigma(u) = \infty$ otherwise. (The use of $\infty$ to denote
+non-recombination nodes is arbitrary, but useful when we extract
+% NOTE: need to be careful about using "local" trees if we're also
+% distinguishing resolved and non-resolved GARGs as being "locally"
+% defined or not. One is topologically local and the other is local
+% along the genome.
+local trees; see below.)
+For each recombination node $u$ we must
 have exactly two edges in $E$ such that $u$ is the child.
@@
 
-% Not sure if we actually include this, but it's good to work it through
-% to make sure the notation works.
-Given a set of sample nodes $S$ we can recover the local tree at
-position $x$ by the following procedure. Let $N$ be a set of
-nodes we are currently constructing the tree at, and let
-$p: \mathbb{N} \rightarrow \mathbb{N}$ be the mapping from child
-node to parent. Firstly, initialise $N \leftarrow S$.
-Then, while $|N| > 0$, we add branches to the
-tree as follows. Let $u$ be an arbitrary element of $N$ (which
-we remove). Then, if $\sigma(u) < x$ we let $j$ be the
-smallest value such that $e_{j,1} = u$, and otherwise
-let $j$ be the largest value such that $e_{j,1} = u$.
-Then, set $v \leftarrow e_{j, 2}$,
-$p_u \leftarrow v$
-and $N \leftarrow N \cup \{v\}$.
-
-The Griffiths ARG encoding is ambiguous because it does not explicitly
-define which of the parent lineages resulting from a recombination
-event carries ancestral material to the left or right of the
-breakpoint. The ambiguity is resolved by adopting a simple rule when
-traversing the graph in order to generate the marginal trees for a
-given position $z$: at a recombination node $u$, we follow the
-first edge in which $u$ is a child if $f(u) \leq z$, and the
-second edge otherwise. This ordering requirement, which straightforward
-% Saying "awkward" too many times here, rephrase.
-to describe can be awkward in practice. For example,
-simulating an ARG in this representation is considerably
+% % Not sure if we actually include this, but it's good to work it through
+% % to make sure the notation works. Maybe something to stick into
+% % an appendix
+% The most fundamental operation that we need to perform on an ARG
+% data structure is extracting the local tree topology at a particular
+% position along the genome. The basic strategy is to traverse
+% the graph upwards, building the tree as we visit each node.
+% At a recombination node, when we have a choice of two parents,
+% we follow the path that corresponds to the position that we
+% are building the tree for. More precisely, let $S$ be the set
+% of sample nodes (usually the leaves of the graph)
+% and $0 \leq x < L$ be a position along the genome, and suppose
+% we are building a mapping $p_u$ that defines the parent of node
+% $u$ at $x$.
+% Let $N$ be a set of
+% nodes we are currently constructing the tree at,
+% and initialise $N \leftarrow S$.
+% Then, while $|N| > 0$, we add branches to the
+% tree as follows. Let $u$ be an arbitrary element of $N$ (which
+% we remove).
+% % This isn't right: we're not handling the case where there's
+% % no edges in E where the child is $u$.
+% Then, if $\sigma(u) < x$ we let $j$ be the
+% smallest value such that $e_{j,1} = u$, and otherwise
+% let $j$ be the largest value such that $e_{j,1} = u$.
+% Then, set $v \leftarrow e_{j, 2}$,
+% $p_u \leftarrow v$
+% and $N \leftarrow N \cup \{v\}$.
+
+The ordering of edges is a vital element of the EARG encoding.
+This is because there is a fundamental ambiguity involved
+in associating recombination breakpoints with \emph{nodes}
+rather than edges,
+and the only way we can break this symmetry is to assign
+some meaning to the order in which the parents are listed
+(\cite{griffiths1997ancestral} refer to the ``left'' and ``right''
+parents of a recombination event). When we want to recover
+the local tree at a position $x$ along the genome,
+we traverse the graph upwards from the leaves. At a particular
+node $u$, if it has one parent we are at a common ancestor
+node and we simply follow that parent. If we are at a
+recombination node, $u$ has two parents; if
+$x$ is less than the breakpoint $\sigma(u)$ we follow
+the first edge, and otherwise follow the second edge.
+This ordering requirement, which although straightforward
+to describe, has some drawbacks in practice. For example,
+simulating an ARG in this representation is
 complicated by the fact that the
 first event to occur may hit the right-hand lineage rather
-than the left.
-Such awkwardnesses can be worked around, of course,
-but in practice methods such has
-\texttt{beagle}~\citep{lyngso2005minimum} and
-\texttt{KwARG}~\citep{ignatieva2021kwarg}
-resolve this problem by
-associating a label ``P'' (prefix) or ``S'' (suffix) with the
-outbound edges of a recombination event.
+than the left (and so we cannot emit edges as they are generated).
+Such issues can be worked around, of course,
+but depending on the ordering of otherwise indistinguishable
+objects is generally problematic. In practice, several
+methods
+explicitly associate information with the outbound edges
+of a recombination event
+to resolve the problem~\citep{lyngso2005minimum,ignatieva2021kwarg}.
 

### 2022-02-17 paper.tex
Try putting the shoulders on the bottom of Fig 3B.
Update

@@
 
-\section*{The ARG as a data structure}
+\section*{The Event ARG data structure}
+% When did we move from the ARG is a process to the ARG is a data structure?
+% What was the historical progression from one to the other?
 Early work on ARG inference focused on the problem of
@@
 
+% What do we mean by an EARG, informally?
 The Griffiths ARG encoding is an elegant and mathematically
@@
 ARG encoding as an ``Event ARG'', or EARG.
-There are many ways in which the EARG structure can be formalised (see
-e.g.~\citep{gusfield2014recombinatorics}); the following is useful
-to illustrate the computational properties of the structure.
-% It would be slightly easier for us if we defined nodes as in N_0, since
-% we use 0 as a node ID, but leaving for now.
-We define an EARG as a tuple $(e, \sigma)$, where $e$ is an ordered
+The basic unit of an EARG are \emph{events} and correspond directly
+to the events that
+occur in the ARG stochastic process described in the previous section.
+Each event corresponds to a node in the graph, and there are therefore
+two types of node: common ancestor and recombination. As in the
+stochastic process, edges represent lineages, and those lineages
+are modified by the events that they pass through.
+
+% What do we mean by an EARG, formally?
+% NOTE we're using the natural numbers but are using 0 as a node in the
+% examples. Probably want to clear this up a bit.
+Formally, we can define an EARG as a tuple $(e, \sigma)$, where $e$ is an ordered
 list of tuples such that $e_j = (c, p)$ defines a directed edge between
@@
 %% Edges
-\draw (s1) |- (s3);
-\draw (s1) |- (s4) |- (s6);
+\draw (s1) -| (s3);
+\draw (s1) -| (s4) |- (s6);
 \draw (s3) |- (s5);
@@
 to describe, has some drawbacks in practice. For example,
+% Surely the meaning is clear from the context here and we don't need
+% qualify the meaning of "ARG"
 simulating an ARG in this representation is
 complicated by the fact that the
-first event to occur may hit the right-hand lineage rather
+first event to occur may hit the lineage carrying the ancestry
+to the right of the breakpoint rather
 than the left (and so we cannot emit edges as they are generated).
@@
 patterns of inheritance without classifying different types of node.
-The second change we must make is to encode
-intervals of ancestry inheritance \emph{explicitly} on graph edges,
-rather than \emph{implicitly} as breakpoints associated with
-recombination events and edge ordering rules.
-That is, given two genomes $p$
-and $c$ where $p$ is the parent genome and $c$ is the child
-(although many generations may have elapsed between the two),
-we associate a set $I$ of intervals of genome which $c$ has
-inherited from $p$.
-These simple changes required to translate from the
-classical Griffiths approach to the proposed encoding
-are shown in Fig.\ref{fig-arg-data-structure}.
-
-\textcolor{red}{
-This falls a bit flat here,  but it's not clear how we're dividing
-the responsibility for explaining the model between the
-different sections.
-Basically, we can't explain everything at
-the end of this section, but we want to give the reader the
-basic idea (which is genuinely simple). Can see how things are
-shaping up after we've got a draft of the rest of the
-sections, I guess.}
+The second change we must make is to associate
+the details of inheritance intervals with \emph{edges} rather than
+as breakpoints associated with recombination events and
+edge ordering rules.
+% That is, given two genomes $p$
+% and $c$ where $p$ is the parent genome and $c$ is the child
+% (although many generations may have elapsed between the two),
+% we associate a set $I$ of intervals of genome which $c$ has
+% inherited from $p$.
+These changes are straightforward, involve no loss of information
+and greatly increase both the expressivity and computational
+power of ancestral recombination graph data structures.
 
 \section*{Genome ARGs}
-
+% We're not the first to talk about ARGs and pedigrees, and we're
+% not doing anything controversial.
 The deep connection between ARGs and pedigrees is of course
@@
 correspond to particular genomes in the pedigree.
-As discussed in the previous section, the definition of the ARG
+As discussed in the previous section, the definition of an ARG
 as a \emph{data structure} in terms of these two events is problematic,
-fundamentally limiting the expressivity of the structure.
+fundamentally limiting the patterns of ancestry that can be described.
 Following \cite{mathieson2020ancestry} we instead define the
@@
 \begin{figure}
+\begin{center}
     \includegraphics[width=\textwidth]{illustrations/pedigree_figure}
+\end{center}
 \caption{\label{fig-pedigree-and-arg}
@@
 denote inheritance. Each edge is annotated with genome coordinates
-defining the specific region(s) inherited. An ARG is a retrospective
-structure defined with
-respect to some set of sample genomes (A--D here), and so only
-genomes directly ancestral to those samples are included.
+defining the specific region(s) inherited.
 (C) This description of inheritance fully defines the
@@
 
-Formally, we can define a GARG as a set of edges $E$, where each
-element of the set is a tuple $(c, p, I)$ such that $c, p \in \mathbb{N}$,
-$c$ is the child node, $p$ is the parent node and $I$ is the set of
-disjoint genomic intervals $(\ell, r]$ over which genome $c$ inherits from $p$ (we may
-also phrase this in terms of the ancestral material of a sample---see
-the next section).
-
-Fig~\ref{fig-pedigree-and-arg} illustrates the basic elements of the model.
+To distinguish from the classical event-based encodings,
+we call this encoding of an ARG data
+structure a ``Genome ARG'', or GARG.
+Fig~\ref{fig-pedigree-and-arg} illustrates the basic elements of a GARG.
 The pedigree is composed of a set of individuals who each carry
@@
 be the recombined product of that parent's two genomes.
-For example, individual $I_1$ has two genomes $A$ and $B$,
+For example, individual $I_1$ in Fig.~\ref{fig-pedigree-and-arg}A
+has two genomes $A$ and $B$,
 inherited from parents $I_3$ and $I_4$. Genome $A$ is the product of
@@
 whereas $B$ was inherited directly from $I_4$'s genome $G$ without
-recombination.
+recombination. The pedigree is shown here for concreteness, but
+is not a necessary part of the model. We can reason directly about the
+genomes without necessarily knowing how they are combined within
+individuals. This ``genome-perspective'' is show in
+Fig.~\ref{fig-pedigree-and-arg}B, which depicts the same graph
+of the ancestry of the sample genomes $A$--$D$. The graph
+topology along with the genome coordinate annotations on the
+edges then uniquely defines the trees at every position
+along the genome (Fig.~\ref{fig-pedigree-and-arg}C).
 
@@
 ancestry resolution.]
+
+Formally, we can define a GARG as a set of edges $E$, where each
+element of the set is a tuple $(c, p, I)$ such that $c, p \in \mathbb{N}$,
+$c$ is the child node, $p$ is the parent node and $I$ is the set of
+disjoint genomic intervals $(\ell, r]$ over which genome $c$ inherits from $p$ (we may
+also phrase this in terms of the ancestral material of a sample---see
+the next section).
+
+% Asking the reader to flick back to figure 2 here, not great.
+The differences between EARGs
+and GARGs are subtle but important, and illustrated in
+Fig.~\ref{fig-arg-data-structure}, which shows a simple ARG
+with three samples and one recombination.
+Fig.~\ref{fig-arg-data-structure}A shows this topology as a
+classical Griffiths-like EARG, and
+Fig.~\ref{fig-arg-data-structure}B shows it as a GARG.
+The two are almost identical, except we model the recombination
+event via the two parental genomes in nodes $3$ and $4$ and explicitly
+store the genomic interval that the child inherited from the
+parent with every edge.
+This approach may seem wasteful since we must store an extra
+node for each recombination and genomic interval information
+with every edge (even though many simply span the full genome).
+However, asymptotically the amount of information stored is the
+same, and as we see in the next section, keeping interval
+information for all edges enables useful generalisations.
+There is also much simplicity to be gained by having
+a single type of node and by treating all edges in the same way.
 
@@
 
+\item We make the point about expressivity of the EARG pretty
+hard, but we don't mention anything about indels and structural
+rearrangements etc. It's probably worth mentioning genome graphs
+somewhere just to fend off this criticism. The high-level point
+would be that it's unlikely to be easier to model this stuff
+using node annotations.
+
 \end{itemize}

### 2022-02-17 argutils/ancestry.py
Tweaks to code
Tweaks

More text

@@ -12,8 +12,6 @@ import numpy as np
 import tskit
 import networkx as nx
 
-from . import viz
-
 NODE_IS_RECOMB = 1 << 1
 
 

### 2022-02-17 illustrations/ARG_edge_annotations.py
Tweaks to Figure 4

@@ -7,6 +7,7 @@ import matplotlib.pyplot as plt
 import networkx as nx
 import numpy as np
 import tskit
+import PIL
 
 current_dir = Path(__file__).parent
 
@@ -57,9 +58,10 @@ def arg_edge_annotations():
                     font_size=16,
                     edge_labels={(e.child, e.parent): f"({e.left:.0f},{e.right:.0f}]" for e in ts.edges() if func1(e) and func2(e)},
                     horizontalalignment=halign,
+                    bbox=dict(boxstyle="round,pad=0.05", ec=(1.0, 1.0, 1.0), fc=(1.0, 1.0, 1.0)),
                 )        
     
-    fig, (ax1, ax2) = plt.subplots(1, 2, figsize=(25,13), sharey=True)
+    fig, (ax1, ax2) = plt.subplots(1, 2, figsize=(26,18), sharey=True)
     
     ax1.set_title("(a) eARG with implicit\nencoding (Wiuf & Hein)")
     ts = argutils.viz.label_nodes(argutils.wh99_example())
@@ -67,13 +69,13 @@ def arg_edge_annotations():
         ts, ax1,
         use_ranked_times=False,
         node_color=mpl.colors.to_hex(plt.cm.tab20(1)),
-        node_size=450,
+        node_size=100,
         max_edge_width=2,
         font_size=14,
         tweak_x={
-            0: 10, 3: 29, 4: 2.5, 7: 20, 11: 10, 12: -18,
-            20: -23, 19: 4, 21: -11, 17: -25, 13: -16,
-            8: 2, 9: -24, 5: -16, 6: -41, 15: 1, 16: -26, 2: -10
+            0: 10, 3: 29, 4: 2.5, 7: 20, 11: 9, 12: -17,
+            20: -22.5, 19: 3.5, 21: -11, 17: -25, 13: -16,
+            8: 2, 9: -24, 5: -16, 6: -41, 15: 0.8, 16: -25.8, 2: -10
         }
     )
     add_edge_labels(ax1, ts, G, pos)
@@ -83,13 +85,106 @@ def arg_edge_annotations():
     pos, G = argutils.viz.draw(
         ts2, ax2, pos=pos,
         node_color=mpl.colors.to_hex(plt.cm.tab20(1)),
-        node_size=450,
+        node_size=200,
         font_size=14,
         #arrows=True,
         draw_edge_widths=True,
     )
     add_edge_labels(ax2, ts2, G, pos)
     
+    
[…]

### 2022-02-17 argutils/viz.py
Figure tweaks to make nodes and labels more visible

@@ -15,6 +15,7 @@ def draw(
     *,
     use_ranked_times=None,
     tweak_x=None,
+    tweak_y=None,
     arrows=False,
     pos=None,
     draw_edge_widths=False,
@@ -40,6 +41,11 @@ def draw(
     tweak_x is a dict of {u: x_adjustment_percent} which allows
     the x position of node u to be hand-adjusted by adding or
     subtracting a percentage of the total x width of the plot
+
+    tweak_y is a dict of {u: y_adjustment_percent} which allows
+    the y position of node u to be hand-adjusted by adding or
+    subtracting a unit of time ranking (only if use_ranked_times
+    is true).
 
     If pos is passed in, it should be a dictionary mapping nodes to
     positions.
@@ -86,6 +92,9 @@ def draw(
             times = ts.tables.nodes.time
             for i, p in pos.items():
                 p[1] = times[i]
+        if tweak_y:
+            for node, tweak_val in tweak_y.items():
+                pos[node] = pos[node] + np.array([0, tweak_val])
     if tweak_x:
         plot_width = np.ptp([x for x, _ in pos.values()])
         for node, tweak_val in tweak_x.items():

### 2022-02-17 illustrations/ARG_edge_annotations.py

@@ -59,15 +59,16 @@ def arg_edge_annotations():
                     edge_labels={(e.child, e.parent): f"({e.left:.0f},{e.right:.0f}]" for e in ts.edges() if func1(e) and func2(e)},
                     horizontalalignment=halign,
                     bbox=dict(boxstyle="round,pad=0.05", ec=(1.0, 1.0, 1.0), fc=(1.0, 1.0, 1.0)),
-                )        
+                )
+
+    fig, (ax1, ax2) = plt.subplots(1, 2, figsize=(20,12), sharey=True)
+    fig.tight_layout()
     
-    fig, (ax1, ax2) = plt.subplots(1, 2, figsize=(26,18), sharey=True)
-    
-    ax1.set_title("(a) eARG with implicit\nencoding (Wiuf & Hein)")
+    ax1.set_title("(a) eARG with implicit\nencoding (Wiuf & Hein)", fontsize="xx-large")
     ts = argutils.viz.label_nodes(argutils.wh99_example())
     pos, G = argutils.viz.draw(
         ts, ax1,
-        use_ranked_times=False,
+        use_ranked_times=True,
         node_color=mpl.colors.to_hex(plt.cm.tab20(1)),
         node_size=100,
         max_edge_width=2,
@@ -75,12 +76,15 @@ def arg_edge_annotations():
         tweak_x={
             0: 10, 3: 29, 4: 2.5, 7: 20, 11: 9, 12: -17,
             20: -22.5, 19: 3.5, 21: -11, 17: -25, 13: -16,
-            8: 2, 9: -24, 5: -16, 6: -41, 15: 0.8, 16: -25.8, 2: -10
+            8: 2, 9: -24, 5: -16, 6: -41, 15: 0.6, 16: -25.6, 2: -10
+        },
+        tweak_y={
+            22: 1, 17: 0.8
         }
     )
     add_edge_labels(ax1, ts, G, pos)
     
-    ax2.set_title("(b) explicit encoding\n(i.e. non-ancestral removed)")
+    ax2.set_title("(b) explicit encoding\n(i.e. non-ancestral removed)", fontsize="xx-large")
     ts2 = argutils.simplify_keeping_all_nodes(ts)
     pos, G = argutils.viz.draw(
         ts2, ax2, pos=pos,
@@ -96,13 +100,8 @@ def arg_edge_annotations():
     # From https://networkx.org/documentation/stable/auto_examples/drawing/plot_custom_node_icons.html
     icons = {
[…]

### 2022-02-18 paper.tex

@@
 also phrase this in terms of the ancestral material of a sample---see
-the next section).
+the next section). Deriving the local tree at a point $x$
+follows largely the same pattern as for an EARG but is somewhat
+simpler because coordinate information is directly associated with
+edges. Starting with the sample nodes $S$ we traverse
+upwards through the graph from child to parent. At each node $u$, we find an
+edge $(c, p, I) \in E$ such that $u = c$ and $x \in I$ (if it
+exists) and move to the next node $p$.
 
@@
 individuals' monoploid genomes
-and their direct patterns of inheritance allows us to describe arbitarily
+and their direct patterns of inheritance allows us to describe arbitrarily
 complex relatedness structures, free from the limitations imposed by
@@
 
-% This is weak. Should we give a better description of the algorithm?
-% Not clear yet, and don't want to spend too much time on it if it's
-% unhelpful.
-Given a set of edges $(u, v, I)$ where node $u$ descends from $v$
-over the set of genomic intervals $I$, and a set of sample nodes $S$
-it is straightforward to construct the local tree at position $x$.
-Starting from the sample nodes in $S$, traverse upwards through the
-graph, such that when we are at a node $u$ we follow the edge
-$(u, v, I)$ where $x \in I$. This is essentially the same procedure as
-described by Griffiths~\citep{griffiths1997ancestral}.
-
-Recovering the tree at a single point along the genome, although
-a fundamental operation, is not the only operation that we may wish
-to perform on an ARG. We can make the operation of efficiently
-constructing \emph{all} trees along the genome much more efficient,
-and gain insight into the structure of the ARG,
-if we perform an additional step and \emph{resolve} the ancestral
-material in the ARG with respect to the samples.
-
-As we saw in the section X, the key to efficiently accessing the
-results of the ARG stochastic process is to reason about
-the flow of ancestral material through the graph using Hudson's
-algorithm. The same basic process allows us to determine the
-exact intervals of ancestral material that are flowing through
-each edge in the graph. Concretely, given a set of edges
-$E = \{(u, v, I)\}$ where $u$ is the child, $v$ is the parent
-and $I$ is the set of genomic intervals passed directly from
-$v$ to $u$, the resolution process with respect to a set of
-samples $S$ produces a set of edges $E' = \{(u, v, I)\}$
-where $I$ is the set of genomic intervals over which there is a
-% weak and not correct - how do we say this properly?
-path from at least one element of $S$ to $u$,
-and $u$ is not an MRCA.
+Given a GARG $E$ and the set of samples $S$ we can \emph{resolve} its
+ancestry by noting that the resolved interval $Ir$ on an edge $(c,p,I)$
+is given by [Rough notes]
+
+\begin{itemize}
+\item The ``ancestry interval'' $A_u$ of a node $u$ is the union of
+all $R_{u, v}$ for all the children $v$ of $u$. (this isn't quite right)
+\item The resolved interval is then the intersection of $A_u$ and the
+original inheritance interval $I_v$ on the edge.
+\item It would be good of if differentiate between the inheritance intervals
+ from the GARG and the ancestry intervals of the resolved GARG.
+\end{itemize}
 
@@
 \end{figure}
+
+% Not sure where this text will end up, but we want it somewhere
+% towards the end.
 
@@
 
+% TODO not clear this belongs in the discussion, but we'll definitely
+% want to make some of these points, even if they are being repeated
+% from earlier section.
+It is difficult to imagine a practical calculation performed on an
+ARG that does not involve either the sequential generation of trees
+along the genome or the resolved ancestry segments.
+Knowing the resolved segments in an ancestry resolved GARG allows
+us to perform many calculations without explicitly generating
+the trees, and lets us work with the ancestry on a genome-by-genome
+basis. We give three examples here. Firstly, generating mutations
+on edges can be performed very quickly on an ancestry resolved
+GARG, since we only need to examine $O(|E|)$ edges rather than
+$O(nt)$ tree branches (assuming $n$ samples and $t$ marginal trees)
+~\citep{baumdicker2021efficient}. If the ancestry were not resolved,
+we would generate a large number of mutations on areas of the edge that
+are non-ancestral. Secondly, computing the likelihood of a given
+ARG realisation under the coalescent with recombination model
+requires that we essentially recapitulate the state of Hudson's
+algorithm (see XXX section) at each node in
+the graph~\citep{baumdicker2021efficient}. Having resolved ancestry
+substantially simplifies this process. Thirdly, computing
+[FIXME describe what we did in tsdate]~\citep{wohns2021unified}.
+
+Sequentially generating the marginal trees along the genome
+is also fundamental, and is necessary whenever we need to
+perform calculations that are contingent on more than just the
+isolated properties of an edge. \cite{kelleher2016efficient}
+showed how all trees can be sequentially generated in
+constant time per tree transition in a fully simplified GARG.
+Furthermore, we can easily reason about how tree topologies
+change (and stay the same), leading to efficient algorithms
+for computing population genetic
+statistics~\citep{kelleher2016efficient,kelleher2018efficient},
+implementing the Li and Stephens
+model~\citep{kelleher2019inferring,wohns2021unified}
+and likely many more.
+
 \bibliographystyle{plainnat}

### 2022-02-18 illustrations/ARG_edge_annotations.py
Changed patty colouring to increase from grey to black when MRCA reached

@@ -8,6 +8,7 @@ import networkx as nx
 import numpy as np
 import tskit
 import PIL
+import string
 
 current_dir = Path(__file__).parent
 
@@ -95,30 +96,18 @@ def arg_edge_annotations():
         draw_edge_widths=True,
     )
     add_edge_labels(ax2, ts2, G, pos)
-    
-    
+
     # From https://networkx.org/documentation/stable/auto_examples/drawing/plot_custom_node_icons.html
     icons = {
         "genome_empty": "node_icons/genome_empty.png",
         "genome_empty_hamburger": "node_icons/genome_empty_hamburger.png",
         "genome_full": "node_icons/genome_full.png",
-        "genome_0-4": "node_icons/genome_0-4.png",
-        "genome_4-7": "node_icons/genome_4-7.png",
-        "genome_0-2": "node_icons/genome_0-2.png",
-        "genome_2-7": "node_icons/genome_2-7.png",
-        "genome_0-2-4-7": "node_icons/genome_0-2-4-7.png",
-        "genome_4-6": "node_icons/genome_4-6.png",
-        "genome_6-7": "node_icons/genome_6-7.png",
-        "genome_0-1": "node_icons/genome_0-1.png",
-        "genome_00-11": "node_icons/genome_00-11.png",
-        "genome_1-7": "node_icons/genome_1-7.png",
-        "genome_00-22-2-7": "node_icons/genome_00-22-2-7.png",
-        "genome_11-22-2-7": "node_icons/genome_11-22-2-7.png",
-        "genome_22-77": "node_icons/genome_22-77.png",
-        "genome_ancestral": "node_icons/genome_ancestral.png",
     }
+    for letter in list(string.ascii_uppercase[3:G.number_of_nodes()]):
+        icons["genome_" + letter] = "node_icons/genome_" + letter + ".png"
     # Load images
     images = {k: PIL.Image.open(fname) for k, fname in icons.items()}
+
     # Panel (a)
     tr_figure = ax1.transData.transform
     tr_axes = fig.transFigure.inverted().transform
@@ -131,43 +120,19 @@ def arg_edge_annotations():
         xa, ya = tr_axes((xf, yf))
         a = plt.axes([xa - icon_center, ya - icon_center, icon_size, icon_size])
         a.imshow(G.nodes[n]["image"])
[…]

### 2022-02-18 paper.tex
Add separate simplification section

@@
 
-\section*{Ancestry resolution and simplification}
+\section*{Ancestry resolution}
 % This is a rough first pass. Will clarify terminology later.
@@
 
+\section*{ARG simplification}
+
+Not all nodes in an ARG are equal. Assuming that we have sequenced the
+genomes of the samples, there are intrinsic limits on what can
+be known about more ancient nodes, determined by the graph topology
+and the flow ancestral material through it.
+\emph{Simplification}~\citep{kelleher2018efficient} is the process
+of removing nodes and re-writing edges in a GARG to remove
+various types of redundancy, while maintaining the crucial properties
+that the meaningful topology in the local trees is not affected,
+and the relationship between nodes in adjacent trees is maintained.
+
+We may have nodes corresponding to genomes that are ancestral to
+the sample along their entire span
+(e.g.\ node $H$ in Fig.~\ref{fig-ancestry-resolution}),
+genomes that carry no genetic material ancestral to the sample at all
+(e.g.\ node $J$ in Fig.~\ref{fig-ancestry-resolution}),
+and every possibility along the continuum between these extremes.
+Clearly, we can only ever hope to infer the fraction of an ancestor's
+genome that is ancestral to the sample.
+
 \begin{figure}
@@
 \includegraphics[width=\linewidth]{illustrations/ARG_recomb_node_deletion}
-\caption{\label{fig-recombination-nodes}
-Recombination nodes are optional in an edge-annotated ARG. (a) A simulated
+\caption{\label{fig-simplification}
+ARG simplification. (a) A simulated
 ARG including all recombination nodes (red). (b) The same ARG having removed both
@@
 \end{figure}
-
-% Not sure where this text will end up, but we want it somewhere
-% towards the end.
 

### 2022-02-18 paper.tex
more text

@@
 % examples. Probably want to clear this up a bit.
-Formally, we can define an EARG as a tuple $(e, \sigma)$, where $e$ is an ordered
-list of tuples such that $e_j = (c, p)$ defines a directed edge between
-child node $c$ and parent $p$ ($c, p \in \mathbb{N}$), and
+Formally, we can define an EARG as a tuple $(e, \sigma)$, where $e$
+defines a directed acyclic graph where $e_j = (c, p)$ describes
+an edge between
+child node $c$ and parent $p$ ($c, p \in \mathbb{N}$).
+Additionally,
 $\sigma: \mathbb{N} \rightarrow \mathbb{N}$
@@
 
-We may have nodes corresponding to genomes that are ancestral to
-the sample along their entire span
-(e.g.\ node $H$ in Fig.~\ref{fig-ancestry-resolution}),
-genomes that carry no genetic material ancestral to the sample at all
-(e.g.\ node $J$ in Fig.~\ref{fig-ancestry-resolution}),
-and every possibility along the continuum between these extremes.
-Clearly, we can only ever hope to infer the fraction of an ancestor's
-genome that is ancestral to the sample.
-
 \begin{figure}
@@
 \caption{\label{fig-simplification}
-ARG simplification. (a) A simulated
-ARG including all recombination nodes (red). (b) The same ARG having removed both
-recombination nodes and nodes of common ancestors in which no genetic
-lineages coalesce (cyan). (c) The sequence of trees corresponding to (b): if we were
-to plot out the trees corresponding to (a) we would have more trees, but the topologies
-and MRCA nodes at each genomic location would be identical. For clarity, nodes are
-plotted by ranked time rather than true time on the y-axis.
-% HYW: note that we also truncate nodes for portions in which they are unary,
-% so that nodes which are coalescent nodes over only some of their length
-% (blue-green) become coalescent nodes over their entire length (green). This
-% can, however, throw away some useful information, for example node 4 has
-% only a single parent edge in (a), but 2 parent edges (one of which appears
-% to bypass node 11) in (b). In fact, this edge should go through node 11, but
-% there may be no way to detect this from real data.
+ARG simplification. (A) An ARG simulated from a diploid Wright-Fisher
+model. Note the complex full-sib relationship between B and C; this
+could not be directly represented in a Griffiths-like EARG encoding
+(see section XXX).
+(B) Simplified to remove all
+% trying out this terminology - can define in the text
+1-connected graph components (e.g., diamonds).
+(C) Remove nodes that are unary in all local trees.
+(D) Rewrite edges to skip any unary nodes in local trees.
+(E) The local trees for ARG (A).
+(F) The local trees for ARG (D).
 }
 \end{figure}
+
+% Q: would it help to talk about "tree vertices" here or would it
+% just be more confusing? I worry that people would think that
+% they are two different things then, and it's really important
+% that they get that a node-is-a-node
+The ideas of ``meaningful'' local tree topologies revolve around the
+presence of ``unary'' nodes: a node in the tree that has exactly
+one child. For example, Fig.~\ref{fig-simplification}E shows the
+local trees along the genome for a simulated ARG
+(Fig.~\ref{fig-simplification}A). Consider the right-most tree,
+and the path above node $D$. Before $D$ meets its common ancestor
+over the interval $(93, 100]$ with the other samples it passes through
+nodes $H, K, N, R$ and $S$, reflecting the corresponding
+traversal through the ARG (see section XXX). Such unary nodes
+in a local tree are not usually considered significant,
+because they provide very little information and are not
+detectable by information from the tree. % weak
+The nodes in a tree usually mark the points where samples find a
+\emph{common} ancestor, rather than focusing on specific
+ancestors. For most purposes, we would consider an edge
+directly joining sample $D$ to the node $T$ to be more
+useful and informative. Simplification is the process of removing
+these uninformative and (for most computational purposes)
+inefficient unary paths.
+
+The first type of simplification that we can perform is to remove
+graph topology that is invisible to the samples. The most well
+known example of such topology is a so-called
+``diamond''~\citep{rasmussen2014genome}
+in which the two parent nodes of a recombination immediately
+join again into a common ancestor (e.g. $K,N,O$ and $R$ in
+Fig.~\ref{fig-simplification}A).
+Unless we are specifically
+interested in the recombination event or ancestral genomes,
+there is no information in this topology and the diamond can b
+e replaced by a single edge. More generally, any
+subgraph which is singly-connected in both the leafward and
+rootward direction (a ``super-diamond'') is non-identifiable and can be
+replaced by a single edge. This definition includes the case
+of a single node that has one inbound and one outbound edge.
+Fig.~\ref{fig-simplification}B shows the result of this type of
+graph-level simplification.
+
+Simplifying away diamonds will remove many unary nodes from the
+local trees, there can still be nodes that are unary in all
+of the local trees. For example, node $H$ is one of the
+parents of the recombinant node $D$ is still present in
+Fig.~\ref{fig-simplification}B after diamond removal, but
+it is unary in all of the local trees since it has an
+in-degree of $1$.
+Fig.~\ref{fig-simplification}C shows the resulting graph
+after removing these nodes.
+
+The remaining nodes are MRCAs of some subset of the samples
+at \emph{some} positions along the genome. We still have
+some unary nodes in the local trees, but these nodes will
+correspond to a coalescence in at least one other
+local tree. For example, node $E$ is unary in the first tree
+of Fig.~\ref{fig-simplification}E, but is either binary
+or ternary in the other trees (recall this is a Wright-Fisher
+simulation).
+
+The final level of simplification is to remove \emph{all}
+unary nodes from the local trees (Fig.~\ref{fig-simplification}D
+and E). [And continue]
+
+% This text might go somewhere
+% We may have nodes corresponding to genomes that are ancestral to
+% the sample along their entire span
+% (e.g.\ node $H$ in Fig.~\ref{fig-ancestry-resolution}),
+% genomes that carry no genetic material ancestral to the sample at all
+% (e.g.\ node $J$ in Fig.~\ref{fig-ancestry-resolution}),
+% and every possibility along the continuum between these extremes.
+% Clearly, we can only ever hope to infer the fraction of an ancestor's
+% genome that is ancestral to the sample.
 

### 2022-02-18 illustrations.py
Added illustrations.py

@@ -1,0 +1,205 @@
+import click
+import argutils
+import matplotlib as mpl
+import matplotlib.pyplot as plt
+
+import io
+import networkx as nx
+import PIL
+import string
+
+
+@click.group()
+def cli():
+    pass
+
+
+@click.command()
+def arg_edge_annotations():
+    def add_edge_labels(ax, ts, G, pos):
+        def edge_not_above_recombinant(edge):
+            return not argutils.is_recombinant(ts.node(edge.child).flags)
+
+        def left_recombinant_edge(edge):
+            if not argutils.is_recombinant(ts.node(edge.child).flags):
+                return False
+            if edge.right == ts.sequence_length:
+                return False
+            return True
+
+        def right_recombinant_edge(edge):
+            if not argutils.is_recombinant(ts.node(edge.child).flags):
+                return False
+            if edge.left == 0:
+                return False
+            if edge.left == 4 and edge.right == 6:  # Hack for this ts
+                return False
+            return True
+
+        for halign, func1 in {
+            "center": edge_not_above_recombinant,
+            "right": left_recombinant_edge,
+            "left": right_recombinant_edge,
+        }.items():
+            for full_edge, func2 in {
+                True: lambda e: e.left == 0 and e.right == 7,
+                False: lambda e: not (e.left == 0 and e.right == 7),
+            }.items():
+                nx.draw_networkx_edge_labels(
+                    G,
+                    pos=pos,
+                    ax=ax,
+                    rotate=False,
+                    font_weight="normal" if full_edge else "bold",
+                    alpha=0.5 if full_edge else None,
+                    font_size=16,
+                    edge_labels={
+                        (e.child, e.parent): f"({e.left:.0f},{e.right:.0f}]"
+                        for e in ts.edges()
+                        if func1(e) and func2(e)
+                    },
+                    horizontalalignment=halign,
+                    bbox=dict(
+                        boxstyle="round,pad=0.05",
+                        ec=(1.0, 1.0, 1.0),
+                        fc=(1.0, 1.0, 1.0),
[…]

### 2022-02-18 illustrations.py
Move simplify example into illustration.py file

@@ -16,6 +16,10 @@ def cli():
 
 @click.command()
 def arg_edge_annotations():
+    """
+    Ancestry propagation via WH99 graph.
+    """
+
     def add_edge_labels(ax, ts, G, pos):
         def edge_not_above_recombinant(edge):
             return not argutils.is_recombinant(ts.node(edge.child).flags)
@@ -199,7 +203,112 @@ def arg_edge_annotations():
         f.write(top_svg)
 
 
+@click.command()
+def arg_node_simplification():
+    """
+    Sequentially simplifying a WF simulation.
+    """
+    seed = 372
+    t_x = {
+        9: 3,
+        6: -8,
+        18: 50,
+        19: -15,
+        17: 5,
+        13: 10,
+        14: 2,
+        10: -10,
+        15: 5,
+        16: -5,
+        7: -5,
+        11: 5,
+        8: 8,
+    }
+
+    ts = argutils.sim_wright_fisher(2, 10, 100, recomb_proba=0.1, seed=seed)
+    labels = {i: string.ascii_uppercase[i] for i in range(len(string.ascii_uppercase))}
+    # relabel the nodes to get samples reading A B C D
+    labels.update({2: "B", 3: "C", 1: "D"})
+    ts = argutils.viz.label_nodes(ts, labels)
+
+    fig, (ax1, ax2, ax3, ax4) = plt.subplots(1, 4, figsize=(12, 5))
+    ax1.set_title("(a) little ARG")
+    ax2.set_title("(b) remove pass through")
+    ax3.set_title("(c) simplify (keep unary in coal)")
+    ax4.set_title("(d) fully simplified")
+    col = mpl.colors.to_hex(plt.cm.tab20(1))
+    pos, G = argutils.viz.draw(
+        ts,
+        ax1,
+        draw_edge_widths=True,
+        use_ranked_times=True,
+        node_color=col,
+        tweak_x=t_x,
+    )
+    ts2, node_map = argutils.simplify_remove_pass_through(
+        ts, repeat=True, map_nodes=True
+    )
+    argutils.viz.draw(
+        ts2,
+        ax2,
+        draw_edge_widths=True,
+        pos={node_map[i]: p for i, p in pos.items()},
+        node_color=col,
+    )
+    ts3, node_map = argutils.simplify_keeping_unary_in_coal(ts, map_nodes=True)
+    argutils.viz.draw(
+        ts3,
+        ax3,
[…]

### 2022-02-18 illustrations/ARG_recomb_node_deletion.py
Swap O and N labels
Fixes #124

@@ -21,12 +21,15 @@ outfile = "ARG_recomb_node_deletion"
 
 def arg_node_simplification():
     seed = 372
-    t_x={9: 3, 6: -8, 18: 50, 19: -15, 17: 5, 13: 10, 14: 2, 10: -10, 15: 5, 16: -5, 7: -5, 11: 5, 8: 8}
+    t_x={
+        9: 3, 6: -8, 18: 50, 19: -15, 17: 5,
+        13: 35, 14: -23, 10: -10, 15: 5, 16: -5,
+        7: -5, 11: 5, 8: 8}
 
     ts = argutils.sim_wright_fisher(2, 10, 100, recomb_proba=0.1, seed=seed)
     labels = {i: string.ascii_uppercase[i] for i in range(len(string.ascii_uppercase))}
     # relabel the nodes to get samples reading A B C D
-    labels.update({2: 'B', 3: 'C', 1: "D"})
+    labels.update({2: 'B', 3: 'C', 1: "D", 14: "N", 13: "O"})
     ts = argutils.viz.label_nodes(ts, labels)
     
     fig, (ax1, ax2, ax3, ax4) = plt.subplots(1, 4, figsize=(12,5))

### 2022-02-18 paper.tex
Correct notation from eARG to gARG

@@
 \caption{\label{fig-ancestry-resolution}
-Implicit vs explicit edge annotations. (a) An eARG equivalent to the encoding
+Implicit vs explicit edge annotations. (a) An gARG equivalent to the encoding
 described by Griffiths can be created by setting all edges to span the full
 genome, apart from edges between two recombining nodes and their recombinant
-child (edge spans in bold). (b) The sample-resolved eARG, where the Hudson
+child (edge spans in bold). (b) The sample-resolved gARG, where the Hudson
 simplification algorithm has shortened edge spans so that they encapsulate only

### 2022-02-21 illustrations.py
Port the pedigree figure.

@@ -1,12 +1,16 @@
-import click
-import argutils
-import matplotlib as mpl
-import matplotlib.pyplot as plt
-
+import pathlib
+import string
 import io
+from types import SimpleNamespace
+
+import click
 import networkx as nx
 import PIL
-import string
+import matplotlib as mpl
+import matplotlib.pyplot as plt
+import tskit
+
+import argutils
 
 
 @click.group()
@@ -307,8 +311,110 @@ def arg_node_simplification():
         f.write(top_svg)
 
 
+@click.command()
+def pedigree():
+    """
+    The ARG as embedded in diploid pedigree.
+    """
+
+    pedigree_svg = pathlib.Path("illustrations/pedigree.svg").read_text()
+    pedigree_svg = pedigree_svg[pedigree_svg.find("<svg") :]
+
+    n = SimpleNamespace()  # convenience labels
+    l = 100
+    tables = tskit.TableCollection(sequence_length=l)
+    tables.nodes.metadata_schema = tskit.MetadataSchema.permissive_json()
+    for gen in range(4):
+        for individual in range(2):
+            i = tables.individuals.add_row()
+            for genome in range(2):
+                label = string.ascii_uppercase[tables.nodes.num_rows]
+                metadata = {
+                    "gender": "male" if individual == 0 else "female",
+                    "name": label,
+                    "genome": "paternal" if genome == 0 else "maternal",
+                }
+                setattr(n, label, tables.nodes.num_rows)
+                tables.nodes.add_row(
+                    flags=tskit.NODE_IS_SAMPLE if gen == 0 else 0,
+                    time=gen,
+                    metadata=metadata,
+                    individual=i,
+                )
+
+    bp = [25, 60]
+    tables.edges.clear()
+    tables.individuals[0] = tables.individuals[0].replace(parents=[2, 3])
+    tables.edges.add_row(child=n.A, parent=n.E, left=0, right=bp[0])
+    tables.edges.add_row(child=n.A, parent=n.F, left=bp[0], right=l)
+    tables.edges.add_row(child=n.B, parent=n.G, left=0, right=l)
+
+    tables.individuals[1] = tables.individuals[1].replace(parents=[2, 3])
[…]

### 2022-02-21 illustrations/ARG_recomb_node_deletion.py
Remove ported illustration scripts.

@@ -1,95 +1,0 @@
-import collections
-import io
-import itertools
-import sys
-import string
-from pathlib import Path
-
-import matplotlib as mpl
-import matplotlib.pyplot as plt
-import numpy as np
-import tskit
-
-current_dir = Path(__file__).parent
-
-# useful modules in the top level dir: we should be able to import this if the script is
-# run from the top level dir, but for some reason it fails, so we just add the path here
-sys.path.append(str(current_dir.parent.absolute()))
-import argutils
-
-outfile = "ARG_recomb_node_deletion"
-
-def arg_node_simplification():
-    seed = 372
-    t_x={
-        9: 3, 6: -8, 18: 50, 19: -15, 17: 5,
-        13: 35, 14: -23, 10: -10, 15: 5, 16: -5,
-        7: -5, 11: 5, 8: 8}
-
-    ts = argutils.sim_wright_fisher(2, 10, 100, recomb_proba=0.1, seed=seed)
-    labels = {i: string.ascii_uppercase[i] for i in range(len(string.ascii_uppercase))}
-    # relabel the nodes to get samples reading A B C D
-    labels.update({2: 'B', 3: 'C', 1: "D", 14: "N", 13: "O"})
-    ts = argutils.viz.label_nodes(ts, labels)
-    
-    fig, (ax1, ax2, ax3, ax4) = plt.subplots(1, 4, figsize=(12,5))
-    ax1.set_title("(a) little ARG")
-    ax2.set_title("(b) remove pass through")
-    ax3.set_title("(c) simplify (keep unary in coal)")
-    ax4.set_title("(d) fully simplified")
-    col = mpl.colors.to_hex(plt.cm.tab20(1))
-    pos, G = argutils.viz.draw(
-        ts, ax1,
-        draw_edge_widths=True,
-        use_ranked_times=True, node_color=col, tweak_x=t_x)
-    ts2, node_map = argutils.simplify_remove_pass_through(ts, repeat=True, map_nodes=True)
-    argutils.viz.draw(
-        ts2, ax2,
-        draw_edge_widths=True,
-        pos={node_map[i]: p for i, p in pos.items()}, node_color=col)
-    ts3, node_map = argutils.simplify_keeping_unary_in_coal(ts, map_nodes=True)
-    argutils.viz.draw(
-        ts3, ax3,
-        draw_edge_widths=True,
[…]

### 2022-02-21 illustrations.py
Remove generated SVGs
The change depending on local versions - remove from git for now
until the figs settle down.

@@ -129,12 +129,12 @@ def arg_edge_annotations():
 
     # From https://networkx.org/documentation/stable/auto_examples/drawing/plot_custom_node_icons.html
     icons = {
-        "genome_empty": "illustrations/node_icons/genome_empty.png",
-        "genome_empty_hamburger": "illustrations/node_icons/genome_empty_hamburger.png",
-        "genome_full": "illustrations/node_icons/genome_full.png",
+        "genome_empty": "illustrations/assets/genome_empty.png",
+        "genome_empty_hamburger": "illustrations/assets/genome_empty_hamburger.png",
+        "genome_full": "illustrations/assets/genome_full.png",
     }
     for letter in list(string.ascii_uppercase[3 : G.number_of_nodes()]):
-        icons["genome_" + letter] = "illustrations/node_icons/genome_" + letter + ".png"
+        icons["genome_" + letter] = "illustrations/assets/genome_" + letter + ".png"
     # Load images
     images = {k: PIL.Image.open(fname) for k, fname in icons.items()}
 
@@ -317,7 +317,7 @@ def pedigree():
     The ARG as embedded in diploid pedigree.
     """
 
-    pedigree_svg = pathlib.Path("illustrations/pedigree.svg").read_text()
+    pedigree_svg = pathlib.Path("illustrations/assets/pedigree.svg").read_text()
     pedigree_svg = pedigree_svg[pedigree_svg.find("<svg") :]
 
     n = SimpleNamespace()  # convenience labels

### 2022-02-21 paper.tex
Rename figure 3, arg-in-pedigree

@@
 \begin{center}
-    \includegraphics[width=\textwidth]{illustrations/pedigree_figure}
+    \includegraphics[width=\textwidth]{illustrations/arg-in-pedigree}
 \end{center}
-\caption{\label{fig-pedigree-and-arg}
+\caption{\label{fig-arg-in-pedigree}
 % TODO caption needs another pass. One the full section has been written
@@
 structure a ``Genome ARG'', or GARG.
-Fig~\ref{fig-pedigree-and-arg} illustrates the basic elements of a GARG.
+Fig~\ref{fig-arg-in-pedigree} illustrates the basic elements of a GARG.
 The pedigree is composed of a set of individuals who each carry
@@
 be the recombined product of that parent's two genomes.
-For example, individual $I_1$ in Fig.~\ref{fig-pedigree-and-arg}A
+For example, individual $I_1$ in Fig.~\ref{fig-arg-in-pedigree}A
 has two genomes $A$ and $B$,
@@
 individuals. This ``genome-perspective'' is show in
-Fig.~\ref{fig-pedigree-and-arg}B, which depicts the same graph
+Fig.~\ref{fig-arg-in-pedigree}B, which depicts the same graph
 of the ancestry of the sample genomes $A$--$D$. The graph
@@
 edges then uniquely defines the trees at every position
-along the genome (Fig.~\ref{fig-pedigree-and-arg}C).
+along the genome (Fig.~\ref{fig-arg-in-pedigree}C).
 
@@
 samples. For example, node $H$ is the direct ancestor of $D$ in
-Fig~\ref{fig-pedigree-and-arg}, and is not the product of either a
+Fig~\ref{fig-arg-in-pedigree}, and is not the product of either a
 coalescence or a recombination. Such nodes would usually be removed
@@
 intervals (representing the precise inheritance patterns).
-For example, in Fig~\ref{fig-pedigree-and-arg}, node $A$ is the
+For example, in Fig~\ref{fig-arg-in-pedigree}, node $A$ is the
 recombinant descendent of $E$ and $F$, inheriting positions

### 2022-02-21 illustrations.py

@@ -312,7 +312,7 @@ def arg_node_simplification():
 
 
 @click.command()
-def pedigree():
+def arg_in_pedigree():
     """
     The ARG as embedded in diploid pedigree.
     """
@@ -408,13 +408,13 @@ def pedigree():
         '"http://www.w3.org/Graphics/SVG/1.1/DTD/svg11.dtd">\n'
     )
     top_svg += "\n".join(svg)
-    with open("illustrations/pedigree_figure.svg", "wt") as f:
+    with open("illustrations/arg-in-pedigree.svg", "wt") as f:
         f.write(top_svg)
 
 
 cli.add_command(arg_edge_annotations)
 cli.add_command(arg_node_simplification)
-cli.add_command(pedigree)
+cli.add_command(arg_in_pedigree)
 
 if __name__ == "__main__":
     cli()

### 2022-02-21 illustrations.py
Rename the simplification figure

@@ -208,7 +208,7 @@ def arg_edge_annotations():
 
 
 @click.command()
-def arg_node_simplification():
+def simplification():
     """
     Sequentially simplifying a WF simulation.
     """
@@ -300,14 +300,12 @@ def arg_node_simplification():
     svg.append('<g transform="translate(0 400) scale(0.83)">' + tree_svg + "</g>")
     svg.append("</g></svg>")
 
-    outfile = "ARG_recomb_node_deletion"
-
     top_svg = (
         '<!DOCTYPE svg PUBLIC "-//W3C//DTD SVG 1.1//EN" '
         '"http://www.w3.org/Graphics/SVG/1.1/DTD/svg11.dtd">\n'
     )
     top_svg += "\n".join(svg)
-    with open(f"illustrations/{outfile}.svg", "wt") as f:
+    with open(f"illustrations/simplification.svg", "wt") as f:
         f.write(top_svg)
 
 
@@ -413,7 +411,7 @@ def arg_in_pedigree():
 
 
 cli.add_command(arg_edge_annotations)
-cli.add_command(arg_node_simplification)
+cli.add_command(simplification)
 cli.add_command(arg_in_pedigree)
 
 if __name__ == "__main__":

### 2022-02-21 illustrations.py
Rename ancestry-resolution figure

@@ -19,9 +19,9 @@ def cli():
 
 
 @click.command()
-def arg_edge_annotations():
+def ancestry_resolution():
     """
-    Ancestry propagation via WH99 graph.
+    Ancestry resolution on the WH99 graph.
     """
 
     def add_edge_labels(ax, ts, G, pos):
@@ -202,8 +202,7 @@ def arg_edge_annotations():
         '"http://www.w3.org/Graphics/SVG/1.1/DTD/svg11.dtd">\n'
     )
     top_svg += "\n".join(svg)
-    outfile = "ARG_edge_annotations"
-    with open(f"illustrations/{outfile}.svg", "wt") as f:
+    with open(f"illustrations/ancestry-resolution.svg", "wt") as f:
         f.write(top_svg)
 
 
@@ -410,9 +409,9 @@ def arg_in_pedigree():
         f.write(top_svg)
 
 
-cli.add_command(arg_edge_annotations)
-cli.add_command(simplification)
 cli.add_command(arg_in_pedigree)
+cli.add_command(ancestry_resolution)
+cli.add_command(simplification)
 
 if __name__ == "__main__":
     cli()

### 2022-02-21 argutils/__init__.py
Work on argweaver impl

@@ -3,3 +3,4 @@ Various utilities for working with args.
 """
 from . ancestry import *
 from . viz import draw
+from . convert import *

### 2022-02-21 argutils/tests.py

@@ -494,3 +494,10 @@ class TestIntervalSet:
         assert s2.is_subset(s1)
         assert not s1.is_subset(s2)
 
+
+class TestConvertArgweaver:
+
+    def test_example(self):
+        ts = argutils.convert_argweaver(open("examples/argweaver.arg"))
+        print(ts)
+

### 2022-02-21 argutils/convert.py
Add the convert module.

@@ -1,0 +1,161 @@
+import collections
+
+import pandas as pd
+import networkx as nx
+import tskit
+
+
+def convert_argweaver(infile, epsilon_scale=1e-5):
+    """
+    Convert an ARGweaver .arg file to a tree sequence. An example .arg file is at
+
+    https://github.com/CshlSiepelLab/argweaver/blob/master/test/data/test_trans/0.arg
+
+    Times are adjusted so that parents are older than children by a small amount
+    `epsilon` which is calculated by taking the largest number of nodes at a single time
+    and dividing the smalles gap between node times by that value, then multiplying by
+    a small value given by `epsilon_scale`.
+    """
+    start, end = next(infile).strip().split()
+    assert start.startswith("start=")
+    start = int(start[len("start=") :])
+    assert end.startswith("end=")
+    end = int(end[len("end=") :])
+    # the "name" field can be a string. Force it to be so, in case it is just numbers
+    df = pd.read_csv(infile, header=0, sep="\t", dtype={"name": str, "parents": str})
+
+    name_to_record = {}
+    for _, row in df.iterrows():
+        row = dict(row)
+        name_to_record[row["name"]] = row
+
+    # Make an nx DiGraph so we can do a topological sort.
+    G = nx.DiGraph()
+    for row in name_to_record.values():
+        child = row["name"]
+        parents = row["parents"]
+        G.add_node(child)
+        if isinstance(parents, str):
+            for parent in row["parents"].split(","):
+                G.add_edge(child, parent)
+
+    tables = tskit.TableCollection(sequence_length=end)
+    tables.nodes.metadata_schema = tskit.MetadataSchema.permissive_json()
+    breakpoints = np.full(len(G), tables.sequence_length)
+    for node in nx.topological_sort(G):
+        record = name_to_record[node]
+        flags = 0
+        if node.startswith("n"):
+            flags = tskit.NODE_IS_SAMPLE
+            assert record["age"] == 0
+            assert record["event"] == "gene"
+            time = 0
+        else:
+            # Use topological sort order for age for the moment.
+            time += 1
+        num_nodes = 2 if record["event"] == "recomb" else 1
+        for _ in range(num_nodes):
+            tsk_node = tables.nodes.add_row(flags=flags, time=time, metadata=record)
+            aw_to_tsk_node[node].append(tsk_node)
+
+    for record in name_to_record.values():
[…]

### 2022-02-21 argutils/__main__.py
Add conversion function for argweaver

@@ -31,7 +31,7 @@ def simulate(num_samples, sequence_length, rho, seed, output):
 def draw(input, output):
     ts = tskit.load(input)
     fig, ax1 = plt.subplots(1, 1, figsize=(5, 6))
-    argutils.draw(ts, ax1)
+    argutils.draw(ts, ax1, draw_edge_widths=True)
     plt.savefig(output)
 
 

### 2022-02-21 argutils/ancestry.py

@@ -1,6 +1,8 @@
 """
 Utilities for generating and converting ARGs in various formats.
 """
+from __future__ import annotations
+
 import collections
 import random
 import math
@@ -674,6 +676,9 @@ def as_resolved_garg(ts):
     return Ep
 
 
+# TODO Need to be clearer here - what are the assumptions about how things
+# are encoded in the tree sequence? We should probably call this
+# as_minimial_arg or something.
 def as_earg(ts):
     """
     Returns the specified unresolved ARG as an EARG (E, sigma).
@@ -737,3 +742,67 @@ def garg_get_tree(E, S, x):
             parent[c] = P[0]
             N.add(P[0])
     return parent
+
+
+@dataclasses.dataclass
+class RecombinationEvent:
+    parent_edges: List[tskit.Edge] = dataclasses.field(default_factory=list)
+    child_edge: tskit.Edge | None = None
+
+
+def earg_to_garg(ts):
+    """
+    Given an EARG encoded in the specified tree sequence, return the
+    corresponding GARG in which we create two parental nodes
+    instead of a single recombination event.
+    """
+    # TODO need to resolve what we mean by the terminology here, the
+    # ts EARG is different to what we're working with above.
+
+    L = ts.sequence_length
+    # Map from recombination nodes to the RecombinationEvent objects.
+    recombination_events = collections.defaultdict(RecombinationEvent)
+
+    for edge in ts.edges():
+        if edge.left != 0 and edge.right != ts.sequence_length:
+            raise ValueError("only edges like (0, x], (x, L] or (0, L]) are allowed")
+        if edge.left != 0 or edge.right != ts.sequence_length:
+            recombination_events[edge.child].parent_edges.append(edge)
+    for edge in ts.edges():
+        if edge.parent in recombination_events:
+            event = recombination_events[edge.parent]
+            assert event.child_edge == None
+            event.child_edge = edge
+
+    # print(recombination_events)
+    tables = ts.dump_tables()
+    tables.edges.clear()
+    for edge in ts.edges():
+        if (
+            edge.child not in recombination_events
+            and edge.parent not in recombination_events
+        ):
+            tables.edges.add_row(edge.left, edge.right, edge.parent, edge.child)
+
+    for u, event in recombination_events.items():
+        assert len(event.parent_edges) == 2
+        pe1, pe2 = event.parent_edges
+        # Rewrite the edges for the first parent, reusing the existing node.
[…]

### 2022-02-21 argutils/convert.py

@@ -2,19 +2,17 @@ import collections
 
 import pandas as pd
 import networkx as nx
+import numpy as np
 import tskit
 
+import argutils
 
-def convert_argweaver(infile, epsilon_scale=1e-5):
+
+def convert_argweaver(infile):
     """
     Convert an ARGweaver .arg file to a tree sequence. An example .arg file is at
 
     https://github.com/CshlSiepelLab/argweaver/blob/master/test/data/test_trans/0.arg
-
-    Times are adjusted so that parents are older than children by a small amount
-    `epsilon` which is calculated by taking the largest number of nodes at a single time
-    and dividing the smalles gap between node times by that value, then multiplying by
-    a small value given by `epsilon_scale`.
     """
     start, end = next(infile).strip().split()
     assert start.startswith("start=")
@@ -28,6 +26,8 @@ def convert_argweaver(infile, epsilon_scale=1e-5):
     for _, row in df.iterrows():
         row = dict(row)
         name_to_record[row["name"]] = row
+    # We could use nx to do this, but we want to be sure the order is correct.
+    parent_map = collections.defaultdict(list)
 
     # Make an nx DiGraph so we can do a topological sort.
     G = nx.DiGraph()
@@ -38,10 +38,12 @@ def convert_argweaver(infile, epsilon_scale=1e-5):
         if isinstance(parents, str):
             for parent in row["parents"].split(","):
                 G.add_edge(child, parent)
+                parent_map[child].append(parent)
 
     tables = tskit.TableCollection(sequence_length=end)
     tables.nodes.metadata_schema = tskit.MetadataSchema.permissive_json()
     breakpoints = np.full(len(G), tables.sequence_length)
+    aw_to_tsk_id = {}
     for node in nx.topological_sort(G):
         record = name_to_record[node]
         flags = 0
@@ -53,109 +55,29 @@ def convert_argweaver(infile, epsilon_scale=1e-5):
         else:
             # Use topological sort order for age for the moment.
             time += 1
-        num_nodes = 2 if record["event"] == "recomb" else 1
-        for _ in range(num_nodes):
-            tsk_node = tables.nodes.add_row(flags=flags, time=time, metadata=record)
-            aw_to_tsk_node[node].append(tsk_node)
+        tsk_id = tables.nodes.add_row(flags=flags, time=time, metadata=record)
+        aw_to_tsk_id[node] = tsk_id
[…]

### 2022-02-21 argutils/tests.py

@@ -201,6 +201,7 @@ def assert_gargs_equal(E1, E2):
         # print(key, I1, I2)
         assert I1 == I2
 
+
 class TestResolvedGARG:
     @pytest.mark.parametrize("seed", range(1, 3))
     @pytest.mark.parametrize(
@@ -351,6 +352,40 @@ def gmrca_example_unresolved():
     return tables.tree_sequence()
 
 
+class TestEARGtoGARG:
+    def test_example(self):
+        # 3.00┊     4   ┊
+        #     ┊  ┏━━┻━┓ ┊
+        # 2.00┊  3    ┃ ┊
+        #     ┊ ┏┻┓   ┃ ┊
+        # 1.00┊ ┃ ┗━2━┛ ┊
+        #     ┊ ┃   ┃   ┊
+        # 0.00┊ 0   1   ┊
+        #
+        tables = tskit.TableCollection(2)
+        tables.nodes.add_row(flags=tskit.NODE_IS_SAMPLE, time=0)
+        tables.nodes.add_row(flags=tskit.NODE_IS_SAMPLE, time=0)
+        tables.nodes.add_row(time=1)
+        tables.nodes.add_row(time=2)
+        tables.nodes.add_row(time=3)
+
+        tables.edges.add_row(0, 2, 2, 1)
+        tables.edges.add_row(0, 2, 3, 0)
+        tables.edges.add_row(0, 2, 4, 3)
+        tables.edges.add_row(0, 1, 3, 2)
+        tables.edges.add_row(1, 2, 4, 2)
+
+        tables.sort()
+        ts = tables.tree_sequence()
+        ts2 = argutils.earg_to_garg(ts)
+        # Can't easily compare the unresolved version because the node IDs
+        # differ.
+        ts3 = ts2.simplify(keep_unary=True)
+        ts3.tables.assert_equals(
+            gmrca_example_resolved().tables, ignore_provenance=True
+        )
+
+
 class TestResolve:
     @pytest.mark.parametrize(
         ["unresolved", "resolved"],
@@ -496,8 +531,12 @@ class TestIntervalSet:
 
 
 class TestConvertArgweaver:
-
     def test_example(self):
-        ts = argutils.convert_argweaver(open("examples/argweaver.arg"))
-        print(ts)
-
+        with open("examples/argweaver.arg") as f:
+            ts = argutils.convert_argweaver(f)
+            print(ts.tables)
+            ts.dump("argweaver_example.trees")
[…]

### 2022-02-24 paper.tex
Rename the simplification figure
Change fig 2 to compare node-vs-edge annotations

@@
 \centering
-\begin{tabular}{cc}
+% \begin{tabular}{cc}
 \begin{tikzpicture}[x=5mm, y=5mm, node distance=2mm and 20mm]
@@
 
-\node [nodelabel,anchor=north west] at ($(-6.5,5)$) {
+\node [nodelabel,anchor=north west] at ($(-8.5,5)$) {
 \begin{tabular}{c|c}
-\multicolumn{2}{c}{Breakpoint}\\
+% \multicolumn{2}{c}{Breakpoint}\\
+node & breakpoint\\
 \hline
-0 & $\infty$ \\
-1 & $\infty$ \\
-2 & $\infty$ \\
+0 & $\varnothing$ \\
+1 & $\varnothing$ \\
+2 & $\varnothing$ \\
 3 & $x$ \\
-4 & $\infty$ \\
-5 & $\infty$ \\
-6 & $\infty$ \\
+4 & $\varnothing$ \\
+5 & $\varnothing$ \\
+6 & $\varnothing$ \\
 \end{tabular}};
@@
 
-\end{tikzpicture}
-&
-\begin{tikzpicture}[x=5mm,y=5mm,node distance=2mm and 20mm]
-\tikzset{greynode/.style={circle,fill,inner sep=1},
-nodelabel/.style={font=\footnotesize}}
-
-\node (s0) [greynode] at (0, 0) {};
-\node (s1) [greynode] at (3, 0) {};
-\node (s2) [greynode] at (6, 0) {};
-\node (s3) [greynode] at (2, 1) {};
-\node (s4) [greynode] at (4, 1) {};
-\node (s5) [greynode] at (1, 2) {};
-\node (s6) [greynode] at (5, 3) {};
-\node (s7) [greynode] at (3, 4) {};
-
-\foreach \u/\lab in {s0/0, s1/1, s2/2} \node[nodelabel,anchor=north] at (\u) {\lab};
-\foreach \u/\lab in {s3/3} \node[nodelabel,anchor=east] at (\u) {\lab};
-\foreach \u/\lab in {s4/4} \node[nodelabel,anchor=west] at (\u) {\lab};
-\foreach \u/\lab in {s5/5} \node[nodelabel,anchor=south west] at (\u) {\lab};
-\foreach \u/\lab in {s6/6} \node[nodelabel,anchor=south east] at (\u) {\lab};
-\foreach \u/\lab in {s7/7} \node[nodelabel,anchor=south] at (\u) {\lab};
-
-%% Edges
-\draw (s1) -| (s3);
-\draw (s1) -| (s4) |- (s6);
-\draw (s3) |- (s5);
-\draw (s0) |- (s5);
-\draw (s2) |- (s6);
-\draw (s5) |- (s7);
-\draw (s6) |- (s7);
-
-\node [nodelabel,anchor=north west] at ($(-5,5)$) {
+\node [nodelabel,anchor=north west] at ($(6,5)$) {
 \begin{tabular}{ll}
@@
 \hline
-$(0, 5)$ & $(0, L]$ \\
-$(1, 3)$ & $(0, x]$ \\
-$(1, 4)$ & $(x, L]$ \\
-$(2, 6)$ & $(0, L]$ \\
-$(3, 5)$ & $(0, L]$ \\
+$(0, 4)$ & $(0, L]$ \\
+$(1, 3)$ & $(0, L]$ \\
+$(2, 5)$ & $(0, L]$ \\
+$(3, 4)$ & $(0, x]$ \\
+$(3, 5)$ & $(x, L]$ \\
 $(4, 6)$ & $(0, L]$ \\
-$(5, 7)$ & $(0, L]$ \\
-$(6, 7)$ & $(0, L]$ \\
+$(5, 6)$ & $(0, L]$ \\
 \end{tabular}};
@@
 \end{tikzpicture}
-\\
-(A) & (B) \\
-\end{tabular}
 \caption{\label{fig-arg-data-structure}
-Encoding an ARG. (A) The classical approach described by Griffiths
+Node vs edge breakpoint annotations.
+The classical approach described by Griffiths
 equates graph nodes with events in the stochastic process,
-and associates recombination breakpoints with nodes. This encoding
-is limited in the ancestral patterns that can be described and ambiguous.
+and associates recombination breakpoint information with \emph{nodes} (left).
+This encoding
+is limited in the pattern of inheritance that can be described
+and requires that edges be ordered to avoid ambiguity.
 [TODO clarify that the function is the same things as having two
 types of nodes]
-(B) Equating graph nodes with \emph{genomes} and attaching the
-interval of genome inherited by the child node with edges removes
-this ambiguity and can describe any pattern of genetic inheritance.
-The most important change between the two encodings is to represent
-a recombination event (node 3 in A) by two parent genomes
-(nodes 3 and 4 in B) and explicitly storing the interval
-that the child inherits from each parent along the two corresponding
-edges. Thus, child $1$ in inherits $(0, x]$ from $3$
-and $(x, L)$ from $4$
-(for recombination breakpoint $x$ and a sequence of length $L$).
-\textcolor{red}{This is very much a draft; let's see how it works with
- the text and other figures and tidy it up in a later revision}
+The alternative is to associate breakpoint annotations
+with \emph{edges} (right).
+Associating the interval of genome inherited by the child node
+from the parent with the corresponding edge, allows
+any pattern genetic inheritance to be described and does
+not required edges to be ordered.
 }
@@
 \centering
-\includegraphics[width=\textwidth]{illustrations/ARG_edge_annotations}
+\includegraphics[width=\textwidth]{illustrations/ancestry-resolution}
 \caption{\label{fig-ancestry-resolution}
@@
 \vspace{5em}
-\includegraphics[width=\linewidth]{illustrations/ARG_recomb_node_deletion}
+\includegraphics[width=\linewidth]{illustrations/simplification}
 \caption{\label{fig-simplification}

### 2022-02-28 paper.tex

@@
 This encoding
-is limited in the pattern of inheritance that can be described
+is limited in the patterns of inheritance that can be described
 and requires that edges be ordered to avoid ambiguity.
@@
 Associating the interval of genome inherited by the child node
-from the parent with the corresponding edge, allows
+from the parent with the corresponding edge allows
 any pattern genetic inheritance to be described and does
-not required edges to be ordered.
+not require edges to be ordered.
 }

### 2022-02-28 paper.tex

@@
 \citep[e.g.][]{griffiths1996ancestral,kuhner2000maximum, nielsen2000estimation,
-fearnhead2001estimating}. These methods met with limited success
+fearnhead2001estimating}.
+% FIXME: need some proper statistical language here to describe why these
+% methods didn't work so well.
+These methods met with limited success
 due to the overwhelmingly large and
-horribly structured [Jere: what's the right way to say this?]
-state space of ARGs. The focus subsequently shifted to
-the more tractable---but still
-NP-hard~\citep{wang2001perfect}---problem of computing
+horribly structured [Jere: what's the right way to say this?] state space of ARGs.
+The focus subsequently shifted to the more
+tractable---but still NP-hard~\citep{wang2001perfect}---problem of computing
 the minimum number of recombinations required
@@
 The distinction between the ARG as a stochastic process
-and the ARG as a data structure became less clear at this point,
-with \citet{minichiello2006mapping} explicitly arguing for
-the data structure interpretation.
+and the ARG as a \emph{data structure}~\citep{minichiello2006mapping}
+became important at this point,
+since parsimonious ARGs have no immediate connection to
+the stochastic process. Indeed, the ARG instances produced by
+the coalescent with recombination are usually highly unparsimonious,
+since they contain many invisible recombinations
+(although those produced under the SMC may be parsimonious, depending
+on the pattern of mutations that is present).
 
@@
 The Griffiths ARG encoding is an elegant and mathematically
-economical description of the coalescent with recombination on a single
-chromosome, but it has significant drawbacks as the basis for a general purpose
+economical description of the coalescent with recombination,
+but it has significant drawbacks as the basis for a general purpose
 storage and interchange format. To avoid confusion, we will refer to this
 ARG encoding as an ``Event ARG'', or EARG.
-The basic unit of an EARG are \emph{events} and correspond directly
-to the events that
+The fundamental units of an EARG are the \emph{events} that correspond directly
+to those that
 occur in the ARG stochastic process described in the previous section.
@@
 are modified by the events that they pass through.
-
-% What do we mean by an EARG, formally?
-% NOTE we're using the natural numbers but are using 0 as a node in the
-% examples. Probably want to clear this up a bit.
-Formally, we can define an EARG as a tuple $(e, \sigma)$, where $e$
-defines a directed acyclic graph where $e_j = (c, p)$ describes
-an edge between
-child node $c$ and parent $p$ ($c, p \in \mathbb{N}$).
-Additionally,
-$\sigma: \mathbb{N} \rightarrow \mathbb{N}$
-is a function mapping nodes to recombination breakpoints, such that $\sigma(u) = x$
-if $u$ is a recombination event with breakpoint $x$ and
-$\sigma(u) = \infty$ otherwise. (The use of $\infty$ to denote
-non-recombination nodes is arbitrary, but useful when we extract
-% NOTE: need to be careful about using "local" trees if we're also
-% distinguishing resolved and non-resolved GARGs as being "locally"
-% defined or not. One is topologically local and the other is local
-% along the genome.
-local trees; see below.)
-For each recombination node $u$ we must
-have exactly two edges in $E$ such that $u$ is the child.
-Note that this description captures only the
-graph topology: if we also wish to know the times of events we need
-an additional function $t(u): \mathbb{N} \rightarrow \mathbb{R}$
-which defines the time of each node.
-An example is given in Fig.~\ref{fig-arg-data-structure}A.
 
@@
 and requires that edges be ordered to avoid ambiguity.
-[TODO clarify that the function is the same things as having two
-types of nodes]
 The alternative is to associate breakpoint annotations
@@
 \end{figure}
+
+% What do we mean by an EARG, formally?
+% NOTE we're using the natural numbers but are using 0 as a node in the
+% examples. Probably want to clear this up a bit.
+Formally, we can define an EARG as a tuple $(e, \sigma)$, where $e$
+is an ordered list of edges defining a directed acyclic graph and
+$\sigma: \mathbb{N} \rightarrow \mathbb{N}$
+is a function mapping nodes to recombination breakpoints,
+such that $\sigma(u) = x$
+if $u$ is a recombination event with breakpoint $x$ and
+$\sigma(u) = \varnothing$ otherwise.
+(Note that this formulation is equivalent to stipulating two
+different ``types'' of node and only storing breakpoints
+with recombination nodes).
+Each edge $e_j = (c, p)$ describes an edge between
+child node $c$ and parent $p$, where $c, p \in \mathbb{N}$.
+For each recombination node $u$ we must
+have exactly two edges in $e$ such that $u$ is the child.
+Note that this description captures only the
+graph topology: if we also wish to know the times of events we need
+an additional function $t(u): \mathbb{N} \rightarrow \mathbb{R}$
+which defines the time of each node.
+An example is given in Fig.~\ref{fig-arg-data-structure}A.
 
@@
 (\cite{griffiths1997ancestral} refer to the ``left'' and ``right''
-parents of a recombination event). When we want to recover
-the local tree at a position $x$ along the genome,
+parents of a recombination event). This is usually described
+in terms of the process in which we extract the local trees
+along the genome from an EARG.
+To recover
+the tree at genome coordinate $x$,
 we traverse the graph upwards from the leaves. At a particular
@@
 node and we simply follow that parent. If we are at a
-recombination node, $u$ has two parents; if
+recombination node, $u$ has two parents: if
 $x$ is less than the breakpoint $\sigma(u)$ we follow
-the first edge, and otherwise follow the second edge.
-This ordering requirement, which although straightforward
-to describe, has some drawbacks in practice. For example,
+the edge to the first parent, and otherwise follow the edge
+to the second parent.
+This ordering requirement, which is straightforward
+to describe, has some practical drawbacks. For example,
 % Surely the meaning is clear from the context here and we don't need
@@
 Fig.~\ref{fig-simplification}B shows the result of this type of
-graph-level simplification.
+graph topology simplification.
 

### 2022-03-01 paper.tex
Update refs

@@
 methods~\citep{rasmussen2014genome,kelleher2019inferring,speidel2019method,
-schaefer2021ancestral,wohns2021unified}
+schaefer2021ancestral,wohns2022unified}
 and data representation~\citep{kelleher2016efficient}
@@
 substantially simplifies this process. Thirdly, computing
-[FIXME describe what we did in tsdate]~\citep{wohns2021unified}.
+[FIXME describe what we did in tsdate]~\citep{wohns2022unified}.
 
@@
 implementing the Li and Stephens
-model~\citep{kelleher2019inferring,wohns2021unified}
+model~\citep{kelleher2019inferring,wohns2022unified}
 and likely many more.

### 2022-03-01 paper.tex
Small update to EARG section

@@
 Fortunately, it is straightforward to generalise the classical Griffiths
-encoding to both simplify implementations and to encompass any
+EARG encoding to both simplify implementations and to encompass any
 possible pattern of genetic inheritance.
-We require only two changes. Firstly, we let nodes in the
-graph model \emph{genomes} rather than \emph{events}. As we
-see in the next section, this allows us to model complex
-patterns of inheritance without classifying different types of node.
-The second change we must make is to associate
+The only change that we require is to associate
 the details of inheritance intervals with \emph{edges} rather than
 as breakpoints associated with recombination events and
-edge ordering rules.
-% That is, given two genomes $p$
-% and $c$ where $p$ is the parent genome and $c$ is the child
-% (although many generations may have elapsed between the two),
-% we associate a set $I$ of intervals of genome which $c$ has
-% inherited from $p$.
-These changes are straightforward, involve no loss of information
-and greatly increase both the expressivity and computational
+edge ordering rules. This is illustrated in Fig~\ref{fig-arg-data-structure}B
+where this alternative edge annotation formulation is
+shown. By associating the interval $(0, x)$ with edge $(3,4)$
+and the interval $(x, L)$ with $(3, 5)$ we make the encoding
+unambigous: we can list the edges in any order, and it
+will describe the same set of local genealogical trees.
+This approach is also more flexible, because any pattern
+of inheritance can be described by these intervals. For
+example, if a gene conversion event occured where a segment
+between between positions $x$ and $y$ was inherited by
+node $3$ from $5$, we would state this by associating
+the intervals $\{(0, x], (y, L]\}$ with $(3,4)$ and
+$\{(x, y]\}$ with $(3,5)$.
+This change is straightforward, involves no loss of information
+and (as we see in later sections)
+greatly increases both the expressivity and computational
 power of ancestral recombination graph data structures.

### 2022-03-01 illustrations.py
Update fig 3 to greyscale
And add in labels to 3b

@@ -313,12 +313,40 @@ def arg_in_pedigree():
     """
     The ARG as embedded in diploid pedigree.
     """
+    def add_edge_labels(ax, ts, G, pos, n):
+        params = {
+            "G": G,
+            "pos": pos,
+            "ax": ax,
+            "rotate": False,
+            "font_size": 12,
+            "bbox": dict(
+                boxstyle="round,pad=0.1",
+                ec=(1.0, 1.0, 1.0),
+                fc=(1.0, 1.0, 1.0)),
+        }
+        edge_labels={(e.child, e.parent): f"({e.left:.0f},{e.right:.0f}]" for e in ts.edges()}
+        full_edges = {k: v for k, v in edge_labels.items() if not argutils.is_recombinant(ts.node(k[0]).flags)}
+        nx.draw_networkx_edge_labels(
+            **params, font_weight="normal", alpha=0.5, edge_labels=full_edges, label_pos=0.6)
+        lab = {k: v for k, v in edge_labels.items() if k[0]==n.A and k[1]==n.E}
+        nx.draw_networkx_edge_labels(
+            **params, font_weight="bold", edge_labels=lab, label_pos=0.5)
+        lab = {k: v for k, v in edge_labels.items() if k[0]==n.A and k[1]==n.F}
+        nx.draw_networkx_edge_labels(
+            **params, font_weight="bold", edge_labels=lab, label_pos=0.7)
+        lab = {k: v for k, v in edge_labels.items() if k[0]==n.C and k[1]==n.E}
+        nx.draw_networkx_edge_labels(
+            **params, font_weight="bold", edge_labels=lab, label_pos=0.3)
+        lab = {k: v for k, v in edge_labels.items() if k[0]==n.C and k[1]==n.F}
+        nx.draw_networkx_edge_labels(
+            **params, font_weight="bold", edge_labels=lab, label_pos=0.5)
 
     pedigree_svg = pathlib.Path("illustrations/assets/pedigree.svg").read_text()
     pedigree_svg = pedigree_svg[pedigree_svg.find("<svg") :]
 
     n = SimpleNamespace()  # convenience labels
-    l = 100
+    l = 10
     tables = tskit.TableCollection(sequence_length=l)
     tables.nodes.metadata_schema = tskit.MetadataSchema.permissive_json()
     for gen in range(4):
@@ -332,14 +360,19 @@ def arg_in_pedigree():
                     "genome": "paternal" if genome == 0 else "maternal",
[…]

### 2022-03-01 illustrations.py
Add labels to fig 3a

@@ -321,7 +321,7 @@ def arg_in_pedigree():
             "rotate": False,
             "font_size": 12,
             "bbox": dict(
-                boxstyle="round,pad=0.1",
+                boxstyle="round,pad=0.05",
                 ec=(1.0, 1.0, 1.0),
                 fc=(1.0, 1.0, 1.0)),
         }
@@ -331,16 +331,16 @@ def arg_in_pedigree():
             **params, font_weight="normal", alpha=0.5, edge_labels=full_edges, label_pos=0.6)
         lab = {k: v for k, v in edge_labels.items() if k[0]==n.A and k[1]==n.E}
         nx.draw_networkx_edge_labels(
-            **params, font_weight="bold", edge_labels=lab, label_pos=0.5)
+            **params, font_weight="bold", edge_labels=lab, label_pos=0.35)
         lab = {k: v for k, v in edge_labels.items() if k[0]==n.A and k[1]==n.F}
         nx.draw_networkx_edge_labels(
-            **params, font_weight="bold", edge_labels=lab, label_pos=0.7)
+            **params, font_weight="bold", edge_labels=lab, label_pos=0.3)
         lab = {k: v for k, v in edge_labels.items() if k[0]==n.C and k[1]==n.E}
         nx.draw_networkx_edge_labels(
-            **params, font_weight="bold", edge_labels=lab, label_pos=0.3)
+            **params, font_weight="bold", edge_labels=lab, label_pos=0.7)
         lab = {k: v for k, v in edge_labels.items() if k[0]==n.C and k[1]==n.F}
         nx.draw_networkx_edge_labels(
-            **params, font_weight="bold", edge_labels=lab, label_pos=0.5)
+            **params, font_weight="bold", edge_labels=lab, label_pos=0.55)
 
     pedigree_svg = pathlib.Path("illustrations/assets/pedigree.svg").read_text()
     pedigree_svg = pedigree_svg[pedigree_svg.find("<svg") :]

### 2022-03-02 paper.tex
Mention optional 3rd change to make a 2-recombination node ARG
Fixes #128 with the rough text Jerome suggested.
Changes for fig3 caption

@@
 % it'll be clearer what we need to say in the caption vs the text.
-The Ancestral Recombination Graph defined as arbitrary inheritance
-of genomes between diploid (or any ploidy; see text) individuals.
-(A) Each individual consists of two genomes, and each genome is
-inherited as a (potentially) recombinant mixture of the individual's parent's
-genomes. (B) Genomes are modelled as nodes in a graph, and the edges
-denote inheritance. Each edge is annotated with genome coordinates
-defining the specific region(s) inherited.
-(C) This description of inheritance fully defines the
+The Ancestral Recombination Graph defined in terms of genomes embedded
+in a pedigree. (a) Diploid individuals, visualised in a highly inbred pedigree and
+labelled $i_1$ to $i_8$, contain both paternal (left) and maternal (right) genomes
+labelled \textsf{A} to \textsf{P}. Black lines show transmission paths connecting
+genomes in the current generation (\textsf{A} to \textsf{D}) with their ancestors.
+Two recombination breakpoints, at positions 3 and 6 mean that genomes \textsf{A}
+and \textsf{C} are independent recombinants of the paternal genomes \textsf{E}
+and \textsf{F}: the region of genome transmitted through each edge is shown in
+these cases. Where genomes merge into a common ancestor, the merged region
+is shown darker, until all genomic regions are fully coalesced in genome \textsf{N}
+(b) The equivalent genome ARG, where there is a single type of node (representing
+a genome) along with annotations on all edges; annotations showing partial
+transmission are highlighted in bold. Recombination, such as that causing the
+breakpoint at position 3, is associated with two parent genomes (\textsf{E} and 
+\textsf{F}) and a child genome (the recombinant, \textsf{A}).
+(c) The edge-annotated genome ARG in (b) fully defines the
 local relationships along the genome, as shown in the local

### 2022-03-02 paper.tex
Change breakpoint numbers to match diagram

@@
 genomes in the current generation (\textsf{A} to \textsf{D}) with their ancestors.
-Two recombination breakpoints, at positions 3 and 6 mean that genomes \textsf{A}
+Two recombination breakpoints, at positions 2 and 7 mean that genomes \textsf{A}
 and \textsf{C} are independent recombinants of the paternal genomes \textsf{E}
@@
 is shown darker, until all genomic regions are fully coalesced in genome \textsf{N}
+in the oldest generation.
 (b) The equivalent genome ARG, where there is a single type of node (representing
@@
 transmission are highlighted in bold. Recombination, such as that causing the
-breakpoint at position 3, is associated with two parent genomes (\textsf{E} and 
-\textsf{F}) and a child genome (the recombinant, \textsf{A}).
+breakpoint at position 7, is associated with two parent genomes (\textsf{E} and 
+\textsf{F}) and a child genome (the recombinant, \textsf{C}).
 (c) The edge-annotated genome ARG in (b) fully defines the

### 2022-03-02 illustrations.py

@@ -372,7 +372,7 @@ def arg_in_pedigree():
                     individual=i,
                 )
 
-    bp = [3, 6]
+    bp = [2, 7]
     tables.edges.clear()
     tables.individuals[0] = tables.individuals[0].replace(parents=[2, 3])
     tables.edges.add_row(child=n.A, parent=n.E, left=0, right=bp[0])

### 2022-03-02 paper.tex
Improve the caption for fig 4
Use "immediate inheritance" vs "sample resolved annotations".

@@
 and \textsf{C} are independent recombinants of the paternal genomes \textsf{E}
-and \textsf{F}: the region of genome transmitted through each edge is shown in
-these cases. Where genomes merge into a common ancestor, the merged region
-is shown darker, until all genomic regions are fully coalesced in genome \textsf{N}
-in the oldest generation.
+and \textsf{F}: in these cases the edge annotations have been shown, indicating the
+region of genome transmitted. Genomes are shaded such that where, backwards in time,
+they merge into a common ancestor, the merged region is darker; this shows that all
+genomic regions are fully coalesced in genome \textsf{N} in the oldest generation.
 (b) The equivalent genome ARG, where there is a single type of node (representing
@@
 \caption{\label{fig-ancestry-resolution}
-Implicit vs explicit edge annotations. (a) An gARG equivalent to the encoding
-described by Griffiths can be created by setting all edges to span the full
-genome, apart from edges between two recombining nodes and their recombinant
-child (edge spans in bold). (b) The sample-resolved gARG, where the Hudson
-simplification algorithm has shortened edge spans so that they encapsulate only
-the upwards transmission of genetic material from the samples A, B, and C;
-plotted line widths are proportional to the genomic span of each edge. The
-example ARG used here is taken from \citet[][fig. 1]{wiuf1999recombination}.
+The \citet[][fig. 1]{wiuf1999recombination} example graph as a gARG, with contrasting
+edge annotation schemes.
+(a) Edge annotations that reflect immediate inheritance, i.e. transmission of all
+genetic material between parent and child. All edges span the full genome from position 0 to 7,
+apart from those linking a recombinant and its two parents (bold).
+(b) Sample-resolved edge annotations: the Hudson simplification algorithm has been used to
+shorten edge spans to reflect only the upwards transmission of genetic
+material from the samples \textsf{A}, \textsf{B}, and \textsf{C}. The depiction of the genome within
+each node is shaded as in figure \ref{fig-arg-in-pedigree}a to show
+coalescing regions of genome back to the grand MRCA genome \textsf{W}.
+In this case, using line widths to reflect the genomic span of each edge can
+help to visualise the relative importance of different transmission pathways. 
 }

### 2022-03-03 paper.tex
Clarified single chr scope and unified Fig1 with text

--- 2022-03-02 paper.tex
@@
 % of ancestry".
+\item Throughout, we consider ARGs defined for a single chromosome comprising
+of a fixed number of discrete sites. Our argument applies equally well to ARGs
+defined for several non-homologous chromosomes \citep{fearnhead2003ancestral,
+koskela2019robust} and to idealised, continuous chromosomes, but addressing 
+all these cases simultaneously would result in needlessly bulky notation.  
 \end{enumerate}
@@
 
-	\draw (0, 0) -- (0, 1.5) -- (-1,1.5) -- (1,1.5);
-	\draw (-1.4, 1.3) -- (-1.1, 1.3) -- (-1.1, 1.2) -- (-1.4, 1.2) -- (-1.4, 1.3);
-	\draw (-1.1, 1.25) -- (-0.6, 1.25);
-	\draw (0.6, 1.25) -- (0.9, 1.25);
-	\draw (0.9, 1.3) -- (1.4, 1.3) -- (1.4, 1.2) -- (0.9, 1.2) -- (0.9, 1.3);
-	\draw (1.2, 1.4) -- (1.2, 1.1);
-	\node [label=below:{\small $x$}] at (1.2, 1.3) {};
+	\draw (0, 0) -- (0, 1.7) -- (-1,1.7) -- (1,1.7);
+	\draw (-1.4, 1.5) -- (-1.1, 1.5) -- (-1.1, 1.4) -- (-1.4, 1.4) -- (-1.4, 1.5);
+	\draw (-1.1, 1.45) -- (-0.6, 1.45);
+	\draw (0.6, 1.45) -- (0.9, 1.45);
+	\draw (0.9, 1.5) -- (1.4, 1.5) -- (1.4, 1.4) -- (0.9, 1.4) -- (0.9, 1.5);
+	\draw (1.2, 1.6) -- (1.2, 1.3);
+	\node [label=below:{\small $x$}] at (1.2, 1.5) {};
 
@@
 
-	\draw (1, 1.5) -- (1, 3.1) -- (0,3.1) -- (2,3.1);
+	\draw (1, 1.7) -- (1, 3.1) -- (0,3.1) -- (2,3.1);
 	\draw (-0.4, 2.85) -- (0.2, 2.85);
@@
 
-	\draw (-1, 1.5) -- (-1, 4.5) -- (0,4.5) -- (0,3.1);
+	\draw (-1, 1.7) -- (-1, 4.5) -- (0,4.5) -- (0,3.1);
 	\node [scale=0.2,label=above left:{\small A}] at (-0.5,4.5) {A};
@@
 	% Edge annotations above each event
-	\node [label=left:{\small$\{[0,L)\};1$}] at (0.2,0.6) {};
-	\node [label=left:{\small$\{[0,L)\};1$}] at (4.2,0.4) {};
-	\node [label=right:{\small$\{[0,L)\};1$}] at (5.8,0.4) {};
-	\node [label=left:{\small$\{[0,L)\};2$}] at (5.2,1.6) {};
-	\node [label=left:{\small$\{[0,v)\};1$}] at (-0.8,2.3) {};
-	\node [label=right:{\small$\{[v,L)\};1$}] at (0.8,2.1) {};
-	\node [label=right:{\small$\{[0,w)\};2$}] at (3.8,3.9) {};
-	\node [label=left:{\small$\{[w,L)\};2$}] at (6.2,5) {};
-	\node [label=right:{\small$\{[x,L)\};1$}] at (-0.2,3.8) {};
-	\node [label=right:{\small$\{[v,x)\};1$}] at (1.8,4.1) {};
-	\node [label=left:{\small$\{[0,v), [x, L)\};1$}] at (-0.3,5.2) {};
-	\node [label=right:{\small$\{[0,v)\};2$}] at (2.8,6.4) {};
-	\node [label=left:{\small$\{[x,L)\};1$}] at (-1.3,7) {};
-	\node [label=right:{\small$\{[0,v)\};1$}] at (0.3,6.4) {};
+	\node [label=left:{\small$\{(0,m,1)\}$}] at (0.2,0.6) {};
+	\node [label=left:{\small$\{(0,m,1)\}$}] at (4.2,0.4) {};
+	\node [label=right:{\small$\{(0,m,1)\}$}] at (5.8,0.4) {};
+	\node [label=left:{\small$\{(0,m,2)\}$}] at (5.2,1.6) {};
+	\node [label=left:{\small$\{(0,v,1)\}$}] at (-0.8,2.3) {};
+	\node [label=right:{\small$\{(v,m,1)\}$}] at (0.8,2.1) {};
+	\node [label=right:{\small$\{(0,w,2)\}$}] at (3.8,3.9) {};
+	\node [label=left:{\small$\{(w,m,2)\}$}] at (6.2,5) {};
+	\node [label=right:{\small$\{(x,m,1)\}$}] at (-0.2,3.8) {};
+	\node [label=right:{\small$\{(v,x,1)\}$}] at (1.8,4.1) {};
+	\node [label=left:{\small$\{(0,v,1), (x,m,1)\}$}] at (-0.3,5.2) {};
+	\node [label=right:{\small$\{(0,v,2)\}$}] at (2.8,6.4) {};
+	\node [label=left:{\small$\{(x,m,1)\}$}] at (-1.3,7) {};
+	\node [label=right:{\small$\{(0,v,1)\}$}] at (0.3,6.4) {};
 
@@
 	\node[label=right:{CA \; RE}] at (7.5, 8.2) {};
-	\node[label=right:{$\binom{2}{2}$ \; $(2 L - x - w) \rho$}] at (7.5, 7.6) {};
-	\node[label=right:{$\binom{4}{2}$ \; $(2 L + 2 v - x - w) \rho$}] at (7.5, 6.6) {};
-	\node[label=right:{$\binom{3}{2}$ \; $(2 L + v - w) \rho$}] at (7.5, 5.7) {};
-	\node[label=right:{$\binom{4}{2}$ \; $(2 L + x - v) \rho$}] at (7.5, 4.95) {};
-	\node[label=right:{$\binom{5}{2}$ \; $2 L \rho$}] at (7.5, 3.8) {};
-	\node[label=right:{$\binom{4}{2}$ \; $2 L \rho$}] at (7.5, 2.75) {};
-	\node[label=right:{$\binom{3}{2}$ \; $2 L \rho$}] at (7.5, 1.95) {};
-	\node[label=right:{$\binom{2}{2}$ \; $2 L \rho$}] at (7.5, 1.35) {};
-	\node[label=right:{$\binom{3}{2}$ \; $3 L \rho$}] at (7.5, 0.6) {};
+	\node[label=right:{$\binom{2}{2}$ \; $\frac{2 m - x - w - 2}{m - 1} \rho$}] at (7.5, 7.6) {};
+	\node[label=right:{$\binom{4}{2}$ \; $\frac{2 m + 2 v - x - w - 3}{m - 1} \rho$}] at (7.5, 6.6) {};
+	\node[label=right:{$\binom{3}{2}$ \; $\frac{2 m + v - w - 3}{m - 1} \rho$}] at (7.5, 5.7) {};
+	\node[label=right:{$\binom{4}{2}$ \; $\frac{2 m + x - v - 4}{m - 1} \rho$}] at (7.5, 4.95) {};
+	\node[label=right:{$\binom{5}{2}$ \; $\frac{2 m - 5}{m - 1} \rho$}] at (7.5, 3.8) {};
+	\node[label=right:{$\binom{4}{2}$ \; $\frac{2 m - 4}{m -1} \rho$}] at (7.5, 2.75) {};
+	\node[label=right:{$\binom{3}{2}$ \; $\frac{2 m - 3}{m - 1} \rho$}] at (7.5, 2.05) {};
+	\node[label=right:{$\binom{2}{2}$ \; $2 \rho$}] at (7.5, 1.45) {};
+	\node[label=right:{$\binom{3}{2}$ \; $3 \rho$}] at (7.5, 0.6) {};
 
@@
 	\draw[color=gray, dashed] (7.5, 1.2) -- (11.1, 1.2);
-	\draw[color=gray, dashed] (7.5, 1.5) -- (11.1, 1.5);
+	\draw[color=gray, dashed] (7.5, 1.7) -- (11.1, 1.7);
 	\draw[color=gray, dashed] (7.5, 2.4) -- (11.1, 2.4);
@@
 
-	\draw (-1, 1.5) -- (-1, 4.5) -- (0,4.5) -- (0,3.1);
+	\draw (-1, 1.7) -- (-1, 4.5) -- (0,4.5) -- (0,3.1);
 	\draw (-0.9, 4.3) -- (-0.6, 4.3) -- (-0.6, 4.2) -- (-0.9, 4.2) -- (-0.9, 4.3);
@@
 	\node[label=right:{CA \; RE}] at (7.5, 8.2) {};
-	\node[label=right:{$\binom{3}{2}$ \; $3 L \rho$}] at (7.5, 7.6) {};
-	\node[label=right:{$\binom{4}{2}$ \; $4 L \rho$}] at (7.5, 6.9) {};
-	\node[label=right:{$\binom{5}{2}$ \; $5 L \rho$}] at (7.5, 6.3) {};
-	\node[label=right:{$\binom{4}{2}$ \; $4 L \rho$}] at (7.5, 5.7) {};
-	\node[label=right:{$\binom{5}{2}$ \; $5 L \rho$}] at (7.5, 4.95) {};
-	\node[label=right:{$\binom{6}{2}$ \; $6 L \rho$}] at (7.5, 4.15) {};
-	\node[label=right:{$\binom{5}{2}$ \; $5 L \rho$}] at (7.5, 3.45) {};
-	\node[label=right:{$\binom{4}{2}$ \; $4 L \rho$}] at (7.5, 2.75) {};
-	\node[label=right:{$\binom{3}{2}$ \; $3 L \rho$}] at (7.5, 1.95) {};
-	\node[label=right:{$\binom{2}{2}$ \; $2 L \rho$}] at (7.5, 1.35) {};
-	\node[label=right:{$\binom{3}{2}$ \; $3 L \rho$}] at (7.5, 0.6) {};
+	\node[label=right:{$\binom{3}{2}$ \; $3 \rho$}] at (7.5, 7.6) {};
+	\node[label=right:{$\binom{4}{2}$ \; $4 \rho$}] at (7.5, 6.9) {};
+	\node[label=right:{$\binom{5}{2}$ \; $5 \rho$}] at (7.5, 6.3) {};
+	\node[label=right:{$\binom{4}{2}$ \; $4 \rho$}] at (7.5, 5.7) {};
+	\node[label=right:{$\binom{5}{2}$ \; $5 \rho$}] at (7.5, 4.95) {};
+	\node[label=right:{$\binom{6}{2}$ \; $6 \rho$}] at (7.5, 4.15) {};
+	\node[label=right:{$\binom{5}{2}$ \; $5 \rho$}] at (7.5, 3.45) {};
+	\node[label=right:{$\binom{4}{2}$ \; $4 \rho$}] at (7.5, 2.75) {};
+	\node[label=right:{$\binom{3}{2}$ \; $3 \rho$}] at (7.5, 2.05) {};
+	\node[label=right:{$\binom{2}{2}$ \; $2 \rho$}] at (7.5, 1.45) {};
+	\node[label=right:{$\binom{3}{2}$ \; $3 \rho$}] at (7.5, 0.6) {};
 
@@
 	\draw[color=gray, dashed] (7.5, 1.2) -- (11, 1.2);
-	\draw[color=gray, dashed] (7.5, 1.5) -- (11, 1.5);
+	\draw[color=gray, dashed] (7.5, 1.7) -- (11, 1.7);
 	\draw[color=gray, dashed] (7.5, 2.4) -- (11, 2.4);
@@
 A realisation of the graph traversed by Hudson's algorithm started from a
-sample of three continuous chromosomes of length $L$ at time $t = 0$, and
+sample of three chromosomes of length $m$ at time $t = 0$, and
 propagated until time $T$. The MRCA on the genetic interval $[v, w)$ is reached
@@
 the specific distribution of ancestral material among lineages, the
-rate of recombination is now simply $k \rho (m - 1)$. At a recombination event
+rate of recombination is now simply $k \rho$. At a recombination event
 we choose a lineage (edge) uniformly, and a
-breakpoint $0 < x < L$ uniformly on its genome. We terminate the edge at a
+breakpoint $0 < x < m$ uniformly on its genome. We terminate the edge at a
 node, record the breakpoint and start two new edges from this node. The process
@@
 These methods met with limited success
-due to the overwhelmingly large and
-horribly structured [Jere: what's the right way to say this?] state space of ARGs.
+due to the overwhelmingly large state space of ARGs, which also has
+no simple geometry or neighbourhood structure that an inference or
+sampling method could exploit.
 The focus subsequently shifted to the more
@@
 model, assuming that the sample size $n$ is much less than the
-population size $N$ and [some assumptions about genome length]
-\textcolor{red}{what is this Jere?}. While these approximations are
+population size $N$, and that the genome is short enough that,
+with high probability, the number of concurrent lineages in the ARG remains
+much smaller than $N$ at each time. While these approximations are
 reasonable

### 2022-03-03 paper.tex
Fix up discussion of fig 3
Also fixes #145

--- 2022-03-02 paper.tex
@@
 fundamentally limiting the patterns of ancestry that can be described.
-Following \cite{mathieson2020ancestry} we instead define the
-ARG data structure \emph{in terms of} an underlying pedigree, so that
-any form of inheritance structure can be modelled.
+Following \cite{mathieson2020ancestry} instead of focussing on events,
+we define the ARG data structure in terms \emph{genomes} and the patterns of
+transmission between them.
 % Bit of duplication here, but good to make the point about being
@@
 Fig~\ref{fig-arg-in-pedigree} illustrates the basic elements of a GARG.
-The pedigree is composed of a set of individuals who each carry
-two \emph{genomes} (we assume diploidy here for concreteness, but there
-is no inherent limitation on ploidy).
-Each genome is inherited from one of the individual's parents, and may
-be the recombined product of that parent's two genomes.
-For example, individual $I_1$ in Fig.~\ref{fig-arg-in-pedigree}A
-has two genomes $A$ and $B$,
-inherited from parents $I_3$ and $I_4$. Genome $A$ is the product of
-recombining $I_3$'s two genomes $E$ and $F$ at position FIXME,
-whereas $B$ was inherited directly from $I_4$'s genome $G$ without
-recombination. The pedigree is shown here for concreteness, but
-is not a necessary part of the model. We can reason directly about the
-genomes without necessarily knowing how they are combined within
-individuals. This ``genome-perspective'' is show in
-Fig.~\ref{fig-arg-in-pedigree}B, which depicts the same graph
-of the ancestry of the sample genomes $A$--$D$. The graph
+For concreteness, Fig.~\ref{fig-arg-in-pedigree}a shows the genomes
+embedded in a inbred diploid pedigree, in which each
+individual carries two genomes (each individual, labelled $i_1$ to $i_8$,
+therefore has a paternal genome on the right and a maternal one on
+the left, although in principle there is no requirement for a specific mating
+system or ploidy). A genome is inherited from one of the individual's parents,
+and may be the recombined product of that parent's two genomes.
+For example, individual $i_1$ has two genomes \textsf{A} and \textsf{B},
+inherited from parents $i_3$ and $i_4$. Genome \textsf{A} is the product of
+recombining $i_3$'s two genomes \textsf{E} and \textsf{F} at position 2,
+whereas \textsf{B} was inherited directly from $i_4$'s genome \textsf{G} without
+recombination. Fig.~\ref{fig-arg-in-pedigree}b illustrates that the patterns of
+transmission between genomes can be followed without necessarily knowing
+how they are combined within individuals. This ``genome-perspective''
+simply depicts the ancestry of the sample genomes $A$--$D$. The graph
 topology along with the genome coordinate annotations on the
 edges then uniquely defines the trees at every position
-along the genome (Fig.~\ref{fig-arg-in-pedigree}C).
+along the genome (Fig.~\ref{fig-arg-in-pedigree}c).
 
@@
 and important consequences. Firstly, the graph may contain nodes that
-do not correspond to any particular event in the ancestry of the
-samples. For example, node $H$ is the direct ancestor of $D$ in
+do not correspond to any particular event in the ancestry of the samples.
+For example, node \textsf{H} is the direct ancestor of \textsf{D} in
 Fig~\ref{fig-arg-in-pedigree}, and is not the product of either a
 coalescence or a recombination. Such nodes would usually be removed
-from the graph so that $D$ descends directly from $K$ (see section XXX)
-but there is no necessity for this from a representational perspective.
-This ability can be useful in practice and could be required if,
+from the graph so that \textsf{D} descends directly from \textsf{K} (see
+section XXX) but there is no necessity for this from a representational
+perspective. This ability can be useful in practice and could be required if,
 % Do we need to spell this out?
@@
 intervals (representing the precise inheritance patterns).
-For example, in Fig~\ref{fig-arg-in-pedigree}, node $A$ is the
-recombinant descendent of $E$ and $F$, inheriting positions
-FIXME from $E$ and genome positions FIXME from $F$. Any particular
+For example, in Fig~\ref{fig-arg-in-pedigree}, node \textsf{C} is the
+recombinant descendent of \textsf{E} and \textsf{F}, inheriting positions
+0 to 7 from $E$ and genome positions 7 to 10 from $F$. Any particular
 node can be both a recombinant (it is the recombined product of
@@
 multiple samples meet at this node; see the next section).
+% Here was can talk about representing the recombination by a
+% recombinant and 2 parent "recombination nodes" rather than
+% a single "recombination node" as in the normal ARG representation.
+% We may also wish to note that in this example the recombination nodes
+% are shared between siblings, and can also be coalescent nodes
+% this ties in with stuff discussed 3 paras below.
 [More stuff here: clarify why we assign coordinates to edges

### 2022-03-03 paper.tex

@@
 
+[We should discuss the relationship between ARG nodes and SPRs
+here. Cite \citep{deng2021distribution} too. Maybe we want to make
+this a separate section, as it's a subtle and important point.]
+
 % This text might go somewhere
@@
 % Should we also mention that most of these infer just the topology, but some also the times?
+
+Espalier~\citep{rasmussen2022espalier} uses Maximum Agreement Forests to
+minimise the differences between local trees estimated using other methods.
 

### 2022-03-03 paper.tex
Move formal gARG definition to earlier in section

@@
 Fig~\ref{fig-arg-in-pedigree} illustrates the basic elements of a GARG.
-The pedigree is composed of a set of individuals who each carry
-two \emph{genomes} (we assume diploidy here for concreteness, but there
-is no inherent limitation on ploidy).
-Each genome is inherited from one of the individual's parents, and may
-be the recombined product of that parent's two genomes.
-For example, individual $I_1$ in Fig.~\ref{fig-arg-in-pedigree}A
-has two genomes $A$ and $B$,
-inherited from parents $I_3$ and $I_4$. Genome $A$ is the product of
-recombining $I_3$'s two genomes $E$ and $F$ at position FIXME,
-whereas $B$ was inherited directly from $I_4$'s genome $G$ without
-recombination. The pedigree is shown here for concreteness, but
-is not a necessary part of the model. We can reason directly about the
-genomes without necessarily knowing how they are combined within
-individuals. This ``genome-perspective'' is show in
-Fig.~\ref{fig-arg-in-pedigree}B, which depicts the same graph
-of the ancestry of the sample genomes $A$--$D$. The graph
+For concreteness, Fig.~\ref{fig-arg-in-pedigree}a shows the genomes
+embedded in a inbred diploid pedigree, in which each
+individual carries two genomes (each individual, labelled $i_1$ to $i_8$,
+therefore has a paternal genome on the right and a maternal one on
+the left, although in principle there is no requirement for a specific mating
+system or ploidy). A genome is inherited from one of the individual's parents,
+and may be the recombined product of that parent's two genomes.
+For example, individual $i_1$ has two genomes \textsf{A} and \textsf{B},
+inherited from parents $i_3$ and $i_4$. Genome \textsf{A} is the product of
+recombining $i_3$'s two genomes \textsf{E} and \textsf{F} at position 2,
+whereas \textsf{B} was inherited directly from $i_4$'s genome \textsf{G} without
+recombination. Fig.~\ref{fig-arg-in-pedigree}b illustrates that the patterns of
+transmission between genomes can be followed without necessarily knowing
+how they are combined within individuals. This ``genome-perspective''
+simply depicts the ancestry of the sample genomes $A$--$D$. The graph
 topology along with the genome coordinate annotations on the
 edges then uniquely defines the trees at every position
-along the genome (Fig.~\ref{fig-arg-in-pedigree}C).
-
-The focus on modelling genomes rather than events has some interesting
-and important consequences. Firstly, the graph may contain nodes that
-do not correspond to any particular event in the ancestry of the
-samples. For example, node $H$ is the direct ancestor of $D$ in
-Fig~\ref{fig-arg-in-pedigree}, and is not the product of either a
-coalescence or a recombination. Such nodes would usually be removed
-from the graph so that $D$ descends directly from $K$ (see section XXX)
-but there is no necessity for this from a representational perspective.
-This ability can be useful in practice and could be required if,
-% Do we need to spell this out?
-for example, we have a deeply sequenced pedigree of
-haplodiploids. % e.g. Honeybees? Be great to cite an example!
-
-Secondly, the way that we model recombination is subtly different.
-Rather than having a single node that represents the \emph{event},
-there are three nodes (representing the recombinant offspring and
-two parental genomes) and two edges with associated genome coordinate
-intervals (representing the precise inheritance patterns).
-For example, in Fig~\ref{fig-arg-in-pedigree}, node $A$ is the
-recombinant descendent of $E$ and $F$, inheriting positions
-FIXME from $E$ and genome positions FIXME from $F$. Any particular
-node can be both a recombinant (it is the recombined product of
-its parent's genomes) and a common ancestor (the ancestry from
-multiple samples meet at this node; see the next section).
-[More stuff here: clarify why we assign coordinates to edges
-rather than nodes. Basically lay out all the stuff that we can
-talk about \emph{without} going in to the details of
-ancestry resolution.]
+along the genome (Fig.~\ref{fig-arg-in-pedigree}c).
 
@@
 exists) and move to the next node $p$.
+
+The focus on modelling genomes rather than events has some interesting
+and important consequences. Firstly, the graph may contain nodes that
+do not correspond to any particular event in the ancestry of the samples.
+For example, node \textsf{H} is the direct ancestor of \textsf{D} in
+Fig~\ref{fig-arg-in-pedigree}, and is not the product of either a
+coalescence or a recombination. Such nodes would usually be removed
+from the graph so that \textsf{D} descends directly from \textsf{K} (see
+section XXX) but there is no necessity for this from a representational
+perspective. This ability can be useful in practice and could be required if,
+% Do we need to spell this out?
+for example, we have a deeply sequenced pedigree of
+haplodiploids. % e.g. Honeybees? Be great to cite an example!
+
+Secondly, the way that we model recombination is subtly different.
+Rather than having a single node that represents the \emph{event},
+there are three nodes (representing the recombinant offspring and
+two parental genomes) and two edges with associated genome coordinate
+intervals (representing the precise inheritance patterns).
+For example, in Fig~\ref{fig-arg-in-pedigree}, node \textsf{C} is the
+recombinant descendent of \textsf{E} and \textsf{F}, inheriting positions
+0 to 7 from $E$ and genome positions 7 to 10 from $F$. Any particular
+node can be both a recombinant (it is the recombined product of
+its parent's genomes) and a common ancestor (the ancestry from
+multiple samples meet at this node; see the next section).
+% Here was can talk about representing the recombination by a
+% recombinant and 2 parent "recombination nodes" rather than
+% a single "recombination node" as in the normal ARG representation.
+% We may also wish to note that in this example the recombination nodes
+% are shared between siblings, and can also be coalescent nodes
+% this ties in with stuff discussed 3 paras below.
+[More stuff here: clarify why we assign coordinates to edges
+rather than nodes. Basically lay out all the stuff that we can
+talk about \emph{without} going in to the details of
+ancestry resolution.]
 

### 2022-03-03 paper.tex
Rephrase explanation of recombination nodes

@@
 Following \cite{mathieson2020ancestry} instead of focussing on events,
-we define the ARG data structure in terms \emph{genomes} and the patterns of
-transmission between them.
+we define the ARG data structure in terms of \emph{genomes} and the
+patterns of transmission between them.
 % Bit of duplication here, but good to make the point about being
@@
 
-The focus on modelling genomes rather than events has some interesting
-and important consequences. Firstly, the graph may contain nodes that
+The focus on genomes rather than events has some interesting
+and important consequences. Firstly, a gARG may contain nodes that
 do not correspond to any particular event in the ancestry of the samples.
@@
 
-Secondly, the way that we model recombination is subtly different.
-Rather than having a single node that represents the \emph{event},
-there are three nodes (representing the recombinant offspring and
-two parental genomes) and two edges with associated genome coordinate
-intervals (representing the precise inheritance patterns).
-For example, in Fig~\ref{fig-arg-in-pedigree}, node \textsf{C} is the
-recombinant descendent of \textsf{E} and \textsf{F}, inheriting positions
-0 to 7 from $E$ and genome positions 7 to 10 from $F$. Any particular
-node can be both a recombinant (it is the recombined product of
-its parent's genomes) and a common ancestor (the ancestry from
-multiple samples meet at this node; see the next section).
-% Here was can talk about representing the recombination by a
-% recombinant and 2 parent "recombination nodes" rather than
-% a single "recombination node" as in the normal ARG representation.
-% We may also wish to note that in this example the recombination nodes
-% are shared between siblings, and can also be coalescent nodes
-% this ties in with stuff discussed 3 paras below.
-[More stuff here: clarify why we assign coordinates to edges
+Secondly, the natural way to represent the process of recombination in a
+gARG is not as a single ``recombination node'' (as in an eARG), but with
+two parent genomes between which recombination takes place (e.g.
+nodes \textsf{E} and \textsf{F} in Fig~\ref{fig-arg-in-pedigree}), and a
+single resulting ``recombinant'' genome (e.g. \textsf{C}). The
+edges that link this recombinant with its two parents carry the information
+concerning breakpoint location that would normally be present on the
+"recombination node" in an eARG. For example, node \textsf{C} in
+Fig~\ref{fig-arg-in-pedigree} inherits genome positions 0 to 7 from \textsf{E}
+and positions 7 to 10 from \textsf{F}, indicating that the breakpoint is at position 7.
+Unlike the eARG, the gARG need not make any distinction
+between common ancestor and recombination nodes: indeed, a
+genome can be simultaneously involved in recombination as well as being
+a common ancestor, as seen most clearly in nodes \textsf{E} and \textsf{F}
+in Fig~\ref{fig-arg-in-pedigree}b. Indeed, in this case, these nodes are involved in
+ \emph{two} separate recombinations, giving rise to the genomes in two siblings.
+
+The gARG method for representing recombinations may seem wasteful,
+since more than one node is required for each recombination,
+and genomic interval information must be stored on every edge.
+However, asymptotically the amount of information stored is the
+same, and there is much simplicity to be gained by having
+a single type of node and by treating all edges in the same way.
+More importantly --- as we see in the next section --- we gain the
+ability to simplify the ARG by removing nodes without affecting the
+local tree topologies, and also the ability to record only the
+regions of genome relevant to the final, sample genomes. This last
+feature turns out to be the key to efficient processing of the ARG.
+
+[Possibly more stuff here: clarify why we assign coordinates to edges
 rather than nodes. Basically lay out all the stuff that we can
@@
 ancestry resolution.]
-
-% Asking the reader to flick back to figure 2 here, not great.
-The differences between EARGs
-and GARGs are subtle but important, and illustrated in
-Fig.~\ref{fig-arg-data-structure}, which shows a simple ARG
-with three samples and one recombination.
-Fig.~\ref{fig-arg-data-structure}A shows this topology as a
-classical Griffiths-like EARG, and
-Fig.~\ref{fig-arg-data-structure}B shows it as a GARG.
-The two are almost identical, except we model the recombination
-event via the two parental genomes in nodes $3$ and $4$ and explicitly
-store the genomic interval that the child inherited from the
-parent with every edge.
-This approach may seem wasteful since we must store an extra
-node for each recombination and genomic interval information
-with every edge (even though many simply span the full genome).
-However, asymptotically the amount of information stored is the
-same, and as we see in the next section, keeping interval
-information for all edges enables useful generalisations.
-There is also much simplicity to be gained by having
-a single type of node and by treating all edges in the same way.
 

### 2022-03-08 paper.tex
Discussion of SPRs

--- 2022-03-03 paper.tex
@@
 
-Formally, we can define a GARG as a set of edges $E$, where each
-element of the set is a tuple $(c, p, I)$ such that $c, p \in \mathbb{N}$,
-$c$ is the child node, $p$ is the parent node and $I$ is the set of
-disjoint genomic intervals $(\ell, r]$ over which genome $c$ inherits from $p$ (we may
-also phrase this in terms of the ancestral material of a sample---see
-the next section). Deriving the local tree at a point $x$
-follows largely the same pattern as for an EARG but is somewhat
-simpler because coordinate information is directly associated with
-edges. Starting with the sample nodes $S$ we traverse
-upwards through the graph from child to parent. At each node $u$, we find an
-edge $(c, p, I) \in E$ such that $u = c$ and $x \in I$ (if it
-exists) and move to the next node $p$.
-
 The focus on modelling genomes rather than events has some interesting
@@
 ancestry resolution.]
+
+Formally, we can define a GARG as a set of edges $E$, where each
+element of the set is a tuple $(c, p, I)$ such that $c, p \in \mathbb{N}$,
+$c$ is the child node, $p$ is the parent node and $I$ is the set of
+disjoint genomic intervals $(\ell, r]$ over which genome $c$ inherits from $p$ (we may
+also phrase this in terms of the ancestral material of a sample---see
+the next section). Deriving the local tree at a point $x$
+follows largely the same pattern as for an EARG but is somewhat
+simpler because coordinate information is directly associated with
+edges. Starting with the sample nodes $S$ we traverse
+upwards through the graph from child to parent. At each node $u$, we find an
+edge $(c, p, I) \in E$ such that $u = c$ and $x \in I$ (if it
+exists) and move to the next node $p$.
 
@@
 
-[We should discuss the relationship between ARG nodes and SPRs
-here. Cite \citep{deng2021distribution} too. Maybe we want to make
-this a separate section, as it's a subtle and important point.]
+\section*{ARGs and SPRs}
+\citet{wiuf1999ancestry,wiuf1999recombination} recast
+the coalescent with recombination
+as a process along the genome rather than backwards in time.
+In this formulation we begin at one end of a chromosome by
+sampling a tree under the standard coalescent, and then reason
+about how this tree is modified by recombination events
+as we move along the genome. Although this algorithm is
+significantly more complicated to simulate than Hudson's
+backwards in time approach, it has led to substantial
+insights into the process of recombination itself
+and to the sequentially Markov Coalescent (SMC)
+approximations~\citep{mcvean2005approximating,marjoram2006fast}
+which are central to many modern methods.
+
+Under the coalescent with recombination (or SMC) each local
+tree along the genome is transformed into the next by
+a ``subtree prune-and-regraft'' (SPR)
+% FIXME there should be earlier refs for SPRs, right?
+operation~\citep{song2003on,song2006properties}.
+The number of SPRs required to transform one leaf-labelled tree
+into another is known as the SPR distance [cite]
+and is computable in O(??) time.
+
+[Discussion of SPRs in an EARG]
+
+The effects of recombination as we move from left-to-right along
+the genome in an unsimplified GARG are easy to interpret.
+Consider the leftmost tree in Fig~\ref{fig-simplification}.
+In the first tree the parent of B is F, and in the second
+tree its parent is E. Since both parents are present as
+unary nodes in the tree, we simply change one edge in
+order to effect the transition.
+A more complex transition occurs when we have a regraft operation
+that occurs above the current root (i.e., the current oldest
+non-unary node). For example, the transition between the
+second and third trees corresponds to node D switching
+from the path through I to H, and we then must add branches
+corresponding to passing through L and meeting at M.
+
+The equivalence between an ARG and a set of local trees
+separated by SPRs is often mentioned [e.g. x y z], but
+it is not obvious how to produce an ARG given (say) a
+collection of newick trees. If the internal nodes are
+labelled, then we can simply input the edges for each
+tree in turn, which can easily be ``squashed'' down
+to a minimal representation. However, internal nodes
+are rarely labelled in newick trees, and therefore
+the process of contructing a graph is much harder.
+In princple, we can contruct an EARG by reasoning
+about the SPRs that must has occured between two
+adjacent trees and postulating the recombination
+nodes that must have occured. However, this process
+is computationally expensive [How expensive? is this
+what Espalier does?] Additionally, unless some extra
+information is provided the time of the recombination
+will not be known. Fundamentally, we must include the
+unary nodes corresponding to recombination parents
+in the trees if we wish to recover an ARG from them.
+
+[ Cite \citep{deng2021distribution} somewhere here.]
 

### 2022-03-08 argutils/convert.py
Convert function for KwARG output

@@ -88,3 +88,46 @@ def convert_argweaver(infile):
     # garg = argutils.earg_to_garg(ts)
 
     return ts.simplify(keep_unary=True)
+
+def convert_kwarg(infile, num_samples, sequence_length):
+    """
+    Convert a KwARG output file to a tree sequence.
+    """
+
+    tables = tskit.TableCollection(sequence_length=sequence_length)
+    tables.nodes.metadata_schema = tskit.MetadataSchema.permissive_json()
+    time = 0.0
+
+    node_ids = {}
+    for n in range(num_samples):
+        node_ids[n] = n
+        tsk_id = tables.nodes.add_row(flags=tskit.NODE_IS_SAMPLE, time=0.0, metadata=n)
+
+    for x in infile:
+        line = x.split()
+        if (line[0] == "Mutation"):
+            site_id = tables.sites.add_row(position=int(line[3]) - 1, ancestral_state='0')
+            tables.mutations.add_row(site=site_id, node=node_ids[int(line[6]) - 1], derived_state='1')
+        elif (line[0] == "Coalescing"):
+            c1 = node_ids[int(line[2]) - 1]
+            c2 = node_ids[int(line[4]) - 1]
+            time += 1
+            tsk_id = tables.nodes.add_row(flags=0, time=time, metadata=line[2])
+            node_ids[int(line[2]) - 1] = tsk_id
+            tables.edges.add_row(0, sequence_length, tsk_id, c1)
+            tables.edges.add_row(0, sequence_length, tsk_id, c2)
+        elif (line[0] == "---->Recombination"):
+            breakpoint = int(line[6][:-1]) - 1
+            n1 = node_ids[int(line[3]) - 1]
+            time += 1
+            tsk_id_1 = tables.nodes.add_row(flags=0, time=time, metadata=line[3])
+            node_ids[int(line[3]) - 1] = tsk_id_1
+            tsk_id_2 = tables.nodes.add_row(flags=0, time=time, metadata=line[11])
+            node_ids[int(line[11]) - 1] = tsk_id_2
+            tables.edges.add_row(0, breakpoint, tsk_id_1, n1)
+            tables.edges.add_row(breakpoint, sequence_length, tsk_id_2, n1)
+
+    tables.sort()
+    ts = tables.tree_sequence()
+
+    return argutils.simplify_keeping_all_nodes(ts)

### 2022-03-08 argutils/tests.py

@@ -538,3 +538,10 @@ class TestConvertArgweaver:
         assert ts.sequence_length == 1000
         assert ts.num_samples == 8
         # TODO more tests
+
+class TestConvertKwARG:
+    def test_example(self):
+        with open("examples/kreitman.kwarg") as f:
+            ts = argutils.convert_kwarg(f, 11, 43)
+        assert ts.sequence_length == 43
+        assert ts.num_samples == 11

### 2022-03-09 paper.tex
Another pass on SPRs

@@
 the coalescent with recombination
-as a process along the genome rather than backwards in time.
+as a stocastic process along the genome rather than backwards in time.
 In this formulation we begin at one end of a chromosome by
@@
 a ``subtree prune-and-regraft'' (SPR)
-% FIXME there should be earlier refs for SPRs, right?
-operation~\citep{song2003on,song2006properties}.
+% Hein et al call it the "subtree transfer" operation but it's
+% the same thing.
+operation~\citep{hein1990reconstructing,song2003on,song2006properties}.
 The number of SPRs required to transform one leaf-labelled tree
-into another is known as the SPR distance [cite]
-and is computable in O(??) time.
+into another is known as the SPR distance
+% Hein 1996 say that SPR distance is NP-hard, but Allen 2001
+% say this is wrong. But then Bordewich went ahead and said
+% yes, actually it is NP hard.
+and is NP-hard to
+compute~\citep{hein1996complexity,allen2001subtree,bordewich2005computational}.
 
@@
 separated by SPRs is often mentioned [e.g. x y z], but
-it is not obvious how to produce an ARG given (say) a
-collection of newick trees. If the internal nodes are
-labelled, then we can simply input the edges for each
-tree in turn, which can easily be ``squashed'' down
-to a minimal representation. However, internal nodes
-are rarely labelled in newick trees, and therefore
-the process of contructing a graph is much harder.
-In princple, we can contruct an EARG by reasoning
-about the SPRs that must has occured between two
-adjacent trees and postulating the recombination
-nodes that must have occured. However, this process
-is computationally expensive [How expensive? is this
-what Espalier does?] Additionally, unless some extra
-information is provided the time of the recombination
-will not be known. Fundamentally, we must include the
-unary nodes corresponding to recombination parents
-in the trees if we wish to recover an ARG from them.
+it is important to note that while extracting local
+trees from an ARG is straightforward, the converse
+process of reconstructing precisely the same ARG from
+these trees is more problematic. If the internal
+nodes are consistently labelled and unary nodes are
+included in the local trees (e.g., Fig~\ref{fig-simplification})
+then it is easy to exactly reconstruct the ARG.
+Otherwise, the problem is much more difficult and it
+may not be possible to uniquely reconstruct the original ARG.
+Without labels for the internal nodes, we are forced to
+to reason about which subtrees remain the same and which
+differ due to the SPR separating adjacent trees in order to
+assign node identity. Without the internal unary nodes marking
+recombinations, we must infer their existence. In
+general, there will not be a unique solution to this problem,
+and so there can be many possible ARGs compatible with
+a given list of
+% want to say "standard phylogenetic" trees here or something, but
+% and they don't have to be binary (just arity > 1), but thought
+% this would be familiar to readers?
+leaf-labelled binary trees.
 

### 2022-03-09 paper.tex
Correct references to figs (a), (b) etc
For clarity I have switched to lowercase subfigure letters, to avoid confusion with the node labels (which are capital letters). This may eventually need to be changed depending on the journal style.

--- 2022-03-08 paper.tex
@@
 }
-\caption{(A)
+\caption{(a)
 A realisation of the graph traversed by Hudson's algorithm started from a
@@
 the recombination rate is $\rho$.
-(B) A corresponding realisation of a big ARG, which augments Hudson's algorithm
+(b) A corresponding realisation of a big ARG, which augments Hudson's algorithm
 by tracking nonancestral lineages. The result is a simpler state space and
@@
 \caption{\label{fig-simplification}
-ARG simplification. (A) An ARG simulated from a diploid Wright-Fisher
+ARG simplification. (a) An ARG and equivalent local trees simulated
+from a diploid Wright-Fisher
 model. Note the complex full-sib relationship between B and C; this
@@
 (see section XXX).
-(B) Simplified to remove all
+(b) Simplified to remove all
 % trying out this terminology - can define in the text
 1-connected graph components (e.g., diamonds).
-(C) Remove nodes that are unary in all local trees.
-(D) Rewrite edges to skip any unary nodes in local trees.
-(E) The local trees for ARG (A).
-(F) The local trees for ARG (D).
+(c) Remove nodes that are unary in all local trees.
+(d) Rewrite edges to skip any unary nodes in local trees.
 }
@@
 presence of ``unary'' nodes: a node in the tree that has exactly
-one child. For example, Fig.~\ref{fig-simplification}E shows the
-local trees along the genome for a simulated ARG
-(Fig.~\ref{fig-simplification}A). Consider the right-most tree,
+one child. For example, Fig.~\ref{fig-simplification} shows a set of
+four simulated gARGs and the local trees that can be constructed from
+them (for simplicity, edge annotations are not shown)
+Consider the right-most tree in Fig.~\ref{fig-simplification}a
 and the path above node $D$. Before $D$ meets its common ancestor
@@
 join again into a common ancestor (e.g. $K,N,O$ and $R$ in
-Fig.~\ref{fig-simplification}A).
+Fig.~\ref{fig-simplification}a).
 Unless we are specifically
@@
 of a single node that has one inbound and one outbound edge.
-Fig.~\ref{fig-simplification}B shows the result of this type of
+Fig.~\ref{fig-simplification}b shows the result of this type of
 graph topology simplification.
@@
 parents of the recombinant node $D$ is still present in
-Fig.~\ref{fig-simplification}B after diamond removal, but
+Fig.~\ref{fig-simplification}b after diamond removal, but
 it is unary in all of the local trees since it has an
 in-degree of $1$.
-Fig.~\ref{fig-simplification}C shows the resulting graph
+Fig.~\ref{fig-simplification}c shows the resulting graph
 after removing these nodes.
@@
 local tree. For example, node $E$ is unary in the first tree
-of Fig.~\ref{fig-simplification}E, but is either binary
+of Fig.~\ref{fig-simplification}a, but is either binary
 or ternary in the other trees (recall this is a Wright-Fisher
@@
 The final level of simplification is to remove \emph{all}
-unary nodes from the local trees (Fig.~\ref{fig-simplification}D
-and E). [And continue]
-
-\section*{ARGs and SPRs}
-\citet{wiuf1999ancestry,wiuf1999recombination} recast
-the coalescent with recombination
-as a process along the genome rather than backwards in time.
-In this formulation we begin at one end of a chromosome by
-sampling a tree under the standard coalescent, and then reason
-about how this tree is modified by recombination events
-as we move along the genome. Although this algorithm is
-significantly more complicated to simulate than Hudson's
-backwards in time approach, it has led to substantial
-insights into the process of recombination itself
-and to the sequentially Markov Coalescent (SMC)
-approximations~\citep{mcvean2005approximating,marjoram2006fast}
-which are central to many modern methods.
-
-Under the coalescent with recombination (or SMC) each local
-tree along the genome is transformed into the next by
-a ``subtree prune-and-regraft'' (SPR)
-% FIXME there should be earlier refs for SPRs, right?
-operation~\citep{song2003on,song2006properties}.
-The number of SPRs required to transform one leaf-labelled tree
-into another is known as the SPR distance [cite]
-and is computable in O(??) time.
-
-[Discussion of SPRs in an EARG]
-
-The effects of recombination as we move from left-to-right along
-the genome in an unsimplified GARG are easy to interpret.
-Consider the leftmost tree in Fig~\ref{fig-simplification}.
-In the first tree the parent of B is F, and in the second
-tree its parent is E. Since both parents are present as
-unary nodes in the tree, we simply change one edge in
-order to effect the transition.
-A more complex transition occurs when we have a regraft operation
-that occurs above the current root (i.e., the current oldest
-non-unary node). For example, the transition between the
-second and third trees corresponds to node D switching
-from the path through I to H, and we then must add branches
-corresponding to passing through L and meeting at M.
-
-The equivalence between an ARG and a set of local trees
-separated by SPRs is often mentioned [e.g. x y z], but
-it is not obvious how to produce an ARG given (say) a
-collection of newick trees. If the internal nodes are
-labelled, then we can simply input the edges for each
-tree in turn, which can easily be ``squashed'' down
-to a minimal representation. However, internal nodes
-are rarely labelled in newick trees, and therefore
-the process of contructing a graph is much harder.
-In princple, we can contruct an EARG by reasoning
-about the SPRs that must has occured between two
-adjacent trees and postulating the recombination
-nodes that must have occured. However, this process
-is computationally expensive [How expensive? is this
-what Espalier does?] Additionally, unless some extra
-information is provided the time of the recombination
-will not be known. Fundamentally, we must include the
-unary nodes corresponding to recombination parents
-in the trees if we wish to recover an ARG from them.
-
-[ Cite \citep{deng2021distribution} somewhere here.]
+unary nodes from the local trees (Fig.~\ref{fig-simplification}d).
+[And continue]
+
+[We should discuss the relationship between ARG nodes and SPRs
+here. Cite \citep{deng2021distribution} too. Maybe we want to make
+this a separate section, as it's a subtle and important point.]
 

### 2022-03-10 paper.tex
Describe implications of fig 3 in 3 parts
As discussed with @jeromekelleher

@@
 % Perhaps we should also mention something about simplification and unknowable
-% nodes, e.g. "show how some of the nodes are not required for useful interpretation
-% of ancestry".
+% nodes, e.g. ``show how some of the nodes are not required for useful interpretation
+% of ancestry''.
 \item Throughout, we consider ARGs defined for a single chromosome comprising
@@
 }
-\caption{(a)
+\caption{(A)
 A realisation of the graph traversed by Hudson's algorithm started from a
@@
 the recombination rate is $\rho$.
-(b) A corresponding realisation of a big ARG, which augments Hudson's algorithm
+(B) A corresponding realisation of a big ARG, which augments Hudson's algorithm
 by tracking nonancestral lineages. The result is a simpler state space and
@@
 % Surely the meaning is clear from the context here and we don't need
-% qualify the meaning of "ARG"
+% qualify the meaning of ``ARG''
 simulating an ARG in this representation is
@@
 Fig~\ref{fig-arg-in-pedigree} illustrates the basic elements of a GARG.
-For concreteness, Fig.~\ref{fig-arg-in-pedigree}a shows the genomes
-embedded in a inbred diploid pedigree, in which each
-individual carries two genomes (each individual, labelled $i_1$ to $i_8$,
-therefore has a paternal genome on the right and a maternal one on
-the left, although in principle there is no requirement for a specific mating
-system or ploidy). A genome is inherited from one of the individual's parents,
+Fig.~\ref{fig-arg-in-pedigree}a shows the genomes
+embedded in an inbred diploid pedigree, in which individuals occur at
+discrete generations and carry two genomes. Each individual (labelled
+$i_1$ to $i_8$) therefore has a paternal genome (on the right) and a maternal one
+(on the left), although the pedigree is only shown for concreteness, and 
+there is no inherent limitation on mating system, ploidy, or age structure.
+A genome is inherited from one of the individual's parents,
 and may be the recombined product of that parent's two genomes.
@@
 whereas \textsf{B} was inherited directly from $i_4$'s genome \textsf{G} without
-recombination. Fig.~\ref{fig-arg-in-pedigree}b illustrates that the patterns of
+recombination. Fig.~\ref{fig-arg-in-pedigree}b shows the bare gARG, illustrating
+that the patterns of
 transmission between genomes can be followed without necessarily knowing
@@
 along the genome (Fig.~\ref{fig-arg-in-pedigree}c).
-
-The focus on modelling genomes rather than events has some interesting
-and important consequences. Firstly, the graph may contain nodes that
-do not correspond to any particular event in the ancestry of the samples.
-For example, node \textsf{H} is the direct ancestor of \textsf{D} in
-Fig~\ref{fig-arg-in-pedigree}, and is not the product of either a
-coalescence or a recombination. Such nodes would usually be removed
-from the graph so that \textsf{D} descends directly from \textsf{K} (see
-section XXX) but there is no necessity for this from a representational
-perspective. This ability can be useful in practice and could be required if,
-% Do we need to spell this out?
-for example, we have a deeply sequenced pedigree of
-haplodiploids. % e.g. Honeybees? Be great to cite an example!
-
-Secondly, the way that we model recombination is subtly different.
-Rather than having a single node that represents the \emph{event},
-there are three nodes (representing the recombinant offspring and
-two parental genomes) and two edges with associated genome coordinate
-intervals (representing the precise inheritance patterns).
-For example, in Fig~\ref{fig-arg-in-pedigree}, node \textsf{C} is the
-recombinant descendent of \textsf{E} and \textsf{F}, inheriting positions
-0 to 7 from $E$ and genome positions 7 to 10 from $F$. Any particular
-node can be both a recombinant (it is the recombined product of
-its parent's genomes) and a common ancestor (the ancestry from
-multiple samples meet at this node; see the next section).
-% Here was can talk about representing the recombination by a
-% recombinant and 2 parent "recombination nodes" rather than
-% a single "recombination node" as in the normal ARG representation.
-% We may also wish to note that in this example the recombination nodes
-% are shared between siblings, and can also be coalescent nodes
-% this ties in with stuff discussed 3 paras below.
-[More stuff here: clarify why we assign coordinates to edges
-rather than nodes. Basically lay out all the stuff that we can
-talk about \emph{without} going in to the details of
-ancestry resolution.]
 
@@
 
-% Asking the reader to flick back to figure 2 here, not great.
-The differences between EARGs
-and GARGs are subtle but important, and illustrated in
-Fig.~\ref{fig-arg-data-structure}, which shows a simple ARG
-with three samples and one recombination.
-Fig.~\ref{fig-arg-data-structure}A shows this topology as a
-classical Griffiths-like EARG, and
-Fig.~\ref{fig-arg-data-structure}B shows it as a GARG.
-The two are almost identical, except we model the recombination
-event via the two parental genomes in nodes $3$ and $4$ and explicitly
-store the genomic interval that the child inherited from the
-parent with every edge.
-This approach may seem wasteful since we must store an extra
-node for each recombination and genomic interval information
-with every edge (even though many simply span the full genome).
+The focus on genomes rather than events has some interesting
+and important ramifications. Firstly, a gARG may contain nodes that
+do not correspond to any particular event in the ancestry of the samples.
+For example, node \textsf{H} is the direct ancestor of \textsf{D} in
+Fig~\ref{fig-arg-in-pedigree}, and is not the product of either a
+coalescence or a recombination. Such nodes would usually be removed
+from the graph so that \textsf{D} descends directly from \textsf{K} (see
+section XXX) but there is no necessity for this from a representational
+perspective. This ability can be useful in practice and could be required if,
+% Do we need to spell this out?
+for example, we have a deeply sequenced pedigree of
+haplodiploids. % e.g. Honeybees? Be great to cite an example!
+
+Secondly, unlike in an eARG, nodes in a gARG are not classified into different
+types. Indeed a single node can have both multiple children
+and multiple parents. This would be the case in Fig~\ref{fig-arg-in-pedigree} if,
+for instance, node \textsf{A} were to have two children.
+% It is a shame that this is not actually shown in fig 3, but I don't think
+% we can do this unless we have 5 generations
+Moreover, a single node can have more than 2 children, and (as we shall see in
+section XXX), more than 2 parents. Another way to state this is that transitions
+between nodes in a gARG can encapsulate multiple events in an eARG.
+
+The final point deals with the way in which recombination is represented, and is
+closely linked to which genomes are chosen as the nodes in a gARG.
+We could potentially choose genomes at any point in the life cycle: for instance,
+we could chose to represent gametes as ARG nodes. However, the most intuitive
+choice is where genomes represent diploid individuals in each generation, as in
+Fig~\ref{fig-arg-in-pedigree}a. In this case recombination
+is not represented by a single ``recombination node'' (as in an eARG), but with
+two parent genomes between which recombination takes place (e.g.
+nodes \textsf{E} and \textsf{F}), and a
+single resulting ``recombinant'' genome (e.g. \textsf{C}). The
+edges that link this recombinant with its two parents carry the information
+concerning breakpoint location that would normally be present on the
+``recombination nod'' in an eARG. For example, node \textsf{C} in
+Fig~\ref{fig-arg-in-pedigree} inherits genome positions 0 to 7 from \textsf{E}
+and positions 7 to 10 from \textsf{F}, indicating that the breakpoint is at position 7.
+
+This method for representing recombinations may seem wasteful,
+since more than one node is required for each recombination,
+and genomic interval information must be stored on every edge.
 However, asymptotically the amount of information stored is the
-same, and as we see in the next section, keeping interval
-information for all edges enables useful generalisations.
-There is also much simplicity to be gained by having
+same, and there is much simplicity to be gained by having
 a single type of node and by treating all edges in the same way.
+More importantly --- as we see in the next section --- we gain the
+ability to simplify the ARG by removing nodes without affecting the
+local tree topologies, and also the ability to record only the
+regions of genome relevant to the final, sample genomes. This last
+feature turns out to be the key to efficient processing of the ARG.
+
+[Possibly more stuff here: clarify why we assign coordinates to edges
+rather than nodes. Basically lay out all the stuff that we can
+talk about \emph{without} going in to the details of
+ancestry resolution.]
 
@@
 % AI: have not mentioned any implicit or un-rooted networks (split, median, haplotype, reticulograms, etc).
-% Also: are we not defining ARGs as "binary networks"? Might be good to specify this here.
+% Also: are we not defining ARGs as ``binary networks''? Might be good to specify this here.
 
@@
 \caption{\label{fig-simplification}
-ARG simplification. (a) An ARG and equivalent local trees simulated
-from a diploid Wright-Fisher
+ARG simplification. (A) An ARG simulated from a diploid Wright-Fisher
 model. Note the complex full-sib relationship between B and C; this
@@
 (see section XXX).
-(b) Simplified to remove all
+(B) Simplified to remove all
 % trying out this terminology - can define in the text
 1-connected graph components (e.g., diamonds).
-(c) Remove nodes that are unary in all local trees.
-(d) Rewrite edges to skip any unary nodes in local trees.
+(C) Remove nodes that are unary in all local trees.
+(D) Rewrite edges to skip any unary nodes in local trees.
+(E) The local trees for ARG (A).
+(F) The local trees for ARG (D).
 }
@@
 
-% Q: would it help to talk about "tree vertices" here or would it
+% Q: would it help to talk about ``tree vertice'' here or would it
 % just be more confusing? I worry that people would think that
@@
 presence of ``unary'' nodes: a node in the tree that has exactly
-one child. For example, Fig.~\ref{fig-simplification} shows a set of
-four simulated gARGs and the local trees that can be constructed from
-them (for simplicity, edge annotations are not shown)
-Consider the right-most tree in Fig.~\ref{fig-simplification}a
+one child. For example, Fig.~\ref{fig-simplification}E shows the
+local trees along the genome for a simulated ARG
+(Fig.~\ref{fig-simplification}A). Consider the right-most tree,
 and the path above node $D$. Before $D$ meets its common ancestor
@@
 join again into a common ancestor (e.g. $K,N,O$ and $R$ in
-Fig.~\ref{fig-simplification}a).
+Fig.~\ref{fig-simplification}A).
 Unless we are specifically
@@
 of a single node that has one inbound and one outbound edge.
-Fig.~\ref{fig-simplification}b shows the result of this type of
+Fig.~\ref{fig-simplification}B shows the result of this type of
 graph topology simplification.
@@
 parents of the recombinant node $D$ is still present in
-Fig.~\ref{fig-simplification}b after diamond removal, but
+Fig.~\ref{fig-simplification}B after diamond removal, but
 it is unary in all of the local trees since it has an
 in-degree of $1$.
-Fig.~\ref{fig-simplification}c shows the resulting graph
+Fig.~\ref{fig-simplification}C shows the resulting graph
 after removing these nodes.
@@
 local tree. For example, node $E$ is unary in the first tree
-of Fig.~\ref{fig-simplification}a, but is either binary
+of Fig.~\ref{fig-simplification}E, but is either binary
 or ternary in the other trees (recall this is a Wright-Fisher
@@
 The final level of simplification is to remove \emph{all}
-unary nodes from the local trees (Fig.~\ref{fig-simplification}d).
-[And continue]
+unary nodes from the local trees (Fig.~\ref{fig-simplification}D
+and E). [And continue]
 
@@
 
-Due to the computational expense of sampling ``little ARGs", state-of-the-art
+Due to the computational expense of sampling ``little ARGs'', state-of-the-art
 Monte Carlo methods typically rely on cheaper, approximate models.

### 2022-03-10 paper.tex

@@
 
-[We should discuss the relationship between ARG nodes and SPRs
-here. Cite \citep{deng2021distribution} too. Maybe we want to make
-this a separate section, as it's a subtle and important point.]
+\section*{ARGs and SPRs}
+\citet{wiuf1999ancestry,wiuf1999recombination} recast
+the coalescent with recombination
+as a stocastic process along the genome rather than backwards in time.
+In this formulation we begin at one end of a chromosome by
+sampling a tree under the standard coalescent, and then reason
+about how this tree is modified by recombination events
+as we move along the genome. Although this algorithm is
+significantly more complicated to simulate than Hudson's
+backwards in time approach, it has led to substantial
+insights into the process of recombination itself
+and to the sequentially Markov Coalescent (SMC)
+approximations~\citep{mcvean2005approximating,marjoram2006fast}
+which are central to many modern methods.
+
+Under the coalescent with recombination (or SMC) each local
+tree along the genome is transformed into the next by
+a ``subtree prune-and-regraft'' (SPR)
+% Hein et al call it the "subtree transfer" operation but it's
+% the same thing.
+operation~\citep{hein1990reconstructing,song2003on,song2006properties}.
+The number of SPRs required to transform one leaf-labelled tree
+into another is known as the SPR distance
+% Hein 1996 say that SPR distance is NP-hard, but Allen 2001
+% say this is wrong. But then Bordewich went ahead and said
+% yes, actually it is NP hard.
+and is NP-hard to
+compute~\citep{hein1996complexity,allen2001subtree,bordewich2005computational}.
+
+[Discussion of SPRs in an EARG]
+
+The effects of recombination as we move from left-to-right along
+the genome in an unsimplified GARG are easy to interpret.
+Consider the leftmost tree in Fig~\ref{fig-simplification}.
+In the first tree the parent of B is F, and in the second
+tree its parent is E. Since both parents are present as
+unary nodes in the tree, we simply change one edge in
+order to effect the transition.
+A more complex transition occurs when we have a regraft operation
+that occurs above the current root (i.e., the current oldest
+non-unary node). For example, the transition between the
+second and third trees corresponds to node D switching
+from the path through I to H, and we then must add branches
+corresponding to passing through L and meeting at M.
+
+The equivalence between an ARG and a set of local trees
+separated by SPRs is often mentioned [e.g. x y z], but
+it is important to note that while extracting local
+trees from an ARG is straightforward, the converse
+process of reconstructing precisely the same ARG from
+these trees is more problematic. If the internal
+nodes are consistently labelled and unary nodes are
+included in the local trees (e.g., Fig~\ref{fig-simplification})
+then it is easy to exactly reconstruct the ARG.
+Otherwise, the problem is much more difficult and it
+may not be possible to uniquely reconstruct the original ARG.
+Without labels for the internal nodes, we are forced to
+to reason about which subtrees remain the same and which
+differ due to the SPR separating adjacent trees in order to
+assign node identity. Without the internal unary nodes marking
+recombinations, we must infer their existence. In
+general, there will not be a unique solution to this problem,
+and so there can be many possible ARGs compatible with
+a given list of
+% want to say "standard phylogenetic" trees here or something, but
+% and they don't have to be binary (just arity > 1), but thought
+% this would be familiar to readers?
+% Jere: Could just say leaf-labelled "marginal trees", "ancestral trees",
+% "local trees", or something like that.
+leaf-labelled binary trees.
+
+[ Cite \citep{deng2021distribution} somewhere here.]
 

### 2022-03-10 paper.tex

@@
 Monte Carlo methods typically rely on cheaper, approximate models.
-The most prominent examples are \texttt{ARGWeaver} \citep{rasmussen2014genome}
+The most prominent examples are
+\texttt{ARGWeaver}~\citep{rasmussen2014genome,hubisz2020mapping,hubisz2020inference}
 and \texttt{Arbores} \citep{heine2018bridging}. The former also utilises a time
@@
 by making use of the tree sequence representation of marginal trees, suitably enriched
-to facilitate the computations needed for MCMC sampling \citep{mahmoudi2021inference}.
+to facilitate the computations needed for MCMC sampling
+\citep{mahmoudi2022bayesian}.
 While not competitive with the scalability of \texttt{ARGWeaver}, \texttt{ARGInfer} extends

### 2022-03-10 argutils/convert.py
Start on fig 6

@@ -101,7 +101,8 @@ def convert_kwarg(infile, num_samples, sequence_length):
     node_ids = {}
     for n in range(num_samples):
         node_ids[n] = n
-        tsk_id = tables.nodes.add_row(flags=tskit.NODE_IS_SAMPLE, time=0.0, metadata=n)
+        tsk_id = tables.nodes.add_row(
+            flags=tskit.NODE_IS_SAMPLE, time=0.0, metadata={"sample": n})
 
     for x in infile:
         line = x.split()
@@ -112,7 +113,7 @@ def convert_kwarg(infile, num_samples, sequence_length):
             c1 = node_ids[int(line[2]) - 1]
             c2 = node_ids[int(line[4]) - 1]
             time += 1
-            tsk_id = tables.nodes.add_row(flags=0, time=time, metadata=line[2])
+            tsk_id = tables.nodes.add_row(flags=0, time=time, metadata={"coal": line[2]})
             node_ids[int(line[2]) - 1] = tsk_id
             tables.edges.add_row(0, sequence_length, tsk_id, c1)
             tables.edges.add_row(0, sequence_length, tsk_id, c2)
@@ -120,9 +121,9 @@ def convert_kwarg(infile, num_samples, sequence_length):
             breakpoint = int(line[6][:-1]) - 1
             n1 = node_ids[int(line[3]) - 1]
             time += 1
-            tsk_id_1 = tables.nodes.add_row(flags=0, time=time, metadata=line[3])
+            tsk_id_1 = tables.nodes.add_row(flags=0, time=time, metadata={"rec": line[3]})
             node_ids[int(line[3]) - 1] = tsk_id_1
-            tsk_id_2 = tables.nodes.add_row(flags=0, time=time, metadata=line[11])
+            tsk_id_2 = tables.nodes.add_row(flags=0, time=time, metadata={"rec": line[11]})
             node_ids[int(line[11]) - 1] = tsk_id_2
             tables.edges.add_row(0, breakpoint, tsk_id_1, n1)
             tables.edges.add_row(breakpoint, sequence_length, tsk_id_2, n1)

### 2022-03-10 illustrations.py

@@ -1,6 +1,7 @@
 import pathlib
 import string
 import io
+import json
 from types import SimpleNamespace
 
 import click
@@ -11,6 +12,7 @@ import matplotlib.pyplot as plt
 import numpy as np
 import scipy.stats
 import tskit
+import tsinfer
 
 import argutils
 
@@ -470,9 +472,68 @@ def arg_in_pedigree():
         f.write(top_svg)
 
 
+@click.command()
+def inference():
+    tree_seqs = {}
+    # KwARG
+    with open("examples/kreitman.kwarg") as f:
+        ts = argutils.convert_kwarg(f, 11, 43)
+        # ts = argutils.simplify_keeping_unary_in_coal(ts)  # in case we want to compare with tsinfer
+        ts = argutils.viz.label_nodes(ts)
+        tree_seqs["KwARG"] = ts
+
+    # Tsinfer
+    with open("examples/kreitman_snp.txt") as file:
+        data = np.array([[int(d) for d in line.strip()] for line in file])
+    with tsinfer.SampleData(sequence_length=data.shape[1]) as sample_data:
+        for pos, sites in enumerate(data.T):
+            sample_data.add_site(pos, sites)
+    ts = tsinfer.infer(sample_data)
+    # For the moment, tsinfer uses byte metadata, so we need to convert it to json
+    tables = ts.dump_tables()
+    tables.nodes.metadata_schema = tskit.MetadataSchema.permissive_json()
+    for n in ts.nodes():
+        tables.nodes[n.id] = tables.nodes[n.id].replace(
+            metadata = json.loads(n.metadata.decode() or "{}"))
+    ts = tables.tree_sequence()
+    tree_seqs["Tsinfer"] = argutils.viz.label_nodes(ts)
+
+    fig, axes = plt.subplots(1, 2, figsize=(10, 5))
+    col = mpl.colors.to_hex(plt.cm.tab20(1))
+    for ax, (name, ts) in zip(axes, tree_seqs.items()):
+        pos, G = argutils.viz.draw(
+            ts, ax,
+            use_ranked_times=True if name == "Tsinfer" else None,
+            draw_edge_widths=True,
+            node_color=col,
+            max_edge_width=2)
+        ax.set_title(name)
+
+    graph_io = io.StringIO()
+    plt.savefig(graph_io, format="svg", bbox_inches="tight")
+    graph_svg = graph_io.getvalue()
[…]

### 2022-03-10 argutils/viz.py
Allow nonsample nodes to be shrunk without labels

@@ -21,6 +21,7 @@ def draw(
     draw_edge_widths=False,
     max_edge_width=5,
     draw_edge_alpha=False,
+    nonsample_node_shrink=None,
     node_size=None,
     font_size=None,
     node_color=None,
@@ -59,6 +60,10 @@ def draw(
     If draw_edge_alpha is True, draw edges with an alpha value equal to their
     genomic span / the total sequence length of the tree sequence.
 
+    If "nonsample_node_shrink" is not None, it should be an integer giving the
+    amount by which nonsample node symbols are reduced; in this case labels on
+    the nodes are omitted
+    
     node_size, font_size, font_color, and node_color are all passed to nx.draw
     directly. In particular this means that node_color can either be a single
     colour for all nodes (e.g. `mpl.colors.to_hex(mpl.pyplot.cm.tab20(1))`)
@@ -69,15 +74,23 @@ def draw(
     """
     if node_size is None:
         node_size = 200
+    if nonsample_node_shrink is not None:
+        # Shrink the nonsample label size
+        node_size_array = np.full(ts.num_nodes, node_size / nonsample_node_shrink)
+        node_size_array[ts.samples()] = node_size
+        node_size = node_size_array
     if font_size is None:
         font_size = 9
     G = convert_nx(ts)
     labels = {}
     for nd in ts.nodes():
-        try:
-            labels[nd.id] = str(nd.metadata["name"])
-        except (TypeError, KeyError):
-            labels[nd.id] = str(nd.id)
+        if nonsample_node_shrink is not None and not nd.is_sample():
+            labels[nd.id] = ""
+        else:
+            try:
+                labels[nd.id] = str(nd.metadata["name"])
+            except (TypeError, KeyError):
+                labels[nd.id] = str(nd.id)
 
     if pos is None:
         pos = nx_get_dot_pos(G, reverse_x_axis)
@@ -129,9 +142,10 @@ def draw(
     edges = nx.draw_networkx_edges(
         G,
         pos,
-        edgelist=G.edges(),
+        edgelist=list(G.edges()),
         arrowstyle="-|>" if arrows else "-",
         ax=ax,
+        node_size=node_size,
         width=edge_widths
     )
     if edge_alpha is not None:

### 2022-03-10 illustrations.py
Do not show nonsample node labels in fig 6

@@ -503,6 +503,7 @@ def inference():
     for ax, (name, ts) in zip(axes, tree_seqs.items()):
         pos, G = argutils.viz.draw(
             ts, ax,
+            nonsample_node_shrink=5,
             use_ranked_times=True if name == "Tsinfer" else None,
             draw_edge_widths=True,
             node_color=col,

### 2022-03-11 illustrations.py
Add files for ARGweaver inference

@@ -230,6 +230,7 @@ def simplification():
         11: 5,
         8: 10,
     }
+
     ts = argutils.sim_wright_fisher(2, 10, 100, recomb_proba=0.1, seed=seed)
     tables = ts.dump_tables()
     rank_times = scipy.stats.rankdata(tables.nodes.time, method="dense")
@@ -476,34 +477,12 @@ def arg_in_pedigree():
 def inference():
     tree_seqs = {}
     # KwARG
-    with open("examples/kreitman.kwarg") as f:
-        ts = argutils.convert_kwarg(f, 11, 43)
-        # ts = argutils.simplify_keeping_unary_in_coal(ts)  # in case we want to compare with tsinfer
-        ts = argutils.viz.label_nodes(ts)
-        tree_seqs["KwARG"] = ts
+    ts = tskit.load("examples/Kreitman_SNP_kwarg.trees")
+    # ts = argutils.simplify_keeping_unary_in_coal(ts)  # in case we want to compare with tsinfer
+    tree_seqs["KwARG"] = argutils.viz.label_nodes(ts)
 
     # Tsinfer
-    with open("examples/kreitman_snp.txt") as file:
-        data = np.array([[int(d) for d in line.strip()] for line in file])
-    with tsinfer.SampleData(sequence_length=data.shape[1]) as sample_data:
-        for pos, sites in enumerate(data.T):
-            sample_data.add_site(pos, sites)
-    ts = tsinfer.infer(sample_data)
-    # For the moment, tsinfer uses byte metadata, so we need to convert it to json
-    tables = ts.dump_tables()
-    tables.nodes.metadata_schema = tskit.MetadataSchema.permissive_json()
-    for n in ts.nodes():
-        tables.nodes[n.id] = tables.nodes[n.id].replace(
-            metadata = json.loads(n.metadata.decode() or "{}"))
-    # temporary hack: remove the ultimate ancestor if it exists
-    oldest_node = np.argmax(tables.nodes.time)
-    if np.sum(tables.edges.parent==oldest_node) == 1:
-        # only a single edge connects to the root. This is a unary "ultimate ancestor"
-        # and can be removed (it will be removed in later tsinfer versions anyway)
-        use = np.arange(tables.nodes.num_rows)
-        use = use[use != oldest_node]
-        tables.subset(use)
-    ts = tables.tree_sequence()
[…]

### 2022-03-15 paper.tex
Start on fig 6

@@
 the coalescent with recombination
-as a stocastic process along the genome rather than backwards in time.
+as a stochastic process along the genome rather than backwards in time.
 In this formulation we begin at one end of a chromosome by
@@
 
+\section*{Evaluation of sampling probabilities}
+
+The conditional sampling probability $p(G | D)$ of an ARG $G$ given an observed
+realization of genetic diversity $D$ at its sampled leaves is a central concept in
+model fitting and selection. Under the Bayesian paradigm, it defines the posterior
+distribution $p(G | D) \propto p(D | G) p(G)$, where $p(D | G)$ is the likelihood
+of the data $D$ given $G$, and $p(G)$ is a prior distribution for the underlying
+ARG $G$. Under the frequentist paradigm, $p(G)$ can be interpreted as a regularizer,
+and $p(G | D)$ as a regularized likelihood. In both cases, $p(G)$ is typically specified
+as the sampling distribution of a stochastic process, such as the coalescent with
+recombination. Essentially all of the model-based inference methods surveyed below
+rely on evaluation of a likelihood and a prior. Two contemporary examples of
+likelihood-based methods are  \cite{mahmoudi2022bayesian, guo2022recombination}.
+
+In essentially all neutral cases in practical use, the likelihood $p(D | G)$ is the law of
+a Poisson process on the ancestral material on the edges of the ARG, which implies
+that mutation processes at two sites are conditionally independent given the
+underlying ancestral trees at those sites. A prototypical example is 
+\citet[Eq.\ (2)]{mahmoudi2022bayesian}. Hence, $p(D | G)$ can be evaluated
+regardless of the precise data structure used to store $G$, provided that it encodes
+at least all marginal ancestral trees along the sampled sequences.  Choosing the
+right data structure can still have a dramatic effect on the efficiency with which
+$p(D | G)$ can be evaluated \citep{mahmoudi2022bayesian}.
+
+In contrast, evaluating $p(G)$ under the coalescent with recombination
+(or related models) requires knowledge of the number of extant lineages and
+the number of links available for recombination (i.e.\ ones at which a recombination
+would split ancestral material) at each time along the ARG
+\citep[Eq.\ (3)]{mahmoudi2022bayesian}. These cannot be uniquely determined
+from just the marginal trees at each site because branches at multiple trees can be
+contained in the same ancestral lineage. Instead, a richer data structure akin to the
+EARG or GARG is needed. Moreover, because EARG does not explicitly store the flow
+of ancestral material along its edges---a necessary piece of information for
+computing the number of effective recombination links at each time---evaluating
+$p(G)$ based on an EARG essentially requires turning it into a GARG at runtime.
+
+The number of extant lineages at a given time is not explicitly represented even
+in a GARG, and needs to be calculated from the node and edge tables.
+For example, any edges with the same child and parent nodes are contained within
+the same lineage. Under the coalescent with recombination, the two edges
+originating from a recombinant child are also contained within a single lineage
+until the time of their parent nodes, at which point they split. Nodes S, T, and U
+in panel (a) of Figure \ref{fig-ancestry-resolution} provide a concrete example
+of the latter: the edges from S to T and S to U are distinct, but correspond to only
+a single lineage in the time interval they span. We emphasize that these
+conventions grouping edges into lineages are not part of the GARG data structure.
+While they are convenient for analysis using the coalescent with recombination,
+other models, such as discrete time Wright--Fisher models with finite populations,
+may benefit from different ways to associate edges with lineages.
+
 \section*{ARG inference methods}
@@
 \begin{figure}
-\vspace{5em}
+\begin{center}
+    \includegraphics[width=\textwidth]{illustrations/inference.pdf}
+\end{center}
 \caption{\label{fig-inferred-args}

### 2022-03-18 argutils/viz.py
Add Relate data

@@ -35,7 +35,7 @@ def draw(
     called "name" exists for the node, it is taken as
     a node label, otherwise the node ID will be used as a label instead.
 
-    If use_ranked times is True, the y axis uses the time ranks, with the
+    If use_ranked_times is True, the y axis uses the time ranks, with the
     same times sharing a rank. If False, it uses the true (tree sequence)
     times. If None, times from the tree sequence are not used and the
     standard dot layout is used.

### 2022-03-18 illustrations.py

@@ -509,6 +509,17 @@ def inference():
     labels = {n.id: n.metadata["name"] if n.is_sample() else "" for n in ts.nodes()}
     tree_seqs["ARGweaver"] = argutils.viz.label_nodes(ts, labels=labels)
 
+    # Relate JBOT
+    ts = tskit.load("examples/Kreitman_SNP_relate_jbot.trees")
+    tables = ts.dump_tables()
+    tables.nodes.metadata_schema = tskit.MetadataSchema.permissive_json()
+    for i, n in enumerate(tables.nodes):
+        tables.nodes[i] = n.replace(metadata={})
+    ts = tables.tree_sequence()
+    # labels not stored by default in the Relate ts metadata, so use the previous ones
+    labels = {n.id: labels[n.id] if n.is_sample() else "" for n in ts.nodes()}
+    tree_seqs["Relate JBOT"] = argutils.viz.label_nodes(ts, labels=labels)
+
     fig, axes = plt.subplots(1, len(tree_seqs), figsize=(10, 5))
     col = mpl.colors.to_hex(plt.cm.tab20(1))
     for ax, (name, ts) in zip(axes, tree_seqs.items()):
@@ -518,6 +529,8 @@ def inference():
             use_ranked_times = None
         if name == "KwARG":
             use_ranked_times = None
+        if name == "Relate JBOT":
+            use_ranked_times = True
         pos, G = argutils.viz.draw(
             ts, ax,
             node_size=30,

### 2022-03-18 make_example_inputs.py

@@ -73,11 +73,18 @@ def cli():
 
 @click.command()
 def tsinfer_input():
+    """
+    Create a ".sample" file in the format required by tsinfer
+    (see https://tsinfer.readthedocs.io/en/latest/tutorial.html)
+    """
     saved_samples = sample_data.copy(path="examples/Kreitman_SNP.samples")
     saved_samples.finalise()
 
 @click.command()
 def kwarg_input():
+    """
+    Create a ".matrix" file in the format required by kwarg
+    """
     # a simple 0/1 sites-by-samples matrix
     with open("examples/Kreitman_SNP.matrix", "wt") as file:
         for row in sample_data.sites_genotypes[:].T:
@@ -86,6 +93,9 @@ def kwarg_input():
 
 @click.command()
 def argweaver_input():
+    """
+    Create a ".sites" file in the format required by ARGweaver
+    """
     with open("examples/Kreitman_SNP.sites", "wt") as file:
         print(
             "NAMES",
@@ -106,9 +116,37 @@ def argweaver_input():
                 file=file,
             )
 
+@click.command()
+def relate_input():
+    """
+    Create a ".haps" file and a ".sample" file in the format required by relate
+    (see https://myersgroup.github.io/relate/input_data.html)
+    """
+    with open("examples/Kreitman_SNP.haps", "wt") as file:
+        for v in sample_data.variants():
+            assert len(v.alleles) == 2
+            print(
+                "2",
+                f"SNP{v.site.id}",
+                int(v.site.position),
+                v.alleles[0],
+                v.alleles[1],
+                " ".join([str(g) for g in v.genotypes]),
+                sep=" ",
+                file=file,
+            )
+    with open("examples/Kreitman_SNP.sample", "wt") as file:
+        print("ID_1 ID_2 missing", file=file)
+        print("0    0    0", file=file)
+        for i in sample_data.individuals():
+            assert len(i.samples) == 1
+            assert i.id == i.samples[0]
+            print(f'{i.metadata["name"].replace("-", "")} NA 0', file=file)
+
 cli.add_command(tsinfer_input)
 cli.add_command(kwarg_input)
 cli.add_command(argweaver_input)
+cli.add_command(relate_input)
 
 if __name__ == "__main__":
     cli()

### 2022-03-18 paper.tex

@@
 \node [nodelabel,anchor=north west] at ($(s3) + (0,0)$) {$x$};
-\foreach \u/\lab in {s0/0, s1/1, s2/2} \node[nodelabel,anchor=north] at (\u) {\lab};
-\foreach \u/\lab in {s4/4} \node[nodelabel,anchor=south west] at (\u) {\lab};
-\foreach \u/\lab in {s5/5} \node[nodelabel,anchor=south east] at (\u) {\lab};
-\foreach \u/\lab in {s3/3, s6/6} \node[nodelabel,anchor=south] at (\u) {\lab};
+\foreach \u/\lab in {s0/$\textsf{A}$, s1/$\textsf{B}$, s2/$\textsf{C}$} \node[nodelabel,anchor=north] at (\u) {\lab};
+\foreach \u/\lab in {s4/$\textsf{E}$} \node[nodelabel,anchor=south west] at (\u) {\lab};
+\foreach \u/\lab in {s5/$\textsf{F}$} \node[nodelabel,anchor=south east] at (\u) {\lab};
+\foreach \u/\lab in {s3/$\textsf{D}$, s6/$\textsf{G}$} \node[nodelabel,anchor=south] at (\u) {\lab};
 
@@
 
-\node [nodelabel,anchor=north west] at ($(-8.5,5)$) {
+\node [nodelabel,anchor=north west] at ($(-10,5)$) {
 \begin{tabular}{c|c}
@@
 \hline
-0 & $\varnothing$ \\
-1 & $\varnothing$ \\
-2 & $\varnothing$ \\
-3 & $x$ \\
-4 & $\varnothing$ \\
-5 & $\varnothing$ \\
-6 & $\varnothing$ \\
+$\textsf{A}$ & $\varnothing$ \\
+$\textsf{B}$ & $\varnothing$ \\
+$\textsf{C}$ & $\varnothing$ \\
+$\textsf{D}$ & $x$ \\
+$\textsf{E}$ & $\varnothing$ \\
+$\textsf{F}$ & $\varnothing$ \\
+$\textsf{G}$ & $\varnothing$ \\
 \end{tabular}};
 
-\node [nodelabel,anchor=north west] at ($(-2.5,5)$) {
+\node [nodelabel,anchor=north west] at ($(-4,5)$) {
 \begin{tabular}{l}
@@
 \hline
-$(0, 4)$ \\
-$(1, 3)$ \\
-$(2, 5)$ \\
-$(3, 4)$ \\
-$(3, 5)$ \\
-$(4, 6)$ \\
-$(5, 6)$ \\
+$(\textsf{A}, \textsf{E})$ \\
+$(\textsf{B}, \textsf{D})$ \\
+$(\textsf{C}, \textsf{F})$ \\
+$(\textsf{D}, \textsf{E})$ \\
+$(\textsf{D}, \textsf{F})$ \\
+$(\textsf{E}, \textsf{G})$ \\
+$(\textsf{F}, \textsf{G})$ \\
 \end{tabular}};
 
-\node [nodelabel,anchor=north west] at ($(6,5)$) {
+\node [nodelabel,anchor=north west] at ($(7,5)$) {
 \begin{tabular}{ll}
@@
 \hline
-$(0, 4)$ & $(0, L]$ \\
-$(1, 3)$ & $(0, L]$ \\
-$(2, 5)$ & $(0, L]$ \\
-$(3, 4)$ & $(0, x]$ \\
-$(3, 5)$ & $(x, L]$ \\
-$(4, 6)$ & $(0, L]$ \\
-$(5, 6)$ & $(0, L]$ \\
+$(\textsf{A}, \textsf{E})$ & $(0, L]$ \\
+$(\textsf{B}, \textsf{D})$ & $(0, L]$ \\
+$(\textsf{C}, \textsf{F})$ & $(0, L]$ \\
+$(\textsf{D}, \textsf{E})$ & $(0, x]$ \\
+$(\textsf{D}, \textsf{F})$ & $(x, L]$ \\
+$(\textsf{E}, \textsf{G})$ & $(0, L]$ \\
+$(\textsf{F}, \textsf{G})$ & $(0, L]$ \\
 \end{tabular}};

### 2022-03-24 illustrations.py
add breakpoints to fig 6

@@ -9,10 +9,10 @@ import networkx as nx
 import PIL
 import matplotlib as mpl
 import matplotlib.pyplot as plt
+import matplotlib.patches as patches
 import numpy as np
 import scipy.stats
 import tskit
-import tsinfer
 
 import argutils
 
@@ -483,6 +483,8 @@ def inference():
     Examples of ARGs produced by various inference algorithms
     """
     tree_seqs = {}
+    # Assign colormap for edges
+    edge_colormap = plt.cm.viridis
 
     # KwARG
     ts = tskit.load("examples/Kreitman_SNP_kwarg.trees")
@@ -519,30 +521,54 @@ def inference():
     labels = {n.id: labels[n.id] if n.is_sample() else "" for n in ts.nodes()}
     tree_seqs["Relate JBOT"] = argutils.viz.label_nodes(ts, labels=labels)
 
-    fig, axes = plt.subplots(1, len(tree_seqs), figsize=(10, 5))
-    for ax, (name, ts) in zip(axes, tree_seqs.items()):
-        tweak_x={}
+    widths = np.ones(4)
+    heights = [1, 0.1]
+    fig, axes = plt.subplots(2, len(tree_seqs), figsize=(10, 6),
+                             gridspec_kw=dict(
+                             width_ratios=widths, height_ratios=heights))
+    tree_seq_positions = []
+    edge_colors_by_ts = {}
+    max_rank_by_ts = {}
+    for ax, (name, ts) in zip(axes[0], tree_seqs.items()):
+        tweak_x = {}
         use_ranked_times = None  # By default use Y axis layout from graphviz dot
         if name == "Tsinfer":
-            tweak_x={
+            tweak_x = {
                 21: 12, 18: -12, 22: 15, 19: 5, 12: -3, 13: 2, 25: 5,
                 23: 0, 11: 12, 23: 52, 5: 110, 25: 30, 20: 5, 24: 10, 17: -5
             }
         if name == "ARGweaver ex.":
-            tweak_x={
+            tweak_x = {
                 54: -10
             }
         if name == "KwARG":
-            tweak_x={
+            tweak_x = {
                 18: 5, 17: 7, 16: 5, 13: -2, 20: -7, 19: -8, 11: -5, 12: -5, 35: -3,
                 22: -5, 23: -5, 33: 10, 27: 5, 32: 8, 29: -2, 28: 2, 26: 5, 25: 3,
             }
[…]

### 2022-03-24 argutils/viz.py
Change colours to be more blue for 4 & 5 parent nodes

@@ -19,8 +19,8 @@ def arity_colors(n_parents):
         return colorcet.cm.CET_I1(255)  # Red
     if n_parents == 1:
         return colorcet.cm.CET_I1(200)  # Yellow
-    # Max out at full blue if 20 or more parents, use log scale
-    multiple_parents_val = (min(np.log2(n_parents), np.log2(10)) - 1) / (np.log2(10) - 1)
+    # Max out at full blue if 6 or more parents; use log scale
+    multiple_parents_val = (min(np.log2(n_parents), np.log2(6)) - 1) / (np.log2(6) - 1)
     return colorcet.cm.CET_I1(100 - int(multiple_parents_val * 100))
 
 

### 2022-03-24 argutils/viz.py
Use dense ranked times

@@ -11,6 +11,7 @@ import matplotlib as mpl
 import networkx as nx
 import numpy as np
 import pydot
+import scipy.stats
 
 
 def arity_colors(n_parents):
@@ -143,8 +144,7 @@ def draw(
 
     if use_ranked_times is not None:
         if use_ranked_times:
-            _, inv = np.unique(ts.tables.nodes.time, return_inverse=True)
-            ranked_times = (np.cumsum(np.bincount(inv)) - 1)[inv]
+            ranked_times = scipy.stats.rankdata(ts.tables.nodes.time, method="dense")
             for i, p in pos.items():
                 p[1] = ranked_times[i]
         else:

### 2022-03-24 paper.tex

@@
 
-Another limitation of the Griffiths ARG encoding illustated in
+Another limitation of the Griffiths ARG encoding illustrated in
 Fig.~\ref{fig-arg-data-structure}A is the inability to represent
@@
 and the interval $(x, L)$ with $(3, 5)$ we make the encoding
-unambigous: we can list the edges in any order, and it
+unambiguous: we can list the edges in any order, and it
 will describe the same set of local genealogical trees.
@@
 of inheritance can be described by these intervals. For
-example, if a gene conversion event occured where a segment
+example, if a gene conversion event occurred where a segment
 between between positions $x$ and $y$ was inherited by

### 2022-03-24 argutils/convert.py
Use alternative Relate conversion script

@@ -144,63 +144,86 @@ def convert_kwarg(infile, num_samples, sequence_length, sample_names=None):
     return argutils.simplify_keeping_all_nodes(ts)
 
 
-def convert_relate_without_times(infile):
+def relate_ts_JBOT_to_ts(ts, additional_equivalents=None):
     """
-    Convert a Relate .anc file to a tree sequence, but with time_units="uncalibrated"
-    which allows us to create an proper ARG rather than a JBOT. See
-    https://myersgroup.github.io/relate/getting_started.html#Output
+    Convert a tree sequence from Relate (converted via relate_lib/bin/Convert, which
+    provides equivalenet via metadata, into a tree sequence in which the nodes
+    have been merged, and new times allocated that are consistent with the ARG
     
-    NUM_HAPLOTYPES 8
-NUM_TREES 7619
-0: 8:(607.26519 0.000 0 86) 13:(2702.54805 0.000 0 28) 9:(2641.61779 0.000 0 41) 11:(3724.67317 3.000 0 28) 9:(2641.61779 7.000 0 41) 12:(5155.21459 5.000 0 21) 10:(3088.44253 4.000 0 28) 8:(607.26519 4.000 0 86) 13:(2095.28286 2.000 0 28) 10:(446.82474 0.000 0 28) 11:(636.23064 0.000 0 28) 12:(1430.54142 0.000 0 21) 14:(1247.46463 0.000 0 21) 14:(3700.13116 2.000 0 21) -1:(0.00000 0.000 0 21) 
-21: 8:(667.04955 0.000 0 86) 11:(2266.08783 0.000 0 28) 9:(2335.63438 0.000 0 41) 12:(3455.60941 3.000 0 28) 9:(2335.63438 7.000 0 41) 14:(7848.96712 
-5.000 21 41) 10:(3236.77021 4.000 0 28) 8:(667.04955 4.000 0 86) 11:(1599.03828 2.000 0 28) 10:(901.13583 0.000 0 28) 12:(218.83920 0.000 0 28) 13:(14
-58.13811 0.000 21 28) 13:(268.61652 0.000 21 28) 14:(4124.74119 5.000 21 41) -1:(0.00000 0.000 21 28) 
-
+    the additional_equivalents parameter is a hack because relate_lib doesn't
+    calculate equivalent edges if the edges are above sample nodes. This is therefore
+    a supplementary dictionary of cases where *nodes* (not edges) in the input
+    tree sequence can be thought as equivalent.
     """
[…]

### 2022-03-24 illustrations.py

@@ -495,7 +495,7 @@ def inference():
     # ARGweaver
     ts = tskit.load("examples/Kreitman_SNP_argweaver.trees")
     labels = {n.id: n.metadata["name"] if n.is_sample() else "" for n in ts.nodes()}
-    tree_seqs["ARGweaver ex."] = argutils.viz.label_nodes(ts, labels=labels)
+    tree_seqs["ARGweaver example"] = argutils.viz.label_nodes(ts, labels=labels)
 
     # Tsinfer
     ts = tskit.load("examples/Kreitman_SNP_tsinfer.trees")
@@ -510,7 +510,7 @@ def inference():
                 labels[n.id] = json.loads(ind_metadata.decode())["name"]
     tree_seqs["Tsinfer"] = argutils.viz.label_nodes(ts, labels=labels)
 
-    # Relate JBOT
+    # Relate JBOT (normally commented out)
     ts = tskit.load("examples/Kreitman_SNP_relate_jbot.trees")
     tables = ts.dump_tables()
     tables.nodes.metadata_schema = tskit.MetadataSchema.permissive_json()
@@ -519,9 +519,20 @@ def inference():
     ts = tables.tree_sequence()
     # labels not stored by default in the Relate ts metadata, so use the previous ones
     labels = {n.id: labels[n.id] if n.is_sample() else "" for n in ts.nodes()}
-    tree_seqs["Relate JBOT"] = argutils.viz.label_nodes(ts, labels=labels)
+    # tree_seqs["Relate JBOT"] = argutils.viz.label_nodes(ts, labels=labels)
 
-    widths = np.ones(4)
+    # Relate non JBOT
+    ts = tskit.load("examples/Kreitman_SNP_relate_merged.trees")
+    tables = ts.dump_tables()
+    tables.nodes.metadata_schema = tskit.MetadataSchema.permissive_json()
+    for i, n in enumerate(tables.nodes):
+        tables.nodes[i] = n.replace(metadata={})
+    ts = tables.tree_sequence()
+    # labels not stored by default in the Relate ts metadata, so use the previous ones
+    labels = {n.id: labels[n.id] if n.is_sample() else "" for n in ts.nodes()}
+    tree_seqs["Relate"] = argutils.viz.label_nodes(ts, labels=labels)
+
+    widths = np.ones(len(tree_seqs))
     heights = [1, 0.1]
     fig, axes = plt.subplots(2, len(tree_seqs), figsize=(10, 6),
[…]

### 2022-03-24 make_example_outputs.py

@@ -126,16 +126,36 @@ def run_relate():
 
     # Convert to JBOT tree sequence format
     subprocess.run([
-        "tools/relate/bin/RelateFileFormats",
+        "tools/relate_lib/bin/Convert",
         "--mode", "ConvertToTreeSequence",
-        "-i", f"{dir}{outfiles}",
-        "-o", f"examples/Kreitman_SNP_relate_jbot"
+        "--anc", f"{dir}{outfiles}.anc",
+        "--mut", f"{dir}{outfiles}.mut",
+        "-o", f"examples/Kreitman_SNP_relate_jbot",
     ])
     
     # Convert to time-uncalibrated tree sequence format
-    #with open("examples/argweaver_output/arg-sample.0.arg") as f:
-    #ts = argutils.convert_relate_without_times(outfiles
-    
+    ts_jbot = tskit.load("examples/Kreitman_SNP_relate_jbot.trees")
+    ts = argutils.relate_ts_JBOT_to_ts(
+        ts_jbot,
+        # Hack here
+        additional_equivalents={
+            24: 14, 34: 14,
+            21: 11,
+            22: 12, 32: 12,
+            23: 13, 33: 13,
+            29: 18,
+            })
+    # See if we can get nicer times
+    try:
+        tables = ts.dump_tables()
+        for n in ts.nodes():
+            tables.nodes[n.id] = n.replace(
+                time=np.mean([v for v in n.metadata["Relate_times"].values()]))
+        tables.sort()
+        ts = tables.tree_sequence()
+    except:
+        print("Can't set plausible Relate time orders, reverting to arbitary")
+    ts.dump("examples/Kreitman_SNP_relate_merged.trees")
     
 cli.add_command(run_tsinfer)
 cli.add_command(run_kwarg)

### 2022-03-27 make_example_outputs.py
Use tsdate and hand-align Fig 6 nodes
All layout can then use the same Y axis (dot layout)

@@ -7,6 +7,7 @@ import os
 import subprocess
 
 import tsinfer
+import tsdate
 import tskit
 import argutils
 import numpy as np
@@ -41,6 +42,7 @@ def run_tsinfer():
         use = use[use != oldest_node]
         tables.subset(use)
     ts = tables.tree_sequence()
+    ts = tsdate.date(ts, mutation_rate = mu, Ne = Ne)
     ts.dump("examples/Kreitman_SNP_tsinfer.trees")
 
 

### 2022-03-28 paper.tex

@@
 \caption{\label{fig-inferred-args}
-ARGs inferred by different methods from a small dataset. Notes about these
-ARGs.
+ARGs inferred by different methods from a 2.7kb region of the \textit{Drosophila melanogaster}
+ADH locus for 11 samples~\citep{kreitman1983nucleotide}. Note that the position of points on the
+Y axis is that chosen for graphical clarity by the graphviz \texttt{dot} algorithm,  rather than being
+a direct measure of time. Additional notes about these ARGs.
 }

### 2022-04-05 illustrations.py
Allow nicer optional plotting of simplified inferences

@@ -482,6 +482,29 @@ def inference():
     """
     Examples of ARGs produced by various inference algorithms
     """
+    def get_edge_colors(ts):
+        """
+        Return a dict of edge colours for nodes and a dict of edge colors for edges
+        """
+        # Sort edges by the age of the child node
+        edges_argsorted = scipy.stats.rankdata(
+            ts.tables.nodes.time[ts.tables.edges.child], method='dense') - 1
+        edge_sorted_dict = {edge_id: sorted_edge_id
+                            for edge_id, sorted_edge_id in enumerate(edges_argsorted)}
+        # Create a dictionary w keys: (parent, child) of each edge; values: edge color
+        edge_colors_by_nodes = {}
+        edge_colors_by_id = {}
+        # Colormap normalization to the number of edges in each tree sequence
+        max_rank_by_ts[name] = np.max(edges_argsorted)
+        norm = mpl.colors.Normalize(vmin=0, vmax=max_rank_by_ts[name])
+        for key_edge, edge in enumerate(ts.edges()):
+            color = argutils.viz.make_color(
+                    edge_colormap(norm(edge_sorted_dict[key_edge])))
+            edge_colors_by_nodes[edge.child, edge.parent] = color
+            edge_colors_by_id[key_edge] = color
+        return edge_colors_by_nodes, edge_colors_by_id
+
+
     tree_seqs = {}
     # Assign colormap for edges
     edge_colormap = plt.cm.viridis
@@ -538,73 +561,59 @@ def inference():
                              gridspec_kw=dict(
                              width_ratios=widths, height_ratios=heights))
     tree_seq_positions = []
-    edge_colors_by_ts = {}
     max_rank_by_ts = {}
     for ax, ax_edges, (name, ts) in zip(axes[0], axes[1], tree_seqs.items()):
-        reverse_x_axis = False
-        tweak_x={}
-        use_ranked_times = None  # By default use Y axis layout from graphviz dot
+        params = dict(
+            # some of these can get overwritten
+            node_size=30,
+            rotated_sample_labels=True,
+            reverse_x_axis = False,
+            draw_edge_widths=True,
+            node_arity_colors=True,
+            max_edge_width=2,
+            tweak_x={},
[…]

### 2022-04-05 illustrations.py
Different example for fig 5

@@ -217,40 +217,24 @@ def simplification():
     """
     Sequentially simplifying a WF simulation.
     """
-    seed = 372
+    seed = 4517  # Chosen to give a diamond, 2 parents + 1 child node and a CA/noncoal
     t_x = {
-        4: 15,
-        12: -5,
-        9: 3,
-        6: -15,
-        18: 15,
-        17: 5,
-        13: 30,
-        14: -20,
-        10: -10,
-        15: 3,
-        16: -13,
-        7: -5,
-        11: 5,
-        8: 10,
+        2: -30, 3: 30, 4: 15, 7: 10, 8: 4, 9: 20, 10: 10,
+        11: 3, 12: 14, 13: 8, 15: 8, 16: 3.5, 17: 15, 18: 20,
     }
 
     ts = argutils.sim_wright_fisher(2, 10, 100, recomb_proba=0.1, seed=seed)
     tables = ts.dump_tables()
     rank_times = scipy.stats.rankdata(tables.nodes.time, method="dense")
     # tweak rank times here for nicer viz
-    rank_times = np.where(rank_times == 1, 0, rank_times)
+    rank_times = np.where(rank_times == 1, 0.5, rank_times)
     tables.nodes.time = rank_times
-    node_order = np.arange(tables.nodes.num_rows)
-    # NB this isn't quite right
-    node_order[[0, 1, 2, 3]] = [2, 1, 3, 0]
-    tables.subset(node_order)
     tables.sort()
     ts=tables.tree_sequence()
 
     labels = {i: string.ascii_uppercase[i] for i in range(len(string.ascii_uppercase))}
     # relabel the nodes to get samples reading A B C D
-    labels.update({0: "B", 2: "A", 3: "C", 1: "D", 4: "F", 5: "E", 13: "O", 14: "N", 7: "I", 8: "H", 10: "L", 11: "K"})
+    labels.update({6: "E", 4: "G", 10: "J", 9: "K", 12: "L", 11: "M", 15: "O", 14: "P"})
     #labels = {i: i for i in range(26)}
     ts1 = argutils.viz.label_nodes(ts, labels)
 

### 2022-04-05 paper.tex
Update caption to fig 5 to match

@@
 \caption{\label{fig-simplification}
-ARG simplification. (A) An ARG simulated from a diploid Wright-Fisher
-model. Note the complex full-sib relationship between B and C; this
-could not be directly represented in a Griffiths-like EARG encoding
+ARG simplification. In all cases the graph visualization is shown on the left
+and the equivalent marginal trees on the right.
+(A) An ARG simulated from a diploid Wright-Fisher
+model. Note that the genome \textsf{K} has three children and the genomes
+\textsf{E} and \textsf{G} simultaneously have multiple parents and multiple children;
+neither could not be directly represented in a Griffiths-like EARG encoding
 (see section XXX).
@@
 % trying out this terminology - can define in the text
-1-connected graph components (e.g., diamonds).
-(C) Remove nodes that are unary in all local trees.
-(D) Rewrite edges to skip any unary nodes in local trees.
-(E) The local trees for ARG (A).
-(F) The local trees for ARG (D).
+1-connected graph components (e.g., diamonds such as \textsf{JLMN}).
+(C) Remove nodes that are unary in all local trees. This removes both recombinant nodes
+that have only one child (e.g. \textsf{N}) and nodes that have more
+than one child (i.e. are ``common ancestors'') but never represent coalescences
+in the local trees (e.g. \textsf{R}).
+(D) Rewrite edges to bypass any unary nodes in local trees. Note that this loses lineage
+information such as the fact that \textsf{K} is connected to \textsf{A} via \textsf{E}; it
+can also create many extra roots.
 }

### 2022-04-06 paper.tex
Change paper description of fig 5
And highlight one of the trees

@@
 
-Given a GARG $E$ and the set of samples $S$ we can \emph{resolve} its
+Given a gARG $E$ and the set of samples $S$ we can \emph{resolve} its
 ancestry by noting that the resolved interval $Ir$ on an edge $(c,p,I)$
@@
 \item It would be good of if differentiate between the inheritance intervals
- from the GARG and the ancestry intervals of the resolved GARG.
+ from the gARG and the ancestry intervals of the resolved gARG.
 \end{itemize}
@@
 \emph{Simplification}~\citep{kelleher2018efficient} is the process
-of removing nodes and re-writing edges in a GARG to remove
+of removing nodes and re-writing edges in a gARG to remove
 various types of redundancy, while maintaining the crucial properties
@@
 and the equivalent marginal trees on the right.
-(A) An ARG simulated from a diploid Wright-Fisher
-model. Note that the genome \textsf{K} has three children and the genomes
+(A) A gARG simulated from a diploid Wright-Fisher
+model. Note that the genome \textsf{K} has three children, and the genomes
 \textsf{E} and \textsf{G} simultaneously have multiple parents and multiple children;
-neither could not be directly represented in a Griffiths-like EARG encoding
-(see section XXX).
+neither could be directly represented in a Griffiths-like eARG encoding
+(see section XXX). The penultimate local tree is highlighted to aid discussion (see section YYY).
 (B) Simplified to remove all
@@
 (D) Rewrite edges to bypass any unary nodes in local trees. Note that this loses lineage
-information such as the fact that \textsf{K} is connected to \textsf{A} via \textsf{E}; it
-can also create many extra roots.
+information such as the fact that \textsf{K} is connected to \textsf{A} via \textsf{E}.
 }
@@
 
-% Q: would it help to talk about ``tree vertice'' here or would it
+% Q: would it help to talk about ``tree vertices'' here or would it
 % just be more confusing? I worry that people would think that
@@
 % that they get that a node-is-a-node
+
 The ideas of ``meaningful'' local tree topologies revolve around the
 presence of ``unary'' nodes: a node in the tree that has exactly
-one child. For example, Fig.~\ref{fig-simplification}E shows the
-local trees along the genome for a simulated ARG
-(Fig.~\ref{fig-simplification}A). Consider the right-most tree,
-and the path above node $D$. Before $D$ meets its common ancestor
-over the interval $(93, 100]$ with the other samples it passes through
-nodes $H, K, N, R$ and $S$, reflecting the corresponding
-traversal through the ARG (see section XXX). Such unary nodes
-in a local tree are not usually considered significant,
+one child. Fig.~\ref{fig-simplification} show a series of successive steps of
+simplification, starting with a complete gARG simulated under a Wright-Fisher
+model, shown both as a graph on the left, and the equivalent set of 6
+local trees on the right (Fig.~\ref{fig-simplification}A). Consider the path above node
+\textsf{A} in the penultimate tree, highlighted in yellow and covering genomic positions
+60 to 70. On this path, before $A$ meets its common ancestor with the other samples at
+node \textsf{Q}, it passes through  \textsf{E},  \textsf{J},  \textsf{M},  \textsf{N} and 
+\textsf{P}, which in this region of the genome are all unary nodes. This reflects the
+passage of this span of genome through the ARG towards the root (see section XXX).
+Such unary nodes in a local tree are not usually considered important,
 because they provide very little information and are not
@@
 The nodes in a tree usually mark the points where samples find a
-\emph{common} ancestor, rather than focusing on specific
-ancestors. For most purposes, we would consider an edge
-directly joining sample $D$ to the node $T$ to be more
-useful and informative. Simplification is the process of removing
-these uninformative and (for most computational purposes)
-inefficient unary paths.
+ \emph{common} ancestor, such as node \textsf{Q} in the highlighted
+tree. For many purposes, we would therefore consider a path directly
+joining sample \textsf{A} to node \textsf{Q} to be of equivalent status to the path
+containing the intermediate unary nodes. Simplification is the process of removing
+these less informative and (for most computational purposes) inefficient nodes
+from the local trees and the corresponding annotated graph.
 
@@
 in which the two parent nodes of a recombination immediately
-join again into a common ancestor (e.g. $K,N,O$ and $R$ in
-Fig.~\ref{fig-simplification}A).
+join again into a common ancestor (e.g. \textsf{J}, \textsf{L}, \textsf{M}
+and  \textsf{N} in Fig.~\ref{fig-simplification}A).
 Unless we are specifically
 interested in the recombination event or ancestral genomes,
-there is no information in this topology and the diamond can b
-e replaced by a single edge. More generally, any
+there is no information in this topology and the diamond can be
+replaced by a single edge. More generally, any
 subgraph which is singly-connected in both the leafward and
@@
 replaced by a single edge. This definition includes the case
-of a single node that has one inbound and one outbound edge.
+of a single node that has one inbound and one outbound edge, such as
+nodes \textsf{F} and \textsf{H}.
 Fig.~\ref{fig-simplification}B shows the result of this type of
@@
 Simplifying away diamonds will remove many unary nodes from the
-local trees, there can still be nodes that are unary in all
-of the local trees. For example, node $H$ is one of the
-parents of the recombinant node $D$ is still present in
-Fig.~\ref{fig-simplification}B after diamond removal, but
-it is unary in all of the local trees since it has an
-in-degree of $1$.
-Fig.~\ref{fig-simplification}C shows the resulting graph
-after removing these nodes.
+local trees, but there can still be nodes that are unary in all
+of the local trees. In particular, a node can represent a recombinant
+with multiple parents in the graph but only a single child (e.g. node \textsf{N}
+in Fig.~\ref{fig-simplification}B), or can represent a common ancestor with
+multiple children in the graph but in which no local coalescence takes place
+(node \textsf{R} in Fig.~\ref{fig-simplification}B).
+% point out that this is why Hudson refers to CA nodes, not coalescence nodes
+Such nodes are not singly connected in the graph, but are nevertheless unary in
+all of the local trees in Fig.~\ref{fig-simplification}B. The operation to remove them,
+the results of which are shown in Fig.~\ref{fig-simplification}C,
+therefore requires knowledge not just of the graph topology, but also of the
+edge annotations. Removal of recombinant nodes can produce graph nodes with
+more than 2 parents (e.g. node \textsf{E}); likewise, removal of
+common ancestor but non-coalescent nodes can produce graph nodes with
+more than 2 children (e.g. node \textsf{S}). These represent a ``stacking up'' of multiple
+often indistinguishable events into a single node, something which is impossible
+to encode in an eARG.
 
@@
 correspond to a coalescence in at least one other
-local tree. For example, node $E$ is unary in the first tree
-of Fig.~\ref{fig-simplification}E, but is either binary
-or ternary in the other trees (recall this is a Wright-Fisher
-simulation).
-
-The final level of simplification is to remove \emph{all}
-unary nodes from the local trees (Fig.~\ref{fig-simplification}D
-and E). [And continue]
+local tree. For example, node  \textsf{K} is unary in the second tree
+of Fig.~\ref{fig-simplification}C, but is either binary
+or ternary in all subsequent trees (recall this is a Wright-Fisher
+simulation). The final level of simplification is to alter the edge annotations
+such that, although no nodes are removed from the graph, \emph{all}
+unary nodes disappear from the local trees (Fig.~\ref{fig-simplification}D).
+It is common for this operation to remove nodes above the top MRCA node
+in the tree, resulting in many more root nodes (here, \textsf{Q} and \textsf{K}
+become roots). [And continue]
+% Note that although this last stage produces much simpler local trees, by
+% removing information about the exact paths taken by lineages through
+% the graph, it loses potentially important information about shared edges
+% between trees. It is not clear whether (c) or (d) leads to a more compact
+% representation of the ARG in terms of number of edges.
 

### 2022-04-06 illustrations.py

@@ -295,15 +295,20 @@ def simplification():
             node_labels={n.id: n.metadata["name"] for n in ts.nodes()},
             x_label=None if i == 3 else "",
             style=(
-                '.x-axis .tick .lab {font-weight: regular; font-size: 12; visibility: hidden} '
-                '.x-axis .tick:first-child .lab, .x-axis .tick:last-child .lab {visibility: visible}'
+                '.x-axis .tick .lab {font-weight: regular; font-size: 12; visibility: hidden} ' +
+                '.x-axis .tick:first-child .lab, .x-axis .tick:last-child .lab {visibility: visible}' +
+                (f'.subfig{i} .background path:nth-child(5) {{fill: #FFFF30; fill-opacity: .5}}' if i==0 else '')
             ),
         )
         svg.append(
             f'<text font-size="2em" font-family="serif" transform="translate(0, {200 * i + 30})">' +
             f'({string.ascii_lowercase[i]})</text>'
         )
-        svg.append(f'<g transform="translate(250 {205 * i}) scale(0.83)">' + tree_svg + "</g>")
+        svg.append(
+            f'<g class="subfig{i}" transform="translate(250 {205 * i}) scale(0.83)">' +
+            tree_svg +
+            "</g>"
+        )
     svg.append("</svg>")
 
 

### 2022-04-06 paper.tex
Experiment with using named labels for section refs

@@
 \usetikzlibrary{calc,positioning}
+\usepackage{hyperref}
 
@@
 \linenumbers
-\title{What is an Ancestral Recombination Graph?}
+\title{A general and efficient representation of Ancestral Recombination Graphs}
 % First authors
@@
 progress. We currently lack a well-defined data-model and shared terminology
-to discuss recombining genealogies,
-leading to basic errors in statements about what different inference methods
-produce as output (see the XXX section below), as well as presenting serious
+to discuss recombining genealogies, leading to basic errors in statements
+about what different inference methods produce as output (see the
+``\nameref{ARG_inference_methods}'' section), as well as presenting serious
 difficulties in either comparing the outputs of different methods or
@@
 Fortunately, it is straightforward to generalise the classical Griffiths
-EARG encoding to both simplify implementations and to encompass any
+eARG encoding to both simplify implementations and to encompass any
 possible pattern of genetic inheritance.
@@
 we call this encoding of an ARG data
-structure a ``Genome ARG'', or GARG.
-Fig~\ref{fig-arg-in-pedigree} illustrates the basic elements of a GARG.
+structure a ``Genome ARG'', or gARG.
+Fig~\ref{fig-arg-in-pedigree} illustrates the basic elements of a gARG.
 Fig.~\ref{fig-arg-in-pedigree}a shows the genomes
@@
 
-Formally, we can define a GARG as a set of edges $E$, where each
+Formally, we can define a gARG as a set of edges $E$, where each
 element of the set is a tuple $(c, p, I)$ such that $c, p \in \mathbb{N}$,
@@
 the next section). Deriving the local tree at a point $x$
-follows largely the same pattern as for an EARG but is somewhat
+follows largely the same pattern as for an eARG but is somewhat
 simpler because coordinate information is directly associated with
@@
 from the graph so that \textsf{D} descends directly from \textsf{K} (see
-section XXX) but there is no necessity for this from a representational
-perspective. This ability can be useful in practice and could be required if,
+the ``\nameref{ARG_simplification}'' section) but there is no necessity for this from
+a representational perspective. This ability can be useful in practice and could be
 % Do we need to spell this out?
-for example, we have a deeply sequenced pedigree of
-haplodiploids. % e.g. Honeybees? Be great to cite an example!
+required if, for example, we have a deeply sequenced pedigree of haplodiploids.
+% e.g. Honeybees? Be great to cite an example!
+% Note by Yan - if we actually have the sequence, then this turns into
+% a sample node, so would be normally included anyway. Perhaps, therefore
+% we should talk about these "pass though" nodes in the context
+% of internal samples.
 
@@
 Moreover, a single node can have more than 2 children, and (as we shall see in
-section XXX), more than 2 parents. Another way to state this is that transitions
-between nodes in a gARG can encapsulate multiple events in an eARG.
+the ``\nameref{ARG_simplification}'' section, more than 2 parents. Another way to
+state this is that transitions between nodes in a gARG can encapsulate multiple
+events in an eARG.
 
@@
 
-\section*{Ancestry resolution}
+\section*{Ancestry resolution}\label{Ancestry_resolution}
 % This is a rough first pass. Will clarify terminology later.
@@
 
-\section*{ARG simplification}
+\section*{ARG simplification}\label{ARG_simplification}
 
@@
 \textsf{P}, which in this region of the genome are all unary nodes. This reflects the
-passage of this span of genome through the ARG towards the root (see section XXX).
+passage of this span of genome through the ARG towards the root (see the
+``\nameref{Ancestry_resolution}'' section).
 Such unary nodes in a local tree are not usually considered important,
@@
 
-\section*{ARG inference methods}
+\section*{ARG inference methods}\label{ARG_inference_methods}
 Using the terminology developed here to classify ARGs, we now review methods developed

### 2022-04-06 paper.tex

@@
 provides the theoretical underpinning for much of contemporary population genetics.
-Hudson extended the coalescent to include
-recombination~\citep{hudson1983properties,hudson1990gene}
+\citet{hudson1983properties,hudson1990gene} extended the coalescent to include
+recombination
 and the model has proved to be an especially powerful basis for
 simulations~\citep{hudson2002generating,baumdicker2021efficient}.
-Hudson's algorithm is concerned with simulating the process of
-the coalescent with recombination as efficiently as possible, and operates by
-tracking segments of ancestral material carried by ancestors as
-we go backwards in time. The details of Hudson's algorithm and how the
-results of the ancestral process are represented are critical to later
-discussions, and so we will explain the process in some detail.
-See~\cite{kelleher2016efficient} for a more detailed
-description of the algorithm, and how it may be implemented efficiently.
-
-\begin{figure}[ht]
-\centering
-% FIXME this is a quick nasty hack to make the figure a bit smaller and
-% prevent flushing all the other figures to the end of the document.
-\scalebox{0.5}{
-\begin{tabular}{cc}
-\begin{tikzpicture}
-	\draw (-0.4, -0.2) -- (0.4, -0.2) -- (0.4, -0.3) -- (-0.4, -0.3) -- (-0.4, -0.2);
-	\draw (-0.1, -0.1) -- (-0.1, -0.4);
-	\node [label=below:{\small $v$}] at (-0.1, -0.2) {};
-	\draw (3.6, -0.2) -- (4.4, -0.2) -- (4.4, -0.3) -- (3.6, -0.3) -- (3.6, -0.2);
-	\draw (5.6, -0.2) -- (6.4, -0.2) -- (6.4, -0.3) -- (5.6, -0.3) -- (5.6, -0.2);
-
-	\draw (4,0) -- (4, 1.2) -- (6, 1.2) -- (6,0);
-	\draw (4.6, 1) -- (5.4, 1) -- (5.4, 0.9) -- (4.6, 0.9) -- (4.6, 1);
-	\draw (5.1, 1.1) -- (5.1, 0.8);
-	\node [label=below:{\small $w$}] at (5.1, 1) {};
-
-	\draw (0, 0) -- (0, 1.7) -- (-1,1.7) -- (1,1.7);
-	\draw (-1.4, 1.5) -- (-1.1, 1.5) -- (-1.1, 1.4) -- (-1.4, 1.4) -- (-1.4, 1.5);
-	\draw (-1.1, 1.45) -- (-0.6, 1.45);
-	\draw (0.6, 1.45) -- (0.9, 1.45);
-	\draw (0.9, 1.5) -- (1.4, 1.5) -- (1.4, 1.4) -- (0.9, 1.4) -- (0.9, 1.5);
-	\draw (1.2, 1.6) -- (1.2, 1.3);
-	\node [label=below:{\small $x$}] at (1.2, 1.5) {};
-
-	\draw (5, 1.2) -- (5, 2.4) -- (4,2.4) -- (6,2.4);
-	\draw (3.6, 2.2) -- (4.1, 2.2) -- (4.1, 2.1) -- (3.6, 2.1) -- (3.6, 2.2);
-	\draw (4.1, 2.15) -- (4.4, 2.15);
-	\draw (5.6, 2.15) -- (6.1, 2.15);
-	\draw (6.1, 2.2) -- (6.4, 2.2) -- (6.4, 2.1) -- (6.1, 2.1) -- (6.1, 2.2);
-
-	\draw (1, 1.7) -- (1, 3.1) -- (0,3.1) -- (2,3.1);
-	\draw (-0.4, 2.85) -- (0.2, 2.85);
-	\draw (0.2, 2.9) -- (0.4, 2.9) -- (0.4, 2.8) -- (0.2, 2.8) -- (0.2, 2.9);
-	\draw (1.6, 2.85) -- (1.9, 2.85);
-	\draw (1.9, 2.9) -- (2.2, 2.9) -- (2.2, 2.8) -- (1.9, 2.8) -- (1.9, 2.9);
-	\draw (2.2, 2.85) -- (2.4, 2.85);
-
-	\draw (-1, 1.7) -- (-1, 4.5) -- (0,4.5) -- (0,3.1);
-	\node [scale=0.2,label=above left:{\small A}] at (-0.5,4.5) {A};
-	\draw (-0.9, 4.3) -- (-0.6, 4.3) -- (-0.6, 4.2) -- (-0.9, 4.2) -- (-0.9, 4.3);
-	\draw (-0.6, 4.25) -- (-0.3, 4.25);
-	\draw (-0.3, 4.3) -- (-0.1, 4.3) -- (-0.1, 4.2) -- (-0.3, 4.2) -- (-0.3, 4.3);
-	\draw (-0.5, 4.4) -- (-0.5, 4.1);
-	\node [label=below:{\small $y$}] at (-0.5, 4.3) {};
-
-	\draw (4,2.4) -- (4,5.4) -- (2,5.4) -- (2,3.1);
-	\node [scale=0.2,label=above left:{\small B}] at (3,5.4) {B};
-	\draw (2.6, 5.2) -- (2.9, 5.2) -- (2.9, 5.1) -- (2.6, 5.1) -- (2.6, 5.2);
-	\draw (2.9, 5.15) -- (3.4, 5.15);
-
-	\draw (-0.5, 4.5) -- (-0.5, 6) -- (-1.5,6) -- (0.5,6);
-	\draw (-1.9, 5.75) -- (-1.3, 5.75);
-	\draw (-1.3, 5.8) -- (-1.1, 5.8) -- (-1.1, 5.7) -- (-1.3, 5.7) -- (-1.3, 5.8);
-	\draw (0.1, 5.8) -- (0.4, 5.8) -- (0.4, 5.7) -- (0.1, 5.7) -- (0.1, 5.8);
-	\draw (0.4, 5.75) -- (0.9, 5.75);
-
-	\draw (0.5, 6) -- (0.5, 7.2) -- (3,7.2) -- (3,5.4);
-	\node [scale=0.2,label=above:{\small C}] at (1.75,7.2) {C};
-	\draw (1.35, 6.95) -- (2.15, 6.95);
-
-	\draw (-1.5, 6) -- (-1.5, 8);
-	\draw (6, 2.4) -- (6, 8);
-
-	% Edge annotations above each event
-	\node [label=left:{\small$\{(0,m,1)\}$}] at (0.2,0.6) {};
-	\node [label=left:{\small$\{(0,m,1)\}$}] at (4.2,0.4) {};
-	\node [label=right:{\small$\{(0,m,1)\}$}] at (5.8,0.4) {};
-	\node [label=left:{\small$\{(0,m,2)\}$}] at (5.2,1.6) {};
-	\node [label=left:{\small$\{(0,v,1)\}$}] at (-0.8,2.3) {};
-	\node [label=right:{\small$\{(v,m,1)\}$}] at (0.8,2.1) {};
-	\node [label=right:{\small$\{(0,w,2)\}$}] at (3.8,3.9) {};
-	\node [label=left:{\small$\{(w,m,2)\}$}] at (6.2,5) {};
-	\node [label=right:{\small$\{(x,m,1)\}$}] at (-0.2,3.8) {};
-	\node [label=right:{\small$\{(v,x,1)\}$}] at (1.8,4.1) {};
-	\node [label=left:{\small$\{(0,v,1), (x,m,1)\}$}] at (-0.3,5.2) {};
-	\node [label=right:{\small$\{(0,v,2)\}$}] at (2.8,6.4) {};
-	\node [label=left:{\small$\{(x,m,1)\}$}] at (-1.3,7) {};
-	\node [label=right:{\small$\{(0,v,1)\}$}] at (0.3,6.4) {};
-
-	% Dashed lines for start and end times
-	\draw[dashed] (-2, 0) -- (11.1, 0);
-	\node [label=left:{$t = 0$}] at (-2,0) {};
-	\draw[dashed] (-2, 8) -- (11.1, 8);
-	\node [label=left:{$t = T$}] at (-2,8) {};
-
-	% Numbers of extant ancestors and links, from top to bottom
-	\node[label=right:{CA \; RE}] at (7.5, 8.2) {};
-	\node[label=right:{$\binom{2}{2}$ \; $\frac{2 m - x - w - 2}{m - 1} \rho$}] at (7.5, 7.6) {};
-	\node[label=right:{$\binom{4}{2}$ \; $\frac{2 m + 2 v - x - w - 3}{m - 1} \rho$}] at (7.5, 6.6) {};
-	\node[label=right:{$\binom{3}{2}$ \; $\frac{2 m + v - w - 3}{m - 1} \rho$}] at (7.5, 5.7) {};
-	\node[label=right:{$\binom{4}{2}$ \; $\frac{2 m + x - v - 4}{m - 1} \rho$}] at (7.5, 4.95) {};
-	\node[label=right:{$\binom{5}{2}$ \; $\frac{2 m - 5}{m - 1} \rho$}] at (7.5, 3.8) {};
-	\node[label=right:{$\binom{4}{2}$ \; $\frac{2 m - 4}{m -1} \rho$}] at (7.5, 2.75) {};
-	\node[label=right:{$\binom{3}{2}$ \; $\frac{2 m - 3}{m - 1} \rho$}] at (7.5, 2.05) {};
-	\node[label=right:{$\binom{2}{2}$ \; $2 \rho$}] at (7.5, 1.45) {};
-	\node[label=right:{$\binom{3}{2}$ \; $3 \rho$}] at (7.5, 0.6) {};
-
-	% Gray dashed lines to visually separate holding times
-	\draw[color=gray, dashed] (7.5, 1.2) -- (11.1, 1.2);
-	\draw[color=gray, dashed] (7.5, 1.7) -- (11.1, 1.7);
-	\draw[color=gray, dashed] (7.5, 2.4) -- (11.1, 2.4);
-	\draw[color=gray, dashed] (7.5, 3.1) -- (11.1, 3.1);
-	\draw[color=gray, dashed] (7.5, 4.5) -- (11.1, 4.5);
-	\draw[color=gray, dashed] (7.5, 5.4) -- (11.1, 5.4);
-	\draw[color=gray, dashed] (7.5, 6) -- (11.1, 6);
-	\draw[color=gray, dashed] (7.5, 7.2) -- (11.1, 7.2);
-\end{tikzpicture}
-&
-\begin{tikzpicture}
-	\draw (-0.4, -0.2) -- (0.4, -0.2) -- (0.4, -0.3) -- (-0.4, -0.3) -- (-0.4, -0.2);
-	\draw (-0.1, -0.1) -- (-0.1, -0.4);
-	\node [label=below:{\small $v$}] at (-0.1, -0.2) {};
-	\draw (3.6, -0.2) -- (4.4, -0.2) -- (4.4, -0.3) -- (3.6, -0.3) -- (3.6, -0.2);
-	\draw (5.6, -0.2) -- (6.4, -0.2) -- (6.4, -0.3) -- (5.6, -0.3) -- (5.6, -0.2);
-
-	\draw (4,0) -- (4, 1.2) -- (6, 1.2) -- (6,0);
-	\draw (4.6, 1) -- (5.4, 1) -- (5.4, 0.9) -- (4.6, 0.9) -- (4.6, 1);
-	\draw (5.1, 1.1) -- (5.1, 0.8);
-	\node [label=below:{\small $w$}] at (5.1, 1) {};
-
-	\draw (0, 0) -- (0, 1.7) -- (-1,1.7) -- (1,1.7);
-	\draw (-1.4, 1.5) -- (-1.1, 1.5) -- (-1.1, 1.4) -- (-1.4, 1.4) -- (-1.4, 1.5);
-	\draw (-1.1, 1.45) -- (-0.6, 1.45);
-	\draw (0.6, 1.45) -- (0.9, 1.45);
-	\draw (0.9, 1.5) -- (1.4, 1.5) -- (1.4, 1.4) -- (0.9, 1.4) -- (0.9, 1.5);
-	\draw (1.2, 1.6) -- (1.2, 1.3);
-	\node [label=below:{\small $x$}] at (1.2, 1.5) {};
-
-	\draw (5, 1.2) -- (5, 2.4) -- (4,2.4) -- (6,2.4);
-	\draw (3.6, 2.2) -- (4.1, 2.2) -- (4.1, 2.1) -- (3.6, 2.1) -- (3.6, 2.2);
-	\draw (4.1, 2.15) -- (4.4, 2.15);
-	\draw (5.6, 2.15) -- (6.1, 2.15);
-	\draw (6.1, 2.2) -- (6.4, 2.2) -- (6.4, 2.1) -- (6.1, 2.1) -- (6.1, 2.2);
-	\draw [color=red](5.8, 2.3) -- (5.8, 2.0);
-	\node [label={[red]below:{\small $z$}}] at (5.8, 2.1) {};
-
-	\draw (1, 1.7) -- (1, 3.1) -- (0,3.1) -- (2,3.1);
-	\draw (-0.4, 2.85) -- (0.2, 2.85);
-	\draw (0.2, 2.9) -- (0.4, 2.9) -- (0.4, 2.8) -- (0.2, 2.8) -- (0.2, 2.9);
-	\draw (1.6, 2.85) -- (1.9, 2.85);
-	\draw (1.9, 2.9) -- (2.2, 2.9) -- (2.2, 2.8) -- (1.9, 2.8) -- (1.9, 2.9);
-	\draw (2.2, 2.85) -- (2.4, 2.85);
-
-	\draw (-1, 1.7) -- (-1, 4.5) -- (0,4.5) -- (0,3.1);
-	\draw (-0.9, 4.3) -- (-0.6, 4.3) -- (-0.6, 4.2) -- (-0.9, 4.2) -- (-0.9, 4.3);
-	\draw (-0.6, 4.25) -- (-0.3, 4.25);
-	\draw (-0.3, 4.3) -- (-0.1, 4.3) -- (-0.1, 4.2) -- (-0.3, 4.2) -- (-0.3, 4.3);
-	\draw (-0.5, 4.4) -- (-0.5, 4.1);
-	\node [label=below:{\small $y$}] at (-0.5, 4.3) {};
-
-	\draw (6, 2.4) -- (6, 3.8) -- (7, 3.8);
-	\draw [color=red](6, 3.8) -- (5, 3.8);
-	\draw (4.6, 3.55) -- (5.4, 3.55);
-	\draw (6.6, 3.55) -- (7.1, 3.55);
-	\draw (7.1, 3.6) -- (7.4, 3.6) -- (7.4, 3.5) -- (7.1, 3.5) -- (7.1, 3.6);
-
-	\draw (4,2.4) -- (4,5.4) -- (2,5.4) -- (2,3.1);
-	\draw (2.6, 5.2) -- (2.9, 5.2) -- (2.9, 5.1) -- (2.6, 5.1) -- (2.6, 5.2);
-	\draw (2.9, 5.15) -- (3.4, 5.15);
-
-	\draw (-0.5, 4.5) -- (-0.5, 6) -- (-1.5,6) -- (0.5,6);
-	\draw (-1.9, 5.75) -- (-1.3, 5.75);
-	\draw (-1.3, 5.8) -- (-1.1, 5.8) -- (-1.1, 5.7) -- (-1.3, 5.7) -- (-1.3, 5.8);
-	\draw (0.1, 5.8) -- (0.4, 5.8) -- (0.4, 5.7) -- (0.1, 5.7) -- (0.1, 5.8);
-	\draw (0.4, 5.75) -- (0.9, 5.75);
-
-	\draw [color=red](5, 3.8) -- (5, 6.6) -- (4, 6.6);
-	\draw (4,6.6) -- (3,6.6) -- (3, 5.4);
-	\draw (3.6, 6.4) -- (3.9, 6.4) -- (3.9, 6.3) -- (3.6, 6.3) -- (3.6, 6.4);
-	\draw (3.9, 6.35) -- (4.4, 6.35);
-
-	\draw (0.5, 6) -- (0.5, 7.2) -- (4,7.2) -- (4,6.6);
-	\draw (1.85, 6.95) -- (2.65, 6.95);
-
-	\draw (-1.5, 6) -- (-1.5, 8);
-	\draw [color=red](2.25, 7.2) -- (2.25, 8);
-	\draw (7, 3.8) -- (7, 8);
-
-	% Dashed lines for start and end times
-	\draw[dashed] (-2, 0) -- (11, 0);
-	\node [label=left:{$t = 0$}] at (-2,0) {};
-	\draw[dashed] (-2, 8) -- (11, 8);
-	\node [label=left:{$t = T$}] at (-2,8) {};
-
-	% Numbers of extant ancestors and links, from top to bottom
-	\node[label=right:{CA \; RE}] at (7.5, 8.2) {};
-	\node[label=right:{$\binom{3}{2}$ \; $3 \rho$}] at (7.5, 7.6) {};
-	\node[label=right:{$\binom{4}{2}$ \; $4 \rho$}] at (7.5, 6.9) {};
-	\node[label=right:{$\binom{5}{2}$ \; $5 \rho$}] at (7.5, 6.3) {};
-	\node[label=right:{$\binom{4}{2}$ \; $4 \rho$}] at (7.5, 5.7) {};
-	\node[label=right:{$\binom{5}{2}$ \; $5 \rho$}] at (7.5, 4.95) {};
-	\node[label=right:{$\binom{6}{2}$ \; $6 \rho$}] at (7.5, 4.15) {};
-	\node[label=right:{$\binom{5}{2}$ \; $5 \rho$}] at (7.5, 3.45) {};
-	\node[label=right:{$\binom{4}{2}$ \; $4 \rho$}] at (7.5, 2.75) {};
-	\node[label=right:{$\binom{3}{2}$ \; $3 \rho$}] at (7.5, 2.05) {};
-	\node[label=right:{$\binom{2}{2}$ \; $2 \rho$}] at (7.5, 1.45) {};
-	\node[label=right:{$\binom{3}{2}$ \; $3 \rho$}] at (7.5, 0.6) {};
-
-	% Gray dashed lines to visually separate holding times
-	\draw[color=gray, dashed] (7.5, 1.2) -- (11, 1.2);
-	\draw[color=gray, dashed] (7.5, 1.7) -- (11, 1.7);
-	\draw[color=gray, dashed] (7.5, 2.4) -- (11, 2.4);
-	\draw[color=gray, dashed] (7.5, 3.1) -- (11, 3.1);
-	\draw[color=gray, dashed] (7.5, 3.8) -- (11, 3.8);
-	\draw[color=gray, dashed] (7.5, 4.5) -- (11, 4.5);
-	\draw[color=gray, dashed] (7.5, 5.4) -- (11, 5.4);
-	\draw[color=gray, dashed] (7.5, 6) -- (11, 6);
-	\draw[color=gray, dashed] (7.5, 6.6) -- (11, 6.6);
-	\draw[color=gray, dashed] (7.5, 7.2) -- (11, 7.2);
-\end{tikzpicture}
-\end{tabular}
-}
-\caption{(A)
-A realisation of the graph traversed by Hudson's algorithm started from a
-sample of three chromosomes of length $m$ at time $t = 0$, and
-propagated until time $T$. The MRCA on the genetic interval $[v, w)$ is reached
-at event B, while that on $[0, v)$ is reached at event C.
-The non-ancestral segment $[v, w)$ above
-A contributes to the rate of effective recombinations because it
-is trapped between ancestral segments. The two columns titled CA and RE
-are the respective rates of mergers and recombinations when
-the recombination rate is $\rho$.
-(B) A corresponding realisation of a big ARG, which augments Hudson's algorithm
-by tracking nonancestral lineages. The result is a simpler state space and
-dynamics, at the cost of extra nodes and edges, highlighted in red, which do
-not affect the local tree at any site.}
-\label{hudson_vs_bigARG}
-\end{figure}
-
-Hudson's algorithm operates by tracking the state of a set of ancestral
-lineages as we go backwards in time.
-% jk-note unclear here about the best choice of notation. We could have
-% lists of tuples (l, r, a), or a tuple of vectors (l_j, r_j, a_j),
-% of a tuple (I_j, a_j), where I is a vector of intervals. Going to
-% stick with the original tuple notation for now.
-%
-% Each lineage consists of a set of
-% equal-length vectors $\mathbf{l}$, $\mathbf{r}$ and $\mathbf{a}$
-% such that for each $1 \leq j \leq \left| l \right|$,
-% $[\mathbf{l}_j, \mathbf{r}_j)$ is a half-closed genomic interval and
-% $\mathbf{a}_j$ is an integer
-% tracking the number of samples the lineage is ancestral to over the interval.
-% Assume that the intervals are sorted from left-to-right so that $\mathbf{l}_1$
-% and $\mathbf{r}_{|\mathbf{r}|}$ are the left- and right-most positions
-% covered by the lineage, respectively.
-Each lineage consists of a list of
-disjoint ancestry segments $(\ell, r, a)$, where
-$[\ell, r)$ is a half-closed genomic interval and $a$ is an integer
-tracking the number of samples to which the lineage is ancestral over that interval.
-(We also usually track the tree node associated with each segment, but
-that is not important for our purposes here so we omit it.)
-If we have $n$ samples and a genome of length $m$, the process begins with $n$ lineages
-of the form $\{(0, m, 1)\}$. The process then works backwards in time from
-the present day as a series of random common ancestor or recombination events.
-Recombination events occur at a rate determined by the amount of ancestral material and
-the way in which it is distributed along a chromosome.
-Let $L$ be the set of ancestors at a given time $t$. Recombination events
-happen at rate $\rho \nu / (m - 1)$ where
-\[
-\nu = \sum_{x \in L}\left( \max_{(\ell, r, a) \in x}r
-    - \min_{(\ell, r, a) \in x}\ell - 1 \right)
-\]
-is the number of available `links' that may be broken. Thus, the rate of
-recombination is determined by the left- and right-most extent of the
-ancestral material carried by each lineage. At a recombination
-event we choose one of these links uniformly and break it. Given a lineage
-$x = [(\ell_j, r_j, a_j)]$ and a breakpoint $k$, we have two lineages
-$x_1$ and $x_2$ such that FILL IN DETAILS
-
-When $k = |L|$ lineages are present, common ancestor events
-occur at rate $\binom{k}{2}$. In a common ancestor event, two lineages
-are chosen uniformly at random and their ancestry segments merged.
-If we have overlapping intervals of ancestry from the two lineages,
-say, $(\ell, r, a_1)$ and $(\ell, r, a_2)$, a
-\emph{coalescence} occurs and ancestor represented by the current event
-will be present as a node (at least) in the marginal trees covering
-the interval $[\ell, r)$. The result of this coalescence is a segment
-$(\ell, r, a_1 + a_2)$, and if $a_1 + a_2 < n$ it is included in the
-ancestry for the new lineage. Otherwise, if $a_1 + a_2 = n$ we know that
-we have found the most recent common ancestor of all samples in
-the interval $[\ell, r)$ and so we do not need to simulate its history any further.
-Nonoverlapping intervals of ancestry from the two lineages are included
-in the resulting lineage without changes. Eventually, as the process continues,
-we find resultant lineages in which all segments have fully coalescenced,
-and so the number of extant lineages gradually dwindles down to zero.
-
-% NOTE: it would probably be helpful to have a drawing of the ARG process
-% in-progress here to help us talk about it. What is the full state
-% that is remembered as we go through the process? It does help to think
-% about the process as choosing edges to merge, not nodes.
+\cite{kelleher2016efficient} provide a description of his algorithm along with
+an efficient implementation, the details of which turn out to be highly relevant in 
+later sections of the current paper (see the ``\nameref{Ancestry_resolution}'' section).
+
 The Ancestral Recombination Graph (ARG) was later introduced by
@@
 the graph structure derived from this process. For clarity, we will refer to the
-process as the Griffiths process, or the coalescent with recombination (CwR),
-and following XX-Who?-XX, we refer to the the structure as the ``big ARG'', as
-opposed to the ``little ARG'' traversed by Hudson's algorithm
-(\citet{wiuf1999recombination} refer to these structures as ARG and HUD respectively).
-
+process as the Griffiths process, or the coalescent with recombination (CwR)
+\citep{minichiello2006mapping}.
+
+% NOTE: it might be helpful to have a drawing of the ARG process
+% in-progress here to help us talk about it. What is the full state
+% that is remembered as we go through the process? It does help to think
+% about the process as choosing edges to merge, not nodes.
+% We could also refer to Fig.~\ref{fig-arg-data-structure} if we want a simple picture
+% to go with this
 In the Griffiths formulation, each edge in the graph corresponds to an extant
 lineage and nodes are events in the process (the initial $n$ leaf nodes are
-``sampling'' events). As in Hudson's
-algorithm, common ancestor events occur at rate $\binom{k}{2}$ when there
+``sampling'' events). Common ancestor events occur at rate $\binom{k}{2}$ when there
 are $k$ lineages present. We choose two lineages (edges) uniformly, and merge them
-into a common ancestor lineage. Recombination events are different
-to Hudson's algorithm, however. Rather than occuring at a rate which depends on
-the specific distribution of ancestral material among lineages, the
-rate of recombination is now simply $k \rho$. At a recombination event
-we choose a lineage (edge) uniformly, and a
+into a common ancestor lineage. Recombination events happen at a rate of
+$k \rho$. Here, we choose a lineage (edge) uniformly, and a
 breakpoint $0 < x < m$ uniformly on its genome. We terminate the edge at a
@@
 recombination nodes provides sufficient information to later recover the marginal
-trees (see the next section for more details and discussion on this point).
-
-The state-space of the Griffiths process is much simpler than Hudson's algorithm,
-which greatly facilitates mathematical reasoning. This simplicity comes at a
-substantial cost, however, if we wish to use it as a practical means of
-simulating recombinant ancestry. The big ARG is vast, and any realisation
-for even moderate levels of recombination is far too large to be of practical
-use. The number of events in the big ARG, all the way back to the GMRCA
-is $O(e^\rho)$~\citep{griffiths1997ancestral}, whereas the number
-of events required to simulate the little ARG is
-$O(\rho^2)$~\citep{hein2004gene,baumdicker2021efficient}.
-This disparity in the number of events in the two formulations is
-because the majority of the events that occur in the big ARG do
-not affect the genetic ancestry of the sample in any way. Recombination
-events that occur outside of ancestral material do not have any bearing
-on the ancestry of the sample, and so the structure is hugely redundant.
-As~\cite{wiuf1999recombination} note,
-``an `ancestral' sequence in the birth and death process
-need not have any genetic material in common with a
-sequence descended from it.''
-
-% Full quote:
-% This process simplifies mathematics on the account that the notion of an
-% ancestor will have a less restrictive meaning than usual:
-% An ``ancestral'' sequence in the birth and death process
-% need not have any genetic material in common with a
-% sequence descended from it.
+trees and calculate sampling probabilities (see the ``\nameref{probabilities}'' section).
 
@@
 reproduction~\citep{gonzalezcasanova2018duality, koskela2019robust}.
+
+\textcolor{red}{TODO: reduce the text below down to one paragraph, which
+defines the relationship between what we define as an ARG and
+what people mean by phylogenetic networks from the literature.
+We probably don't need to get into quite so much detail on
+the different types that people have studies. Essentially,
+we're ending the section with a brief review which situates
+our definition in the literature. I think pulling apart the
+idea of a ``reticulation'' vs what we call a recombinant
+is essential here.}
+
+The term \emph{phylogenetic network} has been broadly accepted to describe the most general
+type of genealogical graph, defined by \citet{huson2010phylogenetic} as ``any graph used to
+represent evolutionary relationships (either abstractly or explicitly) between a set of taxa
+that labels some of its nodes'' (although more narrow and context-dependent definitions can also
+be found in the literature). Explicit rooted phylogenetic networks are rooted DAGs,
+with leaves labelled by the taxa (species, groups, individuals or sequences), and nodes classed
+as either reticulation nodes (with in-degree $\geq 2$) or tree nodes.
+
+Some sub-classes of explicit rooted phylogenetic networks are defined based on the evolutionary event that the
+reticulation nodes capture \citep{huson2010phylogenetic}. In \emph{reassortment} networks, used to
+model the evolution of viruses, reticulation nodes correspond to exchange of
+(non-recombining) genomic segments between viral particles inside a co-infected host.
+In \emph{hybridisation} networks, reticulation nodes correspond to genetic material from two
+different species combining to create a hybrid. In \emph{recombination} networks, the reticulation nodes
+correspond to recombination events (with leaves labelled by sequences and edges labelled by mutations).
+The ARG is a subtype of a recombination network under this definition (restricting to one crossover
+recombination per reticulation node).
+% AI: Not sure about this - do ARGs also encompass multiple crossover recombinations and gene
+% conversion (seems implied above)?
+
+Sub-classes of explicit rooted phylogenetic networks can also be defined by restrictions on their topology.
+In \emph{binary networks}, all internal nodes must have either in-degree two and out-degree one or in-degree
+one and out-degree two (with the root having out-degree two and the leaves in-degree 1) \citep{steel2016phylogeny}.
+In \emph{tree-child networks}, every internal node must have at least one child that is a tree node;
+\emph{tree-sibling networks} are constrained by every reticulation node having at least one sibling that is a
+tree node \citep{cardona2008extended}. Further, sub-types can be defined by restrictions on
+the topology of subgraphs that contain the reticulation loops (cycles containing a reticulation node).
+A reticulation loop that does not share any node with another reticulation loop is termed a gall
+\citep[][p.\ 237]{gusfield2014recombinatorics}; a reticulation network containing only galls is
+called a \emph{galled tree}. Reticulation loops that are joined together in a reticulation network
+by at least one shared node form a blob; a \emph{level-$k$ network} is one where the maximal number
+of reticulation loops inside a blob is $k$ \citep{choy2005computing}. Galled trees are level-1 networks,
+and the tractability afforded by their relatively simple structure results in a number of attractive
+combinatorial properties \citep{wang2001perfect, gusfield2004optimal}. We consider the general case of
+an ARG, without applying any of these restrictions.
+
+% AI: have not mentioned any implicit or un-rooted networks (split, median, haplotype, reticulograms, etc).
+% Also: are we not defining ARGs as ``binary networks''? Might be good to specify this here.
 
@@
 % examples. Probably want to clear this up a bit.
-Formally, we can define an EARG as a tuple $(e, \sigma)$, where $e$
+Formally, we can define an eARG as a tuple $(e, \sigma)$, where $e$
 is an ordered list of edges defining a directed acyclic graph and
@@
 which defines the time of each node.
-An example is given in Fig.~\ref{fig-arg-data-structure}A.
+An example is given in Fig.~\ref{fig-arg-data-structure}.
 
@@
 
-The ordering of edges is a vital element of the EARG encoding.
+The ordering of edges is a vital element of the eARG encoding.
 This is because there is a fundamental ambiguity involved
@@
 in terms of the process in which we extract the local trees
-along the genome from an EARG.
+along the genome from an eARG.
 To recover
@@
 
-\textcolor{red}{TODO: reduce this down to one paragraph, which
-defines the relationship between what we define as an ARG and
-what people mean by phylogenetic networks from the literature.
-We probably don't need to get into quite so much detail on
-the different types that people have studies. Essentially,
-we're ending the section with a brief review which situates
-our definition in the literature. I think pulling apart the
-idea of a ``reticulation'' vs what we call a recombinant
-is essential here.}
-
-The term \emph{phylogenetic network} has been broadly accepted to describe the most general
-type of genealogical graph, defined by \citet{huson2010phylogenetic} as ``any graph used to
-represent evolutionary relationships (either abstractly or explicitly) between a set of taxa
-that labels some of its nodes'' (although more narrow and context-dependent definitions can also
-be found in the literature). Explicit rooted phylogenetic networks are rooted DAGs,
-with leaves labelled by the taxa (species, groups, individuals or sequences), and nodes classed
-as either reticulation nodes (with in-degree $\geq 2$) or tree nodes.
-
-Some sub-classes of explicit rooted phylogenetic networks are defined based on the evolutionary event that the
-reticulation nodes capture \citep{huson2010phylogenetic}. In \emph{reassortment} networks, used to
-model the evolution of viruses, reticulation nodes correspond to exchange of
-(non-recombining) genomic segments between viral particles inside a co-infected host.
-In \emph{hybridisation} networks, reticulation nodes correspond to genetic material from two
-different species combining to create a hybrid. In \emph{recombination} networks, the reticulation nodes
-correspond to recombination events (with leaves labelled by sequences and edges labelled by mutations).
-The ARG is a subtype of a recombination network under this definition (restricting to one crossover
-recombination per reticulation node).
-% AI: Not sure about this - do ARGs also encompass multiple crossover recombinations and gene
-% conversion (seems implied above)?
-
-Sub-classes of explicit rooted phylogenetic networks can also be defined by restrictions on their topology.
-In \emph{binary networks}, all internal nodes must have either in-degree two and out-degree one or in-degree
-one and out-degree two (with the root having out-degree two and the leaves in-degree 1) \citep{steel2016phylogeny}.
-In \emph{tree-child networks}, every internal node must have at least one child that is a tree node;
-\emph{tree-sibling networks} are constrained by every reticulation node having at least one sibling that is a
-tree node \citep{cardona2008extended}. Further, sub-types can be defined by restrictions on
-the topology of subgraphs that contain the reticulation loops (cycles containing a reticulation node).
-A reticulation loop that does not share any node with another reticulation loop is termed a gall
-\citep[][p.\ 237]{gusfield2014recombinatorics}; a reticulation network containing only galls is
-called a \emph{galled tree}. Reticulation loops that are joined together in a reticulation network
-by at least one shared node form a blob; a \emph{level-$k$ network} is one where the maximal number
-of reticulation loops inside a blob is $k$ \citep{choy2005computing}. Galled trees are level-1 networks,
-and the tractability afforded by their relatively simple structure results in a number of attractive
-combinatorial properties \citep{wang2001perfect, gusfield2004optimal}. We consider the general case of
-an ARG, without applying any of these restrictions.
-
-% AI: have not mentioned any implicit or un-rooted networks (split, median, haplotype, reticulograms, etc).
-% Also: are we not defining ARGs as ``binary networks''? Might be good to specify this here.
-
 \section*{Ancestry resolution}\label{Ancestry_resolution}
 % This is a rough first pass. Will clarify terminology later.
+% Need to introduce Hudson's algorithm as concerned with simulating the process of
+% the coalescent with recombination as efficiently as possible, and properly merge in
+% with this section.
+
 The definition of an ARG as a data structure in terms of
@@
 
-We can define a sample-resolved GARG as a tuple $(S, E)$
+We can define a sample-resolved gARG as a tuple $(S, E)$
 where $S \subset \mathbb{N}$ is a set of sample nodes and
@@
 
+% We should talk about ARG and HUD here. Fig~\ref{fig-ancestry-resolution}A shows
+% ARG (the Griffiths graph). However, we have a small issue that subfig B
+% doesn't strictly show HUD, because it doesn't terminate as soon as all samples have coalesced.
+% We could potentially colour the top (non-HUD) section of the plot in a different colour?
+
 Given a gARG $E$ and the set of samples $S$ we can \emph{resolve} its
@@
 
-This process is illustrated in Fig~\ref{fig-ancestry-resolution}.
+This process is illustrated in Fig~\ref{fig-ancestry-resolution}. It is intimately linked to 
+Hudson's algorithm which operates by
+tracking segments of ancestral material carried by ancestors as
+we go backwards in time.
+
+% jk-note unclear here about the best choice of notation. We could have
+% lists of tuples (l, r, a), or a tuple of vectors (l_j, r_j, a_j),
+% of a tuple (I_j, a_j), where I is a vector of intervals. Going to
+% stick with the original tuple notation for now.
+%
+% Each lineage consists of a set of
+% equal-length vectors $\mathbf{l}$, $\mathbf{r}$ and $\mathbf{a}$
+% such that for each $1 \leq j \leq \left| l \right|$,
+% $[\mathbf{l}_j, \mathbf{r}_j)$ is a half-closed genomic interval and
+% $\mathbf{a}_j$ is an integer
+% tracking the number of samples the lineage is ancestral to over the interval.
+% Assume that the intervals are sorted from left-to-right so that $\mathbf{l}_1$
+% and $\mathbf{r}_{|\mathbf{r}|}$ are the left- and right-most positions
+% covered by the lineage, respectively.
+Each lineage consists of a list of
+disjoint ancestry segments $(\ell, r, a)$, where
+$[\ell, r)$ is a half-closed genomic interval and $a$ is an integer
+tracking the number of samples to which the lineage is ancestral over that interval.
+(We also usually track the tree node associated with each segment, but
+that is not important for our purposes here so we omit it.)
+If we have $n$ samples and a genome of length $m$, the process begins with $n$ lineages
+of the form $\{(0, m, 1)\}$. The process then works backwards in time from
+the present day as a series of random common ancestor or recombination events.
+Recombination events occur at a rate determined by the amount of ancestral material and
+the way in which it is distributed along a chromosome.
+Let $L$ be the set of ancestors at a given time $t$. Recombination events
+happen at rate $\rho \nu / (m - 1)$ where
+\[
+\nu = \sum_{x \in L}\left( \max_{(\ell, r, a) \in x}r
+    - \min_{(\ell, r, a) \in x}\ell - 1 \right)
+\]
+is the number of available `links' that may be broken. Thus, the rate of
+recombination is determined by the left- and right-most extent of the
+ancestral material carried by each lineage. At a recombination
+event we choose one of these links uniformly and break it. Given a lineage
+$x = [(\ell_j, r_j, a_j)]$ and a breakpoint $k$, we have two lineages
+$x_1$ and $x_2$ such that FILL IN DETAILS
+
+When $k = |L|$ lineages are present, common ancestor events
+occur at rate $\binom{k}{2}$. In a common ancestor event, two lineages
+are chosen uniformly at random and their ancestry segments merged.
+If we have overlapping intervals of ancestry from the two lineages,
+say, $(\ell, r, a_1)$ and $(\ell, r, a_2)$, a
+\emph{coalescence} occurs and ancestor represented by the current event
+will be present as a node (at least) in the marginal trees covering
+the interval $[\ell, r)$. The result of this coalescence is a segment
+$(\ell, r, a_1 + a_2)$, and if $a_1 + a_2 < n$ it is included in the
+ancestry for the new lineage. Otherwise, if $a_1 + a_2 = n$ we know that
+we have found the most recent common ancestor of all samples in
+the interval $[\ell, r)$ and so we do not need to simulate its history any further.
+Nonoverlapping intervals of ancestry from the two lineages are included
+in the resulting lineage without changes. Eventually, as the process continues,
+we find resultant lineages in which all segments have fully coalescenced,
+and so the number of extant lineages gradually dwindles down to zero.
+
+The state-space of the Griffiths process is much simpler than Hudson's algorithm,
+which greatly facilitates mathematical reasoning. This simplicity comes at a
+substantial cost, however, if we wish to use it as a practical means of
+simulating recombinant ancestry. The big ARG is vast, and any realisation
+for even moderate levels of recombination is far too large to be of practical
+use. The number of events in the big ARG, all the way back to the GMRCA
+is $O(e^\rho)$~\citep{griffiths1997ancestral}, whereas the number
+of events required to simulate the little ARG is
+$O(\rho^2)$~\citep{hein2004gene,baumdicker2021efficient}.
+This disparity in the number of events in the two formulations is
+because the majority of the events that occur in the big ARG do
+not affect the genetic ancestry of the sample in any way. Recombination
+events that occur outside of ancestral material do not have any bearing
+on the ancestry of the sample, and so the structure is hugely redundant.
+As~\cite{wiuf1999recombination} note,
+``an `ancestral' sequence in the birth and death process
+need not have any genetic material in common with a
+sequence descended from it.''
+% Full quote:
+% This process simplifies mathematics on the account that the notion of an
+% ancestor will have a less restrictive meaning than usual:
+% An ``ancestral'' sequence in the birth and death process
+% need not have any genetic material in common with a
+% sequence descended from it.
 
@@
 
-\section*{Evaluation of sampling probabilities}
+\section*{Evaluation of sampling probabilities}\label{probabilities}
+
+\begin{figure}[ht]
+\centering
+% FIXME this is a quick nasty hack to make the figure a bit smaller and
+% prevent flushing all the other figures to the end of the document.
+\scalebox{0.5}{
+\begin{tabular}{cc}
+\begin{tikzpicture}
+	\draw (-0.4, -0.2) -- (0.4, -0.2) -- (0.4, -0.3) -- (-0.4, -0.3) -- (-0.4, -0.2);
+	\draw (-0.1, -0.1) -- (-0.1, -0.4);
+	\node [label=below:{\small $v$}] at (-0.1, -0.2) {};
+	\draw (3.6, -0.2) -- (4.4, -0.2) -- (4.4, -0.3) -- (3.6, -0.3) -- (3.6, -0.2);
+	\draw (5.6, -0.2) -- (6.4, -0.2) -- (6.4, -0.3) -- (5.6, -0.3) -- (5.6, -0.2);
+
+	\draw (4,0) -- (4, 1.2) -- (6, 1.2) -- (6,0);
+	\draw (4.6, 1) -- (5.4, 1) -- (5.4, 0.9) -- (4.6, 0.9) -- (4.6, 1);
+	\draw (5.1, 1.1) -- (5.1, 0.8);
+	\node [label=below:{\small $w$}] at (5.1, 1) {};
+
+	\draw (0, 0) -- (0, 1.7) -- (-1,1.7) -- (1,1.7);
+	\draw (-1.4, 1.5) -- (-1.1, 1.5) -- (-1.1, 1.4) -- (-1.4, 1.4) -- (-1.4, 1.5);
+	\draw (-1.1, 1.45) -- (-0.6, 1.45);
+	\draw (0.6, 1.45) -- (0.9, 1.45);
+	\draw (0.9, 1.5) -- (1.4, 1.5) -- (1.4, 1.4) -- (0.9, 1.4) -- (0.9, 1.5);
+	\draw (1.2, 1.6) -- (1.2, 1.3);
+	\node [label=below:{\small $x$}] at (1.2, 1.5) {};
+
+	\draw (5, 1.2) -- (5, 2.4) -- (4,2.4) -- (6,2.4);
+	\draw (3.6, 2.2) -- (4.1, 2.2) -- (4.1, 2.1) -- (3.6, 2.1) -- (3.6, 2.2);
+	\draw (4.1, 2.15) -- (4.4, 2.15);
+	\draw (5.6, 2.15) -- (6.1, 2.15);
+	\draw (6.1, 2.2) -- (6.4, 2.2) -- (6.4, 2.1) -- (6.1, 2.1) -- (6.1, 2.2);
+
+	\draw (1, 1.7) -- (1, 3.1) -- (0,3.1) -- (2,3.1);
+	\draw (-0.4, 2.85) -- (0.2, 2.85);
+	\draw (0.2, 2.9) -- (0.4, 2.9) -- (0.4, 2.8) -- (0.2, 2.8) -- (0.2, 2.9);
+	\draw (1.6, 2.85) -- (1.9, 2.85);
+	\draw (1.9, 2.9) -- (2.2, 2.9) -- (2.2, 2.8) -- (1.9, 2.8) -- (1.9, 2.9);
+	\draw (2.2, 2.85) -- (2.4, 2.85);
+
+	\draw (-1, 1.7) -- (-1, 4.5) -- (0,4.5) -- (0,3.1);
+	\node [scale=0.2,label=above left:{\small A}] at (-0.5,4.5) {A};
+	\draw (-0.9, 4.3) -- (-0.6, 4.3) -- (-0.6, 4.2) -- (-0.9, 4.2) -- (-0.9, 4.3);
+	\draw (-0.6, 4.25) -- (-0.3, 4.25);
+	\draw (-0.3, 4.3) -- (-0.1, 4.3) -- (-0.1, 4.2) -- (-0.3, 4.2) -- (-0.3, 4.3);
+	\draw (-0.5, 4.4) -- (-0.5, 4.1);
+	\node [label=below:{\small $y$}] at (-0.5, 4.3) {};
+
+	\draw (4,2.4) -- (4,5.4) -- (2,5.4) -- (2,3.1);
+	\node [scale=0.2,label=above left:{\small B}] at (3,5.4) {B};
+	\draw (2.6, 5.2) -- (2.9, 5.2) -- (2.9, 5.1) -- (2.6, 5.1) -- (2.6, 5.2);
+	\draw (2.9, 5.15) -- (3.4, 5.15);
+
+	\draw (-0.5, 4.5) -- (-0.5, 6) -- (-1.5,6) -- (0.5,6);
+	\draw (-1.9, 5.75) -- (-1.3, 5.75);
+	\draw (-1.3, 5.8) -- (-1.1, 5.8) -- (-1.1, 5.7) -- (-1.3, 5.7) -- (-1.3, 5.8);
+	\draw (0.1, 5.8) -- (0.4, 5.8) -- (0.4, 5.7) -- (0.1, 5.7) -- (0.1, 5.8);
+	\draw (0.4, 5.75) -- (0.9, 5.75);
+
+	\draw (0.5, 6) -- (0.5, 7.2) -- (3,7.2) -- (3,5.4);
+	\node [scale=0.2,label=above:{\small C}] at (1.75,7.2) {C};
+	\draw (1.35, 6.95) -- (2.15, 6.95);
+
+	\draw (-1.5, 6) -- (-1.5, 8);
+	\draw (6, 2.4) -- (6, 8);
+
+	% Edge annotations above each event
+	\node [label=left:{\small$\{(0,m,1)\}$}] at (0.2,0.6) {};
+	\node [label=left:{\small$\{(0,m,1)\}$}] at (4.2,0.4) {};
+	\node [label=right:{\small$\{(0,m,1)\}$}] at (5.8,0.4) {};
+	\node [label=left:{\small$\{(0,m,2)\}$}] at (5.2,1.6) {};
+	\node [label=left:{\small$\{(0,v,1)\}$}] at (-0.8,2.3) {};
+	\node [label=right:{\small$\{(v,m,1)\}$}] at (0.8,2.1) {};
+	\node [label=right:{\small$\{(0,w,2)\}$}] at (3.8,3.9) {};
+	\node [label=left:{\small$\{(w,m,2)\}$}] at (6.2,5) {};
+	\node [label=right:{\small$\{(x,m,1)\}$}] at (-0.2,3.8) {};
+	\node [label=right:{\small$\{(v,x,1)\}$}] at (1.8,4.1) {};
+	\node [label=left:{\small$\{(0,v,1), (x,m,1)\}$}] at (-0.3,5.2) {};
+	\node [label=right:{\small$\{(0,v,2)\}$}] at (2.8,6.4) {};
+	\node [label=left:{\small$\{(x,m,1)\}$}] at (-1.3,7) {};
+	\node [label=right:{\small$\{(0,v,1)\}$}] at (0.3,6.4) {};
+
+	% Dashed lines for start and end times
+	\draw[dashed] (-2, 0) -- (11.1, 0);
+	\node [label=left:{$t = 0$}] at (-2,0) {};
+	\draw[dashed] (-2, 8) -- (11.1, 8);
+	\node [label=left:{$t = T$}] at (-2,8) {};
+
+	% Numbers of extant ancestors and links, from top to bottom
+	\node[label=right:{CA \; RE}] at (7.5, 8.2) {};
+	\node[label=right:{$\binom{2}{2}$ \; $\frac{2 m - x - w - 2}{m - 1} \rho$}] at (7.5, 7.6) {};
+	\node[label=right:{$\binom{4}{2}$ \; $\frac{2 m + 2 v - x - w - 3}{m - 1} \rho$}] at (7.5, 6.6) {};
+	\node[label=right:{$\binom{3}{2}$ \; $\frac{2 m + v - w - 3}{m - 1} \rho$}] at (7.5, 5.7) {};
+	\node[label=right:{$\binom{4}{2}$ \; $\frac{2 m + x - v - 4}{m - 1} \rho$}] at (7.5, 4.95) {};
+	\node[label=right:{$\binom{5}{2}$ \; $\frac{2 m - 5}{m - 1} \rho$}] at (7.5, 3.8) {};
+	\node[label=right:{$\binom{4}{2}$ \; $\frac{2 m - 4}{m -1} \rho$}] at (7.5, 2.75) {};
+	\node[label=right:{$\binom{3}{2}$ \; $\frac{2 m - 3}{m - 1} \rho$}] at (7.5, 2.05) {};
+	\node[label=right:{$\binom{2}{2}$ \; $2 \rho$}] at (7.5, 1.45) {};
+	\node[label=right:{$\binom{3}{2}$ \; $3 \rho$}] at (7.5, 0.6) {};
+
+	% Gray dashed lines to visually separate holding times
+	\draw[color=gray, dashed] (7.5, 1.2) -- (11.1, 1.2);
+	\draw[color=gray, dashed] (7.5, 1.7) -- (11.1, 1.7);
+	\draw[color=gray, dashed] (7.5, 2.4) -- (11.1, 2.4);
+	\draw[color=gray, dashed] (7.5, 3.1) -- (11.1, 3.1);
+	\draw[color=gray, dashed] (7.5, 4.5) -- (11.1, 4.5);
+	\draw[color=gray, dashed] (7.5, 5.4) -- (11.1, 5.4);
+	\draw[color=gray, dashed] (7.5, 6) -- (11.1, 6);
+	\draw[color=gray, dashed] (7.5, 7.2) -- (11.1, 7.2);
+\end{tikzpicture}
+&
+\begin{tikzpicture}
+	\draw (-0.4, -0.2) -- (0.4, -0.2) -- (0.4, -0.3) -- (-0.4, -0.3) -- (-0.4, -0.2);
+	\draw (-0.1, -0.1) -- (-0.1, -0.4);
+	\node [label=below:{\small $v$}] at (-0.1, -0.2) {};
+	\draw (3.6, -0.2) -- (4.4, -0.2) -- (4.4, -0.3) -- (3.6, -0.3) -- (3.6, -0.2);
+	\draw (5.6, -0.2) -- (6.4, -0.2) -- (6.4, -0.3) -- (5.6, -0.3) -- (5.6, -0.2);
+
+	\draw (4,0) -- (4, 1.2) -- (6, 1.2) -- (6,0);
+	\draw (4.6, 1) -- (5.4, 1) -- (5.4, 0.9) -- (4.6, 0.9) -- (4.6, 1);
+	\draw (5.1, 1.1) -- (5.1, 0.8);
+	\node [label=below:{\small $w$}] at (5.1, 1) {};
+
+	\draw (0, 0) -- (0, 1.7) -- (-1,1.7) -- (1,1.7);
+	\draw (-1.4, 1.5) -- (-1.1, 1.5) -- (-1.1, 1.4) -- (-1.4, 1.4) -- (-1.4, 1.5);
+	\draw (-1.1, 1.45) -- (-0.6, 1.45);
+	\draw (0.6, 1.45) -- (0.9, 1.45);
+	\draw (0.9, 1.5) -- (1.4, 1.5) -- (1.4, 1.4) -- (0.9, 1.4) -- (0.9, 1.5);
+	\draw (1.2, 1.6) -- (1.2, 1.3);
+	\node [label=below:{\small $x$}] at (1.2, 1.5) {};
+
+	\draw (5, 1.2) -- (5, 2.4) -- (4,2.4) -- (6,2.4);
+	\draw (3.6, 2.2) -- (4.1, 2.2) -- (4.1, 2.1) -- (3.6, 2.1) -- (3.6, 2.2);
+	\draw (4.1, 2.15) -- (4.4, 2.15);
+	\draw (5.6, 2.15) -- (6.1, 2.15);
+	\draw (6.1, 2.2) -- (6.4, 2.2) -- (6.4, 2.1) -- (6.1, 2.1) -- (6.1, 2.2);
+	\draw [color=red](5.8, 2.3) -- (5.8, 2.0);
+	\node [label={[red]below:{\small $z$}}] at (5.8, 2.1) {};
+
+	\draw (1, 1.7) -- (1, 3.1) -- (0,3.1) -- (2,3.1);
+	\draw (-0.4, 2.85) -- (0.2, 2.85);
+	\draw (0.2, 2.9) -- (0.4, 2.9) -- (0.4, 2.8) -- (0.2, 2.8) -- (0.2, 2.9);
+	\draw (1.6, 2.85) -- (1.9, 2.85);
+	\draw (1.9, 2.9) -- (2.2, 2.9) -- (2.2, 2.8) -- (1.9, 2.8) -- (1.9, 2.9);
+	\draw (2.2, 2.85) -- (2.4, 2.85);
+
+	\draw (-1, 1.7) -- (-1, 4.5) -- (0,4.5) -- (0,3.1);
+	\draw (-0.9, 4.3) -- (-0.6, 4.3) -- (-0.6, 4.2) -- (-0.9, 4.2) -- (-0.9, 4.3);
+	\draw (-0.6, 4.25) -- (-0.3, 4.25);
+	\draw (-0.3, 4.3) -- (-0.1, 4.3) -- (-0.1, 4.2) -- (-0.3, 4.2) -- (-0.3, 4.3);
+	\draw (-0.5, 4.4) -- (-0.5, 4.1);
+	\node [label=below:{\small $y$}] at (-0.5, 4.3) {};
+
+	\draw (6, 2.4) -- (6, 3.8) -- (7, 3.8);
+	\draw [color=red](6, 3.8) -- (5, 3.8);
+	\draw (4.6, 3.55) -- (5.4, 3.55);
+	\draw (6.6, 3.55) -- (7.1, 3.55);
+	\draw (7.1, 3.6) -- (7.4, 3.6) -- (7.4, 3.5) -- (7.1, 3.5) -- (7.1, 3.6);
+
+	\draw (4,2.4) -- (4,5.4) -- (2,5.4) -- (2,3.1);
+	\draw (2.6, 5.2) -- (2.9, 5.2) -- (2.9, 5.1) -- (2.6, 5.1) -- (2.6, 5.2);
+	\draw (2.9, 5.15) -- (3.4, 5.15);
+
+	\draw (-0.5, 4.5) -- (-0.5, 6) -- (-1.5,6) -- (0.5,6);
+	\draw (-1.9, 5.75) -- (-1.3, 5.75);
+	\draw (-1.3, 5.8) -- (-1.1, 5.8) -- (-1.1, 5.7) -- (-1.3, 5.7) -- (-1.3, 5.8);
+	\draw (0.1, 5.8) -- (0.4, 5.8) -- (0.4, 5.7) -- (0.1, 5.7) -- (0.1, 5.8);
+	\draw (0.4, 5.75) -- (0.9, 5.75);
+
+	\draw [color=red](5, 3.8) -- (5, 6.6) -- (4, 6.6);
+	\draw (4,6.6) -- (3,6.6) -- (3, 5.4);
+	\draw (3.6, 6.4) -- (3.9, 6.4) -- (3.9, 6.3) -- (3.6, 6.3) -- (3.6, 6.4);
+	\draw (3.9, 6.35) -- (4.4, 6.35);
+
+	\draw (0.5, 6) -- (0.5, 7.2) -- (4,7.2) -- (4,6.6);
+	\draw (1.85, 6.95) -- (2.65, 6.95);
+
+	\draw (-1.5, 6) -- (-1.5, 8);
+	\draw [color=red](2.25, 7.2) -- (2.25, 8);
+	\draw (7, 3.8) -- (7, 8);
+
+	% Dashed lines for start and end times
+	\draw[dashed] (-2, 0) -- (11, 0);
+	\node [label=left:{$t = 0$}] at (-2,0) {};
+	\draw[dashed] (-2, 8) -- (11, 8);
+	\node [label=left:{$t = T$}] at (-2,8) {};
+
+	% Numbers of extant ancestors and links, from top to bottom
+	\node[label=right:{CA \; RE}] at (7.5, 8.2) {};
+	\node[label=right:{$\binom{3}{2}$ \; $3 \rho$}] at (7.5, 7.6) {};
+	\node[label=right:{$\binom{4}{2}$ \; $4 \rho$}] at (7.5, 6.9) {};
+	\node[label=right:{$\binom{5}{2}$ \; $5 \rho$}] at (7.5, 6.3) {};
+	\node[label=right:{$\binom{4}{2}$ \; $4 \rho$}] at (7.5, 5.7) {};
+	\node[label=right:{$\binom{5}{2}$ \; $5 \rho$}] at (7.5, 4.95) {};
+	\node[label=right:{$\binom{6}{2}$ \; $6 \rho$}] at (7.5, 4.15) {};
+	\node[label=right:{$\binom{5}{2}$ \; $5 \rho$}] at (7.5, 3.45) {};
+	\node[label=right:{$\binom{4}{2}$ \; $4 \rho$}] at (7.5, 2.75) {};
+	\node[label=right:{$\binom{3}{2}$ \; $3 \rho$}] at (7.5, 2.05) {};
+	\node[label=right:{$\binom{2}{2}$ \; $2 \rho$}] at (7.5, 1.45) {};
+	\node[label=right:{$\binom{3}{2}$ \; $3 \rho$}] at (7.5, 0.6) {};
+
+	% Gray dashed lines to visually separate holding times
+	\draw[color=gray, dashed] (7.5, 1.2) -- (11, 1.2);
+	\draw[color=gray, dashed] (7.5, 1.7) -- (11, 1.7);
+	\draw[color=gray, dashed] (7.5, 2.4) -- (11, 2.4);
+	\draw[color=gray, dashed] (7.5, 3.1) -- (11, 3.1);
+	\draw[color=gray, dashed] (7.5, 3.8) -- (11, 3.8);
+	\draw[color=gray, dashed] (7.5, 4.5) -- (11, 4.5);
+	\draw[color=gray, dashed] (7.5, 5.4) -- (11, 5.4);
+	\draw[color=gray, dashed] (7.5, 6) -- (11, 6);
+	\draw[color=gray, dashed] (7.5, 6.6) -- (11, 6.6);
+	\draw[color=gray, dashed] (7.5, 7.2) -- (11, 7.2);
+\end{tikzpicture}
+\end{tabular}
+}
+\caption{(A)
+A realisation of the graph traversed by Hudson's algorithm started from a
+sample of three chromosomes of length $m$ at time $t = 0$, and
+propagated until time $T$. The MRCA on the genetic interval $[v, w)$ is reached
+at event B, while that on $[0, v)$ is reached at event C.
+The non-ancestral segment $[v, w)$ above
+A contributes to the rate of effective recombinations because it
+is trapped between ancestral segments. The two columns titled CA and RE
+are the respective rates of mergers and recombinations when
+the recombination rate is $\rho$.
+(B) A corresponding realisation of a big ARG, which augments Hudson's algorithm
+by tracking nonancestral lineages. The result is a simpler state space and
+dynamics, at the cost of extra nodes and edges, highlighted in red, which do
+not affect the local tree at any site.}
+\label{hudson_vs_bigARG}
+\end{figure}
 

### 2022-04-07 paper.tex

@@
 defines a genetic ancestry relationship between two nodes,
-and annotations on those edges define the precise patterns of
-ancestry in intervals along the genome.
-We use this framework to define some classifications and metrics of ARG nodes,
-and discuss the properties of some recently developed inference methods.
+and annotations on those edges define the patterns of
+inheritance in intervals along the genome.
+We show how this definition can be used to generalise the encoding to
+cover both the standard and several simplified versions of ARGs, as well as
+enabling efficient access to local trees along the genome. Additionally, we
+define some classifications and metrics of ARG nodes. Finally, we relate
+these generalisations to the ARGs produced by a number of recently
+developed ancestral inference methods.
 % jk-note: this ends a bit flat - we should say other stuff like we clarify

### 2022-04-07 paper.tex
Minor text changes

@@
 trees~\citep{tajima1983evolutionary,tavare1984line}
+% Do we mean "despite the fact that analysis of genetic data relies on methods
+% which ultimately derive from the inheritance paths defined by genetic trees
+% and the mutations upon them, analysis so far has focussed on summaries ...
 such as allele frequencies and site frequency
@@
 \cite{kelleher2016efficient} provide a description of his algorithm along with
-an efficient implementation, the details of which turn out to be highly relevant in 
+an efficient implementation, the details of which turn out to be highly relevant to 
 later sections of the current paper (see the ``\nameref{Ancestry_resolution}'' section).
 
-The Ancestral Recombination Graph (ARG) was later introduced by
-Griffiths~\citep{griffiths1991two,ethier1990two,griffiths1996ancestral,
-griffiths1997ancestral}, and is a
-closely related and complementary treatment of the same process. Where
-Hudson's algorithm is focused on the efficient \emph{simulation} of the
-process, the original ARG literature is focused on mathematical
-results about the stochastic process. Griffiths and colleagues used the word
+Although Hudson's algorithm was accompanied by a graph
+schematic~\citep[][Fig. 5]{hudson1983properties}, the term
+``Ancestral Recombination Graph'' or ARG, was introduced later by
+Griffiths~\citep{griffiths1991two,ethier1990two,griffiths1996ancestral, griffiths1997ancestral},
+who provided a closely related and complementary treatment.
+Where Hudson's algorithm focusses on efficient \emph{simulation}, the original
+ARG literature is focused on mathematical results about the underlying stochastic
+process. Griffiths and colleagues used the word
 ``ARG'' to refer to both a specific branching-coalescing stochastic process and
-the graph structure derived from this process. For clarity, we will refer to the
-process as the Griffiths process, or the coalescent with recombination (CwR)
+the graph structure derived from this process, but for clarity and following recent usage,
+we will use ARG to refer to a graph structure, and refer to the
+process as the Griffiths process or the coalescent with recombination
 \citep{minichiello2006mapping}.
@@
 happen in finite time because of the linear vs quadratic rates of branching
-and coalescing. The graph structure and breakpoints associated with
+and coalescing. Later, we will show how the graph structure and breakpoints
+associated with
 recombination nodes provides sufficient information to later recover the marginal
-trees and calculate sampling probabilities (see the ``\nameref{probabilities}'' section).
-
-[Review of other work done on the ARG stochastic process here.
-Discuss \citet{wiuf1999ancestry}. Wiuf and Hein's classical pair of papers
-clarified the relationship between Hudson's algorithm and the ARG.
-Explain that ARG simulation is inherently less efficient than Hudson's
-algorithm because we must track all lineages, even those that correspond
-to sections that have coalesced, or we won't have a complete ARG.
-This might be tricky to communicated without a figure, but is an
-esoteric point perhaps not worth spending a figure on.]
+trees and calculate sampling probabilities. However, this process traces out
+a larger graph than the Hudson algorithm (see the ``\nameref{Ancestry_resolution}'' section),
+and is consequently considerably less efficient \citep{wiuf1999ancestry}.
 
@@
 % What was the historical progression from one to the other?
+
+% NB: should this para go into the previous section? It would be
+% clean to start the section with the subsequent para
+% starting "The Griffiths ARG encoding is an elegant and mathematically economical description of the CwR"
 Early work on ARG inference focused on the problem of
@@
 ARG realisations~\citep{song2003parsimonious,song2005efficient,lyngso2005minimum}.
-The distinction between the ARG as a stochastic process
-and the ARG as a \emph{data structure}~\citep{minichiello2006mapping}
-became important at this point,
+This highlighted the distinction between the ARG as a stochastic process
+and the ARG as a \emph{data structure}~\citep{minichiello2006mapping},
 since parsimonious ARGs have no immediate connection to
@@
 since they contain many invisible recombinations
+% Do we need the following caveat?
 (although those produced under the SMC may be parsimonious, depending
@@
 The Griffiths ARG encoding is an elegant and mathematically
-economical description of the coalescent with recombination,
-but it has significant drawbacks as the basis for a general purpose
-storage and interchange format. To avoid confusion, we will refer to this
-ARG encoding as an ``Event ARG'', or EARG.
-The fundamental units of an EARG are the \emph{events} that correspond directly
-to those that
-occur in the ARG stochastic process described in the previous section.
-Each event corresponds to a node in the graph, and there are therefore
-two types of node: common ancestor and recombination. As in the
-stochastic process, edges represent lineages, and those lineages
-are modified by the events that they pass through.
+economical description of the coalescent with recombination, whose
+fundamental units are the  \emph{events} that correspond directly
+to those that occur in the associated stochastic process. We will refer
+to this ARG encoding as an ``Event ARG'', or eARG: 
+an example is given in Fig.~\ref{fig-arg-data-structure}.
+Each node in the graph corresponds to an event: therefore the graph
+contains both recombination nodes (node \textsf{D}) and common ancestor
+nodes (nodes \textsf{E}, \textsf{F}, and \textsf{G}) corresponding to their
+respective event types. Edges represent lineages,
+and those lineages are modified by the events that they pass through.
 
@@
 if $u$ is a recombination event with breakpoint $x$ and
-$\sigma(u) = \varnothing$ otherwise.
-(Note that this formulation is equivalent to stipulating two
-different ``types'' of node and only storing breakpoints
-with recombination nodes).
+$\sigma(u) = \varnothing$ otherwise. (This formulation is equivalent to stipulating that
+a recombination node is of a different ``type'' to other nodes,
+with breakpoints only associated with this specific node type).
 Each edge $e_j = (c, p)$ describes an edge between
@@
 which defines the time of each node.
-An example is given in Fig.~\ref{fig-arg-data-structure}.
 
@@
 % at a time any more.
-The Griffiths ARG encoding is also limited in the types of ancestral
-history that can be described. This ARG encoding is derived from the coalescent
+The eARG is also limited in the types of ancestral history that can
+be described. This ARG encoding is derived from the coalescent
 with recombination, which explicitly assumes that only one type
@@
 much smaller than $N$ at each time. While these approximations are
-reasonable
-to make when studying the Kreitman \emph{Drosphila Melanogaster}
-dataset~\citep{kreitman1983nucleotide} ($n=11$, $N\approx 10^6$,
-at 43 variant sites over x Kb \textcolor{red}{details and citation
-for Ne estimate}) that is classically used for benchmarking
-ARG inference methods [cites],
+reasonable to make for small datasets, such as
+the classic \citet{kreitman1983nucleotide} \emph{Drosophila melanogaster}
+ADH gene dataset (comprising 11 genomes over 2.7 kb of sequence with 43 SNPs,
+as used in the ``\nameref{ARG_inference_methods}'' section),
 they are highly questionable in modern datasets. % absurd really
-For example, there are now many
-human datasets consisting of hundreds of thousands of
+For example, many
+human datasets consist of hundreds of thousands of
 genomes~\citep{bycroft2018genome,karczewski2020mutational,tanjo2021practical},
@@
 
-Another limitation of the Griffiths ARG encoding illustrated in
-Fig.~\ref{fig-arg-data-structure}A is the inability to represent
+A final limitation of the eARG encoding illustrated in
+Fig.~\ref{fig-arg-data-structure} is the inability to represent
 multiple recombinations along a chromosome, gene conversion,
@@
 
+% Potentially useful text from a previous iteration:
+% Wiuf and Hein's classical pair of papers
+% clarified the relationship between Hudson's algorithm and the ARG.
+% ARG simulation is inherently less efficient than Hudson's
+% algorithm because we must track all lineages, even those that correspond
+% to sections that have coalesced, or we won't have a complete ARG.
+
 The state-space of the Griffiths process is much simpler than Hudson's algorithm,

### 2022-04-07 paper.tex
Make node labels lowercase
And add subfigure labels

@@
 % about the process as choosing edges to merge, not nodes.
-% We could also refer to Fig.~\ref{fig-arg-data-structure} if we want a simple picture
+% We could also refer to Fig.~\ref{fig-arg-data-structure}B if we want a simple picture
 % to go with this
@@
 
-\section*{The Event ARG data structure}
+\section*{The Event ARG data structure}\label{eARG}
 % When did we move from the ARG is a process to the ARG is a data structure?
@@
 to this ARG encoding as an ``Event ARG'', or eARG: 
-an example is given in Fig.~\ref{fig-arg-data-structure}.
+an example is given in Fig.~\ref{fig-arg-data-structure}B.
 Each node in the graph corresponds to an event: therefore the graph
-contains both recombination nodes (node \textsf{D}) and common ancestor
-nodes (nodes \textsf{E}, \textsf{F}, and \textsf{G}) corresponding to their
+contains both recombination nodes (node \textsf{d}) and common ancestor
+nodes (nodes \textsf{e}, \textsf{f}, and \textsf{g}) corresponding to their
 respective event types. Edges represent lineages,
@@
 
+\node [anchor=north west] at (-1.5,6) {B};
 \node [nodelabel,anchor=north west] at ($(s3) + (0,0)$) {$x$};
-\foreach \u/\lab in {s0/$\textsf{A}$, s1/$\textsf{B}$, s2/$\textsf{C}$} \node[nodelabel,anchor=north] at (\u) {\lab};
-\foreach \u/\lab in {s4/$\textsf{E}$} \node[nodelabel,anchor=south west] at (\u) {\lab};
-\foreach \u/\lab in {s5/$\textsf{F}$} \node[nodelabel,anchor=south east] at (\u) {\lab};
-\foreach \u/\lab in {s3/$\textsf{D}$, s6/$\textsf{G}$} \node[nodelabel,anchor=south] at (\u) {\lab};
+\foreach \u/\lab in {s0/$\textsf{a}$, s1/$\textsf{b}$, s2/$\textsf{c}$} \node[nodelabel,anchor=north] at (\u) {\lab};
+\foreach \u/\lab in {s4/$\textsf{e}$} \node[nodelabel,anchor=south west] at (\u) {\lab};
+\foreach \u/\lab in {s5/$\textsf{f}$} \node[nodelabel,anchor=south east] at (\u) {\lab};
+\foreach \u/\lab in {s3/$\textsf{d}$, s6/$\textsf{g}$} \node[nodelabel,anchor=south] at (\u) {\lab};
 
@@
 
-\node [nodelabel,anchor=north west] at ($(-10,5)$) {
+\node [anchor=north west] at (-14,6) {A};
+\node [nodelabel,anchor=north west] at ($(-13,5)$) {
 \begin{tabular}{c|c}
@@
 \hline
-$\textsf{A}$ & $\varnothing$ \\
-$\textsf{B}$ & $\varnothing$ \\
-$\textsf{C}$ & $\varnothing$ \\
-$\textsf{D}$ & $x$ \\
-$\textsf{E}$ & $\varnothing$ \\
-$\textsf{F}$ & $\varnothing$ \\
-$\textsf{G}$ & $\varnothing$ \\
+$\textsf{a}$ & $\varnothing$ \\
+$\textsf{b}$ & $\varnothing$ \\
+$\textsf{c}$ & $\varnothing$ \\
+$\textsf{d}$ & $x$ \\
+$\textsf{e}$ & $\varnothing$ \\
+$\textsf{f}$ & $\varnothing$ \\
+$\textsf{g}$ & $\varnothing$ \\
 \end{tabular}};
 
-\node [nodelabel,anchor=north west] at ($(-4,5)$) {
+\node [nodelabel,anchor=north west] at ($(-7,5)$) {
 \begin{tabular}{l}
@@
 \hline
-$(\textsf{A}, \textsf{E})$ \\
-$(\textsf{B}, \textsf{D})$ \\
-$(\textsf{C}, \textsf{F})$ \\
-$(\textsf{D}, \textsf{E})$ \\
-$(\textsf{D}, \textsf{F})$ \\
-$(\textsf{E}, \textsf{G})$ \\
-$(\textsf{F}, \textsf{G})$ \\
+$(\textsf{a}, \textsf{e})$ \\
+$(\textsf{b}, \textsf{d})$ \\
+$(\textsf{c}, \textsf{f})$ \\
+$(\textsf{d}, \textsf{e})$ \\
+$(\textsf{d}, \textsf{f})$ \\
+$(\textsf{e}, \textsf{g})$ \\
+$(\textsf{f}, \textsf{g})$ \\
 \end{tabular}};
 
-\node [nodelabel,anchor=north west] at ($(7,5)$) {
+\node [anchor=north west] at (9,6) {C};
+\node [nodelabel,anchor=north west] at ($(10,5)$) {
 \begin{tabular}{ll}
@@
 \hline
-$(\textsf{A}, \textsf{E})$ & $(0, L]$ \\
-$(\textsf{B}, \textsf{D})$ & $(0, L]$ \\
-$(\textsf{C}, \textsf{F})$ & $(0, L]$ \\
-$(\textsf{D}, \textsf{E})$ & $(0, x]$ \\
-$(\textsf{D}, \textsf{F})$ & $(x, L]$ \\
-$(\textsf{E}, \textsf{G})$ & $(0, L]$ \\
-$(\textsf{F}, \textsf{G})$ & $(0, L]$ \\
+$(\textsf{a}, \textsf{e})$ & $(0, L]$ \\
+$(\textsf{b}, \textsf{d})$ & $(0, L]$ \\
+$(\textsf{c}, \textsf{f})$ & $(0, L]$ \\
+$(\textsf{d}, \textsf{e})$ & $(0, x]$ \\
+$(\textsf{d}, \textsf{f})$ & $(x, L]$ \\
+$(\textsf{e}, \textsf{g})$ & $(0, L]$ \\
+$(\textsf{f}, \textsf{g})$ & $(0, L]$ \\
 \end{tabular}};
@@
 
-% What do we mean by an EARG, formally?
-% NOTE we're using the natural numbers but are using 0 as a node in the
-% examples. Probably want to clear this up a bit.
+% What do we mean by an eARG, formally?
 Formally, we can define an eARG as a tuple $(e, \sigma)$, where $e$
@@
 A final limitation of the eARG encoding illustrated in
-Fig.~\ref{fig-arg-data-structure} is the inability to represent
+Fig.~\ref{fig-arg-data-structure}A is the inability to represent
 multiple recombinations along a chromosome, gene conversion,
@@
 as breakpoints associated with recombination events and
-edge ordering rules. This is illustrated in Fig~\ref{fig-arg-data-structure}B
+edge ordering rules. This is illustrated in Fig~\ref{fig-arg-data-structure}C
 where this alternative edge annotation formulation is
@@
 The Ancestral Recombination Graph defined in terms of genomes embedded
-in a pedigree. (a) Diploid individuals, visualised in a highly inbred pedigree and
-labelled $i_1$ to $i_8$, contain both paternal (left) and maternal (right) genomes
-labelled \textsf{A} to \textsf{P}. Black lines show transmission paths connecting
-genomes in the current generation (\textsf{A} to \textsf{D}) with their ancestors.
-Two recombination breakpoints, at positions 2 and 7 mean that genomes \textsf{A}
-and \textsf{C} are independent recombinants of the paternal genomes \textsf{E}
-and \textsf{F}: in these cases the edge annotations have been shown, indicating the
+in a pedigree. (A) Diploid individuals, visualised in a highly inbred pedigree and
+labelled $I_1$ to $I_8$, contain both paternal (left) and maternal (right) genomes
+labelled \textsf{a} to \textsf{p}. Black lines show transmission paths connecting
+genomes in the current generation (\textsf{a} to \textsf{d}) with their ancestors.
+Two recombination breakpoints, at positions 2 and 7 mean that genomes \textsf{a}
+and \textsf{c} are independent recombinants of the paternal genomes \textsf{e}
+and \textsf{f}: in these cases the edge annotations have been shown, indicating the
 region of genome transmitted. Genomes are shaded such that where, backwards in time,
 they merge into a common ancestor, the merged region is darker; this shows that all
-genomic regions are fully coalesced in genome \textsf{N} in the oldest generation.
-(b) The equivalent genome ARG, where there is a single type of node (representing
+genomic regions are fully coalesced in genome \textsf{n} in the oldest generation.
+(B) The equivalent genome ARG, where there is a single type of node (representing
 a genome) along with annotations on all edges; annotations showing partial
 transmission are highlighted in bold. Recombination, such as that causing the
-breakpoint at position 7, is associated with two parent genomes (\textsf{E} and
-\textsf{F}) and a child genome (the recombinant, \textsf{C}).
-(c) The edge-annotated genome ARG in (b) fully defines the
+breakpoint at position 7, is associated with two parent genomes (\textsf{e} and
+\textsf{f}) and a child genome (the recombinant, \textsf{c}).
+(C) The edge-annotated genome ARG in (B) fully defines the
 local relationships along the genome, as shown in the local
@@
 Fig~\ref{fig-arg-in-pedigree} illustrates the basic elements of a gARG.
-Fig.~\ref{fig-arg-in-pedigree}a shows the genomes
+Fig.~\ref{fig-arg-in-pedigree}A shows the genomes
 embedded in an inbred diploid pedigree, in which individuals occur at
 discrete generations and carry two genomes. Each individual (labelled
-$i_1$ to $i_8$) therefore has a paternal genome (on the right) and a maternal one
+$I_1$ to $I_8$) therefore has a paternal genome (on the right) and a maternal one
 (on the left), although the pedigree is only shown for concreteness, and
@@
 and may be the recombined product of that parent's two genomes.
-For example, individual $i_1$ has two genomes \textsf{A} and \textsf{B},
-inherited from parents $i_3$ and $i_4$. Genome \textsf{A} is the product of
-recombining $i_3$'s two genomes \textsf{E} and \textsf{F} at position 2,
-whereas \textsf{B} was inherited directly from $i_4$'s genome \textsf{G} without
-recombination. Fig.~\ref{fig-arg-in-pedigree}b shows the bare gARG, illustrating
+For example, individual $I_1$ has two genomes \textsf{a} and \textsf{b},
+inherited from parents $I_3$ and $I_4$. Genome \textsf{a} is the product of
+recombining $I_3$'s two genomes \textsf{e} and \textsf{e} at position 2,
+whereas \textsf{b} was inherited directly from $I_4$'s genome \textsf{g} without
+recombination. Fig.~\ref{fig-arg-in-pedigree}B shows the bare gARG, illustrating
 that the patterns of
@@
 how they are combined within individuals. This ``genome-perspective''
-simply depicts the ancestry of the sample genomes $A$--$D$. The graph
+simply depicts the ancestry of the sample genomes \textsf{a}--\textsf{d}. The graph
 topology along with the genome coordinate annotations on the
 edges then uniquely defines the trees at every position
-along the genome (Fig.~\ref{fig-arg-in-pedigree}c).
+along the genome (Fig.~\ref{fig-arg-in-pedigree}C).
 
@@
 do not correspond to any particular event in the ancestry of the samples.
-For example, node \textsf{H} is the direct ancestor of \textsf{D} in
+For example, node \textsf{h} is the direct ancestor of \textsf{d} in
 Fig~\ref{fig-arg-in-pedigree}, and is not the product of either a
 coalescence or a recombination. Such nodes would usually be removed
-from the graph so that \textsf{D} descends directly from \textsf{K} (see
+from the graph so that \textsf{d} descends directly from \textsf{k} (see
 the ``\nameref{ARG_simplification}'' section) but there is no necessity for this from
@@
 and multiple parents. This would be the case in Fig~\ref{fig-arg-in-pedigree} if,
-for instance, node \textsf{A} were to have two children.
+for instance, node \textsf{a} were to have two children.
 % It is a shame that this is not actually shown in fig 3, but I don't think
@@
 Moreover, a single node can have more than 2 children, and (as we shall see in
-the ``\nameref{ARG_simplification}'' section, more than 2 parents. Another way to
+the ``\nameref{ARG_simplification}'' section), more than 2 parents. Another way to
 state this is that transitions between nodes in a gARG can encapsulate multiple
@@
 choice is where genomes represent diploid individuals in each generation, as in
-Fig~\ref{fig-arg-in-pedigree}a. In this case recombination
+Fig~\ref{fig-arg-in-pedigree}A. In this case recombination
 is not represented by a single ``recombination node'' (as in an eARG), but with
 two parent genomes between which recombination takes place (e.g.
-nodes \textsf{E} and \textsf{F}), and a
-single resulting ``recombinant'' genome (e.g. \textsf{C}). The
+nodes \textsf{e} and \textsf{f}), and a
+single resulting ``recombinant'' genome (e.g. \textsf{c}). The
 edges that link this recombinant with its two parents carry the information
 concerning breakpoint location that would normally be present on the
-``recombination nod'' in an eARG. For example, node \textsf{C} in
-Fig~\ref{fig-arg-in-pedigree} inherits genome positions 0 to 7 from \textsf{E}
-and positions 7 to 10 from \textsf{F}, indicating that the breakpoint is at position 7.
+recombination node in an eARG. For example, node \textsf{c} in
+Fig~\ref{fig-arg-in-pedigree} inherits genome positions 0 to 7 from \textsf{e}
+and positions 7 to 10 from \textsf{f}, indicating that the breakpoint is at position 7.
 
@@
 edge annotation schemes.
-(a) Edge annotations that reflect immediate inheritance, i.e. transmission of all
+(A) Edge annotations that reflect immediate inheritance, i.e. transmission of all
 genetic material between parent and child. All edges span the full genome from position 0 to 7,
 apart from those linking a recombinant and its two parents (bold).
-(b) Sample-resolved edge annotations: the Hudson simplification algorithm has been used to
+(B) Sample-resolved edge annotations: the Hudson simplification algorithm has been used to
 shorten edge spans to reflect only the upwards transmission of genetic
-material from the samples \textsf{A}, \textsf{B}, and \textsf{C}. The depiction of the genome within
+material from the samples \textsf{a}, \textsf{b}, and \textsf{c}. The depiction of the genome within
 each node is shaded as in figure \ref{fig-arg-in-pedigree}a to show
-coalescing regions of genome back to the grand MRCA genome \textsf{W}.
+coalescing regions of genome back to the grand MRCA genome \textsf{w}.
 In this case, using line widths to reflect the genomic span of each edge can
@@
 (A) A gARG simulated from a diploid Wright-Fisher
-model. Note that the genome \textsf{K} has three children, and the genomes
-\textsf{E} and \textsf{G} simultaneously have multiple parents and multiple children;
+model. Note that the genome \textsf{k} has three children, and the genomes
+\textsf{e} and \textsf{g} simultaneously have multiple parents and multiple children;
 neither could be directly represented in a Griffiths-like eARG encoding
-(see section XXX). The penultimate local tree is highlighted to aid discussion (see section YYY).
+(see the ``\nameref{eARG}'' section). The penultimate local tree is highlighted to aid discussion.
 (B) Simplified to remove all
 % trying out this terminology - can define in the text
-1-connected graph components (e.g., diamonds such as \textsf{JLMN}).
+1-connected graph components (e.g., diamonds such as \textsf{jlnm}).
 (C) Remove nodes that are unary in all local trees. This removes both recombinant nodes
-that have only one child (e.g. \textsf{N}) and nodes that have more
+that have only one child (e.g. \textsf{n}) and nodes that have more
 than one child (i.e. are ``common ancestors'') but never represent coalescences
-in the local trees (e.g. \textsf{R}).
+in the local trees (e.g. \textsf{r}).
 (D) Rewrite edges to bypass any unary nodes in local trees. Note that this loses lineage
-information such as the fact that \textsf{K} is connected to \textsf{A} via \textsf{E}.
+information such as the fact that \textsf{k} is connected to \textsf{a} via \textsf{e}.
 }
@@
 local trees on the right (Fig.~\ref{fig-simplification}A). Consider the path above node
-\textsf{A} in the penultimate tree, highlighted in yellow and covering genomic positions
+\textsf{a} in the penultimate tree, highlighted in yellow and covering genomic positions
 60 to 70. On this path, before $A$ meets its common ancestor with the other samples at
-node \textsf{Q}, it passes through  \textsf{E},  \textsf{J},  \textsf{M},  \textsf{N} and 
-\textsf{P}, which in this region of the genome are all unary nodes. This reflects the
+node \textsf{q}, it passes through  \textsf{e},  \textsf{j},  \textsf{m},  \textsf{n} and 
+\textsf{p}, which in this region of the genome are all unary nodes. This reflects the
 passage of this span of genome through the ARG towards the root (see the
@@
 The nodes in a tree usually mark the points where samples find a
- \emph{common} ancestor, such as node \textsf{Q} in the highlighted
+ \emph{common} ancestor, such as node \textsf{q} in the highlighted
 tree. For many purposes, we would therefore consider a path directly
-joining sample \textsf{A} to node \textsf{Q} to be of equivalent status to the path
+joining sample \textsf{a} to node \textsf{q} to be of equivalent status to the path
 containing the intermediate unary nodes. Simplification is the process of removing
@@
 in which the two parent nodes of a recombination immediately
-join again into a common ancestor (e.g. \textsf{J}, \textsf{L}, \textsf{M}
-and  \textsf{N} in Fig.~\ref{fig-simplification}A).
+join again into a common ancestor (e.g. \textsf{j}, \textsf{l}, \textsf{m}
+and  \textsf{n} in Fig.~\ref{fig-simplification}A).
 Unless we are specifically
@@
 of a single node that has one inbound and one outbound edge, such as
-nodes \textsf{F} and \textsf{H}.
+nodes \textsf{f} and \textsf{h}.
 Fig.~\ref{fig-simplification}B shows the result of this type of
@@
 of the local trees. In particular, a node can represent a recombinant
-with multiple parents in the graph but only a single child (e.g. node \textsf{N}
+with multiple parents in the graph but only a single child (e.g. node \textsf{n}
 in Fig.~\ref{fig-simplification}B), or can represent a common ancestor with
 multiple children in the graph but in which no local coalescence takes place
-(node \textsf{R} in Fig.~\ref{fig-simplification}B).
+(node \textsf{r} in Fig.~\ref{fig-simplification}B).
 % point out that this is why Hudson refers to CA nodes, not coalescence nodes
@@
 edge annotations. Removal of recombinant nodes can produce graph nodes with
-more than 2 parents (e.g. node \textsf{E}); likewise, removal of
+more than 2 parents (e.g. node \textsf{e}); likewise, removal of
 common ancestor but non-coalescent nodes can produce graph nodes with
-more than 2 children (e.g. node \textsf{S}). These represent a ``stacking up'' of multiple
+more than 2 children (e.g. node \textsf{s}). These represent a ``stacking up'' of multiple
 often indistinguishable events into a single node, something which is impossible
@@
 correspond to a coalescence in at least one other
-local tree. For example, node  \textsf{K} is unary in the second tree
+local tree. For example, node  \textsf{k} is unary in the second tree
 of Fig.~\ref{fig-simplification}C, but is either binary
@@
 It is common for this operation to remove nodes above the top MRCA node
-in the tree, resulting in many more root nodes (here, \textsf{Q} and \textsf{K}
+in the tree, resulting in many more root nodes (here, \textsf{q} and \textsf{k}
 become roots). [And continue]
@@
 The effects of recombination as we move from left-to-right along
-the genome in an unsimplified GARG are easy to interpret.
-Consider the leftmost tree in Fig~\ref{fig-simplification}.
-In the first tree the parent of B is F, and in the second
-tree its parent is E. Since both parents are present as
+the genome in an unsimplified gARG are easy to interpret.
+Consider the highlighed tree in Fig~\ref{fig-simplification}A.
+In it the parent of \textsf{e} is \textsf{j}, whereas in the next tree
+its parent is \textsf{k}. Since both parents are present as
 unary nodes in the tree, we simply change one edge in
 order to effect the transition.
+% Need revising to fit new example
 A more complex transition occurs when we have a regraft operation
@@
 \begin{tikzpicture}
+	\node [anchor=north west] at (-3.5,9.3) {\LARGE \textbf{A}};
 	\draw (-0.4, -0.2) -- (0.4, -0.2) -- (0.4, -0.3) -- (-0.4, -0.3) -- (-0.4, -0.2);
@@
 	\draw (-1, 1.7) -- (-1, 4.5) -- (0,4.5) -- (0,3.1);
-	\node [scale=0.2,label=above left:{\small A}] at (-0.5,4.5) {A};
+	\node [scale=0.2,label=above left:{\small \textsf{a}}] at (-0.5,4.5) {a};
 	\draw (-0.9, 4.3) -- (-0.6, 4.3) -- (-0.6, 4.2) -- (-0.9, 4.2) -- (-0.9, 4.3);
@@
 	\draw (4,2.4) -- (4,5.4) -- (2,5.4) -- (2,3.1);
-	\node [scale=0.2,label=above left:{\small B}] at (3,5.4) {B};
+	\node [scale=0.2,label=above left:{\small \textsf{b}}] at (3,5.4) {b};
 	\draw (2.6, 5.2) -- (2.9, 5.2) -- (2.9, 5.1) -- (2.6, 5.1) -- (2.6, 5.2);
@@
 	\draw (0.5, 6) -- (0.5, 7.2) -- (3,7.2) -- (3,5.4);
-	\node [scale=0.2,label=above:{\small C}] at (1.75,7.2) {C};
+	\node [scale=0.2,label=above:{\small \textsf{c}}] at (1.75,7.2) {c};
 	\draw (1.35, 6.95) -- (2.15, 6.95);
@@
 \begin{tikzpicture}
+	\node [anchor=north west] at (-3.5,9.3) {\LARGE \textbf{B}};
 	\draw (-0.4, -0.2) -- (0.4, -0.2) -- (0.4, -0.3) -- (-0.4, -0.3) -- (-0.4, -0.2);
@@
 propagated until time $T$. The MRCA on the genetic interval $[v, w)$ is reached
-at event B, while that on $[0, v)$ is reached at event C.
+at event \textsf{b}, while that on $[0, v)$ is reached at event \textsf{c}.
 The non-ancestral segment $[v, w)$ above

### 2022-04-07 illustrations.py

@@ -81,7 +81,7 @@ def ancestry_resolution():
     fig, (ax1, ax2) = plt.subplots(1, 2, figsize=(20, 12), sharey=True)
     fig.tight_layout()
 
-    ax1.set_title("(a)", fontsize=32, family="serif")
+    ax1.set_title("A", fontsize=32, family="serif", loc="left")
     ts = argutils.viz.label_nodes(argutils.wh99_example())
     pos, G = argutils.viz.draw(
         ts,
@@ -115,7 +115,7 @@ def ancestry_resolution():
     )
     add_edge_labels(ax1, ts, G, pos)
 
-    ax2.set_title("(b)", fontsize=32, family="serif")
+    ax2.set_title("B", fontsize=32, family="serif", loc="left")
     ts2 = argutils.simplify_keeping_all_nodes(ts)
     pos, G = argutils.viz.draw(
         ts2,
@@ -135,7 +135,7 @@ def ancestry_resolution():
         "genome_empty_hamburger": "illustrations/assets/genome_empty_hamburger.png",
         "genome_full": "illustrations/assets/genome_full.png",
     }
-    for letter in list(string.ascii_uppercase[3 : G.number_of_nodes()]):
+    for letter in list(string.ascii_lowercase[3 : G.number_of_nodes()]):
         icons["genome_" + letter] = "illustrations/assets/genome_" + letter + ".png"
     # Load images
     images = {k: PIL.Image.open(fname) for k, fname in icons.items()}
@@ -155,7 +155,7 @@ def ancestry_resolution():
         a = plt.axes([xa - icon_center1, ya - icon_center2, icon_size1, icon_size2-0.0015])
         a.imshow(G.nodes[n]["image"], aspect='auto', interpolation='spline36')
         a.set_title(
-            string.ascii_uppercase[n],
+            string.ascii_lowercase[n],
             y=0,
             verticalalignment="bottom",
             loc="center",
@@ -173,7 +173,7 @@ def ancestry_resolution():
     for n in [0, 1, 2]:
         G.nodes[n]["image"] = images["genome_full"]
     for n in range(3, G.number_of_nodes()):
-        G.nodes[n]["image"] = images["genome_" + string.ascii_uppercase[n]]
+        G.nodes[n]["image"] = images["genome_" + string.ascii_lowercase[n]]
     G.nodes[9]["image"] = images["genome_empty_hamburger"]
     for n in G.nodes:
         xf, yf = tr_figure(pos[n])
[…]

### 2022-04-07 paper.tex
Update Fig 1 caption to refer to A, B, C

@@
 \caption{\label{fig-arg-data-structure}
-Node vs edge breakpoint annotations.
-The classical approach described by Griffiths
-equates graph nodes with events in the stochastic process,
-and associates recombination breakpoint information with \emph{nodes} (left).
-This encoding
-is limited in the patterns of inheritance that can be described
-and requires that edges be ordered to avoid ambiguity.
-The alternative is to associate breakpoint annotations
-with \emph{edges} (right).
+A classical ARG (subfigure B), contrasting node with edge breakpoint annotations.
+(A) recombination breakpoint information encoded on \emph{nodes}.
+This encoding is limited in the patterns of inheritance that can be
+described and requires that edges be ordered to avoid ambiguity.
+(C) breakpoint annotations associated with \emph{edges}.
 Associating the interval of genome inherited by the child node

### 2022-04-08 paper.tex
More minor tweaks
Fixes #209 and a few readability fixes

@@
 \begin{tabular}{c|c}
-% \multicolumn{2}{c}{Breakpoint}\\
-node & breakpoint\\
+% \multicolumn{2}{c}{Breakpoints}\\
+Nodes & Breakpoints\\
 \hline
@@
 \begin{tabular}{ll}
-Edges\\
+Edges & Intervals\\
 \hline
@@
 any pattern genetic inheritance to be described and does
-not require edges to be ordered.
+not require edges to be ordered. In this example, each edge is associated
+with a single interval, but more generally, multiple non-overlapping intervals
+can be associated with a single edge.
 }
@@
 % What do we mean by an eARG, formally?
-Formally, we can define an eARG as a tuple $(e, \sigma)$, where $e$
+Formally, we can define the classical eARG as a tuple $(e, \sigma)$, where $e$
 is an ordered list of edges defining a directed acyclic graph and
@@
 a recombination node is of a different ``type'' to other nodes,
-with breakpoints only associated with this specific node type).
+with breakpoints only associated with that node type).
 Each edge $e_j = (c, p)$ describes an edge between
@@
 with high probability, the number of concurrent lineages in the ARG remains
-much smaller than $N$ at each time. While these approximations are
+much smaller than $N$ at each time. These approximations are
 reasonable to make for small datasets, such as
@@
 ADH gene dataset (comprising 11 genomes over 2.7 kb of sequence with 43 SNPs,
-as used in the ``\nameref{ARG_inference_methods}'' section),
+see the ``\nameref{ARG_inference_methods}'' section). However,
 they are highly questionable in modern datasets. % absurd really

### 2022-04-12 illustrations.py
Fix illustration code

@@ -136,7 +136,7 @@ def ancestry_resolution():
         "genome_full": "illustrations/assets/genome_full.png",
     }
     for letter in list(string.ascii_lowercase[3 : G.number_of_nodes()]):
-        icons["genome_" + letter] = "illustrations/assets/genome_" + letter + ".png"
+        icons["genome_" + letter] = f"illustrations/assets/genome_{letter.upper()}.png"
     # Load images
     images = {k: PIL.Image.open(fname) for k, fname in icons.items()}
 
@@ -602,7 +602,7 @@ def inference():
         # params["tweak_x"] = None
         # params["reverse_x_axis"] = None
         # argutils.viz.draw(ts, ax, edge_colors=edge_colors_by_node, pos=pos, **params)
-        
+
         ax.set_title(name + f"\n{ts.num_trees} trees")
 
         # Add breakpoints and stacked edges below tree sequences, w same colormap as above
@@ -664,7 +664,7 @@ def inference():
     top_svg += "\n".join(svg)
     with open(f"illustrations/inference.svg", "wt") as f:
         f.write(top_svg)
-     
+
 
 cli.add_command(arg_in_pedigree)
 cli.add_command(ancestry_resolution)

### 2022-04-12 paper.tex
Hide ugly hyperref links

@@
 \usetikzlibrary{calc,positioning}
-\usepackage{hyperref}
+\usepackage[hidelinks]{hyperref}
 
@@
 ``Ancestral Recombination Graph'' or ARG, was introduced later by
-Griffiths~\citep{griffiths1991two,ethier1990two,griffiths1996ancestral, griffiths1997ancestral},
+Griffiths~\citep{griffiths1991two,ethier1990two,griffiths1996ancestral,griffiths1997ancestral},
 who provided a closely related and complementary treatment.
@@
 ARG literature is focused on mathematical results about the underlying stochastic
-process. Griffiths and colleagues used the word
-``ARG'' to refer to both a specific branching-coalescing stochastic process and
-the graph structure derived from this process, but for clarity and following recent usage,
-we will use ARG to refer to a graph structure, and refer to the
-process as the Griffiths process or the coalescent with recombination
-\citep{minichiello2006mapping}.
+process. Griffiths and colleagues used the term
+``ARG'' to refer to both a particular branching-coalescing stochastic process and
+the graph structure produced by this process.
+% but for clarity and following recent usage,
+% we will use ARG to refer to a graph structure, and refer to the
+% process as the Griffiths process or the coalescent with recombination
+% \citep{minichiello2006mapping}.
 
@@
 Early work on ARG inference focused on the problem of
-of inferring parameters of the
+inferring parameters of the
 stochastic process, where the ancestry is regarded as a
@@
 fearnhead2001estimating}.
-% FIXME: need some proper statistical language here to describe why these
-% methods didn't work so well.
 These methods met with limited success
-due to the overwhelmingly large state space of ARGs, which also has
-no simple geometry or neighbourhood structure that an inference or
-sampling method could exploit.
+because the state space of ARGs is overwhelmingly large and
+lacks a simple geometry or neighbourhood structure for inference or
+sampling methods to  exploit.
 The focus subsequently shifted to the more
@@
 since parsimonious ARGs have no immediate connection to
-the stochastic process. Indeed, the ARG instances produced by
-the coalescent with recombination are usually highly unparsimonious,
-since they contain many invisible recombinations
-% Do we need the following caveat?
-(although those produced under the SMC may be parsimonious, depending
-on the pattern of mutations that is present).
+the stochastic process.
 
@@
 the tree at genome coordinate $x$,
-we traverse the graph upwards from the leaves. At a particular
+we traverse the graph rootwards from the leaves. At a particular
 node $u$, if it has one parent we are at a common ancestor
-node and we simply follow that parent. If we are at a
+node and we follow that parent. If we are at a
 recombination node, $u$ has two parents: if
@@
 to the second parent.
-This ordering requirement, which is straightforward
+This ordering requirement, while straightforward
 to describe, has some practical drawbacks. For example,
@@
 to the right of the breakpoint rather
-than the left (and so we cannot emit edges as they are generated).
+than the left, and so we cannot emit edges as they are generated.
 Such issues can be worked around, of course,
@@
 model, assuming that the sample size $n$ is much less than the
-population size $N$, and that the genome is short enough that,
-with high probability, the number of concurrent lineages in the ARG remains
-much smaller than $N$ at each time. These approximations are
+population size $N$, and that the genome is short enough
+that the number of extant lineages remains
+much smaller than $N$ at all times. These approximations are
 reasonable to make for small datasets, such as
@@
 where this alternative edge annotation formulation is
-shown. By associating the interval $(0, x)$ with edge $(3,4)$
-and the interval $(x, L)$ with $(3, 5)$ we make the encoding
+shown. By associating the interval $(0, x)$ with edge $(d,e)$
+and the interval $(x, L)$ with $(d, f)$ we make the encoding
 unambiguous: we can list the edges in any order, and it
@@
 between between positions $x$ and $y$ was inherited by
-node $3$ from $5$, we would state this by associating
-the intervals $\{(0, x], (y, L]\}$ with $(3,4)$ and
-$\{(x, y]\}$ with $(3,5)$.
+node $d$ from $e$, we would state this by associating
+the intervals $\{(0, x], (y, L]\}$ with $(d,e)$ and
+$\{(x, y]\}$ with $(d,f)$.
 This change is straightforward, involves no loss of information
@@
 embedded in an inbred diploid pedigree, in which individuals occur at
-discrete generations and carry two genomes. Each individual (labelled
-$I_1$ to $I_8$) therefore has a paternal genome (on the right) and a maternal one
-(on the left), although the pedigree is only shown for concreteness, and
-there is no inherent limitation on mating system, ploidy, or age structure.
+discrete generations and carry two genomes.
+% Each individual (labelled
+% $I_1$ to $I_8$) therefore has a maternal and paternal genome,
+% although the pedigree is only shown for concreteness, and
+% there is no inherent limitation on mating system, ploidy, or age structure.
 A genome is inherited from one of the individual's parents,
@@
 whereas \textsf{b} was inherited directly from $I_4$'s genome \textsf{g} without
-recombination. Fig.~\ref{fig-arg-in-pedigree}B shows the bare gARG, illustrating
-that the patterns of
-transmission between genomes can be followed without necessarily knowing
-how they are combined within individuals. This ``genome-perspective''
-simply depicts the ancestry of the sample genomes \textsf{a}--\textsf{d}. The graph
+recombination. Fig.~\ref{fig-arg-in-pedigree}B shows the resulting gARG,
+depicting the ancestry of the sample genomes \textsf{a}--\textsf{d}. The graph
 topology along with the genome coordinate annotations on the
@@
 edges. Starting with the sample nodes $S$ we traverse
-upwards through the graph from child to parent. At each node $u$, we find an
-edge $(c, p, I) \in E$ such that $u = c$ and $x \in I$ (if it
-exists) and move to the next node $p$.
+rootwards through the graph from child to parent. At each node $u$, we find an
+edge $(c, p, I) \in E$ such that $u = c$ and $x \in I$
+and move to the next node $p$.
 
@@
 
-This method for representing recombinations may seem wasteful,
-since more than one node is required for each recombination,
-and genomic interval information must be stored on every edge.
-However, asymptotically the amount of information stored is the
-same, and there is much simplicity to be gained by having
-a single type of node and by treating all edges in the same way.
-More importantly --- as we see in the next section --- we gain the
-ability to simplify the ARG by removing nodes without affecting the
-local tree topologies, and also the ability to record only the
-regions of genome relevant to the final, sample genomes. This last
-feature turns out to be the key to efficient processing of the ARG.
-
-[Possibly more stuff here: clarify why we assign coordinates to edges
-rather than nodes. Basically lay out all the stuff that we can
-talk about \emph{without} going in to the details of
-ancestry resolution.]
+% This method for representing recombinations may seem wasteful,
+% since more than one node is required for each recombination,
+% and genomic interval information must be stored on every edge.
+% However, asymptotically the amount of information stored is the
+% same, and there is much simplicity to be gained by having
+% a single type of node and by treating all edges in the same way.
+% More importantly --- as we see in the next section --- we gain the
+% ability to simplify the ARG by removing nodes without affecting the
+% local tree topologies, and also the ability to record only the
+% regions of genome relevant to the final, sample genomes. This last
+% feature turns out to be the key to efficient processing of the ARG.
+
+% [Possibly more stuff here: clarify why we assign coordinates to edges
+% rather than nodes. Basically lay out all the stuff that we can
+% talk about \emph{without} going in to the details of
+% ancestry resolution.]
 

### 2022-05-18 paper.tex
Compressed inference methods section

@@
 
-\subsection*{Parsimony and heuristics}
-The problem of reconstructing ARGs for samples of recombining sequences has been of significant
- interest essentially since the ARG was first defined. Early methods focussed on finding the most
- \emph{parsimonious} ARGs, i.e.\ those minimising the number of posited recombination events in
- the history of the sample \citep{hein1990reconstructing}. Two main approaches to explicitly reconstructing
- parsimonious ARGs then emerged: building the ARG either \emph{backwards-in-time} or \emph{along-the-genome}.
- The former approach, introduced by \citet{lyngso2005minimum} and implemented in the program
- \texttt{Beagle}, starts with the input data matrix and reduces it through row and column operations
- corresponding to coalescence, mutation and recombination events, until the empty matrix is reached;
- the corresponding sequence of events can then be used to construct an ARG from the bottom up. The
- latter approach, introduced by \citet{song2003parsimonious} and first implemented in the program
- \texttt{RecMinPath} \citep{song2005constructing}, starts with an initial local tree at some starting
- position, and refines the tree moving left and right along the genome, changing the topology of the
- tree when necessary through \emph{subtree prune and regraft} (SPR) operations (corresponding to
- recombination events).
+The problem of reconstructing ARGs for samples of recombining sequences has been of 
+interest since the ARG was first defined. Early methods focused on finding \emph{parsimonious} ARGs,
+i.e.\ those with a minimal number of recombination events \citep{hein1990reconstructing}.
+Two main approaches emerged: \emph{backwards-in-time} \citep{lyngso2005minimum} and
+\emph{along-the-genome} \citep{song2003parsimonious, song2005constructing}.
+The former starts with a data matrix and reduces it to an empty matrix through row and column operations
+corresponding to coalescence, mutation, and recombination events, which construct
+an ARG from the bottom up. The latter begins from an initial local tree at a single focal site. Moving the 
+focal site along the genome changes the local tree via a \emph{subtree prune and regraft} (SPR) operation
+whenever a recombination is traversed. Methods based the backwards-in-time approach are described by
+\citet{song2005efficient, wu2008association, thao2019hybrid, ignatieva2021kwarg},
+while \citet{hein1993heuristic, wu2011new, mirzaei2017rent} describe along-the-genome methods.
+\citet{rasmussen2022espalier} focuses on parsimonious fusion of local trees into an ARG, while
+\citet{camara2016inference} is based on topological data analysis.
 
 Reconstructing a most parsimonious ARG for a given data set is NP-hard \citep{wang2001perfect},
-and in order to allow for larger input datasets, subsequent parsimony-based methods have resorted
-to various relaxations and heuristics. Those utilising variants of the backwards-in-time approach
-include the programs \texttt{SHRUB} \citep{song2005efficient}, \texttt{GAMARG} \citep{thao2019hybrid},
-and \texttt{KwARG} \citep{ignatieva2021kwarg}, and the method described by \citet{wu2008association}.
-Tools constructing genealogies along-the-genome include \texttt{RecPars} \citep{hein1993heuristic},
-\texttt{RENT} \citep{wu2011new} and its sequel \texttt{RENT+} \citep{mirzaei2017rent}. Other approaches
-include \texttt{TARGet} \citep{camara2016inference}, which implements a method based on topological
-data analysis. The scalability of these methods varies substantially based on the heuristics in use
-and the underlying data structure, with the most efficient tools limited to analysing hundreds of
-sequences.
-
-A number of heuristic methods have also been developed which do not (explicitly) focus on reconstructing
-the most parsimonious ARGs, but rather aim for computational efficiency while seeking to achieve good
-accuracy of the inferred genealogies. Such methods include \texttt{Margarita} \citep{minichiello2006mapping}
-and the approach of \citet{parida2008estimating}; the more recently developed \texttt{Relate}
-\citep{speidel2019method} and \texttt{tsinfer} \citep{kelleher2019inferring} use a combination of both
-backwards-in-time and along-the-genome approaches, and can be practically used on megabase scale data
-and tens to hundreds of thousands of samples under human-like parameters.
-Another class of methods has recently emerged that use a \emph{threading} approach, constructing ARGs
-by adding sequences to the graph one-by-one. Originally introduced by \citet{rasmussen2014genome} as a
-way of generating MCMC proposals for ARGs, a similar idea was leveraged in \texttt{ARGneedle}
-\citep{zhang2021biobank} to reconstruct ARGs for biobank-scale data.
+so parsimony-based methods resort to heuristics and are limited to analysing hundreds of sequences.
+Hence, a number of methods aim to balance computational efficiency with reconstruction of ``reasonable",
+rather than parsimonious ARGs
+\citep{minichiello2006mapping, parida2008estimating, kelleher2019inferring,  speidel2019method, zhang2021biobank}.
+The latter three methods, as well as the parsimony-based method of \citet{rasmussen2022espalier},
+can be used on megabase scale data and hundreds of thousands of samples under human-like parameters.
+
 % AI: just to note, these papers do not all agree in their definition of an ARG.
@@
 
-Espalier~\citep{rasmussen2022espalier} uses Maximum Agreement Forests to
-minimise the differences between local trees estimated using other methods.
-
-\subsection*{Model-based methods}
-An alternative approach is to treat the ancestry as a latent parameter to be averaged out
+An alternative approach is to treat the ARG as a latent parameter to be averaged out
 by Monte Carlo methods, based either on importance sampling
@@
 or MCMC \citep{kuhner2000maximum, nielsen2000estimation, wang2008bayesian, fallon2013acg}.
-These methods invariably operated on a representation of the ``little ARG'', typically
-generated until the grand MRCA. They are extremely computationally expensive,
-and applicable to at most hundreds of samples consisting of tens of kilobases with
-human-like parameters.
-
-Due to the computational expense of sampling ``little ARGs'', state-of-the-art
-Monte Carlo methods typically rely on cheaper, approximate models.
-The most prominent examples are
-\texttt{ARGWeaver}~\citep{rasmussen2014genome,hubisz2020mapping,hubisz2020inference}
-and \texttt{Arbores} \citep{heine2018bridging}. The former also utilises a time
-discretisation approximation, but scales to dozens of mammal-like genomes.
-The \texttt{ClonalOrigin} method \citep{didelot2010inference,
-medina2020speeding} uses the \texttt{ClonalFrame}
-model and data structure, though it is not as scalable as \texttt{ARGWeaver}.
-
-A very recent MCMC method called \texttt{ARGInfer} has resulted in the first
-improvement in the scalability of Monte Carlo methods for the exact ARG model
-by making use of the tree sequence representation of marginal trees, suitably enriched
-to facilitate the computations needed for MCMC sampling
-\citep{mahmoudi2022bayesian}.
-While not competitive with the scalability of \texttt{ARGWeaver}, \texttt{ARGInfer} extends
-the feasible range of sequence lengths by at least an order of magnitude when
-compared with methods which sample realisations of the exact ``little ARG''.
+These methods operate on representations of the ``little ARG'', and are extremely computationally
+expensive, being applicable to at most hundreds of samples consisting of tens of kilobases with
+human-like parameters. State-of-the-art methods rely on cheaper, approximate models
+\citep{didelot2010inference, heine2018bridging, hubisz2020mapping,hubisz2020inference, medina2020speeding}. 
+The most scalable method, \texttt{ARGWeaver} \citep{rasmussen2014genome}, can be applied to dozens of mammal-like
+genomes \citep{hubisz2020inference}.
+
+Recently, \citet{mahmoudi2022bayesian} have improved the scalability of exact MCMC for ARGs by 
+leveraging the tree sequence representation of marginal trees, suitably enriched to facilitate the
+evaluation of sampling probabilities. While not competitive with \texttt{ARGWeaver}, their method extends
+the feasible range of sequence lengths by at least an order of magnitude.
 

### 2022-06-06 paper.tex
Fixing requirements.txt for pip3

@@
 trees and calculate sampling probabilities. However, this process traces out
-a larger graph than the Hudson algorithm (see the ``\nameref{Ancestry_resolution}'' section),
+a larger graph (the ``big'' ARG) than the Hudson algorithm (the ``little'' ARG; see the ``\nameref{Ancestry_resolution}'' section),
 and is consequently considerably less efficient \citep{wiuf1999ancestry}.
@@
 lineages~\citep{etheridge2009coalescent} or by leveraging perfect simulation
-techniques when recurrent mutation is present\citep{fearnhead2001perfect}.
+techniques when recurrent mutation is present \citep{fearnhead2001perfect}.
 Extensions of the ASG have been developed to frequency-dependent
@@
 that labels some of its nodes'' (although more narrow and context-dependent definitions can also
-be found in the literature). Explicit rooted phylogenetic networks are rooted DAGs,
+be found in the literature). Explicit rooted phylogenetic networks are rooted Directed Acyclic Graphs (DAGs),
 with leaves labelled by the taxa (species, groups, individuals or sequences), and nodes classed
@@
 (often assumed to be $10^4$ in humans).
-% TODO make this better. Ask Gregor for some refs?
-% Basically, Ag datasets make the assumption of isolated
-% rare recombinations absurd.
-Agricultural datasets such as the 1000 bull genomes~\citep{hayes20191000}
-combine genome sequence data with detailed pedigrees, with
-sample data from multiple generations.
+The 1000 bull genomes project~\citep{hayes20191000}
+comprises close to 7000 genomes, which are part of multi-generation pedigrees
+with millions of animals and extensive SNP array genotype data. For example,
+the US dairy cattle database alone comprises more than 6 million animals with SNP array genotypes\footnote{\url{https://queries.uscdcb.com/Genotype/counts.html}},
+while the effective population size in modern dairy cattle breeds is less than 100 and decreasing~\citep{MacLeod2013,Makanjuola2020}.
+This low effective population size is caused by intense selection and
+staggering use of a handful of bulls across the world. The current record
+holder is bull Toystory that has produced 2.4 million doses of semen
+for artificial insemination across 13 years\footnote{\url{https://www.dairyherd.com/news/world-renowned-toystory-bull-has-died}}.
+An extreme sign of these breeding practices is that there are only two ancestral
+Y-chromosome linages present in today's US Holstein dairy breed~\citep{Yue2015}.
+Similar datasets are becoming available also in other livestock species.
+Notably, \citet{RosFreixedes2022} have combined genome
+and SNP array data to infer recombination of genomes for 440,610 pigs within
+7 multi-generation pedigrees~\citep{whalen2018,Johnsson2021,Ros-Freixedes2020}.
+Effective population size in pigs is also less than 100 with each female and male
+routinely producing respectively more than 10 and 5000 progeny each \citep{Hall2016,Porcnic2016}.
 Complete chromosome-level assemblies are now possible
@@
 and ultimately worldwide~\citep{lewin2022earth}.
-While the coalescent model can be suprisingly robust to such
-basic violations to the underlying assumptions~\citep{
+While the coalescent model can be suprisingly robust
+basic violations of the underlying assumptions~\citep{
 wakeley2012gene,bhaskar2014distortion,nelson2020accounting},
@@
 In a Wright-Fisher model, coalescence and recombination can
-occur at the same time in the same genomes, and we can be
-certain that this is also true in the incredibly rich
-datasets that are available today.
+occur at the same time in the same genomes, which is increasingly
+captured in the incredibly rich datasets that are available today.
 
@@
 so we should be hoping to infer it some day''. Presumably people
-already are in bacterial contexts?]
+already are in bacterial contexts? Gregor: I recall this paper from
+Arabidopsis https://www.pnas.org/doi/10.1073/pnas.1211827110]
 To allow us to encode data from these important processes we
@@
 and \textsf{c} are independent recombinants of the paternal genomes \textsf{e}
+% Gregor: there is some overlap between a and c still (the middle region inherited
+% from f) so, a and c are not fully independent!
 and \textsf{f}: in these cases the edge annotations have been shown, indicating the
@@
 they merge into a common ancestor, the merged region is darker; this shows that all
-genomic regions are fully coalesced in genome \textsf{n} in the oldest generation.
+genomic regions fully coalesce in genome \textsf{n} in the oldest generation.
 (B) The equivalent genome ARG, where there is a single type of node (representing
@@
 local relationships along the genome, as shown in the local
-or ``marginal'' trees.
+(``marginal'') trees.
 }
@@
 embedded in an inbred diploid pedigree, in which individuals occur at
-discrete generations and carry two genomes.
-% Each individual (labelled
-% $I_1$ to $I_8$) therefore has a maternal and paternal genome,
-% although the pedigree is only shown for concreteness, and
-% there is no inherent limitation on mating system, ploidy, or age structure.
+discrete generations and carry two genomes. Each individual (labelled
+$I_1$ to $I_8$) therefore has a paternal genome (on the left) and a maternal one
+(on the right), although the pedigree is only shown for concreteness, and
+there is no inherent limitation on mating system, ploidy, or age structure.
 A genome is inherited from one of the individual's parents,
@@
 inherited from parents $I_3$ and $I_4$. Genome \textsf{a} is the product of
-recombining $I_3$'s two genomes \textsf{e} and \textsf{e} at position 2,
+recombining $I_3$'s two genomes \textsf{e} and \textsf{f} at position 2,
 whereas \textsf{b} was inherited directly from $I_4$'s genome \textsf{g} without
@@
 a representational perspective. This ability can be useful in practice and could be
-% Do we need to spell this out?
-required if, for example, we have a deeply sequenced pedigree of haplodiploids.
+% Do we need to spell this out? Gregor: I think so - see below
+required if, for example, we have a deeply sequenced pedigree as in \cite{RosFreixedes2022}.
 % e.g. Honeybees? Be great to cite an example!
@@
 % of internal samples.
+% Gregor: There are no good deeply sequenced pedigrees in honeybees, but sex
+% chromosome behaves like haplodiploid system, also any deep pedigree works here, so I cited our pig
+% work. Also, I think its important to not think of all genomes as current samples - we have many
+% generations of animals with genome or SNP array data and we know pedigree so we can't assume they
+% are all current samples - I believe you made this clear distinction also in the tsdate paper with
+% ancient human samples
 
@@
 recombination node in an eARG. For example, node \textsf{c} in
-Fig~\ref{fig-arg-in-pedigree} inherits genome positions 0 to 7 from \textsf{e}
-and positions 7 to 10 from \textsf{f}, indicating that the breakpoint is at position 7.
+Fig~\ref{fig-arg-in-pedigree} inherits genome positions 0 to 7 from \textsf{f}
+and positions 7 to 10 from \textsf{e}, indicating that the breakpoint is at position 7.
 
@@
 the samples along the genome: do to this, we must construct the
-local (or ``marginal'') trees.
+local (``marginal'') trees.
 
@@
 edge annotation schemes.
-(A) Edge annotations that reflect immediate inheritance, i.e. transmission of all
+(A) Edge annotations that reflect immediate inheritance, i.e. transmission of
 genetic material between parent and child. All edges span the full genome from position 0 to 7,
@@
 for even moderate levels of recombination is far too large to be of practical
-use. The number of events in the big ARG, all the way back to the GMRCA
+use. The number of events in the big ARG all the way back to the GMRCA
 is $O(e^\rho)$~\citep{griffiths1997ancestral}, whereas the number
@@
 be known about more ancient nodes, determined by the graph topology
-and the flow ancestral material through it.
+and the flow of ancestral material through it.
 \emph{Simplification}~\citep{kelleher2018efficient} is the process
@@
 various types of redundancy, while maintaining the crucial properties
-that the meaningful topology in the local trees is not affected,
+that the ``meaningful'' topology in the local trees is not affected,
 and the relationship between nodes in adjacent trees is maintained.
@@
 \textsf{a} in the penultimate tree, highlighted in yellow and covering genomic positions
-60 to 70. On this path, before $A$ meets its common ancestor with the other samples at
+60 to 70. On this path, before \textsf{a} meets its common ancestor with the other samples at
 node \textsf{q}, it passes through  \textsf{e},  \textsf{j},  \textsf{m},  \textsf{n} and
@@
 approximations~\citep{mcvean2005approximating,marjoram2006fast}
-which are central to many modern methods.
+that are central to many modern methods.
 
@@
 rely on evaluation of a likelihood and a prior. Two contemporary examples of
-likelihood-based methods are  \cite{mahmoudi2022bayesian, guo2022recombination}.
-
-In essentially all neutral cases in practical use, the likelihood $p(D | G)$ is the law of
+likelihood-based methods are \cite{mahmoudi2022bayesian} and \cite{guo2022recombination}.
+
+In essentially all neutral cases in practical use, the likelihood $p(D | G)$ is based on
 a Poisson process on the ancestral material on the edges of the ARG, which implies
@@
 contained in the same ancestral lineage. Instead, a richer data structure akin to the
-EARG or GARG is needed. Moreover, because EARG does not explicitly store the flow
+eARG or gARG is needed. Moreover, because eARG does not explicitly store the flow
 of ancestral material along its edges---a necessary piece of information for
 computing the number of effective recombination links at each time---evaluating
-$p(G)$ based on an EARG essentially requires turning it into a GARG at runtime.
+$p(G)$ based on an eARG essentially requires turning it into a gARG at runtime.
 
 The number of extant lineages at a given time is not explicitly represented even
-in a GARG, and needs to be calculated from the node and edge tables.
+in a gARG, and needs to be calculated from the node and edge tables.
 For example, any edges with the same child and parent nodes are contained within
@@
 originating from a recombinant child are also contained within a single lineage
-until the time of their parent nodes, at which point they split. Nodes S, T, and U
+until the time of their parent nodes, at which point they split. Nodes \textsf{s}, \textsf{t}, and \textsf{u}
 in panel (a) of Figure \ref{fig-ancestry-resolution} provide a concrete example
-of the latter: the edges from S to T and S to U are distinct, but correspond to only
+of the latter: the edges from \textsf{s} to \textsf{t} and \textsf{s} to \textsf{u} are distinct, but correspond to only
 a single lineage in the time interval they span. We emphasize that these
-conventions grouping edges into lineages are not part of the GARG data structure.
+conventions grouping edges into lineages are not part of the gARG data structure.
 While they are convenient for analysis using the coalescent with recombination,
@@
 to explicitly \emph{infer} ARGs from sequencing data.
+% Gregor: I expected clear indication which method/tool works with eARGs and gARGs, but I can see this clearly
 
@@
 
-\item We make the point about expressivity of the EARG pretty
+\item We make the point about expressivity of the eARG pretty
 hard, but we don't mention anything about indels and structural
@@
 along the genome or the resolved ancestry segments.
-Knowing the resolved segments in an ancestry resolved GARG allows
+Knowing the resolved segments in an ancestry resolved gARG allows
 us to perform many calculations without explicitly generating
@@
 on edges can be performed very quickly on an ancestry resolved
-GARG, since we only need to examine $O(|E|)$ edges rather than
+gARG, since we only need to examine $O(|E|)$ edges rather than
 $O(nt)$ tree branches (assuming $n$ samples and $t$ marginal trees)
@@
 showed how all trees can be sequentially generated in
-constant time per tree transition in a fully simplified GARG.
+constant time per tree transition in a fully simplified gARG.
 Furthermore, we can easily reason about how tree topologies

### 2022-06-16 paper.tex
Minor edits improving ag dataset part

@@
 comprises close to 7000 genomes, which are part of multi-generation pedigrees
-with millions of animals and extensive SNP array genotype data. For example,
+with millions of animals and extensive SNP array genotype and phenotype data \citep[e.g.][]{Cesarani2022}. For example,
 the US dairy cattle database alone comprises more than 6 million animals with SNP array genotypes\footnote{\url{https://queries.uscdcb.com/Genotype/counts.html}},
@@
 staggering use of a handful of bulls across the world. The current record
-holder is bull Toystory that has produced 2.4 million doses of semen
+holder is the bull Toystory that has produced 2.4 million doses of semen
 for artificial insemination across 13 years\footnote{\url{https://www.dairyherd.com/news/world-renowned-toystory-bull-has-died}}.
@@
 7 multi-generation pedigrees~\citep{whalen2018,Johnsson2021,Ros-Freixedes2020}.
-Effective population size in pigs is also less than 100 with each female and male
-routinely producing respectively more than 10 and 5000 progeny each \citep{Hall2016,Porcnic2016}.
+Effective population size in modern pig breeds is also less than 100 due to
+intense selection and directed reproduction \citep{Hall2016,Porcnic2016}.
 Complete chromosome-level assemblies are now possible
@@
 and ultimately worldwide~\citep{lewin2022earth}.
-While the coalescent model can be suprisingly robust
+While the coalescent model can be suprisingly robust to
 basic violations of the underlying assumptions~\citep{

### 2022-07-11 paper.tex
Figure 6 narrative

@@
 
+\subsection*{ARG reconstruction}
+
 The problem of reconstructing ARGs for samples of recombining sequences has been of
@@
 focal site along the genome changes the local tree via a \emph{subtree prune and regraft} (SPR) operation
-whenever a recombination is traversed. Methods based the backwards-in-time approach are described by
+whenever a recombination is encountered. Methods based the backwards-in-time approach are described by
 \citet{song2005efficient, wu2008association, thao2019hybrid, ignatieva2021kwarg},
@@
 Reconstructing a most parsimonious ARG for a given data set is NP-hard \citep{wang2001perfect},
-so parsimony-based methods resort to heuristics and are limited to analysing hundreds of sequences.
+so parsimony-based methods resort to heuristics and are limited to analysing at most hundreds of sequences.
 Hence, a number of methods aim to balance computational efficiency with reconstruction of ``reasonable",
@@
 the feasible range of sequence lengths by at least an order of magnitude.
+
+\subsection*{Differences in reconstructed ARGs}
+
+There are a vast number of ARGs compatible with a given dataset, so ARG reconstruction tools generally produce different outputs for the same input.
+Figure \ref{fig-inferred-args} shows examples of the ARG topologies produced by four of the tools mentioned above, applied to the classical dataset of 11 \emph{Drosophila melanogaster} sequences of \citet{kreitman1983nucleotide}.
+
+\texttt{KwARG} \citep{ignatieva2021kwarg} produces a parsimonious ARG with seven recombination breakpoints, which has been shown to be the true minimum required to explain the dataset when disallowing recurrent mutation \citep{song2003parsimonious}.
+An ARG drawn from the posterior sample produced by \texttt{ARGweaver} has 17 breakpoints.
+Both tools explicitly record the sequence undergoing each recombination event, and the precise location of the breakpoint along the genome.
+\texttt{Relate} \citep{speidel2019method} and \texttt{tsinfer} \citep{kelleher2019inferring} instead construct a sequence of correlated local trees, but without specifying the precise positions of recombination events (i.e.\ reconstructing the local trees without explicitly specifying the SPR operations required to transition between them).
+The number of inferred trees also depends on the parameter settings; all four tools can allow for the presence of sequencing errors and recurrent mutations, reducing the number of inferred local trees (in this case, based on the input mutation and recombination rate \texttt{Relate} hence posits only two recombination breakpoints).
+
+There is some consistency among the four inferred ARGs (such as the samples FR-F, WA-F and Af-F being grouped into the same clade), but also considerable differences in the order and number of events.
+Note that all four ARGs can be represented in the succinct tree sequence format, with the ARGs produced by \texttt{KwARG} and \texttt{ARGweaver} containing more nodes to record the recombination events (corresponding to Figure 4 stage `B'), and the ARGs output by \texttt{tsinfer} and \texttt{Relate} being, in essence, simplified (Figure 4 stage `D').
 

### 2022-07-13 paper.tex
Incorporating suggestions

@@
 
-\texttt{KwARG} \citep{ignatieva2021kwarg} produces a parsimonious ARG with seven recombination breakpoints, which has been shown to be the true minimum required to explain the dataset when disallowing recurrent mutation \citep{song2003parsimonious}.
+There is some consistency among the four inferred ARGs (such as the samples FR-F, WA-F and Af-F being grouped into the same clade), but also considerable differences in the order and number of events.
+\texttt{KwARG} \citep{ignatieva2021kwarg} produces a parsimonious ARG with seven recombination breakpoints, which has been shown to be the true minimum required to explain the dataset when disallowing recurrent mutation \citep{song2003parsimonious}; this is a lower bound that, in general, significantly underestimates the actual number of recombinations that might have occurred.
 An ARG drawn from the posterior sample produced by \texttt{ARGweaver} has 17 breakpoints.
 Both tools explicitly record the sequence undergoing each recombination event, and the precise location of the breakpoint along the genome.
-\texttt{Relate} \citep{speidel2019method} and \texttt{tsinfer} \citep{kelleher2019inferring} instead construct a sequence of correlated local trees, but without specifying the precise positions of recombination events (i.e.\ reconstructing the local trees without explicitly specifying the SPR operations required to transition between them).
+\texttt{Relate} \citep{speidel2019method} and \texttt{tsinfer} \citep{kelleher2019inferring} do not place recombination events onto the ARG topology, therefore the transitions between local trees (and corresponding SPR operations) are not explicitly inferred.
+\texttt{Relate} first infers local tree topologies and then identifies edges that persist from one tree to the next as a separate step, while \texttt{tsinfer} explicitly constructs nodes that are shared across local trees.
 The number of inferred trees also depends on the parameter settings; all four tools can allow for the presence of sequencing errors and recurrent mutations, reducing the number of inferred local trees (in this case, based on the input mutation and recombination rate \texttt{Relate} hence posits only two recombination breakpoints).
 
-There is some consistency among the four inferred ARGs (such as the samples FR-F, WA-F and Af-F being grouped into the same clade), but also considerable differences in the order and number of events.
-Note that all four ARGs can be represented in the succinct tree sequence format, with the ARGs produced by \texttt{KwARG} and \texttt{ARGweaver} containing more nodes to record the recombination events (corresponding to Figure 4 stage `B'), and the ARGs output by \texttt{tsinfer} and \texttt{Relate} being, in essence, simplified (Figure 4 stage `D').
+Note that all four methods produce output which can be represented in the succinct tree sequence format, and which is consistent with our definition of an ARG as a data structure.
+The ARGs produced by \texttt{KwARG} and \texttt{ARGweaver} contain more nodes to record the recombination events (corresponding to Figure 4 stage `B'), while the ARGs output by \texttt{tsinfer} and \texttt{Relate} being, in essence, simplified (roughly corresponding to Figure 4 stages `C' and 'D', respectively).
 

### 2022-07-13 paper.tex
Tweaks

@@
 Both tools explicitly record the sequence undergoing each recombination event, and the precise location of the breakpoint along the genome.
-\texttt{Relate} \citep{speidel2019method} and \texttt{tsinfer} \citep{kelleher2019inferring} do not place recombination events onto the ARG topology, therefore the transitions between local trees (and corresponding SPR operations) are not explicitly inferred.
+\texttt{Relate} \citep{speidel2019method} and \texttt{tsinfer} \citep{kelleher2019inferring} do not place recombination events onto the ARG topology, therefore the transitions between local trees are not explicitly inferred in terms of SPR operations.
 \texttt{Relate} first infers local tree topologies and then identifies edges that persist from one tree to the next as a separate step, while \texttt{tsinfer} explicitly constructs nodes that are shared across local trees.

### 2022-07-14 paper.tex
Move phylo network discussion to end
Fixes #219

@@
 reproduction~\citep{gonzalezcasanova2018duality, koskela2019robust}.
-
-\textcolor{red}{TODO: reduce the text below down to one paragraph, which
-defines the relationship between what we define as an ARG and
-what people mean by phylogenetic networks from the literature.
-We probably don't need to get into quite so much detail on
-the different types that people have studies. Essentially,
-we're ending the section with a brief review which situates
-our definition in the literature. I think pulling apart the
-idea of a ``reticulation'' vs what we call a recombinant
-is essential here.}
-
-The term \emph{phylogenetic network} has been broadly accepted to describe the most general
-type of genealogical graph, defined by \citet{huson2010phylogenetic} as ``any graph used to
-represent evolutionary relationships (either abstractly or explicitly) between a set of taxa
-that labels some of its nodes'' (although more narrow and context-dependent definitions can also
-be found in the literature). Explicit rooted phylogenetic networks are rooted Directed Acyclic Graphs (DAGs),
-with leaves labelled by the taxa (species, groups, individuals or sequences), and nodes classed
-as either reticulation nodes (with in-degree $\geq 2$) or tree nodes.
-
-Some sub-classes of explicit rooted phylogenetic networks are defined based on the evolutionary event that the
-reticulation nodes capture \citep{huson2010phylogenetic}. In \emph{reassortment} networks, used to
-model the evolution of viruses, reticulation nodes correspond to exchange of
-(non-recombining) genomic segments between viral particles inside a co-infected host.
-In \emph{hybridisation} networks, reticulation nodes correspond to genetic material from two
-different species combining to create a hybrid. In \emph{recombination} networks, the reticulation nodes
-correspond to recombination events (with leaves labelled by sequences and edges labelled by mutations).
-The ARG is a subtype of a recombination network under this definition (restricting to one crossover
-recombination per reticulation node).
-% AI: Not sure about this - do ARGs also encompass multiple crossover recombinations and gene
-% conversion (seems implied above)?
-
-Sub-classes of explicit rooted phylogenetic networks can also be defined by restrictions on their topology.
-In \emph{binary networks}, all internal nodes must have either in-degree two and out-degree one or in-degree
-one and out-degree two (with the root having out-degree two and the leaves in-degree 1) \citep{steel2016phylogeny}.
-In \emph{tree-child networks}, every internal node must have at least one child that is a tree node;
-\emph{tree-sibling networks} are constrained by every reticulation node having at least one sibling that is a
-tree node \citep{cardona2008extended}. Further, sub-types can be defined by restrictions on
-the topology of subgraphs that contain the reticulation loops (cycles containing a reticulation node).
-A reticulation loop that does not share any node with another reticulation loop is termed a gall
-\citep[][p.\ 237]{gusfield2014recombinatorics}; a reticulation network containing only galls is
-called a \emph{galled tree}. Reticulation loops that are joined together in a reticulation network
-by at least one shared node form a blob; a \emph{level-$k$ network} is one where the maximal number
-of reticulation loops inside a blob is $k$ \citep{choy2005computing}. Galled trees are level-1 networks,
-and the tractability afforded by their relatively simple structure results in a number of attractive
-combinatorial properties \citep{wang2001perfect, gusfield2004optimal}. We consider the general case of
-an ARG, without applying any of these restrictions.
-
-% AI: have not mentioned any implicit or un-rooted networks (split, median, haplotype, reticulograms, etc).
-% Also: are we not defining ARGs as ``binary networks''? Might be good to specify this here.
 
@@
 % not doing anything controversial.
-The deep connection between ARGs and pedigrees is of course
+The deep connection between ARGs and pedigrees is
 well known
@@
 
-\section*{Evaluation of sampling probabilities}\label{probabilities}
-
 \begin{figure}[ht]
@@
 
-The conditional sampling probability $p(G | D)$ of an ARG $G$ given an observed
-realization of genetic diversity $D$ at its sampled leaves is a central concept in
-model fitting and selection. Under the Bayesian paradigm, it defines the posterior
-distribution $p(G | D) \propto p(D | G) p(G)$, where $p(D | G)$ is the likelihood
-of the data $D$ given $G$, and $p(G)$ is a prior distribution for the underlying
-ARG $G$. Under the frequentist paradigm, $p(G)$ can be interpreted as a regularizer,
-and $p(G | D)$ as a regularized likelihood. In both cases, $p(G)$ is typically specified
-as the sampling distribution of a stochastic process, such as the coalescent with
-recombination. Essentially all of the model-based inference methods surveyed below
-rely on evaluation of a likelihood and a prior. Two contemporary examples of
-likelihood-based methods are \cite{mahmoudi2022bayesian} and \cite{guo2022recombination}.
-
-In essentially all neutral cases in practical use, the likelihood $p(D | G)$ is based on
-a Poisson process on the ancestral material on the edges of the ARG, which implies
-that mutation processes at two sites are conditionally independent given the
-underlying ancestral trees at those sites. A prototypical example is
-\citet[Eq.\ (2)]{mahmoudi2022bayesian}. Hence, $p(D | G)$ can be evaluated
-regardless of the precise data structure used to store $G$, provided that it encodes
-at least all marginal ancestral trees along the sampled sequences.  Choosing the
-right data structure can still have a dramatic effect on the efficiency with which
-$p(D | G)$ can be evaluated \citep{mahmoudi2022bayesian}.
-
-In contrast, evaluating $p(G)$ under the coalescent with recombination
-(or related models) requires knowledge of the number of extant lineages and
-the number of links available for recombination (i.e.\ ones at which a recombination
-would split ancestral material) at each time along the ARG
-\citep[Eq.\ (3)]{mahmoudi2022bayesian}. These cannot be uniquely determined
-from just the marginal trees at each site because branches at multiple trees can be
-contained in the same ancestral lineage. Instead, a richer data structure akin to the
-eARG or gARG is needed. Moreover, because eARG does not explicitly store the flow
-of ancestral material along its edges---a necessary piece of information for
-computing the number of effective recombination links at each time---evaluating
-$p(G)$ based on an eARG essentially requires turning it into a gARG at runtime.
-
-The number of extant lineages at a given time is not explicitly represented even
-in a gARG, and needs to be calculated from the node and edge tables.
-For example, any edges with the same child and parent nodes are contained within
-the same lineage. Under the coalescent with recombination, the two edges
-originating from a recombinant child are also contained within a single lineage
-until the time of their parent nodes, at which point they split. Nodes \textsf{s}, \textsf{t}, and \textsf{u}
-in panel (a) of Figure \ref{fig-ancestry-resolution} provide a concrete example
-of the latter: the edges from \textsf{s} to \textsf{t} and \textsf{s} to \textsf{u} are distinct, but correspond to only
-a single lineage in the time interval they span. We emphasize that these
-conventions grouping edges into lineages are not part of the gARG data structure.
-While they are convenient for analysis using the coalescent with recombination,
-other models, such as discrete time Wright--Fisher models with finite populations,
-may benefit from different ways to associate edges with lineages.
-
 \section*{ARG inference methods}\label{ARG_inference_methods}
@@
 \citep{griffiths1996ancestral, fearnhead2001estimating, jenkins2011inference}
-or MCMC \citep{kuhner2000maximum, nielsen2000estimation, wang2008bayesian, fallon2013acg}.
+or MCMC \citep{kuhner2000maximum, nielsen2000estimation, wang2008bayesian, fallon2013acg, mahmoudi2022bayesian}.
 These methods operate on representations of the ``little ARG'', and are extremely computationally
-expensive, being applicable to at most hundreds of samples consisting of tens of kilobases with
+expensive, being applicable to at most hundreds of samples consisting of tens or hundreds of kilobases with
 human-like parameters. State-of-the-art methods rely on cheaper, approximate models
@@
 genomes \citep{hubisz2020inference}.
-
-Recently, \citet{mahmoudi2022bayesian} have improved the scalability of exact MCMC for ARGs by
-leveraging the tree sequence representation of marginal trees, suitably enriched to facilitate the
-evaluation of sampling probabilities. While not competitive with \texttt{ARGWeaver}, their method extends
-the feasible range of sequence lengths by at least an order of magnitude.
+A central quantity in all of these sampling methods is the conditional sampling probability 
+$p(G | D) \propto p(D | G) p(G)$ of an ARG $G$ given an observed realization
+of genetic diversity $D$ at its sampled leaves, where $p(D | G)$ is the likelihood
+of the data $D$ given $G$, and $p(G)$ is a Bayesian prior distribution or a 
+frequentist regularizer for the ARG $G$.
+
+The likelihood $p(D | G)$ is effectively always the distribution of
+a Poisson process on the ancestral material along the edges of $G$ \citep[Eq.\ (2)]{mahmoudi2022bayesian}.
+Hence it is typically straightforward to evaluate in principle, but choosing the
+right data structure can still have a dramatic effect on the efficiency of evaluation
+\citep{mahmoudi2022bayesian}. Especially in the Bayesian case, $p(G)$ is typically specified
+as the sampling distribution of a stochastic process, such as the coalescent with
+recombination. Two contemporary examples are
+\cite{mahmoudi2022bayesian, guo2022recombination}.
+Evaluating the distribution of the coalescent with recombination
+(or related models) requires knowledge of the number of extant lineages and
+number of links available for recombination (i.e.\ ones at which a recombination
+would split ancestral material) at each time \citep[Eq.\ (3)]{mahmoudi2022bayesian}.
+These require a data structure from which shared edges on different trees are easy to identify,
+and which encodes recombination events and times explicitly, e.g.\ using nodes.
 
@@
 
+\textcolor{red}{Discussion below on phylogenetic networks: we need to
+reduce the text below down to one paragraph, which
+defines the relationship between what we define as an ARG and
+what people mean by phylogenetic networks from the literature.
+We probably don't need to get into quite so much detail on
+the different types that people have studies. Essentially,
+we're ending the section with a brief review which situates
+our definition in the literature. I think pulling apart the
+idea of a ``reticulation'' vs what we call a recombinant
+is essential here.}
+
+The term \emph{phylogenetic network} has been broadly accepted to describe the most general
+type of genealogical graph, defined by \citet{huson2010phylogenetic} as ``any graph used to
+represent evolutionary relationships (either abstractly or explicitly) between a set of taxa
+that labels some of its nodes'' (although more narrow and context-dependent definitions can also
+be found in the literature). Explicit rooted phylogenetic networks are rooted Directed Acyclic Graphs (DAGs),
+with leaves labelled by the taxa (species, groups, individuals or sequences), and nodes classed
+as either reticulation nodes (with in-degree $\geq 2$) or tree nodes.
+
+Some sub-classes of explicit rooted phylogenetic networks are defined based on the evolutionary event that the
+reticulation nodes capture \citep{huson2010phylogenetic}. In \emph{reassortment} networks, used to
+model the evolution of viruses, reticulation nodes correspond to exchange of
+(non-recombining) genomic segments between viral particles inside a co-infected host.
+In \emph{hybridisation} networks, reticulation nodes correspond to genetic material from two
+different species combining to create a hybrid. In \emph{recombination} networks, the reticulation nodes
+correspond to recombination events (with leaves labelled by sequences and edges labelled by mutations).
+The ARG is a subtype of a recombination network under this definition (restricting to one crossover
+recombination per reticulation node).
+% AI: Not sure about this - do ARGs also encompass multiple crossover recombinations and gene
+% conversion (seems implied above)?
+
+Sub-classes of explicit rooted phylogenetic networks can also be defined by restrictions on their topology.
+In \emph{binary networks}, all internal nodes must have either in-degree two and out-degree one or in-degree
+one and out-degree two (with the root having out-degree two and the leaves in-degree 1) \citep{steel2016phylogeny}.
+In \emph{tree-child networks}, every internal node must have at least one child that is a tree node;
+\emph{tree-sibling networks} are constrained by every reticulation node having at least one sibling that is a
+tree node \citep{cardona2008extended}. Further, sub-types can be defined by restrictions on
+the topology of subgraphs that contain the reticulation loops (cycles containing a reticulation node).
+A reticulation loop that does not share any node with another reticulation loop is termed a gall
+\citep[][p.\ 237]{gusfield2014recombinatorics}; a reticulation network containing only galls is
+called a \emph{galled tree}. Reticulation loops that are joined together in a reticulation network
+by at least one shared node form a blob; a \emph{level-$k$ network} is one where the maximal number
+of reticulation loops inside a blob is $k$ \citep{choy2005computing}. Galled trees are level-1 networks,
+and the tractability afforded by their relatively simple structure results in a number of attractive
+combinatorial properties \citep{wang2001perfect, gusfield2004optimal}. We consider the general case of
+an ARG, without applying any of these restrictions.
+
+% AI: have not mentioned any implicit or un-rooted networks (split, median, haplotype, reticulograms, etc).
+% Also: are we not defining ARGs as ``binary networks''? Might be good to specify this here.
+
 % TODO not clear this belongs in the discussion, but we'll definitely

### 2022-07-15 paper.tex

@@
 reproduction~\citep{gonzalezcasanova2018duality, koskela2019robust}.
+
+\textcolor{red}{TODO: reduce the text below down to one paragraph, which
+defines the relationship between what we define as an ARG and
+what people mean by phylogenetic networks from the literature.
+We probably don't need to get into quite so much detail on
+the different types that people have studies. Essentially,
+we're ending the section with a brief review which situates
+our definition in the literature. I think pulling apart the
+idea of a ``reticulation'' vs what we call a recombinant
+is essential here.}
+
+The term \emph{phylogenetic network} has been broadly accepted to describe the most general
+type of genealogical graph, defined by \citet{huson2010phylogenetic} as ``any graph used to
+represent evolutionary relationships (either abstractly or explicitly) between a set of taxa
+that labels some of its nodes'' (although more narrow and context-dependent definitions can also
+be found in the literature). Explicit rooted phylogenetic networks are rooted Directed Acyclic Graphs (DAGs),
+with leaves labelled by the taxa (species, groups, individuals or sequences), and nodes classed
+as either reticulation nodes (with in-degree $\geq 2$) or tree nodes.
+
+Some sub-classes of explicit rooted phylogenetic networks are defined based on the evolutionary event that the
+reticulation nodes capture \citep{huson2010phylogenetic}. In \emph{reassortment} networks, used to
+model the evolution of viruses, reticulation nodes correspond to exchange of
+(non-recombining) genomic segments between viral particles inside a co-infected host.
+In \emph{hybridisation} networks, reticulation nodes correspond to genetic material from two
+different species combining to create a hybrid. In \emph{recombination} networks, the reticulation nodes
+correspond to recombination events (with leaves labelled by sequences and edges labelled by mutations).
+The ARG is a subtype of a recombination network under this definition (restricting to one crossover
+recombination per reticulation node).
+% AI: Not sure about this - do ARGs also encompass multiple crossover recombinations and gene
+% conversion (seems implied above)?
+
+Sub-classes of explicit rooted phylogenetic networks can also be defined by restrictions on their topology.
+In \emph{binary networks}, all internal nodes must have either in-degree two and out-degree one or in-degree
+one and out-degree two (with the root having out-degree two and the leaves in-degree 1) \citep{steel2016phylogeny}.
+In \emph{tree-child networks}, every internal node must have at least one child that is a tree node;
+\emph{tree-sibling networks} are constrained by every reticulation node having at least one sibling that is a
+tree node \citep{cardona2008extended}. Further, sub-types can be defined by restrictions on
+the topology of subgraphs that contain the reticulation loops (cycles containing a reticulation node).
+A reticulation loop that does not share any node with another reticulation loop is termed a gall
+\citep[][p.\ 237]{gusfield2014recombinatorics}; a reticulation network containing only galls is
+called a \emph{galled tree}. Reticulation loops that are joined together in a reticulation network
+by at least one shared node form a blob; a \emph{level-$k$ network} is one where the maximal number
+of reticulation loops inside a blob is $k$ \citep{choy2005computing}. Galled trees are level-1 networks,
+and the tractability afforded by their relatively simple structure results in a number of attractive
+combinatorial properties \citep{wang2001perfect, gusfield2004optimal}. We consider the general case of
+an ARG, without applying any of these restrictions.
+
+% AI: have not mentioned any implicit or un-rooted networks (split, median, haplotype, reticulograms, etc).
+% Also: are we not defining ARGs as ``binary networks''? Might be good to specify this here.
 
@@
 % not doing anything controversial.
-The deep connection between ARGs and pedigrees is
+The deep connection between ARGs and pedigrees is of course
 well known
@@
 
-\textcolor{red}{Discussion below on phylogenetic networks: we need to
-reduce the text below down to one paragraph, which
-defines the relationship between what we define as an ARG and
-what people mean by phylogenetic networks from the literature.
-We probably don't need to get into quite so much detail on
-the different types that people have studies. Essentially,
-we're ending the section with a brief review which situates
-our definition in the literature. I think pulling apart the
-idea of a ``reticulation'' vs what we call a recombinant
-is essential here.}
-
-The term \emph{phylogenetic network} has been broadly accepted to describe the most general
-type of genealogical graph, defined by \citet{huson2010phylogenetic} as ``any graph used to
-represent evolutionary relationships (either abstractly or explicitly) between a set of taxa
-that labels some of its nodes'' (although more narrow and context-dependent definitions can also
-be found in the literature). Explicit rooted phylogenetic networks are rooted Directed Acyclic Graphs (DAGs),
-with leaves labelled by the taxa (species, groups, individuals or sequences), and nodes classed
-as either reticulation nodes (with in-degree $\geq 2$) or tree nodes.
-
-Some sub-classes of explicit rooted phylogenetic networks are defined based on the evolutionary event that the
-reticulation nodes capture \citep{huson2010phylogenetic}. In \emph{reassortment} networks, used to
-model the evolution of viruses, reticulation nodes correspond to exchange of
-(non-recombining) genomic segments between viral particles inside a co-infected host.
-In \emph{hybridisation} networks, reticulation nodes correspond to genetic material from two
-different species combining to create a hybrid. In \emph{recombination} networks, the reticulation nodes
-correspond to recombination events (with leaves labelled by sequences and edges labelled by mutations).
-The ARG is a subtype of a recombination network under this definition (restricting to one crossover
-recombination per reticulation node).
-% AI: Not sure about this - do ARGs also encompass multiple crossover recombinations and gene
-% conversion (seems implied above)?
-
-Sub-classes of explicit rooted phylogenetic networks can also be defined by restrictions on their topology.
-In \emph{binary networks}, all internal nodes must have either in-degree two and out-degree one or in-degree
-one and out-degree two (with the root having out-degree two and the leaves in-degree 1) \citep{steel2016phylogeny}.
-In \emph{tree-child networks}, every internal node must have at least one child that is a tree node;
-\emph{tree-sibling networks} are constrained by every reticulation node having at least one sibling that is a
-tree node \citep{cardona2008extended}. Further, sub-types can be defined by restrictions on
-the topology of subgraphs that contain the reticulation loops (cycles containing a reticulation node).
-A reticulation loop that does not share any node with another reticulation loop is termed a gall
-\citep[][p.\ 237]{gusfield2014recombinatorics}; a reticulation network containing only galls is
-called a \emph{galled tree}. Reticulation loops that are joined together in a reticulation network
-by at least one shared node form a blob; a \emph{level-$k$ network} is one where the maximal number
-of reticulation loops inside a blob is $k$ \citep{choy2005computing}. Galled trees are level-1 networks,
-and the tractability afforded by their relatively simple structure results in a number of attractive
-combinatorial properties \citep{wang2001perfect, gusfield2004optimal}. We consider the general case of
-an ARG, without applying any of these restrictions.
-
-% AI: have not mentioned any implicit or un-rooted networks (split, median, haplotype, reticulograms, etc).
-% Also: are we not defining ARGs as ``binary networks''? Might be good to specify this here.
-
 % TODO not clear this belongs in the discussion, but we'll definitely

### 2022-07-19 paper.tex
Remove stochastic processes section
Fixes #220

--- 2022-07-14 paper.tex
@@
 
-\section*{Ancestral graphs and stochastic processes}
-The coalescent~\citep{kingman1982coalescent,kingman1982genealogy,
-hudson1983testing, tajima1983evolutionary}
-provides the theoretical underpinning for much of contemporary population genetics.
-\citet{hudson1983properties,hudson1990gene} extended the coalescent to include
-recombination
-and the model has proved to be an especially powerful basis for
-simulations~\citep{hudson2002generating,baumdicker2021efficient}.
-\cite{kelleher2016efficient} provide a description of his algorithm along with
-an efficient implementation, the details of which turn out to be highly relevant to
-later sections of the current paper (see the ``\nameref{Ancestry_resolution}'' section).
-
-Although Hudson's algorithm was accompanied by a graph
-schematic~\citep[][Fig. 5]{hudson1983properties}, the term
-``Ancestral Recombination Graph'' or ARG, was introduced later by
-Griffiths~\citep{griffiths1991two,ethier1990two,griffiths1996ancestral,griffiths1997ancestral},
-who provided a closely related and complementary treatment.
-Where Hudson's algorithm focusses on efficient \emph{simulation}, the original
-ARG literature is focused on mathematical results about the underlying stochastic
-process. Griffiths and colleagues used the term
-``ARG'' to refer to both a particular branching-coalescing stochastic process and
-the graph structure produced by this process.
-% but for clarity and following recent usage,
+% \section*{Ancestral graphs and stochastic processes}
+% The coalescent~\citep{kingman1982coalescent,kingman1982genealogy,
+% hudson1983testing, tajima1983evolutionary}
+% provides the theoretical underpinning for much of contemporary population genetics.
+% \citet{hudson1983properties,hudson1990gene} extended the coalescent to include
+% recombination
+% and the model has proved to be an especially powerful basis for
+% simulations~\citep{hudson2002generating,baumdicker2021efficient}.
+% \cite{kelleher2016efficient} provide a description of his algorithm along with
+% an efficient implementation, the details of which turn out to be highly relevant to
+% later sections of the current paper (see the ``\nameref{Ancestry_resolution}'' section).
+
+% Although Hudson's algorithm was accompanied by a graph
+% schematic~\citep[][Fig. 5]{hudson1983properties}, % but for clarity and following recent usage,
 % we will use ARG to refer to a graph structure, and refer to the
@@
 % to go with this
-In the Griffiths formulation, each edge in the graph corresponds to an extant
-lineage and nodes are events in the process (the initial $n$ leaf nodes are
-``sampling'' events). Common ancestor events occur at rate $\binom{k}{2}$ when there
-are $k$ lineages present. We choose two lineages (edges) uniformly, and merge them
-into a common ancestor lineage. Recombination events happen at a rate of
-$k \rho$. Here, we choose a lineage (edge) uniformly, and a
-breakpoint $0 < x < m$ uniformly on its genome. We terminate the edge at a
-node, record the breakpoint and start two new edges from this node. The process
-then continues until there is only one lineage left (the Grand Most Recent
-Common Ancestor, GMRCA), which is guaranteed to
-happen in finite time because of the linear vs quadratic rates of branching
-and coalescing. Later, we will show how the graph structure and breakpoints
-associated with
-recombination nodes provides sufficient information to later recover the marginal
-trees and calculate sampling probabilities. However, this process traces out
-a larger graph (the ``big'' ARG) than the Hudson algorithm (the ``little'' ARG; see the ``\nameref{Ancestry_resolution}'' section),
-and is consequently considerably less efficient \citep{wiuf1999ancestry}.
-
-Similar graph encodings have also been introduced to model other mechanisms
-which change the local ancestral tree along the genome, such as gene
-conversion~\citep{wiuf2000coalescent} and
-horizontal gene transfer~\citep{baumdicker2014infinitely}.
-[Mention the
-SMC~\citep{mcvean2005approximating,marjoram2006fast}
-and ClonalFrame~\citep{didelot2007inference} approximations somewhere.]
-
-The idea of using branching-coalescing stochastic processes
-to model genetic ancestral trees has also found application in natural selection.
-The Ancestral Selection Graph (ASG)~\citep{krone1997ancestral,neuhauser1997genealogy}
-uses dynamics identical to the ``big'' ARG to simulate an ensemble of correlated
-potential ancestral trees. Weak genic selection is incorporated by sampling a
-true ancestry from the ensemble in a non-uniform way.
-There is no ASG analogue of the more computationally tractable ``little'' ARG,
-though some gains in tractability can be made by considering typed
-lineages~\citep{etheridge2009coalescent} or by leveraging perfect simulation
-techniques when recurrent mutation is present \citep{fearnhead2001perfect}.
-Extensions of the ASG have been developed to frequency-dependent
-selection~\citep{neuhauser1999ancestral, gonzalezcasanova2018duality},
-unlinked chromosomes~\citep{fearnhead2003ancestral}, recombining
-loci whereupon branching is due to both selection and
-recombination~\citep{donnelly1999genealogical}, and high fecundity
-reproduction~\citep{gonzalezcasanova2018duality, koskela2019robust}.
+% In the Griffiths formulation, each edge in the graph corresponds to an extant
+% lineage and nodes are events in the process (the initial $n$ leaf nodes are
+% ``sampling'' events). Common ancestor events occur at rate $\binom{k}{2}$ when there
+% are $k$ lineages present. We choose two lineages (edges) uniformly, and merge them
+% into a common ancestor lineage. Recombination events happen at a rate of
+% $k \rho$. Here, we choose a lineage (edge) uniformly, and a
+% breakpoint $0 < x < m$ uniformly on its genome. We terminate the edge at a
+% node, record the breakpoint and start two new edges from this node. The process
+% then continues until there is only one lineage left (the Grand Most Recent
+% Common Ancestor, GMRCA), which is guaranteed to
+% happen in finite time because of the linear vs quadratic rates of branching
+% and coalescing. Later, we will show how the graph structure and breakpoints
+% associated with
+% recombination nodes provides sufficient information to later recover the marginal
+% trees and calculate sampling probabilities. However, this process traces out
+% a larger graph (the ``big'' ARG) than the Hudson algorithm (the ``little'' ARG; see the ``\nameref{Ancestry_resolution}'' section),
+% and is consequently considerably less efficient \citep{wiuf1999ancestry}.
+
+% Similar graph encodings have also been introduced to model other mechanisms
+% which change the local ancestral tree along the genome, such as gene
+% conversion~\citep{wiuf2000coalescent} and
+% horizontal gene transfer~\citep{baumdicker2014infinitely}.
+% [Mention the
+% SMC~\citep{mcvean2005approximating,marjoram2006fast}
+% and ClonalFrame~\citep{didelot2007inference} approximations somewhere.]
 
 \section*{The Event ARG data structure}\label{eARG}
+Initial work on \emph{simulating} ancestral graphs with recombination was undertaken by
+\citet{hudson1983properties,hudson1990gene}, who extended the classical coalescent 
+framework~\citep{kingman1982coalescent,kingman1982genealogy, hudson1983testing, tajima1983evolutionary}
+to include recombination. Hudson's algorithm has led to powerful simulation
+tools~\citep{hudson2002generating,kelleher2016efficient, baumdicker2021efficient}, and is
+also relevant to efficient analysis, as we shall see later in the
+``\nameref{Ancestry_resolution}'' section.
+
 % When did we move from the ARG is a process to the ARG is a data structure?
@@
 
-% NB: should this para go into the previous section? It would be
-% clean to start the section with the subsequent para
-% starting "The Griffiths ARG encoding is an elegant and mathematically economical description of the CwR"
-Early work on ARG inference focused on the problem of
-inferring parameters of the
-stochastic process, where the ancestry is regarded as a
-latent parameter to be averaged out
-\citep[e.g.][]{griffiths1996ancestral,kuhner2000maximum, nielsen2000estimation,
-fearnhead2001estimating}.
-These methods met with limited success
-because the state space of ARGs is overwhelmingly large and
-lacks a simple geometry or neighbourhood structure for inference or
-sampling methods to  exploit.
-The focus subsequently shifted to the more
-tractable---but still NP-hard~\citep{wang2001perfect}---problem of computing
-the minimum number of recombinations required
-to explain an input dataset [CITATIONS], and inferring the corresponding
-ARG realisations~\citep{song2003parsimonious,song2005efficient,lyngso2005minimum}.
-This highlighted the distinction between the ARG as a stochastic process
-and the ARG as a \emph{data structure}~\citep{minichiello2006mapping},
-since parsimonious ARGs have no immediate connection to
-the stochastic process.
+The precise term ``Ancestral Recombination Graph'' or ARG, was subsequently introduced by
+Griffiths and colleagues~\citep{griffiths1991two,ethier1990two,griffiths1996ancestral,griffiths1997ancestral}.
+Rather than simulation, their work focused on mathematical results about the underlying branching-coalescing
+stochastic process. However, some confusion has arisen from their use of the word ``ARG'' to refer to
+two distinct concepts: the process itself, and the graph structure it generates.
+
+%% NB: I (yan) have commented out the following section as the "latent parameter" quote is
+%% used again in the "ARG reconstruction" section, and may be more detail than is required here
+%% This also loses the description of ARG space being large etc.
+%Early work on ARG inference focused on the problem of
+%inferring parameters of the
+%stochastic process, where the ancestry is regarded as a
+%latent parameter to be averaged out
+%\citep[e.g.][]{griffiths1996ancestral,kuhner2000maximum, nielsen2000estimation,
+%fearnhead2001estimating}.
+%These methods met with limited success
+%because the state space of ARGs is overwhelmingly large and
+%lacks a simple geometry or neighbourhood structure for inference or
+%sampling methods to  exploit.
+%The focus subsequently shifted to the more
+%tractable---but still NP-hard~\citep{wang2001perfect}---problem of computing
+%the minimum number of recombinations required
+%to explain an input dataset [CITATIONS], and inferring the corresponding
+%ARG realisations~\citep{song2003parsimonious,song2005efficient,lyngso2005minimum}.
+%This highlighted the distinction between the ARG as a stochastic process
+%and the ARG as a \emph{data structure}~\citep{minichiello2006mapping},
+%since parsimonious ARGs have no immediate connection to
+%the stochastic process.
+
+The second concept, that of the ARG being a \emph{data structure} was adopted by later authors,
+when focussing on the problem of inferring an ARG-like structure (or a minimum number of
+recombinations [CITATIONS]) without assuming a specific stochastic
+model, for example using parsimony~\citep{song2003parsimonious,song2005efficient,lyngso2005minimum}.
+Following~\citet{minichiello2006mapping}, we adopt this convention, using ARG
+to mean an ancestral graph structure, without reference to the process that generated it.
 
 % What do we mean by an EARG, informally?
-The Griffiths ARG encoding is an elegant and mathematically
-economical description of the coalescent with recombination, whose
-fundamental units are the  \emph{events} that correspond directly
-to those that occur in the associated stochastic process. We will refer
-to this ARG encoding as an ``Event ARG'', or eARG:
+\citet{griffiths1996ancestral} introduced an encoding for the ARG structure, which
+is elegant and mathematically economical. In this encoding, the
+fundamental units correspond directly to \emph{events}
+that occur in the underlying stochastic process they were modelling.
+We will refer to this ARG encoding as an ``event ARG'', or eARG:
 an example is given in Fig.~\ref{fig-arg-data-structure}B.
-Each node in the graph corresponds to an event: therefore the graph
+Since each node in the graph corresponds to an event, the graph
 contains both recombination nodes (node \textsf{d}) and common ancestor
@@
 
+\textcolor{red}{Discussion on ASG ported from earlier (removed) ``stochastic processes'' section}
+
+The idea of using branching-coalescing stochastic processes
+to model genetic ancestral trees has also found application in natural selection.
+The Ancestral Selection Graph (ASG)~\citep{krone1997ancestral,neuhauser1997genealogy}
+uses dynamics identical to the ``big'' ARG to simulate an ensemble of correlated
+potential ancestral trees. Weak genic selection is incorporated by sampling a
+true ancestry from the ensemble in a non-uniform way.
+There is no ASG analogue of the more computationally tractable ``little'' ARG,
+though some gains in tractability can be made by considering typed
+lineages~\citep{etheridge2009coalescent} or by leveraging perfect simulation
+techniques when recurrent mutation is present \citep{fearnhead2001perfect}.
+Extensions of the ASG have been developed to frequency-dependent
+selection~\citep{neuhauser1999ancestral, gonzalezcasanova2018duality},
+unlinked chromosomes~\citep{fearnhead2003ancestral}, recombining
+loci whereupon branching is due to both selection and
+recombination~\citep{donnelly1999genealogical}, and high fecundity
+reproduction~\citep{gonzalezcasanova2018duality, koskela2019robust}.
+
 % TODO not clear this belongs in the discussion, but we'll definitely

### 2022-07-20 paper.tex
Remove "of course"
It's redundant

@@
 % a sample node, so would be normally included anyway. Perhaps, therefore
-% we should talk about these "pass though" nodes in the context
+% we should talk about these ``pass though'' nodes in the context
 % of internal samples.
@@
 \section*{Ancestry resolution}\label{Ancestry_resolution}
-% This is a rough first pass. Will clarify terminology later.
 % Need to introduce Hudson's algorithm as concerned with simulating the process of
@@
 
-The definition of an ARG as a data structure in terms of
-individuals' monoploid genomes
-and their direct patterns of inheritance allows us to describe arbitrarily
-complex relatedness structures, free from the limitations imposed by
-coalescent approximations.
-An ARG is an intrinsically retrospective
-structure, defined with respect to the ancestry of
-a set of sampled genomes.
-The graph structure (the nodes and their connectivity through edges)
-captures a great deal of information about this ancestry,
-but on its own it cannot tell us about the precise relationships between
-the samples along the genome: do to this, we must construct the
-local (``marginal'') trees.
-
-We can define a sample-resolved gARG as a tuple $(S, E)$
-where $S \subset \mathbb{N}$ is a set of sample nodes and
-$E$ is a set of edges. Each element of $E$
-is a tuple $(c, p, I)$ such that $c, p \in \mathbb{N}$,
-$c$ is the child node, $p$ is the parent node and $I$ is the set of
-disjoint genomic intervals $(\ell, r]$
-over which genome $c$ inherits from $p$, and there is a path from
-$c$ to some non-empty subset of $S$ through the local trees.
-% Should probably define how we build the local trees somewhere.
+The edge annotations in Figs~\ref{fig-arg-data-structure} and \ref{fig-arg-in-pedigree}
+describe the regions of genome that have been passed from a parent to its immediate children,
+hence edges either span the entire genome, or all the region to the left (or right) of a breakpoint
+(e.g. Fig~\ref{fig-arg-data-structure}C). In this form of edge annotation, which we call
+``immediate edge annotation'',  % NB - perhaps we can think of a better term than this
+it is inefficient to extract the local trees from the ARG structure.
+
+There is, however, an alternative edge annotation scheme which allows efficient
+storage and extraction of the local trees. It is based on the fact that the ARG
+structure is inherently retrospective, defined as the ancestry of a set of \emph{sample}
+nodes (such as \textsf{a}, \textsf{b}, and \textsf{c} in the examples above). We define a
+``sample-resolved edge annotation'' to be a representation where edge spans
+represent only those parts of the genome that have been inherited by the samples.
+This also allows the removal of nodes which are not genetically ancestral to the samples.
 
@@
 edge annotation schemes.
-(A) Edge annotations that reflect immediate inheritance, i.e. transmission of
+(A) Immediate edge annotation, reflecting ``local'' transmission of all
 genetic material between parent and child. All edges span the full genome from position 0 to 7,
 apart from those linking a recombinant and its two parents (bold).
-(B) Sample-resolved edge annotations: the Hudson simplification algorithm has been used to
+(B) Sample-resolved edge annotation: the Hudson-like simplification algorithm has been used to
 shorten edge spans to reflect only the upwards transmission of genetic
@@
 coalescing regions of genome back to the grand MRCA genome \textsf{w}.
-In this case, using line widths to reflect the genomic span of each edge can
-help to visualise the relative importance of different transmission pathways.
+Line widths reflect the genomic span of each edge, which
+helps visualise the relative importance of different transmission pathways.
 }
@@
 
-% We should talk about ARG and HUD here. Fig~\ref{fig-ancestry-resolution}A shows
-% ARG (the Griffiths graph). However, we have a small issue that subfig B
-% doesn't strictly show HUD, because it doesn't terminate as soon as all samples have coalesced.
-% We could potentially colour the top (non-HUD) section of the plot in a different colour?
-
-Given a gARG $E$ and the set of samples $S$ we can \emph{resolve} its
-ancestry by noting that the resolved interval $Ir$ on an edge $(c,p,I)$
-is given by [Rough notes]
-
-\begin{itemize}
-\item The ``ancestry interval'' $A_u$ of a node $u$ is the union of
-all $R_{u, v}$ for all the children $v$ of $u$. (this isn't quite right)
-\item The resolved interval is then the intersection of $A_u$ and the
-original inheritance interval $I_v$ on the edge.
-\item It would be good of if differentiate between the inheritance intervals
- from the gARG and the ancestry intervals of the resolved gARG.
-\end{itemize}
-
-This process is illustrated in Fig~\ref{fig-ancestry-resolution}. It is intimately linked to
-Hudson's algorithm which operates by
-tracking segments of ancestral material carried by ancestors as
-we go backwards in time.
+Fig~\ref{fig-ancestry-resolution} illustrates the two different edge annotation schemes.
+To obtain the sample-resolved annotation in Fig~\ref{fig-ancestry-resolution}B we must
+traverse backwards in time from sample nodes \textsf{a}, \textsf{b}, and \textsf{c}, using the
+edge intervals to record regions of sample genome that have passed through each route.
+When edges join into a parent node, the parent is allocated the union of the intervals.
+When a child node is connected to two parents, one parent is allocated the edge
+intervals of the child, truncated to include only genomic material to the left of the breakpoint.
+Similarly the other parent is allocated the child's intervals truncated such that they contain
+only regions to the right of the breakpoint (a natural extension exists for multiple
+chromosomal breakpoints).
+
+For example, in Fig~\ref{fig-ancestry-resolution}A the edge between child node \textsf{g} and
+parent node \textsf{r} spans the entire genome from 0..7, reflecting the material
+transmitted immediately between parent and child: i.e. local to those two nodes in the tree.
+However, in Fig~\ref{fig-ancestry-resolution}B the edge has been reduced in extent so that
+it now spans 0..2, the region that eventually ends up in
+sample node \textsf{c}. This sample-resolved edge annotation scheme cannot be
+derived locally: it must be done by traversing backwards in time from the sample nodes
+up through the ARG.
+
+% TECHNICAL DESCRIPTION
+% Given a gARG $E$ and the set of samples $S$ we can \emph{resolve} its
+% ancestry by noting that the resolved interval $Ir$ on an edge $(c,p,I)$
+% is given by [Rough notes]
+% \begin{itemize}
+% \item The ``ancestry interval'' $A_u$ of a node $u$ is the union of
+% all $R_{u, v}$ for all the children $v$ of $u$. (this isn't quite right)
+% \item The resolved interval is then the intersection of $A_u$ and the
+% original inheritance interval $I_v$ on the edge.
+% \item It would be good of if differentiate between the inheritance intervals
+% from the gARG and the ancestry intervals of the resolved gARG.
+% \end{itemize}
+% This process is illustrated in Fig~\ref{fig-ancestry-resolution}. It is intimately linked to
+%Hudson's algorithm which operates by
+% tracking segments of ancestral material carried by ancestors as
+% we go backwards in time.
 
@@
 % covered by the lineage, respectively.
-Each lineage consists of a list of
-disjoint ancestry segments $(\ell, r, a)$, where
-$[\ell, r)$ is a half-closed genomic interval and $a$ is an integer
-tracking the number of samples to which the lineage is ancestral over that interval.
-(We also usually track the tree node associated with each segment, but
-that is not important for our purposes here so we omit it.)
-If we have $n$ samples and a genome of length $m$, the process begins with $n$ lineages
-of the form $\{(0, m, 1)\}$. The process then works backwards in time from
-the present day as a series of random common ancestor or recombination events.
-Recombination events occur at a rate determined by the amount of ancestral material and
-the way in which it is distributed along a chromosome.
-Let $L$ be the set of ancestors at a given time $t$. Recombination events
-happen at rate $\rho \nu / (m - 1)$ where
-\[
-\nu = \sum_{x \in L}\left( \max_{(\ell, r, a) \in x}r
-    - \min_{(\ell, r, a) \in x}\ell - 1 \right)
-\]
-is the number of available `links' that may be broken. Thus, the rate of
-recombination is determined by the left- and right-most extent of the
-ancestral material carried by each lineage. At a recombination
-event we choose one of these links uniformly and break it. Given a lineage
-$x = [(\ell_j, r_j, a_j)]$ and a breakpoint $k$, we have two lineages
-$x_1$ and $x_2$ such that FILL IN DETAILS
-
-When $k = |L|$ lineages are present, common ancestor events
-occur at rate $\binom{k}{2}$. In a common ancestor event, two lineages
-are chosen uniformly at random and their ancestry segments merged.
-If we have overlapping intervals of ancestry from the two lineages,
-say, $(\ell, r, a_1)$ and $(\ell, r, a_2)$, a
-\emph{coalescence} occurs and ancestor represented by the current event
-will be present as a node (at least) in the marginal trees covering
-the interval $[\ell, r)$. The result of this coalescence is a segment
-$(\ell, r, a_1 + a_2)$, and if $a_1 + a_2 < n$ it is included in the
-ancestry for the new lineage. Otherwise, if $a_1 + a_2 = n$ we know that
-we have found the most recent common ancestor of all samples in
-the interval $[\ell, r)$ and so we do not need to simulate its history any further.
-Nonoverlapping intervals of ancestry from the two lineages are included
-in the resulting lineage without changes. Eventually, as the process continues,
-we find resultant lineages in which all segments have fully coalescenced,
-and so the number of extant lineages gradually dwindles down to zero.
+
+When traversing upwards through the ARG like this, it is also helpful to record
+the number of sample nodes to which a given section of genome is ancestral.
+This is indicated in Fig~\ref{fig-ancestry-resolution}B by successively darker shading
+of the regions within each node. When a region has all samples as descendants, it is
+shaded black, and we can say that that region of the genome has fully coalesced.
+
+Formally, we can define a sample-resolved gARG as a tuple $(S, E)$
+where $S \subset \mathbb{N}$ is a set of sample nodes and
+$E$ is a set of edges. Each element of $E$
+is a tuple $(c, p, I)$ such that $c, p \in \mathbb{N}$,
+$c$ is the child node, $p$ is the parent node and $I$ is the set of
+disjoint genomic intervals $(\ell, r]$
+over which genome $c$ inherits from $p$, and there is a path from
+$c$ to some non-empty subset of $S$ through the local trees.
+
+The ARG traversal algorithm which implements this edge trimming and removal process
+is a form of ``simplification'' (cite), and can be thought of as running Hudson's algorithm
+on a fixed tree topology (see appendix XXX).
+
+% NB: below is a discussion of the Hudson algorithm which we might want to 
+% retain in an appendix
+% Each lineage consists of a list of
+% disjoint ancestry segments $(\ell, r, a)$, where
+% $[\ell, r)$ is a half-closed genomic interval and $a$ is an integer
+% tracking the number of samples to which the lineage is ancestral over that interval.
+% (We also usually track the tree node associated with each segment, but
+% that is not important for our purposes here so we omit it.)
+% If we have $n$ samples and a genome of length $m$, the process begins with $n$ lineages
+% of the form $\{(0, m, 1)\}$. The process then works backwards in time from
+% the present day as a series of random common ancestor or recombination events.
+% Recombination events occur at a rate determined by the amount of ancestral material and
+% the way in which it is distributed along a chromosome.
+% Let $L$ be the set of ancestors at a given time $t$. Recombination events
+% happen at rate $\rho \nu / (m - 1)$ where
+% \[
+% \nu = \sum_{x \in L}\left( \max_{(\ell, r, a) \in x}r
+%     - \min_{(\ell, r, a) \in x}\ell - 1 \right)
+% \]
+% is the number of available `links' that may be broken. Thus, the rate of
+% recombination is determined by the left- and right-most extent of the
+% ancestral material carried by each lineage. At a recombination
+% event we choose one of these links uniformly and break it. Given a lineage
+% $x = [(\ell_j, r_j, a_j)]$ and a breakpoint $k$, we have two lineages
+% $x_1$ and $x_2$ such that FILL IN DETAILS
+
+% When $k = |L|$ lineages are present, common ancestor events
+% occur at rate $\binom{k}{2}$. In a common ancestor event, two lineages
+% are chosen uniformly at random and their ancestry segments merged.
+% If we have overlapping intervals of ancestry from the two lineages,
+% say, $(\ell, r, a_1)$ and $(\ell, r, a_2)$, a
+% \emph{coalescence} occurs and ancestor represented by the current event
+% will be present as a node (at least) in the marginal trees covering
+% the interval $[\ell, r)$. The result of this coalescence is a segment
+% $(\ell, r, a_1 + a_2)$, and if $a_1 + a_2 < n$ it is included in the
+% ancestry for the new lineage. Otherwise, if $a_1 + a_2 = n$ we know that
+% we have found the most recent common ancestor of all samples in
+% the interval $[\ell, r)$ and so we do not need to simulate its history any further.
+% Nonoverlapping intervals of ancestry from the two lineages are included
+% in the resulting lineage without changes. Eventually, as the process continues,
+% we find resultant lineages in which all segments have fully coalescenced,
+% and so the number of extant lineages gradually dwindles down to zero.
+
+Several features of Fig 3B are worthy of note. First, since node \textsf{j} is not on
+the ancestral path from the samples it (and its associated edges) can be removed.
+For larger ARGs, the absence of these nodes makes sample-resolved edge annotation
+substantially more efficient than immediate edge annotation. Second, by the time we reach
+nodes \textsf{t} and \textsf{v}, all regions of the genome have fully coalesced, and so
+\textsf{w}, the ``grand MRCA'' can be omitted from the graph, at the cost of having
+multiple root nodes in different regions of the genome. Third, there exist nodes such as
+\textsf{k} with two children (i.e. ``common ancestor'' nodes in Hudson's terminology)
+but in which no ancestral coalescence occurs. Fourth, some breakpoint positions occur
+within regions of genome that are not ancestral (e.g. the breakpoint at position 3 in
+node \textsf{k}). in this case sample-resolved edge annotations lose the exact position of
+the breakpoint, although they retain the information that a breakpoint occurred between
+positions 2 and 4 (which is sufficient for calculating the likelihood, see appendix XXX).
+
+% Should probably define how we build the local trees somewhere.
+TODO: explain why this encoding allows efficient extraction of local trees, as implemented in the tskit library.
+
+% We should talk about ARG and HUD here. Fig~\ref{fig-ancestry-resolution}A shows
+% ARG (the Griffiths graph). However, we have a small issue that subfig B
+% doesn't strictly show HUD, because it doesn't terminate as soon as all samples have coalesced.
+% We could potentially colour the top (non-HUD) section of the plot in a different colour?
+
+TODO: give some feel for how much of a reduction the sample-resolved encoding can give in a large ARG, and
+maybe mention ``big'' vs ``little'' args as may be described in Jere's appendix
 
@@
 
-The state-space of the Griffiths process is much simpler than Hudson's algorithm,
-which greatly facilitates mathematical reasoning. This simplicity comes at a
-substantial cost, however, if we wish to use it as a practical means of
-simulating recombinant ancestry. The big ARG is vast, and any realisation
-for even moderate levels of recombination is far too large to be of practical
-use. The number of events in the big ARG all the way back to the GMRCA
-is $O(e^\rho)$~\citep{griffiths1997ancestral}, whereas the number
-of events required to simulate the little ARG is
-$O(\rho^2)$~\citep{hein2004gene,baumdicker2021efficient}.
-This disparity in the number of events in the two formulations is
-because the majority of the events that occur in the big ARG do
-not affect the genetic ancestry of the sample in any way. Recombination
-events that occur outside of ancestral material do not have any bearing
-on the ancestry of the sample, and so the structure is hugely redundant.
-As~\cite{wiuf1999recombination} note,
-``an `ancestral' sequence in the birth and death process
-need not have any genetic material in common with a
-sequence descended from it.''
+% The state-space of the Griffiths process is much simpler than Hudson's algorithm,
+% which greatly facilitates mathematical reasoning. This simplicity comes at a
+% substantial cost, however, if we wish to use it as a practical means of
+% simulating recombinant ancestry. The big ARG is vast, and any realisation
+% for even moderate levels of recombination is far too large to be of practical
+% use. The number of events in the big ARG all the way back to the GMRCA
+% is $O(e^\rho)$~\citep{griffiths1997ancestral}, whereas the number
+% of events required to simulate the little ARG is
+% $O(\rho^2)$~\citep{hein2004gene,baumdicker2021efficient}.
+% This disparity in the number of events in the two formulations is
+% because the majority of the events that occur in the big ARG do
+% not affect the genetic ancestry of the sample in any way. Recombination
+% events that occur outside of ancestral material do not have any bearing
+% on the ancestry of the sample, and so the structure is hugely redundant.
+% As~\cite{wiuf1999recombination} note,
+% ``an `ancestral' sequence in the birth and death process
+% need not have any genetic material in common with a
+% sequence descended from it.''
 % Full quote:
@@
 a ``subtree prune-and-regraft'' (SPR)
-% Hein et al call it the "subtree transfer" operation but it's
+% Hein et al call it the ``subtree transfer'' operation but it's
 % the same thing.

### 2022-07-27 paper.tex

@@
 This is because there is a fundamental ambiguity involved
-in associating recombination breakpoints with \emph{nodes}
+in associating recombination breakpoints with nodes
 rather than edges,
@@
 that the number of extant lineages remains
-much smaller than $N$ at all times. These approximations are
-reasonable to make for small datasets, such as
-the classic \citet{kreitman1983nucleotide} \emph{Drosophila melanogaster}
-ADH gene dataset (comprising 11 genomes over 2.7 kb of sequence with 43 SNPs,
-see the ``\nameref{ARG_inference_methods}'' section). However,
-they are highly questionable in modern datasets. % absurd really
-For example, many
-human datasets consist of hundreds of thousands of
+much smaller than $N$ at all times.
+While the coalescent model can be suprisingly robust to
+basic violations of the underlying assumptions~\citep{
+wakeley2012gene,bhaskar2014distortion,nelson2020accounting},
+it is not acceptable that our ability to \emph{represent}
+ancestral relationships should be limited by these assumptions.
+In a Wright-Fisher model, coalescence and recombination can
+occur at the same time in the same genomes, which is increasingly
+captured in the incredibly rich datasets that are available today.
+
+Firstly, complete chromosome-level assemblies are now possible
+in humans~\citep{miga2020telomere},
+and projects are under way to obtain high-quality assemblies
+for all eukaryotic species in Britain and Ireland~\citep{darwin2022sequence}
+and ultimately worldwide~\citep{lewin2022earth}. Thus, we can
+expect datasets to span the full continuum from non-recombining
+short genomes to very long genomes with high-levels of recombination
+[rephrase, and some cites?],
+all of which have high-quality whole genome data.
+
+Secondly, the scale of modern datasets presents a major challenge
+to the assumption that $n \ll N_e$.
+Several human datasets now consist of hundreds of thousands of
 genomes~\citep{bycroft2018genome,karczewski2020mutational,tanjo2021practical},
-and thus our sample size is often much \emph{larger} than $N_e$
+and thus our sample size can be substantially \emph{larger} than $N_e$
 (often assumed to be $10^4$ in humans).
@@
 comprises close to 7000 genomes, which are part of multi-generation pedigrees
-with millions of animals and extensive SNP array genotype and phenotype data \citep[e.g.][]{Cesarani2022}. For example,
-the US dairy cattle database alone comprises more than 6 million animals with SNP array genotypes\footnote{\url{https://queries.uscdcb.com/Genotype/counts.html}},
-while the effective population size in modern dairy cattle breeds is less than 100 and decreasing~\citep{MacLeod2013,Makanjuola2020}.
+with millions of animals and extensive SNP array genotype and phenotype
+data \citep[e.g.][]{Cesarani2022}.
+For example,
+the US dairy cattle database alone comprises more than 6 million
+animals with SNP array
+genotypes\footnote{\url{https://queries.uscdcb.com/Genotype/counts.html}},
+while the effective population size in modern dairy cattle breeds is
+less than 100 and decreasing~\citep{MacLeod2013,Makanjuola2020}.
 This low effective population size is caused by intense selection and
@@
 holder is the bull Toystory that has produced 2.4 million doses of semen
-for artificial insemination across 13 years\footnote{\url{https://www.dairyherd.com/news/world-renowned-toystory-bull-has-died}}.
+for artificial insemination across
+13 years\footnote{\url{https://www.dairyherd.com/news/world-renowned-toystory-bull-has-died}}.
 An extreme sign of these breeding practices is that there are only two ancestral
@@
 intense selection and directed reproduction \citep{Hall2016,Porcnic2016}.
-Complete chromosome-level assemblies are now possible
-in humans~\citep{miga2020telomere},
-and projects are under way to obtain high-quality assemblies
-for all eukaryotic species in Britain and Ireland~\citep{darwin2022sequence}
-and ultimately worldwide~\citep{lewin2022earth}.
-While the coalescent model can be suprisingly robust to
-basic violations of the underlying assumptions~\citep{
-wakeley2012gene,bhaskar2014distortion,nelson2020accounting},
-it is not acceptable that our ability to \emph{represent}
-ancestral relationships should be limited by these assumptions.
-In a Wright-Fisher model, coalescence and recombination can
-occur at the same time in the same genomes, which is increasingly
-captured in the incredibly rich datasets that are available today.
 
@@
 example, if a gene conversion event occurred where a segment
-between between positions $x$ and $y$ was inherited by
+between positions $x$ and $y$ was inherited by
 node $d$ from $e$, we would state this by associating
@@
 The deep connection between ARGs and pedigrees is
-well known
+well understood
 \citep[e.g.][]{wakeley2012genetics,gusfield2014recombinatorics,
@@
 As discussed in the previous section, the definition of an ARG
-as a \emph{data structure} in terms of these two events is problematic,
+as a data structure in terms of these two \emph{events} is problematic,
 fundamentally limiting the patterns of ancestry that can be described.
-Following \cite{mathieson2020ancestry} instead of focussing on events,
+Following \cite{mathieson2020ancestry}, instead of focussing on events,
 we define the ARG data structure in terms of \emph{genomes} and the
@@
 % it'll be clearer what we need to say in the caption vs the text.
-The Ancestral Recombination Graph defined in terms of genomes embedded
+An Ancestral Recombination Graph defined in terms of genomes embedded
 in a pedigree. (A) Diploid individuals, visualised in a highly inbred pedigree and
@@
 embedded in an inbred diploid pedigree, in which individuals occur at
-discrete generations and carry two genomes. Each individual (labelled
-$I_1$ to $I_8$) therefore has a paternal genome (on the left) and a maternal one
-(on the right), although the pedigree is only shown for concreteness, and
-there is no inherent limitation on mating system, ploidy, or age structure.
+discrete generations and carry two genomes (maternal and paternal).
+The pedigree is shown here for concreteness, and
+there is no inherent limitation on mating system, ploidy, or age structure
+in the gARG encoding.
 A genome is inherited from one of the individual's parents,
@@
 is not represented by a single ``recombination node'' (as in an eARG), but with
-two parent genomes between which recombination takes place (e.g.
-nodes \textsf{e} and \textsf{f}), and a
-single resulting ``recombinant'' genome (e.g. \textsf{c}). The
+two parent genomes between which recombination takes place
+(e.g.\ nodes \textsf{e} and \textsf{f}), and a
+single resulting ``recombinant'' genome (e.g.\ \textsf{c}). The
 edges that link this recombinant with its two parents carry the information
-concerning breakpoint location that would normally be present on the
+concerning breakpoint location that would be associated with a
 recombination node in an eARG. For example, node \textsf{c} in

### 2022-07-27 paper.tex
Reorder event section
And stick some intermediate headings in to make the logic clearer to ourselves (these can be removed later if need be)

@@
 
+\subsection*{Node versus edge annotation}
+
 % What do we mean by an EARG, informally?
@@
 % qualify the meaning of ``ARG''
-simulating an ARG in this representation is
-complicated by the fact that the
-first event to occur may hit the lineage carrying the ancestry
-to the right of the breakpoint rather
+when using this representation and simulating an ARG backwards in time,
+the first event above a recombination may involve the lineage carrying the
+ancestry to the right of the breakpoint rather
 than the left, and so we cannot emit edges as they are generated.
-Such issues can be worked around, of course,
-but depending on the ordering of otherwise indistinguishable
-objects is generally problematic. In practice, several
-methods
-explicitly associate information with the outbound edges
+Such issues can be worked around, but depending on the ordering of
+otherwise indistinguishable objects is generally problematic. In practice,
+several methods explicitly associate information with the outbound edges
 of a recombination event
@@
 
+Another limitation of the eARG encoding illustrated in
+Fig.~\ref{fig-arg-data-structure}A is the inability to represent
+multiple recombinations along a chromosome, gene conversion,
+or transmission of multiple chromosomes.
+[Sentence saying ``gene conversion is important, actually, cite, cite,
+so we should be hoping to infer it some day''. Presumably people
+already are in bacterial contexts? Gregor: I recall this paper from
+Arabidopsis https://www.pnas.org/doi/10.1073/pnas.1211827110]
+To allow us to encode data from these important processes we
+must either devise workarounds to the event based encoding
+(i.e., model gene conversion as two recombination events) or
+generalise the model to more appropriately reflect the realities
+of modern datasets.
+
+Fortunately, it is straightforward to generalise the classical Griffiths
+eARG encoding to both simplify implementations and to encompass
+alternative patterns of genetic inheritance.
+The only change that we require is to associate
+the details of inheritance intervals with \emph{edges} rather than
+as breakpoints associated with recombination events and
+edge ordering rules. This is illustrated in Fig~\ref{fig-arg-data-structure}C
+where this alternative edge annotation formulation is
+shown. By associating the interval $(0, x)$ with edge $(d,e)$
+and the interval $(x, L)$ with $(d, f)$ we make the encoding
+unambiguous: we can list the edges in any order, and it
+will describe the same set of local genealogical trees.
+This approach is also more flexible, because any pattern
+of inheritance can be described by these intervals. For
+example, if a gene conversion event occurred where a segment
+between positions $x$ and $y$ was inherited by
+node $d$ from $e$, we would state this by associating
+the intervals $\{(0, x], (y, L]\}$ with $(d,e)$ and
+$\{(x, y]\}$ with $(d,f)$.
+This change is straightforward, involves no loss of information
+and (as we see in later sections)
+greatly increases both the expressivity and computational
+power of ancestral recombination graph data structures.
+
 % it's not reasonable to assume that one type of event happens
 % at a time any more.
-The eARG is also limited in the types of ancestral history that can
-be described. This ARG encoding is derived from the coalescent
-with recombination, which explicitly assumes that only one type
+\subsection*{Issues with event-based encoding}
+
+Even with the change to edge annotations, a focus on specific
+types of events can limit the types of ancestral history that can
+be described by an eARG. As it is derived from the coalescent
+with recombination, a standard eARG explicitly assumes that only one type
 of event can occur (i.e., common ancestry or recombination) at a
@@
 
-A final limitation of the eARG encoding illustrated in
-Fig.~\ref{fig-arg-data-structure}A is the inability to represent
-multiple recombinations along a chromosome, gene conversion,
-or transmission of multiple chromosomes.
-[Sentence saying ``gene conversion is important, actually, cite, cite,
-so we should be hoping to infer it some day''. Presumably people
-already are in bacterial contexts? Gregor: I recall this paper from
-Arabidopsis https://www.pnas.org/doi/10.1073/pnas.1211827110]
-To allow us to encode data from these important processes we
-must either devise workarounds to the event based encoding
-(i.e., model gene conversion as two recombination events) or
-generalise the model to more appropriately reflect the realities
-of modern datasets.
-
-Fortunately, it is straightforward to generalise the classical Griffiths
-eARG encoding to both simplify implementations and to encompass any
-possible pattern of genetic inheritance.
-The only change that we require is to associate
-the details of inheritance intervals with \emph{edges} rather than
-as breakpoints associated with recombination events and
-edge ordering rules. This is illustrated in Fig~\ref{fig-arg-data-structure}C
-where this alternative edge annotation formulation is
-shown. By associating the interval $(0, x)$ with edge $(d,e)$
-and the interval $(x, L)$ with $(d, f)$ we make the encoding
-unambiguous: we can list the edges in any order, and it
-will describe the same set of local genealogical trees.
-This approach is also more flexible, because any pattern
-of inheritance can be described by these intervals. For
-example, if a gene conversion event occurred where a segment
-between positions $x$ and $y$ was inherited by
-node $d$ from $e$, we would state this by associating
-the intervals $\{(0, x], (y, L]\}$ with $(d,e)$ and
-$\{(x, y]\}$ with $(d,f)$.
-This change is straightforward, involves no loss of information
-and (as we see in later sections)
-greatly increases both the expressivity and computational
-power of ancestral recombination graph data structures.
+For these reasons, we suggest that instead of defining the nodes in an
+ARG as events, which limits us to certain models of inheritance,
+we use a more flexible interpretation where they are defined as genomes.
 

### 2022-07-31 paper.tex
Light edits

@@
 describe the regions of genome that have been passed from a parent to its immediate children,
-hence edges either span the entire genome, or all the region to the left (or right) of a breakpoint
+hence edges either span the entire genome, or entire region to the left (or right)
+of a breakpoint
 (e.g. Fig~\ref{fig-arg-data-structure}C). In this form of edge annotation, which we call
@@
 
-Fig~\ref{fig-ancestry-resolution} illustrates the two different edge annotation schemes.
-To obtain the sample-resolved annotation in Fig~\ref{fig-ancestry-resolution}B we must
-traverse backwards in time from sample nodes \textsf{a}, \textsf{b}, and \textsf{c}, using the
-edge intervals to record regions of sample genome that have passed through each route.
-When edges join into a parent node, the parent is allocated the union of the intervals.
-When a child node is connected to two parents, one parent is allocated the edge
-intervals of the child, truncated to include only genomic material to the left of the breakpoint.
-Similarly the other parent is allocated the child's intervals truncated such that they contain
-only regions to the right of the breakpoint (a natural extension exists for multiple
-chromosomal breakpoints).
+Fig~\ref{fig-ancestry-resolution} illustrates the two different edge annotation
+schemes. To obtain the sample-resolved annotation in
+Fig~\ref{fig-ancestry-resolution}B we must traverse backwards in time from
+sample nodes \textsf{a}, \textsf{b}, and \textsf{c}, using the edge intervals
+to record regions of sample genomes that have passed through each route. When
+edges join into a parent node, the parent is associated with the union of the
+child intervals. When a child node is connected to two parents, one parent is
+allocated the edge intervals of the child, truncated to include only genomic
+material to the left of the breakpoint. Similarly the other parent is allocated
+the child's intervals truncated such that they contain only regions to the
+right of the breakpoint (a natural extension exists for multiple chromosomal
+breakpoints).
 
@@
 parent node \textsf{r} spans the entire genome from 0..7, reflecting the material
-transmitted immediately between parent and child: i.e. local to those two nodes in the tree.
+transmitted immediately between parent and child.
 However, in Fig~\ref{fig-ancestry-resolution}B the edge has been reduced in extent so that
@@
 For larger ARGs, the absence of these nodes makes sample-resolved edge annotation
-substantially more efficient than immediate edge annotation. Second, by the time we reach
+substantially more efficient than immediate edge annotation.
+Second, by the time we reach
 nodes \textsf{t} and \textsf{v}, all regions of the genome have fully coalesced, and so
 \textsf{w}, the ``grand MRCA'' can be omitted from the graph, at the cost of having
-multiple root nodes in different regions of the genome. Third, there exist nodes such as
+multiple root nodes in different regions of the genome.
+Third, there exist nodes such as
 \textsf{k} with two children (i.e. ``common ancestor'' nodes in Hudson's terminology)
@@
 within regions of genome that are not ancestral (e.g. the breakpoint at position 3 in
-node \textsf{k}). in this case sample-resolved edge annotations lose the exact position of
+node \textsf{k}). In this case sample-resolved edge annotations lose the exact position of
 the breakpoint, although they retain the information that a breakpoint occurred between
@@
 because they provide very little information and are not
-detectable by information from the tree. % weak
+detectable by information from any given individual tree.
 The nodes in a tree usually mark the points where samples find a
@@
 The first type of simplification that we can perform is to remove
-graph topology that is invisible to the samples. The most well
+graph topology that is invisible to the samples. The best
 known example of such topology is a so-called
@@
 in which the two parent nodes of a recombination immediately
-join again into a common ancestor (e.g. \textsf{j}, \textsf{l}, \textsf{m}
+join again into a common ancestor (e.g.~\textsf{j}, \textsf{l}, \textsf{m}
 and  \textsf{n} in Fig.~\ref{fig-simplification}A).
@@
 of the local trees. In particular, a node can represent a recombinant
-with multiple parents in the graph but only a single child (e.g. node \textsf{n}
+with multiple parents in the graph but only a single child (e.g.\ node \textsf{n}
 in Fig.~\ref{fig-simplification}B), or can represent a common ancestor with
@@
 edge annotations. Removal of recombinant nodes can produce graph nodes with
-more than 2 parents (e.g. node \textsf{e}); likewise, removal of
+more than 2 parents (e.g.~node \textsf{e}); likewise, removal of
 common ancestor but non-coalescent nodes can produce graph nodes with
-more than 2 children (e.g. node \textsf{s}). These represent a ``stacking up'' of multiple
-often indistinguishable events into a single node, something which is impossible
-to encode in an eARG.
+more than 2 children (e.g.~node \textsf{s}). These represent a ``stacking up'' of
+multiple---often indistinguishable---events into a single node,
+something which is impossible to encode in an eARG.
 

### 2022-07-31 paper.tex
Tweaks to intro and refs

@@
 and the ancestral relationships between those nodes. A rich literature exists
-on analysing the mathematical properties of these trees [cite?], and numerous
+on analysing the mathematical properties of these
+trees~\citep{steel2016phylogeny}, and numerous
 methods exist to infer them~\citep{felsenstein2004inferring}. The situation is
@@
 Recent breakthroughs in large-scale inference
-methods~\citep{rasmussen2014genome,kelleher2019inferring,speidel2019method,
-schaefer2021ancestral,wohns2022unified}
+methods~\citep{rasmussen2014genome,kelleher2019inferring,speidel2019method,wohns2022unified}
 and data representation~\citep{kelleher2016efficient}
-have raised the genuine prospect of genealogical analysis becoming a standard part
+have raised the realistic prospect of genealogical analysis becoming a standard part
 of the population and statistical genetics toolkit~\citep{hejase2020summary}.
 Applications using these inferred ancestries as input have
-begun to appear~\citep{osmond2021estimating} and many more are sure to
+begun to
+appear~\citep{osmond2021estimating,zhang2021biobank,fan2022genealogical}
+and many more are sure to
 follow~\citep{harris2019database}. This vibrant research area, however,
@@
 difficulties in either comparing the outputs of different methods or
-(from a users perspective) using multiple inference methods in an analysis.
+(from a user's perspective) using multiple inference methods in an analysis.
 
@@
 
-% \section*{Ancestral graphs and stochastic processes}
-% The coalescent~\citep{kingman1982coalescent,kingman1982genealogy,
-% hudson1983testing, tajima1983evolutionary}
-% provides the theoretical underpinning for much of contemporary population genetics.
-% \citet{hudson1983properties,hudson1990gene} extended the coalescent to include
-% recombination
-% and the model has proved to be an especially powerful basis for
-% simulations~\citep{hudson2002generating,baumdicker2021efficient}.
-% \cite{kelleher2016efficient} provide a description of his algorithm along with
-% an efficient implementation, the details of which turn out to be highly relevant to
-% later sections of the current paper (see the ``\nameref{Ancestry_resolution}'' section).
-
-% Although Hudson's algorithm was accompanied by a graph
-% schematic~\citep[][Fig. 5]{hudson1983properties}, % but for clarity and following recent usage,
-% we will use ARG to refer to a graph structure, and refer to the
-% process as the Griffiths process or the coalescent with recombination
-% \citep{minichiello2006mapping}.
-
-% NOTE: it might be helpful to have a drawing of the ARG process
-% in-progress here to help us talk about it. What is the full state
-% that is remembered as we go through the process? It does help to think
-% about the process as choosing edges to merge, not nodes.
-% We could also refer to Fig.~\ref{fig-arg-data-structure}B if we want a simple picture
-% to go with this
-% In the Griffiths formulation, each edge in the graph corresponds to an extant
-% lineage and nodes are events in the process (the initial $n$ leaf nodes are
-% ``sampling'' events). Common ancestor events occur at rate $\binom{k}{2}$ when there
-% are $k$ lineages present. We choose two lineages (edges) uniformly, and merge them
-% into a common ancestor lineage. Recombination events happen at a rate of
-% $k \rho$. Here, we choose a lineage (edge) uniformly, and a
-% breakpoint $0 < x < m$ uniformly on its genome. We terminate the edge at a
-% node, record the breakpoint and start two new edges from this node. The process
-% then continues until there is only one lineage left (the Grand Most Recent
-% Common Ancestor, GMRCA), which is guaranteed to
-% happen in finite time because of the linear vs quadratic rates of branching
-% and coalescing. Later, we will show how the graph structure and breakpoints
-% associated with
-% recombination nodes provides sufficient information to later recover the marginal
-% trees and calculate sampling probabilities. However, this process traces out
-% a larger graph (the ``big'' ARG) than the Hudson algorithm (the ``little'' ARG; see the ``\nameref{Ancestry_resolution}'' section),
-% and is consequently considerably less efficient \citep{wiuf1999ancestry}.
-
-% Similar graph encodings have also been introduced to model other mechanisms
-% which change the local ancestral tree along the genome, such as gene
-% conversion~\citep{wiuf2000coalescent} and
-% horizontal gene transfer~\citep{baumdicker2014infinitely}.
-% [Mention the
-% SMC~\citep{mcvean2005approximating,marjoram2006fast}
-% and ClonalFrame~\citep{didelot2007inference} approximations somewhere.]
+\section*{Ancestral recombination graphs}
+
+The coalescent~\citep{kingman1982coalescent,kingman1982genealogy,
+hudson1983testing, tajima1983evolutionary}
+provides the theoretical underpinning for much of contemporary population genetics.
+\citet{hudson1983properties,hudson1990gene} extended the coalescent to include
+recombination
+and the model has proved to be an especially powerful basis for
+simulations~\citep{hudson2002generating,baumdicker2021efficient}.
+\cite{kelleher2016efficient} provide a description of his algorithm along with
+an efficient implementation, the details of which turn out to be highly relevant to
+later sections of the current paper (see the ``\nameref{Ancestry_resolution}'' section).
+
+Although Hudson's algorithm was accompanied by a graph
+schematic~\citep[][Fig. 5]{hudson1983properties}, % but for clarity and following recent usage,
+we will use ARG to refer to a graph structure, and refer to the
+process as the Griffiths process or the coalescent with recombination
+\citep{minichiello2006mapping}.
+
+NOTE: it might be helpful to have a drawing of the ARG process
+in-progress here to help us talk about it. What is the full state
+that is remembered as we go through the process? It does help to think
+about the process as choosing edges to merge, not nodes.
+We could also refer to Fig.~\ref{fig-arg-data-structure}B if we want a simple picture
+to go with this
+In the Griffiths formulation, each edge in the graph corresponds to an extant
+lineage and nodes are events in the process (the initial $n$ leaf nodes are
+``sampling'' events). Common ancestor events occur at rate $\binom{k}{2}$ when there
+are $k$ lineages present. We choose two lineages (edges) uniformly, and merge them
+into a common ancestor lineage. Recombination events happen at a rate of
+$k \rho$. Here, we choose a lineage (edge) uniformly, and a
+breakpoint $0 < x < m$ uniformly on its genome. We terminate the edge at a
+node, record the breakpoint and start two new edges from this node. The process
+then continues until there is only one lineage left (the Grand Most Recent
+Common Ancestor, GMRCA), which is guaranteed to
+happen in finite time because of the linear vs quadratic rates of branching
+and coalescing. Later, we will show how the graph structure and breakpoints
+associated with
+recombination nodes provides sufficient information to later recover the marginal
+trees and calculate sampling probabilities. However, this process traces out
+a larger graph (the ``big'' ARG) than the Hudson algorithm (the ``little'' ARG; see the ``\nameref{Ancestry_resolution}'' section),
+and is consequently considerably less efficient \citep{wiuf1999ancestry}.
+
+Similar graph encodings have also been introduced to model other mechanisms
+which change the local ancestral tree along the genome, such as gene
+conversion~\citep{wiuf2000coalescent} and
+horizontal gene transfer~\citep{baumdicker2014infinitely}.
+[Mention the
+SMC~\citep{mcvean2005approximating,marjoram2006fast}
+and ClonalFrame~\citep{didelot2007inference} approximations somewhere.]
 
@@
 
+TODO mention \citep{schaefer2021ancestral} somewhere.
+
 % AI: just to note, these papers do not all agree in their definition of an ARG.

### 2022-07-31 paper.tex
Create Ancestral graphs: history section
- Adapt most of content from the Phylogenetic networks section to this
new section or a discussion point.

--- 2022-07-31 paper.tex
@@
 
-% \section*{Ancestral graphs and stochastic processes}
-% The coalescent~\citep{kingman1982coalescent,kingman1982genealogy,
-% hudson1983testing, tajima1983evolutionary}
-% provides the theoretical underpinning for much of contemporary population genetics.
-% \citet{hudson1983properties,hudson1990gene} extended the coalescent to include
-% recombination
-% and the model has proved to be an especially powerful basis for
-% simulations~\citep{hudson2002generating,baumdicker2021efficient}.
-% \cite{kelleher2016efficient} provide a description of his algorithm along with
-% an efficient implementation, the details of which turn out to be highly relevant to
-% later sections of the current paper (see the ``\nameref{Ancestry_resolution}'' section).
+\section*{Ancestral graphs: a brief history}
+The coalescent~\citep{kingman1982coalescent,kingman1982genealogy,
+hudson1983testing, tajima1983evolutionary} models the ancestry of a sample of
+genomes under an idealised population model, and provides the theoretical
+underpinning for much of contemporary population genetics.
+The coalescent is a stochastic process, where each random realisation
+is a genealogical tree describing the genetic
+ancestors of the samples and their inheritance relationships.
+A great strength of the coalescent as a modelling framework
+is its flexibility, and numerous elaborations have been
+incorporated~\citep{hudson1990gene,hein2004gene,wakely2008coalescent}.
+In the first extension to the basic process, \cite{hudson1983properties}
+analysed the effects of recombination on the random genealogical
+trees generated by the coalescent,
+producing several fundamental analytical results
+as well as describing the elegant algorithm
+at the heart of simulation
+programs~\citep{hudson2002generating,kelleher2016efficient,baumdicker2021efficient}
+used in thousands of studies.
+In its original formulation, Hudson's algorithm produced a
+sequence of genealogical trees along the genome.
+Critically, although the correlation structure among these
+local trees was analysed analytically, those correlations
+were not captured explicitly: the ancestry of the sample
+was viewed as a series of disconnected local trees rather than
+as a single cohesive object.
+
+In the 1990s, Griffiths and colleagues revisited the
+coalescent with recombination from a different perspective,
+in which the stochastic process itself is encoded as a graph
+(with particular properties; see section XXX for details),
+and showed how the local trees generated by Hudson's algorithm
+were embedded within
+it~\citep{griffiths1991two,ethier1990two,griffiths1996ancestral,griffiths1997ancestral}.
+They referred to this graph,
+representing both the stochastic process and
+its random realisations, as the Ancestral Recombination Graph (ARG).
+Although mathematically equivalent, it is
+important to note that the Griffiths and Hudson formulations of
+the coalescent with recombination are not identical;
+in particular, a direct implementation of the ARG process
+requires exponential time to simulate
+(see Appendix XXX for details).
+
+Early work on ARG inference focused on the problem of
+inferring parameters of the
+stochastic process, where the ancestry is regarded as a
+latent parameter to be averaged out
+\citep[e.g.][]{griffiths1996ancestral,kuhner2000maximum, nielsen2000estimation,
+fearnhead2001estimating}.
+These methods met with limited success
+because the state space of ARGs is overwhelmingly large and
+lacks a simple geometry or neighbourhood structure for inference or
+sampling methods to  exploit.
+The focus subsequently shifted to the more
+tractable---but still NP-hard~\citep{wang2001perfect}---problem of computing
+the minimum number of recombinations required
+to explain an input dataset~\citep{myers2003bounds}, and inferring the corresponding
+ARG realisations~\citep{song2003parsimonious,song2005efficient,lyngso2005minimum}.
+The previously close correspondance
+between the ARG as a mathematical representation of the coalescent process
+and as concrete realisations of ancestry was now broken,
+since such parsimonious ARGs are extremely unlikely under the
+% JK: I assume this is true; any citations to back it up I wonder?
+coalescent with recombination.
+\cite{minichiello2006mapping} made this separation of the ARG as a
+stochastic process vs the ARG as a data structure explicit, and
+for using the term to mean specific realised ancestral histories.
+This has become a common
+usage~\citep{mathieson2020ancestry,brandt2021evaluation}
+[TODO expand sentence].
+Gusfield~\citep{gusfield2014recombinatorics}
+defines also defines an ARG as a concrete realisation
+and studies instances that satisfy particular
+properties such as ``galled trees''
+\citep{wang2001perfect, gusfield2004optimal}
+as well as examining more general phylogenetic
+networks~\citep{huson2010phylogenetic}.
+
+% JK: This paragraph is weak, and needs some work. Basically want to
+% illustrate the mathematical literature which views ancestral
+% graphs very much as *processes*.
+However, in parallel with this tendancy to decouple the term
+Ancestral Recombination graph from its original defining stochastic process,
+an extensive mathematical literature has developed in
+studying other related stochastic processes encoded via a graph.
+In particular, the Ancestral Selection Graph
+(ASG)~\citep{krone1997ancestral,neuhauser1997genealogy}
+uses a similar approach to model an ensemble of correlated
+potential ancestral trees and has been of considerable
+and ongoing theoretical interest~\citep[e.g.][]{
+neuhauser1999ancestral,
+donnelly1999genealogical,
+fearnhead2001perfect,
+fearnhead2003ancestral,
+etheridge2009coalescent,
+gonzalezcasanova2018duality,
+koskela2019robust}.
+A vital distinction between the ARG (process) and ASG, however, is that there
+is that there is a hard distinction between the stochastic process
+(the graph) and an observable realisation (a single tree) because
+the ASG encodes \emph{potential} ancestral trees.
+Similarly, the Ancestral Gene Transfer Graph
+models horizontal gene transfer~\citep{baumdicker2014infinitely}
+by [FINISH ME. Also, any other ASG-like \emph{processes} to mention?].
+
+The point of this historical overview is to illustrate that the term
+Ancestral Recombination Graph has different meanings in different
+contexts, and in its original form was it was closely coupled to
+the coalescent with recombination. In the following sections we
+show that, although the ARG-as-a-data-structure perspective is
+now the most common interpretation, the \emph{form} of that
+data structure is still closely derived from the original
+stochastic process, and that this form is inflexible
+in the types of genetic inheritance that can be described,
+computationally inefficient, and
+forces spurious precision about the location and timing
+of recombination events.
 
@@
 % and is consequently considerably less efficient \citep{wiuf1999ancestry}.
-
-% Similar graph encodings have also been introduced to model other mechanisms
-% which change the local ancestral tree along the genome, such as gene
-% conversion~\citep{wiuf2000coalescent} and
-% horizontal gene transfer~\citep{baumdicker2014infinitely}.
-% [Mention the
-% SMC~\citep{mcvean2005approximating,marjoram2006fast}
-% and ClonalFrame~\citep{didelot2007inference} approximations somewhere.]
 
@@
 [Sentence saying ``gene conversion is important, actually, cite, cite,
-so we should be hoping to infer it some day''. Presumably people
+so we should be hoping to infer it some day''.
+Cite~\citep{wiuf2000coalescent}.
+Presumably people
 already are in bacterial contexts? Gregor: I recall this paper from
@@
 
+[Mention the
+and ClonalFrame~\citep{didelot2007inference} approximation somewhere.]
+
 Under the coalescent with recombination (or SMC) each local
@@
 can be used on megabase scale data and hundreds of thousands of samples under human-like parameters.
+
+TODO mention \citep{schaefer2021ancestral} somewhere.
 
@@
 
-\item We should discuss other types of ancestry network also.
-The crucial distiction between an ARG and a phylogenetic network (I think?) is
-the presence of position information specifying the location of
-recombination events (and by extension, the complete passage
-of ancestral material the graph). Without position information
-we know that a particular node was an ancestor somewhere along the
-sequence, but not exactly where.
-
-\item We make the point about expressivity of the eARG pretty
-hard, but we don't mention anything about indels and structural
-rearrangements etc. It's probably worth mentioning genome graphs
-somewhere just to fend off this criticism. The high-level point
-would be that it's unlikely to be easier to model this stuff
-using node annotations.
+\item The gARG structure as described here is sufficiently expressive
+to model any form of homologous genetic transmission between individuals,
+and is far more flexible than the classical eARG approach.
+Phylogenetic networks~\citep{huson2010phylogenetic} are a more
+general concept, and specific classes capture dynamics that
+cannot be modelled using an ARG. For example,
+reassortment networks model the exchange of
+(non-recombining) genomic segments between viral particles inside a co-infected host,
+and hybridisation networks, model the genetic material from
+different species combining to create a hybrid.
+Similarly Genome Graphs~\cite{novak2017genome} model complicated things
+within species. Future work should look to capture all of these things
+in a single structure.
 
 \end{itemize}
-
-\textcolor{red}{Discussion below on phylogenetic networks: we need to
-reduce the text below down to one paragraph, which
-defines the relationship between what we define as an ARG and
-what people mean by phylogenetic networks from the literature.
-We probably don't need to get into quite so much detail on
-the different types that people have studies. Essentially,
-we're ending the section with a brief review which situates
-our definition in the literature. I think pulling apart the
-idea of a ``reticulation'' vs what we call a recombinant
-is essential here.}
-
-The term \emph{phylogenetic network} has been broadly accepted to describe the most general
-type of genealogical graph, defined by \citet{huson2010phylogenetic} as ``any graph used to
-represent evolutionary relationships (either abstractly or explicitly) between a set of taxa
-that labels some of its nodes'' (although more narrow and context-dependent definitions can also
-be found in the literature). Explicit rooted phylogenetic networks are rooted Directed Acyclic Graphs (DAGs),
-with leaves labelled by the taxa (species, groups, individuals or sequences), and nodes classed
-as either reticulation nodes (with in-degree $\geq 2$) or tree nodes.
-
-Some sub-classes of explicit rooted phylogenetic networks are defined based on the evolutionary event that the
-reticulation nodes capture \citep{huson2010phylogenetic}. In \emph{reassortment} networks, used to
-model the evolution of viruses, reticulation nodes correspond to exchange of
-(non-recombining) genomic segments between viral particles inside a co-infected host.
-In \emph{hybridisation} networks, reticulation nodes correspond to genetic material from two
-different species combining to create a hybrid. In \emph{recombination} networks, the reticulation nodes
-correspond to recombination events (with leaves labelled by sequences and edges labelled by mutations).
-The ARG is a subtype of a recombination network under this definition (restricting to one crossover
-recombination per reticulation node).
-% AI: Not sure about this - do ARGs also encompass multiple crossover recombinations and gene
-% conversion (seems implied above)?
-
-Sub-classes of explicit rooted phylogenetic networks can also be defined by restrictions on their topology.
-In \emph{binary networks}, all internal nodes must have either in-degree two and out-degree one or in-degree
-one and out-degree two (with the root having out-degree two and the leaves in-degree 1) \citep{steel2016phylogeny}.
-In \emph{tree-child networks}, every internal node must have at least one child that is a tree node;
-\emph{tree-sibling networks} are constrained by every reticulation node having at least one sibling that is a
-tree node \citep{cardona2008extended}. Further, sub-types can be defined by restrictions on
-the topology of subgraphs that contain the reticulation loops (cycles containing a reticulation node).
-A reticulation loop that does not share any node with another reticulation loop is termed a gall
-\citep[][p.\ 237]{gusfield2014recombinatorics}; a reticulation network containing only galls is
-called a \emph{galled tree}. Reticulation loops that are joined together in a reticulation network
-by at least one shared node form a blob; a \emph{level-$k$ network} is one where the maximal number
-of reticulation loops inside a blob is $k$ \citep{choy2005computing}. Galled trees are level-1 networks,
-and the tractability afforded by their relatively simple structure results in a number of attractive
-combinatorial properties \citep{wang2001perfect, gusfield2004optimal}. We consider the general case of
-an ARG, without applying any of these restrictions.
-
-% AI: have not mentioned any implicit or un-rooted networks (split, median, haplotype, reticulograms, etc).
-% Also: are we not defining ARGs as ``binary networks''? Might be good to specify this here.
-
-\textcolor{red}{Discussion on ASG ported from earlier (removed) ``stochastic processes'' section}
-
-The idea of using branching-coalescing stochastic processes
-to model genetic ancestral trees has also found application in natural selection.
-The Ancestral Selection Graph (ASG)~\citep{krone1997ancestral,neuhauser1997genealogy}
-uses dynamics identical to the ``big'' ARG to simulate an ensemble of correlated
-potential ancestral trees. Weak genic selection is incorporated by sampling a
-true ancestry from the ensemble in a non-uniform way.
-There is no ASG analogue of the more computationally tractable ``little'' ARG,
-though some gains in tractability can be made by considering typed
-lineages~\citep{etheridge2009coalescent} or by leveraging perfect simulation
-techniques when recurrent mutation is present \citep{fearnhead2001perfect}.
-Extensions of the ASG have been developed to frequency-dependent
-selection~\citep{neuhauser1999ancestral, gonzalezcasanova2018duality},
-unlinked chromosomes~\citep{fearnhead2003ancestral}, recombining
-loci whereupon branching is due to both selection and
-recombination~\citep{donnelly1999genealogical}, and high fecundity
-reproduction~\citep{gonzalezcasanova2018duality, koskela2019robust}.
 

### 2022-07-31 paper.tex
Create Big and Little ARG appendix
Move related content in there, without actually doing any new writing

@@
 Inferring the genetic relationships between sampled genomes in the form of an
-evolutionary tree is a necessary prerequisite for many analyses of species \citep{rannala2003genetics}, non (or
-rarely) recombining sections of DNA \citep{cann1987mitochondrial,underhill2001annalsofhumangenetics} and viruses \citep{grenfell2004science}.
+evolutionary tree is a necessary prerequisite for many analyses of
+species~\citep{rannala2003genetics}, non-recombining sections of
+DNA~\citep{cann1987mitochondrial,underhill2001annalsofhumangenetics}
+and viruses~\citep{grenfell2004science}.
 A genealogical tree captures what is known and
@@
 % of ancestry''.
-\item Throughout, we consider ARGs defined for a single chromosome comprising
-of a fixed number of discrete sites. Our argument applies equally well to ARGs
-defined for several non-homologous chromosomes \citep{fearnhead2003ancestral,
-koskela2019robust} and to idealised, continuous chromosomes, but addressing
-all these cases simultaneously would result in needlessly bulky notation.
 \end{enumerate}
@@
 requires exponential time to simulate
-(see Appendix XXX for details).
+(see Appendix XXX for details). However, ARGs provided a way
+to describe recombinant ancestry as single object that
+could be inferred in a way
+that was missing from Hudson's treatments.
 
@@
 stochastic process vs the ARG as a data structure explicit, and
-for using the term to mean specific realised ancestral histories.
+argued for using the term to mean specific realised ancestral histories.
 This has become a common
@@
 \citep{wang2001perfect, gusfield2004optimal}
-as well as examining more general phylogenetic
+as well as describing algorithms for more general phylogenetic
 networks~\citep{huson2010phylogenetic}.
@@
 koskela2019robust}.
-A vital distinction between the ARG (process) and ASG, however, is that there
+A vital distinction between the ARG (process) and ASG, however,
 is that there is a hard distinction between the stochastic process
-(the graph) and an observable realisation (a single tree) because
-the ASG encodes \emph{potential} ancestral trees.
+(the graph) and an observable realisation (a single tree) in the ASG
+because the it encodes \emph{potential} ancestral trees.
 Similarly, the Ancestral Gene Transfer Graph
@@
 Ancestral Recombination Graph has different meanings in different
-contexts, and in its original form was it was closely coupled to
+contexts, and in its original form it was closely coupled to
 the coalescent with recombination. In the following sections we
@@
 computationally inefficient, and
-forces spurious precision about the location and timing
+forces complete (and potentially spurious) precision about
+the location and timing
 of recombination events.
-
-% Although Hudson's algorithm was accompanied by a graph
-% schematic~\citep[][Fig. 5]{hudson1983properties}, % but for clarity and following recent usage,
-% we will use ARG to refer to a graph structure, and refer to the
-% process as the Griffiths process or the coalescent with recombination
-% \citep{minichiello2006mapping}.
-
-% NOTE: it might be helpful to have a drawing of the ARG process
-% in-progress here to help us talk about it. What is the full state
-% that is remembered as we go through the process? It does help to think
-% about the process as choosing edges to merge, not nodes.
-% We could also refer to Fig.~\ref{fig-arg-data-structure}B if we want a simple picture
-% to go with this
-% In the Griffiths formulation, each edge in the graph corresponds to an extant
-% lineage and nodes are events in the process (the initial $n$ leaf nodes are
-% ``sampling'' events). Common ancestor events occur at rate $\binom{k}{2}$ when there
-% are $k$ lineages present. We choose two lineages (edges) uniformly, and merge them
-% into a common ancestor lineage. Recombination events happen at a rate of
-% $k \rho$. Here, we choose a lineage (edge) uniformly, and a
-% breakpoint $0 < x < m$ uniformly on its genome. We terminate the edge at a
-% node, record the breakpoint and start two new edges from this node. The process
-% then continues until there is only one lineage left (the Grand Most Recent
-% Common Ancestor, GMRCA), which is guaranteed to
-% happen in finite time because of the linear vs quadratic rates of branching
-% and coalescing. Later, we will show how the graph structure and breakpoints
-% associated with
-% recombination nodes provides sufficient information to later recover the marginal
-% trees and calculate sampling probabilities. However, this process traces out
-% a larger graph (the ``big'' ARG) than the Hudson algorithm (the ``little'' ARG; see the ``\nameref{Ancestry_resolution}'' section),
-% and is consequently considerably less efficient \citep{wiuf1999ancestry}.
+Unless otherwise stated, we will use the ARG-as-a-data-structure
+interpretation for the remainder of this paper.
 
 \section*{The Event ARG data structure}\label{eARG}
-Initial work on \emph{simulating} ancestral graphs with recombination was undertaken by
-\citet{hudson1983properties,hudson1990gene}, who extended the classical coalescent
-framework~\citep{kingman1982coalescent,kingman1982genealogy, hudson1983testing, tajima1983evolutionary}
-to include recombination. Hudson's algorithm has led to powerful simulation
-tools~\citep{hudson2002generating,kelleher2016efficient, baumdicker2021efficient}, and is
-also relevant to efficient analysis, as we shall see later in the
-``\nameref{Ancestry_resolution}'' section.
-
-% When did we move from the ARG is a process to the ARG is a data structure?
-% What was the historical progression from one to the other?
-
-The precise term ``Ancestral Recombination Graph'' or ARG, was subsequently introduced by
-Griffiths and colleagues~\citep{griffiths1991two,ethier1990two,griffiths1996ancestral,griffiths1997ancestral}.
-Rather than simulation, their work focused on mathematical results about the underlying branching-coalescing
-stochastic process. However, some confusion has arisen from their use of the word ``ARG'' to refer to
-two distinct concepts: the process itself, and the graph structure it generates.
-
-%% NB: I (yan) have commented out the following section as the "latent parameter" quote is
-%% used again in the "ARG reconstruction" section, and may be more detail than is required here
-%% This also loses the description of ARG space being large etc.
-%Early work on ARG inference focused on the problem of
-%inferring parameters of the
-%stochastic process, where the ancestry is regarded as a
-%latent parameter to be averaged out
-%\citep[e.g.][]{griffiths1996ancestral,kuhner2000maximum, nielsen2000estimation,
-%fearnhead2001estimating}.
-%These methods met with limited success
-%because the state space of ARGs is overwhelmingly large and
-%lacks a simple geometry or neighbourhood structure for inference or
-%sampling methods to  exploit.
-%The focus subsequently shifted to the more
-%tractable---but still NP-hard~\citep{wang2001perfect}---problem of computing
-%the minimum number of recombinations required
-%to explain an input dataset [CITATIONS], and inferring the corresponding
-%ARG realisations~\citep{song2003parsimonious,song2005efficient,lyngso2005minimum}.
-%This highlighted the distinction between the ARG as a stochastic process
-%and the ARG as a \emph{data structure}~\citep{minichiello2006mapping},
-%since parsimonious ARGs have no immediate connection to
-%the stochastic process.
-
-The second concept, that of the ARG being a \emph{data structure} was adopted by later authors,
-when focussing on the problem of inferring an ARG-like structure (or a minimum number of
-recombinations [CITATIONS]) without assuming a specific stochastic
-model, for example using parsimony~\citep{song2003parsimonious,song2005efficient,lyngso2005minimum}.
-Following~\citet{minichiello2006mapping}, we adopt this convention, using ARG
-to mean an ancestral graph structure, without reference to the process that generated it.
 
@@
 \caption{\label{fig-ancestry-resolution}
-The \citet[][fig. 1]{wiuf1999recombination} example graph as a gARG, with contrasting
+The \citet[][fig.~1]{wiuf1999recombination} example graph as a gARG, with contrasting
 edge annotation schemes.
@@
 
-% NB: below is a discussion of the Hudson algorithm which we might want to
-% retain in an appendix
-% Each lineage consists of a list of
-% disjoint ancestry segments $(\ell, r, a)$, where
-% $[\ell, r)$ is a half-closed genomic interval and $a$ is an integer
-% tracking the number of samples to which the lineage is ancestral over that interval.
-% (We also usually track the tree node associated with each segment, but
-% that is not important for our purposes here so we omit it.)
-% If we have $n$ samples and a genome of length $m$, the process begins with $n$ lineages
-% of the form $\{(0, m, 1)\}$. The process then works backwards in time from
-% the present day as a series of random common ancestor or recombination events.
-% Recombination events occur at a rate determined by the amount of ancestral material and
-% the way in which it is distributed along a chromosome.
-% Let $L$ be the set of ancestors at a given time $t$. Recombination events
-% happen at rate $\rho \nu / (m - 1)$ where
-% \[
-% \nu = \sum_{x \in L}\left( \max_{(\ell, r, a) \in x}r
-%     - \min_{(\ell, r, a) \in x}\ell - 1 \right)
-% \]
-% is the number of available `links' that may be broken. Thus, the rate of
-% recombination is determined by the left- and right-most extent of the
-% ancestral material carried by each lineage. At a recombination
-% event we choose one of these links uniformly and break it. Given a lineage
-% $x = [(\ell_j, r_j, a_j)]$ and a breakpoint $k$, we have two lineages
-% $x_1$ and $x_2$ such that FILL IN DETAILS
-
-% When $k = |L|$ lineages are present, common ancestor events
-% occur at rate $\binom{k}{2}$. In a common ancestor event, two lineages
-% are chosen uniformly at random and their ancestry segments merged.
-% If we have overlapping intervals of ancestry from the two lineages,
-% say, $(\ell, r, a_1)$ and $(\ell, r, a_2)$, a
-% \emph{coalescence} occurs and ancestor represented by the current event
-% will be present as a node (at least) in the marginal trees covering
-% the interval $[\ell, r)$. The result of this coalescence is a segment
-% $(\ell, r, a_1 + a_2)$, and if $a_1 + a_2 < n$ it is included in the
-% ancestry for the new lineage. Otherwise, if $a_1 + a_2 = n$ we know that
-% we have found the most recent common ancestor of all samples in
-% the interval $[\ell, r)$ and so we do not need to simulate its history any further.
-% Nonoverlapping intervals of ancestry from the two lineages are included
-% in the resulting lineage without changes. Eventually, as the process continues,
-% we find resultant lineages in which all segments have fully coalescenced,
-% and so the number of extant lineages gradually dwindles down to zero.
-
 Several features of Fig 3B are worthy of note. First, since node \textsf{j} is not on
@@
 % to sections that have coalesced, or we won't have a complete ARG.
-
-% The state-space of the Griffiths process is much simpler than Hudson's algorithm,
-% which greatly facilitates mathematical reasoning. This simplicity comes at a
-% substantial cost, however, if we wish to use it as a practical means of
-% simulating recombinant ancestry. The big ARG is vast, and any realisation
-% for even moderate levels of recombination is far too large to be of practical
-% use. The number of events in the big ARG all the way back to the GMRCA
-% is $O(e^\rho)$~\citep{griffiths1997ancestral}, whereas the number
-% of events required to simulate the little ARG is
-% $O(\rho^2)$~\citep{hein2004gene,baumdicker2021efficient}.
-% This disparity in the number of events in the two formulations is
-% because the majority of the events that occur in the big ARG do
-% not affect the genetic ancestry of the sample in any way. Recombination
-% events that occur outside of ancestral material do not have any bearing
-% on the ancestry of the sample, and so the structure is hugely redundant.
-% As~\cite{wiuf1999recombination} note,
-% ``an `ancestral' sequence in the birth and death process
-% need not have any genetic material in common with a
-% sequence descended from it.''
-% Full quote:
-% This process simplifies mathematics on the account that the notion of an
-% ancestor will have a less restrictive meaning than usual:
-% An ``ancestral'' sequence in the birth and death process
-% need not have any genetic material in common with a
-% sequence descended from it.
 
@@
 % genome that is ancestral to the sample.
-
-\begin{figure}[ht]
-\centering
-% FIXME this is a quick nasty hack to make the figure a bit smaller and
-% prevent flushing all the other figures to the end of the document.
-\scalebox{0.5}{
-\begin{tabular}{cc}
-\begin{tikzpicture}
-	\node [anchor=north west] at (-3.5,9.3) {\LARGE \textbf{A}};
-	\draw (-0.4, -0.2) -- (0.4, -0.2) -- (0.4, -0.3) -- (-0.4, -0.3) -- (-0.4, -0.2);
-	\draw (-0.1, -0.1) -- (-0.1, -0.4);
-	\node [label=below:{\small $v$}] at (-0.1, -0.2) {};
-	\draw (3.6, -0.2) -- (4.4, -0.2) -- (4.4, -0.3) -- (3.6, -0.3) -- (3.6, -0.2);
-	\draw (5.6, -0.2) -- (6.4, -0.2) -- (6.4, -0.3) -- (5.6, -0.3) -- (5.6, -0.2);
-
-	\draw (4,0) -- (4, 1.2) -- (6, 1.2) -- (6,0);
-	\draw (4.6, 1) -- (5.4, 1) -- (5.4, 0.9) -- (4.6, 0.9) -- (4.6, 1);
-	\draw (5.1, 1.1) -- (5.1, 0.8);
-	\node [label=below:{\small $w$}] at (5.1, 1) {};
-
-	\draw (0, 0) -- (0, 1.7) -- (-1,1.7) -- (1,1.7);
-	\draw (-1.4, 1.5) -- (-1.1, 1.5) -- (-1.1, 1.4) -- (-1.4, 1.4) -- (-1.4, 1.5);
-	\draw (-1.1, 1.45) -- (-0.6, 1.45);
-	\draw (0.6, 1.45) -- (0.9, 1.45);
-	\draw (0.9, 1.5) -- (1.4, 1.5) -- (1.4, 1.4) -- (0.9, 1.4) -- (0.9, 1.5);
-	\draw (1.2, 1.6) -- (1.2, 1.3);
-	\node [label=below:{\small $x$}] at (1.2, 1.5) {};
-
-	\draw (5, 1.2) -- (5, 2.4) -- (4,2.4) -- (6,2.4);
-	\draw (3.6, 2.2) -- (4.1, 2.2) -- (4.1, 2.1) -- (3.6, 2.1) -- (3.6, 2.2);
-	\draw (4.1, 2.15) -- (4.4, 2.15);
-	\draw (5.6, 2.15) -- (6.1, 2.15);
-	\draw (6.1, 2.2) -- (6.4, 2.2) -- (6.4, 2.1) -- (6.1, 2.1) -- (6.1, 2.2);
-
-	\draw (1, 1.7) -- (1, 3.1) -- (0,3.1) -- (2,3.1);
-	\draw (-0.4, 2.85) -- (0.2, 2.85);
-	\draw (0.2, 2.9) -- (0.4, 2.9) -- (0.4, 2.8) -- (0.2, 2.8) -- (0.2, 2.9);
-	\draw (1.6, 2.85) -- (1.9, 2.85);
-	\draw (1.9, 2.9) -- (2.2, 2.9) -- (2.2, 2.8) -- (1.9, 2.8) -- (1.9, 2.9);
-	\draw (2.2, 2.85) -- (2.4, 2.85);
-
-	\draw (-1, 1.7) -- (-1, 4.5) -- (0,4.5) -- (0,3.1);
-	\node [scale=0.2,label=above left:{\small \textsf{a}}] at (-0.5,4.5) {a};
-	\draw (-0.9, 4.3) -- (-0.6, 4.3) -- (-0.6, 4.2) -- (-0.9, 4.2) -- (-0.9, 4.3);
-	\draw (-0.6, 4.25) -- (-0.3, 4.25);
-	\draw (-0.3, 4.3) -- (-0.1, 4.3) -- (-0.1, 4.2) -- (-0.3, 4.2) -- (-0.3, 4.3);
-	\draw (-0.5, 4.4) -- (-0.5, 4.1);
-	\node [label=below:{\small $y$}] at (-0.5, 4.3) {};
-
-	\draw (4,2.4) -- (4,5.4) -- (2,5.4) -- (2,3.1);
-	\node [scale=0.2,label=above left:{\small \textsf{b}}] at (3,5.4) {b};
-	\draw (2.6, 5.2) -- (2.9, 5.2) -- (2.9, 5.1) -- (2.6, 5.1) -- (2.6, 5.2);
-	\draw (2.9, 5.15) -- (3.4, 5.15);
-
-	\draw (-0.5, 4.5) -- (-0.5, 6) -- (-1.5,6) -- (0.5,6);
-	\draw (-1.9, 5.75) -- (-1.3, 5.75);
-	\draw (-1.3, 5.8) -- (-1.1, 5.8) -- (-1.1, 5.7) -- (-1.3, 5.7) -- (-1.3, 5.8);
-	\draw (0.1, 5.8) -- (0.4, 5.8) -- (0.4, 5.7) -- (0.1, 5.7) -- (0.1, 5.8);
-	\draw (0.4, 5.75) -- (0.9, 5.75);
-
-	\draw (0.5, 6) -- (0.5, 7.2) -- (3,7.2) -- (3,5.4);
-	\node [scale=0.2,label=above:{\small \textsf{c}}] at (1.75,7.2) {c};
-	\draw (1.35, 6.95) -- (2.15, 6.95);
-
-	\draw (-1.5, 6) -- (-1.5, 8);
-	\draw (6, 2.4) -- (6, 8);
-
-	% Edge annotations above each event
-	\node [label=left:{\small$\{(0,m,1)\}$}] at (0.2,0.6) {};
-	\node [label=left:{\small$\{(0,m,1)\}$}] at (4.2,0.4) {};
-	\node [label=right:{\small$\{(0,m,1)\}$}] at (5.8,0.4) {};
-	\node [label=left:{\small$\{(0,m,2)\}$}] at (5.2,1.6) {};
-	\node [label=left:{\small$\{(0,v,1)\}$}] at (-0.8,2.3) {};
-	\node [label=right:{\small$\{(v,m,1)\}$}] at (0.8,2.1) {};
-	\node [label=right:{\small$\{(0,w,2)\}$}] at (3.8,3.9) {};
-	\node [label=left:{\small$\{(w,m,2)\}$}] at (6.2,5) {};
-	\node [label=right:{\small$\{(x,m,1)\}$}] at (-0.2,3.8) {};
-	\node [label=right:{\small$\{(v,x,1)\}$}] at (1.8,4.1) {};
-	\node [label=left:{\small$\{(0,v,1), (x,m,1)\}$}] at (-0.3,5.2) {};
-	\node [label=right:{\small$\{(0,v,2)\}$}] at (2.8,6.4) {};
-	\node [label=left:{\small$\{(x,m,1)\}$}] at (-1.3,7) {};
-	\node [label=right:{\small$\{(0,v,1)\}$}] at (0.3,6.4) {};
-
-	% Dashed lines for start and end times
-	\draw[dashed] (-2, 0) -- (11.1, 0);
-	\node [label=left:{$t = 0$}] at (-2,0) {};
-	\draw[dashed] (-2, 8) -- (11.1, 8);
-	\node [label=left:{$t = T$}] at (-2,8) {};
-
-	% Numbers of extant ancestors and links, from top to bottom
-	\node[label=right:{CA \; RE}] at (7.5, 8.2) {};
-	\node[label=right:{$\binom{2}{2}$ \; $\frac{2 m - x - w - 2}{m - 1} \rho$}] at (7.5, 7.6) {};
-	\node[label=right:{$\binom{4}{2}$ \; $\frac{2 m + 2 v - x - w - 3}{m - 1} \rho$}] at (7.5, 6.6) {};
-	\node[label=right:{$\binom{3}{2}$ \; $\frac{2 m + v - w - 3}{m - 1} \rho$}] at (7.5, 5.7) {};
-	\node[label=right:{$\binom{4}{2}$ \; $\frac{2 m + x - v - 4}{m - 1} \rho$}] at (7.5, 4.95) {};
-	\node[label=right:{$\binom{5}{2}$ \; $\frac{2 m - 5}{m - 1} \rho$}] at (7.5, 3.8) {};
-	\node[label=right:{$\binom{4}{2}$ \; $\frac{2 m - 4}{m -1} \rho$}] at (7.5, 2.75) {};
-	\node[label=right:{$\binom{3}{2}$ \; $\frac{2 m - 3}{m - 1} \rho$}] at (7.5, 2.05) {};
-	\node[label=right:{$\binom{2}{2}$ \; $2 \rho$}] at (7.5, 1.45) {};
-	\node[label=right:{$\binom{3}{2}$ \; $3 \rho$}] at (7.5, 0.6) {};
-
-	% Gray dashed lines to visually separate holding times
-	\draw[color=gray, dashed] (7.5, 1.2) -- (11.1, 1.2);
-	\draw[color=gray, dashed] (7.5, 1.7) -- (11.1, 1.7);
-	\draw[color=gray, dashed] (7.5, 2.4) -- (11.1, 2.4);
-	\draw[color=gray, dashed] (7.5, 3.1) -- (11.1, 3.1);
-	\draw[color=gray, dashed] (7.5, 4.5) -- (11.1, 4.5);
-	\draw[color=gray, dashed] (7.5, 5.4) -- (11.1, 5.4);
-	\draw[color=gray, dashed] (7.5, 6) -- (11.1, 6);
-	\draw[color=gray, dashed] (7.5, 7.2) -- (11.1, 7.2);
-\end{tikzpicture}
-&
-\begin{tikzpicture}
-	\node [anchor=north west] at (-3.5,9.3) {\LARGE \textbf{B}};
-	\draw (-0.4, -0.2) -- (0.4, -0.2) -- (0.4, -0.3) -- (-0.4, -0.3) -- (-0.4, -0.2);
-	\draw (-0.1, -0.1) -- (-0.1, -0.4);
-	\node [label=below:{\small $v$}] at (-0.1, -0.2) {};
-	\draw (3.6, -0.2) -- (4.4, -0.2) -- (4.4, -0.3) -- (3.6, -0.3) -- (3.6, -0.2);
-	\draw (5.6, -0.2) -- (6.4, -0.2) -- (6.4, -0.3) -- (5.6, -0.3) -- (5.6, -0.2);
-
-	\draw (4,0) -- (4, 1.2) -- (6, 1.2) -- (6,0);
-	\draw (4.6, 1) -- (5.4, 1) -- (5.4, 0.9) -- (4.6, 0.9) -- (4.6, 1);
-	\draw (5.1, 1.1) -- (5.1, 0.8);
-	\node [label=below:{\small $w$}] at (5.1, 1) {};
-
-	\draw (0, 0) -- (0, 1.7) -- (-1,1.7) -- (1,1.7);
-	\draw (-1.4, 1.5) -- (-1.1, 1.5) -- (-1.1, 1.4) -- (-1.4, 1.4) -- (-1.4, 1.5);
-	\draw (-1.1, 1.45) -- (-0.6, 1.45);
-	\draw (0.6, 1.45) -- (0.9, 1.45);
-	\draw (0.9, 1.5) -- (1.4, 1.5) -- (1.4, 1.4) -- (0.9, 1.4) -- (0.9, 1.5);
-	\draw (1.2, 1.6) -- (1.2, 1.3);
-	\node [label=below:{\small $x$}] at (1.2, 1.5) {};
-
-	\draw (5, 1.2) -- (5, 2.4) -- (4,2.4) -- (6,2.4);
-	\draw (3.6, 2.2) -- (4.1, 2.2) -- (4.1, 2.1) -- (3.6, 2.1) -- (3.6, 2.2);
-	\draw (4.1, 2.15) -- (4.4, 2.15);
-	\draw (5.6, 2.15) -- (6.1, 2.15);
-	\draw (6.1, 2.2) -- (6.4, 2.2) -- (6.4, 2.1) -- (6.1, 2.1) -- (6.1, 2.2);
-	\draw [color=red](5.8, 2.3) -- (5.8, 2.0);
-	\node [label={[red]below:{\small $z$}}] at (5.8, 2.1) {};
-
-	\draw (1, 1.7) -- (1, 3.1) -- (0,3.1) -- (2,3.1);
-	\draw (-0.4, 2.85) -- (0.2, 2.85);
-	\draw (0.2, 2.9) -- (0.4, 2.9) -- (0.4, 2.8) -- (0.2, 2.8) -- (0.2, 2.9);
-	\draw (1.6, 2.85) -- (1.9, 2.85);
-	\draw (1.9, 2.9) -- (2.2, 2.9) -- (2.2, 2.8) -- (1.9, 2.8) -- (1.9, 2.9);
-	\draw (2.2, 2.85) -- (2.4, 2.85);
-
-	\draw (-1, 1.7) -- (-1, 4.5) -- (0,4.5) -- (0,3.1);
-	\draw (-0.9, 4.3) -- (-0.6, 4.3) -- (-0.6, 4.2) -- (-0.9, 4.2) -- (-0.9, 4.3);
-	\draw (-0.6, 4.25) -- (-0.3, 4.25);
-	\draw (-0.3, 4.3) -- (-0.1, 4.3) -- (-0.1, 4.2) -- (-0.3, 4.2) -- (-0.3, 4.3);
-	\draw (-0.5, 4.4) -- (-0.5, 4.1);
-	\node [label=below:{\small $y$}] at (-0.5, 4.3) {};
-
-	\draw (6, 2.4) -- (6, 3.8) -- (7, 3.8);
-	\draw [color=red](6, 3.8) -- (5, 3.8);
-	\draw (4.6, 3.55) -- (5.4, 3.55);
-	\draw (6.6, 3.55) -- (7.1, 3.55);
-	\draw (7.1, 3.6) -- (7.4, 3.6) -- (7.4, 3.5) -- (7.1, 3.5) -- (7.1, 3.6);
-
-	\draw (4,2.4) -- (4,5.4) -- (2,5.4) -- (2,3.1);
-	\draw (2.6, 5.2) -- (2.9, 5.2) -- (2.9, 5.1) -- (2.6, 5.1) -- (2.6, 5.2);
-	\draw (2.9, 5.15) -- (3.4, 5.15);
-
-	\draw (-0.5, 4.5) -- (-0.5, 6) -- (-1.5,6) -- (0.5,6);
-	\draw (-1.9, 5.75) -- (-1.3, 5.75);
-	\draw (-1.3, 5.8) -- (-1.1, 5.8) -- (-1.1, 5.7) -- (-1.3, 5.7) -- (-1.3, 5.8);
-	\draw (0.1, 5.8) -- (0.4, 5.8) -- (0.4, 5.7) -- (0.1, 5.7) -- (0.1, 5.8);
-	\draw (0.4, 5.75) -- (0.9, 5.75);
-
-	\draw [color=red](5, 3.8) -- (5, 6.6) -- (4, 6.6);
-	\draw (4,6.6) -- (3,6.6) -- (3, 5.4);
-	\draw (3.6, 6.4) -- (3.9, 6.4) -- (3.9, 6.3) -- (3.6, 6.3) -- (3.6, 6.4);
-	\draw (3.9, 6.35) -- (4.4, 6.35);
-
-	\draw (0.5, 6) -- (0.5, 7.2) -- (4,7.2) -- (4,6.6);
-	\draw (1.85, 6.95) -- (2.65, 6.95);
-
-	\draw (-1.5, 6) -- (-1.5, 8);
-	\draw [color=red](2.25, 7.2) -- (2.25, 8);
-	\draw (7, 3.8) -- (7, 8);
-
-	% Dashed lines for start and end times
-	\draw[dashed] (-2, 0) -- (11, 0);
-	\node [label=left:{$t = 0$}] at (-2,0) {};
-	\draw[dashed] (-2, 8) -- (11, 8);
-	\node [label=left:{$t = T$}] at (-2,8) {};
-
-	% Numbers of extant ancestors and links, from top to bottom
-	\node[label=right:{CA \; RE}] at (7.5, 8.2) {};
-	\node[label=right:{$\binom{3}{2}$ \; $3 \rho$}] at (7.5, 7.6) {};
-	\node[label=right:{$\binom{4}{2}$ \; $4 \rho$}] at (7.5, 6.9) {};
-	\node[label=right:{$\binom{5}{2}$ \; $5 \rho$}] at (7.5, 6.3) {};
-	\node[label=right:{$\binom{4}{2}$ \; $4 \rho$}] at (7.5, 5.7) {};
-	\node[label=right:{$\binom{5}{2}$ \; $5 \rho$}] at (7.5, 4.95) {};
-	\node[label=right:{$\binom{6}{2}$ \; $6 \rho$}] at (7.5, 4.15) {};
-	\node[label=right:{$\binom{5}{2}$ \; $5 \rho$}] at (7.5, 3.45) {};
-	\node[label=right:{$\binom{4}{2}$ \; $4 \rho$}] at (7.5, 2.75) {};
-	\node[label=right:{$\binom{3}{2}$ \; $3 \rho$}] at (7.5, 2.05) {};
-	\node[label=right:{$\binom{2}{2}$ \; $2 \rho$}] at (7.5, 1.45) {};
-	\node[label=right:{$\binom{3}{2}$ \; $3 \rho$}] at (7.5, 0.6) {};
-
-	% Gray dashed lines to visually separate holding times
-	\draw[color=gray, dashed] (7.5, 1.2) -- (11, 1.2);
-	\draw[color=gray, dashed] (7.5, 1.7) -- (11, 1.7);
-	\draw[color=gray, dashed] (7.5, 2.4) -- (11, 2.4);
-	\draw[color=gray, dashed] (7.5, 3.1) -- (11, 3.1);
-	\draw[color=gray, dashed] (7.5, 3.8) -- (11, 3.8);
-	\draw[color=gray, dashed] (7.5, 4.5) -- (11, 4.5);
-	\draw[color=gray, dashed] (7.5, 5.4) -- (11, 5.4);
-	\draw[color=gray, dashed] (7.5, 6) -- (11, 6);
-	\draw[color=gray, dashed] (7.5, 6.6) -- (11, 6.6);
-	\draw[color=gray, dashed] (7.5, 7.2) -- (11, 7.2);
-\end{tikzpicture}
-\end{tabular}
-}
-\caption{(A)
-A realisation of the graph traversed by Hudson's algorithm started from a
-sample of three chromosomes of length $m$ at time $t = 0$, and
-propagated until time $T$. The MRCA on the genetic interval $[v, w)$ is reached
-at event \textsf{b}, while that on $[0, v)$ is reached at event \textsf{c}.
-The non-ancestral segment $[v, w)$ above
-A contributes to the rate of effective recombinations because it
-is trapped between ancestral segments. The two columns titled CA and RE
-are the respective rates of mergers and recombinations when
-the recombination rate is $\rho$.
-(B) A corresponding realisation of a big ARG, which augments Hudson's algorithm
-by tracking nonancestral lineages. The result is a simpler state space and
-dynamics, at the cost of extra nodes and edges, highlighted in red, which do
-not affect the local tree at any site.}
-\label{hudson_vs_bigARG}
-\end{figure}
 
@@
 
+\item Throughout, we considered ARGs defined for a single chromosome comprising
+of a fixed number of discrete sites. Our argument applies equally well to ARGs
+defined for several non-homologous chromosomes \citep{fearnhead2003ancestral,
+koskela2019robust} and to idealised, continuous chromosomes.
+
 \item The gARG structure as described here is sufficiently expressive
@@
 
+\setcounter{secnumdepth}{2} % Print out appendix section numbers
+
+\appendix
+\section*{Appendix}
+
+\subsection*{The Big and Little ARG}
+\label{app-big-and-little-arg}
+
+\begin{figure}[ht]
+\centering
+% FIXME this is a quick nasty hack to make the figure a bit smaller and
+% prevent flushing all the other figures to the end of the document.
+\scalebox{0.5}{
+\begin{tabular}{cc}
+\begin{tikzpicture}
+	\node [anchor=north west] at (-3.5,9.3) {\LARGE \textbf{A}};
+	\draw (-0.4, -0.2) -- (0.4, -0.2) -- (0.4, -0.3) -- (-0.4, -0.3) -- (-0.4, -0.2);
+	\draw (-0.1, -0.1) -- (-0.1, -0.4);
+	\node [label=below:{\small $v$}] at (-0.1, -0.2) {};
+	\draw (3.6, -0.2) -- (4.4, -0.2) -- (4.4, -0.3) -- (3.6, -0.3) -- (3.6, -0.2);
+	\draw (5.6, -0.2) -- (6.4, -0.2) -- (6.4, -0.3) -- (5.6, -0.3) -- (5.6, -0.2);
+
+	\draw (4,0) -- (4, 1.2) -- (6, 1.2) -- (6,0);
+	\draw (4.6, 1) -- (5.4, 1) -- (5.4, 0.9) -- (4.6, 0.9) -- (4.6, 1);
+	\draw (5.1, 1.1) -- (5.1, 0.8);
+	\node [label=below:{\small $w$}] at (5.1, 1) {};
+
+	\draw (0, 0) -- (0, 1.7) -- (-1,1.7) -- (1,1.7);
+	\draw (-1.4, 1.5) -- (-1.1, 1.5) -- (-1.1, 1.4) -- (-1.4, 1.4) -- (-1.4, 1.5);
+	\draw (-1.1, 1.45) -- (-0.6, 1.45);
+	\draw (0.6, 1.45) -- (0.9, 1.45);
+	\draw (0.9, 1.5) -- (1.4, 1.5) -- (1.4, 1.4) -- (0.9, 1.4) -- (0.9, 1.5);
+	\draw (1.2, 1.6) -- (1.2, 1.3);
+	\node [label=below:{\small $x$}] at (1.2, 1.5) {};
+
+	\draw (5, 1.2) -- (5, 2.4) -- (4,2.4) -- (6,2.4);
+	\draw (3.6, 2.2) -- (4.1, 2.2) -- (4.1, 2.1) -- (3.6, 2.1) -- (3.6, 2.2);
+	\draw (4.1, 2.15) -- (4.4, 2.15);
+	\draw (5.6, 2.15) -- (6.1, 2.15);
+	\draw (6.1, 2.2) -- (6.4, 2.2) -- (6.4, 2.1) -- (6.1, 2.1) -- (6.1, 2.2);
+
+	\draw (1, 1.7) -- (1, 3.1) -- (0,3.1) -- (2,3.1);
+	\draw (-0.4, 2.85) -- (0.2, 2.85);
+	\draw (0.2, 2.9) -- (0.4, 2.9) -- (0.4, 2.8) -- (0.2, 2.8) -- (0.2, 2.9);
+	\draw (1.6, 2.85) -- (1.9, 2.85);
+	\draw (1.9, 2.9) -- (2.2, 2.9) -- (2.2, 2.8) -- (1.9, 2.8) -- (1.9, 2.9);
+	\draw (2.2, 2.85) -- (2.4, 2.85);
+
+	\draw (-1, 1.7) -- (-1, 4.5) -- (0,4.5) -- (0,3.1);
+	\node [scale=0.2,label=above left:{\small \textsf{a}}] at (-0.5,4.5) {a};
+	\draw (-0.9, 4.3) -- (-0.6, 4.3) -- (-0.6, 4.2) -- (-0.9, 4.2) -- (-0.9, 4.3);
+	\draw (-0.6, 4.25) -- (-0.3, 4.25);
+	\draw (-0.3, 4.3) -- (-0.1, 4.3) -- (-0.1, 4.2) -- (-0.3, 4.2) -- (-0.3, 4.3);
+	\draw (-0.5, 4.4) -- (-0.5, 4.1);
+	\node [label=below:{\small $y$}] at (-0.5, 4.3) {};
+
+	\draw (4,2.4) -- (4,5.4) -- (2,5.4) -- (2,3.1);
+	\node [scale=0.2,label=above left:{\small \textsf{b}}] at (3,5.4) {b};
+	\draw (2.6, 5.2) -- (2.9, 5.2) -- (2.9, 5.1) -- (2.6, 5.1) -- (2.6, 5.2);
+	\draw (2.9, 5.15) -- (3.4, 5.15);
+
+	\draw (-0.5, 4.5) -- (-0.5, 6) -- (-1.5,6) -- (0.5,6);
+	\draw (-1.9, 5.75) -- (-1.3, 5.75);
+	\draw (-1.3, 5.8) -- (-1.1, 5.8) -- (-1.1, 5.7) -- (-1.3, 5.7) -- (-1.3, 5.8);
+	\draw (0.1, 5.8) -- (0.4, 5.8) -- (0.4, 5.7) -- (0.1, 5.7) -- (0.1, 5.8);
+	\draw (0.4, 5.75) -- (0.9, 5.75);
+
+	\draw (0.5, 6) -- (0.5, 7.2) -- (3,7.2) -- (3,5.4);
+	\node [scale=0.2,label=above:{\small \textsf{c}}] at (1.75,7.2) {c};
+	\draw (1.35, 6.95) -- (2.15, 6.95);
+
+	\draw (-1.5, 6) -- (-1.5, 8);
+	\draw (6, 2.4) -- (6, 8);
+
+	% Edge annotations above each event
+	\node [label=left:{\small$\{(0,m,1)\}$}] at (0.2,0.6) {};
+	\node [label=left:{\small$\{(0,m,1)\}$}] at (4.2,0.4) {};
+	\node [label=right:{\small$\{(0,m,1)\}$}] at (5.8,0.4) {};
+	\node [label=left:{\small$\{(0,m,2)\}$}] at (5.2,1.6) {};
+	\node [label=left:{\small$\{(0,v,1)\}$}] at (-0.8,2.3) {};
+	\node [label=right:{\small$\{(v,m,1)\}$}] at (0.8,2.1) {};
+	\node [label=right:{\small$\{(0,w,2)\}$}] at (3.8,3.9) {};
+	\node [label=left:{\small$\{(w,m,2)\}$}] at (6.2,5) {};
+	\node [label=right:{\small$\{(x,m,1)\}$}] at (-0.2,3.8) {};
+	\node [label=right:{\small$\{(v,x,1)\}$}] at (1.8,4.1) {};
+	\node [label=left:{\small$\{(0,v,1), (x,m,1)\}$}] at (-0.3,5.2) {};
+	\node [label=right:{\small$\{(0,v,2)\}$}] at (2.8,6.4) {};
+	\node [label=left:{\small$\{(x,m,1)\}$}] at (-1.3,7) {};
+	\node [label=right:{\small$\{(0,v,1)\}$}] at (0.3,6.4) {};
+
+	% Dashed lines for start and end times
+	\draw[dashed] (-2, 0) -- (11.1, 0);
+	\node [label=left:{$t = 0$}] at (-2,0) {};
+	\draw[dashed] (-2, 8) -- (11.1, 8);
+	\node [label=left:{$t = T$}] at (-2,8) {};
+
+	% Numbers of extant ancestors and links, from top to bottom
+	\node[label=right:{CA \; RE}] at (7.5, 8.2) {};
+	\node[label=right:{$\binom{2}{2}$ \; $\frac{2 m - x - w - 2}{m - 1} \rho$}] at (7.5, 7.6) {};
+	\node[label=right:{$\binom{4}{2}$ \; $\frac{2 m + 2 v - x - w - 3}{m - 1} \rho$}] at (7.5, 6.6) {};
+	\node[label=right:{$\binom{3}{2}$ \; $\frac{2 m + v - w - 3}{m - 1} \rho$}] at (7.5, 5.7) {};
+	\node[label=right:{$\binom{4}{2}$ \; $\frac{2 m + x - v - 4}{m - 1} \rho$}] at (7.5, 4.95) {};
+	\node[label=right:{$\binom{5}{2}$ \; $\frac{2 m - 5}{m - 1} \rho$}] at (7.5, 3.8) {};
+	\node[label=right:{$\binom{4}{2}$ \; $\frac{2 m - 4}{m -1} \rho$}] at (7.5, 2.75) {};
+	\node[label=right:{$\binom{3}{2}$ \; $\frac{2 m - 3}{m - 1} \rho$}] at (7.5, 2.05) {};
+	\node[label=right:{$\binom{2}{2}$ \; $2 \rho$}] at (7.5, 1.45) {};
+	\node[label=right:{$\binom{3}{2}$ \; $3 \rho$}] at (7.5, 0.6) {};
+
+	% Gray dashed lines to visually separate holding times
+	\draw[color=gray, dashed] (7.5, 1.2) -- (11.1, 1.2);
+	\draw[color=gray, dashed] (7.5, 1.7) -- (11.1, 1.7);
+	\draw[color=gray, dashed] (7.5, 2.4) -- (11.1, 2.4);
+	\draw[color=gray, dashed] (7.5, 3.1) -- (11.1, 3.1);
+	\draw[color=gray, dashed] (7.5, 4.5) -- (11.1, 4.5);
+	\draw[color=gray, dashed] (7.5, 5.4) -- (11.1, 5.4);
+	\draw[color=gray, dashed] (7.5, 6) -- (11.1, 6);
+	\draw[color=gray, dashed] (7.5, 7.2) -- (11.1, 7.2);
+\end{tikzpicture}
+&
+\begin{tikzpicture}
+	\node [anchor=north west] at (-3.5,9.3) {\LARGE \textbf{B}};
+	\draw (-0.4, -0.2) -- (0.4, -0.2) -- (0.4, -0.3) -- (-0.4, -0.3) -- (-0.4, -0.2);
+	\draw (-0.1, -0.1) -- (-0.1, -0.4);
+	\node [label=below:{\small $v$}] at (-0.1, -0.2) {};
+	\draw (3.6, -0.2) -- (4.4, -0.2) -- (4.4, -0.3) -- (3.6, -0.3) -- (3.6, -0.2);
+	\draw (5.6, -0.2) -- (6.4, -0.2) -- (6.4, -0.3) -- (5.6, -0.3) -- (5.6, -0.2);
+
+	\draw (4,0) -- (4, 1.2) -- (6, 1.2) -- (6,0);
+	\draw (4.6, 1) -- (5.4, 1) -- (5.4, 0.9) -- (4.6, 0.9) -- (4.6, 1);
+	\draw (5.1, 1.1) -- (5.1, 0.8);
+	\node [label=below:{\small $w$}] at (5.1, 1) {};
+
+	\draw (0, 0) -- (0, 1.7) -- (-1,1.7) -- (1,1.7);
+	\draw (-1.4, 1.5) -- (-1.1, 1.5) -- (-1.1, 1.4) -- (-1.4, 1.4) -- (-1.4, 1.5);
+	\draw (-1.1, 1.45) -- (-0.6, 1.45);
+	\draw (0.6, 1.45) -- (0.9, 1.45);
+	\draw (0.9, 1.5) -- (1.4, 1.5) -- (1.4, 1.4) -- (0.9, 1.4) -- (0.9, 1.5);
+	\draw (1.2, 1.6) -- (1.2, 1.3);
+	\node [label=below:{\small $x$}] at (1.2, 1.5) {};
+
+	\draw (5, 1.2) -- (5, 2.4) -- (4,2.4) -- (6,2.4);
+	\draw (3.6, 2.2) -- (4.1, 2.2) -- (4.1, 2.1) -- (3.6, 2.1) -- (3.6, 2.2);
+	\draw (4.1, 2.15) -- (4.4, 2.15);
+	\draw (5.6, 2.15) -- (6.1, 2.15);
+	\draw (6.1, 2.2) -- (6.4, 2.2) -- (6.4, 2.1) -- (6.1, 2.1) -- (6.1, 2.2);
+	\draw [color=red](5.8, 2.3) -- (5.8, 2.0);
+	\node [label={[red]below:{\small $z$}}] at (5.8, 2.1) {};
+
+	\draw (1, 1.7) -- (1, 3.1) -- (0,3.1) -- (2,3.1);
+	\draw (-0.4, 2.85) -- (0.2, 2.85);
+	\draw (0.2, 2.9) -- (0.4, 2.9) -- (0.4, 2.8) -- (0.2, 2.8) -- (0.2, 2.9);
+	\draw (1.6, 2.85) -- (1.9, 2.85);
+	\draw (1.9, 2.9) -- (2.2, 2.9) -- (2.2, 2.8) -- (1.9, 2.8) -- (1.9, 2.9);
+	\draw (2.2, 2.85) -- (2.4, 2.85);
+
+	\draw (-1, 1.7) -- (-1, 4.5) -- (0,4.5) -- (0,3.1);
+	\draw (-0.9, 4.3) -- (-0.6, 4.3) -- (-0.6, 4.2) -- (-0.9, 4.2) -- (-0.9, 4.3);
+	\draw (-0.6, 4.25) -- (-0.3, 4.25);
+	\draw (-0.3, 4.3) -- (-0.1, 4.3) -- (-0.1, 4.2) -- (-0.3, 4.2) -- (-0.3, 4.3);
+	\draw (-0.5, 4.4) -- (-0.5, 4.1);
+	\node [label=below:{\small $y$}] at (-0.5, 4.3) {};
+
+	\draw (6, 2.4) -- (6, 3.8) -- (7, 3.8);
+	\draw [color=red](6, 3.8) -- (5, 3.8);
+	\draw (4.6, 3.55) -- (5.4, 3.55);
+	\draw (6.6, 3.55) -- (7.1, 3.55);
+	\draw (7.1, 3.6) -- (7.4, 3.6) -- (7.4, 3.5) -- (7.1, 3.5) -- (7.1, 3.6);
+
+	\draw (4,2.4) -- (4,5.4) -- (2,5.4) -- (2,3.1);
+	\draw (2.6, 5.2) -- (2.9, 5.2) -- (2.9, 5.1) -- (2.6, 5.1) -- (2.6, 5.2);
+	\draw (2.9, 5.15) -- (3.4, 5.15);
+
+	\draw (-0.5, 4.5) -- (-0.5, 6) -- (-1.5,6) -- (0.5,6);
+	\draw (-1.9, 5.75) -- (-1.3, 5.75);
+	\draw (-1.3, 5.8) -- (-1.1, 5.8) -- (-1.1, 5.7) -- (-1.3, 5.7) -- (-1.3, 5.8);
+	\draw (0.1, 5.8) -- (0.4, 5.8) -- (0.4, 5.7) -- (0.1, 5.7) -- (0.1, 5.8);
+	\draw (0.4, 5.75) -- (0.9, 5.75);
+
+	\draw [color=red](5, 3.8) -- (5, 6.6) -- (4, 6.6);
+	\draw (4,6.6) -- (3,6.6) -- (3, 5.4);
+	\draw (3.6, 6.4) -- (3.9, 6.4) -- (3.9, 6.3) -- (3.6, 6.3) -- (3.6, 6.4);
+	\draw (3.9, 6.35) -- (4.4, 6.35);
+
+	\draw (0.5, 6) -- (0.5, 7.2) -- (4,7.2) -- (4,6.6);
+	\draw (1.85, 6.95) -- (2.65, 6.95);
+
+	\draw (-1.5, 6) -- (-1.5, 8);
+	\draw [color=red](2.25, 7.2) -- (2.25, 8);
+	\draw (7, 3.8) -- (7, 8);
+
+	% Dashed lines for start and end times
+	\draw[dashed] (-2, 0) -- (11, 0);
+	\node [label=left:{$t = 0$}] at (-2,0) {};
+	\draw[dashed] (-2, 8) -- (11, 8);
+	\node [label=left:{$t = T$}] at (-2,8) {};
+
+	% Numbers of extant ancestors and links, from top to bottom
+	\node[label=right:{CA \; RE}] at (7.5, 8.2) {};
+	\node[label=right:{$\binom{3}{2}$ \; $3 \rho$}] at (7.5, 7.6) {};
+	\node[label=right:{$\binom{4}{2}$ \; $4 \rho$}] at (7.5, 6.9) {};
+	\node[label=right:{$\binom{5}{2}$ \; $5 \rho$}] at (7.5, 6.3) {};
+	\node[label=right:{$\binom{4}{2}$ \; $4 \rho$}] at (7.5, 5.7) {};
+	\node[label=right:{$\binom{5}{2}$ \; $5 \rho$}] at (7.5, 4.95) {};
+	\node[label=right:{$\binom{6}{2}$ \; $6 \rho$}] at (7.5, 4.15) {};
+	\node[label=right:{$\binom{5}{2}$ \; $5 \rho$}] at (7.5, 3.45) {};
+	\node[label=right:{$\binom{4}{2}$ \; $4 \rho$}] at (7.5, 2.75) {};
+	\node[label=right:{$\binom{3}{2}$ \; $3 \rho$}] at (7.5, 2.05) {};
+	\node[label=right:{$\binom{2}{2}$ \; $2 \rho$}] at (7.5, 1.45) {};
+	\node[label=right:{$\binom{3}{2}$ \; $3 \rho$}] at (7.5, 0.6) {};
+
+	% Gray dashed lines to visually separate holding times
+	\draw[color=gray, dashed] (7.5, 1.2) -- (11, 1.2);
+	\draw[color=gray, dashed] (7.5, 1.7) -- (11, 1.7);
+	\draw[color=gray, dashed] (7.5, 2.4) -- (11, 2.4);
+	\draw[color=gray, dashed] (7.5, 3.1) -- (11, 3.1);
+	\draw[color=gray, dashed] (7.5, 3.8) -- (11, 3.8);
+	\draw[color=gray, dashed] (7.5, 4.5) -- (11, 4.5);
+	\draw[color=gray, dashed] (7.5, 5.4) -- (11, 5.4);
+	\draw[color=gray, dashed] (7.5, 6) -- (11, 6);
+	\draw[color=gray, dashed] (7.5, 6.6) -- (11, 6.6);
+	\draw[color=gray, dashed] (7.5, 7.2) -- (11, 7.2);
+\end{tikzpicture}
+\end{tabular}
+}
+\caption{(A)
+A realisation of the graph traversed by Hudson's algorithm started from a
+sample of three chromosomes of length $m$ at time $t = 0$, and
+propagated until time $T$. The MRCA on the genetic interval $[v, w)$ is reached
+at event \textsf{b}, while that on $[0, v)$ is reached at event \textsf{c}.
+The non-ancestral segment $[v, w)$ above
+A contributes to the rate of effective recombinations because it
+is trapped between ancestral segments. The two columns titled CA and RE
+are the respective rates of mergers and recombinations when
+the recombination rate is $\rho$.
+(B) A corresponding realisation of a big ARG, which augments Hudson's algorithm
+by tracking nonancestral lineages. The result is a simpler state space and
+dynamics, at the cost of extra nodes and edges, highlighted in red, which do
+not affect the local tree at any site.}
+\label{hudson_vs_bigARG}
+\end{figure}
+
+[JK: this is text roughly related to big vs little ARG taken from various
+iterations]
+
+In the Griffiths formulation, each edge in the graph corresponds to an extant
+lineage and nodes are events in the process (the initial $n$ leaf nodes are
+``sampling'' events). Common ancestor events occur at rate $\binom{k}{2}$ when there
+are $k$ lineages present. We choose two lineages (edges) uniformly, and merge them
+into a common ancestor lineage. Recombination events happen at a rate of
+$k \rho$. Here, we choose a lineage (edge) uniformly, and a
+breakpoint $0 < x < m$ uniformly on its genome. We terminate the edge at a
+node, record the breakpoint and start two new edges from this node. The process
+then continues until there is only one lineage left (the Grand Most Recent
+Common Ancestor, GMRCA), which is guaranteed to
+happen in finite time because of the linear vs quadratic rates of branching
+and coalescing. Later, we will show how the graph structure and breakpoints
+associated with
+recombination nodes provides sufficient information to later recover the marginal
+trees and calculate sampling probabilities. However, this process traces out
+a larger graph (the ``big'' ARG) than the Hudson algorithm (the ``little'' ARG; see the ``\nameref{Ancestry_resolution}'' section),
+and is consequently considerably less efficient \citep{wiuf1999ancestry}.
+
+The state-space of the Griffiths process is much simpler than Hudson's algorithm,
+which greatly facilitates mathematical reasoning. This simplicity comes at a
+substantial cost, however, if we wish to use it as a practical means of
+simulating recombinant ancestry. The big ARG is vast, and any realisation
+for even moderate levels of recombination is far too large to be of practical
+use. The number of events in the big ARG all the way back to the GMRCA
+is $O(e^\rho)$~\citep{griffiths1997ancestral}, whereas the number
+of events required to simulate the little ARG is
+$O(\rho^2)$~\citep{hein2004gene,baumdicker2021efficient}.
+This disparity in the number of events in the two formulations is
+because the majority of the events that occur in the big ARG do
+not affect the genetic ancestry of the sample in any way. Recombination
+events that occur outside of ancestral material do not have any bearing
+on the ancestry of the sample, and so the structure is hugely redundant.
+As~\cite{wiuf1999recombination} note,
+``an `ancestral' sequence in the birth and death process
+need not have any genetic material in common with a
+sequence descended from it.''
+
+% Full quote:
+% This process simplifies mathematics on the account that the notion of an
+% ancestor will have a less restrictive meaning than usual:
+% An ``ancestral'' sequence in the birth and death process
+% need not have any genetic material in common with a
+% sequence descended from it.
+
+% % This is a discussion of Hudson's alg left over from earlier version. We
+% % might wish to adapt some of this when defining the Little ARG.
+
+% Each lineage consists of a list of
+% disjoint ancestry segments $(\ell, r, a)$, where
+% $[\ell, r)$ is a half-closed genomic interval and $a$ is an integer
+% tracking the number of samples to which the lineage is ancestral over that interval.
+% (We also usually track the tree node associated with each segment, but
+% that is not important for our purposes here so we omit it.)
+% If we have $n$ samples and a genome of length $m$, the process begins with $n$ lineages
+% of the form $\{(0, m, 1)\}$. The process then works backwards in time from
+% the present day as a series of random common ancestor or recombination events.
+% Recombination events occur at a rate determined by the amount of ancestral material and
+% the way in which it is distributed along a chromosome.
+% Let $L$ be the set of ancestors at a given time $t$. Recombination events
+% happen at rate $\rho \nu / (m - 1)$ where
+% \[
+% \nu = \sum_{x \in L}\left( \max_{(\ell, r, a) \in x}r
+%     - \min_{(\ell, r, a) \in x}\ell - 1 \right)
+% \]
+% is the number of available `links' that may be broken. Thus, the rate of
+% recombination is determined by the left- and right-most extent of the
+% ancestral material carried by each lineage. At a recombination
+% event we choose one of these links uniformly and break it. Given a lineage
+% $x = [(\ell_j, r_j, a_j)]$ and a breakpoint $k$, we have two lineages
+% $x_1$ and $x_2$ such that FILL IN DETAILS
+
+% When $k = |L|$ lineages are present, common ancestor events
+% occur at rate $\binom{k}{2}$. In a common ancestor event, two lineages
+% are chosen uniformly at random and their ancestry segments merged.
+% If we have overlapping intervals of ancestry from the two lineages,
+% say, $(\ell, r, a_1)$ and $(\ell, r, a_2)$, a
+% \emph{coalescence} occurs and ancestor represented by the current event
+% will be present as a node (at least) in the marginal trees covering
+% the interval $[\ell, r)$. The result of this coalescence is a segment
+% $(\ell, r, a_1 + a_2)$, and if $a_1 + a_2 < n$ it is included in the
+% ancestry for the new lineage. Otherwise, if $a_1 + a_2 = n$ we know that
+% we have found the most recent common ancestor of all samples in
+% the interval $[\ell, r)$ and so we do not need to simulate its history any further.
+% Nonoverlapping intervals of ancestry from the two lineages are included
+% in the resulting lineage without changes. Eventually, as the process continues,
+% we find resultant lineages in which all segments have fully coalescenced,
+% and so the number of extant lineages gradually dwindles down to zero.
+
 \end{document}

### 2022-08-01 paper.tex
Update on parsimony, following review

@@
 ARG realisations~\citep{song2003parsimonious,song2005efficient,lyngso2005minimum}.
+It is important to note that such recombination-parsimony ARGs
+have quite different properties from realisations of the
+(previously synonymous) ARG \emph{process}. Firstly, branch lengths are
+not typically estimated, and node times are required
+order to compute a likelihood under the coalescent with recombination
+(see section XXX). Secondly, even if branch lengths were estimated,
+the realisations would have a very small likelihood since
+the coalescent with recombination is often highly \emph{un}parsimonious
+in terms of recombination events for non-trivial scaled recombination rate.
 The previously close correspondance
 between the ARG as a mathematical representation of the coalescent process
-and as concrete realisations of ancestry was now broken,
-since such parsimonious ARGs are extremely unlikely under the
-% JK: I assume this is true; any citations to back it up I wonder?
-coalescent with recombination.
+and as concrete realisations of ancestry was now therefore broken.
 \cite{minichiello2006mapping} made this separation of the ARG as a

### 2022-08-01 paper.tex
Draft of eARG section

@@
 \usetikzlibrary{calc,positioning}
-\usepackage[hidelinks]{hyperref}
+\usepackage{url}
+% JK: turning this off for the moment as I keep clicking through on links
+% to the bibliography while reading the text and it's intensely annoying.
+% Can reinstate when we're ready to preprint
+% \usepackage[hidelinks]{hyperref}
+
+\newcommand{\noderef}[1]{\textsf{#1}}
+% \newcommand{\msprime}[0]{\texttt{msprime}}
 
@@
 % of ancestry''.
+\item Despite its status as ``the ideal data structure for
+population genomic analysis''~\citep{rasmussen2014genome}, there has been
+little discussion of its computational properties. We discuss the
+relationship between the classical concept of an ARG and the
+recently introduced succinct tree sequence data structure.
 \end{enumerate}
@@
 Similarly, the Ancestral Gene Transfer Graph
+
 models horizontal gene transfer~\citep{baumdicker2014infinitely}
@@
 
-\subsection*{Node versus edge annotation}
-
-% What do we mean by an EARG, informally?
-\citet{griffiths1996ancestral} introduced an encoding for the ARG structure, which
-is elegant and mathematically economical. In this encoding, the
-fundamental units correspond directly to \emph{events}
-that occur in the underlying stochastic process they were modelling.
-We will refer to this ARG encoding as an ``event ARG'', or eARG:
-an example is given in Fig.~\ref{fig-arg-data-structure}B.
-Since each node in the graph corresponds to an event, the graph
-contains both recombination nodes (node \textsf{d}) and common ancestor
-nodes (nodes \textsf{e}, \textsf{f}, and \textsf{g}) corresponding to their
-respective event types. Edges represent lineages,
-and those lineages are modified by the events that they pass through.
+In keeping with its original derivation in terms of a stochastic
+process, the classical definition of an ARG data structure is
+in terms of \emph{events}. Nodes represent
+common ancestor or recombination events that occured in the
+history of some set of sample genomes, and edges represent ancestral
+lineages~\citep{griffiths1996ancestral}.
+In order to distinguish this form of ARG from what we will be discussing
+in later sections we call it an ``event ARG``, or eARG for brevity.
+An example of this form of an eARG
+is given in Fig.~\ref{fig-arg-data-structure}B.
+In a common ancestor event the inbound lineages are merged into a
+single ancestral lineage (e.g., node \noderef{e}
+in Fig.~\ref{fig-arg-data-structure}), and in a recombination
+event a lineage is split into two independent
+ancestral lineages (e.g., node \noderef{d}
+in Fig.~\ref{fig-arg-data-structure}). The final vital
+detail---which distinguishes an arbitrary graph with the required topological
+properties from an ARG---is that we associate a breakpoint with each recombination
+event. This information is sufficient to uniquely
+define the local genealogical trees at every position along the genome,
+which is the basic requirement for a data structure encoding
+genome-wide genetic ancestry.
 
@@
 % What do we mean by an eARG, formally?
-Formally, we can define the classical eARG as a tuple $(e, \sigma)$, where $e$
+To be concrete, we can formally define the
+classical Griffiths eARG as a tuple $(e, \sigma)$, where $e$
 is an ordered list of edges defining a directed acyclic graph and
@@
 if $u$ is a recombination event with breakpoint $x$ and
-$\sigma(u) = \varnothing$ otherwise. (This formulation is equivalent to stipulating that
-a recombination node is of a different ``type'' to other nodes,
-with breakpoints only associated with that node type).
+$\sigma(u) = \varnothing$ otherwise.
+(This formulation is equivalent to having different ``types''
+of node, and that breakpoints are associated with recombination
+nodes only).
 Each edge $e_j = (c, p)$ describes an edge between
@@
 have exactly two edges in $e$ such that $u$ is the child.
-Note that this description captures only the
-graph topology: if we also wish to know the times of events we need
+This encoding of an eARG is equivalent to the output
+produced by programs such as
+\texttt{ARGweaver}~\citep{rasmussen2014genome}
+and \texttt{KwARG}~\citep{ignatieva2021kwarg},
+and example is given in Fig.~\ref{fig-arg-data-structure}A.
+Note that the description above captures only the
+graph topology: if we also wish to know the branch lengths we need
 an additional function $t(u): \mathbb{N} \rightarrow \mathbb{R}$
 which defines the time of each node.
-
-% % Not sure if we actually include this, but it's good to work it through
-% % to make sure the notation works. Maybe something to stick into
-% % an appendix
-% The most fundamental operation that we need to perform on an ARG
-% data structure is extracting the local tree topology at a particular
-% position along the genome. The basic strategy is to traverse
-% the graph upwards, building the tree as we visit each node.
-% At a recombination node, when we have a choice of two parents,
-% we follow the path that corresponds to the position that we
-% are building the tree for. More precisely, let $S$ be the set
-% of sample nodes (usually the leaves of the graph)
-% and $0 \leq x < L$ be a position along the genome, and suppose
-% we are building a mapping $p_u$ that defines the parent of node
-% $u$ at $x$.
-% Let $N$ be a set of
-% nodes we are currently constructing the tree at,
-% and initialise $N \leftarrow S$.
-% Then, while $|N| > 0$, we add branches to the
-% tree as follows. Let $u$ be an arbitrary element of $N$ (which
-% we remove).
-% % This isn't right: we're not handling the case where there's
-% % no edges in E where the child is $u$.
-% Then, if $\sigma(u) < x$ we let $j$ be the
-% smallest value such that $e_{j,1} = u$, and otherwise
-% let $j$ be the largest value such that $e_{j,1} = u$.
-% Then, set $v \leftarrow e_{j, 2}$,
-% $p_u \leftarrow v$
-% and $N \leftarrow N \cup \{v\}$.
 
@@
 in terms of the process in which we extract the local trees
-along the genome from an eARG.
-To recover
-the tree at genome coordinate $x$,
-we traverse the graph rootwards from the leaves. At a particular
+along the genome from an eARG (the most basic downstream
+operation for an ARG data structure).
+To recover the tree at genome coordinate $x$,
+we traverse the graph rootwards from the leaves.
+At a particular
 node $u$, if it has one parent we are at a common ancestor
@@
 than the left, and so we cannot emit edges as they are generated.
-Such issues can be worked around, but depending on the ordering of
+Such issues can be worked around, of course, but depending on the ordering of
 otherwise indistinguishable objects is generally problematic. In practice,
@@
 of a recombination event
-to resolve the problem~\citep{lyngso2005minimum,ignatieva2021kwarg}.
-
-Another limitation of the eARG encoding illustrated in
-Fig.~\ref{fig-arg-data-structure}A is the inability to represent
-multiple recombinations along a chromosome, gene conversion,
-or transmission of multiple chromosomes.
-[Sentence saying ``gene conversion is important, actually, cite, cite,
-so we should be hoping to infer it some day''.
-Cite~\citep{wiuf2000coalescent}.
-Presumably people
-already are in bacterial contexts? Gregor: I recall this paper from
-Arabidopsis https://www.pnas.org/doi/10.1073/pnas.1211827110]
-To allow us to encode data from these important processes we
-must either devise workarounds to the event based encoding
-(i.e., model gene conversion as two recombination events) or
-generalise the model to more appropriately reflect the realities
-of modern datasets.
+to resolve the issue~\citep{lyngso2005minimum,ignatieva2021kwarg}.
+
+% TODO update this when then later sections have been drafted.
+As we see in sections XXX and YYY there are substantial limitations
+of expressivity and computational performance when using this eARG
+encoding. More fundamentally, however, the general approach
+of classifying nodes according to a discrete set of events
+and associating genome information with those nodes is intrinsically
+limiting. As we saw in the previous section, the original ARG
+definitions were derived directly from the coalescent with recombination
+stochastic process, and the types of ancestry that can be
+described are therefore limited by the assumptions of that
+model. While the coalescent as a model can be suprisingly robust to
+basic violations of the underlying assumptions~\citep{
+wakeley2012gene,bhaskar2014distortion,nelson2020accounting},
+it is not acceptable that our ability to \emph{represent}
+ancestral relationships should be limited by these assumptions.
+
+% n << Ne is a silly assumption in today's datasets
+A core assumption of the coalescent is that the sample size $n$
+is much less than the effective population size, $N_e$. This
+assumption ensures that multiple events (i.e., simultaneous
+coalescence and recombination) don't occur at the same
+time, which is reflected in the eARG definitions.
+This once-reasonable assumption has been overtaken by
+the data we have available.
+% This can be tightened up a bit, I think. We may need to drop a
+% few references.
+Several human datasets now consist of hundreds of thousands of
+genomes~\citep{bycroft2018genome,karczewski2020mutational,tanjo2021practical},
+and so sample size is substantially \emph{larger} than $N_e$
+(often assumed to be $10^4$ in humans).
+Agricultural datasets are even more extreme.
+For example,
+the US dairy cattle database alone currently comprises more than 6 million
+animals with SNP array
+genotypes\footnote{\url{https://queries.uscdcb.com/Genotype/counts.html}},
+while the effective population size in modern dairy cattle breeds is
+less than 100 and decreasing~\citep{MacLeod2013,Makanjuola2020}.
+An extreme sign of these breeding practices is that there are only two ancestral
+Y-chromosome linages present in today's US Holstein dairy breed~\citep{Yue2015}.
+Simlarly, the 1000 bull genomes project~\citep{hayes20191000}
+comprises close to 7000 genomes, which are part of multi-generation pedigrees
+with millions of animals and extensive SNP array genotype and phenotype
+data \citep[e.g.][]{Cesarani2022}.
+Such datasets are also available for other species,
+for example, \citet{RosFreixedes2022} have combined genome
+and SNP array data to infer recombination of genomes for 440,610 pigs within
+7 multi-generation pedigrees~\citep{whalen2018,Johnsson2021,Ros-Freixedes2020}.
+Effective population size in modern pig breeds is also less than 100 due to
+intense selection and directed reproduction \citep{Hall2016,Porcnic2016}.
+
+Another assumption of the coalescent with recombination which prevents complex
+combinations of events occuring simultaneously, is that the genome (or
+at least the region under study) is short enough that the number of extant
+lineages remains much smaller than $N_e$ at all times. Whole
+genome sequences have been available for model organisms
+for over a decade now, % True? Citation?
+and indeed complete chromosome-level assemblies are possible
+in humans~\citep{miga2020telomere}.
+Projects are under way to obtain high-quality assemblies
+for all eukaryotic species in Britain and Ireland~\citep{darwin2022sequence}
+and ultimately worldwide~\citep{lewin2022earth}.
+Clearly multiple recombinations (and other complex events) will
+occur in the ancestry of such datasets, and we should not be
+limited in what we can infer by the data structure that we
+use to represent the results.
+
+The eARG encoding consists of two types of event
+(common ancestor and recombination) but nature is not so parsimonious
+and there are many other processes that will affect the
+ancestry of samples.
+For example, gene conversion between homologous
+chromosomes~\citep{wiuf2000coalescent,chen2007gene} is an important
+evolutionary force in many species~\cite[e.g.][]{yang2012great}.
+% Some refs for where people are inferring ancestry in the
+% presence of GC? Isn't this what ClonalFrame etc are all about?
+Similarly, the eARG encoding cannot easily represent multiple
+recombinations on a chromosome, or the transmission of multiple
+chromosomes.
+What we can infer about ancestral histories and
+processes should not be limited by the data structure that we
+use to represent such ancestry. The eARG is simply one way
+to \emph{encode} such history, and, as we have shown,
+one that has significant limitations of expressivity
+derived from its historical origins in the idealised
+coalescent with recombination model.
+
+\section*{Node annotations}
+[FIXME: this is not the intended transition or next step, just putting this
+content here for now.]
 
@@
 power of ancestral recombination graph data structures.
-
-% it's not reasonable to assume that one type of event happens
-% at a time any more.
-\subsection*{Issues with event-based encoding}
-
-Even with the change to edge annotations, a focus on specific
-types of events can limit the types of ancestral history that can
-be described by an eARG. As it is derived from the coalescent
-with recombination, a standard eARG explicitly assumes that only one type
-of event can occur (i.e., common ancestry or recombination) at a
-time. This model is derived as an approximation to a Wright-Fisher
-model, assuming that the sample size $n$ is much less than the
-population size $N$, and that the genome is short enough
-that the number of extant lineages remains
-much smaller than $N$ at all times.
-While the coalescent model can be suprisingly robust to
-basic violations of the underlying assumptions~\citep{
-wakeley2012gene,bhaskar2014distortion,nelson2020accounting},
-it is not acceptable that our ability to \emph{represent}
-ancestral relationships should be limited by these assumptions.
-In a Wright-Fisher model, coalescence and recombination can
-occur at the same time in the same genomes, which is increasingly
-captured in the incredibly rich datasets that are available today.
-
-Firstly, complete chromosome-level assemblies are now possible
-in humans~\citep{miga2020telomere},
-and projects are under way to obtain high-quality assemblies
-for all eukaryotic species in Britain and Ireland~\citep{darwin2022sequence}
-and ultimately worldwide~\citep{lewin2022earth}. Thus, we can
-expect datasets to span the full continuum from non-recombining
-short genomes to very long genomes with high-levels of recombination
-[rephrase, and some cites?],
-all of which have high-quality whole genome data.
-
-Secondly, the scale of modern datasets presents a major challenge
-to the assumption that $n \ll N_e$.
-Several human datasets now consist of hundreds of thousands of
-genomes~\citep{bycroft2018genome,karczewski2020mutational,tanjo2021practical},
-and thus our sample size can be substantially \emph{larger} than $N_e$
-(often assumed to be $10^4$ in humans).
-The 1000 bull genomes project~\citep{hayes20191000}
-comprises close to 7000 genomes, which are part of multi-generation pedigrees
-with millions of animals and extensive SNP array genotype and phenotype
-data \citep[e.g.][]{Cesarani2022}.
-For example,
-the US dairy cattle database alone comprises more than 6 million
-animals with SNP array
-genotypes\footnote{\url{https://queries.uscdcb.com/Genotype/counts.html}},
-while the effective population size in modern dairy cattle breeds is
-less than 100 and decreasing~\citep{MacLeod2013,Makanjuola2020}.
-This low effective population size is caused by intense selection and
-staggering use of a handful of bulls across the world. The current record
-holder is the bull Toystory that has produced 2.4 million doses of semen
-for artificial insemination across
-13 years\footnote{\url{https://www.dairyherd.com/news/world-renowned-toystory-bull-has-died}}.
-An extreme sign of these breeding practices is that there are only two ancestral
-Y-chromosome linages present in today's US Holstein dairy breed~\citep{Yue2015}.
-Similar datasets are becoming available also in other livestock species.
-Notably, \citet{RosFreixedes2022} have combined genome
-and SNP array data to infer recombination of genomes for 440,610 pigs within
-7 multi-generation pedigrees~\citep{whalen2018,Johnsson2021,Ros-Freixedes2020}.
-Effective population size in modern pig breeds is also less than 100 due to
-intense selection and directed reproduction \citep{Hall2016,Porcnic2016}.
-
-For these reasons, we suggest that instead of defining the nodes in an
-ARG as events, which limits us to certain models of inheritance,
-we use a more flexible interpretation where they are defined as genomes.
 

### 2022-08-02 paper.tex
ASG redraft

--- 2022-08-01 paper.tex
@@
 % graphs very much as *processes*.
-However, in parallel with this tendancy to decouple the term
-Ancestral Recombination graph from its original defining stochastic process,
+% Jere: ASG section redrafted, gene transfer left as TODO (because I
+% know essentially nothing about it).
+In parallel with this tendancy to decouple the term
+Ancestral Recombination Graph from its original defining stochastic process,
 an extensive mathematical literature has developed in
@@
 (ASG)~\citep{krone1997ancestral,neuhauser1997genealogy}
-uses a similar approach to model an ensemble of correlated
-potential ancestral trees and has been of considerable
-and ongoing theoretical interest~\citep[e.g.][]{
+uses a similar approach to model natural selection.
+Unlike the ARG process, the ASG imposes a hard distinction between the stochastic process,
+which constructs a random ARG-like graph, and an observable realisation,
+which is a single tree sampled from the graph in a non-uniform way to encode desired
+patterns of natural selection. 
+Constructions of ASG-like stochastic processes encoding various
+forms of selection, often in parallel with recombination or other genetic forces,
+are an area of considerable and ongoing theoretical interest~\citep[e.g.][]{
 neuhauser1999ancestral,
@@
 koskela2019robust}.
-A vital distinction between the ARG (process) and ASG, however,
-is that there is a hard distinction between the stochastic process
-(the graph) and an observable realisation (a single tree) in the ASG
-because the it encodes \emph{potential} ancestral trees.
 Similarly, the Ancestral Gene Transfer Graph
 models horizontal gene transfer~\citep{baumdicker2014infinitely}
-by [FINISH ME. Also, any other ASG-like \emph{processes} to mention?].
+by [FINISH ME.
 

### 2022-08-02 paper.tex

--- 2022-08-01 paper.tex
@@
 koskela2019robust}.
-A vital distinction between the ARG (process) and ASG, however,
-is that there is a hard distinction between the stochastic process
-(the graph) and an observable realisation (a single tree) in the ASG
-because the it encodes \emph{potential} ancestral trees.
 Similarly, the Ancestral Gene Transfer Graph
@@
 models horizontal gene transfer~\citep{baumdicker2014infinitely}
-by [FINISH ME. Also, any other ASG-like \emph{processes} to mention?].
+by [FINISH ME.
 

### 2022-08-03 paper.tex
ARG process appendix

@@
 that was missing from Hudson's treatments.
+An overview of the Hudson and Griffiths stochastic processes is provided in the appendix.
 
@@
 and which encodes recombination events and times explicitly, e.g.\ using nodes.
+We will provide a more detailed discussion on sampling probability evaluation in the appendix.
 
@@
 
+Here we review the two predominant stochastic processes which construct ARGs:
+the ``big" ARG process of \cite{griffiths1997ancestral}, and the ``little" ARG process of
+ \cite{hudson1983properties}. The big ARG process is mathematically simpler
+ but is computationally intractable due to generating a vast number of ancestors
+ which contribute no genetic material to the initial sample.
+The little ARG process avoids non-genetic ancestors at the cost of more complex
+dynamics and state space. We also demonstrate that evaluating the sampling probability
+of either process---a key quantity in many statistical approaches---requires that the
+gARG (or eARG) data structure be interpreted in a model-specific way.
+
+A generic state of the little ARG process consists of a finite collection of lineages $L$,
+each of which is a list of disjoint ancestry segments $(\ell, r, a)$, where
+$[\ell, r)$ is a half-closed genomic interval and $a$ is an integer
+tracking the number of samples to which the lineage is ancestral over that interval.
+We also usually track the node associated with each segment, but
+that is not important for our purposes here so we omit it to lighten notation.
+The initial condition for a sample of $n$ genomes of length $m$ consists of $n$ lineages 
+of the form $\{(0, m, 1)\}$. The process traverses a series of common ancestor and 
+recombination events backwards in time.
+Recombination events happen at rate $\rho \nu / (m - 1)$,
+where $\rho \geq 0$ is a per-genome recombination rate and
+ \[
+ \nu = \sum_{x \in L}\left( \max_{(\ell, r, a) \in x}r
+     - \min_{(\ell, r, a) \in x}\ell - 1 \right)
+ \]
+ is the number of available ``links" surrounded by ancestral material.
+ At a recombination event we choose one of these links uniformly and break it,
+ replacing the original lineage in $L$ with two new lineages containing the ancestral material
+ to the left and right of the break point, respectively.
+
+Common ancestor events occur at rate $\binom{|L|}{2}$.
+In a common ancestor event, two uniformly sampled lineages have their segments
+merged into a single ancestor lineage, which is added to $L$.
+If the lineages have overlapping intervals of ancestry,
+say, $(\ell, r, a_1)$ and $(\ell, r, a_2)$, a
+\emph{coalescence} occurs. The result is a segment
+$(\ell, r, a_1 + a_2)$, and if $a_1 + a_2 < n$ it is included in the
+ancestor lineage. Otherwise, if $a_1 + a_2 = n$, we have found
+the most recent common ancestor of all samples in the interval$[\ell, r)$
+and do not need to simulate its history any further.
+Non-overlapping intervals from the two lineages are included
+ in the ancestor lineage without changes. Eventually,
+we find resultant lineages in which all segments have fully coalesced,
+and so the number of extant lineages gradually falls to zero.
+
+In the Griffiths formulation (the big ARG process), each edge in the graph corresponds to an extant
+lineage and nodes are events in the process. The $n$ initial leaf nodes are
+sampling events. Common ancestor events occur at rate $\binom{|L|}{2}$.
+When a common ancestor event happens, two uniformly chosen lineages
+merge into a common ancestor lineage.
+Recombination events happen at rate $|L| \rho$. Here, we choose a lineage (i.e.\ edge) uniformly,
+and a breakpoint $0 < x < m$ uniformly on its genome. We terminate the edge at a
+node, record the breakpoint, and start two new edges from this node. The process
+then continues until there is only one lineage left (the Grand Most Recent
+Common Ancestor, GMRCA), which is guaranteed to
+happen in finite time because of the linear rate of coalescing vs.\ quadratic rate of branching.
+
+The state-space of the big ARG process is much simpler than that of the little ARG process,
+which greatly facilitates mathematical reasoning. This simplicity comes at a
+substantial cost, however, if we wish to use it as a practical means of
+simulating recombinant ancestries. 
+The number of events in the big ARG all the way back to the GMRCA
+is $O(e^\rho)$~\citep{griffiths1997ancestral}, whereas the number
+of events required to simulate the little ARG is
+$O(\rho^2)$~\citep{hein2004gene,baumdicker2021efficient}.
+This disparity arises because the majority of the events in the big ARG are 
+recombination events which occur outside of ancestral material,
+and this do not have any bearing on the ancestry of the initial sample.
+Because we don't keep track of the distribution of ancestral material during the process,
+we generate a vastly larger graph.
+
+Figure \ref{hudson_vs_bigARG} illustrates the more complex state space
+of the little ARG process, as well as the extra events which occur in the big ARG process.
+Moreover, it depicts the rates of common ancestors and recombination events in each
+interval of time of the realisations.
+In order to evaluate these rates, and hence the sampling probability
+(see e.g.\ \cite[Equation (3)]{mahmoudi2022bayesian}),
+it is necessary to know the number of lineages and number of extant links
+available for recombination in each time interval.
+This information cannot be recovered from the gARG encoding depicted in
+e.g.\ Figure \ref{fig-ancestry-resolution}.
+For example, it is clear that a recombination takes place between nodes \textsf{b} and
+\textsf{d} as well as \textsf{e}, but the exact time of the event is ambiguous,
+and thus so is the number of lineages during the time interval.
+Conventions can be introduced to resolve such ambiguities;
+for the likelihood-based inference algorithms for the coalescent with recombination,
+the two parent nodes are typically created at the time of the recombination event.
+See the appendix of \cite{baumdicker2021efficient} for details of evaluating
+coalescent with recombination likelihoods using this convention.
+This is also the interpretation depicted in
+Figure \ref{hudson_vs_bigARG}, but it means that the two edges above node \textsf{b}
+in Figure \ref{fig-ancestry-resolution} should correspond to only one lineage,
+along which all $m-1$ links are available for recombination.
+The lineage then splits into two at the time of nodes \textsf{d} and \textsf{e}.
+Nodes \textsf{k}, \textsf{l}, and \textsf{m} in Figure \ref{fig-ancestry-resolution}
+demonstrate that the same issue can affect the number of links available for recombination:
+without an external convention, the exact time at which the trapped ancestral material on
+node \textsf{k} ceases to be available for an effective recombination in the little ARG process.
+
 \begin{figure}[ht]
@@
 % prevent flushing all the other figures to the end of the document.
-\scalebox{0.5}{
+\scalebox{0.54}{
 \begin{tabular}{cc}
@@
 	% Dashed lines for start and end times
-	\draw[dashed] (-2, 0) -- (11.1, 0);
+	\draw[dashed] (-2, 0) -- (11, 0);
 	\node [label=left:{$t = 0$}] at (-2,0) {};
-	\draw[dashed] (-2, 8) -- (11.1, 8);
+	\draw[dashed] (-2, 8) -- (11, 8);
 	\node [label=left:{$t = T$}] at (-2,8) {};
@@
 	% Gray dashed lines to visually separate holding times
-	\draw[color=gray, dashed] (7.5, 1.2) -- (11.1, 1.2);
-	\draw[color=gray, dashed] (7.5, 1.7) -- (11.1, 1.7);
-	\draw[color=gray, dashed] (7.5, 2.4) -- (11.1, 2.4);
-	\draw[color=gray, dashed] (7.5, 3.1) -- (11.1, 3.1);
-	\draw[color=gray, dashed] (7.5, 4.5) -- (11.1, 4.5);
-	\draw[color=gray, dashed] (7.5, 5.4) -- (11.1, 5.4);
-	\draw[color=gray, dashed] (7.5, 6) -- (11.1, 6);
-	\draw[color=gray, dashed] (7.5, 7.2) -- (11.1, 7.2);
+	\draw[color=gray, dashed] (7.5, 1.2) -- (11, 1.2);
+	\draw[color=gray, dashed] (7.5, 1.7) -- (11, 1.7);
+	\draw[color=gray, dashed] (7.5, 2.4) -- (11, 2.4);
+	\draw[color=gray, dashed] (7.5, 3.1) -- (11, 3.1);
+	\draw[color=gray, dashed] (7.5, 4.5) -- (11, 4.5);
+	\draw[color=gray, dashed] (7.5, 5.4) -- (11, 5.4);
+	\draw[color=gray, dashed] (7.5, 6) -- (11, 6);
+	\draw[color=gray, dashed] (7.5, 7.2) -- (11, 7.2);
 \end{tikzpicture}
@@
 	% Dashed lines for start and end times
-	\draw[dashed] (-2, 0) -- (11, 0);
+	\draw[dashed] (-2, 0) -- (9.1, 0);
 	\node [label=left:{$t = 0$}] at (-2,0) {};
-	\draw[dashed] (-2, 8) -- (11, 8);
+	\draw[dashed] (-2, 8) -- (9.1, 8);
 	\node [label=left:{$t = T$}] at (-2,8) {};
@@
 	% Gray dashed lines to visually separate holding times
-	\draw[color=gray, dashed] (7.5, 1.2) -- (11, 1.2);
-	\draw[color=gray, dashed] (7.5, 1.7) -- (11, 1.7);
-	\draw[color=gray, dashed] (7.5, 2.4) -- (11, 2.4);
-	\draw[color=gray, dashed] (7.5, 3.1) -- (11, 3.1);
-	\draw[color=gray, dashed] (7.5, 3.8) -- (11, 3.8);
-	\draw[color=gray, dashed] (7.5, 4.5) -- (11, 4.5);
-	\draw[color=gray, dashed] (7.5, 5.4) -- (11, 5.4);
-	\draw[color=gray, dashed] (7.5, 6) -- (11, 6);
-	\draw[color=gray, dashed] (7.5, 6.6) -- (11, 6.6);
-	\draw[color=gray, dashed] (7.5, 7.2) -- (11, 7.2);
+	\draw[color=gray, dashed] (7.5, 1.2) -- (9.1, 1.2);
+	\draw[color=gray, dashed] (7.5, 1.7) -- (9.1, 1.7);
+	\draw[color=gray, dashed] (7.5, 2.4) -- (9.1, 2.4);
+	\draw[color=gray, dashed] (7.5, 3.1) -- (9.1, 3.1);
+	\draw[color=gray, dashed] (7.5, 3.8) -- (9.1, 3.8);
+	\draw[color=gray, dashed] (7.5, 4.5) -- (9.1, 4.5);
+	\draw[color=gray, dashed] (7.5, 5.4) -- (9.1, 5.4);
+	\draw[color=gray, dashed] (7.5, 6) -- (9.1, 6);
+	\draw[color=gray, dashed] (7.5, 6.6) -- (9.1, 6.6);
+	\draw[color=gray, dashed] (7.5, 7.2) -- (9.1, 7.2);
 \end{tikzpicture}
@@
 
-[JK: this is text roughly related to big vs little ARG taken from various
-iterations]
-
-In the Griffiths formulation, each edge in the graph corresponds to an extant
-lineage and nodes are events in the process (the initial $n$ leaf nodes are
-``sampling'' events). Common ancestor events occur at rate $\binom{k}{2}$ when there
-are $k$ lineages present. We choose two lineages (edges) uniformly, and merge them
-into a common ancestor lineage. Recombination events happen at a rate of
-$k \rho$. Here, we choose a lineage (edge) uniformly, and a
-breakpoint $0 < x < m$ uniformly on its genome. We terminate the edge at a
-node, record the breakpoint and start two new edges from this node. The process
-then continues until there is only one lineage left (the Grand Most Recent
-Common Ancestor, GMRCA), which is guaranteed to
-happen in finite time because of the linear vs quadratic rates of branching
-and coalescing. Later, we will show how the graph structure and breakpoints
-associated with
-recombination nodes provides sufficient information to later recover the marginal
-trees and calculate sampling probabilities. However, this process traces out
-a larger graph (the ``big'' ARG) than the Hudson algorithm (the ``little'' ARG; see the ``\nameref{Ancestry_resolution}'' section),
-and is consequently considerably less efficient \citep{wiuf1999ancestry}.
-
-The state-space of the Griffiths process is much simpler than Hudson's algorithm,
-which greatly facilitates mathematical reasoning. This simplicity comes at a
-substantial cost, however, if we wish to use it as a practical means of
-simulating recombinant ancestry. The big ARG is vast, and any realisation
-for even moderate levels of recombination is far too large to be of practical
-use. The number of events in the big ARG all the way back to the GMRCA
-is $O(e^\rho)$~\citep{griffiths1997ancestral}, whereas the number
-of events required to simulate the little ARG is
-$O(\rho^2)$~\citep{hein2004gene,baumdicker2021efficient}.
-This disparity in the number of events in the two formulations is
-because the majority of the events that occur in the big ARG do
-not affect the genetic ancestry of the sample in any way. Recombination
-events that occur outside of ancestral material do not have any bearing
-on the ancestry of the sample, and so the structure is hugely redundant.
-As~\cite{wiuf1999recombination} note,
-``an `ancestral' sequence in the birth and death process
-need not have any genetic material in common with a
-sequence descended from it.''
-
 % Full quote:
@@
 
-% % This is a discussion of Hudson's alg left over from earlier version. We
-% % might wish to adapt some of this when defining the Little ARG.
-
-% Each lineage consists of a list of
-% disjoint ancestry segments $(\ell, r, a)$, where
-% $[\ell, r)$ is a half-closed genomic interval and $a$ is an integer
-% tracking the number of samples to which the lineage is ancestral over that interval.
-% (We also usually track the tree node associated with each segment, but
-% that is not important for our purposes here so we omit it.)
-% If we have $n$ samples and a genome of length $m$, the process begins with $n$ lineages
-% of the form $\{(0, m, 1)\}$. The process then works backwards in time from
-% the present day as a series of random common ancestor or recombination events.
-% Recombination events occur at a rate determined by the amount of ancestral material and
-% the way in which it is distributed along a chromosome.
-% Let $L$ be the set of ancestors at a given time $t$. Recombination events
-% happen at rate $\rho \nu / (m - 1)$ where
-% \[
-% \nu = \sum_{x \in L}\left( \max_{(\ell, r, a) \in x}r
-%     - \min_{(\ell, r, a) \in x}\ell - 1 \right)
-% \]
-% is the number of available `links' that may be broken. Thus, the rate of
-% recombination is determined by the left- and right-most extent of the
-% ancestral material carried by each lineage. At a recombination
-% event we choose one of these links uniformly and break it. Given a lineage
-% $x = [(\ell_j, r_j, a_j)]$ and a breakpoint $k$, we have two lineages
-% $x_1$ and $x_2$ such that FILL IN DETAILS
-
-% When $k = |L|$ lineages are present, common ancestor events
-% occur at rate $\binom{k}{2}$. In a common ancestor event, two lineages
-% are chosen uniformly at random and their ancestry segments merged.
-% If we have overlapping intervals of ancestry from the two lineages,
-% say, $(\ell, r, a_1)$ and $(\ell, r, a_2)$, a
-% \emph{coalescence} occurs and ancestor represented by the current event
-% will be present as a node (at least) in the marginal trees covering
-% the interval $[\ell, r)$. The result of this coalescence is a segment
-% $(\ell, r, a_1 + a_2)$, and if $a_1 + a_2 < n$ it is included in the
-% ancestry for the new lineage. Otherwise, if $a_1 + a_2 = n$ we know that
-% we have found the most recent common ancestor of all samples in
-% the interval $[\ell, r)$ and so we do not need to simulate its history any further.
-% Nonoverlapping intervals of ancestry from the two lineages are included
-% in the resulting lineage without changes. Eventually, as the process continues,
-% we find resultant lineages in which all segments have fully coalescenced,
-% and so the number of extant lineages gradually dwindles down to zero.
-
 \end{document}

### 2022-08-08 paper.tex

@@
 
-\section*{The Event ARG data structure}\label{eARG}
-
+\section*{Event ARGs}\label{eARG}
 In keeping with its original derivation in terms of a stochastic
@@
 
-\section*{Node annotations}
-[FIXME: this is not the intended transition or next step, just putting this
-content here for now.]
-
-Fortunately, it is straightforward to generalise the classical Griffiths
-eARG encoding to both simplify implementations and to encompass
-alternative patterns of genetic inheritance.
-The only change that we require is to associate
-the details of inheritance intervals with \emph{edges} rather than
+\section*{Node vs edge annotations}
+Another aspect of the classical description of an ARG data structure that
+derives from its historical origins as a stochastic
+process is how we encode information about genetic
+inheritance in the graph. The approach taken by Griffiths to
+doing this is concise, elegant: all we need to do is store
+the location of the breakpoint at each recombination event.
+Storing a breakpoint for each recombinatin event is a very
+natural and straightforward thing to do when simulating
+a (``Big''; see Appendix XXX) ARG process.
+Given the graph topology and these breakpoints, we can then
+recover the
+local trees at every point in the genome
+by tracing rootwards from the leaves and choosing the
+appropiate outbound path at recombination nodes.
+However, there are several significant disadvantages
+to this approach: [TODO be concrete here later
+and give pointers to the section where we discuss the
+alternative.]
+
+Fortunately, there is a straightforward change to the
+classical eARG encoding that simplifies
+implementations and greatly increases the flexibility
+of both the types and granularity of inheritance information
+that we can represent.
+The main change that we need to make is to
+associate the
+the details of inheritance with \emph{edges} rather than
 as breakpoints associated with recombination events and
-edge ordering rules. This is illustrated in Fig~\ref{fig-arg-data-structure}C
+edge ordering rules.
+This is illustrated in Fig~\ref{fig-arg-data-structure}C
 where this alternative edge annotation formulation is
@@
 
-% This method for representing recombinations may seem wasteful,
-% since more than one node is required for each recombination,
-% and genomic interval information must be stored on every edge.
-% However, asymptotically the amount of information stored is the
-% same, and there is much simplicity to be gained by having
-% a single type of node and by treating all edges in the same way.
-% More importantly --- as we see in the next section --- we gain the
-% ability to simplify the ARG by removing nodes without affecting the
-% local tree topologies, and also the ability to record only the
-% regions of genome relevant to the final, sample genomes. This last
-% feature turns out to be the key to efficient processing of the ARG.
-
-% [Possibly more stuff here: clarify why we assign coordinates to edges
-% rather than nodes. Basically lay out all the stuff that we can
-% talk about \emph{without} going in to the details of
-% ancestry resolution.]
+For simplicity, throughout the rest of the paper we will contrast the
+classical eARG (as defined in the XXX section) with the gARG definition
+here. However, it is important to note that many of the benefits of the
+gARG encoding that we discuss are due to the encoding of inheritance via
+edges rather than nodes (see section XXX).
+Whether one regards nodes as genomes or events
+is something of a philosophical question, and mostly unimportant
+from a computational perspective.
 

### 2022-08-08 paper.tex
Pass through gARG section

@@
 In order to distinguish this form of ARG from what we will be discussing
-in later sections we call it an ``event ARG``, or eARG for brevity.
+in later sections we refer to it as an ``event ARG``, or eARG for brevity.
 An example of this form of an eARG
@@
 \section*{Genome ARGs}
-% We're not the first to talk about ARGs and pedigrees, and we're
-% not doing anything controversial.
-The deep connection between ARGs and pedigrees is
+As we have seen in previous sections, the focus
+on a limited palette of events makes describing currently-available
+datasets and known patterns of genetic inheritance as an eARG
+quite cumbersome.
+There is no fundamental reason that the patterns of genetic inheritance
+that we wish to describe via an ARG must be in terms
+of discrete \emph{events}, however. Recent discussions have illustrated the
+relationship between ARGs and
+the pedigree~\citep{mathieson2020ancestry,brandt2021evaluation},
+and the deep connections between the structures are
 well understood
 \citep[e.g.][]{wakeley2012genetics,gusfield2014recombinatorics,
-speed2015naturereviewsgenetics}
-and recent discussions have illustrated the
-relationship~\citep{mathieson2020ancestry,brandt2021evaluation}, % more??
-showing how the coalescence and recombination events in an ARG
-correspond to particular genomes in the pedigree.
-As discussed in the previous section, the definition of an ARG
-as a data structure in terms of these two \emph{events} is problematic,
-fundamentally limiting the patterns of ancestry that can be described.
-Following \cite{mathieson2020ancestry}, instead of focussing on events,
-we define the ARG data structure in terms of \emph{genomes} and the
-patterns of transmission between them.
-% Bit of duplication here, but good to make the point about being
-% motivated by real existing datasets not some academic desire to
-% model complexity that isn't there.
-This, coupled with
-the explicit association of genome coordinates with edges, allows
-us to represent even the bewilderingly complex patterns of relatedness
-present in modern datasets.
+speed2015naturereviewsgenetics}.
+Following \cite{mathieson2020ancestry},
+rather defining the inheritance
+of genetic material in terms of \emph{events} as we do in an eARG,
+we can instead define an ARG data structure in terms of
+\emph{genomes} and intervals of genetic inheritance
+along the edges (see the previous section).
+To distinguish from the classical Griffiths eARG,
+we refer to this encoding of an ARG data structure
+as a ``Genome ARG'', or gARG.
+The distinction between nodes representing events
+and genomes may seem pedantic at first, but as we
+see in later sections, we gain a great deal of flexibility
+and computational efficiency by decoupling our representation
+of genetic ancestry from the events that generated it.
 
@@
 they merge into a common ancestor, the merged region is darker; this shows that all
-genomic regions fully coalesce in genome \textsf{n} in the oldest generation.
-(B) The equivalent genome ARG, where there is a single type of node (representing
+genomic regions find a common ancestor in genome \textsf{n} in the oldest generation.
+(B) The corresponding genome ARG, where there is a single type of node (representing
 a genome) along with annotations on all edges; annotations showing partial
-transmission are highlighted in bold. Recombination, such as that causing the
-breakpoint at position 7, is associated with two parent genomes (\textsf{e} and
+transmission are highlighted in bold. Recombination (e.g.,
+the cause of the breakpoint at position 7)
+is associated with two parent genomes (\textsf{e} and
 \textsf{f}) and a child genome (the recombinant, \textsf{c}).
@@
 
-To distinguish from the classical event-based encodings,
-we call this encoding of an ARG data
-structure a ``Genome ARG'', or gARG.
-Fig~\ref{fig-arg-in-pedigree} illustrates the basic elements of a gARG.
-Fig.~\ref{fig-arg-in-pedigree}A shows the genomes
-embedded in an inbred diploid pedigree, in which individuals occur at
-discrete generations and carry two genomes (maternal and paternal).
-The pedigree is shown here for concreteness, and
-there is no inherent limitation on mating system, ploidy, or age structure
+A gARG is a graph in which nodes represent monoploid genomes (i.e.,
+haploids have one genome, diploids two, etc.) and edges represent
+genetic inheritance between an ancestor and a descendant.
+Each edge is annotated (see the previous section) with the coordinates
+of the genomic intervals over which genetic inheritance occurs
+(the next section explores the subtleties of what we mean by
+genetic inheritance).
+Fig~\ref{fig-arg-in-pedigree} illustrates the basic elements of a gARG,
+showing how the genomes embedded in an example (highly inbred)
+pedigree correspond to the nodes in a gARG, along with the corresponding
+local trees.
+A diploid pedigree is shown here for concreteness, but
+there is no limitation on mating system, ploidy, or age structure
 in the gARG encoding.
-A genome is inherited from one of the individual's parents,
+In the diploid setting, a genome is inherited from one
+of the individual's parents,
 and may be the recombined product of that parent's two genomes.
-For example, individual $I_1$ has two genomes \textsf{a} and \textsf{b},
+For example, individual $I_1$ in Fig.~\ref{fig-arg-in-pedigree}
+has two genomes \textsf{a} and \textsf{b},
 inherited from parents $I_3$ and $I_4$. Genome \textsf{a} is the product of
@@
 whereas \textsf{b} was inherited directly from $I_4$'s genome \textsf{g} without
-recombination. Fig.~\ref{fig-arg-in-pedigree}B shows the resulting gARG,
-depicting the ancestry of the sample genomes \textsf{a}--\textsf{d}. The graph
-topology along with the genome coordinate annotations on the
-edges then uniquely defines the trees at every position
-along the genome (Fig.~\ref{fig-arg-in-pedigree}C).
+recombination.
+Arbitrary patterns of genetic inheritance (gene conversions, multiple recombinations,
+etc) can be expressed without difficulty.
 
@@
 also phrase this in terms of the ancestral material of a sample---see
-the next section). Deriving the local tree at a point $x$
+the next section).
+Deriving the local tree at a point $x$
 follows largely the same pattern as for an eARG but is somewhat
@@
 
-The focus on genomes rather than events has some interesting
-and important ramifications. Firstly, a gARG may contain nodes that
+There are some interesting and important consequences to this focus on genomes
+rather than events.
+Firstly, a gARG may contain nodes that are both samples \emph{and}
+internal nodes in the local trees.
+This ability is useful when we have pedigree information
+along with genetic data
+\cite[e.g.][]{hayes20191000,RosFreixedes2022,anderson2022genes},
+where we may have many generations of internal sample nodes
+whose genomes have been sequenced.
+Another situation in which it is important that
+we do not assume samples are all contemporary ``leaf'' nodes
+(or that all internal nodes are either coalescences or recombinations)
+is when we are incorporating ancient genomes into ARG
+inference~\citep{speidel2021inferring,wohns2022unified}.
+Similarly, a gARG may contain nodes that
 do not correspond to any particular event in the ancestry of the samples.
@@
 the ``\nameref{ARG_simplification}'' section) but there is no necessity for this from
-a representational perspective. This ability can be useful in practice and could be
-% Do we need to spell this out? Gregor: I think so - see below
-required if, for example, we have a deeply sequenced pedigree as in \cite{RosFreixedes2022}.
-% e.g. Honeybees? Be great to cite an example!
-% Note by Yan - if we actually have the sequence, then this turns into
-% a sample node, so would be normally included anyway. Perhaps, therefore
-% we should talk about these ``pass though'' nodes in the context
-% of internal samples.
-% Gregor: There are no good deeply sequenced pedigrees in honeybees, but sex
-% chromosome behaves like haplodiploid system, also any deep pedigree works here, so I cited our pig
-% work. Also, I think its important to not think of all genomes as current samples - we have many
-% generations of animals with genome or SNP array data and we know pedigree so we can't assume they
-% are all current samples - I believe you made this clear distinction also in the tsdate paper with
-% ancient human samples
-
-Secondly, unlike in an eARG, nodes in a gARG are not classified into different
-types. Indeed a single node can have both multiple children
-and multiple parents. This would be the case in Fig~\ref{fig-arg-in-pedigree} if,
-for instance, node \textsf{a} were to have two children.
-% It is a shame that this is not actually shown in fig 3, but I don't think
-% we can do this unless we have 5 generations
-Moreover, a single node can have more than 2 children, and (as we shall see in
-the ``\nameref{ARG_simplification}'' section), more than 2 parents. Another way to
-state this is that transitions between nodes in a gARG can encapsulate multiple
-events in an eARG.
-
-The final point deals with the way in which recombination is represented, and is
-closely linked to which genomes are chosen as the nodes in a gARG.
-We could potentially choose genomes at any point in the life cycle: for instance,
-we could chose to represent gametes as ARG nodes. However, the most intuitive
-choice is where genomes represent diploid individuals in each generation, as in
+a representational perspective.
+
+Secondly, because we do not classify nodes by event type
+(common ancestor or recombination) there is no limitation
+on the complexity what can be modelled---multiple recombinations
+and coalescences can happen simultaneously in the same genome.
+In large deeply sequenced pedigrees such complexities are commonplace.
+
+Finally, there is a subtle and important point about how recombination
+is represented: which genome in the organismal lifecycle
+do nodes represent?
+In principle, we can intepret nodes as representing genomes
+at any point in the life cycle [supplementary Figure ?]. Hudson
+views nodes as representing gametes~\citep{hudson1983properties}.
+However, we argue that the most intuitive
+approach is to let nodes represent the genomes present in diploid
+individuals in each generation, as in
 Fig~\ref{fig-arg-in-pedigree}A. In this case recombination
@@
 concerning breakpoint location that would be associated with a
-recombination node in an eARG. For example, node \textsf{c} in
-Fig~\ref{fig-arg-in-pedigree} inherits genome positions 0 to 7 from \textsf{f}
-and positions 7 to 10 from \textsf{e}, indicating that the breakpoint is at position 7.
+recombination node in an eARG. [TODO finish up here with more
+clarity on what this means in terms of the required specifity
+of recombination events, and put in a forward ref to the simplification
+section where we talk about stacking up recombination events
+and their identifiability.]
 
@@
 edges rather than nodes (see section XXX).
-Whether one regards nodes as genomes or events
-is something of a philosophical question, and mostly unimportant
-from a computational perspective.
+Whether one regards nodes as genomes or events is orthogonal,
+and something of a philosophical question which makes little
+difference from a computational perspective.
 
@@
 something which is impossible to encode in an eARG.
+
+% JK: Text moved from an earlier section. See if we want to reuse this here
+% if not already said.
+% Indeed a single node can have both multiple children
+% and multiple parents. This would be the case in Fig~\ref{fig-arg-in-pedigree} if,
+% for instance, node \textsf{a} were to have two children.
+% % It is a shame that this is not actually shown in fig 3, but I don't think
+% % we can do this unless we have 5 generations
+% Moreover, a single node can have more than 2 children, and (as we shall see in
+% the ``\nameref{ARG_simplification}'' section), more than 2 parents. Another way to
+% state this is that transitions between nodes in a gARG can encapsulate multiple
+% events in an eARG.
 
@@
 \begin{itemize}
+\item [Recap quickly distinction bewtween aARG and gARG. Then]
+    This, coupled with
+    the explicit association of genome coordinates with edges, allows
+    us to represent even the bewilderingly complex patterns of relatedness
+    present in modern datasets.
+
 \item It is unfortunate that we are adding yet another meaning to the

### 2022-08-09 paper.tex
Pass through ancestry resolution section

@@
 
-\section*{Ancestry resolution}\label{Ancestry_resolution}
-% Need to introduce Hudson's algorithm as concerned with simulating the process of
-% the coalescent with recombination as efficiently as possible, and properly merge in
-% with this section.
-
-The edge annotations in Figs~\ref{fig-arg-data-structure} and \ref{fig-arg-in-pedigree}
-describe the regions of genome that have been passed from a parent to its immediate children,
-hence edges either span the entire genome, or entire region to the left (or right)
-of a breakpoint
-(e.g. Fig~\ref{fig-arg-data-structure}C). In this form of edge annotation, which we call
-``immediate edge annotation'',  % NB - perhaps we can think of a better term than this
-it is inefficient to extract the local trees from the ARG structure.
-
-There is, however, an alternative edge annotation scheme which allows efficient
-storage and extraction of the local trees. It is based on the fact that the ARG
-structure is inherently retrospective, defined as the ancestry of a set of \emph{sample}
-nodes (such as \textsf{a}, \textsf{b}, and \textsf{c} in the examples above). We define a
-``sample-resolved edge annotation'' to be a representation where edge spans
-represent only those parts of the genome that have been inherited by the samples.
-This also allows the removal of nodes which are not genetically ancestral to the samples.
+\section*{Inheritance and ancestral material}\label{Ancestry_resolution}
+The edges in a gARG represent a path of genetic inheritance from
+parent to child (and ultimately from cell to cell) through some
+number of generations. The edges are ``annotated'' (see section XXX)
+with the coordinates over which inheritance occurs. There are
+two ways in which we can interpret these annotations, and the
+distinction is subtle and important.
+The first interpretation is to define the annotations as
+the genetic intervals inherited by the descendant node from the
+ancestor node. The second is to view the edge annotations as representing
+the ``ancestral material''~\citep{wiuf1999ancestry,wiuf1999recombination}
+of the sample that is propagated through
+the graph from descendant to ancestor.
+These alternatives are illustrated in Fig~\ref{fig-ancestry-resolution}.
 
@@
 \caption{\label{fig-ancestry-resolution}
-The \citet[][fig.~1]{wiuf1999recombination} example graph as a gARG, with contrasting
-edge annotation schemes.
-(A) Immediate edge annotation, reflecting ``local'' transmission of all
-genetic material between parent and child. All edges span the full genome from position 0 to 7,
-apart from those linking a recombinant and its two parents (bold).
-(B) Sample-resolved edge annotation: the Hudson-like simplification algorithm has been used to
-shorten edge spans to reflect only the upwards transmission of genetic
-material from the samples \textsf{a}, \textsf{b}, and \textsf{c}. The depiction of the genome within
-each node is shaded as in figure \ref{fig-arg-in-pedigree}a to show
-coalescing regions of genome back to the grand MRCA genome \textsf{w}.
-Line widths reflect the genomic span of each edge, which
-helps visualise the relative importance of different transmission pathways.
+The \citet[][Fig.~1]{wiuf1999recombination} example graph as a gARG,
+with contrasting
+edge annotation schemes. (A) Immediate edge annotation, reflecting ``local''
+transmission of all genetic material between parent and child. All edges span
+the full genome from position 0 to 7, apart from those linking a recombinant
+and its two parents (bold). (B) Sample-resolved edge annotation: the
+Hudson-like simplification algorithm has been used to shorten edge spans to
+reflect only the upwards transmission of genetic material from the samples
+\textsf{a}, \textsf{b}, and \textsf{c}. The depiction of the genome within each
+node is shaded as in Fig.~\ref{fig-arg-in-pedigree}A to show coalescing
+regions of genome back to the grand MRCA \textsf{w}. Line widths reflect
+the genomic span of each edge, which helps visualise the relative importance of
+different transmission pathways.
 }
@@
 
-Fig~\ref{fig-ancestry-resolution} illustrates the two different edge annotation
-schemes. To obtain the sample-resolved annotation in
-Fig~\ref{fig-ancestry-resolution}B we must traverse backwards in time from
-sample nodes \textsf{a}, \textsf{b}, and \textsf{c}, using the edge intervals
-to record regions of sample genomes that have passed through each route. When
-edges join into a parent node, the parent is associated with the union of the
-child intervals. When a child node is connected to two parents, one parent is
-allocated the edge intervals of the child, truncated to include only genomic
-material to the left of the breakpoint. Similarly the other parent is allocated
-the child's intervals truncated such that they contain only regions to the
-right of the breakpoint (a natural extension exists for multiple chromosomal
-breakpoints).
-
-For example, in Fig~\ref{fig-ancestry-resolution}A the edge between child node \textsf{g} and
-parent node \textsf{r} spans the entire genome from 0..7, reflecting the material
-transmitted immediately between parent and child.
-However, in Fig~\ref{fig-ancestry-resolution}B the edge has been reduced in extent so that
-it now spans 0..2, the region that eventually ends up in
-sample node \textsf{c}. This sample-resolved edge annotation scheme cannot be
-derived locally: it must be done by traversing backwards in time from the sample nodes
-up through the ARG.
-
-% TECHNICAL DESCRIPTION
-% Given a gARG $E$ and the set of samples $S$ we can \emph{resolve} its
-% ancestry by noting that the resolved interval $Ir$ on an edge $(c,p,I)$
-% is given by [Rough notes]
-% \begin{itemize}
-% \item The ``ancestry interval'' $A_u$ of a node $u$ is the union of
-% all $R_{u, v}$ for all the children $v$ of $u$. (this isn't quite right)
-% \item The resolved interval is then the intersection of $A_u$ and the
-% original inheritance interval $I_v$ on the edge.
-% \item It would be good of if differentiate between the inheritance intervals
-% from the gARG and the ancestry intervals of the resolved gARG.
-% \end{itemize}
-% This process is illustrated in Fig~\ref{fig-ancestry-resolution}. It is intimately linked to
-%Hudson's algorithm which operates by
-% tracking segments of ancestral material carried by ancestors as
-% we go backwards in time.
-
-% jk-note unclear here about the best choice of notation. We could have
-% lists of tuples (l, r, a), or a tuple of vectors (l_j, r_j, a_j),
-% of a tuple (I_j, a_j), where I is a vector of intervals. Going to
-% stick with the original tuple notation for now.
-%
-% Each lineage consists of a set of
-% equal-length vectors $\mathbf{l}$, $\mathbf{r}$ and $\mathbf{a}$
-% such that for each $1 \leq j \leq \left| l \right|$,
-% $[\mathbf{l}_j, \mathbf{r}_j)$ is a half-closed genomic interval and
-% $\mathbf{a}_j$ is an integer
-% tracking the number of samples the lineage is ancestral to over the interval.
-% Assume that the intervals are sorted from left-to-right so that $\mathbf{l}_1$
-% and $\mathbf{r}_{|\mathbf{r}|}$ are the left- and right-most positions
-% covered by the lineage, respectively.
-
-When traversing upwards through the ARG like this, it is also helpful to record
-the number of sample nodes to which a given section of genome is ancestral.
-This is indicated in Fig~\ref{fig-ancestry-resolution}B by successively darker shading
-of the regions within each node. When a region has all samples as descendants, it is
-shaded black, and we can say that that region of the genome has fully coalesced.
-
-Formally, we can define a sample-resolved gARG as a tuple $(S, E)$
-where $S \subset \mathbb{N}$ is a set of sample nodes and
-$E$ is a set of edges. Each element of $E$
-is a tuple $(c, p, I)$ such that $c, p \in \mathbb{N}$,
-$c$ is the child node, $p$ is the parent node and $I$ is the set of
-disjoint genomic intervals $(\ell, r]$
-over which genome $c$ inherits from $p$, and there is a path from
-$c$ to some non-empty subset of $S$ through the local trees.
-
-The ARG traversal algorithm which implements this edge trimming and removal process
-is a form of ``simplification'' (cite), and can be thought of as running Hudson's algorithm
-on a fixed tree topology (see appendix XXX).
-
-Several features of Fig 3B are worthy of note. First, since node \textsf{j} is not on
-the ancestral path from the samples it (and its associated edges) can be removed.
-For larger ARGs, the absence of these nodes makes sample-resolved edge annotation
-substantially more efficient than immediate edge annotation.
-Second, by the time we reach
-nodes \textsf{t} and \textsf{v}, all regions of the genome have fully coalesced, and so
-\textsf{w}, the ``grand MRCA'' can be omitted from the graph, at the cost of having
-multiple root nodes in different regions of the genome.
-Third, there exist nodes such as
-\textsf{k} with two children (i.e. ``common ancestor'' nodes in Hudson's terminology)
-but in which no ancestral coalescence occurs. Fourth, some breakpoint positions occur
-within regions of genome that are not ancestral (e.g. the breakpoint at position 3 in
-node \textsf{k}). In this case sample-resolved edge annotations lose the exact position of
-the breakpoint, although they retain the information that a breakpoint occurred between
-positions 2 and 4 (which is sufficient for calculating the likelihood, see appendix XXX).
-
-% Should probably define how we build the local trees somewhere.
-TODO: explain why this encoding allows efficient extraction of local trees, as implemented in the tskit library.
-
-% We should talk about ARG and HUD here. Fig~\ref{fig-ancestry-resolution}A shows
-% ARG (the Griffiths graph). However, we have a small issue that subfig B
-% doesn't strictly show HUD, because it doesn't terminate as soon as all samples have coalesced.
-% We could potentially colour the top (non-HUD) section of the plot in a different colour?
-
-TODO: give some feel for how much of a reduction the sample-resolved encoding can give in a large ARG, and
-maybe mention ``big'' vs ``little'' args as may be described in Jere's appendix
-
-% Potentially useful text from a previous iteration:
-% Wiuf and Hein's classical pair of papers
-% clarified the relationship between Hudson's algorithm and the ARG.
-% ARG simulation is inherently less efficient than Hudson's
-% algorithm because we must track all lineages, even those that correspond
-% to sections that have coalesced, or we won't have a complete ARG.
+ARGs are inherently a retrospective structure defined in terms
+of a set of sample nodes (usually the leaves), and,
+as we discuss in section XXX, storing sample-resolved ancestry
+intervals (Fig~\ref{fig-ancestry-resolution}B) is generally much more
+useful than storing the local parent-child intervals of inheritance
+(Fig~\ref{fig-ancestry-resolution}A).
+Indeed, storing local parent-child
+inheritance intervals for edges is really only of interest because of
+its direct correspondence with the classical approach of associating
+breakpoints with recombination nodes in an eARG. The eARG encoding
+is simply a more limited form, in which the only types of direct
+inheritance we can have are the entire genome, or the intervals
+at either side of a recombination breakpoint.
+
+Fig~\ref{fig-ancestry-resolution}A is an instructive example
+from the literature \citep{wiuf1999recombination} and illustrates
+how a classical eARG can be described as a gARG with local edge annotations
+(directly corresponding to the eARG description), and how we can
+then derive the ancestry resolved edge annotations.
+It is important to note, however, that the example is a sample from
+the so-called ``big'' ARG \emph{process} (see Appendix XXX)
+and has several features that we are unlikely to infer in practice.
+For example, by definition the samples contain no genetic information about
+node \textsf{j} and so it is unlike to be inferred.
+Likewise, is is unlikely we would have information to infer the ``grand MRCA''
+\textsf{w} as all regions of the genome have already reached a
+common ancestor separately
+(in nodes \textsf{o}, \textsf{u}  and \textsf{v}).
+
+% Third, there exist nodes such as
+% \textsf{k} with two children (i.e. ``common ancestor'' nodes in Hudson's terminology)
+% but in which no ancestral coalescence occurs. Fourth, some breakpoint positions occur
+% within regions of genome that are not ancestral (e.g. the breakpoint at position 3 in
+% node \textsf{k}). In this case sample-resolved edge annotations lose the exact position of
+% the breakpoint, although they retain the information that a breakpoint occurred between
+% positions 2 and 4 (which is sufficient for calculating the likelihood, see appendix XXX).
+
+Given the local inheritance edge annotations
+(Fig~\ref{fig-ancestry-resolution}A),
+ we obtain the sample-resolved edge annotation
+(Fig~\ref{fig-ancestry-resolution}B) by traversing the graph backwards in time
+from sample nodes (\textsf{a}, \textsf{b}, \textsf{c}).
+Proceeding in topological order (such that an ancestor node is visited
+after all of its descendants) the ancestral material for a node is the
+union of the ancestral intervals on its inbound edges. We then annotate
+the outbound edges with the intersection of their local inheritance and
+just-computed ancestral intervals for the node.
+For example, node \textsf{d} in Fig~\ref{fig-ancestry-resolution}
+carries ancestral material on the interval $(0, 4]$, since it
+is the left-hand parent of the recombinant \textsf{b}. The edge
+joining \textsf{d} to its ancestor \textsf{h} then carries ancestral
+material on $(0, 4]$ since although \textsf{d} inherits the entire interval
+$(0, 7]$ from \textsf{h}, only $(0, 4]$ is ancestral to the sample.
+This process is directly analogous to Hudson's simulation algorithm
+(Appendix XXX).
+
+These alternatives of recording the direct inheritance of genetic material from
+parent to child vs the propagation of genomic intervals ancestral to the sample
+are analogous to running forwards-time vs backwards-time simulations.
+Indeed, the process of taking a graph with edges annotated with direct parent-child
+inheritance (Fig~\ref{fig-ancestry-resolution}A) and producing the equivalent
+``ancestry resolved'' graph in which only those genomic intervals that are
+ancestral to the sample are retained (Fig~\ref{fig-ancestry-resolution}B)
+is the same key algorithm underlying recent advances in forwards-time simulation.
+\cite{kelleher2018efficient} developed a method in which forwards-time
+simulations are performed by recording all parent-child genetic relationships
+(i.e.~annotating the edges in the simulated pedigree with the genomic
+intervals inherited by a child from a parent) and periodically ``simplifying''
+this structure to retain only ancestral relationships that are relevant
+to the ancestry of the current population. In many situations, recording
+the full genetic ancestry of the population in this way is orders of magnitude faster
+than the alternative of tracking
+sequences~\citep{kelleher2018efficient,haller2018tree}.
 
@@
 one child. Fig.~\ref{fig-simplification} show a series of successive steps of
-simplification, starting with a complete gARG simulated under a Wright-Fisher
+simplification, starting with a complete gARG simulated under a
+(backwards-time) Wright-Fisher
 model, shown both as a graph on the left, and the equivalent set of 6
@@
 ``\nameref{Ancestry_resolution}'' section).
-Such unary nodes in a local tree are not usually considered important,
-because they provide very little information and are not
-detectable by information from any given individual tree.
-The nodes in a tree usually mark the points where samples find a
+Such unary nodes in a local tree are not usually considered important;
+% because they provide little information and are not
+% detectable by information from any given individual tree.
+nodes usually mark the points where samples find a
  \emph{common} ancestor, such as node \textsf{q} in the highlighted
@@
 rootward direction (a ``super-diamond'') is non-identifiable and can be
-replaced by a single edge. This definition includes the case
-of a single node that has one inbound and one outbound edge, such as
+replaced by one edge. This definition includes the case
+of a node that has one inbound and one outbound edge, such as
 nodes \textsf{f} and \textsf{h}.

### 2022-08-10 paper.tex
New sections on efficiency and recombination

@@
 
-\section*{ARGs and SPRs}
-\citet{wiuf1999ancestry,wiuf1999recombination} recast
-the coalescent with recombination
-as a stochastic process along the genome rather than backwards in time.
-In this formulation we begin at one end of a chromosome by
-sampling a tree under the standard coalescent, and then reason
-about how this tree is modified by recombination events
-as we move along the genome. Although this algorithm is
-significantly more complicated to simulate than Hudson's
-backwards in time approach, it has led to substantial
-insights into the process of recombination itself
-and to the sequentially Markov Coalescent (SMC)
-approximations~\citep{mcvean2005approximating,marjoram2006fast}
-that are central to many modern methods.
-
-[Mention the
-and ClonalFrame~\citep{didelot2007inference} approximation somewhere.]
-
-Under the coalescent with recombination (or SMC) each local
-tree along the genome is transformed into the next by
-a ``subtree prune-and-regraft'' (SPR)
+\section*{Relationship between ARGs and trees}
+
+We have seen that the local tree for a given position on
+the genome can be extracted from an ARG by starting at the
+leaves and following the appropriate paths through the graph.
+However, this tells us little about the relationship \emph{between}
+adjacent trees, and what properties the overall emsemble of
+local trees might have.
+A recombination between two adjacent trees causes them to
+differ by a ``subtree prune-and-regraft'' (SPR)
 % Hein et al call it the ``subtree transfer'' operation but it's
@@
 compute~\citep{hein1996complexity,allen2001subtree,bordewich2005computational}.
-
-[Discussion of SPRs in an EARG]
 
@@
 and so there can be many possible ARGs compatible with
-a given list of
-% want to say "standard phylogenetic" trees here or something, but
-% and they don't have to be binary (just arity > 1), but thought
-% this would be familiar to readers?
-% Jere: Could just say leaf-labelled "marginal trees", "ancestral trees",
-% "local trees", or something like that.
-leaf-labelled binary trees.
-
-[ Cite \citep{deng2021distribution} somewhere here.]
-
-% This text might go somewhere
-% We may have nodes corresponding to genomes that are ancestral to
-% the sample along their entire span
-% (e.g.\ node $H$ in Fig.~\ref{fig-ancestry-resolution}),
-% genomes that carry no genetic material ancestral to the sample at all
-% (e.g.\ node $J$ in Fig.~\ref{fig-ancestry-resolution}),
-% and every possibility along the continuum between these extremes.
-% Clearly, we can only ever hope to infer the fraction of an ancestor's
-% genome that is ancestral to the sample.
+a given list of leaf-labelled binary trees.
+This is the approach taken by the Espalier inference
+method~\citep{rasmussen2022espalier}, which finds a
+set of SPRs that reconcile independently inferred
+trees from different regions of the genome
+using.
+
+\section*{Precision of recombination information}
+
+% Not sure where this goes, but there are some critical points
+% that we want to explain and not just throw out there in the
+% discussion.
+A critical distinction between the eARG and gARG encodings
+is the level of precision about the time and location of
+recombination events that is required. In an eARG, we are
+required to be absolutely precise about every recombination
+[explain and contrast with gARG, referring to examples]
+
+For example, \citep{deng2021distribution} state that
+``[the ARG] provides
+all the information about the genealogical history of a sample,
+including the locations of recombination events.''
+
+% JK: This is very rough, just jotting down the basic ideas,
+% will need substantial revision.
+It is not clear that such complete information will always
+(or indeed ever) be available from inferences. We may be
+able to precisely identify the location and timing
+of recent recombinations (particularly if detailed pedigree
+information is available) but beyond this, the information
+required simply isn't present. When sampling from a
+process such as the Sequentially Markov
+Coalescent~\citep{mcvean2005approximating,marjoram2006fast}
+like ARGweaver~\citep{rasmussen2014genome}, it is true that
+we do obtain a fully resolved ARG with precise
+information about recombination, but it is reasonable
+to question how much information from the data is used
+to support older recombinations, in particular.
+
+At the very least, it is not reasonable for the \emph{data
+structure} to require complete precision in order to
+represent some less precise. This is a fundamental
+advantage of the gARG encoding over the classical eARG
+method: we are not forced to posit a precise cause (recombination
+event) for every effect (difference between adjacent trees).
+These differences in precision of recombination
+information are illustrated in the ARGs estimated by different
+methods, as we see in the next section.
 
@@
 
+\section*{Efficiency of computation on ARGs}
+The succinct tree sequence data structure (usually known as a ``tree sequence''
+for brevity, but see later comment on confusion around this point)
+is a concrete encoding of a gARG, as discussed in this article.
+It was originally developed as part of the \texttt{msprime}
+simulator~\cite{kelleher2016efficient} and subsequently ...
+
+[The tree sequence encoding is very efficient for lots of things. Examples]
+
+In retrospect, the choice of terminology was unfortunate, and has led to
+significant confusion. [More]
+
 \section*{Discussion}
@@
 \begin{itemize}
-\item [Recap quickly distinction bewtween aARG and gARG. Then]
+
+\item Recap quickly distinction bewtween aARG and gARG.Then:
     This, coupled with

### 2022-08-10 paper.tex
Move arg inference methods to appendix

@@
 to discuss recombining genealogies, leading to basic errors in statements
-about what different inference methods produce as output (see the
-``\nameref{ARG_inference_methods}'' section), as well as presenting serious
+about what different inference methods produce as output, as well as presenting serious
 difficulties in either comparing the outputs of different methods or
@@
 
-\section*{ARG inference methods}\label{ARG_inference_methods}
-Using the terminology developed here to classify ARGs, we now review methods developed
-to explicitly \emph{infer} ARGs from sequencing data.
-% Gregor: I expected clear indication which method/tool works with eARGs and gARGs, but I can see this clearly
-
-\subsection*{ARG reconstruction}
-
-The problem of reconstructing ARGs for samples of recombining sequences has been of
-interest since the ARG was first defined. Early methods focused on finding \emph{parsimonious} ARGs,
-i.e.\ those with a minimal number of recombination events \citep{hein1990reconstructing}.
-Two main approaches emerged: \emph{backwards-in-time} \citep{lyngso2005minimum} and
-\emph{along-the-genome} \citep{song2003parsimonious, song2005constructing}.
-The former starts with a data matrix and reduces it to an empty matrix through row and column operations
-corresponding to coalescence, mutation, and recombination events, which construct
-an ARG from the bottom up. The latter begins from an initial local tree at a single focal site. Moving the
-focal site along the genome changes the local tree via a \emph{subtree prune and regraft} (SPR) operation
-whenever a recombination is encountered. Methods based the backwards-in-time approach are described by
-\citet{song2005efficient, wu2008association, thao2019hybrid, ignatieva2021kwarg},
-while \citet{hein1993heuristic, wu2011new, mirzaei2017rent} describe along-the-genome methods.
-\citet{rasmussen2022espalier} focuses on parsimonious fusion of local trees into an ARG, while
-\citet{camara2016inference} is based on topological data analysis.
-
-Reconstructing a most parsimonious ARG for a given data set is NP-hard \citep{wang2001perfect},
-so parsimony-based methods resort to heuristics and are limited to analysing at most hundreds of sequences.
-Hence, a number of methods aim to balance computational efficiency with reconstruction of ``reasonable",
-rather than parsimonious ARGs
-\citep{minichiello2006mapping, parida2008estimating, kelleher2019inferring,  speidel2019method, zhang2021biobank}.
-The latter three methods, as well as the parsimony-based method of \citet{rasmussen2022espalier},
-can be used on megabase scale data and hundreds of thousands of samples under human-like parameters.
-
-TODO mention \citep{schaefer2021ancestral} somewhere.
-
-% AI: just to note, these papers do not all agree in their definition of an ARG.
-% I guess this is one of the points of the present paper, but should we mention this fact?
-% Should we also mention that most of these infer just the topology, but some also the times?
-
-An alternative approach is to treat the ARG as a latent parameter to be averaged out
-by Monte Carlo methods, based either on importance sampling
-\citep{griffiths1996ancestral, fearnhead2001estimating, jenkins2011inference}
-or MCMC \citep{kuhner2000maximum, nielsen2000estimation, wang2008bayesian, fallon2013acg, mahmoudi2022bayesian}.
-These methods operate on representations of the ``little ARG'', and are extremely computationally
-expensive, being applicable to at most hundreds of samples consisting of tens or hundreds of kilobases with
-human-like parameters. State-of-the-art methods rely on cheaper, approximate models
-\citep{didelot2010inference, heine2018bridging, hubisz2020mapping,hubisz2020inference, medina2020speeding}.
-The most scalable method, \texttt{ARGWeaver} \citep{rasmussen2014genome}, can be applied to dozens of mammal-like
-genomes \citep{hubisz2020inference}.
-A central quantity in all of these sampling methods is the conditional sampling probability
-$p(G | D) \propto p(D | G) p(G)$ of an ARG $G$ given an observed realization
-of genetic diversity $D$ at its sampled leaves, where $p(D | G)$ is the likelihood
-of the data $D$ given $G$, and $p(G)$ is a Bayesian prior distribution or a
-frequentist regularizer for the ARG $G$.
-
-The likelihood $p(D | G)$ is effectively always the distribution of
-a Poisson process on the ancestral material along the edges of $G$ \citep[Eq.\ (2)]{mahmoudi2022bayesian}.
-Hence it is typically straightforward to evaluate in principle, but choosing the
-right data structure can still have a dramatic effect on the efficiency of evaluation
-\citep{mahmoudi2022bayesian}. Especially in the Bayesian case, $p(G)$ is typically specified
-as the sampling distribution of a stochastic process, such as the coalescent with
-recombination. Two contemporary examples are
-\cite{mahmoudi2022bayesian, guo2022recombination}.
-Evaluating the distribution of the coalescent with recombination
-(or related models) requires knowledge of the number of extant lineages and
-number of links available for recombination (i.e.\ ones at which a recombination
-would split ancestral material) at each time \citep[Eq.\ (3)]{mahmoudi2022bayesian}.
-These require a data structure from which shared edges on different trees are easy to identify,
-and which encodes recombination events and times explicitly, e.g.\ using nodes.
-We will provide a more detailed discussion on sampling probability evaluation in the appendix.
-
 \subsection*{Differences in reconstructed ARGs}
 
-There are a vast number of ARGs compatible with a given dataset, so ARG reconstruction tools generally produce different outputs for the same input.
-Figure \ref{fig-inferred-args} shows examples of the ARG topologies produced by four of the tools mentioned above, applied to the classical dataset of 11 \emph{Drosophila melanogaster} sequences of \citet{kreitman1983nucleotide}.
-
-There is some consistency among the four inferred ARGs (such as the samples FR-F, WA-F and Af-F being grouped into the same clade), but also considerable differences in the order and number of events.
-\texttt{KwARG} \citep{ignatieva2021kwarg} produces a parsimonious ARG with seven recombination breakpoints, which has been shown to be the true minimum required to explain the dataset when disallowing recurrent mutation \citep{song2003parsimonious}; this is a lower bound that, in general, significantly underestimates the actual number of recombinations that might have occurred.
-An ARG drawn from the posterior sample produced by \texttt{ARGweaver} has 17 breakpoints.
-Both tools explicitly record the sequence undergoing each recombination event, and the precise location of the breakpoint along the genome.
-\texttt{Relate} \citep{speidel2019method} and \texttt{tsinfer} \citep{kelleher2019inferring} do not place recombination events onto the ARG topology, therefore the transitions between local trees are not explicitly inferred in terms of SPR operations.
-\texttt{Relate} first infers local tree topologies and then identifies edges that persist from one tree to the next as a separate step, while \texttt{tsinfer} explicitly constructs nodes that are shared across local trees.
-The number of inferred trees also depends on the parameter settings; all four tools can allow for the presence of sequencing errors and recurrent mutations, reducing the number of inferred local trees (in this case, based on the input mutation and recombination rate \texttt{Relate} hence posits only two recombination breakpoints).
-
-Note that all four methods produce output which can be represented in the succinct tree sequence format, and which is consistent with our definition of an ARG as a data structure.
-The ARGs produced by \texttt{KwARG} and \texttt{ARGweaver} contain more nodes to record the recombination events (corresponding to Figure 4 stage `B'), while the ARGs output by \texttt{tsinfer} and \texttt{Relate} being, in essence, simplified (roughly corresponding to Figure 4 stages `C' and 'D', respectively).
-
-\begin{figure}
-\begin{center}
-    \includegraphics[width=\textwidth]{illustrations/inference.pdf}
-\end{center}
-\caption{\label{fig-inferred-args}
-ARGs inferred by different methods from a 2.7kb region of the \textit{Drosophila melanogaster}
-ADH locus for 11 samples~\citep{kreitman1983nucleotide}. Note that the position of points on the
-Y axis is that chosen for graphical clarity by the graphviz \texttt{dot} algorithm,  rather than being
+There are a vast number of ARGs compatible with a given dataset, so ARG
+reconstruction tools generally produce different outputs for the same input.
+Figure \ref{fig-inferred-args} shows examples of the ARG topologies produced by
+four of the tools mentioned above, applied to the classical dataset of 11
+\emph{Drosophila melanogaster} sequences of \citet{kreitman1983nucleotide}.
+
+There is some consistency among the four inferred ARGs (such as the samples
+FR-F, WA-F and Af-F being grouped into the same clade), but also considerable
+differences in the order and number of events. \texttt{KwARG}
+\citep{ignatieva2021kwarg} produces a parsimonious ARG with seven recombination
+breakpoints, which has been shown to be the true minimum required to explain
+the dataset when disallowing recurrent mutation \citep{song2003parsimonious};
+this is a lower bound that, in general, significantly underestimates the actual
+number of recombinations that might have occurred. An ARG drawn from the
+posterior sample produced by \texttt{ARGweaver} has 17 breakpoints. Both tools
+explicitly record the sequence undergoing each recombination event, and the
+precise location of the breakpoint along the genome. \texttt{Relate}
+\citep{speidel2019method} and \texttt{tsinfer} \citep{kelleher2019inferring} do
+not place recombination events onto the ARG topology, therefore the transitions
+between local trees are not explicitly inferred in terms of SPR operations.
+\texttt{Relate} first infers local tree topologies and then identifies edges
+that persist from one tree to the next as a separate step, while
+\texttt{tsinfer} explicitly constructs nodes that are shared across local
+trees. The number of inferred trees also depends on the parameter settings; all
+four tools can allow for the presence of sequencing errors and recurrent
+mutations, reducing the number of inferred local trees (in this case, based on
+the input mutation and recombination rate \texttt{Relate} hence posits only two
+recombination breakpoints).
+
+Note that all four methods produce output which can be represented in the
+succinct tree sequence format, and which is consistent with our definition of
+an ARG as a data structure. The ARGs produced by \texttt{KwARG} and
+\texttt{ARGweaver} contain more nodes to record the recombination events
+(corresponding to Figure 4 stage `B'), while the ARGs output by
+\texttt{tsinfer} and \texttt{Relate} being, in essence, simplified (roughly
+corresponding to Figure 4 stages `C' and 'D', respectively).
+
+\begin{figure} \begin{center}
+\includegraphics[width=\textwidth]{illustrations/inference.pdf} \end{center}
+\caption{\label{fig-inferred-args} ARGs inferred by different methods from a
+2.7kb region of the \textit{Drosophila melanogaster} ADH locus for 11
+samples~\citep{kreitman1983nucleotide}. Note that the position of points on the
+Y axis is that chosen for graphical clarity by the graphviz \texttt{dot}
+algorithm,  rather than being
 a direct measure of time. Additional notes about these ARGs.
@@
 
+\subsection*{Survey of ARG inference methods}
+
+The problem of reconstructing ARGs for samples of recombining sequences has
+been of interest since the ARG was first defined. Early methods focused on
+finding \emph{parsimonious} ARGs, i.e.\ those with a minimal number of
+recombination events \citep{hein1990reconstructing}. Two main approaches
+emerged: \emph{backwards-in-time} \citep{lyngso2005minimum} and
+\emph{along-the-genome} \citep{song2003parsimonious, song2005constructing}. The
+former starts with a data matrix and reduces it to an empty matrix through row
+and column operations corresponding to coalescence, mutation, and recombination
+events, which construct an ARG from the bottom up. The latter begins from an
+initial local tree at a single focal site. Moving the focal site along the
+genome changes the local tree via a \emph{subtree prune and regraft} (SPR)
+operation whenever a recombination is encountered. Methods based the
+backwards-in-time approach are described by \citet{song2005efficient,
+wu2008association, thao2019hybrid, ignatieva2021kwarg}, while
+\citet{hein1993heuristic, wu2011new, mirzaei2017rent} describe along-the-genome
+methods. \citet{rasmussen2022espalier} focuses on parsimonious fusion of local
+trees into an ARG, while \citet{camara2016inference} is based on topological
+data analysis.
+
+Reconstructing a most parsimonious ARG for a given data set is NP-hard
+\citep{wang2001perfect}, so parsimony-based methods resort to heuristics and
+are limited to analysing at most hundreds of sequences. Hence, a number of
+methods aim to balance computational efficiency with reconstruction of
+``reasonable", rather than parsimonious ARGs \citep{minichiello2006mapping,
+parida2008estimating, kelleher2019inferring,  speidel2019method,
+zhang2021biobank}. The latter three methods, as well as the parsimony-based
+method of \citet{rasmussen2022espalier}, can be used on megabase scale data and
+hundreds of thousands of samples under human-like parameters.
+
+TODO mention \citep{schaefer2021ancestral} somewhere.
+
+% AI: just to note, these papers do not all agree in their definition of an
+% ARG. I guess this is one of the points of the present paper, but should we
+% mention this fact? Should we also mention that most of these infer just the
+% topology, but some also the times?
+
+An alternative approach is to treat the ARG as a latent parameter to be
+averaged out by Monte Carlo methods, based either on importance sampling
+\citep{griffiths1996ancestral, fearnhead2001estimating, jenkins2011inference}
+or MCMC \citep{kuhner2000maximum, nielsen2000estimation, wang2008bayesian,
+fallon2013acg, mahmoudi2022bayesian}. These methods operate on representations
+of the ``little ARG'', and are extremely computationally expensive, being
+applicable to at most hundreds of samples consisting of tens or hundreds of
+kilobases with human-like parameters. State-of-the-art methods rely on cheaper,
+approximate models \citep{didelot2010inference, heine2018bridging,
+hubisz2020mapping,hubisz2020inference, medina2020speeding}. The most scalable
+method, \texttt{ARGWeaver} \citep{rasmussen2014genome}, can be applied to
+dozens of mammal-like genomes \citep{hubisz2020inference}. A central quantity
+in all of these sampling methods is the conditional sampling probability $p(G |
+D) \propto p(D | G) p(G)$ of an ARG $G$ given an observed realization of
+genetic diversity $D$ at its sampled leaves, where $p(D | G)$ is the likelihood
+of the data $D$ given $G$, and $p(G)$ is a Bayesian prior distribution or a
+frequentist regularizer for the ARG $G$.
+
+The likelihood $p(D | G)$ is effectively always the distribution of a Poisson
+process on the ancestral material along the edges of $G$ \citep[Eq.\
+(2)]{mahmoudi2022bayesian}. Hence it is typically straightforward to evaluate
+in principle, but choosing the right data structure can still have a dramatic
+effect on the efficiency of evaluation \citep{mahmoudi2022bayesian}. Especially
+in the Bayesian case, $p(G)$ is typically specified as the sampling
+distribution of a stochastic process, such as the coalescent with
+recombination. Two contemporary examples are \cite{mahmoudi2022bayesian,
+guo2022recombination}. Evaluating the distribution of the coalescent with
+recombination (or related models) requires knowledge of the number of extant
+lineages and number of links available for recombination (i.e.\ ones at which a
+recombination would split ancestral material) at each time \citep[Eq.\
+(3)]{mahmoudi2022bayesian}. These require a data structure from which shared
+edges on different trees are easy to identify, and which encodes recombination
+events and times explicitly, e.g.\ using nodes.
+We will provide a more detailed discussion on sampling probability evaluation in the appendix.
 \end{document}

### 2022-08-10 paper.tex
Move method survey to appendix

@@
 
-\subsection*{Differences in reconstructed ARGs}
-
-There are a vast number of ARGs compatible with a given dataset, so ARG
-reconstruction tools generally produce different outputs for the same input.
-Figure \ref{fig-inferred-args} shows examples of the ARG topologies produced by
-four of the tools mentioned above, applied to the classical dataset of 11
-\emph{Drosophila melanogaster} sequences of \citet{kreitman1983nucleotide}.
+\section*{Differences in reconstructed ARGs}
+ARGs inferred by different methods vary greatly in their properties.
+To illustrate this, we examine the inferences made by four
+recent methods on the classical \citet{kreitman1983nucleotide}.
+\emph{Drosophila melanogaster} dataset.
+
+\begin{figure} \begin{center}
+\includegraphics[width=\textwidth]{illustrations/inference.pdf} \end{center}
+\caption{\label{fig-inferred-args} ARGs inferred by different methods from a
+2.7kb region of the \textit{Drosophila melanogaster} ADH locus for 11
+samples~\citep{kreitman1983nucleotide}. Note that the position of points on the
+Y axis is that chosen for graphical clarity by the graphviz \texttt{dot}
+algorithm,  rather than being
+a direct measure of time. Additional notes about these ARGs.
+}
+\end{figure}
 
@@
 corresponding to Figure 4 stages `C' and 'D', respectively).
-
-\begin{figure} \begin{center}
-\includegraphics[width=\textwidth]{illustrations/inference.pdf} \end{center}
-\caption{\label{fig-inferred-args} ARGs inferred by different methods from a
-2.7kb region of the \textit{Drosophila melanogaster} ADH locus for 11
-samples~\citep{kreitman1983nucleotide}. Note that the position of points on the
-Y axis is that chosen for graphical clarity by the graphviz \texttt{dot}
-algorithm,  rather than being
-a direct measure of time. Additional notes about these ARGs.
-}
-\end{figure}
 

### 2022-08-24 paper.tex
Authorship

@@
 \title{A general and efficient representation of Ancestral Recombination Graphs}
+
 % First authors
-\author{Author McAuthorface}
+\author[1]{Yan Wong}
+
+% Second authors
+\author[2,$\star$]{Anastasia Ignatieva}
+\author[3,$\star$]{Jere Koskela}
+
+% Middle Authors
+\author[4]{Gregor Gorjanc}
+\author[5]{Anthony W. Wohns}
+
 % Corresponding
+\author[1,$\dagger$]{Jerome Kelleher}
+
+\affil[1]{Big Data Institute, Li Ka Shing Centre for Health Information and Discovery, University of Oxford, OX3 7LF, UK}
+\affil[2]{Department of Statistics, University of Oxford, OX1 3LB, UK}
+\affil[3]{Department of Statistics, University of Warwick, CV4 7AL, UK}
+\affil[4]{The Roslin Institute and Royal (Dick) School of Veterinary Studies, University of Edinburgh, EH25 9RG, UK}
+\affil[5]{Broad Institute of MIT and Harvard, Cambridge, MA 02142, USA}
+
+\affil[$\star$]{Denotes shared second authorship, listed alphabetically}
+\affil[$\dagger$]{Denotes corresponding author}
 

### 2022-08-25 paper.tex
Minor rephrasing to reduce word count
And also to introduce the idea of local trees early on.

--- 2022-08-10 paper.tex
@@
 and compared to determine their strengths and weaknesses.
-Unfortunately, however, there is substantial confusion over
-terminology and little agreement on shared standards for data interchange,
-significantly hampering progress in this important area.
-This confusion and lack of agreement on basic issues such as what does and
+Unfortunately confusion exists over terminology and there is
+little agreement on shared standards for data interchange,
+significantly hampering progress in the field.
+This lack of clarity and agreement on basic issues such as what does and
 does not constitute an ARG is partly attributable to the historical
@@
 (inherited from the original stochastic process definition) is
-ambiguous, fundamentally limited and an unsuitable basis for data interchange.
+ambiguous, 
+% is it ambiguous? How? Can we simply delete this?
+fundamentally limited and an unsuitable basis for data interchange.
 We provide a simple alternative definition of an ARG in which a node
-corresponds to one of an individual's monoploid genomes, an edge
-% jk-note: this is clumsy
-defines a genetic ancestry relationship between two nodes,
-and annotations on those edges define the patterns of
-inheritance in intervals along the genome.
+corresponds to one of an individual's monoploid genomes and an edge
+defines genetic inheritance between two nodes, with annotations on
+each edge specifying which regions of the genome have
+been inherited via that route.
 We show how this definition can be used to generalise the encoding to
@@
 Inferring the genetic relationships between sampled genomes in the form of an
-evolutionary tree is a necessary prerequisite for many analyses of
+evolutionary or phylogenetic tree is a necessary prerequisite for many analyses of
 species~\citep{rannala2003genetics}, non-recombining sections of
@@
 and viruses~\citep{grenfell2004science}.
-A genealogical tree captures what is known and
-knowable about the genetic ancestry of the sample in an elegant and concise
-way, by postulating a set of nodes that are ancestors common to the samples,
-and the ancestral relationships between those nodes. A rich literature exists
+A phylogenetic tree captures information about the genealogy
+of the sample in an elegant and concise
+way, by postulating a set of ancestors common to the samples
+together with paths of inheritance between them. A rich literature exists
 on analysing the mathematical properties of these
 trees~\citep{steel2016phylogeny}, and numerous
-methods exist to infer them~\citep{felsenstein2004inferring}. The situation is
-quite different in recombining organisms, where the analysis of genetic data
-depends on summaries ultimately derived from the
-trees~\citep{tajima1983evolutionary,tavare1984line}
-% Do we mean "despite the fact that analysis of genetic data relies on methods
-% which ultimately derive from the inheritance paths defined by genetic trees
-% and the mutations upon them, analysis so far has focussed on summaries ...
-such as allele frequencies and site frequency
+methods exist to infer them~\citep{felsenstein2004inferring}.
+However, the genetic genealogy of a recombining organism
+cannot be described as a single evolutionary tree: instead,
+the inheritance of each local section of the genome may be described
+by a (slightly) different tree~\citet{hudson1983properties}. Inferring complex
+genealogies of this nature
+presents profound technical difficulties, meaning tree-based approaches are
+rarely used in population and statistical genetics. Therefore until recently, analysis of
+population genetic data has primarily focussed on summaries derived
+from the local trees~\citep{tajima1983evolutionary,tavare1984line}
+rather than using the underlying genealogy itself. Such
+derived measures include allele frequencies and site frequency
 spectra~\citep{achaz2009frequency,ralph2020efficiently},
-patterns of linkage disequilibrium~\citep{mcvean2002genealogical}, and principal
-components~\citep{mcvean2009genealogical}. Genealogical analysis is equally
-fundamental to the study of recombining organisms, but profound technical
-difficulties have made the routine inference of recombinant genealogies
-impossible, and therefore tree-based analysis methods have been largely
-unavailable to the fields of population and statistical
-genetics.
+patterns of linkage disequilibrium~\citep{mcvean2002genealogical}, and
+principal components~\citep{mcvean2009genealogical}.
 
@@
 of the population and statistical genetics toolkit~\citep{hejase2020summary}.
-Applications using these inferred ancestries as input have
-begun to
+Applications using these inferred ancestries as input have begun to
 appear~\citep{osmond2021estimating,zhang2021biobank,fan2022genealogical}
@@
 follow~\citep{harris2019database}. This vibrant research area, however,
-has a significant difficulty, which, if not addressed, will substantially hamper
-progress. We currently lack a well-defined data-model and shared terminology
-to discuss recombining genealogies, leading to basic errors in statements
+has a significant difficulty, which risks hampering
+progress: we currently lack a well-defined data-model and shared terminology
+to discuss recombining genealogies. This leads to basic errors in statements
 about what different inference methods produce as output, as well as presenting serious
@@
 underpinning for much of contemporary population genetics.
-The coalescent is a stochastic process, where each random realisation
-is a genealogical tree describing the genetic
-ancestors of the samples and their inheritance relationships.
-A great strength of the coalescent as a modelling framework
-is its flexibility, and numerous elaborations have been
-incorporated~\citep{hudson1990gene,hein2004gene,wakely2008coalescent}.
-In the first extension to the basic process, \cite{hudson1983properties}
+It is a stochastic process, where each random realisation
+is a genealogical tree describing the genetic ancestry of the samples.
+Numerous elaborations have been incorporated into the
+basic process~\citep{hudson1990gene,hein2004gene,wakely2008coalescent};
+in particular, early work by \citet{hudson1983properties}
 analysed the effects of recombination on the random genealogical
-trees generated by the coalescent,
-producing several fundamental analytical results
+trees generated by the coalescent. This
+generated several fundamental analytical results
 as well as describing the elegant algorithm
@@
 local trees was analysed analytically, those correlations
-were not captured explicitly: the ancestry of the sample
+were not captured explicitly: the ancestry
 was viewed as a series of disconnected local trees rather than
@@
 inferring parameters of the
-stochastic process, where the ancestry is regarded as a
+stochastic process, where the ancestry was regarded as a
 latent parameter to be averaged out
@@
 have quite different properties from realisations of the
-(previously synonymous) ARG \emph{process}. Firstly, branch lengths are
-not typically estimated, and node times are required
-order to compute a likelihood under the coalescent with recombination
-(see section XXX). Secondly, even if branch lengths were estimated,
+ARG \emph{process} as modelled by Hudson, Griffiths, and colleagues.
+Firstly, branch lengths are not typically estimated,
+% why do we need the following two lines
+and node times are required in order to compute a likelihood under the
+coalescent with recombination (see section XXX).
+Secondly, even if branch lengths were estimated,
 the realisations would have a very small likelihood since
@@
 and as concrete realisations of ancestry was now therefore broken.
-\cite{minichiello2006mapping} made this separation of the ARG as a
-stochastic process vs the ARG as a data structure explicit, and
+\cite{minichiello2006mapping} explicitly distinguished the ARG as a
+stochastic process from the ARG as a data structure, and
 argued for using the term to mean specific realised ancestral histories.
@@
 % know essentially nothing about it).
-In parallel with this tendancy to decouple the term
+In parallel with this tendency to decouple the term
 Ancestral Recombination Graph from its original defining stochastic process,
@@
 the location and timing
-of recombination events.
+of recombination events. % (and coalescent events?)
 Unless otherwise stated, we will use the ARG-as-a-data-structure
@@
 in terms of \emph{events}. Nodes represent
-common ancestor or recombination events that occured in the
+common ancestor or recombination events that occurred in the
 history of some set of sample genomes, and edges represent ancestral
@@
 (This formulation is equivalent to having different ``types''
-of node, and that breakpoints are associated with recombination
+of node, with breakpoints are associated with recombination
 nodes only).
@@
 \texttt{ARGweaver}~\citep{rasmussen2014genome}
-and \texttt{KwARG}~\citep{ignatieva2021kwarg},
-and example is given in Fig.~\ref{fig-arg-data-structure}A.
+and \texttt{KwARG}~\citep{ignatieva2021kwarg}:
+an example is given in Fig.~\ref{fig-arg-data-structure}A.
 Note that the description above captures only the
@@
 in associating recombination breakpoints with nodes
-rather than edges,
+rather than edges: we need to know from which parent the 
 and the only way we can break this symmetry is to assign
@@
 
-% TODO update this when then later sections have been drafted.
+% TODO update this when later sections have been drafted.
 As we see in sections XXX and YYY there are substantial limitations
@@
 Another assumption of the coalescent with recombination which prevents complex
-combinations of events occuring simultaneously, is that the genome (or
+combinations of events occurring simultaneously, is that the genome (or
 at least the region under study) is short enough that the number of extant

### 2022-08-25 paper.tex
Move discussion of ARGs into precision section

@@
 \section*{Precision of recombination information}
+Ancestral recombination graphs are often characterised
+as consisting of a complete record of the genealogical history
+of a sample. For example,
+\cite{rasmussen2014genome} state that an ``ARG provides a record of all
+coalescence and recombination events since the divergence of the sequences
+under study'' and
+according to \cite{deng2021distribution} an ARG
+``provides all the information about the genealogical history of a sample,
+including the locations of recombination events.'' It is worth
+questioning, however, whether such complete and precise information is
+realistically available, and in particular whether our
+\emph{representation} of genealogical history should require
+such precision.
 
@@
 [explain and contrast with gARG, referring to examples]
-
-For example, \citep{deng2021distribution} state that
-``[the ARG] provides
-all the information about the genealogical history of a sample,
-including the locations of recombination events.''
 
@@
 
+It is helpful to consider the difference between fully
+resolved binary trees and those that contain polytomies.
+
+The situation is in fact directly analogous: an ARG that
+contains fully resolved recombination information must have
+at most parents per node.
+
 At the very least, it is not reasonable for the \emph{data
@@
 methods, as we see in the next section.
-
-\section*{Differences in reconstructed ARGs}
-ARGs inferred by different methods vary greatly in their properties.
-To illustrate this, we examine the inferences made by four
-recent methods on the classical \citet{kreitman1983nucleotide}.
-\emph{Drosophila melanogaster} dataset.
 
@@
 \end{figure}
+
+To illustrate this critical point, we inferred ARGs for the
+the \citet{kreitman1983nucleotide} \emph{Drosophila melanogaster} dataset
+(classically used as a benchmark for ARG inference),
+using four recent inference methods and show the results in
+Fig~\ref{fig-inferred-args}).
 
@@
 
-Note that all four methods produce output which can be represented in the
-succinct tree sequence format, and which is consistent with our definition of
-an ARG as a data structure. The ARGs produced by \texttt{KwARG} and
-\texttt{ARGweaver} contain more nodes to record the recombination events
-(corresponding to Figure 4 stage `B'), while the ARGs output by
-\texttt{tsinfer} and \texttt{Relate} being, in essence, simplified (roughly
-corresponding to Figure 4 stages `C' and 'D', respectively).
-
-\section*{Efficiency of computation on ARGs}
+% Note that all four methods produce output which can be represented in the
+% succinct tree sequence format, and which is consistent with our definition of
+% an ARG as a data structure. The ARGs produced by \texttt{KwARG} and
+% \texttt{ARGweaver} contain more nodes to record the recombination events
+% (corresponding to Figure 4 stage `B'), while the ARGs output by
+% \texttt{tsinfer} and \texttt{Relate} being, in essence, simplified (roughly
+% corresponding to Figure 4 stages `C' and 'D', respectively).
+
+\section*{Computational efficiency}
 The succinct tree sequence data structure (usually known as a ``tree sequence''

### 2022-08-25 paper.tex
More text on the continuum of recombination information

@@
 \newcommand{\noderef}[1]{\textsf{#1}}
-% \newcommand{\msprime}[0]{\texttt{msprime}}
+\newcommand{\tsinfer}[0]{\texttt{tsinfer}}
+\newcommand{\kwarg}[0]{\texttt{KwARG}}
+\newcommand{\argweaver}[0]{\texttt{ARGweaver}}
+\newcommand{\relate}[0]{\texttt{Relate}}
+\newcommand{\espalier}[0]{\texttt{Espalier}}
 
@@
 
-\section*{Precision of recombination information}
+\section*{Precision of ancestry information}
 Ancestral recombination graphs are often characterised
@@
 including the locations of recombination events.'' It is worth
-questioning, however, whether such complete and precise information is
+questioning, however, whether such comprehensive information is
 realistically available, and in particular whether our
-\emph{representation} of genealogical history should require
+representation of genealogical history should \emph{require}
 such precision.
@@
 \includegraphics[width=\textwidth]{illustrations/inference.pdf} \end{center}
-\caption{\label{fig-inferred-args} ARGs inferred by different methods from a
-2.7kb region of the \textit{Drosophila melanogaster} ADH locus for 11
-samples~\citep{kreitman1983nucleotide}. Note that the position of points on the
+\caption{\label{fig-inferred-args} ARGs inferred by
+(A) \kwarg~\citep{ignatieva2021kwarg},
+(B) \argweaver~\citep{rasmussen2014genome,hubisz2020inference},
+(C) \tsinfer~\citep{kelleher2019inferring},
+and (D) \relate~\citep{speidel2019method}
+for 11 samples of a 2.7kb region of the \textit{Drosophila melanogaster} ADH
+locus~\citep{kreitman1983nucleotide}. The \argweaver\ example
+is a sample from the inferred posterior distribution.
+Note that the position of points on the
 Y axis is that chosen for graphical clarity by the graphviz \texttt{dot}
@@
 To illustrate this critical point, we inferred ARGs for the
-the \citet{kreitman1983nucleotide} \emph{Drosophila melanogaster} dataset
-(classically used as a benchmark for ARG inference),
+the classical \citet{kreitman1983nucleotide} benchmark dataset
 using four recent inference methods and show the results in
-Fig~\ref{fig-inferred-args}).
-
-There is some consistency among the four inferred ARGs (such as the samples
-FR-F, WA-F and Af-F being grouped into the same clade), but also considerable
-differences in the order and number of events. \texttt{KwARG}
-\citep{ignatieva2021kwarg} produces a parsimonious ARG with seven recombination
-breakpoints, which has been shown to be the true minimum required to explain
-the dataset when disallowing recurrent mutation \citep{song2003parsimonious};
-this is a lower bound that, in general, significantly underestimates the actual
-number of recombinations that might have occurred. An ARG drawn from the
-posterior sample produced by \texttt{ARGweaver} has 17 breakpoints. Both tools
-explicitly record the sequence undergoing each recombination event, and the
-precise location of the breakpoint along the genome. \texttt{Relate}
-\citep{speidel2019method} and \texttt{tsinfer} \citep{kelleher2019inferring} do
-not place recombination events onto the ARG topology, therefore the transitions
-between local trees are not explicitly inferred in terms of SPR operations.
-\texttt{Relate} first infers local tree topologies and then identifies edges
-that persist from one tree to the next as a separate step, while
+Fig~\ref{fig-inferred-args}.
+While there is some consistency among the inferred ARGs (such as the samples
+FR-F, WA-F and Af-F being grouped into the same clade), there
+are also considerable differences. For this dataset, the minimum number
+of recombination breakpoints required in the absence of recurrent mutation
+is 7~\citep{song2003parsimonious}, and the number of inferred breakpoints
+under the input mutation and recombination rates
+varies from 2 (\relate) to 17 (\argweaver).
+The ARGs inferred by the different methods also have major differences
+in how recombination information is represented. Essentially,
+\kwarg\ and \argweaver\ infer a specific causal recombination \emph{event}
+for each breakpoint, and those events are fully
+described in the output graph. On the other hand, \tsinfer\ and
+\relate\ do not infer causal events but instead focus on
+representing the inferred \emph{effect}: breakpoints between the
+trees. It is important to note that this does not imply that
+the methods must then necessarily infer only a sequence of
+unrelated local trees
+% More? See https://github.com/tskit-dev/what-is-an-arg-paper/issues/38
+% for discussion. I'm sure there were other examples of people saying
+% that
+(as claimed by e.g.~\citep{hejase2020summary}),
+and information about recombination and the correlation structure
+between trees is not ``all or nothing''.
+Both methods capture correlation structure between trees:
+\relate\ first infers local tree topologies and then identifies edges
+that persist from one tree to the next as a separate step
+(in a similar manner to \espalier~\citep{rasmussen2022espalier}), while
 \texttt{tsinfer} explicitly constructs nodes that are shared across local
-trees. The number of inferred trees also depends on the parameter settings; all
-four tools can allow for the presence of sequencing errors and recurrent
-mutations, reducing the number of inferred local trees (in this case, based on
-the input mutation and recombination rate \texttt{Relate} hence posits only two
-recombination breakpoints).
-
-% Note that all four methods produce output which can be represented in the
-% succinct tree sequence format, and which is consistent with our definition of
-% an ARG as a data structure. The ARGs produced by \texttt{KwARG} and
-% \texttt{ARGweaver} contain more nodes to record the recombination events
-% (corresponding to Figure 4 stage `B'), while the ARGs output by
-% \texttt{tsinfer} and \texttt{Relate} being, in essence, simplified (roughly
-% corresponding to Figure 4 stages `C' and 'D', respectively).
+trees.
+
+Because the ARGs inferred by \tsinfer\ and \relate\ do not contain
+explicit recombination events we cannot compute a likelihood
+under the coalescent with recombination (see appendix XXX).
+[Some stuff about whether we should see this as necessary or not,
+but maybe we should keep this for the conclusion?]
 

### 2022-08-26 paper.tex
Minor edits

@@
 Such nodes are not singly connected in the graph, but are nevertheless unary in
-all of the local trees in Fig.~\ref{fig-simplification}B. The operation to remove them,
+all of the local trees. The operation to remove them,
 the results of which are shown in Fig.~\ref{fig-simplification}C,
@@
 leaves and following the appropriate paths through the graph.
-However, this tells us little about the relationship \emph{between}
-adjacent trees, and what properties the overall emsemble of
-local trees might have.
+However, this tells us little about the relationship between
+adjacent trees, and what the properties of overall ensemble of
+local trees are.
 A recombination between two adjacent trees causes them to
@@
 the genome in an unsimplified gARG are easy to interpret.
-Consider the highlighed tree in Fig~\ref{fig-simplification}A.
+Consider the highlighted tree in Fig~\ref{fig-simplification}A.
 In it the parent of \textsf{e} is \textsf{j}, whereas in the next tree
@@
 set of SPRs that reconcile independently inferred
-trees from different regions of the genome
-using.
+trees from different regions of the genome.
 
@@
 Ancestral recombination graphs are often characterised
-as consisting of a complete record of the genealogical history
+as consisting of a complete record of the genetic history
 of a sample. For example,
@@
 including the locations of recombination events.'' It is worth
-questioning, however, whether such comprehensive information is
-realistically available, and in particular whether our
+questioning, however, whether such comprehensive information
+will always be available, and in particular whether our
 representation of genealogical history should \emph{require}
-such precision.
+such precision. Simulations will usually generate complete
+information about the history of a sample, of course, but
+we may not have sufficient information to infer such detail.
 
@@
 
-% JK: This is very rough, just jotting down the basic ideas,
-% will need substantial revision.
-It is not clear that such complete information will always
-(or indeed ever) be available from inferences. We may be
-able to precisely identify the location and timing
-of recent recombinations (particularly if detailed pedigree
-information is available) but beyond this, the information
-required simply isn't present. When sampling from a
-process such as the Sequentially Markov
-Coalescent~\citep{mcvean2005approximating,marjoram2006fast}
-like ARGweaver~\citep{rasmussen2014genome}, it is true that
-we do obtain a fully resolved ARG with precise
-information about recombination, but it is reasonable
-to question how much information from the data is used
-to support older recombinations, in particular.
-
-It is helpful to consider the difference between fully
-resolved binary trees and those that contain polytomies.
-
-The situation is in fact directly analogous: an ARG that
-contains fully resolved recombination information must have
-at most parents per node.
-
-At the very least, it is not reasonable for the \emph{data
-structure} to require complete precision in order to
-represent some less precise. This is a fundamental
-advantage of the gARG encoding over the classical eARG
-method: we are not forced to posit a precise cause (recombination
-event) for every effect (difference between adjacent trees).
-These differences in precision of recombination
-information are illustrated in the ARGs estimated by different
-methods, as we see in the next section.
+% % JK: This is very rough, just jotting down the basic ideas,
+% % will need substantial revision.
+% It is not clear that such complete information will always
+% (or indeed ever) be available from inferences. We may be
+% able to precisely identify the location and timing
+% of recent recombinations (particularly if detailed pedigree
+% information is available) but beyond this, the information
+% required simply isn't present. When sampling from a
+% process such as the Sequentially Markov
+% Coalescent~\citep{mcvean2005approximating,marjoram2006fast}
+% like ARGweaver~\citep{rasmussen2014genome}, it is true that
+% we do obtain a fully resolved ARG with precise
+% information about recombination, but it is reasonable
+% to question how much information from the data is used
+% to support older recombinations, in particular.
+
+% It is helpful to consider the difference between fully
+% resolved binary trees and those that contain polytomies.
+
+% The situation is in fact directly analogous: an ARG that
+% contains fully resolved recombination information must have
+% at most parents per node.
+
+% At the very least, it is not reasonable for the \emph{data
+% structure} to require complete precision in order to
+% represent some less precise. This is a fundamental
+% advantage of the gARG encoding over the classical eARG
+% method: we are not forced to posit a precise cause (recombination
+% event) for every effect (difference between adjacent trees).
+% These differences in precision of recombination
+% information are illustrated in the ARGs estimated by different
+% methods, as we see in the next section.
 

### 2022-08-26 paper.tex
Cleanups on history section

@@
 Applications using these inferred ancestries as input have begun to
-appear~\citep{osmond2021estimating,zhang2021biobank,fan2022genealogical}
+appear~\citep{osmond2021estimating,zhang2021biobank,fan2022genealogical,hejase2022deep}
 and many more are sure to
@@
 coalescent with recombination from a different perspective,
-in which the stochastic process itself is encoded as a graph
-(with particular properties; see section XXX for details),
+in which the stochastic process itself is encoded as a graph,
 and showed how the local trees generated by Hudson's algorithm
@@
 to describe recombinant ancestry as single object that
-could be inferred in a way
+could be reasoned about and inferred in a way
 that was missing from Hudson's treatments.
-An overview of the Hudson and Griffiths stochastic processes is provided in the appendix.
 
@@
 the coalescent with recombination is often highly \emph{un}parsimonious
-in terms of recombination events for non-trivial scaled recombination rate.
+in terms of recombination events.
 The previously close correspondance
 between the ARG as a mathematical representation of the coalescent process
-and as concrete realisations of ancestry was now therefore broken.
+and as concrete realisations of ancestry was therefore broken.
 \cite{minichiello2006mapping} explicitly distinguished the ARG as a
@@
 This has become a common
-usage~\citep{mathieson2020ancestry,brandt2021evaluation}
-[TODO expand sentence].
-Gusfield~\citep{gusfield2014recombinatorics}
-defines also defines an ARG as a concrete realisation
-and studies instances that satisfy particular
-properties such as ``galled trees''
-\citep{wang2001perfect, gusfield2004optimal}
-as well as describing algorithms for more general phylogenetic
-networks~\citep{huson2010phylogenetic}.
+usage~\citep[e.g.][]{gusfield2014recombinatorics,mathieson2020ancestry,brandt2021evaluation},
+although not universal, with authors also using the term
+in the original stochastic
+process sense~\citep[e.g.][]{nordborg2000linkage,birkner2013ancestral,
+wilton2015smc,griffiths2016coalescent}.
+
+% Gusfield~\citep{gusfield2014recombinatorics}
+% defines also defines an ARG as a concrete realisation
+% and studies instances that satisfy particular
+% properties such as ``galled trees''
+% \citep{wang2001perfect, gusfield2004optimal}
+% as well as describing algorithms for more general phylogenetic
+% networks~\citep{huson2010phylogenetic}.
 
@@
 In parallel with this tendency to decouple the term
-Ancestral Recombination Graph from its original defining stochastic process,
+Ancestral Recombination Graph from the stochastic process,
 an extensive mathematical literature has developed in
@@
 koskela2019robust}.
-Similarly, the Ancestral Gene Transfer Graph
-
-models horizontal gene transfer~\citep{baumdicker2014infinitely}
-by [FINISH ME.
 
@@
 now the most common interpretation, the \emph{form} of that
-data structure is still closely derived from the original
+data structure is still closely linked to the original
 stochastic process, and that this form is inflexible
@@
 
-\section*{ARG simplification}\label{ARG_simplification}
-
-Not all nodes in an ARG are equal. Assuming that we have sequenced the
-genomes of the samples, there are intrinsic limits on what can
-be known about more ancient nodes, determined by the graph topology
-and the flow of ancestral material through it.
-\emph{Simplification}~\citep{kelleher2018efficient} is the process
-of removing nodes and re-writing edges in a gARG to remove
-various types of redundancy, while maintaining the crucial properties
-that the ``meaningful'' topology in the local trees is not affected,
-and the relationship between nodes in adjacent trees is maintained.
-
-\begin{figure}
-\centering
-\vspace{5em}
-\includegraphics[width=\linewidth]{illustrations/simplification}
-\caption{\label{fig-simplification}
-ARG simplification. In all cases the graph visualization is shown on the left
-and the equivalent marginal trees on the right.
-(A) A gARG simulated from a diploid Wright-Fisher
-model. Note that the genome \textsf{k} has three children, and the genomes
-\textsf{e} and \textsf{g} simultaneously have multiple parents and multiple children;
-neither could be directly represented in a Griffiths-like eARG encoding
-(see the ``\nameref{eARG}'' section). The penultimate local tree is highlighted to aid discussion.
-(B) Simplified to remove all
-% trying out this terminology - can define in the text
-1-connected graph components (e.g., diamonds such as \textsf{jlnm}).
-(C) Remove nodes that are unary in all local trees. This removes both recombinant nodes
-that have only one child (e.g. \textsf{n}) and nodes that have more
-than one child (i.e. are ``common ancestors'') but never represent coalescences
-in the local trees (e.g. \textsf{r}).
-(D) Rewrite edges to bypass any unary nodes in local trees. Note that this loses lineage
-information such as the fact that \textsf{k} is connected to \textsf{a} via \textsf{e}.
-}
-\end{figure}
-
-% Q: would it help to talk about ``tree vertices'' here or would it
-% just be more confusing? I worry that people would think that
-% they are two different things then, and it's really important
-% that they get that a node-is-a-node
-
-The ideas of ``meaningful'' local tree topologies revolve around the
-presence of ``unary'' nodes: a node in the tree that has exactly
-one child. Fig.~\ref{fig-simplification} show a series of successive steps of
-simplification, starting with a complete gARG simulated under a
-(backwards-time) Wright-Fisher
-model, shown both as a graph on the left, and the equivalent set of 6
-local trees on the right (Fig.~\ref{fig-simplification}A). Consider the path above node
-\textsf{a} in the penultimate tree, highlighted in yellow and covering genomic positions
-60 to 70. On this path, before \textsf{a} meets its common ancestor with the other samples at
-node \textsf{q}, it passes through  \textsf{e},  \textsf{j},  \textsf{m},  \textsf{n} and
-\textsf{p}, which in this region of the genome are all unary nodes. This reflects the
-passage of this span of genome through the ARG towards the root (see the
-``\nameref{Ancestry_resolution}'' section).
-Such unary nodes in a local tree are not usually considered important;
-% because they provide little information and are not
-% detectable by information from any given individual tree.
-nodes usually mark the points where samples find a
- \emph{common} ancestor, such as node \textsf{q} in the highlighted
-tree. For many purposes, we would therefore consider a path directly
-joining sample \textsf{a} to node \textsf{q} to be of equivalent status to the path
-containing the intermediate unary nodes. Simplification is the process of removing
-these less informative and (for most computational purposes) inefficient nodes
-from the local trees and the corresponding annotated graph.
-
-The first type of simplification that we can perform is to remove
-graph topology that is invisible to the samples. The best
-known example of such topology is a so-called
-``diamond''~\citep{rasmussen2014genome}
-in which the two parent nodes of a recombination immediately
-join again into a common ancestor (e.g.~\textsf{j}, \textsf{l}, \textsf{m}
-and  \textsf{n} in Fig.~\ref{fig-simplification}A).
-Unless we are specifically
-interested in the recombination event or ancestral genomes,
-there is no information in this topology and the diamond can be
-replaced by a single edge. More generally, any
-subgraph which is singly-connected in both the leafward and
-rootward direction (a ``super-diamond'') is non-identifiable and can be
-replaced by one edge. This definition includes the case
-of a node that has one inbound and one outbound edge, such as
-nodes \textsf{f} and \textsf{h}.
-Fig.~\ref{fig-simplification}B shows the result of this type of
-graph topology simplification.
-
-Simplifying away diamonds will remove many unary nodes from the
-local trees, but there can still be nodes that are unary in all
-of the local trees. In particular, a node can represent a recombinant
-with multiple parents in the graph but only a single child (e.g.\ node \textsf{n}
-in Fig.~\ref{fig-simplification}B), or can represent a common ancestor with
-multiple children in the graph but in which no local coalescence takes place
-(node \textsf{r} in Fig.~\ref{fig-simplification}B).
-% point out that this is why Hudson refers to CA nodes, not coalescence nodes
-Such nodes are not singly connected in the graph, but are nevertheless unary in
-all of the local trees. The operation to remove them,
-the results of which are shown in Fig.~\ref{fig-simplification}C,
-therefore requires knowledge not just of the graph topology, but also of the
-edge annotations. Removal of recombinant nodes can produce graph nodes with
-more than 2 parents (e.g.~node \textsf{e}); likewise, removal of
-common ancestor but non-coalescent nodes can produce graph nodes with
-more than 2 children (e.g.~node \textsf{s}). These represent a ``stacking up'' of
-multiple---often indistinguishable---events into a single node,
-something which is impossible to encode in an eARG.
-
-% JK: Text moved from an earlier section. See if we want to reuse this here
-% if not already said.
-% Indeed a single node can have both multiple children
-% and multiple parents. This would be the case in Fig~\ref{fig-arg-in-pedigree} if,
-% for instance, node \textsf{a} were to have two children.
-% % It is a shame that this is not actually shown in fig 3, but I don't think
-% % we can do this unless we have 5 generations
-% Moreover, a single node can have more than 2 children, and (as we shall see in
-% the ``\nameref{ARG_simplification}'' section), more than 2 parents. Another way to
-% state this is that transitions between nodes in a gARG can encapsulate multiple
-% events in an eARG.
-
-The remaining nodes are MRCAs of some subset of the samples
-at \emph{some} positions along the genome. We still have
-some unary nodes in the local trees, but these nodes will
-correspond to a coalescence in at least one other
-local tree. For example, node  \textsf{k} is unary in the second tree
-of Fig.~\ref{fig-simplification}C, but is either binary
-or ternary in all subsequent trees (recall this is a Wright-Fisher
-simulation). The final level of simplification is to alter the edge annotations
-such that, although no nodes are removed from the graph, \emph{all}
-unary nodes disappear from the local trees (Fig.~\ref{fig-simplification}D).
-It is common for this operation to remove nodes above the top MRCA node
-in the tree, resulting in many more root nodes (here, \textsf{q} and \textsf{k}
-become roots). [And continue]
-% Note that although this last stage produces much simpler local trees, by
-% removing information about the exact paths taken by lineages through
-% the graph, it loses potentially important information about shared edges
-% between trees. It is not clear whether (c) or (d) leads to a more compact
-% representation of the ARG in terms of number of edges.
-
-\section*{Relationship between ARGs and trees}
-
-We have seen that the local tree for a given position on
-the genome can be extracted from an ARG by starting at the
-leaves and following the appropriate paths through the graph.
-However, this tells us little about the relationship between
-adjacent trees, and what the properties of overall ensemble of
-local trees are.
-A recombination between two adjacent trees causes them to
-differ by a ``subtree prune-and-regraft'' (SPR)
-% Hein et al call it the ``subtree transfer'' operation but it's
-% the same thing.
-operation~\citep{hein1990reconstructing,song2003on,song2006properties}.
-The number of SPRs required to transform one leaf-labelled tree
-into another is known as the SPR distance
-% Hein 1996 say that SPR distance is NP-hard, but Allen 2001
-% say this is wrong. But then Bordewich went ahead and said
-% yes, actually it is NP hard.
-and is NP-hard to
-compute~\citep{hein1996complexity,allen2001subtree,bordewich2005computational}.
-
-The effects of recombination as we move from left-to-right along
-the genome in an unsimplified gARG are easy to interpret.
-Consider the highlighted tree in Fig~\ref{fig-simplification}A.
-In it the parent of \textsf{e} is \textsf{j}, whereas in the next tree
-its parent is \textsf{k}. Since both parents are present as
-unary nodes in the tree, we simply change one edge in
-order to effect the transition.
-% Need revising to fit new example
-A more complex transition occurs when we have a regraft operation
-that occurs above the current root (i.e., the current oldest
-non-unary node). For example, the transition between the
-second and third trees corresponds to node D switching
-from the path through I to H, and we then must add branches
-corresponding to passing through L and meeting at M.
-
-The equivalence between an ARG and a set of local trees
-separated by SPRs is often mentioned [e.g. x y z], but
-it is important to note that while extracting local
-trees from an ARG is straightforward, the converse
-process of reconstructing precisely the same ARG from
-these trees is more problematic. If the internal
-nodes are consistently labelled and unary nodes are
-included in the local trees (e.g., Fig~\ref{fig-simplification})
-then it is easy to exactly reconstruct the ARG.
-Otherwise, the problem is much more difficult and it
-may not be possible to uniquely reconstruct the original ARG.
-Without labels for the internal nodes, we are forced to
-to reason about which subtrees remain the same and which
-differ due to the SPR separating adjacent trees in order to
-assign node identity. Without the internal unary nodes marking
-recombinations, we must infer their existence. In
-general, there will not be a unique solution to this problem,
-and so there can be many possible ARGs compatible with
-a given list of leaf-labelled binary trees.
-This is the approach taken by the Espalier inference
-method~\citep{rasmussen2022espalier}, which finds a
-set of SPRs that reconcile independently inferred
-trees from different regions of the genome.
-
 \section*{Precision of ancestry information}
@@
 
+\section*{ARG simplification}\label{ARG_simplification}
+
+Not all nodes in an ARG are equal. Assuming that we have sequenced the
+genomes of the samples, there are intrinsic limits on what can
+be known about more ancient nodes, determined by the graph topology
+and the flow of ancestral material through it.
+\emph{Simplification}~\citep{kelleher2018efficient} is the process
+of removing nodes and re-writing edges in a gARG to remove
+various types of redundancy, while maintaining the crucial properties
+that the ``meaningful'' topology in the local trees is not affected,
+and the relationship between nodes in adjacent trees is maintained.
+
+\begin{figure}
+\centering
+\vspace{5em}
+\includegraphics[width=\linewidth]{illustrations/simplification}
+\caption{\label{fig-simplification}
+ARG simplification. In all cases the graph visualization is shown on the left
+and the equivalent marginal trees on the right.
+(A) A gARG simulated from a diploid Wright-Fisher
+model. Note that the genome \textsf{k} has three children, and the genomes
+\textsf{e} and \textsf{g} simultaneously have multiple parents and multiple children;
+neither could be directly represented in a Griffiths-like eARG encoding
+(see the ``\nameref{eARG}'' section). The penultimate local tree is highlighted to aid discussion.
+(B) Simplified to remove all
+% trying out this terminology - can define in the text
+1-connected graph components (e.g., diamonds such as \textsf{jlnm}).
+(C) Remove nodes that are unary in all local trees. This removes both recombinant nodes
+that have only one child (e.g. \textsf{n}) and nodes that have more
+than one child (i.e. are ``common ancestors'') but never represent coalescences
+in the local trees (e.g. \textsf{r}).
+(D) Rewrite edges to bypass any unary nodes in local trees. Note that this loses lineage
+information such as the fact that \textsf{k} is connected to \textsf{a} via \textsf{e}.
+}
+\end{figure}
+
+% Q: would it help to talk about ``tree vertices'' here or would it
+% just be more confusing? I worry that people would think that
+% they are two different things then, and it's really important
+% that they get that a node-is-a-node
+
+The ideas of ``meaningful'' local tree topologies revolve around the
+presence of ``unary'' nodes: a node in the tree that has exactly
+one child. Fig.~\ref{fig-simplification} show a series of successive steps of
+simplification, starting with a complete gARG simulated under a
+(backwards-time) Wright-Fisher
+model, shown both as a graph on the left, and the equivalent set of 6
+local trees on the right (Fig.~\ref{fig-simplification}A). Consider the path above node
+\textsf{a} in the penultimate tree, highlighted in yellow and covering genomic positions
+60 to 70. On this path, before \textsf{a} meets its common ancestor with the other samples at
+node \textsf{q}, it passes through  \textsf{e},  \textsf{j},  \textsf{m},  \textsf{n} and
+\textsf{p}, which in this region of the genome are all unary nodes. This reflects the
+passage of this span of genome through the ARG towards the root (see the
+``\nameref{Ancestry_resolution}'' section).
+Such unary nodes in a local tree are not usually considered important;
+% because they provide little information and are not
+% detectable by information from any given individual tree.
+nodes usually mark the points where samples find a
+ \emph{common} ancestor, such as node \textsf{q} in the highlighted
+tree. For many purposes, we would therefore consider a path directly
+joining sample \textsf{a} to node \textsf{q} to be of equivalent status to the path
+containing the intermediate unary nodes. Simplification is the process of removing
+these less informative and (for most computational purposes) inefficient nodes
+from the local trees and the corresponding annotated graph.
+
+The first type of simplification that we can perform is to remove
+graph topology that is invisible to the samples. The best
+known example of such topology is a so-called
+``diamond''~\citep{rasmussen2014genome}
+in which the two parent nodes of a recombination immediately
+join again into a common ancestor (e.g.~\textsf{j}, \textsf{l}, \textsf{m}
+and  \textsf{n} in Fig.~\ref{fig-simplification}A).
+Unless we are specifically
+interested in the recombination event or ancestral genomes,
+there is no information in this topology and the diamond can be
+replaced by a single edge. More generally, any
+subgraph which is singly-connected in both the leafward and
+rootward direction (a ``super-diamond'') is non-identifiable and can be
+replaced by one edge. This definition includes the case
+of a node that has one inbound and one outbound edge, such as
+nodes \textsf{f} and \textsf{h}.
+Fig.~\ref{fig-simplification}B shows the result of this type of
+graph topology simplification.
+
+Simplifying away diamonds will remove many unary nodes from the
+local trees, but there can still be nodes that are unary in all
+of the local trees. In particular, a node can represent a recombinant
+with multiple parents in the graph but only a single child (e.g.\ node \textsf{n}
+in Fig.~\ref{fig-simplification}B), or can represent a common ancestor with
+multiple children in the graph but in which no local coalescence takes place
+(node \textsf{r} in Fig.~\ref{fig-simplification}B).
+% point out that this is why Hudson refers to CA nodes, not coalescence nodes
+Such nodes are not singly connected in the graph, but are nevertheless unary in
+all of the local trees. The operation to remove them,
+the results of which are shown in Fig.~\ref{fig-simplification}C,
+therefore requires knowledge not just of the graph topology, but also of the
+edge annotations. Removal of recombinant nodes can produce graph nodes with
+more than 2 parents (e.g.~node \textsf{e}); likewise, removal of
+common ancestor but non-coalescent nodes can produce graph nodes with
+more than 2 children (e.g.~node \textsf{s}). These represent a ``stacking up'' of
+multiple---often indistinguishable---events into a single node,
+something which is impossible to encode in an eARG.
+
+% JK: Text moved from an earlier section. See if we want to reuse this here
+% if not already said.
+% Indeed a single node can have both multiple children
+% and multiple parents. This would be the case in Fig~\ref{fig-arg-in-pedigree} if,
+% for instance, node \textsf{a} were to have two children.
+% % It is a shame that this is not actually shown in fig 3, but I don't think
+% % we can do this unless we have 5 generations
+% Moreover, a single node can have more than 2 children, and (as we shall see in
+% the ``\nameref{ARG_simplification}'' section), more than 2 parents. Another way to
+% state this is that transitions between nodes in a gARG can encapsulate multiple
+% events in an eARG.
+
+The remaining nodes are MRCAs of some subset of the samples
+at \emph{some} positions along the genome. We still have
+some unary nodes in the local trees, but these nodes will
+correspond to a coalescence in at least one other
+local tree. For example, node  \textsf{k} is unary in the second tree
+of Fig.~\ref{fig-simplification}C, but is either binary
+or ternary in all subsequent trees (recall this is a Wright-Fisher
+simulation). The final level of simplification is to alter the edge annotations
+such that, although no nodes are removed from the graph, \emph{all}
+unary nodes disappear from the local trees (Fig.~\ref{fig-simplification}D).
+It is common for this operation to remove nodes above the top MRCA node
+in the tree, resulting in many more root nodes (here, \textsf{q} and \textsf{k}
+become roots). [And continue]
+% Note that although this last stage produces much simpler local trees, by
+% removing information about the exact paths taken by lineages through
+% the graph, it loses potentially important information about shared edges
+% between trees. It is not clear whether (c) or (d) leads to a more compact
+% representation of the ARG in terms of number of edges.
+
+\section*{Relationship between ARGs and trees}
+
+We have seen that the local tree for a given position on
+the genome can be extracted from an ARG by starting at the
+leaves and following the appropriate paths through the graph.
+However, this tells us little about the relationship between
+adjacent trees, and what the properties of overall ensemble of
+local trees are.
+A recombination between two adjacent trees causes them to
+differ by a ``subtree prune-and-regraft'' (SPR)
+% Hein et al call it the ``subtree transfer'' operation but it's
+% the same thing.
+operation~\citep{hein1990reconstructing,song2003on,song2006properties}.
+The number of SPRs required to transform one leaf-labelled tree
+into another is known as the SPR distance
+% Hein 1996 say that SPR distance is NP-hard, but Allen 2001
+% say this is wrong. But then Bordewich went ahead and said
+% yes, actually it is NP hard.
+and is NP-hard to
+compute~\citep{hein1996complexity,allen2001subtree,bordewich2005computational}.
+
+The effects of recombination as we move from left-to-right along
+the genome in an unsimplified gARG are easy to interpret.
+Consider the highlighted tree in Fig~\ref{fig-simplification}A.
+In it the parent of \textsf{e} is \textsf{j}, whereas in the next tree
+its parent is \textsf{k}. Since both parents are present as
+unary nodes in the tree, we simply change one edge in
+order to effect the transition.
+% Need revising to fit new example
+A more complex transition occurs when we have a regraft operation
+that occurs above the current root (i.e., the current oldest
+non-unary node). For example, the transition between the
+second and third trees corresponds to node D switching
+from the path through I to H, and we then must add branches
+corresponding to passing through L and meeting at M.
+
+The equivalence between an ARG and a set of local trees
+separated by SPRs is often mentioned [e.g. x y z], but
+it is important to note that while extracting local
+trees from an ARG is straightforward, the converse
+process of reconstructing precisely the same ARG from
+these trees is more problematic. If the internal
+nodes are consistently labelled and unary nodes are
+included in the local trees (e.g., Fig~\ref{fig-simplification})
+then it is easy to exactly reconstruct the ARG.
+Otherwise, the problem is much more difficult and it
+may not be possible to uniquely reconstruct the original ARG.
+Without labels for the internal nodes, we are forced to
+to reason about which subtrees remain the same and which
+differ due to the SPR separating adjacent trees in order to
+assign node identity. Without the internal unary nodes marking
+recombinations, we must infer their existence. In
+general, there will not be a unique solution to this problem,
+and so there can be many possible ARGs compatible with
+a given list of leaf-labelled binary trees.
+This is the approach taken by the Espalier inference
+method~\citep{rasmussen2022espalier}, which finds a
+set of SPRs that reconcile independently inferred
+trees from different regions of the genome.
+
 \section*{Computational efficiency}

### 2022-08-26 paper.tex
Tweak text on Ne >> n violations

@@
 
-% JK: This paragraph is weak, and needs some work. Basically want to
-% illustrate the mathematical literature which views ancestral
-% graphs very much as *processes*.
-% Jere: ASG section redrafted, gene transfer left as TODO (because I
-% know essentially nothing about it).
 In parallel with this tendency to decouple the term
@@
 an extensive mathematical literature has developed in
-studying other related stochastic processes encoded via a graph.
+studying other related stochastic processes described as a graph.
 In particular, the Ancestral Selection Graph
@@
 koskela2019robust}.
+% Could mention the Ancestral Gene Transfer graph here too, within
+% something like "Similarly the Ancestral Gene Transfer Graph [cite]
+% and [other ancestral graphs, if they exist?] model genealogical
+% processes in bacteria as a graph.
 
@@
 in later sections we refer to it as an ``event ARG``, or eARG for brevity.
-An example of this form of an eARG
+An example of an eARG
 is given in Fig.~\ref{fig-arg-data-structure}B.
@@
 detail---which distinguishes an arbitrary graph with the required topological
-properties from an ARG---is that we associate a breakpoint with each recombination
+properties from an eARG---is that we associate a breakpoint with each recombination
 event. This information is sufficient to uniquely
@@
 $\sigma(u) = \varnothing$ otherwise.
-(This formulation is equivalent to having different ``types''
-of node, with breakpoints are associated with recombination
+(This is equivalent to having different ``types''
+of node, where breakpoints are associated with recombination
 nodes only).
@@
 child node $c$ and parent $p$, where $c, p \in \mathbb{N}$.
-For each recombination node $u$ we must
-have exactly two edges in $e$ such that $u$ is the child.
+% For each recombination node $u$ we must
+% have exactly two edges in $e$ such that $u$ is the child.
 This encoding of an eARG is equivalent to the output
-produced by programs such as
+of programs such as
 \texttt{ARGweaver}~\citep{rasmussen2014genome}
-and \texttt{KwARG}~\citep{ignatieva2021kwarg}:
-an example is given in Fig.~\ref{fig-arg-data-structure}A.
+and \texttt{KwARG}~\citep{ignatieva2021kwarg},
+and an example is given in Fig.~\ref{fig-arg-data-structure}A.
 Note that the description above captures only the
@@
 This is because there is a fundamental ambiguity involved
-in associating recombination breakpoints with nodes
-rather than edges: we need to know from which parent the
-and the only way we can break this symmetry is to assign
+in associating recombination breakpoints with nodes:
+we need to know from which parent the node inherited
+the segments of genome to the left and right of the breakpoint.
+The simplest way to break this symmetry is to assign
 some meaning to the order in which the parents are listed
@@
 data \citep[e.g.][]{Cesarani2022}.
-Such datasets are also available for other species,
-for example, \citet{RosFreixedes2022} have combined genome
-and SNP array data to infer recombination of genomes for 440,610 pigs within
-7 multi-generation pedigrees~\citep{whalen2018,Johnsson2021,Ros-Freixedes2020}.
+In another example, genome and SNP array data for 440,610 individuals within
+7 multi-generation pedigrees~\citep{whalen2018,Johnsson2021,Ros-Freixedes2020}
+were combined to infer recombination in pigs~\citep{RosFreixedes2022}.
 Effective population size in modern pig breeds is also less than 100 due to

### 2022-08-26 paper.tex
Move more text around and trim down

@@
 
-\section*{Node vs edge annotations}
-Another aspect of the classical description of an ARG data structure that
-derives from its historical origins as a stochastic
-process is how we encode information about genetic
-inheritance in the graph. The approach taken by Griffiths to
-doing this is concise, elegant: all we need to do is store
-the location of the breakpoint at each recombination event.
-Storing a breakpoint for each recombinatin event is a very
-natural and straightforward thing to do when simulating
-a (``Big''; see Appendix XXX) ARG process.
-Given the graph topology and these breakpoints, we can then
-recover the
-local trees at every point in the genome
-by tracing rootwards from the leaves and choosing the
-appropiate outbound path at recombination nodes.
-However, there are several significant disadvantages
-to this approach: [TODO be concrete here later
-and give pointers to the section where we discuss the
-alternative.]
-
-Fortunately, there is a straightforward change to the
-classical eARG encoding that simplifies
-implementations and greatly increases the flexibility
-of both the types and granularity of inheritance information
-that we can represent.
-The main change that we need to make is to
-associate the
-the details of inheritance with \emph{edges} rather than
-as breakpoints associated with recombination events and
-edge ordering rules.
-This is illustrated in Fig~\ref{fig-arg-data-structure}C
-where this alternative edge annotation formulation is
-shown. By associating the interval $(0, x)$ with edge $(d,e)$
-and the interval $(x, L)$ with $(d, f)$ we make the encoding
-unambiguous: we can list the edges in any order, and it
-will describe the same set of local genealogical trees.
-This approach is also more flexible, because any pattern
-of inheritance can be described by these intervals. For
-example, if a gene conversion event occurred where a segment
-between positions $x$ and $y$ was inherited by
-node $d$ from $e$, we would state this by associating
-the intervals $\{(0, x], (y, L]\}$ with $(d,e)$ and
-$\{(x, y]\}$ with $(d,f)$.
-This change is straightforward, involves no loss of information
-and (as we see in later sections)
-greatly increases both the expressivity and computational
-power of ancestral recombination graph data structures.
+% \section*{Node vs edge annotations}
+% Another aspect of the classical description of an ARG data structure that
+% derives from its historical origins as a stochastic
+% process is how we encode information about genetic
+% inheritance in the graph. The approach taken by Griffiths to
+% doing this is elegant: all we need to do is store
+% the location of the breakpoint at each recombination event.
+% % Storing a breakpoint for each recombinatin event is a very
+% % natural and straightforward thing to do when simulating
+% % an  ARG process.
+% Given the graph topology and these breakpoints, we can then
+% recover the
+% local trees at every point in the genome
+% by tracing rootwards from the leaves and choosing the
+% appropriate outbound path at recombination nodes.
+% However, there are several significant disadvantages
+% to this approach: [TODO be concrete here later
+% and give pointers to the section where we discuss the
+% alternative.]
+
+% Fortunately, there is a straightforward change to the
+% classical eARG encoding that simplifies
+% implementations and greatly increases the flexibility
+% of both the types and granularity of inheritance information
+% that we can represent.
+% The main change that we need to make is to associate
+% the details of inheritance with \emph{edges} rather than
+% as breakpoints associated with recombination events and
+% edge ordering rules.
+% This is illustrated in Fig~\ref{fig-arg-data-structure}C
+% where this alternative edge annotation formulation is
+% shown. By associating the interval $(0, x)$ with edge $(d,e)$
+% and the interval $(x, L)$ with $(d, f)$ we make the encoding
+% unambiguous: we can list the edges in any order, and it
+% will describe the same set of local genealogical trees.
+% This approach is also more flexible, because any pattern
+% of inheritance can be described by these intervals. For
+% example, if a gene conversion event occurred where a segment
+% between positions $x$ and $y$ was inherited by
+% node $d$ from $e$, we would state this by associating
+% the intervals $\{(0, x], (y, L]\}$ with $(d,e)$ and
+% $\{(x, y]\}$ with $(d,f)$.
+% This change is straightforward, involves no loss of information
+% and (as we see in later sections)
+% greatly increases both the expressivity and computational
+% power of ancestral recombination graph data structures.
 
 \section*{Genome ARGs}
-As we have seen in previous sections, the focus
+As we have seen in the previous section, the focus
 on a limited palette of events makes describing currently-available
 datasets and known patterns of genetic inheritance as an eARG
-quite cumbersome.
+cumbersome at best.
 There is no fundamental reason that the patterns of genetic inheritance
@@
 \emph{genomes} and intervals of genetic inheritance
-along the edges (see the previous section).
+along the edges.
 To distinguish from the classical Griffiths eARG,
@@
 genetic inheritance between an ancestor and a descendant.
-Each edge is annotated (see the previous section) with the coordinates
+Each edge is annotated with the coordinates
 of the genomic intervals over which genetic inheritance occurs
@@
 a representational perspective.
-
 Secondly, because we do not classify nodes by event type
@@
 In large deeply sequenced pedigrees such complexities are commonplace.
-
 Finally, there is a subtle and important point about how recombination
@@
 gARG encoding that we discuss are due to the encoding of inheritance via
-edges rather than nodes (see section XXX).
+edges rather than nodes.
 Whether one regards nodes as genomes or events is orthogonal,
@@
 parent to child (and ultimately from cell to cell) through some
-number of generations. The edges are ``annotated'' (see section XXX)
-with the coordinates over which inheritance occurs. There are
+number of generations. The edges are ``annotated''
+with the genome coordinates over which inheritance occurs. There are
 two ways in which we can interpret these annotations, and the
 distinction is subtle and important.
-The first interpretation is to define the annotations as
-the genetic intervals inherited by the descendant node from the
+The first interpretation is to regard the annotations as describing
+the complete set of genomic intervals inherited by the descendant node from the
 ancestor node. The second is to view the edge annotations as representing
@@
 the so-called ``big'' ARG \emph{process} (see Appendix XXX)
-and has several features that we are unlikely to infer in practice.
+and has several features that we are unlikely to simulate or infer in practice.
 For example, by definition the samples contain no genetic information about
-node \textsf{j} and so it is unlike to be inferred.
+node \textsf{j} and so it is unlikey to be inferred.
 Likewise, is is unlikely we would have information to infer the ``grand MRCA''
@@
 
+\begin{figure}
+\centering
+\vspace{5em}
+\includegraphics[width=\linewidth]{illustrations/simplification}
+\caption{\label{fig-simplification}
+ARG simplification. In all cases the graph visualization is shown on the left
+and the equivalent marginal trees on the right.
+(A) A gARG simulated from a diploid Wright-Fisher
+model. Note that the genome \textsf{k} has three children, and the genomes
+\textsf{e} and \textsf{g} simultaneously have multiple parents and multiple
+children.
+The penultimate local tree is highlighted to aid discussion.
+(B) Simplified to remove all
+% trying out this terminology - can define in the text
+1-connected graph components (e.g., diamonds such as \textsf{jlnm}).
+(C) Remove nodes that are unary in all local trees. This removes both recombinant nodes
+that have only one child (e.g. \textsf{n}) and nodes that have more
+than one child (i.e. are ``common ancestors'') but never represent coalescences
+in the local trees (e.g. \textsf{r}).
+(D) Rewrite edges to bypass any unary nodes in local trees. Note that this loses lineage
+information such as the fact that \textsf{k} is connected to \textsf{a} via \textsf{e}.
+}
+\end{figure}
+
+\section*{Simplification}\label{ARG_simplification}
+ARG simplification~\citep{kelleher2018efficient} is the process
+of removing nodes and re-writing edges in a gARG to remove
+various types of redundancy, while maintaining the crucial properties
+that the ``meaningful'' topology in the local trees is not affected,
+and the relationship between nodes in adjacent trees is maintained.
+The idea of meaningful local tree topologies is based on the
+presence of ``unary'' nodes: a node in the tree that has exactly
+one child. Fig.~\ref{fig-simplification} show a series of successive steps of
+simplification, starting with a complete gARG simulated under a
+(backwards-time) Wright-Fisher
+model, shown both as a graph on the left, and the equivalent set of 6
+local trees on the right (Fig.~\ref{fig-simplification}A). Consider the path above node
+\textsf{a} in the penultimate tree, highlighted in yellow and covering genomic positions
+60 to 70. On this path, before \textsf{a} meets its common ancestor with the other samples at
+node \textsf{q}, it passes through  \textsf{e},  \textsf{j},  \textsf{m},  \textsf{n} and
+\textsf{p}, which in this region of the genome are all unary nodes. This reflects the
+passage of this span of genome through the ARG towards the root.
+Such unary nodes in a local tree are not usually considered important;
+% because they provide little information and are not
+% detectable by information from any given individual tree.
+nodes usually mark the points where samples find a
+ \emph{common} ancestor, such as node \textsf{q} in the highlighted
+tree. For many purposes, we would therefore consider a path directly
+joining sample \textsf{a} to node \textsf{q} to be equivalent to the path
+containing the intermediate unary nodes. Simplification is the process of removing
+these less informative nodes from the local trees by modifying the gARG.
+
+The first type of simplification that we can perform is to remove
+graph topology that is invisible to the samples. The best
+known example of such topology is a so-called
+``diamond''~\citep{rasmussen2014genome}
+in which the two parent nodes of a recombination immediately
+join again into a common ancestor (e.g.~\textsf{j}, \textsf{l}, \textsf{m}
+and  \textsf{n} in Fig.~\ref{fig-simplification}A).
+Unless we are specifically
+interested in the recombination event or ancestral genomes,
+there is no information in this topology and the diamond can be
+replaced by a single edge. More generally, any
+subgraph that is singly-connected in both the leafward and
+rootward direction (a ``super-diamond'') is non-identifiable and can be
+replaced by one edge. This definition includes the case
+of a node that has one inbound and one outbound edge, such as
+nodes \textsf{f} and \textsf{h}.
+Fig.~\ref{fig-simplification}B shows the result of this type of
+graph topology simplification.
+
+Simplifying away diamonds will remove many unary nodes from the
+local trees, but there can still be nodes that are unary in all
+of the local trees. In particular, a node can represent a recombinant
+with multiple parents in the graph but only a single child (e.g.\ node \textsf{n}
+in Fig.~\ref{fig-simplification}B), or can represent a common ancestor with
+multiple children in the graph but in which no local coalescence takes place
+(node \textsf{r} in Fig.~\ref{fig-simplification}B).
+% point out that this is why Hudson refers to CA nodes, not coalescence nodes
+Such nodes are not singly connected in the graph, but are nevertheless unary in
+all of the local trees.
+The operation to remove them
+therefore requires knowledge not just of the graph topology but also of the
+edge annotations.
+As we see in Fig.~\ref{fig-simplification}C,
+removal of recombinant nodes can produce graph nodes with
+more than two parents (e.g.~node \textsf{e}); and likewise, removal of
+common ancestor but non-coalescent nodes can produce graph nodes with
+more than two children (e.g.~node \textsf{s}). These represent a ``stacking up'' of
+multiple---often indistinguishable---events into a single node,
+something which is impossible to encode in an eARG.
+
+% JK: Text moved from an earlier section. See if we want to reuse this here
+% if not already said.
+% Indeed a single node can have both multiple children
+% and multiple parents. This would be the case in Fig~\ref{fig-arg-in-pedigree} if,
+% for instance, node \textsf{a} were to have two children.
+% % It is a shame that this is not actually shown in fig 3, but I don't think
+% % we can do this unless we have 5 generations
+% Moreover, a single node can have more than 2 children, and (as we shall see in
+% the ``\nameref{ARG_simplification}'' section), more than 2 parents. Another way to
+% state this is that transitions between nodes in a gARG can encapsulate multiple
+% events in an eARG.
+
+The remaining nodes are MRCAs of some subset of the samples
+at \emph{some} positions along the genome. We still have
+some unary nodes in the local trees, but these nodes will
+correspond to a coalescence in at least one other
+local tree. For example, node  \textsf{k} is unary in the second tree
+of Fig.~\ref{fig-simplification}C, but is either binary
+or ternary in all subsequent trees (recall this is a Wright-Fisher
+simulation). The final level of simplification is to alter the edge annotations
+such that, although no nodes are removed from the graph, \emph{all}
+unary nodes disappear from the local trees (Fig.~\ref{fig-simplification}D).
+Note that although this last stage produces simpler local trees, by
+removing information about the exact paths taken by lineages through
+the graph, it removes potentially useful information about shared edges
+between trees. The extent to which this information can be inferred
+and utilised is an interesting open question.
+
 \section*{Precision of ancestry information}
+% Not all nodes in an ARG are equal. Assuming that we have sequenced the
+% genomes of the samples, there are intrinsic limits on what can
+% be known about more ancient nodes, determined by the graph topology
+% and the flow of ancestral material through it.
+
 Ancestral recombination graphs are often characterised
@@
 % that
-(as claimed by e.g.~\citep{hejase2020summary}),
+(as claimed by e.g.~\citet{hejase2020summary}),
 and information about recombination and the correlation structure
@@
 but maybe we should keep this for the conclusion?]
-
-\section*{ARG simplification}\label{ARG_simplification}
-
-Not all nodes in an ARG are equal. Assuming that we have sequenced the
-genomes of the samples, there are intrinsic limits on what can
-be known about more ancient nodes, determined by the graph topology
-and the flow of ancestral material through it.
-\emph{Simplification}~\citep{kelleher2018efficient} is the process
-of removing nodes and re-writing edges in a gARG to remove
-various types of redundancy, while maintaining the crucial properties
-that the ``meaningful'' topology in the local trees is not affected,
-and the relationship between nodes in adjacent trees is maintained.
-
-\begin{figure}
-\centering
-\vspace{5em}
-\includegraphics[width=\linewidth]{illustrations/simplification}
-\caption{\label{fig-simplification}
-ARG simplification. In all cases the graph visualization is shown on the left
-and the equivalent marginal trees on the right.
-(A) A gARG simulated from a diploid Wright-Fisher
-model. Note that the genome \textsf{k} has three children, and the genomes
-\textsf{e} and \textsf{g} simultaneously have multiple parents and multiple children;
-neither could be directly represented in a Griffiths-like eARG encoding
-(see the ``\nameref{eARG}'' section). The penultimate local tree is highlighted to aid discussion.
-(B) Simplified to remove all
-% trying out this terminology - can define in the text
-1-connected graph components (e.g., diamonds such as \textsf{jlnm}).
-(C) Remove nodes that are unary in all local trees. This removes both recombinant nodes
-that have only one child (e.g. \textsf{n}) and nodes that have more
-than one child (i.e. are ``common ancestors'') but never represent coalescences
-in the local trees (e.g. \textsf{r}).
-(D) Rewrite edges to bypass any unary nodes in local trees. Note that this loses lineage
-information such as the fact that \textsf{k} is connected to \textsf{a} via \textsf{e}.
-}
-\end{figure}
-
-% Q: would it help to talk about ``tree vertices'' here or would it
-% just be more confusing? I worry that people would think that
-% they are two different things then, and it's really important
-% that they get that a node-is-a-node
-
-The ideas of ``meaningful'' local tree topologies revolve around the
-presence of ``unary'' nodes: a node in the tree that has exactly
-one child. Fig.~\ref{fig-simplification} show a series of successive steps of
-simplification, starting with a complete gARG simulated under a
-(backwards-time) Wright-Fisher
-model, shown both as a graph on the left, and the equivalent set of 6
-local trees on the right (Fig.~\ref{fig-simplification}A). Consider the path above node
-\textsf{a} in the penultimate tree, highlighted in yellow and covering genomic positions
-60 to 70. On this path, before \textsf{a} meets its common ancestor with the other samples at
-node \textsf{q}, it passes through  \textsf{e},  \textsf{j},  \textsf{m},  \textsf{n} and
-\textsf{p}, which in this region of the genome are all unary nodes. This reflects the
-passage of this span of genome through the ARG towards the root (see the
-``\nameref{Ancestry_resolution}'' section).
-Such unary nodes in a local tree are not usually considered important;
-% because they provide little information and are not
-% detectable by information from any given individual tree.
-nodes usually mark the points where samples find a
- \emph{common} ancestor, such as node \textsf{q} in the highlighted
-tree. For many purposes, we would therefore consider a path directly
-joining sample \textsf{a} to node \textsf{q} to be of equivalent status to the path
-containing the intermediate unary nodes. Simplification is the process of removing
-these less informative and (for most computational purposes) inefficient nodes
-from the local trees and the corresponding annotated graph.
-
-The first type of simplification that we can perform is to remove
-graph topology that is invisible to the samples. The best
-known example of such topology is a so-called
-``diamond''~\citep{rasmussen2014genome}
-in which the two parent nodes of a recombination immediately
-join again into a common ancestor (e.g.~\textsf{j}, \textsf{l}, \textsf{m}
-and  \textsf{n} in Fig.~\ref{fig-simplification}A).
-Unless we are specifically
-interested in the recombination event or ancestral genomes,
-there is no information in this topology and the diamond can be
-replaced by a single edge. More generally, any
-subgraph which is singly-connected in both the leafward and
-rootward direction (a ``super-diamond'') is non-identifiable and can be
-replaced by one edge. This definition includes the case
-of a node that has one inbound and one outbound edge, such as
-nodes \textsf{f} and \textsf{h}.
-Fig.~\ref{fig-simplification}B shows the result of this type of
-graph topology simplification.
-
-Simplifying away diamonds will remove many unary nodes from the
-local trees, but there can still be nodes that are unary in all
-of the local trees. In particular, a node can represent a recombinant
-with multiple parents in the graph but only a single child (e.g.\ node \textsf{n}
-in Fig.~\ref{fig-simplification}B), or can represent a common ancestor with
-multiple children in the graph but in which no local coalescence takes place
-(node \textsf{r} in Fig.~\ref{fig-simplification}B).
-% point out that this is why Hudson refers to CA nodes, not coalescence nodes
-Such nodes are not singly connected in the graph, but are nevertheless unary in
-all of the local trees. The operation to remove them,
-the results of which are shown in Fig.~\ref{fig-simplification}C,
-therefore requires knowledge not just of the graph topology, but also of the
-edge annotations. Removal of recombinant nodes can produce graph nodes with
-more than 2 parents (e.g.~node \textsf{e}); likewise, removal of
-common ancestor but non-coalescent nodes can produce graph nodes with
-more than 2 children (e.g.~node \textsf{s}). These represent a ``stacking up'' of
-multiple---often indistinguishable---events into a single node,
-something which is impossible to encode in an eARG.
-
-% JK: Text moved from an earlier section. See if we want to reuse this here
-% if not already said.
-% Indeed a single node can have both multiple children
-% and multiple parents. This would be the case in Fig~\ref{fig-arg-in-pedigree} if,
-% for instance, node \textsf{a} were to have two children.
-% % It is a shame that this is not actually shown in fig 3, but I don't think
-% % we can do this unless we have 5 generations
-% Moreover, a single node can have more than 2 children, and (as we shall see in
-% the ``\nameref{ARG_simplification}'' section), more than 2 parents. Another way to
-% state this is that transitions between nodes in a gARG can encapsulate multiple
-% events in an eARG.
-
-The remaining nodes are MRCAs of some subset of the samples
-at \emph{some} positions along the genome. We still have
-some unary nodes in the local trees, but these nodes will
-correspond to a coalescence in at least one other
-local tree. For example, node  \textsf{k} is unary in the second tree
-of Fig.~\ref{fig-simplification}C, but is either binary
-or ternary in all subsequent trees (recall this is a Wright-Fisher
-simulation). The final level of simplification is to alter the edge annotations
-such that, although no nodes are removed from the graph, \emph{all}
-unary nodes disappear from the local trees (Fig.~\ref{fig-simplification}D).
-It is common for this operation to remove nodes above the top MRCA node
-in the tree, resulting in many more root nodes (here, \textsf{q} and \textsf{k}
-become roots). [And continue]
-% Note that although this last stage produces much simpler local trees, by
-% removing information about the exact paths taken by lineages through
-% the graph, it loses potentially important information about shared edges
-% between trees. It is not clear whether (c) or (d) leads to a more compact
-% representation of the ARG in terms of number of edges.
 

### 2022-08-26 paper.tex
Move SPR section content elsewhere

@@
 
+% Needs another pass
+An important consequence of simplifying local trees to remove
+unary nodes is that we may lose the ability to identify
+the ``subtree prune-and-regraft'' (SPR) operations
+by which they differ~\citep{hein1990reconstructing,song2003on,song2006properties}.
+% The effects of recombination as we move from left-to-right along
+% the genome in the complete simulated gARG
+% of Fig~\ref{fig-simplification}A are easy to interpret,
+% but as more redundancy is removed from the graph
+% the
+As explained above, we can have multiple recombination and
+common ancestor events stacking up, and therefore
+many possible SPR operations. Deducing the SPRs between
+trees is NP-hard in
+general~\citep{hein1996complexity,allen2001subtree,bordewich2005computational},
+but effective heuristics have been proposed~\citep{rasmussen2022espalier}.
+
+% The equivalence between an ARG and a set of local trees
+% separated by SPRs is often mentioned [e.g. x y z], but
+% it is important to note that while extracting local
+% trees from an ARG is straightforward, the converse
+% process of reconstructing precisely the same ARG from
+% these trees is more problematic. If the internal
+% nodes are consistently labelled and unary nodes are
+% included in the local trees (e.g., Fig~\ref{fig-simplification})
+% then it is easy to exactly reconstruct the ARG.
+% Otherwise, the problem is much more difficult and it
+% may not be possible to uniquely reconstruct the original ARG.
+% Without labels for the internal nodes, we are forced to
+% to reason about which subtrees remain the same and which
+% differ due to the SPR separating adjacent trees in order to
+% assign node identity. Without the internal unary nodes marking
+% recombinations, we must infer their existence. In
+% general, there will not be a unique solution to this problem,
+% and so there can be many possible ARGs compatible with
+% a given list of leaf-labelled binary trees.
+% This is the approach taken by the Espalier inference
+% method~\citep{rasmussen2022espalier}, which finds a
+% set of SPRs that reconcile independently inferred
+% trees from different regions of the genome.
+
 \section*{Precision of ancestry information}
@@
 
-\section*{Relationship between ARGs and trees}
-
-We have seen that the local tree for a given position on
-the genome can be extracted from an ARG by starting at the
-leaves and following the appropriate paths through the graph.
-However, this tells us little about the relationship between
-adjacent trees, and what the properties of overall ensemble of
-local trees are.
-A recombination between two adjacent trees causes them to
-differ by a ``subtree prune-and-regraft'' (SPR)
-% Hein et al call it the ``subtree transfer'' operation but it's
-% the same thing.
-operation~\citep{hein1990reconstructing,song2003on,song2006properties}.
-The number of SPRs required to transform one leaf-labelled tree
-into another is known as the SPR distance
-% Hein 1996 say that SPR distance is NP-hard, but Allen 2001
-% say this is wrong. But then Bordewich went ahead and said
-% yes, actually it is NP hard.
-and is NP-hard to
-compute~\citep{hein1996complexity,allen2001subtree,bordewich2005computational}.
-
-The effects of recombination as we move from left-to-right along
-the genome in an unsimplified gARG are easy to interpret.
-Consider the highlighted tree in Fig~\ref{fig-simplification}A.
-In it the parent of \textsf{e} is \textsf{j}, whereas in the next tree
-its parent is \textsf{k}. Since both parents are present as
-unary nodes in the tree, we simply change one edge in
-order to effect the transition.
-% Need revising to fit new example
-A more complex transition occurs when we have a regraft operation
-that occurs above the current root (i.e., the current oldest
-non-unary node). For example, the transition between the
-second and third trees corresponds to node D switching
-from the path through I to H, and we then must add branches
-corresponding to passing through L and meeting at M.
-
-The equivalence between an ARG and a set of local trees
-separated by SPRs is often mentioned [e.g. x y z], but
-it is important to note that while extracting local
-trees from an ARG is straightforward, the converse
-process of reconstructing precisely the same ARG from
-these trees is more problematic. If the internal
-nodes are consistently labelled and unary nodes are
-included in the local trees (e.g., Fig~\ref{fig-simplification})
-then it is easy to exactly reconstruct the ARG.
-Otherwise, the problem is much more difficult and it
-may not be possible to uniquely reconstruct the original ARG.
-Without labels for the internal nodes, we are forced to
-to reason about which subtrees remain the same and which
-differ due to the SPR separating adjacent trees in order to
-assign node identity. Without the internal unary nodes marking
-recombinations, we must infer their existence. In
-general, there will not be a unique solution to this problem,
-and so there can be many possible ARGs compatible with
-a given list of leaf-labelled binary trees.
-This is the approach taken by the Espalier inference
-method~\citep{rasmussen2022espalier}, which finds a
-set of SPRs that reconcile independently inferred
-trees from different regions of the genome.
+% The equivalence between an ARG and a set of local trees
+% separated by SPRs is often mentioned [e.g. x y z], but
+% it is important to note that while extracting local
+% trees from an ARG is straightforward, the converse
+% process of reconstructing precisely the same ARG from
+% these trees is more problematic. If the internal
+% nodes are consistently labelled and unary nodes are
+% included in the local trees (e.g., Fig~\ref{fig-simplification})
+% then it is easy to exactly reconstruct the ARG.
+% Otherwise, the problem is much more difficult and it
+% may not be possible to uniquely reconstruct the original ARG.
+% Without labels for the internal nodes, we are forced to
+% to reason about which subtrees remain the same and which
+% differ due to the SPR separating adjacent trees in order to
+% assign node identity. Without the internal unary nodes marking
+% recombinations, we must infer their existence. In
+% general, there will not be a unique solution to this problem,
+% and so there can be many possible ARGs compatible with
+% a given list of leaf-labelled binary trees.
+% This is the approach taken by the Espalier inference
+% method~\citep{rasmussen2022espalier}, which finds a
+% set of SPRs that reconcile independently inferred
+% trees from different regions of the genome.
 

### 2022-08-30 paper.tex
More text moving
Tweaks to simplification section

@@
 However, the genetic genealogy of a recombining organism
-cannot be described as a single evolutionary tree: instead,
-the inheritance of each local section of the genome may be described
-by a (slightly) different tree~\citet{hudson1983properties}. Inferring complex
+cannot be captured by a single tree: instead,
+the inheritance of each section of the genome may be described
+by a (slightly) different local tree~\citep{hudson1983properties}. Inferring complex
 genealogies of this nature
-presents profound technical difficulties, meaning tree-based approaches are
-rarely used in population and statistical genetics. Therefore until recently, analysis of
-population genetic data has primarily focussed on summaries derived
+presents profound technical difficulties, meaning genealogy-based approaches have
+not been widely used in population and statistical genetics. Until recently, analysis of
+population genetic data has therefore primarily focussed on summaries derived
 from the local trees~\citep{tajima1983evolutionary,tavare1984line}
-rather than using the underlying genealogy itself. Such
+rather than using the full underlying genealogy itself. Such
 derived measures include allele frequencies and site frequency
@@
 and data representation~\citep{kelleher2016efficient}
-have raised the realistic prospect of genealogical analysis becoming a standard part
+have raised the realistic prospect of genealogy-based analysis becoming a standard part
 of the population and statistical genetics toolkit~\citep{hejase2020summary}.
-Applications using these inferred ancestries as input have begun to
+Applications using these inferred ancestries as inputs have begun to
 appear~\citep{osmond2021estimating,zhang2021biobank,fan2022genealogical,hejase2022deep}
@@
 progress: we currently lack a well-defined data-model and shared terminology
-to discuss recombining genealogies. This leads to basic errors in statements
+to discuss genealogies in the presence of recombination. This leads to basic errors in statements
 about what different inference methods produce as output, as well as presenting serious
@@
 It is a stochastic process, where each random realisation
-is a genealogical tree describing the genetic ancestry of the samples.
-Numerous elaborations have been incorporated into the
-basic process~\citep{hudson1990gene,hein2004gene,wakely2008coalescent};
-in particular, early work by \citet{hudson1983properties}
-analysed the effects of recombination on the random genealogical
-trees generated by the coalescent. This
-generated several fundamental analytical results
+is a genealogical tree describing the genetic ancestry of the sample.
+Numerous extensions to the model have been proposed~\citep{hudson1990gene,hein2004gene,wakely2008coalescent}, incorporating various types of evolutionary processes. 
+In particular, \citet{hudson1983properties}
+analysed the effects of recombination on coalescent trees, generating several fundamental analytical results,
 as well as describing the elegant algorithm
-at the heart of simulation
-programs~\citep{hudson2002generating,kelleher2016efficient,baumdicker2021efficient}
-used in thousands of studies.
+still at the heart of widely-used simulation
+programs~\citep{hudson2002generating,kelleher2016efficient,baumdicker2021efficient}.
 In its original formulation, Hudson's algorithm produced a
 sequence of genealogical trees along the genome.
-Critically, although the correlation structure among these
+Crucially, although the correlation structure among these
 local trees was analysed analytically, those correlations
 were not captured explicitly: the ancestry
-was viewed as a series of disconnected local trees rather than
+was viewed as a set of disconnected local trees rather than
 as a single cohesive object.
@@
 coalescent with recombination from a different perspective,
-in which the stochastic process itself is encoded as a graph,
-and showed how the local trees generated by Hudson's algorithm
-were embedded within
-it~\citep{griffiths1991two,ethier1990two,griffiths1996ancestral,griffiths1997ancestral}.
-They referred to this graph,
-representing both the stochastic process and
-its random realisations, as the Ancestral Recombination Graph (ARG).
+formulating it as a stochastic process where each realisation 
+is a genealogy encoded as a graph,
+and showed that the local trees generated by Hudson's algorithm
+are embedded within this graph~\citep{griffiths1991two,ethier1990two,griffiths1996ancestral,griffiths1997ancestral}.
+They referred to both the stochastic process and
+its random realisations as the Ancestral Recombination Graph (ARG).
 Although mathematically equivalent, it is
@@
 (see Appendix XXX for details). However, ARGs provided a way
-to describe recombinant ancestry as single object that
-could be reasoned about and inferred in a way
-that was missing from Hudson's treatments.
-
-Early work on ARG inference focused on the problem of
-inferring parameters of the
-stochastic process, where the ancestry was regarded as a
-latent parameter to be averaged out
-\citep[e.g.][]{griffiths1996ancestral,kuhner2000maximum, nielsen2000estimation,
-fearnhead2001estimating}.
-These methods met with limited success
-because the state space of ARGs is overwhelmingly large and
-lacks a simple geometry or neighbourhood structure for inference or
-sampling methods to  exploit.
-The focus subsequently shifted to the more
-tractable---but still NP-hard~\citep{wang2001perfect}---problem of computing
-the minimum number of recombinations required
-to explain an input dataset~\citep{myers2003bounds}, and inferring the corresponding
-ARG realisations~\citep{song2003parsimonious,song2005efficient,lyngso2005minimum}.
-It is important to note that such recombination-parsimony ARGs
-have quite different properties from realisations of the
-ARG \emph{process} as modelled by Hudson, Griffiths, and colleagues.
-Firstly, branch lengths are not typically estimated,
-% why do we need the following two lines
-and node times are required in order to compute a likelihood under the
-coalescent with recombination (see section XXX).
-Secondly, even if branch lengths were estimated,
-the realisations would have a very small likelihood since
-the coalescent with recombination is often highly \emph{un}parsimonious
-in terms of recombination events.
-The previously close correspondance
-between the ARG as a mathematical representation of the coalescent process
-and as concrete realisations of ancestry was therefore broken.
-\cite{minichiello2006mapping} explicitly distinguished the ARG as a
-stochastic process from the ARG as a data structure, and
-argued for using the term to mean specific realised ancestral histories.
-This has become a common
-usage~\citep[e.g.][]{gusfield2014recombinatorics,mathieson2020ancestry,brandt2021evaluation},
-although not universal, with authors also using the term
-in the original stochastic
-process sense~\citep[e.g.][]{nordborg2000linkage,birkner2013ancestral,
-wilton2015smc,griffiths2016coalescent}.
-
-% Gusfield~\citep{gusfield2014recombinatorics}
-% defines also defines an ARG as a concrete realisation
-% and studies instances that satisfy particular
-% properties such as ``galled trees''
-% \citep{wang2001perfect, gusfield2004optimal}
-% as well as describing algorithms for more general phylogenetic
-% networks~\citep{huson2010phylogenetic}.
-
-In parallel with this tendency to decouple the term
-Ancestral Recombination Graph from the stochastic process,
-an extensive mathematical literature has developed in
-studying other related stochastic processes described as a graph.
+to reason about and infer recombinant ancestry as a single object,
+which was not possible within Hudson's framework.
+
+Subsequent work on ARGs proceeded in broadly three main directions: focussing on (1) 
+exploring the mathematical properties of the coalescent with recombination and 
+related stochastic processes, (2) inferring evolutionary parameters under 
+(approximations to) this model, either with or without explicitly reconstructing the 
+genealogy of the sample, and (3) treating the ARG as a discrete graph, ignoring the 
+generating stochastic process, and studying its properties from a computational and 
+algorithmic perspective.
+
+% Mathematical treatment of the CwR and ARGs (and ARG-like stochastic processes)
+An extensive body of work has developed in
+studying the coalescent with recombination 
+% Can we add a couple of the most well known CwR examples here
+and other related 
+graph-valued stochastic processes from a mathematical perspective.
 In particular, the Ancestral Selection Graph
 (ASG)~\citep{krone1997ancestral,neuhauser1997genealogy}
-uses a similar approach to model natural selection.
+uses a similar approach to model natural selection instead of recombination.
 Unlike the ARG process, the ASG imposes a hard distinction between the stochastic process,
 which constructs a random ARG-like graph, and an observable realisation,
-which is a single tree sampled from the graph in a non-uniform way to encode desired
-patterns of natural selection.
+which is a single tree sampled from the graph in a non-uniform way to encode
+desired patterns of natural selection.
 Constructions of ASG-like stochastic processes encoding various
@@
 
-The point of this historical overview is to illustrate that the term
-Ancestral Recombination Graph has different meanings in different
-contexts, and in its original form it was closely coupled to
-the coalescent with recombination. In the following sections we
+% Inference of parameters and explicit ARG reconstruction
+Early work on inference under the coalescent with recombination
+focused on the problem of
+inferring the parameters of the
+stochastic process, where the ancestry was regarded as a
+latent parameter to be averaged out
+\citep[e.g.][]{griffiths1996ancestral,kuhner2000maximum, nielsen2000estimation,
+fearnhead2001estimating}.
+These methods met with limited success
+because the state space of ARGs is overwhelmingly large, and
+lacks a simple geometry or neighbourhood structure for inference or
+sampling methods to  exploit.
+Several breakthroughs in this direction were achieved through
+formulating simplified but more tractable approximations to the full 
+model~\citep{mcvean2005approximating,marjoram2006fast,li2011inference,
+paul2011accurate,schiffels2014inferring}.
+The related problem of \emph{sampling} genealogies compatible with a given 
+dataset under the coalescent with recombination also proved notoriously difficult
+computationally; progress in explicitly inferring genealogies at scale 
+has similarly been achieved through resorting to principled
+approximations~\citep{rasmussen2014genome,mahmoudi2022bayesian},
+or moving away from the coalescent with recombination altogether and seeking
+to infer a single plausible ARG~\citep[e.g.][]{speidel2019method} or ARG
+topology~\citep[e.g.][]{minichiello2006mapping,kelleher2019inferring}.
+
+% Optimisation problems to do with ARGs and other phylogenetic networks
+There has also been substantial interest in formulating and answering 
+fundamental questions about properties
+of the ARG as a discrete graph structure, abstracting away from the
+generating process and focussing on the ARG topology without considering 
+branch lengths. 
+The first prominent problem was calculating (lower bounds on) the minimum number of
+recombinations required to reconstruct a valid genealogy for a given
+sample~\citep{myers2003bounds}, and constructing the corresponding 
+minimal (parsimonious)
+ARGs~\citep{song2003parsimonious,song2005efficient,lyngso2005minimum}.
+These problems are NP-hard in general~\citep{wang2001perfect}, and progress has
+been achieved through studying various constrained special cases of ARGs~\citep[e.g.][]{gusfield2004optimal} and
+other more general types of phylogenetic networks~\citep{huson2010phylogenetic}. The
+focus has been on algorithmic and combinatorial results, that largely do not
+have direct relevance to the inference problems described above.
+
+%It is important to note that parsimonious ARGs
+%have quite different properties from realisations of the
+%ARG \emph{process} as modelled by Hudson, Griffiths, and colleagues.
+%Firstly, branch lengths are not typically estimated,
+%% why do we need the following two lines
+%and node times are required in order to compute a likelihood under the
+%coalescent with recombination (see section XXX).
+%Secondly, even if branch lengths were estimated,
+%the realisations would have a very small likelihood since
+%the coalescent with recombination is often highly \emph{un}parsimonious
+%in terms of recombination events.
+
+The goal of this historical overview is to illustrate that the meaning of the term ``ARG" now strongly
+depends on the context in which it is used, and can mean both the 
+stochastic process generating genealogies in the presence of 
+recombination~\citep[e.g.][]{nordborg2000linkage,birkner2013ancestral,
+wilton2015smc,griffiths2016coalescent}, 
+as well as, more commonly, the concrete realisations of the 
+ancestry~\citep[e.g.][]{gusfield2014recombinatorics,mathieson2020ancestry,brandt2021evaluation}.
+\cite{minichiello2006mapping} explicitly distinguished the ARG as a
+stochastic process from the ARG as a data structure, and
+argued for using the term to mean specific realised ancestral histories.
+In the following sections we
 show that, although the ARG-as-a-data-structure perspective is
@@
 ARG simplification. In all cases the graph visualization is shown on the left
-and the equivalent marginal trees on the right.
+and the equivalent local trees on the right.
 (A) A gARG simulated from a diploid Wright-Fisher
@@
 of removing nodes and re-writing edges in a gARG to remove
-various types of redundancy, while maintaining the crucial properties
-that the ``meaningful'' topology in the local trees is not affected,
-and the relationship between nodes in adjacent trees is maintained.
+various types of redundancy,
+retaining the ``meaningful'' topology within local trees
+and maintaining the identity of nodes across different trees.
 The idea of meaningful local tree topologies is based on the
@@
 model, shown both as a graph on the left, and the equivalent set of 6
-local trees on the right (Fig.~\ref{fig-simplification}A). Consider the path above node
-\textsf{a} in the penultimate tree, highlighted in yellow and covering genomic positions
-60 to 70. On this path, before \textsf{a} meets its common ancestor with the other samples at
+local trees on the right.
+For example, consider the path above node
+\textsf{a} in the highlighted tree (covering positions 60 to 70)
+in Fig.~\ref{fig-simplification}A.
+Before \textsf{a} meets its common ancestor with the other samples at
 node \textsf{q}, it passes through  \textsf{e},  \textsf{j},  \textsf{m},  \textsf{n} and
-\textsf{p}, which in this region of the genome are all unary nodes. This reflects the
+\textsf{p}, which in this region of the genome are all unary nodes.
+This reflects the
 passage of this span of genome through the ARG towards the root.
@@
  \emph{common} ancestor, such as node \textsf{q} in the highlighted
-tree. For many purposes, we would therefore consider a path directly
+tree. For most purposes, we would therefore consider a path directly
 joining sample \textsf{a} to node \textsf{q} to be equivalent to the path
@@
 more than two children (e.g.~node \textsf{s}). These represent a ``stacking up'' of
-multiple---often indistinguishable---events into a single node,
-something which is impossible to encode in an eARG.
+multiple---often indistinguishable---events into a single node.
+% something which is impossible to encode in an eARG.
 
@@
 simulation). The final level of simplification is to alter the edge annotations
-such that, although no nodes are removed from the graph, \emph{all}
+such that, although no nodes are removed from the graph, all
 unary nodes disappear from the local trees (Fig.~\ref{fig-simplification}D).
@@
 removing information about the exact paths taken by lineages through
-the graph, it removes potentially useful information about shared edges
+the graph, we lose potentially useful information about shared edges
 between trees. The extent to which this information can be inferred
@@
 but effective heuristics have been proposed~\citep{rasmussen2022espalier}.
+
+% JK TODO: finish up this section with a quick version of the
+% para here. Basically say, people often say an ARG is a
+% set of local trees cite cite, but it's not as simple as that as we
+% can see from the example (probably).
 
@@
 
-% The equivalence between an ARG and a set of local trees
-% separated by SPRs is often mentioned [e.g. x y z], but
-% it is important to note that while extracting local
-% trees from an ARG is straightforward, the converse
-% process of reconstructing precisely the same ARG from
-% these trees is more problematic. If the internal
-% nodes are consistently labelled and unary nodes are
-% included in the local trees (e.g., Fig~\ref{fig-simplification})
-% then it is easy to exactly reconstruct the ARG.
-% Otherwise, the problem is much more difficult and it
-% may not be possible to uniquely reconstruct the original ARG.
-% Without labels for the internal nodes, we are forced to
-% to reason about which subtrees remain the same and which
-% differ due to the SPR separating adjacent trees in order to
-% assign node identity. Without the internal unary nodes marking
-% recombinations, we must infer their existence. In
-% general, there will not be a unique solution to this problem,
-% and so there can be many possible ARGs compatible with
-% a given list of leaf-labelled binary trees.
-% This is the approach taken by the Espalier inference
-% method~\citep{rasmussen2022espalier}, which finds a
-% set of SPRs that reconcile independently inferred
-% trees from different regions of the genome.
-
 \section*{Computational efficiency}

### 2022-08-30 paper.tex
Create "Full ARG" section

@@
 % TODO update this when later sections have been drafted.
-As we see in sections XXX and YYY there are substantial limitations
-of expressivity and computational performance when using this eARG
-encoding. More fundamentally, however, the general approach
-of classifying nodes according to a discrete set of events
-and associating genome information with those nodes is intrinsically
-limiting. As we saw in the previous section, the original ARG
-definitions were derived directly from the coalescent with recombination
-stochastic process, and the types of ancestry that can be
-described are therefore limited by the assumptions of that
-model. While the coalescent as a model can be suprisingly robust to
-basic violations of the underlying assumptions~\citep{
-wakeley2012gene,bhaskar2014distortion,nelson2020accounting},
-it is not acceptable that our ability to \emph{represent}
-ancestral relationships should be limited by these assumptions.
-
-% n << Ne is a silly assumption in today's datasets
-A core assumption of the coalescent is that the sample size $n$
-is much less than the effective population size, $N_e$. This
-assumption ensures that multiple events (i.e., simultaneous
-coalescence and recombination) don't occur at the same
-time, which is reflected in the eARG definitions.
-This once-reasonable assumption has been overtaken by
-the data we have available.
-% This can be tightened up a bit, I think. We may need to drop a
-% few references.
-Several human datasets now consist of hundreds of thousands of
-genomes~\citep{bycroft2018genome,karczewski2020mutational,tanjo2021practical},
-and so sample size is substantially \emph{larger} than $N_e$
-(often assumed to be $10^4$ in humans).
-Agricultural datasets are even more extreme.
-For example,
-the US dairy cattle database alone currently comprises more than 6 million
-animals with SNP array
-genotypes\footnote{\url{https://queries.uscdcb.com/Genotype/counts.html}},
-while the effective population size in modern dairy cattle breeds is
-less than 100 and decreasing~\citep{MacLeod2013,Makanjuola2020}.
-An extreme sign of these breeding practices is that there are only two ancestral
-Y-chromosome linages present in today's US Holstein dairy breed~\citep{Yue2015}.
-Simlarly, the 1000 bull genomes project~\citep{hayes20191000}
-comprises close to 7000 genomes, which are part of multi-generation pedigrees
-with millions of animals and extensive SNP array genotype and phenotype
-data \citep[e.g.][]{Cesarani2022}.
-In another example, genome and SNP array data for 440,610 individuals within
-7 multi-generation pedigrees~\citep{whalen2018,Johnsson2021,Ros-Freixedes2020}
-were combined to infer recombination in pigs~\citep{RosFreixedes2022}.
-Effective population size in modern pig breeds is also less than 100 due to
-intense selection and directed reproduction \citep{Hall2016,Porcnic2016}.
-
-Another assumption of the coalescent with recombination which prevents complex
-combinations of events occurring simultaneously, is that the genome (or
-at least the region under study) is short enough that the number of extant
-lineages remains much smaller than $N_e$ at all times. Whole
-genome sequences have been available for model organisms
-for over a decade now, % True? Citation?
-and indeed complete chromosome-level assemblies are possible
-in humans~\citep{miga2020telomere}.
-Projects are under way to obtain high-quality assemblies
-for all eukaryotic species in Britain and Ireland~\citep{darwin2022sequence}
-and ultimately worldwide~\citep{lewin2022earth}.
-Clearly multiple recombinations (and other complex events) will
-occur in the ancestry of such datasets, and we should not be
-limited in what we can infer by the data structure that we
-use to represent the results.
+% As we see in sections XXX and YYY there are substantial limitations
+% of expressivity and computational performance when using this eARG
+% encoding. More fundamentally, however, the general approach
+% of classifying nodes according to a discrete set of events
+% and associating genome information with those nodes is intrinsically
+% limiting. As we saw in the previous section, the original ARG
+% definitions were derived directly from the coalescent with recombination
+% stochastic process, and the types of ancestry that can be
+% described are therefore limited by the assumptions of that
+% model.
 
@@
 
+\section*{When do we need ``full'' ARGs?}
+% The full ancestral recombination graph (ARG) is a structure that encodes all
+% coalescence and recombination events resulting from the stochastic process of
+% the coalescent with recombination.
+\citet{brandt2021evaluation} define a ``full'' ARG as
+``a structure that encodes all
+coalescence and recombination events resulting from the stochastic process of
+the coalescent with recombination'' (thus going against the general
+historial trend discussed in the XXX section).
+\citet{rasmussen2022espalier} have a less restrictive definition, in
+which a ``full'' ARG is one that contains complete information about
+recombination events, without necessarily being tied to a particular
+stochastic process.
+While [something positive], it is important to be clear on
+what the overall goal of such precise inferences are, and
+on the realistic limitations are on when they can be made.
+
+The main reasons for wanting precise estimates of all recombination
+events are to either compute a likelihood of the inferred ARG
+under a statistic model, or to study the recombinants in
+detail~\citet{rasmussen2022espalier}.
+% Explain that we're really interested in the details of recombination
+% itself with a small number of samples.
+In the latter case, we might
+have a small number of samples~\citep{guo2022recombination}.
+
+% If we want to use the coalescent, then really the assumptions
+% need to make sense.
+While the coalescent as a model can be suprisingly robust to
+basic violations of the underlying assumptions~\citep{
+wakeley2012gene,bhaskar2014distortion,nelson2020accounting},
+% n << Ne is a silly assumption in today's datasets
+A core assumption of the coalescent is that the sample size $n$
+is much less than the effective population size, $N_e$. This
+assumption ensures that multiple events (i.e., simultaneous
+coalescence and recombination) don't occur at the same
+time, which is reflected in the eARG definitions.
+This once-reasonable assumption has been overtaken by
+the data we have available.
+% This can be tightened up a bit, I think. We may need to drop a
+% few references.
+Several human datasets now consist of hundreds of thousands of
+genomes~\citep{bycroft2018genome,karczewski2020mutational,tanjo2021practical},
+and so sample size is substantially \emph{larger} than $N_e$
+(often assumed to be $10^4$ in humans).
+Agricultural datasets are even more extreme.
+For example,
+the US dairy cattle database alone currently comprises more than 6 million
+animals with SNP array
+genotypes\footnote{\url{https://queries.uscdcb.com/Genotype/counts.html}},
+while the effective population size in modern dairy cattle breeds is
+less than 100 and decreasing~\citep{MacLeod2013,Makanjuola2020}.
+An extreme sign of these breeding practices is that there are only two ancestral
+Y-chromosome linages present in today's US Holstein dairy breed~\citep{Yue2015}.
+Simlarly, the 1000 bull genomes project~\citep{hayes20191000}
+comprises close to 7000 genomes, which are part of multi-generation pedigrees
+with millions of animals and extensive SNP array genotype and phenotype
+data \citep[e.g.][]{Cesarani2022}.
+In another example, genome and SNP array data for 440,610 individuals within
+7 multi-generation pedigrees~\citep{whalen2018,Johnsson2021,Ros-Freixedes2020}
+were combined to infer recombination in pigs~\citep{RosFreixedes2022}.
+Effective population size in modern pig breeds is also less than 100 due to
+intense selection and directed reproduction \citep{Hall2016,Porcnic2016}.
+
+Another assumption of the coalescent with recombination which prevents complex
+combinations of events occurring simultaneously, is that the genome (or
+at least the region under study) is short enough that the number of extant
+lineages remains much smaller than $N_e$ at all times. Whole
+genome sequences have been available for model organisms
+for over a decade now, % True? Citation?
+and indeed complete chromosome-level assemblies are possible
+in humans~\citep{miga2020telomere}.
+Projects are under way to obtain high-quality assemblies
+for all eukaryotic species in Britain and Ireland~\citep{darwin2022sequence}
+and ultimately worldwide~\citep{lewin2022earth}.
+% Clearly multiple recombinations (and other complex events) will
+% occur in the ancestry of such datasets, and we should not be
+% limited in what we can infer by the data structure that we
+% use to represent the results.
+
+Such datasets cannot be reasonably be described by
+the coalescent as a statical model, and so more non-parametric
+methods like \tsinfer\ and \relate\ would seem more appropriate.
+[More]
+
 \section*{Computational efficiency}
@@
     present in modern datasets.
+
+\item In general, discussions tend to contrast an ARG
+on one hand as holding \emph{complete} information about recombination
+(and therefore between-tree correlation structure)
+and a sequence of local trees on the other (with, by implication,
+no direct information about between-tree correlations).
+A key goal of this article is to show that this is a false dichotomy:
+a \emph{continuum} of information about recombination and between-tree
+correlation structure can be represented and inferred, and all
+such levels of information are useful, even if only for computational
+efficiency.
 
@@
 looked at either some form of pairwise comparison of trees along the genome
-using tree metrics, or looked at the distributions of TMRCAs. No
+using tree metrics as proposed by \citet{kuhner2015assessing},
+or looked at the distributions of TMRCAs. No
 one has succeeded in performing any comparisons of ARGs beyond
@@
 TODO mention \citep{schaefer2021ancestral} somewhere.
+Also \citep{kuhner2017consensus}.
 

### 2022-08-31 paper.tex
Clarify narrative in several places.

@@
 recombination.
+\item In the following sections we
+show that, although the ARG-as-a-data-structure perspective is
+now the most common interpretation, the \emph{form} of that
+data structure is still closely linked to the original
+stochastic process, and that this form is inflexible
+in the types of genetic inheritance that can be described,
+computationally inefficient, and
+forces complete (and potentially spurious) precision about
+the location and timing
+of recombination events. % (and coalescent events?)
 \item Suggest a definition of ARGs that is free of these limitations,
@@
 argued for using the term to mean specific realised ancestral histories.
-In the following sections we
-show that, although the ARG-as-a-data-structure perspective is
-now the most common interpretation, the \emph{form} of that
-data structure is still closely linked to the original
-stochastic process, and that this form is inflexible
-in the types of genetic inheritance that can be described,
-computationally inefficient, and
-forces complete (and potentially spurious) precision about
-the location and timing
-of recombination events. % (and coalescent events?)
 Unless otherwise stated, we will use the ARG-as-a-data-structure
@@
 in terms of the process in which we extract the local trees
-along the genome from an eARG (the most basic downstream
-operation for an ARG data structure).
+along the genome from an eARG.
 To recover the tree at genome coordinate $x$,
@@
 when using this representation and simulating an ARG backwards in time,
-the first event above a recombination may involve the lineage carrying the
+the first event older than a recombination may involve the lineage carrying the
 ancestry to the right of the breakpoint rather
@@
 genetic inheritance).
-Fig~\ref{fig-arg-in-pedigree} illustrates the basic elements of a gARG,
+Fig.~\ref{fig-arg-in-pedigree} illustrates the basic elements of a gARG,
 showing how the genomes embedded in an example (highly inbred)
@@
 
+% This paragraph is a bit weak and muddled, needs a few more passes.
+% In particular, the "Finally" bit doesn't really seem to follow
+% from the others. Is there a more direct way we can address these points ?
 There are some interesting and important consequences to this focus on genomes
@@
 (common ancestor or recombination) there is no limitation
-on the complexity what can be modelled---multiple recombinations
-and coalescences can happen simultaneously in the same genome.
-In large deeply sequenced pedigrees such complexities are commonplace.
+on the complexity that can be modelled---multiple recombinations
+and coalescences can happen simultaneously in the same genome,
+a common occurance in large deeply sequenced
+pedigrees~\citep{hayes20191000,RosFreixedes2022}.
 Finally, there is a subtle and important point about how recombination
@@
 do nodes represent?
-In principle, we can intepret nodes as representing genomes
+In principle, we can interpret nodes as representing genomes
 at any point in the life cycle [supplementary Figure ?]. Hudson
@@
 The idea of meaningful local tree topologies is based on the
-presence of ``unary'' nodes: a node in the tree that has exactly
+presence of ``unary'' nodes: a tree node that has exactly
 one child. Fig.~\ref{fig-simplification} show a series of successive steps of
@@
 The first type of simplification that we can perform is to remove
-graph topology that is invisible to the samples. The best
-known example of such topology is a so-called
+graph topology that is invisible to the samples.
+An example of such topology is a
 ``diamond''~\citep{rasmussen2014genome}
@@
 events are to either compute a likelihood of the inferred ARG
-under a statistic model, or to study the recombinants in
+under a statistical model, or to study the recombinants in
 detail~\citet{rasmussen2022espalier}.
@@
 
+Beyond cases in which visual inspection of the trees and the
+effects of recombination is feasible, the main application
+of having completely precise estimates of recombination events is
+to evaluate the likelihood of the estimate under some
+statistical model. The only statistical model available
+is the coalescent with recombination
+and its approximations~\citep{mcvean2005approximating,marjoram2006fast}.
+In this case, the utility of a ``full`` ARG for some dataset
+is limited by how well the dataset fits the assumptions
+of the model.
+
 % If we want to use the coalescent, then really the assumptions
 % need to make sense.
-While the coalescent as a model can be suprisingly robust to
-basic violations of the underlying assumptions~\citep{
-wakeley2012gene,bhaskar2014distortion,nelson2020accounting},
 % n << Ne is a silly assumption in today's datasets
 A core assumption of the coalescent is that the sample size $n$
-is much less than the effective population size, $N_e$. This
-assumption ensures that multiple events (i.e., simultaneous
-coalescence and recombination) don't occur at the same
-time, which is reflected in the eARG definitions.
+is much less than the effective population size, $N_e$.
+% This
+% assumption ensures that multiple events (i.e., simultaneous
+% coalescence and recombination) don't occur at the same
+% time, which is reflected in the eARG definitions.
 This once-reasonable assumption has been overtaken by
@@
 intense selection and directed reproduction \citep{Hall2016,Porcnic2016}.
-
-Another assumption of the coalescent with recombination which prevents complex
-combinations of events occurring simultaneously, is that the genome (or
+Another core assumption of the coalescent model is that the genome (or
 at least the region under study) is short enough that the number of extant
@@
 % use to represent the results.
+
+While the coalescent as a model can be suprisingly robust to
+basic violations of the underlying assumptions~\citep{
+wakeley2012gene,bhaskar2014distortion,nelson2020accounting}, [something
+something]
 

### 2022-09-01 paper.tex
Paragraph on downstream apps for full ARGs

@@
 
+% Do downstream applications actually make use of such full ARGs?
+Downstream applications using estimated ARGs to date rarely
+make use of such complete recombination information.
+The \relate\ selection statistic~\citep{speidel2019method}
+computes a $p$-value for selection [FIXME] for each
+local tree independently by comparing the rates of coalescence
+at various time-points. For cases like strong positive selection,
+where we expect there to be a clear signal in the local tree,
+such tree-by-tree methods can be simple and powerful.
+In their method
+for estimating dispersal rates and the locations of genetic
+ancestors,
+\citep{osmond2021estimating} downsample trees along the genome
+so that they can be regarded as approximately independent.
+The SIA method for detecting selection~\citep{hejase2022deep}
+encodes local trees as a set of lineage counts at discrete
+time intervals, and uses these as feature for a
+type of machine learning algorithm
+that takes ``temporal'' correlations into account. Thus,
+while SIA does use some of the information about local tree correlation,
+clearly much of the detail about recombination
+events in a ``full'' ARG is lost.
+The \texttt{tsdate} algorithm (and related naive estimator
+for ancestral location) uses much more between-tree
+information~\citep{wohns2022unified}, explicitly using the gARG
+encoding to reason about ancestral nodes.
+A major strength of this approach is that it can be used on
+any gARG, ranging from estimates containing relatively little
+information about recombination (e.g., as estimated by \relate)
+to ``full'' ARGs containing complete recombination information
+(e.g., as estimated by \kwarg), and will fully utilise
+any between-tree correlation information that is present.
+
 \section*{Computational efficiency}

### 2022-09-01 paper.tex
Add shipilina et al and some places to mention

@@
 While [something positive], it is important to be clear on
-what the overall goal of such precise inferences are, and
-on the realistic limitations are on when they can be made.
+what the overall goal of such precise inferences are,
+what the limitations on when they can be made are,
+and their downstream utility in applications.
 
@@
 
+% Is it practical to estimate full ARGs?
+[Not particularly \citep{shipilina2022origin}]
+
 % Do downstream applications actually make use of such full ARGs?
@@
 
+\item This idea of a continuum of information about recombination
+being inferrable and representable is consistent with the ideas
+discussed by~\cite{shipilina2022origin}. There a haplotype block
+is synonymous with coalescent nodes in the ARG, which we can generalise
+as nodes in a gARG with at least two children. In principle, these
+nodes are identifiable by the mutations that occur on the outbound
+edges of the node.
+Their definition of an ARG is interesting, as they emphasise that it
+is composed of events, but illustate the genomes. [Nice to mention
+this connection somehere if we can, see Fig B1]
+
 \item It is unfortunate that we are adding yet another meaning to the

### 2022-09-01 paper.tex
Restructuring and editing history of ARGs section

--- 2022-08-31 paper.tex
@@
 recombination.
-\item In the following sections we
-show that, although the ARG-as-a-data-structure perspective is
-now the most common interpretation, the \emph{form} of that
-data structure is still closely linked to the original
-stochastic process, and that this form is inflexible
-in the types of genetic inheritance that can be described,
-computationally inefficient, and
-forces complete (and potentially spurious) precision about
-the location and timing
-of recombination events. % (and coalescent events?)
 \item Suggest a definition of ARGs that is free of these limitations,
@@
 
-% Mathematical treatment of the CwR and ARGs (and ARG-like stochastic processes)
+\paragraph*{Properties of the coalescent with recombination and related models}
 An extensive body of work has developed in
@@
 
-% Inference of parameters and explicit ARG reconstruction
+\paragraph*{Inference of evolutionary parameters and genealogies}
 Early work on inference under the coalescent with recombination
@@
 
-% Optimisation problems to do with ARGs and other phylogenetic networks
+\paragraph*{Optimisation problems for phylogenetic networks}
 There has also been substantial interest in formulating and answering 
@@
 
-The goal of this historical overview is to illustrate that the meaning of the term ``ARG" now strongly
+\paragraph*{} The goal of this historical overview is to illustrate that the meaning of the term ``ARG" now strongly
 depends on the context in which it is used, and can mean both the 
@@
 argued for using the term to mean specific realised ancestral histories.
+In the following sections we
+show that, although the ARG-as-a-data-structure perspective is
+now the most common interpretation, the \emph{form} of that
+data structure is still closely linked to the original
+stochastic process, and that this form is inflexible
+in the types of genetic inheritance that can be described,
+computationally inefficient, and
+forces complete (and potentially spurious) precision about
+the location and timing
+of recombination events. % (and coalescent events?)
 Unless otherwise stated, we will use the ARG-as-a-data-structure
@@
 in terms of the process in which we extract the local trees
-along the genome from an eARG.
+along the genome from an eARG (the most basic downstream
+operation for an ARG data structure).
 To recover the tree at genome coordinate $x$,
@@
 when using this representation and simulating an ARG backwards in time,
-the first event older than a recombination may involve the lineage carrying the
+the first event above a recombination may involve the lineage carrying the
 ancestry to the right of the breakpoint rather
@@
 % TODO update this when later sections have been drafted.
-% As we see in sections XXX and YYY there are substantial limitations
-% of expressivity and computational performance when using this eARG
-% encoding. More fundamentally, however, the general approach
-% of classifying nodes according to a discrete set of events
-% and associating genome information with those nodes is intrinsically
-% limiting. As we saw in the previous section, the original ARG
-% definitions were derived directly from the coalescent with recombination
-% stochastic process, and the types of ancestry that can be
-% described are therefore limited by the assumptions of that
-% model.
+As we see in sections XXX and YYY there are substantial limitations
+of expressivity and computational performance when using this eARG
+encoding. More fundamentally, however, the general approach
+of classifying nodes according to a discrete set of events
+and associating genome information with those nodes is intrinsically
+limiting. As we saw in the previous section, the original ARG
+definitions were derived directly from the coalescent with recombination
+stochastic process, and the types of ancestry that can be
+described are therefore limited by the assumptions of that
+model. While the coalescent as a model can be suprisingly robust to
+basic violations of the underlying assumptions~\citep{
+wakeley2012gene,bhaskar2014distortion,nelson2020accounting},
+it is not acceptable that our ability to \emph{represent}
+ancestral relationships should be limited by these assumptions.
+
+% n << Ne is a silly assumption in today's datasets
+A core assumption of the coalescent is that the sample size $n$
+is much less than the effective population size, $N_e$. This
+assumption ensures that multiple events (i.e., simultaneous
+coalescence and recombination) don't occur at the same
+time, which is reflected in the eARG definitions.
+This once-reasonable assumption has been overtaken by
+the data we have available.
+% This can be tightened up a bit, I think. We may need to drop a
+% few references.
+Several human datasets now consist of hundreds of thousands of
+genomes~\citep{bycroft2018genome,karczewski2020mutational,tanjo2021practical},
+and so sample size is substantially \emph{larger} than $N_e$
+(often assumed to be $10^4$ in humans).
+Agricultural datasets are even more extreme.
+For example,
+the US dairy cattle database alone currently comprises more than 6 million
+animals with SNP array
+genotypes\footnote{\url{https://queries.uscdcb.com/Genotype/counts.html}},
+while the effective population size in modern dairy cattle breeds is
+less than 100 and decreasing~\citep{MacLeod2013,Makanjuola2020}.
+An extreme sign of these breeding practices is that there are only two ancestral
+Y-chromosome linages present in today's US Holstein dairy breed~\citep{Yue2015}.
+Simlarly, the 1000 bull genomes project~\citep{hayes20191000}
+comprises close to 7000 genomes, which are part of multi-generation pedigrees
+with millions of animals and extensive SNP array genotype and phenotype
+data \citep[e.g.][]{Cesarani2022}.
+In another example, genome and SNP array data for 440,610 individuals within
+7 multi-generation pedigrees~\citep{whalen2018,Johnsson2021,Ros-Freixedes2020}
+were combined to infer recombination in pigs~\citep{RosFreixedes2022}.
+Effective population size in modern pig breeds is also less than 100 due to
+intense selection and directed reproduction \citep{Hall2016,Porcnic2016}.
+
+Another assumption of the coalescent with recombination which prevents complex
+combinations of events occurring simultaneously, is that the genome (or
+at least the region under study) is short enough that the number of extant
+lineages remains much smaller than $N_e$ at all times. Whole
+genome sequences have been available for model organisms
+for over a decade now, % True? Citation?
+and indeed complete chromosome-level assemblies are possible
+in humans~\citep{miga2020telomere}.
+Projects are under way to obtain high-quality assemblies
+for all eukaryotic species in Britain and Ireland~\citep{darwin2022sequence}
+and ultimately worldwide~\citep{lewin2022earth}.
+Clearly multiple recombinations (and other complex events) will
+occur in the ancestry of such datasets, and we should not be
+limited in what we can infer by the data structure that we
+use to represent the results.
 
@@
 genetic inheritance).
-Fig.~\ref{fig-arg-in-pedigree} illustrates the basic elements of a gARG,
+Fig~\ref{fig-arg-in-pedigree} illustrates the basic elements of a gARG,
 showing how the genomes embedded in an example (highly inbred)
@@
 
-% This paragraph is a bit weak and muddled, needs a few more passes.
-% In particular, the "Finally" bit doesn't really seem to follow
-% from the others. Is there a more direct way we can address these points ?
 There are some interesting and important consequences to this focus on genomes
@@
 (common ancestor or recombination) there is no limitation
-on the complexity that can be modelled---multiple recombinations
-and coalescences can happen simultaneously in the same genome,
-a common occurance in large deeply sequenced
-pedigrees~\citep{hayes20191000,RosFreixedes2022}.
+on the complexity what can be modelled---multiple recombinations
+and coalescences can happen simultaneously in the same genome.
+In large deeply sequenced pedigrees such complexities are commonplace.
 Finally, there is a subtle and important point about how recombination
@@
 do nodes represent?
-In principle, we can interpret nodes as representing genomes
+In principle, we can intepret nodes as representing genomes
 at any point in the life cycle [supplementary Figure ?]. Hudson
@@
 of removing nodes and re-writing edges in a gARG to remove
-various types of redundancy,
-retaining the ``meaningful'' topology within local trees
-and maintaining the identity of nodes across different trees.
+various types of redundancy, while maintaining the crucial properties
+that the ``meaningful'' topology in the local trees is not affected,
+and the relationship between nodes in adjacent trees is maintained.
 The idea of meaningful local tree topologies is based on the
-presence of ``unary'' nodes: a tree node that has exactly
+presence of ``unary'' nodes: a node in the tree that has exactly
 one child. Fig.~\ref{fig-simplification} show a series of successive steps of
@@
 model, shown both as a graph on the left, and the equivalent set of 6
-local trees on the right.
-For example, consider the path above node
-\textsf{a} in the highlighted tree (covering positions 60 to 70)
-in Fig.~\ref{fig-simplification}A.
-Before \textsf{a} meets its common ancestor with the other samples at
+local trees on the right (Fig.~\ref{fig-simplification}A). Consider the path above node
+\textsf{a} in the penultimate tree, highlighted in yellow and covering genomic positions
+60 to 70. On this path, before \textsf{a} meets its common ancestor with the other samples at
 node \textsf{q}, it passes through  \textsf{e},  \textsf{j},  \textsf{m},  \textsf{n} and
-\textsf{p}, which in this region of the genome are all unary nodes.
-This reflects the
+\textsf{p}, which in this region of the genome are all unary nodes. This reflects the
 passage of this span of genome through the ARG towards the root.
@@
  \emph{common} ancestor, such as node \textsf{q} in the highlighted
-tree. For most purposes, we would therefore consider a path directly
+tree. For many purposes, we would therefore consider a path directly
 joining sample \textsf{a} to node \textsf{q} to be equivalent to the path
@@
 The first type of simplification that we can perform is to remove
-graph topology that is invisible to the samples.
-An example of such topology is a
+graph topology that is invisible to the samples. The best
+known example of such topology is a so-called
 ``diamond''~\citep{rasmussen2014genome}
@@
 more than two children (e.g.~node \textsf{s}). These represent a ``stacking up'' of
-multiple---often indistinguishable---events into a single node.
-% something which is impossible to encode in an eARG.
+multiple---often indistinguishable---events into a single node,
+something which is impossible to encode in an eARG.
 
@@
 simulation). The final level of simplification is to alter the edge annotations
-such that, although no nodes are removed from the graph, all
+such that, although no nodes are removed from the graph, \emph{all}
 unary nodes disappear from the local trees (Fig.~\ref{fig-simplification}D).
@@
 removing information about the exact paths taken by lineages through
-the graph, we lose potentially useful information about shared edges
+the graph, it removes potentially useful information about shared edges
 between trees. The extent to which this information can be inferred
@@
 
-\section*{When do we need ``full'' ARGs?}
-% The full ancestral recombination graph (ARG) is a structure that encodes all
-% coalescence and recombination events resulting from the stochastic process of
-% the coalescent with recombination.
-\citet{brandt2021evaluation} define a ``full'' ARG as
-``a structure that encodes all
-coalescence and recombination events resulting from the stochastic process of
-the coalescent with recombination'' (thus going against the general
-historial trend discussed in the XXX section).
-\citet{rasmussen2022espalier} have a less restrictive definition, in
-which a ``full'' ARG is one that contains complete information about
-recombination events, without necessarily being tied to a particular
-stochastic process.
-While [something positive], it is important to be clear on
-what the overall goal of such precise inferences are, and
-on the realistic limitations are on when they can be made.
-
-The main reasons for wanting precise estimates of all recombination
-events are to either compute a likelihood of the inferred ARG
-under a statistical model, or to study the recombinants in
-detail~\citet{rasmussen2022espalier}.
-% Explain that we're really interested in the details of recombination
-% itself with a small number of samples.
-In the latter case, we might
-have a small number of samples~\citep{guo2022recombination}.
-
-Beyond cases in which visual inspection of the trees and the
-effects of recombination is feasible, the main application
-of having completely precise estimates of recombination events is
-to evaluate the likelihood of the estimate under some
-statistical model. The only statistical model available
-is the coalescent with recombination
-and its approximations~\citep{mcvean2005approximating,marjoram2006fast}.
-In this case, the utility of a ``full`` ARG for some dataset
-is limited by how well the dataset fits the assumptions
-of the model.
-
-% If we want to use the coalescent, then really the assumptions
-% need to make sense.
-% n << Ne is a silly assumption in today's datasets
-A core assumption of the coalescent is that the sample size $n$
-is much less than the effective population size, $N_e$.
-% This
-% assumption ensures that multiple events (i.e., simultaneous
-% coalescence and recombination) don't occur at the same
-% time, which is reflected in the eARG definitions.
-This once-reasonable assumption has been overtaken by
-the data we have available.
-% This can be tightened up a bit, I think. We may need to drop a
-% few references.
-Several human datasets now consist of hundreds of thousands of
-genomes~\citep{bycroft2018genome,karczewski2020mutational,tanjo2021practical},
-and so sample size is substantially \emph{larger} than $N_e$
-(often assumed to be $10^4$ in humans).
-Agricultural datasets are even more extreme.
-For example,
-the US dairy cattle database alone currently comprises more than 6 million
-animals with SNP array
-genotypes\footnote{\url{https://queries.uscdcb.com/Genotype/counts.html}},
-while the effective population size in modern dairy cattle breeds is
-less than 100 and decreasing~\citep{MacLeod2013,Makanjuola2020}.
-An extreme sign of these breeding practices is that there are only two ancestral
-Y-chromosome linages present in today's US Holstein dairy breed~\citep{Yue2015}.
-Simlarly, the 1000 bull genomes project~\citep{hayes20191000}
-comprises close to 7000 genomes, which are part of multi-generation pedigrees
-with millions of animals and extensive SNP array genotype and phenotype
-data \citep[e.g.][]{Cesarani2022}.
-In another example, genome and SNP array data for 440,610 individuals within
-7 multi-generation pedigrees~\citep{whalen2018,Johnsson2021,Ros-Freixedes2020}
-were combined to infer recombination in pigs~\citep{RosFreixedes2022}.
-Effective population size in modern pig breeds is also less than 100 due to
-intense selection and directed reproduction \citep{Hall2016,Porcnic2016}.
-Another core assumption of the coalescent model is that the genome (or
-at least the region under study) is short enough that the number of extant
-lineages remains much smaller than $N_e$ at all times. Whole
-genome sequences have been available for model organisms
-for over a decade now, % True? Citation?
-and indeed complete chromosome-level assemblies are possible
-in humans~\citep{miga2020telomere}.
-Projects are under way to obtain high-quality assemblies
-for all eukaryotic species in Britain and Ireland~\citep{darwin2022sequence}
-and ultimately worldwide~\citep{lewin2022earth}.
-% Clearly multiple recombinations (and other complex events) will
-% occur in the ancestry of such datasets, and we should not be
-% limited in what we can infer by the data structure that we
-% use to represent the results.
-
-While the coalescent as a model can be suprisingly robust to
-basic violations of the underlying assumptions~\citep{
-wakeley2012gene,bhaskar2014distortion,nelson2020accounting}, [something
-something]
-
-Such datasets cannot be reasonably be described by
-the coalescent as a statical model, and so more non-parametric
-methods like \tsinfer\ and \relate\ would seem more appropriate.
-[More]
-
 \section*{Computational efficiency}
@@
     present in modern datasets.
-
-\item In general, discussions tend to contrast an ARG
-on one hand as holding \emph{complete} information about recombination
-(and therefore between-tree correlation structure)
-and a sequence of local trees on the other (with, by implication,
-no direct information about between-tree correlations).
-A key goal of this article is to show that this is a false dichotomy:
-a \emph{continuum} of information about recombination and between-tree
-correlation structure can be represented and inferred, and all
-such levels of information are useful, even if only for computational
-efficiency.
 
@@
 looked at either some form of pairwise comparison of trees along the genome
-using tree metrics as proposed by \citet{kuhner2015assessing},
-or looked at the distributions of TMRCAs. No
+using tree metrics, or looked at the distributions of TMRCAs. No
 one has succeeded in performing any comparisons of ARGs beyond
@@
 TODO mention \citep{schaefer2021ancestral} somewhere.
-Also \citep{kuhner2017consensus}.
 

### 2022-09-01 paper.tex
Tweaks

--- 2022-09-01 paper.tex
@@
 
+% What is the goal of having fully precise ARGs?
+% 1) studying the actual recombination events;
+% 2) evaluating the likelihoods
 The main reasons for wanting precise estimates of all recombination
@@
 under a statistical model, or to study the recombinants in
-detail~\citet{rasmussen2022espalier}.
+detail~\citep{rasmussen2022espalier}.
 % Explain that we're really interested in the details of recombination
@@
 is much less than the effective population size, $N_e$.
-% This
-% assumption ensures that multiple events (i.e., simultaneous
-% coalescence and recombination) don't occur at the same
-% time, which is reflected in the eARG definitions.
-This once-reasonable assumption has been overtaken by
-the data we have available.
-% This can be tightened up a bit, I think. We may need to drop a
-% few references.
 Several human datasets now consist of hundreds of thousands of
@@
 and ultimately worldwide~\citep{lewin2022earth}.
-% Clearly multiple recombinations (and other complex events) will
-% occur in the ancestry of such datasets, and we should not be
-% limited in what we can infer by the data structure that we
-% use to represent the results.
 
@@
 basic violations of the underlying assumptions~\citep{
-wakeley2012gene,bhaskar2014distortion,nelson2020accounting}, [something
-something]
-
+wakeley2012gene,bhaskar2014distortion,nelson2020accounting}
+it is reasonable to question the validity of the model in
+representing datasets such as UK Biobank.
+[Also mention somehow that $n\ll N_e$ is only one problem -
+we're also ignoring the myriad other evolutionary forces.]
 Such datasets cannot be reasonably be described by
@@
 % Is it practical to estimate full ARGs?
-[Not particularly \citep{shipilina2022origin}]
+% [Not particularly \citep{shipilina2022origin} - but maybe ARGneedle
+% has done this for UKB? Need to check]
 

### 2022-09-01 paper.tex
Deleting paragraph headings

--- 2022-09-01 paper.tex
@@
 
-\paragraph*{Properties of the coalescent with recombination and related models}
 An extensive body of work has developed in
@@
 
-\paragraph*{Inference of evolutionary parameters and genealogies}
 Early work on inference under the coalescent with recombination
@@
 
-\paragraph*{Optimisation problems for phylogenetic networks}
 There has also been substantial interest in formulating and answering 
@@
 
-\paragraph*{} The goal of this historical overview is to illustrate that the meaning of the term ``ARG" now strongly
+The goal of this historical overview is to illustrate that the meaning of the term ``ARG" now strongly
 depends on the context in which it is used, and can mean both the 

### 2022-09-13 paper.tex
Comments with paragraph summaries
Merge eARG and gARG sections

--- 2022-09-01 paper.tex
@@
 
-\section*{Event ARGs}\label{eARG}
+\section*{Event vs Genome ARGs}\label{eARG}
 In keeping with its original derivation in terms of a stochastic
 process, the classical definition of an ARG data structure is
-in terms of \emph{events}. Nodes represent
+in terms of events.
+Nodes represent
 common ancestor or recombination events that occurred in the
@@
 lineages~\citep{griffiths1996ancestral}.
-In order to distinguish this form of ARG from what we will be discussing
-in later sections we refer to it as an ``event ARG``, or eARG for brevity.
-An example of an eARG
-is given in Fig.~\ref{fig-arg-data-structure}B.
 In a common ancestor event the inbound lineages are merged into a
-single ancestral lineage (e.g., node \noderef{e}
-in Fig.~\ref{fig-arg-data-structure}), and in a recombination
+single ancestral lineage, and in a recombination
 event a lineage is split into two independent
-ancestral lineages (e.g., node \noderef{d}
-in Fig.~\ref{fig-arg-data-structure}). The final vital
-detail---which distinguishes an arbitrary graph with the required topological
-properties from an eARG---is that we associate a breakpoint with each recombination
-event. This information is sufficient to uniquely
-define the local genealogical trees at every position along the genome,
-which is the basic requirement for a data structure encoding
-genome-wide genetic ancestry.
+ancestral lineages.
+% The final vital
+% detail is that we associate a breakpoint with each recombination
+% event.
+% This information is sufficient to uniquely
+% define the local genealogical trees at every position along the genome,
+% which is the basic requirement for a data structure encoding
+% genome-wide genetic ancestry.
+As illustrated in Fig.~\ref{fig-arg-data-structure}, this information
+is usually concretely encoded by associating a recombination breakpoint
+(which is null for common ancestor events) and an ordered list of parents
+with each node. We refer to such a  data structure as an
+``event ARG'' (eARG).
+To recover the local tree at genome coordinate $x$ (the most
+fundamental operation required for an ARG data structure),
+we traverse the graph rootwards from the leaves.
+At a particular
+node $u$, if it has one parent we are at a common ancestor
+node and we follow that parent. If we are at a
+recombination node, $u$ has two parents: if
+$x$ is less than the breakpoint we follow
+the edge to the first parent, and otherwise follow the edge
+to the second parent.
+The order in which parents are listed at a recombination node is
+therefore significant, telling us
+from which parent the segments of genome to the left and right of the breakpoint
+were inherited.
+
+% This ordering requirement, while straightforward
+% to describe, has some practical drawbacks. For example,
+% % Surely the meaning is clear from the context here and we don't need
+% % qualify the meaning of ``ARG''
+% when using this representation and simulating an ARG backwards in time,
+% the first event older than a recombination may involve the lineage carrying the
+% ancestry to the right of the breakpoint rather
+% than the left, and so we cannot emit edges as they are generated.
+% Such issues can be worked around, of course, but depending on the ordering of
+% otherwise indistinguishable objects is generally problematic.
 
@@
 
-\node [anchor=north west] at (-1.5,6) {B};
+% \node [anchor=north west] at (0,6) {A};
 \node [nodelabel,anchor=north west] at ($(s3) + (0,0)$) {$x$};
@@
 
-\node [anchor=north west] at (-14,6) {A};
-\node [nodelabel,anchor=north west] at ($(-13,5)$) {
-\begin{tabular}{c|c}
+% \node [anchor=north west] at (9,6) {B};
+\node [nodelabel,anchor=north west] at ($(10,5)$) {
+\begin{tabular}{c|c|l}
 % \multicolumn{2}{c}{Breakpoints}\\
-Nodes & Breakpoints\\
+Node & Breakpoint & Parents\\
 \hline
-$\textsf{a}$ & $\varnothing$ \\
-$\textsf{b}$ & $\varnothing$ \\
-$\textsf{c}$ & $\varnothing$ \\
-$\textsf{d}$ & $x$ \\
-$\textsf{e}$ & $\varnothing$ \\
-$\textsf{f}$ & $\varnothing$ \\
-$\textsf{g}$ & $\varnothing$ \\
+$\noderef{a}$ & $\varnothing$ & [\noderef{e}]\\
+$\noderef{b}$ & $\varnothing$ & [\noderef{d}] \\
+$\noderef{c}$ & $\varnothing$ & [\noderef{f}]\\
+$\noderef{d}$ & $x$ & [\noderef{e}, \noderef{f}] \\
+$\noderef{e}$ & $\varnothing$ & [\noderef{g}]\\
+$\noderef{f}$ & $\varnothing$ & [\noderef{g}]\\
+$\noderef{g}$ & $\varnothing$ & []\\
 \end{tabular}};
 
-\node [nodelabel,anchor=north west] at ($(-7,5)$) {
-\begin{tabular}{l}
-Edges\\
-\hline
-$(\textsf{a}, \textsf{e})$ \\
-$(\textsf{b}, \textsf{d})$ \\
-$(\textsf{c}, \textsf{f})$ \\
-$(\textsf{d}, \textsf{e})$ \\
-$(\textsf{d}, \textsf{f})$ \\
-$(\textsf{e}, \textsf{g})$ \\
-$(\textsf{f}, \textsf{g})$ \\
-\end{tabular}};
-
-\node [anchor=north west] at (9,6) {C};
-\node [nodelabel,anchor=north west] at ($(10,5)$) {
-\begin{tabular}{ll}
-Edges & Intervals\\
-\hline
-$(\textsf{a}, \textsf{e})$ & $(0, L]$ \\
-$(\textsf{b}, \textsf{d})$ & $(0, L]$ \\
-$(\textsf{c}, \textsf{f})$ & $(0, L]$ \\
-$(\textsf{d}, \textsf{e})$ & $(0, x]$ \\
-$(\textsf{d}, \textsf{f})$ & $(x, L]$ \\
-$(\textsf{e}, \textsf{g})$ & $(0, L]$ \\
-$(\textsf{f}, \textsf{g})$ & $(0, L]$ \\
-\end{tabular}};
+% \node [anchor=north west] at (9,6) {C};
+% \node [nodelabel,anchor=north west] at ($(10,5)$) {
+% \begin{tabular}{ll}
+% Edges & Intervals\\
+% \hline
+% $(\textsf{a}, \textsf{e})$ & $(0, L]$ \\
+% $(\textsf{b}, \textsf{d})$ & $(0, L]$ \\
+% $(\textsf{c}, \textsf{f})$ & $(0, L]$ \\
+% $(\textsf{d}, \textsf{e})$ & $(0, x]$ \\
+% $(\textsf{d}, \textsf{f})$ & $(x, L]$ \\
+% $(\textsf{e}, \textsf{g})$ & $(0, L]$ \\
+% $(\textsf{f}, \textsf{g})$ & $(0, L]$ \\
+% \end{tabular}};
 
@@
 \caption{\label{fig-arg-data-structure}
-A classical ARG (subfigure B), contrasting node with edge breakpoint annotations.
-(A) recombination breakpoint information encoded on \emph{nodes}.
-This encoding is limited in the patterns of inheritance that can be
-described and requires that edges be ordered to avoid ambiguity.
-(C) breakpoint annotations associated with \emph{edges}.
-Associating the interval of genome inherited by the child node
-from the parent with the corresponding edge allows
-any pattern genetic inheritance to be described and does
-not require edges to be ordered. In this example, each edge is associated
-with a single interval, but more generally, multiple non-overlapping intervals
-can be associated with a single edge.
+A graphical depiction of a classical event ARG and a concrete
+encoding of the information. Node \noderef{d} is a recombination
+event as it is associated with a non-null breakpoint, and has
+two parents. Nodes \noderef{e}, \noderef{f} and \noderef{g} are
+common ancestor events.
 }
@@
 
-% What do we mean by an eARG, formally?
-To be concrete, we can formally define the
-classical Griffiths eARG as a tuple $(e, \sigma)$, where $e$
-is an ordered list of edges defining a directed acyclic graph and
-$\sigma: \mathbb{N} \rightarrow \mathbb{N}$
-is a function mapping nodes to recombination breakpoints,
-such that $\sigma(u) = x$
-if $u$ is a recombination event with breakpoint $x$ and
-$\sigma(u) = \varnothing$ otherwise.
-(This is equivalent to having different ``types''
-of node, where breakpoints are associated with recombination
-nodes only).
-Each edge $e_j = (c, p)$ describes an edge between
-child node $c$ and parent $p$, where $c, p \in \mathbb{N}$.
-% For each recombination node $u$ we must
-% have exactly two edges in $e$ such that $u$ is the child.
-This encoding of an eARG is equivalent to the output
-of programs such as
-\texttt{ARGweaver}~\citep{rasmussen2014genome}
-and \texttt{KwARG}~\citep{ignatieva2021kwarg},
-and an example is given in Fig.~\ref{fig-arg-data-structure}A.
-Note that the description above captures only the
-graph topology: if we also wish to know the branch lengths we need
-an additional function $t(u): \mathbb{N} \rightarrow \mathbb{R}$
-which defines the time of each node.
-
-The ordering of edges is a vital element of the eARG encoding.
-This is because there is a fundamental ambiguity involved
-in associating recombination breakpoints with nodes:
-we need to know from which parent the node inherited
-the segments of genome to the left and right of the breakpoint.
-The simplest way to break this symmetry is to assign
-some meaning to the order in which the parents are listed
-(\cite{griffiths1997ancestral} refer to the ``left'' and ``right''
-parents of a recombination event). This is usually described
-in terms of the process in which we extract the local trees
-along the genome from an eARG.
-To recover the tree at genome coordinate $x$,
-we traverse the graph rootwards from the leaves.
-At a particular
-node $u$, if it has one parent we are at a common ancestor
-node and we follow that parent. If we are at a
-recombination node, $u$ has two parents: if
-$x$ is less than the breakpoint $\sigma(u)$ we follow
-the edge to the first parent, and otherwise follow the edge
-to the second parent.
-This ordering requirement, while straightforward
-to describe, has some practical drawbacks. For example,
-% Surely the meaning is clear from the context here and we don't need
-% qualify the meaning of ``ARG''
-when using this representation and simulating an ARG backwards in time,
-the first event older than a recombination may involve the lineage carrying the
-ancestry to the right of the breakpoint rather
-than the left, and so we cannot emit edges as they are generated.
-Such issues can be worked around, of course, but depending on the ordering of
-otherwise indistinguishable objects is generally problematic. In practice,
-several methods explicitly associate information with the outbound edges
-of a recombination event
-to resolve the issue~\citep{lyngso2005minimum,ignatieva2021kwarg}.
+% % What do we mean by an eARG, formally?
+% To be concrete, we can formally define the
+% classical Griffiths eARG as a tuple $(e, \sigma)$, where $e$
+% is an ordered list of edges defining a directed acyclic graph and
+% $\sigma: \mathbb{N} \rightarrow \mathbb{N}$
+% is a function mapping nodes to recombination breakpoints,
+% such that $\sigma(u) = x$
+% if $u$ is a recombination event with breakpoint $x$ and
+% $\sigma(u) = \varnothing$ otherwise.
+% (This is equivalent to having different ``types''
+% of node, where breakpoints are associated with recombination
+% nodes only).
+% Each edge $e_j = (c, p)$ describes an edge between
+% child node $c$ and parent $p$, where $c, p \in \mathbb{N}$.
+% % For each recombination node $u$ we must
+% % have exactly two edges in $e$ such that $u$ is the child.
+% This encoding of an eARG is equivalent to the output
+% of programs such as
+% \texttt{ARGweaver}~\citep{rasmussen2014genome}
+% and \texttt{KwARG}~\citep{ignatieva2021kwarg},
+% and an example is given in Fig.~\ref{fig-arg-data-structure}A.
+% Note that the description above captures only the
+% graph topology: if we also wish to know the branch lengths we need
+% an additional function $t(u): \mathbb{N} \rightarrow \mathbb{R}$
+% which defines the time of each node.
+
+% In practice,
+% several methods explicitly associate information with the outbound edges
+% of a recombination event
+% to resolve the issue~\citep{lyngso2005minimum,ignatieva2021kwarg}.
 
@@
 
-The eARG encoding consists of two types of event
-(common ancestor and recombination) but nature is not so parsimonious
-and there are many other processes that will affect the
-ancestry of samples.
-For example, gene conversion between homologous
-chromosomes~\citep{wiuf2000coalescent,chen2007gene} is an important
-evolutionary force in many species~\cite[e.g.][]{yang2012great}.
-% Some refs for where people are inferring ancestry in the
-% presence of GC? Isn't this what ClonalFrame etc are all about?
-Similarly, the eARG encoding cannot easily represent multiple
-recombinations on a chromosome, or the transmission of multiple
-chromosomes.
-What we can infer about ancestral histories and
-processes should not be limited by the data structure that we
-use to represent such ancestry. The eARG is simply one way
-to \emph{encode} such history, and, as we have shown,
-one that has significant limitations of expressivity
-derived from its historical origins in the idealised
-coalescent with recombination model.
+% JK: Removing stuff about the limitations of the encoding here
+% because it's reasonably straightforward to see an extension in
+% which the number of breakpoints is in less than the number of
+% parents, and it starts muddying the waters a bit.
+
+% The eARG encoding consists of two types of event
+% (common ancestor and recombination) but nature is not so parsimonious
+% and there are many other processes that will affect the
+% ancestry of samples.
+% For example, gene conversion between homologous
+% chromosomes~\citep{wiuf2000coalescent,chen2007gene} is an important
+% evolutionary force in many species~\cite[e.g.][]{yang2012great}.
+% % Some refs for where people are inferring ancestry in the
+% % presence of GC? Isn't this what ClonalFrame etc are all about?
+% Similarly, the eARG encoding cannot easily represent multiple
+% recombinations on a chromosome, or the transmission of multiple
+% chromosomes.
+% What we can infer about ancestral histories and
+% processes should not be limited by the data structure that we
+% use to represent such ancestry. The eARG is simply one way
+% to \emph{encode} such history, and, as we have shown,
+% one that has significant limitations of expressivity
+% derived from its historical origins in the idealised
+% coalescent with recombination model.
 
@@
 
-\section*{Genome ARGs}
-As we have seen in the previous section, the focus
-on a limited palette of events makes describing currently-available
-datasets and known patterns of genetic inheritance as an eARG
-cumbersome at best.
-There is no fundamental reason that the patterns of genetic inheritance
-that we wish to describe via an ARG must be in terms
-of discrete \emph{events}, however. Recent discussions have illustrated the
-relationship between ARGs and
-the pedigree~\citep{mathieson2020ancestry,brandt2021evaluation},
-and the deep connections between the structures are
-well understood
-\citep[e.g.][]{wakeley2012genetics,gusfield2014recombinatorics,
-speed2015naturereviewsgenetics}.
-Following \cite{mathieson2020ancestry},
-rather defining the inheritance
-of genetic material in terms of \emph{events} as we do in an eARG,
-we can instead define an ARG data structure in terms of
-\emph{genomes} and intervals of genetic inheritance
-along the edges.
-To distinguish from the classical Griffiths eARG,
-we refer to this encoding of an ARG data structure
-as a ``Genome ARG'', or gARG.
-The distinction between nodes representing events
-and genomes may seem pedantic at first, but as we
-see in later sections, we gain a great deal of flexibility
-and computational efficiency by decoupling our representation
-of genetic ancestry from the events that generated it.
+% \section*{Genome ARGs}
+There is no fundamental reason, however, that the data structure we use
+to describe patterns of genetic inheritance must be in terms
+of specific causal events that generated those patterns.
+Let us define a genome as the complete set of genetic material that a child
+inherits from one parent (i.e., haploids have one genome, diploids two, etc.).
+Then, a ``genome ARG'' (gARG) is a graph in which nodes represent
+monoploid genomes  and edges represent
+genetic inheritance between an ancestor and a descendant,
+where each edge is annotated with the coordinates
+of the genomic intervals over which genetic inheritance occurs
+(the next section explores the subtleties of what we mean by
+genetic inheritance). To recover the local tree for a genomic
+position $x$, we traverse the graph rootwards from the leaves
+as in an eARG. At a particular node, we then follow the outbound
+edge associated with the genomic interval containing $x$.
+
+% Recent discussions have illustrated the
+% relationship between ARGs and
+% pedigrees~\citep{mathieson2020ancestry,brandt2021evaluation},
+% and the deep connections between the structures are
+% well understood
+% \citep[e.g.][]{wakeley2012genetics,gusfield2014recombinatorics,
+% speed2015naturereviewsgenetics}.
+Fig.~\ref{fig-arg-in-pedigree} illustrates the basic elements of a gARG,
+showing how the genomes embedded in an example (highly inbred)
+pedigree correspond to the nodes in a gARG, along with the corresponding
+local trees.
+A diploid pedigree is shown here for concreteness, but
+there is no limitation on mating system, ploidy, or age structure
+in the gARG encoding.
+In the diploid setting, a genome is inherited from one
+of the individual's parents,
+and may be the recombined product of that parent's two genomes.
+For example, individual $I_1$ in Fig.~\ref{fig-arg-in-pedigree}
+has two genomes \textsf{a} and \textsf{b},
+inherited from parents $I_3$ and $I_4$. Genome \textsf{a} is the product of
+recombining $I_3$'s two genomes \textsf{e} and \textsf{f} at position 2,
+whereas \textsf{b} was inherited directly from $I_4$'s genome \textsf{g} without
+recombination.
+Arbitrary patterns of genetic inheritance (gene conversions, multiple recombinations,
+etc) can be expressed without difficulty.
+
+% To distinguish from the classical Griffiths eARG,
+% we refer to this encoding of an ARG data structure
+% as a ``Genome ARG'', or gARG.
+% The distinction between nodes representing events
+% and genomes may seem pedantic at first, but as we
+% see in later sections, we gain a great deal of flexibility
+% and computational efficiency by decoupling our representation
+% of genetic ancestry from the events that generated it.
 
@@
 
-A gARG is a graph in which nodes represent monoploid genomes (i.e.,
-haploids have one genome, diploids two, etc.) and edges represent
-genetic inheritance between an ancestor and a descendant.
-Each edge is annotated with the coordinates
-of the genomic intervals over which genetic inheritance occurs
-(the next section explores the subtleties of what we mean by
-genetic inheritance).
-Fig.~\ref{fig-arg-in-pedigree} illustrates the basic elements of a gARG,
-showing how the genomes embedded in an example (highly inbred)
-pedigree correspond to the nodes in a gARG, along with the corresponding
-local trees.
-A diploid pedigree is shown here for concreteness, but
-there is no limitation on mating system, ploidy, or age structure
-in the gARG encoding.
-In the diploid setting, a genome is inherited from one
-of the individual's parents,
-and may be the recombined product of that parent's two genomes.
-For example, individual $I_1$ in Fig.~\ref{fig-arg-in-pedigree}
-has two genomes \textsf{a} and \textsf{b},
-inherited from parents $I_3$ and $I_4$. Genome \textsf{a} is the product of
-recombining $I_3$'s two genomes \textsf{e} and \textsf{f} at position 2,
-whereas \textsf{b} was inherited directly from $I_4$'s genome \textsf{g} without
-recombination.
-Arbitrary patterns of genetic inheritance (gene conversions, multiple recombinations,
-etc) can be expressed without difficulty.
-
-Formally, we can define a gARG as a set of edges $E$, where each
-element of the set is a tuple $(c, p, I)$ such that $c, p \in \mathbb{N}$,
-$c$ is the child node, $p$ is the parent node and $I$ is the set of
-disjoint genomic intervals $(\ell, r]$ over which genome $c$ inherits from $p$ (we may
-also phrase this in terms of the ancestral material of a sample---see
-the next section).
-Deriving the local tree at a point $x$
-follows largely the same pattern as for an eARG but is somewhat
-simpler because coordinate information is directly associated with
-edges. Starting with the sample nodes $S$ we traverse
-rootwards through the graph from child to parent. At each node $u$, we find an
-edge $(c, p, I) \in E$ such that $u = c$ and $x \in I$
-and move to the next node $p$.
+% Formally, we can define a gARG as a set of edges $E$, where each
+% element of the set is a tuple $(c, p, I)$ such that $c, p \in \mathbb{N}$,
+% $c$ is the child node, $p$ is the parent node and $I$ is the set of
+% disjoint genomic intervals $(\ell, r]$ over which genome $c$ inherits from $p$ (we may
+% also phrase this in terms of the ancestral material of a sample---see
+% the next section).
+% Deriving the local tree at a point $x$
+% follows largely the same pattern as for an eARG but is somewhat
+% simpler because coordinate information is directly associated with
+% edges. Starting with the sample nodes $S$ we traverse
+% rootwards through the graph from child to parent. At each node $u$, we find an
+% edge $(c, p, I) \in E$ such that $u = c$ and $x \in I$
+% and move to the next node $p$.
 
@@
 % from the others. Is there a more direct way we can address these points ?
-There are some interesting and important consequences to this focus on genomes
-rather than events.
-Firstly, a gARG may contain nodes that are both samples \emph{and}
-internal nodes in the local trees.
-This ability is useful when we have pedigree information
-along with genetic data
-\cite[e.g.][]{hayes20191000,RosFreixedes2022,anderson2022genes},
-where we may have many generations of internal sample nodes
-whose genomes have been sequenced.
-Another situation in which it is important that
-we do not assume samples are all contemporary ``leaf'' nodes
-(or that all internal nodes are either coalescences or recombinations)
-is when we are incorporating ancient genomes into ARG
-inference~\citep{speidel2021inferring,wohns2022unified}.
-Similarly, a gARG may contain nodes that
-do not correspond to any particular event in the ancestry of the samples.
-For example, node \textsf{h} is the direct ancestor of \textsf{d} in
-Fig~\ref{fig-arg-in-pedigree}, and is not the product of either a
-coalescence or a recombination. Such nodes would usually be removed
-from the graph so that \textsf{d} descends directly from \textsf{k} (see
-the ``\nameref{ARG_simplification}'' section) but there is no necessity for this from
-a representational perspective.
-Secondly, because we do not classify nodes by event type
-(common ancestor or recombination) there is no limitation
-on the complexity that can be modelled---multiple recombinations
-and coalescences can happen simultaneously in the same genome,
-a common occurance in large deeply sequenced
-pedigrees~\citep{hayes20191000,RosFreixedes2022}.
-Finally, there is a subtle and important point about how recombination
-is represented: which genome in the organismal lifecycle
-do nodes represent?
-In principle, we can interpret nodes as representing genomes
-at any point in the life cycle [supplementary Figure ?]. Hudson
-views nodes as representing gametes~\citep{hudson1983properties}.
-However, we argue that the most intuitive
-approach is to let nodes represent the genomes present in diploid
-individuals in each generation, as in
-Fig~\ref{fig-arg-in-pedigree}A. In this case recombination
-is not represented by a single ``recombination node'' (as in an eARG), but with
-two parent genomes between which recombination takes place
-(e.g.\ nodes \textsf{e} and \textsf{f}), and a
-single resulting ``recombinant'' genome (e.g.\ \textsf{c}). The
-edges that link this recombinant with its two parents carry the information
-concerning breakpoint location that would be associated with a
-recombination node in an eARG. [TODO finish up here with more
-clarity on what this means in terms of the required specifity
-of recombination events, and put in a forward ref to the simplification
-section where we talk about stacking up recombination events
-and their identifiability.]
-
+% There are some interesting and important consequences to this focus on genomes
+% rather than events.
+
+% Firstly, a gARG may contain nodes that are both samples \emph{and}
+% internal nodes in the local trees.
+% This ability is useful when we have pedigree information
+% along with genetic data
+% \cite[e.g.][]{hayes20191000,RosFreixedes2022,anderson2022genes},
+% where we may have many generations of internal sample nodes
+% whose genomes have been sequenced.
+% Another situation in which it is important that
+% we do not assume samples are all contemporary ``leaf'' nodes
+% (or that all internal nodes are either coalescences or recombinations)
+% is when we are incorporating ancient genomes into ARG
+% inference~\citep{speidel2021inferring,wohns2022unified}.
+% Similarly, a gARG may contain nodes that
+% do not correspond to any particular event in the ancestry of the samples.
+% For example, node \textsf{h} is the direct ancestor of \textsf{d} in
+% Fig~\ref{fig-arg-in-pedigree}, and is not the product of either a
+% coalescence or a recombination. Such nodes would usually be removed
+% from the graph so that \textsf{d} descends directly from \textsf{k} (see
+% the ``\nameref{ARG_simplification}'' section) but there is no necessity for this from
+% a representational perspective.
+% Secondly, because we do not classify nodes by event type
+% (common ancestor or recombination) there is no limitation
+% on the complexity that can be modelled---multiple recombinations
+% and coalescences can happen simultaneously in the same genome,
+% a common occurance in large deeply sequenced
+% pedigrees~\citep{hayes20191000,RosFreixedes2022}.
+% Finally, there is a subtle and important point about how recombination
+% is represented: which genome in the organismal lifecycle
+% do nodes represent?
+% In principle, we can interpret nodes as representing genomes
+% at any point in the life cycle [supplementary Figure ?]. Hudson
+% views nodes as representing gametes~\citep{hudson1983properties}.
+% However, we argue that the most intuitive
+% approach is to let nodes represent the genomes present in diploid
+% individuals in each generation, as in
+% Fig~\ref{fig-arg-in-pedigree}A. In this case recombination
+% is not represented by a single ``recombination node'' (as in an eARG), but with
+% two parent genomes between which recombination takes place
+% (e.g.\ nodes \textsf{e} and \textsf{f}), and a
+% single resulting ``recombinant'' genome (e.g.\ \textsf{c}). The
+% edges that link this recombinant with its two parents carry the information
+% concerning breakpoint location that would be associated with a
+% recombination node in an eARG. [TODO finish up here with more
+% clarity on what this means in terms of the required specifity
+% of recombination events, and put in a forward ref to the simplification
+% section where we talk about stacking up recombination events
+% and their identifiability.]
+
+The eARG and gARG definitions provided here are apparently very
+similar, but, as we shall see, there are many advantages to decoupling
+our representation of patterns of genetic inheritance
+from the events that generated them.
 For simplicity, throughout the rest of the paper we will contrast the
-classical eARG (as defined in the XXX section) with the gARG definition
-here. However, it is important to note that many of the benefits of the
+classical eARG with the gARG definition here.
+However, it is important to note that many of the benefits of the
 gARG encoding that we discuss are due to the encoding of inheritance via
 edges rather than nodes.
-Whether one regards nodes as genomes or events is orthogonal,
-and something of a philosophical question which makes little
-difference from a computational perspective.
+Whether one regards nodes as genomes or events is essentially
+a matter of interpretation, and unimportant for many purposes.
 
@@
 The edges in a gARG represent a path of genetic inheritance from
-parent to child (and ultimately from cell to cell) through some
-number of generations. The edges are ``annotated''
+ancestor to descendant through some
+number of generations (ultimately from cell to cell). The edges are ``annotated''
 with the genome coordinates over which inheritance occurs. There are

### 2022-09-14 paper.tex
Refine

@@
 to describe patterns of genetic inheritance must be in terms
-of specific causal events that generated those patterns.
-Let us define a genome as the complete set of genetic material that a child
+of specific causal events. As we demonstrate in subsequent sections,
+there are significant advantages to using
+a slightly different formulation, which
+we call a ``genome ARG'' (gARG) to distinguish it from the classical
+eARG described above.
+First, let us define a genome as the complete set of genetic material that a child
 inherits from one parent (i.e., haploids have one genome, diploids two, etc.).
-Then, a ``genome ARG'' (gARG) is a graph in which nodes represent
+Then, a gARG is a graph in which nodes represent
 monoploid genomes  and edges represent
@@
 position $x$, we traverse the graph rootwards from the leaves
-as in an eARG. At a particular node, we then follow the outbound
+(as in an eARG) and at a particular node, we follow the outbound
 edge associated with the genomic interval containing $x$.
-
 % Recent discussions have illustrated the
@@
 % speed2015naturereviewsgenetics}.
-Fig.~\ref{fig-arg-in-pedigree} illustrates the basic elements of a gARG,
-showing how the genomes embedded in an example (highly inbred)
+Fig.~\ref{fig-arg-in-pedigree}
+shows how the genomes embedded in an example (highly inbred)
 pedigree correspond to the nodes in a gARG, along with the corresponding
 local trees.
-A diploid pedigree is shown here for concreteness, but
-there is no limitation on mating system, ploidy, or age structure
-in the gARG encoding.
-In the diploid setting, a genome is inherited from one
+In this diploid setting
+(there is no limitation on mating system, ploidy, or age structure
+in the gARG encoding),
+a genome is inherited from one
 of the individual's parents,
@@
 
-The eARG and gARG definitions provided here are apparently very
-similar, but, as we shall see, there are many advantages to decoupling
-our representation of patterns of genetic inheritance
-from the events that generated them.
-For simplicity, throughout the rest of the paper we will contrast the
-classical eARG with the gARG definition here.
-However, it is important to note that many of the benefits of the
-gARG encoding that we discuss are due to the encoding of inheritance via
-edges rather than nodes.
-Whether one regards nodes as genomes or events is essentially
-a matter of interpretation, and unimportant for many purposes.
+% The reader may arrive at the end of this section going, er, so what?
+% Try to tackle this head on by getting to the heart of the distinction.
+The difference between eARGs and gARGs are subtle, and whether
+we choose to interpret nodes as events or genomes
+is often of little practical importance.
+However, as we show in the remainder of this article,
+there are substantial practical gains
+from using the gARG encoding, allowing differential
+levels of information about the recombination process to be
+represented in a computionally efficient way.
+The key to this is not in how we interpret nodes, but in how
+we represent the process of genetic inheritance.
+The critical distinction is that
+the gARG encoding allows us to associate
+genetic inheritance in non-contiguous chunks of genome
+between an ancestor and a descendant.
 

### 2022-09-14 paper.tex

@@
 a slightly different formulation, which
-we call a ``genome ARG'' (gARG) to distinguish it from the classical
-eARG described above.
+we call a ``genome ARG'' (gARG).
 First, let us define a genome as the complete set of genetic material that a child
@@
 etc) can be expressed without difficulty.
-
-% To distinguish from the classical Griffiths eARG,
-% we refer to this encoding of an ARG data structure
-% as a ``Genome ARG'', or gARG.
-% The distinction between nodes representing events
-% and genomes may seem pedantic at first, but as we
-% see in later sections, we gain a great deal of flexibility
-% and computational efficiency by decoupling our representation
-% of genetic ancestry from the events that generated it.
 
@@
 represented in a computionally efficient way.
-The key to this is not in how we interpret nodes, but in how
+The key to this is not so much in how we interpret nodes
+(see Appendix XXX), but specifically in how
 we represent the process of genetic inheritance.
 The critical distinction is that
-the gARG encoding allows us to associate
-genetic inheritance in non-contiguous chunks of genome
-between an ancestor and a descendant.
+the gARG encoding allows us to directly describe
+genetic inheritance in multiple non-contiguous genome intervals
+between an ancestor and a descendant, something that
+cannot be done by storing recombination breakpoints.
 

### 2022-09-21 paper.tex
Remove description of W&H node u as a MRCA
I think the only common ancestors that are present in the simplified TS will be o and v
Some trivial edits and comments

@@
 % Mathematical treatment of the CwR and ARGs (and ARG-like stochastic processes)
-An extensive body of work has developed in
+An extensive body of work has been developed from
 studying the coalescent with recombination
@@
 wilton2015smc,griffiths2016coalescent},
-as well as, more commonly, the concrete realisations of the
-ancestry~\citep[e.g.][]{gusfield2014recombinatorics,mathieson2020ancestry,brandt2021evaluation}.
+as well as, more commonly, the concrete realisation of ancestry from a
+process~\citep[e.g.][]{gusfield2014recombinatorics,mathieson2020ancestry,brandt2021evaluation}.
 \cite{minichiello2006mapping} explicitly distinguished the ARG as a
@@
 % one that has significant limitations of expressivity
-% derived from its historical origins in the idealised
-% coalescent with recombination model.
+% derived from its historical origins in the idealised model of
+% coalescent with recombination.
 
@@
 Arbitrary patterns of genetic inheritance (gene conversions, multiple recombinations,
-etc) can be expressed without difficulty.
+etc) can be expressed without difficulty using this encoding.
 
@@
 we choose to interpret nodes as events or genomes
-is often of little practical importance.
+might seem of little practical importance.
 However, as we show in the remainder of this article,
@@
 common ancestor separately
-(in nodes \textsf{o}, \textsf{u}  and \textsf{v}).
+(in nodes \textsf{o} and \textsf{v}).
 
@@
 parent to child vs the propagation of genomic intervals ancestral to the sample
-are analogous to running forwards-time vs backwards-time simulations.
+are analogous to running forwards-in-time vs backwards-in-time simulations.
 Indeed, the process of taking a graph with edges annotated with direct parent-child
@@
 ancestral to the sample are retained (Fig~\ref{fig-ancestry-resolution}B)
-is the same key algorithm underlying recent advances in forwards-time simulation.
-\cite{kelleher2018efficient} developed a method in which forwards-time
+is the same key algorithm underlying recent advances in forwards-in-time simulation.
+\cite{kelleher2018efficient} developed a method in which forwards-in-time
 simulations are performed by recording all parent-child genetic relationships
@@
 to the ancestry of the current population. In many situations, recording
-the full genetic ancestry of the population in this way is orders of magnitude faster
+the full genetic ancestry of the current population in this way is orders of magnitude faster
 than the alternative of tracking
@@
 \section*{Simplification}\label{ARG_simplification}
+% I think a sentence is needed here on why "so much" text is devoted to simplification.
+% How does simplification connect to eARG vs gARG and practical benefits of gARG over eARG
+% and why do you now talk about unary nodes and a bit polytomies? Hold my hand here please;)
 ARG simplification~\citep{kelleher2018efficient} is the process
@@
 presence of ``unary'' nodes: a tree node that has exactly
+% should presence above be absence?
 one child. Fig.~\ref{fig-simplification} show a series of successive steps of
 simplification, starting with a complete gARG simulated under a
-(backwards-time) Wright-Fisher
+(backwards-in-time) Wright-Fisher
 model, shown both as a graph on the left, and the equivalent set of 6
@@
 An extreme sign of these breeding practices is that there are only two ancestral
-Y-chromosome linages present in today's US Holstein dairy breed~\citep{Yue2015}.
+Y-chromosome lineages present in today's US Holstein dairy breed~\citep{Yue2015}.
 Simlarly, the 1000 bull genomes project~\citep{hayes20191000}

### 2022-09-28 argutils/ancestry.py
Get the burgers working

@@ -309,6 +309,46 @@ def remove_unused_nodes(ts):
     return ts.subset(used_nodes)
 
 
+def remove_edges_above_local_roots(ts):
+    # use simplify() to find the roots
+    ts_simp, node_map = ts.simplify(map_nodes=True)
+    inverted_node_map = {n: i for i, n in enumerate(node_map)}
+    edges_to_prune = {}
+    for intvl, orig_tree, tree in ts.coiterate(ts_simp):
+        for root in tree.roots:
+            orig_node = inverted_node_map[root]
+            if orig_tree.parent(orig_node) != tskit.NULL:
+                edge_above = orig_tree.edge(orig_node)
+                if edge_above in edges_to_prune:
+                    intervals = edges_to_prune[edge_above]
+                    if intervals[-1].right == intvl.left:
+                        intervals[-1] = tskit.Interval(intervals[-1].left, intvl.right)
+                    else:
+                        intervals.append(intvl)
+                else:
+                    edges_to_prune[edge_above] = [intvl]
+    # Now split the edges to prune by their intervals
+    tables = ts.dump_tables()
+    edges = tables.edges
+    for edge_id, intervals in edges_to_prune.items():
+        deleted_edge = tables.edges[edge_id]
+        if deleted_edge.left < intervals[0].left:
+            edges.append(deleted_edge.replace(right=intervals[0].left))
+        for i in range(len(intervals)):
+            try:
+                edges.append(deleted_edge.replace(
+                    left=intervals[i].right, right=intervals[i+1].left))
+            except IndexError:
+                # last interval to delete
+                if deleted_edge.right > intervals[i].right:
+                    edges.append(deleted_edge.replace(left=intervals[i].right))
+    keep = np.ones(edges.num_rows, dtype=bool)
+    keep[list(edges_to_prune.keys())] = False
+    tables.edges.replace_with(edges[keep])
+    tables.sort()
+    return tables.tree_sequence()        
+            
+   
 def simplify_keeping_all_nodes(ts):
     """
     Run the Hudson algorithm to convert from an implicit to an explicit edge encoding,
@@ -573,7 +613,7 @@ def wh99_example(one_node_recombination=None):
         nonlocal t
         t += 1
         if one_node_recombination:
-            parent = nodes.add_row(time=t)
+            parent = nodes.add_row(time=t, flags=NODE_IS_RECOMB)
             try:
                 child, left, right = child
             except TypeError:

### 2022-09-28 argutils/viz.py

@@ -293,8 +293,8 @@ def convert_nx(ts):
     edges = collections.defaultdict(list)
     for edge in ts.edges():
         edges[(edge.child, edge.parent)].append((edge.left, edge.right))
-    for node in ts.nodes():
-        G.add_node(node.id, time=node.time, flags=node.flags)
+    for n in ts.nodes():
+        G.add_node(n.id, time=n.time, flags=n.flags, label=n.metadata.get("name", None))
     for edge, intervals in edges.items():
         G.add_edge(*edge, intervals=intervals)
     return G

### 2022-09-28 illustrations.py

@@ -27,99 +27,139 @@ def ancestry_resolution():
     """
     Ancestry resolution on the WH99 graph.
     """
+    def plot_ellipse(ts, ax, xa, ya, burger_segments=None, trim_left=0):
+        # trim_left is a hack to not show stuff above the root
+        icon_size1 = (ax.get_xlim()[1] - ax.get_xlim()[0]) * 0.00029
+        icon_center1 = icon_size1 / 2.0
+        icon_size2 = icon_size1 * 0.9
+        icon_center2 = icon_size2 / 2.0
+        a = plt.axes([xa - icon_center1 - 0.005/2, ya - icon_center2+0.005/2, icon_size1 + 0.005, icon_size2-0.005])
+        icon = patches.Ellipse(
+            (0.5, 0.5),
+            0.96,
+            0.96,
+            linewidth=2,
+            edgecolor='k',
+            facecolor='white',
+        )
+        a.add_patch(icon)
+        a.axis("off")
+        if burger_segments is not None:
+            icon_size1 = icon_size1 * 0.83
+            icon_center1 = icon_size1 / 2.0
+            icon_size2 = icon_size1 * 0.4
+            icon_center2 = icon_size2 / 2.0
+            a2 = plt.axes([xa - icon_center1 - 0.005/2, ya - icon_center2+0.005/2, icon_size1 + 0.005, icon_size2-0.005])
+            for segment in burger_segments:
+                icon = patches.Rectangle(
+                    (segment[0].left/ts.sequence_length, 0),
+                    segment[0].right/ts.sequence_length,
+                    1,
+                    facecolor=str(1-segment[1]/ts.num_samples),
+                    )
+                a2.add_patch(icon)
+            a2.add_patch(patches.Rectangle(
+                (trim_left, 0),
+                1 - trim_left,
+                1,
+                edgecolor="k",
+                facecolor='none',
+                linewidth=2))
+            a2.axis("off")
+            
+        return a
+
+    def plot_rect(ax, xa, ya, scale=(1, 1), color="k"):
+        icon_size1 = (ax.get_xlim()[1] - ax.get_xlim()[0]) * 0.0002 * scale[0]
+        icon_center1 = icon_size1 / 2.0
+        icon_size2 = icon_size1 * 1.3 * scale[1]
+        icon_center2 = icon_size2 / 2.0
+        a = plt.axes([xa - icon_center1, ya - icon_center2, icon_size1, icon_size2])
[…]

### 2022-09-29 paper.tex

@@
 number of generations (ultimately from cell to cell). The edges are ``annotated''
-with the genome coordinates over which inheritance occurs. There are
-two ways in which we can interpret these annotations, and the
+with the genome coordinates over which inheritance occurs.
+Fig~\ref{fig-ancestry-resolution} is an instructive example
+from the literature \citep{wiuf1999recombination} and illustrates
+how a classical eARG (Fig~\ref{fig-ancestry-resolution}A) can be described as
+a gARG with edge annotations (Figs~\ref{fig-ancestry-resolution}B~and~C).
+
+There are two distinct annotation schemes possible, and the
 distinction is subtle and important.
-The first interpretation is to regard the annotations as describing
+The first scheme is to use the annotations to describe
 the complete set of genomic intervals inherited by the descendant node from the
-ancestor node. The second is to view the edge annotations as representing
+ancestor node (Fig~\ref{fig-ancestry-resolution}B). The second is to use the edge annotations to describe
 the ``ancestral material''~\citep{wiuf1999ancestry,wiuf1999recombination}
 of the sample that is propagated through
-the graph from descendant to ancestor.
-These alternatives are illustrated in Fig~\ref{fig-ancestry-resolution}.
+the graph from descendant to ancestor (Fig~\ref{fig-ancestry-resolution}C).
 
@@
 \caption{\label{fig-ancestry-resolution}
-The \citet[][Fig.~1]{wiuf1999recombination} example graph as a gARG,
-with contrasting
-edge annotation schemes. (A) Immediate edge annotation, reflecting ``local''
-transmission of all genetic material between parent and child. All edges span
-the full genome from position 0 to 7, apart from those linking a recombinant
-and its two parents (bold). (B) Sample-resolved edge annotation: the
-Hudson-like simplification algorithm has been used to shorten edge spans to
+The \citet[][Fig.~1]{wiuf1999recombination} example graph as an eARG and
+two gARGs with alternative edge annotation schemes.
+(A) A conventional Griffiths eARG. Square nodes represent events, with
+recombinations (in red) annotated with the position of a single breakpoint
+along the 7-unit long chromosome.
+(B) A gARG with direct-inheritance edge annotations. Oval nodes represent
+genomes and annotations reflect ``local'' transmission of all genetic material
+between parent and child. Edges that do not span the full genome are visually
+highlighted, showing how breakpoint positions can be encoded.
+(C) A gARG with sample-resolved edge annotations. The Hudson-like simplification
+algorithm has been used to shorten edge spans to
 reflect only the upwards transmission of genetic material from the samples
@@
 node is shaded as in Fig.~\ref{fig-arg-in-pedigree}A to show coalescing
-regions of genome back to the grand MRCA \textsf{w}. Line widths reflect
-the genomic span of each edge, which helps visualise the relative importance of
-different transmission pathways.
+regions of genome back to the MRCA nodes \textsf{k} (regions 0 to 2) and \textsf{p}
+(regions 2 to 7). Line widths reflect the genomic span of each edge, highlighting
+the relative importance of different transmission pathways.
 }
@@
 as we discuss in section XXX, storing sample-resolved ancestry
-intervals (Fig~\ref{fig-ancestry-resolution}B) is generally much more
+intervals (Fig~\ref{fig-ancestry-resolution}C) is generally much more
 useful than storing the local parent-child intervals of inheritance
-(Fig~\ref{fig-ancestry-resolution}A).
+(Fig~\ref{fig-ancestry-resolution}B).
 Indeed, storing local parent-child
@@
 its direct correspondence with the classical approach of associating
-breakpoints with recombination nodes in an eARG. The eARG encoding
-is simply a more limited form, in which the only types of direct
+breakpoints with recombination nodes in an eARG (Fig~\ref{fig-ancestry-resolution}A).
+The eARG encoding is simply a more limited form, in which the only types of direct
 inheritance we can have are the entire genome, or the intervals
@@
 
-Fig~\ref{fig-ancestry-resolution}A is an instructive example
-from the literature \citep{wiuf1999recombination} and illustrates
-how a classical eARG can be described as a gARG with local edge annotations
-(directly corresponding to the eARG description), and how we can
-then derive the ancestry resolved edge annotations.
-It is important to note, however, that the example is a sample from
-the so-called ``big'' ARG \emph{process} (see Appendix XXX)
-and has several features that we are unlikely to simulate or infer in practice.
-For example, by definition the samples contain no genetic information about
-node \textsf{j} and so it is unlikey to be inferred.
+Note that this example is a sample from the so-called \emph{``big'' ARG process}
+(see Appendix XXX) and has several features that we are unlikely to simulate or
+infer in practice. For example, by definition the samples contain no genetic information about
+nodes \textsf{g} and nodes \textsf{j} and so neither of those nodes are likely to be inferred.
 Likewise, is is unlikely we would have information to infer the ``grand MRCA''
-\textsf{w} as all regions of the genome have already reached a
-common ancestor separately
-(in nodes \textsf{o} and \textsf{v}).
+\textsf{q} as all regions of the genome have already reached a
+common ancestor separately (in nodes \textsf{k} and \textsf{p}).
 
@@
 Given the local inheritance edge annotations
-(Fig~\ref{fig-ancestry-resolution}A),
+(Fig~\ref{fig-ancestry-resolution}B),
  we obtain the sample-resolved edge annotation
-(Fig~\ref{fig-ancestry-resolution}B) by traversing the graph backwards in time
+(Fig~\ref{fig-ancestry-resolution}C) by traversing the graph backwards in time
 from sample nodes (\textsf{a}, \textsf{b}, \textsf{c}).
@@
 union of the ancestral intervals on its inbound edges. We then annotate
-the outbound edges with the intersection of their local inheritance and
-just-computed ancestral intervals for the node.
-For example, node \textsf{d} in Fig~\ref{fig-ancestry-resolution}
-carries ancestral material on the interval $(0, 4]$, since it
-is the left-hand parent of the recombinant \textsf{b}. The edge
-joining \textsf{d} to its ancestor \textsf{h} then carries ancestral
-material on $(0, 4]$ since although \textsf{d} inherits the entire interval
-$(0, 7]$ from \textsf{h}, only $(0, 4]$ is ancestral to the sample.
-This process is directly analogous to Hudson's simulation algorithm
+the outbound edges with the intersection of their local inheritance intervals
+and the just-computed ancestral intervals for the node.
+For example, node \textsf{i} in Fig~\ref{fig-ancestry-resolution}C
+carries ancestral material on the intervals $(0, 2]$ and $(4, 7]$. The edge
+joining \textsf{i} to its ancestor \textsf{k} then carries ancestral
+material on $(0, 2]$ since although \textsf{i} directly inherits interval
+$(0, 3]$ from \textsf{k} (Fig~\ref{fig-ancestry-resolution}B), only $(0, 2]$ is
+ancestral to the sample. This process is directly analogous to Hudson's
+simulation algorithm
 (Appendix XXX).
@@
 Indeed, the process of taking a graph with edges annotated with direct parent-child
-inheritance (Fig~\ref{fig-ancestry-resolution}A) and producing the equivalent
+inheritance (Fig~\ref{fig-ancestry-resolution}B) and producing the equivalent
 ``ancestry resolved'' graph in which only those genomic intervals that are
-ancestral to the sample are retained (Fig~\ref{fig-ancestry-resolution}B)
+ancestral to the sample are retained (Fig~\ref{fig-ancestry-resolution}C)
 is the same key algorithm underlying recent advances in forwards-in-time simulation.
@@
 this structure to retain only ancestral relationships that are relevant
-to the ancestry of the current population. In many situations, recording
-the full genetic ancestry of the current population in this way is orders of magnitude faster
+to the ancestry of the current population. In many situations, using this approach to
+record the full genetic ancestry of the current population is orders of magnitude faster
 than the alternative of tracking

### 2022-09-29 paper.tex

@@
 
-There are two distinct annotation schemes possible, and the
-distinction is subtle and important.
-The first scheme is to use the annotations to describe
-the complete set of genomic intervals inherited by the descendant node from the
-ancestor node (Fig~\ref{fig-ancestry-resolution}B). The second is to use the edge annotations to describe
-the ``ancestral material''~\citep{wiuf1999ancestry,wiuf1999recombination}
+There are two distinct annotation schemes possible, and the distinction is subtle
+and important. The first scheme is to use the annotations to describe the complete
+set of genomic intervals directly inherited by the descendant node from the ancestor
+node (Fig~\ref{fig-ancestry-resolution}B). The second is to use the edge annotations to
+describe the ``ancestral material''~\citep{wiuf1999ancestry,wiuf1999recombination}
 of the sample that is propagated through
@@
 intervals (Fig~\ref{fig-ancestry-resolution}C) is generally much more
-useful than storing the local parent-child intervals of inheritance
+useful than storing the parent-child intervals recording direct inheritance
 (Fig~\ref{fig-ancestry-resolution}B).
-Indeed, storing local parent-child
-inheritance intervals for edges is really only of interest because of
-its direct correspondence with the classical approach of associating
+Indeed, storing direct inheritance intervals for edges is really only of interest because of
+its correspondence with the classical approach of associating
 breakpoints with recombination nodes in an eARG (Fig~\ref{fig-ancestry-resolution}A).
@@
 
-Given the local inheritance edge annotations
+Given the direct-inheritance edge annotations
 (Fig~\ref{fig-ancestry-resolution}B),
@@
 Proceeding in topological order (such that an ancestor node is visited
-after all of its descendants) the ancestral material for a node is the
-union of the ancestral intervals on its inbound edges. We then annotate
-the outbound edges with the intersection of their local inheritance intervals
-and the just-computed ancestral intervals for the node.
+after all of its descendants) the sample-resolved material for a node is the
+union of the sample-resolved intervals on its inbound edges. We then annotate
+the outbound edges with the intersection of their direct-inheritance intervals
+and the just-computed sample-resolved intervals for the node.
 For example, node \textsf{i} in Fig~\ref{fig-ancestry-resolution}C
-carries ancestral material on the intervals $(0, 2]$ and $(4, 7]$. The edge
-joining \textsf{i} to its ancestor \textsf{k} then carries ancestral
+carries sample-resolved material on the intervals $(0, 2]$ and $(4, 7]$. The edge
+joining \textsf{i} to its ancestor \textsf{k} then carries
 material on $(0, 2]$ since although \textsf{i} directly inherits interval

### 2022-09-29 paper.tex
Rejig and condense text about "big" ARGs etc
I have tried to make the link between sample-resolving big ARGs, and (more importantly) forwards simulations, more explicit, as well as giving a tee-up into the simplification section.

@@
 the graph from descendant to ancestor (Fig~\ref{fig-ancestry-resolution}C).
+As we discuss in section XXX, this second scheme has many advantages
+in terms of efficiency.
 
@@
 (A) A conventional Griffiths eARG. Square nodes represent events, with
-recombinations (in red) annotated with the position of a single breakpoint
-along the 7-unit long chromosome.
+each recombination (red) annotated with the position of a single breakpoint
+along the 7-unit-long chromosome.
 (B) A gARG with direct-inheritance edge annotations. Oval nodes represent
@@
 reflect only the upwards transmission of genetic material from the samples
-\textsf{a}, \textsf{b}, and \textsf{c}. The depiction of the genome within each
+\textsf{a}, \textsf{b}, and \textsf{c}; very few edges then span the entire genome.
+The depiction of the genome within each
 node is shaded as in Fig.~\ref{fig-arg-in-pedigree}A to show coalescing
@@
 (regions 2 to 7). Line widths reflect the genomic span of each edge, highlighting
-the relative importance of different transmission pathways.
+the relative importance of different transmission pathways to the samples.
 }
@@
 
-ARGs are inherently a retrospective structure defined in terms
-of a set of sample nodes (usually the leaves), and,
-as we discuss in section XXX, storing sample-resolved ancestry
-intervals (Fig~\ref{fig-ancestry-resolution}C) is generally much more
-useful than storing the parent-child intervals recording direct inheritance
-(Fig~\ref{fig-ancestry-resolution}B).
-Indeed, storing direct inheritance intervals for edges is really only of interest because of
-its correspondence with the classical approach of associating
-breakpoints with recombination nodes in an eARG (Fig~\ref{fig-ancestry-resolution}A).
-The eARG encoding is simply a more limited form, in which the only types of direct
-inheritance we can have are the entire genome, or the intervals
-at either side of a recombination breakpoint.
-
-Note that this example is a sample from the so-called \emph{``big'' ARG process}
-(see Appendix XXX) and has several features that we are unlikely to simulate or
-infer in practice. For example, by definition the samples contain no genetic information about
-nodes \textsf{g} and nodes \textsf{j} and so neither of those nodes are likely to be inferred.
-Likewise, is is unlikely we would have information to infer the ``grand MRCA''
-\textsf{q} as all regions of the genome have already reached a
-common ancestor separately (in nodes \textsf{k} and \textsf{p}).
+%ARGs are inherently a retrospective structure defined in terms
+%of a set of sample nodes (usually the leaves), and,
+%as we discuss in section XXX, storing sample-resolved ancestry
+%intervals (Fig~\ref{fig-ancestry-resolution}C) is generally much more
+%useful than storing the parent-child intervals recording direct inheritance
+%(Fig~\ref{fig-ancestry-resolution}B).
+%Indeed, storing direct inheritance intervals for edges is really only of interest because of
+%its correspondence with the classical approach of associating
+%breakpoints with recombination nodes in an eARG (Fig~\ref{fig-ancestry-resolution}A).
+%The eARG encoding is simply a more limited form, in which the only types of direct
+%inheritance we can have are the entire genome, or the intervals
+%at either side of a recombination breakpoint.
+
+%Note that this example is a sample from the so-called \emph{``big'' ARG process}
+%(see Appendix XXX) and has several features that we are unlikely to simulate or
+%infer in practice. For example, by definition the samples contain no genetic information about
+%nodes \textsf{g} and nodes \textsf{j} and so neither of those nodes are likely to be inferred.
+%Likewise, is is unlikely we would have information to infer the ``grand MRCA''
+%\textsf{q} as all regions of the genome have already reached a
+%common ancestor separately (in nodes \textsf{k} and \textsf{p}).
 
@@
 $(0, 3]$ from \textsf{k} (Fig~\ref{fig-ancestry-resolution}B), only $(0, 2]$ is
-ancestral to the sample. This process is directly analogous to Hudson's
-simulation algorithm
+ancestral to the sample. For each node, we also keep track of the number of
+descendant samples over each genomic interval, and stop tracking that
+interval once the lineages from all samples have coalesced. For example,
+the edge from \textsf{k} to \textsf{n} omits the region $(0,2]$ because it has
+already fully coalesced (indicated by a fully black genomic region).
+This process is directly analogous to Hudson's simulation algorithm
 (Appendix XXX).
 
-These alternatives of recording the direct inheritance of genetic material from
-parent to child vs the propagation of genomic intervals ancestral to the sample
-are analogous to running forwards-in-time vs backwards-in-time simulations.
-Indeed, the process of taking a graph with edges annotated with direct parent-child
-inheritance (Fig~\ref{fig-ancestry-resolution}B) and producing the equivalent
-``ancestry resolved'' graph in which only those genomic intervals that are
-ancestral to the sample are retained (Fig~\ref{fig-ancestry-resolution}C)
-is the same key algorithm underlying recent advances in forwards-in-time simulation.
+In practice, the backwards-in-time Griffiths process that produces
+ARGs such as that shown in Figs~{fig-ancestry-resolution}A~and~B
+is rarely simulated, because the resulting ``big ARG'' (Appendix XXX) is
+dominated by nodes such as \textsf{g} and \textsf{j} which contain information
+about inheritance in non-ancestral material, as well as nodes such as
+\textsf{o} and \textsf{q} which track details of ancestry even once the
+local genomic region has fully coalesced. Instead, simulating
+ARGs from the coalescent with recombination is usually done by
+running Hudson's algorithm to produce a sample-resolved ARG directly.
+Moreover, when inferring ARGs from genetic data, there is no information
+in the sample set that can be used to resolve these nodes, so that they
+are essentially unknowable.
+
+Nevertheless, there is a practical use for converting a direct-inheritance
+gARG to a sample resolved version. It is the key algorithm underlying recent
+advances in forwards-in-time simulation.
 \cite{kelleher2018efficient} developed a method in which forwards-in-time
@@
 record the full genetic ancestry of the current population is orders of magnitude faster
-than the alternative of tracking
-sequences~\citep{kelleher2018efficient,haller2018tree}.
+than the alternative of tracking neutral genetic variation directly~\citep{kelleher2018efficient,haller2018tree}.
+
+It is worth noting that nodes \textsf{g},  \textsf{j}, \textsf{o} and \textsf{q}
+in Fig~\ref{fig-ancestry-resolution}C are redundant in that they can
+be removed without reducing our knowledge of the genetic
+ancestry of the samples. Removing nodes,
+both redundant ones and others, is part of a process we call
+``simplification''.
 

### 2022-09-29 paper.tex
Remove explicit listing of redundant nodes
And move to the caption, so that people don't have to refer back to the figure too much.

@@
 levels of information about the recombination process to be
-represented in a computionally efficient way.
+represented in a computationally efficient way.
 The key to this is not so much in how we interpret nodes
@@
 (regions 2 to 7). Line widths reflect the genomic span of each edge, highlighting
-the relative importance of different transmission pathways to the samples.
+the relative importance of different transmission pathways to the samples. Nodes
+\textsf{g},  \textsf{j}, \textsf{o} and \textsf{q} are redundant in that they correspond
+to events which have no impact on the genetic ancestry of the samples.
 }
@@
 
-It is worth noting that nodes \textsf{g},  \textsf{j}, \textsf{o} and \textsf{q}
-in Fig~\ref{fig-ancestry-resolution}C are redundant in that they can
-be removed without reducing our knowledge of the genetic
-ancestry of the samples. Removing nodes,
+It is worth noting the process of sample resolution can reveal that certain ARG
+nodes are \emph{redundant}: if they are removed and the edges adjusted to
+pas directly from the node's child to the node's parent, it does not reduce our
+knowledge of the samples' genetic ancestry. Removing nodes,
 both redundant ones and others, is part of a process we call

### 2022-09-29 argutils/viz.py
Change colour scheme

@@ -17,12 +17,12 @@ import scipy.stats
 def arity_colors(n_parents):
     assert n_parents >= 0
     if n_parents == 0:
-        return colorcet.cm.CET_I1(255)  # Red
+        return colorcet.cm.CET_I1(0)  # Blue
     if n_parents == 1:
-        return colorcet.cm.CET_I1(200)  # Yellow
+        return colorcet.cm.CET_I1(100)  # Green
     # Max out at full blue if 6 or more parents; use log scale
     multiple_parents_val = (min(np.log2(n_parents), np.log2(6)) - 1) / (np.log2(6) - 1)
-    return colorcet.cm.CET_I1(100 - int(multiple_parents_val * 100))
+    return colorcet.cm.CET_I1(200 + int(multiple_parents_val * 100))
 
 
 def make_color(rgb, lighten=0):

### 2022-09-29 illustrations.py

@@ -39,7 +39,7 @@ def ancestry_resolution():
             0.96,
             0.96,
             linewidth=2,
-            edgecolor='k',
+            edgecolor='tab:green',
             facecolor='white',
         )
         a.add_patch(icon)
@@ -176,7 +176,7 @@ def ancestry_resolution():
     edges = ts.tables.edges    
     for n in G.nodes:
         if G.nodes[n]["flags"] & argutils.ancestry.NODE_IS_RECOMB:
-            col = 'r'
+            col = 'tab:red'
             size = (1, 0.8)
             breaks = (edges.left[edges.child == n], edges.right[edges.child == n])
             breaks = np.unique(np.concatenate(breaks))
@@ -185,7 +185,7 @@ def ancestry_resolution():
             assert breaks[2] == ts.sequence_length
             breakpoint = breaks[1]
         else:
-            col = 'k' if ts.node(n).is_sample() else 'g'
+            col = 'k' if ts.node(n).is_sample() else 'tab:blue'
             size = (0.5, 1)
             breakpoint = None
         xf, yf = tr_figure(pos[n])

### 2022-10-04 paper.tex
complete fig 5 caption

@@
 locus~\citep{kreitman1983nucleotide}. The \argweaver\ example
-is a sample from the inferred posterior distribution.
+is a sample from the inferred posterior distribution. In each ARG,
+colours are assigned to edges based on the time of the edge's child node: 
+lighter colours are assigned to edges with older child
+nodes, darker colours are assigned to edges with younger child nodes
 Note that the position of points on the
-Y axis is that chosen for graphical clarity by the graphviz \texttt{dot}
-algorithm,  rather than being
-a direct measure of time. Additional notes about these ARGs.
+Y axis is chosen for graphical clarity by the graphviz \texttt{dot}
+algorithm, rather than as a direct measure of time.
+Line width corresponds to the genomic span of the edge. Node colour
+refers to the number of parents of the node: red is zero, brown is one, 
+green to blue are two to N parents. The brightness
+of the node is determined by the genomic span over which
+the node has two or more children divided by the span where it has any
+children. Edges are visualised below each ARG, where each row is an edge 
+and the x axis reflects genomic position. The width of edges
+thus corresponds to their genomic span. Edges
+are ordered by the age of their child node, from oldest to youngest
+moving down the page, with colours matching the edge colours in the ARG. 
+Vertical lines indicate breakpoints between marginal trees.
 }

### 2022-10-04 paper.tex
Add a caption
Involves reversing the coordinates in Fig 4

--- 2022-09-29 paper.tex
@@
 ARG simplification. In all cases the graph visualization is shown on the left
-and the equivalent local trees on the right.
+and the equivalent local trees on the right. Since multiple parents are indicative of
+recombination(s), nodes are coloured by the number of parents. and shaded
+according to the proportion of their span over which they are coalescent.
 (A) A gARG simulated from a diploid Wright-Fisher
@@
 children.
-The penultimate local tree is highlighted to aid discussion.
 (B) Simplified to remove all
@@
 For example, consider the path above node
-\textsf{a} in the highlighted tree (covering positions 60 to 70)
+\textsf{a} in the first tree (covering positions 30 to 40)
 in Fig.~\ref{fig-simplification}A.
@@
 \textsf{p}, which in this region of the genome are all unary nodes.
-This reflects the
-passage of this span of genome through the ARG towards the root.
+This reflects the passage of this span of genome through the ARG towards the root.
 Such unary nodes in a local tree are not usually considered important;
@@
 % detectable by information from any given individual tree.
-nodes usually mark the points where samples find a
- \emph{common} ancestor, such as node \textsf{q} in the highlighted
-tree. For most purposes, we would therefore consider a path directly
+nodes usually mark the points where samples coalesce in a
+ \emph{common} ancestor. For most purposes, we would therefore consider a path directly
 joining sample \textsf{a} to node \textsf{q} to be equivalent to the path

### 2022-10-04 argutils/viz.py

@@ -20,9 +20,12 @@ def arity_colors(n_parents):
         return colorcet.cm.CET_I1(0)  # Blue
     if n_parents == 1:
         return colorcet.cm.CET_I1(100)  # Green
-    # Max out at full blue if 6 or more parents; use log scale
-    multiple_parents_val = (min(np.log2(n_parents), np.log2(6)) - 1) / (np.log2(6) - 1)
-    return colorcet.cm.CET_I1(200 + int(multiple_parents_val * 100))
+    if n_parents == 2:
+        return colorcet.cm.CET_I1(190)  # Orange
+    if n_parents == 3:
+        return colorcet.cm.CET_I1(220)  # orange-red
+    if n_parents == 4:
+        return colorcet.cm.CET_I1(255)  # "Red"
 
 
 def make_color(rgb, lighten=0):
@@ -120,7 +123,7 @@ def draw(
 
     """
     if node_size is None:
-        node_size = 200
+        node_size = 180
     if nonsample_node_shrink is not None:
         # Shrink the nonsample label size
         node_size_array = np.full(ts.num_nodes, node_size / nonsample_node_shrink)
@@ -216,6 +219,7 @@ def draw(
         node_shape="o",
         node_size=node_size,
         font_size=font_size,
+        linewidths=2,
         ax=ax,
         edgelist=[],
     )

### 2022-10-05 paper.tex

--- 2022-10-04 paper.tex
@@
 the node has two or more children divided by the span where it has any
-children. Edges are visualised below each ARG, where each row is an edge 
+children. The genomic intervals associated with edges are visualised below
+each ARG, where each row is an edge 
 and the x axis reflects genomic position. The width of edges
@@
 moving down the page, with colours matching the edge colours in the ARG. 
-Vertical lines indicate breakpoints between marginal trees.
+Vertical lines indicate breakpoints between local trees.
 }

### 2022-10-15 paper.tex
Fixed rates in ARG description

@@
 Common Ancestor, GMRCA), which is guaranteed to
-happen in finite time because of the linear rate of coalescing vs.\ quadratic rate of branching.
+happen in finite time because of the quadratic rate of coalescing vs.\ linear rate of branching.
 

### 2023-08-07 paper.tex
Tidy up "history" section a bit

@@
 % substantial revision.
+% https://www.overleaf.com/5594399339wckggzwhnfvv
 \begin{abstract}
-It has recently become possible to infer genetic ancestry in the presence of
-recombination at scale for the first time, enabling many
+Recent breakthroughs have made it possible to infer genetic ancestry in
+the presence of recombination at scale, enabling many
 downstream applications in population and statistical genetics.
@@
 \section*{Introduction}
-Inferring the genetic relationships between sampled genomes in the form of an
-evolutionary or phylogenetic tree is a necessary prerequisite for many analyses of
+Inferring the genetic relationships between sampled genomes in the form of a
+phylogenetic tree is a necessary prerequisite for many analyses of
 species~\citep{rannala2003genetics}, non-recombining sections of
@@
 the inheritance of each section of the genome may be described
-by a (slightly) different local tree~\citep{hudson1983properties}. Inferring complex
+by a (slightly) different local tree~\citep{hudson1983properties}.
+Inferring complex
 genealogies of this nature
 presents profound technical difficulties, meaning genealogy-based approaches have
-not been widely used in population and statistical genetics. Until recently, analysis of
+not been widely used in population and statistical genetics.
+Until recently, analysis of
 population genetic data has therefore primarily focussed on summaries derived
@@
 is a genealogical tree describing the genetic ancestry of the sample.
-Numerous extensions to the model have been proposed~\citep{hudson1990gene,hein2004gene,wakely2008coalescent}, incorporating various types of evolutionary processes.
-In particular, \citet{hudson1983properties}
-analysed the effects of recombination on coalescent trees, generating several fundamental analytical results,
-as well as describing the elegant algorithm
-still at the heart of widely-used simulation
-programs~\citep{hudson2002generating,kelleher2016efficient,baumdicker2021efficient}.
-In its original formulation, Hudson's algorithm produced a
-sequence of genealogical trees along the genome.
-Crucially, although the correlation structure among these
-local trees was analysed analytically, those correlations
-were not captured explicitly: the ancestry
-was viewed as a set of disconnected local trees rather than
-as a single cohesive object.
-
+Numerous extensions to the model have been
+proposed~\citep{hudson1990gene,hein2004gene,wakely2008coalescent},
+incorporating many evolutionary processes.
+\citet{hudson1983properties}
+first incorporated recombination into the coalescent process,
+providing several fundamental analytical results
+and describing the basic simulation algorithm, still in
+widespread use~\citep{hudson2002generating,kelleher2016efficient,
+baumdicker2021efficient}.
 In the 1990s, Griffiths and colleagues revisited the
@@
 formulating it as a stochastic process where each realisation
-is a genealogy encoded as a graph,
-and showed that the local trees generated by Hudson's algorithm
-are embedded within this graph~\citep{griffiths1991two,ethier1990two,griffiths1996ancestral,griffiths1997ancestral}.
+is encoded as a graph~\citep{griffiths1991two,ethier1990two,
+griffiths1996ancestral,griffiths1997ancestral}.
 They referred to both the stochastic process and
@@
 in particular, a direct implementation of the ARG process
-requires exponential time to simulate
+as originally described requires exponential time to simulate
 (see Appendix XXX for details). However, ARGs provided a way
 to reason about and infer recombinant ancestry as a single object,
-which was not possible within Hudson's framework.
+which was not possible within Hudson's framework, which emphasised
+the collection of local trees along the genome resulting from recombination.
 

### 2023-08-07 paper.tex
Move history to appendix
Closes #287

@@
 
-\section*{Ancestral graphs: a brief history}
-The coalescent~\citep{kingman1982coalescent,kingman1982genealogy,
-hudson1983testing, tajima1983evolutionary} models the ancestry of a sample of
-genomes under an idealised population model, and provides the theoretical
-underpinning for much of contemporary population genetics.
-It is a stochastic process, where each random realisation
-is a genealogical tree describing the genetic ancestry of the sample.
-Numerous extensions to the model have been
-proposed~\citep{hudson1990gene,hein2004gene,wakely2008coalescent},
-incorporating many evolutionary processes.
-\citet{hudson1983properties}
-first incorporated recombination into the coalescent process,
-providing several fundamental analytical results
-and describing the basic simulation algorithm, still in
-widespread use~\citep{hudson2002generating,kelleher2016efficient,
-baumdicker2021efficient}.
-In the 1990s, Griffiths and colleagues revisited the
-coalescent with recombination from a different perspective,
-formulating it as a stochastic process where each realisation
-is encoded as a graph~\citep{griffiths1991two,ethier1990two,
-griffiths1996ancestral,griffiths1997ancestral}.
-They referred to both the stochastic process and
-its random realisations as the Ancestral Recombination Graph (ARG).
-Although mathematically equivalent, it is
-important to note that the Griffiths and Hudson formulations of
-the coalescent with recombination are not identical;
-in particular, a direct implementation of the ARG process
-as originally described requires exponential time to simulate
-(see Appendix XXX for details). However, ARGs provided a way
-to reason about and infer recombinant ancestry as a single object,
-which was not possible within Hudson's framework, which emphasised
-the collection of local trees along the genome resulting from recombination.
-
-Subsequent work on ARGs proceeded in broadly three main directions: focussing on (1)
-exploring the mathematical properties of the coalescent with recombination and
-related stochastic processes, (2) inferring evolutionary parameters under
-(approximations to) this model, either with or without explicitly reconstructing the
-genealogy of the sample, and (3) treating the ARG as a discrete graph, ignoring the
-generating stochastic process, and studying its properties from a computational and
-algorithmic perspective.
-
-% Mathematical treatment of the CwR and ARGs (and ARG-like stochastic processes)
-An extensive body of work has been developed from
-studying the coalescent with recombination
-% Can we add a couple of the most well known CwR examples here
-and other related
-graph-valued stochastic processes from a mathematical perspective.
-In particular, the Ancestral Selection Graph
-(ASG)~\citep{krone1997ancestral,neuhauser1997genealogy}
-uses a similar approach to model natural selection instead of recombination.
-Unlike the ARG process, the ASG imposes a hard distinction between the stochastic process,
-which constructs a random ARG-like graph, and an observable realisation,
-which is a single tree sampled from the graph in a non-uniform way to encode
-desired patterns of natural selection.
-Constructions of ASG-like stochastic processes encoding various
-forms of selection, often in parallel with recombination or other genetic forces,
-are an area of considerable and ongoing theoretical interest~\citep[e.g.][]{
-neuhauser1999ancestral,
-donnelly1999genealogical,
-fearnhead2001perfect,
-fearnhead2003ancestral,
-etheridge2009coalescent,
-gonzalezcasanova2018duality,
-koskela2019robust}.
-% Could mention the Ancestral Gene Transfer graph here too, within
-% something like "Similarly the Ancestral Gene Transfer Graph [cite]
-% and [other ancestral graphs, if they exist?] model genealogical
-% processes in bacteria as a graph.
-
-% Inference of parameters and explicit ARG reconstruction
-Early work on inference under the coalescent with recombination
-focused on the problem of
-inferring the parameters of the
-stochastic process, where the ancestry was regarded as a
-latent parameter to be averaged out
-\citep[e.g.][]{griffiths1996ancestral,kuhner2000maximum, nielsen2000estimation,
-fearnhead2001estimating}.
-These methods met with limited success
-because the state space of ARGs is overwhelmingly large, and
-lacks a simple geometry or neighbourhood structure for inference or
-sampling methods to  exploit.
-Several breakthroughs in this direction were achieved through
-formulating simplified but more tractable approximations to the full
-model~\citep{mcvean2005approximating,marjoram2006fast,li2011inference,
-paul2011accurate,schiffels2014inferring}.
-The related problem of \emph{sampling} genealogies compatible with a given
-dataset under the coalescent with recombination also proved notoriously difficult
-computationally; progress in explicitly inferring genealogies at scale
-has similarly been achieved through resorting to principled
-approximations~\citep{rasmussen2014genome,mahmoudi2022bayesian},
-or moving away from the coalescent with recombination altogether and seeking
-to infer a single plausible ARG~\citep[e.g.][]{speidel2019method} or ARG
-topology~\citep[e.g.][]{minichiello2006mapping,kelleher2019inferring}.
-
-% Optimisation problems to do with ARGs and other phylogenetic networks
-There has also been substantial interest in formulating and answering
-fundamental questions about properties
-of the ARG as a discrete graph structure, abstracting away from the
-generating process and focussing on the ARG topology without considering
-branch lengths.
-The first prominent problem was calculating (lower bounds on) the minimum number of
-recombinations required to reconstruct a valid genealogy for a given
-sample~\citep{myers2003bounds}, and constructing the corresponding
-minimal (parsimonious)
-ARGs~\citep{song2003parsimonious,song2005efficient,lyngso2005minimum}.
-These problems are NP-hard in general~\citep{wang2001perfect}, and progress has
-been achieved through studying various constrained special cases of ARGs~\citep[e.g.][]{gusfield2004optimal} and
-other more general types of phylogenetic networks~\citep{huson2010phylogenetic}. The
-focus has been on algorithmic and combinatorial results, that largely do not
-have direct relevance to the inference problems described above.
-
-%It is important to note that parsimonious ARGs
-%have quite different properties from realisations of the
-%ARG \emph{process} as modelled by Hudson, Griffiths, and colleagues.
-%Firstly, branch lengths are not typically estimated,
-%% why do we need the following two lines
-%and node times are required in order to compute a likelihood under the
-%coalescent with recombination (see section XXX).
-%Secondly, even if branch lengths were estimated,
-%the realisations would have a very small likelihood since
-%the coalescent with recombination is often highly \emph{un}parsimonious
-%in terms of recombination events.
-
-The goal of this historical overview is to illustrate that the meaning of the term ``ARG" now strongly
-depends on the context in which it is used, and can mean both the
-stochastic process generating genealogies in the presence of
-recombination~\citep[e.g.][]{nordborg2000linkage,birkner2013ancestral,
-wilton2015smc,griffiths2016coalescent},
-as well as, more commonly, the concrete realisation of ancestry from a
-process~\citep[e.g.][]{gusfield2014recombinatorics,mathieson2020ancestry,brandt2021evaluation}.
-\cite{minichiello2006mapping} explicitly distinguished the ARG as a
-stochastic process from the ARG as a data structure, and
-argued for using the term to mean specific realised ancestral histories.
-Unless otherwise stated, we will use the ARG-as-a-data-structure
-interpretation for the remainder of this paper.
-
 \section*{Event vs Genome ARGs}\label{eARG}
 In keeping with its original derivation in terms of a stochastic
-process, the classical definition of an ARG data structure is
+process (see Appendix~\ref{sec-arg-history}), the classical definition of an
+ARG data structure is
 in terms of events.
@@
 
+% TODO revisit number here
+\section*{Appendix}
 \appendix
-\section*{Appendix}
+
+\section{Ancestral graphs: a brief history}
+\label{sec-arg-history}
+The coalescent~\citep{kingman1982coalescent,kingman1982genealogy,
+hudson1983testing, tajima1983evolutionary} models the ancestry of a sample of
+genomes under an idealised population model, and provides the theoretical
+underpinning for much of contemporary population genetics.
+It is a stochastic process, where each random realisation
+is a genealogical tree describing the genetic ancestry of the sample.
+Numerous extensions to the model have been
+proposed~\citep{hudson1990gene,hein2004gene,wakely2008coalescent},
+incorporating many evolutionary processes.
+\citet{hudson1983properties}
+first incorporated recombination into the coalescent process,
+providing several fundamental analytical results
+and describing the basic simulation algorithm, still in
+widespread use~\citep{hudson2002generating,kelleher2016efficient,
+baumdicker2021efficient}.
+In the 1990s, Griffiths and colleagues revisited the
+coalescent with recombination from a different perspective,
+formulating it as a stochastic process where each realisation
+is encoded as a graph~\citep{griffiths1991two,ethier1990two,
+griffiths1996ancestral,griffiths1997ancestral}.
+They referred to both the stochastic process and
+its random realisations as the Ancestral Recombination Graph (ARG).
+Although mathematically equivalent, it is
+important to note that the Griffiths and Hudson formulations of
+the coalescent with recombination are not identical;
+in particular, a direct implementation of the ARG process
+as originally described requires exponential time to simulate
+(see Appendix XXX for details). However, ARGs provided a way
+to reason about and infer recombinant ancestry as a single object,
+which was not possible within Hudson's framework, which emphasised
+the collection of local trees along the genome resulting from recombination.
+
+Subsequent work on ARGs proceeded in broadly three main directions: focussing on (1)
+exploring the mathematical properties of the coalescent with recombination and
+related stochastic processes, (2) inferring evolutionary parameters under
+(approximations to) this model, either with or without explicitly reconstructing the
+genealogy of the sample, and (3) treating the ARG as a discrete graph, ignoring the
+generating stochastic process, and studying its properties from a computational and
+algorithmic perspective.
+
+% Mathematical treatment of the CwR and ARGs (and ARG-like stochastic processes)
+An extensive body of work has been developed from
+studying the coalescent with recombination
+% Can we add a couple of the most well known CwR examples here
+and other related
+graph-valued stochastic processes from a mathematical perspective.
+In particular, the Ancestral Selection Graph
+(ASG)~\citep{krone1997ancestral,neuhauser1997genealogy}
+uses a similar approach to model natural selection instead of recombination.
+Unlike the ARG process, the ASG imposes a hard distinction between the stochastic process,
+which constructs a random ARG-like graph, and an observable realisation,
+which is a single tree sampled from the graph in a non-uniform way to encode
+desired patterns of natural selection.
+Constructions of ASG-like stochastic processes encoding various
+forms of selection, often in parallel with recombination or other genetic forces,
+are an area of considerable and ongoing theoretical interest~\citep[e.g.][]{
+neuhauser1999ancestral,
+donnelly1999genealogical,
+fearnhead2001perfect,
+fearnhead2003ancestral,
+etheridge2009coalescent,
+gonzalezcasanova2018duality,
+koskela2019robust}.
+% Could mention the Ancestral Gene Transfer graph here too, within
+% something like "Similarly the Ancestral Gene Transfer Graph [cite]
+% and [other ancestral graphs, if they exist?] model genealogical
+% processes in bacteria as a graph.
+
+% Inference of parameters and explicit ARG reconstruction
+Early work on inference under the coalescent with recombination
+focused on the problem of
+inferring the parameters of the
+stochastic process, where the ancestry was regarded as a
+latent parameter to be averaged out
+\citep[e.g.][]{griffiths1996ancestral,kuhner2000maximum, nielsen2000estimation,
+fearnhead2001estimating}.
+These methods met with limited success
+because the state space of ARGs is overwhelmingly large, and
+lacks a simple geometry or neighbourhood structure for inference or
+sampling methods to  exploit.
+Several breakthroughs in this direction were achieved through
+formulating simplified but more tractable approximations to the full
+model~\citep{mcvean2005approximating,marjoram2006fast,li2011inference,
+paul2011accurate,schiffels2014inferring}.
+The related problem of \emph{sampling} genealogies compatible with a given
+dataset under the coalescent with recombination also proved notoriously difficult
+computationally; progress in explicitly inferring genealogies at scale
+has similarly been achieved through resorting to principled
+approximations~\citep{rasmussen2014genome,mahmoudi2022bayesian},
+or moving away from the coalescent with recombination altogether and seeking
+to infer a single plausible ARG~\citep[e.g.][]{speidel2019method} or ARG
+topology~\citep[e.g.][]{minichiello2006mapping,kelleher2019inferring}.
+
+% Optimisation problems to do with ARGs and other phylogenetic networks
+There has also been substantial interest in formulating and answering
+fundamental questions about properties
+of the ARG as a discrete graph structure, abstracting away from the
+generating process and focussing on the ARG topology without considering
+branch lengths.
+The first prominent problem was calculating (lower bounds on) the minimum number of
+recombinations required to reconstruct a valid genealogy for a given
+sample~\citep{myers2003bounds}, and constructing the corresponding
+minimal (parsimonious)
+ARGs~\citep{song2003parsimonious,song2005efficient,lyngso2005minimum}.
+These problems are NP-hard in general~\citep{wang2001perfect}, and progress has
+been achieved through studying various constrained special cases of ARGs~\citep[e.g.][]{gusfield2004optimal} and
+other more general types of phylogenetic networks~\citep{huson2010phylogenetic}. The
+focus has been on algorithmic and combinatorial results, that largely do not
+have direct relevance to the inference problems described above.
+
+%It is important to note that parsimonious ARGs
+%have quite different properties from realisations of the
+%ARG \emph{process} as modelled by Hudson, Griffiths, and colleagues.
+%Firstly, branch lengths are not typically estimated,
+%% why do we need the following two lines
+%and node times are required in order to compute a likelihood under the
+%coalescent with recombination (see section XXX).
+%Secondly, even if branch lengths were estimated,
+%the realisations would have a very small likelihood since
+%the coalescent with recombination is often highly \emph{un}parsimonious
+%in terms of recombination events.
+
+The goal of this historical overview is to illustrate that the meaning of the term ``ARG" now strongly
+depends on the context in which it is used, and can mean both the
+stochastic process generating genealogies in the presence of
+recombination~\citep[e.g.][]{nordborg2000linkage,birkner2013ancestral,
+wilton2015smc,griffiths2016coalescent},
+as well as, more commonly, the concrete realisation of ancestry from a
+process~\citep[e.g.][]{gusfield2014recombinatorics,mathieson2020ancestry,brandt2021evaluation}.
+\cite{minichiello2006mapping} explicitly distinguished the ARG as a
+stochastic process from the ARG as a data structure, and
+argued for using the term to mean specific realised ancestral histories.
+Unless otherwise stated, we will use the ARG-as-a-data-structure
+interpretation for the remainder of this paper.
 

### 2023-08-07 paper.tex

@@
 to describe patterns of genetic inheritance must be in terms
-of specific causal events. As we demonstrate in subsequent sections,
+of specific causal events. As demonstrated in subsequent sections,
 there are significant advantages to using
@@
 and may be the recombined product of that parent's two genomes.
+% NOTE: does this example really help? Rather than spelling out what
+% we can already see, perhaps we should point out the non-obvious
+% thing that we have decoupled from "events"
 For example, individual $I_1$ in Fig.~\ref{fig-arg-in-pedigree}
@@
 might seem of little practical importance.
-However, as we show in the remainder of this article,
-there are substantial practical gains
+However, there are substantial practical benefits
 from using the gARG encoding, allowing differential
@@
 In practice, the backwards-in-time Griffiths process that produces
-ARGs such as that shown in Figs~{fig-ancestry-resolution}A~and~B
+ARGs such as that shown in Figs~\ref{fig-ancestry-resolution}A~and~B
 is rarely simulated, because the resulting ``big ARG'' (Appendix XXX) is
@@
 
-It is worth noting the process of sample resolution can reveal that certain ARG
-nodes are \emph{redundant}: if they are removed and the edges adjusted to
-pas directly from the node's child to the node's parent, it does not reduce our
-knowledge of the samples' genetic ancestry. Removing nodes,
-both redundant ones and others, is part of a process we call
-``simplification''.
+% It is worth noting the process of sample resolution can reveal that certain ARG
+% nodes are \emph{redundant}: if they are removed and the edges adjusted to
+% pas directly from the node's child to the node's parent, it does not reduce our
+% knowledge of the samples' genetic ancestry. Removing nodes,
+% both redundant ones and others, is part of a process we call
+% ``simplification''.
 
@@
 
-Because the ARGs inferred by \tsinfer\ and \relate\ do not contain
-explicit recombination events we cannot compute a likelihood
-under the coalescent with recombination (see appendix XXX).
-[Some stuff about whether we should see this as necessary or not,
-but maybe we should keep this for the conclusion?]
+% Because the ARGs inferred by \tsinfer\ and \relate\ do not contain
+% explicit recombination events we cannot compute a likelihood
+% under the coalescent with recombination (see appendix XXX).
+% [Some stuff about whether we should see this as necessary or not,
+% but maybe we should keep this for the conclusion?]
 

### 2023-08-08 paper.tex

@@
 
-% JK: this is a rough first pass for a slightly different paper. Needs
-% substantial revision.
-% https://www.overleaf.com/5594399339wckggzwhnfvv
+% FIXME this is weak, but gives the basic outline
 \begin{abstract}
-Recent breakthroughs have made it possible to infer genetic ancestry in
+Recent breakthroughs have made it possible to infer genetic genealogies in
 the presence of recombination at scale, enabling many
 downstream applications in population and statistical genetics.
-Such recombinant genetic ancestry is usually
-referred to as an Ancestral Recombination Graph, or ARG.
-% Note: slight repetition here with first sentence.
-There are now multiple methods that can infer ARGs at a practical scale
-and it is therefore vital that these methods can be systematically evaluated
-and compared to determine their strengths and weaknesses.
-Unfortunately confusion exists over terminology and there is
-little agreement on shared standards for data interchange,
-significantly hampering progress in the field.
-This lack of clarity and agreement on basic issues such as what does and
-does not constitute an ARG is partly attributable to the historical
-development of the term.
-Originally rigorously defined as
-a stochastic process (the coalescent with recombination) together with its
-representation, an ARG is now understood to refer to any
-concrete realisation of a recombinant genetic ancestry.
-We show that the standard description of an ARG
-in terms of common ancestor and recombination events
-(inherited from the original stochastic process definition) is
-ambiguous,
-% is it ambiguous? How? Can we simply delete this?
-fundamentally limited and an unsuitable basis for data interchange.
-We provide a simple alternative definition of an ARG in which a node
-corresponds to one of an individual's monoploid genomes and an edge
-defines genetic inheritance between two nodes, with annotations on
-each edge specifying which regions of the genome have
-been inherited via that route.
-We show how this definition can be used to generalise the encoding to
-cover both the standard and several simplified versions of ARGs, as well as
-enabling efficient access to local trees along the genome. Additionally, we
-define some classifications and metrics of ARG nodes. Finally, we relate
-these generalisations to the ARGs produced by a number of recently
-developed ancestral inference methods.
-% jk-note: this ends a bit flat - we should say other stuff like we clarify
-% how the ARG relates the pedigree, etc, etc. Let's go through again when
-% we have a reasonable draft.
+The structure representing such recombinant genetic ancestry
+is usually referred to as an Ancestral Recombination Graph (ARG),
+although there is some confusion about the interpretation and little
+agreement on the concrete details.
+We propose a concrete definition of ARGs that is suitable for modern
+datasets, in terms of genomes and their patterns of inheritance,
+and contrast this with the classical event-based definition which
+makes several limiting assumptions.
+We show that this new definition enables fine-grained
+levels of precision about ancestral processes to be represented,
+and forms the basis of an efficient computational framework.
 \end{abstract}

### 2023-08-08 paper.tex
Update

@@
 
-\section*{Event vs Genome ARGs}\label{eARG}
-In keeping with its original derivation in terms of a stochastic
-process (see Appendix~\ref{sec-arg-history}), the classical definition of an
-ARG data structure is
-in terms of events.
-Nodes represent
-common ancestor or recombination events that occurred in the
-history of some set of sample genomes, and edges represent ancestral
-lineages~\citep{griffiths1996ancestral}.
-In a common ancestor event the inbound lineages are merged into a
-single ancestral lineage, and in a recombination
-event a lineage is split into two independent
-ancestral lineages.
-% The final vital
-% detail is that we associate a breakpoint with each recombination
-% event.
-% This information is sufficient to uniquely
-% define the local genealogical trees at every position along the genome,
-% which is the basic requirement for a data structure encoding
-% genome-wide genetic ancestry.
-As illustrated in Fig.~\ref{fig-arg-data-structure}, this information
-is usually concretely encoded by associating a recombination breakpoint
-(which is null for common ancestor events) and an ordered list of parents
-with each node. We refer to such a  data structure as an
-``event ARG'' (eARG).
-To recover the local tree at genome coordinate $x$ (the most
-fundamental operation required for an ARG data structure),
-we traverse the graph rootwards from the leaves.
-At a particular
-node $u$, if it has one parent we are at a common ancestor
-node and we follow that parent. If we are at a
-recombination node, $u$ has two parents: if
-$x$ is less than the breakpoint we follow
-the edge to the first parent, and otherwise follow the edge
-to the second parent.
-The order in which parents are listed at a recombination node is
-therefore significant, telling us
-from which parent the segments of genome to the left and right of the breakpoint
-were inherited.
+% \section*{Genome ARGs}\label{eARG}
+% In keeping with its original derivation in terms of a stochastic
+% process (see Appendix~\ref{sec-arg-history}), the classical definition of an
+% ARG data structure is
+% in terms of events.
+% Nodes represent
+% common ancestor or recombination events that occurred in the
+% history of some set of sample genomes, and edges represent ancestral
+% lineages~\citep{griffiths1996ancestral}.
+% In a common ancestor event the inbound lineages are merged into a
+% single ancestral lineage, and in a recombination
+% event a lineage is split into two independent
+% ancestral lineages.
+% % The final vital
+% % detail is that we associate a breakpoint with each recombination
+% % event.
+% % This information is sufficient to uniquely
+% % define the local genealogical trees at every position along the genome,
+% % which is the basic requirement for a data structure encoding
+% % genome-wide genetic ancestry.
+% As illustrated in Fig.~\ref{fig-arg-data-structure}, this information
+% is usually concretely encoded by associating a recombination breakpoint
+% (which is null for common ancestor events) and an ordered list of parents
+% with each node. We refer to such a  data structure as an
+% ``event ARG'' (eARG).
+% To recover the local tree at genome coordinate $x$ (the most
+% fundamental operation required for an ARG data structure),
+% we traverse the graph rootwards from the leaves.
+% At a particular
+% node $u$, if it has one parent we are at a common ancestor
+% node and we follow that parent. If we are at a
+% recombination node, $u$ has two parents: if
+% $x$ is less than the breakpoint we follow
+% the edge to the first parent, and otherwise follow the edge
+% to the second parent.
+% The order in which parents are listed at a recombination node is
+% therefore significant, telling us
+% from which parent the segments of genome to the left and right of the breakpoint
+% were inherited.
 
@@
 
-\begin{figure}
-\centering
-% \begin{tabular}{cc}
-\begin{tikzpicture}[x=5mm, y=5mm, node distance=2mm and 20mm]
-\tikzset{greynode/.style={circle,fill,inner sep=1},
-nodelabel/.style={font=\footnotesize}}
-
-\node (s0) [greynode] at (0, 0) {};
-\node (s1) [greynode] at (3, 0) {};
-\node (s2) [greynode] at (6, 0) {};
-\node (s3) [greynode] at (3, 1) {};
-\node (s4) [greynode] at (1, 2) {};
-\node (s5) [greynode] at (5, 3) {};
-\node (s6) [greynode] at (3, 4) {};
-
-% \node [anchor=north west] at (0,6) {A};
-\node [nodelabel,anchor=north west] at ($(s3) + (0,0)$) {$x$};
-\foreach \u/\lab in {s0/$\textsf{a}$, s1/$\textsf{b}$, s2/$\textsf{c}$} \node[nodelabel,anchor=north] at (\u) {\lab};
-\foreach \u/\lab in {s4/$\textsf{e}$} \node[nodelabel,anchor=south west] at (\u) {\lab};
-\foreach \u/\lab in {s5/$\textsf{f}$} \node[nodelabel,anchor=south east] at (\u) {\lab};
-\foreach \u/\lab in {s3/$\textsf{d}$, s6/$\textsf{g}$} \node[nodelabel,anchor=south] at (\u) {\lab};
-
-%% Edges
-\draw (s1) -- (s3);
-\draw (s0) |- (s4);
-\draw (s4) -- (2,2) |- (s3);
-\draw (s4) |- (s6);
-\draw (s3) -- (4,1) |- (s5);
-\draw (s2) |- (s5);
-\draw (s5) |- (s6);
-
-% \node [anchor=north west] at (9,6) {B};
-\node [nodelabel,anchor=north west] at ($(10,5)$) {
-\begin{tabular}{c|c|l}
-% \multicolumn{2}{c}{Breakpoints}\\
-Node & Breakpoint & Parents\\
-\hline
-$\noderef{a}$ & $\varnothing$ & [\noderef{e}]\\
-$\noderef{b}$ & $\varnothing$ & [\noderef{d}] \\
-$\noderef{c}$ & $\varnothing$ & [\noderef{f}]\\
-$\noderef{d}$ & $x$ & [\noderef{e}, \noderef{f}] \\
-$\noderef{e}$ & $\varnothing$ & [\noderef{g}]\\
-$\noderef{f}$ & $\varnothing$ & [\noderef{g}]\\
-$\noderef{g}$ & $\varnothing$ & []\\
-\end{tabular}};
-
-% \node [anchor=north west] at (9,6) {C};
+% \begin{figure}
+% \centering
+% % \begin{tabular}{cc}
+% \begin{tikzpicture}[x=5mm, y=5mm, node distance=2mm and 20mm]
+% \tikzset{greynode/.style={circle,fill,inner sep=1},
+% nodelabel/.style={font=\footnotesize}}
+
+% \node (s0) [greynode] at (0, 0) {};
+% \node (s1) [greynode] at (3, 0) {};
+% \node (s2) [greynode] at (6, 0) {};
+% \node (s3) [greynode] at (3, 1) {};
+% \node (s4) [greynode] at (1, 2) {};
+% \node (s5) [greynode] at (5, 3) {};
+% \node (s6) [greynode] at (3, 4) {};
+
+% % \node [anchor=north west] at (0,6) {A};
+% \node [nodelabel,anchor=north west] at ($(s3) + (0,0)$) {$x$};
+% \foreach \u/\lab in {s0/$\textsf{a}$, s1/$\textsf{b}$, s2/$\textsf{c}$} \node[nodelabel,anchor=north] at (\u) {\lab};
+% \foreach \u/\lab in {s4/$\textsf{e}$} \node[nodelabel,anchor=south west] at (\u) {\lab};
+% \foreach \u/\lab in {s5/$\textsf{f}$} \node[nodelabel,anchor=south east] at (\u) {\lab};
+% \foreach \u/\lab in {s3/$\textsf{d}$, s6/$\textsf{g}$} \node[nodelabel,anchor=south] at (\u) {\lab};
+
+% %% Edges
+% \draw (s1) -- (s3);
+% \draw (s0) |- (s4);
+% \draw (s4) -- (2,2) |- (s3);
+% \draw (s4) |- (s6);
+% \draw (s3) -- (4,1) |- (s5);
+% \draw (s2) |- (s5);
+% \draw (s5) |- (s6);
+
+% % \node [anchor=north west] at (9,6) {B};
 % \node [nodelabel,anchor=north west] at ($(10,5)$) {
-% \begin{tabular}{ll}
-% Edges & Intervals\\
+% \begin{tabular}{c|c|l}
+% % \multicolumn{2}{c}{Breakpoints}\\
+% Node & Breakpoint & Parents\\
 % \hline
-% $(\textsf{a}, \textsf{e})$ & $(0, L]$ \\
-% $(\textsf{b}, \textsf{d})$ & $(0, L]$ \\
-% $(\textsf{c}, \textsf{f})$ & $(0, L]$ \\
-% $(\textsf{d}, \textsf{e})$ & $(0, x]$ \\
-% $(\textsf{d}, \textsf{f})$ & $(x, L]$ \\
-% $(\textsf{e}, \textsf{g})$ & $(0, L]$ \\
-% $(\textsf{f}, \textsf{g})$ & $(0, L]$ \\
+% $\noderef{a}$ & $\varnothing$ & [\noderef{e}]\\
+% $\noderef{b}$ & $\varnothing$ & [\noderef{d}] \\
+% $\noderef{c}$ & $\varnothing$ & [\noderef{f}]\\
+% $\noderef{d}$ & $x$ & [\noderef{e}, \noderef{f}] \\
+% $\noderef{e}$ & $\varnothing$ & [\noderef{g}]\\
+% $\noderef{f}$ & $\varnothing$ & [\noderef{g}]\\
+% $\noderef{g}$ & $\varnothing$ & []\\
 % \end{tabular}};
 
-\end{tikzpicture}
-\caption{\label{fig-arg-data-structure}
-A graphical depiction of a classical event ARG and a concrete
-encoding of the information. Node \noderef{d} is a recombination
-event as it is associated with a non-null breakpoint, and has
-two parents. Nodes \noderef{e}, \noderef{f} and \noderef{g} are
-common ancestor events.
-}
-\end{figure}
+% % \node [anchor=north west] at (9,6) {C};
+% % \node [nodelabel,anchor=north west] at ($(10,5)$) {
+% % \begin{tabular}{ll}
+% % Edges & Intervals\\
+% % \hline
+% % $(\textsf{a}, \textsf{e})$ & $(0, L]$ \\
+% % $(\textsf{b}, \textsf{d})$ & $(0, L]$ \\
+% % $(\textsf{c}, \textsf{f})$ & $(0, L]$ \\
+% % $(\textsf{d}, \textsf{e})$ & $(0, x]$ \\
+% % $(\textsf{d}, \textsf{f})$ & $(x, L]$ \\
+% % $(\textsf{e}, \textsf{g})$ & $(0, L]$ \\
+% % $(\textsf{f}, \textsf{g})$ & $(0, L]$ \\
+% % \end{tabular}};
+
+% \end{tikzpicture}
+% \caption{\label{fig-arg-data-structure}
+% A graphical depiction of a classical event ARG and a concrete
+% encoding of the information. Node \noderef{d} is a recombination
+% event as it is associated with a non-null breakpoint, and has
+% two parents. Nodes \noderef{e}, \noderef{f} and \noderef{g} are
+% common ancestor events.
+% }
+% \end{figure}
 
@@
 
-% \section*{Genome ARGs}
-There is no fundamental reason, however, that the data structure we use
-to describe patterns of genetic inheritance must be in terms
-of specific causal events. As demonstrated in subsequent sections,
-there are significant advantages to using
-a slightly different formulation, which
-we call a ``genome ARG'' (gARG).
+\section*{Genome ARGs}
+% There is no fundamental reason, however, that the data structure we use
+% to describe patterns of genetic inheritance must be in terms
+% of specific causal events. As demonstrated in subsequent sections,
+% there are significant advantages to using
+% a slightly different formulation, which
+% we call a ``genome ARG'' (gARG).
+In this section we define a concrete definition of an ARG in terms
+of particular genomes and their patterns of inheritance, which
+to avoid confusion, we refer to as a ``genome ARG``, or gARG.
+(We contrast this with the classical definition derived from the
+coalescent process in later sections.)
 First, let us define a genome as the complete set of genetic material that a child
@@
 (the next section explores the subtleties of what we mean by
-genetic inheritance). To recover the local tree for a genomic
-position $x$, we traverse the graph rootwards from the leaves
-(as in an eARG) and at a particular node, we follow the outbound
-edge associated with the genomic interval containing $x$.
+genetic inheritance).
+% To recover the local tree for a genomic
+% position $x$, we traverse the graph rootwards from the leaves
+% (as in an eARG) and at a particular node, we follow the outbound
+% edge associated with the genomic interval containing $x$.
 % Recent discussions have illustrated the
@@
 In this diploid setting
-(there is no limitation on mating system, ploidy, or age structure
-in the gARG encoding),
+(there is no limitation on mating system, ploidy, or age structure)
 a genome is inherited from one
@@
 
-% Formally, we can define a gARG as a set of edges $E$, where each
-% element of the set is a tuple $(c, p, I)$ such that $c, p \in \mathbb{N}$,
-% $c$ is the child node, $p$ is the parent node and $I$ is the set of
-% disjoint genomic intervals $(\ell, r]$ over which genome $c$ inherits from $p$ (we may
-% also phrase this in terms of the ancestral material of a sample---see
-% the next section).
-% Deriving the local tree at a point $x$
-% follows largely the same pattern as for an eARG but is somewhat
-% simpler because coordinate information is directly associated with
-% edges. Starting with the sample nodes $S$ we traverse
-% rootwards through the graph from child to parent. At each node $u$, we find an
-% edge $(c, p, I) \in E$ such that $u = c$ and $x \in I$
-% and move to the next node $p$.
+Formally, we can define a gARG as a set of edges $E$, where each
+element of the set is a tuple $(c, p, I)$ such that $c, p \in \mathbb{N}$,
+$c$ is the child node, $p$ is the parent node and $I$ is the set of
+disjoint genomic intervals $(\ell, r]$ over which genome $c$ inherits from $p$ (we may
+also phrase this in terms of the ancestral material of a sample---see
+the next section).
+% To recover the local tree for a genomic
+% position $x$, we traverse the graph rootwards from the leaves
+% (as in an eARG) and at a particular node, we follow the outbound
+% edge associated with the genomic interval containing $x$.
+To recover the local tree at genome coordinate $x$ (the most
+fundamental operation required for an ARG data structure; see
+Section XXX),
+we traverse the graph rootwards from the leaves.
+Starting with the sample nodes $S$ we traverse
+rootwards through the graph from child to parent. At each node $u$, we find an
+edge $(c, p, I) \in E$ such that $u = c$ and $x \in I$
+and move to the next node $p$.
+
+[COMMENT ON FIG 1C, and unary nodes in the trees]
 
@@
 % Try to tackle this head on by getting to the heart of the distinction.
-The difference between eARGs and gARGs are subtle, and whether
-we choose to interpret nodes as events or genomes
-might seem of little practical importance.
-However, there are substantial practical benefits
-from using the gARG encoding, allowing differential
-levels of information about the recombination process to be
-represented in a computationally efficient way.
-The key to this is not so much in how we interpret nodes
-(see Appendix XXX), but specifically in how
-we represent the process of genetic inheritance.
-The critical distinction is that
-the gARG encoding allows us to directly describe
-genetic inheritance in multiple non-contiguous genome intervals
-between an ancestor and a descendant, something that
-cannot be done by storing recombination breakpoints.
+% The difference between eARGs and gARGs are subtle, and whether
+% we choose to interpret nodes as events or genomes
+% might seem of little practical importance.
+% However, there are substantial practical benefits
+% from using the gARG encoding, allowing differential
+% levels of information about the recombination process to be
+% represented in a computationally efficient way.
+% The key to this is not so much in how we interpret nodes
+% (see Appendix XXX), but specifically in how
+% we represent the process of genetic inheritance.
+% The critical distinction is that
+% the gARG encoding allows us to directly describe
+% genetic inheritance in multiple non-contiguous genome intervals
+% between an ancestor and a descendant, something that
+% cannot be done by storing recombination breakpoints.
 

### 2023-08-09 paper.tex
More meat in the Genome ARGs section

@@
 local trees.
-In this diploid setting
-(there is no limitation on mating system, ploidy, or age structure)
+In this diploid example
 a genome is inherited from one
@@
 recombination.
-Arbitrary patterns of genetic inheritance (gene conversions, multiple recombinations,
-etc) can be expressed without difficulty using this encoding.
 
@@
 % it'll be clearer what we need to say in the caption vs the text.
-An Ancestral Recombination Graph defined in terms of genomes embedded
-in a pedigree. (A) Diploid individuals, visualised in a highly inbred pedigree and
-labelled $I_1$ to $I_8$, contain both paternal (left) and maternal (right) genomes
+An example gARG embedded in a pedigree.
+(A) Diploid individuals, visualised in a highly inbred pedigree and
+labelled $I_1$ to $I_8$, contain both paternal  and maternal  genomes
 labelled \textsf{a} to \textsf{p}. Black lines show transmission paths connecting
 genomes in the current generation (\textsf{a} to \textsf{d}) with their ancestors.
-Two recombination breakpoints, at positions 2 and 7 mean that genomes \textsf{a}
-and \textsf{c} are independent recombinants of the paternal genomes \textsf{e}
+Genomes \textsf{a} and \textsf{c} are independent recombinants of
+the paternal genomes \textsf{e}
 % Gregor: there is some overlap between a and c still (the middle region inherited
 % from f) so, a and c are not fully independent!
-and \textsf{f}: in these cases the edge annotations have been shown, indicating the
-region of genome transmitted. Genomes are shaded such that where, backwards in time,
-they merge into a common ancestor, the merged region is darker; this shows that all
-genomic regions find a common ancestor in genome \textsf{n} in the oldest generation.
-(B) The corresponding genome ARG, where there is a single type of node (representing
-a genome) along with annotations on all edges; annotations showing partial
-transmission are highlighted in bold. Recombination (e.g.,
+and \textsf{f}, and regions of genome transmitted are shown.
+Genomes are shaded such that where, backwards in time,
+they merge into a common ancestor, the merged region is darker.
+(B) The corresponding genome ARG along with annotations on all edges;
+annotations showing partial
+genome transmission are highlighted in bold. Recombination (e.g.,
 the cause of the breakpoint at position 7)
@@
 
-Formally, we can define a gARG as a set of edges $E$, where each
-element of the set is a tuple $(c, p, I)$ such that $c, p \in \mathbb{N}$,
-$c$ is the child node, $p$ is the parent node and $I$ is the set of
+Formally, we can define a gARG as a set of nodes $N$ and edges $E$, where each
+element of $E$ is a tuple $(c, p, I)$ such that
+$c\in N$ is the child node, $p\in N$ is the parent node and $I$ is the set of
 disjoint genomic intervals $(\ell, r]$ over which genome $c$ inherits from $p$ (we may
@@
 edge $(c, p, I) \in E$ such that $u = c$ and $x \in I$
-and move to the next node $p$.
-
-[COMMENT ON FIG 1C, and unary nodes in the trees]
+and move to the next node $p$. The efficiency of this process is discussed
+in Section XXX).
+
+There are some interesting consequences to these definitions, exemplified
+in Figure~\ref{fig-arg-in-pedigree}. Firstly,
+there is
+is no limitation on mating system, ploidy, or age structure. By making the
+basic unit of analysis an individual's \emph{genomes} (i.e., two per individuals
+for diploids), any pattern of inheritance among mixed ploidy can be
+accommodated.
+Secondly, because we only record the \emph{intervals} of inheritance
+between genomes (and not the causal \emph{processes}),
+arbitrary patterns of genetic inheritance such as gene conversions,
+multiple recombinations and so on can be directly expressed.
+Thirdly, ``local`` trees that we generate for a particular position on the
+genome are not explicitly concerned with ``coalescence'' (nodes at which
+one or more samples find a common ancestor); for example, node $e$
+is unary in the left-most tree of Figure~\ref{fig-arg-in-pedigree},
+denoting the fact that $e$ the ancestor of $a$ over the interval $(0, 2]$,
+but no coalescence occurs in this interval (see section XXX for more
+discussion). Fourth, recombination is modelled in terms of
+its outcomes in terms of inheritance intervals among parental genomes;
+thus, we will have a child node (the recombinant; e.g. node $a$
+Figure~\ref{fig-arg-in-pedigree}) and two (or more; see section XXX)
+paternal genomes which this child has inherited segments from.
+Finally, gARGs are not defined in terms of the
+coalescent with recombination (or any other) stochastic process.
+Thus, the gARG encoding is focused on faithfully representing the
+\emph{outcome} of the evolutionary processes of genetic inheritance and
+recombination, and is agnostic to the details of those processes.
 

### 2023-08-09 paper.tex
More updates

@@
 with the genome coordinates over which inheritance occurs.
-Fig~\ref{fig-ancestry-resolution} is an instructive example
-from the literature \citep{wiuf1999recombination} and illustrates
-how a classical eARG (Fig~\ref{fig-ancestry-resolution}A) can be described as
-a gARG with edge annotations (Figs~\ref{fig-ancestry-resolution}B~and~C).
-
-There are two distinct annotation schemes possible, and the distinction is subtle
-and important. The first scheme is to use the annotations to describe the complete
-set of genomic intervals directly inherited by the descendant node from the ancestor
-node (Fig~\ref{fig-ancestry-resolution}B). The second is to use the edge annotations to
-describe the ``ancestral material''~\citep{wiuf1999ancestry,wiuf1999recombination}
-of the sample that is propagated through
-the graph from descendant to ancestor (Fig~\ref{fig-ancestry-resolution}C).
-As we discuss in section XXX, this second scheme has many advantages
-in terms of efficiency.
-
+There are two different ways in which we use these inheritance
+annotations, and the distinction is subtle and important. The difference
+is illustrated in Fig~\ref{fig-ancestry-resolution}.
+
+% TODO majorly condense this caption, it's much too long.
 \begin{figure}
@@
 \caption{\label{fig-ancestry-resolution}
+TODO DELETE A? Or keep it? Maybe this would help some hard-core eARGers?
 The \citet[][Fig.~1]{wiuf1999recombination} example graph as an eARG and
 two gARGs with alternative edge annotation schemes.
-(A) A conventional Griffiths eARG. Square nodes represent events, with
-each recombination (red) annotated with the position of a single breakpoint
-along the 7-unit-long chromosome.
-(B) A gARG with direct-inheritance edge annotations. Oval nodes represent
+% (A) A conventional Griffiths eARG. Square nodes represent events, with
+% each recombination (red) annotated with the position of a single breakpoint
+% along the 7-unit-long chromosome.
+(A) A gARG with edges annotated by . Oval nodes represent
 genomes and annotations reflect ``local'' transmission of all genetic material
@@
 
+The simplest interpretation of the inheritance intervals $I$ for
+an ancestor genome $v$ and descendant $u$, is that $I$ defines the
+regions of genome that $u$ inherited from $v$. For example,
+node $o$ is the recombinant descendant of nodes $p$ and $q$ in
+Figure~\ref{fig-ancestry-resolution}A, which shows that it in inherited
+everything to the left of position $1$ from $p$ and the rest of
+its genome from node $q$.
+Thus, these edge annotations are concerned only with the genetic relationship
+between the nodein question, and are not affected by the rest of the graph.
+We will thefore refer to inheritance annotations of this type as ``local inheritance
+annotations''. Graphs of this type result from forwards-time
+simulations~\citep{kelleher2018efficient,haller2018tree} or the
+``Big ARG'' stochastic process (Appendix XXX).
+
+ARGs are most often thought of as a \emph{retrospective} structure however,
+defined with respect to some set of (usually, contemporary) sample genomes.
+In this case we are not explicitly interested in the
+entire interval of genome descendant $u$ inherited from ancestor $v$,
+but rather the subset of that interval that is \emph{ancestral to our
+sample}.
+Fig~\ref{fig-ancestry-resolution}B shows the graph that we obtain when
+reduce the annotation intervals to those that are ancestral to the
+sample (note that node $q$ is now ommitted because the MRCA of the sample
+has already been found for the full genome at node $p$).
+Tracking such
+``ancestral material''~\citep{wiuf1999ancestry,wiuf1999recombination}
+is usually much more useful and, indeed, all that is possible to
+have information about when inferring ARGs from contemporary samples.
+We refer to a gARG with inheritance annotations of this type as
+``sample resolved''.
+
+% Fig~\ref{fig-ancestry-resolution} is an instructive example
+% from the literature \citep{wiuf1999recombination} and illustrates
+% how a classical eARG (Fig~\ref{fig-ancestry-resolution}A) can be described as
+% a gARG with edge annotations (Figs~\ref{fig-ancestry-resolution}B~and~C).
+
+% There are two distinct annotation schemes possible, and the distinction is subtle
+% and important. The first scheme is to use the annotations to describe the complete
+% set of genomic intervals directly inherited by the descendant node from the ancestor
+% node (Fig~\ref{fig-ancestry-resolution}B). The second is to use the edge annotations to
+% describe the ``ancestral material''~\citep{wiuf1999ancestry,wiuf1999recombination}
+% of the sample that is propagated through
+% the graph from descendant to ancestor (Fig~\ref{fig-ancestry-resolution}C).
+% As we discuss in section XXX, this second scheme has many advantages
+% in terms of efficiency.
+
 %ARGs are inherently a retrospective structure defined in terms
@@
 
-Given the direct-inheritance edge annotations
+Given a gARG with direct-inheritance edge annotations
 (Fig~\ref{fig-ancestry-resolution}B),
@@
 already fully coalesced (indicated by a fully black genomic region).
-This process is directly analogous to Hudson's simulation algorithm
-(Appendix XXX).
-
-In practice, the backwards-in-time Griffiths process that produces
-ARGs such as that shown in Figs~\ref{fig-ancestry-resolution}A~and~B
-is rarely simulated, because the resulting ``big ARG'' (Appendix XXX) is
-dominated by nodes such as \textsf{g} and \textsf{j} which contain information
-about inheritance in non-ancestral material, as well as nodes such as
-\textsf{o} and \textsf{q} which track details of ancestry even once the
-local genomic region has fully coalesced. Instead, simulating
-ARGs from the coalescent with recombination is usually done by
-running Hudson's algorithm to produce a sample-resolved ARG directly.
-Moreover, when inferring ARGs from genetic data, there is no information
-in the sample set that can be used to resolve these nodes, so that they
-are essentially unknowable.
-
-Nevertheless, there is a practical use for converting a direct-inheritance
-gARG to a sample resolved version. It is the key algorithm underlying recent
-advances in forwards-in-time simulation.
-\cite{kelleher2018efficient} developed a method in which forwards-in-time
-simulations are performed by recording all parent-child genetic relationships
-(i.e.~annotating the edges in the simulated pedigree with the genomic
-intervals inherited by a child from a parent) and periodically ``simplifying''
-this structure to retain only ancestral relationships that are relevant
-to the ancestry of the current population. In many situations, using this approach to
-record the full genetic ancestry of the current population is orders of magnitude faster
-than the alternative of tracking neutral genetic variation directly~\citep{kelleher2018efficient,haller2018tree}.
+
+% In practice, the backwards-in-time Griffiths process that produces
+% ARGs such as that shown in Figs~\ref{fig-ancestry-resolution}A~and~B
+% is rarely simulated, because the resulting ``big ARG'' (Appendix XXX) is
+% dominated by nodes such as \textsf{g} and \textsf{j} which contain information
+% about inheritance in non-ancestral material, as well as nodes such as
+% \textsf{o} and \textsf{q} which track details of ancestry even once the
+% local genomic region has fully coalesced. Instead, simulating
+% ARGs from the coalescent with recombination is usually done by
+% running Hudson's algorithm to produce a sample-resolved ARG directly.
+% Moreover, when inferring ARGs from genetic data, there is no information
+% in the sample set that can be used to resolve these nodes, so that they
+% are essentially unknowable.
+
+The process of converting a gARG with local-inheritance annotations
+to its sample-resolved equivalent is also known as ``simplification'',
+and is the key algorithm underlying recent
+advances in forwards-in-time
+simulation~\citep{kelleher2018efficient,haller2018tree}.
+It has many useful properties, as we discuss in the next section.
 
@@
 
+% FIXME putting this here for now just to see if this will work.
+
+\section*{Event ARGs}\label{eARG}
+In keeping with its original derivation in terms of a stochastic
+process (see Appendix~\ref{sec-arg-history}), the classical definition of an
+ARG data structure is
+in terms of events.
+Nodes represent
+common ancestor or recombination events that occurred in the
+history of some set of sample genomes, and edges represent ancestral
+lineages~\citep{griffiths1996ancestral}.
+In a common ancestor event the inbound lineages are merged into a
+single ancestral lineage, and in a recombination
+event a lineage is split into two independent
+ancestral lineages.
+% The final vital
+% detail is that we associate a breakpoint with each recombination
+% event.
+% This information is sufficient to uniquely
+% define the local genealogical trees at every position along the genome,
+% which is the basic requirement for a data structure encoding
+% genome-wide genetic ancestry.
+As illustrated in Fig.~\ref{fig-arg-data-structure}, this information
+is usually concretely encoded by associating a recombination breakpoint
+(which is null for common ancestor events) and an ordered list of parents
+with each node. We refer to such a  data structure as an
+``event ARG'' (eARG).
+To recover the local tree at genome coordinate $x$ (the most
+fundamental operation required for an ARG data structure),
+we traverse the graph rootwards from the leaves.
+At a particular
+node $u$, if it has one parent we are at a common ancestor
+node and we follow that parent. If we are at a
+recombination node, $u$ has two parents: if
+$x$ is less than the breakpoint we follow
+the edge to the first parent, and otherwise follow the edge
+to the second parent.
+The order in which parents are listed at a recombination node is
+therefore significant, telling us
+from which parent the segments of genome to the left and right of the breakpoint
+were inherited.
+
+This ordering requirement, while straightforward
+to describe, has some practical drawbacks. For example,
+% Surely the meaning is clear from the context here and we don't need
+% qualify the meaning of ``ARG''
+when using this representation and simulating an ARG backwards in time,
+the first event older than a recombination may involve the lineage carrying the
+ancestry to the right of the breakpoint rather
+than the left, and so we cannot emit edges as they are generated.
+Such issues can be worked around, of course, but depending on the ordering of
+otherwise indistinguishable objects is generally problematic.
+
+\begin{figure}
+\centering
+% \begin{tabular}{cc}
+\begin{tikzpicture}[x=5mm, y=5mm, node distance=2mm and 20mm]
+\tikzset{greynode/.style={circle,fill,inner sep=1},
+nodelabel/.style={font=\footnotesize}}
+
+\node (s0) [greynode] at (0, 0) {};
+\node (s1) [greynode] at (3, 0) {};
+\node (s2) [greynode] at (6, 0) {};
+\node (s3) [greynode] at (3, 1) {};
+\node (s4) [greynode] at (1, 2) {};
+\node (s5) [greynode] at (5, 3) {};
+\node (s6) [greynode] at (3, 4) {};
+
+% \node [anchor=north west] at (0,6) {A};
+\node [nodelabel,anchor=north west] at ($(s3) + (0,0)$) {$x$};
+\foreach \u/\lab in {s0/$\textsf{a}$, s1/$\textsf{b}$, s2/$\textsf{c}$} \node[nodelabel,anchor=north] at (\u) {\lab};
+\foreach \u/\lab in {s4/$\textsf{e}$} \node[nodelabel,anchor=south west] at (\u) {\lab};
+\foreach \u/\lab in {s5/$\textsf{f}$} \node[nodelabel,anchor=south east] at (\u) {\lab};
+\foreach \u/\lab in {s3/$\textsf{d}$, s6/$\textsf{g}$} \node[nodelabel,anchor=south] at (\u) {\lab};
+
+%% Edges
+\draw (s1) -- (s3);
+\draw (s0) |- (s4);
+\draw (s4) -- (2,2) |- (s3);
+\draw (s4) |- (s6);
+\draw (s3) -- (4,1) |- (s5);
+\draw (s2) |- (s5);
+\draw (s5) |- (s6);
+
+% \node [anchor=north west] at (9,6) {B};
+\node [nodelabel,anchor=north west] at ($(10,5)$) {
+\begin{tabular}{c|c|l}
+% \multicolumn{2}{c}{Breakpoints}\\
+Node & Breakpoint & Parents\\
+\hline
+$\noderef{a}$ & $\varnothing$ & [\noderef{e}]\\
+$\noderef{b}$ & $\varnothing$ & [\noderef{d}] \\
+$\noderef{c}$ & $\varnothing$ & [\noderef{f}]\\
+$\noderef{d}$ & $x$ & [\noderef{e}, \noderef{f}] \\
+$\noderef{e}$ & $\varnothing$ & [\noderef{g}]\\
+$\noderef{f}$ & $\varnothing$ & [\noderef{g}]\\
+$\noderef{g}$ & $\varnothing$ & []\\
+\end{tabular}};
+
+% \node [anchor=north west] at (9,6) {C};
+% \node [nodelabel,anchor=north west] at ($(10,5)$) {
+% \begin{tabular}{ll}
+% Edges & Intervals\\
+% \hline
+% $(\textsf{a}, \textsf{e})$ & $(0, L]$ \\
+% $(\textsf{b}, \textsf{d})$ & $(0, L]$ \\
+% $(\textsf{c}, \textsf{f})$ & $(0, L]$ \\
+% $(\textsf{d}, \textsf{e})$ & $(0, x]$ \\
+% $(\textsf{d}, \textsf{f})$ & $(x, L]$ \\
+% $(\textsf{e}, \textsf{g})$ & $(0, L]$ \\
+% $(\textsf{f}, \textsf{g})$ & $(0, L]$ \\
+% \end{tabular}};
+
+\end{tikzpicture}
+\caption{\label{fig-arg-data-structure}
+A graphical depiction of a classical event ARG and a concrete
+encoding of the information. Node \noderef{d} is a recombination
+event as it is associated with a non-null breakpoint, and has
+two parents. Nodes \noderef{e}, \noderef{f} and \noderef{g} are
+common ancestor events.
+}
+\end{figure}
+
+% What do we mean by an eARG, formally?
+To be concrete, we can formally define the
+classical Griffiths eARG as a tuple $(e, \sigma)$, where $e$
+is an ordered list of edges defining a directed acyclic graph and
+$\sigma: \mathbb{N} \rightarrow \mathbb{N}$
+is a function mapping nodes to recombination breakpoints,
+such that $\sigma(u) = x$
+if $u$ is a recombination event with breakpoint $x$ and
+$\sigma(u) = \varnothing$ otherwise.
+(This is equivalent to having different ``types''
+of node, where breakpoints are associated with recombination
+nodes only).
+Each edge $e_j = (c, p)$ describes an edge between
+child node $c$ and parent $p$, where $c, p \in \mathbb{N}$.
+% For each recombination node $u$ we must
+% have exactly two edges in $e$ such that $u$ is the child.
+This encoding of an eARG is equivalent to the output
+of programs such as
+\texttt{ARGweaver}~\citep{rasmussen2014genome}
+and \texttt{KwARG}~\citep{ignatieva2021kwarg},
+and an example is given in Fig.~\ref{fig-arg-data-structure}A.
+Note that the description above captures only the
+graph topology: if we also wish to know the branch lengths we need
+an additional function $t(u): \mathbb{N} \rightarrow \mathbb{R}$
+which defines the time of each node.
+
 \subsection*{The Big and Little ARG}

### 2023-08-09 paper.tex

@@
 Applications using these inferred ancestries as inputs have begun to
-appear~\citep{osmond2021estimating,zhang2021biobank,fan2022genealogical,hejase2022deep}
+appear~\citep{osmond2021estimating,zhang2023biobank,fan2022genealogical,hejase2022deep}
 and many more are sure to
@@
 
+% NEEDS more work!
 \section*{Precision of ancestry information}
-% Not all nodes in an ARG are equal. Assuming that we have sequenced the
-% genomes of the samples, there are intrinsic limits on what can
-% be known about more ancient nodes, determined by the graph topology
-% and the flow of ancestral material through it.
-
 Ancestral recombination graphs are often characterised
@@
 
+% The full ancestral recombination graph (ARG) is a structure that encodes all
+% coalescence and recombination events resulting from the stochastic process of
+% the coalescent with recombination.
+\citet{brandt2021evaluation} define a ``full'' ARG as
+``a structure that encodes all
+coalescence and recombination events resulting from the stochastic process of
+the coalescent with recombination'' (thus going against the general
+historial trend discussed in the XXX section).
+\citet{rasmussen2022espalier} have a less restrictive definition, in
+which a ``full'' ARG is one that contains complete information about
+recombination events, without necessarily being tied to a particular
+stochastic process.
+While [something positive], it is important to be clear on
+what the overall goal of such precise inferences are,
+what the limitations on when they can be made are,
+and their downstream utility in applications.
+
+% What is the goal of having fully precise ARGs?
+% 1) studying the actual recombination events;
+% 2) evaluating the likelihoods
+The main reasons for wanting precise estimates of all recombination
+events are to either compute a likelihood of the inferred ARG
+under a statistical model, or to study the recombinants in
+detail~\citep{rasmussen2022espalier}.
+% Explain that we're really interested in the details of recombination
+% itself with a small number of samples.
+In the latter case, we might
+have a small number of samples~\citep{guo2022recombination}.
+
+Beyond cases in which visual inspection of the trees and the
+effects of recombination is feasible, the main application
+of having completely precise estimates of recombination events is
+to evaluate the likelihood of the estimate under some
+statistical model. The only statistical model available
+is the coalescent with recombination
+and its approximations~\citep{mcvean2005approximating,marjoram2006fast}.
+In this case, the utility of a ``full`` ARG for some dataset
+is limited by how well the dataset fits the assumptions
+of the model.
+
 % Not sure where this goes, but there are some critical points
@@
 % discussion.
-A critical distinction between the eARG and gARG encodings
-is the level of precision about the time and location of
-recombination events that is required. In an eARG, we are
-required to be absolutely precise about every recombination
-[explain and contrast with gARG, referring to examples]
-
-% % JK: This is very rough, just jotting down the basic ideas,
-% % will need substantial revision.
-% It is not clear that such complete information will always
-% (or indeed ever) be available from inferences. We may be
-% able to precisely identify the location and timing
-% of recent recombinations (particularly if detailed pedigree
-% information is available) but beyond this, the information
-% required simply isn't present. When sampling from a
-% process such as the Sequentially Markov
-% Coalescent~\citep{mcvean2005approximating,marjoram2006fast}
-% like ARGweaver~\citep{rasmussen2014genome}, it is true that
-% we do obtain a fully resolved ARG with precise
-% information about recombination, but it is reasonable
-% to question how much information from the data is used
-% to support older recombinations, in particular.
-
-% It is helpful to consider the difference between fully
-% resolved binary trees and those that contain polytomies.
-
-% The situation is in fact directly analogous: an ARG that
-% contains fully resolved recombination information must have
-% at most parents per node.
-
-% At the very least, it is not reasonable for the \emph{data
-% structure} to require complete precision in order to
-% represent some less precise. This is a fundamental
-% advantage of the gARG encoding over the classical eARG
-% method: we are not forced to posit a precise cause (recombination
-% event) for every effect (difference between adjacent trees).
-% These differences in precision of recombination
-% information are illustrated in the ARGs estimated by different
-% methods, as we see in the next section.
+% A critical distinction between the eARG and gARG encodings
+% is the level of precision about the time and location of
+% recombination events that is required. In an eARG, we are
+% required to be absolutely precise about every recombination
+% [explain and contrast with gARG, referring to examples]
+
+% JK: This is very rough, just jotting down the basic ideas,
+% will need substantial revision.
+It is not clear that such complete information will always
+(or indeed ever) be available from inferences. We may be
+able to precisely identify the location and timing
+of recent recombinations (particularly if detailed pedigree
+information is available) but beyond this, the information
+required simply isn't present. When sampling from a
+process such as the Sequentially Markov
+Coalescent~\citep{mcvean2005approximating,marjoram2006fast}
+like ARGweaver~\citep{rasmussen2014genome}, it is true that
+we do obtain a fully resolved ARG with precise
+information about recombination, but it is reasonable
+to question how much information from the data is used
+to support older recombinations, in particular.
+
+It is helpful to consider the difference between fully
+resolved binary trees and those that contain polytomies.
+
+The situation is in fact directly analogous: an ARG that
+contains fully resolved recombination information must have
+at most parents per node.
+
+At the very least, it is not reasonable for the \emph{data
+structure} to require complete precision in order to
+represent some less precise. This is a fundamental
+advantage of the gARG encoding over the classical eARG
+method: we are not forced to posit a precise cause (recombination
+event) for every effect (difference between adjacent trees).
+These differences in precision of recombination
+information are illustrated in the ARGs estimated by different
+methods, as we see in the next section.
 
@@
 
-To illustrate this critical point, we inferred ARGs for the
+To illustrate these points and to emphasise the heterogeneity
+of the ARGs inferred by current methods, we inferred ARGs for the
 the classical \citet{kreitman1983nucleotide} benchmark dataset
@@
 \texttt{tsinfer} explicitly constructs nodes that are shared across local
-trees.
+trees. See \cite{ignatieva2023distribution} for a detailed discussion
+and comparsion of inference methods on this property.
 
@@
 
-\section*{When do we need ``full'' ARGs?}
-% The full ancestral recombination graph (ARG) is a structure that encodes all
+% \section*{When do we need ``full'' ARGs?}
+% % The full ancestral recombination graph (ARG) is a structure that encodes all
+% % coalescence and recombination events resulting from the stochastic process of
+% % the coalescent with recombination.
+% \citet{brandt2021evaluation} define a ``full'' ARG as
+% ``a structure that encodes all
 % coalescence and recombination events resulting from the stochastic process of
-% the coalescent with recombination.
-\citet{brandt2021evaluation} define a ``full'' ARG as
-``a structure that encodes all
-coalescence and recombination events resulting from the stochastic process of
-the coalescent with recombination'' (thus going against the general
-historial trend discussed in the XXX section).
-\citet{rasmussen2022espalier} have a less restrictive definition, in
-which a ``full'' ARG is one that contains complete information about
-recombination events, without necessarily being tied to a particular
-stochastic process.
-While [something positive], it is important to be clear on
-what the overall goal of such precise inferences are,
-what the limitations on when they can be made are,
-and their downstream utility in applications.
-
-% What is the goal of having fully precise ARGs?
-% 1) studying the actual recombination events;
-% 2) evaluating the likelihoods
-The main reasons for wanting precise estimates of all recombination
-events are to either compute a likelihood of the inferred ARG
-under a statistical model, or to study the recombinants in
-detail~\citep{rasmussen2022espalier}.
-% Explain that we're really interested in the details of recombination
-% itself with a small number of samples.
-In the latter case, we might
-have a small number of samples~\citep{guo2022recombination}.
-
-Beyond cases in which visual inspection of the trees and the
-effects of recombination is feasible, the main application
-of having completely precise estimates of recombination events is
-to evaluate the likelihood of the estimate under some
-statistical model. The only statistical model available
-is the coalescent with recombination
-and its approximations~\citep{mcvean2005approximating,marjoram2006fast}.
-In this case, the utility of a ``full`` ARG for some dataset
-is limited by how well the dataset fits the assumptions
-of the model.
-
-% If we want to use the coalescent, then really the assumptions
-% need to make sense.
-% n << Ne is a silly assumption in today's datasets
-A core assumption of the coalescent is that the sample size $n$
-is much less than the effective population size, $N_e$.
-Several human datasets now consist of hundreds of thousands of
-genomes~\citep{bycroft2018genome,karczewski2020mutational,tanjo2021practical},
-and so sample size is substantially \emph{larger} than $N_e$
-(often assumed to be $10^4$ in humans).
-Agricultural datasets are even more extreme.
-For example,
-the US dairy cattle database alone currently comprises more than 6 million
-animals with SNP array
-genotypes\footnote{\url{https://queries.uscdcb.com/Genotype/counts.html}},
-while the effective population size in modern dairy cattle breeds is
-less than 100 and decreasing~\citep{MacLeod2013,Makanjuola2020}.
-An extreme sign of these breeding practices is that there are only two ancestral
-Y-chromosome lineages present in today's US Holstein dairy breed~\citep{Yue2015}.
-Simlarly, the 1000 bull genomes project~\citep{hayes20191000}
-comprises close to 7000 genomes, which are part of multi-generation pedigrees
-with millions of animals and extensive SNP array genotype and phenotype
-data \citep[e.g.][]{Cesarani2022}.
-In another example, genome and SNP array data for 440,610 individuals within
-7 multi-generation pedigrees~\citep{whalen2018,Johnsson2021,Ros-Freixedes2020}
-were combined to infer recombination in pigs~\citep{RosFreixedes2022}.
-Effective population size in modern pig breeds is also less than 100 due to
-intense selection and directed reproduction \citep{Hall2016,Porcnic2016}.
-Another core assumption of the coalescent model is that the genome (or
-at least the region under study) is short enough that the number of extant
-lineages remains much smaller than $N_e$ at all times. Whole
-genome sequences have been available for model organisms
-for over a decade now, % True? Citation?
-and indeed complete chromosome-level assemblies are possible
-in humans~\citep{miga2020telomere}.
-Projects are under way to obtain high-quality assemblies
-for all eukaryotic species in Britain and Ireland~\citep{darwin2022sequence}
-and ultimately worldwide~\citep{lewin2022earth}.
-
-While the coalescent as a model can be suprisingly robust to
-basic violations of the underlying assumptions~\citep{
-wakeley2012gene,bhaskar2014distortion,nelson2020accounting}
-it is reasonable to question the validity of the model in
-representing datasets such as UK Biobank.
-[Also mention somehow that $n\ll N_e$ is only one problem -
-we're also ignoring the myriad other evolutionary forces.]
-Such datasets cannot be reasonably be described by
-the coalescent as a statical model, and so more non-parametric
-methods like \tsinfer\ and \relate\ would seem more appropriate.
-[More]
-
-% Is it practical to estimate full ARGs?
-% [Not particularly \citep{shipilina2022origin} - but maybe ARGneedle
-% has done this for UKB? Need to check]
-
-% Do downstream applications actually make use of such full ARGs?
-Downstream applications using estimated ARGs to date rarely
-make use of such complete recombination information.
-The \relate\ selection statistic~\citep{speidel2019method}
-computes a $p$-value for selection [FIXME] for each
-local tree independently by comparing the rates of coalescence
-at various time-points. For cases like strong positive selection,
-where we expect there to be a clear signal in the local tree,
-such tree-by-tree methods can be simple and powerful.
-In their method
-for estimating dispersal rates and the locations of genetic
-ancestors,
-\citep{osmond2021estimating} downsample trees along the genome
-so that they can be regarded as approximately independent.
-The SIA method for detecting selection~\citep{hejase2022deep}
-encodes local trees as a set of lineage counts at discrete
-time intervals, and uses these as feature for a
-type of machine learning algorithm
-that takes ``temporal'' correlations into account. Thus,
-while SIA does use some of the information about local tree correlation,
-clearly much of the detail about recombination
-events in a ``full'' ARG is lost.
-The \texttt{tsdate} algorithm (and related naive estimator
-for ancestral location) uses much more between-tree
-information~\citep{wohns2022unified}, explicitly using the gARG
-encoding to reason about ancestral nodes.
-A major strength of this approach is that it can be used on
-any gARG, ranging from estimates containing relatively little
-information about recombination (e.g., as estimated by \relate)
-to ``full'' ARGs containing complete recombination information
-(e.g., as estimated by \kwarg), and will fully utilise
-any between-tree correlation information that is present.
-
-\section*{Computational efficiency}
+% the coalescent with recombination'' (thus going against the general
+% historial trend discussed in the XXX section).
+% \citet{rasmussen2022espalier} have a less restrictive definition, in
+% which a ``full'' ARG is one that contains complete information about
+% recombination events, without necessarily being tied to a particular
+% stochastic process.
+% While [something positive], it is important to be clear on
+% what the overall goal of such precise inferences are,
+% what the limitations on when they can be made are,
+% and their downstream utility in applications.
+
+% % What is the goal of having fully precise ARGs?
+% % 1) studying the actual recombination events;
+% % 2) evaluating the likelihoods
+% The main reasons for wanting precise estimates of all recombination
+% events are to either compute a likelihood of the inferred ARG
+% under a statistical model, or to study the recombinants in
+% detail~\citep{rasmussen2022espalier}.
+% % Explain that we're really interested in the details of recombination
+% % itself with a small number of samples.
+% In the latter case, we might
+% have a small number of samples~\citep{guo2022recombination}.
+
+% Beyond cases in which visual inspection of the trees and the
+% effects of recombination is feasible, the main application
+% of having completely precise estimates of recombination events is
+% to evaluate the likelihood of the estimate under some
+% statistical model. The only statistical model available
+% is the coalescent with recombination
+% and its approximations~\citep{mcvean2005approximating,marjoram2006fast}.
+% In this case, the utility of a ``full`` ARG for some dataset
+% is limited by how well the dataset fits the assumptions
+% of the model.
+
+% % If we want to use the coalescent, then really the assumptions
+% % need to make sense.
+% % n << Ne is a silly assumption in today's datasets
+% A core assumption of the coalescent is that the sample size $n$
+% is much less than the effective population size, $N_e$.
+% Several human datasets now consist of hundreds of thousands of
+% genomes~\citep{bycroft2018genome,karczewski2020mutational,tanjo2021practical},
+% and so sample size is substantially \emph{larger} than $N_e$
+% (often assumed to be $10^4$ in humans).
+% Agricultural datasets are even more extreme.
+% For example,
+% the US dairy cattle database alone currently comprises more than 6 million
+% animals with SNP array
+% genotypes\footnote{\url{https://queries.uscdcb.com/Genotype/counts.html}},
+% while the effective population size in modern dairy cattle breeds is
+% less than 100 and decreasing~\citep{MacLeod2013,Makanjuola2020}.
+% An extreme sign of these breeding practices is that there are only two ancestral
+% Y-chromosome lineages present in today's US Holstein dairy breed~\citep{Yue2015}.
+% Simlarly, the 1000 bull genomes project~\citep{hayes20191000}
+% comprises close to 7000 genomes, which are part of multi-generation pedigrees
+% with millions of animals and extensive SNP array genotype and phenotype
+% data \citep[e.g.][]{Cesarani2022}.
+% In another example, genome and SNP array data for 440,610 individuals within
+% 7 multi-generation pedigrees~\citep{whalen2018,Johnsson2021,Ros-Freixedes2020}
+% were combined to infer recombination in pigs~\citep{RosFreixedes2022}.
+% Effective population size in modern pig breeds is also less than 100 due to
+% intense selection and directed reproduction \citep{Hall2016,Porcnic2016}.
+% Another core assumption of the coalescent model is that the genome (or
+% at least the region under study) is short enough that the number of extant
+% lineages remains much smaller than $N_e$ at all times. Whole
+% genome sequences have been available for model organisms
+% for over a decade now, % True? Citation?
+% and indeed complete chromosome-level assemblies are possible
+% in humans~\citep{miga2020telomere}.
+% Projects are under way to obtain high-quality assemblies
+% for all eukaryotic species in Britain and Ireland~\citep{darwin2022sequence}
+% and ultimately worldwide~\citep{lewin2022earth}.
+
+% While the coalescent as a model can be suprisingly robust to
+% basic violations of the underlying assumptions~\citep{
+% wakeley2012gene,bhaskar2014distortion,nelson2020accounting}
+% it is reasonable to question the validity of the model in
+% representing datasets such as UK Biobank.
+% [Also mention somehow that $n\ll N_e$ is only one problem -
+% we're also ignoring the myriad other evolutionary forces.]
+% Such datasets cannot be reasonably be described by
+% the coalescent as a statical model, and so more non-parametric
+% methods like \tsinfer\ and \relate\ would seem more appropriate.
+% [More]
+
+% % Is it practical to estimate full ARGs?
+% % [Not particularly \citep{shipilina2022origin} - but maybe ARGneedle
+% % has done this for UKB? Need to check]
+
+% % Do downstream applications actually make use of such full ARGs?
+% Downstream applications using estimated ARGs to date rarely
+% make use of such complete recombination information.
+% The \relate\ selection statistic~\citep{speidel2019method}
+% computes a $p$-value for selection [FIXME] for each
+% local tree independently by comparing the rates of coalescence
+% at various time-points. For cases like strong positive selection,
+% where we expect there to be a clear signal in the local tree,
+% such tree-by-tree methods can be simple and powerful.
+% In their method
+% for estimating dispersal rates and the locations of genetic
+% ancestors,
+% \citep{osmond2021estimating} downsample trees along the genome
+% so that they can be regarded as approximately independent.
+% The SIA method for detecting selection~\citep{hejase2022deep}
+% encodes local trees as a set of lineage counts at discrete
+% time intervals, and uses these as feature for a
+% type of machine learning algorithm
+% that takes ``temporal'' correlations into account. Thus,
+% while SIA does use some of the information about local tree correlation,
+% clearly much of the detail about recombination
+% events in a ``full'' ARG is lost.
+% The \texttt{tsdate} algorithm (and related naive estimator
+% for ancestral location) uses much more between-tree
+% information~\citep{wohns2022unified}, explicitly using the gARG
+% encoding to reason about ancestral nodes.
+% A major strength of this approach is that it can be used on
+% any gARG, ranging from estimates containing relatively little
+% information about recombination (e.g., as estimated by \relate)
+% to ``full'' ARGs containing complete recombination information
+% (e.g., as estimated by \kwarg), and will fully utilise
+% any between-tree correlation information that is present.
+
+\section*{Computational efficiency and data interchange}
+ARG inference methods have been successfully applied to vast datasets,
+including 500,000 genotyped humans~\citep{kelleher2019inferring,zhang2023biobank}
+from UK Biobank~\citep{bycroft2018genome}
+and over a million SARS-Cov-2 whole genomes~\citep{zhan2023towards}.
+Even larger datasets are
+available~\citep[e.g.][]{halldorsson2022sequences} % TODO more
+and it seems inevitable that ARGs will be inferred for them.
+The efficiency with which we can store and analyse these inferred
+ARGs is therefore of great importantance, and a key factor
+in the ultimate success of downstream ARG-based methods.
+Another key factor in the how important ARG-based methods become
+across the wide range of predicted application
+areas [CITE commentary pieces, tsinfer, ARGneedle any others?]
+is how well inference methods and downstream analysis applications
+interoperate. [Genetics doesn't have a good record here, let's
+try not to make the same mistakes again.]
+
+The gARG encoding discussed and concretely defined in this manuscript
+leads to highly efficient storage and processing of ARG data,
+and has already been in use for several years.
 The succinct tree sequence data structure (usually known as a ``tree sequence''
-for brevity, but see later comment on confusion around this point)
+for brevity, but see below for confusion around this point)
 is a concrete encoding of a gARG, as discussed in this article.
 It was originally developed as part of the \texttt{msprime}
-simulator~\cite{kelleher2016efficient} and subsequently ...
-
-[The tree sequence encoding is very efficient for lots of things. Examples]
-
-In retrospect, the choice of terminology was unfortunate, and has led to
-significant confusion. [More]
+simulator~\cite{kelleher2016efficient} and subsequently been
+extended and applied to forward-time
+simulations~\citep{kelleher2018efficient,haller2018tree},
+calculation of population genetics statistics~\citep{ralph2020efficiently}.
+The succinct tree sequence encoding extends the basic definition
+of a gARG provided here by stipulating a
+simple tabular representation and nodes and edges,
+and also defining a concise and lossless representation of
+sequence variation using the ``site'' and  ``mutation'' tables.
+
+The \texttt{tskit} library~\citep{tskit2023tskit} is a liberally
+licensed open source toolkit that provides a comprehensive suite
+of tools for working with gARGs (encoded as a succinct tree sequence).
+Based on core functionality written
+in C, it provides interfaces in C, Python and Rust.
+Tskit is mature software, widely used in population genetics, and
+has been incorporated into several downstream
+applications~\citep[e.g.,][]{haller2019slim,speidel2019method,
+adrion2020community,
+terasaki2021geonomics,
+baumdicker2021efficient,
+fan2022genealogical,korfmann2022weak,
+mahmoudi2022bayesian,petr2022slendr,rasmussen2022espalier,
+zhang2023biobank}.
+
+In retrospect, the choice of terminology around the succinct
+tree sequence data structure was unfortunate, and has led to
+significant confusion. [Discuss, and finish on a positive note]
 
@@
 parida2008estimating, kelleher2019inferring,  speidel2019method,
-zhang2021biobank}. The latter three methods, as well as the parsimony-based
+zhang2023biobank}. The latter three methods, as well as the parsimony-based
 method of \citet{rasmussen2022espalier}, can be used on megabase scale data and

### 2023-08-10 paper.tex

@@
 in the ultimate success of downstream ARG-based methods.
-Another key factor in the how important ARG-based methods become
+Another key factor in how important ARG-based methods become
 across the wide range of predicted application
-areas [CITE commentary pieces, tsinfer, ARGneedle any others?]
+% TODO more?
+areas~\citep{rasmussen2014genome,harris2019database,
+hejase2020summary,schaefer2021ancestral,
+fan2022genealogical,wohns2022unified,harris2023using}
 is how well inference methods and downstream analysis applications
-interoperate. [Genetics doesn't have a good record here, let's
-try not to make the same mistakes again.]
+interoperate. Without a well-defined and efficient shared format
+for ARG data interchange,
+users will suffer from problems of converting data from one programs
+output to another, an unfortunate hallmark of population genetics
+software for many years~\citep{excoffier2006computer}.
+While current methods tend to tightly couple downstream analysis
+of the inferred ARG with the inference
+itself within the same software package
+% FIXME-  more explicit?
+(e.g.\ Relate, ARGneedle)
+this approach scales poorly in terms of software development
+effort, and is ultimately not compatible with the widespread use
+of ARGs for routine data analysis, and a healthy and diverse
+software ecosystem.
 
@@
 It was originally developed as part of the \texttt{msprime}
-simulator~\cite{kelleher2016efficient} and subsequently been
+simulator~\citep{kelleher2016efficient} and subsequently been
 extended and applied to forward-time
@@
 sequence variation using the ``site'' and  ``mutation'' tables.
-
 The \texttt{tskit} library~\citep{tskit2023tskit} is a liberally
@@
 tree sequence data structure was unfortunate, and has led to
-significant confusion. [Discuss, and finish on a positive note]
+significant confusion. As first introduced by~\cite{kelleher2016efficient}
+a ``tree sequence`` was defined as the set of ``coalescence records``
+that is output by Hudson's simulation algorithm.
+This was then generalised to use the
+tabular encoding described above and formally described as a ``succinct
+tree sequence''~\citep{kelleher2018efficient}.
+The ``succinct'' prefix here is intended to connect
+with the concept of succinct data structures
+\citep[e.g.][]{gog2014theory},
+which have near-optimal
+space usage but still support efficient retrieval and computation.
+The concrete data structures of the succinct tree sequence, precisely
+defined and with a particular focus on computational efficiency
+did not have an obvious connection to the ARG definitions
+provided by Griffiths and
+colleagues~\citep{griffiths1991two,ethier1990two,
+griffiths1996ancestral,griffiths1997ancestral}, and with the lack
+of any other precise definitions, it seemed best to avoid the
+term to prevent confusion. Unfortunately, however, the opposite
+has occured. In describing the output of \tsinfer\ as a ``tree
+sequence''~\citep{kelleher2019inferring} many
+have---not unreasonably, but incorrectly---concluded that the
+output is a collection of \emph{independent}
+% MORE?
+% https://github.com/tskit-dev/what-is-an-arg-paper/issues/38
+trees~\citep[e.g.,][]{hejase2020summary,ignatieva2021kwarg}.
+% Could say more here, but maybe we'll have a note about terminology
+% in the discussion
 

### 2023-08-10 paper.tex
Whittle down discussion items

@@
 between genomes (and not the causal \emph{processes}),
-arbitrary patterns of genetic inheritance such as gene conversions,
-multiple recombinations and so on can be directly expressed.
+arbitrary patterns of genetic inheritance such as
+gene conversions,
+multiple recombinations
+across different chromosomes~\citep{fearnhead2003ancestral,koskela2019robust},
+and so on can be directly expressed.
+
 Thirdly, ``local`` trees that we generate for a particular position on the
@@
 ``ancestral material''~\citep{wiuf1999ancestry,wiuf1999recombination}
-is usually much more useful and, indeed, all that is possible to
+is much more useful~\citep{mcgill2013graphml,arenas2013importance}
+and, indeed, all that is possible to
 have information about when inferring ARGs from contemporary samples.
@@
 ``sample resolved''.
+% Keeping track of resolved ancestral material is one of the distinguishing
+% features of ARGs vs phylogenetic
+% networks~\citep{mcgill2013graphml,arenas2013importance}, but
+% is actually incompatible with the Griffiths gARG encoding (it does this
+% implicitly, I guess)
 
@@
 software ecosystem.
+% FIXME not sure where this goes
+Earlier efforts to standardise ARG interchange
+did not succeed~\citep{cardona2008extended,mcgill2013graphml}.
 
@@
 
+% FIXME this needs more work and doesn't really fit into the
+% narrative right now, but we do need to make this point.
+The key insight that makes the succinct tree sequence encoding
+an efficient substrate for defining analysis algorithms is that
+it allows us to generate the local trees along the genome
+in a way that allows us to reason about the \emph{differences}
+between those trees.
+Sequentially generating the marginal trees along the genome
+is also fundamental, and is necessary whenever we need to
+perform calculations that are contingent on more than just the
+isolated properties of an edge. \cite{kelleher2016efficient}
+showed how all trees can be sequentially generated in
+constant time per tree transition in a fully simplified gARG.
+Furthermore, we can easily reason about how tree topologies
+change (and stay the same), leading to efficient algorithms
+for computing population genetic
+statistics~\citep{kelleher2016efficient,kelleher2018efficient},
+implementing the Li and Stephens
+model~\citep{kelleher2019inferring,wohns2022unified}
+and likely many more.
+
 In retrospect, the choice of terminology around the succinct
@@
 next point also.)
+
 \item Terminology based around an inferred ancestry being a ``true''
@@
 discuss and compare.
-\item tskit can represent any ARG.
-\item Interchange is a real problem, and if we're going to realise the
-potential for ARGs in genomics we must solve it. Efforts to standardise
-phylogenetic networks haven't done very well (~\citep{cardona2008extended}
-has 89 citations). The only real effort to standardise an ARG format
-based on extending GraphML~\citep{mcgill2013graphml} has 11 citations.
-GraphML~\citep{mcgill2013graphml} has some of the same key ideas
-as tskit though, in particular allowing the annotation of ``live sites''
-to an edge as a set of disjoint intervals that hold ancestral material.
-
-\item There are real computational disadvantages to the Griffiths ARG
-data structure. Having two types of node (three really, as we need
-``sample'' nodes which are also treated differently) complicates algorithms,
-and makes it difficult to reason about the structure of the ARG as a
-graph. For example, there's some weirdness about mutations over
-recombination nodes discussed by Gusfield [Ana?]
 
@@
 
-\item Throughout, we considered ARGs defined for a single chromosome comprising
-of a fixed number of discrete sites. Our argument applies equally well to ARGs
-defined for several non-homologous chromosomes \citep{fearnhead2003ancestral,
-koskela2019robust} and to idealised, continuous chromosomes.
-
-\item The gARG structure as described here is sufficiently expressive
-to model any form of homologous genetic transmission between individuals,
-and is far more flexible than the classical eARG approach.
-Phylogenetic networks~\citep{huson2010phylogenetic} are a more
-general concept, and specific classes capture dynamics that
-cannot be modelled using an ARG. For example,
-reassortment networks model the exchange of
-(non-recombining) genomic segments between viral particles inside a co-infected host,
-and hybridisation networks, model the genetic material from
-different species combining to create a hybrid.
-Similarly Genome Graphs~\cite{novak2017genome} model complicated things
-within species. Future work should look to capture all of these things
-in a single structure.
-
 \end{itemize}
-
-% TODO not clear this belongs in the discussion, but we'll definitely
-% want to make some of these points, even if they are being repeated
-% from earlier section.
-It is difficult to imagine a practical calculation performed on an
-ARG that does not involve either the sequential generation of trees
-along the genome or the resolved ancestry segments.
-Knowing the resolved segments in an ancestry resolved gARG allows
-us to perform many calculations without explicitly generating
-the trees, and lets us work with the ancestry on a genome-by-genome
-basis. We give three examples here. Firstly, generating mutations
-on edges can be performed very quickly on an ancestry resolved
-gARG, since we only need to examine $O(|E|)$ edges rather than
-$O(nt)$ tree branches (assuming $n$ samples and $t$ marginal trees)
-~\citep{baumdicker2021efficient}. If the ancestry were not resolved,
-we would generate a large number of mutations on areas of the edge that
-are non-ancestral. Secondly, computing the likelihood of a given
-ARG realisation under the coalescent with recombination model
-requires that we essentially recapitulate the state of Hudson's
-algorithm (see XXX section) at each node in
-the graph~\citep{baumdicker2021efficient}. Having resolved ancestry
-substantially simplifies this process. Thirdly, computing
-[FIXME describe what we did in tsdate]~\citep{wohns2022unified}.
-
-Sequentially generating the marginal trees along the genome
-is also fundamental, and is necessary whenever we need to
-perform calculations that are contingent on more than just the
-isolated properties of an edge. \cite{kelleher2016efficient}
-showed how all trees can be sequentially generated in
-constant time per tree transition in a fully simplified gARG.
-Furthermore, we can easily reason about how tree topologies
-change (and stay the same), leading to efficient algorithms
-for computing population genetic
-statistics~\citep{kelleher2016efficient,kelleher2018efficient},
-implementing the Li and Stephens
-model~\citep{kelleher2019inferring,wohns2022unified}
-and likely many more.
 

### 2023-08-10 paper.tex
More fleshing out of discussion

@@
 recently introduced succinct tree sequence data structure.
+
+\item This, coupled with
+    the explicit association of genome coordinates with edges, allows
+    us to represent even the bewilderingly complex patterns of relatedness
+    present in modern datasets.
+
 \end{enumerate}
@@
 It is not clear that such complete information will always
-(or indeed ever) be available from inferences. We may be
+(or indeed ever) be available from inferences~\cite{shipilina2022origin}.
+ We may be
 able to precisely identify the location and timing
@@
 
+% This idea of a continuum of information about recombination
+% being inferrable and representable is consistent with the ideas
+% discussed by~\cite{shipilina2022origin}. There a haplotype block
+% is synonymous with coalescent nodes in the ARG, which we can generalise
+% as nodes in a gARG with at least two children. In principle, these
+% nodes are identifiable by the mutations that occur on the outbound
+% edges of the node.
+
 It is helpful to consider the difference between fully
@@
 methods, as we see in the next section.
+
+In general, discussions tend to contrast an ARG
+on one hand as holding \emph{complete} information about recombination
+(and therefore between-tree correlation structure)
+and a sequence of local trees on the other (with, by implication,
+no direct information about between-tree correlations).
+This is a false dichotomy:
+a \emph{continuum} of information about recombination and between-tree
+correlation structure can be represented and inferred, and all
+such levels of information are useful, even if only for computational
+efficiency.
 
@@
 \section*{Discussion}
-
-Discussion points (may or may not be used, certainly would be heavily revised)
-
-\begin{itemize}
-
-\item Recap quickly distinction bewtween aARG and gARG.Then:
-    This, coupled with
-    the explicit association of genome coordinates with edges, allows
-    us to represent even the bewilderingly complex patterns of relatedness
-    present in modern datasets.
-
-\item In general, discussions tend to contrast an ARG
-on one hand as holding \emph{complete} information about recombination
-(and therefore between-tree correlation structure)
-and a sequence of local trees on the other (with, by implication,
-no direct information about between-tree correlations).
-A key goal of this article is to show that this is a false dichotomy:
-a \emph{continuum} of information about recombination and between-tree
-correlation structure can be represented and inferred, and all
-such levels of information are useful, even if only for computational
-efficiency.
-
-\item This idea of a continuum of information about recombination
-being inferrable and representable is consistent with the ideas
-discussed by~\cite{shipilina2022origin}. There a haplotype block
-is synonymous with coalescent nodes in the ARG, which we can generalise
-as nodes in a gARG with at least two children. In principle, these
-nodes are identifiable by the mutations that occur on the outbound
-edges of the node.
-Their definition of an ARG is interesting, as they emphasise that it
-is composed of events, but illustate the genomes. [Nice to mention
-this connection somehere if we can, see Fig B1]
-
-\item It is unfortunate that we are adding yet another meaning to the
-term ARG here, but all we're really doing is making the implicit
-assumptions that others have made (i.e., describing the output of
-Relate, tsinfer etc as ARGs) and making them concrete. We need to have a
-term to describe all of these things, and since people are already using
-ARG for this, we might as well make it rigorous. We can then move away
-from the unfruitful discussion about which methods infer a ``true'' or
-``real'' ARG (none do, really, under the original definitions) and
-instead start discussing the \emph{properties} of these ARGs. (See
-next point also.)
-
-\item Terminology based around an inferred ancestry being a ``true''
-ARG is impoverished and misleading. The only reasonable interpretation
-of a ARG being true or not is whether it's a sample from the coalescent.
-Most inference methods (and all that scale well) do not. In any case,
-the coalescent is just one model, and a highly idealised one. It is
-exceedingly unlikely that the actual ARG describing the ancestry
-of any biological population is ``true'' ARG in this sense. The terminology is
-impoverished because it reduces the possibilities available
-to something either being an ARG or not, whereas there are many different
-interesting properties of these structures that we should be able to
-discuss and compare.
-
-\item Comparison metrics are a problem. All evaluations has essentially
-looked at either some form of pairwise comparison of trees along the genome
-using tree metrics as proposed by \citet{kuhner2015assessing},
-or looked at the distributions of TMRCAs. No
-one has succeeded in performing any comparisons of ARGs beyond
-regarding them as a sequence of independent trees. If we genuinely believe
-that ARGs are more than this, then we need to devise some metrics that
-consider more than the local tree structure.
-
-\end{itemize}
+Recent breakthroughs have made ARG inference methods practical
+for the first time, and there has been a surge of interest
+in inference methods, their evaluation and application.
+The prospect of ARGs being used routinely in population
+and statistical genetics applications is tantalising,
+but in reality there is  substantial work to be done to
+enable this, and much remains to be done.
+
+A necessary first step towards this goal is some terminological
+clarity. As discussed in Appendix XXX, the term ARG has several
+subtly different interpretations depending on context. The
+trend to decouple the ARG from its origins as a stochastic
+process and instead use the term as a more general representation of any
+recombinant genetic ancestry seems most useful, which we have
+tried to clarify and systematise here. Thus
+we can think of an ARG as being any structure that encodes the
+reticulate genetic ancestry of a sample of colinear sequences under
+the influence of recombination. The ``genome'' ARG encoding
+made explicit here is one way we can concretely
+define such recombinant ancestry, which we have shown is both
+flexible and efficient.
+
+The flexiblity of the gARG encoding contrasts with the classical
+``event'' ARG (eARG), which, because of its origins in an idealised
+mathematical model, is more limited in what can be described.
+The necessity of complete precision under the eARG encoding has been
+conflated with the idea of a ``true'' or ``full'' ARG.
+[A discussion of how realistic inferring a completely precise ARG is]
+[Do downstream applications use full ARGs? usually no]
+
+% % Do downstream applications actually make use of such full ARGs?
+% Downstream applications using estimated ARGs to date rarely
+% make use of such complete recombination information.
+% The \relate\ selection statistic~\citep{speidel2019method}
+% computes a $p$-value for selection [FIXME] for each
+% local tree independently by comparing the rates of coalescence
+% at various time-points. For cases like strong positive selection,
+% where we expect there to be a clear signal in the local tree,
+% such tree-by-tree methods can be simple and powerful.
+% In their method
+% for estimating dispersal rates and the locations of genetic
+% ancestors,
+% \citep{osmond2021estimating} downsample trees along the genome
+% so that they can be regarded as approximately independent.
+% The SIA method for detecting selection~\citep{hejase2022deep}
+% encodes local trees as a set of lineage counts at discrete
+% time intervals, and uses these as feature for a
+% type of machine learning algorithm
+% that takes ``temporal'' correlations into account. Thus,
+% while SIA does use some of the information about local tree correlation,
+% clearly much of the detail about recombination
+% events in a ``full'' ARG is lost.
+% The \texttt{tsdate} algorithm (and related naive estimator
+% for ancestral location) uses much more between-tree
+% information~\citep{wohns2022unified}, explicitly using the gARG
+% encoding to reason about ancestral nodes.
+% A major strength of this approach is that it can be used on
+% any gARG, ranging from estimates containing relatively little
+% information about recombination (e.g., as estimated by \relate)
+% to ``full'' ARGs containing complete recombination information
+% (e.g., as estimated by \kwarg), and will fully utilise
+% any between-tree correlation information that is present.
+
+Fully decoupling the general concept of an ARG from the coalescent
+with recombination stochastic process is an important step.
+While the coalescent with recombination is a very useful and
+robust model~\citep{wakeley2012gene,bhaskar2014distortion,nelson2020accounting}
+many modern datasets have properties that grossly
+violate its assumptions.
+For example, several human datasets now consist of hundreds of thousands of
+genomes~\citep{bycroft2018genome,karczewski2020mutational,tanjo2021practical},
+substantially \emph{larger} than the usually assumed $N_e$ values.
+Agricultural datasets are an even more extreme departure from coalescent
+assumptions, with large samples embedded in
+multi-generational pedigrees~\citep{hayes20191000,Ros-Freixedes2020}
+and effective population sizes of 100 and even
+less~\citep{MacLeod2013,Makanjuola2020,Hall2016,Porcnic2016}.
+Finally, the COVID-19 pandemic has provided an example of a dataset that
+is clearly a very poor fit for the coalescent with recombination, but which,
+nevertheless, we would like to estimate ARGs.
+\cite{zhan2023towards} inferred an ARG for 1.27 million SARS-Cov-2 whole
+genomes, and showed how it accurately captures several known properties of
+the pathogen's evolution, and can efficiently and concisely
+encode this vast dataset.
+
+This view of ARGs,
+decoupled from generative models and
+without the hard requirement
+of complete precision on all historical events, may help improve
+methods for inference method evaluation. In the majority of cases,
+ARG inference accuracy is evaluated by comparing some truth simulations
+to the inferred ARGs by the pairwise comparison of local trees along the genome
+using tree metrics as originally proposed by \citet{kuhner2015assessing}.
+While this provides valuable insights into the method's performance,
+it is not clear how well these tree distance metrics reflect
+performance on real data.
+In comparing tree-by-tree along the genome, the effects of recombination
+structure is incorporated in a very indirect manner.
+Tree distance metrics usually have $O(n^2)$ time complexity, and so cannot be
+applied to the very large sample sizes currently of interest.
+\citet{deng2021distribution} derived the distribution of the expected
+waiting distances to tree changes.
+[ \citet{ignatieva2023distribution} DID SOMETHING ELSE]
+\citet{brandt2021evaluation} compared the
+distributions of time to the most recent common ancestor across several
+methods, which is a useful addition but only indirectly captures
+topological comparisons.
+% Not sure this is the right place for this sentence
+All of these comparisons and evaluations have been based on
+highly idealised simulations, and the effects richness of real data
+(beyond simplistic error models) are almost entirely unknown.
+A suite of true ARG distance metrics, that take into account more of the
+global topology and can be applied to large ARGs, would be a
+valuable addition to the field.
 

### 2023-08-10 paper.tex
Splice back in minimal eARG section

@@
 
-\section*{Inheritance and ancestral material}\label{Ancestry_resolution}
-The edges in a gARG represent a path of genetic inheritance from
-ancestor to descendant through some
-number of generations (ultimately from cell to cell). The edges are ``annotated''
-with the genome coordinates over which inheritance occurs.
-There are two different ways in which we use these inheritance
-annotations, and the distinction is subtle and important. The difference
-is illustrated in Fig~\ref{fig-ancestry-resolution}.
-
-% TODO majorly condense this caption, it's much too long.
+\section*{Event ARGs}\label{eARG}
+We are mainly interested in exploring the properties of gARGs in
+this manuscript, but in order to contrast with the existing
+methods we concretely define an ``event ARG'', or eARG.
+
 \begin{figure}
 \centering
-\includegraphics[width=\textwidth]{illustrations/ancestry-resolution}
-\caption{\label{fig-ancestry-resolution}
-TODO DELETE A? Or keep it? Maybe this would help some hard-core eARGers?
-The \citet[][Fig.~1]{wiuf1999recombination} example graph as an eARG and
-two gARGs with alternative edge annotation schemes.
-% (A) A conventional Griffiths eARG. Square nodes represent events, with
-% each recombination (red) annotated with the position of a single breakpoint
-% along the 7-unit-long chromosome.
-(A) A gARG with edges annotated by . Oval nodes represent
-genomes and annotations reflect ``local'' transmission of all genetic material
-between parent and child. Edges that do not span the full genome are visually
-highlighted, showing how breakpoint positions can be encoded.
-(C) A gARG with sample-resolved edge annotations. The Hudson-like simplification
-algorithm has been used to shorten edge spans to
-reflect only the upwards transmission of genetic material from the samples
-\textsf{a}, \textsf{b}, and \textsf{c}; very few edges then span the entire genome.
-The depiction of the genome within each
-node is shaded as in Fig.~\ref{fig-arg-in-pedigree}A to show coalescing
-regions of genome back to the MRCA nodes \textsf{k} (regions 0 to 2) and \textsf{p}
-(regions 2 to 7). Line widths reflect the genomic span of each edge, highlighting
-the relative importance of different transmission pathways to the samples. Nodes
-\textsf{g},  \textsf{j}, \textsf{o} and \textsf{q} are redundant in that they correspond
-to events which have no impact on the genetic ancestry of the samples.
+% \begin{tabular}{cc}
+\begin{tikzpicture}[x=5mm, y=5mm, node distance=2mm and 20mm]
+\tikzset{greynode/.style={circle,fill,inner sep=1},
+nodelabel/.style={font=\footnotesize}}
+
+\node (s0) [greynode] at (0, 0) {};
+\node (s1) [greynode] at (3, 0) {};
+\node (s2) [greynode] at (6, 0) {};
+\node (s3) [greynode] at (3, 1) {};
+\node (s4) [greynode] at (1, 2) {};
+\node (s5) [greynode] at (5, 3) {};
+\node (s6) [greynode] at (3, 4) {};
+
+% \node [anchor=north west] at (0,6) {A};
+\node [nodelabel,anchor=north west] at ($(s3) + (0,0)$) {$x$};
+\foreach \u/\lab in {s0/$\textsf{a}$, s1/$\textsf{b}$, s2/$\textsf{c}$} \node[nodelabel,anchor=north] at (\u) {\lab};
+\foreach \u/\lab in {s4/$\textsf{e}$} \node[nodelabel,anchor=south west] at (\u) {\lab};
+\foreach \u/\lab in {s5/$\textsf{f}$} \node[nodelabel,anchor=south east] at (\u) {\lab};
+\foreach \u/\lab in {s3/$\textsf{d}$, s6/$\textsf{g}$} \node[nodelabel,anchor=south] at (\u) {\lab};
+
+%% Edges
+\draw (s1) -- (s3);
+\draw (s0) |- (s4);
+\draw (s4) -- (2,2) |- (s3);
+\draw (s4) |- (s6);
+\draw (s3) -- (4,1) |- (s5);
+\draw (s2) |- (s5);
+\draw (s5) |- (s6);
+
+% \node [anchor=north west] at (9,6) {B};
+\node [nodelabel,anchor=north west] at ($(10,5)$) {
+\begin{tabular}{c|c|l}
+% \multicolumn{2}{c}{Breakpoints}\\
+Node & Breakpoint & Parents\\
+\hline
+$\noderef{a}$ & $\varnothing$ & [\noderef{e}]\\
+$\noderef{b}$ & $\varnothing$ & [\noderef{d}] \\
+$\noderef{c}$ & $\varnothing$ & [\noderef{f}]\\
+$\noderef{d}$ & $x$ & [\noderef{e}, \noderef{f}] \\
+$\noderef{e}$ & $\varnothing$ & [\noderef{g}]\\
+$\noderef{f}$ & $\varnothing$ & [\noderef{g}]\\
+$\noderef{g}$ & $\varnothing$ & []\\
+\end{tabular}};
+
+% \node [anchor=north west] at (9,6) {C};
+% \node [nodelabel,anchor=north west] at ($(10,5)$) {
+% \begin{tabular}{ll}
+% Edges & Intervals\\
+% \hline
+% $(\textsf{a}, \textsf{e})$ & $(0, L]$ \\
+% $(\textsf{b}, \textsf{d})$ & $(0, L]$ \\
+% $(\textsf{c}, \textsf{f})$ & $(0, L]$ \\
+% $(\textsf{d}, \textsf{e})$ & $(0, x]$ \\
+% $(\textsf{d}, \textsf{f})$ & $(x, L]$ \\
+% $(\textsf{e}, \textsf{g})$ & $(0, L]$ \\
+% $(\textsf{f}, \textsf{g})$ & $(0, L]$ \\
+% \end{tabular}};
+
+\end{tikzpicture}
+\caption{\label{fig-arg-data-structure}
+A graphical depiction of a classical event ARG and a concrete
+encoding of the information. Node \noderef{d} is a recombination
+event as it is associated with a non-null breakpoint, and has
+two parents. Nodes \noderef{e}, \noderef{f} and \noderef{g} are
+common ancestor events.
 }
@@
 
-The simplest interpretation of the inheritance intervals $I$ for
-an ancestor genome $v$ and descendant $u$, is that $I$ defines the
-regions of genome that $u$ inherited from $v$. For example,
-node $o$ is the recombinant descendant of nodes $p$ and $q$ in
-Figure~\ref{fig-ancestry-resolution}A, which shows that it in inherited
-everything to the left of position $1$ from $p$ and the rest of
-its genome from node $q$.
-Thus, these edge annotations are concerned only with the genetic relationship
-between the nodein question, and are not affected by the rest of the graph.
-We will thefore refer to inheritance annotations of this type as ``local inheritance
-annotations''. Graphs of this type result from forwards-time
-simulations~\citep{kelleher2018efficient,haller2018tree} or the
-``Big ARG'' stochastic process (Appendix XXX).
-
-ARGs are most often thought of as a \emph{retrospective} structure however,
-defined with respect to some set of (usually, contemporary) sample genomes.
-In this case we are not explicitly interested in the
-entire interval of genome descendant $u$ inherited from ancestor $v$,
-but rather the subset of that interval that is \emph{ancestral to our
-sample}.
-Fig~\ref{fig-ancestry-resolution}B shows the graph that we obtain when
-reduce the annotation intervals to those that are ancestral to the
-sample (note that node $q$ is now ommitted because the MRCA of the sample
-has already been found for the full genome at node $p$).
-Tracking such
-``ancestral material''~\citep{wiuf1999ancestry,wiuf1999recombination}
-is much more useful~\citep{mcgill2013graphml,arenas2013importance}
-and, indeed, all that is possible to
-have information about when inferring ARGs from contemporary samples.
-We refer to a gARG with inheritance annotations of this type as
-``sample resolved''.
-% Keeping track of resolved ancestral material is one of the distinguishing
-% features of ARGs vs phylogenetic
-% networks~\citep{mcgill2013graphml,arenas2013importance}, but
-% is actually incompatible with the Griffiths gARG encoding (it does this
-% implicitly, I guess)
-
-% Fig~\ref{fig-ancestry-resolution} is an instructive example
-% from the literature \citep{wiuf1999recombination} and illustrates
-% how a classical eARG (Fig~\ref{fig-ancestry-resolution}A) can be described as
-% a gARG with edge annotations (Figs~\ref{fig-ancestry-resolution}B~and~C).
-
-% There are two distinct annotation schemes possible, and the distinction is subtle
-% and important. The first scheme is to use the annotations to describe the complete
-% set of genomic intervals directly inherited by the descendant node from the ancestor
-% node (Fig~\ref{fig-ancestry-resolution}B). The second is to use the edge annotations to
-% describe the ``ancestral material''~\citep{wiuf1999ancestry,wiuf1999recombination}
-% of the sample that is propagated through
-% the graph from descendant to ancestor (Fig~\ref{fig-ancestry-resolution}C).
-% As we discuss in section XXX, this second scheme has many advantages
-% in terms of efficiency.
-
-%ARGs are inherently a retrospective structure defined in terms
-%of a set of sample nodes (usually the leaves), and,
-%as we discuss in section XXX, storing sample-resolved ancestry
-%intervals (Fig~\ref{fig-ancestry-resolution}C) is generally much more
-%useful than storing the parent-child intervals recording direct inheritance
-%(Fig~\ref{fig-ancestry-resolution}B).
-%Indeed, storing direct inheritance intervals for edges is really only of interest because of
-%its correspondence with the classical approach of associating
-%breakpoints with recombination nodes in an eARG (Fig~\ref{fig-ancestry-resolution}A).
-%The eARG encoding is simply a more limited form, in which the only types of direct
-%inheritance we can have are the entire genome, or the intervals
-%at either side of a recombination breakpoint.
-
-%Note that this example is a sample from the so-called \emph{``big'' ARG process}
-%(see Appendix XXX) and has several features that we are unlikely to simulate or
-%infer in practice. For example, by definition the samples contain no genetic information about
-%nodes \textsf{g} and nodes \textsf{j} and so neither of those nodes are likely to be inferred.
-%Likewise, is is unlikely we would have information to infer the ``grand MRCA''
-%\textsf{q} as all regions of the genome have already reached a
-%common ancestor separately (in nodes \textsf{k} and \textsf{p}).
-
-% Third, there exist nodes such as
-% \textsf{k} with two children (i.e. ``common ancestor'' nodes in Hudson's terminology)
-% but in which no ancestral coalescence occurs. Fourth, some breakpoint positions occur
-% within regions of genome that are not ancestral (e.g. the breakpoint at position 3 in
-% node \textsf{k}). In this case sample-resolved edge annotations lose the exact position of
-% the breakpoint, although they retain the information that a breakpoint occurred between
-% positions 2 and 4 (which is sufficient for calculating the likelihood, see appendix XXX).
-
-Given a gARG with direct-inheritance edge annotations
-(Fig~\ref{fig-ancestry-resolution}B),
- we obtain the sample-resolved edge annotation
-(Fig~\ref{fig-ancestry-resolution}C) by traversing the graph backwards in time
-from sample nodes (\textsf{a}, \textsf{b}, \textsf{c}).
-Proceeding in topological order (such that an ancestor node is visited
-after all of its descendants) the sample-resolved material for a node is the
-union of the sample-resolved intervals on its inbound edges. We then annotate
-the outbound edges with the intersection of their direct-inheritance intervals
-and the just-computed sample-resolved intervals for the node.
-For example, node \textsf{i} in Fig~\ref{fig-ancestry-resolution}C
-carries sample-resolved material on the intervals $(0, 2]$ and $(4, 7]$. The edge
-joining \textsf{i} to its ancestor \textsf{k} then carries
-material on $(0, 2]$ since although \textsf{i} directly inherits interval
-$(0, 3]$ from \textsf{k} (Fig~\ref{fig-ancestry-resolution}B), only $(0, 2]$ is
-ancestral to the sample. For each node, we also keep track of the number of
-descendant samples over each genomic interval, and stop tracking that
-interval once the lineages from all samples have coalesced. For example,
-the edge from \textsf{k} to \textsf{n} omits the region $(0,2]$ because it has
-already fully coalesced (indicated by a fully black genomic region).
-
-% In practice, the backwards-in-time Griffiths process that produces
-% ARGs such as that shown in Figs~\ref{fig-ancestry-resolution}A~and~B
-% is rarely simulated, because the resulting ``big ARG'' (Appendix XXX) is
-% dominated by nodes such as \textsf{g} and \textsf{j} which contain information
-% about inheritance in non-ancestral material, as well as nodes such as
-% \textsf{o} and \textsf{q} which track details of ancestry even once the
-% local genomic region has fully coalesced. Instead, simulating
-% ARGs from the coalescent with recombination is usually done by
-% running Hudson's algorithm to produce a sample-resolved ARG directly.
-% Moreover, when inferring ARGs from genetic data, there is no information
-% in the sample set that can be used to resolve these nodes, so that they
-% are essentially unknowable.
-
-The process of converting a gARG with local-inheritance annotations
-to its sample-resolved equivalent is also known as ``simplification'',
-and is the key algorithm underlying recent
-advances in forwards-in-time
-simulation~\citep{kelleher2018efficient,haller2018tree}.
-It has many useful properties, as we discuss in the next section.
-
-% It is worth noting the process of sample resolution can reveal that certain ARG
-% nodes are \emph{redundant}: if they are removed and the edges adjusted to
-% pas directly from the node's child to the node's parent, it does not reduce our
-% knowledge of the samples' genetic ancestry. Removing nodes,
-% both redundant ones and others, is part of a process we call
-% ``simplification''.
-
-\begin{figure}
-\centering
-\vspace{5em}
-\includegraphics[width=\linewidth]{illustrations/simplification}
-\caption{\label{fig-simplification}
-ARG simplification. In all cases the graph visualization is shown on the left
-and the equivalent local trees on the right. Since multiple parents are indicative of
-recombination(s), nodes are coloured by the number of parents. and shaded
-according to the proportion of their span over which they are coalescent.
-(A) A gARG simulated from a diploid Wright-Fisher
-model. Note that the genome \textsf{k} has three children, and the genomes
-\textsf{e} and \textsf{g} simultaneously have multiple parents and multiple
-children.
-(B) Simplified to remove all
-% trying out this terminology - can define in the text
-1-connected graph components (e.g., diamonds such as \textsf{jlnm}).
-(C) Remove nodes that are unary in all local trees. This removes both recombinant nodes
-that have only one child (e.g. \textsf{n}) and nodes that have more
-than one child (i.e. are ``common ancestors'') but never represent coalescences
-in the local trees (e.g. \textsf{r}).
-(D) Rewrite edges to bypass any unary nodes in local trees. Note that this loses lineage
-information such as the fact that \textsf{k} is connected to \textsf{a} via \textsf{e}.
-}
-\end{figure}
-
-\section*{Simplification}\label{ARG_simplification}
-% I think a sentence is needed here on why "so much" text is devoted to simplification.
-% How does simplification connect to eARG vs gARG and practical benefits of gARG over eARG
-% and why do you now talk about unary nodes and a bit polytomies? Hold my hand here please;)
-ARG simplification~\citep{kelleher2018efficient} is the process
-of removing nodes and re-writing edges in a gARG to remove
-various types of redundancy,
-retaining the ``meaningful'' topology within local trees
-and maintaining the identity of nodes across different trees.
-The idea of meaningful local tree topologies is based on the
-presence of ``unary'' nodes: a tree node that has exactly
-% should presence above be absence?
-one child. Fig.~\ref{fig-simplification} show a series of successive steps of
-simplification, starting with a complete gARG simulated under a
-(backwards-in-time) Wright-Fisher
-model, shown both as a graph on the left, and the equivalent set of 6
-local trees on the right.
-For example, consider the path above node
-\textsf{a} in the first tree (covering positions 30 to 40)
-in Fig.~\ref{fig-simplification}A.
-Before \textsf{a} meets its common ancestor with the other samples at
-node \textsf{q}, it passes through \textsf{e}, \textsf{j}, \textsf{m}, \textsf{n} and
-\textsf{p}, which in this region of the genome are all unary nodes.
-This reflects the passage of this span of genome through the ARG towards the root.
-Such unary nodes in a local tree are not usually considered important;
-% because they provide little information and are not
-% detectable by information from any given individual tree.
-nodes usually mark the points where samples coalesce in a
- \emph{common} ancestor. For most purposes, we would therefore consider a path directly
-joining sample \textsf{a} to node \textsf{q} to be equivalent to the path
-containing the intermediate unary nodes. Simplification is the process of removing
-these less informative nodes from the local trees by modifying the gARG.
-
-The first type of simplification that we can perform is to remove
-graph topology that is invisible to the samples.
-An example of such topology is a
-``diamond''~\citep{rasmussen2014genome}
-in which the two parent nodes of a recombination immediately
-join again into a common ancestor (e.g.~\textsf{j}, \textsf{l}, \textsf{m}
-and \textsf{n} in Fig.~\ref{fig-simplification}A).
-Unless we are specifically
-interested in the recombination event or ancestral genomes,
-there is no information in this topology and the diamond can be
-replaced by a single edge. More generally, any
-subgraph that is singly-connected in both the leafward and
-rootward direction (a ``super-diamond'') is non-identifiable and can be
-replaced by one edge. This definition includes the case
-of a node that has one inbound and one outbound edge, such as
-nodes \textsf{f} and \textsf{h}.
-Fig.~\ref{fig-simplification}B shows the result of this type of
-graph topology simplification.
-
-Simplifying away diamonds will remove many unary nodes from the
-local trees, but there can still be nodes that are unary in all
-of the local trees. In particular, a node can represent a recombinant
-with multiple parents in the graph but only a single child (e.g.\ node \textsf{n}
-in Fig.~\ref{fig-simplification}B), or can represent a common ancestor with
-multiple children in the graph but in which no local coalescence takes place
-(node \textsf{r} in Fig.~\ref{fig-simplification}B).
-% point out that this is why Hudson refers to CA nodes, not coalescence nodes
-Such nodes are not singly connected in the graph, but are nevertheless unary in
-all of the local trees.
-The operation to remove them
-therefore requires knowledge not just of the graph topology but also of the
-edge annotations.
-As we see in Fig.~\ref{fig-simplification}C,
-removal of recombinant nodes can produce graph nodes with
-more than two parents (e.g.~node \textsf{e}); and likewise, removal of
-common ancestor but non-coalescent nodes can produce graph nodes with
-more than two children (e.g.~node \textsf{s}). These represent a ``stacking up'' of
-multiple---often indistinguishable---events into a single node.
-% something which is impossible to encode in an eARG.
-
-% JK: Text moved from an earlier section. See if we want to reuse this here
-% if not already said.
-% Indeed a single node can have both multiple children
-% and multiple parents. This would be the case in Fig~\ref{fig-arg-in-pedigree} if,
-% for instance, node \textsf{a} were to have two children.
-% % It is a shame that this is not actually shown in fig 3, but I don't think
-% % we can do this unless we have 5 generations
-% Moreover, a single node can have more than 2 children, and (as we shall see in
-% the ``\nameref{ARG_simplification}'' section), more than 2 parents. Another way to
-% state this is that transitions between nodes in a gARG can encapsulate multiple
-% events in an eARG.
-
-The remaining nodes are MRCAs of some subset of the samples
-at \emph{some} positions along the genome. We still have
-some unary nodes in the local trees, but these nodes will
-correspond to a coalescence in at least one other
-local tree. For example, node  \textsf{k} is unary in the second tree
-of Fig.~\ref{fig-simplification}C, but is either binary
-or ternary in all subsequent trees (recall this is a Wright-Fisher
-simulation). The final level of simplification is to alter the edge annotations
-such that, although no nodes are removed from the graph, all
-unary nodes disappear from the local trees (Fig.~\ref{fig-simplification}D).
-Note that although this last stage produces simpler local trees, by
-removing information about the exact paths taken by lineages through
-the graph, we lose potentially useful information about shared edges
-between trees. The extent to which this information can be inferred
-and utilised is an interesting open question.
-
-% Needs another pass
-An important consequence of simplifying local trees to remove
-unary nodes is that we may lose the ability to identify
-the ``subtree prune-and-regraft'' (SPR) operations
-by which they differ~\citep{hein1990reconstructing,song2003on,song2006properties}.
-% The effects of recombination as we move from left-to-right along
-% the genome in the complete simulated gARG
-% of Fig~\ref{fig-simplification}A are easy to interpret,
-% but as more redundancy is removed from the graph
-% the
-As explained above, we can have multiple recombination and
-common ancestor events stacking up, and therefore
-many possible SPR operations. Deducing the SPRs between
-trees is NP-hard in
-general~\citep{hein1996complexity,allen2001subtree,bordewich2005computational},
-but effective heuristics have been proposed~\citep{rasmussen2022espalier}.
-
-% JK TODO: finish up this section with a quick version of the
-% para here. Basically say, people often say an ARG is a
-% set of local trees cite cite, but it's not as simple as that as we
-% can see from the example (probably).
-
-% The equivalence between an ARG and a set of local trees
-% separated by SPRs is often mentioned [e.g. x y z], but
-% it is important to note that while extracting local
-% trees from an ARG is straightforward, the converse
-% process of reconstructing precisely the same ARG from
-% these trees is more problematic. If the internal
-% nodes are consistently labelled and unary nodes are
-% included in the local trees (e.g., Fig~\ref{fig-simplification})
-% then it is easy to exactly reconstruct the ARG.
-% Otherwise, the problem is much more difficult and it
-% may not be possible to uniquely reconstruct the original ARG.
-% Without labels for the internal nodes, we are forced to
-% to reason about which subtrees remain the same and which
-% differ due to the SPR separating adjacent trees in order to
-% assign node identity. Without the internal unary nodes marking
-% recombinations, we must infer their existence. In
-% general, there will not be a unique solution to this problem,
-% and so there can be many possible ARGs compatible with
-% a given list of leaf-labelled binary trees.
-% This is the approach taken by the Espalier inference
-% method~\citep{rasmussen2022espalier}, which finds a
-% set of SPRs that reconcile independently inferred
-% trees from different regions of the genome.
-
-% NEEDS more work!
-\section*{Precision of ancestry information}
-Ancestral recombination graphs are often characterised
-as consisting of a complete record of the genetic history
-of a sample. For example,
-\cite{rasmussen2014genome} state that an ``ARG provides a record of all
-coalescence and recombination events since the divergence of the sequences
-under study'' and
-according to \cite{deng2021distribution} an ARG
-``provides all the information about the genealogical history of a sample,
-including the locations of recombination events.'' It is worth
-questioning, however, whether such comprehensive information
-will always be available, and in particular whether our
-representation of genealogical history should \emph{require}
-such precision. Simulations will usually generate complete
-information about the history of a sample, of course, but
-we may not have sufficient information to infer such detail.
-
-% The full ancestral recombination graph (ARG) is a structure that encodes all
-% coalescence and recombination events resulting from the stochastic process of
-% the coalescent with recombination.
-\citet{brandt2021evaluation} define a ``full'' ARG as
-``a structure that encodes all
-coalescence and recombination events resulting from the stochastic process of
-the coalescent with recombination'' (thus going against the general
-historial trend discussed in the XXX section).
-\citet{rasmussen2022espalier} have a less restrictive definition, in
-which a ``full'' ARG is one that contains complete information about
-recombination events, without necessarily being tied to a particular
-stochastic process.
-While [something positive], it is important to be clear on
-what the overall goal of such precise inferences are,
-what the limitations on when they can be made are,
-and their downstream utility in applications.
-
-% What is the goal of having fully precise ARGs?
-% 1) studying the actual recombination events;
-% 2) evaluating the likelihoods
-The main reasons for wanting precise estimates of all recombination
-events are to either compute a likelihood of the inferred ARG
-under a statistical model, or to study the recombinants in
-detail~\citep{rasmussen2022espalier}.
-% Explain that we're really interested in the details of recombination
-% itself with a small number of samples.
-In the latter case, we might
-have a small number of samples~\citep{guo2022recombination}.
-
-Beyond cases in which visual inspection of the trees and the
-effects of recombination is feasible, the main application
-of having completely precise estimates of recombination events is
-to evaluate the likelihood of the estimate under some
-statistical model. The only statistical model available
-is the coalescent with recombination
-and its approximations~\citep{mcvean2005approximating,marjoram2006fast}.
-In this case, the utility of a ``full`` ARG for some dataset
-is limited by how well the dataset fits the assumptions
-of the model.
-
-% Not sure where this goes, but there are some critical points
-% that we want to explain and not just throw out there in the
-% discussion.
-% A critical distinction between the eARG and gARG encodings
-% is the level of precision about the time and location of
-% recombination events that is required. In an eARG, we are
-% required to be absolutely precise about every recombination
-% [explain and contrast with gARG, referring to examples]
-
-% JK: This is very rough, just jotting down the basic ideas,
-% will need substantial revision.
-It is not clear that such complete information will always
-(or indeed ever) be available from inferences~\cite{shipilina2022origin}.
- We may be
-able to precisely identify the location and timing
-of recent recombinations (particularly if detailed pedigree
-information is available) but beyond this, the information
-required simply isn't present. When sampling from a
-process such as the Sequentially Markov
-Coalescent~\citep{mcvean2005approximating,marjoram2006fast}
-like ARGweaver~\citep{rasmussen2014genome}, it is true that
-we do obtain a fully resolved ARG with precise
-information about recombination, but it is reasonable
-to question how much information from the data is used
-to support older recombinations, in particular.
-
-% This idea of a continuum of information about recombination
-% being inferrable and representable is consistent with the ideas
-% discussed by~\cite{shipilina2022origin}. There a haplotype block
-% is synonymous with coalescent nodes in the ARG, which we can generalise
-% as nodes in a gARG with at least two children. In principle, these
-% nodes are identifiable by the mutations that occur on the outbound
-% edges of the node.
-
-It is helpful to consider the difference between fully
-resolved binary trees and those that contain polytomies.
-
-The situation is in fact directly analogous: an ARG that
-contains fully resolved recombination information must have
-at most parents per node.
-
-At the very least, it is not reasonable for the \emph{data
-structure} to require complete precision in order to
-represent some less precise. This is a fundamental
-advantage of the gARG encoding over the classical eARG
-method: we are not forced to posit a precise cause (recombination
-event) for every effect (difference between adjacent trees).
-These differences in precision of recombination
-information are illustrated in the ARGs estimated by different
-methods, as we see in the next section.
-
-In general, discussions tend to contrast an ARG
-on one hand as holding \emph{complete} information about recombination
-(and therefore between-tree correlation structure)
-and a sequence of local trees on the other (with, by implication,
-no direct information about between-tree correlations).
-This is a false dichotomy:
-a \emph{continuum} of information about recombination and between-tree
-correlation structure can be represented and inferred, and all
-such levels of information are useful, even if only for computational
-efficiency.
-
-\begin{figure} \begin{center}
-\includegraphics[width=\textwidth]{illustrations/inference.pdf} \end{center}
-\caption{\label{fig-inferred-args} ARGs inferred by
-(A) \kwarg~\citep{ignatieva2021kwarg},
-(B) \argweaver~\citep{rasmussen2014genome,hubisz2020inference},
-(C) \tsinfer~\citep{kelleher2019inferring},
-and (D) \relate~\citep{speidel2019method}
-for 11 samples of a 2.7kb region of the \textit{Drosophila melanogaster} ADH
-locus~\citep{kreitman1983nucleotide}. The \argweaver\ example
-is a sample from the inferred posterior distribution. In each ARG,
-colours are assigned to edges based on the time of the edge's child node:
-lighter colours are assigned to edges with older child
-nodes, darker colours are assigned to edges with younger child nodes
-Note that the position of points on the
-Y axis is chosen for graphical clarity by the graphviz \texttt{dot}
-algorithm, rather than as a direct measure of time.
-Line width corresponds to the genomic span of the edge. Node colour
-refers to the number of parents of the node: red is zero, brown is one,
-green to blue are two to N parents. The brightness
-of the node is determined by the genomic span over which
-the node has two or more children divided by the span where it has any
-children. The genomic intervals associated with edges are visualised below
-each ARG, where each row is an edge
-and the x axis reflects genomic position. The width of edges
-thus corresponds to their genomic span. Edges
-are ordered by the age of their child node, from oldest to youngest
-moving down the page, with colours matching the edge colours in the ARG.
-Vertical lines indicate breakpoints between local trees.
-}
-\end{figure}
-
-To illustrate these points and to emphasise the heterogeneity
-of the ARGs inferred by current methods, we inferred ARGs for the
-the classical \citet{kreitman1983nucleotide} benchmark dataset
-using four recent inference methods and show the results in
-Fig~\ref{fig-inferred-args}.
-While there is some consistency among the inferred ARGs (such as the samples
-FR-F, WA-F and Af-F being grouped into the same clade), there
-are also considerable differences. For this dataset, the minimum number
-of recombination breakpoints required in the absence of recurrent mutation
-is 7~\citep{song2003parsimonious}, and the number of inferred breakpoints
-under the input mutation and recombination rates
-varies from 2 (\relate) to 17 (\argweaver).
-The ARGs inferred by the different methods also have major differences
-in how recombination information is represented. Essentially,
-\kwarg\ and \argweaver\ infer a specific causal recombination \emph{event}
-for each breakpoint, and those events are fully
-described in the output graph. On the other hand, \tsinfer\ and
-\relate\ do not infer causal events but instead focus on
-representing the inferred \emph{effect}: breakpoints between the
-trees. It is important to note that this does not imply that
-the methods must then necessarily infer only a sequence of
-unrelated local trees
-% More? See https://github.com/tskit-dev/what-is-an-arg-paper/issues/38
-% for discussion. I'm sure there were other examples of people saying
-% that
-(as claimed by e.g.~\citet{hejase2020summary}),
-and information about recombination and the correlation structure
-between trees is not ``all or nothing''.
-Both methods capture correlation structure between trees:
-\relate\ first infers local tree topologies and then identifies edges
-that persist from one tree to the next as a separate step
-(in a similar manner to \espalier~\citep{rasmussen2022espalier}), while
-\texttt{tsinfer} explicitly constructs nodes that are shared across local
-trees. See \cite{ignatieva2023distribution} for a detailed discussion
-and comparsion of inference methods on this property.
-
-% Because the ARGs inferred by \tsinfer\ and \relate\ do not contain
-% explicit recombination events we cannot compute a likelihood
-% under the coalescent with recombination (see appendix XXX).
-% [Some stuff about whether we should see this as necessary or not,
-% but maybe we should keep this for the conclusion?]
-
-% \section*{When do we need ``full'' ARGs?}
-% % The full ancestral recombination graph (ARG) is a structure that encodes all
-% % coalescence and recombination events resulting from the stochastic process of
-% % the coalescent with recombination.
-% \citet{brandt2021evaluation} define a ``full'' ARG as
-% ``a structure that encodes all
-% coalescence and recombination events resulting from the stochastic process of
-% the coalescent with recombination'' (thus going against the general
-% historial trend discussed in the XXX section).
-% \citet{rasmussen2022espalier} have a less restrictive definition, in
-% which a ``full'' ARG is one that contains complete information about
-% recombination events, without necessarily being tied to a particular
-% stochastic process.
-% While [something positive], it is important to be clear on
-% what the overall goal of such precise inferences are,
-% what the limitations on when they can be made are,
-% and their downstream utility in applications.
-
-% % What is the goal of having fully precise ARGs?
-% % 1) studying the actual recombination events;
-% % 2) evaluating the likelihoods
-% The main reasons for wanting precise estimates of all recombination
-% events are to either compute a likelihood of the inferred ARG
-% under a statistical model, or to study the recombinants in
-% detail~\citep{rasmussen2022espalier}.
-% % Explain that we're really interested in the details of recombination
-% % itself with a small number of samples.
-% In the latter case, we might
-% have a small number of samples~\citep{guo2022recombination}.
-
-% Beyond cases in which visual inspection of the trees and the
-% effects of recombination is feasible, the main application
-% of having completely precise estimates of recombination events is
-% to evaluate the likelihood of the estimate under some
-% statistical model. The only statistical model available
-% is the coalescent with recombination
-% and its approximations~\citep{mcvean2005approximating,marjoram2006fast}.
-% In this case, the utility of a ``full`` ARG for some dataset
-% is limited by how well the dataset fits the assumptions
-% of the model.
-
-% % If we want to use the coalescent, then really the assumptions
-% % need to make sense.
-% % n << Ne is a silly assumption in today's datasets
-% A core assumption of the coalescent is that the sample size $n$
-% is much less than the effective population size, $N_e$.
-% Several human datasets now consist of hundreds of thousands of
-% genomes~\citep{bycroft2018genome,karczewski2020mutational,tanjo2021practical},
-% and so sample size is substantially \emph{larger} than $N_e$
-% (often assumed to be $10^4$ in humans).
-% Agricultural datasets are even more extreme.
-% For example,
-% the US dairy cattle database alone currently comprises more than 6 million
-% animals with SNP array
-% genotypes\footnote{\url{https://queries.uscdcb.com/Genotype/counts.html}},
-% while the effective population size in modern dairy cattle breeds is
-% less than 100 and decreasing~\citep{MacLeod2013,Makanjuola2020}.
-% An extreme sign of these breeding practices is that there are only two ancestral
-% Y-chromosome lineages present in today's US Holstein dairy breed~\citep{Yue2015}.
-% Simlarly, the 1000 bull genomes project~\citep{hayes20191000}
-% comprises close to 7000 genomes, which are part of multi-generation pedigrees
-% with millions of animals and extensive SNP array genotype and phenotype
-% data \citep[e.g.][]{Cesarani2022}.
-% In another example, genome and SNP array data for 440,610 individuals within
-% 7 multi-generation pedigrees~\citep{whalen2018,Johnsson2021,Ros-Freixedes2020}
-% were combined to infer recombination in pigs~\citep{RosFreixedes2022}.
-% Effective population size in modern pig breeds is also less than 100 due to
-% intense selection and directed reproduction \citep{Hall2016,Porcnic2016}.
-% Another core assumption of the coalescent model is that the genome (or
-% at least the region under study) is short enough that the number of extant
-% lineages remains much smaller than $N_e$ at all times. Whole
-% genome sequences have been available for model organisms
-% for over a decade now, % True? Citation?
-% and indeed complete chromosome-level assemblies are possible
-% in humans~\citep{miga2020telomere}.
-% Projects are under way to obtain high-quality assemblies
-% for all eukaryotic species in Britain and Ireland~\citep{darwin2022sequence}
-% and ultimately worldwide~\citep{lewin2022earth}.
-
-% While the coalescent as a model can be suprisingly robust to
-% basic violations of the underlying assumptions~\citep{
-% wakeley2012gene,bhaskar2014distortion,nelson2020accounting}
-% it is reasonable to question the validity of the model in
-% representing datasets such as UK Biobank.
-% [Also mention somehow that $n\ll N_e$ is only one problem -
-% we're also ignoring the myriad other evolutionary forces.]
-% Such datasets cannot be reasonably be described by
-% the coalescent as a statical model, and so more non-parametric
-% methods like \tsinfer\ and \relate\ would seem more appropriate.
-% [More]
-
-% % Is it practical to estimate full ARGs?
-% % [Not particularly \citep{shipilina2022origin} - but maybe ARGneedle
-% % has done this for UKB? Need to check]
-
-% % Do downstream applications actually make use of such full ARGs?
-% Downstream applications using estimated ARGs to date rarely
-% make use of such complete recombination information.
-% The \relate\ selection statistic~\citep{speidel2019method}
-% computes a $p$-value for selection [FIXME] for each
-% local tree independently by comparing the rates of coalescence
-% at various time-points. For cases like strong positive selection,
-% where we expect there to be a clear signal in the local tree,
-% such tree-by-tree methods can be simple and powerful.
-% In their method
-% for estimating dispersal rates and the locations of genetic
-% ancestors,
-% \citep{osmond2021estimating} downsample trees along the genome
-% so that they can be regarded as approximately independent.
-% The SIA method for detecting selection~\citep{hejase2022deep}
-% encodes local trees as a set of lineage counts at discrete
-% time intervals, and uses these as feature for a
-% type of machine learning algorithm
-% that takes ``temporal'' correlations into account. Thus,
-% while SIA does use some of the information about local tree correlation,
-% clearly much of the detail about recombination
-% events in a ``full'' ARG is lost.
-% The \texttt{tsdate} algorithm (and related naive estimator
-% for ancestral location) uses much more between-tree
-% information~\citep{wohns2022unified}, explicitly using the gARG
-% encoding to reason about ancestral nodes.
-% A major strength of this approach is that it can be used on
-% any gARG, ranging from estimates containing relatively little
-% information about recombination (e.g., as estimated by \relate)
-% to ``full'' ARGs containing complete recombination information
-% (e.g., as estimated by \kwarg), and will fully utilise
-% any between-tree correlation information that is present.
-
-\section*{Computational efficiency and data interchange}
-ARG inference methods have been successfully applied to vast datasets,
-including 500,000 genotyped humans~\citep{kelleher2019inferring,zhang2023biobank}
-from UK Biobank~\citep{bycroft2018genome}
-and over a million SARS-Cov-2 whole genomes~\citep{zhan2023towards}.
-Even larger datasets are
-available~\citep[e.g.][]{halldorsson2022sequences} % TODO more
-and it seems inevitable that ARGs will be inferred for them.
-The efficiency with which we can store and analyse these inferred
-ARGs is therefore of great importantance, and a key factor
-in the ultimate success of downstream ARG-based methods.
-Another key factor in how important ARG-based methods become
-across the wide range of predicted application
-% TODO more?
-areas~\citep{rasmussen2014genome,harris2019database,
-hejase2020summary,schaefer2021ancestral,
-fan2022genealogical,wohns2022unified,harris2023using}
-is how well inference methods and downstream analysis applications
-interoperate. Without a well-defined and efficient shared format
-for ARG data interchange,
-users will suffer from problems of converting data from one programs
-output to another, an unfortunate hallmark of population genetics
-software for many years~\citep{excoffier2006computer}.
-While current methods tend to tightly couple downstream analysis
-of the inferred ARG with the inference
-itself within the same software package
-% FIXME-  more explicit?
-(e.g.\ Relate, ARGneedle)
-this approach scales poorly in terms of software development
-effort, and is ultimately not compatible with the widespread use
-of ARGs for routine data analysis, and a healthy and diverse
-software ecosystem.
-% FIXME not sure where this goes
-Earlier efforts to standardise ARG interchange
-did not succeed~\citep{cardona2008extended,mcgill2013graphml}.
-
-The gARG encoding discussed and concretely defined in this manuscript
-leads to highly efficient storage and processing of ARG data,
-and has already been in use for several years.
-The succinct tree sequence data structure (usually known as a ``tree sequence''
-for brevity, but see below for confusion around this point)
-is a concrete encoding of a gARG, as discussed in this article.
-It was originally developed as part of the \texttt{msprime}
-simulator~\citep{kelleher2016efficient} and subsequently been
-extended and applied to forward-time
-simulations~\citep{kelleher2018efficient,haller2018tree},
-calculation of population genetics statistics~\citep{ralph2020efficiently}.
-The succinct tree sequence encoding extends the basic definition
-of a gARG provided here by stipulating a
-simple tabular representation and nodes and edges,
-and also defining a concise and lossless representation of
-sequence variation using the ``site'' and  ``mutation'' tables.
-The \texttt{tskit} library~\citep{tskit2023tskit} is a liberally
-licensed open source toolkit that provides a comprehensive suite
-of tools for working with gARGs (encoded as a succinct tree sequence).
-Based on core functionality written
-in C, it provides interfaces in C, Python and Rust.
-Tskit is mature software, widely used in population genetics, and
-has been incorporated into several downstream
-applications~\citep[e.g.,][]{haller2019slim,speidel2019method,
-adrion2020community,
-terasaki2021geonomics,
-baumdicker2021efficient,
-fan2022genealogical,korfmann2022weak,
-mahmoudi2022bayesian,petr2022slendr,rasmussen2022espalier,
-zhang2023biobank}.
-
-% FIXME this needs more work and doesn't really fit into the
-% narrative right now, but we do need to make this point.
-The key insight that makes the succinct tree sequence encoding
-an efficient substrate for defining analysis algorithms is that
-it allows us to generate the local trees along the genome
-in a way that allows us to reason about the \emph{differences}
-between those trees.
-Sequentially generating the marginal trees along the genome
-is also fundamental, and is necessary whenever we need to
-perform calculations that are contingent on more than just the
-isolated properties of an edge. \cite{kelleher2016efficient}
-showed how all trees can be sequentially generated in
-constant time per tree transition in a fully simplified gARG.
-Furthermore, we can easily reason about how tree topologies
-change (and stay the same), leading to efficient algorithms
-for computing population genetic
-statistics~\citep{kelleher2016efficient,kelleher2018efficient},
-implementing the Li and Stephens
-model~\citep{kelleher2019inferring,wohns2022unified}
-and likely many more.
-
-In retrospect, the choice of terminology around the succinct
-tree sequence data structure was unfortunate, and has led to
-significant confusion. As first introduced by~\cite{kelleher2016efficient}
-a ``tree sequence`` was defined as the set of ``coalescence records``
-that is output by Hudson's simulation algorithm.
-This was then generalised to use the
-tabular encoding described above and formally described as a ``succinct
-tree sequence''~\citep{kelleher2018efficient}.
-The ``succinct'' prefix here is intended to connect
-with the concept of succinct data structures
-\citep[e.g.][]{gog2014theory},
-which have near-optimal
-space usage but still support efficient retrieval and computation.
-The concrete data structures of the succinct tree sequence, precisely
-defined and with a particular focus on computational efficiency
-did not have an obvious connection to the ARG definitions
-provided by Griffiths and
-colleagues~\citep{griffiths1991two,ethier1990two,
-griffiths1996ancestral,griffiths1997ancestral}, and with the lack
-of any other precise definitions, it seemed best to avoid the
-term to prevent confusion. Unfortunately, however, the opposite
-has occured. In describing the output of \tsinfer\ as a ``tree
-sequence''~\citep{kelleher2019inferring} many
-have---not unreasonably, but incorrectly---concluded that the
-output is a collection of \emph{independent}
-% MORE?
-% https://github.com/tskit-dev/what-is-an-arg-paper/issues/38
-trees~\citep[e.g.,][]{hejase2020summary,ignatieva2021kwarg}.
-% Could say more here, but maybe we'll have a note about terminology
-% in the discussion
-
-\section*{Discussion}
-Recent breakthroughs have made ARG inference methods practical
-for the first time, and there has been a surge of interest
-in inference methods, their evaluation and application.
-The prospect of ARGs being used routinely in population
-and statistical genetics applications is tantalising,
-but in reality there is  substantial work to be done to
-enable this, and much remains to be done.
-
-A necessary first step towards this goal is some terminological
-clarity. As discussed in Appendix XXX, the term ARG has several
-subtly different interpretations depending on context. The
-trend to decouple the ARG from its origins as a stochastic
-process and instead use the term as a more general representation of any
-recombinant genetic ancestry seems most useful, which we have
-tried to clarify and systematise here. Thus
-we can think of an ARG as being any structure that encodes the
-reticulate genetic ancestry of a sample of colinear sequences under
-the influence of recombination. The ``genome'' ARG encoding
-made explicit here is one way we can concretely
-define such recombinant ancestry, which we have shown is both
-flexible and efficient.
-
-The flexiblity of the gARG encoding contrasts with the classical
-``event'' ARG (eARG), which, because of its origins in an idealised
-mathematical model, is more limited in what can be described.
-The necessity of complete precision under the eARG encoding has been
-conflated with the idea of a ``true'' or ``full'' ARG.
-[A discussion of how realistic inferring a completely precise ARG is]
-[Do downstream applications use full ARGs? usually no]
-
-% % Do downstream applications actually make use of such full ARGs?
-% Downstream applications using estimated ARGs to date rarely
-% make use of such complete recombination information.
-% The \relate\ selection statistic~\citep{speidel2019method}
-% computes a $p$-value for selection [FIXME] for each
-% local tree independently by comparing the rates of coalescence
-% at various time-points. For cases like strong positive selection,
-% where we expect there to be a clear signal in the local tree,
-% such tree-by-tree methods can be simple and powerful.
-% In their method
-% for estimating dispersal rates and the locations of genetic
-% ancestors,
-% \citep{osmond2021estimating} downsample trees along the genome
-% so that they can be regarded as approximately independent.
-% The SIA method for detecting selection~\citep{hejase2022deep}
-% encodes local trees as a set of lineage counts at discrete
-% time intervals, and uses these as feature for a
-% type of machine learning algorithm
-% that takes ``temporal'' correlations into account. Thus,
-% while SIA does use some of the information about local tree correlation,
-% clearly much of the detail about recombination
-% events in a ``full'' ARG is lost.
-% The \texttt{tsdate} algorithm (and related naive estimator
-% for ancestral location) uses much more between-tree
-% information~\citep{wohns2022unified}, explicitly using the gARG
-% encoding to reason about ancestral nodes.
-% A major strength of this approach is that it can be used on
-% any gARG, ranging from estimates containing relatively little
-% information about recombination (e.g., as estimated by \relate)
-% to ``full'' ARGs containing complete recombination information
-% (e.g., as estimated by \kwarg), and will fully utilise
-% any between-tree correlation information that is present.
-
-Fully decoupling the general concept of an ARG from the coalescent
-with recombination stochastic process is an important step.
-While the coalescent with recombination is a very useful and
-robust model~\citep{wakeley2012gene,bhaskar2014distortion,nelson2020accounting}
-many modern datasets have properties that grossly
-violate its assumptions.
-For example, several human datasets now consist of hundreds of thousands of
-genomes~\citep{bycroft2018genome,karczewski2020mutational,tanjo2021practical},
-substantially \emph{larger} than the usually assumed $N_e$ values.
-Agricultural datasets are an even more extreme departure from coalescent
-assumptions, with large samples embedded in
-multi-generational pedigrees~\citep{hayes20191000,Ros-Freixedes2020}
-and effective population sizes of 100 and even
-less~\citep{MacLeod2013,Makanjuola2020,Hall2016,Porcnic2016}.
-Finally, the COVID-19 pandemic has provided an example of a dataset that
-is clearly a very poor fit for the coalescent with recombination, but which,
-nevertheless, we would like to estimate ARGs.
-\cite{zhan2023towards} inferred an ARG for 1.27 million SARS-Cov-2 whole
-genomes, and showed how it accurately captures several known properties of
-the pathogen's evolution, and can efficiently and concisely
-encode this vast dataset.
-
-This view of ARGs,
-decoupled from generative models and
-without the hard requirement
-of complete precision on all historical events, may help improve
-methods for inference method evaluation. In the majority of cases,
-ARG inference accuracy is evaluated by comparing some truth simulations
-to the inferred ARGs by the pairwise comparison of local trees along the genome
-using tree metrics as originally proposed by \citet{kuhner2015assessing}.
-While this provides valuable insights into the method's performance,
-it is not clear how well these tree distance metrics reflect
-performance on real data.
-In comparing tree-by-tree along the genome, the effects of recombination
-structure is incorporated in a very indirect manner.
-Tree distance metrics usually have $O(n^2)$ time complexity, and so cannot be
-applied to the very large sample sizes currently of interest.
-\citet{deng2021distribution} derived the distribution of the expected
-waiting distances to tree changes.
-[ \citet{ignatieva2023distribution} DID SOMETHING ELSE]
-\citet{brandt2021evaluation} compared the
-distributions of time to the most recent common ancestor across several
-methods, which is a useful addition but only indirectly captures
-topological comparisons.
-% Not sure this is the right place for this sentence
-All of these comparisons and evaluations have been based on
-highly idealised simulations, and the effects richness of real data
-(beyond simplistic error models) are almost entirely unknown.
-A suite of true ARG distance metrics, that take into account more of the
-global topology and can be applied to large ARGs, would be a
-valuable addition to the field.
-
-\bibliographystyle{plainnat}
-\bibliography{paper}
-
-\setcounter{secnumdepth}{2} % Print out appendix section numbers
-
-% TODO revisit number here
-\section*{Appendix}
-\appendix
-
-\section{Ancestral graphs: a brief history}
-\label{sec-arg-history}
-The coalescent~\citep{kingman1982coalescent,kingman1982genealogy,
-hudson1983testing, tajima1983evolutionary} models the ancestry of a sample of
-genomes under an idealised population model, and provides the theoretical
-underpinning for much of contemporary population genetics.
-It is a stochastic process, where each random realisation
-is a genealogical tree describing the genetic ancestry of the sample.
-Numerous extensions to the model have been
-proposed~\citep{hudson1990gene,hein2004gene,wakely2008coalescent},
-incorporating many evolutionary processes.
-\citet{hudson1983properties}
-first incorporated recombination into the coalescent process,
-providing several fundamental analytical results
-and describing the basic simulation algorithm, still in
-widespread use~\citep{hudson2002generating,kelleher2016efficient,
-baumdicker2021efficient}.
-In the 1990s, Griffiths and colleagues revisited the
-coalescent with recombination from a different perspective,
-formulating it as a stochastic process where each realisation
-is encoded as a graph~\citep{griffiths1991two,ethier1990two,
-griffiths1996ancestral,griffiths1997ancestral}.
-They referred to both the stochastic process and
-its random realisations as the Ancestral Recombination Graph (ARG).
-Although mathematically equivalent, it is
-important to note that the Griffiths and Hudson formulations of
-the coalescent with recombination are not identical;
-in particular, a direct implementation of the ARG process
-as originally described requires exponential time to simulate
-(see Appendix XXX for details). However, ARGs provided a way
-to reason about and infer recombinant ancestry as a single object,
-which was not possible within Hudson's framework, which emphasised
-the collection of local trees along the genome resulting from recombination.
-
-Subsequent work on ARGs proceeded in broadly three main directions: focussing on (1)
-exploring the mathematical properties of the coalescent with recombination and
-related stochastic processes, (2) inferring evolutionary parameters under
-(approximations to) this model, either with or without explicitly reconstructing the
-genealogy of the sample, and (3) treating the ARG as a discrete graph, ignoring the
-generating stochastic process, and studying its properties from a computational and
-algorithmic perspective.
-
-% Mathematical treatment of the CwR and ARGs (and ARG-like stochastic processes)
-An extensive body of work has been developed from
-studying the coalescent with recombination
-% Can we add a couple of the most well known CwR examples here
-and other related
-graph-valued stochastic processes from a mathematical perspective.
-In particular, the Ancestral Selection Graph
-(ASG)~\citep{krone1997ancestral,neuhauser1997genealogy}
-uses a similar approach to model natural selection instead of recombination.
-Unlike the ARG process, the ASG imposes a hard distinction between the stochastic process,
-which constructs a random ARG-like graph, and an observable realisation,
-which is a single tree sampled from the graph in a non-uniform way to encode
-desired patterns of natural selection.
-Constructions of ASG-like stochastic processes encoding various
-forms of selection, often in parallel with recombination or other genetic forces,
-are an area of considerable and ongoing theoretical interest~\citep[e.g.][]{
-neuhauser1999ancestral,
-donnelly1999genealogical,
-fearnhead2001perfect,
-fearnhead2003ancestral,
-etheridge2009coalescent,
-gonzalezcasanova2018duality,
-koskela2019robust}.
-% Could mention the Ancestral Gene Transfer graph here too, within
-% something like "Similarly the Ancestral Gene Transfer Graph [cite]
-% and [other ancestral graphs, if they exist?] model genealogical
-% processes in bacteria as a graph.
-
-% Inference of parameters and explicit ARG reconstruction
-Early work on inference under the coalescent with recombination
-focused on the problem of
-inferring the parameters of the
-stochastic process, where the ancestry was regarded as a
-latent parameter to be averaged out
-\citep[e.g.][]{griffiths1996ancestral,kuhner2000maximum, nielsen2000estimation,
-fearnhead2001estimating}.
-These methods met with limited success
-because the state space of ARGs is overwhelmingly large, and
-lacks a simple geometry or neighbourhood structure for inference or
-sampling methods to  exploit.
-Several breakthroughs in this direction were achieved through
-formulating simplified but more tractable approximations to the full
-model~\citep{mcvean2005approximating,marjoram2006fast,li2011inference,
-paul2011accurate,schiffels2014inferring}.
-The related problem of \emph{sampling} genealogies compatible with a given
-dataset under the coalescent with recombination also proved notoriously difficult
-computationally; progress in explicitly inferring genealogies at scale
-has similarly been achieved through resorting to principled
-approximations~\citep{rasmussen2014genome,mahmoudi2022bayesian},
-or moving away from the coalescent with recombination altogether and seeking
-to infer a single plausible ARG~\citep[e.g.][]{speidel2019method} or ARG
-topology~\citep[e.g.][]{minichiello2006mapping,kelleher2019inferring}.
-
-% Optimisation problems to do with ARGs and other phylogenetic networks
-There has also been substantial interest in formulating and answering
-fundamental questions about properties
-of the ARG as a discrete graph structure, abstracting away from the
-generating process and focussing on the ARG topology without considering
-branch lengths.
-The first prominent problem was calculating (lower bounds on) the minimum number of
-recombinations required to reconstruct a valid genealogy for a given
-sample~\citep{myers2003bounds}, and constructing the corresponding
-minimal (parsimonious)
-ARGs~\citep{song2003parsimonious,song2005efficient,lyngso2005minimum}.
-These problems are NP-hard in general~\citep{wang2001perfect}, and progress has
-been achieved through studying various constrained special cases of ARGs~\citep[e.g.][]{gusfield2004optimal} and
-other more general types of phylogenetic networks~\citep{huson2010phylogenetic}. The
-focus has been on algorithmic and combinatorial results, that largely do not
-have direct relevance to the inference problems described above.
-
-%It is important to note that parsimonious ARGs
-%have quite different properties from realisations of the
-%ARG \emph{process} as modelled by Hudson, Griffiths, and colleagues.
-%Firstly, branch lengths are not typically estimated,
-%% why do we need the following two lines
-%and node times are required in order to compute a likelihood under the
-%coalescent with recombination (see section XXX).
-%Secondly, even if branch lengths were estimated,
-%the realisations would have a very small likelihood since
-%the coalescent with recombination is often highly \emph{un}parsimonious
-%in terms of recombination events.
-
-The goal of this historical overview is to illustrate that the meaning of the term ``ARG" now strongly
-depends on the context in which it is used, and can mean both the
-stochastic process generating genealogies in the presence of
-recombination~\citep[e.g.][]{nordborg2000linkage,birkner2013ancestral,
-wilton2015smc,griffiths2016coalescent},
-as well as, more commonly, the concrete realisation of ancestry from a
-process~\citep[e.g.][]{gusfield2014recombinatorics,mathieson2020ancestry,brandt2021evaluation}.
-\cite{minichiello2006mapping} explicitly distinguished the ARG as a
-stochastic process from the ARG as a data structure, and
-argued for using the term to mean specific realised ancestral histories.
-Unless otherwise stated, we will use the ARG-as-a-data-structure
-interpretation for the remainder of this paper.
-
-% FIXME putting this here for now just to see if this will work.
-
-\section*{Event ARGs}\label{eARG}
 In keeping with its original derivation in terms of a stochastic
@@
 
-This ordering requirement, while straightforward
-to describe, has some practical drawbacks. For example,
-% Surely the meaning is clear from the context here and we don't need
-% qualify the meaning of ``ARG''
-when using this representation and simulating an ARG backwards in time,
-the first event older than a recombination may involve the lineage carrying the
-ancestry to the right of the breakpoint rather
-than the left, and so we cannot emit edges as they are generated.
-Such issues can be worked around, of course, but depending on the ordering of
-otherwise indistinguishable objects is generally problematic.
-
+[TODO list the salient points that we want to contrast against
+the gARG. We don't want this to seem like a straw man]
+
+\section*{Inheritance and ancestral material}\label{Ancestry_resolution}
+The edges in a gARG represent a path of genetic inheritance from
+ancestor to descendant through some
+number of generations (ultimately from cell to cell). The edges are ``annotated''
+with the genome coordinates over which inheritance occurs.
+There are two different ways in which we use these inheritance
+annotations, and the distinction is subtle and important. The difference
+is illustrated in Fig~\ref{fig-ancestry-resolution}.
+
+% TODO majorly condense this caption, it's much too long.
 \begin{figure}
 \centering
-% \begin{tabular}{cc}
-\begin{tikzpicture}[x=5mm, y=5mm, node distance=2mm and 20mm]
-\tikzset{greynode/.style={circle,fill,inner sep=1},
-nodelabel/.style={font=\footnotesize}}
-
-\node (s0) [greynode] at (0, 0) {};
-\node (s1) [greynode] at (3, 0) {};
-\node (s2) [greynode] at (6, 0) {};
-\node (s3) [greynode] at (3, 1) {};
-\node (s4) [greynode] at (1, 2) {};
-\node (s5) [greynode] at (5, 3) {};
-\node (s6) [greynode] at (3, 4) {};
-
-% \node [anchor=north west] at (0,6) {A};
-\node [nodelabel,anchor=north west] at ($(s3) + (0,0)$) {$x$};
-\foreach \u/\lab in {s0/$\textsf{a}$, s1/$\textsf{b}$, s2/$\textsf{c}$} \node[nodelabel,anchor=north] at (\u) {\lab};
-\foreach \u/\lab in {s4/$\textsf{e}$} \node[nodelabel,anchor=south west] at (\u) {\lab};
-\foreach \u/\lab in {s5/$\textsf{f}$} \node[nodelabel,anchor=south east] at (\u) {\lab};
-\foreach \u/\lab in {s3/$\textsf{d}$, s6/$\textsf{g}$} \node[nodelabel,anchor=south] at (\u) {\lab};
-
-%% Edges
-\draw (s1) -- (s3);
-\draw (s0) |- (s4);
-\draw (s4) -- (2,2) |- (s3);
-\draw (s4) |- (s6);
-\draw (s3) -- (4,1) |- (s5);
-\draw (s2) |- (s5);
-\draw (s5) |- (s6);
-
-% \node [anchor=north west] at (9,6) {B};
-\node [nodelabel,anchor=north west] at ($(10,5)$) {
-\begin{tabular}{c|c|l}
-% \multicolumn{2}{c}{Breakpoints}\\
-Node & Breakpoint & Parents\\
-\hline
-$\noderef{a}$ & $\varnothing$ & [\noderef{e}]\\
-$\noderef{b}$ & $\varnothing$ & [\noderef{d}] \\
-$\noderef{c}$ & $\varnothing$ & [\noderef{f}]\\
-$\noderef{d}$ & $x$ & [\noderef{e}, \noderef{f}] \\
-$\noderef{e}$ & $\varnothing$ & [\noderef{g}]\\
-$\noderef{f}$ & $\varnothing$ & [\noderef{g}]\\
-$\noderef{g}$ & $\varnothing$ & []\\
-\end{tabular}};
-
-% \node [anchor=north west] at (9,6) {C};
-% \node [nodelabel,anchor=north west] at ($(10,5)$) {
-% \begin{tabular}{ll}
-% Edges & Intervals\\
-% \hline
-% $(\textsf{a}, \textsf{e})$ & $(0, L]$ \\
-% $(\textsf{b}, \textsf{d})$ & $(0, L]$ \\
-% $(\textsf{c}, \textsf{f})$ & $(0, L]$ \\
-% $(\textsf{d}, \textsf{e})$ & $(0, x]$ \\
-% $(\textsf{d}, \textsf{f})$ & $(x, L]$ \\
-% $(\textsf{e}, \textsf{g})$ & $(0, L]$ \\
-% $(\textsf{f}, \textsf{g})$ & $(0, L]$ \\
-% \end{tabular}};
-
-\end{tikzpicture}
-\caption{\label{fig-arg-data-structure}
-A graphical depiction of a classical event ARG and a concrete
-encoding of the information. Node \noderef{d} is a recombination
-event as it is associated with a non-null breakpoint, and has
-two parents. Nodes \noderef{e}, \noderef{f} and \noderef{g} are
-common ancestor events.
+\includegraphics[width=\textwidth]{illustrations/ancestry-resolution}
+\caption{\label{fig-ancestry-resolution}
+TODO DELETE A? Or keep it? Maybe this would help some hard-core eARGers?
+The \citet[][Fig.~1]{wiuf1999recombination} example graph as an eARG and
+two gARGs with alternative edge annotation schemes.
+% (A) A conventional Griffiths eARG. Square nodes represent events, with
+% each recombination (red) annotated with the position of a single breakpoint
+% along the 7-unit-long chromosome.
+(A) A gARG with edges annotated by . Oval nodes represent
+genomes and annotations reflect ``local'' transmission of all genetic material
+between parent and child. Edges that do not span the full genome are visually
+highlighted, showing how breakpoint positions can be encoded.
+(C) A gARG with sample-resolved edge annotations. The Hudson-like simplification
+algorithm has been used to shorten edge spans to
+reflect only the upwards transmission of genetic material from the samples
+\textsf{a}, \textsf{b}, and \textsf{c}; very few edges then span the entire genome.
+The depiction of the genome within each
+node is shaded as in Fig.~\ref{fig-arg-in-pedigree}A to show coalescing
+regions of genome back to the MRCA nodes \textsf{k} (regions 0 to 2) and \textsf{p}
+(regions 2 to 7). Line widths reflect the genomic span of each edge, highlighting
+the relative importance of different transmission pathways to the samples. Nodes
+\textsf{g},  \textsf{j}, \textsf{o} and \textsf{q} are redundant in that they correspond
+to events which have no impact on the genetic ancestry of the samples.
 }
@@
 
-% What do we mean by an eARG, formally?
-To be concrete, we can formally define the
-classical Griffiths eARG as a tuple $(e, \sigma)$, where $e$
-is an ordered list of edges defining a directed acyclic graph and
-$\sigma: \mathbb{N} \rightarrow \mathbb{N}$
-is a function mapping nodes to recombination breakpoints,
-such that $\sigma(u) = x$
-if $u$ is a recombination event with breakpoint $x$ and
-$\sigma(u) = \varnothing$ otherwise.
-(This is equivalent to having different ``types''
-of node, where breakpoints are associated with recombination
-nodes only).
-Each edge $e_j = (c, p)$ describes an edge between
-child node $c$ and parent $p$, where $c, p \in \mathbb{N}$.
-% For each recombination node $u$ we must
-% have exactly two edges in $e$ such that $u$ is the child.
-This encoding of an eARG is equivalent to the output
-of programs such as
-\texttt{ARGweaver}~\citep{rasmussen2014genome}
-and \texttt{KwARG}~\citep{ignatieva2021kwarg},
-and an example is given in Fig.~\ref{fig-arg-data-structure}A.
-Note that the description above captures only the
-graph topology: if we also wish to know the branch lengths we need
-an additional function $t(u): \mathbb{N} \rightarrow \mathbb{R}$
-which defines the time of each node.
+The simplest interpretation of the inheritance intervals $I$ for
+an ancestor genome $v$ and descendant $u$, is that $I$ defines the
+regions of genome that $u$ inherited from $v$. For example,
+node $o$ is the recombinant descendant of nodes $p$ and $q$ in
+Figure~\ref{fig-ancestry-resolution}A, which shows that it in inherited
+everything to the left of position $1$ from $p$ and the rest of
+its genome from node $q$.
+Thus, these edge annotations are concerned only with the genetic relationship
+between the nodein question, and are not affected by the rest of the graph.
+We will thefore refer to inheritance annotations of this type as ``local inheritance
+annotations''. Graphs of this type result from forwards-time
+simulations~\citep{kelleher2018efficient,haller2018tree} or the
+``Big ARG'' stochastic process (Appendix XXX).
+
+ARGs are most often thought of as a \emph{retrospective} structure however,
+defined with respect to some set of (usually, contemporary) sample genomes.
+In this case we are not explicitly interested in the
+entire interval of genome descendant $u$ inherited from ancestor $v$,
+but rather the subset of that interval that is \emph{ancestral to our
+sample}.
+Fig~\ref{fig-ancestry-resolution}B shows the graph that we obtain when
+reduce the annotation intervals to those that are ancestral to the
+sample (note that node $q$ is now ommitted because the MRCA of the sample
+has already been found for the full genome at node $p$).
+Tracking such
+``ancestral material''~\citep{wiuf1999ancestry,wiuf1999recombination}
+is much more useful~\citep{mcgill2013graphml,arenas2013importance}
+and, indeed, all that is possible to
+have information about when inferring ARGs from contemporary samples.
+We refer to a gARG with inheritance annotations of this type as
+``sample resolved''.
+% Keeping track of resolved ancestral material is one of the distinguishing
+% features of ARGs vs phylogenetic
+% networks~\citep{mcgill2013graphml,arenas2013importance}, but
+% is actually incompatible with the Griffiths gARG encoding (it does this
+% implicitly, I guess)
+
+% Fig~\ref{fig-ancestry-resolution} is an instructive example
+% from the literature \citep{wiuf1999recombination} and illustrates
+% how a classical eARG (Fig~\ref{fig-ancestry-resolution}A) can be described as
+% a gARG with edge annotations (Figs~\ref{fig-ancestry-resolution}B~and~C).
+
+% There are two distinct annotation schemes possible, and the distinction is subtle
+% and important. The first scheme is to use the annotations to describe the complete
+% set of genomic intervals directly inherited by the descendant node from the ancestor
+% node (Fig~\ref{fig-ancestry-resolution}B). The second is to use the edge annotations to
+% describe the ``ancestral material''~\citep{wiuf1999ancestry,wiuf1999recombination}
+% of the sample that is propagated through
+% the graph from descendant to ancestor (Fig~\ref{fig-ancestry-resolution}C).
+% As we discuss in section XXX, this second scheme has many advantages
+% in terms of efficiency.
+
+%ARGs are inherently a retrospective structure defined in terms
+%of a set of sample nodes (usually the leaves), and,
+%as we discuss in section XXX, storing sample-resolved ancestry
+%intervals (Fig~\ref{fig-ancestry-resolution}C) is generally much more
+%useful than storing the parent-child intervals recording direct inheritance
+%(Fig~\ref{fig-ancestry-resolution}B).
+%Indeed, storing direct inheritance intervals for edges is really only of interest because of
+%its correspondence with the classical approach of associating
+%breakpoints with recombination nodes in an eARG (Fig~\ref{fig-ancestry-resolution}A).
+%The eARG encoding is simply a more limited form, in which the only types of direct
+%inheritance we can have are the entire genome, or the intervals
+%at either side of a recombination breakpoint.
+
+%Note that this example is a sample from the so-called \emph{``big'' ARG process}
+%(see Appendix XXX) and has several features that we are unlikely to simulate or
+%infer in practice. For example, by definition the samples contain no genetic information about
+%nodes \textsf{g} and nodes \textsf{j} and so neither of those nodes are likely to be inferred.
+%Likewise, is is unlikely we would have information to infer the ``grand MRCA''
+%\textsf{q} as all regions of the genome have already reached a
+%common ancestor separately (in nodes \textsf{k} and \textsf{p}).
+
+% Third, there exist nodes such as
+% \textsf{k} with two children (i.e. ``common ancestor'' nodes in Hudson's terminology)
+% but in which no ancestral coalescence occurs. Fourth, some breakpoint positions occur
+% within regions of genome that are not ancestral (e.g. the breakpoint at position 3 in
+% node \textsf{k}). In this case sample-resolved edge annotations lose the exact position of
+% the breakpoint, although they retain the information that a breakpoint occurred between
+% positions 2 and 4 (which is sufficient for calculating the likelihood, see appendix XXX).
+
+Given a gARG with direct-inheritance edge annotations
+(Fig~\ref{fig-ancestry-resolution}B),
+ we obtain the sample-resolved edge annotation
+(Fig~\ref{fig-ancestry-resolution}C) by traversing the graph backwards in time
+from sample nodes (\textsf{a}, \textsf{b}, \textsf{c}).
+Proceeding in topological order (such that an ancestor node is visited
+after all of its descendants) the sample-resolved material for a node is the
+union of the sample-resolved intervals on its inbound edges. We then annotate
+the outbound edges with the intersection of their direct-inheritance intervals
+and the just-computed sample-resolved intervals for the node.
+For example, node \textsf{i} in Fig~\ref{fig-ancestry-resolution}C
+carries sample-resolved material on the intervals $(0, 2]$ and $(4, 7]$. The edge
+joining \textsf{i} to its ancestor \textsf{k} then carries
+material on $(0, 2]$ since although \textsf{i} directly inherits interval
+$(0, 3]$ from \textsf{k} (Fig~\ref{fig-ancestry-resolution}B), only $(0, 2]$ is
+ancestral to the sample. For each node, we also keep track of the number of
+descendant samples over each genomic interval, and stop tracking that
+interval once the lineages from all samples have coalesced. For example,
+the edge from \textsf{k} to \textsf{n} omits the region $(0,2]$ because it has
+already fully coalesced (indicated by a fully black genomic region).
+
+% In practice, the backwards-in-time Griffiths process that produces
+% ARGs such as that shown in Figs~\ref{fig-ancestry-resolution}A~and~B
+% is rarely simulated, because the resulting ``big ARG'' (Appendix XXX) is
+% dominated by nodes such as \textsf{g} and \textsf{j} which contain information
+% about inheritance in non-ancestral material, as well as nodes such as
+% \textsf{o} and \textsf{q} which track details of ancestry even once the
+% local genomic region has fully coalesced. Instead, simulating
+% ARGs from the coalescent with recombination is usually done by
+% running Hudson's algorithm to produce a sample-resolved ARG directly.
+% Moreover, when inferring ARGs from genetic data, there is no information
+% in the sample set that can be used to resolve these nodes, so that they
+% are essentially unknowable.
+
+The process of converting a gARG with local-inheritance annotations
+to its sample-resolved equivalent is also known as ``simplification'',
+and is the key algorithm underlying recent
+advances in forwards-in-time
+simulation~\citep{kelleher2018efficient,haller2018tree}.
+It has many useful properties, as we discuss in the next section.
+
+% It is worth noting the process of sample resolution can reveal that certain ARG
+% nodes are \emph{redundant}: if they are removed and the edges adjusted to
+% pas directly from the node's child to the node's parent, it does not reduce our
+% knowledge of the samples' genetic ancestry. Removing nodes,
+% both redundant ones and others, is part of a process we call
+% ``simplification''.
+
+\begin{figure}
+\centering
+\vspace{5em}
+\includegraphics[width=\linewidth]{illustrations/simplification}
+\caption{\label{fig-simplification}
+ARG simplification. In all cases the graph visualization is shown on the left
+and the equivalent local trees on the right. Since multiple parents are indicative of
+recombination(s), nodes are coloured by the number of parents. and shaded
+according to the proportion of their span over which they are coalescent.
+(A) A gARG simulated from a diploid Wright-Fisher
+model. Note that the genome \textsf{k} has three children, and the genomes
+\textsf{e} and \textsf{g} simultaneously have multiple parents and multiple
+children.
+(B) Simplified to remove all
+% trying out this terminology - can define in the text
+1-connected graph components (e.g., diamonds such as \textsf{jlnm}).
+(C) Remove nodes that are unary in all local trees. This removes both recombinant nodes
+that have only one child (e.g. \textsf{n}) and nodes that have more
+than one child (i.e. are ``common ancestors'') but never represent coalescences
+in the local trees (e.g. \textsf{r}).
+(D) Rewrite edges to bypass any unary nodes in local trees. Note that this loses lineage
+information such as the fact that \textsf{k} is connected to \textsf{a} via \textsf{e}.
+}
+\end{figure}
+
+\section*{Simplification}\label{ARG_simplification}
+% I think a sentence is needed here on why "so much" text is devoted to simplification.
+% How does simplification connect to eARG vs gARG and practical benefits of gARG over eARG
+% and why do you now talk about unary nodes and a bit polytomies? Hold my hand here please;)
+ARG simplification~\citep{kelleher2018efficient} is the process
+of removing nodes and re-writing edges in a gARG to remove
+various types of redundancy,
+retaining the ``meaningful'' topology within local trees
+and maintaining the identity of nodes across different trees.
+The idea of meaningful local tree topologies is based on the
+presence of ``unary'' nodes: a tree node that has exactly
+% should presence above be absence?
+one child. Fig.~\ref{fig-simplification} show a series of successive steps of
+simplification, starting with a complete gARG simulated under a
+(backwards-in-time) Wright-Fisher
+model, shown both as a graph on the left, and the equivalent set of 6
+local trees on the right.
+For example, consider the path above node
+\textsf{a} in the first tree (covering positions 30 to 40)
+in Fig.~\ref{fig-simplification}A.
+Before \textsf{a} meets its common ancestor with the other samples at
+node \textsf{q}, it passes through \textsf{e}, \textsf{j}, \textsf{m}, \textsf{n} and
+\textsf{p}, which in this region of the genome are all unary nodes.
+This reflects the passage of this span of genome through the ARG towards the root.
+Such unary nodes in a local tree are not usually considered important;
+% because they provide little information and are not
+% detectable by information from any given individual tree.
+nodes usually mark the points where samples coalesce in a
+ \emph{common} ancestor. For most purposes, we would therefore consider a path directly
+joining sample \textsf{a} to node \textsf{q} to be equivalent to the path
+containing the intermediate unary nodes. Simplification is the process of removing
+these less informative nodes from the local trees by modifying the gARG.
+
+The first type of simplification that we can perform is to remove
+graph topology that is invisible to the samples.
+An example of such topology is a
+``diamond''~\citep{rasmussen2014genome}
+in which the two parent nodes of a recombination immediately
+join again into a common ancestor (e.g.~\textsf{j}, \textsf{l}, \textsf{m}
+and \textsf{n} in Fig.~\ref{fig-simplification}A).
+Unless we are specifically
+interested in the recombination event or ancestral genomes,
+there is no information in this topology and the diamond can be
+replaced by a single edge. More generally, any
+subgraph that is singly-connected in both the leafward and
+rootward direction (a ``super-diamond'') is non-identifiable and can be
+replaced by one edge. This definition includes the case
+of a node that has one inbound and one outbound edge, such as
+nodes \textsf{f} and \textsf{h}.
+Fig.~\ref{fig-simplification}B shows the result of this type of
+graph topology simplification.
+
+Simplifying away diamonds will remove many unary nodes from the
+local trees, but there can still be nodes that are unary in all
+of the local trees. In particular, a node can represent a recombinant
+with multiple parents in the graph but only a single child (e.g.\ node \textsf{n}
+in Fig.~\ref{fig-simplification}B), or can represent a common ancestor with
+multiple children in the graph but in which no local coalescence takes place
+(node \textsf{r} in Fig.~\ref{fig-simplification}B).
+% point out that this is why Hudson refers to CA nodes, not coalescence nodes
+Such nodes are not singly connected in the graph, but are nevertheless unary in
+all of the local trees.
+The operation to remove them
+therefore requires knowledge not just of the graph topology but also of the
+edge annotations.
+As we see in Fig.~\ref{fig-simplification}C,
+removal of recombinant nodes can produce graph nodes with
+more than two parents (e.g.~node \textsf{e}); and likewise, removal of
+common ancestor but non-coalescent nodes can produce graph nodes with
+more than two children (e.g.~node \textsf{s}). These represent a ``stacking up'' of
+multiple---often indistinguishable---events into a single node.
+% something which is impossible to encode in an eARG.
+
+% JK: Text moved from an earlier section. See if we want to reuse this here
+% if not already said.
+% Indeed a single node can have both multiple children
+% and multiple parents. This would be the case in Fig~\ref{fig-arg-in-pedigree} if,
+% for instance, node \textsf{a} were to have two children.
+% % It is a shame that this is not actually shown in fig 3, but I don't think
+% % we can do this unless we have 5 generations
+% Moreover, a single node can have more than 2 children, and (as we shall see in
+% the ``\nameref{ARG_simplification}'' section), more than 2 parents. Another way to
+% state this is that transitions between nodes in a gARG can encapsulate multiple
+% events in an eARG.
+
+The remaining nodes are MRCAs of some subset of the samples
+at \emph{some} positions along the genome. We still have
+some unary nodes in the local trees, but these nodes will
+correspond to a coalescence in at least one other
+local tree. For example, node  \textsf{k} is unary in the second tree
+of Fig.~\ref{fig-simplification}C, but is either binary
+or ternary in all subsequent trees (recall this is a Wright-Fisher
+simulation). The final level of simplification is to alter the edge annotations
+such that, although no nodes are removed from the graph, all
+unary nodes disappear from the local trees (Fig.~\ref{fig-simplification}D).
+Note that although this last stage produces simpler local trees, by
+removing information about the exact paths taken by lineages through
+the graph, we lose potentially useful information about shared edges
+between trees. The extent to which this information can be inferred
+and utilised is an interesting open question.
+
+% Needs another pass
+An important consequence of simplifying local trees to remove
+unary nodes is that we may lose the ability to identify
+the ``subtree prune-and-regraft'' (SPR) operations
+by which they differ~\citep{hein1990reconstructing,song2003on,song2006properties}.
+% The effects of recombination as we move from left-to-right along
+% the genome in the complete simulated gARG
+% of Fig~\ref{fig-simplification}A are easy to interpret,
+% but as more redundancy is removed from the graph
+% the
+As explained above, we can have multiple recombination and
+common ancestor events stacking up, and therefore
+many possible SPR operations. Deducing the SPRs between
+trees is NP-hard in
+general~\citep{hein1996complexity,allen2001subtree,bordewich2005computational},
+but effective heuristics have been proposed~\citep{rasmussen2022espalier}.
+
+% JK TODO: finish up this section with a quick version of the
+% para here. Basically say, people often say an ARG is a
+% set of local trees cite cite, but it's not as simple as that as we
+% can see from the example (probably).
+
+% The equivalence between an ARG and a set of local trees
+% separated by SPRs is often mentioned [e.g. x y z], but
+% it is important to note that while extracting local
+% trees from an ARG is straightforward, the converse
+% process of reconstructing precisely the same ARG from
+% these trees is more problematic. If the internal
+% nodes are consistently labelled and unary nodes are
+% included in the local trees (e.g., Fig~\ref{fig-simplification})
+% then it is easy to exactly reconstruct the ARG.
+% Otherwise, the problem is much more difficult and it
+% may not be possible to uniquely reconstruct the original ARG.
+% Without labels for the internal nodes, we are forced to
+% to reason about which subtrees remain the same and which
+% differ due to the SPR separating adjacent trees in order to
+% assign node identity. Without the internal unary nodes marking
+% recombinations, we must infer their existence. In
+% general, there will not be a unique solution to this problem,
+% and so there can be many possible ARGs compatible with
+% a given list of leaf-labelled binary trees.
+% This is the approach taken by the Espalier inference
+% method~\citep{rasmussen2022espalier}, which finds a
+% set of SPRs that reconcile independently inferred
+% trees from different regions of the genome.
+
+% NEEDS more work!
+\section*{Precision of ancestry information}
+Ancestral recombination graphs are often characterised
+as consisting of a complete record of the genetic history
+of a sample. For example,
+\cite{rasmussen2014genome} state that an ``ARG provides a record of all
+coalescence and recombination events since the divergence of the sequences
+under study'' and
+according to \cite{deng2021distribution} an ARG
+``provides all the information about the genealogical history of a sample,
+including the locations of recombination events.'' It is worth
+questioning, however, whether such comprehensive information
+will always be available, and in particular whether our
+representation of genealogical history should \emph{require}
+such precision. Simulations will usually generate complete
+information about the history of a sample, of course, but
+we may not have sufficient information to infer such detail.
+
+% The full ancestral recombination graph (ARG) is a structure that encodes all
+% coalescence and recombination events resulting from the stochastic process of
+% the coalescent with recombination.
+\citet{brandt2021evaluation} define a ``full'' ARG as
+``a structure that encodes all
+coalescence and recombination events resulting from the stochastic process of
+the coalescent with recombination'' (thus going against the general
+historial trend discussed in the XXX section).
+\citet{rasmussen2022espalier} have a less restrictive definition, in
+which a ``full'' ARG is one that contains complete information about
+recombination events, without necessarily being tied to a particular
+stochastic process.
+While [something positive], it is important to be clear on
+what the overall goal of such precise inferences are,
+what the limitations on when they can be made are,
+and their downstream utility in applications.
+
+% What is the goal of having fully precise ARGs?
+% 1) studying the actual recombination events;
+% 2) evaluating the likelihoods
+The main reasons for wanting precise estimates of all recombination
+events are to either compute a likelihood of the inferred ARG
+under a statistical model, or to study the recombinants in
+detail~\citep{rasmussen2022espalier}.
+% Explain that we're really interested in the details of recombination
+% itself with a small number of samples.
+In the latter case, we might
+have a small number of samples~\citep{guo2022recombination}.
+
+Beyond cases in which visual inspection of the trees and the
+effects of recombination is feasible, the main application
+of having completely precise estimates of recombination events is
+to evaluate the likelihood of the estimate under some
+statistical model. The only statistical model available
+is the coalescent with recombination
+and its approximations~\citep{mcvean2005approximating,marjoram2006fast}.
+In this case, the utility of a ``full`` ARG for some dataset
+is limited by how well the dataset fits the assumptions
+of the model.
+
+% Not sure where this goes, but there are some critical points
+% that we want to explain and not just throw out there in the
+% discussion.
+% A critical distinction between the eARG and gARG encodings
+% is the level of precision about the time and location of
+% recombination events that is required. In an eARG, we are
+% required to be absolutely precise about every recombination
+% [explain and contrast with gARG, referring to examples]
+
+% JK: This is very rough, just jotting down the basic ideas,
+% will need substantial revision.
+It is not clear that such complete information will always
+(or indeed ever) be available from inferences~\cite{shipilina2022origin}.
+ We may be
+able to precisely identify the location and timing
+of recent recombinations (particularly if detailed pedigree
+information is available) but beyond this, the information
+required simply isn't present. When sampling from a
+process such as the Sequentially Markov
+Coalescent~\citep{mcvean2005approximating,marjoram2006fast}
+like ARGweaver~\citep{rasmussen2014genome}, it is true that
+we do obtain a fully resolved ARG with precise
+information about recombination, but it is reasonable
+to question how much information from the data is used
+to support older recombinations, in particular.
+
+% This idea of a continuum of information about recombination
+% being inferrable and representable is consistent with the ideas
+% discussed by~\cite{shipilina2022origin}. There a haplotype block
+% is synonymous with coalescent nodes in the ARG, which we can generalise
+% as nodes in a gARG with at least two children. In principle, these
+% nodes are identifiable by the mutations that occur on the outbound
+% edges of the node.
+
+It is helpful to consider the difference between fully
+resolved binary trees and those that contain polytomies.
+
+The situation is in fact directly analogous: an ARG that
+contains fully resolved recombination information must have
+at most parents per node.
+
+At the very least, it is not reasonable for the \emph{data
+structure} to require complete precision in order to
+represent some less precise. This is a fundamental
+advantage of the gARG encoding over the classical eARG
+method: we are not forced to posit a precise cause (recombination
+event) for every effect (difference between adjacent trees).
+These differences in precision of recombination
+information are illustrated in the ARGs estimated by different
+methods, as we see in the next section.
+
+In general, discussions tend to contrast an ARG
+on one hand as holding \emph{complete} information about recombination
+(and therefore between-tree correlation structure)
+and a sequence of local trees on the other (with, by implication,
+no direct information about between-tree correlations).
+This is a false dichotomy:
+a \emph{continuum} of information about recombination and between-tree
+correlation structure can be represented and inferred, and all
+such levels of information are useful, even if only for computational
+efficiency.
+
+\begin{figure} \begin{center}
+\includegraphics[width=\textwidth]{illustrations/inference.pdf} \end{center}
+\caption{\label{fig-inferred-args} ARGs inferred by
+(A) \kwarg~\citep{ignatieva2021kwarg},
+(B) \argweaver~\citep{rasmussen2014genome,hubisz2020inference},
+(C) \tsinfer~\citep{kelleher2019inferring},
+and (D) \relate~\citep{speidel2019method}
+for 11 samples of a 2.7kb region of the \textit{Drosophila melanogaster} ADH
+locus~\citep{kreitman1983nucleotide}. The \argweaver\ example
+is a sample from the inferred posterior distribution. In each ARG,
+colours are assigned to edges based on the time of the edge's child node:
+lighter colours are assigned to edges with older child
+nodes, darker colours are assigned to edges with younger child nodes
+Note that the position of points on the
+Y axis is chosen for graphical clarity by the graphviz \texttt{dot}
+algorithm, rather than as a direct measure of time.
+Line width corresponds to the genomic span of the edge. Node colour
+refers to the number of parents of the node: red is zero, brown is one,
+green to blue are two to N parents. The brightness
+of the node is determined by the genomic span over which
+the node has two or more children divided by the span where it has any
+children. The genomic intervals associated with edges are visualised below
+each ARG, where each row is an edge
+and the x axis reflects genomic position. The width of edges
+thus corresponds to their genomic span. Edges
+are ordered by the age of their child node, from oldest to youngest
+moving down the page, with colours matching the edge colours in the ARG.
+Vertical lines indicate breakpoints between local trees.
+}
+\end{figure}
+
+To illustrate these points and to emphasise the heterogeneity
+of the ARGs inferred by current methods, we inferred ARGs for the
+the classical \citet{kreitman1983nucleotide} benchmark dataset
+using four recent inference methods and show the results in
+Fig~\ref{fig-inferred-args}.
+While there is some consistency among the inferred ARGs (such as the samples
+FR-F, WA-F and Af-F being grouped into the same clade), there
+are also considerable differences. For this dataset, the minimum number
+of recombination breakpoints required in the absence of recurrent mutation
+is 7~\citep{song2003parsimonious}, and the number of inferred breakpoints
+under the input mutation and recombination rates
+varies from 2 (\relate) to 17 (\argweaver).
+The ARGs inferred by the different methods also have major differences
+in how recombination information is represented. Essentially,
+\kwarg\ and \argweaver\ infer a specific causal recombination \emph{event}
+for each breakpoint, and those events are fully
+described in the output graph. On the other hand, \tsinfer\ and
+\relate\ do not infer causal events but instead focus on
+representing the inferred \emph{effect}: breakpoints between the
+trees. It is important to note that this does not imply that
+the methods must then necessarily infer only a sequence of
+unrelated local trees
+% More? See https://github.com/tskit-dev/what-is-an-arg-paper/issues/38
+% for discussion. I'm sure there were other examples of people saying
+% that
+(as claimed by e.g.~\citet{hejase2020summary}),
+and information about recombination and the correlation structure
+between trees is not ``all or nothing''.
+Both methods capture correlation structure between trees:
+\relate\ first infers local tree topologies and then identifies edges
+that persist from one tree to the next as a separate step
+(in a similar manner to \espalier~\citep{rasmussen2022espalier}), while
+\texttt{tsinfer} explicitly constructs nodes that are shared across local
+trees. See \cite{ignatieva2023distribution} for a detailed discussion
+and comparsion of inference methods on this property.
+
+% Because the ARGs inferred by \tsinfer\ and \relate\ do not contain
+% explicit recombination events we cannot compute a likelihood
+% under the coalescent with recombination (see appendix XXX).
+% [Some stuff about whether we should see this as necessary or not,
+% but maybe we should keep this for the conclusion?]
+
+% \section*{When do we need ``full'' ARGs?}
+% % The full ancestral recombination graph (ARG) is a structure that encodes all
+% % coalescence and recombination events resulting from the stochastic process of
+% % the coalescent with recombination.
+% \citet{brandt2021evaluation} define a ``full'' ARG as
+% ``a structure that encodes all
+% coalescence and recombination events resulting from the stochastic process of
+% the coalescent with recombination'' (thus going against the general
+% historial trend discussed in the XXX section).
+% \citet{rasmussen2022espalier} have a less restrictive definition, in
+% which a ``full'' ARG is one that contains complete information about
+% recombination events, without necessarily being tied to a particular
+% stochastic process.
+% While [something positive], it is important to be clear on
+% what the overall goal of such precise inferences are,
+% what the limitations on when they can be made are,
+% and their downstream utility in applications.
+
+% % What is the goal of having fully precise ARGs?
+% % 1) studying the actual recombination events;
+% % 2) evaluating the likelihoods
+% The main reasons for wanting precise estimates of all recombination
+% events are to either compute a likelihood of the inferred ARG
+% under a statistical model, or to study the recombinants in
+% detail~\citep{rasmussen2022espalier}.
+% % Explain that we're really interested in the details of recombination
+% % itself with a small number of samples.
+% In the latter case, we might
+% have a small number of samples~\citep{guo2022recombination}.
+
+% Beyond cases in which visual inspection of the trees and the
+% effects of recombination is feasible, the main application
+% of having completely precise estimates of recombination events is
+% to evaluate the likelihood of the estimate under some
+% statistical model. The only statistical model available
+% is the coalescent with recombination
+% and its approximations~\citep{mcvean2005approximating,marjoram2006fast}.
+% In this case, the utility of a ``full`` ARG for some dataset
+% is limited by how well the dataset fits the assumptions
+% of the model.
+
+% % If we want to use the coalescent, then really the assumptions
+% % need to make sense.
+% % n << Ne is a silly assumption in today's datasets
+% A core assumption of the coalescent is that the sample size $n$
+% is much less than the effective population size, $N_e$.
+% Several human datasets now consist of hundreds of thousands of
+% genomes~\citep{bycroft2018genome,karczewski2020mutational,tanjo2021practical},
+% and so sample size is substantially \emph{larger} than $N_e$
+% (often assumed to be $10^4$ in humans).
+% Agricultural datasets are even more extreme.
+% For example,
+% the US dairy cattle database alone currently comprises more than 6 million
+% animals with SNP array
+% genotypes\footnote{\url{https://queries.uscdcb.com/Genotype/counts.html}},
+% while the effective population size in modern dairy cattle breeds is
+% less than 100 and decreasing~\citep{MacLeod2013,Makanjuola2020}.
+% An extreme sign of these breeding practices is that there are only two ancestral
+% Y-chromosome lineages present in today's US Holstein dairy breed~\citep{Yue2015}.
+% Simlarly, the 1000 bull genomes project~\citep{hayes20191000}
+% comprises close to 7000 genomes, which are part of multi-generation pedigrees
+% with millions of animals and extensive SNP array genotype and phenotype
+% data \citep[e.g.][]{Cesarani2022}.
+% In another example, genome and SNP array data for 440,610 individuals within
+% 7 multi-generation pedigrees~\citep{whalen2018,Johnsson2021,Ros-Freixedes2020}
+% were combined to infer recombination in pigs~\citep{RosFreixedes2022}.
+% Effective population size in modern pig breeds is also less than 100 due to
+% intense selection and directed reproduction \citep{Hall2016,Porcnic2016}.
+% Another core assumption of the coalescent model is that the genome (or
+% at least the region under study) is short enough that the number of extant
+% lineages remains much smaller than $N_e$ at all times. Whole
+% genome sequences have been available for model organisms
+% for over a decade now, % True? Citation?
+% and indeed complete chromosome-level assemblies are possible
+% in humans~\citep{miga2020telomere}.
+% Projects are under way to obtain high-quality assemblies
+% for all eukaryotic species in Britain and Ireland~\citep{darwin2022sequence}
+% and ultimately worldwide~\citep{lewin2022earth}.
+
+% While the coalescent as a model can be suprisingly robust to
+% basic violations of the underlying assumptions~\citep{
+% wakeley2012gene,bhaskar2014distortion,nelson2020accounting}
+% it is reasonable to question the validity of the model in
+% representing datasets such as UK Biobank.
+% [Also mention somehow that $n\ll N_e$ is only one problem -
+% we're also ignoring the myriad other evolutionary forces.]
+% Such datasets cannot be reasonably be described by
+% the coalescent as a statical model, and so more non-parametric
+% methods like \tsinfer\ and \relate\ would seem more appropriate.
+% [More]
+
+% % Is it practical to estimate full ARGs?
+% % [Not particularly \citep{shipilina2022origin} - but maybe ARGneedle
+% % has done this for UKB? Need to check]
+
+% % Do downstream applications actually make use of such full ARGs?
+% Downstream applications using estimated ARGs to date rarely
+% make use of such complete recombination information.
+% The \relate\ selection statistic~\citep{speidel2019method}
+% computes a $p$-value for selection [FIXME] for each
+% local tree independently by comparing the rates of coalescence
+% at various time-points. For cases like strong positive selection,
+% where we expect there to be a clear signal in the local tree,
+% such tree-by-tree methods can be simple and powerful.
+% In their method
+% for estimating dispersal rates and the locations of genetic
+% ancestors,
+% \citep{osmond2021estimating} downsample trees along the genome
+% so that they can be regarded as approximately independent.
+% The SIA method for detecting selection~\citep{hejase2022deep}
+% encodes local trees as a set of lineage counts at discrete
+% time intervals, and uses these as feature for a
+% type of machine learning algorithm
+% that takes ``temporal'' correlations into account. Thus,
+% while SIA does use some of the information about local tree correlation,
+% clearly much of the detail about recombination
+% events in a ``full'' ARG is lost.
+% The \texttt{tsdate} algorithm (and related naive estimator
+% for ancestral location) uses much more between-tree
+% information~\citep{wohns2022unified}, explicitly using the gARG
+% encoding to reason about ancestral nodes.
+% A major strength of this approach is that it can be used on
+% any gARG, ranging from estimates containing relatively little
+% information about recombination (e.g., as estimated by \relate)
+% to ``full'' ARGs containing complete recombination information
+% (e.g., as estimated by \kwarg), and will fully utilise
+% any between-tree correlation information that is present.
+
+\section*{Computational efficiency and data interchange}
+ARG inference methods have been successfully applied to vast datasets,
+including 500,000 genotyped humans~\citep{kelleher2019inferring,zhang2023biobank}
+from UK Biobank~\citep{bycroft2018genome}
+and over a million SARS-Cov-2 whole genomes~\citep{zhan2023towards}.
+Even larger datasets are
+available~\citep[e.g.][]{halldorsson2022sequences} % TODO more
+and it seems inevitable that ARGs will be inferred for them.
+The efficiency with which we can store and analyse these inferred
+ARGs is therefore of great importantance, and a key factor
+in the ultimate success of downstream ARG-based methods.
+Another key factor in how important ARG-based methods become
+across the wide range of predicted application
+% TODO more?
+areas~\citep{rasmussen2014genome,harris2019database,
+hejase2020summary,schaefer2021ancestral,
+fan2022genealogical,wohns2022unified,harris2023using}
+is how well inference methods and downstream analysis applications
+interoperate. Without a well-defined and efficient shared format
+for ARG data interchange,
+users will suffer from problems of converting data from one programs
+output to another, an unfortunate hallmark of population genetics
+software for many years~\citep{excoffier2006computer}.
+While current methods tend to tightly couple downstream analysis
+of the inferred ARG with the inference
+itself within the same software package
+% FIXME-  more explicit?
+(e.g.\ Relate, ARGneedle)
+this approach scales poorly in terms of software development
+effort, and is ultimately not compatible with the widespread use
+of ARGs for routine data analysis, and a healthy and diverse
+software ecosystem.
+% FIXME not sure where this goes
+Earlier efforts to standardise ARG interchange
+did not succeed~\citep{cardona2008extended,mcgill2013graphml}.
+
+The gARG encoding discussed and concretely defined in this manuscript
+leads to highly efficient storage and processing of ARG data,
+and has already been in use for several years.
+The succinct tree sequence data structure (usually known as a ``tree sequence''
+for brevity, but see below for confusion around this point)
+is a concrete encoding of a gARG, as discussed in this article.
+It was originally developed as part of the \texttt{msprime}
+simulator~\citep{kelleher2016efficient} and subsequently been
+extended and applied to forward-time
+simulations~\citep{kelleher2018efficient,haller2018tree},
+calculation of population genetics statistics~\citep{ralph2020efficiently}.
+The succinct tree sequence encoding extends the basic definition
+of a gARG provided here by stipulating a
+simple tabular representation and nodes and edges,
+and also defining a concise and lossless representation of
+sequence variation using the ``site'' and  ``mutation'' tables.
+The \texttt{tskit} library~\citep{tskit2023tskit} is a liberally
+licensed open source toolkit that provides a comprehensive suite
+of tools for working with gARGs (encoded as a succinct tree sequence).
+Based on core functionality written
+in C, it provides interfaces in C, Python and Rust.
+Tskit is mature software, widely used in population genetics, and
+has been incorporated into several downstream
+applications~\citep[e.g.,][]{haller2019slim,speidel2019method,
+adrion2020community,
+terasaki2021geonomics,
+baumdicker2021efficient,
+fan2022genealogical,korfmann2022weak,
+mahmoudi2022bayesian,petr2022slendr,rasmussen2022espalier,
+zhang2023biobank}.
+
+% FIXME this needs more work and doesn't really fit into the
+% narrative right now, but we do need to make this point.
+The key insight that makes the succinct tree sequence encoding
+an efficient substrate for defining analysis algorithms is that
+it allows us to generate the local trees along the genome
+in a way that allows us to reason about the \emph{differences}
+between those trees.
+Sequentially generating the marginal trees along the genome
+is also fundamental, and is necessary whenever we need to
+perform calculations that are contingent on more than just the
+isolated properties of an edge. \cite{kelleher2016efficient}
+showed how all trees can be sequentially generated in
+constant time per tree transition in a fully simplified gARG.
+Furthermore, we can easily reason about how tree topologies
+change (and stay the same), leading to efficient algorithms
+for computing population genetic
+statistics~\citep{kelleher2016efficient,kelleher2018efficient},
+implementing the Li and Stephens
+model~\citep{kelleher2019inferring,wohns2022unified}
+and likely many more.
+
+In retrospect, the choice of terminology around the succinct
+tree sequence data structure was unfortunate, and has led to
+significant confusion. As first introduced by~\cite{kelleher2016efficient}
+a ``tree sequence`` was defined as the set of ``coalescence records``
+that is output by Hudson's simulation algorithm.
+This was then generalised to use the
+tabular encoding described above and formally described as a ``succinct
+tree sequence''~\citep{kelleher2018efficient}.
+The ``succinct'' prefix here is intended to connect
+with the concept of succinct data structures
+\citep[e.g.][]{gog2014theory},
+which have near-optimal
+space usage but still support efficient retrieval and computation.
+The concrete data structures of the succinct tree sequence, precisely
+defined and with a particular focus on computational efficiency
+did not have an obvious connection to the ARG definitions
+provided by Griffiths and
+colleagues~\citep{griffiths1991two,ethier1990two,
+griffiths1996ancestral,griffiths1997ancestral}, and with the lack
+of any other precise definitions, it seemed best to avoid the
+term to prevent confusion. Unfortunately, however, the opposite
+has occured. In describing the output of \tsinfer\ as a ``tree
+sequence''~\citep{kelleher2019inferring} many
+have---not unreasonably, but incorrectly---concluded that the
+output is a collection of \emph{independent}
+% MORE?
+% https://github.com/tskit-dev/what-is-an-arg-paper/issues/38
+trees~\citep[e.g.,][]{hejase2020summary,ignatieva2021kwarg}.
+% Could say more here, but maybe we'll have a note about terminology
+% in the discussion
+
+\section*{Discussion}
+Recent breakthroughs have made ARG inference methods practical
+for the first time, and there has been a surge of interest
+in inference methods, their evaluation and application.
+The prospect of ARGs being used routinely in population
+and statistical genetics applications is tantalising,
+but in reality there is  substantial work to be done to
+enable this, and much remains to be done.
+
+A necessary first step towards this goal is some terminological
+clarity. As discussed in Appendix XXX, the term ARG has several
+subtly different interpretations depending on context. The
+trend to decouple the ARG from its origins as a stochastic
+process and instead use the term as a more general representation of any
+recombinant genetic ancestry seems most useful, which we have
+tried to clarify and systematise here. Thus
+we can think of an ARG as being any structure that encodes the
+reticulate genetic ancestry of a sample of colinear sequences under
+the influence of recombination. The ``genome'' ARG encoding
+made explicit here is one way we can concretely
+define such recombinant ancestry, which we have shown is both
+flexible and efficient.
+
+The flexiblity of the gARG encoding contrasts with the classical
+``event'' ARG (eARG), which, because of its origins in an idealised
+mathematical model, is more limited in what can be described.
+The necessity of complete precision under the eARG encoding has been
+conflated with the idea of a ``true'' or ``full'' ARG.
+[A discussion of how realistic inferring a completely precise ARG is]
+[Do downstream applications use full ARGs? usually no]
+
+% % Do downstream applications actually make use of such full ARGs?
+% Downstream applications using estimated ARGs to date rarely
+% make use of such complete recombination information.
+% The \relate\ selection statistic~\citep{speidel2019method}
+% computes a $p$-value for selection [FIXME] for each
+% local tree independently by comparing the rates of coalescence
+% at various time-points. For cases like strong positive selection,
+% where we expect there to be a clear signal in the local tree,
+% such tree-by-tree methods can be simple and powerful.
+% In their method
+% for estimating dispersal rates and the locations of genetic
+% ancestors,
+% \citep{osmond2021estimating} downsample trees along the genome
+% so that they can be regarded as approximately independent.
+% The SIA method for detecting selection~\citep{hejase2022deep}
+% encodes local trees as a set of lineage counts at discrete
+% time intervals, and uses these as feature for a
+% type of machine learning algorithm
+% that takes ``temporal'' correlations into account. Thus,
+% while SIA does use some of the information about local tree correlation,
+% clearly much of the detail about recombination
+% events in a ``full'' ARG is lost.
+% The \texttt{tsdate} algorithm (and related naive estimator
+% for ancestral location) uses much more between-tree
+% information~\citep{wohns2022unified}, explicitly using the gARG
+% encoding to reason about ancestral nodes.
+% A major strength of this approach is that it can be used on
+% any gARG, ranging from estimates containing relatively little
+% information about recombination (e.g., as estimated by \relate)
+% to ``full'' ARGs containing complete recombination information
+% (e.g., as estimated by \kwarg), and will fully utilise
+% any between-tree correlation information that is present.
+
+Fully decoupling the general concept of an ARG from the coalescent
+with recombination stochastic process is an important step.
+While the coalescent with recombination is a very useful and
+robust model~\citep{wakeley2012gene,bhaskar2014distortion,nelson2020accounting}
+many modern datasets have properties that grossly
+violate its assumptions.
+For example, several human datasets now consist of hundreds of thousands of
+genomes~\citep{bycroft2018genome,karczewski2020mutational,tanjo2021practical},
+substantially \emph{larger} than the usually assumed $N_e$ values.
+Agricultural datasets are an even more extreme departure from coalescent
+assumptions, with large samples embedded in
+multi-generational pedigrees~\citep{hayes20191000,Ros-Freixedes2020}
+and effective population sizes of 100 and even
+less~\citep{MacLeod2013,Makanjuola2020,Hall2016,Porcnic2016}.
+Finally, the COVID-19 pandemic has provided an example of a dataset that
+is clearly a very poor fit for the coalescent with recombination, but which,
+nevertheless, we would like to estimate ARGs.
+\cite{zhan2023towards} inferred an ARG for 1.27 million SARS-Cov-2 whole
+genomes, and showed how it accurately captures several known properties of
+the pathogen's evolution, and can efficiently and concisely
+encode this vast dataset.
+
+This view of ARGs,
+decoupled from generative models and
+without the hard requirement
+of complete precision on all historical events, may help improve
+methods for inference method evaluation. In the majority of cases,
+ARG inference accuracy is evaluated by comparing some truth simulations
+to the inferred ARGs by the pairwise comparison of local trees along the genome
+using tree metrics as originally proposed by \citet{kuhner2015assessing}.
+While this provides valuable insights into the method's performance,
+it is not clear how well these tree distance metrics reflect
+performance on real data.
+In comparing tree-by-tree along the genome, the effects of recombination
+structure is incorporated in a very indirect manner.
+Tree distance metrics usually have $O(n^2)$ time complexity, and so cannot be
+applied to the very large sample sizes currently of interest.
+\citet{deng2021distribution} derived the distribution of the expected
+waiting distances to tree changes.
+[ \citet{ignatieva2023distribution} DID SOMETHING ELSE]
+\citet{brandt2021evaluation} compared the
+distributions of time to the most recent common ancestor across several
+methods, which is a useful addition but only indirectly captures
+topological comparisons.
+% Not sure this is the right place for this sentence
+All of these comparisons and evaluations have been based on
+highly idealised simulations, and the effects richness of real data
+(beyond simplistic error models) are almost entirely unknown.
+A suite of true ARG distance metrics, that take into account more of the
+global topology and can be applied to large ARGs, would be a
+valuable addition to the field.
+
+\bibliographystyle{plainnat}
+\bibliography{paper}
+
+\setcounter{secnumdepth}{2} % Print out appendix section numbers
+
+% TODO revisit number here
+\section*{Appendix}
+\appendix
+
+\section{Ancestral graphs: a brief history}
+\label{sec-arg-history}
+The coalescent~\citep{kingman1982coalescent,kingman1982genealogy,
+hudson1983testing, tajima1983evolutionary} models the ancestry of a sample of
+genomes under an idealised population model, and provides the theoretical
+underpinning for much of contemporary population genetics.
+It is a stochastic process, where each random realisation
+is a genealogical tree describing the genetic ancestry of the sample.
+Numerous extensions to the model have been
+proposed~\citep{hudson1990gene,hein2004gene,wakely2008coalescent},
+incorporating many evolutionary processes.
+\citet{hudson1983properties}
+first incorporated recombination into the coalescent process,
+providing several fundamental analytical results
+and describing the basic simulation algorithm, still in
+widespread use~\citep{hudson2002generating,kelleher2016efficient,
+baumdicker2021efficient}.
+In the 1990s, Griffiths and colleagues revisited the
+coalescent with recombination from a different perspective,
+formulating it as a stochastic process where each realisation
+is encoded as a graph~\citep{griffiths1991two,ethier1990two,
+griffiths1996ancestral,griffiths1997ancestral}.
+They referred to both the stochastic process and
+its random realisations as the Ancestral Recombination Graph (ARG).
+Although mathematically equivalent, it is
+important to note that the Griffiths and Hudson formulations of
+the coalescent with recombination are not identical;
+in particular, a direct implementation of the ARG process
+as originally described requires exponential time to simulate
+(see Appendix XXX for details). However, ARGs provided a way
+to reason about and infer recombinant ancestry as a single object,
+which was not possible within Hudson's framework, which emphasised
+the collection of local trees along the genome resulting from recombination.
+
+Subsequent work on ARGs proceeded in broadly three main directions: focussing on (1)
+exploring the mathematical properties of the coalescent with recombination and
+related stochastic processes, (2) inferring evolutionary parameters under
+(approximations to) this model, either with or without explicitly reconstructing the
+genealogy of the sample, and (3) treating the ARG as a discrete graph, ignoring the
+generating stochastic process, and studying its properties from a computational and
+algorithmic perspective.
+
+% Mathematical treatment of the CwR and ARGs (and ARG-like stochastic processes)
+An extensive body of work has been developed from
+studying the coalescent with recombination
+% Can we add a couple of the most well known CwR examples here
+and other related
+graph-valued stochastic processes from a mathematical perspective.
+In particular, the Ancestral Selection Graph
+(ASG)~\citep{krone1997ancestral,neuhauser1997genealogy}
+uses a similar approach to model natural selection instead of recombination.
+Unlike the ARG process, the ASG imposes a hard distinction between the stochastic process,
+which constructs a random ARG-like graph, and an observable realisation,
+which is a single tree sampled from the graph in a non-uniform way to encode
+desired patterns of natural selection.
+Constructions of ASG-like stochastic processes encoding various
+forms of selection, often in parallel with recombination or other genetic forces,
+are an area of considerable and ongoing theoretical interest~\citep[e.g.][]{
+neuhauser1999ancestral,
+donnelly1999genealogical,
+fearnhead2001perfect,
+fearnhead2003ancestral,
+etheridge2009coalescent,
+gonzalezcasanova2018duality,
+koskela2019robust}.
+% Could mention the Ancestral Gene Transfer graph here too, within
+% something like "Similarly the Ancestral Gene Transfer Graph [cite]
+% and [other ancestral graphs, if they exist?] model genealogical
+% processes in bacteria as a graph.
+
+% Inference of parameters and explicit ARG reconstruction
+Early work on inference under the coalescent with recombination
+focused on the problem of
+inferring the parameters of the
+stochastic process, where the ancestry was regarded as a
+latent parameter to be averaged out
+\citep[e.g.][]{griffiths1996ancestral,kuhner2000maximum, nielsen2000estimation,
+fearnhead2001estimating}.
+These methods met with limited success
+because the state space of ARGs is overwhelmingly large, and
+lacks a simple geometry or neighbourhood structure for inference or
+sampling methods to  exploit.
+Several breakthroughs in this direction were achieved through
+formulating simplified but more tractable approximations to the full
+model~\citep{mcvean2005approximating,marjoram2006fast,li2011inference,
+paul2011accurate,schiffels2014inferring}.
+The related problem of \emph{sampling} genealogies compatible with a given
+dataset under the coalescent with recombination also proved notoriously difficult
+computationally; progress in explicitly inferring genealogies at scale
+has similarly been achieved through resorting to principled
+approximations~\citep{rasmussen2014genome,mahmoudi2022bayesian},
+or moving away from the coalescent with recombination altogether and seeking
+to infer a single plausible ARG~\citep[e.g.][]{speidel2019method} or ARG
+topology~\citep[e.g.][]{minichiello2006mapping,kelleher2019inferring}.
+
+% Optimisation problems to do with ARGs and other phylogenetic networks
+There has also been substantial interest in formulating and answering
+fundamental questions about properties
+of the ARG as a discrete graph structure, abstracting away from the
+generating process and focussing on the ARG topology without considering
+branch lengths.
+The first prominent problem was calculating (lower bounds on) the minimum number of
+recombinations required to reconstruct a valid genealogy for a given
+sample~\citep{myers2003bounds}, and constructing the corresponding
+minimal (parsimonious)
+ARGs~\citep{song2003parsimonious,song2005efficient,lyngso2005minimum}.
+These problems are NP-hard in general~\citep{wang2001perfect}, and progress has
+been achieved through studying various constrained special cases of ARGs~\citep[e.g.][]{gusfield2004optimal} and
+other more general types of phylogenetic networks~\citep{huson2010phylogenetic}. The
+focus has been on algorithmic and combinatorial results, that largely do not
+have direct relevance to the inference problems described above.
+
+%It is important to note that parsimonious ARGs
+%have quite different properties from realisations of the
+%ARG \emph{process} as modelled by Hudson, Griffiths, and colleagues.
+%Firstly, branch lengths are not typically estimated,
+%% why do we need the following two lines
+%and node times are required in order to compute a likelihood under the
+%coalescent with recombination (see section XXX).
+%Secondly, even if branch lengths were estimated,
+%the realisations would have a very small likelihood since
+%the coalescent with recombination is often highly \emph{un}parsimonious
+%in terms of recombination events.
+
+The goal of this historical overview is to illustrate that the meaning of the term ``ARG" now strongly
+depends on the context in which it is used, and can mean both the
+stochastic process generating genealogies in the presence of
+recombination~\citep[e.g.][]{nordborg2000linkage,birkner2013ancestral,
+wilton2015smc,griffiths2016coalescent},
+as well as, more commonly, the concrete realisation of ancestry from a
+process~\citep[e.g.][]{gusfield2014recombinatorics,mathieson2020ancestry,brandt2021evaluation}.
+\cite{minichiello2006mapping} explicitly distinguished the ARG as a
+stochastic process from the ARG as a data structure, and
+argued for using the term to mean specific realised ancestral histories.
+Unless otherwise stated, we will use the ARG-as-a-data-structure
+interpretation for the remainder of this paper.
+
+% FIXME putting this here for now just to see if this will work.
+
+% \section*{Event ARGs}\label{eARG}
+% In keeping with its original derivation in terms of a stochastic
+% process (see Appendix~\ref{sec-arg-history}), the classical definition of an
+% ARG data structure is
+% in terms of events.
+% Nodes represent
+% common ancestor or recombination events that occurred in the
+% history of some set of sample genomes, and edges represent ancestral
+% lineages~\citep{griffiths1996ancestral}.
+% In a common ancestor event the inbound lineages are merged into a
+% single ancestral lineage, and in a recombination
+% event a lineage is split into two independent
+% ancestral lineages.
+% % The final vital
+% % detail is that we associate a breakpoint with each recombination
+% % event.
+% % This information is sufficient to uniquely
+% % define the local genealogical trees at every position along the genome,
+% % which is the basic requirement for a data structure encoding
+% % genome-wide genetic ancestry.
+% As illustrated in Fig.~\ref{fig-arg-data-structure}, this information
+% is usually concretely encoded by associating a recombination breakpoint
+% (which is null for common ancestor events) and an ordered list of parents
+% with each node. We refer to such a  data structure as an
+% ``event ARG'' (eARG).
+% To recover the local tree at genome coordinate $x$ (the most
+% fundamental operation required for an ARG data structure),
+% we traverse the graph rootwards from the leaves.
+% At a particular
+% node $u$, if it has one parent we are at a common ancestor
+% node and we follow that parent. If we are at a
+% recombination node, $u$ has two parents: if
+% $x$ is less than the breakpoint we follow
+% the edge to the first parent, and otherwise follow the edge
+% to the second parent.
+% The order in which parents are listed at a recombination node is
+% therefore significant, telling us
+% from which parent the segments of genome to the left and right of the breakpoint
+% were inherited.
+
+% This ordering requirement, while straightforward
+% to describe, has some practical drawbacks. For example,
+% % Surely the meaning is clear from the context here and we don't need
+% % qualify the meaning of ``ARG''
+% when using this representation and simulating an ARG backwards in time,
+% the first event older than a recombination may involve the lineage carrying the
+% ancestry to the right of the breakpoint rather
+% than the left, and so we cannot emit edges as they are generated.
+% Such issues can be worked around, of course, but depending on the ordering of
+% otherwise indistinguishable objects is generally problematic.
+
+% \begin{figure}
+% \centering
+% % \begin{tabular}{cc}
+% \begin{tikzpicture}[x=5mm, y=5mm, node distance=2mm and 20mm]
+% \tikzset{greynode/.style={circle,fill,inner sep=1},
+% nodelabel/.style={font=\footnotesize}}
+
+% \node (s0) [greynode] at (0, 0) {};
+% \node (s1) [greynode] at (3, 0) {};
+% \node (s2) [greynode] at (6, 0) {};
+% \node (s3) [greynode] at (3, 1) {};
+% \node (s4) [greynode] at (1, 2) {};
+% \node (s5) [greynode] at (5, 3) {};
+% \node (s6) [greynode] at (3, 4) {};
+
+% % \node [anchor=north west] at (0,6) {A};
+% \node [nodelabel,anchor=north west] at ($(s3) + (0,0)$) {$x$};
+% \foreach \u/\lab in {s0/$\textsf{a}$, s1/$\textsf{b}$, s2/$\textsf{c}$} \node[nodelabel,anchor=north] at (\u) {\lab};
+% \foreach \u/\lab in {s4/$\textsf{e}$} \node[nodelabel,anchor=south west] at (\u) {\lab};
+% \foreach \u/\lab in {s5/$\textsf{f}$} \node[nodelabel,anchor=south east] at (\u) {\lab};
+% \foreach \u/\lab in {s3/$\textsf{d}$, s6/$\textsf{g}$} \node[nodelabel,anchor=south] at (\u) {\lab};
+
+% %% Edges
+% \draw (s1) -- (s3);
+% \draw (s0) |- (s4);
+% \draw (s4) -- (2,2) |- (s3);
+% \draw (s4) |- (s6);
+% \draw (s3) -- (4,1) |- (s5);
+% \draw (s2) |- (s5);
+% \draw (s5) |- (s6);
+
+% % \node [anchor=north west] at (9,6) {B};
+% \node [nodelabel,anchor=north west] at ($(10,5)$) {
+% \begin{tabular}{c|c|l}
+% % \multicolumn{2}{c}{Breakpoints}\\
+% Node & Breakpoint & Parents\\
+% \hline
+% $\noderef{a}$ & $\varnothing$ & [\noderef{e}]\\
+% $\noderef{b}$ & $\varnothing$ & [\noderef{d}] \\
+% $\noderef{c}$ & $\varnothing$ & [\noderef{f}]\\
+% $\noderef{d}$ & $x$ & [\noderef{e}, \noderef{f}] \\
+% $\noderef{e}$ & $\varnothing$ & [\noderef{g}]\\
+% $\noderef{f}$ & $\varnothing$ & [\noderef{g}]\\
+% $\noderef{g}$ & $\varnothing$ & []\\
+% \end{tabular}};
+
+% % \node [anchor=north west] at (9,6) {C};
+% % \node [nodelabel,anchor=north west] at ($(10,5)$) {
+% % \begin{tabular}{ll}
+% % Edges & Intervals\\
+% % \hline
+% % $(\textsf{a}, \textsf{e})$ & $(0, L]$ \\
+% % $(\textsf{b}, \textsf{d})$ & $(0, L]$ \\
+% % $(\textsf{c}, \textsf{f})$ & $(0, L]$ \\
+% % $(\textsf{d}, \textsf{e})$ & $(0, x]$ \\
+% % $(\textsf{d}, \textsf{f})$ & $(x, L]$ \\
+% % $(\textsf{e}, \textsf{g})$ & $(0, L]$ \\
+% % $(\textsf{f}, \textsf{g})$ & $(0, L]$ \\
+% % \end{tabular}};
+
+% \end{tikzpicture}
+% \caption{\label{fig-arg-data-structure}
+% A graphical depiction of a classical event ARG and a concrete
+% encoding of the information. Node \noderef{d} is a recombination
+% event as it is associated with a non-null breakpoint, and has
+% two parents. Nodes \noderef{e}, \noderef{f} and \noderef{g} are
+% common ancestor events.
+% }
+% \end{figure}
+
+% % What do we mean by an eARG, formally?
+% To be concrete, we can formally define the
+% classical Griffiths eARG as a tuple $(e, \sigma)$, where $e$
+% is an ordered list of edges defining a directed acyclic graph and
+% $\sigma: \mathbb{N} \rightarrow \mathbb{N}$
+% is a function mapping nodes to recombination breakpoints,
+% such that $\sigma(u) = x$
+% if $u$ is a recombination event with breakpoint $x$ and
+% $\sigma(u) = \varnothing$ otherwise.
+% (This is equivalent to having different ``types''
+% of node, where breakpoints are associated with recombination
+% nodes only).
+% Each edge $e_j = (c, p)$ describes an edge between
+% child node $c$ and parent $p$, where $c, p \in \mathbb{N}$.
+% % For each recombination node $u$ we must
+% % have exactly two edges in $e$ such that $u$ is the child.
+% This encoding of an eARG is equivalent to the output
+% of programs such as
+% \texttt{ARGweaver}~\citep{rasmussen2014genome}
+% and \texttt{KwARG}~\citep{ignatieva2021kwarg},
+% and an example is given in Fig.~\ref{fig-arg-data-structure}A.
+% Note that the description above captures only the
+% graph topology: if we also wish to know the branch lengths we need
+% an additional function $t(u): \mathbb{N} \rightarrow \mathbb{R}$
+% which defines the time of each node.
 

### 2023-08-10 paper.tex
Pass at intro

@@
 
-\textbf{Keywords:} Ancestral Recombination Graphs
+\textbf{Keywords:} Ancestral recombination graphs
 
 \section*{Introduction}
-Inferring the genetic relationships between sampled genomes in the form of a
-phylogenetic tree is a necessary prerequisite for many analyses of
-species~\citep{rannala2003genetics}, non-recombining sections of
-DNA~\citep{cann1987mitochondrial,underhill2001annalsofhumangenetics}
-and viruses~\citep{grenfell2004science}.
-A phylogenetic tree captures information about the genealogy
-of the sample in an elegant and concise
-way, by postulating a set of ancestors common to the samples
-together with paths of inheritance between them. A rich literature exists
-on analysing the mathematical properties of these
-trees~\citep{steel2016phylogeny}, and numerous
-methods exist to infer them~\citep{felsenstein2004inferring}.
-However, the genetic genealogy of a recombining organism
-cannot be captured by a single tree: instead,
-the inheritance of each section of the genome may be described
-by a (slightly) different local tree~\citep{hudson1983properties}.
-Inferring complex
-genealogies of this nature
-presents profound technical difficulties, meaning genealogy-based approaches have
-not been widely used in population and statistical genetics.
-Until recently, analysis of
-population genetic data has therefore primarily focussed on summaries derived
-from the local trees~\citep{tajima1983evolutionary,tavare1984line}
-rather than using the full underlying genealogy itself. Such
-derived measures include allele frequencies and site frequency
-spectra~\citep{achaz2009frequency,ralph2020efficiently},
-patterns of linkage disequilibrium~\citep{mcvean2002genealogical}, and
-principal components~\citep{mcvean2009genealogical}.
-
-Recent breakthroughs in large-scale inference
-methods~\citep{rasmussen2014genome,kelleher2019inferring,speidel2019method,wohns2022unified}
-and data representation~\citep{kelleher2016efficient}
-have raised the realistic prospect of genealogy-based analysis becoming a standard part
+% Inferring the genetic relationships between sampled genomes in the form of a
+% phylogenetic tree is a necessary prerequisite for many analyses of
+% species~\citep{rannala2003genetics}, non-recombining sections of
+% DNA~\citep{cann1987mitochondrial,underhill2001annalsofhumangenetics}
+% and viruses~\citep{grenfell2004science}.
+% A phylogenetic tree captures information about the genealogy
+% of the sample in an elegant and concise
+% way, by postulating a set of ancestors common to the samples
+% together with paths of inheritance between them. A rich literature exists
+% on analysing the mathematical properties of these
+% trees~\citep{steel2016phylogeny}, and numerous
+% methods exist to infer them~\citep{felsenstein2004inferring}.
+% However, the genetic genealogy of a recombining organism
+% cannot be captured by a single tree: instead,
+% the inheritance of each section of the genome may be described
+% by a (slightly) different local tree~\citep{hudson1983properties}.
+% Inferring complex
+% genealogies of this nature
+% presents profound technical difficulties, meaning genealogy-based approaches have
+% not been widely used in population and statistical genetics.
+% Until recently, analysis of
+% population genetic data has therefore primarily focussed on summaries derived
+% from the local trees~\citep{tajima1983evolutionary,tavare1984line}
+% rather than using the full underlying genealogy itself. Such
+% derived measures include allele frequencies and site frequency
+% spectra~\citep{achaz2009frequency,ralph2020efficiently},
+% patterns of linkage disequilibrium~\citep{mcvean2002genealogical}, and
+% principal components~\citep{mcvean2009genealogical}.
+Inferring the detailed genetic genealogy of a set of sequences
+under the influence of recombination,
+usually known as an Ancestral Recombination Graph (ARG), is a long-standing
+goal in genetics.
+Recent breakthroughs
+in large-scale inference
+methods~\citep{rasmussen2014genome,kelleher2019inferring,speidel2019method,
+schaefer2021ancestral,wohns2022unified,zhang2023biobank,zhan2023towards}
+and data representation~\citep{kelleher2016efficient,kelleher2018efficient}
+have raised the realistic prospect of ARG-based analysis becoming a standard part
 of the population and statistical genetics toolkit~\citep{hejase2020summary}.
 Applications using these inferred ancestries as inputs have begun to
-appear~\citep{osmond2021estimating,zhang2023biobank,fan2022genealogical,hejase2022deep}
+appear~\citep{osmond2021estimating,fan2022genealogical,hejase2022deep,zhang2023biobank}
 and many more are sure to
-follow~\citep{harris2019database}. This vibrant research area, however,
-has a significant difficulty, which risks hampering
-progress: we currently lack a well-defined data-model and shared terminology
-to discuss genealogies in the presence of recombination. This leads to basic errors in statements
-about what different inference methods produce as output, as well as presenting serious
-difficulties in either comparing the outputs of different methods or
-(from a user's perspective) using multiple inference methods in an analysis.
-
-In this paper we [turn into proper round up paragraph later]
-\begin{enumerate}
-\item Discuss the classical Ancestral Recombination Graph formulation,
-and show how it derives from (and is limited by) the coalescent with
-recombination.
-\item In the following sections we
-show that, although the ARG-as-a-data-structure perspective is
-now the most common interpretation, the \emph{form} of that
-data structure is still closely linked to the original
-stochastic process, and that this form is inflexible
-in the types of genetic inheritance that can be described,
-computationally inefficient, and
-forces complete (and potentially spurious) precision about
-the location and timing
-of recombination events. % (and coalescent events?)
-\item Suggest a definition of ARGs that is free of these limitations,
-where the formulation is derived from a (normally diploid) pedigree rather than from
-coalescent approximations, and where the genetic
-material which is transmitted is described by annotations attached to
-edges rather than nodes.
-\item Suggest a classification of ARG nodes that helps us to understand
-the properties of the structures inferred by different methods, and illustrate
-with some examples.
-% Perhaps we should also mention something about simplification and unknowable
-% nodes, e.g. ``show how some of the nodes are not required for useful interpretation
-% of ancestry''.
-\item Despite its status as ``the ideal data structure for
-population genomic analysis''~\citep{rasmussen2014genome}, there has been
-little discussion of its computational properties. We discuss the
-relationship between the classical concept of an ARG and the
-recently introduced succinct tree sequence data structure.
-
-\item This, coupled with
-    the explicit association of genome coordinates with edges, allows
-    us to represent even the bewilderingly complex patterns of relatedness
-    present in modern datasets.
-
-\end{enumerate}
+follow~\citep{harris2019database,harris2023using}.
+
+% TODO need to mention ARG-as-a-data-structure to clarify.
+% What is the problem?
+Although it is widely accepted that ARGs are important, there is significant
+confusion about what, precisely, an ARG \emph{is}. [THIS IS BAD, WHY?]
+We also lack a shared format for interchanging ARG data [which is bad
+for users, progress]. [Because we don't really agree on what an ARG is,
+we don't really know what properties it should have, when we infer it].
+
+% What do we do?
+Here we precisely define an encoding for ARGs based on specific genomes
+and their intervals of genetic inheritance. We contrast this ``gARG''
+encoding with the classical approach derived from the coalescent
+with recombination stochastic process of Griffiths
+and colleagues~\citep{griffiths1991two,ethier1990two,
+griffiths1996ancestral,griffiths1997ancestral}. As this is based
+on recording evolutionary events, we refer to it as the event ARG, or
+``eARG'' encoding.
+We show that the gARG encoding is both flexible and efficient,
+and can represent varying levels of precision about historical
+events.
+[MORE]
+
+% and because it is focused on encoding the \emph{outcome}
+% of evolutionary events rather than the details of the events themselves,
+
+% In this paper we [turn into proper round up paragraph later]
+% \begin{enumerate}
+% \item Discuss the classical Ancestral Recombination Graph formulation,
+% and show how it derives from (and is limited by) the coalescent with
+% recombination.
+% \item In the following sections we
+% show that, although the ARG-as-a-data-structure perspective is
+% now the most common interpretation, the \emph{form} of that
+% data structure is still closely linked to the original
+% stochastic process, and that this form is inflexible
+% in the types of genetic inheritance that can be described,
+% computationally inefficient, and
+% forces complete (and potentially spurious) precision about
+% the location and timing
+% of recombination events. % (and coalescent events?)
+% \item Suggest a definition of ARGs that is free of these limitations,
+% where the formulation is derived from a (normally diploid) pedigree rather than from
+% coalescent approximations, and where the genetic
+% material which is transmitted is described by annotations attached to
+% edges rather than nodes.
+% \item Suggest a classification of ARG nodes that helps us to understand
+% the properties of the structures inferred by different methods, and illustrate
+% with some examples.
+
+% \item This, coupled with
+%     the explicit association of genome coordinates with edges, allows
+%     us to represent even the bewilderingly complex patterns of relatedness
+%     present in modern datasets.
+
+% \end{enumerate}
 

### 2023-08-24 paper.tex
Abstract pass

@@
 agreement on the concrete details.
-We propose a concrete definition of ARGs that is suitable for modern
-datasets, in terms of genomes and their patterns of inheritance,
-and contrast this with the classical event-based definition which
-makes several limiting assumptions.
-We show that this new definition enables fine-grained
-levels of precision about ancestral processes to be represented,
+We propose a concrete definition of ARGs
+in terms of genomes and their intervals of genetic inheritance (gARG)
+and contrast this with the classical event-based definition (eARG).
+The eARG definition is based on the coalescent with recombination
+stochastic process, and we show how the assumptions of that process
+place limitations on the patterns of inheritance that can be
+represented, and that these assumptions are
+% Better wording
+substantially broken
+in modern large-scale datasets.
+Representing genetic ancestry
+with an eARG also requires complete precision about the ordering
+of all recombination events, which we will rarely have sufficient
+information to infer.
+We show, in contrast, that the gARG definition
+can fully capture the richness of modern large-scale datasets,
+enables fine-grained
+levels of precision about recombination to be represented,
 and forms the basis of an efficient computational framework.

### 2023-08-24 paper.tex
Add Prospective vs Retrospective section

@@
 % principal components~\citep{mcvean2009genealogical}.
-Inferring the detailed genetic genealogy of a set of sequences
+Estimating the detailed genetic genealogy of a set of sequences
 under the influence of recombination,
@@
 Applications using these inferred ancestries as inputs have begun to
-appear~\citep{osmond2021estimating,fan2022genealogical,hejase2022deep,zhang2023biobank}
+appear~\citep{osmond2021estimating,fan2022genealogical,hejase2022deep,zhang2023biobank,
+% TODO update when natgen paper comes out
+nowbandegani2022extremely}
 and many more are sure to
@@
 of particular genomes and their patterns of inheritance, which
-to avoid confusion, we refer to as a ``genome ARG``, or gARG.
-(We contrast this with the classical definition derived from the
-coalescent process in later sections.)
+we refer to as a ``genome ARG``, or gARG.
 First, let us define a genome as the complete set of genetic material that a child
@@
 where each edge is annotated with the coordinates
-of the genomic intervals over which genetic inheritance occurs
-(the next section explores the subtleties of what we mean by
-genetic inheritance).
+of the genomic intervals over which genetic inheritance occurs.
+We can therefore view a gARG for a given set of individuals
+as a subset of their pedigree, but, crucially,
+where each individual is represented by two nodes (for diploids).
 % To recover the local tree for a genomic
@@
 % speed2015naturereviewsgenetics}.
-Fig.~\ref{fig-arg-in-pedigree}
+This is illustrated in Fig.~\ref{fig-arg-in-pedigree},
+which
 shows how the genomes embedded in an example (highly inbred)
-pedigree correspond to the nodes in a gARG, along with the corresponding
-local trees.
+pedigree correspond to the nodes in a gARG.
 In this diploid example
@@
 and may be the recombined product of that parent's two genomes.
-% NOTE: does this example really help? Rather than spelling out what
-% we can already see, perhaps we should point out the non-obvious
-% thing that we have decoupled from "events"
 For example, individual $I_1$ in Fig.~\ref{fig-arg-in-pedigree}
@@
 labelled $I_1$ to $I_8$, contain both paternal  and maternal  genomes
-labelled \textsf{a} to \textsf{p}. Black lines show transmission paths connecting
+labelled \textsf{a} to \textsf{p}. Black lines show inheritance paths connecting
 genomes in the current generation (\textsf{a} to \textsf{d}) with their ancestors.
@@
 % from f) so, a and c are not fully independent!
-and \textsf{f}, and regions of genome transmitted are shown.
+and \textsf{f}, and regions of genome inherited are shown.
 Genomes are shaded such that where, backwards in time,
 they merge into a common ancestor, the merged region is darker.
-(B) The corresponding genome ARG along with annotations on all edges;
-annotations showing partial
-genome transmission are highlighted in bold. Recombination (e.g.,
-the cause of the breakpoint at position 7)
-is associated with two parent genomes (\textsf{e} and
-\textsf{f}) and a child genome (the recombinant, \textsf{c}).
-(C) The edge-annotated genome ARG in (B) fully defines the
-local relationships along the genome, as shown in the local
-(``marginal'') trees.
+(B) The corresponding gARG along with inheritance annotations on all edges
+(partial inheritance in bold).
+% Recombination (e.g.,
+% the cause of the breakpoint at position 7)
+% is associated with two parent genomes (\textsf{e} and
+% \textsf{f}) and a child genome (the recombinant, \textsf{c}).
+% (C) The edge-annotated genome ARG in (B) fully defines the
+(C) The corresponding local (``marginal'') trees.
 }
 \end{figure}
+
+The edges in a gARG specify that genetic inheritance occurred between an
+ancestor and a descendant, but the connectivity information alone
+does not tell us which \emph{parts} of their genomes were inherited.
 
@@
 
-\section*{Inheritance and ancestral material}\label{Ancestry_resolution}
+% \section*{Inheritance and ancestral material}\label{Ancestry_resolution}
+\section*{Prospective vs restrospective ARGs}
+ARGs are usually thought of as a restrospective structure, in which we
+reason about the genomes that are genetic ancestors of some set of
+sample genomes. Figure~\ref{fig-arg-in-pedigree}
+is retrospective because it only includes information
+about genetic inheritance relative to the sample individuals
+$I_1$ and $I_2$ (genomes \textsf{a}--\textsf{d}).
+Genetic inheritance must have occured between (e.g.)
+$I_8$ and her daughter $I_6$, but this is not recorded because
+% this is probably not true, given we know the complete pedigree?
+we have no information about it, given these sampled genomes.
+Event ARGs, given their roots in the coalescent with recombination,
+are inherently retrospective. They define the evolutionary
+events that have occured in the genetic ancestry of a sample.
+% This is weak, rephrase.
+An event ARG must have some set of samples, or there
+will be nothing to define events in terms of.
+
+The gARG representation, on the other hand,
+can be either retrospective or \emph{prospective}.
 The edges in a gARG represent a path of genetic inheritance from
 ancestor to descendant through some
-number of generations (ultimately from cell to cell). The edges are ``annotated''
+number of generations (ultimately from cell to cell),
+and are annotated
 with the genome coordinates over which inheritance occurs.
-There are two different ways in which we use these inheritance
-annotations, and the distinction is subtle and important. The difference
-is illustrated in Fig~\ref{fig-ancestry-resolution}.
+In a prospective gARG, the inheritance intervals $I$
+on an edge joining ancestor genome $v$ and descendant $u$
+defines \emph{all} regions of genome that $u$ inherited from $v$.
+Thus, these  inheritance intervals are
+concerned only with the genetic relationship
+between the nodes in question, and are not defined with respect
+to any given set of samples. We are simply recording the
+direct genetic inheritance between two genomes
+(in the simplest case, between a parent and a child).
+
+Prospective ARGs occur in individual-based forwards-time
+simulations~\citep{kelleher2018efficient,haller2018tree},
+where the genetic inheritance between genomes is recorded
+exhaustively and periodically
+``simplified'' to retain only
+the ancestry relevant to the current population.
+\citet{kelleher2018efficient} refer to the structure
+recording all such forwards-in-time genetic relationships
+as an ``embellished pedigree`` (and suggest the term ``nedigree'').
+Describing this instead as a ``prospective ARG'', and emphasising
+the use of the same data structures and direct connection with
+the usual retrospective interpretation of an ARG, may be
+more useful.
+The ``simplify'' algorithm which converts a prospective
+ARG into a retrospective one is a powerful tool,
+with other applications as we discuss in the following sections.
+
+\section*{Ancestry resolution}\label{Ancestry_resolution}
+
+These genomic intervals, directly ancestral to the samples
+and  scattered across a set of lineages at some time in the past,
+are usually known as the
+``ancestral material''~\citep{wiuf1999ancestry,wiuf1999recombination}
+of the sample.
 
@@
 \caption{\label{fig-ancestry-resolution}
-TODO DELETE A? Or keep it? Maybe this would help some hard-core eARGers?
 The \citet[][Fig.~1]{wiuf1999recombination} example graph as an eARG and
@@
 
-The simplest interpretation of the inheritance intervals $I$ for
-an ancestor genome $v$ and descendant $u$, is that $I$ defines the
-regions of genome that $u$ inherited from $v$. For example,
-node $o$ is the recombinant descendant of nodes $p$ and $q$ in
-Figure~\ref{fig-ancestry-resolution}A, which shows that it in inherited
-everything to the left of position $1$ from $p$ and the rest of
-its genome from node $q$.
-Thus, these edge annotations are concerned only with the genetic relationship
-between the nodein question, and are not affected by the rest of the graph.
-We will thefore refer to inheritance annotations of this type as ``local inheritance
-annotations''. Graphs of this type result from forwards-time
-simulations~\citep{kelleher2018efficient,haller2018tree} or the
-``Big ARG'' stochastic process (Appendix XXX).
-
 ARGs are most often thought of as a \emph{retrospective} structure however,
@@
 ``sample resolved''.
+
 % Keeping track of resolved ancestral material is one of the distinguishing

### 2023-08-24 paper.tex

@@
 The ``simplify'' algorithm which converts a prospective
-ARG into a retrospective one is a powerful tool,
-with other applications as we discuss in the following sections.
-
-\section*{Ancestry resolution}\label{Ancestry_resolution}
-
-These genomic intervals, directly ancestral to the samples
-and  scattered across a set of lineages at some time in the past,
-are usually known as the
+gARG into a retrospective one is a powerful tool,
+with several other applications
+as we discuss in the following sections.
+
+\section*{Converting an eARG to a gARG}
+The process of converting an eARG to a gARG is very similar to
+simplifying a prospective gARG with respect to a set of samples,
+and highlights some important differences between the encodings.
+The process involves two steps, and is illustrated in
+Fig~\ref{fig-ancestry-resolution}.
+
+\begin{figure}
+\centering
+% TODO update the source of this figure to make less tall. It's taking
+% up far too much vspace at in native coords at the moment.
+\includegraphics[width=0.75\textwidth]{illustrations/ancestry-resolution}
+\caption{\label{fig-ancestry-resolution}
+Converting the \citet[][Fig.~1]{wiuf1999recombination} example graph
+to a gARG. (A) The original eARG; square nodes represent events, with
+each recombination (red) showing the corresponding breakpoint.
+(B) The corresponding gARG with breakpoints directly converted to
+edges annotated with inheritance intervals.
+(C) The sample-resolved gARG resulting from simplifying with respect
+to the samples, $a$, $b$, and $c$. Dashed lines show edges that are
+no longer present (in practice, nodes $g$ and $q$ would also be removed).
+Coalescence with respect to the sample is indicated by shaded bars, as
+in Fig~\ref{fig-arg-in-pedigree}A.
+Line thickness is proportional to the genomic span of each edge.
+}
+\end{figure}
+
+The first step is straightforward
+(Figure~\ref{fig-ancestry-resolution}B).
+We duplicate the nodes and
+edges from the input eARG, and the add inheritance annotations
+to the output gARG's edges. If the node is a common ancestor node,
+we annotate the single outbound edge with the interval $(0,L]$.
+If the node is a recombination node with a breakpoint $x$,
+we annotate the outbound edges with the intervals $(0, x]$
+and $(x, L]$.
+These inheritance annotations are clearly in one-to-one
+correspondance with the information in the input eARG,
+but they are a superset of the
 ``ancestral material''~\citep{wiuf1999ancestry,wiuf1999recombination}
@@
 
-% TODO majorly condense this caption, it's much too long.
-\begin{figure}
-\centering
-\includegraphics[width=\textwidth]{illustrations/ancestry-resolution}
-\caption{\label{fig-ancestry-resolution}
-The \citet[][Fig.~1]{wiuf1999recombination} example graph as an eARG and
-two gARGs with alternative edge annotation schemes.
-% (A) A conventional Griffiths eARG. Square nodes represent events, with
-% each recombination (red) annotated with the position of a single breakpoint
-% along the 7-unit-long chromosome.
-(A) A gARG with edges annotated by . Oval nodes represent
-genomes and annotations reflect ``local'' transmission of all genetic material
-between parent and child. Edges that do not span the full genome are visually
-highlighted, showing how breakpoint positions can be encoded.
-(C) A gARG with sample-resolved edge annotations. The Hudson-like simplification
-algorithm has been used to shorten edge spans to
-reflect only the upwards transmission of genetic material from the samples
-\textsf{a}, \textsf{b}, and \textsf{c}; very few edges then span the entire genome.
-The depiction of the genome within each
-node is shaded as in Fig.~\ref{fig-arg-in-pedigree}A to show coalescing
-regions of genome back to the MRCA nodes \textsf{k} (regions 0 to 2) and \textsf{p}
-(regions 2 to 7). Line widths reflect the genomic span of each edge, highlighting
-the relative importance of different transmission pathways to the samples. Nodes
-\textsf{g},  \textsf{j}, \textsf{o} and \textsf{q} are redundant in that they correspond
-to events which have no impact on the genetic ancestry of the samples.
-}
-\end{figure}
-
-ARGs are most often thought of as a \emph{retrospective} structure however,
-defined with respect to some set of (usually, contemporary) sample genomes.
-In this case we are not explicitly interested in the
-entire interval of genome descendant $u$ inherited from ancestor $v$,
-but rather the subset of that interval that is \emph{ancestral to our
-sample}.
-Fig~\ref{fig-ancestry-resolution}B shows the graph that we obtain when
-reduce the annotation intervals to those that are ancestral to the
-sample (note that node $q$ is now ommitted because the MRCA of the sample
-has already been found for the full genome at node $p$).
-Tracking such
-``ancestral material''~\citep{wiuf1999ancestry,wiuf1999recombination}
-is much more useful~\citep{mcgill2013graphml,arenas2013importance}
-and, indeed, all that is possible to
-have information about when inferring ARGs from contemporary samples.
-We refer to a gARG with inheritance annotations of this type as
-``sample resolved''.
-
-% Keeping track of resolved ancestral material is one of the distinguishing
-% features of ARGs vs phylogenetic
-% networks~\citep{mcgill2013graphml,arenas2013importance}, but
-% is actually incompatible with the Griffiths gARG encoding (it does this
-% implicitly, I guess)
+Ancestral material is a critical concept in a retrospective ARG,
+and refers to the genomic intervals directly ancestral to the samples
+scattered across lineages (the edges of the graph)
+at some time in the past. At recombination
+events, ancestral material is split between the ancestral lineages.
+As genome segments coalesce in a common
+ancestor the total amount ancestral material is reduced, until
+eventually as we traverse pastwards through the graph, all
+samples reach a most recent common ancestor at all locations
+along the genome. The process of transmitting ancestral material
+along the edges of an ARG is directly analogous to Hudson's
+simulation algorithm for the coalescent with
+recombination~\citep{hudson1983testing,kelleher2016efficient}.
+
+Fig~\ref{fig-ancestry-resolution}C shows the graph that we obtain
+by resolving the ancestral material transmitted along each edge
+(using the ``simplify'' algorithm), and highlights some important
+differences between the eARG and gARG encodings.
+Firstly, we can see that some nodes and edges are removed entirely
+from the graph. The ``grand MRCA'' $q$ is ommitted from the
+fully resolved gARG because all segments of the genome have
+fully coalesced before it is reached. Likewise, the edge
+between $g$ and $j$ is omitted because the recombination
+event at position $5$ fell in non-ancestral material.
+(This example of an eARG
+from \citet{wiuf1999recombination} is something of a straw-man,
+in that it is highly unlikely that an inference or simulation
+program would produce an ARG that contains nodes or edges
+that are not ancestral to the sample.)
+More generally, the gARG encoding allows for ``local'' inspection
+(with respect to a set of samples)
+of an ARG in a way that is not possible by storing
+common ancestor and recombination events in an eARG. Because
+the ancestral material is stored with each edge, the effects
+of multiple events can be reasoned about at once. Many algorithms
+that we wish to perform on an ARG will require resolving
+the genetic material with respect to a sample.
+% this is weak, but would be good to emphasise this point.
+% you just end up running simplify lots of times, in your
+% downstream algs
+The gARG encoding
+allows us to do this once and to store the result, whereas
+the eARG encoding requires us to repeat the process.
+
+%(see Appendix XXX) and has several features that we are unlikely to simulate or
+%infer in practice. For example, by definition the samples contain no genetic information about
+%nodes \textsf{g} and nodes \textsf{j} and so neither of those nodes are likely to be inferred.
+%Likewise, is is unlikely we would have information to infer the ``grand MRCA''
+%\textsf{q} as all regions of the genome have already reached a
+%common ancestor separately (in nodes \textsf{k} and \textsf{p}).
+
+% Third, there exist nodes such as
+% \textsf{k} with two children (i.e. ``common ancestor'' nodes in Hudson's terminology)
+% but in which no ancestral coalescence occurs. Fourth, some breakpoint positions occur
+% within regions of genome that are not ancestral (e.g. the breakpoint at position 3 in
+% node \textsf{k}). In this case sample-resolved edge annotations lose the exact position of
+% the breakpoint, although they retain the information that a breakpoint occurred between
+% positions 2 and 4 (which is sufficient for calculating the likelihood, see appendix XXX).
+
+% reduce the annotation intervals to those that are ancestral to the
+% sample (note that node $q$ is now ommitted because the MRCA of the sample
+% has already been found for the full genome at node $p$).
+% Tracking such
+% ``ancestral material''~\citep{wiuf1999ancestry,wiuf1999recombination}
+% is much more useful~\citep{mcgill2013graphml,arenas2013importance}
+% and, indeed, all that is possible to
+% have information about when inferring ARGs from contemporary samples.
+% We refer to a gARG with inheritance annotations of this type as
+% ``sample resolved''.
 
@@
 
-Given a gARG with direct-inheritance edge annotations
-(Fig~\ref{fig-ancestry-resolution}B),
- we obtain the sample-resolved edge annotation
-(Fig~\ref{fig-ancestry-resolution}C) by traversing the graph backwards in time
-from sample nodes (\textsf{a}, \textsf{b}, \textsf{c}).
-Proceeding in topological order (such that an ancestor node is visited
-after all of its descendants) the sample-resolved material for a node is the
-union of the sample-resolved intervals on its inbound edges. We then annotate
-the outbound edges with the intersection of their direct-inheritance intervals
-and the just-computed sample-resolved intervals for the node.
-For example, node \textsf{i} in Fig~\ref{fig-ancestry-resolution}C
-carries sample-resolved material on the intervals $(0, 2]$ and $(4, 7]$. The edge
-joining \textsf{i} to its ancestor \textsf{k} then carries
-material on $(0, 2]$ since although \textsf{i} directly inherits interval
-$(0, 3]$ from \textsf{k} (Fig~\ref{fig-ancestry-resolution}B), only $(0, 2]$ is
-ancestral to the sample. For each node, we also keep track of the number of
-descendant samples over each genomic interval, and stop tracking that
-interval once the lineages from all samples have coalesced. For example,
-the edge from \textsf{k} to \textsf{n} omits the region $(0,2]$ because it has
-already fully coalesced (indicated by a fully black genomic region).
+% Given a gARG with direct-inheritance edge annotations
+% (Fig~\ref{fig-ancestry-resolution}B),
+%  we obtain the sample-resolved edge annotation
+% (Fig~\ref{fig-ancestry-resolution}C) by traversing the graph backwards in time
+% from sample nodes (\textsf{a}, \textsf{b}, \textsf{c}).
+% Proceeding in topological order (such that an ancestor node is visited
+% after all of its descendants) the sample-resolved material for a node is the
+% union of the sample-resolved intervals on its inbound edges. We then annotate
+% the outbound edges with the intersection of their direct-inheritance intervals
+% and the just-computed sample-resolved intervals for the node.
+% For example, node \textsf{i} in Fig~\ref{fig-ancestry-resolution}C
+% carries sample-resolved material on the intervals $(0, 2]$ and $(4, 7]$. The edge
+% joining \textsf{i} to its ancestor \textsf{k} then carries
+% material on $(0, 2]$ since although \textsf{i} directly inherits interval
+% $(0, 3]$ from \textsf{k} (Fig~\ref{fig-ancestry-resolution}B), only $(0, 2]$ is
+% ancestral to the sample. For each node, we also keep track of the number of
+% descendant samples over each genomic interval, and stop tracking that
+% interval once the lineages from all samples have coalesced. For example,
+% the edge from \textsf{k} to \textsf{n} omits the region $(0,2]$ because it has
+% already fully coalesced (indicated by a fully black genomic region).
 
@@
 
-The process of converting a gARG with local-inheritance annotations
-to its sample-resolved equivalent is also known as ``simplification'',
-and is the key algorithm underlying recent
-advances in forwards-in-time
-simulation~\citep{kelleher2018efficient,haller2018tree}.
-It has many useful properties, as we discuss in the next section.
+% The process of converting a gARG with local-inheritance annotations
+% to its sample-resolved equivalent is also known as ``simplification'',
+% and is the key algorithm underlying recent
+% advances in forwards-in-time
+% simulation~\citep{kelleher2018efficient,haller2018tree}.
+% It has many useful properties, as we discuss in the next section.
 
@@
 % ``simplification''.
-
-\begin{figure}
-\centering
-\vspace{5em}
-\includegraphics[width=\linewidth]{illustrations/simplification}
-\caption{\label{fig-simplification}
-ARG simplification. In all cases the graph visualization is shown on the left
-and the equivalent local trees on the right. Since multiple parents are indicative of
-recombination(s), nodes are coloured by the number of parents. and shaded
-according to the proportion of their span over which they are coalescent.
-(A) A gARG simulated from a diploid Wright-Fisher
-model. Note that the genome \textsf{k} has three children, and the genomes
-\textsf{e} and \textsf{g} simultaneously have multiple parents and multiple
-children.
-(B) Simplified to remove all
-% trying out this terminology - can define in the text
-1-connected graph components (e.g., diamonds such as \textsf{jlnm}).
-(C) Remove nodes that are unary in all local trees. This removes both recombinant nodes
-that have only one child (e.g. \textsf{n}) and nodes that have more
-than one child (i.e. are ``common ancestors'') but never represent coalescences
-in the local trees (e.g. \textsf{r}).
-(D) Rewrite edges to bypass any unary nodes in local trees. Note that this loses lineage
-information such as the fact that \textsf{k} is connected to \textsf{a} via \textsf{e}.
-}
-\end{figure}
 
@@
 these less informative nodes from the local trees by modifying the gARG.
+
+\begin{figure}
+\centering
+\vspace{5em}
+\includegraphics[width=\linewidth]{illustrations/simplification}
+\caption{\label{fig-simplification}
+ARG simplification. In all cases the graph visualization is shown on the left
+and the equivalent local trees on the right. Since multiple parents are indicative of
+recombination(s), nodes are coloured by the number of parents. and shaded
+according to the proportion of their span over which they are coalescent.
+(A) A gARG simulated from a diploid Wright-Fisher
+model. Note that the genome \textsf{k} has three children, and the genomes
+\textsf{e} and \textsf{g} simultaneously have multiple parents and multiple
+children.
+(B) Simplified to remove all
+% trying out this terminology - can define in the text
+1-connected graph components (e.g., diamonds such as \textsf{jlnm}).
+(C) Remove nodes that are unary in all local trees. This removes both recombinant nodes
+that have only one child (e.g. \textsf{n}) and nodes that have more
+than one child (i.e. are ``common ancestors'') but never represent coalescences
+in the local trees (e.g. \textsf{r}).
+(D) Rewrite edges to bypass any unary nodes in local trees. Note that this loses lineage
+information such as the fact that \textsf{k} is connected to \textsf{a} via \textsf{e}.
+}
+\end{figure}
 

### 2023-08-25 paper.tex
eARG conversion tweaks

@@
 % this is probably not true, given we know the complete pedigree?
-we have no information about it, given these sampled genomes.
+we have no information about it, given these samples.
 Event ARGs, given their roots in the coalescent with recombination,
@@
 recording all such forwards-in-time genetic relationships
-as an ``embellished pedigree`` (and suggest the term ``nedigree'').
-Describing this instead as a ``prospective ARG'', and emphasising
-the use of the same data structures and direct connection with
+as an ``embellished pedigree`` (and suggest the term ``nedigree''
+as a shorthard).
+Describing this instead as a ``prospective ARG'', emphasising
+the use of the same data structures and the direct connection with
 the usual retrospective interpretation of an ARG, may be
@@
 event at position $5$ fell in non-ancestral material.
-(This example of an eARG
-from \citet{wiuf1999recombination} is something of a straw-man,
-in that it is highly unlikely that an inference or simulation
-program would produce an ARG that contains nodes or edges
-that are not ancestral to the sample.)
-More generally, the gARG encoding allows for ``local'' inspection
-(with respect to a set of samples)
+More generally, we can see that the fully resolved
+gARG encoding of Fig~\ref{fig-ancestry-resolution}C
+allows for ``local'' inspection
 of an ARG in a way that is not possible by storing
@@
 allows us to do this once and to store the result, whereas
-the eARG encoding requires us to repeat the process.
+the eARG encoding requires us to repeat the process on
+each traversal.
+
+It is important to note here that the \citet{wiuf1999recombination} eARG
+example used here is something of a straw man, in that it is highly unlikely
+that an inference or simulation program would produce a retrospective
+ARG that contains
+nodes or edges that are not ancestral to the sample. Nonetheless, it
+is an instructive example from the literature which highlights several
+important properties of ARGs, and the general point about
+the need to resolve ancestral material ``on the fly'' for eARG traversals
+holds.
 

### 2023-08-25 paper.tex
Add args and trees section

@@
 Line thickness is proportional to the genomic span of each edge.
+Nodes representing recombination events have been retained
+for clarity, but could be removed by simplification if
+desired.
 }
@@
 
-\section*{Simplification}\label{ARG_simplification}
-% I think a sentence is needed here on why "so much" text is devoted to simplification.
-% How does simplification connect to eARG vs gARG and practical benefits of gARG over eARG
-% and why do you now talk about unary nodes and a bit polytomies? Hold my hand here please;)
-ARG simplification~\citep{kelleher2018efficient} is the process
+\section*{Levels of simplification}\label{ARG_simplification}
+ARG simplification is a powerful tool, and can do much more
+besides removing unreachable nodes and edges in a prospective ARG
+and resolving ancestral material in an eARG.
+In general, we can think of
+ARG simplification as the process
 of removing nodes and re-writing edges in a gARG to remove
-various types of redundancy,
-retaining the ``meaningful'' topology within local trees
-and maintaining the identity of nodes across different trees.
-The idea of meaningful local tree topologies is based on the
-presence of ``unary'' nodes: a tree node that has exactly
-% should presence above be absence?
-one child. Fig.~\ref{fig-simplification} show a series of successive steps of
+various types of redundancy.
+Many of the types of redundancy revolve around the presence
+of ``unary'' nodes in the local trees
+(a tree node that has exactly one child).
+
+Fig.~\ref{fig-simplification} show a series of successive steps of
 simplification, starting with a complete gARG simulated under a
-(backwards-in-time) Wright-Fisher
-model, shown both as a graph on the left, and the equivalent set of 6
+backwards-in-time Wright-Fisher
+model, shown both as a graph on the left, and the corresponding sequence of 6
 local trees on the right.
@@
 joining sample \textsf{a} to node \textsf{q} to be equivalent to the path
-containing the intermediate unary nodes. Simplification is the process of removing
-these less informative nodes from the local trees by modifying the gARG.
+containing the intermediate unary nodes.
 
@@
 \caption{\label{fig-simplification}
-ARG simplification. In all cases the graph visualization is shown on the left
-and the equivalent local trees on the right. Since multiple parents are indicative of
-recombination(s), nodes are coloured by the number of parents. and shaded
+Levels of ARG simplification.
+In all cases the graph visualization is shown on the left
+and the corresponding local trees on the right.
+Nodes are coloured by the number of parents and shaded
 according to the proportion of their span over which they are coalescent.
@@
 
-The first type of simplification that we can perform is to remove
+The first level of simplification that we can perform is to remove
 graph topology that is invisible to the samples.
@@
 Unless we are specifically
-interested in the recombination event or ancestral genomes,
+interested in the recombination event or these ancestral genomes,
 there is no information in this topology and the diamond can be
@@
 in Fig.~\ref{fig-simplification}B), or can represent a common ancestor with
-multiple children in the graph but in which no local coalescence takes place
+multiple children in the graph but in which no coalescence takes place
+in the local trees
 (node \textsf{r} in Fig.~\ref{fig-simplification}B).
@@
 therefore requires knowledge not just of the graph topology but also of the
-edge annotations.
+ancestral material associated with the edges.
 As we see in Fig.~\ref{fig-simplification}C,
@@
 
-% Needs another pass
 An important consequence of simplifying local trees to remove
-unary nodes is that we may lose the ability to identify
-the ``subtree prune-and-regraft'' (SPR) operations
-by which they differ~\citep{hein1990reconstructing,song2003on,song2006properties}.
-% The effects of recombination as we move from left-to-right along
-% the genome in the complete simulated gARG
-% of Fig~\ref{fig-simplification}A are easy to interpret,
-% but as more redundancy is removed from the graph
-% the
-As explained above, we can have multiple recombination and
-common ancestor events stacking up, and therefore
-many possible SPR operations. Deducing the SPRs between
-trees is NP-hard in
-general~\citep{hein1996complexity,allen2001subtree,bordewich2005computational},
-but effective heuristics have been proposed~\citep{rasmussen2022espalier}.
+unary nodes is that we lose information about recombination
+events. This is related to the amount of \emph{precision} about
+recombination events that we store, which is the topic of the next
+section.
+
+\section*{Equivalance of ARGs and the set of local trees}
+An ARG is sometimes referred to as being equivalent to a set of
+local trees along the genome [TODO citation], but the connection
+between the two is not as simple as it may seem.
+In order for the mapping to be one-to-one, so that the original
+gARG can be exactly recovered from a set of local trees and
+their genome coordinates,
+some specific properties of the ARG and the trees are required.
+
+Firstly, the internal nodes in the local trees must be labelled
+(i.e.\ marked with the identity of the corresponding ARG node).
+If the nodes are not labelled, we must infer their identity
+from the tree topology, which is both
+expensive to compute and imprecise. For example
+we could infer that node \textsf{k} is shared in the second
+and third trees of in Fig.~\ref{fig-simplification}D
+because it subtends identical subtrees in both,
+but it would not be straightforward to infer that it
+is also the root of the first tree.
+% JK: Does this need to be said?
+% (One may use the time of a node as a proxy for identity,
+% but it is not reliable as we can have multiple nodes at the
+% same time, e.g. with time discretisation.)
+
+The precision of the information about recombination that
+can be represented in the trees depends the ARG itself.
+Secondly, the local trees must contain ``unary'' nodes.
+[Some discussion].
+With these two conditions met, there is a direct correspondance
+between a gARG and the trees.
+
+These properties have interesting consequences when we
+consider the inference strategy of esimating local trees
+independenty and subsequently stitching them together
+into an ARG. Two recent methods take this approach,
+\relate\ and
+\texttt{espalier}~\citep{rasmussen2022espalier},
+and have different strategies for reconciling the local
+trees into an ARG. [Relate does something fairly lightweight,
+but by default imprecise.]
+Espalier, on the other hand, is specifically in interested
+in the recombination events and therefore attempts to infer
+the precise ``subtree prune-and-regraft'' (SPR)
+operations~\citep{hein1990reconstructing,song2003on,song2006properties}
+using heuristics~\citep{rasmussen2022espalier}.
+Inferring the SPRs between
+leaf-labelled trees is NP-hard in
+general~\citep{hein1996complexity,allen2001subtree,bordewich2005computational}.
 

### 2023-08-25 paper.tex
Partial updates on precision of recomb info sectio

@@
 
-An important consequence of simplifying local trees to remove
-unary nodes is that we lose information about recombination
+An important consequence of simplifying ARGs to remove
+unary nodes in local trees is that we lose some information
+about recombination
 events. This is related to the amount of \emph{precision} about
@@
 
-\section*{Equivalance of ARGs and the set of local trees}
-An ARG is sometimes referred to as being equivalent to a set of
-local trees along the genome [TODO citation], but the connection
-between the two is not as simple as it may seem.
-In order for the mapping to be one-to-one, so that the original
-gARG can be exactly recovered from a set of local trees and
-their genome coordinates,
-some specific properties of the ARG and the trees are required.
-
-Firstly, the internal nodes in the local trees must be labelled
-(i.e.\ marked with the identity of the corresponding ARG node).
-If the nodes are not labelled, we must infer their identity
-from the tree topology, which is both
-expensive to compute and imprecise. For example
-we could infer that node \textsf{k} is shared in the second
-and third trees of in Fig.~\ref{fig-simplification}D
-because it subtends identical subtrees in both,
-but it would not be straightforward to infer that it
-is also the root of the first tree.
-% JK: Does this need to be said?
-% (One may use the time of a node as a proxy for identity,
-% but it is not reliable as we can have multiple nodes at the
-% same time, e.g. with time discretisation.)
-
-The precision of the information about recombination that
-can be represented in the trees depends the ARG itself.
-Secondly, the local trees must contain ``unary'' nodes.
-[Some discussion].
-With these two conditions met, there is a direct correspondance
-between a gARG and the trees.
-
-These properties have interesting consequences when we
-consider the inference strategy of esimating local trees
-independenty and subsequently stitching them together
-into an ARG. Two recent methods take this approach,
-\relate\ and
-\texttt{espalier}~\citep{rasmussen2022espalier},
-and have different strategies for reconciling the local
-trees into an ARG. [Relate does something fairly lightweight,
-but by default imprecise.]
-Espalier, on the other hand, is specifically in interested
-in the recombination events and therefore attempts to infer
-the precise ``subtree prune-and-regraft'' (SPR)
-operations~\citep{hein1990reconstructing,song2003on,song2006properties}
-using heuristics~\citep{rasmussen2022espalier}.
-Inferring the SPRs between
-leaf-labelled trees is NP-hard in
-general~\citep{hein1996complexity,allen2001subtree,bordewich2005computational}.
-
-% JK TODO: finish up this section with a quick version of the
-% para here. Basically say, people often say an ARG is a
-% set of local trees cite cite, but it's not as simple as that as we
-% can see from the example (probably).
-
-% The equivalence between an ARG and a set of local trees
-% separated by SPRs is often mentioned [e.g. x y z], but
-% it is important to note that while extracting local
-% trees from an ARG is straightforward, the converse
-% process of reconstructing precisely the same ARG from
-% these trees is more problematic. If the internal
-% nodes are consistently labelled and unary nodes are
-% included in the local trees (e.g., Fig~\ref{fig-simplification})
-% then it is easy to exactly reconstruct the ARG.
-% Otherwise, the problem is much more difficult and it
-% may not be possible to uniquely reconstruct the original ARG.
-% Without labels for the internal nodes, we are forced to
-% to reason about which subtrees remain the same and which
-% differ due to the SPR separating adjacent trees in order to
-% assign node identity. Without the internal unary nodes marking
-% recombinations, we must infer their existence. In
-% general, there will not be a unique solution to this problem,
-% and so there can be many possible ARGs compatible with
-% a given list of leaf-labelled binary trees.
-% This is the approach taken by the Espalier inference
-% method~\citep{rasmussen2022espalier}, which finds a
-% set of SPRs that reconcile independently inferred
-% trees from different regions of the genome.
-
-% NEEDS more work!
-\section*{Precision of ancestry information}
-Ancestral recombination graphs are often characterised
-as consisting of a complete record of the genetic history
-of a sample. For example,
-\cite{rasmussen2014genome} state that an ``ARG provides a record of all
-coalescence and recombination events since the divergence of the sequences
-under study'' and
-according to \cite{deng2021distribution} an ARG
-``provides all the information about the genealogical history of a sample,
-including the locations of recombination events.'' It is worth
-questioning, however, whether such comprehensive information
-will always be available, and in particular whether our
-representation of genealogical history should \emph{require}
-such precision. Simulations will usually generate complete
-information about the history of a sample, of course, but
-we may not have sufficient information to infer such detail.
-
-% The full ancestral recombination graph (ARG) is a structure that encodes all
+\section*{Precision of recombination information}
+% The ARGs in Fig.~\ref{fig-simplification} represent different
+% levels of detail about the same ancestral history. They represent
+% the same set of recombination and common ancestor events,
+As illustated in Fig.~\ref{fig-simplification}, successive levels
+of ARG simplification reduce the amount of information about the
+history of the sample that is stored. Some of the information lost,
+for example, ``diamond'' removal (Fig.~\ref{fig-simplification}A),
+seems like a reasonable tradeoff for a simpler structure.
+The consequences of other simplifications, however, are
+more subtle and relate directly to what can be known about
+recombination events and the levels of precision that
+we should seek to infer about them.
+
+The ARGs in Fig.~\ref{fig-simplification} contain different
+numbers of local trees ($6$, $5$, $5$ and $4$ for A through
+D, respectively). When we move from A to B the local trees
+for the intervals $(44,61]$ and $(61,87]$ are merged because
+the only differences between them are their journeys through
+nodes \noderef{l} and \noderef{m}. These nodes that participated
+in the diamond are removed from the ARG, and we have lost
+all information about the corresponding recombination at
+position 61. Other nodes (e.g.\ \noderef{o} and \noderef{p})
+have also been removed but these represent the \emph{parents}
+of recombinants. The recombinant nodes themselves
+(e.g. \noderef{n}) are still present, and represent precise
+information about the time, genomic location and tree
+branches involved
+in the recombination event.
+
+Fig.~\ref{fig-simplification}C has the same number of local trees
+as Fig.~\ref{fig-simplification}B, but has less precise information
+about recombination. Continuing the previous example, node
+\noderef{n} has been removed from the graph because it was unary
+in all of the local trees; its outbound edges to \noderef{s}
+and \noderef{q} have effectively been ``pushed down''
+to \noderef{e} (which is retained because it is the coalescent
+parent of \noderef{a} and \noderef{b} over the interval
+$(44, 100]$). We
+have therefore lost precision about
+the \emph{timing} of this recombination event, and know only
+that it must have occured between the times of node \noderef{e}
+and \noderef{q}.
+
+Fig.~\ref{fig-simplification}D removes all unary nodes from the
+local trees, and further reduces the precision of
+recombination information. Node \noderef{e} has not been
+removed from the graph because it is coalescent in the
+final tree, but we no longer know that the recombination
+event at position 30 was ancestral to it, or have
+any indication of its timings. Furthermore,
+trees for $(44, 87]$ and $(87, 100]$ were only distinguishable
+by the passage of the former tree through nodes \noderef{e}
+and \noderef{q}, and so the recombination on node \noderef{g}
+at position 87 has been lost entirely.
+
+% The example in Fig.~\ref{fig-simplification} is usually
+% referred to as ``fully simplified'', and is the default output
+% of simulation programs like \msprime. While there is significant
+% precision about recombination events lost
+
+The point of these examples is to emphasise that
+information about recombination in an ARG is not ``all
+or nothing'': there is a continuuum of detail
+about recombination, in terms of genomic location, timing
+the ancestral lineages involved, that we can describe.
+While there has been a recent tendancy to emphasise the
+importance of ``full ARGs''
+\citep[e.g.][]{deng2021distribution,brandt2021evaluation,
+rasmussen2022espalier}
+which contain complete information about all recombination
+events, it is reasonable to question the value of
+such precise information and whether it is knowable.
+[SOME MORE STUFF]
+
+% Ancestral recombination graphs are often characterised
+% as consisting of a complete record of the genetic history
+% of a sample. For example,
+% \cite{rasmussen2014genome} state that an ``ARG provides a record of all
+% coalescence and recombination events since the divergence of the sequences
+% under study'' and
+% according to \cite{deng2021distribution} an ARG
+% ``provides all the information about the genealogical history of a sample,
+% including the locations of recombination events.'' It is worth
+% questioning, however, whether such comprehensive information
+% will always be available, and in particular whether our
+% representation of genealogical history should \emph{require}
+% such precision. Simulations will usually generate complete
+% information about the history of a sample, of course, but
+% we may not have sufficient information to infer such detail.
+
+% % The full ancestral recombination graph (ARG) is a structure that encodes all
+% % coalescence and recombination events resulting from the stochastic process of
+% % the coalescent with recombination.
+% \citet{brandt2021evaluation} define a ``full'' ARG as
+% ``a structure that encodes all
 % coalescence and recombination events resulting from the stochastic process of
-% the coalescent with recombination.
-\citet{brandt2021evaluation} define a ``full'' ARG as
-``a structure that encodes all
-coalescence and recombination events resulting from the stochastic process of
-the coalescent with recombination'' (thus going against the general
-historial trend discussed in the XXX section).
-\citet{rasmussen2022espalier} have a less restrictive definition, in
-which a ``full'' ARG is one that contains complete information about
-recombination events, without necessarily being tied to a particular
-stochastic process.
-While [something positive], it is important to be clear on
-what the overall goal of such precise inferences are,
-what the limitations on when they can be made are,
-and their downstream utility in applications.
-
-% What is the goal of having fully precise ARGs?
-% 1) studying the actual recombination events;
-% 2) evaluating the likelihoods
-The main reasons for wanting precise estimates of all recombination
-events are to either compute a likelihood of the inferred ARG
-under a statistical model, or to study the recombinants in
-detail~\citep{rasmussen2022espalier}.
-% Explain that we're really interested in the details of recombination
-% itself with a small number of samples.
-In the latter case, we might
-have a small number of samples~\citep{guo2022recombination}.
-
-Beyond cases in which visual inspection of the trees and the
-effects of recombination is feasible, the main application
-of having completely precise estimates of recombination events is
-to evaluate the likelihood of the estimate under some
-statistical model. The only statistical model available
-is the coalescent with recombination
-and its approximations~\citep{mcvean2005approximating,marjoram2006fast}.
-In this case, the utility of a ``full`` ARG for some dataset
-is limited by how well the dataset fits the assumptions
-of the model.
-
-% Not sure where this goes, but there are some critical points
-% that we want to explain and not just throw out there in the
-% discussion.
-% A critical distinction between the eARG and gARG encodings
-% is the level of precision about the time and location of
-% recombination events that is required. In an eARG, we are
-% required to be absolutely precise about every recombination
-% [explain and contrast with gARG, referring to examples]
-
-% JK: This is very rough, just jotting down the basic ideas,
-% will need substantial revision.
-It is not clear that such complete information will always
-(or indeed ever) be available from inferences~\cite{shipilina2022origin}.
- We may be
-able to precisely identify the location and timing
-of recent recombinations (particularly if detailed pedigree
-information is available) but beyond this, the information
-required simply isn't present. When sampling from a
-process such as the Sequentially Markov
-Coalescent~\citep{mcvean2005approximating,marjoram2006fast}
-like ARGweaver~\citep{rasmussen2014genome}, it is true that
-we do obtain a fully resolved ARG with precise
-information about recombination, but it is reasonable
-to question how much information from the data is used
-to support older recombinations, in particular.
-
-% This idea of a continuum of information about recombination
-% being inferrable and representable is consistent with the ideas
-% discussed by~\cite{shipilina2022origin}. There a haplotype block
-% is synonymous with coalescent nodes in the ARG, which we can generalise
-% as nodes in a gARG with at least two children. In principle, these
-% nodes are identifiable by the mutations that occur on the outbound
-% edges of the node.
-
-It is helpful to consider the difference between fully
-resolved binary trees and those that contain polytomies.
-
-The situation is in fact directly analogous: an ARG that
-contains fully resolved recombination information must have
-at most parents per node.
-
-At the very least, it is not reasonable for the \emph{data
-structure} to require complete precision in order to
-represent some less precise. This is a fundamental
-advantage of the gARG encoding over the classical eARG
-method: we are not forced to posit a precise cause (recombination
-event) for every effect (difference between adjacent trees).
-These differences in precision of recombination
-information are illustrated in the ARGs estimated by different
-methods, as we see in the next section.
-
-In general, discussions tend to contrast an ARG
-on one hand as holding \emph{complete} information about recombination
-(and therefore between-tree correlation structure)
-and a sequence of local trees on the other (with, by implication,
-no direct information about between-tree correlations).
-This is a false dichotomy:
-a \emph{continuum} of information about recombination and between-tree
-correlation structure can be represented and inferred, and all
-such levels of information are useful, even if only for computational
-efficiency.
+% the coalescent with recombination'' (thus going against the general
+% historial trend discussed in the XXX section).
+% \citet{rasmussen2022espalier} have a less restrictive definition, in
+% which a ``full'' ARG is one that contains complete information about
+% recombination events, without necessarily being tied to a particular
+% stochastic process.
+% While [something positive], it is important to be clear on
+% what the overall goal of such precise inferences are,
+% what the limitations on when they can be made are,
+% and their downstream utility in applications.
+
+% % What is the goal of having fully precise ARGs?
+% % 1) studying the actual recombination events;
+% % 2) evaluating the likelihoods
+% The main reasons for wanting precise estimates of all recombination
+% events are to either compute a likelihood of the inferred ARG
+% under a statistical model, or to study the recombinants in
+% detail~\citep{rasmussen2022espalier}.
+% % Explain that we're really interested in the details of recombination
+% % itself with a small number of samples.
+% In the latter case, we might
+% have a small number of samples~\citep{guo2022recombination}.
+
+% Beyond cases in which visual inspection of the trees and the
+% effects of recombination is feasible, the main application
+% of having completely precise estimates of recombination events is
+% to evaluate the likelihood of the estimate under some
+% statistical model. The only statistical model available
+% is the coalescent with recombination
+% and its approximations~\citep{mcvean2005approximating,marjoram2006fast}.
+% In this case, the utility of a ``full`` ARG for some dataset
+% is limited by how well the dataset fits the assumptions
+% of the model.
+
+% % Not sure where this goes, but there are some critical points
+% % that we want to explain and not just throw out there in the
+% % discussion.
+% % A critical distinction between the eARG and gARG encodings
+% % is the level of precision about the time and location of
+% % recombination events that is required. In an eARG, we are
+% % required to be absolutely precise about every recombination
+% % [explain and contrast with gARG, referring to examples]
+
+% % JK: This is very rough, just jotting down the basic ideas,
+% % will need substantial revision.
+% It is not clear that such complete information will always
+% (or indeed ever) be available from inferences~\cite{shipilina2022origin}.
+%  We may be
+% able to precisely identify the location and timing
+% of recent recombinations (particularly if detailed pedigree
+% information is available) but beyond this, the information
+% required simply isn't present. When sampling from a
+% process such as the Sequentially Markov
+% Coalescent~\citep{mcvean2005approximating,marjoram2006fast}
+% like ARGweaver~\citep{rasmussen2014genome}, it is true that
+% we do obtain a fully resolved ARG with precise
+% information about recombination, but it is reasonable
+% to question how much information from the data is used
+% to support older recombinations, in particular.
+
+% % This idea of a continuum of information about recombination
+% % being inferrable and representable is consistent with the ideas
+% % discussed by~\cite{shipilina2022origin}. There a haplotype block
+% % is synonymous with coalescent nodes in the ARG, which we can generalise
+% % as nodes in a gARG with at least two children. In principle, these
+% % nodes are identifiable by the mutations that occur on the outbound
+% % edges of the node.
+
+% It is helpful to consider the difference between fully
+% resolved binary trees and those that contain polytomies.
+
+% The situation is in fact directly analogous: an ARG that
+% contains fully resolved recombination information must have
+% at most parents per node.
+
+% At the very least, it is not reasonable for the \emph{data
+% structure} to require complete precision in order to
+% represent some less precise. This is a fundamental
+% advantage of the gARG encoding over the classical eARG
+% method: we are not forced to posit a precise cause (recombination
+% event) for every effect (difference between adjacent trees).
+% These differences in precision of recombination
+% information are illustrated in the ARGs estimated by different
+% methods, as we see in the next section.
+
+% In general, discussions tend to contrast an ARG
+% on one hand as holding \emph{complete} information about recombination
+% (and therefore between-tree correlation structure)
+% and a sequence of local trees on the other (with, by implication,
+% no direct information about between-tree correlations).
+% This is a false dichotomy:
+% a \emph{continuum} of information about recombination and between-tree
+% correlation structure can be represented and inferred, and all
+% such levels of information are useful, even if only for computational
+% efficiency.
 
@@
 the methods must then necessarily infer only a sequence of
-unrelated local trees
-% More? See https://github.com/tskit-dev/what-is-an-arg-paper/issues/38
-% for discussion. I'm sure there were other examples of people saying
-% that
-(as claimed by e.g.~\citet{hejase2020summary}),
-and information about recombination and the correlation structure
-between trees is not ``all or nothing''.
-Both methods capture correlation structure between trees:
-\relate\ first infers local tree topologies and then identifies edges
-that persist from one tree to the next as a separate step
-(in a similar manner to \espalier~\citep{rasmussen2022espalier}), while
-\texttt{tsinfer} explicitly constructs nodes that are shared across local
-trees. See \cite{ignatieva2023distribution} for a detailed discussion
-and comparsion of inference methods on this property.
-
-% Because the ARGs inferred by \tsinfer\ and \relate\ do not contain
-% explicit recombination events we cannot compute a likelihood
-% under the coalescent with recombination (see appendix XXX).
-% [Some stuff about whether we should see this as necessary or not,
-% but maybe we should keep this for the conclusion?]
+\emph{unrelated} local trees (i.e., with either unlabelled internal
+nodes or differently labelled nodes in each tree; see the
+next section).
+The ARGs inferred by \relate\ and \tsinfer\ both contain substantial
+information about node sharing across trees, and hence capturing
+the important correlation structures.
+
+% \relate\ by first inferring local topologies and then identifies edges
+% that persist from one tree, while
+% local trees only emerge from \tsinfer\ as a consequence of the copying
+% patterns inferred among ancestral haplotypes.
+
+% % More? See https://github.com/tskit-dev/what-is-an-arg-paper/issues/38
+% % for discussion. I'm sure there were other examples of people saying
+% % that
+% (as claimed by e.g.~\citet{hejase2020summary}),
+% and information about recombination and the correlation structure
+% between trees is not ``all or nothing''.
+% Both methods capture correlation structure between trees:
+% \relate\ first infers local tree topologies and then identifies edges
+% that persist from one tree to the next as a separate step, while
+% \texttt{tsinfer} explicitly constructs nodes that are shared across local
+% trees.
+% % Because the ARGs inferred by \tsinfer\ and \relate\ do not contain
+% % explicit recombination events we cannot compute a likelihood
+% % under the coalescent with recombination (see appendix XXX).
+% % [Some stuff about whether we should see this as necessary or not,
+% % but maybe we should keep this for the conclusion?]
+
+\section*{Equivalance of ARGs and the set of local trees}
+An ARG is sometimes referred to as being equivalent to a set of
+local trees along the genome [TODO citation], but the connection
+between the two is not as simple as it may seem.
+In order for the mapping to be one-to-one, so that the original
+gARG can be exactly recovered from a set of local trees and
+their genome coordinates,
+some specific properties of the ARG and the trees are required.
+
+Firstly, the internal nodes in the local trees must be labelled
+(i.e.\ marked with the identity of the corresponding ARG node).
+If the nodes are not labelled, we must infer their identity
+from the tree topology, which is both
+expensive to compute and imprecise. For example
+we could infer that node \textsf{k} is shared in the second
+and third trees of in Fig.~\ref{fig-simplification}D
+because it subtends identical subtrees in both,
+but it would not be straightforward to infer that it
+is also the root of the first tree.
+% JK: Does this need to be said?
+% (One may use the time of a node as a proxy for identity,
+% but it is not reliable as we can have multiple nodes at the
+% same time, e.g. with time discretisation.)
+
+The precision of the information about recombination that
+can be represented in the trees depends the ARG itself.
+Secondly, the local trees must contain ``unary'' nodes.
+[Some discussion].
+With these two conditions met, there is a direct correspondance
+between a gARG and the trees.
+
+These properties have interesting consequences when we
+consider the inference strategy of esimating local trees
+independenty and subsequently stitching them together
+into an ARG. Two recent methods take this approach,
+\relate\ and
+\texttt{espalier}~\citep{rasmussen2022espalier},
+and have different strategies for reconciling the local
+trees into an ARG. [Relate does something fairly lightweight,
+but by default imprecise.]
+Espalier, on the other hand, is specifically in interested
+in the recombination events and therefore attempts to infer
+the precise ``subtree prune-and-regraft'' (SPR)
+operations~\citep{hein1990reconstructing,song2003on,song2006properties}
+using heuristics~\citep{rasmussen2022espalier}.
+Inferring the SPRs between
+leaf-labelled trees is NP-hard in
+general~\citep{hein1996complexity,allen2001subtree,bordewich2005computational}.
+
+% JK TODO: finish up this section with a quick version of the
+% para here. Basically say, people often say an ARG is a
+% set of local trees cite cite, but it's not as simple as that as we
+% can see from the example (probably).
+
+% The equivalence between an ARG and a set of local trees
+% separated by SPRs is often mentioned [e.g. x y z], but
+% it is important to note that while extracting local
+% trees from an ARG is straightforward, the converse
+% process of reconstructing precisely the same ARG from
+% these trees is more problematic. If the internal
+% nodes are consistently labelled and unary nodes are
+% included in the local trees (e.g., Fig~\ref{fig-simplification})
+% then it is easy to exactly reconstruct the ARG.
+% Otherwise, the problem is much more difficult and it
+% may not be possible to uniquely reconstruct the original ARG.
+% Without labels for the internal nodes, we are forced to
+% to reason about which subtrees remain the same and which
+% differ due to the SPR separating adjacent trees in order to
+% assign node identity. Without the internal unary nodes marking
+% recombinations, we must infer their existence. In
+% general, there will not be a unique solution to this problem,
+% and so there can be many possible ARGs compatible with
+% a given list of leaf-labelled binary trees.
+% This is the approach taken by the Espalier inference
+% method~\citep{rasmussen2022espalier}, which finds a
+% set of SPRs that reconcile independently inferred
+% trees from different regions of the genome.
 
@@
 hejase2020summary,schaefer2021ancestral,
-fan2022genealogical,wohns2022unified,harris2023using}
+fan2022genealogical,wohns2022unified,harris2023using,
+nowbandegani2022extremely}
 is how well inference methods and downstream analysis applications

### 2023-08-29 paper.tex

@@
 \section*{Converting an eARG to a gARG}
-The process of converting an eARG to a gARG is very similar to
-simplifying a prospective gARG with respect to a set of samples,
-and highlights some important differences between the encodings.
+Any eARG can be converted to a gARG without loss of information,
+and the conversion process is very similar to
+simplifying a prospective gARG with respect to a set of samples.
 The process involves two steps, and is illustrated in
 Fig~\ref{fig-ancestry-resolution}.
+The first step is straightforward
+(Figure~\ref{fig-ancestry-resolution}B).
+We duplicate the nodes and
+edges from the input eARG, and the add inheritance annotations
+to the output gARG's edges. If the node is a common ancestor node,
+we annotate the single outbound edge with the interval $(0,L]$,
+for a genome of length $L$.
+If the node is a recombination event with a breakpoint $x$,
+we annotate the outbound edges with the intervals $(0, x]$
+and $(x, L]$.
+These inheritance annotations are clearly in one-to-one
+correspondance with the information in the input eARG,
+but they are a superset of the
+``ancestral material''~\citep{wiuf1999ancestry,wiuf1999recombination}
+of the sample.
 
@@
 
-The first step is straightforward
-(Figure~\ref{fig-ancestry-resolution}B).
-We duplicate the nodes and
-edges from the input eARG, and the add inheritance annotations
-to the output gARG's edges. If the node is a common ancestor node,
-we annotate the single outbound edge with the interval $(0,L]$.
-If the node is a recombination node with a breakpoint $x$,
-we annotate the outbound edges with the intervals $(0, x]$
-and $(x, L]$.
-These inheritance annotations are clearly in one-to-one
-correspondance with the information in the input eARG,
-but they are a superset of the
-``ancestral material''~\citep{wiuf1999ancestry,wiuf1999recombination}
-of the sample.
-
 Ancestral material is a critical concept in a retrospective ARG,
@@
 Firstly, we can see that some nodes and edges are removed entirely
-from the graph. The ``grand MRCA'' $q$ is ommitted from the
+from the graph. The ``grand MRCA'' \noderef{q} is omitted from the
 fully resolved gARG because all segments of the genome have
 fully coalesced before it is reached. Likewise, the edge
-between $g$ and $j$ is omitted because the recombination
-event at position $5$ fell in non-ancestral material.
+between \noderef{g} and \noderef{j} is omitted because the recombination
+event at position $5$ (represented by node \noderef{g})
+fell in non-ancestral material.
 More generally, we can see that the fully resolved
@@
 common ancestor and recombination events in an eARG. Because
-the ancestral material is stored with each edge, the effects
-of multiple events can be reasoned about at once. Many algorithms
+the ancestral material is stored with each edge, the
+cumulative effects of events over time can be reasoned
+about at once. Many computations
 that we wish to perform on an ARG will require resolving
@@
 
-Fig.~\ref{fig-simplification} show a series of successive steps of
-simplification, starting with a complete gARG simulated under a
-backwards-in-time Wright-Fisher
-model, shown both as a graph on the left, and the corresponding sequence of 6
-local trees on the right.
-For example, consider the path above node
-\textsf{a} in the first tree (covering positions 30 to 40)
-in Fig.~\ref{fig-simplification}A.
-Before \textsf{a} meets its common ancestor with the other samples at
-node \textsf{q}, it passes through \textsf{e}, \textsf{j}, \textsf{m}, \textsf{n} and
-\textsf{p}, which in this region of the genome are all unary nodes.
-This reflects the passage of this span of genome through the ARG towards the root.
-Such unary nodes in a local tree are not usually considered important;
-% because they provide little information and are not
-% detectable by information from any given individual tree.
-nodes usually mark the points where samples coalesce in a
- \emph{common} ancestor. For most purposes, we would therefore consider a path directly
-joining sample \textsf{a} to node \textsf{q} to be equivalent to the path
-containing the intermediate unary nodes.
-
 \begin{figure}
@@
 (A) A gARG simulated from a diploid Wright-Fisher
-model. Note that the genome \textsf{k} has three children, and the genomes
-\textsf{e} and \textsf{g} simultaneously have multiple parents and multiple
+model. Note that the genome \noderef{k} has three children, and the genomes
+\noderef{e} and \noderef{g} simultaneously have multiple parents and multiple
 children.
 (B) Simplified to remove all
-% trying out this terminology - can define in the text
-1-connected graph components (e.g., diamonds such as \textsf{jlnm}).
+singly-connected graph components (e.g., diamonds such as \noderef{jlnm}).
 (C) Remove nodes that are unary in all local trees. This removes both recombinant nodes
-that have only one child (e.g. \textsf{n}) and nodes that have more
+that have only one child (e.g.\ \noderef{n}) and nodes that have more
 than one child (i.e. are ``common ancestors'') but never represent coalescences
-in the local trees (e.g. \textsf{r}).
+in the local trees (e.g.\ \noderef{r}).
 (D) Rewrite edges to bypass any unary nodes in local trees. Note that this loses lineage
-information such as the fact that \textsf{k} is connected to \textsf{a} via \textsf{e}.
+information such as the fact that \noderef{k} is connected to \noderef{a} via \noderef{e}.
 }
 \end{figure}
+
+Fig.~\ref{fig-simplification} show a series of successive steps of
+simplification, starting with a complete [TODO be more concrete about
+what is stored here; ``complete'' isn't very helpful] gARG simulated under a
+backwards-in-time Wright-Fisher
+model, shown both as a graph on the left, and the corresponding sequence of
+local trees on the right.
+In Fig.~\ref{fig-simplification}A the local trees (right)
+contain many unary nodes, and as we simplify more and more
+the local trees contain fewer
+unary nodes (Fig.~\ref{fig-simplification}B,C) until
+we reach Fig.~\ref{fig-simplification}D in which the local trees
+have no unary nodes.
+Unary nodes in a local tree arise from
+genetic inheritance
+passing through an ARG node in which no coalescence
+occurs in the corresponding genome interval.
+The distinction between the ``common ancestry'' of two or more genomes
+in an ancestral genome and the ``coalescence'' which may or may
+not occur in the local trees is
+important~\citep{hudson1983testing,kelleher2016efficient}.
+Consider \noderef{e} in in Fig.~\ref{fig-simplification}A,
+for example. We can see from the graph that it is a common
+ancestor of  samples \noderef{a} and \noderef{b}, but
+it does not correspond to any coalescence in the
+local trees to the left of position $44$, and is a unary
+node in these three trees. Unary nodes in the local trees
+arise from a variety of processes, and would not usually
+be considered significant. We are usually interested in tree
+nodes that have at least two children, because these
+are in principle detectable from mutational differences
+arising in the respective subtrees.
+
+% \noderef{a} in the leftmost tree (covering positions 0 to 30)
+% Consider, for example, the path ancestral to node
+% \noderef{a} in the leftmost tree (covering positions 0 to 30)
+% in Fig.~\ref{fig-simplification}A.
+% We can see from the graph that
+% % Before \noderef{a} meets its common ancestor with the other samples at
+% % node \noderef{q}, it passes through \noderef{e}, \noderef{j}, \noderef{m}, \noderef{n} and
+% % \noderef{p}, which in this region of the genome are all unary nodes.
+% % Node \noderef{e}
+
+% Such unary nodes in a local tree are not usually considered important;
+% % because they provide little information and are not
+% % detectable by information from any given individual tree.
+% nodes usually mark the points where samples coalesce in a
+%  \emph{common} ancestor. For most purposes, we would therefore consider a path directly
+% joining sample \noderef{a} to node \noderef{q} to be equivalent to the path
+% containing the intermediate unary nodes.
 
@@
 in which the two parent nodes of a recombination immediately
-join again into a common ancestor (e.g.~\textsf{j}, \textsf{l}, \textsf{m}
-and \textsf{n} in Fig.~\ref{fig-simplification}A).
+join again into a common ancestor (e.g.~\noderef{j}, \noderef{l}, \noderef{m}
+and \noderef{n} in Fig.~\ref{fig-simplification}A).
 Unless we are specifically
@@
 of a node that has one inbound and one outbound edge, such as
-nodes \textsf{f} and \textsf{h}.
+nodes \noderef{f} and \noderef{h}.
 Fig.~\ref{fig-simplification}B shows the result of this type of
@@
 of the local trees. In particular, a node can represent a recombinant
-with multiple parents in the graph but only a single child (e.g.\ node \textsf{n}
+with multiple parents in the graph but only a single child (e.g.\ node \noderef{n}
 in Fig.~\ref{fig-simplification}B), or can represent a common ancestor with
@@
 in the local trees
-(node \textsf{r} in Fig.~\ref{fig-simplification}B).
+(node \noderef{r} in Fig.~\ref{fig-simplification}B).
 % point out that this is why Hudson refers to CA nodes, not coalescence nodes
@@
 removal of recombinant nodes can produce graph nodes with
-more than two parents (e.g.~node \textsf{e}); and likewise, removal of
+more than two parents (e.g.~node \noderef{e}); and likewise, removal of
 common ancestor but non-coalescent nodes can produce graph nodes with
-more than two children (e.g.~node \textsf{s}). These represent a ``stacking up'' of
+more than two children (e.g.~node \noderef{s}). These represent a ``stacking up'' of
 multiple---often indistinguishable---events into a single node.
@@
 % and multiple parents. This would be the case in Fig~\ref{fig-arg-in-pedigree} if,
-% for instance, node \textsf{a} were to have two children.
+% for instance, node \noderef{a} were to have two children.
 % % It is a shame that this is not actually shown in fig 3, but I don't think
@@
 correspond to a coalescence in at least one other
-local tree. For example, node  \textsf{k} is unary in the second tree
+local tree. For example, node  \noderef{k} is unary in the second tree
 of Fig.~\ref{fig-simplification}C, but is either binary
@@
 for the intervals $(44,61]$ and $(61,87]$ are merged because
-the only differences between them are their journeys through
+the only differences between them are their paths through
 nodes \noderef{l} and \noderef{m}. These nodes that participated
@@
 information about recombination in an ARG is not ``all
-or nothing'': there is a continuuum of detail
-about recombination, in terms of genomic location, timing
-the ancestral lineages involved, that we can describe.
+or nothing'': there is a gradation of detail
+about recombination, in terms of genomic location, timing,
+and the ancestral lineages involved.
 While there has been a recent tendancy to emphasise the

### 2023-08-30 argutils/convert.py
Add fix and better output for ARGweaver conversion

@@ -10,7 +10,8 @@ import argutils
 
 def convert_argweaver(infile):
     """
-    Convert an ARGweaver .arg file to a tree sequence. An example .arg file is at
+    Convert an ARGweaver .arg file to a tree sequence. `infile` should be a filehandle,
+    e.g. as returned by the `open` command. An example .arg file is at
 
     https://github.com/CshlSiepelLab/argweaver/blob/master/test/data/test_trans/0.arg
     """
@@ -21,7 +22,9 @@ def convert_argweaver(infile):
     end = int(end[len("end=") :])
     # the "name" field can be a string. Force it to be so, in case it is just numbers
     df = pd.read_csv(infile, header=0, sep="\t", dtype={"name": str, "parents": str})
-
+    for col in ("name", "parents", "age"):
+        if col not in df.columns:
+            raise ValueError(f"Column {col} not found in ARGweaver file")
     name_to_record = {}
     for _, row in df.iterrows():
         row = dict(row)
@@ -48,23 +51,31 @@ def convert_argweaver(infile):
     aw_to_tsk_id = {}
     min_time_diff = min(np.diff(sorted(time_map.keys())))
     epsilon = min_time_diff / 1e6
-    for node in nx.lexicographical_topological_sort(G):
-        record = name_to_record[node]
-        flags = 0
-        # Sample nodes are marked as "gene" events
-        if record["event"] == "gene":
-            flags = tskit.NODE_IS_SAMPLE
-            assert record["age"] == 0
-            time = record["age"]
-        else:
-            time = time_map[record["age"]]
-            # Argweaver allows age of parent and child to be the same, so we
-            # need to add epsilons to enforce parent_age > child_age
-            time_map[record["age"]] += epsilon
-        tsk_id = tables.nodes.add_row(flags=flags, time=time, metadata=record)
-        aw_to_tsk_id[node] = tsk_id
-        if record["event"] == "recomb":
-            breakpoints[tsk_id] = record["pos"]
+    try:
+        for node in nx.lexicographical_topological_sort(G):
+            record = name_to_record[node]
+            flags = 0
+            # Sample nodes are marked as "gene" events
+            if record["event"] == "gene":
+                flags = tskit.NODE_IS_SAMPLE
+                assert record["age"] == 0
+                time = record["age"]
+            else:
+                if record["age"] == 0:
+                    time_map[record["age"]] += epsilon
+                time = time_map[record["age"]]
[…]

### 2023-08-30 paper.tex
Refine

@@
 appear~\citep{osmond2021estimating,fan2022genealogical,hejase2022deep,zhang2023biobank,
-% TODO update when natgen paper comes out
-nowbandegani2022extremely}
+nowbandegani2023extremely}
 and many more are sure to
@@
 recording all such forwards-in-time genetic relationships
-as an ``embellished pedigree`` (and suggest the term ``nedigree''
-as a shorthard).
+as an ``embellished pedigree`` (and suggest the term ``nedigree'' as a shorthand).
 Describing this instead as a ``prospective ARG'', emphasising
@@
 more useful.
-The ``simplify'' algorithm which converts a prospective
-gARG into a retrospective one is a powerful tool,
-with several other applications
-as we discuss in the following sections.
+The ``simplify'' algorithm, which prunes away parts of the graph that 
+are not ancestral to a sample, is the key enabling factor for 
+this approach to forwards-time simulation and is
+described in detail by
+\citet{kelleher2018efficient}. It is a 
+powerful tool with many other applications,
+some of which we discuss in the following sections.
 
@@
 The process involves two steps, and is illustrated in
-Fig~\ref{fig-ancestry-resolution}.
+Fig.~\ref{fig-ancestry-resolution}.
 The first step is straightforward
-(Figure~\ref{fig-ancestry-resolution}B).
+(Fig.~\ref{fig-ancestry-resolution}B).
 We duplicate the nodes and
@@
 (C) The sample-resolved gARG resulting from simplifying with respect
-to the samples, $a$, $b$, and $c$. Dashed lines show edges that are
-no longer present (in practice, nodes $g$ and $q$ would also be removed).
+to the samples, \noderef{a}, \noderef{b}, and \noderef{c}. Dashed lines show edges that are
+no longer present (in practice, nodes \noderef{g} and \noderef{q} would also be removed).
 Coalescence with respect to the sample is indicated by shaded bars, as
@@
 ancestor the total amount ancestral material is reduced, until
-eventually as we traverse pastwards through the graph, all
-samples reach a most recent common ancestor at all locations
+eventually, as we traverse pastwards through the graph, all
+samples reach a most recent common ancestor at every location
 along the genome. The process of transmitting ancestral material
@@
 
-Fig~\ref{fig-ancestry-resolution}C shows the graph that we obtain
+Fig.~\ref{fig-ancestry-resolution}C shows the graph that we obtain
 by resolving the ancestral material transmitted along each edge
-(using the ``simplify'' algorithm), and highlights some important
+using the ``simplify'' algorithm~\citep{kelleher2018efficient}, 
+and highlights some important
 differences between the eARG and gARG encodings.
 Firstly, we can see that some nodes and edges are removed entirely
-from the graph. The ``grand MRCA'' \noderef{q} is omitted from the
-fully resolved gARG because all segments of the genome have
+from the graph by this conversion process. 
+The ``grand MRCA'' \noderef{q} is omitted from the
+sample resolved gARG because all segments of the genome have
 fully coalesced before it is reached. Likewise, the edge
@@
 fell in non-ancestral material.
-More generally, we can see that the fully resolved
-gARG encoding of Fig~\ref{fig-ancestry-resolution}C
+More generally, we can see that the sample resolved
+retrospective gARG of Fig.~\ref{fig-ancestry-resolution}C
 allows for ``local'' inspection
@@
 cumulative effects of events over time can be reasoned
-about at once. Many computations
+about, without first ``replaying'' those events. Many computations
 that we wish to perform on an ARG will require resolving
-the genetic material with respect to a sample.
+the ancestral material with respect to a sample.
 % this is weak, but would be good to emphasise this point.
@@
 allows us to do this once and to store the result, whereas
-the eARG encoding requires us to repeat the process on
-each traversal.
+the eARG encoding requires us to repeat the process 
+each time.
 
 It is important to note here that the \citet{wiuf1999recombination} eARG
-example used here is something of a straw man, in that it is highly unlikely
+in Fig.~\ref{fig-ancestry-resolution} is something of a straw man, 
+in that it is unlikely
 that an inference or simulation program would produce a retrospective
 ARG that contains
-nodes or edges that are not ancestral to the sample. Nonetheless, it
+nodes or edges that are not ancestral to the sample
+(see the discussion of the ``Big ARG'' stochastic process in Appendix XXX,
+however). Nonetheless, it
 is an instructive example from the literature which highlights several
@@
 fan2022genealogical,wohns2022unified,harris2023using,
-nowbandegani2022extremely}
+nowbandegani2023extremely}
 is how well inference methods and downstream analysis applications
@@
 mahmoudi2022bayesian,petr2022slendr,rasmussen2022espalier,
-zhang2023biobank}.
+zhang2023biobank,nowbandegani2023extremely}.
 
@@
 for computing population genetic
-statistics~\citep{kelleher2016efficient,kelleher2018efficient},
+statistics~\citep{kelleher2016efficient,ralph2020efficiently},
 implementing the Li and Stephens
@@
 significant confusion. As first introduced by~\cite{kelleher2016efficient}
-a ``tree sequence`` was defined as the set of ``coalescence records``
+a ``tree sequence`` was defined as the set of ``coalescence records''
 that is output by Hudson's simulation algorithm.

### 2023-08-30 paper.tex
Update

@@
 ARG simplification as the process
-of removing nodes and re-writing edges in a gARG to remove
-various types of redundancy.
-Many of the types of redundancy revolve around the presence
+of removing nodes and re-writing edges (and their inheritance annotations)
+to remove various types of redundancy.
+The redundancy that we are interested in 
+revolves around the presence
 of ``unary'' nodes in the local trees
-(a tree node that has exactly one child).
+(a tree node that has exactly one child), which arise in several different ways.
+Unary nodes are not a standard feature of evolutionary trees;
+we are usually interested
+nodes that have at least two children, because these
+are in principle detectable from mutational differences
+arising in the respective subtrees.
+A tree branch may correspond to many generations 
+of individuals (or indeed cell divisions) and there is usually
+very little information about these intermediates, or utility
+in modelling them. 
+% FIXME the bridge between the next few paragraphs is a bit weird, in 
+% that we've just been talking about how interesting unary nodes are
+% but then go on about how to remove them. Just needs a bit of smoothing
+% out I think.
+ARGs, however, are different, and there 
+can be significant information about evolutionary processes
+provided by unary nodes in the local trees.
 
@@
 In Fig.~\ref{fig-simplification}A the local trees (right)
-contain many unary nodes, and as we simplify more and more
+contain many unary nodes, and as we successively simplify
 the local trees contain fewer
 unary nodes (Fig.~\ref{fig-simplification}B,C) until
-we reach Fig.~\ref{fig-simplification}D in which the local trees
+we reach Fig.~\ref{fig-simplification}D, where the local trees
 have no unary nodes.
-Unary nodes in a local tree arise from
-genetic inheritance
-passing through an ARG node in which no coalescence
-occurs in the corresponding genome interval.
+A node is unary in a local tree covering the genome interval $I$
+if genetic inheritance passes through that ARG node,
+but no coalescence occurs in the interval $I$.
 The distinction between the ``common ancestry'' of two or more genomes
@@
 it does not correspond to any coalescence in the
-local trees to the left of position $44$, and is a unary
-node in these three trees. Unary nodes in the local trees
-arise from a variety of processes, and would not usually
-be considered significant. We are usually interested in tree
-nodes that have at least two children, because these
-are in principle detectable from mutational differences
-arising in the respective subtrees.
+local trees to the left of position $44$, and is a therefore a 
+unary node in these three trees.
 
@@
 history of the sample that is stored. Some of the information lost,
-for example, ``diamond'' removal (Fig.~\ref{fig-simplification}A),
+e.g.\ ``diamond'' removal (Fig.~\ref{fig-simplification}A),
 seems like a reasonable tradeoff for a simpler structure.
@@
 
+% or nothing'': there is a gradation of detail
+% about recombination, in terms of genomic location, timing,
+% and the ancestral lineages involved.
+% that vary levels of precision about reccomb
+
 % The example in Fig.~\ref{fig-simplification} is usually
@@
 
-The point of these examples is to emphasise that
-information about recombination in an ARG is not ``all
-or nothing'': there is a gradation of detail
-about recombination, in terms of genomic location, timing,
-and the ancestral lineages involved.
-While there has been a recent tendancy to emphasise the
-importance of ``full ARGs''
-\citep[e.g.][]{deng2021distribution,brandt2021evaluation,
-rasmussen2022espalier}
-which contain complete information about all recombination
-events, it is reasonable to question the value of
-such precise information and whether it is knowable.
-[SOME MORE STUFF]
+% The point of these examples is to emphasise that
+% information about recombination in an ARG is not ``all
+% or nothing'': there is a gradation of detail
+% about recombination, in terms of genomic location, timing,
+% and the ancestral lineages involved.
+% While there has been a recent tendancy to emphasise the
+% importance of ``full ARGs''
+% \citep[e.g.][]{deng2021distribution,brandt2021evaluation,
+% rasmussen2022espalier}
+% which contain complete information about all recombination
+% events, it is reasonable to question the value of
+% such precise information and whether it is knowable.
+% [SOME MORE STUFF]
 
@@
 \caption{\label{fig-inferred-args} ARGs inferred by
-(A) \kwarg~\citep{ignatieva2021kwarg},
-(B) \argweaver~\citep{rasmussen2014genome,hubisz2020inference},
-(C) \tsinfer~\citep{kelleher2019inferring},
-and (D) \relate~\citep{speidel2019method}
+(A) \kwarg, (B) \argweaver, (C) \tsinfer,
+and (D) \relate\
 for 11 samples of a 2.7kb region of the \textit{Drosophila melanogaster} ADH
-locus~\citep{kreitman1983nucleotide}. The \argweaver\ example
+locus~\citep{kreitman1983nucleotide}. 
+The \argweaver\ example
 is a sample from the inferred posterior distribution. In each ARG,
-colours are assigned to edges based on the time of the edge's child node:
-lighter colours are assigned to edges with older child
-nodes, darker colours are assigned to edges with younger child nodes
-Note that the position of points on the
-Y axis is chosen for graphical clarity by the graphviz \texttt{dot}
-algorithm, rather than as a direct measure of time.
+colours are assigned to edges based on the time of the edge's child node
+(lighter: older; darker: younger).
+Position on the X and Y axes for the graph visualization is arbitrary.
+% Note that the position of points on the
+% Y axis is chosen for graphical clarity by the graphviz \texttt{dot}
+% algorithm, rather than as a direct measure of time.
 Line width corresponds to the genomic span of the edge. Node colour
-refers to the number of parents of the node: red is zero, brown is one,
-green to blue are two to N parents. The brightness
-of the node is determined by the genomic span over which
-the node has two or more children divided by the span where it has any
-children. The genomic intervals associated with edges are visualised below
+refers to the number of parents of the node.
+The fill colour intensity of a node is determined by the genomic span over which
+it has two or more children divided by the span where it has any
+children (the ``coalescent span''). 
+The genomic intervals associated with edges are shown below
 each ARG, where each row is an edge
-and the x axis reflects genomic position. The width of edges
-thus corresponds to their genomic span. Edges
-are ordered by the age of their child node, from oldest to youngest
-moving down the page, with colours matching the edge colours in the ARG.
+and the X axis reflects genomic position. 
+% Edges
+% are ordered by the age of their child node, from oldest to youngest
+% moving down the page, with colours matching the edge colours in the ARG.
 Vertical lines indicate breakpoints between local trees.
@@
 
-To illustrate these points and to emphasise the heterogeneity
-of the ARGs inferred by current methods, we inferred ARGs for the
-the classical \citet{kreitman1983nucleotide} benchmark dataset
-using four recent inference methods and show the results in
-Fig~\ref{fig-inferred-args}.
+The example of successive simplification of a simulated ARG
+in Fig.~\ref{fig-simplification} illustrates the point
+that is is possible to \emph{store} varying levels of precision
+about recombination in an ARG. 
+It is also possible, of course, to \emph{infer} varying levels of precision about 
+recombination.
+To illustrate this, and to demonstrate the structural 
+heterogeneity of the ARGs inferred  by current methods,
+Fig.~\ref{fig-inferred-args} shows the results ARG inference
+% To illustrate these points and to emphasise the heterogeneity
+% of the ARGs inferred by current methods, we inferred ARGs for the
+for the classical \citet{kreitman1983nucleotide} benchmark dataset
+using  \kwarg~\citep{ignatieva2021kwarg},
+\argweaver~\citep{rasmussen2014genome,hubisz2020inference},
+\tsinfer~\citep{kelleher2019inferring},
+and \relate~\citep{speidel2019method}.
 While there is some consistency among the inferred ARGs (such as the samples
@@
 varies from 2 (\relate) to 17 (\argweaver).
+% TODO this needs a bit of rewriting with the rest of the paper in mind
 The ARGs inferred by the different methods also have major differences
-in how recombination information is represented. Essentially,
+in how recombination information is represented. 
+Essentially,
 \kwarg\ and \argweaver\ infer a specific causal recombination \emph{event}
@@
 The ARGs inferred by \relate\ and \tsinfer\ both contain substantial
-information about node sharing across trees, and hence capturing
-the important correlation structures.
-
-% \relate\ by first inferring local topologies and then identifies edges
+information about node sharing across trees, and hence lie somewhere 
+in between a sequence of unrelated trees on one extreme and a 
+completely precise event ARG on the other, in terms of the precision
+inferred about recombination.
+
+% \relate\ "postorder" first inferring local topologies and then identifies edges
 % that persist from one tree, while

### 2023-08-30 paper.tex

@@
 \includegraphics[width=\textwidth]{illustrations/inference.pdf} \end{center}
-\caption{\label{fig-inferred-args} ARGs inferred by
+\caption{\label{fig-inferred-args} Inference results for
 (A) \kwarg, (B) \argweaver, (C) \tsinfer,
@@
 The \argweaver\ example
-is a sample from the inferred posterior distribution. In each ARG,
-colours are assigned to edges based on the time of the edge's child node
+is a sample from the inferred posterior distribution.
+Edge colours are based on the time of the edge's child node
 (lighter: older; darker: younger).
+Line width corresponds to the genomic span of the edge.
 Position on the X and Y axes for the graph visualization is arbitrary.
-% Note that the position of points on the
-% Y axis is chosen for graphical clarity by the graphviz \texttt{dot}
-% algorithm, rather than as a direct measure of time.
-Line width corresponds to the genomic span of the edge. Node colour
+ Node colour
 refers to the number of parents of the node.
@@
 and the X axis reflects genomic position. 
-% Edges
-% are ordered by the age of their child node, from oldest to youngest
-% moving down the page, with colours matching the edge colours in the ARG.
 Vertical lines indicate breakpoints between local trees.
@@
 % TODO this needs a bit of rewriting with the rest of the paper in mind
+% kwarg and argweaver are inferring eARGs, but we need to be careful
+% not to confuse people.
 The ARGs inferred by the different methods also have major differences
@@
 The ARGs inferred by \relate\ and \tsinfer\ both contain substantial
-information about node sharing across trees, and hence lie somewhere 
+information about node sharing across trees, and hence,
+in terms of the level of detail inferred about recombination,
+lie somewhere 
 in between a sequence of unrelated trees on one extreme and a 
-completely precise event ARG on the other, in terms of the precision
-inferred about recombination.
+completely precise event ARG on the other.
 
@@
 
-\section*{Equivalance of ARGs and the set of local trees}
-An ARG is sometimes referred to as being equivalent to a set of
-local trees along the genome [TODO citation], but the connection
-between the two is not as simple as it may seem.
-In order for the mapping to be one-to-one, so that the original
-gARG can be exactly recovered from a set of local trees and
-their genome coordinates,
-some specific properties of the ARG and the trees are required.
-
-Firstly, the internal nodes in the local trees must be labelled
-(i.e.\ marked with the identity of the corresponding ARG node).
-If the nodes are not labelled, we must infer their identity
-from the tree topology, which is both
-expensive to compute and imprecise. For example
-we could infer that node \textsf{k} is shared in the second
-and third trees of in Fig.~\ref{fig-simplification}D
-because it subtends identical subtrees in both,
-but it would not be straightforward to infer that it
-is also the root of the first tree.
-% JK: Does this need to be said?
-% (One may use the time of a node as a proxy for identity,
-% but it is not reliable as we can have multiple nodes at the
-% same time, e.g. with time discretisation.)
-
-The precision of the information about recombination that
-can be represented in the trees depends the ARG itself.
-Secondly, the local trees must contain ``unary'' nodes.
-[Some discussion].
-With these two conditions met, there is a direct correspondance
-between a gARG and the trees.
-
-These properties have interesting consequences when we
-consider the inference strategy of esimating local trees
-independenty and subsequently stitching them together
-into an ARG. Two recent methods take this approach,
-\relate\ and
-\texttt{espalier}~\citep{rasmussen2022espalier},
-and have different strategies for reconciling the local
-trees into an ARG. [Relate does something fairly lightweight,
-but by default imprecise.]
-Espalier, on the other hand, is specifically in interested
-in the recombination events and therefore attempts to infer
-the precise ``subtree prune-and-regraft'' (SPR)
-operations~\citep{hein1990reconstructing,song2003on,song2006properties}
-using heuristics~\citep{rasmussen2022espalier}.
-Inferring the SPRs between
-leaf-labelled trees is NP-hard in
-general~\citep{hein1996complexity,allen2001subtree,bordewich2005computational}.
+% \section*{Equivalance of ARGs and the set of local trees}
+% An ARG is sometimes referred to as being equivalent to a set of
+% local trees along the genome [TODO citation], but the connection
+% between the two is not as simple as it may seem.
+% In order for the mapping to be one-to-one, so that the original
+% gARG can be exactly recovered from a set of local trees and
+% their genome coordinates,
+% some specific properties of the ARG and the trees are required.
+
+% Firstly, the internal nodes in the local trees must be labelled
+% (i.e.\ marked with the identity of the corresponding ARG node).
+% If the nodes are not labelled, we must infer their identity
+% from the tree topology, which is both
+% expensive to compute and imprecise. For example
+% we could infer that node \textsf{k} is shared in the second
+% and third trees of in Fig.~\ref{fig-simplification}D
+% because it subtends identical subtrees in both,
+% but it would not be straightforward to infer that it
+% is also the root of the first tree.
+% % JK: Does this need to be said?
+% % (One may use the time of a node as a proxy for identity,
+% % but it is not reliable as we can have multiple nodes at the
+% % same time, e.g. with time discretisation.)
+
+% The precision of the information about recombination that
+% can be represented in the trees depends the ARG itself.
+% Secondly, the local trees must contain ``unary'' nodes.
+% [Some discussion].
+% With these two conditions met, there is a direct correspondance
+% between a gARG and the trees.
+
+% These properties have interesting consequences when we
+% consider the inference strategy of esimating local trees
+% independenty and subsequently stitching them together
+% into an ARG. Two recent methods take this approach,
+% \relate\ and
+% \texttt{espalier}~\citep{rasmussen2022espalier},
+% and have different strategies for reconciling the local
+% trees into an ARG. [Relate does something fairly lightweight,
+% but by default imprecise.]
+% Espalier, on the other hand, is specifically in interested
+% in the recombination events and therefore attempts to infer
+% the precise ``subtree prune-and-regraft'' (SPR)
+% operations~\citep{hein1990reconstructing,song2003on,song2006properties}
+% using heuristics~\citep{rasmussen2022espalier}.
+% Inferring the SPRs between
+% leaf-labelled trees is NP-hard in
+% general~\citep{hein1996complexity,allen2001subtree,bordewich2005computational}.
 
@@
 A necessary first step towards this goal is some terminological
-clarity. As discussed in Appendix XXX, the term ARG has several
+clarity. As discussed in Appendix XXX, the term Ancestral Recombination
+Graph has several
 subtly different interpretations depending on context. The
@@
 flexible and efficient. 
-
-The flexiblity of the gARG encoding contrasts with the classical
+The flexibility of the gARG encoding contrasts with the classical
 ``event'' ARG (eARG), which, because of its origins in an idealised
 mathematical model, is more limited in what can be described.
-The necessity of complete precision under the eARG encoding has been
-conflated with the idea of a ``true'' or ``full'' ARG.
-[A discussion of how realistic inferring a completely precise ARG is]
-[Do downstream applications use full ARGs? usually no]
+
+The necessity of complete precision about all recombination
+events under the eARG encoding has been
+conflated with the idea of a ``true'' or ``full'' ARG. [Cites]
+For example, \citet{brandt2021evaluation} define a ``full'' ARG as
+``a structure that encodes all
+coalescence and recombination events''. While there are certainly cases
+when inferring such fully precise ARGs is useful, for example when 
+studying the details of a small number of recombination
+events~\citep{rasmussen2022espalier}, it is worth considering
+whether sufficient information exists to support such inferences
+and indeed whether this precision is required (or indeed useful
+for) downstream applications.
+
+[Discuss information about recomb]
+\citet{shipilina2022origin}
 
 % % Do downstream applications actually make use of such full ARGs?
-% Downstream applications using estimated ARGs to date rarely
-% make use of such complete recombination information.
-% The \relate\ selection statistic~\citep{speidel2019method}
-% computes a $p$-value for selection [FIXME] for each
-% local tree independently by comparing the rates of coalescence
-% at various time-points. For cases like strong positive selection,
-% where we expect there to be a clear signal in the local tree,
-% such tree-by-tree methods can be simple and powerful.
-% In their method
-% for estimating dispersal rates and the locations of genetic
-% ancestors,
-% \citep{osmond2021estimating} downsample trees along the genome
-% so that they can be regarded as approximately independent.
-% The SIA method for detecting selection~\citep{hejase2022deep}
-% encodes local trees as a set of lineage counts at discrete
-% time intervals, and uses these as feature for a
-% type of machine learning algorithm
-% that takes ``temporal'' correlations into account. Thus,
-% while SIA does use some of the information about local tree correlation,
-% clearly much of the detail about recombination
-% events in a ``full'' ARG is lost.
-% The \texttt{tsdate} algorithm (and related naive estimator
-% for ancestral location) uses much more between-tree
-% information~\citep{wohns2022unified}, explicitly using the gARG
-% encoding to reason about ancestral nodes.
+Downstream applications using estimated ARGs to date rarely
+make use of such complete recombination information.
+The \relate\ selection statistic~\citep{speidel2019method}
+computes a $p$-value for selection [FIXME] for each
+local tree independently by comparing the rates of coalescence
+at various time-points. For cases like strong positive selection,
+where we expect there to be a clear signal in the local tree,
+such tree-by-tree methods can be simple and powerful.
+In their method
+for estimating dispersal rates and the locations of genetic
+ancestors,
+\citep{osmond2021estimating} downsample trees along the genome
+so that they can be regarded as approximately independent.
+The SIA method for detecting selection~\citep{hejase2022deep}
+encodes local trees as a set of lineage counts at discrete
+time intervals, and uses these as feature for a
+type of machine learning algorithm
+that takes ``temporal'' correlations into account. Thus,
+while SIA does use some of the information about local tree correlation,
+clearly much of the detail about recombination
+events in a ``full'' ARG is lost.
+The \texttt{tsdate} algorithm (and related naive estimator
+for ancestral location) uses much more between-tree
+information~\citep{wohns2022unified}, explicitly using the gARG
+encoding to reason about ancestral nodes.
+[MORE applications]
+
 % A major strength of this approach is that it can be used on

### 2023-08-31 paper.tex
Wilders comments

@@
 \author[4]{Gregor Gorjanc}
-\author[5]{Anthony W. Wohns}
+\author[5,6]{Anthony W. Wohns}
 
@@
 \affil[5]{Broad Institute of MIT and Harvard, Cambridge, MA 02142, USA}
+\affil[6]{Department of Genetics, Stanford University School of Medicine, Stanford, CA 94305, USA}
 
@@
 events.
+% Wilder: important to get this point into the intro
+We argue that while adopting the general concept of an ARG data structure
+is useful, it is important that we then fully decouple this data structure
+from the original stochastic process.
 [MORE]
@@
 % we call a ``genome ARG'' (gARG).
-In this section we define a concrete definition of an ARG in terms
+In this section we provide a  concrete definition of an ARG in terms
 of particular genomes and their patterns of inheritance, which
@@
 there is
-is no limitation on mating system, ploidy, or age structure. By making the
+no limitation on mating system, ploidy, or age structure. By making the
 basic unit of analysis an individual's \emph{genomes} (i.e., two per individuals
@@
 and so on can be directly expressed.
-
-Thirdly, ``local`` trees that we generate for a particular position on the
+% Wilder: give forward ref to section on efficiency here, mention tree sequence
+
+Thirdly, ``local'' trees that we generate for a particular position on the
 genome are not explicitly concerned with ``coalescence'' (nodes at which
@@
 paternal genomes which this child has inherited segments from.
+% TODO rephrase this, it'll confuse people. Just need to reiterate the
+% points about events, and maybe lead into the next section
 Finally, gARGs are not defined in terms of the
@@
 ancestor to descendant through some
+% Wilder: finds the cell-to-cell confusing. Maybe drop?
 number of generations (ultimately from cell to cell),
@@
 correspond to a coalescence in at least one other
-local tree. For example, node  \noderef{k} is unary in the second tree
+local tree. For example, node  \noderef{k} is unary in the fourth tree
 of Fig.~\ref{fig-simplification}C, but is either binary
@@
 but in reality there is  substantial work to be done to
-enable this, and much remains to be done.
+enable this.
 

### 2023-09-01 paper.tex
Fill out discussion a bit

@@
 inherits from one parent (i.e., haploids have one genome, diploids two, etc.).
-Then, a gARG is a graph in which nodes represent
+In this usage a ``genome'' is equivalent to the definition of a ``haplotype'' used by
+\cite{shipilina2023origin}; we avoid term haplotype because 
+it already has many interpretations, and in particular carries the implication of 
+specific mutational patterns [CITES] and the existance of a particular 
+haplotype over many generations. A ``genome'' in our usage corresponds to the 
+genetic material present in a particular individual or cell.
+We will also use the term ``genome`` in its 
+more common sense of ``the genome'' of a species, denoting the full set 
+of chromosomes and their coordinate spaces, and hope that the distinction
+will be clear from the context.
+We consider only patterns of genetic inheritance, with mutational 
+processes being entirely separate.
+
+A gARG is then a graph in which nodes represent
 monoploid genomes and edges represent
@@
 
+\cite{mathieson2020ancestry} have a slightly different definition of an ARG,
+which they say ``is a subset of the pedigree... contain[ing] only those edges along which
+inherited segments of DNA have been transmitted, and only those nodes
+corresponding to ancestors in which there was a recombination or coalescence
+event.'' Thus, while their definition is decoupled from the coalescent
+process it is defined by events.
+
 % \section*{Inheritance and ancestral material}\label{Ancestry_resolution}
@@
 % It is not clear that such complete information will always
-% (or indeed ever) be available from inferences~\cite{shipilina2022origin}.
+% (or indeed ever) be available from inferences~\cite{shipilina2023origin}.
 %  We may be
@@
 % % being inferrable and representable is consistent with the ideas
-% % discussed by~\cite{shipilina2022origin}. There a haplotype block
+% % discussed by~\cite{shipilina2023origin}. There a haplotype block
 % % is synonymous with coalescent nodes in the ARG, which we can generalise
@@
 
-% \section*{Equivalance of ARGs and the set of local trees}
-% An ARG is sometimes referred to as being equivalent to a set of
-% local trees along the genome [TODO citation], but the connection
-% between the two is not as simple as it may seem.
-% In order for the mapping to be one-to-one, so that the original
-% gARG can be exactly recovered from a set of local trees and
-% their genome coordinates,
-% some specific properties of the ARG and the trees are required.
-
-% Firstly, the internal nodes in the local trees must be labelled
-% (i.e.\ marked with the identity of the corresponding ARG node).
-% If the nodes are not labelled, we must infer their identity
-% from the tree topology, which is both
-% expensive to compute and imprecise. For example
-% we could infer that node \textsf{k} is shared in the second
-% and third trees of in Fig.~\ref{fig-simplification}D
-% because it subtends identical subtrees in both,
-% but it would not be straightforward to infer that it
-% is also the root of the first tree.
-% % JK: Does this need to be said?
-% % (One may use the time of a node as a proxy for identity,
-% % but it is not reliable as we can have multiple nodes at the
-% % same time, e.g. with time discretisation.)
-
-% The precision of the information about recombination that
-% can be represented in the trees depends the ARG itself.
-% Secondly, the local trees must contain ``unary'' nodes.
-% [Some discussion].
-% With these two conditions met, there is a direct correspondance
-% between a gARG and the trees.
+\section*{Equivalance of ARGs and the set of local trees}
+An ARG is sometimes referred to as being equivalent to a set of
+local trees along the genome [TODO citation], but the connection
+between the two is not as simple as it may seem.
+In order for the mapping to be one-to-one, so that the original
+gARG can be exactly recovered from a set of local trees and
+their genome coordinates,
+some specific properties of the ARG and the trees are required.
+
+Firstly, the internal nodes in the local trees must be labelled
+(i.e.\ marked with the identity of the corresponding ARG node).
+If the nodes are not labelled, we must infer their identity
+from the tree topology, which is both
+expensive to compute and imprecise. For example
+we could infer that node \textsf{k} is shared in the second
+and third trees of in Fig.~\ref{fig-simplification}D
+because it subtends identical subtrees in both,
+but it would not be straightforward to infer that it
+is also the root of the first tree.
+% JK: Does this need to be said?
+% (One may use the time of a node as a proxy for identity,
+% but it is not reliable as we can have multiple nodes at the
+% same time, e.g. with time discretisation.)
+
+The precision of the information about recombination that
+can be represented in the trees depends the ARG itself.
+Secondly, the local trees must contain ``unary'' nodes.
+[Some discussion].
+With these two conditions met, there is a direct correspondance
+between a gARG and the trees.
+
+\cite{shipilina2023origin} discuss the same ideas, and note that 
+``full ARG ... contains more information than the series of tree
+sequences along the genome''. They view the ``full ARG'' as an
+eARG, and the ``series of tree sequences'' is their 
+% jk: probably should delete this. It *is* muddled though
+(somewhat muddled) 
+term for the sequence of local trees along the genome. We can see
+that their statment is true, \emph{if} the local trees contain
+no unary nodes. However, as we have shown, if the local trees 
+contain labelled unary nodes, then  there is precisely as 
+much information in them as an ARG. Whether such unary nodes
+are \emph{observable} is an entirely separate question,
+of course.
 
@@
 whether sufficient information exists to support such inferences
-and indeed whether this precision is required (or indeed useful
-for) downstream applications.
-
-[Discuss information about recomb]
-\citet{shipilina2022origin}
+and indeed whether this precision is required, or useful,
+for downstream applications.
+
+The amount of information about recombination that is inferrable 
+is the subject of much research
+has a long history [summarise this in a couple of sentences?]
+\cite{shipilina2023origin} discuss the idea of a ``haplotype block''
+(equivalent to the ``bricked tree sequence'' of
+\cite{nowbandegani2023extremely}), where we consider the unique sets of 
+samples that coalesce at a node in the ARG over a particular 
+genome interval, and the fundamental limitations they place 
+on ARG inference.
 
 % % Do downstream applications actually make use of such full ARGs?
-Downstream applications using estimated ARGs to date rarely
+Besides the inherent limitations that exist on inferring fully precise ARGs
+from data,
+we should also consider the value that such precise estimates provide
+to downstream applications. 
+To date, few downstream applications using estimated ARGs 
 make use of such complete recombination information.
@@
 local tree independently by comparing the rates of coalescence
-at various time-points. For cases like strong positive selection,
-where we expect there to be a clear signal in the local tree,
-such tree-by-tree methods can be simple and powerful.
+at various time-points. 
 In their method
@@
 [MORE applications]
-
-% A major strength of this approach is that it can be used on
-% any gARG, ranging from estimates containing relatively little
-% information about recombination (e.g., as estimated by \relate)
-% to ``full'' ARGs containing complete recombination information
-% (e.g., as estimated by \kwarg), and will fully utilise
-% any between-tree correlation information that is present.
+The main application for fully precise ARGs thus far has been
+to be to compute a likelihood under the coalescent with 
+recombination~\citep[e.g.][]{kuhner2000maximum,mahmoudi2022bayesian}, 
+which currently requires the details of all recombination
+events to be known.
 
@@
 with recombination stochastic process is an important step.
-While the coalescent with recombination is a very useful and
+While it has proven to be a useful and
 robust model~\citep{wakeley2012gene,bhaskar2014distortion,nelson2020accounting}

### 2023-09-01 paper.tex
Intro draft

@@
 Although it is widely accepted that ARGs are important, there is significant
-confusion about what, precisely, an ARG \emph{is}. [THIS IS BAD, WHY?]
-We also lack a shared format for interchanging ARG data [which is bad
-for users, progress]. [Because we don't really agree on what an ARG is,
-we don't really know what properties it should have, when we infer it].
+confusion about what, precisely, an ARG \emph{is}. 
+Originally, the ARG was defined as a variant 
+of the coalescent with recombination stochastic 
+process~\citep{griffiths1991two,ethier1990two,griffiths1996ancestral,griffiths1997ancestral},
+but has subsequently come to be seen as a data
+structure~\citep{minichiello2006mapping}, i.e., a means of 
+describing the \emph{outcome} of some such ancestral process
+(the distinction, however, is not clear cut and subfields use the term
+differently; see Appendix XXX).
+ARG inference was for many years synonymous with estimating realisations
+of the coalescent with recombination (with, it must be said, limited success),
+and the distinction between process and realisation was not especially important.
+The recent introduction of two methods, \relate~\citep{speidel2019method}
+and \tsinfer~\citep{kelleher2019inferring}, that estimate structures 
+sharing many properties of this classical idea of an ARG at previously
+unthinkable scale, has led to a surge of interest in the area.
+Although both \cite{speidel2019method} and \cite{kelleher2019inferring}
+avoid describing the output of the methods as an ARG, the general 
+consensus has emerged that they belong to a growing class of 
+ARG inference methods~\citep[MORE][]{mathieson2020ancestry},
+and we therefore need a more general definition of an ``ARG'' than the classical
+Griffiths formulation.
+
+A general term to describe the detailed genetic genealogy of a sample 
+that encompasses the output of all these methods is indeed useful, but 
+there are substantial problems emerging from the current usage of the term 
+ARG (and its implicit concrete definition in the Griffiths formulation).
+The problem is much deeper than some pedantic discussion over terminology,
+and directly informs what we should be attempting to estimate, how we 
+evaluate inference methods, and what the limitations on the precision
+and accuracy of our inferences are.
+We currently lack a well-defined target for inference, and therefore there is 
+disagreement on how these inference should be evaluated, and 
+what the criteria for success should be.
+We also lack a shared format for interchanging ARG data between
+programs, with the assorted negative consequences which
+geneticists are only too familiar with~\citep{excoffier2006computer}.
+Finally, and perhaps most importantly, and overly narrow definition of ARGs
+leads to potentially missed opportunities in their application and usage.
 
 % What do we do?
-Here we precisely define an encoding for ARGs based on specific genomes
-and their intervals of genetic inheritance. We contrast this ``gARG''
-encoding with the classical approach derived from the coalescent
-with recombination stochastic process of Griffiths
-and colleagues~\citep{griffiths1991two,ethier1990two,
-griffiths1996ancestral,griffiths1997ancestral}. As this is based
-on recording evolutionary events, we refer to it as the event ARG, or
-``eARG'' encoding.
-We show that the gARG encoding is both flexible and efficient,
-and can represent varying levels of precision about historical
-events.
-% Wilder: important to get this point into the intro
-We argue that while adopting the general concept of an ARG data structure
-is useful, it is important that we then fully decouple this data structure
-from the original stochastic process.
-[MORE]
+In this paper we attempt to address these problems by 
+providing a precise definition of an ARG that is general enough to encompass
+the output of all currently available inference methods. 
+The detailed genetic genealogy of a sample is defined in terms 
+of based on specific genomes
+and their intervals of genetic inheritance, which we refer to as 
+a ``genome ARG'', or ``gARG''.
+This contrasts with the classical Griffiths approach which 
+is based on recording evolutionary events, and we therefore refer to as 
+the ``event ARG'', or ``eARG'' encoding.
+We show that the gARG encoding has a range of advantages over the 
+eARG approach, being both flexible and efficient.
+In particular, we argue that the goal of inferring fully precise 
+ARGs---a necessity under the eARG encoding---should be critically 
+examined as part of the process of fully decoupling our ARG data 
+structure from the coalescent with recombination model.
 
@@
 In this section we provide a  concrete definition of an ARG in terms
-of particular genomes and their patterns of inheritance, which
+of specific genomes and their patterns of inheritance, which
 we refer to as a ``genome ARG``, or gARG.
@@
 local trees along the genome [TODO citation], but the connection
-between the two is not as simple as it may seem.
+between the two is subtle.
 In order for the mapping to be one-to-one, so that the original

### 2023-09-01 paper.tex
Update

@@
 and \tsinfer~\citep{kelleher2019inferring}, that estimate structures 
-sharing many properties of this classical idea of an ARG at previously
+sharing many properties of this classical idea of an ARG, but at previously
 unthinkable scale, has led to a surge of interest in the area.
@@
 it already has many interpretations, and in particular carries the implication of 
-specific mutational patterns [CITES] and the existance of a particular 
+specific mutational patterns [CITES] and the persistance of a particular 
 haplotype over many generations. A ``genome'' in our usage corresponds to the 
@@
 of the genomic intervals over which genetic inheritance occurs.
+\footnote{We use the standard graph theoretic sense of
+an ``edge'' here, rather than the somewhat specialised interpretation of \cite{shipilina2023origin}.}
 We can therefore view a gARG for a given set of individuals
-as a subset of their pedigree, but, crucially,
+as a subset of their pedigree, but
 where each individual is represented by two nodes (for diploids).
-% To recover the local tree for a genomic
-% position $x$, we traverse the graph rootwards from the leaves
-% (as in an eARG) and at a particular node, we follow the outbound
-% edge associated with the genomic interval containing $x$.
-% Recent discussions have illustrated the
-% relationship between ARGs and
-% pedigrees~\citep{mathieson2020ancestry,brandt2021evaluation},
-% and the deep connections between the structures are
-% well understood
-% \citep[e.g.][]{wakeley2012genetics,gusfield2014recombinatorics,
-% speed2015naturereviewsgenetics}.
 This is illustrated in Fig.~\ref{fig-arg-in-pedigree},
@@
 The edges in a gARG specify that genetic inheritance occurred between an
-ancestor and a descendant, but the connectivity information alone
+ancestor and a descendant, but the graph connectivity 
 does not tell us which \emph{parts} of their genomes were inherited.
-
+In order to capture the effects of recombination (and any other process
+that results in the partial inheritance of a descendant genome 
+from an ancestor) we must ``annotate'' the edges with the genome
+coordinates over which inheritance occurs.
+Continuing the example of genome \textsf{a} in Fig.~\ref{fig-arg-in-pedigree},
+the recombination at position 2 means that it inherits the (half-closed) genome
+interval $(0, 2]$ from \textsf{e} and it inherits $(2, 10]$ from \textsf{f}.
+These intervals are shown attached to the corresponding graph edges.
+
+% JK: dunno if this will make some readers switch off, but I feel it's 
+% important to show that this is a formal model and the concrete definition
+% will help some readers?
 Formally, we can define a gARG as a set of nodes $N$ and edges $E$, where each
-element of $E$ is a tuple $(c, p, I)$ such that
-$c\in N$ is the child node, $p\in N$ is the parent node and $I$ is the set of
-disjoint genomic intervals $(\ell, r]$ over which genome $c$ inherits from $p$ (we may
-also phrase this in terms of the ancestral material of a sample---see
-the next section).
-% To recover the local tree for a genomic
-% position $x$, we traverse the graph rootwards from the leaves
-% (as in an eARG) and at a particular node, we follow the outbound
-% edge associated with the genomic interval containing $x$.
-To recover the local tree at genome coordinate $x$ (the most
-fundamental operation required for an ARG data structure; see
-Section XXX),
-we traverse the graph rootwards from the leaves.
-Starting with the sample nodes $S$ we traverse
-rootwards through the graph from child to parent. At each node $u$, we find an
-edge $(c, p, I) \in E$ such that $u = c$ and $x \in I$
-and move to the next node $p$. The efficiency of this process is discussed
-in Section XXX).
+element of $E$ is a tuple $(c, p, I)$ such that $c, p \in N$ are the child and
+parent nodes and $I$ is the set of disjoint genomic intervals $(\ell, r]$ over
+which genome $c$ inherits from $p$. The most fundamental operation for
+an ARG data structure is the process of recovering the corresponding sequence
+of local trees along the genome. To generate the local tree at genome
+coordinate $x$, we traverse the graph rootwards from the leaves. 
+For concreteness, we can describe this tree as an oriented forest $\pi$
+\citep{kelleher2013coalescent,kelleher2014coalescent,kelleher2016efficient}.
+At each node
+$u$, we find an edge $(c, p, I) \in E$ such that $u = c$ and $x \in I$.
+We then set $\pi_c \leftarrow p$,  and move to the next node $p$.
+At the end of this process, $\pi$ fully describes the local tree 
+at position $x$.
 

### 2023-09-01 paper.tex

@@
 accommodated.
-Secondly, because we only record the \emph{intervals} of inheritance
-between genomes (and not the causal \emph{processes}),
+Secondly, because we record the intervals of inheritance
+between genomes (and not causal \emph{events}),
 arbitrary patterns of genetic inheritance such as
@@
 % Wilder: give forward ref to section on efficiency here, mention tree sequence
-
 Thirdly, ``local'' trees that we generate for a particular position on the
 genome are not explicitly concerned with ``coalescence'' (nodes at which
-one or more samples find a common ancestor); for example, node $e$
+one or more samples find a common ancestor); for example, node \noderef{e}
 is unary in the left-most tree of Figure~\ref{fig-arg-in-pedigree},
-denoting the fact that $e$ the ancestor of $a$ over the interval $(0, 2]$,
+denoting the fact that \noderef{e} the ancestor of \noderef{a} over the interval $(0, 2]$,
 but no coalescence occurs in this interval (see section XXX for more
@@
 its outcomes in terms of inheritance intervals among parental genomes;
-thus, we will have a child node (the recombinant; e.g. node $a$
+thus, we will have a child node (the recombinant; e.g. node \noderef{a}
 Figure~\ref{fig-arg-in-pedigree}) and two (or more; see section XXX)
@@
 % genome-wide genetic ancestry.
-As illustrated in Fig.~\ref{fig-arg-data-structure}, this information
-is usually concretely encoded by associating a recombination breakpoint
-(which is null for common ancestor events) and an ordered list of parents
-with each node. We refer to such a  data structure as an
-``event ARG'' (eARG).
-To recover the local tree at genome coordinate $x$ (the most
-fundamental operation required for an ARG data structure),
-we traverse the graph rootwards from the leaves.
-At a particular
-node $u$, if it has one parent we are at a common ancestor
-node and we follow that parent. If we are at a
-recombination node, $u$ has two parents: if
-$x$ is less than the breakpoint we follow
-the edge to the first parent, and otherwise follow the edge
-to the second parent.
-The order in which parents are listed at a recombination node is
-therefore significant, telling us
-from which parent the segments of genome to the left and right of the breakpoint
-were inherited.
+% As illustrated in Fig.~\ref{fig-arg-data-structure}, this information
+% is usually concretely encoded by associating a recombination breakpoint
+% (which is null for common ancestor events) and an ordered list of parents
+% with each node. We refer to such a  data structure as an
+% ``event ARG'' (eARG).
+% To recover the local tree at genome coordinate $x$ (the most
+% fundamental operation required for an ARG data structure),
+% we traverse the graph rootwards from the leaves.
+% At a particular
+% node $u$, if it has one parent we are at a common ancestor
+% node and we follow that parent. If we are at a
+% recombination node, $u$ has two parents: if
+% $x$ is less than the breakpoint we follow
+% the edge to the first parent, and otherwise follow the edge
+% to the second parent.
+% The order in which parents are listed at a recombination node is
+% therefore significant, telling us
+% from which parent the segments of genome to the left and right of the breakpoint
+% were inherited.
 

### 2023-09-01 paper.tex
Some updates

--- 2023-09-01 paper.tex
@@
 differently; see Appendix XXX).
-ARG inference was for many years synonymous with estimating realisations
+ARG inference was for many years synonymous [FIXME: this isn't really true,
+what about parsimony ARGs?]
+with estimating realisations
 of the coalescent with recombination (with, it must be said, limited success),
@@
 it already has many interpretations, and in particular carries the implication of 
-specific mutational patterns [CITES] and the persistance of a particular 
-haplotype over many generations. A ``genome'' in our usage corresponds to the 
+specific mutational patterns [CITES] which persist over 
+many generations. A ``genome'' in our usage corresponds to the 
 genetic material present in a particular individual or cell.
@@
 genetic inheritance between an ancestor and a descendant,
-where each edge is annotated with the coordinates
+where each 
+edge\footnote{We use the standard graph theoretic sense of
+an ``edge'' here, rather than the somewhat specialised interpretation of \cite{shipilina2023origin}.}
+ is annotated with the coordinates
 of the genomic intervals over which genetic inheritance occurs.
-\footnote{We use the standard graph theoretic sense of
-an ``edge'' here, rather than the somewhat specialised interpretation of \cite{shipilina2023origin}.}
 We can therefore view a gARG for a given set of individuals
@@
 \section*{Event ARGs}\label{eARG}
-We are mainly interested in exploring the properties of gARGs in
-this manuscript, but in order to contrast with the existing
-methods we concretely define an ``event ARG'', or eARG.
+In keeping with its original derivation in terms of a stochastic
+process (see Appendix~\ref{sec-arg-history}), the classical definition of an
+ARG data structure is
+in terms of events.
+Nodes represent
+common ancestor or recombination events that occurred in the
+history of some set of sample genomes, and edges represent ancestral
+lineages~\citep{griffiths1996ancestral}.
+In a common ancestor event the inbound lineages are merged into a
+single ancestral lineage, and in a recombination
+event a lineage is split into two independent
+ancestral lineages.
+The final vital
+detail is that we associate a \emph{breakpoint} with each recombination
+event, and this information is sufficient to uniquely
+define the local genealogical trees at every position along the genome.
+
+As illustrated in Fig.~\ref{fig-arg-data-structure}, this information
+is usually concretely encoded by associating a recombination breakpoint
+(which is null for common ancestor events) and an ordered list of parents
+with each node. We refer to such a  data structure as an
+``event ARG'' (eARG).
+
+% To recover the local tree at genome coordinate $x$ (the most
+% fundamental operation required for an ARG data structure),
+% we traverse the graph rootwards from the leaves.
+% At a particular
+% node $u$, if it has one parent we are at a common ancestor
+% node and we follow that parent. If we are at a
+% recombination node, $u$ has two parents: if
+% $x$ is less than the breakpoint we follow
+% the edge to the first parent, and otherwise follow the edge
+% to the second parent.
+% The order in which parents are listed at a recombination node is
+% therefore significant, telling us
+% from which parent the segments of genome to the left and right of the breakpoint
+% were inherited.
+
+% [TODO list the salient points that we want to contrast against
+% the gARG. We don't want this to seem like a straw man]
+
+% Griffiths and colleagues~\citep{griffiths1997ancestral} originally defined 
+% the ARG as analog of the Kingman coalescent~\citep{kingman1982coalescent},
+% in which the ancestry of a sample of $n$ genomes under idealised 
+% conditions is modelled. One of the key assumptions of this model is that 
+% the sample size $n$ is much less than the population size $N$, which means 
+% that the probability of two multiple concurrent events (i.e., coalescence
+% and recombination events) is negligible.
 
@@
 
-In keeping with its original derivation in terms of a stochastic
-process (see Appendix~\ref{sec-arg-history}), the classical definition of an
-ARG data structure is
-in terms of events.
-Nodes represent
-common ancestor or recombination events that occurred in the
-history of some set of sample genomes, and edges represent ancestral
-lineages~\citep{griffiths1996ancestral}.
-In a common ancestor event the inbound lineages are merged into a
-single ancestral lineage, and in a recombination
-event a lineage is split into two independent
-ancestral lineages.
-% The final vital
-% detail is that we associate a breakpoint with each recombination
-% event.
-% This information is sufficient to uniquely
-% define the local genealogical trees at every position along the genome,
-% which is the basic requirement for a data structure encoding
-% genome-wide genetic ancestry.
-As illustrated in Fig.~\ref{fig-arg-data-structure}, this information
-is usually concretely encoded by associating a recombination breakpoint
-(which is null for common ancestor events) and an ordered list of parents
-with each node. We refer to such a  data structure as an
-``event ARG'' (eARG).
-To recover the local tree at genome coordinate $x$ (the most
-fundamental operation required for an ARG data structure),
-we traverse the graph rootwards from the leaves.
-At a particular
-node $u$, if it has one parent we are at a common ancestor
-node and we follow that parent. If we are at a
-recombination node, $u$ has two parents: if
-$x$ is less than the breakpoint we follow
-the edge to the first parent, and otherwise follow the edge
-to the second parent.
-The order in which parents are listed at a recombination node is
-therefore significant, telling us
-from which parent the segments of genome to the left and right of the breakpoint
-were inherited.
-
-[TODO list the salient points that we want to contrast against
-the gARG. We don't want this to seem like a straw man]
+[TODO: what do we need to cover here?]
+
+[What is the ARG definition provided by the parsimony ARG literature?]
+
+[What is Gusfields ARG definition? Presumably it's still an eARG, but with a
+quirky dependence on SNP data?]
 

### 2023-09-14 paper.tex
Updates from Overleaf

@@
 \author[2,$\star$]{Anastasia Ignatieva}
-\author[3,$\star$]{Jere Koskela}
+\author[3,4,$\star$]{Jere Koskela}
 
 % Middle Authors
-\author[4]{Gregor Gorjanc}
-\author[5,6]{Anthony W. Wohns}
+\author[5]{Gregor Gorjanc}
+\author[6,7]{Anthony W. Wohns}
 
@@
 \affil[2]{Department of Statistics, University of Oxford, OX1 3LB, UK}
-\affil[3]{Department of Statistics, University of Warwick, CV4 7AL, UK}
-\affil[4]{The Roslin Institute and Royal (Dick) School of Veterinary Studies, University of Edinburgh, EH25 9RG, UK}
-\affil[5]{Broad Institute of MIT and Harvard, Cambridge, MA 02142, USA}
-\affil[6]{Department of Genetics, Stanford University School of Medicine, Stanford, CA 94305, USA}
+\affil[3]{School of Mathematics, Statistics and Physics, Newcastle University, NE1 7RU, UK}
+\affil[4]{Department of Statistics, University of Warwick, CV4 7AL, UK}
+\affil[5]{The Roslin Institute and Royal (Dick) School of Veterinary Studies, University of Edinburgh, EH25 9RG, UK}
+\affil[6]{Broad Institute of MIT and Harvard, Cambridge, MA 02142, USA}
+\affil[7]{Department of Genetics, Stanford University School of Medicine, Stanford, CA 94305, USA}
 
@@
 \begin{abstract}
-Recent breakthroughs have made it possible to infer genetic genealogies in
+New developments have made it possible to infer genetic genealogies in
 the presence of recombination at scale, enabling many
@@
 % principal components~\citep{mcvean2009genealogical}.
-Estimating the detailed genetic genealogy of a set of sequences
+Estimating the detailed genealogy of a set of genome sequences
 under the influence of recombination,
@@
 of the population and statistical genetics toolkit~\citep{hejase2020summary}.
-Applications using these inferred ancestries as inputs have begun to
+Applications using these ancestries as inputs have begun to
 appear~\citep{osmond2021estimating,fan2022genealogical,hejase2022deep,zhang2023biobank,
@@
 confusion about what, precisely, an ARG \emph{is}. 
-Originally, the ARG was defined as a variant 
-of the coalescent with recombination stochastic 
-process~\citep{griffiths1991two,ethier1990two,griffiths1996ancestral,griffiths1997ancestral},
-but has subsequently come to be seen as a data
-structure~\citep{minichiello2006mapping}, i.e., a means of 
-describing the \emph{outcome} of some such ancestral process
-(the distinction, however, is not clear cut and subfields use the term
-differently; see Appendix XXX).
-ARG inference was for many years synonymous [FIXME: this isn't really true,
-what about parsimony ARGs?]
-with estimating realisations
-of the coalescent with recombination (with, it must be said, limited success),
-and the distinction between process and realisation was not especially important.
-The recent introduction of two methods, \relate~\citep{speidel2019method}
-and \tsinfer~\citep{kelleher2019inferring}, that estimate structures 
-sharing many properties of this classical idea of an ARG, but at previously
-unthinkable scale, has led to a surge of interest in the area.
-Although both \cite{speidel2019method} and \cite{kelleher2019inferring}
-avoid describing the output of the methods as an ARG, the general 
-consensus has emerged that they belong to a growing class of 
-ARG inference methods~\citep[MORE][]{mathieson2020ancestry},
-and we therefore need a more general definition of an ``ARG'' than the classical
-Griffiths formulation.
-
-A general term to describe the detailed genetic genealogy of a sample 
-that encompasses the output of all these methods is indeed useful, but 
-there are substantial problems emerging from the current usage of the term 
-ARG (and its implicit concrete definition in the Griffiths formulation).
-The problem is much deeper than some pedantic discussion over terminology,
-and directly informs what we should be attempting to estimate, how we 
-evaluate inference methods, and what the limitations on the precision
-and accuracy of our inferences are.
-We currently lack a well-defined target for inference, and therefore there is 
-disagreement on how these inference should be evaluated, and 
+Originally, the ARG was based on a stochastic 
+process~\citep{griffiths1991two,ethier1990two,griffiths1996ancestral,griffiths1997ancestral} which modelled the coalescent with recombination \citep{hudson1983properties}.
+Subsequently, it has come instead to be thought of as a data
+structure~\citep{minichiello2006mapping}, i.e.\ a means of 
+describing the sequence of correlated local trees that are the \emph{outcome} of some such ancestral process.
+However, the distinction between process and structure is not clear cut and subfields use the term
+differently (see Appendix XXX).
+%% Sentence deleted: point made already
+% ARG inference was for many years synonymous [FIXME: this isn't really true,
+% what about parsimony ARGs?]
+% with estimating realisations
+% of the coalescent with recombination (with, it must be said, limited success),
+% and the distinction between process and realisation was not especially important.
+
+%% Repeat of first para
+% The recent introduction of two methods, \relate~\citep{speidel2019method}
+%and \tsinfer~\citep{kelleher2019inferring}, that estimate structures 
+%sharing many properties of this classical idea of an ARG, but at previously
+%unthinkable scale, has led to a surge of interest in the area.
+% 
+%
+%Although both \cite{speidel2019method} and \cite{kelleher2019inferring}
+%avoid describing the output of the methods as an ARG, 
+%and we therefore need a more general definition of an ``ARG'' than the classical
+%Griffiths formulation.
+
+The word ``ARG'' is now widely used to refer to genetic genealogies in general~\citep[MORE][]{mathieson2020ancestry}, including
+varied outputs from modern simulation and
+inference methods~\citep{rasmussen2014genome, palamara2016argon, haller2018tree, kelleher2019inferring, speidel2019method, baumdicker2021efficient, zhang2023biobank}.
+However, this broad usage means there is disagreement on the exact object being defined. This causes problems, for example,
+not just when evaluating inferred ARGs, but even when defining
 what the criteria for success should be.
-We also lack a shared format for interchanging ARG data between
-programs, with the assorted negative consequences which
-geneticists are only too familiar with~\citep{excoffier2006computer}.
-Finally, and perhaps most importantly, and overly narrow definition of ARGs
-leads to potentially missed opportunities in their application and usage.
+It also means there is no shared format for interchanging ARG data between
+programs, resulting in the assorted negative consequences with which
+geneticists are only too familiar~\citep{excoffier2006computer}.
+Perhaps most importantly, some definitions of ARGs (especially those linked to a specific stochastic process) are overly narrow, which
+can lead to missed opportunities in their application and usage.
+Such problems go much deeper than pedantic discussion over terminology;
+they directly inform what we should be attempting to estimate, how we 
+assess success, and the expected limitations on the precision
+and accuracy of our inferences.
 
@@
 providing a precise definition of an ARG that is general enough to encompass
-the output of all currently available inference methods. 
-The detailed genetic genealogy of a sample is defined in terms 
-of based on specific genomes
-and their intervals of genetic inheritance, which we refer to as 
+the output of all popular simulation and inference methods, yet simple enough to be encoded via a conventional mathematical graph. 
+%% Yan: I added the above as both the ARGON and the Shipilova definitions
+%% use non-conventional graph structures for their encoding (e.g. hypergraphs, weird edges, etc)
+We follow the well established principle that it is advantageous to decouple the presence of mutations from the genealogical structure itself \citep[contra][chap.~3]{gusfield2014recombinatorics}.
+We propose a definition of genetic genealogy purely in terms of specific genomes
+and their intervals of genetic inheritance, and refer to this as 
 a ``genome ARG'', or ``gARG''.
-This contrasts with the classical Griffiths approach which 
-is based on recording evolutionary events, and we therefore refer to as 
-the ``event ARG'', or ``eARG'' encoding.
+This contrasts with the classical Griffiths approach, 
+which is based on recording evolutionary events, and which we call 
+an ``event ARG'', or ``eARG''.
 We show that the gARG encoding has a range of advantages over the 
@@
 In particular, we argue that the goal of inferring fully precise 
-ARGs---a necessity under the eARG encoding---should be critically 
-examined as part of the process of fully decoupling our ARG data 
-structure from the coalescent with recombination model.
+ARGs---a necessity under the eARG encoding---should be 
+% weird phrasing for a criticism?
+critically examined
+%
+as part of the process of fully decoupling our ARG data 
+structure from any specific generative model.
 
@@
 % % \hline
-% % $(\textsf{a}, \textsf{e})$ & $(0, L]$ \\
-% % $(\textsf{b}, \textsf{d})$ & $(0, L]$ \\
-% % $(\textsf{c}, \textsf{f})$ & $(0, L]$ \\
-% % $(\textsf{d}, \textsf{e})$ & $(0, x]$ \\
-% % $(\textsf{d}, \textsf{f})$ & $(x, L]$ \\
-% % $(\textsf{e}, \textsf{g})$ & $(0, L]$ \\
-% % $(\textsf{f}, \textsf{g})$ & $(0, L]$ \\
+% % $(\textsf{a}, \textsf{e})$ & $[0, L)$ \\
+% % $(\textsf{b}, \textsf{d})$ & $[0, L)$ \\
+% % $(\textsf{c}, \textsf{f})$ & $[0, L)$ \\
+% % $(\textsf{d}, \textsf{e})$ & $[0, x)$ \\
+% % $(\textsf{d}, \textsf{f})$ & $[x, L)$ \\
+% % $(\textsf{e}, \textsf{g})$ & $[0, L)$ \\
+% % $(\textsf{f}, \textsf{g})$ & $[0, L)$ \\
 % % \end{tabular}};
@@
 % node $d$ from $e$, we would state this by associating
-% the intervals $\{(0, x], (y, L]\}$ with $(d,e)$ and
-% $\{(x, y]\}$ with $(d,f)$.
+% the intervals $\{[0, x), [y, L)\}$ with $(d,e)$ and
+% $\{[x, y)\}$ with $(d,f)$.
 % This change is straightforward, involves no loss of information
@@
 \section*{Genome ARGs}
-% There is no fundamental reason, however, that the data structure we use
-% to describe patterns of genetic inheritance must be in terms
-% of specific causal events. As demonstrated in subsequent sections,
-% there are significant advantages to using
-% a slightly different formulation, which
-% we call a ``genome ARG'' (gARG).
-In this section we provide a  concrete definition of an ARG in terms
-of specific genomes and their patterns of inheritance, which
-we refer to as a ``genome ARG``, or gARG.
-First, let us define a genome as the complete set of genetic material that a child
-inherits from one parent (i.e., haploids have one genome, diploids two, etc.).
-In this usage a ``genome'' is equivalent to the definition of a ``haplotype'' used by
-\cite{shipilina2023origin}; we avoid term haplotype because 
+
+We define a genome as the complete set of genetic material that a child
+inherits from one parent (i.e.\ haploids have one genome, diploids two, etc.).
+In this usage a ``genome'' is equivalent to the definition of a ``haplotype'' used by~\cite{shipilina2023origin}, which we avoid because 
 it already has many interpretations, and in particular carries the implication of 
-specific mutational patterns [CITES] which persist over 
+specific mutational patterns [CITES] that persist over 
 many generations. A ``genome'' in our usage corresponds to the 
-genetic material present in a particular individual or cell.
+haploid genetic material present in a particular individual or cell.
 We will also use the term ``genome`` in its 
@@
 will be clear from the context.
-We consider only patterns of genetic inheritance, with mutational 
-processes being entirely separate.
 
 A gARG is then a graph in which nodes represent
-monoploid genomes and edges represent
-genetic inheritance between an ancestor and a descendant,
-where each 
-edge\footnote{We use the standard graph theoretic sense of
-an ``edge'' here, rather than the somewhat specialised interpretation of \cite{shipilina2023origin}.}
- is annotated with the coordinates
-of the genomic intervals over which genetic inheritance occurs.
+haploid genomes, edges represent
+genetic inheritance between an ancestor and a descendant, and
+annotations specify coordinates of the genomic intervals
+over which genetic inheritance occurs.
 We can therefore view a gARG for a given set of individuals
-as a subset of their pedigree, but
-where each individual is represented by two nodes (for diploids).
-This is illustrated in Fig.~\ref{fig-arg-in-pedigree},
-which
-shows how the genomes embedded in an example (highly inbred)
-pedigree correspond to the nodes in a gARG.
-In this diploid example
-a genome is inherited from one
+as a subset of their pedigree.
+An example of this for diploids (so each individual is represented by two nodes) is illustrated in Fig.~\ref{fig-arg-in-pedigree}.
+Each genome is inherited from one
 of the individual's parents,
@@
 genomes in the current generation (\textsf{a} to \textsf{d}) with their ancestors.
-Genomes \textsf{a} and \textsf{c} are independent recombinants of
+Genomes \textsf{a} and \textsf{c} are the product of two independent recombination events between
 the paternal genomes \textsf{e}
-% Gregor: there is some overlap between a and c still (the middle region inherited
-% from f) so, a and c are not fully independent!
 and \textsf{f}, and regions of genome inherited are shown.
@@
 that results in the partial inheritance of a descendant genome 
-from an ancestor) we must ``annotate'' the edges with the genome
+from an ancestor) we ``annotate'' the edges with the genome
 coordinates over which inheritance occurs.
@@
 the recombination at position 2 means that it inherits the (half-closed) genome
-interval $(0, 2]$ from \textsf{e} and it inherits $(2, 10]$ from \textsf{f}.
-These intervals are shown attached to the corresponding graph edges.
+interval $[0, 2)$ from \textsf{e} and it inherits $[2, 10)$ from \textsf{f}; these intervals are shown attached to the corresponding graph edges.
+Alternative gARG encodings split transmitted intervals and associate each with a new node, resulting in a single genome being represented by a group of multiple nodes (or ``blocks'' \citep{palamara2016argon}), with implications for efficiency of storage.
 
@@
 % will help some readers?
-Formally, we can define a gARG as a set of nodes $N$ and edges $E$, where each
+Formally, our gARG encoding is defined by sets of nodes $N$ and edges $E$, where each
 element of $E$ is a tuple $(c, p, I)$ such that $c, p \in N$ are the child and
-parent nodes and $I$ is the set of disjoint genomic intervals $(\ell, r]$ over
+parent nodes and $I$ is the set of disjoint genomic intervals $[\ell, r)$ over
 which genome $c$ inherits from $p$. The most fundamental operation for
 an ARG data structure is the process of recovering the corresponding sequence
-of local trees along the genome. To generate the local tree at genome
-coordinate $x$, we traverse the graph rootwards from the leaves. 
-For concreteness, we can describe this tree as an oriented forest $\pi$
-\citep{kelleher2013coalescent,kelleher2014coalescent,kelleher2016efficient}.
-At each node
-$u$, we find an edge $(c, p, I) \in E$ such that $u = c$ and $x \in I$.
-We then set $\pi_c \leftarrow p$,  and move to the next node $p$.
-At the end of this process, $\pi$ fully describes the local tree 
-at position $x$.
-
-There are some interesting consequences to these definitions, exemplified
-in Figure~\ref{fig-arg-in-pedigree}. Firstly,
+of local trees along the genome. Concretely, we can describe a local tree
+at position $x$ as an oriented forest $\pi = ( \pi_1, \hdots, \pi_{|N|} )$, where the entry $\pi_j$ is the index
+of the parent of $j$ and $\pi_j = \emptyset$ if $j$ is not present in the 
+tree~\citep{kelleher2013coalescent,kelleher2014coalescent,kelleher2016efficient}.
+We use a forest to describe the local ancestry at a given position so that we allow for cases where not all leaves have found a most recent common ancestor, resulting in multiple roots.
+To construct $\pi$ at position $x$, set $\pi_i = \emptyset$ for $i \in \{1, \hdots, |N|\}$. 
+Then, traversing over all nodes of the tree,
+at each node
+$u$, find an edge $(c, p, I) \in E$ such that $u = c$ and $x \in I$, and set
+$\pi_c \leftarrow p$.
+At the end of this process, $\pi$ fully describes the local tree.
+
+There are some interesting consequences to these definitions. Firstly,
 there is
 no limitation on mating system, ploidy, or age structure. By making the
-basic unit of analysis an individual's \emph{genomes} (i.e., two per individuals
+basic unit of analysis an individual's \emph{genomes} (i.e.\ two per individuals
 for diploids), any pattern of inheritance among mixed ploidy can be
@@
 gene conversions,
-multiple recombinations
-across different chromosomes~\citep{fearnhead2003ancestral,koskela2019robust},
-and so on can be directly expressed.
+multiple simultaneous recombinations (e.g.\ Figure~\ref{fig-arg-in-pedigree}), and so on can be directly expressed.
 % Wilder: give forward ref to section on efficiency here, mention tree sequence
@@
 one or more samples find a common ancestor); for example, node \noderef{e}
-is unary in the left-most tree of Figure~\ref{fig-arg-in-pedigree},
-denoting the fact that \noderef{e} the ancestor of \noderef{a} over the interval $(0, 2]$,
-but no coalescence occurs in this interval (see section XXX for more
+is present but not coalescent in the left-most tree of Figure~\ref{fig-arg-in-pedigree} (see the ``Precision of recombination information'' section for more
 discussion). Fourth, recombination is modelled in terms of
 its outcomes in terms of inheritance intervals among parental genomes;
-thus, we will have a child node (the recombinant; e.g. node \noderef{a}
+thus we will have a child node (the recombinant; e.g. node \noderef{a}
 Figure~\ref{fig-arg-in-pedigree}) and two (or more; see section XXX)
-paternal genomes which this child has inherited segments from.
-% TODO rephrase this, it'll confuse people. Just need to reiterate the
+paternal genomes from which this child has inherited segments.
+% TODO rephrase the above, it'll confuse people. Just need to reiterate the
 % points about events, and maybe lead into the next section
@@
 recombination, and is agnostic to the details of those processes.
-
-% This paragraph is a bit weak and muddled, needs a few more passes.
+% The paragraph above is a bit weak and muddled, needs a few more passes.
 % In particular, the "Finally" bit doesn't really seem to follow
 % from the others. Is there a more direct way we can address these points ?
+
+In this section we have considered the nodes in a gARG as separate from events in a generative process.
+Nevertheless, gARGs are commonly constructed to include only those genomes associated with specific events.
+For example, \cite{mathieson2020ancestry} define an ARG as ``a subset of the pedigree... contain[ing] only those edges along which inherited segments of DNA have been transmitted, and only those nodes
+corresponding to ancestors in which there was a recombination or coalescence
+event.''. In an eARG however, events are even more explicitly embedded in the data structure.
+
 % There are some interesting and important consequences to this focus on genomes
@@
 \section*{Event ARGs}\label{eARG}
-In keeping with its original derivation in terms of a stochastic
-process (see Appendix~\ref{sec-arg-history}), the classical definition of an
-ARG data structure is
-in terms of events.
-Nodes represent
+In keeping with its original derivation in terms of a backward-time stochastic
+process (see Appendix~\ref{sec-arg-history}), the classical definition of an ARG data structure is in terms of events, and we therefore refer to it as an ``event ARG'' (eARG).
+In an eARG, nodes represent
 common ancestor or recombination events that occurred in the
@@
 In a common ancestor event the inbound lineages are merged into a
-single ancestral lineage, and in a recombination
+single ancestral lineage with one parent, whereas in a recombination
 event a lineage is split into two independent
-ancestral lineages.
-The final vital
-detail is that we associate a \emph{breakpoint} with each recombination
-event, and this information is sufficient to uniquely
-define the local genealogical trees at every position along the genome.
-
-As illustrated in Fig.~\ref{fig-arg-data-structure}, this information
-is usually concretely encoded by associating a recombination breakpoint
-(which is null for common ancestor events) and an ordered list of parents
-with each node. We refer to such a  data structure as an
-``event ARG'' (eARG).
-
-% To recover the local tree at genome coordinate $x$ (the most
-% fundamental operation required for an ARG data structure),
-% we traverse the graph rootwards from the leaves.
-% At a particular
-% node $u$, if it has one parent we are at a common ancestor
-% node and we follow that parent. If we are at a
-% recombination node, $u$ has two parents: if
-% $x$ is less than the breakpoint we follow
-% the edge to the first parent, and otherwise follow the edge
-% to the second parent.
-% The order in which parents are listed at a recombination node is
-% therefore significant, telling us
-% from which parent the segments of genome to the left and right of the breakpoint
-% were inherited.
-
-% [TODO list the salient points that we want to contrast against
-% the gARG. We don't want this to seem like a straw man]
+ancestral lineages, and a list of two parents is required.
+As illustrated in Figure~\ref{fig-arg-data-structure},
+the regions of the genome which are transmitted are usually
+specified by associating recombination nodes with a breakpoint.
+In other words, eARGs traditionally
+carry positional information on their nodes, rather than
+storing genome positions as edge annotations.
+
+To recover the local tree at position $x$,
+requires traversing the graph rootwards from the leaves.
+When we encounter a recombination node during traversal
+if $x$ is less than the breakpoint we follow
+the edge to the first parent, and otherwise follow the edge
+to the second parent. This means the order in which parents are
+listed is important. Alternatively, additional information can be incorporated to distinguish left and right parents \citep[e.g.][Chap.\ 3]{gusfield2014recombinatorics}.
+
+% Jere & Yan: we thought about going into how simulation is difficult
+% with just one recombination node due to needing an order before
+% one is available during runtime. But since that can be solved 
+% fairly trivially with two recombination nodes, we decided it was
+% more fiddly than it's worth. 
+% However, see https://github.com/tskit-dev/what-is-an-arg-paper/discussions/41#discussioncomment-2035609 for additional complexities
 
@@
 
-% \node [anchor=north west] at (0,6) {A};
+\node [anchor=north west] at (0,6) {A};
 \node [nodelabel,anchor=north west] at ($(s3) + (0,0)$) {$x$};
@@
 
-% \node [anchor=north west] at (9,6) {B};
+\node [anchor=north west] at (8,6) {B};
 \node [nodelabel,anchor=north west] at ($(10,5)$) {
@@
 
-% \node [anchor=north west] at (9,6) {C};
-% \node [nodelabel,anchor=north west] at ($(10,5)$) {
-% \begin{tabular}{ll}
+\node [anchor=north west] at (20,6) {C};
+\node [nodelabel,anchor=north west] at ($(10,5)$) {
+\begin{tabular}{ll}
 % Edges & Intervals\\
-% \hline
-% $(\textsf{a}, \textsf{e})$ & $(0, L]$ \\
-% $(\textsf{b}, \textsf{d})$ & $(0, L]$ \\
-% $(\textsf{c}, \textsf{f})$ & $(0, L]$ \\
-% $(\textsf{d}, \textsf{e})$ & $(0, x]$ \\
-% $(\textsf{d}, \textsf{f})$ & $(x, L]$ \\
-% $(\textsf{e}, \textsf{g})$ & $(0, L]$ \\
-% $(\textsf{f}, \textsf{g})$ & $(0, L]$ \\
-% \end{tabular}};
+\hline
+$(\textsf{a}, \textsf{e})$ & $[0, L)$ \\
+$(\textsf{b}, \textsf{d})$ & $[0, L)$ \\
+$(\textsf{c}, \textsf{f})$ & $[0, L)$ \\
+$(\textsf{d}, \textsf{e})$ & $[0, x)$ \\
+$(\textsf{d}, \textsf{f})$ & $[x, L)$ \\
+$(\textsf{e}, \textsf{g})$ & $[0, L)$ \\
+$(\textsf{f}, \textsf{g})$ & $[0, L)$ \\
+\end{tabular}};
 
@@
 \caption{\label{fig-arg-data-structure}
-A graphical depiction of a classical event ARG and a concrete
-encoding of the information. Node \noderef{d} is a recombination
-event as it is associated with a non-null breakpoint, and has
-two parents. Nodes \noderef{e}, \noderef{f} and \noderef{g} are
+A graphical depiction of a classical event ARG (left) and a corresponding encoding of the information (right). For uniformity, rather than defining alternative node types, we
+identify recombination nodes (e.g.\ \noderef{d}) as those with a non-null breakpoint; these also have two parents. Nodes \noderef{e}, \noderef{f} and \noderef{g} are
 common ancestor events.
@@
 
-[TODO: what do we need to cover here?]
+Storing only breakpoint positions, rather than transmitted regions,
+makes it hard to perform genetic analyses (such as extracting local trees) 
+without traversing the entire graph.
+% "For this reason"?
+Some eARG encodings, such as \citet{shipilina2023origin}, have a specialised definition of edges, which places them outside the scope of the standard graph-theoretic representation.
+%% Is the above true?? It the reason for this to allow efficient tree extraction? Are there other examples where encodings have been
+%% built on top of an eARG-like node annotation scheme (ARGON?) and made efficient somehow?
 
@@
 
-[What is Gusfields ARG definition? Presumably it's still an eARG, but with a
-quirky dependence on SNP data?]
-
-\cite{mathieson2020ancestry} have a slightly different definition of an ARG,
-which they say ``is a subset of the pedigree... contain[ing] only those edges along which
-inherited segments of DNA have been transmitted, and only those nodes
-corresponding to ancestors in which there was a recombination or coalescence
-event.'' Thus, while their definition is decoupled from the coalescent
-process it is defined by events.
+%[What is Gusfields ARG definition? Presumably it's still an eARG, but with a
+%quirky dependence on SNP data?] DONE
+
+Encoding genome positions on edges, rather than breakpoints on the nodes of an eARG, has important implications for efficiency, especially in forward-time simulations (see XXX)
 
 % \section*{Inheritance and ancestral material}\label{Ancestry_resolution}
-\section*{Prospective vs restrospective ARGs}
-ARGs are usually thought of as a restrospective structure, in which we
+\section*{Prospective vs retrospective ARGs}
+ARGs are usually thought of as a retrospective structure, in which we
 reason about the genomes that are genetic ancestors of some set of
@@
 $I_1$ and $I_2$ (genomes \textsf{a}--\textsf{d}).
-Genetic inheritance must have occured between (e.g.)
+Genetic inheritance must have occurred between (e.g.)
 $I_8$ and her daughter $I_6$, but this is not recorded because
@@
 are inherently retrospective. They define the evolutionary
-events that have occured in the genetic ancestry of a sample.
+events that have occurred in the genetic ancestry of a sample.
 % This is weak, rephrase.
@@
 
-Prospective ARGs occur in individual-based forwards-time
+Prospective ARGs arise in individual-based forwards-time
 simulations~\citep{kelleher2018efficient,haller2018tree},
@@
 
-\section*{Converting an eARG to a gARG}
+\section*{From eARGs to sample-resolved gARGs}
 Any eARG can be converted to a gARG without loss of information,
@@
 to the output gARG's edges. If the node is a common ancestor node,
-we annotate the single outbound edge with the interval $(0,L]$,
+we annotate the single outbound edge with the interval $[0,L)$,
 for a genome of length $L$.
 If the node is a recombination event with a breakpoint $x$,
-we annotate the outbound edges with the intervals $(0, x]$
-and $(x, L]$.
+we annotate the outbound edges with the intervals $[0, x)$
+and $[x, L)$.
 These inheritance annotations are clearly in one-to-one
@@
 As genome segments coalesce in a common
-ancestor the total amount ancestral material is reduced, until
-eventually, as we traverse pastwards through the graph, all
+ancestor the total amount of ancestral material is reduced, until
+eventually, as we traverse rootwards through the graph, all
 samples reach a most recent common ancestor at every location
@@
 
-It is important to note here that the \citet{wiuf1999recombination} eARG
+Note that the \citet{wiuf1999recombination} eARG
 in Fig.~\ref{fig-ancestry-resolution} is something of a straw man,
-in that it is unlikely
-that an inference or simulation program would produce a retrospective
-ARG that contains
-nodes or edges that are not ancestral to the sample
+in that typically an inference or backward-time simulation program would only produce an 
+ARG containing
+nodes or edges ancestral to the sample
 (see the discussion of the ``Big ARG'' stochastic process in Appendix XXX,
@@
 % For example, node \textsf{i} in Fig~\ref{fig-ancestry-resolution}C
-% carries sample-resolved material on the intervals $(0, 2]$ and $(4, 7]$. The edge
+% carries sample-resolved material on the intervals $[0, 2)$ and $[4, 7)$. The edge
 % joining \textsf{i} to its ancestor \textsf{k} then carries
-% material on $(0, 2]$ since although \textsf{i} directly inherits interval
-% $(0, 3]$ from \textsf{k} (Fig~\ref{fig-ancestry-resolution}B), only $(0, 2]$ is
+% material on $[0, 2)$ since although \textsf{i} directly inherits interval
+% $[0, 3)$ from \textsf{k} (Fig~\ref{fig-ancestry-resolution}B), only $[0, 2)$ is
 % ancestral to the sample. For each node, we also keep track of the number of
@@
 % interval once the lineages from all samples have coalesced. For example,
-% the edge from \textsf{k} to \textsf{n} omits the region $(0,2]$ because it has
+% the edge from \textsf{k} to \textsf{n} omits the region $[0,2)$ because it has
 % already fully coalesced (indicated by a fully black genomic region).
@@
 revolves around the presence
-of ``unary'' nodes in the local trees
-(a tree node that has exactly one child), which arise in several different ways.
-Unary nodes are not a standard feature of evolutionary trees;
-we are usually interested
+of ``unary'' regions for nodes
+(where a node has exactly one child over a particular genomic region). These arise in several different ways.
+Nodes with one child are not a standard feature of evolutionary trees;
+we are usually interested in
 nodes that have at least two children, because these
@@
 can be significant information about evolutionary processes
-provided by unary nodes in the local trees.
+provided by nodes which appear unary in one or more local trees.
 
@@
 according to the proportion of their span over which they are coalescent.
-(A) A gARG simulated from a diploid Wright-Fisher
-model. Note that the genome \noderef{k} has three children, and the genomes
-\noderef{e} and \noderef{g} simultaneously have multiple parents and multiple
-children.
+(A) A gARG simulated from a backward-time diploid Wright-Fisher
+model: genome \noderef{k} has three children, and the genomes
+\noderef{e} and \noderef{g} simultaneously have multiple parents and multiple children.
 (B) Simplified to remove all
 singly-connected graph components (e.g., diamonds such as \noderef{jlnm}).
-(C) Remove nodes that are unary in all local trees. This removes both recombinant nodes
-that have only one child (e.g.\ \noderef{n}) and nodes that have more
-than one child (i.e. are ``common ancestors'') but never represent coalescences
-in the local trees (e.g.\ \noderef{r}).
-(D) Rewrite edges to bypass any unary nodes in local trees. Note that this loses lineage
-information such as the fact that \noderef{k} is connected to \noderef{a} via \noderef{e}.
+(C) Remove nodes that never represent coalescences, i.e. are unary everywhere (e.g.\ \noderef{n} and \noderef{r}).
+(D) Rewrite edges to bypass any unary nodes in local trees.
 }
@@
 Fig.~\ref{fig-simplification} show a series of successive steps of
-simplification, starting with a complete [TODO be more concrete about
-what is stored here; ``complete'' isn't very helpful] gARG simulated under a
-backwards-in-time Wright-Fisher
-model, shown both as a graph on the left, and the corresponding sequence of
+simplification, starting with a sample-resolved but unsimplified gARG, simulated under a
+backward-time Wright-Fisher
+model. We show the graph on the left, and the corresponding sequence of
 local trees on the right.
@@
 between trees. The extent to which this information can be inferred
-and utilised is an interesting open question.
+and utilised is an interesting open question [CITE Ralph in prep.].
 
@@
 \section*{Precision of recombination information}
+\label{sec-precision}
 % The ARGs in Fig.~\ref{fig-simplification} represent different
@@
 D, respectively). When we move from A to B the local trees
-for the intervals $(44,61]$ and $(61,87]$ are merged because
+for the intervals $[44,61)$ and $[61,87)$ are merged because
 the only differences between them are their paths through
@@
 parent of \noderef{a} and \noderef{b} over the interval
-$(44, 100]$). We
+$[44, 100)$). We
 have therefore lost precision about
@@
 any indication of its timings. Furthermore,
-trees for $(44, 87]$ and $(87, 100]$ were only distinguishable
+trees for $[44, 87)$ and $[87, 100)$ were only distinguishable
 by the passage of the former tree through nodes \noderef{e}
@@
 in Fig.~\ref{fig-simplification} illustrates the point
-that is is possible to \emph{store} varying levels of precision
-about recombination in an ARG.
-It is also possible, of course, to \emph{infer} varying levels of precision about
+that is is possible to \emph{store} arbitrary levels of precision
+about recombination in a gARG.
+It is also possible to \emph{infer} various levels of precision about
 recombination.
@@
 [Some discussion].
-With these two conditions met, there is a direct correspondance
+With these two conditions met, there is a direct correspondence
 between a gARG and the trees.
@@
 term for the sequence of local trees along the genome. We can see
-that their statment is true, \emph{if} the local trees contain
+that their statement is true, \emph{if} the local trees contain
 no unary nodes. However, as we have shown, if the local trees 
@@
 nevertheless, we would like to estimate ARGs.
-\cite{zhan2023towards} inferred an ARG for 1.27 million SARS-Cov-2 whole
+\cite{zhan2023towards} inferred an ARG for 1.27 million SARS-CoV-2 whole
 genomes, and showed how it accurately captures several known properties of

### 2023-09-14 paper.tex
Updates from Overleaf

@@
 geneticists are only too familiar~\citep{excoffier2006computer}.
-Perhaps most importantly, some definitions of ARGs (especially those linked to a specific stochastic process) are overly narrow, which
+Perhaps most importantly some definitions of ARGs, especially those linked to a specific stochastic process, are overly narrow which
 can lead to missed opportunities in their application and usage.
@@
 haploid genetic material present in a particular individual or cell.
-We will also use the term ``genome`` in its 
+We will also use the term ``genome'' in its
 more common sense of ``the genome'' of a species, denoting the full set
@@
 
-The edges in a gARG specify that genetic inheritance occurred between an
-ancestor and a descendant, but the graph connectivity 
+The topology of a gARG specifies that genetic inheritance occurred between particular
+ancestors and descendants, but the graph connectivity
 does not tell us which \emph{parts} of their genomes were inherited.

### 2023-09-14 paper.tex

@@
 Originally, the ARG was based on a stochastic
-process~\citep{griffiths1991two,ethier1990two,griffiths1996ancestral,griffiths1997ancestral} which modelled the coalescent with recombination \citep{hudson1983properties}.
+process~\citep{griffiths1991two,ethier1990two,griffiths1996ancestral,griffiths1997ancestral}
+which modelled the coalescent with recombination \citep{hudson1983properties}.
 Subsequently, it has come instead to be thought of as a data
 structure~\citep{minichiello2006mapping}, i.e.\ a means of
-describing the sequence of correlated local trees that are the \emph{outcome} of some such ancestral process.
-However, the distinction between process and structure is not clear cut and subfields use the term
+describing the sequence of correlated local trees that
+are the \emph{outcome} of some such ancestral process.
+However, the distinction between process and structure is not clear cut
+and subfields use the term
 differently (see Appendix XXX).
@@
 
-The word ``ARG'' is now widely used to refer to genetic genealogies in general~\citep[MORE][]{mathieson2020ancestry}, including
+The word ``ARG'' is now widely used to refer to genetic genealogies in
+general~\citep[MORE][]{mathieson2020ancestry}, including
 varied outputs from modern simulation and
-inference methods~\citep{rasmussen2014genome, palamara2016argon, haller2018tree, kelleher2019inferring, speidel2019method, baumdicker2021efficient, zhang2023biobank}.
-However, this broad usage means there is disagreement on the exact object being defined. This causes problems, for example,
+inference methods~\citep{rasmussen2014genome, palamara2016argon, haller2018tree,
+kelleher2019inferring, speidel2019method, baumdicker2021efficient, zhang2023biobank}.
+However, this broad usage means there is disagreement on the
+exact object being defined. This causes problems, for example,
 not just when evaluating inferred ARGs, but even when defining
@@
 geneticists are only too familiar~\citep{excoffier2006computer}.
-Perhaps most importantly some definitions of ARGs, especially those linked to a specific stochastic process, are overly narrow which
+Perhaps most importantly, some definitions of ARGs
+(especially those linked to a specific stochastic process) are overly narrow, which
 can lead to missed opportunities in their application and usage.
@@
 providing a precise definition of an ARG that is general enough to encompass
-the output of all popular simulation and inference methods, yet simple enough to be encoded via a conventional mathematical graph.
+the output of all popular simulation and inference methods,
+yet simple enough to be encoded via a conventional mathematical graph.
 %% Yan: I added the above as both the ARGON and the Shipilova definitions
 %% use non-conventional graph structures for their encoding (e.g. hypergraphs, weird edges, etc)
-We follow the well established principle that it is advantageous to decouple the presence of mutations from the genealogical structure itself \citep[contra][chap.~3]{gusfield2014recombinatorics}.
+We follow the well established principle that it is advantageous
+to decouple the presence of mutations from the genealogical structure
+itself \citep[contra][chap.~3]{gusfield2014recombinatorics}.
 We propose a definition of genetic genealogy purely in terms of specific genomes
@@
 inherits from one parent (i.e.\ haploids have one genome, diploids two, etc.).
-In this usage a ``genome'' is equivalent to the definition of a ``haplotype'' used by~\cite{shipilina2023origin}, which we avoid because
+In this usage a ``genome'' is equivalent to the definition of a
+``haplotype'' used by~\cite{shipilina2023origin}, which we avoid because
 it already has many interpretations, and in particular carries the implication of
@@
 over which genetic inheritance occurs.
+% FIXME WE're missing the important distinction here that the pedigree
+% has one node per ind, whereas a gARG has two nodes.
 We can therefore view a gARG for a given set of individuals
 as a subset of their pedigree.
-An example of this for diploids (so each individual is represented by two nodes) is illustrated in Fig.~\ref{fig-arg-in-pedigree}.
+An example of this for diploids
+(so each individual is represented by two nodes)
+is illustrated in Fig.~\ref{fig-arg-in-pedigree}.
 Each genome is inherited from one
@@
 genomes in the current generation (\textsf{a} to \textsf{d}) with their ancestors.
-Genomes \textsf{a} and \textsf{c} are the product of two independent recombination events between
+Genomes \textsf{a} and \textsf{c} are the product of two independent
+recombination events between
 the paternal genomes \textsf{e}
@@
 the recombination at position 2 means that it inherits the (half-closed) genome
-interval $[0, 2)$ from \textsf{e} and it inherits $[2, 10)$ from \textsf{f}; these intervals are shown attached to the corresponding graph edges.
-Alternative gARG encodings split transmitted intervals and associate each with a new node, resulting in a single genome being represented by a group of multiple nodes (or ``blocks'' \citep{palamara2016argon}), with implications for efficiency of storage.
-
-% JK: dunno if this will make some readers switch off, but I feel it's
+interval $[0, 2)$ from \textsf{e} and it inherits $[2, 10)$ from \textsf{f};
+these intervals are shown attached to the corresponding graph edges.
+
 % important to show that this is a formal model and the concrete definition
 % will help some readers?
-Formally, our gARG encoding is defined by sets of nodes $N$ and edges $E$, where each
-element of $E$ is a tuple $(c, p, I)$ such that $c, p \in N$ are the child and
+% FIXME - the oriented forest definition is wrong, we need to be more
+% concrete about the fact that nodes are integers.
+Formally, we define a gARG as a sets of nodes $N$ and edges $E$.
+Each element of $E$ is a tuple $(c, p, I)$ such that $c, p \in N$ are the child and
 parent nodes and $I$ is the set of disjoint genomic intervals $[\ell, r)$ over
@@
 of local trees along the genome. Concretely, we can describe a local tree
-at position $x$ as an oriented forest $\pi = ( \pi_1, \hdots, \pi_{|N|} )$, where the entry $\pi_j$ is the index
+at position $x$ as an oriented forest $\pi = ( \pi_1, \hdots, \pi_{|N|} )$,
+where the entry $\pi_j$ is the index
 of the parent of $j$ and $\pi_j = \emptyset$ if $j$ is not present in the
 tree~\citep{kelleher2013coalescent,kelleher2014coalescent,kelleher2016efficient}.
-We use a forest to describe the local ancestry at a given position so that we allow for cases where not all leaves have found a most recent common ancestor, resulting in multiple roots.
-To construct $\pi$ at position $x$, set $\pi_i = \emptyset$ for $i \in \{1, \hdots, |N|\}$.
+We use a forest to describe the local ancestry at a given
+position so that we allow for cases where not all leaves have
+found a most recent common ancestor, resulting in multiple roots.
+To construct $\pi$ at position $x$, set
+$\pi_i = \emptyset$ for $i \in \{1, \hdots, |N|\}$.
+% FIXME - this is wrong
 Then, traversing over all nodes of the tree,
@@
 
-In this section we have considered the nodes in a gARG as separate from events in a generative process.
-Nevertheless, gARGs are commonly constructed to include only those genomes associated with specific events.
-For example, \cite{mathieson2020ancestry} define an ARG as ``a subset of the pedigree... contain[ing] only those edges along which inherited segments of DNA have been transmitted, and only those nodes
-corresponding to ancestors in which there was a recombination or coalescence
-event.''. In an eARG however, events are even more explicitly embedded in the data structure.
+[FIXME find somewhere natural for this. The last clause is too vague - what
+are the implications?]
+Alternative gARG encodings split transmitted intervals and associate each with
+a new node, resulting in a single genome being represented by a group of
+multiple nodes (or ``blocks'' \citep{palamara2016argon}), with implications for
+efficiency of storage.
+
+In this section we have considered the nodes in a gARG as separate from events
+in a generative process. Nevertheless, gARGs are commonly constructed to
+include only those genomes associated with specific events. For example,
+\cite{mathieson2020ancestry} define an ARG as ``a subset of the pedigree...
+contain[ing] only those edges along which inherited segments of DNA have been
+transmitted, and only those nodes corresponding to ancestors in which there was
+a recombination or coalescence event.''. In an eARG however, events are even
+more explicitly embedded in the data structure.
 
@@
 
+% FIXME this isn't worth getting into
 Storing only breakpoint positions, rather than transmitted regions,
@@
 % "For this reason"?
-Some eARG encodings, such as \citet{shipilina2023origin}, have a specialised definition of edges, which places them outside the scope of the standard graph-theoretic representation.
 %% Is the above true?? It the reason for this to allow efficient tree extraction? Are there other examples where encodings have been
@@
 
-\section*{From eARGs to sample-resolved gARGs}
+% JK: previous heading was jargony - this is clearer
+\section*{Converting an eARG to a gARG}
 Any eARG can be converted to a gARG without loss of information,
@@
 (C) The sample-resolved gARG resulting from simplifying with respect
-to the samples, \noderef{a}, \noderef{b}, and \noderef{c}. Dashed lines show edges that are
+to the samples, \noderef{a}, \noderef{b}, and \noderef{c}.
+Dashed lines show edges that are
 no longer present (in practice, nodes \noderef{g} and \noderef{q} would also be removed).
@@
 revolves around the presence
-of ``unary'' regions for nodes
-(where a node has exactly one child over a particular genomic region). These arise in several different ways.
+``unary'' nodes in the local trees
+(where a node has exactly one child over a particular genomic region).
+% "unary regions" is jargon - just dropping it in here like this doesn't
+% clarify. We can't assume that people are used to switching between
+% graphs and trees.
+% of ``unary'' regions for nodes
+% (where a node has exactly one child over a particular genomic region).
+These arise in several different ways.
 Nodes with one child are not a standard feature of evolutionary trees;
@@
 can be significant information about evolutionary processes
-provided by nodes which appear unary in one or more local trees.
+provided by nodes that are unary in one or more local trees.
 
@@
 singly-connected graph components (e.g., diamonds such as \noderef{jlnm}).
-(C) Remove nodes that never represent coalescences, i.e. are unary everywhere (e.g.\ \noderef{n} and \noderef{r}).
+(C) Remove nodes that never represent coalescences,
+i.e.\ are unary everywhere (e.g.\ \noderef{n} and \noderef{r}).
 (D) Rewrite edges to bypass any unary nodes in local trees.
@@
 
-Fig.~\ref{fig-simplification} show a series of successive steps of
-simplification, starting with a sample-resolved but unsimplified gARG, simulated under a
+Fig.~\ref{fig-simplification} shows a series of successive steps of
+simplification,
+% FIXME this doesn't explain anything - what did we store in the simulation?
+starting with a sample-resolved but unsimplified gARG, simulated under a
 backward-time Wright-Fisher
@@
 important~\citep{hudson1983testing,kelleher2016efficient}.
-Consider \noderef{e} in in Fig.~\ref{fig-simplification}A,
+Consider \noderef{e} in Fig.~\ref{fig-simplification}A,
 for example. We can see from the graph that it is a common

### 2023-09-15 paper.tex

@@
 
+\section*{SARS-CoV-2 ARGs: a case study}
+% What is SARS-CoV-2 and why is it interesting here? What's the dataset?
+% What are its basic evolutionary properties, summarised for a
+% population geneticist?
+Recombination via template switching is an important
+factor in the evolution of
+SARS-CoV-2~\citep{vaninsberghe2021recombinant,jackson2021generation,ignatieva2022ongoing}.
+With multiple deeply divergent lineages circulating globally,
+coinfection (and the potential for recombination)
+is relatively likely~\citep{bal2022detection}.
+The COVID-19 pandemic led to collection of whole-genome viral data
+at unprecented scale, with the GISAID database~\citep{shu2017gisaid}
+currently holding over 15 million sequences [FIXME]. Detection
+and characterisation of recombination is of profound interest
+for [reasons XYZ].
+
+% Summary of sc2ts paper. We inferred some big ole ARGs, with an
+% unprecedented degree of detail about recombination (outside, I guess,
+% of agri datasets). Summary of what's previously known about sc2 recomb.
+\citet{zhan2023towards} inferred two large ARGs for SARS-CoV-2 based
+on GISAID data, one containing 1.27 million sequences sampled up to
+June 30, 2021 (the ``Wide ARG'', and the second
+(more sparsely sampled) consisting of 657K sequences sampled up to June 30,
+2022 (the ``Long ARG''). Recombination is rare in both ARGs, with $<0.3\%$
+of the nodes being inferred to be recombinants (and quite likely
+a substantial number of false positives caused by poor data quality).
+Nonetheless, this provides thousands of automatically detected
+recombination events, the characterisation of which is usually a manual
+and labour intensive process. The recombinations events detected
+and fully characterised in these large ARGs fit well with previous
+which exhaustively investigate specific
+recombinants~\citep{jackson2021generation}.
+[Not sure what I'm trying to say here;
+what does a PopGen person who knows very little about SARS2 need to
+know here to given them context?]
+
+% What's the point of this case study? To emphasise that we
+% need to systematically represent topological uncertainty in ARGs.
+The pandemic scale SARS-CoV-2 dataset is an interesting case study
+for a number of reasons.
+Firstly, fundamental properties of the data underscore the
+need for topological uncertainty through polytomies
+and stacked recombination [more concise way of saying this?].
+The scale of sequencing effort at the height
+of the pandemic means that there were at times tens of thousands
+of sequences collected in a single day. This, combined with
+the slow rate of mutation, means that we will often have hundreds
+of identical sequences collected on the same day.
+This makes polytomies an unescapable reality, and a
+fully-resolved binary tree is clearly false precision in this case.
+At the other extreme, ``unary'' nodes are also a common and
+necessary property of such densely sampled data [explain and
+give numbers?]
+Even with this exquisite level of detail we are
+often unable to fully time-resolve recombination events. While
+in some cases a recombination event with two parental lineages
+can be quite precisely located in space and time (e.g. XA was in Wales
+around [when]), in others three or more quite distinct parental
+lineages indicate multiple unobserved recombinations.
+[JK: I'm just making stuff up here: I'm assuming that there's an
+example of a complex X lineages that people haven't figured the
+exact history of. Just seeing if this makes a good narrative
+before filling in the details.]
+While this
+may be seen as a limitation of the \texttt{sc2ts} method,
+concerted efforts to disentangle the origins of [some example
+of a complicated X lineage] have not succeeded in shedding
+much extra light, and truthfully representing the uncertainty
+around these events seems a more useful approach than
+ arbitrary resolution. Encoding the dataset as an eARG,
+with a specific event for each recombination breakpoint,
+would require such false precision.
+
+An interesting property of the SARS-CoV-2 dataset is that
+it highlights the uncertainty in the genomic position of
+recombination breakpoints. [Some discussion about why
+we can never know where the breakpoint really happened, and
+some follow-up to note that this is even more acute in
+standard popgen datasets where we're thinking about ancient
+recombs].
+This is a limitation of the
+gARG encoding proposed here, which requires that we specify
+exact intervals of inheritance and therefore precise
+recombination breakpoints.
+An approach in which we can systematically reason about the
+uncertainty of inheritance intervals as they propogate
+through an ARG
+would be an interesting avenue for future research.
+
 \section*{Discussion}

### 2023-09-15 paper.tex
Fixup intro

@@
 % principal components~\citep{mcvean2009genealogical}.
-Estimating the detailed genealogy of a set of genome sequences
+Estimating the genetic genealogy of a set of genome sequences
 under the influence of recombination,
@@
 of the population and statistical genetics toolkit~\citep{hejase2020summary}.
-Applications using these ancestries as inputs have begun to
+Applications using inferred ARGs as input have begun to
 appear~\citep{osmond2021estimating,fan2022genealogical,hejase2022deep,zhang2023biobank,
@@
 confusion about what, precisely, an ARG \emph{is}.
-Originally, the ARG was based on a stochastic
-process~\citep{griffiths1991two,ethier1990two,griffiths1996ancestral,griffiths1997ancestral}
-which modelled the coalescent with recombination \citep{hudson1983properties}.
-Subsequently, it has come instead to be thought of as a data
-structure~\citep{minichiello2006mapping}, i.e.\ a means of
-describing the sequence of correlated local trees that
-are the \emph{outcome} of some such ancestral process.
+Originally, ARGs were defined by Griffiths and colleagues as an alternative
+formulation of the coalescent with recombination~\citep{hudson1983properties},
+where the stochastic process of coalescence and recombination
+among ancestral lineages is formalised as a
+graph~\citep{griffiths1991two,ethier1990two,griffiths1996ancestral,griffiths1997ancestral}.
+Subsequently, ARGs have come to be thought of as a data
+structure~\citep{minichiello2006mapping}, i.e.\ describing
+the \emph{outcome} of such a random process or an inference
+of the genetic genealogy of a sample of genome sequences.
 However, the distinction between process and structure is not clear cut
@@
 geneticists are only too familiar~\citep{excoffier2006computer}.
-Perhaps most importantly, some definitions of ARGs
-(especially those linked to a specific stochastic process) are overly narrow, which
+Perhaps most importantly, the direct linkage to
+the original stochastic process definition leads to an overly
+overly narrow interpretation of ARGs, which is significantly
+out of step with modern methods, and
 can lead to missed opportunities in their application and usage.
@@
 providing a precise definition of an ARG that is general enough to encompass
-the output of all popular simulation and inference methods,
-yet simple enough to be encoded via a conventional mathematical graph.
-%% Yan: I added the above as both the ARGON and the Shipilova definitions
-%% use non-conventional graph structures for their encoding (e.g. hypergraphs, weird edges, etc)
-We follow the well established principle that it is advantageous
-to decouple the presence of mutations from the genealogical structure
-itself \citep[contra][chap.~3]{gusfield2014recombinatorics}.
-We propose a definition of genetic genealogy purely in terms of specific genomes
+the output of all popular simulation and inference methods.
+% We follow the well established principle that it is advantageous
+% to decouple the presence of mutations from the genealogical structure
+% itself \citep[contra][chap.~3]{gusfield2014recombinatorics}.
+We propose a definition of genetic genealogy in terms of specific genomes
 and their intervals of genetic inheritance, and refer to this as
@@
 This contrasts with the classical Griffiths approach,
-which is based on recording evolutionary events, and which we call
+based on recording evolutionary events, which we call
 an ``event ARG'', or ``eARG''.
 We show that the gARG encoding has a range of advantages over the
-eARG approach, being both flexible and efficient.
-In particular, we argue that the goal of inferring fully precise
-ARGs---a necessity under the eARG encoding---should be
-% weird phrasing for a criticism?
-critically examined
-%
-as part of the process of fully decoupling our ARG data
-structure from any specific generative model.
+eARG approach, being flexible, efficient and, in particular,
+allowing us to systematically encode uncertainty about the
+time ordering of recombination events.
+
+% We examineV}
+% In particular, we argue that the goal of inferring fully precise
+% ARGs---a necessity under the eARG encoding---should be
+% % weird phrasing for a criticism?
+% critically examined
+% %
+% as part of the process of fully decoupling our ARG data
+% structure from any specific generative model.
 
@@
 In keeping with its original derivation in terms of a backward-time stochastic
-process (see Appendix~\ref{sec-arg-history}), the classical definition of an ARG data structure is in terms of events, and we therefore refer to it as an ``event ARG'' (eARG).
+process (see Appendix~\ref{sec-arg-history}),
+the classical definition of an ARG data structure is in terms of evolutionary
+\emph{events}.
+
+and we therefore refer to it as an ``event ARG'' (eARG).
 In an eARG, nodes represent
@@
 $I_1$ and $I_2$ (genomes \textsf{a}--\textsf{d}).
-Genetic inheritance must have occurred between (e.g.)
-$I_8$ and her daughter $I_6$, but this is not recorded because
+Genetic transmission must have occurred between (e.g.)
+$I_8$ and her daughter $I_6$, but this is omitted because
 % this is probably not true, given we know the complete pedigree?
-we have no information about it, given these samples.
-Event ARGs, given their roots in the coalescent with recombination,
-are inherently retrospective. They define the evolutionary
-events that have occurred in the genetic ancestry of a sample.
-% This is weak, rephrase.
-An event ARG must have some set of samples, or there
-will be nothing to define events in terms of.
+it is not relevant to the genetic ancestry of the samples
 

### 2023-09-15 paper.tex
Rewrite gARG

--- 2023-09-15 paper.tex
@@
 Perhaps most importantly, the direct linkage to
-the original stochastic process definition leads to an overly
+the original stochastic process definition leads to an
 overly narrow interpretation of ARGs, which is significantly
@@
 \section*{Genome ARGs}
-
 We define a genome as the complete set of genetic material that a child
@@
 specific mutational patterns [CITES] that persist over
-many generations. A ``genome'' in our usage corresponds to the
-haploid genetic material present in a particular individual or cell.
+many generations.
+% A ``genome'' in our usage corresponds to the
+% haploid genetic material present in a particular individual or cell.
 We will also use the term ``genome'' in its
@@
 
-A gARG is then a graph in which nodes represent
-haploid genomes, edges represent
-genetic inheritance between an ancestor and a descendant, and
-annotations specify coordinates of the genomic intervals
-over which genetic inheritance occurs.
-% FIXME WE're missing the important distinction here that the pedigree
-% has one node per ind, whereas a gARG has two nodes.
-We can therefore view a gARG for a given set of individuals
-as a subset of their pedigree.
-An example of this for diploids
-(so each individual is represented by two nodes)
-is illustrated in Fig.~\ref{fig-arg-in-pedigree}.
-Each genome is inherited from one
-of the individual's parents,
-and may be the recombined product of that parent's two genomes.
-For example, individual $I_1$ in Fig.~\ref{fig-arg-in-pedigree}
-has two genomes \textsf{a} and \textsf{b},
-inherited from parents $I_3$ and $I_4$. Genome \textsf{a} is the product of
-recombining $I_3$'s two genomes \textsf{e} and \textsf{f} at position 2,
-whereas \textsf{b} was inherited directly from $I_4$'s genome \textsf{g} without
-recombination.
+A genome ARG (gARG) is a graph in which nodes represent
+haploid genomes and edges represent
+genetic inheritance between an ancestor and a descendant.
+% This has two purposes: clarify that we're not talking about Shipilina
+% et als edges, and also get "Graph theory" in here
+(Note that we are using the term ``edge'' in its standard
+graph-theoretic sense
+and not the specialised interpretation of
+\cite{shipilina2023origin}.)
+The topology of a gARG specifies that genetic inheritance
+occurred between particular
+ancestors and descendants, but the graph connectivity
+does not tell us which \emph{parts} of their genomes were inherited.
+In order to capture the effects of recombination
+(and other processes that lead to the partial genome inheritance)
+ we ``annotate'' the edges with the genome
+coordinates over which inheritance occurs.
+We can define a gARG formally as follows.
+% TODO we should change the notation for inheritance intervals here,
+% as we use I to index individuals.
+Let $N \subset \mathbb{N}$ be the set of nodes representing
+the genomes in the gARG, and $E$ be the set of edges.
+We will usually define a set of nodes $S \subseteq N$
+representing the sampled genomes
+(but see section XXX for some subtlety around this point).
+Each element of $E$ is a tuple $(c, p, I)$ such that $c, p \in N$ are the child and
+parent nodes and $I$ is the set of disjoint genomic intervals $[\ell, r)$ over
+which genome $c$ inherits from $p$. Thus, each topological connection between
+ a parent and child in the graph is ``annotated'' with set of
+inheritance intervals $I$.
 
@@
 
-The topology of a gARG specifies that genetic inheritance occurred between particular
-ancestors and descendants, but the graph connectivity
-does not tell us which \emph{parts} of their genomes were inherited.
-In order to capture the effects of recombination (and any other process
-that results in the partial inheritance of a descendant genome
-from an ancestor) we ``annotate'' the edges with the genome
-coordinates over which inheritance occurs.
-Continuing the example of genome \textsf{a} in Fig.~\ref{fig-arg-in-pedigree},
-the recombination at position 2 means that it inherits the (half-closed) genome
-interval $[0, 2)$ from \textsf{e} and it inherits $[2, 10)$ from \textsf{f};
+As illustrated in Fig.~\ref{fig-arg-in-pedigree},
+the gARG for a given set of individuals is embedded in their pedigree.
+In this diploid example, each genome is inherited from one
+of the individual's parents,
+and may be the recombined product of that parent's two genomes.
+For example, individual $I_1$ in Fig.~\ref{fig-arg-in-pedigree}
+has two genomes \noderef{a} and \noderef{b},
+inherited from parents $I_3$ and $I_4$.
+Genome \noderef{b} was inherited directly from $I_4$'s genome \noderef{g} without
+recombination, whereas
+\noderef{a} is the product of
+recombining $I_3$'s two genomes \noderef{e} and \noderef{f} at position 2.
+This recombination at means \noderef{a} inherits the (half-closed) genome
+interval $[0, 2)$ from \noderef{e} and it inherits $[2, 10)$ from \noderef{f};
 these intervals are shown attached to the corresponding graph edges.
 
-% important to show that this is a formal model and the concrete definition
-% will help some readers?
-% FIXME - the oriented forest definition is wrong, we need to be more
-% concrete about the fact that nodes are integers.
-Formally, we define a gARG as a sets of nodes $N$ and edges $E$.
-Each element of $E$ is a tuple $(c, p, I)$ such that $c, p \in N$ are the child and
-parent nodes and $I$ is the set of disjoint genomic intervals $[\ell, r)$ over
-which genome $c$ inherits from $p$. The most fundamental operation for
-an ARG data structure is the process of recovering the corresponding sequence
-of local trees along the genome. Concretely, we can describe a local tree
-at position $x$ as an oriented forest $\pi = ( \pi_1, \hdots, \pi_{|N|} )$,
-where the entry $\pi_j$ is the index
-of the parent of $j$ and $\pi_j = \emptyset$ if $j$ is not present in the
-tree~\citep{kelleher2013coalescent,kelleher2014coalescent,kelleher2016efficient}.
-We use a forest to describe the local ancestry at a given
-position so that we allow for cases where not all leaves have
-found a most recent common ancestor, resulting in multiple roots.
-To construct $\pi$ at position $x$, set
-$\pi_i = \emptyset$ for $i \in \{1, \hdots, |N|\}$.
-% FIXME - this is wrong
-Then, traversing over all nodes of the tree,
-at each node
-$u$, find an edge $(c, p, I) \in E$ such that $u = c$ and $x \in I$, and set
-$\pi_c \leftarrow p$.
-At the end of this process, $\pi$ fully describes the local tree.
-
-There are some interesting consequences to these definitions. Firstly,
-there is
-no limitation on mating system, ploidy, or age structure. By making the
-basic unit of analysis an individual's \emph{genomes} (i.e.\ two per individuals
-for diploids), any pattern of inheritance among mixed ploidy can be
-accommodated.
+% The most fundamental operation for
+% an ARG data structure is to recover the embedded local tree
+% at some position $x$ along the genome.
+The relationship between an ARG and the corresponding local trees
+is somewhat subtle (see section XXX), and it useful to have
+a concrete definition and process for reconstruction.
+We can describe a local tree as an
+oriented tree~\cite[p.\ 461]{knuth11combinatorial},
+a sequence of integers $\pi_1\pi_2\dots$ where $\pi_u$ denotes the parent of
+node $u$~\citep{kelleher2013coalescent,kelleher2014coalescent,kelleher2016efficient}.
+If $u$ is a root
+% I'm not sure if this is a helpful clarification, and whether it'll just
+% confuse people. It depends on some tricky details.
+(or not reachable from the samples)
+in the local tree at position $x$, $\pi_u = -1$.
+Technically, this structure is a ``forest'', as it allows us to have
+multiple disconnected trees, which occurs when all samples have not
+found a most recent common ancestor. Oriented trees also allow nodes
+to have  any number of children from one (see section XXX for more
+discussion on such ``unary nodes'' in the local trees), to arbitrarily
+large polytomies.
+To recover the local tree $\pi$ at position $x$  then, we begin by
+setting $\pi_u = -1$ for each $u\in N$. Then, for each sample
+node in $S$ we trace its path rootwards through the
+ARG for position $x$, and record this path in $\pi$.
+Specifically, at a given node $u$,
+we an edge $(c, p, I) \in E$ such that $u = c$ and $x \in I$, and set
+$\pi_c \leftarrow p$. The set $u \leftarrow p$, and repeat
+until either $\pi_u \neq -1$ (indicating we have traversed this section
+of the ARG already on the path from another sample) or there
+is no matching outbound edge (indicating we are at a root).
+
+There are some important consequences to these definitions.
+Firstly, there is
+no limitation on mating system, ploidy, or age structure in a gARG.
+By making the
+basic unit of analysis an individual's genomes,
+any pattern of inheritance among mixed ploidy can be accommodated.
 Secondly, because we record the intervals of inheritance
-between genomes (and not causal \emph{events}),
+between genomes,
 arbitrary patterns of genetic inheritance such as
 gene conversions,
-multiple simultaneous recombinations (e.g.\ Figure~\ref{fig-arg-in-pedigree}), and so on can be directly expressed.
+multiple simultaneous recombinations
+(e.g.\ Figure~\ref{fig-arg-in-pedigree}),
+and so on can be directly expressed.
 % Wilder: give forward ref to section on efficiency here, mention tree sequence
-Thirdly, ``local'' trees that we generate for a particular position on the
-genome are not explicitly concerned with ``coalescence'' (nodes at which
-one or more samples find a common ancestor); for example, node \noderef{e}
-is present but not coalescent in the left-most tree of Figure~\ref{fig-arg-in-pedigree} (see the ``Precision of recombination information'' section for more
-discussion). Fourth, recombination is modelled in terms of
+Thirdly, local trees that we generate for a particular position on the
+genome can have nodes of any arity, including ``unary''
+nodes with a single child (see Section XXX).
+% are not explicitly concerned with ``coalescence'' (nodes at which
+% one or more samples find a common ancestor); for example, node \noderef{e}
+% is present but not coalescent in the left-most tree of
+% Figure~\ref{fig-arg-in-pedigree}
+% (see section XXX for more discussion).
+Fourth, recombination is modelled in terms of
 its outcomes in terms of inheritance intervals among parental genomes;
@@
 
-[FIXME find somewhere natural for this. The last clause is too vague - what
-are the implications?]
-Alternative gARG encodings split transmitted intervals and associate each with
-a new node, resulting in a single genome being represented by a group of
-multiple nodes (or ``blocks'' \citep{palamara2016argon}), with implications for
-efficiency of storage.
-
-In this section we have considered the nodes in a gARG as separate from events
-in a generative process. Nevertheless, gARGs are commonly constructed to
-include only those genomes associated with specific events. For example,
-\cite{mathieson2020ancestry} define an ARG as ``a subset of the pedigree...
-contain[ing] only those edges along which inherited segments of DNA have been
-transmitted, and only those nodes corresponding to ancestors in which there was
-a recombination or coalescence event.''. In an eARG however, events are even
-more explicitly embedded in the data structure.
+% In this section we have considered the nodes in a gARG as separate from events
+% in a generative process. Nevertheless, gARGs are commonly constructed to
+% include only those genomes associated with specific events. For example,
+% \cite{mathieson2020ancestry} define an ARG as ``a subset of the pedigree...
+% contain[ing] only those edges along which inherited segments of DNA have been
+% transmitted, and only those nodes corresponding to ancestors in which there was
+% a recombination or coalescence event.''. In an eARG however, events are even
+% more explicitly embedded in the data structure.
 
@@
 (equivalent to the ``bricked tree sequence'' of
-\cite{nowbandegani2023extremely}), where we consider the unique sets of
+\cite{nowbandegani2023extremely},
+% TODO look this up. I expect it's basically the same??
+and the block-based ARG data structure of
+\citep{palamara2016argon}),
+here we consider the unique sets of
 samples that coalesce at a node in the ARG over a particular
@@
 on ARG inference.
+
+% [FIXME find somewhere natural for this. The last clause is too vague - what
+% are the implications?]
+% Alternative gARG encodings split transmitted intervals and associate each with
+% a new node, resulting in a single genome being represented by a group of
+% multiple nodes (or ``blocks'' \citep{palamara2016argon}), with implications for
+% efficiency of storage.
 

### 2023-09-15 paper.tex
Add sequence-of-trees sentence
Start on eARG section

@@
 goal in genetics.
+% TODO say this better.
+From a high-level, ARGs describe the sequence of correlated genealogical
+trees along the genome, induced by the different paths of
+genetic inheritance caused by recombination.
 Recent breakthroughs
@@
 % (C) The edge-annotated genome ARG in (B) fully defines the
-(C) The corresponding local (``marginal'') trees.
+(C) The corresponding local trees.
 }
@@
 % at some position $x$ along the genome.
-The relationship between an ARG and the corresponding local trees
+The relationship between an ARG and the corresponding local
+% TODO do we mention "marginal" anywhere? Might not be worth the
+% effort, surely "local" is clear?
 is somewhat subtle (see section XXX), and it useful to have
@@
 \section*{Event ARGs}\label{eARG}
-In keeping with its original derivation in terms of a backward-time stochastic
-process (see Appendix~\ref{sec-arg-history}),
-the classical definition of an ARG data structure is in terms of evolutionary
-\emph{events}.
-
+Genome ARGs model the inheritance between specific genomes,
+annotating the edges of the graph with the intervals
+of genetic inheritance. It is primarily concerned with
+describing relationships and the details of inheritance
+between pairs of genomes.
+In contrast, the classical ARG data structure is defined
+by evolutionary \emph{events},
 and we therefore refer to it as an ``event ARG'' (eARG).
-In an eARG, nodes represent
-common ancestor or recombination events that occurred in the
-history of some set of sample genomes, and edges represent ancestral
-lineages~\citep{griffiths1996ancestral}.
+In an eARG we usually have two different types
+of node,
+representing common ancestor and recombination events that occurred in the
+history of some set of sample genomes
+and edges represent ancestral lineages~\citep{griffiths1996ancestral}.
 In a common ancestor event the inbound lineages are merged into a
@@
 event a lineage is split into two independent
-ancestral lineages, and a list of two parents is required.
-As illustrated in Figure~\ref{fig-arg-data-structure},
+ancestral lineages with an associated breakpoint.
+Figure~\ref{fig-arg-data-structure} shows an example of a classical
+eARG with three samples (\noderef{a}, \noderef{b}, and \noderef{c})
+and a single recombination event, along with an example concrete
+encoding of the data.  There is a single recombination event
+at node \noderef{d} with a breakpoint at position $x$. We
+assume that \noderef{d} inherits genetic material to the
+left of $x$ from \noderef{e} and to the right of $x$ from \noderef{f}.
+We use the standard convention here
+that the order of parents
+is significant, but additional information can be incorporated to
+distinguish left and right parents
+\citep[e.g.][Chap.\ 3]{gusfield2014recombinatorics}.
+
+Beyone the interpretation of nodes as representing events or
+genomes (which may reasonably be seen as somewhat philosophical),
+the essential difference between a gARG and a classical eARG
+[JK WIP from here:]
+is that eARGs define the
+
+carry positional information on their nodes, rather than
+storing genome positions as edge annotations.
+
 the regions of the genome which are transmitted are usually
 specified by associating recombination nodes with a breakpoint.
-In other words, eARGs traditionally
+In other words, eARGs
 carry positional information on their nodes, rather than
@@
 to the second parent. This means the order in which parents are
-listed is important. Alternatively, additional information can be incorporated to distinguish left and right parents \citep[e.g.][Chap.\ 3]{gusfield2014recombinatorics}.
+listed is important.
 
@@
 % However, see https://github.com/tskit-dev/what-is-an-arg-paper/discussions/41#discussioncomment-2035609 for additional complexities
-
-% Griffiths and colleagues~\citep{griffiths1997ancestral} originally defined
-% the ARG as analog of the Kingman coalescent~\citep{kingman1982coalescent},
-% in which the ancestry of a sample of $n$ genomes under idealised
-% conditions is modelled. One of the key assumptions of this model is that
-% the sample size $n$ is much less than the population size $N$, which means
-% that the probability of two multiple concurrent events (i.e., coalescence
-% and recombination events) is negligible.
 
@@
 
-\node [anchor=north west] at (20,6) {C};
-\node [nodelabel,anchor=north west] at ($(10,5)$) {
-\begin{tabular}{ll}
-% Edges & Intervals\\
-\hline
-$(\textsf{a}, \textsf{e})$ & $[0, L)$ \\
-$(\textsf{b}, \textsf{d})$ & $[0, L)$ \\
-$(\textsf{c}, \textsf{f})$ & $[0, L)$ \\
-$(\textsf{d}, \textsf{e})$ & $[0, x)$ \\
-$(\textsf{d}, \textsf{f})$ & $[x, L)$ \\
-$(\textsf{e}, \textsf{g})$ & $[0, L)$ \\
-$(\textsf{f}, \textsf{g})$ & $[0, L)$ \\
-\end{tabular}};
+% \node [anchor=north west] at (20,6) {C};
+% \node [nodelabel,anchor=north west] at ($(10,5)$) {
+% \begin{tabular}{ll}
+% % Edges & Intervals\\
+% \hline
+% $(\textsf{a}, \textsf{e})$ & $[0, L)$ \\
+% $(\textsf{b}, \textsf{d})$ & $[0, L)$ \\
+% $(\textsf{c}, \textsf{f})$ & $[0, L)$ \\
+% $(\textsf{d}, \textsf{e})$ & $[0, x)$ \\
+% $(\textsf{d}, \textsf{f})$ & $[x, L)$ \\
+% $(\textsf{e}, \textsf{g})$ & $[0, L)$ \\
+% $(\textsf{f}, \textsf{g})$ & $[0, L)$ \\
+% \end{tabular}};
 
@@
 \caption{\label{fig-arg-data-structure}
-A graphical depiction of a classical event ARG (left) and a corresponding encoding of the information (right). For uniformity, rather than defining alternative node types, we
-identify recombination nodes (e.g.\ \noderef{d}) as those with a non-null breakpoint; these also have two parents. Nodes \noderef{e}, \noderef{f} and \noderef{g} are
-common ancestor events.
+A graphical depiction of a classical event ARG (left) and a corresponding
+encoding of the information (right).
+Recombination nodes (e.g.\ \noderef{d}) have a non-null breakpoint
+and two parents.
+Nodes \noderef{e}, \noderef{f} and \noderef{g} are common ancestor events.
 }
@@
 between those trees.
-Sequentially generating the marginal trees along the genome
+Sequentially generating the local trees along the genome
 is also fundamental, and is necessary whenever we need to

### 2023-09-18 paper.tex
Front-load ARG definitions

@@
 Each element of $E$ is a tuple $(c, p, I)$ such that $c, p \in N$ are the child and
-parent nodes and $I$ is the set of disjoint genomic intervals $[\ell, r)$ over
+parent nodes and $I$ is the set of disjoint genomic intervals
+%$[\ell, r)$
+over
 which genome $c$ inherits from $p$. Thus, each topological connection between
@@
 % at some position $x$ along the genome.
-The relationship between an ARG and the corresponding local
+The relationship between an ARG and the corresponding local trees
 % TODO do we mention "marginal" anywhere? Might not be worth the
 % effort, surely "local" is clear?
-is somewhat subtle (see section XXX), and it useful to have
-a concrete definition and process for reconstruction.
+is somewhat subtle (see section XXX), and it is useful to have
+a concrete definition and explicit process for reconstruction.
 We can describe a local tree as an
@@
 Specifically, at a given node $u$,
-we an edge $(c, p, I) \in E$ such that $u = c$ and $x \in I$, and set
-$\pi_c \leftarrow p$. The set $u \leftarrow p$, and repeat
+we an find edge $(c, p, I) \in E$ such that $u = c$ and $x \in I$, and set
+$\pi_c \leftarrow p$. We then set $u \leftarrow p$, and repeat
 until either $\pi_u \neq -1$ (indicating we have traversed this section
@@
 and we therefore refer to it as an ``event ARG'' (eARG).
+The difference between gARGs and eARGs is subtle and the distinctions
+at times rather porous. In essense, our main goal of introducing the
+term ``Event ARG'' is to contrast the flexibility of the genome ARG
+view with current practice. It is therefore useful to consider
+the definition of an ARG from recent work.
+
+% Todo this is rough. Ideally we want a bit of commentary running through
+% the quotations.
+\cite{mathieson2020ancestry} define an ARG as ``a subset of the pedigree...
+contain[ing] only those edges along which inherited segments of DNA have been
+transmitted, and only those nodes corresponding to ancestors in which there was
+a recombination or coalescence event.''
+\citet{shipilina2023origin} emphasise the reality  ``The ARG describes the complete
+ancestry of a sample of genomes through a series of real coalescence and
+recombination events''. In contrast,
+\citet{brandt2021evaluation} tie the definition of an ARG explicitly
+to a generative process:
+``The full ancestral recombination graph (ARG) is a structure that encodes all
+coalescence and recombination events resulting from the stochastic process of
+the coalescent with recombination.''
+In contrast ~\citet{gusfield2014recombinatorics} defines an ARG as [FIXME]
+
+The common features of these definitions is that they are expressed
+in terms of evolutionary events, emphasise the exhaustiveness
+of the record (full, complete), and they do not specify the mechanism
+by which partial genetic inheritance occurs. Just as the graph topology
+in a gARG defines the paths along which genetic inheritance occurs, some
+additional information is required to define the actual regions of genome
+inherited.
+
+% % The full ancestral recombination graph (ARG) is a structure that encodes all
+% % coalescence and recombination events resulting from the stochastic process of
+% % the coalescent with recombination.
+
+% historial trend discussed in the XXX section).
+% \citet{rasmussen2022espalier} have a less restrictive definition, in
+% which a ``full'' ARG is one that contains complete information about
+% recombination events, without necessarily being tied to a particular
+% stochastic process.
+% While [something positive], it is important to be clear on
+% what the overall goal of such precise inferences are,
+% what the limitations on when they can be made are,
+% and their downstream utility in applications.
+
 In an eARG we usually have two different types

### 2023-09-18 paper.tex
Added a few clarifying notes

@@
 The relationship between an ARG and the corresponding local trees
-% TODO do we mention "marginal" anywhere? Might not be worth the
-% effort, surely "local" is clear?
 is somewhat subtle (see section XXX), and it is useful to have
@@
 
+% Papers that do not discuss how positions work:
+% - Mathieson and Scally: no discussion of how it works concretely (but quite
+%   close to gARG, in some ways)
+%
+
 The common features of these definitions is that they are expressed
@@
 
-% % The full ancestral recombination graph (ARG) is a structure that encodes all
-% % coalescence and recombination events resulting from the stochastic process of
-% % the coalescent with recombination.
-
-% historial trend discussed in the XXX section).
-% \citet{rasmussen2022espalier} have a less restrictive definition, in
-% which a ``full'' ARG is one that contains complete information about
-% recombination events, without necessarily being tied to a particular
-% stochastic process.
-% While [something positive], it is important to be clear on
-% what the overall goal of such precise inferences are,
-% what the limitations on when they can be made are,
-% and their downstream utility in applications.
-
-In an eARG we usually have two different types
+In the classical view of an eARG we usually have two different types
 of node,
@@
 provided by nodes that are unary in one or more local trees.
+[FIXME this is clumsy; should mention somewhere, though.]
+\citep{mathieson2020ancestry} show examples in which the ARGs 
+contain unary nodes, and state that paths are ``implicitly passing through points representing specific
+individuals.''
+% Full quote from fig caption:
+% Genetic ancestry in the form of the ARG for a single individual. Combining
+% genetic ancestry from different positions leads to a graph, incorporating all
+% realized genetic ancestry paths, implicitly passing through points representing
+% specific individuals.
 
@@
 some specific properties of the ARG and the trees are required.
+Note that many authors use the term ``marginal''
+trees~\citep[e.g.][]{minichiello2006mapping} [more]
+which we avoid due the potential confusion with statistical
+marginal distrubtions [FIXME Gregor, what's the right wording here?]
 

### 2023-09-19 paper.tex
Upadte

@@
 In contrast ~\citet{gusfield2014recombinatorics} defines an ARG as [FIXME]
-
-% Papers that do not discuss how positions work:
-% - Mathieson and Scally: no discussion of how it works concretely (but quite
-%   close to gARG, in some ways)
-%
-
 The common features of these definitions is that they are expressed
-in terms of evolutionary events, emphasise the exhaustiveness
-of the record (full, complete), and they do not specify the mechanism
-by which partial genetic inheritance occurs. Just as the graph topology
-in a gARG defines the paths along which genetic inheritance occurs, some
-additional information is required to define the actual regions of genome
-inherited.
-
-In the classical view of an eARG we usually have two different types
-of node,
-representing common ancestor and recombination events that occurred in the
-history of some set of sample genomes
-and edges represent ancestral lineages~\citep{griffiths1996ancestral}.
-In a common ancestor event the inbound lineages are merged into a
-single ancestral lineage with one parent, whereas in a recombination
-event a lineage is split into two independent
-ancestral lineages with an associated breakpoint.
+in terms of evolutionary events, and emphasise the exhaustiveness
+of the record (full, complete). We will examine the consequences of both
+of these properties in different contexts throughout the remainder of
+this article.
+
+Another common feature of these definitions is that they do not explicitly
+discuss the mechanism by which genealogical trees vary along the genome.
+The topology of an eARG defines the common ancestor and
+recombination events that occured,
+but the graph structure alone is not sufficient to define the local trees,
+or equivalently, how ``ancestral material'' (Section XXX) flows along
+the edges of the ARG.
+% % Papers that do not discuss how positions work:
+% % - Mathieson and Scally: no discussion of how it works concretely (but quite
+% %   close to gARG, in some ways)
+% %
+% The common features of these definitions is that they are expressed
+% in terms of evolutionary events, emphasise the exhaustiveness
+% of the record (full, complete), and they do not specify the mechanism
+% by which partial genetic inheritance occurs. Just as the graph topology
+% in a gARG defines the paths along which genetic inheritance occurs, some
+% additional information is required to define the actual regions of genome
+% inherited.
+Classically, this is done by having two different types
+of node:
+common ancestor nodes in which the inbound lineages are merged into a
+single ancestral lineage with one parent, and recombination
+nodes in which a single lineage is split into two independent
+ancestral lineages.
+Recombination nodes are then
+annotated with
+the corresponding crossover breakpoint~\citep{griffiths1996ancestral}.
+% In a common ancestor event the inbound lineages are merged into a
+% single ancestral lineage with one parent, whereas in a recombination
+% event a lineage is split into two independent
+% ancestral lineages with an associated breakpoint.
 Figure~\ref{fig-arg-data-structure} shows an example of a classical
@@
 left of $x$ from \noderef{e} and to the right of $x$ from \noderef{f}.
-We use the standard convention here
+We use a common convention here
 that the order of parents
-is significant, but additional information can be incorporated to
+is significant, but additional information is sometimes incorporated to
 distinguish left and right parents
-\citep[e.g.][Chap.\ 3]{gusfield2014recombinatorics}.
+\citep[e.g.][]{gusfield2014recombinatorics,ignatieva2021kwarg}.
 
@@
 \caption{\label{fig-arg-data-structure}
-A graphical depiction of a classical event ARG (left) and a corresponding
+A classical event ARG (left) and a corresponding
 encoding of the information (right).

### 2023-09-20 paper.tex
Attempt to smooth the narrative

@@
 \section*{Event ARGs}\label{eARG}
-Genome ARGs model the inheritance between specific genomes,
-annotating the edges of the graph with the intervals
-of genetic inheritance. It is primarily concerned with
-describing relationships and the details of inheritance
-between pairs of genomes.
-In contrast, the classical ARG data structure is defined
-by evolutionary \emph{events},
-and we therefore refer to it as an ``event ARG'' (eARG).
-The difference between gARGs and eARGs is subtle and the distinctions
-at times rather porous. In essense, our main goal of introducing the
-term ``Event ARG'' is to contrast the flexibility of the genome ARG
-view with current practice. It is therefore useful to consider
-the definition of an ARG from recent work.
-
-% Todo this is rough. Ideally we want a bit of commentary running through
-% the quotations.
-\cite{mathieson2020ancestry} define an ARG as ``a subset of the pedigree...
+Genome ARGs are concerned with
+describing the details of genetic inheritance
+between specific pairs of genomes.
+In contrast, ARGs are classically seem as being defined
+by the evolutionary \emph{events} in the history of a sample.
+The difference between gARGs and this prevalent view of
+what defines an ARGs is subtle and the distinctions sometimes porous.
+It is therefore useful to first survey some representative
+definitions of
+an ARG from the literature, to allow us to capture the common features
+of these definitions
+(see Appendix XXX for a full review).
+
+\citet{brandt2021evaluation} follow the original Griffiths definition of
+an ARG by coupling it directly to a generative process, stating
+that the ``full ancestral recombination graph (ARG) is a structure that encodes all
+coalescence and recombination events resulting from the stochastic process of
+the coalescent with recombination.''
+\citet{shipilina2023origin} emphasise the true existance of the
+corresponding events (rather than the generative process) by saying that
+``[t]he ARG describes the complete
+ancestry of a sample of genomes through a series of real coalescence and
+recombination events.''
+Following this trend to decouple the concept of an ARG from
+generative processes~\citep{minichiello2006mapping},
+\cite{mathieson2020ancestry}
+define an ARG as ``a subset of the pedigree...
 contain[ing] only those edges along which inherited segments of DNA have been
@@
 a recombination or coalescence event.''
-\citet{shipilina2023origin} emphasise the reality  ``The ARG describes the complete
-ancestry of a sample of genomes through a series of real coalescence and
-recombination events''. In contrast,
-\citet{brandt2021evaluation} tie the definition of an ARG explicitly
-to a generative process:
-``The full ancestral recombination graph (ARG) is a structure that encodes all
-coalescence and recombination events resulting from the stochastic process of
-the coalescent with recombination.''
-In contrast ~\citet{gusfield2014recombinatorics} defines an ARG as [FIXME]
-The common features of these definitions is that they are expressed
-in terms of evolutionary events, and emphasise the exhaustiveness
-of the record (full, complete). We will examine the consequences of both
-of these properties in different contexts throughout the remainder of
-this article.
-
+In contrast ~\citet{gusfield2014recombinatorics} defines an ARG as [TODO]
+
+% Papers that do not discuss how positions work:
+% - Mathieson and Scally: no discussion of how it works concretely (but quite
+%   close to gARG, in some ways)
+% TODO list out some of these as examples of where people don't bother
+% getting into this important detail?
+The common feature of these definitions is that they are expressed
+in terms of evolutionary \emph{events}, and so for the purposes of
+discussion we can label
+this prevailing interpretation as an ``Event ARG'', or eARG.
+(Many definitions also emphasise
+the necessity of exhaustively capturing \emph{all} events, which we
+examine in later sections.)
 Another common feature of these definitions is that they do not explicitly
@@
 the edges of the ARG.
-% % Papers that do not discuss how positions work:
-% % - Mathieson and Scally: no discussion of how it works concretely (but quite
-% %   close to gARG, in some ways)
-% %
-% The common features of these definitions is that they are expressed
-% in terms of evolutionary events, emphasise the exhaustiveness
-% of the record (full, complete), and they do not specify the mechanism
-% by which partial genetic inheritance occurs. Just as the graph topology
-% in a gARG defines the paths along which genetic inheritance occurs, some
-% additional information is required to define the actual regions of genome
-% inherited.
-Classically, this is done by having two different types
-of node:
+This is assumed to follow the mechanism described by
+Griffiths and colleagues in which we have
+two different ``types'' of node in the graph:
 common ancestor nodes in which the inbound lineages are merged into a
@@
 the corresponding crossover breakpoint~\citep{griffiths1996ancestral}.
-% In a common ancestor event the inbound lineages are merged into a
-% single ancestral lineage with one parent, whereas in a recombination
-% event a lineage is split into two independent
-% ancestral lineages with an associated breakpoint.
 Figure~\ref{fig-arg-data-structure} shows an example of a classical
@@
 that the order of parents
-is significant, but additional information is sometimes incorporated to
+% TODO list papers where people use ordering? ARGweaver I think?
+is significant,
+but additional information is sometimes incorporated to
 distinguish left and right parents

### 2023-09-20 paper.tex
Stab at getting eARG section finished

@@
 among ancestral lineages is formalised as a
+% Note: please don't add "mathematical graph". If a reader doesn't know what a
+% graph is they're not going to understand this paper.
 graph~\citep{griffiths1991two,ethier1990two,griffiths1996ancestral,griffiths1997ancestral}.
@@
 between specific pairs of genomes.
-In contrast, ARGs are classically seem as being defined
+In contrast, ARGs are classically seem defined
 by the evolutionary \emph{events} in the history of a sample.
 The difference between gARGs and this prevalent view of
-what defines an ARGs is subtle and the distinctions sometimes porous.
+what defines an ARG is subtle and the distinctions sometimes porous.
 It is therefore useful to first survey some representative
@@
 left of $x$ from \noderef{e} and to the right of $x$ from \noderef{f}.
-We use a common convention here
-that the order of parents
-% TODO list papers where people use ordering? ARGweaver I think?
-is significant,
-but additional information is sometimes incorporated to
+As shown in Figure~\ref{fig-arg-data-structure}B,
+the local trees are embedded in the graph, and can be reconstructed
+in essentially the same manner as outlined the previous section for
+gARGs. The only difference is that decisions about which outbound
+edge to follow during the rootward traversal are taken at recombination
+nodes, following
+either the common convention that the order of parents
+is significant~\citep[e.g.][]{griffiths1991two},
+or by incorporating additional information to
 distinguish left and right parents
-\citep[e.g.][]{gusfield2014recombinatorics,ignatieva2021kwarg}.
-
-Beyone the interpretation of nodes as representing events or
-genomes (which may reasonably be seen as somewhat philosophical),
-the essential difference between a gARG and a classical eARG
-[JK WIP from here:]
-is that eARGs define the
-
-carry positional information on their nodes, rather than
-storing genome positions as edge annotations.
-
-the regions of the genome which are transmitted are usually
-specified by associating recombination nodes with a breakpoint.
-In other words, eARGs
-carry positional information on their nodes, rather than
-storing genome positions as edge annotations.
-
-To recover the local tree at position $x$,
-requires traversing the graph rootwards from the leaves.
-When we encounter a recombination node during traversal
-if $x$ is less than the breakpoint we follow
-the edge to the first parent, and otherwise follow the edge
-to the second parent. This means the order in which parents are
-listed is important.
-
-% Jere & Yan: we thought about going into how simulation is difficult
-% with just one recombination node due to needing an order before
-% one is available during runtime. But since that can be solved
-% fairly trivially with two recombination nodes, we decided it was
-% more fiddly than it's worth.
-% However, see https://github.com/tskit-dev/what-is-an-arg-paper/discussions/41#discussioncomment-2035609 for additional complexities
+into the
+encoding~\citep[e.g.][]{gusfield2014recombinatorics,ignatieva2021kwarg}.
 
@@
 
-\node [anchor=north west] at (8,6) {B};
+\node [anchor=north west] at (8,6) {C};
 \node [nodelabel,anchor=north west] at ($(10,5)$) {
@@
 
-% \node [anchor=north west] at (20,6) {C};
-% \node [nodelabel,anchor=north west] at ($(10,5)$) {
-% \begin{tabular}{ll}
-% % Edges & Intervals\\
-% \hline
-% $(\textsf{a}, \textsf{e})$ & $[0, L)$ \\
-% $(\textsf{b}, \textsf{d})$ & $[0, L)$ \\
-% $(\textsf{c}, \textsf{f})$ & $[0, L)$ \\
-% $(\textsf{d}, \textsf{e})$ & $[0, x)$ \\
-% $(\textsf{d}, \textsf{f})$ & $[x, L)$ \\
-% $(\textsf{e}, \textsf{g})$ & $[0, L)$ \\
-% $(\textsf{f}, \textsf{g})$ & $[0, L)$ \\
-% \end{tabular}};
-
 \end{tikzpicture}
 \caption{\label{fig-arg-data-structure}
-A classical event ARG (left) and a corresponding
-encoding of the information (right).
-Recombination nodes (e.g.\ \noderef{d}) have a non-null breakpoint
+A classical event ARG. (A) Standard graph depiction with
+breakpoint $x$ associated with the recombination node \noderef{d}.
+Nodes \noderef{e}, \noderef{f} and \noderef{g} are common ancestor events.
+(B) Corresponding local trees to the left and right of breakpoint $x$.
+(C) A concrete encoding of the information (right).
+Recombination nodes have a non-null breakpoint
 and two parents.
-Nodes \noderef{e}, \noderef{f} and \noderef{g} are common ancestor events.
 }
@@
 
-% FIXME this isn't worth getting into
-Storing only breakpoint positions, rather than transmitted regions,
-makes it hard to perform genetic analyses (such as extracting local trees)
-without traversing the entire graph.
-% "For this reason"?
-%% Is the above true?? It the reason for this to allow efficient tree extraction? Are there other examples where encodings have been
-%% built on top of an eARG-like node annotation scheme (ARGON?) and made efficient somehow?
-
-[What is the ARG definition provided by the parsimony ARG literature?]
-
-%[What is Gusfields ARG definition? Presumably it's still an eARG, but with a
-%quirky dependence on SNP data?] DONE
-
-Encoding genome positions on edges, rather than breakpoints on the nodes of an eARG, has important implications for efficiency, especially in forward-time simulations (see XXX)
-
-% \section*{Inheritance and ancestral material}\label{Ancestry_resolution}
+Whether we regard nodes as events or genomes is something of a
+philosophical question, and in most senses the two are equivalent
+(but see the discussion on ``unary nodes'' in section XXX).
+From a concrete data-structure encoding perspective, the only real
+difference between an eARG and a gARG is that we store recombination
+\emph{breakpoints} at nodes of out-degree two (recombination nodes)
+in an eARG and we store \emph{inheritance intervals} on the edges
+of a gARG. A key advantage of associating inheritance intervals with
+edges in a gARG is that it allows us to describe evolutionary events other
+than simple crossovers, such as multiple recombinations and
+gene-conversions. By explicitly stating the exact regions of genome
+that are transmitted along an edge, we generalise the representation
+to encompass any process that leads to the differential inheritance of
+genetic material along the genome.
+
+While the additional generality provided by the gARG representation
+is useful, it is in reality a small change that does not
+lead to significantly different time or space complexity in computational
+tasks. The vitally important and fundamental distinction between
+a gARG and the classical eARG perspective is that gARGs do not
+\emph{require} us to posit explicit recombination events in order
+to describe a dataset, but
+instead provide a means of encoding the effects of
+multiple events, and to systematically describe the uncertainty around them
+(see Section XXX).
+As we see in the following sections, freeing ourselves from the requirement
+of recording specific events and
+the additional generality of the gARG encoding
+has many advantages. It also raises questions about what our goals
+for inferring ARGs should be, and how we evaluate the success
+of such inferences.
+
 \section*{Prospective vs retrospective ARGs}
-ARGs are usually thought of as a retrospective structure, in which we
-reason about the genomes that are genetic ancestors of some set of
-sample genomes. Figure~\ref{fig-arg-in-pedigree}
+Event ARGs, as defined in the previous section as a way of encompassing
+the range of definitions provided in the literature, are
+inherently \emph{retrospective}.
+Event ARGs are defined in terms
+of the events that occur in the history of some set of sampled
+sequences.
+For example, Figure~\ref{fig-arg-in-pedigree}
 is retrospective because it only includes information
@@
 Genetic transmission must have occurred between (e.g.)
-$I_8$ and her daughter $I_6$, but this is omitted because
+$I_8$ and her daughter $I_6$, but this is not recorded because
 % this is probably not true, given we know the complete pedigree?
-it is not relevant to the genetic ancestry of the samples
-
-The gARG representation, on the other hand,
+it is not relevant to the genetic ancestry of the samples.
+
+Genome ARGs, on the hand,
 can be either retrospective or \emph{prospective}.
@@
 Note that many authors use the term ``marginal''
-trees~\citep[e.g.][]{minichiello2006mapping} [more]
+trees~\citep[e.g.][]{minichiello2006mapping,griffiths1996ancestral} [more]
 which we avoid due the potential confusion with statistical

### 2023-09-20 paper.tex
Trip intro commented out stuff

@@
 \section*{Introduction}
-% Inferring the genetic relationships between sampled genomes in the form of a
-% phylogenetic tree is a necessary prerequisite for many analyses of
-% species~\citep{rannala2003genetics}, non-recombining sections of
-% DNA~\citep{cann1987mitochondrial,underhill2001annalsofhumangenetics}
-% and viruses~\citep{grenfell2004science}.
-% A phylogenetic tree captures information about the genealogy
-% of the sample in an elegant and concise
-% way, by postulating a set of ancestors common to the samples
-% together with paths of inheritance between them. A rich literature exists
-% on analysing the mathematical properties of these
-% trees~\citep{steel2016phylogeny}, and numerous
-% methods exist to infer them~\citep{felsenstein2004inferring}.
-% However, the genetic genealogy of a recombining organism
-% cannot be captured by a single tree: instead,
-% the inheritance of each section of the genome may be described
-% by a (slightly) different local tree~\citep{hudson1983properties}.
-% Inferring complex
-% genealogies of this nature
-% presents profound technical difficulties, meaning genealogy-based approaches have
-% not been widely used in population and statistical genetics.
-% Until recently, analysis of
-% population genetic data has therefore primarily focussed on summaries derived
-% from the local trees~\citep{tajima1983evolutionary,tavare1984line}
-% rather than using the full underlying genealogy itself. Such
-% derived measures include allele frequencies and site frequency
-% spectra~\citep{achaz2009frequency,ralph2020efficiently},
-% patterns of linkage disequilibrium~\citep{mcvean2002genealogical}, and
-% principal components~\citep{mcvean2009genealogical}.
 Estimating the genetic genealogy of a set of genome sequences
@@
 differently (see Appendix XXX).
-%% Sentence deleted: point made already
-% ARG inference was for many years synonymous [FIXME: this isn't really true,
-% what about parsimony ARGs?]
-% with estimating realisations
-% of the coalescent with recombination (with, it must be said, limited success),
-% and the distinction between process and realisation was not especially important.
-
-%% Repeat of first para
-% The recent introduction of two methods, \relate~\citep{speidel2019method}
-%and \tsinfer~\citep{kelleher2019inferring}, that estimate structures
-%sharing many properties of this classical idea of an ARG, but at previously
-%unthinkable scale, has led to a surge of interest in the area.
-%
-%
-%Although both \cite{speidel2019method} and \cite{kelleher2019inferring}
-%avoid describing the output of the methods as an ARG,
-%and we therefore need a more general definition of an ``ARG'' than the classical
-%Griffiths formulation.
 
@@
 the output of all popular simulation and inference methods.
-% We follow the well established principle that it is advantageous
-% to decouple the presence of mutations from the genealogical structure
-% itself \citep[contra][chap.~3]{gusfield2014recombinatorics}.
 We propose a definition of genetic genealogy in terms of specific genomes
@@
 
-% We examineV}
-% In particular, we argue that the goal of inferring fully precise
-% ARGs---a necessity under the eARG encoding---should be
-% % weird phrasing for a criticism?
-% critically examined
-% %
-% as part of the process of fully decoupling our ARG data
-% structure from any specific generative model.
-
-% and because it is focused on encoding the \emph{outcome}
-% of evolutionary events rather than the details of the events themselves,
-
-% In this paper we [turn into proper round up paragraph later]
-% \begin{enumerate}
-% \item Discuss the classical Ancestral Recombination Graph formulation,
-% and show how it derives from (and is limited by) the coalescent with
-% recombination.
-% \item In the following sections we
-% show that, although the ARG-as-a-data-structure perspective is
-% now the most common interpretation, the \emph{form} of that
-% data structure is still closely linked to the original
-% stochastic process, and that this form is inflexible
-% in the types of genetic inheritance that can be described,
-% computationally inefficient, and
-% forces complete (and potentially spurious) precision about
-% the location and timing
-% of recombination events. % (and coalescent events?)
-% \item Suggest a definition of ARGs that is free of these limitations,
-% where the formulation is derived from a (normally diploid) pedigree rather than from
-% coalescent approximations, and where the genetic
-% material which is transmitted is described by annotations attached to
-% edges rather than nodes.
-% \item Suggest a classification of ARG nodes that helps us to understand
-% the properties of the structures inferred by different methods, and illustrate
-% with some examples.
-
-% \item This, coupled with
-%     the explicit association of genome coordinates with edges, allows
-%     us to represent even the bewilderingly complex patterns of relatedness
-%     present in modern datasets.
-
-% \end{enumerate}
-
-% \section*{Genome ARGs}\label{eARG}
-% In keeping with its original derivation in terms of a stochastic
-% process (see Appendix~\ref{sec-arg-history}), the classical definition of an
-% ARG data structure is
-% in terms of events.
-% Nodes represent
-% common ancestor or recombination events that occurred in the
-% history of some set of sample genomes, and edges represent ancestral
-% lineages~\citep{griffiths1996ancestral}.
-% In a common ancestor event the inbound lineages are merged into a
-% single ancestral lineage, and in a recombination
-% event a lineage is split into two independent
-% ancestral lineages.
-% % The final vital
-% % detail is that we associate a breakpoint with each recombination
-% % event.
-% % This information is sufficient to uniquely
-% % define the local genealogical trees at every position along the genome,
-% % which is the basic requirement for a data structure encoding
-% % genome-wide genetic ancestry.
-% As illustrated in Fig.~\ref{fig-arg-data-structure}, this information
-% is usually concretely encoded by associating a recombination breakpoint
-% (which is null for common ancestor events) and an ordered list of parents
-% with each node. We refer to such a  data structure as an
-% ``event ARG'' (eARG).
-% To recover the local tree at genome coordinate $x$ (the most
-% fundamental operation required for an ARG data structure),
-% we traverse the graph rootwards from the leaves.
-% At a particular
-% node $u$, if it has one parent we are at a common ancestor
-% node and we follow that parent. If we are at a
-% recombination node, $u$ has two parents: if
-% $x$ is less than the breakpoint we follow
-% the edge to the first parent, and otherwise follow the edge
-% to the second parent.
-% The order in which parents are listed at a recombination node is
-% therefore significant, telling us
-% from which parent the segments of genome to the left and right of the breakpoint
-% were inherited.
-
-% This ordering requirement, while straightforward
-% to describe, has some practical drawbacks. For example,
-% % Surely the meaning is clear from the context here and we don't need
-% % qualify the meaning of ``ARG''
-% when using this representation and simulating an ARG backwards in time,
-% the first event older than a recombination may involve the lineage carrying the
-% ancestry to the right of the breakpoint rather
-% than the left, and so we cannot emit edges as they are generated.
-% Such issues can be worked around, of course, but depending on the ordering of
-% otherwise indistinguishable objects is generally problematic.
-
-% \begin{figure}
-% \centering
-% % \begin{tabular}{cc}
-% \begin{tikzpicture}[x=5mm, y=5mm, node distance=2mm and 20mm]
-% \tikzset{greynode/.style={circle,fill,inner sep=1},
-% nodelabel/.style={font=\footnotesize}}
-
-% \node (s0) [greynode] at (0, 0) {};
-% \node (s1) [greynode] at (3, 0) {};
-% \node (s2) [greynode] at (6, 0) {};
-% \node (s3) [greynode] at (3, 1) {};
-% \node (s4) [greynode] at (1, 2) {};
-% \node (s5) [greynode] at (5, 3) {};
-% \node (s6) [greynode] at (3, 4) {};
-
-% % \node [anchor=north west] at (0,6) {A};
-% \node [nodelabel,anchor=north west] at ($(s3) + (0,0)$) {$x$};
-% \foreach \u/\lab in {s0/$\textsf{a}$, s1/$\textsf{b}$, s2/$\textsf{c}$} \node[nodelabel,anchor=north] at (\u) {\lab};
-% \foreach \u/\lab in {s4/$\textsf{e}$} \node[nodelabel,anchor=south west] at (\u) {\lab};
-% \foreach \u/\lab in {s5/$\textsf{f}$} \node[nodelabel,anchor=south east] at (\u) {\lab};
-% \foreach \u/\lab in {s3/$\textsf{d}$, s6/$\textsf{g}$} \node[nodelabel,anchor=south] at (\u) {\lab};
-
-% %% Edges
-% \draw (s1) -- (s3);
-% \draw (s0) |- (s4);
-% \draw (s4) -- (2,2) |- (s3);
-% \draw (s4) |- (s6);
-% \draw (s3) -- (4,1) |- (s5);
-% \draw (s2) |- (s5);
-% \draw (s5) |- (s6);
-
-% % \node [anchor=north west] at (9,6) {B};
-% \node [nodelabel,anchor=north west] at ($(10,5)$) {
-% \begin{tabular}{c|c|l}
-% % \multicolumn{2}{c}{Breakpoints}\\
-% Node & Breakpoint & Parents\\
-% \hline
-% $\noderef{a}$ & $\varnothing$ & [\noderef{e}]\\
-% $\noderef{b}$ & $\varnothing$ & [\noderef{d}] \\
-% $\noderef{c}$ & $\varnothing$ & [\noderef{f}]\\
-% $\noderef{d}$ & $x$ & [\noderef{e}, \noderef{f}] \\
-% $\noderef{e}$ & $\varnothing$ & [\noderef{g}]\\
-% $\noderef{f}$ & $\varnothing$ & [\noderef{g}]\\
-% $\noderef{g}$ & $\varnothing$ & []\\
-% \end{tabular}};
-
-% % \node [anchor=north west] at (9,6) {C};
-% % \node [nodelabel,anchor=north west] at ($(10,5)$) {
-% % \begin{tabular}{ll}
-% % Edges & Intervals\\
-% % \hline
-% % $(\textsf{a}, \textsf{e})$ & $[0, L)$ \\
-% % $(\textsf{b}, \textsf{d})$ & $[0, L)$ \\
-% % $(\textsf{c}, \textsf{f})$ & $[0, L)$ \\
-% % $(\textsf{d}, \textsf{e})$ & $[0, x)$ \\
-% % $(\textsf{d}, \textsf{f})$ & $[x, L)$ \\
-% % $(\textsf{e}, \textsf{g})$ & $[0, L)$ \\
-% % $(\textsf{f}, \textsf{g})$ & $[0, L)$ \\
-% % \end{tabular}};
-
-% \end{tikzpicture}
-% \caption{\label{fig-arg-data-structure}
-% A graphical depiction of a classical event ARG and a concrete
-% encoding of the information. Node \noderef{d} is a recombination
-% event as it is associated with a non-null breakpoint, and has
-% two parents. Nodes \noderef{e}, \noderef{f} and \noderef{g} are
-% common ancestor events.
-% }
-% \end{figure}
-
-% % What do we mean by an eARG, formally?
-% To be concrete, we can formally define the
-% classical Griffiths eARG as a tuple $(e, \sigma)$, where $e$
-% is an ordered list of edges defining a directed acyclic graph and
-% $\sigma: \mathbb{N} \rightarrow \mathbb{N}$
-% is a function mapping nodes to recombination breakpoints,
-% such that $\sigma(u) = x$
-% if $u$ is a recombination event with breakpoint $x$ and
-% $\sigma(u) = \varnothing$ otherwise.
-% (This is equivalent to having different ``types''
-% of node, where breakpoints are associated with recombination
-% nodes only).
-% Each edge $e_j = (c, p)$ describes an edge between
-% child node $c$ and parent $p$, where $c, p \in \mathbb{N}$.
-% % For each recombination node $u$ we must
-% % have exactly two edges in $e$ such that $u$ is the child.
-% This encoding of an eARG is equivalent to the output
-% of programs such as
-% \texttt{ARGweaver}~\citep{rasmussen2014genome}
-% and \texttt{KwARG}~\citep{ignatieva2021kwarg},
-% and an example is given in Fig.~\ref{fig-arg-data-structure}A.
-% Note that the description above captures only the
-% graph topology: if we also wish to know the branch lengths we need
-% an additional function $t(u): \mathbb{N} \rightarrow \mathbb{R}$
-% which defines the time of each node.
-
-% In practice,
-% several methods explicitly associate information with the outbound edges
-% of a recombination event
-% to resolve the issue~\citep{lyngso2005minimum,ignatieva2021kwarg}.
-
-% TODO update this when later sections have been drafted.
-% As we see in sections XXX and YYY there are substantial limitations
-% of expressivity and computational performance when using this eARG
-% encoding. More fundamentally, however, the general approach
-% of classifying nodes according to a discrete set of events
-% and associating genome information with those nodes is intrinsically
-% limiting. As we saw in the previous section, the original ARG
-% definitions were derived directly from the coalescent with recombination
-% stochastic process, and the types of ancestry that can be
-% described are therefore limited by the assumptions of that
-% model.
-
-% JK: Removing stuff about the limitations of the encoding here
-% because it's reasonably straightforward to see an extension in
-% which the number of breakpoints is in less than the number of
-% parents, and it starts muddying the waters a bit.
-
-% The eARG encoding consists of two types of event
-% (common ancestor and recombination) but nature is not so parsimonious
-% and there are many other processes that will affect the
-% ancestry of samples.
-% For example, gene conversion between homologous
-% chromosomes~\citep{wiuf2000coalescent,chen2007gene} is an important
-% evolutionary force in many species~\cite[e.g.][]{yang2012great}.
-% % Some refs for where people are inferring ancestry in the
-% % presence of GC? Isn't this what ClonalFrame etc are all about?
-% Similarly, the eARG encoding cannot easily represent multiple
-% recombinations on a chromosome, or the transmission of multiple
-% chromosomes.
-% What we can infer about ancestral histories and
-% processes should not be limited by the data structure that we
-% use to represent such ancestry. The eARG is simply one way
-% to \emph{encode} such history, and, as we have shown,
-% one that has significant limitations of expressivity
-% derived from its historical origins in the idealised model of
-% coalescent with recombination.
-
-% \section*{Node vs edge annotations}
-% Another aspect of the classical description of an ARG data structure that
-% derives from its historical origins as a stochastic
-% process is how we encode information about genetic
-% inheritance in the graph. The approach taken by Griffiths to
-% doing this is elegant: all we need to do is store
-% the location of the breakpoint at each recombination event.
-% % Storing a breakpoint for each recombinatin event is a very
-% % natural and straightforward thing to do when simulating
-% % an  ARG process.
-% Given the graph topology and these breakpoints, we can then
-% recover the
-% local trees at every point in the genome
-% by tracing rootwards from the leaves and choosing the
-% appropriate outbound path at recombination nodes.
-% However, there are several significant disadvantages
-% to this approach: [TODO be concrete here later
-% and give pointers to the section where we discuss the
-% alternative.]
-
-% Fortunately, there is a straightforward change to the
-% classical eARG encoding that simplifies
-% implementations and greatly increases the flexibility
-% of both the types and granularity of inheritance information
-% that we can represent.
-% The main change that we need to make is to associate
-% the details of inheritance with \emph{edges} rather than
-% as breakpoints associated with recombination events and
-% edge ordering rules.
-% This is illustrated in Fig~\ref{fig-arg-data-structure}C
-% where this alternative edge annotation formulation is
-% shown. By associating the interval $(0, x)$ with edge $(d,e)$
-% and the interval $(x, L)$ with $(d, f)$ we make the encoding
-% unambiguous: we can list the edges in any order, and it
-% will describe the same set of local genealogical trees.
-% This approach is also more flexible, because any pattern
-% of inheritance can be described by these intervals. For
-% example, if a gene conversion event occurred where a segment
-% between positions $x$ and $y$ was inherited by
-% node $d$ from $e$, we would state this by associating
-% the intervals $\{[0, x), [y, L)\}$ with $(d,e)$ and
-% $\{[x, y)\}$ with $(d,f)$.
-% This change is straightforward, involves no loss of information
-% and (as we see in later sections)
-% greatly increases both the expressivity and computational
-% power of ancestral recombination graph data structures.
-
 \section*{Genome ARGs}
@@
 many generations.
+We are not concerned here with mutational processes or observed sequences,
+% FIXME topological isn't the right word...
+but consider only processes of inheritance,
+following the standard practise in coalescent theory.
 % A ``genome'' in our usage corresponds to the

### 2023-09-20 paper.tex
More tidy up

@@
 following the standard practise in coalescent theory.
-% A ``genome'' in our usage corresponds to the
-% haploid genetic material present in a particular individual or cell.
 We will also use the term ``genome'' in its
@@
 \begin{center}
-    \includegraphics[width=\textwidth]{illustrations/arg-in-pedigree}
+    % FIXME making this 0.75 textwidth for now because it's too tall.
+    % https://github.com/tskit-dev/what-is-an-arg-paper/issues/321
+    \includegraphics[width=0.75\textwidth]{illustrations/arg-in-pedigree}
 \end{center}
@@
 % it'll be clearer what we need to say in the caption vs the text.
-An example gARG embedded in a pedigree.
+An example genome ARG embedded in a pedigree.
 (A) Diploid individuals, visualised in a highly inbred pedigree and
@@
 (partial inheritance in bold).
-% Recombination (e.g.,
-% the cause of the breakpoint at position 7)
-% is associated with two parent genomes (\textsf{e} and
-% \textsf{f}) and a child genome (the recombinant, \textsf{c}).
-% (C) The edge-annotated genome ARG in (B) fully defines the
 (C) The corresponding local trees.
@@
 nodes with a single child (see Section XXX).
-% are not explicitly concerned with ``coalescence'' (nodes at which
-% one or more samples find a common ancestor); for example, node \noderef{e}
-% is present but not coalescent in the left-most tree of
-% Figure~\ref{fig-arg-in-pedigree}
-% (see section XXX for more discussion).
 Fourth, recombination is modelled in terms of
@@
 % from the others. Is there a more direct way we can address these points ?
-
-% In this section we have considered the nodes in a gARG as separate from events
-% in a generative process. Nevertheless, gARGs are commonly constructed to
-% include only those genomes associated with specific events. For example,
-% \cite{mathieson2020ancestry} define an ARG as ``a subset of the pedigree...
-% contain[ing] only those edges along which inherited segments of DNA have been
-% transmitted, and only those nodes corresponding to ancestors in which there was
-% a recombination or coalescence event.''. In an eARG however, events are even
-% more explicitly embedded in the data structure.
-
-% There are some interesting and important consequences to this focus on genomes
-% rather than events.
 
@@
 % and their identifiability.]
-
-% The reader may arrive at the end of this section going, er, so what?
-% Try to tackle this head on by getting to the heart of the distinction.
-% The difference between eARGs and gARGs are subtle, and whether
-% we choose to interpret nodes as events or genomes
-% might seem of little practical importance.
-% However, there are substantial practical benefits
-% from using the gARG encoding, allowing differential
-% levels of information about the recombination process to be
-% represented in a computationally efficient way.
-% The key to this is not so much in how we interpret nodes
-% (see Appendix XXX), but specifically in how
-% we represent the process of genetic inheritance.
-% The critical distinction is that
-% the gARG encoding allows us to directly describe
-% genetic inheritance in multiple non-contiguous genome intervals
-% between an ancestor and a descendant, something that
-% cannot be done by storing recombination breakpoints.
 
@@
 \begin{figure}
+% FIXME add local trees in B
+% https://github.com/tskit-dev/what-is-an-arg-paper/issues/320
 \centering
@@
 lead to significantly different time or space complexity in computational
-tasks. The vitally important and fundamental distinction between
-a gARG and the classical eARG perspective is that gARGs do not
-\emph{require} us to posit explicit recombination events in order
-to describe a dataset, but
-instead provide a means of encoding the effects of
-multiple events, and to systematically describe the uncertainty around them
-(see Section XXX).
+tasks.
+The vitally important and fundamental distinction between
+a gARG and an eARG is that gARGs
+allow us to directly describe
+genetic inheritance in multiple non-contiguous genome intervals
+between an ancestor and a descendant, which cannot be done
+by specifying recombination events and breakpoints.
+Thus, we have the means of encoding the effects of
+multiple \emph{simultaneous} events, and to
+systematically describe the uncertainty around their
+ordering (see Section XXX).
 As we see in the following sections, freeing ourselves from the requirement
@@
 
-% JK: previous heading was jargony - this is clearer
 \section*{Converting an eARG to a gARG}

### 2023-09-20 paper.tex
Tidy up section on Wiuf & Hein ARG
Rescale WH figure

@@
 % up far too much vspace at in native coords at the moment.
-\includegraphics[width=0.75\textwidth]{illustrations/ancestry-resolution}
+\includegraphics[width=0.6\textwidth]{illustrations/ancestry-resolution}
 \caption{\label{fig-ancestry-resolution}
@@
 
-%(see Appendix XXX) and has several features that we are unlikely to simulate or
-%infer in practice. For example, by definition the samples contain no genetic information about
-%nodes \textsf{g} and nodes \textsf{j} and so neither of those nodes are likely to be inferred.
-%Likewise, is is unlikely we would have information to infer the ``grand MRCA''
-%\textsf{q} as all regions of the genome have already reached a
-%common ancestor separately (in nodes \textsf{k} and \textsf{p}).
-
-% Third, there exist nodes such as
-% \textsf{k} with two children (i.e. ``common ancestor'' nodes in Hudson's terminology)
-% but in which no ancestral coalescence occurs. Fourth, some breakpoint positions occur
-% within regions of genome that are not ancestral (e.g. the breakpoint at position 3 in
-% node \textsf{k}). In this case sample-resolved edge annotations lose the exact position of
-% the breakpoint, although they retain the information that a breakpoint occurred between
-% positions 2 and 4 (which is sufficient for calculating the likelihood, see appendix XXX).
-
-% reduce the annotation intervals to those that are ancestral to the
-% sample (note that node $q$ is now ommitted because the MRCA of the sample
-% has already been found for the full genome at node $p$).
-% Tracking such
-% ``ancestral material''~\citep{wiuf1999ancestry,wiuf1999recombination}
-% is much more useful~\citep{mcgill2013graphml,arenas2013importance}
-% and, indeed, all that is possible to
-% have information about when inferring ARGs from contemporary samples.
-% We refer to a gARG with inheritance annotations of this type as
-% ``sample resolved''.
-
-% Fig~\ref{fig-ancestry-resolution} is an instructive example
-% from the literature \citep{wiuf1999recombination} and illustrates
-% how a classical eARG (Fig~\ref{fig-ancestry-resolution}A) can be described as
-% a gARG with edge annotations (Figs~\ref{fig-ancestry-resolution}B~and~C).
-
-% There are two distinct annotation schemes possible, and the distinction is subtle
-% and important. The first scheme is to use the annotations to describe the complete
-% set of genomic intervals directly inherited by the descendant node from the ancestor
-% node (Fig~\ref{fig-ancestry-resolution}B). The second is to use the edge annotations to
-% describe the ``ancestral material''~\citep{wiuf1999ancestry,wiuf1999recombination}
-% of the sample that is propagated through
-% the graph from descendant to ancestor (Fig~\ref{fig-ancestry-resolution}C).
-% As we discuss in section XXX, this second scheme has many advantages
-% in terms of efficiency.
-
-%ARGs are inherently a retrospective structure defined in terms
-%of a set of sample nodes (usually the leaves), and,
-%as we discuss in section XXX, storing sample-resolved ancestry
-%intervals (Fig~\ref{fig-ancestry-resolution}C) is generally much more
-%useful than storing the parent-child intervals recording direct inheritance
-%(Fig~\ref{fig-ancestry-resolution}B).
-%Indeed, storing direct inheritance intervals for edges is really only of interest because of
-%its correspondence with the classical approach of associating
-%breakpoints with recombination nodes in an eARG (Fig~\ref{fig-ancestry-resolution}A).
-%The eARG encoding is simply a more limited form, in which the only types of direct
-%inheritance we can have are the entire genome, or the intervals
-%at either side of a recombination breakpoint.
-
-%Note that this example is a sample from the so-called \emph{``big'' ARG process}
-%(see Appendix XXX) and has several features that we are unlikely to simulate or
-%infer in practice. For example, by definition the samples contain no genetic information about
-%nodes \textsf{g} and nodes \textsf{j} and so neither of those nodes are likely to be inferred.
-%Likewise, is is unlikely we would have information to infer the ``grand MRCA''
-%\textsf{q} as all regions of the genome have already reached a
-%common ancestor separately (in nodes \textsf{k} and \textsf{p}).
-
-% Third, there exist nodes such as
-% \textsf{k} with two children (i.e. ``common ancestor'' nodes in Hudson's terminology)
-% but in which no ancestral coalescence occurs. Fourth, some breakpoint positions occur
-% within regions of genome that are not ancestral (e.g. the breakpoint at position 3 in
-% node \textsf{k}). In this case sample-resolved edge annotations lose the exact position of
-% the breakpoint, although they retain the information that a breakpoint occurred between
-% positions 2 and 4 (which is sufficient for calculating the likelihood, see appendix XXX).
-
-% Given a gARG with direct-inheritance edge annotations
-% (Fig~\ref{fig-ancestry-resolution}B),
-%  we obtain the sample-resolved edge annotation
-% (Fig~\ref{fig-ancestry-resolution}C) by traversing the graph backwards in time
-% from sample nodes (\textsf{a}, \textsf{b}, \textsf{c}).
-% Proceeding in topological order (such that an ancestor node is visited
-% after all of its descendants) the sample-resolved material for a node is the
-% union of the sample-resolved intervals on its inbound edges. We then annotate
-% the outbound edges with the intersection of their direct-inheritance intervals
-% and the just-computed sample-resolved intervals for the node.
-% For example, node \textsf{i} in Fig~\ref{fig-ancestry-resolution}C
-% carries sample-resolved material on the intervals $[0, 2)$ and $[4, 7)$. The edge
-% joining \textsf{i} to its ancestor \textsf{k} then carries
-% material on $[0, 2)$ since although \textsf{i} directly inherits interval
-% $[0, 3)$ from \textsf{k} (Fig~\ref{fig-ancestry-resolution}B), only $[0, 2)$ is
-% ancestral to the sample. For each node, we also keep track of the number of
-% descendant samples over each genomic interval, and stop tracking that
-% interval once the lineages from all samples have coalesced. For example,
-% the edge from \textsf{k} to \textsf{n} omits the region $[0,2)$ because it has
-% already fully coalesced (indicated by a fully black genomic region).
-
-% In practice, the backwards-in-time Griffiths process that produces
-% ARGs such as that shown in Figs~\ref{fig-ancestry-resolution}A~and~B
-% is rarely simulated, because the resulting ``big ARG'' (Appendix XXX) is
-% dominated by nodes such as \textsf{g} and \textsf{j} which contain information
-% about inheritance in non-ancestral material, as well as nodes such as
-% \textsf{o} and \textsf{q} which track details of ancestry even once the
-% local genomic region has fully coalesced. Instead, simulating
-% ARGs from the coalescent with recombination is usually done by
-% running Hudson's algorithm to produce a sample-resolved ARG directly.
-% Moreover, when inferring ARGs from genetic data, there is no information
-% in the sample set that can be used to resolve these nodes, so that they
-% are essentially unknowable.
-
-% The process of converting a gARG with local-inheritance annotations
-% to its sample-resolved equivalent is also known as ``simplification'',
-% and is the key algorithm underlying recent
-% advances in forwards-in-time
-% simulation~\citep{kelleher2018efficient,haller2018tree}.
-% It has many useful properties, as we discuss in the next section.
-
-% It is worth noting the process of sample resolution can reveal that certain ARG
-% nodes are \emph{redundant}: if they are removed and the edges adjusted to
-% pas directly from the node's child to the node's parent, it does not reduce our
-% knowledge of the samples' genetic ancestry. Removing nodes,
-% both redundant ones and others, is part of a process we call
-% ``simplification''.
-
 \section*{Levels of simplification}\label{ARG_simplification}
@@
 (where a node has exactly one child over a particular genomic region).
-% "unary regions" is jargon - just dropping it in here like this doesn't
-% clarify. We can't assume that people are used to switching between
-% graphs and trees.
-% of ``unary'' regions for nodes
-% (where a node has exactly one child over a particular genomic region).
 These arise in several different ways.
@@
 \centering
-\vspace{5em}
-\includegraphics[width=\linewidth]{illustrations/simplification}
+\includegraphics[width=\textwidth]{illustrations/simplification}
 \caption{\label{fig-simplification}
@@
 
-% \noderef{a} in the leftmost tree (covering positions 0 to 30)
-% Consider, for example, the path ancestral to node
-% \noderef{a} in the leftmost tree (covering positions 0 to 30)
-% in Fig.~\ref{fig-simplification}A.
-% We can see from the graph that
-% % Before \noderef{a} meets its common ancestor with the other samples at
-% % node \noderef{q}, it passes through \noderef{e}, \noderef{j}, \noderef{m}, \noderef{n} and
-% % \noderef{p}, which in this region of the genome are all unary nodes.
-% % Node \noderef{e}
-
-% Such unary nodes in a local tree are not usually considered important;
-% % because they provide little information and are not
-% % detectable by information from any given individual tree.
-% nodes usually mark the points where samples coalesce in a
-%  \emph{common} ancestor. For most purposes, we would therefore consider a path directly
-% joining sample \noderef{a} to node \noderef{q} to be equivalent to the path
-% containing the intermediate unary nodes.
-
 The first level of simplification that we can perform is to remove
@@
 multiple---often indistinguishable---events into a single node.
-% something which is impossible to encode in an eARG.
 
@@
 between trees. The extent to which this information can be inferred
-and utilised is an interesting open question [CITE Ralph in prep.].
+and utilised is an interesting open question.
+% I don't think there's an actual manuscript or title to refer to here
+% [CITE Ralph in prep.].
 
@@
 at position 87 has been lost entirely.
-
-% or nothing'': there is a gradation of detail
-% about recombination, in terms of genomic location, timing,
-% and the ancestral lineages involved.
-% that vary levels of precision about reccomb
-
-% The example in Fig.~\ref{fig-simplification} is usually
-% referred to as ``fully simplified'', and is the default output
-% of simulation programs like \msprime. While there is significant
-% precision about recombination events lost
 
@@
 
-% % Not sure where this goes, but there are some critical points
-% % that we want to explain and not just throw out there in the
-% % discussion.
-% % A critical distinction between the eARG and gARG encodings
-% % is the level of precision about the time and location of
-% % recombination events that is required. In an eARG, we are
-% % required to be absolutely precise about every recombination
-% % [explain and contrast with gARG, referring to examples]
-
 % % JK: This is very rough, just jotting down the basic ideas,
@@
 % resolved binary trees and those that contain polytomies.
-
 % The situation is in fact directly analogous: an ARG that
 % contains fully resolved recombination information must have
-% at most parents per node.
-
-% At the very least, it is not reasonable for the \emph{data
-% structure} to require complete precision in order to
-% represent some less precise. This is a fundamental
-% advantage of the gARG encoding over the classical eARG
-% method: we are not forced to posit a precise cause (recombination
-% event) for every effect (difference between adjacent trees).
-% These differences in precision of recombination
-% information are illustrated in the ARGs estimated by different
-% methods, as we see in the next section.
+% at two most parents per node.
 
@@
 Note that many authors use the term ``marginal''
-trees~\citep[e.g.][]{minichiello2006mapping,griffiths1996ancestral} [more]
+trees~\citep[e.g.][]{griffiths1996ancestral,minichiello2006mapping} [more]
 which we avoid due the potential confusion with statistical
@@
 % trees from different regions of the genome.
-
-% \section*{When do we need ``full'' ARGs?}
-% % The full ancestral recombination graph (ARG) is a structure that encodes all
-% % coalescence and recombination events resulting from the stochastic process of
-% % the coalescent with recombination.
-% \citet{brandt2021evaluation} define a ``full'' ARG as
-% ``a structure that encodes all
-% coalescence and recombination events resulting from the stochastic process of
-% the coalescent with recombination'' (thus going against the general
-% historial trend discussed in the XXX section).
-% \citet{rasmussen2022espalier} have a less restrictive definition, in
-% which a ``full'' ARG is one that contains complete information about
-% recombination events, without necessarily being tied to a particular
-% stochastic process.
-% While [something positive], it is important to be clear on
-% what the overall goal of such precise inferences are,
-% what the limitations on when they can be made are,
-% and their downstream utility in applications.
-
-% % What is the goal of having fully precise ARGs?
-% % 1) studying the actual recombination events;
-% % 2) evaluating the likelihoods
-% The main reasons for wanting precise estimates of all recombination
-% events are to either compute a likelihood of the inferred ARG
-% under a statistical model, or to study the recombinants in
-% detail~\citep{rasmussen2022espalier}.
-% % Explain that we're really interested in the details of recombination
-% % itself with a small number of samples.
-% In the latter case, we might
-% have a small number of samples~\citep{guo2022recombination}.
-
-% Beyond cases in which visual inspection of the trees and the
-% effects of recombination is feasible, the main application
-% of having completely precise estimates of recombination events is
-% to evaluate the likelihood of the estimate under some
-% statistical model. The only statistical model available
-% is the coalescent with recombination
-% and its approximations~\citep{mcvean2005approximating,marjoram2006fast}.
-% In this case, the utility of a ``full`` ARG for some dataset
-% is limited by how well the dataset fits the assumptions
-% of the model.
 
@@
 % [More]
-
-% % Is it practical to estimate full ARGs?
-% % [Not particularly \citep{shipilina2022origin} - but maybe ARGneedle
-% % has done this for UKB? Need to check]
 

### 2023-09-21 paper.tex
Pass through latter sections

@@
 but consider only processes of inheritance,
-following the standard practise in coalescent theory.
+following the standard practice in coalescent theory.
 We will also use the term ``genome'' in its
@@
 large polytomies.
-To recover the local tree $\pi$ at position $x$  then, we begin by
+To recover the local tree $\pi$ at position $x$, we begin by
 setting $\pi_u = -1$ for each $u\in N$. Then, for each sample
@@
 and \relate~\citep{speidel2019method}.
-While there is some consistency among the inferred ARGs (such as the samples
-FR-F, WA-F and Af-F being grouped into the same clade), there
-are also considerable differences. For this dataset, the minimum number
+% While there is some consistency among the inferred ARGs (such as the samples
+% FR-F, WA-F and Af-F being grouped into the same clade), there
+% are also considerable differences.
+The minimum number
 of recombination breakpoints required in the absence of recurrent mutation
-is 7~\citep{song2003parsimonious}, and the number of inferred breakpoints
+is 7 for this dataset~\citep{song2003parsimonious},
+and the number of inferred breakpoints
 under the input mutation and recombination rates
@@
 Essentially,
-\kwarg\ and \argweaver\ infer a specific causal recombination \emph{event}
-for each breakpoint, and those events are fully
-described in the output graph. On the other hand, \tsinfer\ and
-\relate\ do not infer causal events but instead focus on
-representing the inferred \emph{effect}: breakpoints between the
-trees. It is important to note that this does not imply that
-the methods must then necessarily infer only a sequence of
+\kwarg\ and \argweaver\ infer complete eARGs (section XXX),
+which contain specific causal recombination \emph{events}
+for each breakpoint.
+On the other hand, \tsinfer\ and
+\relate\ do not infer causal events but rather
+infer gARGs with differing levels of precision about recombination.
+The bottom row of Fig~\ref{fig-inferred-args} shows the extent along
+the genome to which graph edges are shared between multiple trees.
+We can see, for example, that the \relate\ trees ``share'' less topology
+than other methods, but it certainly does not infer a series of
 \emph{unrelated} local trees (i.e., with either unlabelled internal
@@
 The ARGs inferred by \relate\ and \tsinfer\ both contain substantial
-information about node sharing across trees, and hence,
-in terms of the level of detail inferred about recombination,
-lie somewhere
-in between a sequence of unrelated trees on one extreme and a
-completely precise event ARG on the other.
+information about node sharing across trees; for example,
+both infer that Af-f, Fr-f and Wa-f share a common ancestor along
+the entire sequence.
+Therefore, they contain a level of detail about recombination
+that lies somewhere between a sequence of unrelated local trees on one
+extreme and a completely precise event ARG on the other.
 
@@
 Recent breakthroughs have made ARG inference methods practical
-for the first time, and there has been a surge of interest
+at scale for the first time, and there has been a surge of interest
 in inference methods, their evaluation and application.
@@
 Graph has several
-subtly different interpretations depending on context. The
-trend to decouple the ARG from its origins as a stochastic
+subtly different interpretations depending on context.
+The trend to decouple ARGs from their origins as a stochastic
 process and instead use the term as a more general representation of any
@@
 genomes~\citep{bycroft2018genome,karczewski2020mutational,tanjo2021practical},
-substantially \emph{larger} than the usually assumed $N_e$ values.
+meaning that sample size is substantially larger than the
+usually assumed $N_e$ values.
 Agricultural datasets are an even more extreme departure from coalescent
-assumptions, with large samples embedded in
+assumptions, with hundreds of thousands of samples embedded in
 multi-generational pedigrees~\citep{hayes20191000,Ros-Freixedes2020}
@@
 without the hard requirement
-of complete precision on all historical events, may help improve
-methods for inference method evaluation. In the majority of cases,
+of complete precision on all historical events, may help
+to clarify inference goals and improve
+methods for evaluation.
+In the majority of cases,
 ARG inference accuracy is evaluated by comparing some truth simulations
 to the inferred ARGs by the pairwise comparison of local trees along the genome
-using tree metrics as originally proposed by \citet{kuhner2015assessing}.
-While this provides valuable insights into the method's performance,
-it is not clear how well these tree distance metrics reflect
-performance on real data.
+using tree distance
+metrics~\citep[e.g.][]{robinson1981comparison,kendall2016mapping}
+as originally proposed by \citet{kuhner2015assessing}.
+% While this provides valuable insights into the method's performance,
+% it is not clear how well these tree distance metrics reflect
+% performance on real data.
 In comparing tree-by-tree along the genome, the effects of recombination
-structure is incorporated in a very indirect manner.
+structure is incorporated in a indirect manner, and the tree
+metrics assume unlabelled nodes, effectively discarding information
+about node sharing between trees.
 Tree distance metrics usually have $O(n^2)$ time complexity, and so cannot be
 applied to the very large sample sizes currently of interest.
-\citet{deng2021distribution} derived the distribution of the expected
-waiting distances to tree changes.
-[ \citet{ignatieva2023distribution} DID SOMETHING ELSE]
+A recent trend has been to move away from such tree distance
+based approaches, and to examine more properties of the inferred ARGs.
 \citet{brandt2021evaluation} compared the
 distributions of time to the most recent common ancestor across several
-methods, which is a useful addition but only indirectly captures
-topological comparisons.
+methods, which captures many important aspects of dated ARGs
+but neglects [SOME OTHER THINGS].
+\citet{deng2021distribution} derived the distribution of the expected
+waiting distances to tree changes, and compared this to the
+outputs of several methods.
+[ \citet{ignatieva2023distribution} DID SOMETHING ELSE]
 % Not sure this is the right place for this sentence
-All of these comparisons and evaluations have been based on
-highly idealised simulations, and the effects richness of real data
-(beyond simplistic error models) are almost entirely unknown.
-A suite of true ARG distance metrics, that take into account more of the
+All of these evaluations have been based on
+ground truth data from highly idealised simulations
+with small sample sizes,
+and the effects of the richness of real data
+(beyond simplistic error models)
+on modern biobank scale datasets
+are almost entirely unknown.
+A suite of true ARG evaluation metrics, that take into account more of the
 global topology and can be applied to large ARGs, would be a
 valuable addition to the field.
+
+[NOTE This is a pedantic and slightly bullshitty to end up on,
+just writing it down to see if it works. Should probably cut.]
+A final point to note, is that
+if we are using the term ARG
+to loosely refer to genetic genealogies of varying
+levels of detail, either inferred or simulated,
+it makes little sense to refer ``the'' Ancestral Recombination
+Graph (i.e, in  the singular, like ``the tree'' or
+``the phylogeny'')
+% TODO add more if we keep this
+as favoured by some~\citep[e.g.][]{
+arenas2013importance,hubisz2020inference,zhang2023biobank}.
+While there is of course a single ``true'' ARG that describes
+the actual history of a sample, in reality there will be astronically
+many ARGs that are compatible with the data, even under
+parsimony conditions or when sampling from a model.
+There are many different levels of granularity at which we
+can reason about recombination, and many different ARGs.
+We should embrace this diversity.
 

### 2023-09-21 paper.tex
Address some comments

--- 2023-09-21 paper.tex
@@
 % TODO say this better.
-From a high-level, ARGs describe the sequence of correlated genealogical
-trees along the genome, induced by the different paths of
-genetic inheritance caused by recombination.
+From a high-level, an ARG describes the different paths of genetic inheritance 
+caused by recombination, and encodes a sequence of correlated genealogical 
+trees along the genome.
 Recent breakthroughs
@@
 If $u$ is a root
-% I'm not sure if this is a helpful clarification, and whether it'll just
+% I'm not sure if this reachability stuff is a helpful clarification, and whether it'll just
 % confuse people. It depends on some tricky details.
 (or not reachable from the samples)
-in the local tree at position $x$, $\pi_u = -1$.
+in the local tree at position $x$, $\pi_u = -1$ (by convention).
 Technically, this structure is a ``forest'', as it allows us to have
@@
 \section*{Event ARGs}\label{eARG}
-Genome ARGs are concerned with
+Genome ARGs are composed by 
 describing the details of genetic inheritance

### 2023-09-21 paper.tex
Change to refer to palaramara for "The" ARG
Drop "in retrospect" para
Closes #308

@@
 extreme and a completely precise event ARG on the other.
-
-% \relate\ "postorder" first inferring local topologies and then identifies edges
-% that persist from one tree, while
-% local trees only emerge from \tsinfer\ as a consequence of the copying
-% patterns inferred among ancestral haplotypes.
-
-% % More? See https://github.com/tskit-dev/what-is-an-arg-paper/issues/38
-% % for discussion. I'm sure there were other examples of people saying
-% % that
-% (as claimed by e.g.~\citet{hejase2020summary}),
-% and information about recombination and the correlation structure
-% between trees is not ``all or nothing''.
-% Both methods capture correlation structure between trees:
-% \relate\ first infers local tree topologies and then identifies edges
-% that persist from one tree to the next as a separate step, while
-% \texttt{tsinfer} explicitly constructs nodes that are shared across local
-% trees.
-% % Because the ARGs inferred by \tsinfer\ and \relate\ do not contain
-% % explicit recombination events we cannot compute a likelihood
-% % under the coalescent with recombination (see appendix XXX).
-% % [Some stuff about whether we should see this as necessary or not,
-% % but maybe we should keep this for the conclusion?]
+The output of these programs is certainly not a collection of
+\emph{unrelated} local trees, as implied by,
+e.g.,
+\citet{hejase2020summary} when they state that \tsinfer\
+``does not explicitly infer an ARG but rather a sequence of local
+gene trees''.
+
+% More cites? Not worth harping on about now?
+% https://github.com/tskit-dev/what-is-an-arg-paper/issues/38
 
@@
 
-% These properties have interesting consequences when we
-% consider the inference strategy of esimating local trees
-% independenty and subsequently stitching them together
-% into an ARG. Two recent methods take this approach,
-% \relate\ and
-% \texttt{espalier}~\citep{rasmussen2022espalier},
-% and have different strategies for reconciling the local
-% trees into an ARG. [Relate does something fairly lightweight,
-% but by default imprecise.]
-% Espalier, on the other hand, is specifically in interested
-% in the recombination events and therefore attempts to infer
-% the precise ``subtree prune-and-regraft'' (SPR)
-% operations~\citep{hein1990reconstructing,song2003on,song2006properties}
-% using heuristics~\citep{rasmussen2022espalier}.
-% Inferring the SPRs between
-% leaf-labelled trees is NP-hard in
-% general~\citep{hein1996complexity,allen2001subtree,bordewich2005computational}.
-
-% JK TODO: finish up this section with a quick version of the
-% para here. Basically say, people often say an ARG is a
-% set of local trees cite cite, but it's not as simple as that as we
+These properties have interesting consequences when we
+consider the inference strategy of estimating local trees
+independently and subsequently stitching them together
+into an ARG.
+Two recent methods take this approach,
+\relate\ and
+\espalier~\citep{rasmussen2022espalier},
+and have different strategies for reconciling the local trees into an ARG.
+\relate\ first infers local tree topologies, and the identifies
+edges that persist from tree to tree by reasoning about the sets of samples
+subtended. It does not attempt to be exhaustive.
+\espalier, on the other hand, is specifically interested
+in the details of recombination events (among small samples)
+and therefore attempts to infer
+the precise subtree prune-and-regraft (SPR)
+operations~\citep{hein1990reconstructing,song2003on,song2006properties}
+induced by recombination.
+Inferring the SPRs between leaf-labelled trees is
+NP-hard~\citep{hein1996complexity,allen2001subtree,bordewich2005computational},
+but it is unclear what the complexity is when there
+is a degree of internal node sharing between trees.
+
 % can see from the example (probably).
@@
 The succinct tree sequence data structure (usually known as a ``tree sequence''
-for brevity, but see below for confusion around this point)
+for brevity)
 is a concrete encoding of a gARG, as discussed in this article.
@@
 and likely many more.
-
-In retrospect, the choice of terminology around the succinct
-tree sequence data structure was unfortunate, and has led to
-significant confusion. As first introduced by~\cite{kelleher2016efficient}
-a ``tree sequence`` was defined as the set of ``coalescence records''
-that is output by Hudson's simulation algorithm.
-This was then generalised to use the
-tabular encoding described above and formally described as a ``succinct
-tree sequence''~\citep{kelleher2018efficient}.
-The ``succinct'' prefix here is intended to connect
-with the concept of succinct data structures
-\citep[e.g.][]{gog2014theory},
-which have near-optimal
-space usage but still support efficient retrieval and computation.
-The concrete data structures of the succinct tree sequence, precisely
-defined and with a particular focus on computational efficiency
-did not have an obvious connection to the ARG definitions
-provided by Griffiths and
-colleagues~\citep{griffiths1991two,ethier1990two,
-griffiths1996ancestral,griffiths1997ancestral}, and with the lack
-of any other precise definitions, it seemed best to avoid the
-term to prevent confusion. Unfortunately, however, the opposite
-has occured. In describing the output of \tsinfer\ as a ``tree
-sequence''~\citep{kelleher2019inferring} many
-have---not unreasonably, but incorrectly---concluded that the
-output is a collection of \emph{independent}
-% MORE?
-% https://github.com/tskit-dev/what-is-an-arg-paper/issues/38
-trees~\citep[e.g.,][]{hejase2020summary,ignatieva2021kwarg}.
-% Could say more here, but maybe we'll have a note about terminology
-% in the discussion
 
@@
 as favoured by some~\citep[e.g.][]{
-arenas2013importance,hubisz2020inference,zhang2023biobank}.
+arenas2013importance,palamara2016argon,hubisz2020inference}.
 While there is of course a single ``true'' ARG that describes

### 2023-09-21 paper.tex
Cleanup
Update

@@
 
-% can see from the example (probably).
-
 % The equivalence between an ARG and a set of local trees
@@
 
+\section*{Computational efficiency and data interchange}
+ARG inference methods have been successfully applied to vast datasets,
+including 500,000 genotyped humans~\citep{kelleher2019inferring,zhang2023biobank}
+from UK Biobank~\citep{bycroft2018genome}
+and over a million SARS-Cov-2 whole genomes~\citep{zhan2023towards}.
+Even larger datasets are
+available~\citep[e.g.][]{halldorsson2022sequences} % TODO more
+and it seems inevitable that ARGs will be inferred for them.
+The efficiency with which we can store and analyse these inferred
+ARGs is therefore of great importantance, and a key factor
+in the ultimate success of downstream ARG-based methods.
+Another key factor in how important ARG-based methods become
+across the wide range of predicted application
+% TODO more?
+areas~\citep{rasmussen2014genome,harris2019database,
+hejase2020summary,schaefer2021ancestral,
+fan2022genealogical,wohns2022unified,harris2023using,
+nowbandegani2023extremely}
+is how well inference methods and downstream analysis applications
+interoperate. Without a well-defined and efficient shared format
+for ARG data interchange,
+users will suffer from problems of converting data from one programs
+output to another, an unfortunate hallmark of population genetics
+software for many years~\citep{excoffier2006computer}.
+While current methods tend to tightly couple downstream analysis
+of the inferred ARG with the inference
+itself within the same software package
+% FIXME-  more explicit?
+(e.g.\ Relate, ARGneedle)
+this approach scales poorly in terms of software development
+effort, and is ultimately not compatible with the widespread use
+of ARGs for routine data analysis, and a healthy and diverse
+software ecosystem.
+% FIXME not sure where this goes
+Earlier efforts to standardise ARG interchange
+did not succeed~\citep{cardona2008extended,mcgill2013graphml}.
+
+The gARG encoding discussed and concretely defined in this manuscript
+leads to highly efficient storage and processing of ARG data,
+and has already been in use for several years.
+The succinct tree sequence data structure (usually known as a ``tree sequence''
+for brevity)
+is a concrete encoding of a gARG, as discussed in this article.
+It was originally developed as part of the \texttt{msprime}
+simulator~\citep{kelleher2016efficient} and subsequently been
+extended and applied to forward-time
+simulations~\citep{kelleher2018efficient,haller2018tree},
+calculation of population genetics statistics~\citep{ralph2020efficiently}.
+The succinct tree sequence encoding extends the basic definition
+of a gARG provided here by stipulating a
+simple tabular representation and nodes and edges,
+and also defining a concise and lossless representation of
+sequence variation using the ``site'' and  ``mutation'' tables.
+The \texttt{tskit} library~\citep{tskit2023tskit} is a liberally
+licensed open source toolkit that provides a comprehensive suite
+of tools for working with gARGs (encoded as a succinct tree sequence).
+Based on core functionality written
+in C, it provides interfaces in C, Python and Rust.
+Tskit is mature software, widely used in population genetics, and
+has been incorporated into several downstream
+applications~\citep[e.g.,][]{haller2019slim,speidel2019method,
+adrion2020community,
+terasaki2021geonomics,
+baumdicker2021efficient,
+fan2022genealogical,korfmann2022weak,
+mahmoudi2022bayesian,petr2022slendr,rasmussen2022espalier,
+zhang2023biobank,nowbandegani2023extremely}.
+
+% FIXME this needs more work and doesn't really fit into the
+% narrative right now, but we do need to make this point.
+The key insight that makes the succinct tree sequence encoding
+an efficient substrate for defining analysis algorithms is that
+it allows us to generate the local trees along the genome
+in a way that allows us to reason about the \emph{differences}
+between those trees.
+Sequentially generating the local trees along the genome
+is also fundamental, and is necessary whenever we need to
+perform calculations that are contingent on more than just the
+isolated properties of an edge. \cite{kelleher2016efficient}
+showed how all trees can be sequentially generated in
+constant time per tree transition in a fully simplified gARG.
+Furthermore, we can easily reason about how tree topologies
+change (and stay the same), leading to efficient algorithms
+for computing population genetic
+statistics~\citep{kelleher2016efficient,ralph2020efficiently},
+implementing the Li and Stephens
+model~\citep{kelleher2019inferring,wohns2022unified}
+and likely many more.
+
+\section*{SARS-CoV-2 ARGs: a case study}
+% What is SARS-CoV-2 and why is it interesting here? What's the dataset?
+% What are its basic evolutionary properties, summarised for a
+% population geneticist?
+Recombination via template switching is an important
+factor in the evolution of
+SARS-CoV-2~\citep{vaninsberghe2021recombinant,jackson2021generation,ignatieva2022ongoing}.
+With multiple deeply divergent lineages circulating globally,
+coinfection (and the potential for recombination)
+is relatively likely~\citep{bal2022detection}.
+The COVID-19 pandemic led to collection of whole-genome viral data
+at unprecented scale, with the GISAID database~\citep{shu2017gisaid}
+currently holding over 15 million sequences [FIXME]. Detection
+and characterisation of recombination is of profound interest
+for [reasons XYZ].
+
+% Summary of sc2ts paper. We inferred some big ole ARGs, with an
+% unprecedented degree of detail about recombination (outside, I guess,
+% of agri datasets). Summary of what's previously known about sc2 recomb.
+\citet{zhan2023towards} inferred two large ARGs for SARS-CoV-2 based
+on GISAID data, one containing 1.27 million sequences sampled up to
+June 30, 2021 (the ``Wide ARG'', and the second
+(more sparsely sampled) consisting of 657K sequences sampled up to June 30,
+2022 (the ``Long ARG''). Recombination is rare in both ARGs, with $<0.3\%$
+of the nodes being inferred to be recombinants (and quite likely
+a substantial number of false positives caused by poor data quality).
+Nonetheless, this provides thousands of automatically detected
+recombination events, the characterisation of which is usually a manual
+and labour intensive process. The recombinations events detected
+and fully characterised in these large ARGs fit well with previous
+which exhaustively investigate specific
+recombinants~\citep{jackson2021generation}.
+[Not sure what I'm trying to say here;
+what does a PopGen person who knows very little about SARS2 need to
+know here to given them context?]
+
+% What's the point of this case study? To emphasise that we
+% need to systematically represent topological uncertainty in ARGs.
+The pandemic scale SARS-CoV-2 dataset is an interesting case study
+for a number of reasons.
+Firstly, fundamental properties of the data underscore the
+need for topological uncertainty through polytomies
+and stacked recombination [more concise way of saying this?].
+The scale of sequencing effort at the height
+of the pandemic means that there were at times tens of thousands
+of sequences collected in a single day. This, combined with
+the slow rate of mutation, means that we will often have hundreds
+of identical sequences collected on the same day.
+This makes polytomies an unescapable reality, and a
+fully-resolved binary tree is clearly false precision in this case.
+At the other extreme, ``unary'' nodes are also a common and
+necessary property of such densely sampled data [explain and
+give numbers?]
+Even with this exquisite level of detail we are
+often unable to fully time-resolve recombination events. While
+in some cases a recombination event with two parental lineages
+can be quite precisely located in space and time (e.g. XA was in Wales
+around [when]), in others three or more quite distinct parental
+lineages indicate multiple unobserved recombinations.
+[JK: I'm just making stuff up here: I'm assuming that there's an
+example of a complex X lineages that people haven't figured the
+exact history of. Just seeing if this makes a good narrative
+before filling in the details.]
+While this
+may be seen as a limitation of the \texttt{sc2ts} method,
+concerted efforts to disentangle the origins of [some example
+of a complicated X lineage] have not succeeded in shedding
+much extra light, and truthfully representing the uncertainty
+around these events seems a more useful approach than
+ arbitrary resolution. Encoding the dataset as an eARG,
+with a specific event for each recombination breakpoint,
+would require such false precision.
+
+An interesting property of the SARS-CoV-2 dataset is that
+it highlights the uncertainty in the genomic position of
+recombination breakpoints. [Some discussion about why
+we can never know where the breakpoint really happened, and
+some follow-up to note that this is even more acute in
+standard popgen datasets where we're thinking about ancient
+recombs].
+This is a limitation of the
+gARG encoding proposed here, which requires that we specify
+exact intervals of inheritance and therefore precise
+recombination breakpoints.
+An approach in which we can systematically reason about the
+uncertainty of inheritance intervals as they propogate
+through an ARG
+would be an interesting avenue for future research.
+
+%%%% Some old text about the coalescent we might still want to plunder
 % % If we want to use the coalescent, then really the assumptions
@@
 
-% While the coalescent as a model can be suprisingly robust to
-% basic violations of the underlying assumptions~\citep{
-% wakeley2012gene,bhaskar2014distortion,nelson2020accounting}
-% it is reasonable to question the validity of the model in
-% representing datasets such as UK Biobank.
-% [Also mention somehow that $n\ll N_e$ is only one problem -
-% we're also ignoring the myriad other evolutionary forces.]
-% Such datasets cannot be reasonably be described by
-% the coalescent as a statical model, and so more non-parametric
-% methods like \tsinfer\ and \relate\ would seem more appropriate.
-% [More]
-
-% % Do downstream applications actually make use of such full ARGs?
-% Downstream applications using estimated ARGs to date rarely
-% make use of such complete recombination information.
-% The \relate\ selection statistic~\citep{speidel2019method}
-% computes a $p$-value for selection [FIXME] for each
-% local tree independently by comparing the rates of coalescence
-% at various time-points. For cases like strong positive selection,
-% where we expect there to be a clear signal in the local tree,
-% such tree-by-tree methods can be simple and powerful.
-% In their method
-% for estimating dispersal rates and the locations of genetic
-% ancestors,
-% \citep{osmond2021estimating} downsample trees along the genome
-% so that they can be regarded as approximately independent.
-% The SIA method for detecting selection~\citep{hejase2022deep}
-% encodes local trees as a set of lineage counts at discrete
-% time intervals, and uses these as feature for a
-% type of machine learning algorithm
-% that takes ``temporal'' correlations into account. Thus,
-% while SIA does use some of the information about local tree correlation,
-% clearly much of the detail about recombination
-% events in a ``full'' ARG is lost.
-% The \texttt{tsdate} algorithm (and related naive estimator
-% for ancestral location) uses much more between-tree
-% information~\citep{wohns2022unified}, explicitly using the gARG
-% encoding to reason about ancestral nodes.
-% A major strength of this approach is that it can be used on
-% any gARG, ranging from estimates containing relatively little
-% information about recombination (e.g., as estimated by \relate)
-% to ``full'' ARGs containing complete recombination information
-% (e.g., as estimated by \kwarg), and will fully utilise
-% any between-tree correlation information that is present.
-
-\section*{Computational efficiency and data interchange}
-ARG inference methods have been successfully applied to vast datasets,
-including 500,000 genotyped humans~\citep{kelleher2019inferring,zhang2023biobank}
-from UK Biobank~\citep{bycroft2018genome}
-and over a million SARS-Cov-2 whole genomes~\citep{zhan2023towards}.
-Even larger datasets are
-available~\citep[e.g.][]{halldorsson2022sequences} % TODO more
-and it seems inevitable that ARGs will be inferred for them.
-The efficiency with which we can store and analyse these inferred
-ARGs is therefore of great importantance, and a key factor
-in the ultimate success of downstream ARG-based methods.
-Another key factor in how important ARG-based methods become
-across the wide range of predicted application
-% TODO more?
-areas~\citep{rasmussen2014genome,harris2019database,
-hejase2020summary,schaefer2021ancestral,
-fan2022genealogical,wohns2022unified,harris2023using,
-nowbandegani2023extremely}
-is how well inference methods and downstream analysis applications
-interoperate. Without a well-defined and efficient shared format
-for ARG data interchange,
-users will suffer from problems of converting data from one programs
-output to another, an unfortunate hallmark of population genetics
-software for many years~\citep{excoffier2006computer}.
-While current methods tend to tightly couple downstream analysis
-of the inferred ARG with the inference
-itself within the same software package
-% FIXME-  more explicit?
-(e.g.\ Relate, ARGneedle)
-this approach scales poorly in terms of software development
-effort, and is ultimately not compatible with the widespread use
-of ARGs for routine data analysis, and a healthy and diverse
-software ecosystem.
-% FIXME not sure where this goes
-Earlier efforts to standardise ARG interchange
-did not succeed~\citep{cardona2008extended,mcgill2013graphml}.
-
-The gARG encoding discussed and concretely defined in this manuscript
-leads to highly efficient storage and processing of ARG data,
-and has already been in use for several years.
-The succinct tree sequence data structure (usually known as a ``tree sequence''
-for brevity)
-is a concrete encoding of a gARG, as discussed in this article.
-It was originally developed as part of the \texttt{msprime}
-simulator~\citep{kelleher2016efficient} and subsequently been
-extended and applied to forward-time
-simulations~\citep{kelleher2018efficient,haller2018tree},
-calculation of population genetics statistics~\citep{ralph2020efficiently}.
-The succinct tree sequence encoding extends the basic definition
-of a gARG provided here by stipulating a
-simple tabular representation and nodes and edges,
-and also defining a concise and lossless representation of
-sequence variation using the ``site'' and  ``mutation'' tables.
-The \texttt{tskit} library~\citep{tskit2023tskit} is a liberally
-licensed open source toolkit that provides a comprehensive suite
-of tools for working with gARGs (encoded as a succinct tree sequence).
-Based on core functionality written
-in C, it provides interfaces in C, Python and Rust.
-Tskit is mature software, widely used in population genetics, and
-has been incorporated into several downstream
-applications~\citep[e.g.,][]{haller2019slim,speidel2019method,
-adrion2020community,
-terasaki2021geonomics,
-baumdicker2021efficient,
-fan2022genealogical,korfmann2022weak,
-mahmoudi2022bayesian,petr2022slendr,rasmussen2022espalier,
-zhang2023biobank,nowbandegani2023extremely}.
-
-% FIXME this needs more work and doesn't really fit into the
-% narrative right now, but we do need to make this point.
-The key insight that makes the succinct tree sequence encoding
-an efficient substrate for defining analysis algorithms is that
-it allows us to generate the local trees along the genome
-in a way that allows us to reason about the \emph{differences}
-between those trees.
-Sequentially generating the local trees along the genome
-is also fundamental, and is necessary whenever we need to
-perform calculations that are contingent on more than just the
-isolated properties of an edge. \cite{kelleher2016efficient}
-showed how all trees can be sequentially generated in
-constant time per tree transition in a fully simplified gARG.
-Furthermore, we can easily reason about how tree topologies
-change (and stay the same), leading to efficient algorithms
-for computing population genetic
-statistics~\citep{kelleher2016efficient,ralph2020efficiently},
-implementing the Li and Stephens
-model~\citep{kelleher2019inferring,wohns2022unified}
-and likely many more.
-
-\section*{SARS-CoV-2 ARGs: a case study}
-% What is SARS-CoV-2 and why is it interesting here? What's the dataset?
-% What are its basic evolutionary properties, summarised for a
-% population geneticist?
-Recombination via template switching is an important
-factor in the evolution of
-SARS-CoV-2~\citep{vaninsberghe2021recombinant,jackson2021generation,ignatieva2022ongoing}.
-With multiple deeply divergent lineages circulating globally,
-coinfection (and the potential for recombination)
-is relatively likely~\citep{bal2022detection}.
-The COVID-19 pandemic led to collection of whole-genome viral data
-at unprecented scale, with the GISAID database~\citep{shu2017gisaid}
-currently holding over 15 million sequences [FIXME]. Detection
-and characterisation of recombination is of profound interest
-for [reasons XYZ].
-
-% Summary of sc2ts paper. We inferred some big ole ARGs, with an
-% unprecedented degree of detail about recombination (outside, I guess,
-% of agri datasets). Summary of what's previously known about sc2 recomb.
-\citet{zhan2023towards} inferred two large ARGs for SARS-CoV-2 based
-on GISAID data, one containing 1.27 million sequences sampled up to
-June 30, 2021 (the ``Wide ARG'', and the second
-(more sparsely sampled) consisting of 657K sequences sampled up to June 30,
-2022 (the ``Long ARG''). Recombination is rare in both ARGs, with $<0.3\%$
-of the nodes being inferred to be recombinants (and quite likely
-a substantial number of false positives caused by poor data quality).
-Nonetheless, this provides thousands of automatically detected
-recombination events, the characterisation of which is usually a manual
-and labour intensive process. The recombinations events detected
-and fully characterised in these large ARGs fit well with previous
-which exhaustively investigate specific
-recombinants~\citep{jackson2021generation}.
-[Not sure what I'm trying to say here;
-what does a PopGen person who knows very little about SARS2 need to
-know here to given them context?]
-
-% What's the point of this case study? To emphasise that we
-% need to systematically represent topological uncertainty in ARGs.
-The pandemic scale SARS-CoV-2 dataset is an interesting case study
-for a number of reasons.
-Firstly, fundamental properties of the data underscore the
-need for topological uncertainty through polytomies
-and stacked recombination [more concise way of saying this?].
-The scale of sequencing effort at the height
-of the pandemic means that there were at times tens of thousands
-of sequences collected in a single day. This, combined with
-the slow rate of mutation, means that we will often have hundreds
-of identical sequences collected on the same day.
-This makes polytomies an unescapable reality, and a
-fully-resolved binary tree is clearly false precision in this case.
-At the other extreme, ``unary'' nodes are also a common and
-necessary property of such densely sampled data [explain and
-give numbers?]
-Even with this exquisite level of detail we are
-often unable to fully time-resolve recombination events. While
-in some cases a recombination event with two parental lineages
-can be quite precisely located in space and time (e.g. XA was in Wales
-around [when]), in others three or more quite distinct parental
-lineages indicate multiple unobserved recombinations.
-[JK: I'm just making stuff up here: I'm assuming that there's an
-example of a complex X lineages that people haven't figured the
-exact history of. Just seeing if this makes a good narrative
-before filling in the details.]
-While this
-may be seen as a limitation of the \texttt{sc2ts} method,
-concerted efforts to disentangle the origins of [some example
-of a complicated X lineage] have not succeeded in shedding
-much extra light, and truthfully representing the uncertainty
-around these events seems a more useful approach than
- arbitrary resolution. Encoding the dataset as an eARG,
-with a specific event for each recombination breakpoint,
-would require such false precision.
-
-An interesting property of the SARS-CoV-2 dataset is that
-it highlights the uncertainty in the genomic position of
-recombination breakpoints. [Some discussion about why
-we can never know where the breakpoint really happened, and
-some follow-up to note that this is even more acute in
-standard popgen datasets where we're thinking about ancient
-recombs].
-This is a limitation of the
-gARG encoding proposed here, which requires that we specify
-exact intervals of inheritance and therefore precise
-recombination breakpoints.
-An approach in which we can systematically reason about the
-uncertainty of inheritance intervals as they propogate
-through an ARG
-would be an interesting avenue for future research.
-
 \section*{Discussion}
@@
 has a long history [summarise this in a couple of sentences?]
+[Cite: \citep{myers2002detection,hayman2023recoverability}
+% This might be where we mention Fisher's junctions and IBD?
+% https://github.com/tskit-dev/what-is-an-arg-paper/issues/121
 \cite{shipilina2023origin} discuss the idea of a ``haplotype block''
@@
 % multiple nodes (or ``blocks'' \citep{palamara2016argon}), with implications for
-% efficiency of storage.
 

### 2023-09-22 paper.tex
Edits from Jere, Yan and Jerome

--- 2023-09-21 paper.tex
@@
 structure~\citep{minichiello2006mapping}, i.e.\ describing
-the \emph{outcome} of such a random process or an inference
+the outcome of such a random process or an inference
 of the genetic genealogy of a sample of genome sequences.
-However, the distinction between process and structure is not clear cut
+However, the distinction between process and data structure is not clear cut
 and subfields use the term
@@
 kelleher2019inferring, speidel2019method, baumdicker2021efficient, zhang2023biobank}.
-However, this broad usage means there is disagreement on the
+This broad usage means there is disagreement on the
 exact object being defined. This causes problems, for example,
-not just when evaluating inferred ARGs, but even when defining
+when evaluating inferred ARGs, and even when defining
 what the criteria for success should be.
@@
 providing a precise definition of an ARG that is general enough to encompass
-the output of all popular simulation and inference methods.
+the output of popular simulation and inference methods.
 We propose a definition of genetic genealogy in terms of specific genomes
@@
 (or not reachable from the samples)
-in the local tree at position $x$, $\pi_u = -1$ (by convention).
+in the local tree at position $x$, then $\pi_u = -1$ (by convention).
 Technically, this structure is a ``forest'', as it allows us to have
@@
 multiple simultaneous recombinations
-(e.g.\ Figure~\ref{fig-arg-in-pedigree}),
+(e.g.\ Fig.~\ref{fig-arg-in-pedigree}),
 and so on can be directly expressed.
@@
 its outcomes in terms of inheritance intervals among parental genomes;
-thus we will have a child node (the recombinant; e.g. node \noderef{a}
-Figure~\ref{fig-arg-in-pedigree}) and two (or more; see section XXX)
+thus we will have a child node (the recombinant; e.g.\ node \noderef{a}
+Fig.~\ref{fig-arg-in-pedigree}) and two (or more; see section XXX)
 paternal genomes from which this child has inherited segments.
@@
 describing the details of genetic inheritance
-between specific pairs of genomes.
+between specific ancestor and descendant genomes.
 In contrast, ARGs are classically seem defined
@@
 a recombination or coalescence event.''
-In contrast ~\citet{gusfield2014recombinatorics} defines an ARG as [TODO]
-
+In contrast, ~\citet{gusfield2014recombinatorics} defines an ARG as [TODO]
+x
 % Papers that do not discuss how positions work:
@@
 the edges of the ARG.
-This is assumed to follow the mechanism described by
+This is usually assumed to follow the mechanism described by
 Griffiths and colleagues in which we have
@@
 the corresponding crossover breakpoint~\citep{griffiths1996ancestral}.
-Figure~\ref{fig-arg-data-structure} shows an example of a classical
+Fig.~\ref{fig-arg-data-structure} shows an example of a classical
 eARG with three samples (\noderef{a}, \noderef{b}, and \noderef{c})
@@
 left of $x$ from \noderef{e} and to the right of $x$ from \noderef{f}.
-As shown in Figure~\ref{fig-arg-data-structure}B,
+As shown in Fig.~\ref{fig-arg-data-structure}B,
 the local trees are embedded in the graph, and can be reconstructed
@@
 sequences.
-For example, Figure~\ref{fig-arg-in-pedigree}
+For example, Fig.~\ref{fig-arg-in-pedigree}
 is retrospective because it only includes information
@@
 The redundancy that we are interested in
-revolves around the presence
-``unary'' nodes in the local trees
+revolves around the presence of nodes which appear
+``unary'' in one or more local trees
 (where a node has exactly one child over a particular genomic region).
@@
 [FIXME this is clumsy; should mention somewhere, though.]
-\citep{mathieson2020ancestry} show examples in which the ARGs
+\citet{mathieson2020ancestry} show examples in which the ARGs
 contain unary nodes, and state that paths are ``implicitly passing through points representing specific
@@
 % the same set of recombination and common ancestor events,
-As illustated in Fig.~\ref{fig-simplification}, successive levels
+As illustrated in Fig.~\ref{fig-simplification}, successive levels
 of ARG simplification reduce the amount of information about the
@@
 While there is of course a single ``true'' ARG that describes
-the actual history of a sample, in reality there will be astronically
-many ARGs that are compatible with the data, even under
-parsimony conditions or when sampling from a model.
-There are many different levels of granularity at which we
-can reason about recombination, and many different ARGs.
+the actual history of a sample, 
+under parsimony conditions or when sampling from a model there 
+will be astronomically
+many ARGs that are equally likely given the data.
+There are also many different levels of granularity at which we
+can reason about recombination, 
+and consequently many different types of ARG.
 We should embrace this diversity.

### 2023-09-22 paper.tex
Change I -> D

@@
 overly narrow interpretation of ARGs, which is significantly
-out of step with modern methods, and
+out of step with modern methods and large-scale datasets, and
 can lead to missed opportunities in their application and usage.
@@
 inherits from one parent (i.e.\ haploids have one genome, diploids two, etc.).
-In this usage a ``genome'' is equivalent to the definition of a
-``haplotype'' used by~\cite{shipilina2023origin}, which we avoid because
-it already has many interpretations, and in particular carries the implication of
-specific mutational patterns [CITES] that persist over
-many generations.
+% Talking about "haplotypes" probably just confuses more
+% In this usage a ``genome'' is equivalent to the definition of a
+% ``haplotype'' used by~\cite{shipilina2023origin}, which we avoid because
+% it already has many interpretations, and in particular carries the implication of
+% specific mutational patterns [CITES] that persist over
+% many generations.
 We are not concerned here with mutational processes or observed sequences,
-% FIXME topological isn't the right word...
 but consider only processes of inheritance,
@@
 % it'll be clearer what we need to say in the caption vs the text.
+[FIXME Update fig to make wider and less tall, and change $I$ to $D$.
+Also possibly add some tabular encoding of gARG data.]
 An example genome ARG embedded in a pedigree.
 (A) Diploid individuals, visualised in a highly inbred pedigree and
-labelled $I_1$ to $I_8$, contain both paternal and maternal  genomes
+labelled $D_1$ to $D_8$,
+contain both paternal and maternal  genomes
 labelled \textsf{a} to \textsf{p}. Black lines show inheritance paths connecting
@@
 and may be the recombined product of that parent's two genomes.
-For example, individual $I_1$ in Fig.~\ref{fig-arg-in-pedigree}
+For example, diploid individual $D_1$ in Fig.~\ref{fig-arg-in-pedigree}
 has two genomes \noderef{a} and \noderef{b},
-inherited from parents $I_3$ and $I_4$.
-Genome \noderef{b} was inherited directly from $I_4$'s genome \noderef{g} without
+inherited from parents $D_3$ and $D_4$.
+Genome \noderef{b} was inherited directly from $D_4$'s genome \noderef{g} without
 recombination, whereas
 \noderef{a} is the product of
-recombining $I_3$'s two genomes \noderef{e} and \noderef{f} at position 2.
+recombining $D_3$'s two genomes \noderef{e} and \noderef{f} at position 2.
 This recombination at means \noderef{a} inherits the (half-closed) genome
@@
 about genetic inheritance relative to the sample individuals
-$I_1$ and $I_2$ (genomes \textsf{a}--\textsf{d}).
+$D_1$ and $D_2$ (genomes \textsf{a}--\textsf{d}).
 Genetic transmission must have occurred between (e.g.)
-$I_8$ and her daughter $I_6$, but this is not recorded because
+$D_8$ and her daughter $D_6$, but this is not recorded because
 % this is probably not true, given we know the complete pedigree?
@@
 ``the phylogeny'')
-% TODO add more if we keep this
+% TODO add more if we keep this. Anyone who says "the ARG is"
 as favoured by some~\citep[e.g.][]{

### 2023-09-22 paper.tex
Apply Gregors comments

@@
 recombination, whereas
-\noderef{a} is the product of
-recombining $D_3$'s two genomes \noderef{e} and \noderef{f} at position 2.
-This recombination at means \noderef{a} inherits the (half-closed) genome
-interval $[0, 2)$ from \noderef{e} and it inherits $[2, 10)$ from \noderef{f};
-these intervals are shown attached to the corresponding graph edges.
+\noderef{a} is the recombinant product of
+$D_2$'s two genomes \noderef{e} and \noderef{f} crossing over at position 2.
+This recombination means \noderef{a} inherits the (half-closed) genome
+interval $[0, 2)$ from \noderef{e} and $[2, 10)$ from \noderef{f}.
+These intervals are shown attached to the corresponding graph edges.
 
@@
 % at some position $x$ along the genome.
+[NOTE: we could move this to the later section currently
+called "the equivalence of ARGS and the local trees",and
+refocus a bit to "ARGs and local trees".]
 The relationship between an ARG and the corresponding local trees
-is somewhat subtle (see section XXX), and it is useful to have
+is subtle (see section XXX), and it is useful to have
 a concrete definition and explicit process for reconstruction.
@@
 in the local tree at position $x$, then $\pi_u = -1$ (by convention).
-Technically, this structure is a ``forest'', as it allows us to have
-multiple disconnected trees, which occurs when all samples have not
-found a most recent common ancestor. Oriented trees also allow nodes
-to have  any number of children from one (see section XXX for more
+% JK: this confused some people, so best left out?
+% Technically, this structure is a ``forest'', as it allows us to have
+% multiple disconnected trees, which occurs when all samples have not
+% found a most recent common ancestor.
+Oriented trees allow nodes
+to have  any number of children; from one (see section XXX for more
 discussion on such ``unary nodes'' in the local trees), to arbitrarily
@@
 basic unit of analysis an individual's genomes,
-any pattern of inheritance among mixed ploidy can be accommodated.
+any pattern of inheritance among mixed ploidy (e.g.\ haplodiploid
+species) can be accommodated.
 Secondly, because we record the intervals of inheritance
@@
 nodes with a single child (see Section XXX).
+% TODO should rephrase this from parent/child to ancestor/descendant?
 Fourth, recombination is modelled in terms of
@@
 Thus, the gARG encoding is focused on faithfully representing the
-\emph{outcome} of the evolutionary processes of genetic inheritance and
-recombination, and is agnostic to the details of those processes.
+\emph{outcome} of the evolutionary processes of genetic inheritance
+with recombination,
+and is agnostic to the details of those processes.
 % The paragraph above is a bit weak and muddled, needs a few more passes.
@@
 by specifying recombination events and breakpoints.
+% FIXME this sentence is problematic. We should recast to talk about
+% the outcomes of events, not talk about simulateousness (which will
+% confuse people
 Thus, we have the means of encoding the effects of
@@
 $D_1$ and $D_2$ (genomes \textsf{a}--\textsf{d}).
-Genetic transmission must have occurred between (e.g.)
+% Gregor & Jerome think that "inheritance" is clearer here than "transmission"
+Genetic inheritance must have occurred between (e.g.)
 $D_8$ and her daughter $D_6$, but this is not recorded because
@@
 to any given set of samples. We are simply recording the
-direct genetic inheritance between two genomes
+genetic inheritance between two genomes
 (in the simplest case, between a parent and a child).
@@
 Ancestral material is a critical concept in a retrospective ARG,
-and refers to the genomic intervals directly ancestral to the samples
+and refers to the genomic intervals ancestral to the samples
 scattered across lineages (the edges of the graph)
@@
 In Fig.~\ref{fig-simplification}A the local trees (right)
-contain many unary nodes, and as we successively simplify
+contain many unary nodes, and as we successively simplify,
 the local trees contain fewer
@@
 it does not correspond to any coalescence in the
-local trees to the left of position $44$, and is a therefore a
-unary node in these three trees.
+local trees to the left of position $44$, and is therefore
+unary in these three trees.
 
@@
 trees~\citep[e.g.][]{griffiths1996ancestral,minichiello2006mapping} [more]
-which we avoid due the potential confusion with statistical
-marginal distrubtions [FIXME Gregor, what's the right wording here?]
+which we avoid due the potential confusion with the
+statistical concept of a marginal distributions indicating
+an integrated tree over many positions.
 
@@
 (somewhat muddled)
-term for the sequence of local trees along the genome. We can see
-that their statement is true, \emph{if} the local trees contain
+term for the sequence of local trees along the genome.
+Their statement is true, \emph{if} the local trees are
+unlabelled and contain
 no unary nodes. However, as we have shown, if the local trees
-contain labelled unary nodes, then  there is precisely as
+contain labelled internal (possibly unary) nodes,
+then  there is precisely as
 much information in them as an ARG. Whether such unary nodes
@@
 simulations~\citep{kelleher2018efficient,haller2018tree},
-calculation of population genetics statistics~\citep{ralph2020efficiently}.
+inference from data~\citep{kelleher2019inferring,wohns2022unified},
+and calculation of population genetics statistics~\citep{ralph2020efficiently}.
 The succinct tree sequence encoding extends the basic definition
@@
 Finally, the COVID-19 pandemic has provided an example of a dataset that
-is clearly a very poor fit for the coalescent with recombination, but which,
+is clearly a very poor fit for the coalescent with recombination,
+but for which,
 nevertheless, we would like to estimate ARGs.

### 2023-09-22 paper.tex
New local trees section

@@
 These intervals are shown attached to the corresponding graph edges.
-
-% The most fundamental operation for
-% an ARG data structure is to recover the embedded local tree
-% at some position $x$ along the genome.
-[NOTE: we could move this to the later section currently
-called "the equivalence of ARGS and the local trees",and
-refocus a bit to "ARGs and local trees".]
-The relationship between an ARG and the corresponding local trees
-is subtle (see section XXX), and it is useful to have
-a concrete definition and explicit process for reconstruction.
-We can describe a local tree as an
-oriented tree~\cite[p.\ 461]{knuth11combinatorial},
-a sequence of integers $\pi_1\pi_2\dots$ where $\pi_u$ denotes the parent of
-node $u$~\citep{kelleher2013coalescent,kelleher2014coalescent,kelleher2016efficient}.
-If $u$ is a root
-% I'm not sure if this reachability stuff is a helpful clarification, and whether it'll just
-% confuse people. It depends on some tricky details.
-(or not reachable from the samples)
-in the local tree at position $x$, then $\pi_u = -1$ (by convention).
-% JK: this confused some people, so best left out?
-% Technically, this structure is a ``forest'', as it allows us to have
-% multiple disconnected trees, which occurs when all samples have not
-% found a most recent common ancestor.
-Oriented trees allow nodes
-to have  any number of children; from one (see section XXX for more
-discussion on such ``unary nodes'' in the local trees), to arbitrarily
-large polytomies.
-To recover the local tree $\pi$ at position $x$, we begin by
-setting $\pi_u = -1$ for each $u\in N$. Then, for each sample
-node in $S$ we trace its path rootwards through the
-ARG for position $x$, and record this path in $\pi$.
-Specifically, at a given node $u$,
-we an find edge $(c, p, I) \in E$ such that $u = c$ and $x \in I$, and set
-$\pi_c \leftarrow p$. We then set $u \leftarrow p$, and repeat
-until either $\pi_u \neq -1$ (indicating we have traversed this section
-of the ARG already on the path from another sample) or there
-is no matching outbound edge (indicating we are at a root).
 
@@
 
+\section*{ARGs and local trees}
+Local trees, describing the genetic genealogy of a sample
+at a particular location along the genome, are said to be
+``embedded'' within in ARG [CITES], and the relationship
+between the two is subtle and important.
+% An ARG is sometimes referred to as being equivalent to a set of
+% local trees along the genome [TODO citation], but the connection
+% between the two is subtle.
+% In order for the mapping to be one-to-one, so that the original
+% gARG can be exactly recovered from a set of local trees and
+% their genome coordinates,
+% some specific properties of the ARG and the trees are required.
+Note that many authors use the term ``marginal''
+trees~\citep[e.g.][]{griffiths1996ancestral,minichiello2006mapping} [more]
+which we avoid due the potential confusion with the
+statistical concept of a marginal distributions indicating
+an integrated tree over many positions.
+
+It is useful to have
+a concrete definition and explicit process for reconstruction.
+We can describe a local tree as an
+oriented tree~\cite[p.\ 461]{knuth11combinatorial},
+a sequence of integers $\pi_1\pi_2\dots$ where $\pi_u$ denotes the parent of
+node $u$~\citep{kelleher2013coalescent,kelleher2014coalescent,kelleher2016efficient}.
+If $u$ is a root
+% I'm not sure if this reachability stuff is a helpful clarification, and whether it'll just
+% confuse people. It depends on some tricky details.
+(or not reachable from the samples)
+in the local tree at position $x$, then $\pi_u = -1$ (by convention).
+% JK: this confused some people, so best left out?
+% Technically, this structure is a ``forest'', as it allows us to have
+% multiple disconnected trees, which occurs when all samples have not
+% found a most recent common ancestor.
+Oriented trees allow nodes
+to have  any number of children; from one (see section XXX for more
+discussion on such ``unary nodes'' in the local trees), to arbitrarily
+large polytomies.
+To recover the local tree $\pi$ at position $x$, we begin by
+setting $\pi_u = -1$ for each $u\in N$. Then, for each sample
+node in $S$ we trace its path rootwards through the
+ARG for position $x$, and record this path in $\pi$.
+Specifically, at a given node $u$,
+we an find edge $(c, p, I) \in E$ such that $u = c$ and $x \in I$, and set
+$\pi_c \leftarrow p$. We then set $u \leftarrow p$, and repeat
+until either $\pi_u \neq -1$ (indicating we have traversed this section
+of the ARG already on the path from another sample) or there
+is no matching outbound edge (indicating we are at a root).
+
+The nodes in a local tree encoded
+in this way are \emph{labelled} (i.e.\ carry the identity
+of the corresponding ARG node), and this is fundamentally
+important to how a local tree relates to its parent ARG.
+If we are given the sequence of local trees for a gARG
+encoded as an oriented tree
+and along with the genome interval covered by each tree,
+then we can recover precisely the same ARG. Thus, there
+is a one-to-one correspondance between an ARG and
+the sequence of local trees that it encodes.
+
+The internal nodes of trees, however, are usually
+seen as \emph{unlabelled}, since they most often correspond
+to anonymous hypothetical ancestors. Internal
+ARG nodes likewise correspond to anonymous hypothetical
+ancestors, but if we do not label them we lose the fact that
+the \emph{same} ancestors persist across multiple trees.
+Without labelled internal tree nodes,
+we must infer their identity
+from the tree topology, which is both
+expensive to compute and imprecise. For example
+we could infer that node \textsf{k} is shared in the second
+and third trees of in Fig.~\ref{fig-simplification}D
+because it subtends identical subtrees in both,
+but it would not be straightforward to infer that it
+is also the root of the first tree.
+
+\cite{shipilina2023origin} discuss the same ideas, and note that
+``full ARG ... contains more information than the series of tree
+sequences along the genome''. They view the ``full ARG'' as an
+eARG, and the ``series of tree sequences'' is their
+% jk: probably should delete this. It *is* muddled though
+(somewhat muddled)
+term for the sequence of local trees along the genome.
+Their statement is true, \emph{if} the local trees are
+unlabelled and contain
+no unary nodes. However, as we have shown, if the local trees
+contain labelled internal (possibly unary) nodes,
+then  there is precisely as
+much information in them as an ARG. Whether such unary nodes
+are \emph{observable} is an entirely separate question,
+of course.
+
+These properties have interesting consequences when we
+consider the inference strategy of estimating local trees
+independently and subsequently stitching them together
+into an ARG.
+Two recent methods take this approach,
+\relate\ and
+\espalier~\citep{rasmussen2022espalier},
+and have different strategies for reconciling the local trees into an ARG.
+\relate\ first infers local tree topologies, and the identifies
+edges that persist from tree to tree by reasoning about the sets of samples
+subtended. It does not attempt to be exhaustive.
+\espalier, on the other hand, is specifically interested
+in the details of recombination events (among small samples)
+and therefore attempts to infer
+the precise subtree prune-and-regraft (SPR)
+operations~\citep{hein1990reconstructing,song2003on,song2006properties}
+induced by recombination.
+Inferring the SPRs between leaf-labelled trees is
+NP-hard~\citep{hein1996complexity,allen2001subtree,bordewich2005computational},
+but it is unclear what the complexity is when there
+is a degree of internal node sharing between trees.
+
+% The equivalence between an ARG and a set of local trees
+% separated by SPRs is often mentioned [e.g. x y z], but
+% it is important to note that while extracting local
+% trees from an ARG is straightforward, the converse
+% process of reconstructing precisely the same ARG from
+% these trees is more problematic. If the internal
+% nodes are consistently labelled and unary nodes are
+% included in the local trees (e.g., Fig~\ref{fig-simplification})
+% then it is easy to exactly reconstruct the ARG.
+% Otherwise, the problem is much more difficult and it
+% may not be possible to uniquely reconstruct the original ARG.
+% Without labels for the internal nodes, we are forced to
+% to reason about which subtrees remain the same and which
+% differ due to the SPR separating adjacent trees in order to
+% assign node identity. Without the internal unary nodes marking
+% recombinations, we must infer their existence. In
+% general, there will not be a unique solution to this problem,
+% and so there can be many possible ARGs compatible with
+% a given list of leaf-labelled binary trees.
+% This is the approach taken by the Espalier inference
+% method~\citep{rasmussen2022espalier}, which finds a
+% set of SPRs that reconcile independently inferred
+% trees from different regions of the genome.
+
 \section*{Levels of simplification}\label{ARG_simplification}
@@
 
-\section*{Equivalance of ARGs and the set of local trees}
-An ARG is sometimes referred to as being equivalent to a set of
-local trees along the genome [TODO citation], but the connection
-between the two is subtle.
-In order for the mapping to be one-to-one, so that the original
-gARG can be exactly recovered from a set of local trees and
-their genome coordinates,
-some specific properties of the ARG and the trees are required.
-Note that many authors use the term ``marginal''
-trees~\citep[e.g.][]{griffiths1996ancestral,minichiello2006mapping} [more]
-which we avoid due the potential confusion with the
-statistical concept of a marginal distributions indicating
-an integrated tree over many positions.
-
-Firstly, the internal nodes in the local trees must be labelled
-(i.e.\ marked with the identity of the corresponding ARG node).
-If the nodes are not labelled, we must infer their identity
-from the tree topology, which is both
-expensive to compute and imprecise. For example
-we could infer that node \textsf{k} is shared in the second
-and third trees of in Fig.~\ref{fig-simplification}D
-because it subtends identical subtrees in both,
-but it would not be straightforward to infer that it
-is also the root of the first tree.
-% JK: Does this need to be said?
-% (One may use the time of a node as a proxy for identity,
-% but it is not reliable as we can have multiple nodes at the
-% same time, e.g. with time discretisation.)
-
-The precision of the information about recombination that
-can be represented in the trees depends the ARG itself.
-Secondly, the local trees must contain ``unary'' nodes.
-[Some discussion].
-With these two conditions met, there is a direct correspondence
-between a gARG and the trees.
-
-\cite{shipilina2023origin} discuss the same ideas, and note that
-``full ARG ... contains more information than the series of tree
-sequences along the genome''. They view the ``full ARG'' as an
-eARG, and the ``series of tree sequences'' is their
-% jk: probably should delete this. It *is* muddled though
-(somewhat muddled)
-term for the sequence of local trees along the genome.
-Their statement is true, \emph{if} the local trees are
-unlabelled and contain
-no unary nodes. However, as we have shown, if the local trees
-contain labelled internal (possibly unary) nodes,
-then  there is precisely as
-much information in them as an ARG. Whether such unary nodes
-are \emph{observable} is an entirely separate question,
-of course.
-
-These properties have interesting consequences when we
-consider the inference strategy of estimating local trees
-independently and subsequently stitching them together
-into an ARG.
-Two recent methods take this approach,
-\relate\ and
-\espalier~\citep{rasmussen2022espalier},
-and have different strategies for reconciling the local trees into an ARG.
-\relate\ first infers local tree topologies, and the identifies
-edges that persist from tree to tree by reasoning about the sets of samples
-subtended. It does not attempt to be exhaustive.
-\espalier, on the other hand, is specifically interested
-in the details of recombination events (among small samples)
-and therefore attempts to infer
-the precise subtree prune-and-regraft (SPR)
-operations~\citep{hein1990reconstructing,song2003on,song2006properties}
-induced by recombination.
-Inferring the SPRs between leaf-labelled trees is
-NP-hard~\citep{hein1996complexity,allen2001subtree,bordewich2005computational},
-but it is unclear what the complexity is when there
-is a degree of internal node sharing between trees.
-
-% The equivalence between an ARG and a set of local trees
-% separated by SPRs is often mentioned [e.g. x y z], but
-% it is important to note that while extracting local
-% trees from an ARG is straightforward, the converse
-% process of reconstructing precisely the same ARG from
-% these trees is more problematic. If the internal
-% nodes are consistently labelled and unary nodes are
-% included in the local trees (e.g., Fig~\ref{fig-simplification})
-% then it is easy to exactly reconstruct the ARG.
-% Otherwise, the problem is much more difficult and it
-% may not be possible to uniquely reconstruct the original ARG.
-% Without labels for the internal nodes, we are forced to
-% to reason about which subtrees remain the same and which
-% differ due to the SPR separating adjacent trees in order to
-% assign node identity. Without the internal unary nodes marking
-% recombinations, we must infer their existence. In
-% general, there will not be a unique solution to this problem,
-% and so there can be many possible ARGs compatible with
-% a given list of leaf-labelled binary trees.
-% This is the approach taken by the Espalier inference
-% method~\citep{rasmussen2022espalier}, which finds a
-% set of SPRs that reconcile independently inferred
-% trees from different regions of the genome.
-
 \section*{Computational efficiency and data interchange}

### 2023-10-05 paper.tex
Draft of first half of disco

@@
 enable this.
-
-A necessary first step towards this goal is some terminological
-clarity. As discussed in Appendix XXX, the term Ancestral Recombination
+A necessary first step is a degree of  terminological clarity.
+As discussed in Appendix XXX, the term Ancestral Recombination
 Graph has several
@@
 mathematical model, is more limited in what can be described.
+
+Fully decoupling the general concept of an ARG from the coalescent
+with recombination stochastic process is an important step.
+While it has proven to be a useful and
+robust
+model~\citep{wakeley2012gene,bhaskar2014distortion,nelson2020accounting},
+many modern datasets have properties that grossly
+violate its assumptions.
+For example, several human datasets now consist of hundreds of thousands of
+% TODO cite GeL paper
+genomes~\citep{bycroft2018genome,karczewski2020mutational,tanjo2021practical},
+and so sample size is substantially \emph{larger} than the
+usually assumed $N_e$ values.
+Agricultural datasets are an even more extreme departure from coalescent
+assumptions, with hundreds of thousands of samples embedded in
+multi-generational pedigrees~\citep{hayes20191000,Ros-Freixedes2020}
+and effective population sizes of 100 and even
+less~\citep{MacLeod2013,Makanjuola2020,Hall2016,Porcnic2016}.
+Even if sufficiently complex demographic models~\citep{gower2022demes}
+encompassing hundreds of populations, explosive growth rates and
+myriad interconnections of migration, were somehow estimated and
+provided as input, ARGs sampled from
+the coalescent cannot capture complexities of family structure which
+make, for example, polytomies a necessary feature of local trees.
+
+Given that there is no single model that can capture the complexities
+of modern datasets across species (and taxa; see discussion of SARS-CoV-2
+below), we cannot rely on sampling from a distribution to capture
+the uncertainty in our ARG estimates. Indeed, even if the coalescent
+were a suitable model for datasets such as UKB and we had a means of
+sampling from its distribution of ARGs at the scale of hundreds of thousands
+of samples, we could only explore the tiniest corner such an  incompehensibly
+large space of possibilities.
+It is therefore important to systematically
+describe and utilise uncertainty about ARG inference.
+One approach, enabled by the gARG encoding described here,
+is to use polytomies and stacked recombinations as direct topological
+descriptions of uncertainty. Other methods to capture, for example,
+uncertainty about node ages and breakpoint location, is an important
+aspect of future work.
 
@@
 
-Fully decoupling the general concept of an ARG from the coalescent
-with recombination stochastic process is an important step.
-While it has proven to be a useful and
-robust model~\citep{wakeley2012gene,bhaskar2014distortion,nelson2020accounting}
-many modern datasets have properties that grossly
-violate its assumptions.
-For example, several human datasets now consist of hundreds of thousands of
-genomes~\citep{bycroft2018genome,karczewski2020mutational,tanjo2021practical},
-meaning that sample size is substantially larger than the
-usually assumed $N_e$ values.
-Agricultural datasets are an even more extreme departure from coalescent
-assumptions, with hundreds of thousands of samples embedded in
-multi-generational pedigrees~\citep{hayes20191000,Ros-Freixedes2020}
-and effective population sizes of 100 and even
-less~\citep{MacLeod2013,Makanjuola2020,Hall2016,Porcnic2016}.
 Finally, the COVID-19 pandemic has provided an example of a dataset that

### 2023-10-05 paper.tex
Further notes on uncertainty

@@
 
-The necessity of complete precision about all recombination
-events under the eARG encoding has been
-conflated with the idea of a ``true'' or ``full'' ARG. [Cites]
-For example, \citet{brandt2021evaluation} define a ``full'' ARG as
-``a structure that encodes all
-coalescence and recombination events''. While there are certainly cases
-when inferring such fully precise ARGs is useful, for example when
-studying the details of a small number of recombination
-events~\citep{rasmussen2022espalier}, it is worth considering
-whether sufficient information exists to support such inferences
-and indeed whether this precision is required, or useful,
-for downstream applications.
-
-The amount of information about recombination that is inferrable
-is the subject of much research
-has a long history [summarise this in a couple of sentences?]
-[Cite: \citep{myers2002detection,hayman2023recoverability}
+Uncertainty is a central element of inference about recombination,
+and, even with small samples, we can only expect to uniquely identify
+lineages and breakpoints in special cases.
+[summarise info inferable about recomb in few sentenances here?
+Cite: \citep{myers2002detection,hayman2023recoverability}]
 % This might be where we mention Fisher's junctions and IBD?
 % https://github.com/tskit-dev/what-is-an-arg-paper/issues/121
-\cite{shipilina2023origin} discuss the idea of a ``haplotype block''
-(equivalent to the ``bricked tree sequence'' of
-\cite{nowbandegani2023extremely},
-% TODO look this up. I expect it's basically the same??
-and the block-based ARG data structure of
-\citep{palamara2016argon}),
-here we consider the unique sets of
-samples that coalesce at a node in the ARG over a particular
-genome interval, and the fundamental limitations they place
-on ARG inference.
+The fact that the classical ``event'' ARG (eARG) encoding
+\emph{requires} completely precise estimates about recombination
+is therefore a fundamental limitation, and a central message
+of this paper.
+
+% The necessity of complete precision about all recombination
+% events under the eARG encoding has been
+% conflated with the idea of a ``true'' or ``full'' ARG. [Cites]
+% For example, \citet{brandt2021evaluation} define a ``full'' ARG as
+% ``a structure that encodes all
+% coalescence and recombination events''. While there are certainly cases
+% when inferring such fully precise ARGs is useful, for example when
+% studying the details of a small number of recombination
+% events~\citep{rasmussen2022espalier}, it is worth considering
+% whether sufficient information exists to support such inferences
+% and indeed whether this precision is required, or useful,
+% for downstream applications.
+
+% \cite{shipilina2023origin} discuss the idea of a ``haplotype block''
+% (equivalent to the ``bricked tree sequence'' of
+% \cite{nowbandegani2023extremely},
+% % TODO look this up. I expect it's basically the same??
+% and the block-based ARG data structure of
+% \citep{palamara2016argon}),
+% here we consider the unique sets of
+% samples that coalesce at a node in the ARG over a particular
+% genome interval, and the fundamental limitations they place
+% on ARG inference.
 
@@
 % % Do downstream applications actually make use of such full ARGs?
-Besides the inherent limitations that exist on inferring fully precise ARGs
-from data,
+Besides the inherent limitations that exist on inferring fully
+precise ARGs from data,
 we should also consider the value that such precise estimates provide
@@
 To date, few downstream applications using estimated ARGs
-make use of such complete recombination information.
-The \relate\ selection statistic~\citep{speidel2019method}
-computes a $p$-value for selection [FIXME] for each
-local tree independently by comparing the rates of coalescence
+make use of complete recombination information.
+Most applications work by examining local trees independently.
+For example, the \relate\ selection statistic~\citep{speidel2019method}
+computes selection $p$-values by comparing the rates of coalescence
 at various time-points.
@@
 clearly much of the detail about recombination
-events in a ``full'' ARG is lost.
+events in an ARG is lost.
 The \texttt{tsdate} algorithm (and related naive estimator
@@
 encoding to reason about ancestral nodes.
-[MORE applications]
 The main application for fully precise ARGs thus far has been

### 2023-10-05 paper.tex
Drop pedantic "the ARG" ending

@@
 
-[NOTE This is a pedantic and slightly bullshitty to end up on,
-just writing it down to see if it works. Should probably cut.]
-A final point to note, is that
-if we are using the term ARG
-to loosely refer to genetic genealogies of varying
-levels of detail, either inferred or simulated,
-it makes little sense to refer ``the'' Ancestral Recombination
-Graph (i.e, in  the singular, like ``the tree'' or
-``the phylogeny'')
-% TODO add more if we keep this. Anyone who says "the ARG is"
-as favoured by some~\citep[e.g.][]{
-arenas2013importance,palamara2016argon,hubisz2020inference}.
-While there is of course a single ``true'' ARG that describes
-the actual history of a sample,
-under parsimony conditions or when sampling from a model there
-will be astronomically
-many ARGs that are equally likely given the data.
-There are also many different levels of granularity at which we
-can reason about recombination,
-and consequently many different types of ARG.
-We should embrace this diversity.
-
 \bibliographystyle{plainnat}

### 2023-10-05 paper.tex

@@
 
-\section*{SARS-CoV-2 ARGs: a case study}
-% What is SARS-CoV-2 and why is it interesting here? What's the dataset?
-% What are its basic evolutionary properties, summarised for a
-% population geneticist?
-Recombination via template switching is an important
-factor in the evolution of
-SARS-CoV-2~\citep{vaninsberghe2021recombinant,jackson2021generation,ignatieva2022ongoing}.
-With multiple deeply divergent lineages circulating globally,
-coinfection (and the potential for recombination)
-is relatively likely~\citep{bal2022detection}.
-The COVID-19 pandemic led to collection of whole-genome viral data
-at unprecented scale, with the GISAID database~\citep{shu2017gisaid}
-currently holding over 15 million sequences [FIXME]. Detection
-and characterisation of recombination is of profound interest
-for [reasons XYZ].
-
-% Summary of sc2ts paper. We inferred some big ole ARGs, with an
-% unprecedented degree of detail about recombination (outside, I guess,
-% of agri datasets). Summary of what's previously known about sc2 recomb.
-\citet{zhan2023towards} inferred two large ARGs for SARS-CoV-2 based
-on GISAID data, one containing 1.27 million sequences sampled up to
-June 30, 2021 (the ``Wide ARG'', and the second
-(more sparsely sampled) consisting of 657K sequences sampled up to June 30,
-2022 (the ``Long ARG''). Recombination is rare in both ARGs, with $<0.3\%$
-of the nodes being inferred to be recombinants (and quite likely
-a substantial number of false positives caused by poor data quality).
-Nonetheless, this provides thousands of automatically detected
-recombination events, the characterisation of which is usually a manual
-and labour intensive process. The recombinations events detected
-and fully characterised in these large ARGs fit well with previous
-which exhaustively investigate specific
-recombinants~\citep{jackson2021generation}.
-[Not sure what I'm trying to say here;
-what does a PopGen person who knows very little about SARS2 need to
-know here to given them context?]
-
-% What's the point of this case study? To emphasise that we
-% need to systematically represent topological uncertainty in ARGs.
-The pandemic scale SARS-CoV-2 dataset is an interesting case study
-for a number of reasons.
-Firstly, fundamental properties of the data underscore the
-need for topological uncertainty through polytomies
-and stacked recombination [more concise way of saying this?].
-The scale of sequencing effort at the height
-of the pandemic means that there were at times tens of thousands
-of sequences collected in a single day. This, combined with
-the slow rate of mutation, means that we will often have hundreds
-of identical sequences collected on the same day.
-This makes polytomies an unescapable reality, and a
-fully-resolved binary tree is clearly false precision in this case.
-At the other extreme, ``unary'' nodes are also a common and
-necessary property of such densely sampled data [explain and
-give numbers?]
-Even with this exquisite level of detail we are
-often unable to fully time-resolve recombination events. While
-in some cases a recombination event with two parental lineages
-can be quite precisely located in space and time (e.g. XA was in Wales
-around [when]), in others three or more quite distinct parental
-lineages indicate multiple unobserved recombinations.
-[JK: I'm just making stuff up here: I'm assuming that there's an
-example of a complex X lineages that people haven't figured the
-exact history of. Just seeing if this makes a good narrative
-before filling in the details.]
-While this
-may be seen as a limitation of the \texttt{sc2ts} method,
-concerted efforts to disentangle the origins of [some example
-of a complicated X lineage] have not succeeded in shedding
-much extra light, and truthfully representing the uncertainty
-around these events seems a more useful approach than
- arbitrary resolution. Encoding the dataset as an eARG,
-with a specific event for each recombination breakpoint,
-would require such false precision.
-
-An interesting property of the SARS-CoV-2 dataset is that
-it highlights the uncertainty in the genomic position of
-recombination breakpoints. [Some discussion about why
-we can never know where the breakpoint really happened, and
-some follow-up to note that this is even more acute in
-standard popgen datasets where we're thinking about ancient
-recombs].
-This is a limitation of the
-gARG encoding proposed here, which requires that we specify
-exact intervals of inheritance and therefore precise
-recombination breakpoints.
-An approach in which we can systematically reason about the
-uncertainty of inheritance intervals as they propogate
-through an ARG
-would be an interesting avenue for future research.
-
 %%%% Some old text about the coalescent we might still want to plunder
@@
 
-% The necessity of complete precision about all recombination
-% events under the eARG encoding has been
-% conflated with the idea of a ``true'' or ``full'' ARG. [Cites]
-% For example, \citet{brandt2021evaluation} define a ``full'' ARG as
-% ``a structure that encodes all
-% coalescence and recombination events''. While there are certainly cases
-% when inferring such fully precise ARGs is useful, for example when
-% studying the details of a small number of recombination
-% events~\citep{rasmussen2022espalier}, it is worth considering
-% whether sufficient information exists to support such inferences
-% and indeed whether this precision is required, or useful,
-% for downstream applications.
-
-% \cite{shipilina2023origin} discuss the idea of a ``haplotype block''
-% (equivalent to the ``bricked tree sequence'' of
-% \cite{nowbandegani2023extremely},
-% % TODO look this up. I expect it's basically the same??
-% and the block-based ARG data structure of
-% \citep{palamara2016argon}),
-% here we consider the unique sets of
-% samples that coalesce at a node in the ARG over a particular
-% genome interval, and the fundamental limitations they place
-% on ARG inference.
-
-% [FIXME find somewhere natural for this. The last clause is too vague - what
-% are the implications?]
-% Alternative gARG encodings split transmitted intervals and associate each with
-% a new node, resulting in a single genome being represented by a group of
-% multiple nodes (or ``blocks'' \citep{palamara2016argon}), with implications for
-
 % % Do downstream applications actually make use of such full ARGs?
@@
 
-Finally, the COVID-19 pandemic has provided an example of a dataset that
-is clearly a very poor fit for the coalescent with recombination,
-but for which,
-nevertheless, we would like to estimate ARGs.
-\cite{zhan2023towards} inferred an ARG for 1.27 million SARS-CoV-2 whole
-genomes, and showed how it accurately captures several known properties of
-the pathogen's evolution, and can efficiently and concisely
-encode this vast dataset.
+% Note: haven't said anything about the coalescent here. Do we need to?
+The advantages of a model-free representation that naturally
+incorporates uncertainty are well illustrated by the
+ARGs inferred for SARS-CoV-2 by~\cite{zhan2023towards}.
+The COVID-19 pandemic led to collection of whole-genome viral data
+at unprecented scale, with the GISAID database~\citep{shu2017gisaid}
+holding over 15 million sequences as of [FIXME].
+Recombination, while rare, is an important
+factor in the evolution of
+SARS-CoV-2~\citep{vaninsberghe2021recombinant,jackson2021generation,ignatieva2022ongoing}
+with recombinants potentially gaining fitness by combining
+mutations from deeply diverged lineages [cite].
+\citet{zhan2023towards} inferred two large ARGs for SARS-CoV-2 based
+on GISAID data, one containing 1.27 million sequences sampled up to
+June 30, 2021 and the other (more sparsely sampled) consisting of
+657K sequences sampled up to June 30, 2022.
+Sequencing effort at the height of the pandemic produced
+tens of thousands of sequences collected in a single day,
+with hundreds being identical.
+Polytomies are therefore and unescapable reality, and a
+fully-resolved binary tree would clearly be false precision.
+Because of the longitudinal data collection, there is no
+requirement that samples must be leaf nodes, and samples
+will frequently descend from other (older) samples.
+Unary nodes (a node in a local tree with exactly one child)
+are also a common feature, when a sample is inferred to have
+exactly one direct descendant.
+Despite the abundance of data and the rarity of recombination,
+the parental lineages cannot be precisely identified
+in many cases. For example, many are inferred to have three
+or more parental lineages (in agreement with the
+analysis resulting in the Pango designation of the recombinant
+lineage); in most cases, these will indicate two or more
+independent recombination events.
+Faithfully representing the uncertainty
+around these events seems a more useful approach than
+the arbitrary resolution which would be required,
+if encoded as an eARG.
+
+% An interesting property of the SARS-CoV-2 dataset is that
+% it highlights the uncertainty in the genomic position of
+% recombination breakpoints. [Some discussion about why
+% we can never know where the breakpoint really happened, and
+% some follow-up to note that this is even more acute in
+% standard popgen datasets where we're thinking about ancient
+% recombs].
+% This is a limitation of the
+% gARG encoding proposed here, which requires that we specify
+% exact intervals of inheritance and therefore precise
+% recombination breakpoints.
+% An approach in which we can systematically reason about the
+% uncertainty of inheritance intervals as they propogate
+% through an ARG
+% would be an interesting avenue for future research.
 
@@
 
+[TODO finish up with a positive note about interchange, and make the point
+clearly that tskit is an ARG library. Splice in the stuff about haplotype
+blocks and ARGON format here somewhere --- basically this is a derived product
+of the gARG encoding, so while there are some computation advantages it's not
+fundamendal to the actual representation of the thing]
+
+% \cite{shipilina2023origin} discuss the idea of a ``haplotype block''
+% (equivalent to the ``bricked tree sequence'' of
+% \cite{nowbandegani2023extremely},
+% % TODO look this up. I expect it's basically the same??
+% and the block-based ARG data structure of
+% \citep{palamara2016argon}),
+% here we consider the unique sets of
+% samples that coalesce at a node in the ARG over a particular
+% genome interval, and the fundamental limitations they place
+% on ARG inference.
+% [FIXME find somewhere natural for this. The last clause is too vague - what
+% are the implications?]
+% Alternative gARG encodings split transmitted intervals and associate each with
+% a new node, resulting in a single genome being represented by a group of
+% multiple nodes (or ``blocks'' \citep{palamara2016argon}), with implications for
+
 \bibliographystyle{plainnat}

### 2023-10-06 paper.tex
Gregor's updates

@@
 tried to clarify and systematise here. Thus
-we can think of an ARG as being any structure that encodes the
+we can think of an ARG as any structure that encodes the
 reticulate genetic ancestry of a sample of colinear sequences under
@@
 The flexibility of the gARG encoding contrasts with the classical
-``event'' ARG (eARG), which, because of its origins in an idealised
-mathematical model, is more limited in what can be described.
+``event'' ARG (eARG), which is more limited in what can be described, because of its origins in an idealised mathematical model.
 
@@
 provided as input, ARGs sampled from
-the coalescent cannot capture complexities of family structure which
-make, for example, polytomies a necessary feature of local trees.
+the coalescent cannot capture the complexities of realistic family structures.
+There are many species with sizeable to large number of progeny, making polytomies a necessary feature of local trees.
+Flexible ARG encoding should accommodate such cases.
 
@@
 the uncertainty in our ARG estimates. Indeed, even if the coalescent
-were a suitable model for datasets such as UKB and we had a means of
+was a suitable model for datasets such as UKB and we had a means of
 sampling from its distribution of ARGs at the scale of hundreds of thousands
-of samples, we could only explore the tiniest corner such an  incompehensibly
+of samples, we could only explore the tiniest corner of such an incompehensibly
 large space of possibilities.
-It is therefore important to systematically
+Therefore, it is important to systematically
 describe and utilise uncertainty about ARG inference.
@@
 descriptions of uncertainty. Other methods to capture, for example,
-uncertainty about node ages and breakpoint location, is an important
+uncertainty about node ages and breakpoint locations, is an important
 aspect of future work.
@@
 For example, the \relate\ selection statistic~\citep{speidel2019method}
-computes selection $p$-values by comparing the rates of coalescence
-at various time-points.
+computes selection $p$-values by comparing the coalescence rates at various time points.
 In their method
@@
 The \texttt{tsdate} algorithm (and related naive estimator
-for ancestral location) uses much more between-tree
+of ancestral location) uses much more between-tree
 information~\citep{wohns2022unified}, explicitly using the gARG
@@
 The main application for fully precise ARGs thus far has been
-to be to compute a likelihood under the coalescent with
+to compute a likelihood under the coalescent with
 recombination~\citep[e.g.][]{kuhner2000maximum,mahmoudi2022bayesian},
@@
 The advantages of a model-free representation that naturally
-incorporates uncertainty are well illustrated by the
+incorporates uncertainty about the ordering of events in and ARG are well illustrated by the
 ARGs inferred for SARS-CoV-2 by~\cite{zhan2023towards}.
 The COVID-19 pandemic led to collection of whole-genome viral data
-at unprecented scale, with the GISAID database~\citep{shu2017gisaid}
-holding over 15 million sequences as of [FIXME].
+at unprecented scale. The GISAID database~\citep{shu2017gisaid}
+holds over 15 million sequences as of [FIXME].
 Recombination, while rare, is an important
@@
 fully-resolved binary tree would clearly be false precision.
-Because of the longitudinal data collection, there is no
+Because of longitudinal data collection, there is no
 requirement that samples must be leaf nodes, and samples
@@
 without the hard requirement
-of complete precision on all historical events, may help
-to clarify inference goals and improve
+of complete precision on all historical events, may clarify inference goals and improve
 methods for evaluation.
-In the majority of cases,
-ARG inference accuracy is evaluated by comparing some truth simulations
+In most cases,
+the accuracy of ARG inference is evaluated by comparing some truth simulations
 to the inferred ARGs by the pairwise comparison of local trees along the genome
@@
 In comparing tree-by-tree along the genome, the effects of recombination
-structure is incorporated in a indirect manner, and the tree
+structure is incorporated in an indirect manner, and the tree
 metrics assume unlabelled nodes, effectively discarding information
 about node sharing between trees.
-Tree distance metrics usually have $O(n^2)$ time complexity, and so cannot be
+Tree distance metrics usually have $O(n^2)$ time complexity and therefore cannot be
 applied to the very large sample sizes currently of interest.
-A recent trend has been to move away from such tree distance
-based approaches, and to examine more properties of the inferred ARGs.
+A recent trend has been to move away from such tree distance-based approaches and to examine more properties of the inferred ARGs.
 \citet{brandt2021evaluation} compared the
@@
 (beyond simplistic error models)
-on modern biobank scale datasets
+on modern biobank scale or multi-generational pedigree datasets
 are almost entirely unknown.
@@
 fundamendal to the actual representation of the thing]
+In summary, we advocate for decoupling between generative models that give rise to observed data and data formats that can faithfully and realistically encode these data.
+We have shown that ``genome'' ARG (gARG) encoding enables flexible encoding of today's datasets with realistic levels of precision.
+Importantly, genome ARG encoding should facilitate easier data interchange between different tools, which will in turn support active research in this area.
 

### 2023-10-06 paper.tex
Tweaks

@@
 of the gARG encoding, so while there are some computation advantages it's not
-fundamendal to the actual representation of the thing]
+fundamental to the actual representation of the thing]
 In summary, we advocate for decoupling between generative models that give rise to observed data and data formats that can faithfully and realistically encode these data.

### 2023-10-06 paper.tex
Make Fig 3 fit on half a page

--- 2023-10-05 paper.tex
@@
 % up far too much vspace at in native coords at the moment.
-\includegraphics[width=0.6\textwidth]{illustrations/ancestry-resolution}
+\includegraphics[width=\textwidth]{illustrations/ancestry-resolution}
 \caption{\label{fig-ancestry-resolution}
-Converting the \citet[][Fig.~1]{wiuf1999recombination} example graph
+Converting the \citet[][Fig.~1]{wiuf1999recombination} example
 to a gARG. (A) The original eARG; square nodes represent events, with
-each recombination (red) showing the corresponding breakpoint.
+each recombination (red) containing a breakpoint.
 (B) The corresponding gARG with breakpoints directly converted to
@@
 Dashed lines show edges that are
-no longer present (in practice, nodes \noderef{g} and \noderef{q} would also be removed).
+no longer present (in practice, nodes \noderef{g}, \noderef{j}, and \noderef{q} would also be removed).
 Coalescence with respect to the sample is indicated by shaded bars, as
@@
 Line thickness is proportional to the genomic span of each edge.
-Nodes representing recombination events have been retained
+Nodes representing recombination events are retained
 for clarity, but could be removed by simplification if

### 2023-10-06 paper.tex
Implement most of Gideon and Alex's comments

--- 2023-10-06 paper.tex
@@
 % TODO say this better.
-From a high-level, an ARG describes the different paths of genetic inheritance
+From a high level, an ARG describes the different paths of genetic inheritance
 caused by recombination, and encodes a sequence of correlated genealogical
@@
 graph~\citep{griffiths1991two,ethier1990two,griffiths1996ancestral,griffiths1997ancestral}.
-Subsequently, ARGs have come to be thought of as a data
+Subsequently, an ARG  has come to be thought of as a data
 structure~\citep{minichiello2006mapping}, i.e.\ describing
@@
 and accuracy of our inferences.
+% Thanks Gideon and Alex!
+If not addressed, these problems will
+impede progress on applying genetic genealogical
+information to understand the biological world.
+
+% Evidence of confusion to cite above in "problem" para?
+% https://github.com/tskit-dev/what-is-an-arg-paper/issues/334
+% The output of these programs is certainly not a collection of
+% \emph{unrelated} local trees, as implied by,
+% e.g.,
+% \citet{hejase2020summary} when they state that \tsinfer\
+% ``does not explicitly infer an ARG but rather a sequence of local
+% gene trees''.
+% More cites? Not worth harping on about now?
+% https://github.com/tskit-dev/what-is-an-arg-paper/issues/38
 
@@
 multiple simultaneous recombinations
-(e.g.\ Fig.~\ref{fig-arg-in-pedigree}),
 and so on can be directly expressed.
@@
 Thirdly, local trees that we generate for a particular position on the
-genome can have nodes of any arity, including ``unary''
+genome can have nodes of any arity (number of child nodes), including ``unary''
 nodes with a single child (see Section XXX).
@@
 between specific ancestor and descendant genomes.
-In contrast, ARGs are classically seem defined
+In contrast, ARGs are classically defined
 by the evolutionary \emph{events} in the history of a sample.
 The difference between gARGs and this prevalent view of
-what defines an ARG is subtle and the distinctions sometimes porous.
+what defines an ARG is subtle and difficult to distinguish.
 It is therefore useful to first survey some representative
-definitions of
-an ARG from the literature, to allow us to capture the common features
-of these definitions
+recent definitions of
+an ARG from the literature, to allow us to capture their common features
 (see Appendix XXX for a full review).
@@
 the coalescent with recombination.''
-\citet{shipilina2023origin} emphasise the true existance of the
+\citet{shipilina2023origin} emphasise the true existence of the
 corresponding events (rather than the generative process) by saying that
@@
 recombination events.''
-Following this trend to decouple the concept of an ARG from
-generative processes~\citep{minichiello2006mapping},
 \cite{mathieson2020ancestry}
@@
 a recombination or coalescence event.''
-In contrast, ~\citet{gusfield2014recombinatorics} defines an ARG as [TODO]
-x
+% In contrast, ~\citet{gusfield2014recombinatorics} defines an ARG as [TODO]
 % Papers that do not discuss how positions work:
@@
 % getting into this important detail?
+
 The common feature of these definitions is that they are expressed
@@
 $D_1$ and $D_2$ (genomes \textsf{a}--\textsf{d}).
+That is, only the inheritance
+history of the sample genomes is recorded in the ARG.
 % Gregor & Jerome think that "inheritance" is clearer here than "transmission"
@@
 $D_8$ and her daughter $D_6$, but this is not recorded because
-% this is probably not true, given we know the complete pedigree?
-it is not relevant to the genetic ancestry of the samples.
+it is not relevant to the genetic ancestry of the sample
 
@@
 described in detail by
-\citet{kelleher2018efficient}. It is a
-powerful tool with many other applications,
-some of which we discuss in the following sections.
+\citet{kelleher2018efficient}.
+It is a powerful tool for simulation and analysis, as it
+allows to efficiently subset ARGs both in terms of the
+samples and levels of detail about recombination retained,
+as we discuss in the following sections.
 
@@
 We duplicate the nodes and
-edges from the input eARG, and the add inheritance annotations
+edges from the input eARG, and then add inheritance annotations
 to the output gARG's edges. If the node is a common ancestor node,
@@
 which we avoid due the potential confusion with the
-statistical concept of a marginal distributions indicating
-an integrated tree over many positions.
+statistical concept of marginal distributions
+(which could lead to interpretation as an integrated tree over many positions).
 
@@
 If we are given the sequence of local trees for a gARG
-encoded as an oriented tree
-and along with the genome interval covered by each tree,
+encoded as an oriented tree (or in any other form in which
+the tree nodes are explicitly labelled with the identity of the
+corresponding ARG node)
+along with the genome interval covered by each tree,
 then we can recover precisely the same ARG. Thus, there
@@
 
-The internal nodes of trees, however, are usually
-seen as \emph{unlabelled}, since they most often correspond
+The internal nodes of trees, however, are usually \emph{unlabelled},
+because they most often correspond
 to anonymous hypothetical ancestors. Internal
@@
 
-\cite{shipilina2023origin} discuss the same ideas, and note that
+\cite{shipilina2023origin} discuss the same ideas,
+and note that
 ``full ARG ... contains more information than the series of tree
-sequences along the genome''. They view the ``full ARG'' as an
-eARG, and the ``series of tree sequences'' is their
-% jk: probably should delete this. It *is* muddled though
-(somewhat muddled)
-term for the sequence of local trees along the genome.
-Their statement is true, \emph{if} the local trees are
-unlabelled and contain
-no unary nodes. However, as we have shown, if the local trees
+sequences along the genome''.
+Their statement that an ARG
+contains more information than its local trees is true only if
+the local trees are unlabeled.
+However, as we have shown, if the local trees
 contain labelled internal (possibly unary) nodes,
 then  there is precisely as
-much information in them as an ARG. Whether such unary nodes
-are \emph{observable} is an entirely separate question,
-of course.
+much information in them as an ARG.
+Whether such unary nodes are \emph{observable}
+is an entirely separate question, of course.
 
@@
 and have different strategies for reconciling the local trees into an ARG.
-\relate\ first infers local tree topologies, and the identifies
+\relate\ first infers local tree topologies, and then identifies
 edges that persist from tree to tree by reasoning about the sets of samples
@@
 The redundancy that we are interested in
-revolves around the presence of nodes which appear
+revolves around the presence of nodes that appear
 ``unary'' in one or more local trees
@@
 more than two children (e.g.~node \noderef{s}). These represent a ``stacking up'' of
-multiple---often indistinguishable---events into a single node.
+the \emph{effects} of multiple evolutionary events in a single node (genome), and the
+ARG no longer contains the intermediate genomes representing those events.
 
@@
 of Fig.~\ref{fig-simplification}C, but is either binary
-or ternary in all subsequent trees (recall this is a Wright-Fisher
+or ternary in all other local trees (recall this is a Wright-Fisher
 simulation). The final level of simplification is to alter the edge annotations
@@
 Essentially,
-\kwarg\ and \argweaver\ infer complete eARGs (section XXX),
-which contain specific causal recombination \emph{events}
-for each breakpoint.
+\kwarg\ and \argweaver\ infer eARGs that contain
+specific causal recombination \emph{events} for each breakpoint.
 On the other hand, \tsinfer\ and
@@
 extreme and a completely precise event ARG on the other.
-The output of these programs is certainly not a collection of
-\emph{unrelated} local trees, as implied by,
-e.g.,
-\citet{hejase2020summary} when they state that \tsinfer\
-``does not explicitly infer an ARG but rather a sequence of local
-gene trees''.
-
-% More cites? Not worth harping on about now?
-% https://github.com/tskit-dev/what-is-an-arg-paper/issues/38
 

### 2023-10-06 paper.tex

@@
 
+\section*{Acknowledgements}
+We are grateful to Gideon Bradburd and Alex Lewanski for helpful discussions
+and comments on the manuscript.
+
 \bibliographystyle{plainnat}

### 2023-10-06 paper.tex
REmove "straw man" reference

@@
 Note that the \citet{wiuf1999recombination} eARG
-in Fig.~\ref{fig-ancestry-resolution} is something of a straw man,
-in that typically an inference or backward-time simulation program would only produce an
-ARG containing
-nodes or edges ancestral to the sample
-(see the discussion of the ``Big ARG'' stochastic process in Appendix XXX,
-however). Nonetheless, it
-is an instructive example from the literature which highlights several
+in Fig.~\ref{fig-ancestry-resolution} is not particularly
+representative, because inference or simulation methods will usually
+only generate ARGs containing nodes and edges ancestral to the sample
+(see the discussion of the ``Big ARG'' stochastic process in Appendix XXX, however).
+Nonetheless, it is an instructive example from the literature which highlights several
 important properties of ARGs, and the general point about

### 2023-10-06 paper.tex
Add description of missing box regions in Fig 3

--- 2023-10-06 paper.tex
@@
 Coalescence with respect to the sample is indicated by shaded bars, as
-in Fig~\ref{fig-arg-in-pedigree}A.
+in Fig~\ref{fig-arg-in-pedigree}A; nodes \noderef{n}, \noderef{o}, \noderef{p}, \noderef{q} have incomplete
+bars showing that local ancestry of entirely coalesced regions is omitted.
 Line thickness is proportional to the genomic span of each edge.

### 2023-10-06 paper.tex

--- 2023-10-06 paper.tex
@@
 \usepackage{tikz}
-\usetikzlibrary{calc,positioning}
+\usetikzlibrary{math,calc,positioning}
 \usepackage{url}
@@
 \begin{figure}
-% FIXME add local trees in B
-% https://github.com/tskit-dev/what-is-an-arg-paper/issues/320
 \centering
-% \begin{tabular}{cc}
+\tikzmath{\x1 =0; \x2=8;\xx=12.5; \x3=14; \xt=18;}
 \begin{tikzpicture}[x=5mm, y=5mm, node distance=2mm and 20mm]
@@
 
-\node (s0) [greynode] at (0, 0) {};
-\node (s1) [greynode] at (3, 0) {};
-\node (s2) [greynode] at (6, 0) {};
-\node (s3) [greynode] at (3, 1) {};
-\node (s4) [greynode] at (1, 2) {};
-\node (s5) [greynode] at (5, 3) {};
-\node (s6) [greynode] at (3, 4) {};
-
-\node [anchor=north west] at (0,6) {A};
-\node [nodelabel,anchor=north west] at ($(s3) + (0,0)$) {$x$};
-\foreach \u/\lab in {s0/$\textsf{a}$, s1/$\textsf{b}$, s2/$\textsf{c}$} \node[nodelabel,anchor=north] at (\u) {\lab};
-\foreach \u/\lab in {s4/$\textsf{e}$} \node[nodelabel,anchor=south west] at (\u) {\lab};
-\foreach \u/\lab in {s5/$\textsf{f}$} \node[nodelabel,anchor=south east] at (\u) {\lab};
-\foreach \u/\lab in {s3/$\textsf{d}$, s6/$\textsf{g}$} \node[nodelabel,anchor=south] at (\u) {\lab};
-
-%% Edges
+\node [anchor=north west] at (\x1,6) {A};
+\node [anchor=north west] at (\x2,6) {B};
+\node [anchor=north west] at (\xt,6) {C};
+
+%%% (A) ARG
+
+\node (s0) [greynode] at (\x1 + 0, 0) {};
+\node (s1) [greynode] at (\x1 + 3, 0) {};
+\node (s2) [greynode] at (\x1 + 6, 0) {};
+\node (s3) [greynode] at (\x1 + 3, 1) {};
+\node (s4) [greynode] at (\x1 + 1, 2) {};
+\node (s5) [greynode] at (\x1 + 5, 3) {};
+\node (s6) [greynode] at (\x1 + 3, 4) {};
+
 \draw (s1) -- (s3);
 \draw (s0) |- (s4);
-\draw (s4) -- (2,2) |- (s3);
+\draw (s4) -- (\x1 + 2,2) |- (s3);
 \draw (s4) |- (s6);
-\draw (s3) -- (4,1) |- (s5);
+\draw (s3) -- (\x1 + 4,1) |- (s5);
 \draw (s2) |- (s5);
@@
 
-\node [anchor=north west] at (8,6) {C};
-\node [nodelabel,anchor=north west] at ($(10,5)$) {
+%%% (B) Trees
+\node (l0) [greynode] at (\x2 + 0, 0) {};
+\node (l1) [greynode] at (\x2 + 2, 0) {};
+\node (l2) [greynode] at (\x2 + 3, 0) {};
+\node (l3) [greynode] at (\x2 + 1, 2) {};
+\node (l4) [greynode] at (\x2 + 2, 4) {};
+
+\draw (l0) |- (l3);
+\draw (l1) |- (l3);
+\draw (l2) |- (l4);
+\draw (l3) |- (l4);
+
+\node (r0) [greynode] at (\x3 + 0, 0) {};
+\node (r1) [greynode] at (\x3 + 1, 0) {};
+\node (r2) [greynode] at (\x3 + 3, 0) {};
+\node (r3) [greynode] at (\x3 + 2, 3) {};
+\node (r4) [greynode] at (\x3 + 1, 4) {};
+
+\draw (r0) |- (r4);
+\draw (r1) |- (r3);
+\draw (r2) |- (r3);
+\draw (r3) |- (r4);
+
+\foreach \u/\lab in {
+        s0/$\textsf{a}$, s1/$\textsf{b}$, s2/$\textsf{c}$,
+        r0/$\textsf{a}$, r1/$\textsf{b}$, r2/$\textsf{c}$,
+        l0/$\textsf{a}$, l1/$\textsf{b}$, l2/$\textsf{c}$}
+    \node[nodelabel,anchor=south] at ([yshift=-12pt]\u) {\lab};
+
+\foreach \u/\lab in {
+        s6/$\textsf{g}$,
+        l4/$\textsf{g}$,
+        r4/$\textsf{g}$}
+    \node[nodelabel,anchor=south] at (\u) {\lab};
+
+\node [nodelabel,anchor=north west] at ($(s3) + (\x1 + 0,0)$) {$x$};
+\foreach \u/\lab in {
+        s4/$\textsf{e}$,
+        l3/$\textsf{e}$}
+    \node[nodelabel,anchor=south west] at (\u) {\lab};
+\foreach \u/\lab in {
+        s5/$\textsf{f}$,
+        r3/$\textsf{f}$}
+    \node[nodelabel,anchor=south east] at (\u) {\lab};
+\foreach \u/\lab in {s3/$\textsf{d}$, s6/$\textsf{g}$}
+    \node[nodelabel,anchor=south] at (\u) {\lab};
+
+\draw[dashed] (\xx,0) -- (\xx, 4);
+\node[nodelabel,anchor=north] at (\xx,0) {$x$};
+
+%%% (C) Encoding
+\node [nodelabel,anchor=north west] at ($(\xt,5)$) {
 \begin{tabular}{c|c|l}

### 2023-10-06 paper.tex
Minor updates

@@
 Nodes \noderef{e}, \noderef{f} and \noderef{g} are common ancestor events.
-(B) Corresponding local trees to the left and right of breakpoint $x$.
+(B) Corresponding local trees to the left and right of breakpoint $x$
+(note these are shown in the conventional form in which only coalescences
+within the local tree are included; see section XXX for a discussion
+of this important point).
 (C) A concrete encoding of the information (right).
@@
 % confuse people
-Thus, we have the means of encoding the effects of
-multiple \emph{simultaneous} events, and to
+Thus, we have the means of encoding the \emph{effects} of
+multiple evolutionary events, and to
 systematically describe the uncertainty around their
@@
 Fig.~\ref{fig-ancestry-resolution}.
-The first step is straightforward
-(Fig.~\ref{fig-ancestry-resolution}B).
-We duplicate the nodes and
+The first step is straightforward (Fig.~\ref{fig-ancestry-resolution}B):
+we duplicate the nodes and
 edges from the input eARG, and then add inheritance annotations
@@
 If the node is a recombination event with a breakpoint $x$,
-we annotate the outbound edges with the intervals $[0, x)$
-and $[x, L)$.
+we annotate the two outbound edges with the intervals $[0, x)$
+and $[x, L)$, respectively.
 These inheritance annotations are clearly in one-to-one
@@
 Two recent methods take this approach,
-\relate\ and
+\relate~\citep{speidel2019method} and
 \espalier~\citep{rasmussen2022espalier},
@@
 
-\subsection*{The Big and Little ARG}
+\section{The Big and Little ARG}
 \label{app-big-and-little-arg}
@@
 
-\subsection*{Survey of ARG inference methods}
+\section{Survey of ARG inference methods}
 

### 2023-10-06 paper.tex

@@
 Local trees, describing the genetic genealogy of a sample
-at a particular location along the genome, are said to be
-``embedded'' within in ARG [CITES], and the relationship
+at a particular location along the genome, are
+embedded within in ARG, and the relationship
 between the two is subtle and important.
@@
 It is useful to have
-a concrete definition and explicit process for reconstruction.
+a concrete definition and explicit process for local tree extraction.
 We can describe a local tree as an
@@
 
-The internal nodes of trees, however, are usually \emph{unlabelled},
+The internal nodes of trees, however, are usually
+considered to be \emph{unlabelled},
 because they most often correspond
-to anonymous hypothetical ancestors. Internal
-ARG nodes likewise correspond to anonymous hypothetical
+to anonymous hypothetical ancestors.
+Internal ARG nodes likewise correspond to anonymous hypothetical
 ancestors, but if we do not label them we lose the fact that
@@
 \cite{shipilina2023origin} discuss the same ideas,
-and note that
+and note that the
 ``full ARG ... contains more information than the series of tree
@@
 contains more information than its local trees is true only if
-the local trees are unlabeled.
+the local trees are unlabeled and/or do not contain unary nodes.
 However, as we have shown, if the local trees
@@
 A tree branch may correspond to many generations
-of individuals (or indeed cell divisions) and there is usually
+of individuals
+% Following feedback from Wilder, this is more likely to confuse
+% people unless we properly explain
+% (or indeed cell divisions)
+and there is usually
 very little information about these intermediates, or utility
@@
 provided by nodes that are unary in one or more local trees.
-[FIXME this is clumsy; should mention somewhere, though.]
+% Not sure what to do with this - should mention though.
 \citet{mathieson2020ancestry} show examples in which the ARGs
-contain unary nodes, and state that paths are ``implicitly passing through points representing specific
-individuals.''
+contain unary nodes, and state that paths are
+``implicitly passing through points representing specific individuals.''
 % Full quote from fig caption:
@@
 \emph{unrelated} local trees (i.e., with either unlabelled internal
-nodes or differently labelled nodes in each tree; see the
-next section).
+nodes or differently labelled nodes in each tree).
 The ARGs inferred by \relate\ and \tsinfer\ both contain substantial
@@
 the entire sequence.
-Therefore, they contain a level of detail about recombination
+They contain a level of detail about recombination
 that lies somewhere between a sequence of unrelated local trees on one
@@
 of the inferred ARG with the inference
-itself within the same software package
-% FIXME-  more explicit?
-(e.g.\ Relate, ARGneedle)
+itself within the same software package,
 this approach scales poorly in terms of software development
@@
 software ecosystem.
-% FIXME not sure where this goes
-Earlier efforts to standardise ARG interchange
-did not succeed~\citep{cardona2008extended,mcgill2013graphml}.
 
@@
 sequence variation using the ``site'' and  ``mutation'' tables.
-The \texttt{tskit} library~\citep{tskit2023tskit} is a liberally
+(Technically, edges in tree sequence terminology would be better
+described as ``edge-intervals'', as each describes a single continguous
+interval of genome inheritance between a pair of nodes. This
+denormalisation of the gARG data model is for efficiency purposes.)
+The \texttt{tskit} library is a liberally
 licensed open source toolkit that provides a comprehensive suite
@@
 fundamental to the actual representation of the thing]
+
+Earlier efforts to standardise ARG interchange
+did not succeed~\citep{cardona2008extended,mcgill2013graphml}.
+
 In summary, we advocate for decoupling between generative models that give rise to observed data and data formats that can faithfully and realistically encode these data.

### 2023-10-06 paper.tex
Make Figs 1 & 2 shorter but still readable

--- 2023-10-06 paper.tex
@@
 \usepackage{tikz}
-\usetikzlibrary{math,calc,positioning}
+\usetikzlibrary{calc,positioning}
 \usepackage{url}
@@
 \begin{center}
-    % FIXME making this 0.75 textwidth for now because it's too tall.
-    % https://github.com/tskit-dev/what-is-an-arg-paper/issues/321
-    \includegraphics[width=0.75\textwidth]{illustrations/arg-in-pedigree}
+    \includegraphics[width=\textwidth]{illustrations/arg-in-pedigree}
 \end{center}
@@
 % it'll be clearer what we need to say in the caption vs the text.
-[FIXME Update fig to make wider and less tall, and change $I$ to $D$.
-Also possibly add some tabular encoding of gARG data.]
 An example genome ARG embedded in a pedigree.
@@
 the paternal genomes \textsf{e}
-and \textsf{f}, and regions of genome inherited are shown.
+and \textsf{f} (with breakpoints in red), and regions of genome inherited are shown.
 Genomes are shaded such that where, backwards in time,
@@
 \begin{figure}
+% FIXME add local trees in B
+% https://github.com/tskit-dev/what-is-an-arg-paper/issues/320
 \centering
-\tikzmath{\x1 =0; \x2=8;\xx=12.5; \x3=14; \xt=18;}
+% \begin{tabular}{cc}
 \begin{tikzpicture}[x=5mm, y=5mm, node distance=2mm and 20mm]
@@
 
-\node [anchor=north west] at (\x1,6) {A};
-\node [anchor=north west] at (\x2,6) {B};
-\node [anchor=north west] at (\xt,6) {C};
-
-%%% (A) ARG
-
-\node (s0) [greynode] at (\x1 + 0, 0) {};
-\node (s1) [greynode] at (\x1 + 3, 0) {};
-\node (s2) [greynode] at (\x1 + 6, 0) {};
-\node (s3) [greynode] at (\x1 + 3, 1) {};
-\node (s4) [greynode] at (\x1 + 1, 2) {};
-\node (s5) [greynode] at (\x1 + 5, 3) {};
-\node (s6) [greynode] at (\x1 + 3, 4) {};
-
+\node (s0) [greynode] at (0, 0) {};
+\node (s1) [greynode] at (3, 0) {};
+\node (s2) [greynode] at (6, 0) {};
+\node (s3) [greynode] at (3, 1) {};
+\node (s4) [greynode] at (1, 2) {};
+\node (s5) [greynode] at (5, 3) {};
+\node (s6) [greynode] at (3, 4) {};
+
+\node [anchor=north west] at (0,6) {A};
+\node [nodelabel,anchor=north west] at ($(s3) + (0,0)$) {$x$};
+\foreach \u/\lab in {s0/$\textsf{a}$, s1/$\textsf{b}$, s2/$\textsf{c}$} \node[nodelabel,anchor=north] at (\u) {\lab};
+\foreach \u/\lab in {s4/$\textsf{e}$} \node[nodelabel,anchor=south west] at (\u) {\lab};
+\foreach \u/\lab in {s5/$\textsf{f}$} \node[nodelabel,anchor=south east] at (\u) {\lab};
+\foreach \u/\lab in {s3/$\textsf{d}$, s6/$\textsf{g}$} \node[nodelabel,anchor=south] at (\u) {\lab};
+
+%% Edges
 \draw (s1) -- (s3);
 \draw (s0) |- (s4);
-\draw (s4) -- (\x1 + 2,2) |- (s3);
+\draw (s4) -- (2,2) |- (s3);
 \draw (s4) |- (s6);
-\draw (s3) -- (\x1 + 4,1) |- (s5);
+\draw (s3) -- (4,1) |- (s5);
 \draw (s2) |- (s5);
@@
 
-%%% (B) Trees
-\node (l0) [greynode] at (\x2 + 0, 0) {};
-\node (l1) [greynode] at (\x2 + 2, 0) {};
-\node (l2) [greynode] at (\x2 + 3, 0) {};
-\node (l3) [greynode] at (\x2 + 1, 2) {};
-\node (l4) [greynode] at (\x2 + 2, 4) {};
-
-\draw (l0) |- (l3);
-\draw (l1) |- (l3);
-\draw (l2) |- (l4);
-\draw (l3) |- (l4);
-
-\node (r0) [greynode] at (\x3 + 0, 0) {};
-\node (r1) [greynode] at (\x3 + 1, 0) {};
-\node (r2) [greynode] at (\x3 + 3, 0) {};
-\node (r3) [greynode] at (\x3 + 2, 3) {};
-\node (r4) [greynode] at (\x3 + 1, 4) {};
-
-\draw (r0) |- (r4);
-\draw (r1) |- (r3);
-\draw (r2) |- (r3);
-\draw (r3) |- (r4);
-
-\foreach \u/\lab in {
-        s0/$\textsf{a}$, s1/$\textsf{b}$, s2/$\textsf{c}$,
-        r0/$\textsf{a}$, r1/$\textsf{b}$, r2/$\textsf{c}$,
-        l0/$\textsf{a}$, l1/$\textsf{b}$, l2/$\textsf{c}$}
-    \node[nodelabel,anchor=south] at ([yshift=-12pt]\u) {\lab};
-
-\foreach \u/\lab in {
-        s6/$\textsf{g}$,
-        l4/$\textsf{g}$,
-        r4/$\textsf{g}$}
-    \node[nodelabel,anchor=south] at (\u) {\lab};
-
-\node [nodelabel,anchor=north west] at ($(s3) + (\x1 + 0,0)$) {$x$};
-\foreach \u/\lab in {
-        s4/$\textsf{e}$,
-        l3/$\textsf{e}$}
-    \node[nodelabel,anchor=south west] at (\u) {\lab};
-\foreach \u/\lab in {
-        s5/$\textsf{f}$,
-        r3/$\textsf{f}$}
-    \node[nodelabel,anchor=south east] at (\u) {\lab};
-\foreach \u/\lab in {s3/$\textsf{d}$, s6/$\textsf{g}$}
-    \node[nodelabel,anchor=south] at (\u) {\lab};
-
-\draw[dashed] (\xx,0) -- (\xx, 4);
-\node[nodelabel,anchor=north] at (\xx,0) {$x$};
-
-%%% (C) Encoding
-\node [nodelabel,anchor=north west] at ($(\xt,5)$) {
+\node [anchor=north west] at (8,6) {C};
+\node [nodelabel,anchor=north west] at ($(10,5)$) {
 \begin{tabular}{c|c|l}
@@
 Note that the \citet{wiuf1999recombination} eARG
-in Fig.~\ref{fig-ancestry-resolution} is not particularly
-representative, because inference or simulation methods will usually
-only generate ARGs containing nodes and edges ancestral to the sample
-(see the discussion of the ``Big ARG'' stochastic process in Appendix XXX, however).
-Nonetheless, it is an instructive example from the literature which highlights several
+in Fig.~\ref{fig-ancestry-resolution} is something of a straw man,
+in that typically an inference or backward-time simulation program would only produce an
+ARG containing
+nodes or edges ancestral to the sample
+(see the discussion of the ``Big ARG'' stochastic process in Appendix XXX,
+however). Nonetheless, it
+is an instructive example from the literature which highlights several
 important properties of ARGs, and the general point about

### 2023-10-06 illustrations.py

@@ -107,7 +107,7 @@ def ancestry_resolution():
                 rotate=False,
                 font_weight="normal" if full else "bold",
                 alpha=0.5 if full else None,
-                font_size=16,
+                font_size=24,
                 edge_labels={k: l for k, l in lab.items() if (l == full_edge) is full},
                 horizontalalignment="center",
                 bbox=dict(
@@ -129,6 +129,7 @@ def ancestry_resolution():
     pos, G = argutils.viz.draw(
         ts,
         ax1,
+        tweak_x={7: 15, 3: -12, 4: -8, 6: -6},
         node_color=mpl.colors.to_hex(plt.cm.tab20(1)),
         node_size=100,
         max_edge_width=2,
@@ -181,7 +182,7 @@ def ancestry_resolution():
     for n in G.nodes:
         if G.nodes[n]["flags"] & argutils.ancestry.NODE_IS_RECOMB:
             col = 'tab:red'
-            size = (1, 1)
+            size = (1.2, 1.2)
             breaks = (edges.left[edges.child == n], edges.right[edges.child == n])
             breaks = np.unique(np.concatenate(breaks))
             assert len(breaks == 3)
@@ -200,7 +201,7 @@ def ancestry_resolution():
                 0.4, 0.2,
                 str(int(breakpoint)),
                 fontname="Trebuchet MS",
-                fontsize=30,
+                fontsize=40,
             )
 
     # Panel (b)
@@ -216,7 +217,7 @@ def ancestry_resolution():
             fontname="Trebuchet MS",
             verticalalignment="bottom",
             loc="center",
-            fontsize=20,
+            fontsize=24,
         )
 
     # Panel (c)
@@ -560,12 +561,14 @@ def arg_in_pedigree():
 
     ts_used = ts  # argutils.remove_unused_nodes(ts)  # use this to remove unused nodes
 
-    fig, ax = plt.subplots(1, 1, figsize=(3.8, 5))
-    col = mpl.colors.to_hex(plt.cm.tab20(1))
+    fig, ax = plt.subplots(1, 1, figsize=(3.8, 4.35))
+    col = mpl.colors.to_hex("lightgray")
     pos, G = argutils.viz.draw(
         ts_used, ax,
         reverse_x_axis=True,
         draw_edge_widths=False,
+        node_size=400,
+        font_size=14,
         node_color=col,
[…]

### 2023-10-06 illustrations.py
Fig gap at bottom of simplification fig
Fixes #323

@@ -423,7 +423,7 @@ def simplification():
     graph1_svg = graph1_svg[graph1_svg.find("<svg") :]    
 
     svg = [
-        '<svg width="900" height="900" xmlns="http://www.w3.org/2000/svg" '
+        '<svg width="900" height="820" xmlns="http://www.w3.org/2000/svg" '
         'xmlns:xlink="http://www.w3.org/1999/xlink">',
         "<style>.tree-sequence text {font-family: sans-serif}</style>"
     ]

### 2023-10-06 paper.tex
Updates from Overleaf

--- 2023-10-06 paper.tex
@@
 \documentclass{article}
+\usepackage[letterpaper,top=2cm,bottom=2.5cm,left=3cm,right=3cm,marginparwidth=1.75cm]{geometry}
 \usepackage[round]{natbib}
@@
 \usepackage{url}
+\usepackage[symbol]{footmisc}
 % JK: turning this off for the moment as I keep clicking through on links
@@
 \linenumbers
-\title{A general and efficient representation of Ancestral Recombination Graphs}
-
-% First authors
-\author[1]{Yan Wong}
-
-% Second authors
-\author[2,$\star$]{Anastasia Ignatieva}
-\author[3,4,$\star$]{Jere Koskela}
-
-% Middle Authors
-\author[5]{Gregor Gorjanc}
-\author[6,7]{Anthony W. Wohns}
-
-% Corresponding
-\author[1,$\dagger$]{Jerome Kelleher}
-
-\affil[1]{Big Data Institute, Li Ka Shing Centre for Health Information and Discovery, University of Oxford, OX3 7LF, UK}
-\affil[2]{Department of Statistics, University of Oxford, OX1 3LB, UK}
-\affil[3]{School of Mathematics, Statistics and Physics, Newcastle University, NE1 7RU, UK}
-\affil[4]{Department of Statistics, University of Warwick, CV4 7AL, UK}
-\affil[5]{The Roslin Institute and Royal (Dick) School of Veterinary Studies, University of Edinburgh, EH25 9RG, UK}
-\affil[6]{Broad Institute of MIT and Harvard, Cambridge, MA 02142, USA}
-\affil[7]{Department of Genetics, Stanford University School of Medicine, Stanford, CA 94305, USA}
-
-\affil[$\star$]{Denotes shared second authorship, listed alphabetically}
-\affil[$\dagger$]{Denotes corresponding author}
+\title{\vspace{-1.5em} \bf A general and efficient representation of ancestral recombination graphs}
+
+% % First authors
+% \author[1]{Yan Wong}
+
+% % Second authors
+% \author[2,$\star$]{Anastasia Ignatieva}
+% \author[3,4,$\star$]{Jere Koskela}
+
+% % Middle Authors
+% \author[5]{Gregor Gorjanc}
+% \author[6,7]{Anthony W. Wohns}
+
+% % Corresponding
+% \author[1,$\dagger$]{Jerome Kelleher}
+
+% \affil[1]{Big Data Institute, Li Ka Shing Centre for Health Information and Discovery, University of Oxford, OX3 7LF, UK}
+% \affil[2]{Department of Statistics, University of Oxford, OX1 3LB, UK}
+% \affil[3]{School of Mathematics, Statistics and Physics, Newcastle University, NE1 7RU, UK}
+% \affil[4]{Department of Statistics, University of Warwick, CV4 7AL, UK}
+% \affil[5]{The Roslin Institute and Royal (Dick) School of Veterinary Studies, University of Edinburgh, EH25 9RG, UK}
+% \affil[6]{Broad Institute of MIT and Harvard, Cambridge, MA 02142, USA}
+% \affil[7]{Department of Genetics, Stanford University School of Medicine, Stanford, CA 94305, USA}
+
+% \affil[$\star$]{Denotes shared second authorship, listed alphabetically}
+% \affil[$\dagger$]{Denotes corresponding author}
+
+\author[1]{Yan~Wong}
+\author[2,3$\star$]{Anastasia~Ignatieva}
+\author[4,5$\star$]{Jere~Koskela}
+\author[6]{Gregor~Gorjanc}
+\author[7,8]{Anthony~W.~Wohns}
+\author[1$\dagger$]{Jerome~Kelleher}
+\affil[ ]{\mbox{}\vspace{-2.5em}}
 
 \maketitle
+
+\setlength{\skip\footins}{1em}
+\setlength{\footnotemargin}{0.5em}
+\renewcommand{\thefootnote}{\arabic{footnote}}
+\footnotetext[1]{Big Data Institute, Li Ka Shing Centre for Health
+Information and Discovery, University of Oxford, UK}
+\footnotetext[2]{School of Mathematics and Statistics, University of Glasgow, UK}
+\footnotetext[3]{Department of Statistics, University of Oxford, UK}
+\footnotetext[4]{School of Mathematics, Statistics and Physics, Newcastle University, UK}
+\footnotetext[5]{Department of Statistics, University of Warwick, UK}
+\footnotetext[6]{The Roslin Institute and Royal (Dick) School of Veterinary Studies, University of Edinburgh, UK}
+\footnotetext[7]{Broad Institute of MIT and Harvard, Cambridge, USA}
+\footnotetext[8]{Department of Genetics, Stanford University School of Medicine, Stanford, USA}
+\renewcommand{\thefootnote}{\fnsymbol{footnote}}
+\footnotetext[1]{Shared second authorship, listed alphabetically}
+\footnotetext[2]{Correspondence: jerome.kelleher@bdi.ox.ac.uk}
+
+\vspace{-1em}
 
@@
 The structure representing such recombinant genetic ancestry
-is usually referred to as an Ancestral Recombination Graph (ARG),
+is usually referred to as an ancestral recombination graph (ARG),
 although there is some confusion about the interpretation and little
@@
 \section*{Discussion}
-Recent breakthroughs have made ARG inference methods practical
-at scale for the first time, and there has been a surge of interest
-in inference methods, their evaluation and application.
-The prospect of ARGs being used routinely in population
-and statistical genetics applications is tantalising,
+Recent breakthroughs have finally made large-scale ARG inference
+feasible in practice, leading to a surge of interest
+in development of ARG-based inference methods, their evaluation 
+and application.
+The prospect of ARGs being used routinely within population
+and statistical genetics is tantalising,
 but in reality there is substantial work to be done to
@@
 A necessary first step is a degree of terminological clarity.
-As discussed in Appendix XXX, the term Ancestral Recombination
-Graph has several
-subtly different interpretations depending on context.
-The trend to decouple ARGs from their origins as a stochastic
+As discussed in Appendix XXX, the term ``ancestral recombination
+graph" has several
+subtly different interpretations, depending on context.
+The trend to decouple ARGs from their original definiton 
+as a stochastic
 process and instead use the term as a more general representation of any
@@
 sampling from its distribution of ARGs at the scale of hundreds of thousands
-of samples, we could only explore the tiniest corner of such an incompehensibly
+of samples, we could only explore the tiniest corner of such an incomprehensibly
 large space of possibilities.
@@
 describe and utilise uncertainty about ARG inference.
-One approach, enabled by the gARG encoding described here,
-is to use polytomies and stacked recombinations as direct topological
-descriptions of uncertainty. Other methods to capture, for example,
+One approach, enabled by the gARG encoding described here, is to allow
+nodes to have (rootward) in-degree greater than two (polytomies
+representing uncertainty over the ordering of coalescence events) or 
+out-degree greater than two (representing uncertainty over the ordering
+of multiple recombination events). Other methods to capture, for example,
 uncertainty about node ages and breakpoint locations, is an important

### 2023-10-06 illustrations.py
Switch back to letters for fig 4 labels

@@ -373,8 +373,8 @@ def simplification():
 
     labels = {i: i for i in range(len(string.ascii_lowercase))}
     # relabel the nodes to get samples reading A B C D
-    #labels.update({6: "e", 4: "g", 10: "j", 9: "k", 12: "l", 11: "m", 15: "o", 14: "p"})
-    labels = {i: i for i in range(26)}
+    labels.update({6: "e", 4: "g", 10: "j", 9: "k", 12: "l", 11: "m", 15: "o", 14: "p"})
+    #labels = {i: i for i in range(26)}  # uncommment to get node IDs for re-positioning
     ts1 = argutils.viz.label_nodes(ts, labels)
 
     fig, (ax1, ax2, ax3, ax4) = plt.subplots(4, 1, figsize=(3, 12))

### 2023-10-06 paper.tex

@@
 although there is some confusion about the interpretation and little
-agreement on the concrete details.
+agreement on concrete details.
 We propose a concrete definition of ARGs
@@
 and contrast this with the classical event-based definition (eARG).
-The eARG definition is based on the coalescent with recombination
-stochastic process, and we show how the assumptions of that process
+The eARG definition is based on the stochatic process of the coalescent with recombination,
+and we show how the assumptions of that process
 place limitations on the patterns of inheritance that can be
-represented, and that these assumptions are
-% Better wording
-substantially broken
-in modern large-scale datasets.
-Representing genetic ancestry
-with an eARG also requires complete precision about the ordering
+represented, which are inadequate for modern large-scale datasets.
+The eARG representation also requires complete precision about the ordering
 of all recombination events, which we will rarely have sufficient
@@
 programs, resulting in the assorted negative consequences with which
-geneticists are only too familiar~\citep{excoffier2006computer}.
+geneticists are only too familiar~\citep{excoffier2006computer,gower2022demes}.
 Perhaps most importantly, the direct linkage to
@@
 \section*{ARGs and local trees}
+% TODO mention Yamato here? I'm sure there were some interesting points 
+% about local trees in that paper
 Local trees, describing the genetic genealogy of a sample
@@
 i.e.\ are unary everywhere (e.g.\ \noderef{n} and \noderef{r}).
-(D) Rewrite edges to bypass any unary nodes in local trees.
+(D) Rewrite edges to bypass nodes where they appear unary in local trees.
 }
@@
 The efficiency with which we can store and analyse these inferred
-ARGs is therefore of great importantance, and a key factor
+ARGs is therefore of great importance, and a key factor
 in the ultimate success of downstream ARG-based methods.
@@
 and likely many more.
+
+[TODO discussion of haplotype block based storage
+like ARGON. Basically: this is useful for some applications,
+but ultimately it's a derived property of the actual ARG
+topology that can be recovered efficiently, as demonstrated by 
+\cite{nowbandegani2023extremely}]
+
+% \cite{shipilina2023origin} discuss the idea of a ``haplotype block''
+% (equivalent to the ``bricked tree sequence'' of
+% \cite{nowbandegani2023extremely},
+% % TODO look this up. I expect it's basically the same??
+% and the block-based ARG data structure of
+% \citep{palamara2016argon}),
+% here we consider the unique sets of
+% samples that coalesce at a node in the ARG over a particular
+% genome interval, and the fundamental limitations they place
+% on ARG inference.
+% [FIXME find somewhere natural for this. The last clause is too vague - what
+% are the implications?]
+% Alternative gARG encodings split transmitted intervals and associate each with
+% a new node, resulting in a single genome being represented by a group of
+% multiple nodes (or ``blocks'' \citep{palamara2016argon}), with implications for
 
@@
 subtly different interpretations, depending on context.
-The trend to decouple ARGs from their original definiton 
-as a stochastic
-process and instead use the term as a more general representation of any
-recombinant genetic ancestry seems most useful, which we have
-tried to clarify and systematise here. Thus
+The trend to decouple ARGs from their original definition 
+within the context of stochastic
+processes and instead use the term as a more general representation of any
+recombinant genetic ancestry seems useful, and we have
+tried to clarify and systematise it here. Thus
 we can think of an ARG as any structure that encodes the
 reticulate genetic ancestry of a sample of colinear sequences under
-the influence of recombination. The ``genome'' ARG encoding
+the influence of recombination. The ``genome'' ARG (gARG) encoding
 made explicit here is one way we can concretely
@@
 violate its assumptions.
-For example, several human datasets now consist of hundreds of thousands of
+For instance, several human datasets now consist of hundreds of thousands of
 % TODO cite GeL paper
@@
 the uncertainty in our ARG estimates. Indeed, even if the coalescent
-was a suitable model for datasets such as UKB and we had a means of
+were a suitable model for datasets such as UKB and we had a means of
 sampling from its distribution of ARGs at the scale of hundreds of thousands
@@
 out-degree greater than two (representing uncertainty over the ordering
-of multiple recombination events). Other methods to capture, for example,
-uncertainty about node ages and breakpoint locations, is an important
+of multiple recombination events). Development of other methods to capture, for example,
+uncertainty about node ages and recombination breakpoint positions, is an important
 aspect of future work.
@@
 we should also consider the value that such precise estimates provide
-to downstream applications.
+for downstream applications.
 To date, few downstream applications using estimated ARGs
@@
 Most applications work by examining local trees independently.
-For example, the \relate\ selection statistic~\citep{speidel2019method}
-computes selection $p$-values by comparing the coalescence rates at various time points.
+For example, the \relate\ selection test~\citep{speidel2019method}
+computes selection $p$-values by computing clade size probabilities conditional on the timing of coalescence events in a given local tree.
 In their method
@@
 % Note: haven't said anything about the coalescent here. Do we need to?
-The advantages of a model-free representation that naturally
-incorporates uncertainty about the ordering of events in and ARG are well illustrated by the
-ARGs inferred for SARS-CoV-2 by~\cite{zhan2023towards}.
-The COVID-19 pandemic led to collection of whole-genome viral data
-at unprecented scale. The GISAID database~\citep{shu2017gisaid}
-holds over 15 million sequences as of [FIXME].
-Recombination, while rare, is an important
-factor in the evolution of
-SARS-CoV-2~\citep{vaninsberghe2021recombinant,jackson2021generation,ignatieva2022ongoing}
-with recombinants potentially gaining fitness by combining
-mutations from deeply diverged lineages [cite].
-\citet{zhan2023towards} inferred two large ARGs for SARS-CoV-2 based
-on GISAID data, one containing 1.27 million sequences sampled up to
-June 30, 2021 and the other (more sparsely sampled) consisting of
-657K sequences sampled up to June 30, 2022.
-Sequencing effort at the height of the pandemic produced
-tens of thousands of sequences collected in a single day,
-with hundreds being identical.
-Polytomies are therefore and unescapable reality, and a
-fully-resolved binary tree would clearly be false precision.
-Because of longitudinal data collection, there is no
-requirement that samples must be leaf nodes, and samples
-will frequently descend from other (older) samples.
-Unary nodes (a node in a local tree with exactly one child)
-are also a common feature, when a sample is inferred to have
-exactly one direct descendant.
-Despite the abundance of data and the rarity of recombination,
-the parental lineages cannot be precisely identified
-in many cases. For example, many are inferred to have three
-or more parental lineages (in agreement with the
-analysis resulting in the Pango designation of the recombinant
-lineage); in most cases, these will indicate two or more
-independent recombination events.
-Faithfully representing the uncertainty
-around these events seems a more useful approach than
-the arbitrary resolution which would be required,
-if encoded as an eARG.
-
-% An interesting property of the SARS-CoV-2 dataset is that
-% it highlights the uncertainty in the genomic position of
-% recombination breakpoints. [Some discussion about why
-% we can never know where the breakpoint really happened, and
-% some follow-up to note that this is even more acute in
-% standard popgen datasets where we're thinking about ancient
-% recombs].
-% This is a limitation of the
-% gARG encoding proposed here, which requires that we specify
-% exact intervals of inheritance and therefore precise
-% recombination breakpoints.
-% An approach in which we can systematically reason about the
-% uncertainty of inheritance intervals as they propogate
-% through an ARG
-% would be an interesting avenue for future research.
+The advantages of a model-agnostic representation that naturally
+incorporates uncertainty about the ordering of events in an ARG are well illustrated by~\cite{zhan2023towards}, who
+inferred ARGs using millions of SARS-CoV-2 sequences from the
+GISAID database~\citep{shu2017gisaid}. In contrast to typical human sequencing datasets, the SARS-CoV-2 data is
+sampled continuously through time, sometimes with tens of thousands of sequences collected per day, with relatively little genetic diversity to
+distinguish them. The reconstructed ARGs thus contain polytomies (representing
+both the uncertainty around coalescence times and the presence of true
+multiple mergers), non-leaf sample nodes (sequences with descendants
+also present in the dataset) and unary nodes (samples that have
+one other sampled descendant). Recombination is known
+to be an important factor in the evolution of SARS-CoV-2~\citep{vaninsberghe2021recombinant,jackson2021generation,ignatieva2022ongoing},
+and the ARGs also contain recombination nodes. 
+Because parental sequences
+are generally never sampled themselves, and often 
+a recombinant strain is the product of multiple recombination events,
+uncertainty around this is captured by recording the ancestry
+of each part of the recombinant sequence without arbitrarily assigning
+times or orderings for these events. This demonstrates the flexibility of the gARG
+encoding in capturing recombinant genealogies regardless of the underlying
+(possibly unknowable) generative evolutionary process.
 
@@
 In most cases,
-the accuracy of ARG inference is evaluated by comparing some truth simulations
-to the inferred ARGs by the pairwise comparison of local trees along the genome
+ARG inference is evaluated by simulating data from a known ground truth ARG, and comparing this to the inferred version via pairwise comparison of local trees along the genome, for example
 using tree distance
-metrics~\citep[e.g.][]{robinson1981comparison,kendall2016mapping}
-as originally proposed by \citet{kuhner2015assessing}.
+metrics~\citep[e.g.][]{robinson1981comparison,kendall2016mapping},
+as described by \citet{kuhner2015assessing}.
 % While this provides valuable insights into the method's performance,
@@
 In comparing tree-by-tree along the genome, the effects of recombination
-structure is incorporated in an indirect manner, and the tree
-metrics assume unlabelled nodes, effectively discarding information
-about node sharing between trees.
+are only incorporated in an indirect manner through the correlations
+between the local trees, rather than through directly taking into account
+the persistence of nodes and edges across multiple trees.
 Tree distance metrics usually have $O(n^2)$ time complexity and therefore cannot be
@@
 A recent trend has been to move away from such tree distance-based approaches and to examine more properties of the inferred ARGs.
-\citet{brandt2021evaluation} compared the
-distributions of time to the most recent common ancestor across several
-methods, which captures many important aspects of dated ARGs
-but neglects [SOME OTHER THINGS].
-\citet{deng2021distribution} derived the distribution of the expected
-waiting distances to tree changes, and compared this to the
-outputs of several methods.
-[ \citet{ignatieva2023distribution} DID SOMETHING ELSE]
+\citet{brandt2021evaluation} focussed on the
+distributions of pairwise MRCA times; \citet{deng2021distribution} derived an approximation to the distribution of the 
+waiting distances between local trees; \citet{ignatieva2023distribution}
+derived the distribution of the genomic span of an edge and that of a
+clade of samples. In each case, simulation studies demonstrated 
+substantial differences between these quantities in simulated and reconstructed ARGs,
+that were not captured using tree-by-tree comparisons.
 % Not sure this is the right place for this sentence
-All of these evaluations have been based on
-ground truth data from highly idealised simulations
+These evaluations have been based on
+ground truth data from highly idealised simulations, some 
 with small sample sizes,
@@
 
-[TODO finish up with a positive note about interchange, and make the point
-clearly that tskit is an ARG library. Splice in the stuff about haplotype
-blocks and ARGON format here somewhere --- basically this is a derived product
-of the gARG encoding, so while there are some computation advantages it's not
-fundamental to the actual representation of the thing]
-
-Earlier efforts to standardise ARG interchange
-did not succeed~\citep{cardona2008extended,mcgill2013graphml}.
-
-In summary, we advocate for decoupling between generative models that give rise to observed data and data formats that can faithfully and realistically encode these data.
-We have shown that ``genome'' ARG (gARG) encoding enables flexible encoding of today's datasets with realistic levels of precision.
-Importantly, genome ARG encoding should facilitate easier data interchange between different tools, which will in turn support active research in this area.
-
-% \cite{shipilina2023origin} discuss the idea of a ``haplotype block''
-% (equivalent to the ``bricked tree sequence'' of
-% \cite{nowbandegani2023extremely},
-% % TODO look this up. I expect it's basically the same??
-% and the block-based ARG data structure of
-% \citep{palamara2016argon}),
-% here we consider the unique sets of
-% samples that coalesce at a node in the ARG over a particular
-% genome interval, and the fundamental limitations they place
-% on ARG inference.
-% [FIXME find somewhere natural for this. The last clause is too vague - what
-% are the implications?]
-% Alternative gARG encodings split transmitted intervals and associate each with
-% a new node, resulting in a single genome being represented by a group of
-% multiple nodes (or ``blocks'' \citep{palamara2016argon}), with implications for
+Interest in ARG inference methods and downstream applications is 
+burgeoning, with exciting developments arriving at ever-increasing pace.
+Without agreement on basic terminology and some standardisation
+on data formats, however, the ARG revolution may falter.
+Here, we have advocated
+for decoupling the data structures used to represent ARGs
+from generative processes, and have shown that the 
+gARG encoding enables flexible description of 
+contemporary datasets with varying levels of precision and integral
+handling of some important forms of temporal uncertainty.
+For ARG based methods to achieve mainstream status, we require
+a rich supporting software ecosystem. 
+Ideally, this would comprise a wide range of different 
+inference methods specialised to different organisms, types 
+and scales of data and 
+% I'm thinking about things like Espalier specifically wanting
+% to characterise the recombs, vs Relate mainly wanting to 
+% get good local trees
+specific inference goals, sharing a common ARG interchange format. 
+The output of these diverse inference methods could then be 
+processed by numerous downstream applications, 
+enabling users to freely compare the outputs of different 
+inference methods on their data, and the result of different
+analyses on these inferences.
+Earlier efforts to standardise ARG interchange shared 
+this vision, 
+but did not succeed~\citep{cardona2008extended,mcgill2013graphml}.
+Today, however, we have the benefit of a pre-existing ARG
+interchange format, already widely used and accompanied 
+by high-quality, high-performance software (see Section XXX for details). 
+Adopting the \texttt{tskit} library 
+as a community standard,
+utilising its rich suite of operations,
+would substantially accelerate progress, and help 
+establish ARG methods as mainstream genetics.
 
@@
 fundamental questions about properties
-of the ARG as a discrete graph structure, abstracting away from the
-generating process and focussing on the ARG topology without considering
-branch lengths.
+of the ARG as a discrete graph structure, focussing on the ARG topology without considering
+either branch lengths or indeed the generating process.
 The first prominent problem was calculating (lower bounds on) the minimum number of
@@
 other more general types of phylogenetic networks~\citep{huson2010phylogenetic}. The
-focus has been on algorithmic and combinatorial results, that largely do not
+focus has been on algorithmic and combinatorial results that largely do not
 have direct relevance to the inference problems described above.
@@
 depends on the context in which it is used, and can mean both the
-stochastic process generating genealogies in the presence of
+stochastic process that generates genealogies in the presence of
 recombination~\citep[e.g.][]{nordborg2000linkage,birkner2013ancestral,
@@
 \cite{minichiello2006mapping} explicitly distinguished the ARG as a
-stochastic process from the ARG as a data structure, and
-argued for using the term to mean specific realised ancestral histories.
-Unless otherwise stated, we will use the ARG-as-a-data-structure
-interpretation for the remainder of this paper.
+stochastic process from the ARG as a data structure. We have followed them in
+using the term in the latter sense, to mean a specific realised ancestral history.
 
@@
 available for recombination in each time interval.
-This information cannot be recovered from the gARG encoding depicted in
-e.g.\ Figure \ref{fig-ancestry-resolution}.
-For example, it is clear that a recombination takes place between nodes \textsf{b} and
+This may not be obvious.  example, in the gARG encoding depicted in Figure \ref{fig-ancestry-resolution}
+it is clear that a recombination takes place between nodes \textsf{b} and
 \textsf{d} as well as \textsf{e}, but the exact time of the event is ambiguous,
-and thus so is the number of lineages during the time interval.
+and thus so is the number of lineages during the time interval. This information cannot be recovered
+from the gARG encoding 
 Conventions can be introduced to resolve such ambiguities;

### 2023-10-06 paper.tex
tweaks

@@
 although there is some confusion about the interpretation and little
-agreement on concrete details.
+agreement on the concrete details.
 We propose a concrete definition of ARGs
@@
 and contrast this with the classical event-based definition (eARG).
-The eARG definition is based on the stochatic process of the coalescent with recombination,
+The eARG definition is based on the coalescent with recombination
+stochastic process,
 and we show how the assumptions of that process
@@
 These evaluations have been based on
-ground truth data from highly idealised simulations, some 
-with small sample sizes,
-and the effects of the richness of real data
-(beyond simplistic error models)
-on modern biobank scale or multi-generational pedigree datasets
-are almost entirely unknown.
-A suite of true ARG evaluation metrics, that take into account more of the
+ground truth data from highly idealised simulations, with sample sizes limited to at most a few thousand (typically much fewer).
+The effects of the richness of real data
+on modern biobank-scale or multi-generational pedigree datasets
+are almost entirely unknown (beyond studies using very simplistic error models).
+A suite of ARG evaluation metrics, that take into account more of the
 global topology and can be applied to large ARGs, would be a
@@
 available for recombination in each time interval.
-This may not be obvious.  example, in the gARG encoding depicted in Figure \ref{fig-ancestry-resolution}
+This may not be obvious. For example, in the gARG encoding depicted in Figure \ref{fig-ancestry-resolution}C
 it is clear that a recombination takes place between nodes \textsf{b} and
 \textsf{d} as well as \textsf{e}, but the exact time of the event is ambiguous,
-and thus so is the number of lineages during the time interval. This information cannot be recovered
-from the gARG encoding 
-Conventions can be introduced to resolve such ambiguities;
+and thus so is the number of lineages during the time interval. In fact, this
+information cannot be recovered from the gARG encoding use in that figure,
+so alternative conventions must be introduced to resolve such ambiguities;
 for the likelihood-based inference algorithms for the coalescent with recombination,

### 2023-10-06 illustrations.py
Actually fix the numeric labels, whoops.

@@ -371,7 +371,7 @@ def simplification():
     tables.sort()
     ts=tables.tree_sequence()
 
-    labels = {i: i for i in range(len(string.ascii_lowercase))}
+    labels = {i: a for i, a in enumerate(string.ascii_lowercase)}
     # relabel the nodes to get samples reading A B C D
     labels.update({6: "e", 4: "g", 10: "j", 9: "k", 12: "l", 11: "m", 15: "o", 14: "p"})
     #labels = {i: i for i in range(26)}  # uncommment to get node IDs for re-positioning

### 2023-10-06 paper.tex
Updates from Overleaf

@@
 
-\linenumbers
 \title{\vspace{-1.5em} \bf A general and efficient representation of ancestral recombination graphs}
@@
 \renewcommand{\thefootnote}{\fnsymbol{footnote}}
-\footnotetext[1]{Shared second authorship, listed alphabetically}
+\footnotetext[1]{Joint second author, listed alphabetically}
 \footnotetext[2]{Correspondence: jerome.kelleher@bdi.ox.ac.uk}
@@
 
+\linenumbers
 \section*{Introduction}
@@
 appear~\citep{osmond2021estimating,fan2022genealogical,hejase2022deep,zhang2023biobank,
-nowbandegani2023extremely}
+nowbandegani2023extremely,ignatieva2023distribution}
 and many more are sure to
@@
 The word ``ARG'' is now widely used to refer to genetic genealogies in
-general~\citep[MORE][]{mathieson2020ancestry}, including
+general~\citep[][AND OTHERS]{mathieson2020ancestry}, including
 varied outputs from modern simulation and
@@
 % What do we do?
-In this paper we attempt to address these problems by
+In this paper, we attempt to address these problems by
 providing a precise definition of an ARG that is general enough to encompass
@@
 discussion we can label
-this prevailing interpretation as an ``Event ARG'', or eARG.
+this prevailing interpretation as an ``event ARG'', or eARG.
 (Many definitions also emphasise
@@
 % confuse people
-Thus, we have the means of encoding the \emph{effects} of
+Thus, we have the means to encode the \emph{effects} of
 multiple evolutionary events, and to
@@
 Note that many authors use the term ``marginal''
-trees~\citep[e.g.][]{griffiths1996ancestral,minichiello2006mapping} [more]
+trees~\citep[e.g.][AND OTHERS]{griffiths1996ancestral,minichiello2006mapping}
 which we avoid due the potential confusion with the
@@
 then we can recover precisely the same ARG. Thus, there
-is a one-to-one correspondance between an ARG and
+is a one-to-one correspondence between an ARG and
 the sequence of local trees that it encodes.
@@
 and note that the
-``full ARG ... contains more information than the series of tree
+``full ARG...~contains more information than the series of tree
 sequences along the genome''.
@@
 nodes that have at least two children, because these
-are in principle detectable from mutational differences
+are, in principle, detectable from mutational differences
 arising in the respective subtrees.
@@
 The first level of simplification that we can perform is to remove
-graph topology that is invisible to the samples.
+parts of the graph topology that are invisible to the samples.
 An example of such topology is a
@@
 common ancestor but non-coalescent nodes can produce graph nodes with
-more than two children (e.g.~node \noderef{s}). These represent a ``stacking up'' of
+more than two children (e.g.~node \noderef{s}). These represent merging
 the \emph{effects} of multiple evolutionary events in a single node (genome), and the
@@
 Another key factor in how important ARG-based methods become
-across the wide range of predicted application
+across the wide range of possible applications
 % TODO more?
@@
 fan2022genealogical,wohns2022unified,harris2023using,
-nowbandegani2023extremely}
+nowbandegani2023extremely,ignatieva2023duration}
 is how well inference methods and downstream analysis applications
@@
 are almost entirely unknown (beyond studies using very simplistic error models).
-A suite of ARG evaluation metrics, that take into account more of the
-global topology and can be applied to large ARGs, would be a
-valuable addition to the field.
+The development of ARG evaluation metrics that take into account more of the
+global topology and can be applied to large ARGs would be a
+valuable and timely addition to the field.
 
@@
 handling of some important forms of temporal uncertainty.
-For ARG based methods to achieve mainstream status, we require
+For ARG-based methods to achieve mainstream status, we require
 a rich supporting software ecosystem. 
@@
 \cite{minichiello2006mapping} explicitly distinguished the ARG as a
-stochastic process from the ARG as a data structure. We have followed them in
-using the term in the latter sense, to mean a specific realised ancestral history.
+stochastic process from the ARG as a data structure. We have followed them and
+used the term in the latter sense, to mean a specific realised ancestral history.
 
@@
 available for recombination in each time interval.
-This may not be obvious. For example, in the gARG encoding depicted in Figure \ref{fig-ancestry-resolution}C
+Some representations may not provide this information.
+For example, in the gARG encoding depicted in Figure \ref{fig-ancestry-resolution}C
 it is clear that a recombination takes place between nodes \textsf{b} and
@@
 and thus so is the number of lineages during the time interval. In fact, this
-information cannot be recovered from the gARG encoding use in that figure,
-so alternative conventions must be introduced to resolve such ambiguities;
-for the likelihood-based inference algorithms for the coalescent with recombination,
-the two parent nodes are typically created at the time of the recombination event.
-See the appendix of \cite{baumdicker2021efficient} for details of evaluating
+information cannot be recovered from the gARG encoding used in Figure \ref{fig-ancestry-resolution}C,
+and alternative conventions must be introduced to resolve such ambiguities.
+For the likelihood-based inference algorithms for the coalescent with recombination,
+it is sufficient to create \emph{two} gARG nodes at the time of the recombination event:
+see the appendix of \cite{baumdicker2021efficient} for details of evaluating
 coalescent with recombination likelihoods using this convention.
@@
 distribution of a stochastic process, such as the coalescent with
-recombination. Two contemporary examples are \cite{mahmoudi2022bayesian,
-guo2022recombination}. Evaluating the distribution of the coalescent with
+recombination. Two contemporary examples are \citet{mahmoudi2022bayesian} and
+\citet{guo2022recombination}. Evaluating the distribution of the coalescent with
 recombination (or related models) requires knowledge of the number of extant

### 2023-10-06 paper.tex
Various updates from Yan, Jerome & Ana

@@
 
-% FIXME this is weak, but gives the basic outline
+% FIXME abstract needs some rewriting, we're placing too much emphasis
+% on the coalescent there.
 \begin{abstract}
@@
 and contrast this with the classical event-based definition (eARG).
-The eARG definition is based on the coalescent with recombination
-stochastic process,
+The eARG definition is derived from on a stochastic process, 
+the coalescent with recombination, 
 and we show how the assumptions of that process
@@
 If not addressed, these problems will
-impede progress on applying genetic genealogical
+impede progress in applying genetic genealogical
 information to understand the biological world.
@@
 time ordering of recombination events.
+For a recent and more biologically oriented introduction to
+the topic, see \citet{lewanski2023era}.
 
@@
 We can define a gARG formally as follows.
-% TODO we should change the notation for inheritance intervals here,
-% as we use I to index individuals.
 Let $N \subset \mathbb{N}$ be the set of nodes representing
@@
 In contrast, ARGs are classically defined
-by the evolutionary \emph{events} in the history of a sample.
+by the evolutionary \emph{events} in the history of a sample of genomes.
 The difference between gARGs and this prevalent view of
@@
 The topology of an eARG defines the common ancestor and
-recombination events that occured,
+recombination events that occurred,
 but the graph structure alone is not sufficient to define the local trees,
@@
 consider the inference strategy of estimating local trees
-independently and subsequently stitching them together
+sequentially along the genome, and subsequently stitching them together
 into an ARG.
@@
 common ancestor but non-coalescent nodes can produce graph nodes with
-more than two children (e.g.~node \noderef{s}). These represent merging
-the \emph{effects} of multiple evolutionary events in a single node (genome), and the
+more than two children (e.g.~node \noderef{s}). These represent the 
+merged
+\emph{effects} of multiple evolutionary events in a single node (genome), and the
 ARG no longer contains the intermediate genomes representing those events.
@@
 fan2022genealogical,wohns2022unified,harris2023using,
-nowbandegani2023extremely,ignatieva2023duration}
+nowbandegani2023extremely,ignatieva2023distribution}
 is how well inference methods and downstream analysis applications
@@
 for ARG data interchange,
-users will suffer from problems of converting data from one programs
+users will suffer from problems with converting data from one program's
 output to another, an unfortunate hallmark of population genetics
@@
 It was originally developed as part of the \texttt{msprime}
-simulator~\citep{kelleher2016efficient} and subsequently been
+simulator~\citep{kelleher2016efficient} and has subsequently been
 extended and applied to forward-time
@@
 mahmoudi2022bayesian,petr2022slendr,rasmussen2022espalier,
-zhang2023biobank,nowbandegani2023extremely}.
+zhang2023biobank,nowbandegani2023extremely,ignatieva2023distribution}.
 
@@
 and likely many more.
-
-[TODO discussion of haplotype block based storage
-like ARGON. Basically: this is useful for some applications,
-but ultimately it's a derived property of the actual ARG
-topology that can be recovered efficiently, as demonstrated by 
-\cite{nowbandegani2023extremely}]
 
@@
 with recombination stochastic process is an important step.
-While it has proven to be a useful and
+While this has proven to be a useful and
 robust
@@
 lineages and breakpoints in special cases.
+% Keeping comments here - otherwise this para is clearly missing stuff
 [summarise info inferable about recomb in few sentenances here?
@@
 For example, the \relate\ selection test~\citep{speidel2019method}
-computes selection $p$-values by computing clade size probabilities conditional on the timing of coalescence events in a given local tree.
+obtains $p$-values by computing clade size probabilities conditional on the timing of coalescence events in a given local tree.
 In their method
@@
 utilising its rich suite of operations,
-would substantially accelerate progress, and help 
-establish ARG methods as mainstream genetics.
+may accelerate progress and help to
+establish ARG methods within mainstream genetics.
 

### 2023-10-06 paper.tex

@@
 and contrast this with the classical event-based definition (eARG).
-The eARG definition is derived from on a stochastic process, 
-the coalescent with recombination, 
-and we show how the assumptions of that process
-place limitations on the patterns of inheritance that can be
-represented, which are inadequate for modern large-scale datasets.
-The eARG representation also requires complete precision about the ordering
-of all recombination events, which we will rarely have sufficient
-information to infer.
+We show that eARGs are limited in the patterns
+of genetic inheritance that can be represented,
+and require complete precision about all recombination events.
 We show, in contrast, that the gARG definition
@@
 For a recent and more biologically oriented introduction to
-the topic, see \citet{lewanski2023era}.
+ARGs, see \citet{lewanski2023era}.
 

### 2023-10-08 paper.tex
Various edits from Yan and Ana

@@
 although there is some confusion about the interpretation and little
-agreement on the concrete details.
+agreement on specific details.
 We propose a concrete definition of ARGs
@@
 
-The common feature of these definitions is that they are expressed
+% NOTE: important to emphasise TWO things here, events and vagueness
+% actual encoding mechanism (i.e. how trees fit into the ARG)
+A common feature of these definitions is that they are expressed
 in terms of evolutionary \emph{events}, and so for the purposes of
@@
 i.e.\ are unary everywhere (e.g.\ \noderef{n} and \noderef{r}).
-(D) Rewrite edges to bypass nodes where they appear unary in local trees.
+(D) Rewrite edges to bypass nodes where they are unary in local trees.
 }
@@
 
-Uncertainty is a central element of inference about recombination,
-and, even with small samples, we can only expect to uniquely identify
-lineages and breakpoints in special cases.
-% Keeping comments here - otherwise this para is clearly missing stuff
-[summarise info inferable about recomb in few sentenances here?
-Cite: \citep{myers2002detection,hayman2023recoverability}]
+The timing, positions, and even the number of recombination events is generally
+not possible to infer precisely from sequencing data. For ARGs under 
+coalescent-based models, the proportion of recombination events that change the
+ARG topology grows very slowly with sample size \citep{hein2004gene}, and of those
+events only a small proportion are actually detectable from the data, assuming
+human-like mutation rates \citep{myers2002detection,hayman2023recoverability}.
+Even when a recombination event \emph{is} detectable, its timing and breakpoint
+positions can only be inferred approximately, depending on how much information
+can be elucidated from mutations in the surrounding genomic region. For general
+ARGs, for instance those representing pedigrees, or genealogies of organisms with
+low mutation rates and complex patterns of recombination, such inference can be
+even more challenging. The fact that the eARG encoding \emph{requires} completely
+precise estimates about recombination is therefore a fundamental limitation, which
+is a central message of this paper.
 % This might be where we mention Fisher's junctions and IBD?
 % https://github.com/tskit-dev/what-is-an-arg-paper/issues/121
-The fact that the classical ``event'' ARG (eARG) encoding
-\emph{requires} completely precise estimates about recombination
-is therefore a fundamental limitation, and a central message
-of this paper.
 
@@
 fundamental questions about properties
-of the ARG as a discrete graph structure, focussing on the ARG topology without considering
+of the ARG as a discrete graph structure, focusing on the ARG topology without considering
 either branch lengths or indeed the generating process.
@@
 other more general types of phylogenetic networks~\citep{huson2010phylogenetic}. The
-focus has been on algorithmic and combinatorial results that largely do not
-have direct relevance to the inference problems described above.
+focus has been on algorithmic and combinatorial results~\citep{gusfield2014recombinatorics}
+that are often not of
+direct relevance to the inference problems described above.
 
@@
 
-% FIXME putting this here for now just to see if this will work.
-
-% \section*{Event ARGs}\label{eARG}
-% In keeping with its original derivation in terms of a stochastic
-% process (see Appendix~\ref{sec-arg-history}), the classical definition of an
-% ARG data structure is
-% in terms of events.
-% Nodes represent
-% common ancestor or recombination events that occurred in the
-% history of some set of sample genomes, and edges represent ancestral
-% lineages~\citep{griffiths1996ancestral}.
-% In a common ancestor event the inbound lineages are merged into a
-% single ancestral lineage, and in a recombination
-% event a lineage is split into two independent
-% ancestral lineages.
-% % The final vital
-% % detail is that we associate a breakpoint with each recombination
-% % event.
-% % This information is sufficient to uniquely
-% % define the local genealogical trees at every position along the genome,
-% % which is the basic requirement for a data structure encoding
-% % genome-wide genetic ancestry.
-% As illustrated in Fig.~\ref{fig-arg-data-structure}, this information
-% is usually concretely encoded by associating a recombination breakpoint
-% (which is null for common ancestor events) and an ordered list of parents
-% with each node. We refer to such a  data structure as an
-% ``event ARG'' (eARG).
-% To recover the local tree at genome coordinate $x$ (the most
-% fundamental operation required for an ARG data structure),
-% we traverse the graph rootwards from the leaves.
-% At a particular
-% node $u$, if it has one parent we are at a common ancestor
-% node and we follow that parent. If we are at a
-% recombination node, $u$ has two parents: if
-% $x$ is less than the breakpoint we follow
-% the edge to the first parent, and otherwise follow the edge
-% to the second parent.
-% The order in which parents are listed at a recombination node is
-% therefore significant, telling us
-% from which parent the segments of genome to the left and right of the breakpoint
-% were inherited.
-
-% This ordering requirement, while straightforward
-% to describe, has some practical drawbacks. For example,
-% % Surely the meaning is clear from the context here and we don't need
-% % qualify the meaning of ``ARG''
-% when using this representation and simulating an ARG backwards in time,
-% the first event older than a recombination may involve the lineage carrying the
-% ancestry to the right of the breakpoint rather
-% than the left, and so we cannot emit edges as they are generated.
-% Such issues can be worked around, of course, but depending on the ordering of
-% otherwise indistinguishable objects is generally problematic.
-
-% \begin{figure}
-% \centering
-% % \begin{tabular}{cc}
-% \begin{tikzpicture}[x=5mm, y=5mm, node distance=2mm and 20mm]
-% \tikzset{greynode/.style={circle,fill,inner sep=1},
-% nodelabel/.style={font=\footnotesize}}
-
-% \node (s0) [greynode] at (0, 0) {};
-% \node (s1) [greynode] at (3, 0) {};
-% \node (s2) [greynode] at (6, 0) {};
-% \node (s3) [greynode] at (3, 1) {};
-% \node (s4) [greynode] at (1, 2) {};
-% \node (s5) [greynode] at (5, 3) {};
-% \node (s6) [greynode] at (3, 4) {};
-
-% % \node [anchor=north west] at (0,6) {A};
-% \node [nodelabel,anchor=north west] at ($(s3) + (0,0)$) {$x$};
-% \foreach \u/\lab in {s0/$\textsf{a}$, s1/$\textsf{b}$, s2/$\textsf{c}$} \node[nodelabel,anchor=north] at (\u) {\lab};
-% \foreach \u/\lab in {s4/$\textsf{e}$} \node[nodelabel,anchor=south west] at (\u) {\lab};
-% \foreach \u/\lab in {s5/$\textsf{f}$} \node[nodelabel,anchor=south east] at (\u) {\lab};
-% \foreach \u/\lab in {s3/$\textsf{d}$, s6/$\textsf{g}$} \node[nodelabel,anchor=south] at (\u) {\lab};
-
-% %% Edges
-% \draw (s1) -- (s3);
-% \draw (s0) |- (s4);
-% \draw (s4) -- (2,2) |- (s3);
-% \draw (s4) |- (s6);
-% \draw (s3) -- (4,1) |- (s5);
-% \draw (s2) |- (s5);
-% \draw (s5) |- (s6);
-
-% % \node [anchor=north west] at (9,6) {B};
-% \node [nodelabel,anchor=north west] at ($(10,5)$) {
-% \begin{tabular}{c|c|l}
-% % \multicolumn{2}{c}{Breakpoints}\\
-% Node & Breakpoint & Parents\\
-% \hline
-% $\noderef{a}$ & $\varnothing$ & [\noderef{e}]\\
-% $\noderef{b}$ & $\varnothing$ & [\noderef{d}] \\
-% $\noderef{c}$ & $\varnothing$ & [\noderef{f}]\\
-% $\noderef{d}$ & $x$ & [\noderef{e}, \noderef{f}] \\
-% $\noderef{e}$ & $\varnothing$ & [\noderef{g}]\\
-% $\noderef{f}$ & $\varnothing$ & [\noderef{g}]\\
-% $\noderef{g}$ & $\varnothing$ & []\\
-% \end{tabular}};
-
-% % \node [anchor=north west] at (9,6) {C};
-% % \node [nodelabel,anchor=north west] at ($(10,5)$) {
-% % \begin{tabular}{ll}
-% % Edges & Intervals\\
-% % \hline
-% % $(\textsf{a}, \textsf{e})$ & $[0, L)$ \\
-% % $(\textsf{b}, \textsf{d})$ & $[0, L)$ \\
-% % $(\textsf{c}, \textsf{f})$ & $[0, L)$ \\
-% % $(\textsf{d}, \textsf{e})$ & $[0, x)$ \\
-% % $(\textsf{d}, \textsf{f})$ & $[x, L)$ \\
-% % $(\textsf{e}, \textsf{g})$ & $[0, L)$ \\
-% % $(\textsf{f}, \textsf{g})$ & $[0, L)$ \\
-% % \end{tabular}};
-
-% \end{tikzpicture}
-% \caption{\label{fig-arg-data-structure}
-% A graphical depiction of a classical event ARG and a concrete
-% encoding of the information. Node \noderef{d} is a recombination
-% event as it is associated with a non-null breakpoint, and has
-% two parents. Nodes \noderef{e}, \noderef{f} and \noderef{g} are
-% common ancestor events.
-% }
-% \end{figure}
-
-% % What do we mean by an eARG, formally?
-% To be concrete, we can formally define the
-% classical Griffiths eARG as a tuple $(e, \sigma)$, where $e$
-% is an ordered list of edges defining a directed acyclic graph and
-% $\sigma: \mathbb{N} \rightarrow \mathbb{N}$
-% is a function mapping nodes to recombination breakpoints,
-% such that $\sigma(u) = x$
-% if $u$ is a recombination event with breakpoint $x$ and
-% $\sigma(u) = \varnothing$ otherwise.
-% (This is equivalent to having different ``types''
-% of node, where breakpoints are associated with recombination
-% nodes only).
-% Each edge $e_j = (c, p)$ describes an edge between
-% child node $c$ and parent $p$, where $c, p \in \mathbb{N}$.
-% % For each recombination node $u$ we must
-% % have exactly two edges in $e$ such that $u$ is the child.
-% This encoding of an eARG is equivalent to the output
-% of programs such as
-% \texttt{ARGweaver}~\citep{rasmussen2014genome}
-% and \texttt{KwARG}~\citep{ignatieva2021kwarg},
-% and an example is given in Fig.~\ref{fig-arg-data-structure}A.
-% Note that the description above captures only the
-% graph topology: if we also wish to know the branch lengths we need
-% an additional function $t(u): \mathbb{N} \rightarrow \mathbb{R}$
-% which defines the time of each node.
-
 \section{The Big and Little ARG}
 \label{app-big-and-little-arg}
-
 Here we review the two predominant stochastic processes which construct ARGs:

### 2023-10-10 paper.tex
Tweak
Fig 6 fixed

@@
 
-Genome ARGs, on the hand,
-can be either retrospective or \emph{prospective}.
+There is no particular reason, however, that ARGs must be 
+retrospective. They can also be \emph{prospective}.
 The edges in a gARG represent a path of genetic inheritance from
@@
 ancestor lineage. Otherwise, if $a_1 + a_2 = n$, we have found
-the most recent common ancestor of all samples in the interval$[\ell, r)$
+the most recent common ancestor of all samples in the interval $[\ell, r)$
 and do not need to simulate its history any further.
@@
 In order to evaluate these rates, and hence the sampling probability
-(see e.g.\ \cite[Equation (3)]{mahmoudi2022bayesian}),
+\cite[see e.g.][Equation (3)]{mahmoudi2022bayesian},
 it is necessary to know the number of lineages and number of extant links
@@
 	\draw (0.9, 1.5) -- (1.4, 1.5) -- (1.4, 1.4) -- (0.9, 1.4) -- (0.9, 1.5);
-	\draw (1.2, 1.6) -- (1.2, 1.3);
-	\node [label=below:{\small $x$}] at (1.2, 1.5) {};
+	\draw (1.1, 1.6) -- (1.1, 1.3);
+	\node [label=below:{\small $w$}] at (1.1, 1.5) {};
 
@@
 	\draw (1, 1.7) -- (1, 3.1) -- (0,3.1) -- (2,3.1);
-	\draw (-0.4, 2.85) -- (0.2, 2.85);
-	\draw (0.2, 2.9) -- (0.4, 2.9) -- (0.4, 2.8) -- (0.2, 2.8) -- (0.2, 2.9);
+	\draw (-0.4, 2.85) -- (0.1, 2.85);
+	\draw (0.1, 2.9) -- (0.4, 2.9) -- (0.4, 2.8) -- (0.1, 2.8) -- (0.1, 2.9);
 	\draw (1.6, 2.85) -- (1.9, 2.85);
-	\draw (1.9, 2.9) -- (2.2, 2.9) -- (2.2, 2.8) -- (1.9, 2.8) -- (1.9, 2.9);
-	\draw (2.2, 2.85) -- (2.4, 2.85);
+	\draw (1.9, 2.9) -- (2.1, 2.9) -- (2.1, 2.8) -- (1.9, 2.8) -- (1.9, 2.9);
+	\draw (2.1, 2.85) -- (2.4, 2.85);
 
@@
 	\draw (-0.9, 4.3) -- (-0.6, 4.3) -- (-0.6, 4.2) -- (-0.9, 4.2) -- (-0.9, 4.3);
-	\draw (-0.6, 4.25) -- (-0.3, 4.25);
-	\draw (-0.3, 4.3) -- (-0.1, 4.3) -- (-0.1, 4.2) -- (-0.3, 4.2) -- (-0.3, 4.3);
+	\draw (-0.6, 4.25) -- (-0.4, 4.25);
+	\draw (-0.4, 4.3) -- (-0.1, 4.3) -- (-0.1, 4.2) -- (-0.4, 4.2) -- (-0.4, 4.3);
 	\draw (-0.5, 4.4) -- (-0.5, 4.1);
-	\node [label=below:{\small $y$}] at (-0.5, 4.3) {};
+	\node [label=below:{\small $x$}] at (-0.5, 4.3) {};
 
@@
 	\draw (-0.5, 4.5) -- (-0.5, 6) -- (-1.5,6) -- (0.5,6);
-	\draw (-1.9, 5.75) -- (-1.3, 5.75);
-	\draw (-1.3, 5.8) -- (-1.1, 5.8) -- (-1.1, 5.7) -- (-1.3, 5.7) -- (-1.3, 5.8);
+	\draw (-1.9, 5.75) -- (-1.4, 5.75);
+	\draw (-1.4, 5.8) -- (-1.1, 5.8) -- (-1.1, 5.7) -- (-1.4, 5.7) -- (-1.4, 5.8);
 	\draw (0.1, 5.8) -- (0.4, 5.8) -- (0.4, 5.7) -- (0.1, 5.7) -- (0.1, 5.8);
@@
 	\node [label=left:{\small$\{(w,m,2)\}$}] at (6.2,5) {};
-	\node [label=right:{\small$\{(x,m,1)\}$}] at (-0.2,3.8) {};
-	\node [label=right:{\small$\{(v,x,1)\}$}] at (1.8,4.1) {};
-	\node [label=left:{\small$\{(0,v,1), (x,m,1)\}$}] at (-0.3,5.2) {};
+	\node [label=right:{\small$\{(w,m,1)\}$}] at (-0.2,3.8) {};
+	\node [label=right:{\small$\{(v,w,1)\}$}] at (1.8,4.1) {};
+	\node [label=left:{\small$\{(0,v,1), (w,m,1)\}$}] at (-0.3,5.2) {};
 	\node [label=right:{\small$\{(0,v,2)\}$}] at (2.8,6.4) {};
-	\node [label=left:{\small$\{(x,m,1)\}$}] at (-1.3,7) {};
+	\node [label=left:{\small$\{(w,m,1)\}$}] at (-1.3,7) {};
 	\node [label=right:{\small$\{(0,v,1)\}$}] at (0.3,6.4) {};
@@
 	\node[label=right:{CA \; RE}] at (7.5, 8.2) {};
-	\node[label=right:{$\binom{2}{2}$ \; $\frac{2 m - x - w - 2}{m - 1} \rho$}] at (7.5, 7.6) {};
-	\node[label=right:{$\binom{4}{2}$ \; $\frac{2 m + 2 v - x - w - 3}{m - 1} \rho$}] at (7.5, 6.6) {};
+	\node[label=right:{$\binom{2}{2}$ \; $\frac{2 m - 2 w - 2}{m - 1} \rho$}] at (7.5, 7.6) {};
+	\node[label=right:{$\binom{4}{2}$ \; $\frac{2 m + 2 v - 2 w - 4}{m - 1} \rho$}] at (7.5, 6.6) {};
 	\node[label=right:{$\binom{3}{2}$ \; $\frac{2 m + v - w - 3}{m - 1} \rho$}] at (7.5, 5.7) {};
-	\node[label=right:{$\binom{4}{2}$ \; $\frac{2 m + x - v - 4}{m - 1} \rho$}] at (7.5, 4.95) {};
+	\node[label=right:{$\binom{4}{2}$ \; $\frac{2 m + w - v - 4}{m - 1} \rho$}] at (7.5, 4.95) {};
 	\node[label=right:{$\binom{5}{2}$ \; $\frac{2 m - 5}{m - 1} \rho$}] at (7.5, 3.8) {};
@@
 	\draw [color=red](5.8, 2.3) -- (5.8, 2.0);
-	\node [label={[red]below:{\small $z$}}] at (5.8, 2.1) {};
+	\node [label={[red]below:{\small $y$}}] at (5.8, 2.1) {};
 
 	\draw (1, 1.7) -- (1, 3.1) -- (0,3.1) -- (2,3.1);
-	\draw (-0.4, 2.85) -- (0.2, 2.85);
-	\draw (0.2, 2.9) -- (0.4, 2.9) -- (0.4, 2.8) -- (0.2, 2.8) -- (0.2, 2.9);
+	\draw (-0.4, 2.85) -- (0.1, 2.85);
+	\draw (0.1, 2.9) -- (0.4, 2.9) -- (0.4, 2.8) -- (0.1, 2.8) -- (0.1, 2.9);
 	\draw (1.6, 2.85) -- (1.9, 2.85);
-	\draw (1.9, 2.9) -- (2.2, 2.9) -- (2.2, 2.8) -- (1.9, 2.8) -- (1.9, 2.9);
-	\draw (2.2, 2.85) -- (2.4, 2.85);
+	\draw (1.9, 2.9) -- (2.1, 2.9) -- (2.1, 2.8) -- (1.9, 2.8) -- (1.9, 2.9);
+	\draw (2.1, 2.85) -- (2.4, 2.85);
 
@@
 A realisation of the graph traversed by Hudson's algorithm started from a
-sample of three chromosomes of length $m$ at time $t = 0$, and
+sample of three chromosomes with $m$ discrete sites each at time $t = 0$, and
 propagated until time $T$. The MRCA on the genetic interval $[v, w)$ is reached
@@
 The non-ancestral segment $[v, w)$ above
-A contributes to the rate of effective recombinations because it
+\textsf{a} contributes to the rate of effective recombinations because it
 is trapped between ancestral segments. The two columns titled CA and RE
@@
 dynamics, at the cost of extra nodes and edges, highlighted in red, which do
-not affect the local tree at any site.}
+not affect the local tree at any site.
+Recombination positions are labelled alphabetically in time, and their ordering
+along the genome is $y < v < x < w$, of which the first only appears in panel B.
+There are two separate recombination events at link $w$.
+}
 \label{hudson_vs_bigARG}

### 2023-10-10 paper.tex
Edits from yan

--- 2023-10-08 paper.tex
@@
 Nodes are coloured by the number of parents and shaded
-according to the proportion of their span over which they are coalescent.
+according to the proportion of their span over which they are coalescent
+(legend in Fig.~\ref{fig-inferred-args})
 (A) A gARG simulated from a backward-time diploid Wright-Fisher
@@
 i.e.\ are unary everywhere (e.g.\ \noderef{n} and \noderef{r}).
-(D) Rewrite edges to bypass nodes where they are unary in local trees.
+(D) Rewrite edges to bypass nodes where they are locally unary.
 }

### 2023-10-12 paper.tex
Tweak
Pass at locally unary nodes section

@@
 
-A genome ARG (gARG) is a graph in which nodes represent
+A genome ARG (gARG) is a directed acyclic graph in which nodes represent
 haploid genomes and edges represent
@@
 An example genome ARG embedded in a pedigree.
-(A) Diploid individuals, visualised in a highly inbred pedigree and
+(A) Diploid individuals (blue), visualised in a highly inbred pedigree and
 labelled $D_1$ to $D_8$,
@@
 Genomes \textsf{a} and \textsf{c} are the product of two independent
-recombination events between
+recombination events (red) between
 the paternal genomes \textsf{e}
-and \textsf{f} (with breakpoints in red), and regions of genome inherited are shown.
+and \textsf{f}, and regions of genome inherited are shown.
 Genomes are shaded such that where, backwards in time,
@@
 the gARG for a given set of individuals is embedded in their pedigree.
-In this diploid example, each genome is inherited from one
-of the individual's parents,
-and may be the recombined product of that parent's two genomes.
-For example, diploid individual $D_1$ in Fig.~\ref{fig-arg-in-pedigree}
-has two genomes \noderef{a} and \noderef{b},
-inherited from parents $D_3$ and $D_4$.
-Genome \noderef{b} was inherited directly from $D_4$'s genome \noderef{g} without
+In this figure, diploid individuals are labelled $D_1, D_2...$, and
+rather than denoting their constituent genomes using the natural numbers
+we use lowercase alphabetical labels, to avoid confusion with genomic
+positions. Individual $D_1$ is therefore composed
+of two genomes \noderef{a} and \noderef{b}, which are inherited from its
+two parents, $D_3$ and $D_4$. Each inherited genome may be the recombined product
+of the genomes contained within a parent. In this example,
+genome \noderef{b} was inherited directly from $D_4$'s genome \noderef{g} without
 recombination, whereas
@@
 the corresponding crossover breakpoint~\citep{griffiths1996ancestral}.
-Fig.~\ref{fig-arg-data-structure} shows an example of a classical
+Fig.~\ref{fig-event-arg} shows an example of a classical
 eARG with three samples (\noderef{a}, \noderef{b}, and \noderef{c})
@@
 left of $x$ from \noderef{e} and to the right of $x$ from \noderef{f}.
-As shown in Fig.~\ref{fig-arg-data-structure}B,
+As shown in Fig.~\ref{fig-event-arg}B,
 the local trees are embedded in the graph, and can be reconstructed
@@
 \end{tikzpicture}
-\caption{\label{fig-arg-data-structure}
+\caption{\label{fig-event-arg}
 A classical event ARG. (A) Standard graph depiction with
@@
 From a concrete data-structure encoding perspective, the only real
-difference between an eARG and a gARG is that we store recombination
+difference between an eARG and a gARG is that in an eARG we store recombination
 \emph{breakpoints} at nodes of out-degree two (recombination nodes)
-in an eARG and we store \emph{inheritance intervals} on the edges
-of a gARG. A key advantage of associating inheritance intervals with
+whereas in a gARG we store \emph{inheritance intervals} on edges.
+A key advantage of associating inheritance intervals with
 edges in a gARG is that it allows us to describe evolutionary events other
@@
 \citet{kelleher2018efficient}.
-It is a powerful tool for simulation and analysis, as it
-allows to efficiently subset ARGs both in terms of the
-samples and levels of detail about recombination retained,
+It is a powerful tool for simulation and analysis as it
+allows efficient subsetting of ARGs, with the ability to
+remove both samples and certain levels of detail about recombination,
 as we discuss in the following sections.
@@
 oriented tree~\cite[p.\ 461]{knuth11combinatorial},
-a sequence of integers $\pi_1\pi_2\dots$ where $\pi_u$ denotes the parent of
+a sequence of integers $\pi_1, \pi_2\dots$ where $\pi_u$ denotes the parent of
 node $u$~\citep{kelleher2013coalescent,kelleher2014coalescent,kelleher2016efficient}.
@@
 Oriented trees allow nodes
-to have  any number of children; from one (see section XXX for more
-discussion on such ``unary nodes'' in the local trees), to arbitrarily
+to have  any number of children; from one (see the next section for more
+discussion on such ``unary nodes''), to arbitrarily
 large polytomies.
@@
 Specifically, at a given node $u$,
-we an find edge $(c, p, I) \in E$ such that $u = c$ and $x \in I$, and set
+we find an edge $(c, p, I) \in E$ such that $u = c$ and $x \in I$, and set
 $\pi_c \leftarrow p$. We then set $u \leftarrow p$, and repeat
@@
 expensive to compute and imprecise. For example
-we could infer that node \textsf{k} is shared in the second
-and third trees of in Fig.~\ref{fig-simplification}D
-because it subtends identical subtrees in both,
-but it would not be straightforward to infer that it
-is also the root of the first tree.
-
+using only the tree topologies it would not be straightforward to
+infer that node \noderef{i} in the first and last trees of
+Fig.~\ref{fig-arg-in-pedigree}C is the same node.
 \cite{shipilina2023origin} discuss the same ideas,
@@
 contains more information than its local trees is true only if
-the local trees are unlabeled and/or do not contain unary nodes.
-However, as we have shown, if the local trees
-contain labelled internal (possibly unary) nodes,
+the local trees are unlabeled and/or do not contain unary nodes
+(see the next section for discussion on this point).
+However, if the local trees
+contain labelled internal nodes,
 then  there is precisely as
 much information in them as an ARG.
-Whether such unary nodes are \emph{observable}
-is an entirely separate question, of course.
 
@@
 % trees from different regions of the genome.
+
+\section*{Locally unary nodes}
+The arity of a node is the number of child nodes it has. There is an
+important distinction between the number of children a node has
+in the graph and the number of children it has in the local trees,
+because a (graph) child node is only a child in the local tree for position
+$x$ if $x$ is in the inheritance interval for the corresponding edge.
+Thus, for a given node $u$ and a genome position $x$ the arity
+of $u$ in that local tree is less than or equal to the graph arity of
+$u$. The arity
+of a node in the local trees can therefore change as we move along
+the genome.
+
+This distinction between graph and local tree arity is mainly
+of interest when we consider \emph{unary} nodes, those that have
+a single child.
+Returning to the example in Fig~\ref{fig-arg-in-pedigree}, nodes
+\noderef{g} and \noderef{h} are graph-unary, and are consequently
+also unary in the local trees.
+On the other hand, node \noderef{f} has two children
+in the graph (Fig~\ref{fig-arg-in-pedigree}B), but is only binary
+in the local tree covering the interval $[2, 7)$,
+representing the coalescence of samples \noderef{a} and \noderef{c}
+in this region of genome. Over the interval $[0, 2)$ no coalescence occurs,
+but we still record the fact that genome \noderef{c} inherits from \noderef{f}
+in the local tree. Thus, \noderef{c} has a single child in this
+interval and it is unary in this local tree.
+In another example, \noderef{e} is binary in the graph, but is
+unary in each of the local trees it is present in.
+This is because \noderef{a} and \noderef{c} are both recombinant
+offspring of \noderef{e} and \noderef{f}, and,
+in this example,
+\noderef{a} inherits from
+\noderef{e} over the interval $[0, 2)$,
+and \noderef{c} inherits from \noderef{e} over $[7, 10)$.
+The intervals over which
+we know that a node inherits from another, but where there
+is no local coalescence, are marked by unary nodes in the
+local trees.
+
+Outside of the context of simulation, graph-unary nodes are
+only like to occur in longitudinal datasets where genetic
+data is sampled at many timepoints and recombination
+is rare (e.g.~SARS-Cov-2; see Discussion).
+\emph{Locally} unary nodes, those which have one child
+in one or more local trees, are a common and important feature
+of many different types of ARG, and a recurring topic
+in this article. Without these nodes marking the passage
+of ancestral material through specific ancestors, the local trees
+lack information about events other than local coalescence.
+% Not sure if switching example will confuse people here, but would
+% like to contrast with example where we don't use unary nodes to
+% emphasise the point that we lose info
+For example, the local trees for the classical event ARG
+depicted in Fig.~\ref{fig-event-arg}B follow the usual conventions
+and do not include any information about the recombination
+that occurred at node \noderef{d}. Given these two local trees
+in isolation
+% (without further information about the corresponding ARG)
+we lack specific information about the recombination.
+Explicitly recording that node \noderef{d} lies on the
+branch joining \noderef{b} to \noderef{e} in the left hand
+tree, and \noderef{b} to \noderef{f} in the right hand tree
+resolves all ambiguity, and makes the collection of local
+trees exactly equivalent to the corresponding ARG (see previous section).
+Unary nodes are a vital link between ARGs and local trees, and we
+cannot fully reason about how a local tree is embedded in an ARG
+without them.
 
@@
 
-\begin{figure}[ht]
+\begin{figure}[t]
 \centering

### 2023-10-12 paper.tex
Unary nodes

@@
 
+Nodes with one child are not a standard feature of evolutionary trees
+or the ARG literature
+(although \citet{mathieson2020ancestry} show examples in which the ARGs
+contain unary nodes, and state that paths are
+``implicitly passing through points representing specific individuals.'').
+We are usually interested in
+nodes that have at least two children because these
+are in principle detectable from mutational differences
+arising in the respective subtrees.
+A tree branch may correspond to many generations
+of individuals (or indeed cell divisions; see Appendix XXX)
+and there is usually
+little information about these intermediates, or utility
+in modelling them.
+As we see in the next section, however, there is a range of
+different ways in which nodes can be unary in the local trees
+of an ARG. Furthermore, ARGs inferred by current methods contain
+locally unary nodes from several of them (see section XXX).
+
 \section*{Levels of simplification}\label{ARG_simplification}
-ARG simplification is a powerful tool, and can do much more
-besides removing unreachable nodes and edges in a prospective ARG
+ARG simplification is a powerful tool, and can do far more
+than removing unreachable nodes and edges in a prospective ARG
 and resolving ancestral material in an eARG.
@@
 The redundancy that we are interested in
-revolves around the presence of nodes that appear
-``unary'' in one or more local trees
-(where a node has exactly one child over a particular genomic region).
-These arise in several different ways.
-Nodes with one child are not a standard feature of evolutionary trees;
-we are usually interested in
-nodes that have at least two children, because these
-are, in principle, detectable from mutational differences
-arising in the respective subtrees.
-A tree branch may correspond to many generations
-of individuals
-% Following feedback from Wilder, this is more likely to confuse
-% people unless we properly explain
-% (or indeed cell divisions)
-and there is usually
-very little information about these intermediates, or utility
-in modelling them.
-% FIXME the bridge between the next few paragraphs is a bit weird, in
-% that we've just been talking about how interesting unary nodes are
-% but then go on about how to remove them. Just needs a bit of smoothing
-% out I think.
-ARGs, however, are different, and there
-can be significant information about evolutionary processes
-provided by nodes that are unary in one or more local trees.
-% Not sure what to do with this - should mention though.
-\citet{mathieson2020ancestry} show examples in which the ARGs
-contain unary nodes, and state that paths are
-``implicitly passing through points representing specific individuals.''
-% Full quote from fig caption:
-% Genetic ancestry in the form of the ARG for a single individual. Combining
-% genetic ancestry from different positions leads to a graph, incorporating all
-% realized genetic ancestry paths, implicitly passing through points representing
-% specific individuals.
+% We are interested in both graph and local unary, so qualifying this
+% would just make it less readable, IMO
+revolves around the presence unary nodes (see previous section),
+which we illustrate by an example simulation.
+[THIS SIMULATION...]
 
@@
 local trees on the right.
+
 In Fig.~\ref{fig-simplification}A the local trees (right)
@@
 
-% JK: Text moved from an earlier section. See if we want to reuse this here
-% if not already said.
-% Indeed a single node can have both multiple children
-% and multiple parents. This would be the case in Fig~\ref{fig-arg-in-pedigree} if,
-% for instance, node \noderef{a} were to have two children.
-% % It is a shame that this is not actually shown in fig 3, but I don't think
-% % we can do this unless we have 5 generations
-% Moreover, a single node can have more than 2 children, and (as we shall see in
-% the ``\nameref{ARG_simplification}'' section), more than 2 parents. Another way to
-% state this is that transitions between nodes in a gARG can encapsulate multiple
-% events in an eARG.
-
 The remaining nodes are MRCAs of some subset of the samples
@@
 the graph, we lose potentially useful information about shared edges
-between trees. The extent to which this information can be inferred
-and utilised is an interesting open question.
-% I don't think there's an actual manuscript or title to refer to here
-% [CITE Ralph in prep.].
+between trees.
+The \texttt{msprime} simulator, and the version of Hudson's algorithm described
+by~\citet{kelleher2016efficient}, produces ARGs
+that are fully simplified (i.e., contain no locally unary nodes).
+It is not difficult, however, to update
+these methods to record information about the passage of ancestral
+material through genomes under a range conditions.
 

### 2023-10-12 paper.tex
Yans updates (applying comments from Andrew)

--- 2023-10-10 paper.tex
@@
 expensive to compute and imprecise. For example
-we could infer that node \textsf{k} is shared in the second
-and third trees of in Fig.~\ref{fig-simplification}D
-because it subtends identical subtrees in both,
-but it would not be straightforward to infer that it
-is also the root of the first tree.
+using only the tree topologies it would not be straightforward to
+infer that node \noderef{i} in the first and last trees of
+Fig.~.\ref{fig-arg-in-pedigree}C is the same node.
 

### 2023-10-12 illustrations.py
Change KwARG to 1-re-node

@@ -160,8 +160,7 @@ def ancestry_resolution():
         width=1,
     )
 
-    ts_simp = argutils.simplify_keeping_all_nodes(
-        argutils.remove_edges_above_local_roots(ts))
+    ts_simp = argutils.remove_edges_above_local_roots(ts).simplify(filter_nodes=False, keep_unary=True)
     pos, G = argutils.viz.draw(
         ts_simp,
         ax3,
@@ -842,7 +841,7 @@ def inference():
         'xmlns:xlink="http://www.w3.org/1999/xlink">'
     ]
     svg.append('<g transform="">' + graph_svg[graph_svg.find("<svg"):] + "</g>")
-    svg.append('<g transform="translate(160, 30) scale(0.5)">' + legend_svg() + "</g>")
+    svg.append('<g transform="translate(150, 30) scale(0.5)">' + legend_svg() + "</g>")
     svg.append('</svg>')
     top_svg = (
         '<!DOCTYPE svg PUBLIC "-//W3C//DTD SVG 1.1//EN" '

### 2023-10-12 make_example_outputs.py

@@ -59,8 +59,19 @@ def run_kwarg():
                 s.id: sample_data.individual(s.individual).metadata["name"]
                 for s in sample_data.samples()
             },
+            two_re_nodes=True
         )
-        # Label the samples
+        ts.dump("examples/Kreitman_SNP_kwarg-2RE.trees")
+    with open("examples/kreitman.kwarg") as f:
+        ts = argutils.convert_kwarg(
+            f,
+            11,
+            43,
+            sample_names={
+                s.id: sample_data.individual(s.individual).metadata["name"]
+                for s in sample_data.samples()
+            },
+        )
         ts.dump("examples/Kreitman_SNP_kwarg.trees")
 
 

### 2023-10-12 paper.tex

--- 2023-10-12 paper.tex
@@
 \section*{Locally unary nodes}
-The arity of a node is the number of child nodes it has. There is an
+We can define the arity of a node as the number of child nodes it has.
+There is an
 important distinction between the number of children a node has
@@
 revolves around the presence unary nodes (see previous section),
-which we illustrate by an example simulation.
-[THIS SIMULATION...]
+which we illustrate by an example simulated ARG, which
+undergoes a series of successive steps of simplification
+in Fig.~\ref{fig-simplification}.
 
@@
 Levels of ARG simplification.
-In all cases the graph visualization is shown on the left
-and the corresponding local trees on the right.
-Nodes are coloured by the number of parents and shaded
-according to the proportion of their span over which they are coalescent
-(legend in Fig.~\ref{fig-inferred-args})
-(A) A gARG simulated from a backward-time diploid Wright-Fisher
-model: genome \noderef{k} has three children, and the genomes
-\noderef{e} and \noderef{g} simultaneously have multiple parents and multiple children.
-(B) Simplified to remove all
+(A) An example ARG simulated from a diploid Wright-Fisher model.
+(B) Remove all
 singly-connected graph components (e.g., diamonds such as \noderef{jlnm}).
@@
 i.e.\ are unary everywhere (e.g.\ \noderef{n} and \noderef{r}).
-(D) Rewrite edges to bypass nodes where they are locally unary.
+(D) Rewrite edges to bypass nodes in local trees in which they are unary.
+In each case, the graph is shown on the left
+and corresponding local trees on the right.
+Graph nodes are coloured by the number of parents and shaded
+according to the proportion of their span over which they are coalescent;
+see the text for more details.
 }
@@
 
-Fig.~\ref{fig-simplification} shows a series of successive steps of
-simplification,
-% FIXME this doesn't explain anything - what did we store in the simulation?
-starting with a sample-resolved but unsimplified gARG, simulated under a
-backward-time Wright-Fisher
-model. We show the graph on the left, and the corresponding sequence of
-local trees on the right.
-
+The ARG in Fig.~\ref{fig-simplification}A is the output of a
+backwards-time Wright-Fisher simulation for a sample of 2 diploid
+individuals (population size $N=10$), and follows a similar process
+to the methods described by~\cite{nelson2020accounting}.
+As we proceed backwards in time, generation by generation, the
+extant lineages chose parents randomly.
+With a certain probability recombination occurs, and the ancestral
+material of a lineage is split between the two parental
+genomes. Local coalesence occurs
+when lineages with overlapping ancestral material choose the same
+parent genome.
+Note that in this simulation we do not explicitly
+model recombination \emph{events} via an ARG node, but simply record
+the \emph{effects} of a recombination via outbound edges to
+the parent individual's two genomes.
+Thus, a recombinant node may also correspond to a coalescence
+(e.g.~node \noderef{g} in Fig.~\ref{fig-simplification}).
+The distinction of
+using a single node to represent
+a recombination event (e.g.~Fig~\ref{fig-ancestry-resolution}),
+or two to represent the parent genomes is usually not important
+(either is possible in the gARG encoding)
+and the most convenient approach will vary by application.
+Note also that polytomies are a natural feature of such
+a Wright-Fisher model
+(e.g.~node \noderef{k} in Fig.~\ref{fig-simplification}).
+
+Fig.~\ref{fig-simplification} shows four ARGs (A--D), with
+the graph depicted on the left and corresponding sequence
+of local trees on the right. The graph visualisations have
+some novel features, which require some explanation.
+Firstly, edge weights (the thickness of the lines joining
+nodes) correspond to the size of the inheritance intervals
+they are annotated with. This allows us to distinguish
+edges that persist across many local trees from those that are
+less influential (contrast the edge
+$(\noderef{g}, \noderef{h})$
+with $(\noderef{g}, \noderef{i})$
+in Fig.~\ref{fig-simplification}A).
+% FIXME this needs some work
+Secondly, node colours denote the number of parents that they
+have in the graph, allowing us to easily see roots (those
+with zero parents), recombinanants (those with two parents)
+and more complex situations arising from simplification (see below).
+Thirdly, the shading intensity of a node denotes the
+fraction of the node's span (the length of genome in which it
+is reachable from the samples in the local trees)
+over which it is coalescent (has more than one child).
+% This is useful why? I can't come up with a quick way of saying
+% this. In some senses it's a measure of identifiability?
+
+Returning to the main topic of this section,
+Fig.~\ref{fig-simplification}A is the direct output of a Wright-Fisher
+simulation in which we retain all nodes involved in recombination
+or common ancestry events. This is the true history, and contains
+a very high level of detail, some of which may be considered
+redundant (or, from another perspective, unobservable).
 In Fig.~\ref{fig-simplification}A the local trees (right)
@@
 (lighter: older; darker: younger).
-Line width corresponds to the genomic span of the edge.
 Position on the X and Y axes for the graph visualization is arbitrary.
- Node colour
-refers to the number of parents of the node.
-The fill colour intensity of a node is determined by the genomic span over which
-it has two or more children divided by the span where it has any
-children (the ``coalescent span'').
+Line width and node colour are as described in Fig.\ref{fig-simplification}.
 The genomic intervals associated with edges are shown below

### 2023-10-13 paper.tex
Gregor's updates

@@
 and many more are sure to
-follow~\citep{harris2019database,harris2023using}.
+follow~\citep{harris2019database,harris2023using}. \citet{lewanski2023era} give a more biologically oriented introduction to ARGs.
 
@@
 structure~\citep{minichiello2006mapping}, i.e.\ describing
-the outcome of such a random process or an inference
+the outcome of such a stochastic process or of an inference
 of the genetic genealogy of a sample of genome sequences.
@@
 kelleher2019inferring, speidel2019method, baumdicker2021efficient, zhang2023biobank}.
-This broad usage means there is disagreement on the
+This broad usage means there is a lack of consensus on the
 exact object being defined. This causes problems, for example,
@@
 If not addressed, these problems will
-impede progress in applying genetic genealogical
-information to understand the biological world.
+impede progress in applying genetic genealogies to understand the biological world \citep{lewanski2023era}.
 
@@
 allowing us to systematically encode uncertainty about the
-time ordering of recombination events.
-For a recent and more biologically oriented introduction to
-ARGs, see \citet{lewanski2023era}.
+the precise details of recombination events.
+
+TODO: Provide a map of the paper's sections and how they tie together in support of gARG as a way forward.
 
@@
 % This has two purposes: clarify that we're not talking about Shipilina
-% et als edges, and also get "Graph theory" in here
+% et al's edges, and also get "Graph theory" in here
 (Note that we are using the term ``edge'' in its standard
@@
 which genome $c$ inherits from $p$. Thus, each topological connection between
- a parent and child in the graph is ``annotated'' with set of
+ a parent and child node in the graph is ``annotated'' with a set of
 inheritance intervals $I$. 
+Here, the terms parent and child are used in graph sense; 
+these nodes can respectively represent an ancestor and a descendant genome, 
+which can be genetically separated by multiple meioses. 
+We will use these two sets of terms interchangeably.
 
@@
 \caption{\label{fig-arg-in-pedigree}
-% TODO caption needs another pass. One the full section has been written
+% TODO caption needs another pass. Once the full section has been written
 % it'll be clearer what we need to say in the caption vs the text.
-An example genome ARG embedded in a pedigree.
+An example genome ARG (gARG) embedded in a pedigree.
 (A) Diploid individuals (blue), visualised in a highly inbred pedigree and
@@
 the paternal genomes \textsf{e}
-and \textsf{f}, and regions of genome inherited are shown.
+and \textsf{f}, and regions of genome inherited are shown with shaded colour.
 Genomes are shaded such that where, backwards in time,
@@
 we use lowercase alphabetical labels, to avoid confusion with genomic
-positions. Individual $D_1$ is therefore composed
-of two genomes \noderef{a} and \noderef{b}, which are inherited from its
+positions and associated intervals. Individual $D_1$ is therefore composed
+of genomes \noderef{a} and \noderef{b}, which are inherited from its
 two parents, $D_3$ and $D_4$. Each inherited genome may be the recombined product
@@
 recombination, whereas
-\noderef{a} is the recombinant product of
-$D_2$'s two genomes \noderef{e} and \noderef{f} crossing over at position 2.
-This recombination means \noderef{a} inherits the (half-closed) genome
-interval $[0, 2)$ from \noderef{e} and $[2, 10)$ from \noderef{f}.
+genome \noderef{a} is the recombinant of
+$D_2$'s genomes \noderef{e} and \noderef{f} crossing over at position 2.
+Specifically, genome \noderef{a} inherited the (half-closed) 
+interval $[0, 2)$ from genome \noderef{e} and $[2, 10)$ from genome \noderef{f}.
 These intervals are shown attached to the corresponding graph edges. 
-
-There are some important consequences to these definitions.
+The figure shows the annotated pedigree with realised inheritance of genomes 
+between generations, the corresponding gARG, and finally the corresponding 
+sequence of local trees along the genome. 
+The local trees represent the genealogy of genomes \noderef{a}--\noderef{d}, 
+which changes between local trees due to recombination. 
+The local trees span the three genome regions delineated 
+by the two recombinations that gave rise to these genomes.
+% Following the above definitions, corresponding data for this gARG is shown in Supplementary? Table XXX.
+
+These definitions follow naturally from the outcome of events 
+associated with the inheritance of DNA between generations. 
+As such,
+there are some important consequences to these definitions.
 Firstly, there is
-no limitation on mating system, ploidy, or age structure in a gARG.
+no limitation on a mating system, ploidy, or age structure in a gARG.
 By making the
-basic unit of analysis an individual's genomes,
+basic unit an individual's genomes,
 any pattern of inheritance among mixed ploidy (e.g.\ haplodiploid
@@
 between genomes,
-arbitrary patterns of genetic inheritance such as
+arbitrary patterns of genetic inheritance can be directly expressed, such as
 gene conversions,
-multiple simultaneous recombinations
-and so on can be directly expressed.
+multiple simultaneous recombinations,
+and so on.
 % Wilder: give forward ref to section on efficiency here, mention tree sequence
@@
 Fourth, recombination is modelled in terms of
-its outcomes in terms of inheritance intervals among parental genomes;
-thus we will have a child node (the recombinant; e.g.\ node \noderef{a}
+its inputs (parental genomes) and output (child genome) with associated inheritance intervals.
+Thus we will have a child node (the recombinant; e.g.\ node \noderef{a}
 Fig.~\ref{fig-arg-in-pedigree}) and two (or more; see section XXX)
-paternal genomes from which this child has inherited segments.
+parental nodes from which this child has inherited genome segments.
 % TODO rephrase the above, it'll confuse people. Just need to reiterate the
@@
 Thus, the gARG encoding is focused on faithfully representing the
-\emph{outcome} of the evolutionary processes of genetic inheritance
+\emph{outcome} of the processes of genetic inheritance
 with recombination,
 and is agnostic to the details of those processes.
-% The paragraph above is a bit weak and muddled, needs a few more passes.
-% In particular, the "Finally" bit doesn't really seem to follow
-% from the others. Is there a more direct way we can address these points ?
 
@@
 \section*{Event ARGs}\label{eARG}
-Genome ARGs are composed by
+Genome ARGs are defined by
 describing the details of genetic inheritance
-between specific ancestor and descendant genomes.
+between ancestor and descendant genomes.
 In contrast, ARGs are classically defined
-by the evolutionary \emph{events} in the history of a sample of genomes.
-The difference between gARGs and this prevalent view of
-what defines an ARG is subtle and difficult to distinguish.
+by describing the \emph{events} in the history of a sample of genomes.
+The difference between gARGs and the prevalent views of eARGs
+% what defines an ARG
+are subtle and difficult to distinguish.
 It is therefore useful to first survey some representative
@@
 Another common feature of these definitions is that they do not explicitly
-discuss the mechanism by which genealogical trees vary along the genome.
+discuss the mechanism by which local trees vary along the genome.
 The topology of an eARG defines the common ancestor and
@@
 Fig.~\ref{fig-event-arg} shows an example of a classical
-eARG with three samples (\noderef{a}, \noderef{b}, and \noderef{c})
-and a single recombination event, along with an example concrete
-encoding of the data.  There is a single recombination event
+eARG with three sample genomes (\noderef{a}, \noderef{b}, and \noderef{c})
+and a single recombination event, 
+% Trying to emphasise here that this is just an example data encoding. 
+% I can imagine
+% people nitpicking about how this specific way of doing it is bad
+% for reason X.
+along with an example encoding of the data. 
+There is a single recombination event
 at node \noderef{d} with a breakpoint at position $x$. We
@@
 the local trees are embedded in the graph, and can be reconstructed
-in essentially the same manner as outlined the previous section for
+in essentially the same manner as outlined in the previous section for
 gARGs. The only difference is that decisions about which outbound
@@
 encoding~\citep[e.g.][]{gusfield2014recombinatorics,ignatieva2021kwarg}. 
+% JK: not sure what you're getting at here Gregor? The ordering matters 
+% because you have to write each event down precisely.
+% Gregor: Importantly, the order of events matters to WHAT!?
 
@@
 A key advantage of associating inheritance intervals with
-edges in a gARG is that it allows us to describe evolutionary events other
+edges in a gARG is that it allows us to describe genetic inheritance events other
 than simple crossovers, such as multiple recombinations and
-gene-conversions. By explicitly stating the exact regions of genome
+gene conversions. By explicitly stating the exact regions of genome
 that are transmitted along an edge, we generalise the representation
@@
 % confuse people
-Thus, we have the means to encode the \emph{effects} of
-multiple evolutionary events, and to
+Thus, we have the means to encode the \emph{outcome} of
+multiple genetic inheritance events, and to
 systematically describe the uncertainty around their
 ordering (see Section XXX).
-As we see in the following sections, freeing ourselves from the requirement
+As we describe in the following sections, freeing ourselves from the requirement
 of recording specific events and
 the additional generality of the gARG encoding
-has many advantages. It also raises questions about what our goals
+has significant advantages. It also raises questions about what our goals
 for inferring ARGs should be, and how we evaluate the success
@@
 Event ARGs are defined in terms
-of the events that occur in the history of some set of sampled
-sequences.
+of the events that occured in the history of some set of sampled
+genomes.
 For example, Fig.~\ref{fig-arg-in-pedigree}
@@
 about genetic inheritance relative to the sample individuals
-$D_1$ and $D_2$ (genomes \textsf{a}--\textsf{d}).
+$D_1$ and $D_2$ (genomes \noderef{a}--\noderef{d}).
 That is, only the inheritance
-history of the sample genomes is recorded in the ARG.
+history of the sample genomes \noderef{a}--\noderef{d} is recorded in the ARG.
 % Gregor & Jerome think that "inheritance" is clearer here than "transmission"
@@
 $D_8$ and her daughter $D_6$, but this is not recorded because
-it is not relevant to the genetic ancestry of the sample
+it is not relevant to the genetic ancestry of the sample genomes.
 
@@
 ancestor to descendant through some
-% Wilder: finds the cell-to-cell confusing. Maybe drop?
-number of generations (ultimately from cell to cell),
+number of generations (ultimately from cell to cell; see Appendix XXX),
 and are annotated
-with the genome coordinates over which inheritance occurs.
+with genome coordinates over which inheritance occurs.
 In a prospective gARG, the inheritance intervals $I$
@@
 Fig.~\ref{fig-ancestry-resolution}.
-The first step is straightforward (Fig.~\ref{fig-ancestry-resolution}B):
+The first step is straightforward (Fig.~\ref{fig-ancestry-resolution}A,B):
 we duplicate the nodes and
@@
 If the node is a recombination event with a breakpoint $x$,
-we annotate the two outbound edges with the intervals $[0, x)$
-and $[x, L)$, respectively.
-These inheritance annotations are clearly in one-to-one
-correspondance with the information in the input eARG,
+we annotate the two outbound edges respectively with the intervals $[0, x)$ and $[x, L)$.
+These inheritance interval annotations are clearly in one-to-one
+correspondence with the information in the input eARG,
 but they are a superset of the
@@
 \centering
-% TODO update the source of this figure to make less tall. It's taking
-% up far too much vspace at in native coords at the moment.
 \includegraphics[width=\textwidth]{illustrations/ancestry-resolution}
@@
 (C) The sample-resolved gARG resulting from simplifying with respect
-to the samples, \noderef{a}, \noderef{b}, and \noderef{c}.
+to the sample genomes, \noderef{a}, \noderef{b}, and \noderef{c}.
 Dashed lines show edges that are
@@
 The ``grand MRCA'' \noderef{q} is omitted from the
-sample resolved gARG because all segments of the genome have
+sample-resolved gARG because all segments of the genome have
 fully coalesced before it is reached. Likewise, the edge
@@
 The gARG encoding
-allows us to do this once and to store the result, whereas
-the eARG encoding requires us to repeat the process
+allows us to do this X once and to store the result, whereas
+the eARG encoding requires us to repeat the process of X
 each time.
@@
 at a particular location along the genome, are
-embedded within in ARG, and the relationship
+embedded within an ARG, and the relationship
 between the two is subtle and important.
@@
 trees~\citep[e.g.][AND OTHERS]{griffiths1996ancestral,minichiello2006mapping}
-which we avoid due the potential confusion with the
+which we avoid due to the potential confusion with the
 statistical concept of marginal distributions
@@
 in the local tree at position $x$, then $\pi_u = -1$ (by convention).
-% JK: this confused some people, so best left out?
-% Technically, this structure is a ``forest'', as it allows us to have
-% multiple disconnected trees, which occurs when all samples have not
-% found a most recent common ancestor.
-Oriented trees allow nodes
-to have  any number of children; from one (see the next section for more
-discussion on such ``unary nodes''), to arbitrarily
-large polytomies.
+% This sentence needs another pass
+A node in an oriented tree can have 
+zero, one (see the next section for discussion of ``unary'' nodes), 
+or many children (a node with more than two children is known as a polytomy).
 To recover the local tree $\pi$ at position $x$, we begin by
@@
 of the corresponding ARG node), and this is fundamentally
-important to how a local tree relates to its parent ARG.
+important to how a local tree relates to its ARG.
 If we are given the sequence of local trees for a gARG
@@
 then there is precisely as
-much information in them as an ARG.
+much information in them as in an ARG.
 
@@
 \espalier, on the other hand, is specifically interested
-in the details of recombination events (among small samples)
+in the details of recombination events (among a small number of samples)
 and therefore attempts to infer
@@
 Returning to the example in Fig~\ref{fig-arg-in-pedigree}, nodes
-\noderef{g} and \noderef{h} are graph-unary, and are consequently
-also unary in the local trees.
+\noderef{g} and \noderef{h} are ARG-unary (Fig~\ref{fig-arg-in-pedigree}B), and are consequently
+also unary in the local trees (Fig~\ref{fig-arg-in-pedigree}C).
 On the other hand, node \noderef{f} has two children
-in the graph (Fig~\ref{fig-arg-in-pedigree}B), but is only binary
+in the graph, but is binary only
 in the local tree covering the interval $[2, 7)$,
@@
 in the local tree. Thus, \noderef{c} has a single child in this
-interval and it is unary in this local tree.
+interval and it is hence unary in this local tree.
 In another example, \noderef{e} is binary in the graph, but is
@@
 
-Outside of the context of simulation, graph-unary nodes are
+Outside of the context of simulation, ARG-unary nodes are
 only like to occur in longitudinal datasets where genetic
@@
 and there is usually
-little information about these intermediates, or utility
+little information about these intermediates, and hence limited utility
 in modelling them.
@@
 % would just make it less readable, IMO
-revolves around the presence unary nodes (see previous section),
-which we illustrate by an example simulated ARG, which
-undergoes a series of successive steps of simplification
+revolves around the presence unary nodes (see previous section).
+We illustrate this successive removal of redundancy 
+through a series simplification steps
 in Fig.~\ref{fig-simplification}.
@@
 The ARG in Fig.~\ref{fig-simplification}A is the output of a
-backwards-time Wright-Fisher simulation for a sample of 2 diploid
+backwards-time Wright-Fisher simulation for a sample of two diploid
 individuals (population size $N=10$), and follows a similar process
@@
 model recombination \emph{events} via an ARG node, but simply record
-the \emph{effects} of a recombination via outbound edges to
-the parent individual's two genomes.
+the \emph{outcome} of a recombination via outbound edges to
+the parent's two genomes.
 Thus, a recombinant node may also correspond to a coalescence
@@
 Fig.~\ref{fig-simplification} shows four ARGs (A--D), with
-the graph depicted on the left and corresponding sequence
+the graph depicted on the left and the corresponding sequence
 of local trees on the right. The graph visualisations have
@@
 Firstly, edge weights (the thickness of the lines joining
-nodes) correspond to the size of the inheritance intervals
+nodes) correspond to the length of the inheritance intervals
 they are annotated with. This allows us to distinguish
@@
 % this. In some senses it's a measure of identifiability?
+% JK: Hmm, not sure we want to jump into inferability here?
+All these features will impact what kind of an ARG we can 
+hope to infer from sampled genomes.
 
@@
 these methods to record information about the passage of ancestral
-material through genomes under a range conditions.
+material through genomes under a range of conditions.
 
@@
 events. This is related to the amount of \emph{precision} about
-recombination events that we store, which is the topic of the next
+recombination events that we store and 
+can hope to infer from sampled genomes, which is the topic of the next
 section.
@@
 The ARGs in Fig.~\ref{fig-simplification} contain different
-numbers of local trees ($6$, $5$, $5$ and $4$ for A through
-D, respectively). When we move from A to B the local trees
+numbers of local trees ($6$, $5$, $5$ and $4$ respectively for A through
+D). When we move from A to B the local trees
 for the intervals $[44,61)$ and $[61,87)$ are merged because
@@
 the \emph{timing} of this recombination event, and know only
-that it must have occured between the times of node \noderef{e}
+that it must have occurred between the times of node \noderef{e}
 and \noderef{q}.
@@
 % and the ancestral lineages involved.
-% While there has been a recent tendancy to emphasise the
+% While there has been a recent tendency to emphasise the
 % importance of ``full ARGs''
@@
 % events, it is reasonable to question the value of
-% such precise information and whether it is knowable.
+% such precise information and whether it is inferrable.
 % [SOME MORE STUFF]
@@
 % information about the history of a sample, of course, but
-% we may not have sufficient information to infer such detail.
+% we may not have sufficient information to infer such detail from sampled genomes.
 
@@
 heterogeneity of the ARGs inferred  by current methods,
-Fig.~\ref{fig-inferred-args} shows the results ARG inference
+Fig.~\ref{fig-inferred-args} shows the results of ARG inference
 % To illustrate these points and to emphasise the heterogeneity
@@
 \kwarg\ and \argweaver\ infer eARGs that contain
-specific causal recombination \emph{events} for each breakpoint.
+specific recombination \emph{events} for each breakpoint.
 On the other hand, \tsinfer\ and
-\relate\ do not infer causal events but rather
-infer gARGs with differing levels of precision about recombination.
+\relate\ do not infer events but rather
+infer gARGs with by differing levels of precision about recombination.
 The bottom row of Fig~\ref{fig-inferred-args} shows the extent along
@@
 (Technically, edges in tree sequence terminology would be better
-described as ``edge-intervals'', as each describes a single continguous
+described as ``edge-intervals'', as each describes a single contiguous
 interval of genome inheritance between a pair of nodes. This
@@
 The timing, positions, and even the number of recombination events is generally
-not possible to infer precisely from sequencing data. For ARGs under
+not possible to infer precisely from genome sequencing data. For ARGs under
 coalescent-based models, the proportion of recombination events that change the
@@
 low mutation rates and complex patterns of recombination, such inference can be
-even more challenging. The fact that the eARG encoding \emph{requires} completely
+even more challenging. The fact that the eARG encoding \emph{requires} 
 precise estimates about recombination is therefore a fundamental limitation, which
@@
 The advantages of a model-agnostic representation that naturally
-incorporates uncertainty about the ordering of events in an ARG are well illustrated by~\cite{zhan2023towards}, who
+incorporates uncertainty about the ordering of events in an ARG are well-illustrated by~\cite{zhan2023towards}, who
 inferred ARGs using millions of SARS-CoV-2 sequences from the
@@
 \section*{Acknowledgements}
-We are grateful to Gideon Bradburd and Alex Lewanski for helpful discussions
-and comments on the manuscript.
+We are grateful to Nick Barton, Gideon Bradburd, Alex Lewanski and Andrew Vaughan 
+for helpful discussions and comments on the manuscript.
+
+% JK: I'd usually use "JK was supported by XX", but this form seems fine too?
+% I don't have an opinion about it, if others think something else is more 
+% appropriate.
+GG acknowledges support from the BBSRC ISP grant to The Roslin Institute (BBS/E/D/30002275, BBS/E/RL/230001A, BBS/E/RL/230001C).
+Jerome Kelleher acknowledges support from the Robertson Foundation.
+Jere Koskela was supported by EPSRC research grant EP/V049208/1.
 

### 2023-10-13 paper.tex
Minor updates

@@
 of genetic inheritance that can be represented,
-and require complete precision about all recombination events.
+and require a precise description of the details of all recombination events.
 We show, in contrast, that the gARG definition
@@
 If not addressed, these problems will
-impede progress in applying genetic genealogies to understand the biological world \citep{lewanski2023era}.
+impede progress in applying genetic genealogies to understand the biological world.
 
@@
 The gARG encoding
-allows us to do this X once and to store the result, whereas
-the eARG encoding requires us to repeat the process of X
+allows us to perform this resolution of ancestral material once 
+and to store the result, 
+whereas the eARG encoding requires us to repeat the process 
 each time.
@@
 interval of genome inheritance between a pair of nodes. This
-denormalisation of the gARG data model is for efficiency purposes.)
+denormalisation of the gARG data model in \texttt{tskit} is for efficiency purposes.)
 The \texttt{tskit} library is a liberally

### 2023-10-13 paper.tex
Update local trees section

@@
 
-\title{\vspace{-1.5em} \bf A general and efficient representation of ancestral recombination graphs}
-
-% % First authors
-% \author[1]{Yan Wong}
-
-% % Second authors
-% \author[2,$\star$]{Anastasia Ignatieva}
-% \author[3,4,$\star$]{Jere Koskela}
-
-% % Middle Authors
-% \author[5]{Gregor Gorjanc}
-% \author[6,7]{Anthony W. Wohns}
-
-% % Corresponding
-% \author[1,$\dagger$]{Jerome Kelleher}
-
-% \affil[1]{Big Data Institute, Li Ka Shing Centre for Health Information and Discovery, University of Oxford, OX3 7LF, UK}
-% \affil[2]{Department of Statistics, University of Oxford, OX1 3LB, UK}
-% \affil[3]{School of Mathematics, Statistics and Physics, Newcastle University, NE1 7RU, UK}
-% \affil[4]{Department of Statistics, University of Warwick, CV4 7AL, UK}
-% \affil[5]{The Roslin Institute and Royal (Dick) School of Veterinary Studies, University of Edinburgh, EH25 9RG, UK}
-% \affil[6]{Broad Institute of MIT and Harvard, Cambridge, MA 02142, USA}
-% \affil[7]{Department of Genetics, Stanford University School of Medicine, Stanford, CA 94305, USA}
-
-% \affil[$\star$]{Denotes shared second authorship, listed alphabetically}
-% \affil[$\dagger$]{Denotes corresponding author}
+\title{\vspace{-1.5em} \bf
+A general and efficient representation of ancestral recombination graphs}
 
@@
 
-% FIXME abstract needs some rewriting, we're placing too much emphasis
-% on the coalescent there.
 \begin{abstract}
@@
 and accuracy of our inferences.
-% Thanks Gideon and Alex!
+% Thanks for the phrase Gideon and Alex!
 If not addressed, these problems will
@@
 
-It is useful to have
+To make the following discussions precise, we require
 a concrete definition and explicit process for local tree extraction.
@@
 zero, one (see the next section for discussion of ``unary'' nodes),
-or many children (a node with more than two children is known as a polytomy).
+or many children.
 To recover the local tree $\pi$ at position $x$, we begin by
@@
 the tree nodes are explicitly labelled with the identity of the
-corresponding ARG node)
+corresponding ARG node),
 along with the genome interval covered by each tree,
-then we can recover precisely the same ARG. Thus, there
+then we can recover precisely the same ARG. Thus, under this
+definition, there
 is a one-to-one correspondence between an ARG and
@@
 
-The internal nodes of trees, however, are usually
-considered to be \emph{unlabelled},
-because they most often correspond
-to anonymous hypothetical ancestors.
-Internal ARG nodes likewise correspond to anonymous hypothetical
-ancestors, but if we do not label them we lose the fact that
-the \emph{same} ancestors persist across multiple trees.
-Without labelled internal tree nodes,
-we must infer their identity
-from the tree topology, which is both
-expensive to compute and imprecise. For example
-using only the tree topologies it would not be straightforward to
-infer that node \noderef{i} in the first and last trees of
-Fig.~\ref{fig-arg-in-pedigree}C is the same node.
+This is not the prevailing view, however.
+\cite{kuhner2017consensus} argue that the
+``interval-tree'' representation of
+an ARG (the local trees and the genome intervals they cover)
+``does not contain all of the information in the underlying ARG: it lacks the
+number of recombinations occuring at each site, the times at which
+recombinations occurred, and the specific sequences involved as recombination
+partners.''
 \cite{shipilina2023origin} discuss the same ideas,
@@
 sequences along the genome''.
-Their statement that an ARG
-contains more information than its local trees is true only if
-the local trees are unlabeled and/or do not contain unary nodes
-(see the next section for discussion on this point).
-However, if the local trees
-contain labelled internal nodes,
-then there is precisely as
+These statements that an ARG
+contains more information than its local trees is only true
+if we use a representation of the local trees that discards
+information about those local trees that can be derived from the ARG.
+The internal nodes of evolutionary trees are usually
+considered to be \emph{unlabelled}
+because they most often correspond
+to anonymous hypothetical ancestors.
+Internal ARG nodes may likewise correspond to anonymous hypothetical
+ancestors, but if we do not label the tree nodes we lose the fact that
+the \emph{same} ancestors persist across multiple trees.
+Besides having labelled nodes, the local trees must contain
+unary nodes (see the next section) marking recombinations
+and other processes that lead to lineages passing through a given
+ARG node without coalescence in the local trees (see section XXX).
+If the local trees contain labelled internal nodes, precisely
+describing the corresponding path through the graph for each
+genome interval, then there is exactly as
 much information in them as in an ARG.
@@
 consider the inference strategy of estimating local trees
-sequentially along the genome, and subsequently stitching them together
-into an ARG.
+% JK: Relate is inferring trees independently within segments,
+% it's just using a dynamic algorithm to decide what those segments
+% are, right? That's how it differs from things like Espalier which
+% assume a set of input recomb breakpoints.
+independently within genome segments,
+and subsequently stitching them together into an ARG.
 Two recent methods take this approach,
@@
 is a degree of internal node sharing between trees.
-
-% The equivalence between an ARG and a set of local trees
-% separated by SPRs is often mentioned [e.g. x y z], but
-% it is important to note that while extracting local
-% trees from an ARG is straightforward, the converse
-% process of reconstructing precisely the same ARG from
-% these trees is more problematic. If the internal
-% nodes are consistently labelled and unary nodes are
-% included in the local trees (e.g., Fig~\ref{fig-simplification})
-% then it is easy to exactly reconstruct the ARG.
-% Otherwise, the problem is much more difficult and it
-% may not be possible to uniquely reconstruct the original ARG.
-% Without labels for the internal nodes, we are forced to
-% to reason about which subtrees remain the same and which
-% differ due to the SPR separating adjacent trees in order to
-% assign node identity. Without the internal unary nodes marking
-% recombinations, we must infer their existence. In
-% general, there will not be a unique solution to this problem,
-% and so there can be many possible ARGs compatible with
-% a given list of leaf-labelled binary trees.
-% This is the approach taken by the Espalier inference
-% method~\citep{rasmussen2022espalier}, which finds a
-% set of SPRs that reconcile independently inferred
-% trees from different regions of the genome.
 
@@
 In most cases,
-ARG inference is evaluated by simulating data from a known ground truth ARG, and comparing this to the inferred version via pairwise comparison of local trees along the genome, for example
+ARG inference is evaluated by simulating data from a known ground truth ARG,
+and comparing this to the inferred version via pairwise comparison of
+local trees along the genome, for example
 using tree distance
@@
 \citep{griffiths1996ancestral, fearnhead2001estimating, jenkins2011inference}
-or MCMC \citep{kuhner2000maximum, nielsen2000estimation, wang2008bayesian,
-fallon2013acg, mahmoudi2022bayesian}. These methods operate on representations
+or MCMC \citep{kuhner2000maximum, kuhner2006lamarc, nielsen2000estimation, wang2008bayesian,
+wang2009population, fallon2013acg, mahmoudi2022bayesian}.
+These methods operate on representations
 of the ``little ARG'', and are extremely computationally expensive, being
@@
 method, \texttt{ARGWeaver} \citep{rasmussen2014genome}, can be applied to
-dozens of mammal-like genomes \citep{hubisz2020inference}. A central quantity
-in all of these sampling methods is the conditional sampling probability $p(G |
-D) \propto p(D | G) p(G)$ of an ARG $G$ given an observed realization of
-genetic diversity $D$ at its sampled leaves, where $p(D | G)$ is the likelihood
-of the data $D$ given $G$, and $p(G)$ is a Bayesian prior distribution or a
-frequentist regularizer for the ARG $G$.
-
-The likelihood $p(D | G)$ is effectively always the distribution of a Poisson
-process on the ancestral material along the edges of $G$ \citep[Eq.\
-(2)]{mahmoudi2022bayesian}. Hence it is typically straightforward to evaluate
-in principle, but choosing the right data structure can still have a dramatic
-effect on the efficiency of evaluation \citep{mahmoudi2022bayesian}. Especially
-in the Bayesian case, $p(G)$ is typically specified as the sampling
-distribution of a stochastic process, such as the coalescent with
-recombination. Two contemporary examples are \citet{mahmoudi2022bayesian} and
-\citet{guo2022recombination}. Evaluating the distribution of the coalescent with
-recombination (or related models) requires knowledge of the number of extant
-lineages and number of links available for recombination (i.e.\ ones at which a
-recombination would split ancestral material) at each time \citep[Eq.\
-(3)]{mahmoudi2022bayesian}. These require a data structure from which shared
-edges on different trees are easy to identify, and which encodes recombination
-events and times explicitly, e.g.\ using nodes.
-We will provide a more detailed discussion on sampling probability evaluation in the appendix.
+dozens of mammal-like genomes \citep{hubisz2020inference}.
+
+Methods to sample ARGs generate a ``cloud'' of estimates
+\citep{kuhner2017consensus} develop
+
+% Not sure we want this content now? We're not really talking about
+% evaluation of liklihoods any more.
+% A central quantity
+% in all of these sampling methods is the conditional sampling probability $p(G |
+% D) \propto p(D | G) p(G)$ of an ARG $G$ given an observed realization of
+% genetic diversity $D$ at its sampled leaves, where $p(D | G)$ is the likelihood
+% of the data $D$ given $G$, and $p(G)$ is a Bayesian prior distribution or a
+% frequentist regularizer for the ARG $G$.
+
+% The likelihood $p(D | G)$ is effectively always the distribution of a Poisson
+% process on the ancestral material along the edges of $G$ \citep[Eq.\
+% (2)]{mahmoudi2022bayesian}. Hence it is typically straightforward to evaluate
+% in principle, but choosing the right data structure can still have a dramatic
+% effect on the efficiency of evaluation \citep{mahmoudi2022bayesian}. Especially
+% in the Bayesian case, $p(G)$ is typically specified as the sampling
+% distribution of a stochastic process, such as the coalescent with
+% recombination. Two contemporary examples are \citet{mahmoudi2022bayesian} and
+% \citet{guo2022recombination}. Evaluating the distribution of the coalescent with
+% recombination (or related models) requires knowledge of the number of extant
+% lineages and number of links available for recombination (i.e.\ ones at which a
+% recombination would split ancestral material) at each time \citep[Eq.\
+% (3)]{mahmoudi2022bayesian}. These require a data structure from which shared
+% edges on different trees are easy to identify, and which encodes recombination
+% events and times explicitly, e.g.\ using nodes.
+
 \end{document}

### 2023-10-13 paper.tex

@@
 the persistence of nodes and edges across multiple trees.
-Tree distance metrics usually have $O(n^2)$ time complexity and therefore cannot be
+The performance of tree distance metrics varies by
+application~\citep{kuhner2015assessing}, and the correct approach
+to handling subtleties such as polytomies
+is an open question~\citep{kelleher2019inferring,zhang2023biobank}.
+% Is this true? These guys claim RF takes linear time:
+% https://bmcgenomics.biomedcentral.com/articles/10.1186/s12864-020-07011-0
+Tree distance metrics usually have at least $O(n^2)$ time complexity
+and therefore cannot be
 applied to the very large sample sizes currently of interest.
-A recent trend has been to move away from such tree distance-based approaches and to examine more properties of the inferred ARGs.
+A recent trend has been to move away from such
+tree distance-based approaches and to examine more properties of the inferred ARGs.
 \citet{brandt2021evaluation} focussed on the
-distributions of pairwise MRCA times; \citet{deng2021distribution} derived an approximation to the distribution of the
-waiting distances between local trees; \citet{ignatieva2023distribution}
+distributions of pairwise MRCA times;
+\citet{deng2021distribution} derived an approximation to the distribution of the
+waiting distances between local trees;
+\citet{ignatieva2023distribution}
 derived the distribution of the genomic span of an edge and that of a
@@
 
-Methods to sample ARGs generate a ``cloud'' of estimates
-\citep{kuhner2017consensus} develop
+Methods to sample ARGs generate a ``cloud'' of estimates, and
+\citep{kuhner2017consensus} provide an approach to generate a
+set of consensus breakpoints and local trees from
+such a cloud.
+The approach is based on examining the recombination breakpoints
+in all of the input ARGS, and including those that are
+in at least $k$ of the input ARGs (with some additional
+filtering criteria) in the output.
+Within the resulting intervals, a consensus
+local tree is then generated using standard phylogenetic methods.
 

### 2023-10-13 paper.tex
Some refs

@@
 \section*{ARGs and local trees}
-% TODO mention Yamato here? I'm sure there were some interesting points
-% about local trees in that paper
 Local trees, describing the genetic genealogy of a sample
@@
 between the two is subtle and important.
-% An ARG is sometimes referred to as being equivalent to a set of
-% local trees along the genome [TODO citation], but the connection
-% between the two is subtle.
-% In order for the mapping to be one-to-one, so that the original
-% gARG can be exactly recovered from a set of local trees and
-% their genome coordinates,
-% some specific properties of the ARG and the trees are required.
 Note that many authors use the term ``marginal''
-trees~\citep[e.g.][AND OTHERS]{griffiths1996ancestral,minichiello2006mapping}
+% could list loads here, but not much point.
+trees~\citep[e.g.][]{griffiths1996ancestral,minichiello2006mapping,kelleher2016efficient}
 which we avoid due to the potential confusion with the
 statistical concept of marginal distributions
-(which could lead to interpretation as an integrated tree over many positions).
+(potentially leading to interpretation as an integrated tree over many positions).
 

### 2023-10-13 paper.tex
Fill in AND OTHERS

@@
 and many more are sure to
-follow~\citep{harris2019database,harris2023using}. \citet{lewanski2023era} give a more biologically oriented introduction to ARGs.
+follow~\citep{harris2019database,harris2023using}.
+See \citet{lewanski2023era} for a biologically oriented introduction to ARGs.
 
@@
 The word ``ARG'' is now widely used to refer to genetic genealogies in
-general~\citep[][AND OTHERS]{mathieson2020ancestry}, including
+general~\citep[e.g.][]{mathieson2020ancestry,hejase2020summary,
+schaefer2021ancestral,harris2023using,zhang2023biobank}, including
 varied outputs from modern simulation and

### 2023-10-13 paper.tex
Cite clonal origin paper
Closes #318

@@
 methods aim to balance computational efficiency with reconstruction of
-``reasonable", rather than parsimonious ARGs \citep{minichiello2006mapping,
+``reasonable'', rather than parsimonious ARGs or sampling from a
+distribution under a population genetics model
+\citep{minichiello2006mapping,
 parida2008estimating, kelleher2019inferring,  speidel2019method,
-zhang2023biobank}. The latter three methods, as well as the parsimony-based
-method of \citet{rasmussen2022espalier}, can be used on megabase scale data and
-hundreds of thousands of samples under human-like parameters.
-
-TODO mention \citep{schaefer2021ancestral} somewhere.
+schaefer2021ancestral, zhang2023biobank}.
 
@@
 or MCMC \citep{kuhner2000maximum, kuhner2006lamarc, nielsen2000estimation, wang2008bayesian,
-wang2009population, fallon2013acg, mahmoudi2022bayesian}.
+wang2009population, fallon2013acg, vaughan2017inferring, mahmoudi2022bayesian}.
 These methods operate on representations
-of the ``little ARG'', and are extremely computationally expensive, being
+of the ``little ARG'', and are computationally expensive, being
 applicable to at most hundreds of samples consisting of tens or hundreds of
-kilobases with human-like parameters. State-of-the-art methods rely on cheaper,
+kilobases with human-like parameters.
+State-of-the-art methods rely on cheaper,
 approximate models \citep{didelot2010inference, heine2018bridging,
@@
 Methods to sample ARGs generate a ``cloud'' of estimates, and
-\citep{kuhner2017consensus} provide an approach to generate a
+\cite{kuhner2017consensus} provide an approach to generate a
 set of consensus breakpoints and local trees from

### 2023-10-13 paper.tex
Add cell lines text

--- 2023-10-13 paper.tex
@@
 \section*{ARGs and local trees}
+% TODO mention Yamato here? I'm sure there were some interesting points
+% about local trees in that paper
 Local trees, describing the genetic genealogy of a sample
@@
 between the two is subtle and important.
+% An ARG is sometimes referred to as being equivalent to a set of
+% local trees along the genome [TODO citation], but the connection
+% between the two is subtle.
+% In order for the mapping to be one-to-one, so that the original
+% gARG can be exactly recovered from a set of local trees and
+% their genome coordinates,
+% some specific properties of the ARG and the trees are required.
 Note that many authors use the term ``marginal''
-% could list loads here, but not much point.
-trees~\citep[e.g.][]{griffiths1996ancestral,minichiello2006mapping,kelleher2016efficient}
+trees~\citep[e.g.][AND OTHERS]{griffiths1996ancestral,minichiello2006mapping}
 which we avoid due to the potential confusion with the
 statistical concept of marginal distributions
-(potentially leading to interpretation as an integrated tree over many positions).
+(which could lead to interpretation as an integrated tree over many positions).
 
@@
 
+\section{Cell lineages and ARGs}
+\label{cell-lineages-and-args}
+As we have argued, an important benefit of the gARG representation is that it
+can concatenate multiple events by encoding the outcome of their effects
+on a single node. In the case of recombination events, this manifests itself as
+a single node with three or more parents (e.g.\ Fig.~\ref{fig-simplification}C, node \noderef{e}).
+Here we make the argument that the same principle underlies ARGs containing
+coalescence points that involve three or more children. These ``polytomies'' are generated not only
+by Wright-Fisher dynamics such as those in Fig.~\ref{fig-simplification}C (see node \noderef{k}), but also
+by life histories that involve ``burst'' reproduction events, which are often modelled by a lambda
+coalecent process [AND OTHERS]\citep{wakely2008coalescent}.
+
+At the base of this concept is a subtle and important point about which genomes the nodes in a gARG represent. 
+In principle, we can interpret nodes as representing genomes
+at any point in the life cycle, for instance, \citet{hudson1983properties} views nodes as representing gametes,
+whereas in Fig.~\ref{fig-arg-in-pedigree}A, we present them as genomes within a diploid individual.
+In reality, however, recombination and replication of DNA occur at a cellular level,
+and correspond to the processes of meiosis and mitosis respectively. A comprehensive gARG would therefore
+trace the process of inheritance not between diploid individuals, but between each cell during the
+life cycle of an organism. This is illustrated in Fig.~\ref{fig-cell-lines}, where we show a simplified
+schematic of cellular inheritance in a diploid population over four partially overlapping generations.
+In this figure, thick black lines trace an ARG that describes the ancestry of the lower arm of the
+four sampled chromosomes. The potential ARG nodes are shown as green chromosomal regions, and the ``events''
+underlying the ARG are shown as red and blue squares.
+
+The unusual feature of this particular example is that three ARG lineages coalesce in one individual, $D_10$
+(who has three children). We know that the underlying coalescence process must consist of two successive bifurcations
+as the only known method of reproducing DNA is by (semi-conservative) duplication. However, if we take each node in the
+ARG to represents the generalised genome of an individual, rather than the specific genome of a cell within an individual,
+we can only represent this as a instantanous trifucation or polytomy.
+
+There is biological significance to this cellular-level inheritance process. Were a mutation to occur
+during the first cell division of $D_10$, the two gametes produced by the cells
+in the left half of $D_10$ could share a mutation not present in the right hand gamete. In other words,
+the relationship between three (or more) gametes that originate, without recombination, from the same parental
+chromosomes will nevertheless be strictly bifurcating; two of the gametes have a closer relationship with the
+third more distant. Of course, unless mutations occur between successive rounds of cell division, the between-cell
+relationships will be impossible to infer from the data, with subsequent implications for
+using the eARG representation to store the results of inference.
+
+As with concatenating multiple recombination events, the gARG representation allows us to concatenate within-individual
+coalescence events such that we are not obliged to represent precisely the underlying cellular process. In humans and
+other multicellular diploids, it seems most intuitive to take the nodes in the gARG to represent two haploid genomes within
+a newly fertilised diploid individual. If this individual has 3 or more children, and these children feature in the
+eventual ancestry of the samples, the gARG approach outlined in this paper means that they can be legitimately represented
+by a polytomy in the local trees.
+\begin{figure}
+\begin{center}
+    \includegraphics[width=0.8\textwidth]{illustrations/cell-lines}
+\end{center}
+\caption{\label{fig-cell-lines}
+Cellular inheritance of a single chromosome in a diploid population.
+Individuals (blue) contain diploid cells (white circles enclosing a homologous pair of chromosomes).
+For clarity, only two rounds of mitotic germ-line cell division are shown per individual, and
+meiosis is not illustrated in detail.
+Lines show prospective inheritance paths for all chromosomes. Solid lines show all possible
+retrospective ancestry paths for four chosen chromosomes (indicated by square black ``sampling events'')
+sampled from 3 diploid individuals ($D_1$, $D_2$, $D_3$) in the current generation.
+Ancestral recombination events and coalescence events are shown as red and blue squares respectively.
+The ARG paths for the lower arm of the sampled chromosomes (green) is highlighted as a
+think solid line and involves a single recombination event and four coalescence events
+(highlighted as deep red and blue squares within individuals $D_5$, $D_10$, and $D_13$).
+ARG lineages also show gametic genomes, contained within shaded circles.
+As in Fig.~\ref{fig-arg-in-pedigree}A, inherited regions within the sampled chromosome arm are
+shaded by the number of descendant samples.
+}
+\end{figure}
+
 \end{document}

### 2023-10-13 paper.tex
Tweaks to appendix

@@
 ancestor to descendant through some
-number of generations (ultimately from cell to cell; see Appendix XXX),
+number of generations
+% Wilder found cell-to-cell a little confusing here, we could delete
+% if necessary (it's tackled later on)
+(ultimately from cell to cell; see Appendix~\ref{cell-lineages-and-args}),
 and are annotated
@@
 A tree branch may correspond to many generations
-of individuals (or indeed cell divisions; see Appendix XXX)
+of individuals (or indeed cell divisions; see Appendix \ref{cell-lineages-and-args})
 and there is usually
@@
 
-At the base of this concept is a subtle and important point about which genomes the nodes in a gARG represent. 
-In principle, we can interpret nodes as representing genomes
-at any point in the life cycle, for instance, \citet{hudson1983properties} views nodes as representing gametes,
-whereas in Fig.~\ref{fig-arg-in-pedigree}A, we present them as genomes within a diploid individual.
-In reality, however, recombination and replication of DNA occur at a cellular level,
-and correspond to the processes of meiosis and mitosis respectively. A comprehensive gARG would therefore
-trace the process of inheritance not between diploid individuals, but between each cell during the
-life cycle of an organism. This is illustrated in Fig.~\ref{fig-cell-lines}, where we show a simplified
-schematic of cellular inheritance in a diploid population over four partially overlapping generations.
-In this figure, thick black lines trace an ARG that describes the ancestry of the lower arm of the
-four sampled chromosomes. The potential ARG nodes are shown as green chromosomal regions, and the ``events''
-underlying the ARG are shown as red and blue squares.
-
-The unusual feature of this particular example is that three ARG lineages coalesce in one individual, $D_10$
-(who has three children). We know that the underlying coalescence process must consist of two successive bifurcations
-as the only known method of reproducing DNA is by (semi-conservative) duplication. However, if we take each node in the
-ARG to represents the generalised genome of an individual, rather than the specific genome of a cell within an individual,
-we can only represent this as a instantanous trifucation or polytomy.
-
-There is biological significance to this cellular-level inheritance process. Were a mutation to occur
-during the first cell division of $D_10$, the two gametes produced by the cells
-in the left half of $D_10$ could share a mutation not present in the right hand gamete. In other words,
-the relationship between three (or more) gametes that originate, without recombination, from the same parental
-chromosomes will nevertheless be strictly bifurcating; two of the gametes have a closer relationship with the
-third more distant. Of course, unless mutations occur between successive rounds of cell division, the between-cell
-relationships will be impossible to infer from the data, with subsequent implications for
-using the eARG representation to store the results of inference.
-
-As with concatenating multiple recombination events, the gARG representation allows us to concatenate within-individual
-coalescence events such that we are not obliged to represent precisely the underlying cellular process. In humans and
-other multicellular diploids, it seems most intuitive to take the nodes in the gARG to represent two haploid genomes within
-a newly fertilised diploid individual. If this individual has 3 or more children, and these children feature in the
-eventual ancestry of the samples, the gARG approach outlined in this paper means that they can be legitimately represented
-by a polytomy in the local trees.
 \begin{figure}
@@
 think solid line and involves a single recombination event and four coalescence events
-(highlighted as deep red and blue squares within individuals $D_5$, $D_10$, and $D_13$).
+(highlighted as deep red and blue squares within individuals $D_5$, $D_{10}$, and $D_{13}$).
 ARG lineages also show gametic genomes, contained within shaded circles.
@@
 
+At the base of this concept is a subtle and important point about which genomes the nodes in a gARG represent. 
+In principle, we can interpret nodes as representing genomes
+at any point in the life cycle, for instance, \citet{hudson1983properties} views nodes as representing gametes,
+whereas in Fig.~\ref{fig-arg-in-pedigree}A, we present them as genomes within a diploid individual.
+In reality, however, recombination and replication of DNA occur at a cellular level,
+and correspond to the processes of meiosis and mitosis respectively. A comprehensive gARG would therefore
+trace the process of inheritance not between diploid individuals, but between each cell during the
+life cycle of an organism. This is illustrated in Fig.~\ref{fig-cell-lines}, where we show a simplified
+schematic of cellular inheritance in a diploid population over four partially overlapping generations.
+In this figure, thick black lines trace an ARG that describes the ancestry of the lower arm of the
+four sampled chromosomes. The potential ARG nodes are shown as green chromosomal regions, and the ``events''
+underlying the ARG are shown as red and blue squares.
+
+The unusual feature of this particular example is that three ARG lineages coalesce in one individual, $D_{10}$
+(who has three children). We know that the underlying coalescence process must consist of two successive bifurcations
+as the only known method of reproducing DNA is by (semi-conservative) duplication. However, if we take each node in the
+ARG to represents the generalised genome of an individual, rather than the specific genome of a cell within an individual,
+we can only represent this as a instantaneous trifurcation or polytomy.
+
+There is biological significance to this cellular-level inheritance process. Were a mutation to occur
+during the first cell division of $D_{10}$, the two gametes produced by the cells
+in the left half of $D_{10}$ could share a mutation not present in the right hand gamete. In other words,
+the relationship between three (or more) gametes that originate, without recombination, from the same parental
+chromosomes will nevertheless be strictly bifurcating; two of the gametes have a closer relationship with the
+third more distant. Of course, unless mutations occur between successive rounds of cell division, the between-cell
+relationships will be impossible to infer from the data.
+
+As with concatenating multiple recombination events, the gARG representation allows us to concatenate within-individual
+coalescence events such that we are not obliged to represent precisely the underlying cellular process. In humans and
+other multicellular diploids, it seems most intuitive to take the nodes in the gARG to represent two haploid genomes within
+a newly fertilised diploid individual. If this individual has 3 or more children, and these children feature in the
+eventual ancestry of the samples, the gARG approach outlined in this paper means that they can be legitimately represented
+by a polytomy in the local trees.
+
 \end{document}

### 2023-10-13 paper.tex
Tweaks

@@
 trees exactly equivalent to the corresponding ARG (see previous section).
-Unary nodes are a vital link between ARGs and local trees, and we
+Locally unary nodes are a vital link between ARGs and local trees, and we
 cannot fully reason about how a local tree is embedded in an ARG
@@
 (although \citet{mathieson2020ancestry} show examples in which the ARGs
-contain unary nodes, and state that paths are
+contain locally unary nodes, and state that paths are
 ``implicitly passing through points representing specific individuals.'').
@@
 As we see in the next section, however, there is a range of
-different ways in which nodes can be unary in the local trees
-of an ARG. Furthermore, ARGs inferred by current methods contain
-locally unary nodes from several of them (see section XXX).
+different ways in which ARG nodes can be locally unary. Furthermore, 
+several current ARG inference methods infer locally unary nodes (see section XXX).
 
@@
 by life histories that involve ``burst'' reproduction events, which are often modelled by a lambda
-coalecent process [AND OTHERS]\citep{wakely2008coalescent}.
+coalescent process [AND OTHERS]\citep{wakely2008coalescent}.
 
@@
 In this figure, thick black lines trace an ARG that describes the ancestry of the lower arm of the
-four sampled chromosomes. The potential ARG nodes are shown as green chromosomal regions, and the ``events''
-underlying the ARG are shown as red and blue squares.
-
-The unusual feature of this particular example is that three ARG lineages coalesce in one individual, $D_{10}$
-(who has three children). We know that the underlying coalescence process must consist of two successive bifurcations
-as the only known method of reproducing DNA is by (semi-conservative) duplication. However, if we take each node in the
-ARG to represents the generalised genome of an individual, rather than the specific genome of a cell within an individual,
-we can only represent this as a instantaneous trifurcation or polytomy.
+four sampled chromosomes. The ``events'' underlying the ARG are shown as red and blue squares,
+and the potential ARG nodes are shown as green chromosomal regions. Note that a number of these nodes are
+``pass-through'' nodes with one parent and one child, and in practice these would normally be omitted.
+
+The unusual feature of this particular example is that three ARG lineages coalesce in one historical individual,
+$D_10$, whose three children all happen to be ancestors of the sampled chromosomes.
+We can see that the process that generates these three child lineages actually consists of two successive bifurcations:
+necessitated by the fact that the only known method of reproducing DNA is by (semi-conservative) duplication.
+However, if we take each node in the ARG to represents the generalised genome of an individual (rather than the specific
+genome of a cell within an individual) we end up representing this as a instantaneous trifucation or polytomy.
 
@@
 
-As with concatenating multiple recombination events, the gARG representation allows us to concatenate within-individual
-coalescence events such that we are not obliged to represent precisely the underlying cellular process. In humans and
-other multicellular diploids, it seems most intuitive to take the nodes in the gARG to represent two haploid genomes within
-a newly fertilised diploid individual. If this individual has 3 or more children, and these children feature in the
-eventual ancestry of the samples, the gARG approach outlined in this paper means that they can be legitimately represented
-by a polytomy in the local trees.
+% As with concatenating multiple recombination events, the gARG representation allows us to concatenate within-individual
+% coalescence events such that we are not obliged to represent precisely the underlying cellular process. In humans and
+% other multicellular diploids, it seems most intuitive to take the nodes in the gARG to represent two haploid genomes within
+% a newly fertilised diploid individual. If this individual has 3 or more children, and these children feature in the
+% eventual ancestry of the samples, the gARG approach outlined in this paper means that they can be legitimately represented
+% by a polytomy in the local trees.
 

### 2023-10-13 paper.tex
Tweaks to gARG section

@@
 
+% JK: this para need a major update, it's quite weak right now.
 The word ``ARG'' is now widely used to refer to genetic genealogies in
 general~\citep[e.g.][]{mathieson2020ancestry,hejase2020summary,
-schaefer2021ancestral,harris2023using,zhang2023biobank}, including
-varied outputs from modern simulation and
+schaefer2021ancestral,harris2023using,zhang2023biobank},
+encompassing the varied outputs from modern simulation and
 inference methods~\citep{rasmussen2014genome, palamara2016argon, haller2018tree,
@@
 inherits from one parent (i.e.\ haploids have one genome, diploids two, etc.).
-% Talking about "haplotypes" probably just confuses more
-% In this usage a ``genome'' is equivalent to the definition of a
-% ``haplotype'' used by~\cite{shipilina2023origin}, which we avoid because
-% it already has many interpretations, and in particular carries the implication of
-% specific mutational patterns [CITES] that persist over
-% many generations.
-We are not concerned here with mutational processes or observed sequences,
-but consider only processes of inheritance,
-following the standard practice in coalescent theory.
 We will also use the term ``genome'' in its
@@
 will be clear from the context.
+We are not concerned here with mutational processes or observed sequences,
+but consider only processes of inheritance,
+following the standard practice in coalescent theory.
 
@@
 genetic inheritance between an ancestor and a descendant.
-% This has two purposes: clarify that we're not talking about Shipilina
-% et al's edges, and also get "Graph theory" in here
-(Note that we are using the term ``edge'' in its standard
-graph-theoretic sense
-and not the specialised interpretation of
-\cite{shipilina2023origin}.)
 The topology of a gARG specifies that genetic inheritance
@@
 parent nodes and $I$ is the set of disjoint genomic intervals
-%$[\ell, r)$
-over
-which genome $c$ inherits from $p$. Thus, each topological connection between
+over which genome $c$ inherits from $p$.
+Thus, each topological connection between
 a parent and child node in the graph is ``annotated'' with a set of
@@
 We will use these two sets of terms interchangeably.
+
+In many settings, nodes are dated, i.e.\
+each node $u\in N$ is associated with a time $\tau_u$.
+However, the topological relationships
+defined by the directed graph structure is sufficient for
+many applications, and we will assume that node dates are not
+known here.
+% This should be obvious, but seems like it needs to be said
+In practical settings, we will wish to associate additional
+metadata such as a sample identifiers or quality-control metrics
+with nodes; it is therefore best to think of the
+integers used here in the definition of a node as an \emph{identifier},
+with which arbitrary additional information can be associated.
 
@@
 the gARG for a given set of individuals is embedded in their pedigree.
-In this figure, diploid individuals are labelled $D_1, D_2...$, and
-rather than denoting their constituent genomes using the natural numbers
-we use lowercase alphabetical labels, to avoid confusion with genomic
-positions and associated intervals. Individual $D_1$ is therefore composed
+The pedigree relationship between the diploid individuals is
+shown, along with their constituent genomes and paths
+of genetic inheritance.
+We label nodes using using lowercase alphabetical letters
+rather than integer identifiers to avoid confusion with genomic
+positions and associated intervals.
+Thus, individual $D_1$ is composed
 of genomes \noderef{a} and \noderef{b}, which are inherited from its
@@
 The unusual feature of this particular example is that three ARG lineages coalesce in one historical individual,
-$D_10$, whose three children all happen to be ancestors of the sampled chromosomes.
+$D_{10}$, whose three children all happen to be ancestors of the sampled chromosomes.
 We can see that the process that generates these three child lineages actually consists of two successive bifurcations:

### 2023-10-13 argutils/viz.py
Make the inference plots a bit less tall
To fit text below, which seems to have been pushed off.

@@ -231,8 +231,8 @@ def draw(
             labels=labels,
             ax=ax,
         )
-        y_below_extra = 0.018 * np.diff(ax.get_ylim())  # 2% of the y range
         if rotated_sample_labels:
+            y_below_extra = 0.0145 * np.diff(ax.get_ylim())  # 2% of the y range
             for nd, t in text.items():
                 if is_sample[nd]:
                     x, y = t.get_position()

### 2023-10-13 paper.tex
Yan's edits

@@
 goal in genetics.
-% TODO say this better.
-From a high level, an ARG describes the different paths of genetic inheritance
+Broadly speaking, an ARG describes the different paths of genetic inheritance
 caused by recombination, and encodes a sequence of correlated genealogical
@@
 structure~\citep{minichiello2006mapping}, i.e.\ describing
-the outcome of such a stochastic process or of an inference
-of the genetic genealogy of a sample of genome sequences.
+the outcome of such a stochastic process or a
+genetic genealogy inferred from a sample of genome sequences.
 However, the distinction between process and data structure is not clear cut
@@
 This broad usage means there is a lack of consensus on the
-exact object being defined. This causes problems, for example,
+exact object being defined, which causes problems, for example,
 when evaluating inferred ARGs, and even when defining
@@
 geneticists are only too familiar~\citep{excoffier2006computer,gower2022demes}.
-Perhaps most importantly, the direct linkage to
-the original stochastic process definition leads to an
+Perhaps most importantly, definitions linked to
+the original stochastic process lead to an
 overly narrow interpretation of ARGs, which is significantly
@@
 The pedigree relationship between the diploid individuals is
-shown, along with their constituent genomes and paths
-of genetic inheritance.
-We label nodes using using lowercase alphabetical letters
+shown, along with their paths
+of genetic inheritance and constituent genomes (labelled
+using using lowercase alphabetical letters
 rather than integer identifiers to avoid confusion with genomic
-positions and associated intervals.
+intervals).
 Thus, individual $D_1$ is composed
@@
 trees exactly equivalent to the corresponding ARG (see previous section).
-Locally unary nodes are a vital link between ARGs and local trees, and we
+Unary nodes are a vital link between ARGs and local trees, and we
 cannot fully reason about how a local tree is embedded in an ARG
@@
 (although \citet{mathieson2020ancestry} show examples in which the ARGs
-contain locally unary nodes, and state that paths are
+contain unary nodes, and state that paths are
 ``implicitly passing through points representing specific individuals.'').
@@
 
+\section{Additional plots}
+
+\begin{figure}[h]
+	\begin{center}
+		\includegraphics[width=0.8\textwidth]{illustrations/simplification-with-edges.pdf}
+	\end{center}
+	\caption{\label{fig-simplification-with-edges}
+	The four levels of ARG simplification from Fig.~\ref{fig-simplification}A--D, with edge intervals shown.
+	}
+\end{figure}
+
 \end{document}

### 2023-10-13 paper.tex
Tweak figs
Second pass at prospective/retrospective
To address Andrew's comment, my thoughts, and (partially) Gideon's comments about application

@@
 \section*{Prospective vs retrospective ARGs}
-Event ARGs, as defined in the previous section as a way of encompassing
-the range of definitions provided in the literature, are
-inherently \emph{retrospective}.
-Event ARGs are defined in terms
-of the events that occured in the history of some set of sampled
-genomes.
-For example, Fig.~\ref{fig-arg-in-pedigree}
+The ARG concept arose from a \emph{retrospective} view of inheritance.
+For example the definition of an eARG, as discussed above, is invariably
+couched in terms of the events that occured in the history of some set of sampled
+genomes. Similarly, the genome ARG in Fig.~\ref{fig-arg-in-pedigree}
 is retrospective because it only includes information
@@
 There is no particular reason, however, that ARGs must be
-retrospective. They can also be \emph{prospective}.
-The edges in a gARG represent a path of genetic inheritance from
-ancestor to descendant through some
-number of generations
-% Wilder found cell-to-cell a little confusing here, we could delete
-% if necessary (it's tackled later on)
-(ultimately from cell to cell; see Appendix~\ref{cell-lineages-and-args}),
-and are annotated
-with genome coordinates over which inheritance occurs.
+retrospective. Recent advances in individual-based forward-time simulation
+~\citep{kelleher2018efficient,haller2018tree} depend upon the ability to
+record gARGs in a \emph{prospective} manner.
 In a prospective gARG, the inheritance intervals $I$
 on an edge joining ancestor genome $v$ and descendant $u$
-defines \emph{all} regions of genome that $u$ inherited from $v$.
+define \emph{all} regions of genome that $u$ inherited from $v$.
 Thus, these inheritance intervals are
@@
 between the nodes in question, and are not defined with respect
-to any given set of samples. We are simply recording the
+to any given set of samples. They simply record the immediate
 genetic inheritance between two genomes
 (in the simplest case, between a parent and a child).
-
-Prospective ARGs arise in individual-based forwards-time
-simulations~\citep{kelleher2018efficient,haller2018tree},
-where the genetic inheritance between genomes is recorded
-exhaustively and periodically
-``simplified'' to retain only
-the ancestry relevant to the current population.
 \citet{kelleher2018efficient} refer to the structure
@@
 more useful.
+
+In a prospective ARG, the exhastive recording of all inheritance intervals from generation
+to generation leads to a rapid build-up of information, much of which
+may not be relevant to the genetic ancestry of the current population. 
+Redundant nodes and inheritance intervals can therefore be periodically
+``simplified'' to retain only
+the ancestry relevant to the current population.
 The ``simplify'' algorithm, which prunes away parts of the graph that
@@
 
-\section{Additional plots}
-
-\begin{figure}[h]
+\clearpage
+\section*{Supplementary figures}
+
+% TODO add label and change numbering to S1, etc
+\begin{figure}[ht]
 	\begin{center}

### 2023-10-14 paper.tex
Updates from Yan and gregor

--- 2023-10-13 paper.tex
@@
 eARG approach, being flexible, efficient and, in particular,
-allowing us to systematically encode uncertainty about the
+allowing us to systematically encode uncertainty about
 the precise details of recombination events.
 
-TODO: Provide a map of the paper's sections and how they tie together in support of gARG as a way forward.
+% TODO: Provide a map of the paper's sections and how they tie together in support of gARG as a way forward.
+% Here is a rough draft - it would be good to state the most important message (in one sentence or so) from each of the sections to help with frontloading key messages
+% Genome ARGs &
+% Event ARGs
+In the following, we define the genome ARG (gARG) and contrast it with the event ARGs (eARG).
+% Retrospective vs prospective
+% TODO: This needs revision - see also https://github.com/tskit-dev/what-is-an-arg-paper/pull/380
+We continue with a description of the usually held retrospective view of ARGs, but note that ARGs can also be viewed as prospective, particularly in forwards-time simulations.
+In the context of forwards-time simulations, we mention the simplification operation, which removes ancestral genome material that was not inherited.
+% Converting and eARG to gARG &
+% ARGs and local trees &
+% Locally unary nodes &
+% Levels of simplification
+This simplification is closely related to operations needed to convert eARG and gARG, which opens a rich set of details about ARGs and local trees, local unary nodes, and levels of simplification.
+% Precision of recombination information
+We then highlight the important difference between the required level of precision about recombination information for eARGs compared to gARGs.
+% Computational efficiency and data interchange
+The way this recombination information is stored in eARGs and gARGs has important implications for computational efficiency, which is critical if we want to use ARGs for modern genomic datasets.
+% Discussion
+We wrap with a discussion highlighting that WHAT-IS-THE-KEY-SINGLE-MESSAGE-NOT-COVERED-ABOVE?
+Finally, the literature on ARGs and associated aspects is vast, particularly due to different definitions and implementations.
+We cover some important aspects in appendices, including a brief history of ancestral graphs (Appendix XXX), description of the big and little ARGs (Appendix ZZZ), survey of ARG inference methods (Appendix YYY), and ARGs at an individual vs cell lineage level (Appendix WWW).
 
@@
 Here, the terms parent and child are used in graph sense;
-these nodes can respectively represent an ancestor and a descendant genome,
-which can be genetically separated by multiple meioses.
+these nodes can respectively represent ancestor and descendant genomes,
+which can be separated by multiple generations.
 We will use these two sets of terms interchangeably.
@@
 In practical settings, we will wish to associate additional
-metadata such as a sample identifiers or quality-control metrics
+metadata such as sample identifiers or quality-control metrics
 with nodes; it is therefore best to think of the
@@
 of genetic inheritance and constituent genomes (labelled
-using using lowercase alphabetical letters
+with lowercase alphabetical letters
 rather than integer identifiers to avoid confusion with genomic
@@
 The local trees span the three genome regions delineated
-by the two recombinations that gave rise to these genomes.
+by the two recombination breakpoints that gave rise to these genomes.
 % Following the above definitions, corresponding data for this gARG is shown in Supplementary? Table XXX.
@@
 
-To make the following discussions precise, we require
+To make the following discussions precise, we provide
 a concrete definition and explicit process for local tree extraction.
@@
 (or not reachable from the samples)
-in the local tree at position $x$, then $\pi_u = -1$ (by convention).
+in the local tree at position $x$, then we adopt the convention that $\pi_u = -1$.
 % This sentence needs another pass
 A node in an oriented tree can have
-zero, one (see the next section for discussion of ``unary'' nodes),
+zero, one,
 or many children.
@@
 ARG node without coalescence in the local trees (see section XXX).
-If the local trees contain labelled internal nodes, precisely
+If the local trees contain labelled internal nodes, they precisely
 describing the corresponding path through the graph for each
-genome interval, then there is exactly as
+genome interval and there is exactly as
 much information in them as in an ARG.
@@
 In another example, \noderef{e} is binary in the graph, but is
-unary in each of the local trees it is present in.
+unary in each of the local trees in which it is present.
 This is because \noderef{a} and \noderef{c} are both recombinant
@@
 % would just make it less readable, IMO
-revolves around the presence unary nodes (see previous section).
+revolves around the presence of unary nodes (see previous section).
 We illustrate this successive removal of redundancy
@@
 provided as input, ARGs sampled from
-the coalescent cannot capture the complexities of realistic family structures.
-There are many species with sizeable to large number of progeny, making polytomies a necessary feature of local trees.
+the coalescent cannot capture the complexities of realistic family structures, which often involve overlapping generations and extreme variation in lifetime reproductive success.
 Flexible ARG encoding should accommodate such cases.
@@
 
-% JK: I'd usually use "JK was supported by XX", but this form seems fine too?
-% I don't have an opinion about it, if others think something else is more
-% appropriate.
-GG acknowledges support from the BBSRC ISP grant to The Roslin Institute (BBS/E/D/30002275, BBS/E/RL/230001A, BBS/E/RL/230001C).
+% FIXME: simplest to order this alphabetically by surname?
+% Anastasia TODO
+Jere Koskela was supported by EPSRC research grant EP/V049208/1.
+Gregor Gorjanc acknowledges support from the BBSRC ISP grant to The Roslin Institute (BBS/E/D/30002275, BBS/E/RL/230001A, and BBS/E/RL/230001C).
+% Wilder TODO
 Jerome Kelleher acknowledges support from the Robertson Foundation.
-Jere Koskela was supported by EPSRC research grant EP/V049208/1.
 
@@
 
-At the base of this concept is a subtle and important point about which genomes the nodes in a gARG represent.
+At the base of this concept is a subtle and important point about which genomes are represented by gARG nodes.
 In principle, we can interpret nodes as representing genomes
@@
 In reality, however, recombination and replication of DNA occur at a cellular level,
-and correspond to the processes of meiosis and mitosis respectively. A comprehensive gARG would therefore
-trace the process of inheritance not between diploid individuals, but between each cell during the
-life cycle of an organism. This is illustrated in Fig.~\ref{fig-cell-lines}, where we show a simplified
+and respectively correspond to the processes of meiosis and mitosis. Therefore, a comprehensive gARG would not
+trace genetic inheritance between diploid individuals, but between cells during the
+life cycle of these individuals. This is illustrated in Fig.~\ref{fig-cell-lines}, where we show a simplified
 schematic of cellular inheritance in a diploid population over four partially overlapping generations.
@@
 and the potential ARG nodes are shown as green chromosomal regions. Note that a number of these nodes are
-``pass-through'' nodes with one parent and one child, and in practice these would normally be omitted.
+``pass-through'' nodes with one parent and one child, which would normally be omitted.
 

### 2023-10-14 paper.tex

@@
 
-In a prospective ARG, the exhastive recording of all inheritance intervals from generation
+The exhastive recording of full inheritance intervals from generation
 to generation leads to a rapid build-up of information, much of which
 may not be relevant to the genetic ancestry of the current population.
-Redundant nodes and inheritance intervals can therefore be periodically
-``simplified'' to retain only
-the ancestry relevant to the current population.
-The ``simplify'' algorithm, which prunes away parts of the graph that
-are not ancestral to a sample, is the key enabling factor for
-this approach to forwards-time simulation and is
-described in detail by
-\citet{kelleher2018efficient}.
-It is a powerful tool for simulation and analysis as it
-allows efficient subsetting of ARGs, with the ability to
-remove both samples and certain levels of detail about recombination,
-as we discuss in the following sections.
+This redundancy can be periodically removed using the ``simplify'' algorithm,
+described in detail by \citet{kelleher2018efficient}. Simplification
+allows efficient subsetting of ARGs (removing both samples and certain
+levels of detail about recombination, as discussed in Section XXX),
+and is also the key enabling factor when incorporating propective ARGs into
+forward-time simulation.
+The core algorithm involves the concept of
+``ancestral material''~\citep{wiuf1999ancestry,wiuf1999recombination}:
+the genomic intervals ancestral to a set of samples,
+scattered across lineages (the edges of the graph)
+at some time in the past. In the next section, we outline how
+tracing ancestral material allows the identification and removal of nodes
+and inheritance intervals, such that
+only genetic ancestry relevant to the current population is retained.
 
 \section*{Converting an eARG to a gARG}
-Any eARG can be converted to a gARG without loss of information,
-and the conversion process is very similar to
-simplifying a prospective gARG with respect to a set of samples.
-The process involves two steps, and is illustrated in
-Fig.~\ref{fig-ancestry-resolution}.
-The first step is straightforward (Fig.~\ref{fig-ancestry-resolution}A,B):
-we duplicate the nodes and
-edges from the input eARG, and then add inheritance annotations
-to the output gARG's edges. If the node is a common ancestor node,
+Any eARG can be converted to a gARG without loss of information.
+The process is illustrated in Fig.~\ref{fig-ancestry-resolution} (A to B):
+we convert all event nodes to genome nodes and add inheritance annotations
+to the gARG's edges. If the node is a common ancestor node,
 we annotate the single outbound edge with the interval $[0,L)$,
@@
 These inheritance interval annotations are clearly in one-to-one
-correspondence with the information in the input eARG,
-but they are a superset of the
-``ancestral material''~\citep{wiuf1999ancestry,wiuf1999recombination}
-of the sample.
+correspondence with the information in the input eARG.
 
@@
 Converting the \citet[][Fig.~1]{wiuf1999recombination} example
-to a gARG. (A) The original eARG; square nodes represent events, with
+to a sample-resolved gARG. (A) The original eARG; square nodes represent events, with
 each recombination (red) containing a breakpoint.
@@
 
-Ancestral material is a critical concept in a retrospective ARG,
-and refers to the genomic intervals ancestral to the samples
-scattered across lineages (the edges of the graph)
-at some time in the past. At recombination
-events, ancestral material is split between the ancestral lineages.
+This way of specifying inheritance intervals generates annotations equivalent to
+those produced in a prospective gARG, and which is similarly amenable to ``simplification'',
+by considering the genetic material that is ancestral to the samples \citep{kelleher2018efficient}.
+The outcome of this process, which is
+directly analogous to Hudson's simulation algorithm for the coalescent with
+recombination~\citep{hudson1983testing,kelleher2016efficient}, is shown in
+Fig.~\ref{fig-ancestry-resolution}C. We traverse from the samples rootwards through the graph
+keeping track of which material is ancestral to the samples.
+At recombination events, ancestral material is split between the ancestral lineages.
 As genome segments coalesce in a common
 ancestor the total amount of ancestral material is reduced, until
-eventually, as we traverse rootwards through the graph, all
-samples reach a most recent common ancestor at every location
-along the genome. The process of transmitting ancestral material
-along the edges of an ARG is directly analogous to Hudson's
-simulation algorithm for the coalescent with
-recombination~\citep{hudson1983testing,kelleher2016efficient}.
-
-Fig.~\ref{fig-ancestry-resolution}C shows the graph that we obtain
-by resolving the ancestral material transmitted along each edge
-using the ``simplify'' algorithm~\citep{kelleher2018efficient},
-and highlights some important
-differences between the eARG and gARG encodings.
-Firstly, we can see that some nodes and edges are removed entirely
-from the graph by this conversion process.
+eventually all samples reach a most recent common ancestor at every location
+along the genome.
+
+Fig.~\ref{fig-ancestry-resolution}C therefore represents a \emph{sample-resolved} gARG,
+% useful to say that we don't *need* to simulate an eARG and convert to a 
+% sample-resolved form, but can simulate a sample-resolved gARG directly
+of the sort that can be simulated directly by the Hudson algorithm.
+This differs in some important ways from its equivalent eARG (Fig.~\ref{fig-ancestry-resolution}A).
+Firstly, we can see that some nodes and edges have been removed entirely
+from the graph.
 The ``grand MRCA'' \noderef{q} is omitted from the
@@
 More generally, we can see that the sample resolved
-retrospective gARG of Fig.~\ref{fig-ancestry-resolution}C
+gARG of Fig.~\ref{fig-ancestry-resolution}C
 allows for ``local'' inspection

### 2023-10-14 paper.tex
Fix whitespace

@@
 \linenumbers
-\section*{Introduction}
+\section{Introduction}
+\label{sec-intro}
 Estimating the genetic genealogy of a set of genome sequences
@@
 and subfields use the term
-differently (see Appendix XXX).
+differently (see Appendix~\ref{sec-arg-history}).
 
@@
 % TODO: This needs revision - see also https://github.com/tskit-dev/what-is-an-arg-paper/pull/380
-We continue with a description of the usually held retrospective view of ARGs, but note that ARGs can also be viewed as prospective, particularly in forwards-time simulations.
-In the context of forwards-time simulations, we mention the simplification operation, which removes ancestral genome material that was not inherited.
+We continue with a description of the usually held retrospective view of ARGs,
+but note that ARGs can also be viewed as prospective, particularly in forwards-time
+simulations. In the context of forwards-time simulations, we mention the
+simplification operation, which removes ancestral genome material that was not
+inherited.
 % Converting and eARG to gARG &
@@
 % Levels of simplification
-This simplification is closely related to operations needed to convert eARG and gARG, which opens a rich set of details about ARGs and local trees, local unary nodes, and levels of simplification.
+This simplification is closely related to operations needed to convert eARG and gARG,
+which opens a rich set of details about ARGs and local trees, local unary nodes, and
+levels of simplification.
 % Precision of recombination information
-We then highlight the important difference between the required level of precision about recombination information for eARGs compared to gARGs.
+We then highlight the important difference between the required level of precision
+about recombination information for eARGs compared to gARGs.
 % Computational efficiency and data interchange
-The way this recombination information is stored in eARGs and gARGs has important implications for computational efficiency, which is critical if we want to use ARGs for modern genomic datasets.
+The way this recombination information is stored in eARGs and gARGs has important
+implications for computational efficiency, which is critical if we want to use ARGs
+for modern genomic datasets.
 % Discussion
 We wrap with a discussion highlighting that WHAT-IS-THE-KEY-SINGLE-MESSAGE-NOT-COVERED-ABOVE?
-Finally, the literature on ARGs and associated aspects is vast, particularly due to different definitions and implementations.
-We cover some important aspects in appendices, including a brief history of ancestral graphs (Appendix XXX), description of the big and little ARGs (Appendix ZZZ), survey of ARG inference methods (Appendix YYY), and ARGs at an individual vs cell lineage level (Appendix WWW).
-
-\section*{Genome ARGs}
+Finally, the literature on ARGs and associated aspects is vast, particularly
+due to different definitions and implementations. We cover some important aspects
+in appendices, including
+a brief history of ancestral graphs (Appendix~\ref{sec-arg-history}),
+description of the big and little ARGs (Appendix~\ref{sec-big-and-little-arg}),
+survey of ARG inference methods (Appendix~\ref{sec-survey-arg-infer}), and
+ARGs at an individual vs cell lineage level (Appendix~\ref{sec-cell-lineages-and-args}).
+
+\section{Genome ARGs}
+\label{sec-gARG}
 We define a genome as the complete set of genetic material that a child
@@
 representing the sampled genomes
-(but see section XXX for some subtlety around this point).
+(but see Section~\ref{sec-retro-pro} for some subtlety around this point).
 Each element of $E$ is a tuple $(c, p, I)$ such that $c, p \in N$ are the child and
@@
 contain both paternal and maternal  genomes
-labelled \textsf{a} to \textsf{p}. Black lines show inheritance paths connecting
-genomes in the current generation (\textsf{a} to \textsf{d}) with their ancestors.
-Genomes \textsf{a} and \textsf{c} are the product of two independent
+labelled \noderef{a} to \noderef{p}. Black lines show inheritance paths connecting
+genomes in the current generation (\noderef{a} to \noderef{d}) with their ancestors.
+Genomes \noderef{a} and \noderef{c} are the product of two independent
 recombination events (red) between
-the paternal genomes \textsf{e}
-and \textsf{f}, and regions of genome inherited are shown with shaded colour.
+the paternal genomes \noderef{e}
+and \noderef{f}, and regions of genome inherited are shown with shaded colour.
 Genomes are shaded such that where, backwards in time,
@@
 by the two recombination breakpoints that gave rise to these genomes.
-% Following the above definitions, corresponding data for this gARG is shown in Supplementary? Table XXX.
+Following the above definitions, corresponding data encoding for
+this gARG is shown in Tab.~\ref{tab-gARG-data} (Appendix~\ref{sec-gARG-data}).
 
@@
 genome can have nodes of any arity (number of child nodes), including ``unary''
-nodes with a single child (see Section XXX).
+nodes with a single child (see Section~\ref{sec-locally-unary-edges}).
 % TODO should rephrase this from parent/child to ancestor/descendant?
@@
 Thus we will have a child node (the recombinant; e.g.\ node \noderef{a}
-Fig.~\ref{fig-arg-in-pedigree}) and two (or more; see section XXX)
+Fig.~\ref{fig-arg-in-pedigree}) and two (or more; see Section~\ref{sec-ARG-simplification})
+% TODO: check above if we want to refer to sec-ARG-simplification or some other!
 parental nodes from which this child has inherited genome segments.
@@
 % do not correspond to any particular event in the ancestry of the samples.
-% For example, node \textsf{h} is the direct ancestor of \textsf{d} in
+% For example, node \noderef{h} is the direct ancestor of \noderef{d} in
 % Fig~\ref{fig-arg-in-pedigree}, and is not the product of either a
 % coalescence or a recombination. Such nodes would usually be removed
-% from the graph so that \textsf{d} descends directly from \textsf{k} (see
+% from the graph so that \noderef{d} descends directly from \noderef{k} (see
 % the ``\nameref{ARG_simplification}'' section) but there is no necessity for this from
@@
 % two parent genomes between which recombination takes place
-% (e.g.\ nodes \textsf{e} and \textsf{f}), and a
-% single resulting ``recombinant'' genome (e.g.\ \textsf{c}). The
+% (e.g.\ nodes \noderef{e} and \noderef{f}), and a
+% single resulting ``recombinant'' genome (e.g.\ \noderef{c}). The
 % edges that link this recombinant with its two parents carry the information
@@
 
-\section*{Event ARGs}\label{eARG}
+\section{Event ARGs}
+\label{sec-eARG}
 Genome ARGs are defined by
@@
 an ARG from the literature, to allow us to capture their common features
-(see Appendix XXX for a full review).
+(see Appendix~\ref{sec-arg-history} for a full review).
 
@@
 but the graph structure alone is not sufficient to define the local trees,
-or equivalently, how ``ancestral material'' (Section XXX) flows along
-the edges of the ARG.
+or equivalently, how ``ancestral material'' flows along
+the edges of the ARG (see Section~\ref{sec-eARG-to-gARG}).
+% TODO: check above if we want to refer to sec-eARG-to-gARG or some other!
 This is usually assumed to follow the mechanism described by
@@
 \foreach \u/\lab in {
-        s0/$\textsf{a}$, s1/$\textsf{b}$, s2/$\textsf{c}$,
-        r0/$\textsf{a}$, r1/$\textsf{b}$, r2/$\textsf{c}$,
-        l0/$\textsf{a}$, l1/$\textsf{b}$, l2/$\textsf{c}$}
+        s0/$\noderef{a}$, s1/$\noderef{b}$, s2/$\noderef{c}$,
+        r0/$\noderef{a}$, r1/$\noderef{b}$, r2/$\noderef{c}$,
+        l0/$\noderef{a}$, l1/$\noderef{b}$, l2/$\noderef{c}$}
     \node[nodelabel,anchor=south] at ([yshift=-12pt]\u) {\lab};
@@
 \foreach \u/\lab in {
-        s6/$\textsf{g}$,
-        l4/$\textsf{g}$,
-        r4/$\textsf{g}$}
+        s6/$\noderef{g}$,
+        l4/$\noderef{g}$,
+        r4/$\noderef{g}$}
     \node[nodelabel,anchor=south] at (\u) {\lab};
@@
 \foreach \u/\lab in {
-        s4/$\textsf{e}$,
-        l3/$\textsf{e}$}
+        s4/$\noderef{e}$,
+        l3/$\noderef{e}$}
     \node[nodelabel,anchor=south west] at (\u) {\lab};
 \foreach \u/\lab in {
-        s5/$\textsf{f}$,
-        r3/$\textsf{f}$}
+        s5/$\noderef{f}$,
+        r3/$\noderef{f}$}
     \node[nodelabel,anchor=south east] at (\u) {\lab};
-\foreach \u/\lab in {s3/$\textsf{d}$, s6/$\textsf{g}$}
+\foreach \u/\lab in {s3/$\noderef{d}$, s6/$\noderef{g}$}
     \node[nodelabel,anchor=south] at (\u) {\lab};
@@
 (note these are shown in the conventional form in which only coalescences
-within the local tree are included; see section XXX for a discussion
-of this important point).
-(C) A concrete encoding of the information (right).
+within the local tree are included; see Section~\ref{sec-ARG-and-local-trees}
+% TODO: check above if we want to refer to sec-ARG-and-local-trees or some other!
+for a discussion of this important point).
+(C) A concrete encoding of the information.
 Recombination nodes have a non-null breakpoint
@@
 philosophical question, and in most senses the two are equivalent
-(but see the discussion on ``unary nodes'' in section XXX).
+(but see the discussion on ``unary nodes'' in Section~\ref{sec-locally-unary-edges}).
 From a concrete data-structure encoding perspective, the only real
@@
 systematically describe the uncertainty around their
-ordering (see Section XXX).
+ordering (see Section~\ref{sec-precision}).
 As we describe in the following sections, freeing ourselves from the requirement
@@
 
-\section*{Prospective vs retrospective ARGs}
+\section{Retrospective vs prospective ARGs}
+\label{sec-retro-pro}
 The ARG concept arose from a \emph{retrospective} view of inheritance.
@@
 allows efficient subsetting of ARGs (removing both samples and certain
-levels of detail about recombination, as discussed in Section XXX),
+levels of detail about recombination, as discussed in Section~\ref{sec-ARG-simplification}),
 and is also the key enabling factor when incorporating propective ARGs into
@@
 
-\section*{Converting an eARG to a gARG}
+\section{Converting an eARG to a gARG}
+\label{sec-eARG-to-gARG}
 Any eARG can be converted to a gARG without loss of information.
@@
 only generate ARGs containing nodes and edges ancestral to the sample
-(see the discussion of the ``Big ARG'' stochastic process in Appendix XXX, however).
+(see the discussion of the ``Big ARG'' stochastic process in
+Appendix~\ref{sec-big-and-little-arg}, however).
 Nonetheless, it is an instructive example from the literature which highlights several
@@
 
-\section*{ARGs and local trees}
+\section{ARGs and local trees}
+\label{sec-ARG-and-local-trees}
 Local trees, describing the genetic genealogy of a sample
@@
 Besides having labelled nodes, the local trees must contain
-unary nodes (see the next section) marking recombinations
+unary nodes marking recombinations
 and other processes that lead to lineages passing through a given
-ARG node without coalescence in the local trees (see section XXX).
+ARG node without coalescence in the local trees
+(see Section~\ref{sec-locally-unary-edges}).
 If the local trees contain labelled internal nodes, they precisely
@@
 
-\section*{Locally unary nodes}
+\section{Locally unary nodes}
+\label{sec-locally-unary-edges}
 We can define the arity of a node as the number of child nodes it has.
@@
 A tree branch may correspond to many generations
-of individuals (or indeed cell divisions; see Appendix \ref{cell-lineages-and-args})
+of individuals (or indeed cell divisions; see Appendix \ref{sec-cell-lineages-and-args})
 and there is usually
@@
 different ways in which ARG nodes can be locally unary. Furthermore,
-several current ARG inference methods infer locally unary nodes (see section XXX).
-
-\section*{Levels of simplification}\label{ARG_simplification}
+several current ARG inference methods infer locally unary nodes
+(see Section~\ref{sec-precision}).
+
+\section{Levels of simplification}
+\label{sec-ARG-simplification}
 ARG simplification is a powerful tool, and can do far more
@@
 
-\section*{Precision of recombination information}
+\section{Precision of recombination information}
 \label{sec-precision}
@@
 
-\section*{Computational efficiency and data interchange}
+\section{Computational efficiency and data interchange}
+\label{sec-efficiency}
 ARG inference methods have been successfully applied to vast datasets,
@@
 
-\section*{Discussion}
+\section{Discussion}
+\label{sec-discussion}
 Recent breakthroughs have finally made large-scale ARG inference
@@
 A necessary first step is a degree of terminological clarity.
-As discussed in Appendix XXX, the term ``ancestral recombination
-graph" has several
+As discussed in Appendix~\ref{sec-arg-history}, the term
+``ancestral recombination graph" has several
 subtly different interpretations, depending on context.
@@
 interchange format, already widely used and accompanied
-by high-quality, high-performance software (see Section XXX for details).
+by high-quality, high-performance software
+(see Section~\ref{sec-efficiency} for details).
 Adopting the \texttt{tskit} library
@@
 
-% FIXME: simplest to order this alphabetically by surname?
-% Anastasia TODO
-Jere Koskela was supported by EPSRC research grant EP/V049208/1.
+% Ordered alphabetically by surname
 Gregor Gorjanc acknowledges support from the BBSRC ISP grant to The Roslin Institute (BBS/E/D/30002275, BBS/E/RL/230001A, and BBS/E/RL/230001C).
-% Wilder TODO
+% Anastasia Ignatieva acknowledges support from TODO
 Jerome Kelleher acknowledges support from the Robertson Foundation.
+Jere Koskela acknowledges support from EPSRC research grant EP/V049208/1.
+% Anthony Wilder Wohns acknowledges support from TODO
 
@@
 \appendix
+
+\setcounter{table}{0}
+\setcounter{figure}{0}
+\renewcommand{\thetable}{A\arabic{table}}
+\renewcommand{\thefigure}{A\arabic{figure}}
 
@@
 as originally described requires exponential time to simulate
-(see Appendix XXX for details). However, ARGs provided a way
+(see Appendix~\ref{big-and-little-arg} for details). However, ARGs provided a way
 to reason about and infer recombinant ancestry as a single object,
@@
 \section{The Big and Little ARG}
-\label{app-big-and-little-arg}
+\label{sec-big-and-little-arg}
 Here we review the two predominant stochastic processes which construct ARGs:
@@
 For example, in the gARG encoding depicted in Figure \ref{fig-ancestry-resolution}C
-it is clear that a recombination takes place between nodes \textsf{b} and
-\textsf{d} as well as \textsf{e}, but the exact time of the event is ambiguous,
+it is clear that a recombination takes place between nodes \noderef{b} and
+\noderef{d} as well as \noderef{e}, but the exact time of the event is ambiguous,
 and thus so is the number of lineages during the time interval. In fact, this
@@
 This is also the interpretation depicted in
-Figure \ref{hudson_vs_bigARG}, but it means that the two edges above node \textsf{b}
+Figure \ref{hudson_vs_bigARG}, but it means that the two edges above node \noderef{b}
 in Figure \ref{fig-ancestry-resolution} should correspond to only one lineage,
 along which all $m-1$ links are available for recombination.
-The lineage then splits into two at the time of nodes \textsf{d} and \textsf{e}.
-Nodes \textsf{k}, \textsf{l}, and \textsf{m} in Figure \ref{fig-ancestry-resolution}
+The lineage then splits into two at the time of nodes \noderef{d} and \noderef{e}.
+Nodes \noderef{k}, \noderef{l}, and \noderef{m} in Figure \ref{fig-ancestry-resolution}
 demonstrate that the same issue can affect the number of links available for recombination:
 without an external convention, the exact time at which the trapped ancestral material on
-node \textsf{k} ceases to be available for an effective recombination in the little ARG process.
+node \noderef{k} ceases to be available for an effective recombination in the little ARG process.
 
@@
 	\draw (-1, 1.7) -- (-1, 4.5) -- (0,4.5) -- (0,3.1);
-	\node [scale=0.2,label=above left:{\small \textsf{a}}] at (-0.5,4.5) {a};
+	\node [scale=0.2,label=above left:{\small \noderef{a}}] at (-0.5,4.5) {a};
 	\draw (-0.9, 4.3) -- (-0.6, 4.3) -- (-0.6, 4.2) -- (-0.9, 4.2) -- (-0.9, 4.3);
@@
 	\draw (4,2.4) -- (4,5.4) -- (2,5.4) -- (2,3.1);
-	\node [scale=0.2,label=above left:{\small \textsf{b}}] at (3,5.4) {b};
+	\node [scale=0.2,label=above left:{\small \noderef{b}}] at (3,5.4) {b};
 	\draw (2.6, 5.2) -- (2.9, 5.2) -- (2.9, 5.1) -- (2.6, 5.1) -- (2.6, 5.2);
@@
 	\draw (0.5, 6) -- (0.5, 7.2) -- (3,7.2) -- (3,5.4);
-	\node [scale=0.2,label=above:{\small \textsf{c}}] at (1.75,7.2) {c};
+	\node [scale=0.2,label=above:{\small \noderef{c}}] at (1.75,7.2) {c};
 	\draw (1.35, 6.95) -- (2.15, 6.95);
@@
 propagated until time $T$. The MRCA on the genetic interval $[v, w)$ is reached
-at event \textsf{b}, while that on $[0, v)$ is reached at event \textsf{c}.
+at event \noderef{b}, while that on $[0, v)$ is reached at event \noderef{c}.
 The non-ancestral segment $[v, w)$ above
-\textsf{a} contributes to the rate of effective recombinations because it
+\noderef{a} contributes to the rate of effective recombinations because it
 is trapped between ancestral segments. The two columns titled CA and RE
@@
 \section{Survey of ARG inference methods}
+\label{sec-survey-arg-infer}
 
@@
 \section{Cell lineages and ARGs}
-\label{cell-lineages-and-args}
+\label{sec-cell-lineages-and-args}
 As we have argued, an important benefit of the gARG representation is that it
@@
 \clearpage
-\section*{Supplementary figures}
-
-% TODO add label and change numbering to S1, etc
+\section{Data encoding of the genome ARG}
+\label{sec-gARG-data}
+
+\begin{table}[ht]
+    \caption{\label{tab-gARG-data}
+    Data encoding of the genome ARG shown in Fig.~\ref{fig-arg-in-pedigree} with the
+	node table on the left (having node and time columns) and the edge table on the
+	right (having child, parent, and intervals columns).
+    }
+    \begin{center}
+        \begin{tabular}{c|ccc|c|c}
+			\multicolumn{2}{c}{Node table}& ~ &\multicolumn{3}{c}{Edge table}\\
+			\cline{1-2} \cline{4-6}
+            Node & Time & ~ & Child & Parent & Intervals\\
+            \cline{1-2} \cline{4-6}
+            $\noderef{a}$ & 0 & ~ & $\noderef{a}$ & $\noderef{e}$ & $\{[0,2)\}$\\
+            $\noderef{b}$ & 0 & ~ & $\noderef{a}$ & $\noderef{f}$ & $\{[2,10)\}$\\
+            $\noderef{c}$ & 0 & ~ & $\noderef{b}$ & $\noderef{g}$ & $\{[0,10)\}$\\
+            $\noderef{d}$ & 0 & ~ & $\noderef{c}$ & $\noderef{f}$ & $\{[0,7)\}$\\
+            $\noderef{e}$ & 1 & ~ & $\noderef{c}$ & $\noderef{e}$ & $\{[7,10)\}$\\
+            $\noderef{f}$ & 1 & ~ & $\noderef{d}$ & $\noderef{h}$ & $\{[0,10)\}$\\
+            $\noderef{g}$ & 1 & ~ & $\noderef{e}$ & $\noderef{i}$ & $\{[0,10)\}$\\
+            $\noderef{i}$ & 1 & ~ & $\noderef{f}$ & $\noderef{k}$ & $\{[0,10)\}$\\
+            $\noderef{k}$ & 2 & ~ & $\noderef{g}$ & $\noderef{i}$ & $\{[0,10)\}$\\
+            $\noderef{n}$ & 3 & ~ & $\noderef{h}$ & $\noderef{k}$ & $\{[0,10)\}$\\
+            \multicolumn{2}{c}{~}& ~ & $\noderef{i}$ & $\noderef{n}$ & $\{[0,10)\}$\\ 
+            \multicolumn{2}{c}{~}& ~ & $\noderef{k}$ & $\noderef{n}$ & $\{[0,10)\}$
+        \end{tabular}
+    \end{center}
+\end{table}
+
+\clearpage
+\section{TODO}
+\label{sec-x}
+
+% TODO: we don't seem to cite this figure at all!?
+
 \begin{figure}[ht]

### 2023-10-15 paper.tex
Minor tweaks and notes

@@
 
+% TODO add a few sentences here on why defining nodes concretely as
+% integers has significant advantages.
+% 1) leads to direct link between the ARG and local trees in terms
+%    of a well-defined combinatorial object (local trees section)
+% 2) Definitions can be implemented directly in an efficient way
+%    (computational efficiency section.
 In many settings, nodes are dated, i.e.\
@@
 two parents, $D_3$ and $D_4$. Each inherited genome may be the recombined product
-of the genomes contained within a parent. In this example,
+of the genomes
+% FIXME "contained within" is a poor phrase here, use "constituent" or something
+contained within a parent.
+In this example,
 genome \noderef{b} was inherited directly from $D_4$'s genome \noderef{g} without
 recombination, whereas
-genome \noderef{a} is the recombinant of
+genome \noderef{a} is the recombinant product of
 $D_2$'s genomes \noderef{e} and \noderef{f} crossing over at position 2.
@@
 associated with the inheritance of DNA between generations.
+% FIXME this sentance doesn't follow well from the first, needs a
+% different bridge before listing important properties.
+% Maybe the "listicle" format here isn't right anyway.
 As such,
@@
 % Wilder: give forward ref to section on efficiency here, mention tree sequence
+% TODO, should probably drop this point.
 Thirdly, local trees that we generate for a particular position on the
@@
 \label{sec-eARG}
+% FIXME - this intro doesn't work at all now - we haven't defined what an
+% eARG *is* before discussing it.
+% Maybe best to skip this entirely, and make the quoted definitions with
+% a thin intro the way to start the section.
 Genome ARGs are defined by
@@
 
+% TODO reference forward to the cell ARG appendix. Ultimately
+% every event corresponds to a particular genome in a particular
+% individual or cell.
 Whether we regard nodes as events or genomes is something of a

### 2023-10-15 argutils/viz.py
Make symbols round in fig 1

@@ -59,6 +59,7 @@ def draw(
     font_size=None,
     node_color=None,
     font_color=None,
+    node_symbols="os",
     reverse_x_axis=None,
 ):
     """
@@ -118,6 +119,10 @@ def draw(
     directly. In particular this means that node_color can either be a single
     colour for all nodes (e.g. `mpl.colors.to_hex(mpl.pyplot.cm.tab20(1))`)
     or a list of colours (which is ignored if node_arity_colors is True)
+
+    node_symbols is a string of length 2, where the first character is the
+    matplotlib symbol used for nonsample nodes, and the second is the symbol
+    used for sample nodes. By default these are circles and squares ("o" and "s")
 
     if reverse_x_axis is True, the graph is reflected horizontally
 
@@ -200,7 +205,7 @@ def draw(
             arity_edge_color[i] = make_color(col)
             node_color[i] = make_color(col, lighten= 1 - (c_spans[n] / spans[n]))
 
-    for use, shape, size in zip((is_sample, ~is_sample), ("s", "o"), (node_size*0.8, node_size)):
+    for use, shape, size in zip((~is_sample, is_sample), node_symbols, node_size * np.array([1, 0.8])):
         try:
             node_col = node_color[use]
         except TypeError:

### 2023-10-16 paper.tex
Notes
Updates from Overleaf

@@
 
-The exhastive recording of full inheritance intervals from generation
+The exhaustive recording of full inheritance intervals from generation
 to generation leads to a rapid build-up of information, much of which
@@
 
+% TODO Need to make clear here what came from where. Basic combinatorial object
+% cite Knuth. Application to genetic ancestry, sim papers. Need to look
+% up what's said in favour of this integer based approach in which paper
+% and refer to tat.
 To make the following discussions precise, we provide
@@
 in the local tree at position $x$, then we adopt the convention that $\pi_u = -1$.
-% This sentence needs another pass
-A node in an oriented tree can have
-zero, one,
-or many children.
+% TODO Give example of an oriented tree from one of the figs?
+% TODO
+% A node in an oriented tree can have
+% zero, one,
+% or many children.
+
 To recover the local tree $\pi$ at position $x$, we begin by
@@
 If the local trees contain labelled internal nodes, they precisely
-describing the corresponding path through the graph for each
+describe the corresponding path through the graph for each
 genome interval and there is exactly as
@@
 consider the inference strategy of estimating local trees
-% JK: Relate is inferring trees independently within segments,
-% it's just using a dynamic algorithm to decide what those segments
-% are, right? That's how it differs from things like Espalier which
-% assume a set of input recomb breakpoints.
+% TODO rephrase this. Ana doesn't like the way it implies that the
+% trees are independent, because info from mutations outside the
+% window are shared. However, each tree *inference* is independent,
+% whether information is used multiple times in different trees or not
 independently within genome segments,
@@
 We illustrate this successive removal of redundancy
-through a series simplification steps
-in Fig.~\ref{fig-simplification}.
+through a series of simplification steps
+in Fig.~\ref{fig-simplification};
+the graph is shown on the left and the corresponding sequence
+of local trees on the right.
 
@@
 material of a lineage is split between the two parental
-genomes. Local coalesence occurs
+genomes. Local coalescence occurs
 when lineages with overlapping ancestral material choose the same
@@
 
-Fig.~\ref{fig-simplification} shows four ARGs (A--D), with
-the graph depicted on the left and the corresponding sequence
-of local trees on the right. The graph visualisations have
+The graph visualisations in Fig.~\ref{fig-simplification} have
 some novel features, which require some explanation.
@@
 have in the graph, allowing us to easily see roots (those
-with zero parents), recombinanants (those with two parents)
+with zero parents), recombinants (those with two parents)
 and more complex situations arising from simplification (see below).
@@
 ancestors,
-\citep{osmond2021estimating} downsample trees along the genome
+\cite{osmond2021estimating} downsample trees along the genome
 so that they can be regarded as approximately independent.
@@
 as originally described requires exponential time to simulate
-(see Appendix~\ref{big-and-little-arg} for details). However, ARGs provided a way
+(see Appendix~\ref{sec-big-and-little-arg} for details). However, ARGs provided a way
 to reason about and infer recombinant ancestry as a single object,

### 2023-10-16 paper.tex
Various edits

@@
 \label{sec-locally-unary-edges}
+% TODO needs intro, giving some bit of warm-up for the reader.
 We can define the arity of a node as the number of child nodes it has.
@@
 because a (graph) child node is only a child in the local tree for position
+% FIXME state this formally, as both Yan and Jere misunderstood here.
 $x$ if $x$ is in the inheritance interval for the corresponding edge.
@@
 interval and it is hence unary in this local tree.
+% TODO note this is CA but not coalescent, and refer to where we discuss this.
 In another example, \noderef{e} is binary in the graph, but is
@@
 
+% TODO add explanation of the bottom row.
+% https://github.com/tskit-dev/what-is-an-arg-paper/issues/386
 The example of successive simplification of a simulated ARG

### 2023-10-16 paper.tex

@@
 Following the above definitions, corresponding data encoding for
-this gARG is shown in Tab.~\ref{tab-gARG-data} (Appendix~\ref{sec-gARG-data}).
+this gARG is shown in Tab.~\ref{tab-gARG-data}.
 
@@
 and corresponding local trees on the right.
+% Repetition here, but I'm thinking about the casual "I'm just reading
+% the captions reader" who might not realise that there's an essential
+% ingredient missing from the graph version
+In the interest of visual clarity, inheritance intervals are not shown
+on the graph edges; Supplementary Fig.~\ref{fig-simplification-with-edges}
+shows the graphs with these inheritance intervals included.
 Graph nodes are coloured by the number of parents and shaded
@@
 \clearpage
-\section{Data encoding of the genome ARG}
-\label{sec-gARG-data}
+\renewcommand\thefigure{S\arabic{figure}}
+\setcounter{figure}{0}
+\renewcommand\thetable{S\arabic{table}}
+\setcounter{table}{0}
+
+\section*{Supplementary Material}
 
 \begin{table}[ht]
-    \caption{\label{tab-gARG-data}
-    Data encoding of the genome ARG shown in Fig.~\ref{fig-arg-in-pedigree} with the
-	node table on the left (having node and time columns) and the edge table on the
-	right (having child, parent, and intervals columns).
-    }
     \begin{center}
@@
     \end{center}
+    % NOTE: I know  the "correct" placement for table captions is on top, but I
+    % always find this quite weird (since many of our figs are quite "table
+    % like" anyway. If anyone feels strongly about putting the caption above
+    % the table please just move it back.
+    \caption{\label{tab-gARG-data}
+    Data encoding of the genome ARG shown in Fig.~\ref{fig-arg-in-pedigree} with the
+	node table on the left (having node and time columns) and the edge table on the
+	right (having child, parent, and intervals columns).
+    }
 \end{table}
 
-\clearpage
-\section{TODO}
-\label{sec-x}
-
-% TODO: we don't seem to cite this figure at all!?
-
-\begin{figure}[ht]
+\begin{figure}[]
 	\begin{center}
@@
 	\caption{\label{fig-simplification-with-edges}
-	The four levels of ARG simplification from Fig.~\ref{fig-simplification}A--D, with edge intervals shown.
+	Example ARGs Fig.~\ref{fig-simplification}A--D,
+    with edges annotated with inheritance intervals.
 	}

### 2023-10-16 paper.tex
tweak
Add Fan et al 2023
Closes  #389

@@
 appear~\citep{osmond2021estimating,fan2022genealogical,hejase2022deep,zhang2023biobank,
-nowbandegani2023extremely,ignatieva2023distribution}
+nowbandegani2023extremely,ignatieva2023distribution,fan2023likelihood}
 and many more are sure to
@@
 general~\citep[e.g.][]{mathieson2020ancestry,hejase2020summary,
-schaefer2021ancestral,harris2023using,zhang2023biobank},
+schaefer2021ancestral,harris2023using,zhang2023biobank,fan2023likelihood},
 encompassing the varied outputs from modern simulation and
@@
 mahmoudi2022bayesian,petr2022slendr,rasmussen2022espalier,
-zhang2023biobank,nowbandegani2023extremely,ignatieva2023distribution}.
+zhang2023biobank,nowbandegani2023extremely,ignatieva2023distribution,
+fan2023likelihood}.
 
@@
 for downstream applications.
-To date, few downstream applications using estimated ARGs
-make use of complete recombination information.
+% To date, few make use such detailed recombination information.
 Most applications work by examining local trees independently.
 For example, the \relate\ selection test~\citep{speidel2019method}
-obtains $p$-values by computing clade size probabilities conditional on the timing of coalescence events in a given local tree.
+obtains $p$-values by computing clade size probabilities conditional
+on the timing of coalescence events in a given local tree.
 In their method
@@
 so that they can be regarded as approximately independent.
+Similarly, \cite{fan2023likelihood} compute the likelihood
+of an ARG under a particular demographic model as the product
+over a sample of widely-separated local trees, assumed to be independent.
 The SIA method for detecting selection~\citep{hejase2022deep}
@@
 
-\begin{figure}[]
+\begin{figure}
 	\begin{center}

### 2023-10-16 paper.tex
yan's upates

@@
 For instance, several human datasets now consist of hundreds of thousands of
-% TODO cite GeL paper
-genomes~\citep{bycroft2018genome,karczewski2020mutational,tanjo2021practical},
+genomes~\citep{turnbull2018hundred, bycroft2018genome,karczewski2020mutational,tanjo2021practical},
 and so sample size is substantially \emph{larger} than the

### 2023-10-16 make_example_outputs.py
Tidy the inference plots
and don't use tsdate

@@ -7,7 +7,6 @@ import os
 import subprocess
 
 import tsinfer
-import tsdate
 import tskit
 import argutils
 import numpy as np
@@ -25,24 +24,22 @@ def cli():
 @click.command()
 def run_tsinfer():
     sample_data = tsinfer.load("examples/Kreitman_SNP.samples")
-    ts = tsinfer.infer(sample_data)
+    ts = tsinfer.infer(sample_data, num_threads=0)
     # For the moment, tsinfer uses byte metadata, so we need to convert it to json
     # so that we can label the nodes easily
     tables = ts.dump_tables()
     tables.nodes.metadata_schema = tskit.MetadataSchema.permissive_json()
+    time_map = {n.time: n.time for n in ts.nodes()}
     for n in ts.nodes():
         tables.nodes[n.id] = tables.nodes[n.id].replace(
-            metadata = json.loads(n.metadata.decode() or "{}"))
-    # temporary hack: remove the ultimate ancestor if it exists
-    oldest_node = np.argmax(tables.nodes.time)
-    if np.sum(tables.edges.parent==oldest_node) == 1:
-        # only a single edge connects to the root. This is a unary "ultimate ancestor"
-        # and can be removed (it will be removed in later tsinfer versions anyway)
-        use = np.arange(tables.nodes.num_rows)
-        use = use[use != oldest_node]
-        tables.subset(use)
+            metadata = json.loads(n.metadata.decode() or "{}"),
+            time = time_map[n.time],
+        )
+        if not n.is_sample():
+            time_map[n.time] = np.nextafter(time_map[n.time], np.inf)
+    # break ties by node order
+    tables.sort()
     ts = tables.tree_sequence()
-    ts = tsdate.date(ts, mutation_rate = mu, Ne = Ne)
     ts.dump("examples/Kreitman_SNP_tsinfer.trees")
 
 

### 2023-10-17 paper.tex
Add SCAR paper
Closes #392

@@
 schaefer2021ancestral,wohns2022unified,zhang2023biobank,zhan2023towards}
-and data representation~\citep{kelleher2016efficient,kelleher2018efficient}
+% and data representation~\citep{kelleher2016efficient,kelleher2018efficient}
 have raised the realistic prospect of ARG-based analysis becoming a standard part
@@
 Applications using inferred ARGs as input have begun to
-appear~\citep{osmond2021estimating,fan2022genealogical,hejase2022deep,zhang2023biobank,
-nowbandegani2023extremely,ignatieva2023distribution,fan2023likelihood}
+appear~\citep{osmond2021estimating,
+fan2022genealogical,
+hejase2022deep,
+guo2022recombination,
+zhang2023biobank,
+nowbandegani2023extremely,
+ignatieva2023distribution,
+fan2023likelihood}
 and many more are sure to
 follow~\citep{harris2019database,harris2023using}.
+% TODO finish this sentence with "and more reasons why ARGs are cool"
 See \citet{lewanski2023era} for a biologically oriented introduction to ARGs.
@@
 across the wide range of possible applications
-% TODO more?
+% TODO get rid of this, it's a duplication of the introduction
 areas~\citep{rasmussen2014genome,harris2019database,
@@
 baumdicker2021efficient,
-fan2022genealogical,korfmann2022weak,
-mahmoudi2022bayesian,petr2022slendr,rasmussen2022espalier,
-zhang2023biobank,nowbandegani2023extremely,ignatieva2023distribution,
+fan2022genealogical,
+guo2022recombination,
+korfmann2022weak,
+mahmoudi2022bayesian,
+petr2022slendr,
+rasmussen2022espalier,
+zhang2023biobank,
+nowbandegani2023extremely,
+ignatieva2023distribution,
 fan2023likelihood}.
@@
 to compute a likelihood under the coalescent with
-recombination~\citep[e.g.][]{kuhner2000maximum,mahmoudi2022bayesian},
+recombination~\citep[e.g.][]{kuhner2000maximum,mahmoudi2022bayesian,
+guo2022recombination},
 which currently requires the details of all recombination
@@
 % Ordered alphabetically by surname
-Gregor Gorjanc acknowledges support from the BBSRC ISP grant to The Roslin Institute (BBS/E/D/30002275, BBS/E/RL/230001A, and BBS/E/RL/230001C).
+Gregor Gorjanc acknowledges support from the BBSRC ISP grant to The Roslin Institute
+(BBS/E/D/30002275, BBS/E/RL/230001A, and BBS/E/RL/230001C).
 % Anastasia Ignatieva acknowledges support from TODO
+% TODO Jerome: add EPSRC and NIH grants.
 Jerome Kelleher acknowledges support from the Robertson Foundation.

### 2023-10-20 illustrations.py
added Genome position (kb) axis label

@@ -786,6 +786,7 @@ def inference():
             node_arity_colors=True,
             max_edge_width=2,
             tweak_x={},
+            font_size=8.5,
             #use_ranked_times = False,  # By default use Y axis layout from graphviz dot
         )
         path = pathlib.Path(f"illustrations/assets/{name}.json")
@@ -881,8 +882,10 @@ def inference():
             ax_edges.add_patch(rect)
 
     graph_io = io.StringIO()
+    # Add the central label for all plots
+    fig.text(0.5, 0.015, "Genomic position (kb)", ha="center", va="center")
     fig.tight_layout()
-    fig.subplots_adjust(bottom=0)
+    fig.subplots_adjust(bottom=0.035)
     plt.savefig(graph_io, format="svg")
     graph_svg = graph_io.getvalue()
     plt.close()

### 2023-10-20 illustrations.py
Add x-axis labels to inference plot

@@ -768,7 +768,7 @@ def inference():
     widths = np.ones(len(tree_seqs))
     heights = [1, 0.11]
     fig, axes = plt.subplots(
-        2, len(tree_seqs), figsize=(10, 5.2),
+        2, len(tree_seqs), figsize=(10, 5.55),
         gridspec_kw=dict(width_ratios=widths, height_ratios=heights),
         # autoscale_on=False
     )
@@ -838,14 +838,15 @@ def inference():
         # params["tweak_x"] = None
         # params["reverse_x_axis"] = None
         # argutils.viz.draw(ts, ax, edge_colors=edge_colors_by_node, pos=pos, **params)
-        ax.text(270, -35, lab, fontsize=20, family="serif")
+        ax.text(260, -38, lab, fontsize=20, family="serif")
 
         ax_edges.set_title(f"{name} ({subtitle})", y=1.05)
 
         # Add breakpoints and stacked edges below tree sequences, w same colormap as above
-        ax_edges.axis("off")
-        left_coord = 0.05
-        width = 0.9
+        #ax_edges.axis("off")
+        ax_edges.get_yaxis().set_visible(False)
+        left_coord = 0
+        width = 1
 
         edges = {}
         times = ts.tables.nodes.time
@@ -879,17 +880,18 @@ def inference():
                 (1 / ts.num_edges), linewidth=1,
                 facecolor=edge_colors_by_id[edge_id])
             ax_edges.add_patch(rect)
-
+        ax_edges.tick_params(axis='x', which='major', pad=1, length=2) 
+        ax_edges.set_xticks([0, 1], ["-63", "2357"], fontsize=8)
+        ax_edges.set_xlabel("Genomic position (bp)", fontsize=8, labelpad=-7)
     graph_io = io.StringIO()
     fig.tight_layout()
-    fig.subplots_adjust(bottom=0)
     plt.savefig(graph_io, format="svg")
     graph_svg = graph_io.getvalue()
     plt.close()
 
     
     svg = [
-        '<svg width="960" height="480" xmlns="http://www.w3.org/2000/svg" '
+        '<svg width="960" height="500" xmlns="http://www.w3.org/2000/svg" '
         'xmlns:xlink="http://www.w3.org/1999/xlink">'
     ]
     svg.append('<g transform="translate(0, -19)">' + graph_svg[graph_svg.find("<svg"):] + "</g>")

### 2023-10-20 paper.tex
Fill in Example inferred ARGs section

--- 2023-10-16 paper.tex
@@
 \usepackage{subcaption}
+\usepackage{siunitx}
 \usepackage{tikz}
@@
 appear~\citep{osmond2021estimating,fan2022genealogical,hejase2022deep,zhang2023biobank,
-nowbandegani2023extremely,ignatieva2023distribution,fan2023likelihood}
+nowbandegani2023extremely,ignatieva2023distribution}
 and many more are sure to
@@
 general~\citep[e.g.][]{mathieson2020ancestry,hejase2020summary,
-schaefer2021ancestral,harris2023using,zhang2023biobank,fan2023likelihood},
+schaefer2021ancestral,harris2023using,zhang2023biobank},
 encompassing the varied outputs from modern simulation and
@@
 
+\section{Example inferred ARGs}
+\label{sec-example-inferred-args}
+
 \begin{figure} \begin{center}
 \includegraphics[width=\textwidth]{illustrations/inference.pdf} \end{center}
-\caption{\label{fig-inferred-args} Inference results for
-(A) \kwarg, (B) \argweaver, (C) \tsinfer,
-and (D) \relate\
-for 11 samples of a 2.7kb region of the \textit{Drosophila melanogaster} ADH
-locus~\citep{kreitman1983nucleotide}.
-The \argweaver\ example
-is a sample from the inferred posterior distribution.
+\caption{\label{fig-inferred-args} Inference of an 11 sample ARG over a 2.7kb
+region of the \textit{Drosophila melanogaster} ADH locus~\citep{kreitman1983nucleotide}.
+Results for four different methods based on
+(A) parsimony, (B) sampling the SMC$\prime$, (C) copying from
+inferred ancestral haplotypes, and (D) correlated local tree estimation.
 Edge colours are based on the time of the edge's child node
@@
 Position on the X and Y axes for the graph visualization is arbitrary.
-Line width and node colour are as described in Fig.\ref{fig-simplification}.
-The genomic intervals associated with edges are shown below
-each ARG, where each row is an edge
-and the X axis reflects genomic position.
-Vertical lines indicate breakpoints between local trees.
+Line width and node colour are as described in Fig.~\ref{fig-simplification}.
+The graphic below each ARG shows the genomic intervals for each graph edge in the
+corresponding ARGs: edges are horizontal lines, stacked in time order
+(edges with youngest children at the bottom), with vertical dashed lines indicating
+breakpoints between local trees.
 }
@@
 
+The scalability gains made by recent ARG inference methods have been, in part, due to inferring
+lower levels of precision about recombination. Indeed, a common statement in the literature is
+that modern tools such as \tsinfer \citep{kelleher2019inferring} and \relate \citep{speidel2019method}
+construct not an ARG but "only" a sequence of local trees \[cite examples from \#38\]. While this statement
+is correct inasmuch as neither method attempts to infer the explicit recombination events that cause
+adjacent trees to differ, it is certainly not the case that they infer independent local trees.
+In fact, both methods take advantage of genome-wide mutation information to construct genealogies,
+and as a result nodes and edges persist across multiple local trees.
+Their results can therefore be thought of as an inheritance graph, and represented in the gARG format.
+
+To illustrate the varying levels of information that can be captured in a gARG,
+Fig.~\ref{fig-inferred-args} presents graph representations of genealogies output by two tools that
+\emph{do} explicitly infer recombination events, namely \kwarg \citep{ignatieva2021kwarg} and \argweaver
+\citep{rasmussen2014genome}, and two which do not (tsinfer and Relate). All inferences are based on the
+same underlying dataset from \citet{kreitman1983nucleotide}, a standard benchmark in the classical
+ARG literature, for which the minimum number of recombination breakpoints in the absence of recurrent
+mutation is 7~\citep{song2003parsimonious}. This dataset comprises 43 biallelic SNPs spanning 2.4Kb of the
+\emph{D.~melanogaster} ADH locus on chromosome 2L. Where required for inference purposes we
+assume a mutation rate of \SI{5.49e-9}{site^{-1}generation^{-1}} and a recombination rate of
+\SI{8.4e-9}{site^{-1}generation^{-1}} [CITE stdpopsim] and a constant effective population size of 1000,000.
+The version numbers of inference software, together with the specific inference parameters
+needed to reproduce the ARGs, are available from our GitHub repository.
+
+Note that both \kwarg and \argweaver produce many ARGs
+(those shown in Fig.~\ref{fig-inferred-args} A\& B are arbitrarily selected examples).
+This stems from the fact that \kwarg searches the space of plausible ARGs,
+outputting minimal ones using heuristics, and ARGweaver samples from the posterior distribution over
+ARGs given the data and a fixed model (the SMC$\prime$, an approximation of the full coalescent
+with recombination). In contrast, \tsinfer and \relate (Fig.~\ref{fig-inferred-args} C \& D) are
+deterministic methods, aiming to find a single ``best guess'' ARG.
+The ability of a gARG to encode the effects of recombination, rather
+than having to precisely specify each event, is a important contributor to this aim,
+as most recombination events cannot be uniquely identified from the data.
+Indeed, there is considerable variation in the number of recombinations inferred by
+different methods, with e.g.~\argweaver (Fig.~\ref{fig-inferred-args} B) suggesting
+far more than are required under minimal parsimony assumptions (Fig.~\ref{fig-inferred-args} A).
+
+A rough feel for the amount of recombination in each ARG is given by the distribution
+of number of parents per node, as highlighted by the node colour scheme.
+In Fig.~\ref{fig-inferred-args} A \& B, each recombination node has exactly 2 parents,
+corresponding to a single recombination event with a specific breakpoint. Assuming that each
+breakpoint occurs a a different location (the "infinite breakpoints" model), then the
+number of breakpoints should exactly equal the number of 2-parent (brown) nodes in the plot.
+In contrast, Figure Fig.~\ref{fig-inferred-args} C \& D have ARG nodes with greater
+than two parents, indicating deliberate imprecision: these are cases where multiple recombination events
+have been ``pushed down'' onto a single more recent node. Note, however, that the number
+of breakpoints in these two ARGs is smaller than the number of ARG nodes with multiple parents,
+because those nodes frequently share breakpoint positions. The sharing of breakpoint locations
+between multiple-parent nodes indicates that in \tsinfer and \relate ARGs it may be
+difficult to condense each transition between trees into a set of simple SPR moves.
+
+Shading by coalescent span reveals that many non-recombinant (green) nodes are
+``locally unary'' (see section \ref{sec-locally-unary-edges}). In other words,
+in some regions of the genome, these nodes are present in the ancestry but do
+not correspond to coalescence. As argued previously, this can provide potentially useful
+information about shared ancestry. Note however, that such unary segments are not inferred
+in the \relate ARG: other than the samples it has entirely solid nodes.
+Only the \tsinfer results have nodes with more than two children (polytomies);
+in Appendix \ref{sec-cell-lineages-and-args} we argue that this reflects
+uncertainty in the order of coalescent events, in an equivalent
+manner to the lack of precision in recombination nodes with more
+than two parents. The clearest example is the polytomy involving the nodes
+\noderef{Fr-F}, \noderef{Wa-F}, and \noderef{Af-F}, which happen to
+have identical sequences. \kwarg, \argweaver, and \relate require
+bifurcating trees, so are forced
+to pick an arbitrary order of branching
+(hence Fig.~\ref{fig-inferred-args} A and Fig.~\ref{fig-inferred-args} B disagree on the
+relationship between these samples). TsThe number of nodes in Fig.~\ref{fig-inferred-args} C
+that have more than two children also hints at considerable uncertainty in
+the order of branching elsewhere in the ARG.
 % TODO add explanation of the bottom row.
 % https://github.com/tskit-dev/what-is-an-arg-paper/issues/386
-The example of successive simplification of a simulated ARG
-in Fig.~\ref{fig-simplification} illustrates the point
-that is is possible to \emph{store} arbitrary levels of precision
-about recombination in a gARG.
-It is also possible to \emph{infer} various levels of precision about
-recombination.
-To illustrate this, and to demonstrate the structural
-heterogeneity of the ARGs inferred  by current methods,
-Fig.~\ref{fig-inferred-args} shows the results of ARG inference
+%The example of successive simplification of a simulated ARG
+%in Fig.~\ref{fig-simplification} illustrates the point
+%that is is possible to \emph{store} arbitrary levels of precision
+%about recombination in a gARG.
+%It is also possible to \emph{infer} various levels of precision about
+%recombination.
+%To illustrate this, and to demonstrate the structural
+%heterogeneity of the ARGs inferred  by current methods,
+%Fig.~\ref{fig-inferred-args} shows the results of ARG inference
 % To illustrate these points and to emphasise the heterogeneity
 % of the ARGs inferred by current methods, we inferred ARGs for the
-for the classical \citet{kreitman1983nucleotide} benchmark dataset
-using  \kwarg~\citep{ignatieva2021kwarg},
-\argweaver~\citep{rasmussen2014genome,hubisz2020inference},
-\tsinfer~\citep{kelleher2019inferring},
-and \relate~\citep{speidel2019method}.
+%for the classical \citet{kreitman1983nucleotide} benchmark dataset
+%using  \kwarg~\citep{ignatieva2021kwarg},
+%\argweaver~\citep{rasmussen2014genome,hubisz2020inference},
+%\tsinfer~\citep{kelleher2019inferring},
+%and \relate~\citep{speidel2019method}.
 % While there is some consistency among the inferred ARGs (such as the samples
@@
 % are also considerable differences.
-The minimum number
-of recombination breakpoints required in the absence of recurrent mutation
-is 7 for this dataset~\citep{song2003parsimonious},
-and the number of inferred breakpoints
-under the input mutation and recombination rates
-varies from 2 (\relate) to 17 (\argweaver).
+%The minimum number
+%of recombination breakpoints required in the absence of recurrent mutation
+%is 7 for this dataset~\citep{song2003parsimonious},
+%and the number of inferred breakpoints
+%under the input mutation and recombination rates
+%varies from 2 (\relate) to 17 (\argweaver).
 % TODO this needs a bit of rewriting with the rest of the paper in mind
@@
 % not to confuse people.
-The ARGs inferred by the different methods also have major differences
-in how recombination information is represented.
-Essentially,
-\kwarg\ and \argweaver\ infer eARGs that contain
-specific recombination \emph{events} for each breakpoint.
-On the other hand, \tsinfer\ and
-\relate\ do not infer events but rather
-infer gARGs with by differing levels of precision about recombination.
+%The ARGs inferred by the different methods also have major differences
+%in how recombination information is represented.
+%Essentially,
+%\kwarg\ and \argweaver\ infer eARGs that contain
+%specific recombination \emph{events} for each breakpoint.
+%On the other hand, \tsinfer\ and
+%\relate\ do not infer events but rather
+%infer gARGs with by differing levels of precision about recombination.
+
 The bottom row of Fig~\ref{fig-inferred-args} shows the extent along
@@
 than other methods, but it certainly does not infer a series of
-\emph{unrelated} local trees (i.e., with either unlabelled internal
-nodes or differently labelled nodes in each tree).
+\emph{unrelated} local trees
+%(i.e., with either unlabelled internal
+%nodes or differently labelled nodes in each tree).
 The ARGs inferred by \relate\ and \tsinfer\ both contain substantial
@@
 mahmoudi2022bayesian,petr2022slendr,rasmussen2022espalier,
-zhang2023biobank,nowbandegani2023extremely,ignatieva2023distribution,
-fan2023likelihood}.
+zhang2023biobank,nowbandegani2023extremely,ignatieva2023distribution}.
 
@@
 for downstream applications.
-% To date, few make use such detailed recombination information.
+To date, few downstream applications using estimated ARGs
+make use of complete recombination information.
 Most applications work by examining local trees independently.
 For example, the \relate\ selection test~\citep{speidel2019method}
-obtains $p$-values by computing clade size probabilities conditional
-on the timing of coalescence events in a given local tree.
+obtains $p$-values by computing clade size probabilities conditional on the timing of coalescence events in a given local tree.
 In their method
@@
 so that they can be regarded as approximately independent.
-Similarly, \cite{fan2023likelihood} compute the likelihood
-of an ARG under a particular demographic model as the product
-over a sample of widely-separated local trees, assumed to be independent.
 The SIA method for detecting selection~\citep{hejase2022deep}

### 2023-10-20 paper.tex
Change 2.7kb to 2.4kb
Because we now cut off the tail of the sequence, where there is no site information. This makes it easier to display default tsinfer outputs, which are trimmed to the first and last site.

@@
 \includegraphics[width=\textwidth]{illustrations/inference.pdf} \end{center}
-\caption{\label{fig-inferred-args} Inference of an 11 sample ARG over a 2.7kb
+\caption{\label{fig-inferred-args} Inference of an 11 sample ARG over a 2.4kb
 region of the \textit{Drosophila melanogaster} ADH locus~\citep{kreitman1983nucleotide}.

### 2023-10-20 paper.tex

@@
 Applications using inferred ARGs as input have begun to
-appear~\citep{osmond2021estimating,fan2022genealogical,hejase2022deep,zhang2023biobank,
-nowbandegani2023extremely,ignatieva2023distribution}
+appear~\citep{osmond2021estimating,
+fan2022genealogical,
+hejase2022deep,
+guo2022recombination,
+zhang2023biobank,
+nowbandegani2023extremely,
+ignatieva2023distribution,
+fan2023likelihood}
 and many more are sure to
 follow~\citep{harris2019database,harris2023using}.
+% TODO finish this sentence with "and more reasons why ARGs are cool"
 See \citet{lewanski2023era} for a biologically oriented introduction to ARGs.
@@
 baumdicker2021efficient,
-fan2022genealogical,korfmann2022weak,
-mahmoudi2022bayesian,petr2022slendr,rasmussen2022espalier,
-zhang2023biobank,nowbandegani2023extremely,ignatieva2023distribution}.
+fan2022genealogical,
+guo2022recombination,
+korfmann2022weak,
+mahmoudi2022bayesian,
+petr2022slendr,
+rasmussen2022espalier,
+zhang2023biobank,
+nowbandegani2023extremely,
+ignatieva2023distribution,
+fan2023likelihood}.
 

### 2023-10-20 paper.tex
Fix minor latex issues.

@@
 lower levels of precision about recombination. Indeed, a common statement in the literature is
-that modern tools such as \tsinfer \citep{kelleher2019inferring} and \relate \citep{speidel2019method}
-construct not an ARG but "only" a sequence of local trees \[cite examples from \#38\]. While this statement
+that modern tools such as \tsinfer~\citep{kelleher2019inferring} and \relate~\citep{speidel2019method}
+construct not an ARG but ``only'' a sequence of local 
+trees [TODO cite examples from \#38]. While this statement
 is correct inasmuch as neither method attempts to infer the explicit recombination events that cause
@@
 Fig.~\ref{fig-inferred-args} presents graph representations of genealogies output by two tools that
-\emph{do} explicitly infer recombination events, namely \kwarg \citep{ignatieva2021kwarg} and \argweaver
-\citep{rasmussen2014genome}, and two which do not (tsinfer and Relate). All inferences are based on the
+\emph{do} explicitly infer recombination events, namely \kwarg~\citep{ignatieva2021kwarg} and 
+\argweaver~\citep{rasmussen2014genome}, 
+and two which do not (\tsinfer\ and \relate). 
+All inferences are based on the
 same underlying dataset from \citet{kreitman1983nucleotide}, a standard benchmark in the classical
@@
 \emph{D.~melanogaster} ADH locus on chromosome 2L. Where required for inference purposes we
-assume a mutation rate of \SI{5.49e-9}{site^{-1}generation^{-1}} and a recombination rate of
-\SI{8.4e-9}{site^{-1}generation^{-1}} [CITE stdpopsim] and a constant effective population size of 1000,000.
+assume a mutation rate of 
+% per site per generation may be easier to follow?
+\SI{5.49e-9}{site^{-1}generation^{-1}} and a recombination rate of
+\SI{8.4e-9}{site^{-1}generation^{-1}} [CITE stdpopsim] and a constant effective population size of 
+1,000,000.
 The version numbers of inference software, together with the specific inference parameters
@@
 
-Note that both \kwarg and \argweaver produce many ARGs
+Note that both \kwarg\ and \argweaver\ produce many ARGs
 (those shown in Fig.~\ref{fig-inferred-args} A\& B are arbitrarily selected examples).
-This stems from the fact that \kwarg searches the space of plausible ARGs,
-outputting minimal ones using heuristics, and ARGweaver samples from the posterior distribution over
+This stems from the fact that \kwarg\ searches the space of plausible ARGs,
+outputting minimal ones using heuristics, and \argweaver\ samples 
+from the posterior distribution over
 ARGs given the data and a fixed model (the SMC$\prime$, an approximation of the full coalescent
-with recombination). In contrast, \tsinfer and \relate (Fig.~\ref{fig-inferred-args} C \& D) are
+with recombination) [TODO CITE]. 
+In contrast, \tsinfer\ and \relate\ (Fig.~\ref{fig-inferred-args} C \& D) are
 deterministic methods, aiming to find a single ``best guess'' ARG.
@@
 Indeed, there is considerable variation in the number of recombinations inferred by
-different methods, with e.g.~\argweaver (Fig.~\ref{fig-inferred-args} B) suggesting
+different methods, with e.g.~\argweaver\ (Fig.~\ref{fig-inferred-args} B) suggesting
 far more than are required under minimal parsimony assumptions (Fig.~\ref{fig-inferred-args} A).
@@
 because those nodes frequently share breakpoint positions. The sharing of breakpoint locations
-between multiple-parent nodes indicates that in \tsinfer and \relate ARGs it may be
+between multiple-parent nodes indicates that in \tsinfer\ and \relate\ ARGs it may be
 difficult to condense each transition between trees into a set of simple SPR moves.
@@
 information about shared ancestry. Note however, that such unary segments are not inferred
-in the \relate ARG: other than the samples it has entirely solid nodes.
-Only the \tsinfer results have nodes with more than two children (polytomies);
+in the \relate\ ARG: other than the samples it has entirely solid nodes.
+Only the \tsinfer\ results have nodes with more than two children (polytomies);
 in Appendix \ref{sec-cell-lineages-and-args} we argue that this reflects
@@
 \noderef{Fr-F}, \noderef{Wa-F}, and \noderef{Af-F}, which happen to
-have identical sequences. \kwarg, \argweaver, and \relate require
+have identical sequences. \kwarg, \argweaver, and \relate\ require
 bifurcating trees, so are forced
@@
 (hence Fig.~\ref{fig-inferred-args} A and Fig.~\ref{fig-inferred-args} B disagree on the
-relationship between these samples). TsThe number of nodes in Fig.~\ref{fig-inferred-args} C
+relationship between these samples). 
+The number of nodes in Fig.~\ref{fig-inferred-args} C
 that have more than two children also hints at considerable uncertainty in

### 2023-10-20 paper.tex
Yans updates

@@
 \includegraphics[width=\textwidth]{illustrations/inference.pdf} \end{center}
-\caption{\label{fig-inferred-args} Inference of an 11 sample ARG over a 2.4kb
-region of the \textit{Drosophila melanogaster} ADH locus~\citep{kreitman1983nucleotide}.
+\caption{\label{fig-inferred-args} Inference of sample-resolved ARGs for 11
+\textit{Drosophila melanogaster} genetic sequences over a 2.4kb
+region of the ADH locus~\citep{kreitman1983nucleotide}.
 Results for four different methods based on
-(A) parsimony, (B) sampling the SMC$\prime$, (C) copying from
+(A) parsimony, (B) sampling the SMC$^\prime$, (C) copying from
 inferred ancestral haplotypes, and (D) correlated local tree estimation.
@@
 (lighter: older; darker: younger).
-Position on the X and Y axes for the graph visualization is arbitrary.
-Line width and node colour are as described in Fig.~\ref{fig-simplification}.
-The graphic below each ARG shows the genomic intervals for each graph edge in the
-corresponding ARGs: edges are horizontal lines, stacked in time order
-(edges with youngest children at the bottom), with vertical dashed lines indicating
-breakpoints between local trees.
+X and Y positions of graph nodes are arbitrary;
+line width and node colour are as described in Fig.~\ref{fig-simplification}.
+Graphics on the bottom row show the genomic positions, relative to the start of the ADH gene,
+for each graph edge from the corresponding ARG. Edge intervals are drawn as horizontal lines,
+stacked in time order (edges with youngest children at the bottom);
+vertical dashed lines indicate breakpoints between local trees.
 }
@@
 construct not an ARG but ``only'' a sequence of local 
-trees [TODO cite examples from \#38]. While this statement
+trees \citep{hejase2020summary, ignatieva2021kwarg}. While this statement
 is correct inasmuch as neither method attempts to infer the explicit recombination events that cause
@@
 % per site per generation may be easier to follow?
-\SI{5.49e-9}{site^{-1}generation^{-1}} and a recombination rate of
-\SI{8.4e-9}{site^{-1}generation^{-1}} [CITE stdpopsim] and a constant effective population size of 
-1,000,000.
+\SI{5.49e-9}{per site per generation} and a recombination rate of
+\SI{8.4e-9}{per site per generation} \citep{lauterbur2022expanding}, together with
+a constant effective population size of 1,000,000.
 The version numbers of inference software, together with the specific inference parameters
@@
 from the posterior distribution over
-ARGs given the data and a fixed model (the SMC$\prime$, an approximation of the full coalescent
-with recombination) [TODO CITE]. 
+ARGs given the data and a fixed model (the SMC$^\prime$, an approximation of the full coalescent
+with recombination \citep{marjoram2006fast}).
 In contrast, \tsinfer\ and \relate\ (Fig.~\ref{fig-inferred-args} C \& D) are

### 2023-10-20 paper.tex
Yans updates

@@
 from the graph.
-The ``grand MRCA'' \noderef{q} is omitted from the
+The ``grand MRCA'' \noderef{q} is normally omitted from the
 sample-resolved gARG because all segments of the genome have
@@
 and more complex situations arising from simplification (see below).
-Thirdly, the shading intensity of a node denotes the
+Thirdly, the shading intensity of a node denotes the ``coalescent span'', the
 fraction of the node's span (the length of genome in which it
 is reachable from the samples in the local trees)
-over which it is coalescent (has more than one child).
+over which it is coalescent (has more than one child). Nodes which
+are never locally unary therefore have a coalescent span of 100\%,
+whereas nodes in which ancestral material never coalesces have a
+coalescent span of 0\% (such nodes will be extremely difficult to
+identify from the distribution of shared mutations in the samples).
 % This is useful why? I can't come up with a quick way of saying
@@
 line width and node colour are as described in Fig.~\ref{fig-simplification}.
-Graphics on the bottom row show the genomic positions, relative to the start of the ADH gene,
+Graphics on the bottom row show the genome positions, relative to the start of the ADH gene,
 for each graph edge from the corresponding ARG. Edge intervals are drawn as horizontal lines,
@@
 To illustrate the varying levels of information that can be captured in a gARG,
-Fig.~\ref{fig-inferred-args} presents graph representations of genealogies output by two tools that
+Fig.~\ref{fig-inferred-args} presents graphical representations of genealogies output by two tools that
 \emph{do} explicitly infer recombination events, namely \kwarg~\citep{ignatieva2021kwarg} and 
@@
 a constant effective population size of 1,000,000.
-The version numbers of inference software, together with the specific inference parameters
-needed to reproduce the ARGs, are available from our GitHub repository.
+To focus on the ARG topology rather than estimates of node times, the Y position of each node
+is arbitrary, taken from the ranked order of nodes output by each software method.
+The version numbers of inference software, the specific inference parameters
+needed to reproduce the ARGs, and (where necessary) functions to convert outputs to the
+\texttt{tskit} gARG format, are available from our GitHub repository.
 
@@
 from the posterior distribution over
-ARGs given the data and a fixed model (the SMC$^\prime$, an approximation of the full coalescent
-with recombination \citep{marjoram2006fast}).
+ARGs given the data and a fixed model (the SMC$^\prime$ \citep{marjoram2006fast},
+an approximation of the full coalescent with recombination).
 In contrast, \tsinfer\ and \relate\ (Fig.~\ref{fig-inferred-args} C \& D) are
@@
 
+% Here we describe recombination features
 A rough feel for the amount of recombination in each ARG is given by the distribution
@@
 than two parents, indicating deliberate imprecision: these are cases where multiple recombination events
-have been ``pushed down'' onto a single more recent node. Note, however, that the number
+have been ``pushed down'' onto a single more recent node.
+Note, however, that the number
 of breakpoints in these two ARGs is smaller than the number of ARG nodes with multiple parents,
@@
 
-Shading by coalescent span reveals that many non-recombinant (green) nodes are
-``locally unary'' (see section \ref{sec-locally-unary-edges}). In other words,
-in some regions of the genome, these nodes are present in the ancestry but do
-not correspond to coalescence. As argued previously, this can provide potentially useful
+% Here we move on to coalescence
+As for coalescence in these inferred ARGs, the shading (``coalescent span'')
+reveals that many non-recombinant (green) nodes are
+locally unary (see section \ref{sec-locally-unary-edges}). In other words,
+in some regions of the genome these nodes are present in the ancestry without
+corresponding to coalescence of ancestral material. As argued previously, this can provide useful
 information about shared ancestry. Note however, that such unary segments are not inferred
@@
 uncertainty in the order of coalescent events, in an equivalent
-manner to the lack of precision in recombination nodes with more
+manner to the imprecision associated with recombination nodes having more
 than two parents. The clearest example is the polytomy involving the nodes
@@
 the order of branching elsewhere in the ARG.
-% TODO add explanation of the bottom row.
-% https://github.com/tskit-dev/what-is-an-arg-paper/issues/386
+
+%% possible TODO - mention something about multiple local roots in
+% tsinfer and Relate versus the other two.
+
 %The example of successive simplification of a simulated ARG
@@
 
+% TODO add additional explanation of the bottom row, if enough isn't written below
+% https://github.com/tskit-dev/what-is-an-arg-paper/issues/386
 The bottom row of Fig~\ref{fig-inferred-args} shows the extent along
 the genome to which graph edges are shared between multiple trees.
-We can see, for example, that the \relate\ trees ``share'' less topology
-than other methods, but it certainly does not infer a series of
-\emph{unrelated} local trees
+We can see that both \tsinfer and (to a lesser extent) \relate have edges
+that span multiple tree boundaries, indicating that they certainly
+do not infer a series of \emph{unrelated} local trees.
 %(i.e., with either unlabelled internal
 %nodes or differently labelled nodes in each tree).
-The ARGs inferred by \relate\ and \tsinfer\ both contain substantial
-information about node sharing across trees; for example,
-both infer that Af-f, Fr-f and Wa-f share a common ancestor along
-the entire sequence.
-They contain a level of detail about recombination
+For example, both infer that \noderef{Af-f}, \noderef{Fr-f} and \noderef{Wa-f}
+share a common ancestor along
+the entire sequence. They do not, however, show extensive node sharing in the
+oldest time periods, reflecting the fact that they do not encode topology
+above the local tree roots (i.e. they do not go back to a
+"grand MRCA"). This is also seen in the fact they their ARGs contain multiple roots
+(nodes with zero parents, in blue), whereas \kwarg and \argweaver have a
+single ARG root.
+These two methods contain a level of detail about recombination
 that lies somewhere between a sequence of unrelated local trees on one

### 2023-10-20 paper.tex
Yans updates

@@
 % Event ARGs
-In the following, we define the genome ARG (gARG) and contrast it with the event ARGs (eARG).
+In the following, we define a genome ARG (gARG) and contrast it with event ARGs (eARGs).
 % Retrospective vs prospective
@@
 of number of parents per node, as highlighted by the node colour scheme.
-In Fig.~\ref{fig-inferred-args} A \& B, each recombination node has exactly 2 parents,
-corresponding to a single recombination event with a specific breakpoint. Assuming that each
-breakpoint occurs a a different location (the "infinite breakpoints" model), then the
+In Fig.~\ref{fig-inferred-args} A \& B, each recombination event is tagged by
+a node with exactly 2 parents and one child. Assuming that the breakpoint for each event
+occurs at a different location (the "infinite breakpoints" model), then the
 number of breakpoints should exactly equal the number of 2-parent (brown) nodes in the plot.
-In contrast, Figure Fig.~\ref{fig-inferred-args} C \& D have ARG nodes with greater
-than two parents, indicating deliberate imprecision: these are cases where multiple recombination events
-have been ``pushed down'' onto a single more recent node.
-Note, however, that the number
-of breakpoints in these two ARGs is smaller than the number of ARG nodes with multiple parents,
-because those nodes frequently share breakpoint positions. The sharing of breakpoint locations
-between multiple-parent nodes indicates that in \tsinfer\ and \relate\ ARGs it may be
+In contrast, Figure Fig.~\ref{fig-inferred-args} C \& D have ARG nodes with multiple
+parents and multiple children. These indicate deliberate imprecision, where one
+or more recombinations have been ``pushed down'' onto a more recent, identifiable node.
+Several of these nodes have more than two parents (in red), representing
+multiple recombinations inferred to have occurred in an unresolvable order.
+Nevertheless, the number of breakpoints in these two ARGs is smaller than the
+number of multiple-parent ARG nodes, showing that several multiple-parent nodes
+must share breakpoint positions. The sharing of breakpoint locations
+between these nodes indicates that in \tsinfer\ and \relate\ ARGs it may be
 difficult to condense each transition between trees into a set of simple SPR moves. 
+Overall, Fig.~\ref{fig-inferred-args} shows that \tsinfer\ and \relate\ ARGs contain a level
+of detail about recombination that lies somewhere between a sequence of unrelated local
+trees on one extreme and an ARG with precisely specified recombination events on the other.
 
 % Here we move on to coalescence
-As for coalescence in these inferred ARGs, the shading (``coalescent span'')
+As for inferred coalescences, the shading (``coalescent span'')
 reveals that many non-recombinant (green) nodes are
@@
 in the \relate\ ARG: other than the samples it has entirely solid nodes.
-Only the \tsinfer\ results have nodes with more than two children (polytomies);
+Another clear difference involves coalescent nodes with more than two children (polytomies);
 in Appendix \ref{sec-cell-lineages-and-args} we argue that this reflects
@@
 manner to the imprecision associated with recombination nodes having more
-than two parents. The clearest example is the polytomy involving the nodes
-\noderef{Fr-F}, \noderef{Wa-F}, and \noderef{Af-F}, which happen to
-have identical sequences. \kwarg, \argweaver, and \relate\ require
-bifurcating trees, so are forced
-to pick an arbitrary order of branching
+than two parents. Only \tsinfer\ creates such polytomies, the clearest of which
+involves the nodes \noderef{Fr-F}, \noderef{Wa-F}, and \noderef{Af-F}. These
+happen to have identical sequences, but because \kwarg, \argweaver, and \relate\ require
+bifurcating trees, they are forced to pick an arbitrary order of branching
 (hence Fig.~\ref{fig-inferred-args} A and Fig.~\ref{fig-inferred-args} B disagree on the
 relationship between these samples). 
-The number of nodes in Fig.~\ref{fig-inferred-args} C
-that have more than two children also hints at considerable uncertainty in
+More generally, the number of nodes in Fig.~\ref{fig-inferred-args} C
+that have more than two children hints at considerable uncertainty in
 the order of branching elsewhere in the ARG.
@@
 the genome to which graph edges are shared between multiple trees.
-We can see that both \tsinfer and (to a lesser extent) \relate have edges
-that span multiple tree boundaries, indicating that they certainly
-do not infer a series of \emph{unrelated} local trees.
+This illustrates that both \tsinfer\ and (to a lesser extent) \relate\ have edges
+that span multiple tree boundaries, indicating that they are not inferring
+a series of \emph{unrelated} local trees.
 %(i.e., with either unlabelled internal
@@
 share a common ancestor along
-the entire sequence. They do not, however, show extensive node sharing in the
-oldest time periods, reflecting the fact that they do not encode topology
-above the local tree roots (i.e. they do not go back to a
-"grand MRCA"). This is also seen in the fact they their ARGs contain multiple roots
-(nodes with zero parents, in blue), whereas \kwarg and \argweaver have a
-single ARG root.
-These two methods contain a level of detail about recombination
-that lies somewhere between a sequence of unrelated local trees on one
-extreme and a completely precise event ARG on the other.
+the entire sequence. However, in comparison to \kwarg\ and \argweaver\ neither
+method results in extensive node sharing in the
+oldest time periods. This primarily reflects the fact that they do not encode topology
+above the local MRCAs, also indicated by multiple roots in the ARG
+(nodes with zero parents, in blue), compared to the
+single ARG root in the \kwarg\ and \argweaver\ genealogies. In terms of inference,
+\tsinfer\ and \relate\ show a more pragmatic representation: we do not expect
+genetic variation to be very informative about topological relationships
+between nodes that that are older than the local MRCAs.
 

### 2023-10-20 paper.tex
Drop some changes that will be nuked

@@
 \label{sec-intro}
+% Para 1: ARGs are cool!
 Estimating the genetic genealogy of a set of genome sequences
@@
 follow~\citep{harris2019database,harris2023using}.
-% TODO finish this sentence with "and more reasons why ARGs are cool"
-See \citet{lewanski2023era} for a biologically oriented introduction to ARGs.
-
-% TODO need to mention ARG-as-a-data-structure to clarify.
-% What is the problem?
+See \citet{lewanski2023era} for a biologically oriented introduction to ARGs
+and further discussion on the potential impacts.
+
+% Para 2: What are they, though?
 Although it is widely accepted that ARGs are important, there is significant
@@
 structure~\citep{minichiello2006mapping}, i.e.\ describing
-the outcome of such a stochastic process or a
-genetic genealogy inferred from a sample of genome sequences.
-However, the distinction between process and data structure is not clear cut
+the \emph{outcome} of such a random process or an
+% Note on earlier version: introducing a different term for ARG here is just confusing
+inference from a sample of genome sequences.
+The distinction between stochastic process
+and data structure is not clear cut, however,
 and subfields use the term
 differently (see Appendix~\ref{sec-arg-history}).
-
-% JK: this para need a major update, it's quite weak right now.
-The word ``ARG'' is now widely used to refer to genetic genealogies in
-general~\citep[e.g.][]{mathieson2020ancestry,hejase2020summary,
+% pedantically, ARG is not really a "word"
+The term ``ARG'' is now most often used in quite a general
+sense~\citep[e.g.][]{mathieson2020ancestry,hejase2020summary,
 schaefer2021ancestral,harris2023using,zhang2023biobank,fan2023likelihood},
+as a data structure
 encompassing the varied outputs from modern simulation and
@@
 kelleher2019inferring, speidel2019method, baumdicker2021efficient, zhang2023biobank}.
-This broad usage means there is a lack of consensus on the
-exact object being defined, which causes problems, for example,
-when evaluating inferred ARGs, and even when defining
-what the criteria for success should be.
-It also means there is no shared format for interchanging ARG data between
-programs, resulting in the assorted negative consequences with which
-geneticists are only too familiar~\citep{excoffier2006computer,gower2022demes}.
-Perhaps most importantly, definitions linked to
-the original stochastic process lead to an
-overly narrow interpretation of ARGs, which is significantly
-out of step with modern methods and large-scale datasets, and
-can lead to missed opportunities in their application and usage.
-Such problems go much deeper than pedantic discussion over terminology;
-they directly inform what we should be attempting to estimate, how we
-assess success, and the expected limitations on the precision
+This broad usage seems a useful generalisation,
+but is not compatible with the usual concrete definition of an ARG.
+This mismatch has led to a range of problems that, if unchecked,
+may impede progress in applying ARGs to understand the biological world.
+
+% Para 3: what is the problem?
+% FIXME current version isn't great, but it's better than previous effort.
+[FIXME this paragrph needs a substantial update, need to reread the
+eARG and gARG sections again to be clear what we say we're doing.
+We should focus on the concrete interchange aspects rather than
+coalescenty type things.]
+The mismatch between the classical Griffiths definition of an ARG and the
+modern interpretation is the source of substantial confusion,
+and, we argue, a significant limiting factor for the field.
+% actual outputs of these modern methods is a central issue.
+As this definition is essentially a reformulation of the
+coalescent with recombination (henceforth ``the coalescent''; we
+will assume the incorporation of recombination for the purposes of
+this article) in graph-theoretic form,
+the distinction between the simplifying assumptions of the model
+and the necessary requirements for a data structure are not clear.
+The consequences of conflating model with data structure
+are subtle, varied, and deep.
+% TODO rephrase, not great currentyl
+We argue that the prevailing definition of an ARG
+imposes unneccesary limitations, and that with a broader definition
+we can reassess what we should be attempting to estimate,
+how we measure success, and the expected limitations on the precision
 and accuracy of our inferences.
-% Thanks for the phrase Gideon and Alex!
-If not addressed, these problems will
-impede progress in applying genetic genealogies to understand the biological world.
-
-% Evidence of confusion to cite above in "problem" para?
-% https://github.com/tskit-dev/what-is-an-arg-paper/issues/334
-% The output of these programs is certainly not a collection of
-% \emph{unrelated} local trees, as implied by,
-% e.g.,
-% \citet{hejase2020summary} when they state that \tsinfer\
-% ``does not explicitly infer an ARG but rather a sequence of local
-% gene trees''.
-% More cites? Not worth harping on about now?
-% https://github.com/tskit-dev/what-is-an-arg-paper/issues/38
 
@@
 % Event ARGs
-In the following, we define a genome ARG (gARG) and contrast it with event ARGs (eARGs).
-% Retrospective vs prospective
-% TODO: This needs revision - see also https://github.com/tskit-dev/what-is-an-arg-paper/pull/380
-We continue with a description of the usually held retrospective view of ARGs,
-but note that ARGs can also be viewed as prospective, particularly in forwards-time
-simulations. In the context of forwards-time simulations, we mention the
-simplification operation, which removes ancestral genome material that was not
-inherited.
+In the following, we first define the genome ARG encoding
+and contrast it with the event ARG approach.
+We then discuss the important concept of ancestral material in
+Section~\ref{sec-ancestral-material}
+and [DO SOME STUFF].
 % Converting and eARG to gARG &
@@
 more common sense of ``the genome'' of a species, denoting the full set
-of chromosomes and their coordinate spaces, and hope that the distinction
+of chromosomes, and hope that the distinction
 will be clear from the context.
@@
 following the standard practice in coalescent theory.
+We also do not consider structural variation, and assume that all
+samples and ancestors share the same genome coordinate space.
 
@@
 Let $N \subset \mathbb{N}$ be the set of nodes representing
-the genomes in the gARG, and $E$ be the set of edges.
-We will usually define a set of nodes $S \subseteq N$
-representing the sampled genomes
-(but see Section~\ref{sec-retro-pro} for some subtlety around this point).
-Each element of $E$ is a tuple $(c, p, I)$ such that $c, p \in N$ are the child and
+the genomes in the gARG, $S \subseteq N$ be the set of sampled genomes.
+% (but see Section~\ref{sec-retro-pro} for some subtlety around this point).
+Then, $E$ is the set of edges, where each element
+is a tuple $(c, p, I)$ such that $c, p \in N$ are the child and
 parent nodes and $I$ is the set of disjoint genomic intervals
@@
 Thus, each topological connection between
-a parent and child node in the graph is ``annotated'' with a set of
+a parent and child node in the graph is annotated with a set of
 inheritance intervals $I$.
 Here, the terms parent and child are used in graph sense;
-these nodes can respectively represent ancestor and descendant genomes,
+these nodes respectively represent ancestor and descendant genomes,
 which can be separated by multiple generations.
@@
 
-% TODO add a few sentences here on why defining nodes concretely as
-% integers has significant advantages.
-% 1) leads to direct link between the ARG and local trees in terms
-%    of a well-defined combinatorial object (local trees section)
-% 2) Definitions can be implemented directly in an efficient way
-%    (computational efficiency section.
-In many settings, nodes are dated, i.e.\
-each node $u\in N$ is associated with a time $\tau_u$.
-However, the topological relationships
-defined by the directed graph structure is sufficient for
-many applications, and we will assume that node dates are not
-known here.
+In many settings, nodes are dated, i.e.\ each
+node $u\in N$ is associated with a time $\tau_u$.
+The topological relationships
+defined by the directed graph structure, however,
+ is sufficient for many applications,
+and we will assume that node dates are not known here.
 % This should be obvious, but seems like it needs to be said
@@
 with which arbitrary additional information can be associated.
+Concretely defining nodes as integers in this way has some
+significant advantages. In particular, it provides a direct link
+between an ARG and its local trees in terms of a well-defined
+combinatorial object (Section~\ref{sec-ARG-and-local-trees}),
+as well as enabling straightforward connection between
+definitions and practical implementation (Section~\ref{sec-efficiency}).
 
@@
 the gARG for a given set of individuals is embedded in their pedigree.
-The pedigree relationship between the diploid individuals is
+The pedigree relationships between the diploid individuals is
 shown, along with their paths
@@
 The figure shows the annotated pedigree with realised inheritance of genomes
-between generations, the corresponding gARG, and finally the corresponding
-sequence of local trees along the genome.
-The local trees represent the genealogy of genomes \noderef{a}--\noderef{d},
-which changes between local trees due to recombination.
+between generations (A), the corresponding gARG (B), and finally the corresponding
+sequence of local trees  along the
+genome (C).
+% Jerome: this feels unecessary and "genealogy" is often interpreted as
+% "pedigree" so needs to be used very carefully.
+% The local trees represent the genealogy of genomes \noderef{a}--\noderef{d},
+% which changes between local trees due to recombination.
 The local trees span the three genome regions delineated
 by the two recombination breakpoints that gave rise to these genomes.
-Following the above definitions, corresponding data encoding for
-this gARG is shown in Tab.~\ref{tab-gARG-data}.
-
-These definitions follow naturally from the outcome of events
-associated with the inheritance of DNA between generations.
-% FIXME this sentance doesn't follow well from the first, needs a
-% different bridge before listing important properties.
-% Maybe the "listicle" format here isn't right anyway.
-As such,
-there are some important consequences to these definitions.
-Firstly, there is
-no limitation on a mating system, ploidy, or age structure in a gARG.
-By making the
-basic unit an individual's genomes,
-any pattern of inheritance among mixed ploidy (e.g.\ haplodiploid
-species) can be accommodated.
-Secondly, because we record the intervals of inheritance
-between genomes,
-arbitrary patterns of genetic inheritance can be directly expressed, such as
-gene conversions,
-multiple simultaneous recombinations,
-and so on.
-% Wilder: give forward ref to section on efficiency here, mention tree sequence
-% TODO, should probably drop this point.
-Thirdly, local trees that we generate for a particular position on the
-genome can have nodes of any arity (number of child nodes), including ``unary''
-nodes with a single child (see Section~\ref{sec-locally-unary-edges}).
-% TODO should rephrase this from parent/child to ancestor/descendant?
-Fourth, recombination is modelled in terms of
-its inputs (parental genomes) and output (child genome) with associated inheritance intervals.
-Thus we will have a child node (the recombinant; e.g.\ node \noderef{a}
-Fig.~\ref{fig-arg-in-pedigree}) and two (or more; see Section~\ref{sec-ARG-simplification})
-% TODO: check above if we want to refer to sec-ARG-simplification or some other!
-parental nodes from which this child has inherited genome segments.
-% TODO rephrase the above, it'll confuse people. Just need to reiterate the
-% points about events, and maybe lead into the next section
-Finally, gARGs are not defined in terms of the
-coalescent with recombination (or any other) stochastic process.
-Thus, the gARG encoding is focused on faithfully representing the
-\emph{outcome} of the processes of genetic inheritance
-with recombination,
-and is agnostic to the details of those processes.
+See Section~\ref{sec-ARG-and-local-trees} for further details on
+how the local trees are embedded in an ARG.
+% Following the above definitions, corresponding data encoding for
+% this gARG is shown in Tab.~\ref{tab-gARG-data}.
+
+% These definitions follow naturally from the outcome of events
+% associated with the inheritance of DNA between generations.
+% % FIXME this sentance doesn't follow well from the first, needs a
+% % different bridge before listing important properties.
+% % Maybe the "listicle" format here isn't right anyway.
+% As such,
+% there are some important consequences to these definitions.
+% Firstly, there is
+% no limitation on a mating system, ploidy, or age structure in a gARG.
+% By making the
+% basic unit an individual's genomes,
+% any pattern of inheritance among mixed ploidy (e.g.\ haplodiploid
+% species) can be accommodated.
+% Secondly, because we record the intervals of inheritance
+% between genomes,
+% arbitrary patterns of genetic inheritance can be directly expressed, such as
+% gene conversions,
+% multiple simultaneous recombinations,
+% and so on.
+% % Wilder: give forward ref to section on efficiency here, mention tree sequence
+% % TODO, should probably drop this point.
+% Thirdly, local trees that we generate for a particular position on the
+% genome can have nodes of any arity (number of child nodes), including ``unary''
+% nodes with a single child (see Section~\ref{sec-locally-unary-edges}).
+% % TODO should rephrase this from parent/child to ancestor/descendant?
+% Fourth, recombination is modelled in terms of
+% its inputs (parental genomes) and output (child genome) with associated inheritance intervals.
+% Thus we will have a child node (the recombinant; e.g.\ node \noderef{a}
+% Fig.~\ref{fig-arg-in-pedigree}) and two (or more; see Section~\ref{sec-ARG-simplification})
+% % TODO: check above if we want to refer to sec-ARG-simplification or some other!
+% parental nodes from which this child has inherited genome segments.
+% % TODO rephrase the above, it'll confuse people. Just need to reiterate the
+% % points about events, and maybe lead into the next section
+% Finally, gARGs are not defined in terms of the
+% coalescent with recombination (or any other) stochastic process.
+% Thus, the gARG encoding is focused on faithfully representing the
+% \emph{outcome} of the processes of genetic inheritance
+% with recombination,
+% and is agnostic to the details of those processes.
 
@@
 \label{sec-eARG}
-% FIXME - this intro doesn't work at all now - we haven't defined what an
-% eARG *is* before discussing it.
-% Maybe best to skip this entirely, and make the quoted definitions with
-% a thin intro the way to start the section.
-Genome ARGs are defined by
-describing the details of genetic inheritance
-between ancestor and descendant genomes.
-In contrast, ARGs are classically defined
-by describing the \emph{events} in the history of a sample of genomes.
-The difference between gARGs and the prevalent views of eARGs
-% what defines an ARG
-are subtle and difficult to distinguish.
-It is therefore useful to first survey some representative
-recent definitions of
-an ARG from the literature, to allow us to capture their common features
-(see Appendix~\ref{sec-arg-history} for a full review).
-
-\citet{brandt2021evaluation} follow the original Griffiths definition of
-an ARG by coupling it directly to a generative process, stating
-that the ``full ancestral recombination graph (ARG) is a structure that encodes all
-coalescence and recombination events resulting from the stochastic process of
-the coalescent with recombination.''
-\citet{shipilina2023origin} emphasise the true existence of the
-corresponding events (rather than the generative process) by saying that
-``[t]he ARG describes the complete
-ancestry of a sample of genomes through a series of real coalescence and
-recombination events.''
-\cite{mathieson2020ancestry}
-define an ARG as ``a subset of the pedigree...
-contain[ing] only those edges along which inherited segments of DNA have been
-transmitted, and only those nodes corresponding to ancestors in which there was
-a recombination or coalescence event.''
-% In contrast, ~\citet{gusfield2014recombinatorics} defines an ARG as [TODO]
-% Papers that do not discuss how positions work:
-% - Mathieson and Scally: no discussion of how it works concretely (but quite
-%   close to gARG, in some ways)
-% TODO list out some of these as examples of where people don't bother
-% getting into this important detail?
-
-% NOTE: important to emphasise TWO things here, events and vagueness
-% actual encoding mechanism (i.e. how trees fit into the ARG)
-A common feature of these definitions is that they are expressed
-in terms of evolutionary \emph{events}, and so for the purposes of
-discussion we can label
-this prevailing interpretation as an ``event ARG'', or eARG.
-(Many definitions also emphasise
-the necessity of exhaustively capturing \emph{all} events, which we
-examine in later sections.)
-Another common feature of these definitions is that they do not explicitly
-discuss the mechanism by which local trees vary along the genome.
-The topology of an eARG defines the common ancestor and
-recombination events that occurred,
-but the graph structure alone is not sufficient to define the local trees,
-or equivalently, how ``ancestral material'' flows along
-the edges of the ARG (see Section~\ref{sec-eARG-to-gARG}).
-% TODO: check above if we want to refer to sec-eARG-to-gARG or some other!
-This is usually assumed to follow the mechanism described by
-Griffiths and colleagues in which we have
-two different ``types'' of node in the graph:
-common ancestor nodes in which the inbound lineages are merged into a
-single ancestral lineage with one parent, and recombination
-nodes in which a single lineage is split into two independent
-ancestral lineages.
-Recombination nodes are then
-annotated with
-the corresponding crossover breakpoint~\citep{griffiths1996ancestral}.
-Fig.~\ref{fig-event-arg} shows an example of a classical
-eARG with three sample genomes (\noderef{a}, \noderef{b}, and \noderef{c})
-and a single recombination event,
-% Trying to emphasise here that this is just an example data encoding.
-% I can imagine
-% people nitpicking about how this specific way of doing it is bad
-% for reason X.
-along with an example encoding of the data.
-There is a single recombination event
-at node \noderef{d} with a breakpoint at position $x$. We
-assume that \noderef{d} inherits genetic material to the
-left of $x$ from \noderef{e} and to the right of $x$ from \noderef{f}.
-As shown in Fig.~\ref{fig-event-arg}B,
-the local trees are embedded in the graph, and can be reconstructed
-in essentially the same manner as outlined in the previous section for
-gARGs. The only difference is that decisions about which outbound
-edge to follow during the rootward traversal are taken at recombination
-nodes, following
-either the common convention that the order of parents
-is significant~\citep[e.g.][]{griffiths1991two},
-or by incorporating additional information to
-distinguish left and right parents
-into the
-encoding~\citep[e.g.][]{gusfield2014recombinatorics,ignatieva2021kwarg}.
-% JK: not sure what you're getting at here Gregor? The ordering matters
-% because you have to write each event down precisely.
-% Gregor: Importantly, the order of events matters to WHAT!?
+% % I feel like sophisticated readers would at this point be saying,
+% % ah, yeah, that's basically what an ARG is just in different words;
+% % what's the point of this paper?
+The genome ARG framework defined in the previous section is
+in many ways a clarification of existing treatments
+\cite[e.g.][]{mathieson2020ancestry,shipilina2023origin},
+adding concrete details to the mechanisms by which
+differential inheritance of genetic material are modelled.
+% This is definitely needed for the Nicks, who are always thinking
+% about the deeper processes and are quite happy being loosey-goosey
+% with representation details
+Note that here, and throughout, we are not questioning the form of the
+actual ancestral processes that occur in nature, but rather how we
+\emph{represent} such realised histories in a practical sense.
+% Again, need to remind the reader that there's an actual point
+% to all this getting stuck into the details.
+These practical details,
+as demonstrated through the remainder of the article, have important
+consequences not only for how methods exchange information about
+simulated and inferred ARGs, but more fundamentally in
+how we set our goals for inference and evaluate the success of results.
+
+% section is in many ways simply a clarification of existing definitions,
+% it does provide substantial advantages (as shown in later sections).
+% To allow us to concretely discuss and characterise the prevailing
+% definition of an ARG, we begin by surveying some
+% recent definitions from the literature.
+% \citet{brandt2021evaluation} follow the original Griffiths definition of
+% an ARG by coupling it directly to a generative process, stating
+% that the ``full ancestral recombination graph (ARG) is a structure that encodes all
+% coalescence and recombination events resulting from the stochastic process of
+% the coalescent with recombination.''
+% \citet{shipilina2023origin} emphasise the true existence of the
+% corresponding events (rather than the generative process) by saying that
+% ``[t]he ARG describes the complete
+% ancestry of a sample of genomes through a series of real coalescence and
+% recombination events.''
+% \cite{mathieson2020ancestry}
+% define an ARG as ``a subset of the pedigree...
+% contain[ing] only those edges along which inherited segments of DNA have been
+% transmitted, and only those nodes corresponding to ancestors in which there was
+% a recombination or coalescence event.''
+
+% The prevailing view, captured in these representative definitions, is that an
+% ARG is composed of \emph{events}, and so we will loosely label this approach as
+% ``event ARGs'', or ``eARGs''. Whether we view nodes in an ARG as events or
+% genomes is essentially a philosophical question and in most senses the two are
+% equivalent (see Appendix~\ref{sec-cell-lineages-and-args}, however, for a discussion
+% of some subtle issues around \emph{which} genome corresponds to events
+% of various types).
+% How we represent and reason about the differential inheritance of
+% genetic matierial \emph{between} nodes in an ARG, however, is of fundamental
+% importance, and the central theme of this article.
+% % Need to make this point strongly for the Nicks. Not sure this is the
+% % right place, though. Maybe it's good to remind them again here though...
+% Note that here, and throughout, we are not questioning the form of the
+% actual ancestral processes that occur in nature, but rather how we
+% \emph{represent} such realised histories in a practical sense.
 
@@
 \node [anchor=north west] at (\x2,6) {B};
-\node [anchor=north west] at (\xt,6) {C};
+% \node [anchor=north west] at (\xt,6) {C};
 
@@
 
-%%% (C) Encoding
-\node [nodelabel,anchor=north west] at ($(\xt,5)$) {
-\begin{tabular}{c|c|l}
-% \multicolumn{2}{c}{Breakpoints}\\
-Node & Breakpoint & Parents\\
-\hline
-$\noderef{a}$ & $\varnothing$ & [\noderef{e}]\\
-$\noderef{b}$ & $\varnothing$ & [\noderef{d}] \\
-$\noderef{c}$ & $\varnothing$ & [\noderef{f}]\\
-$\noderef{d}$ & $x$ & [\noderef{e}, \noderef{f}] \\
-$\noderef{e}$ & $\varnothing$ & [\noderef{g}]\\
-$\noderef{f}$ & $\varnothing$ & [\noderef{g}]\\
-$\noderef{g}$ & $\varnothing$ & []\\
-\end{tabular}};
+% %%% (C) Encoding
+% \node [nodelabel,anchor=north west] at ($(\xt,5)$) {
+% \begin{tabular}{c|c|l}
+% % \multicolumn{2}{c}{Breakpoints}\\
+% Node & Breakpoint & Parents\\
+% \hline
+% $\noderef{a}$ & $\varnothing$ & [\noderef{e}]\\
+% $\noderef{b}$ & $\varnothing$ & [\noderef{d}] \\
+% $\noderef{c}$ & $\varnothing$ & [\noderef{f}]\\
+% $\noderef{d}$ & $x$ & [\noderef{e}, \noderef{f}] \\
+% $\noderef{e}$ & $\varnothing$ & [\noderef{g}]\\
+% $\noderef{f}$ & $\varnothing$ & [\noderef{g}]\\
+% $\noderef{g}$ & $\varnothing$ & []\\
+% \end{tabular}};
 
@@
 for a discussion of this important point).
-(C) A concrete encoding of the information.
-Recombination nodes have a non-null breakpoint
-and two parents.
 }
 \end{figure}
+
+% The prevailing view is that ARGs are composed of events, and so we
+% characterise this view
+% Besides viewing ARGs as being composed of events, another
+% feature of the definitions given above is that they do not
+% explicitly describe
+% % TODO check this! Do they talk about how trees are embedded?
+% the details of how such events are represented.
+Where the concrete details of an ARG data structure
+(the encoding) are
+% TODO more?
+specified~\citep[e.g.][]{
+rasmussen2014genome,gusfield2014recombinatorics,ignatieva2021kwarg}
+they follow the approach described by Griffiths and colleagues.
+Because it is focused on recording the details of events,
+we will refer to this Griffiths encoding as an ``Event ARG'', or eARG.
+In an eARG, we have two different types of node in the graph
+representing the common ancestor and recombination events
+in the history of a sample.
+At common ancestor nodes, the inbound lineages merge into a
+single ancestral lineage with one parent, and at recombination
+nodes a single lineage is split into two independent
+ancestral lineages. Recombination nodes are annotated with
+the corresponding crossover breakpoint, and these breakpoints
+are used to reconstruct the local trees.
+This is done by tracing rootwards through the graph from the samples,
+taking decisions about which outbound edge to follow through
+recombination nodes based on the breakpoint
+position~\citep{griffiths1996ancestral}.
+Fig.~\ref{fig-event-arg} shows an example of a classical
+eARG with three sample genomes (\noderef{a}, \noderef{b}, and \noderef{c})
+and a single recombination event (node \noderef{d}) with a breakpoint
+at position $x$.
+We assume here that \noderef{d} inherits genetic material to the
+left of $x$ from \noderef{e} and to the right of $x$ from \noderef{f}.
+
+While the Griffiths approach of annotating recombination nodes
+with a breakpoint in an eARG is a concise and elegant way of describing
+realisations of the coalescent, it has some significant
+issues when applied to the complexities of present-day datasets.
+Because we are explicitly modelling the \emph{properties} of events,
+the encoding must be extended to incorporate
+different types of event that are not simple crossover recombination
+or common ancestry.
+Processes such as gene conversion, for example, can in principle
+be included, but either require additional node types
+or (e.g.) multiple recombination nodes joined by a zero-length edge.
+Modelling multiple recombinant offspring of the same parents
+(while keeping track of their identity as siblings)
+as occurs
+in Wright-Fisher models (e.g.\ Fig.~\ref{fig-arg-in-pedigree}),
+or densely sampled multi-generational
+data~\citep[e.g.][]{turnbull2018hundred,hayes20191000,RosFreixedes2022}
+% circumlocutions isn't technically correct, but I like the roundabout
+% overly verbose feel it gives.
+would require more involved circumlocutions.
+These and other problems can be worked around
+(although, we would ideally like everyone to use the
+\emph{same} workarounds if methods are to interoperate; see the Discussion),
+but it is much simpler to use the gARG encoding described
+in the previous section.
+By recording the \emph{outcome} of events in terms of the
+intervals of genetic inheritance on edges instead of recording the
+properties of the events themselves, we remove the need to explicitly
+model different types of event in the data structure.
+More fundamentally, because gARGs
+allow us to directly describe
+genetic inheritance in multiple non-contiguous genome intervals
+between an ancestor and a descendant,
+we can encode the outcome of
+\emph{multiple} genetic inheritance events, and to
+systematically describe the uncertainty around their
+ordering (see Section~\ref{sec-precision}).
+
+There is also a certain clarity gained by explicitly modeling nodes
+in the inheritance graph as an individual's genomes.
+Outside of the context of a
+% The point being, events are perfectly well defined in the models
+% but not in reality
+mathematical model, an ``event'' is a slippery concept.
+For example, \emph{which} genome along a chain of cell divisions should be
+regarded as the one where an event occured
+(Appendix~\ref{sec-cell-lineages-and-args}),
+or whether multiple recombinations
+within a single individual should be regarded as one or multiple events are
+debateable.
+% If we want to have an ARG software ecosystem then such fireside
+% discussions won't help.
+From the perspective of a concrete data structure,
+ideally forming the basis of an ecosystem of interoperable
+inference and analysis methods, such debates
+are unproductive.
+
+% By regarding nodes as an individual's genomes and
+% explicitly annotating edges with the intervals of genetic inheritance
+% between those genomes, we can capture the \emph{outcome} of a
+% range of different types of event without needing to explicitly model
+% the details of these events in the ARG data structure.
+% More fundamentally, because gARGs
+% allow us to directly describe
+% genetic inheritance in multiple non-contiguous genome intervals
+% between an ancestor and a descendant,
+% we can encode the outcome of
+% \emph{multiple} genetic inheritance events, and to
+% systematically describe the uncertainty around their
+% ordering (see Section~\ref{sec-precision}).
+
+% While the additional generality provided by the gARG representation
+% is useful, it is in reality a small change that does not
+% lead to significantly different time or space complexity in computational
+% tasks.
+% The vitally important and fundamental distinction between
+% a gARG and an eARG is that gARGs
+% allow us to directly describe
+% genetic inheritance in multiple non-contiguous genome intervals
+% between an ancestor and a descendant, which cannot be done
+% by specifying recombination events and breakpoints.
+% % FIXME this sentence is problematic. We should recast to talk about
+% % the outcomes of events, not talk about simulateousness (which will
+% % confuse people
+% Thus, we have the means to encode the \emph{outcome} of
+% multiple genetic inheritance events, and to
+% systematically describe the uncertainty around their
+% ordering (see Section~\ref{sec-precision}).
+% As we describe in the following sections, freeing ourselves from the requirement
+% of recording specific events and
+% the additional generality of the gARG encoding
+% has significant advantages. It also raises questions about what our goals
+% for inferring ARGs should be, and how we evaluate the success
+% of such inferences.
+
+% A common feature of these definitions is that they are expressed
+% in terms of evolutionary \emph{events}, and so for the purposes of
+% discussion we can label
+% this prevailing interpretation as an ``event ARG'', or eARG.
+% (Many definitions also emphasise
+% the necessity of exhaustively capturing \emph{all} events, which we
+% examine in later sections.)
+% Another common feature of these definitions is that they do not explicitly
+% discuss the mechanism by which local trees vary along the genome.
+% The topology of an eARG defines the common ancestor and
+% recombination events that occurred,
+% but the graph structure alone is not sufficient to define the local trees,
+% or equivalently, how ``ancestral material'' flows along
+% the edges of the ARG (see Section~\ref{sec-eARG-to-gARG}).
 
@@
 % individual or cell.
-Whether we regard nodes as events or genomes is something of a
-philosophical question, and in most senses the two are equivalent
-(but see the discussion on ``unary nodes'' in Section~\ref{sec-locally-unary-edges}).
-From a concrete data-structure encoding perspective, the only real
-difference between an eARG and a gARG is that in an eARG we store recombination
-\emph{breakpoints} at nodes of out-degree two (recombination nodes)
-whereas in a gARG we store \emph{inheritance intervals} on edges.
-A key advantage of associating inheritance intervals with
-edges in a gARG is that it allows us to describe genetic inheritance events other
-than simple crossovers, such as multiple recombinations and
-gene conversions. By explicitly stating the exact regions of genome
-that are transmitted along an edge, we generalise the representation
-to encompass any process that leads to the differential inheritance of
-genetic material along the genome.
-
-While the additional generality provided by the gARG representation
-is useful, it is in reality a small change that does not
-lead to significantly different time or space complexity in computational
-tasks.
-The vitally important and fundamental distinction between
-a gARG and an eARG is that gARGs
-allow us to directly describe
-genetic inheritance in multiple non-contiguous genome intervals
-between an ancestor and a descendant, which cannot be done
-by specifying recombination events and breakpoints.
-% FIXME this sentence is problematic. We should recast to talk about
-% the outcomes of events, not talk about simulateousness (which will
-% confuse people
-Thus, we have the means to encode the \emph{outcome} of
-multiple genetic inheritance events, and to
-systematically describe the uncertainty around their
-ordering (see Section~\ref{sec-precision}).
-As we describe in the following sections, freeing ourselves from the requirement
-of recording specific events and
-the additional generality of the gARG encoding
-has significant advantages. It also raises questions about what our goals
-for inferring ARGs should be, and how we evaluate the success
-of such inferences.
-
-\section{Retrospective vs prospective ARGs}
-\label{sec-retro-pro}
-The ARG concept arose from a \emph{retrospective} view of inheritance.
-For example the definition of an eARG, as discussed above, is invariably
-couched in terms of the events that occured in the history of some set of sampled
-genomes. Similarly, the genome ARG in Fig.~\ref{fig-arg-in-pedigree}
-is retrospective because it only includes information
-about genetic inheritance relative to the sample individuals
-$D_1$ and $D_2$ (genomes \noderef{a}--\noderef{d}).
-That is, only the inheritance
-history of the sample genomes \noderef{a}--\noderef{d} is recorded in the ARG.
-% Gregor & Jerome think that "inheritance" is clearer here than "transmission"
-Genetic inheritance must have occurred between (e.g.)
-$D_8$ and her daughter $D_6$, but this is not recorded because
-it is not relevant to the genetic ancestry of the sample genomes.
-
-There is no particular reason, however, that ARGs must be
-retrospective. Recent advances in individual-based forward-time simulation
-~\citep{kelleher2018efficient,haller2018tree} depend upon the ability to
-record gARGs in a \emph{prospective} manner.
-In a prospective gARG, the inheritance intervals $I$
-on an edge joining ancestor genome $v$ and descendant $u$
-define \emph{all} regions of genome that $u$ inherited from $v$.
-Thus, these inheritance intervals are
-concerned only with the genetic relationship
-between the nodes in question, and are not defined with respect
-to any given set of samples. They simply record the immediate
-genetic inheritance between two genomes
-(in the simplest case, between a parent and a child).
-\citet{kelleher2018efficient} refer to the structure
-recording all such forwards-in-time genetic relationships
-as an ``embellished pedigree`` (and suggest the term ``nedigree'' as a shorthand).
-Describing this instead as a ``prospective ARG'', emphasising
-the use of the same data structures and the direct connection with
-the usual retrospective interpretation of an ARG, may be
-more useful.
-
-The exhaustive recording of full inheritance intervals from generation
-to generation leads to a rapid build-up of information, much of which
-may not be relevant to the genetic ancestry of the current population.
-This redundancy can be periodically removed using the ``simplify'' algorithm,
-described in detail by \citet{kelleher2018efficient}. Simplification
-allows efficient subsetting of ARGs (removing both samples and certain
-levels of detail about recombination, as discussed in Section~\ref{sec-ARG-simplification}),
-and is also the key enabling factor when incorporating propective ARGs into
+% Whether we regard nodes as events or genomes is something of a
+% philosophical question, and in most senses the two are equivalent
+% (but see the discussion on ``unary nodes'' in Section~\ref{sec-locally-unary-edges}).
+% From a concrete data-structure encoding perspective, the only real
+% difference between an eARG and a gARG is that in an eARG we store recombination
+% \emph{breakpoints} at nodes of out-degree two (recombination nodes)
+% whereas in a gARG we store \emph{inheritance intervals} on edges.
+% A key advantage of associating inheritance intervals with
+% edges in a gARG is that it allows us to describe genetic inheritance events other
+% than simple crossovers, such as multiple recombinations and
+% gene conversions. By explicitly stating the exact regions of genome
+% that are transmitted along an edge, we generalise the representation
+% to encompass any process that leads to the differential inheritance of
+% genetic material along the genome.
+
+% While the additional generality provided by the gARG representation
+% is useful, it is in reality a small change that does not
+% lead to significantly different time or space complexity in computational
+% tasks.
+% The vitally important and fundamental distinction between
+% a gARG and an eARG is that gARGs
+% allow us to directly describe
+% genetic inheritance in multiple non-contiguous genome intervals
+% between an ancestor and a descendant, which cannot be done
+% by specifying recombination events and breakpoints.
+% % FIXME this sentence is problematic. We should recast to talk about
+% % the outcomes of events, not talk about simulateousness (which will
+% % confuse people
+% Thus, we have the means to encode the \emph{outcome} of
+% multiple genetic inheritance events, and to
+% systematically describe the uncertainty around their
+% ordering (see Section~\ref{sec-precision}).
+% As we describe in the following sections, freeing ourselves from the requirement
+% of recording specific events and
+% the additional generality of the gARG encoding
+% has significant advantages. It also raises questions about what our goals
+% for inferring ARGs should be, and how we evaluate the success
+% of such inferences.
+
+\section{Ancestral material and sample resolution}
+\label{sec-ancestral-material}
+Ancestral material~\citep{wiuf1999ancestry,wiuf1999recombination},
+denoting the genomic intervals ancestral to a set of samples
+on the edges of an ARG, is a key concept. In
+Fig.~\ref{fig-arg-in-pedigree}, for example, we have
+four sample genomes, \noderef{a}--\noderef{d}. As we
+trace their genetic ancestry into the previous generation
+(\noderef{e}--\noderef{h}), we can think of the ancestral
+material of the sample propagating through the graph
+backwards in time. In the interval $[2, 7)$, there is a
+local coalescence where nodes \noderef{a} and \noderef{c}
+find common ancestor in \noderef{f}. Thus, in this interval,
+the total number of genome segments that are ancestral to the
+sample is reduced from four to three. Note that in the same
+interval, node \noderef{e} is not ancestral to any
+of the samples: it carries \emph{non ancestral} material.
+This process of local coalescence continues through the
+graph, until all samples reach their most recent common
+ancestor in node \noderef{n}.
+
+The process of tracking local coalescences and updating
+segments of ancestral material is a core element of
+Hudson's seminal simulation
+algorithm~\citep{hudson1983testing,kelleher2016efficient},
+and the key distinguishing feature between the
+``big'' and ``little'' ARG stochastic processes
+(see Appendix~\ref{sec-big-and-little-arg}).
+The ability to \emph{store} the resolved ancestral material
+for a sample
+in an ARG is also a key distinction between the eARG and gARG
+encodings. Because an eARG stores only the graph topology and
+recombination breakpoints, there is no way to locally
+ascertain the ancestral material without traversing the graph
+rootwards from the samples,
+resolving the effects of recombination and common ancestor events.
+
+Efficiently propagating and resolving ancestral material for
+a sample through a pre-existing graph is a well-studied problem,
+and central to recent advances in individual-based forwards-time
+simulations~\citep{kelleher2018efficient,haller2018tree}.
+In contrast to the usual ``retrospective'' view of ARGs
+discussed so far, these methods record an ARG forwards in
+time in a ``prospective'' manner. Genetic inheritance relationships
+are recorded exhaustively, generation-by-generation
+leading to a rapid build-up of information, much of which
+will not be relevant to the genetic ancestry of the current population.
+This redundancy is periodically removed using the ``simplify''
+algorithm~\citep{kelleher2018efficient}, which propagates and
+resolves ancestral material.
+Simplification allows efficient subsetting of ARGs (removing both samples and
+certain levels of detail about recombination, as discussed in
+Section~\ref{sec-ARG-simplification}), and is also the key enabling factor when
+incorporating propective ARGs into
 forward-time simulation.
-The core algorithm involves the concept of
-``ancestral material''~\citep{wiuf1999ancestry,wiuf1999recombination}:
-the genomic intervals ancestral to a set of samples,
-scattered across lineages (the edges of the graph)
-at some time in the past. In the next section, we outline how
-tracing ancestral material allows the identification and removal of nodes
-and inheritance intervals, such that
-only genetic ancestry relevant to the current population is retained.
-
-\section{Converting an eARG to a gARG}
-\label{sec-eARG-to-gARG}
-Any eARG can be converted to a gARG without loss of information.
-The process is illustrated in Fig.~\ref{fig-ancestry-resolution} (A to B):
-we convert all event nodes to genome nodes and add inheritance annotations
-to the gARG's edges. If the node is a common ancestor node,
-we annotate the single outbound edge with the interval $[0,L)$,
-for a genome of length $L$.
-If the node is a recombination event with a breakpoint $x$,
-we annotate the two outbound edges respectively with the intervals $[0, x)$ and $[x, L)$.
-These inheritance interval annotations are clearly in one-to-one
-correspondence with the information in the input eARG.
 
@@
 
-This way of specifying inheritance intervals generates annotations equivalent to
-those produced in a prospective gARG, and which is similarly amenable to ``simplification'',
-by considering the genetic material that is ancestral to the samples \citep{kelleher2018efficient}.
-The outcome of this process, which is
-directly analogous to Hudson's simulation algorithm for the coalescent with
-recombination~\citep{hudson1983testing,kelleher2016efficient}, is shown in
-Fig.~\ref{fig-ancestry-resolution}C. We traverse from the samples rootwards through the graph
-keeping track of which material is ancestral to the samples.
-At recombination events, ancestral material is split between the ancestral lineages.
-As genome segments coalesce in a common
-ancestor the total amount of ancestral material is reduced, until
-eventually all samples reach a most recent common ancestor at every location
-along the genome.
-
-Fig.~\ref{fig-ancestry-resolution}C therefore represents a \emph{sample-resolved} gARG,
-% useful to say that we don't *need* to simulate an eARG and convert to a
-% sample-resolved form, but can simulate a sample-resolved gARG directly
-of the sort that can be simulated directly by the Hudson algorithm.
-This differs in some important ways from its equivalent eARG (Fig.~\ref{fig-ancestry-resolution}A).
+Any eARG can be converted to a sample-resolved gARG using the same machinery,
+and the two-step process is illustrated in Fig.~\ref{fig-ancestry-resolution}.
+The first step is to take the input eARG (Fig.~\ref{fig-ancestry-resolution}A)
+and convert all event nodes to genome nodes and add inheritance annotations to
+the gARG's edges (Fig.~\ref{fig-ancestry-resolution}B).
+If the node is a common ancestor node, we annotate the single
+outbound edge with the interval $[0,L)$, for a genome of length $L$. If the
+node is a recombination event with a breakpoint $x$, we annotate the two
+outbound edges respectively with the intervals $[0, x)$ and $[x, L)$. These
+inheritance interval annotations are clearly in one-to-one correspondence with
+the information in the input eARG. They are also also analogous to the
+inheritance intervals we get on the edges in a prospective ARG
+produced by a forwards-time simulation, which are concerned with recording
+the local genetic relationship between a parent and child genome and not
+necessarily minimal in terms of the resolved ancestral material of a sample.
+Thus, the final step is to use the ``simplify'' algorithm to perform this
+resolution~(Fig.\ref{fig-ancestry-resolution}C).
+
+The sample-resolved gARG of Fig.~\ref{fig-ancestry-resolution}C
+differs in some important ways to the
+original eARG (Fig.~\ref{fig-ancestry-resolution}A).
 Firstly, we can see that some nodes and edges have been removed entirely
@@
 holds.
+
+% \section{Retrospective vs prospective ARGs}
+% \label{sec-retro-pro}
+% The ARG concept arose from a \emph{retrospective} view of inheritance.
+% For example the definition of an eARG, as discussed above, is invariably
+% couched in terms of the events that occured in the history of some set of sampled
+% genomes. Similarly, the genome ARG in Fig.~\ref{fig-arg-in-pedigree}
+% is retrospective because it only includes information
+% about genetic inheritance relative to the sample individuals
+% $D_1$ and $D_2$ (genomes \noderef{a}--\noderef{d}).
+% That is, only the inheritance
+% history of the sample genomes \noderef{a}--\noderef{d} is recorded in the ARG.
+% % Gregor & Jerome think that "inheritance" is clearer here than "transmission"
+% Genetic inheritance must have occurred between (e.g.)
+% $D_8$ and her daughter $D_6$, but this is not recorded because
+% it is not relevant to the genetic ancestry of the sample genomes.
+
+% There is no particular reason, however, that ARGs must be
+% retrospective. Recent advances in individual-based forward-time simulation
+% ~\citep{kelleher2018efficient,haller2018tree} depend upon the ability to
+% record gARGs in a \emph{prospective} manner.
+% In a prospective gARG, the inheritance intervals $I$
+% on an edge joining ancestor genome $v$ and descendant $u$
+% define \emph{all} regions of genome that $u$ inherited from $v$.
+% Thus, these inheritance intervals are
+% concerned only with the genetic relationship
+% between the nodes in question, and are not defined with respect
+% to any given set of samples. They simply record the immediate
+% genetic inheritance between two genomes
+% (in the simplest case, between a parent and a child).
+% \citet{kelleher2018efficient} refer to the structure
+% recording all such forwards-in-time genetic relationships
+% as an ``embellished pedigree`` (and suggest the term ``nedigree'' as a shorthand).
+% Describing this instead as a ``prospective ARG'', emphasising
+% the use of the same data structures and the direct connection with
+% the usual retrospective interpretation of an ARG, may be
+% more useful.
+
+% The exhaustive recording of full inheritance intervals from generation
+% to generation leads to a rapid build-up of information, much of which
+% may not be relevant to the genetic ancestry of the current population.
+% This redundancy can be periodically removed using the ``simplify'' algorithm,
+% described in detail by \citet{kelleher2018efficient}. Simplification
+% allows efficient subsetting of ARGs (removing both samples and certain
+% levels of detail about recombination, as discussed in Section~\ref{sec-ARG-simplification}),
+% and is also the key enabling factor when incorporating propective ARGs into
+% forward-time simulation.
+% The core algorithm involves the concept of
+% ``ancestral material''~\citep{wiuf1999ancestry,wiuf1999recombination}:
+% the genomic intervals ancestral to a set of samples,
+% scattered across lineages (the edges of the graph)
+% at some time in the past. In the next section, we outline how
+% tracing ancestral material allows the identification and removal of nodes
+% and inheritance intervals, such that
+% only genetic ancestry relevant to the current population is retained.
 
@@
 in the local tree. Thus, \noderef{c} has a single child in this
-interval and it is hence unary in this local tree.
+interval and it is unary in this local tree.
 % TODO note this is CA but not coalescent, and refer to where we discuss this.
@@
 data is sampled at many timepoints and recombination
-is rare (e.g.~SARS-Cov-2; see Discussion).
+is rare (e.g.~SARS-CoV-2; see Discussion).
 \emph{Locally} unary nodes, those which have one child
@@
 (hence Fig.~\ref{fig-inferred-args} A and Fig.~\ref{fig-inferred-args} B disagree on the
-relationship between these samples). 
 More generally, the number of nodes in Fig.~\ref{fig-inferred-args} C
@@
 Even when a recombination event \emph{is} detectable, its timing and breakpoint
-positions can only be inferred approximately, depending on how much information
+position can only be inferred approximately, depending on how much information
 can be elucidated from mutations in the surrounding genomic region. For general
@@
 The main application for fully precise ARGs thus far has been
-to compute a likelihood under the coalescent with
-recombination~\citep[e.g.][]{kuhner2000maximum,mahmoudi2022bayesian,
+to compute a likelihood under the
+coalescent~\citep[e.g.][]{kuhner2000maximum,mahmoudi2022bayesian,
 guo2022recombination},
@@
 
-\begin{table}[ht]
-    \begin{center}
-        \begin{tabular}{c|ccc|c|c}
-			\multicolumn{2}{c}{Node table}& ~ &\multicolumn{3}{c}{Edge table}\\
-			\cline{1-2} \cline{4-6}
-            Node & Time & ~ & Child & Parent & Intervals\\
-            \cline{1-2} \cline{4-6}
-            $\noderef{a}$ & 0 & ~ & $\noderef{a}$ & $\noderef{e}$ & $\{[0,2)\}$\\
-            $\noderef{b}$ & 0 & ~ & $\noderef{a}$ & $\noderef{f}$ & $\{[2,10)\}$\\
-            $\noderef{c}$ & 0 & ~ & $\noderef{b}$ & $\noderef{g}$ & $\{[0,10)\}$\\
-            $\noderef{d}$ & 0 & ~ & $\noderef{c}$ & $\noderef{f}$ & $\{[0,7)\}$\\
-            $\noderef{e}$ & 1 & ~ & $\noderef{c}$ & $\noderef{e}$ & $\{[7,10)\}$\\
-            $\noderef{f}$ & 1 & ~ & $\noderef{d}$ & $\noderef{h}$ & $\{[0,10)\}$\\
-            $\noderef{g}$ & 1 & ~ & $\noderef{e}$ & $\noderef{i}$ & $\{[0,10)\}$\\
-            $\noderef{i}$ & 1 & ~ & $\noderef{f}$ & $\noderef{k}$ & $\{[0,10)\}$\\
-            $\noderef{k}$ & 2 & ~ & $\noderef{g}$ & $\noderef{i}$ & $\{[0,10)\}$\\
-            $\noderef{n}$ & 3 & ~ & $\noderef{h}$ & $\noderef{k}$ & $\{[0,10)\}$\\
-            \multicolumn{2}{c}{~}& ~ & $\noderef{i}$ & $\noderef{n}$ & $\{[0,10)\}$\\
-            \multicolumn{2}{c}{~}& ~ & $\noderef{k}$ & $\noderef{n}$ & $\{[0,10)\}$
-        \end{tabular}
-    \end{center}
-    % NOTE: I know  the "correct" placement for table captions is on top, but I
-    % always find this quite weird (since many of our figs are quite "table
-    % like" anyway. If anyone feels strongly about putting the caption above
-    % the table please just move it back.
-    \caption{\label{tab-gARG-data}
-    Data encoding of the genome ARG shown in Fig.~\ref{fig-arg-in-pedigree} with the
-	node table on the left (having node and time columns) and the edge table on the
-	right (having child, parent, and intervals columns).
-    }
-\end{table}
+% \begin{table}[ht]
+%     \begin{center}
+%         \begin{tabular}{c|ccc|c|c}
+% 			\multicolumn{2}{c}{Node table}& ~ &\multicolumn{3}{c}{Edge table}\\
+% 			\cline{1-2} \cline{4-6}
+%             Node & Time & ~ & Child & Parent & Intervals\\
+%             \cline{1-2} \cline{4-6}
+%             $\noderef{a}$ & 0 & ~ & $\noderef{a}$ & $\noderef{e}$ & $\{[0,2)\}$\\
+%             $\noderef{b}$ & 0 & ~ & $\noderef{a}$ & $\noderef{f}$ & $\{[2,10)\}$\\
+%             $\noderef{c}$ & 0 & ~ & $\noderef{b}$ & $\noderef{g}$ & $\{[0,10)\}$\\
+%             $\noderef{d}$ & 0 & ~ & $\noderef{c}$ & $\noderef{f}$ & $\{[0,7)\}$\\
+%             $\noderef{e}$ & 1 & ~ & $\noderef{c}$ & $\noderef{e}$ & $\{[7,10)\}$\\
+%             $\noderef{f}$ & 1 & ~ & $\noderef{d}$ & $\noderef{h}$ & $\{[0,10)\}$\\
+%             $\noderef{g}$ & 1 & ~ & $\noderef{e}$ & $\noderef{i}$ & $\{[0,10)\}$\\
+%             $\noderef{i}$ & 1 & ~ & $\noderef{f}$ & $\noderef{k}$ & $\{[0,10)\}$\\
+%             $\noderef{k}$ & 2 & ~ & $\noderef{g}$ & $\noderef{i}$ & $\{[0,10)\}$\\
+%             $\noderef{n}$ & 3 & ~ & $\noderef{h}$ & $\noderef{k}$ & $\{[0,10)\}$\\
+%             \multicolumn{2}{c}{~}& ~ & $\noderef{i}$ & $\noderef{n}$ & $\{[0,10)\}$\\
+%             \multicolumn{2}{c}{~}& ~ & $\noderef{k}$ & $\noderef{n}$ & $\{[0,10)\}$
+%         \end{tabular}
+%     \end{center}
+%     % NOTE: I know  the "correct" placement for table captions is on top, but I
+%     % always find this quite weird (since many of our figs are quite "table
+%     % like" anyway. If anyone feels strongly about putting the caption above
+%     % the table please just move it back.
+%     \caption{\label{tab-gARG-data}
+%     Data encoding of the genome ARG shown in Fig.~\ref{fig-arg-in-pedigree} with the
+% 	node table on the left (having node and time columns) and the edge table on the
+% 	right (having child, parent, and intervals columns).
+%     }
+% \end{table}
 

### 2023-10-22 illustrations/ARGweaver_example.py
Update all the tool outputs for inference

@@ -1,97 +1,0 @@
-import collections
-import io
-import sys
-from pathlib import Path
-
-import networkx as nx
-import matplotlib.pyplot as plt
-
-# useful scripts in the top level dir
-sys.path.append(str(Path(__file__).parent.parent.absolute() / "utils"))
-import convert
-import ts_process
-import argdraw
-
-ts = convert.arg_to_ts(io.StringIO(
-"""\
-start=0	end=1000
-name	event	age	pos	parents	children
-1	recomb	132.704741329	918	5,16	n2
-2	coal	441.51496637	0	11	n5,n1
-3	recomb	441.51496637	271	26,25	n6
-4	recomb	1160.13100175	670	14,7	n0
-5	coal	1160.13100175	0	10	n3,1
-6	recomb	2832.38458802	59	18,9	n7
-7	recomb	2832.38458802	782	8,14	4
-8	coal	2832.38458802	0	15	7,n4
-9	recomb	2832.38458802	714	10,20	6
-10	coal	2832.38458802	0	21	9,5
-11	recomb	2832.38458802	254	12,13	2
-12	recomb	2832.38458802	7	13,16	11
-13	coal	2832.38458802	0	17	12,11
-14	coal	2832.38458802	0	20	7,4
-15	recomb	2832.38458802	826	29,28	8
-16	coal	2832.38458802	0	23	12,1
-17	recomb	6723.79797033	695	27,18	13
-18	coal	6723.79797033	0	19	17,6
-19	recomb	6723.79797033	41	28,22	18
-20	coal	6723.79797033	0	26	14,9
-21	recomb	6723.79797033	484	35,23	10
-22	recomb	6723.79797033	834	24,25	19
-23	coal	6723.79797033	0	30	21,16
-24	recomb	6723.79797033	222	39,29	22
-25	coal	6723.79797033	0	37	22,3
-26	coal	6723.79797033	0	32	20,3
-27	recomb	6723.79797033	6	40,30	17
-28	coal	6723.79797033	0	31	19,15
-29	coal	6723.79797033	0	31	24,15
-30	coal	6723.79797033	0	33	23,27
-31	coal	6723.79797033	0	34	29,28
-32	recomb	6723.79797033	474	33,45	26
-33	coal	6723.79797033	0	41	30,32
-34	recomb	6723.79797033	710	36,37	31
[…]

### 2023-10-22 make_example_outputs.py

@@ -150,14 +150,16 @@ def run_relate():
     subprocess.run(params, cwd=dir)
 
     # Convert to JBOT tree sequence format
-    subprocess.run([
+    params = [
         "tools/relate_lib/bin/Convert",
         "--mode", "ConvertToTreeSequence",
         "--anc", f"{dir}{outfiles}.anc",
         "--mut", f"{dir}{outfiles}.mut",
         "-o", f"examples/Kreitman_SNP_relate_jbot",
         *conf["tools"]["RELATELIB_PARAMS"].split(),
-    ])
+    ]
+    print(f"running `{' '.join(params)}`")
+    subprocess.run(params)
     
     # Convert to time-uncalibrated tree sequence format
     ts_jbot = tskit.load("examples/Kreitman_SNP_relate_jbot.trees")

### 2023-10-23 paper.tex
Yans updates

@@
 follow~\citep{harris2019database,harris2023using}.
-See \citet{lewanski2023era} for a biologically oriented introduction to ARGs
-and further discussion on the potential impacts.
+\citet{lewanski2023era} provide a biologically oriented introduction to ARGs
+and further discuss their potential impacts.
 
@@
 are subtle, varied, and deep.
-% TODO rephrase, not great currentyl
+% TODO rephrase, not great currently
 We argue that the prevailing definition of an ARG
-imposes unneccesary limitations, and that with a broader definition
+imposes unnecessary limitations, and that with a broader definition
 we can reassess what we should be attempting to estimate,
@@
 % Levels of simplification
-This simplification is closely related to operations needed to convert eARG and gARG,
+This simplification is closely related to operations needed to convert eARGs to
+various forms of gARGs,
 which opens a rich set of details about ARGs and local trees, local unary nodes, and
@@
 % Anastasia Ignatieva acknowledges support from TODO
-% TODO Jerome: add EPSRC and NIH grants.
+% TODO Jerome: add EPSRC and NIH HG011395-0 grants.
 Jerome Kelleher acknowledges support from the Robertson Foundation.

### 2023-10-23 paper.tex
Remove SI units

@@
 \usepackage{subcaption}
-\usepackage{siunitx}
 \usepackage{tikz}
@@
 assume a mutation rate of
-% per site per generation may be easier to follow?
-\SI{5.49e-9}{per site per generation} and a recombination rate of
-\SI{8.4e-9}{per site per generation} \citep{lauterbur2022expanding}, together with
+$5.49\times10^{-9}$ per site per generation
+and a recombination rate of
+$8.4\times10^{-9}$ per site per generation
+\citep{adrion2020community,lauterbur2022expanding},
+together with
 a constant effective population size of 1,000,000.

### 2023-10-23 paper.tex
Yans updates

@@
 integers used here in the definition of a node as an \emph{identifier},
-with which arbitrary additional information can be associated.
-Concretely defining nodes as integers in this way has some
+with which arbitrary additional information can be associated.Concretely defining nodes as integers in this way has some
 significant advantages. In particular, it provides a direct link
@@
 Note that both \kwarg\ and \argweaver\ produce many ARGs
-(those shown in Fig.~\ref{fig-inferred-args} A\& B are arbitrarily selected examples).
+(those shown in Fig.~\ref{fig-inferred-args}A,B are arbitrarily selected examples).
 This stems from the fact that \kwarg\ searches the space of plausible ARGs,
@@
 an approximation of the full coalescent with recombination).
-In contrast, \tsinfer\ and \relate\ (Fig.~\ref{fig-inferred-args} C \& D) are
+In contrast, \tsinfer\ and \relate\ (Fig.~\ref{fig-inferred-args}C,D) are
 deterministic methods, aiming to find a single ``best guess'' ARG.
@@
 of number of parents per node, as highlighted by the node colour scheme.
-In Fig.~\ref{fig-inferred-args} A \& B, each recombination event is tagged by
+In Fig.~\ref{fig-inferred-args}A,B, each recombination event is tagged by
 a node with exactly 2 parents and one child. Assuming that the breakpoint for each event
@@
 number of breakpoints should exactly equal the number of 2-parent (brown) nodes in the plot.
-In contrast, Figure Fig.~\ref{fig-inferred-args} C \& D have ARG nodes with multiple
+In contrast, Figure Fig.~\ref{fig-inferred-args}C,D have ARG nodes with multiple
 parents and multiple children. These indicate deliberate imprecision, where one
@@
 bifurcating trees, they are forced to pick an arbitrary order of branching
-(hence Fig.~\ref{fig-inferred-args} A and Fig.~\ref{fig-inferred-args} B disagree on the
+(hence Fig.~\ref{fig-inferred-args}A,B disagrees with Fig.~\ref{fig-inferred-args}D
+in this order)
 More generally, the number of nodes in Fig.~\ref{fig-inferred-args} C

### 2023-10-23 paper.tex
Some edits from Gregor

@@
 We define a genome as the complete set of genetic material that a child
-inherits from one parent (i.e.\ haploids have one genome, diploids two, etc.).
+inherits from one parent. A diploid individual  (we assume diploids here
+for clarity, but the definitions apply to organisms of arbitrary ploidy)
+% Q: will someone nitpick that we're assuming diploid obligate sexual species?
+therefore carries two genomes, one inherited from each parent.
 We will also use the term ``genome'' in its
@@
  we ``annotate'' the edges with the genome
-coordinates over which inheritance occurs.
+coordinates over which inheritance occured.
 We can define a gARG formally as follows.
@@
 In practical settings, we will wish to associate additional
-metadata such as sample identifiers or quality-control metrics
-with nodes; it is therefore best to think of the
+metadata with nodes such as sample identifiers or quality-control metrics;
+It is therefore best to think of the
 integers used here in the definition of a node as an \emph{identifier},
-with which arbitrary additional information can be associated.Concretely defining nodes as integers in this way has some
+with which arbitrary additional information can be associated.
+Concretely defining nodes as integers in this way has some
 significant advantages. In particular, it provides a direct link
@@
 the gARG for a given set of individuals is embedded in their pedigree.
-The pedigree relationships between the diploid individuals is
-shown, along with their paths
-of genetic inheritance and constituent genomes (labelled
-with lowercase alphabetical letters
-rather than integer identifiers to avoid confusion with genomic
-intervals).
+The figure shows the pedigree of eight diploid individuals, along with paths
+of genetic inheritance and constituent genomes.
+To avoid confusion with genomic intervals, nodes are
+labelled with lowercase alphabetical letters
+rather than integer identifiers.
 Thus, individual $D_1$ is composed
@@
 \cite[e.g.][]{mathieson2020ancestry,shipilina2023origin},
-adding concrete details to the mechanisms by which
-differential inheritance of genetic material are modelled.
+adding concrete details to describe the
+differential inheritance of genetic material between genomes.
 % This is definitely needed for the Nicks, who are always thinking
@@
 actual ancestral processes that occur in nature, but rather how we
-\emph{represent} such realised histories in a practical sense.
+\emph{represent} the outcomse of such processes in a practical sense.
 % Again, need to remind the reader that there's an actual point
@@
 ancestral lineages. Recombination nodes are annotated with
-the corresponding crossover breakpoint, and these breakpoints
-are used to reconstruct the local trees.
+the corresponding crossover breakpoints. These breakpoints
+are used to construct the local trees.
 This is done by tracing rootwards through the graph from the samples,
@@
 be included, but either require additional node types
-or (e.g.) multiple recombination nodes joined by a zero-length edge.
+or, for instance, multiple recombination nodes joined by a zero-length edge.
 Modelling multiple recombinant offspring of the same parents
@@
 inheritance interval annotations are clearly in one-to-one correspondence with
-the information in the input eARG. They are also also analogous to the
+the information in the input eARG. They are also analogous to the
 inheritance intervals we get on the edges in a prospective ARG

### 2023-10-23 paper.tex
Ana's updates

--- 2023-10-23 paper.tex
@@
 integers used here in the definition of a node as an \emph{identifier},
-with which arbitrary additional information can be associated.Concretely defining nodes as integers in this way has some
+with which arbitrary additional information can be associated. Concretely defining nodes as integers in this way has some
 significant advantages. In particular, it provides a direct link
@@
 the gARG for a given set of individuals is embedded in their pedigree.
-The pedigree relationships between the diploid individuals is
+The pedigree relationships between the diploid individuals are
 shown, along with their paths

### 2023-10-23 paper.tex
Gregor comments
standardise Big and Little ARG

--- 2023-10-23 paper.tex
@@
 a brief history of ancestral graphs (Appendix~\ref{sec-arg-history}),
-description of the big and little ARGs (Appendix~\ref{sec-big-and-little-arg}),
+description of the Big and Little ARGs (Appendix~\ref{sec-big-and-little-arg}),
 survey of ARG inference methods (Appendix~\ref{sec-survey-arg-infer}), and
@@
 In practical settings, we will wish to associate additional
-metadata with nodes such as sample identifiers or quality-control metrics;
+metadata with nodes such as sample identifiers or quality-control metrics.
 It is therefore best to think of the
@@
 actual ancestral processes that occur in nature, but rather how we
-\emph{represent} the outcomse of such processes in a practical sense.
+\emph{represent} the outcomes of such processes in a practical sense.
 % Again, need to remind the reader that there's an actual point
@@
 \label{sec-ancestral-material}
-Ancestral material~\citep{wiuf1999ancestry,wiuf1999recombination},
-denoting the genomic intervals ancestral to a set of samples
-on the edges of an ARG, is a key concept. In
-Fig.~\ref{fig-arg-in-pedigree}, for example, we have
+Ancestral material~\citep{wiuf1999ancestry,wiuf1999recombination}
+is a key concept in understanding the overall inheritance structure
+of an ARG. It denotes the genomic intervals ancestral to a set of samples
+on the edges of an ARG.
+In Fig.~\ref{fig-arg-in-pedigree}, for example, we have
 four sample genomes, \noderef{a}--\noderef{d}. As we
@@
 (\noderef{e}--\noderef{h}), we can think of the ancestral
-material of the sample propagating through the graph
+material of the samples propagating through the graph
 backwards in time. In the interval $[2, 7)$, there is a
 local coalescence where nodes \noderef{a} and \noderef{c}
-find common ancestor in \noderef{f}. Thus, in this interval,
+find a common ancestor in \noderef{f}. Thus, in this interval,
 the total number of genome segments that are ancestral to the
@@
 segments of ancestral material is a core element of
-Hudson's seminal simulation
+Hudson's seminal simulation simulation
 algorithm~\citep{hudson1983testing,kelleher2016efficient},
 and the key distinguishing feature between the
-``big'' and ``little'' ARG stochastic processes
+``Big'' and ``Little'' ARG stochastic processes
 (see Appendix~\ref{sec-big-and-little-arg}).
@@
 from the graph.
-The ``grand MRCA'' \noderef{q} is normally omitted from the
+The ``grand MRCA'' \noderef{q} is omitted from the
 sample-resolved gARG because all segments of the genome have
@@
 construct not an ARG but ``only'' a sequence of local
-trees \citep{hejase2020summary, ignatieva2021kwarg}. While this statement
+trees \citep[e.g.][]{hejase2020summary, brandt2021evaluation, ignatieva2021kwarg}. While this statement
 is correct inasmuch as neither method attempts to infer the explicit recombination events that cause
@@
 from the posterior distribution over
-ARGs given the data and a fixed model (the SMC$^\prime$ \citep{marjoram2006fast},
+ARGs given the data and a fixed model (a discretised version of the SMC$^\prime$ \citep{marjoram2006fast},
 an approximation of the full coalescent with recombination).
@@
 a node with exactly 2 parents and one child. Assuming that the breakpoint for each event
-occurs at a different location (the "infinite breakpoints" model), then the
+occurs at a different location (the ``infinite breakpoints" model), then the
 number of breakpoints should exactly equal the number of 2-parent (brown) nodes in the plot.
@@
 an efficient substrate for defining analysis algorithms is that
-it allows us to generate the local trees along the genome
+it allows us to recover the local trees along the genome
 in a way that allows us to reason about the \emph{differences}
@@
 the ``big" ARG process of \cite{griffiths1997ancestral}, and the ``little" ARG process of
- \cite{hudson1983properties}. The big ARG process is mathematically simpler
+ \cite{hudson1983properties}. The Big ARG process is mathematically simpler
  but is computationally intractable due to generating a vast number of ancestors
  which contribute no genetic material to the initial sample.
-The little ARG process avoids non-genetic ancestors at the cost of more complex
+The Little ARG process avoids non-genetic ancestors at the cost of more complex
 dynamics and state space. We also demonstrate that evaluating the sampling probability
@@
 
-A generic state of the little ARG process consists of a finite collection of lineages $L$,
+A generic state of the Little ARG process consists of a finite collection of lineages $L$,
 each of which is a list of disjoint ancestry segments $(\ell, r, a)$, where
@@
 
-In the Griffiths formulation (the big ARG process), each edge in the graph corresponds to an extant
+In the Griffiths formulation (the Big ARG process), each edge in the graph corresponds to an extant
 lineage and nodes are events in the process. The $n$ initial leaf nodes are
@@
 
-The state-space of the big ARG process is much simpler than that of the little ARG process,
+The state-space of the Big ARG process is much simpler than that of the Little ARG process,
 which greatly facilitates mathematical reasoning. This simplicity comes at a
@@
 simulating recombinant ancestries.
-The number of events in the big ARG all the way back to the GMRCA
+The number of events in the Big ARG all the way back to the GMRCA
 is $O(e^\rho)$~\citep{griffiths1997ancestral}, whereas the number
-of events required to simulate the little ARG is
+of events required to simulate the Little ARG is
 $O(\rho^2)$~\citep{hein2004gene,baumdicker2021efficient}.
-This disparity arises because the majority of the events in the big ARG are
+This disparity arises because the majority of the events in the Big ARG are
 recombination events which occur outside of ancestral material,
@@
 Figure \ref{hudson_vs_bigARG} illustrates the more complex state space
-of the little ARG process, as well as the extra events which occur in the big ARG process.
+of the Little ARG process, as well as the extra events which occur in the Big ARG process.
 Moreover, it depicts the rates of common ancestors and recombination events in each
@@
 without an external convention, the exact time at which the trapped ancestral material on
-node \noderef{k} ceases to be available for an effective recombination in the little ARG process.
+node \noderef{k} ceases to be available for an effective recombination in the Little ARG process.
 
@@
 the recombination rate is $\rho$.
-(B) A corresponding realisation of a big ARG, which augments Hudson's algorithm
+(B) A corresponding realisation of a Big ARG, which augments Hudson's algorithm
 by tracking nonancestral lineages. The result is a simpler state space and
@@
 These methods operate on representations
-of the ``little ARG'', and are computationally expensive, being
+of the ``Little ARG'', and are computationally expensive, being
 applicable to at most hundreds of samples consisting of tens or hundreds of

### 2023-10-23 paper.tex
More gregor updates

@@
 representing the coalescence of samples \noderef{a} and \noderef{c}
-in this region of genome. Over the interval $[0, 2)$ no coalescence occurs,
+in this genome region. Over the interval $[0, 2)$ no coalescence occurs,
 but we still record the fact that genome \noderef{c} inherits from \noderef{f}
@@
 Outside of the context of simulation, ARG-unary nodes are
-only like to occur in longitudinal datasets where genetic
+only likely to occur in longitudinal datasets where genetic
 data is sampled at many timepoints and recombination
@@
 
+% FIXME: move to top of section
 Nodes with one child are not a standard feature of evolutionary trees
@@
 The graph visualisations in Fig.~\ref{fig-simplification} have
-some novel features, which require some explanation.
+three novel features.
 Firstly, edge weights (the thickness of the lines joining
@@
 \caption{\label{fig-inferred-args} Inference of sample-resolved ARGs for 11
-\textit{Drosophila melanogaster} genetic sequences over a 2.4kb
+\textit{Drosophila melanogaster} DNA sequences over a 2.4kb
 region of the ADH locus~\citep{kreitman1983nucleotide}.
 Results for four different methods based on
+% FIXME: this is the wrong place to be summarising. Name the methods
+% here and summarise in the text
 (A) parsimony, (B) sampling the SMC$^\prime$, (C) copying from
@@
 The ability of a gARG to encode the effects of recombination, rather
-than having to precisely specify each event, is a important contributor to this aim,
+than having to precisely specify each event, is an important contributor to this aim,
 as most recombination events cannot be uniquely identified from the data.
@@
 a node with exactly 2 parents and one child. Assuming that the breakpoint for each event
-occurs at a different location (the ``infinite breakpoints" model), then the
+% TODO "infinite breakpoints" is goign to confuse people. What are we
+% clarifying here? Citation necessary if we keep.
+occurs at a different location (the ``infinite breakpoints'' model), then the
 number of breakpoints should exactly equal the number of 2-parent (brown) nodes in the plot.
@@
 %nodes or differently labelled nodes in each tree).
-For example, both infer that \noderef{Af-f}, \noderef{Fr-f} and \noderef{Wa-f}
+For example, both infer that \noderef{Af-f}, \noderef{Fr-f}, and \noderef{Wa-f}
 share a common ancestor along
@@
 single ARG root in the \kwarg\ and \argweaver\ genealogies. In terms of inference,
-\tsinfer\ and \relate\ show a more pragmatic representation: we do not expect
+\tsinfer\ and \relate\ show a pragmatic representation: we do not expect
 genetic variation to be very informative about topological relationships
@@
 The flexibility of the gARG encoding contrasts with the classical
-``event'' ARG (eARG), which is more limited in what can be described, because of its origins in an idealised mathematical model.
+``event'' ARG (eARG), which is more limited in what can be described,
+because of its origins in an idealised mathematical model,
+and the assumption of having complete information about
+recombination events.
 
@@
 even more challenging. The fact that the eARG encoding \emph{requires}
-precise estimates about recombination is therefore a fundamental limitation, which
+precise information about recombination is therefore a fundamental limitation, which
 is a central message of this paper.

### 2023-10-23 paper.tex
tweak

--- 2023-10-23 paper.tex
@@
 In practical settings, we will wish to associate additional
-metadata such as sample identifiers or quality-control metrics
-with nodes; it is therefore best to think of the
+metadata with nodes such as sample identifiers or quality-control metrics.
+It is therefore best to think of the
 integers used here in the definition of a node as an \emph{identifier},
-with which arbitrary additional information can be associated. Concretely defining nodes as integers in this way has some
+with which arbitrary additional information can be associated.
+Concretely defining nodes as integers in this way has some
 significant advantages. In particular, it provides a direct link
@@
 the gARG for a given set of individuals is embedded in their pedigree.
-The pedigree relationships between the diploid individuals are
-shown, along with their paths
-of genetic inheritance and constituent genomes (labelled
-with lowercase alphabetical letters
-rather than integer identifiers to avoid confusion with genomic
-intervals).
+The figure shows the pedigree of eight diploid individuals, along with paths
+of genetic inheritance and constituent genomes.
+To avoid confusion with genomic intervals, nodes are
+labelled with lowercase alphabetical letters
+rather than integer identifiers.
 Thus, individual $D_1$ is composed
@@
 actual ancestral processes that occur in nature, but rather how we
-\emph{represent} such realised histories in a practical sense.
+\emph{represent} the outcomes of such processes in a practical sense.
 % Again, need to remind the reader that there's an actual point

### 2023-10-23 paper.tex
Fix

--- 2023-10-23 paper.tex
@@
 segments of ancestral material is a core element of
-Hudson's seminal simulation simulation
+Hudson's seminal simulation
 algorithm~\citep{hudson1983testing,kelleher2016efficient},

### 2023-10-23 paper.tex
Update

@@
 and the two-step process is illustrated in Fig.~\ref{fig-ancestry-resolution}.
-The first step is to take the input eARG (Fig.~\ref{fig-ancestry-resolution}A)
-and convert all event nodes to genome nodes and add inheritance annotations to
-the gARG's edges (Fig.~\ref{fig-ancestry-resolution}B).
-If the node is a common ancestor node, we annotate the single
+The first step is to take the input eARG (Fig.~\ref{fig-ancestry-resolution}A),
+duplicate its graph topology, and then add inheritance annotations
+to each of the gARG's edges (Fig.~\ref{fig-ancestry-resolution}B).
+If a given node is a common ancestor event, we annotate the single
 outbound edge with the interval $[0,L)$, for a genome of length $L$. If the
@@
 the information in the input eARG. They are also analogous to the
-inheritance intervals we get on the edges in a prospective ARG
+inheritance intervals we get on the edges in a prospective gARG
 produced by a forwards-time simulation, which are concerned with recording
@@
 Thus, the final step is to use the ``simplify'' algorithm to perform this
-resolution~(Fig.\ref{fig-ancestry-resolution}C).
+resolution (Fig.~\ref{fig-ancestry-resolution}C).
 
@@
 cumulative effects of events over time can be reasoned
-about, without first ``replaying'' those events. Many computations
+about, without first ``replaying'' those events.
+Many computations
 that we wish to perform on an ARG will require resolving
@@
 The gARG encoding
-allows us to perform this resolution of ancestral material once
+allows us to perform this once
 and to store the result,
@@
 \label{sec-ARG-and-local-trees}
-Local trees, describing the genetic genealogy of a sample
-at a particular location along the genome, are
-embedded within an ARG, and the relationship
-between the two is subtle and important.
-Note that many authors use the term ``marginal''
-% could list loads here, but not much point.
-trees~\citep[e.g.][]{griffiths1996ancestral,minichiello2006mapping,kelleher2016efficient}
-which we avoid due to the potential confusion with the
-statistical concept of marginal distributions
-(potentially leading to interpretation as an integrated tree over many positions).
-
-% TODO Need to make clear here what came from where. Basic combinatorial object
-% cite Knuth. Application to genetic ancestry, sim papers. Need to look
-% up what's said in favour of this integer based approach in which paper
-% and refer to tat.
-To make the following discussions precise, we provide
-a concrete definition and explicit process for local tree extraction.
-We can describe a local tree as an
-oriented tree~\cite[p.\ 461]{knuth11combinatorial},
-a sequence of integers $\pi_1, \pi_2\dots$ where $\pi_u$ denotes the parent of
-node $u$~\citep{kelleher2013coalescent,kelleher2014coalescent,kelleher2016efficient}.
-If $u$ is a root
-% I'm not sure if this reachability stuff is a helpful clarification, and whether it'll just
-% confuse people. It depends on some tricky details.
-(or not reachable from the samples)
-in the local tree at position $x$, then we adopt the convention that $\pi_u = -1$.
-% TODO Give example of an oriented tree from one of the figs?
-% TODO
-% A node in an oriented tree can have
-% zero, one,
-% or many children.
-
-To recover the local tree $\pi$ at position $x$, we begin by
-setting $\pi_u = -1$ for each $u\in N$. Then, for each sample
+% (We use ``local'' rather than
+% ``marginal''
+% % could list loads here, but not much point.
+% trees~\citep[e.g.][]{griffiths1996ancestral,minichiello2006mapping,kelleher2016efficient}
+% to avoid the potential confusion with the
+% statistical concept of marginal distributions.)
+A fundamental property of genetics is that a
+% TODO is this (genome) clarification helpful or confusing?
+given DNA nucleotide is inherited from exactly one parent (genome),
+both at an organismal and cell-by-cell level
+(Appendix~\ref{sec-cell-lineages-and-args}).
+These chains of single-parent inheritance over time give rise,
+by definition, to a tree structure.
+As a result of recombination, adjacent nucleotides can have
+different chains of inheritance, and
+an ARG encodes the entire ensemble of local trees along the
+genome for a given set of sample genomes.
+The relationship between an ARG and its corresponding
+local trees is subtle and important. To make the following
+discussions precise, we will define a concrete combinatorial encoding
+for local trees and an explicit procedure for constructing them.
+
+Oriented trees provide a
+convenient formalism to capture these relationships in a well-defined
+combinatorial object.
+% There's no need for a comma
+Let $\pi_1\dots\pi_n$ be a sequence of integers, such that $\pi_u$
+denotes the parent of node $u$, and $\pi_u = 0$ if $u$ is a
+root~\cite[p.\ 461]{knuth11combinatorial}.
+This encoding is particularly useful to describe
+evolutionary trees because parent-child relationships are
+important but the ordering of children at a node is
+not~\citep{kelleher2013coalescent,kelleher2014coalescent,
+kelleher2016efficient}.
+Thus, for a given gARG with nodes $\{1, \dots, n\}$ and
+edges $E$, we recover the local tree as follows.
+We begin by
+setting $\pi_u = 0$ for each $1 \leq u \leq n$. Then, for each sample
 node in $S$ we trace its path rootwards through the
@@
 $\pi_c \leftarrow p$. We then set $u \leftarrow p$, and repeat
-until either $\pi_u \neq -1$ (indicating we have traversed this section
+until either $\pi_u \neq 0$ (indicating we have traversed this section
 of the ARG already on the path from another sample) or there
 is no matching outbound edge (indicating we are at a root).
-
-The nodes in a local tree encoded
-in this way are \emph{labelled} (i.e.\ carry the identity
-of the corresponding ARG node), and this is fundamentally
-important to how a local tree relates to its ARG.
+Note that the trees for an ARG are ``sparse''~\citep{kelleher2016efficient},
+because many ancestral nodes will not be reachable from the
+samples at a given position.
+
+This combinatorial approach provides at least
+one novel insight into the relationship between ARGs and local trees.
 If we are given the sequence of local trees for a gARG
-encoded as an oriented tree (or in any other form in which
-the tree nodes are explicitly labelled with the identity of the
-corresponding ARG node),
-along with the genome interval covered by each tree,
-then we can recover precisely the same ARG. Thus, under this
-definition, there
+encoded as an oriented tree along with the genome interval
+covered by each tree, then we can recover precisely the same ARG.
+% I guess we should be a bit more formal here?
+Thus, under this definition, there
 is a one-to-one correspondence between an ARG and
@@
 sequences along the genome''.
+% Could cite some more? E.g. KwARG paper says
+% "We note that ARGs contain more information than local trees"
+% If even Ana and Jotun are wrong about this it then it must be
+% significant! ;)
 These statements that an ARG
@@
 information about those local trees that can be derived from the ARG.
-The internal nodes of evolutionary trees are usually
-considered to be \emph{unlabelled}
+
+There are two properties of how evolutionary trees are conventionally
+represented that lead to this
+% Yes, OK, we can rephrase. It is a misunderstanding though
+misunderstanding of the relationship between local trees and an ARG.
+Firstly, the internal nodes of evolutionary trees are usually
+considered to be \emph{unlabelled}.
+Defining ARG nodes as integers and using the oriented tree encoding
+explicitly labels internal nodes, and makes the relationship between
+tree and ARG nodes explicit.
+
+[WIP]
+
 because they most often correspond

### 2023-10-24 paper.tex

@@
 % statistical concept of marginal distributions.)
+The relationship between an ARG and its corresponding
+local trees is subtle and important.
 A fundamental property of genetics is that a
-% TODO is this (genome) clarification helpful or confusing?
-given DNA nucleotide is inherited from exactly one parent (genome),
+given DNA nucleotide is inherited from exactly one parent genome,
 both at an organismal and cell-by-cell level
@@
 an ARG encodes the entire ensemble of local trees along the
-genome for a given set of sample genomes.
-The relationship between an ARG and its corresponding
-local trees is subtle and important. To make the following
-discussions precise, we will define a concrete combinatorial encoding
-for local trees and an explicit procedure for constructing them.
+% Said "sample of genomes" originally, but using term "genome" in two
+% different ways in the same sentence then.
+genome for a given sample.
+% This is for Yan - to me this is a given, but maybe some people feel
+% the following is overly formal?
+Precisely defining the process by which local trees are extracted
+from an ARG is essential to our understanding of how they
+are related, and we require a concrete mathematical structure
+to describe the local trees.
 
 Oriented trees provide a
-convenient formalism to capture these relationships in a well-defined
-combinatorial object.
+convenient formalism to capture these parent-child relationships
+in a well-defined combinatorial object.
 % There's no need for a comma
@@
 Thus, for a given gARG with nodes $\{1, \dots, n\}$ and
-edges $E$, we recover the local tree as follows.
-We begin by
-setting $\pi_u = 0$ for each $1 \leq u \leq n$. Then, for each sample
-node in $S$ we trace its path rootwards through the
+edges $E$ (Section~\ref{sec-gARG}), we recover the local tree
+at position $x$ as follows.
+We begin by setting $\pi_u = 0$ for each $1 \leq u \leq n$.
+Then, for each sample node in $S$ we trace its path rootwards through the
 ARG for position $x$, and record this path in $\pi$.
@@
 is no matching outbound edge (indicating we are at a root).
-Note that the trees for an ARG are ``sparse''~\citep{kelleher2016efficient},
+Note that the local trees for an ARG are ``sparse''~\citep{kelleher2016efficient},
 because many ancestral nodes will not be reachable from the
@@
 covered by each tree, then we can recover precisely the same ARG.
-% I guess we should be a bit more formal here?
+% We do this site-by-site for simplicity, but hopefully it's obvious
+% that this isn't necessary?
+More formally, suppose we are given the local tree $\pi^x_1\dots\pi^x_n$
+for each nucleotide position $1 \leq x \leq L$ on a genome of length $L$.
+Then, the edges of the ``local ARG'' for this tree is given by
+$E^x = \{(u, \pi_u^x, \{x\} \mid \pi_u^x \neq 0\}$. Since the ARG
+edges are defined by $(c, p, I)$ tuples, where the set $I$ defines
+the positions over which node $c$ inherits from parent $p$, we can
+then simply combine the ``local ARGs'' for each position $x$
+to recover precisely the same set of edges as the original ARG.
 Thus, under this definition, there
@@
 These statements that an ARG
-contains more information than its local trees is only true
+contains more information than its local trees are only true
 if we use a representation of the local trees that discards
@@
 represented that lead to this
-% Yes, OK, we can rephrase. It is a misunderstanding though
-misunderstanding of the relationship between local trees and an ARG.
+% We can rephrase... suggestions?
+disagreement about the relationship between local trees and an ARG.
 Firstly, the internal nodes of evolutionary trees are usually
 considered to be \emph{unlabelled}.
-Defining ARG nodes as integers and using the oriented tree encoding
-explicitly labels internal nodes, and makes the relationship between
-tree and ARG nodes explicit.
-
-[WIP]
-
+Evolutionary trees usually regard internal nodes as unlabelled
 because they most often correspond
@@
 ancestors, but if we do not label the tree nodes we lose the fact that
-the \emph{same} ancestors persist across multiple trees.
-Besides having labelled nodes, the local trees must contain
-unary nodes marking recombinations
-and other processes that lead to lineages passing through a given
-ARG node without coalescence in the local trees
-(see Section~\ref{sec-locally-unary-edges}).
-If the local trees contain labelled internal nodes, they precisely
-describe the corresponding path through the graph for each
-genome interval and there is exactly as
-much information in them as in an ARG.
-
-These properties have interesting consequences when we
-consider the inference strategy of estimating local trees
-% TODO rephrase this. Ana doesn't like the way it implies that the
-% trees are independent, because info from mutations outside the
-% window are shared. However, each tree *inference* is independent,
-% whether information is used multiple times in different trees or not
-independently within genome segments,
-and subsequently stitching them together into an ARG.
-Two recent methods take this approach,
-\relate~\citep{speidel2019method} and
-\espalier~\citep{rasmussen2022espalier},
-and have different strategies for reconciling the local trees into an ARG.
-\relate\ first infers local tree topologies, and then identifies
-edges that persist from tree to tree by reasoning about the sets of samples
-subtended. It does not attempt to be exhaustive.
-\espalier, on the other hand, is specifically interested
-in the details of recombination events (among a small number of samples)
-and therefore attempts to infer
+the \emph{same} ancestors sometimes persist across multiple trees.
+Defining ARG nodes as integers and using the oriented tree encoding
+explicitly labels internal nodes, and makes the relationship between
+tree and ARG nodes clear.
+
+The second property of how evolutionary trees
+are conventionally represented that is unhelpful in the context of ARGs is their
+focus on branching points (coalescences), i.e.\
+nodes that have two or more children.
+As the introductory paragraph of this section emphasised,
+parent-child relationships are what fundamentally define a tree,
+and branching points can be seen as incidental. This is reflected
+by the oriented tree encoding which simply stores the local
+parent-child relationships, and does not, for example,
+directly tell us how many children a particular node has.
+The local tree at a given position records the \emph{path} through
+the ARG, and if we discard sections of that path that are not
+branching points in the local tree, we lose information about the ARG.
+We return to this important point in the following two sections.
+
+It is important that we make the distinction here between the local
+trees that we can derive from a known ARG (as just discussed),
+and the ARGs that we can derive from a sequence of \emph{estimated}
+local trees.
+The recent
+\espalier~\citep{rasmussen2022espalier} ARG inference method
+is illustrative in this context.
+\espalier\ works by splitting an input sequence alignment into
+segments that are assumed to be non-recombining. Within
+each segment, an initial local tree is estimated using
+standard phylogenetic methods. By necessity, these local
+trees will contain internal nodes that are unlabelled and
+consist only of branching points: there is no information
+shared between the independent tree estimation steps
+across segments. Part of the task of stitching
+these trees together
+into an ARG is then, essentially, to generate labels for
+the internal nodes, and decide which nodes persist
+across multiple local trees.
+\espalier\ is specifically interested
+in the details of recombination events,
+and also attempts to infer
 the precise subtree prune-and-regraft (SPR)
 operations~\citep{hein1990reconstructing,song2003on,song2006properties}
-induced by recombination.
+induced by recombination between these partially reconciled local trees.
 Inferring the SPRs between leaf-labelled trees is
@@
 is a degree of internal node sharing between trees.
+% I don't really know if this is true, it's just a way of wrapping the
+% section up by saying that other things we thought were settled might
+% be worth thinking about again.
+The combinatorial formulation of ARGs and local trees provided here
+may help clarify these fundamental questions.
 
@@
 \label{sec-locally-unary-edges}
-% TODO needs intro, giving some bit of warm-up for the reader.
-We can define the arity of a node as the number of child nodes it has.
-There is an
-important distinction between the number of children a node has
-in the graph and the number of children it has in the local trees,
-because a (graph) child node is only a child in the local tree for position
-% FIXME state this formally, as both Yan and Jere misunderstood here.
-$x$ if $x$ is in the inheritance interval for the corresponding edge.
-Thus, for a given node $u$ and a genome position $x$ the arity
-of $u$ in that local tree is less than or equal to the graph arity of
-$u$. The arity
-of a node in the local trees can therefore change as we move along
-the genome.
-
-This distinction between graph and local tree arity is mainly
+As discussed in the previous section, the local tree at a given position
+$x$ is best seen as the path through the ARG at that position, defined
+by the oriented tree $\pi^x_1\dots\pi^x_n$. This path does not directly
+contain information about branching points, and so it is helpful
+to define the ``arity''.
+The ``local arity'' of a node is the number of children it has
+in the local tree at position $x$, i.e., $a^x_u = |\{v : \pi^x_v = u\}|$
+for each $1 \leq u \leq n$. The ``ARG arity'' of a node $u$ is the
+number of children it has in the graph topology, i.e.,
+$a_u = |\{v: (v, u, I) \in E\}|$. Thus, the local arity is less than
+or equal to the ARG arity (more precisely, $0 \leq a^x_u \leq a_u$),
+and the local arity of a node may change as we move along the genome.
+
+This distinction between ARG and local tree arity is mainly
 of interest when we consider \emph{unary} nodes, those that have
@@
 Returning to the example in Fig~\ref{fig-arg-in-pedigree}, nodes
-\noderef{g} and \noderef{h} are ARG-unary (Fig~\ref{fig-arg-in-pedigree}B), and are consequently
+\noderef{g} and \noderef{h}
+are ARG-unary (Fig~\ref{fig-arg-in-pedigree}B), and are consequently
 also unary in the local trees (Fig~\ref{fig-arg-in-pedigree}C).
@@
 in one or more local trees, are a common and important feature
-of many different types of ARG, and a recurring topic
-in this article. Without these nodes marking the passage
+of many different types of ARG.
+As discussed in the previous section,
+without these nodes marking the passage
 of ancestral material through specific ancestors, the local trees
@@
 
-% FIXME: move to top of section
 Nodes with one child are not a standard feature of evolutionary trees
@@
 different ways in which ARG nodes can be locally unary. Furthermore,
-several current ARG inference methods infer locally unary nodes
+several inference methods infer ARGs containing locally unary nodes
 (see Section~\ref{sec-precision}).

### 2023-10-24 paper.tex

@@
 
-\section{Computational efficiency and data interchange}
+\section{Implementation and efficiency}
 \label{sec-efficiency}
-ARG inference methods have been successfully applied to vast datasets,
-including 500,000 genotyped humans~\citep{kelleher2019inferring,zhang2023biobank}
-from UK Biobank~\citep{bycroft2018genome}
-and over a million SARS-Cov-2 whole genomes~\citep{zhan2023towards}.
-Even larger datasets are
-available~\citep[e.g.][]{halldorsson2022sequences} % TODO more
-and it seems inevitable that ARGs will be inferred for them.
-The efficiency with which we can store and analyse these inferred
-ARGs is therefore of great importance, and a key factor
-in the ultimate success of downstream ARG-based methods.
-Another key factor in how important ARG-based methods become
-across the wide range of possible applications
-% TODO get rid of this, it's a duplication of the introduction
-areas~\citep{rasmussen2014genome,harris2019database,
-hejase2020summary,schaefer2021ancestral,
-fan2022genealogical,wohns2022unified,harris2023using,
-nowbandegani2023extremely,ignatieva2023distribution}
-is how well inference methods and downstream analysis applications
-interoperate. Without a well-defined and efficient shared format
-for ARG data interchange,
-users will suffer from problems with converting data from one program's
-output to another, an unfortunate hallmark of population genetics
-software for many years~\citep{excoffier2006computer}.
-While current methods tend to tightly couple downstream analysis
-of the inferred ARG with the inference
-itself within the same software package,
-this approach scales poorly in terms of software development
-effort, and is ultimately not compatible with the widespread use
-of ARGs for routine data analysis, and a healthy and diverse
-software ecosystem.
-
-The gARG encoding discussed and concretely defined in this manuscript
+The gARG encoding discussed defined here
 leads to highly efficient storage and processing of ARG data,
 and has already been in use for several years.
-The succinct tree sequence data structure (usually known as a ``tree sequence''
-for brevity)
-is a concrete encoding of a gARG, as discussed in this article.
+The succinct tree sequence data structure
+(usually known as a ``tree sequence'' for brevity)
+is a practical gARG implementation focused on efficiency.
 It was originally developed as part of the \texttt{msprime}
@@
 simple tabular representation and nodes and edges,
-and also defining a concise and lossless representation of
+and also defining a concise representation of
 sequence variation using the ``site'' and  ``mutation'' tables.
-(Technically, edges in tree sequence terminology would be better
-described as ``edge-intervals'', as each describes a single contiguous
-interval of genome inheritance between a pair of nodes. This
-denormalisation of the gARG data model in \texttt{tskit} is for efficiency purposes.)
+The key property of the succinct tree sequence encoding
+that makes it an efficient substrate for defining analysis
+algorithms is that it allows us to sequentially
+recover the local trees along the genome very efficiently,
+and in a way that allows us to reason about the \emph{differences}
+between those trees~\citep{kelleher2016efficient,ralph2020efficiently}.
+
 The \texttt{tskit} library is a liberally
@@
 fan2023likelihood}.
-
-% FIXME this needs more work and doesn't really fit into the
-% narrative right now, but we do need to make this point.
-The key insight that makes the succinct tree sequence encoding
-an efficient substrate for defining analysis algorithms is that
-it allows us to recover the local trees along the genome
-in a way that allows us to reason about the \emph{differences}
-between those trees.
-Sequentially generating the local trees along the genome
-is also fundamental, and is necessary whenever we need to
-perform calculations that are contingent on more than just the
-isolated properties of an edge. \cite{kelleher2016efficient}
-showed how all trees can be sequentially generated in
-constant time per tree transition in a fully simplified gARG.
-Furthermore, we can easily reason about how tree topologies
-change (and stay the same), leading to efficient algorithms
-for computing population genetic
-statistics~\citep{kelleher2016efficient,ralph2020efficiently},
-implementing the Li and Stephens
-model~\citep{kelleher2019inferring,wohns2022unified}
-and likely many more.
+The technical details of \texttt{tskit}, and how it provides an
+efficient and portable platform for ARG-based analysis, are beyond
+the scope of this manuscript.
+In the interest of avoiding confusion, however, we list a
+few minor details in which the formal details of gARGs
+provided here differ from their practical implementation in
+\texttt{tskit}.
+Firstly, ``edges'' in tree sequence terminology would be perhaps better be
+described as ``edge-intervals'', as each describes a single contiguous
+interval of genome inheritance between a pair of nodes.
+This denormalisation of the gARG data model in is for efficiency purposes.
+Secondly, $-1$ is used to denote the presence of
+roots in an oriented tree, rather than $0$ as used here for notational
+simplicity.
+
+% Other representations of ARGs, beyond straightforward implementations
+% of a Griffiths-style
+% eARG~\citep[e.g.][]{ignatieva2021kwarg,rasmussen2014genome}
+% and \texttt{tskit}'s gARG format, are of course possible.
+% The ARGON simulator~\citep{palamara2016argon}, for example,
+% stores ARGs
+% ARG-Needle~\citep{zhan2023towards}, for example, uses a format based
+% on hap
 
@@
 For instance, several human datasets now consist of hundreds of thousands of
-genomes~\citep{turnbull2018hundred, bycroft2018genome,karczewski2020mutational,tanjo2021practical},
+genomes~\citep{turnbull2018hundred, bycroft2018genome,
+karczewski2020mutational,tanjo2021practical,
+halldorsson2022sequences},
 and so sample size is substantially \emph{larger} than the
@@
 on data formats, however, the ARG revolution may falter.
-Here, we have advocated
-for decoupling the data structures used to represent ARGs
-from generative processes, and have shown that the
-gARG encoding enables flexible description of
-contemporary datasets with varying levels of precision and integral
-handling of some important forms of temporal uncertainty.
 For ARG-based methods to achieve mainstream status, we require
@@
 analyses on these inferences.
-Earlier efforts to standardise ARG interchange shared
-this vision,
+Earlier efforts to standardise ARG interchange shared this vision,
 but did not succeed~\citep{cardona2008extended,mcgill2013graphml}.
-Today, however, we have the benefit of a pre-existing ARG
-interchange format, already widely used and accompanied
-by high-quality, high-performance software
-(see Section~\ref{sec-efficiency} for details).
-Adopting the \texttt{tskit} library
-as a community standard,
-utilising its rich suite of operations,
+Without a well-defined and efficient shared format
+for ARG data interchange,
+users will suffer from problems with converting data from one program's
+output to another, an unfortunate hallmark of population genetics
+software for many years~\citep{excoffier2006computer}.
+Current methods tend to tightly couple downstream analysis
+of the inferred ARG with the inference
+itself within the same software package,
+which is ultimately not compatible with the widespread use
+of ARGs for routine data analysis, and a healthy and diverse software ecosystem.
+The gARG encoding described here is a significant generalisation
+of classical concepts, capable of describing even the bewildering complexity
+of contemporary datasets and describing important forms of temporal uncertainty.
+The \texttt{tskit} library is a high-quality open-source gARG implementation,
+with proven efficiency and
+scalability~\citep[e.g.][]{anderson2022genes,zhan2023towards},
+that is already in widespread use.
+Adopting it as a community standard,
+and utilising its rich suite of operations,
 may accelerate progress and help to

### 2023-10-25 paper.tex
Tweaks

@@
 \newcommand{\espalier}[0]{\texttt{Espalier}}
+\newcommand{\arbores}[0]{\texttt{Arbores}}
 
@@
 In order to capture the effects of recombination
-(and other processes that lead to the partial genome inheritance)
+(and other processes that lead to partial genome inheritance)
  we ``annotate'' the edges with the genome
@@
 We can define a gARG formally as follows.
-Let $N \subset \mathbb{N}$ be the set of nodes representing
-the genomes in the gARG, $S \subseteq N$ be the set of sampled genomes.
-% (but see Section~\ref{sec-retro-pro} for some subtlety around this point).
+Let $N = \{1, \dots, n\}$ be the set of nodes representing
+the genomes in the gARG, 
+and  $S \subseteq N$ be the set of sampled genomes.
 Then, $E$ is the set of edges, where each element
@@
 with which arbitrary additional information can be associated.
+% We can delete then next two sentances if we want, but nodes are
+% integers and this is a vital element of the model.
 Concretely defining nodes as integers in this way has some
@@
 local trees.
-The recent
-\espalier~\citep{rasmussen2022espalier} ARG inference method
-is illustrative in this context.
-\espalier\ works by splitting an input sequence alignment into
+The ARG inference methods \arbores~\citep{heine2018bridging} and
+\espalier~\citep{rasmussen2022espalier} are illustrative in this context.
+Both begin by splitting an input sequence alignment into
 segments that are assumed to be non-recombining. Within
@@
 the internal nodes, and decide which nodes persist
-across multiple local trees.
+across multiple local trees. \arbores~bridges the gaps 
+between the inferred local trees via a coalescent-based 
+sampling scheme.
 \espalier\ is specifically interested
@@
 interval of genome inheritance between a pair of nodes.
-This denormalisation of the gARG data model in is for efficiency purposes.
+This denormalisation of the gARG data model is for efficiency purposes.
 Secondly, $-1$ is used to denote the presence of

### 2023-10-25 paper.tex
minor tidyup of some appendices

@@
 as originally described requires exponential time to simulate
-(see Appendix~\ref{sec-big-and-little-arg} for details). However, ARGs provided a way
+(see Appendix~\ref{sec-big-and-little-arg} for details).
+However, ARGs provided a way
 to reason about and infer recombinant ancestry as a single object,
-which was not possible within Hudson's framework, which emphasised
-the collection of local trees along the genome resulting from recombination.
-
-Subsequent work on ARGs proceeded in broadly three main directions: focussing on (1)
-exploring the mathematical properties of the coalescent with recombination and
-related stochastic processes, (2) inferring evolutionary parameters under
+in a way that not possible within Hudson's framework, which emphasised
+instead the collection of local trees along the genome
+resulting from recombination.
+
+Subsequent work on ARGs proceeded in broadly three main directions:
+(1) exploring the mathematical properties of the coalescent with recombination and
+related stochastic processes;
+(2) inferring evolutionary parameters under
 (approximations to) this model, either with or without explicitly reconstructing the
-genealogy of the sample, and (3) treating the ARG as a discrete graph, ignoring the
+genealogy of the sample;
+and (3) treating the ARG as a discrete graph, ignoring the
 generating stochastic process, and studying its properties from a computational and
@@
 or moving away from the coalescent with recombination altogether and seeking
-to infer a single plausible ARG~\citep[e.g.][]{speidel2019method} or ARG
-topology~\citep[e.g.][]{minichiello2006mapping,kelleher2019inferring}.
+to infer a single plausible ARG~\citep[e.g.][]{
+minichiello2006mapping,kelleher2019inferring,speidel2019method}.
 
@@
 process~\citep[e.g.][]{gusfield2014recombinatorics,mathieson2020ancestry,brandt2021evaluation}.
-\cite{minichiello2006mapping} explicitly distinguished the ARG as a
-stochastic process from the ARG as a data structure. We have followed them and
-used the term in the latter sense, to mean a specific realised ancestral history.
 
@@
 \label{sec-big-and-little-arg}
-Here we review the two predominant stochastic processes which construct ARGs:
-the ``big" ARG process of \cite{griffiths1997ancestral}, and the ``little" ARG process of
+Here we review two important stochastic processes that construct ARGs:
+the ``Big" ARG process of \cite{griffiths1997ancestral}, and the ``Little" ARG process of
  \cite{hudson1983properties}. The Big ARG process is mathematically simpler
@@
 recombination events which occur outside of ancestral material,
-and this do not have any bearing on the ancestry of the initial sample.
+and these do not have any bearing on the ancestry of the initial sample.
 Because we don't keep track of the distribution of ancestral material during the process,
@@
 
-Figure \ref{hudson_vs_bigARG} illustrates the more complex state space
-of the Little ARG process, as well as the extra events which occur in the Big ARG process.
-Moreover, it depicts the rates of common ancestors and recombination events in each
-interval of time of the realisations.
-In order to evaluate these rates, and hence the sampling probability
-\cite[see e.g.][Equation (3)]{mahmoudi2022bayesian},
-it is necessary to know the number of lineages and number of extant links
-available for recombination in each time interval.
-Some representations may not provide this information.
-For example, in the gARG encoding depicted in Figure \ref{fig-ancestry-resolution}C
-it is clear that a recombination takes place between nodes \noderef{b} and
-\noderef{d} as well as \noderef{e}, but the exact time of the event is ambiguous,
-and thus so is the number of lineages during the time interval. In fact, this
-information cannot be recovered from the gARG encoding used in Figure \ref{fig-ancestry-resolution}C,
-and alternative conventions must be introduced to resolve such ambiguities.
-For the likelihood-based inference algorithms for the coalescent with recombination,
-it is sufficient to create \emph{two} gARG nodes at the time of the recombination event:
-see the appendix of \cite{baumdicker2021efficient} for details of evaluating
-coalescent with recombination likelihoods using this convention.
-This is also the interpretation depicted in
-Figure \ref{hudson_vs_bigARG}, but it means that the two edges above node \noderef{b}
-in Figure \ref{fig-ancestry-resolution} should correspond to only one lineage,
-along which all $m-1$ links are available for recombination.
-The lineage then splits into two at the time of nodes \noderef{d} and \noderef{e}.
-Nodes \noderef{k}, \noderef{l}, and \noderef{m} in Figure \ref{fig-ancestry-resolution}
-demonstrate that the same issue can affect the number of links available for recombination:
-without an external convention, the exact time at which the trapped ancestral material on
-node \noderef{k} ceases to be available for an effective recombination in the Little ARG process.
-
-\begin{figure}[t]
+\begin{figure}
 \centering
@@
 
+Figure \ref{hudson_vs_bigARG} illustrates the more complex state space
+of the Little ARG process, as well as the extra events which occur in the Big ARG process.
+Moreover, it depicts the rates of common ancestors and recombination events in each
+interval of time of the realisations.
+In order to evaluate these rates, and hence the sampling probability
+\cite[see e.g.][Equation (3)]{mahmoudi2022bayesian},
+it is necessary to know the number of lineages and number of extant links
+available for recombination in each time interval.
+Some representations may not provide this information.
+For example, in the gARG encoding depicted in Figure \ref{fig-ancestry-resolution}C
+it is clear that a recombination takes place between nodes \noderef{b} and
+\noderef{d} as well as \noderef{e}, but the exact time of the event is ambiguous,
+and thus so is the number of lineages during the time interval. In fact, this
+information cannot be recovered from the gARG encoding used in Figure \ref{fig-ancestry-resolution}C,
+and alternative conventions must be introduced to resolve such ambiguities.
+For the likelihood-based inference algorithms for the coalescent with recombination,
+it is sufficient to create \emph{two} gARG nodes at the time of the recombination event:
+see the appendix of \cite{baumdicker2021efficient} for details of evaluating
+coalescent with recombination likelihoods using this convention.
+This is also the interpretation depicted in
+Figure \ref{hudson_vs_bigARG}, but it means that the two edges above node \noderef{b}
+in Figure \ref{fig-ancestry-resolution} should correspond to only one lineage,
+along which all $m-1$ links are available for recombination.
+The lineage then splits into two at the time of nodes \noderef{d} and \noderef{e}.
+Nodes \noderef{k}, \noderef{l}, and \noderef{m} in Figure \ref{fig-ancestry-resolution}
+demonstrate that the same issue can affect the number of links available for recombination:
+without an external convention, the exact time at which the trapped ancestral material on
+node \noderef{k} ceases to be available for an effective recombination in the Little ARG process.
+
 % Full quote:
@@
 finding \emph{parsimonious} ARGs, i.e.\ those with a minimal number of
-recombination events \citep{hein1990reconstructing}. Two main approaches
-emerged: \emph{backwards-in-time} \citep{lyngso2005minimum} and
-\emph{along-the-genome} \citep{song2003parsimonious, song2005constructing}. The
-former starts with a data matrix and reduces it to an empty matrix through row
+recombination events~\citep{hein1990reconstructing}. Two main approaches
+emerged: ``backwards-in-time''~\citep{lyngso2005minimum} and
+``along-the-genome'' \citep{song2003parsimonious, song2005constructing}.
+Backwards-in-time approaches
+start with a data matrix and reduce it to an empty matrix through row
 and column operations corresponding to coalescence, mutation, and recombination
-events, which construct an ARG from the bottom up. The latter begins from an
-initial local tree at a single focal site. Moving the focal site along the
-genome changes the local tree via a \emph{subtree prune and regraft} (SPR)
-operation whenever a recombination is encountered. Methods based the
-backwards-in-time approach are described by \citet{song2005efficient,
-wu2008association, thao2019hybrid, ignatieva2021kwarg}, while
-\citet{hein1993heuristic, wu2011new, mirzaei2017rent} describe along-the-genome
-methods. \citet{rasmussen2022espalier} focuses on parsimonious fusion of local
-trees into an ARG, while \citet{camara2016inference} is based on topological
-data analysis.
-
-Reconstructing a most parsimonious ARG for a given data set is NP-hard
+events, which construct an ARG from the bottom up~\citep{song2005efficient,
+wu2008association,thao2019hybrid,ignatieva2021kwarg}.
+Along-the-genome approaches
+begin from an initial local tree at a single focal site.
+Moving the focal site along the
+genome changes the local tree via a subtree prune and regraft
+operation whenever a recombination is encountered~\citep{hein1993heuristic,
+wu2011new, mirzaei2017rent}.
+\citet{rasmussen2022espalier} focus on parsimonious fusion of local
+trees into an ARG, while the method described by
+\citet{camara2016inference} is based on topological data analysis.
+% The "most" is understood here, surely
+Reconstructing a parsimonious ARG for a given data set is NP-hard
 \citep{wang2001perfect}, so parsimony-based methods resort to heuristics and
@@
 methods aim to balance computational efficiency with reconstruction of
-``reasonable'', rather than parsimonious ARGs or sampling from a
-distribution under a population genetics model
+``reasonable'', rather than parsimonious ARGs
+% distribution under a population genetics model
 \citep{minichiello2006mapping,
@@
 
-% \begin{table}[ht]
-%     \begin{center}
-%         \begin{tabular}{c|ccc|c|c}
-% 			\multicolumn{2}{c}{Node table}& ~ &\multicolumn{3}{c}{Edge table}\\
-% 			\cline{1-2} \cline{4-6}
-%             Node & Time & ~ & Child & Parent & Intervals\\
-%             \cline{1-2} \cline{4-6}
-%             $\noderef{a}$ & 0 & ~ & $\noderef{a}$ & $\noderef{e}$ & $\{[0,2)\}$\\
-%             $\noderef{b}$ & 0 & ~ & $\noderef{a}$ & $\noderef{f}$ & $\{[2,10)\}$\\
-%             $\noderef{c}$ & 0 & ~ & $\noderef{b}$ & $\noderef{g}$ & $\{[0,10)\}$\\
-%             $\noderef{d}$ & 0 & ~ & $\noderef{c}$ & $\noderef{f}$ & $\{[0,7)\}$\\
-%             $\noderef{e}$ & 1 & ~ & $\noderef{c}$ & $\noderef{e}$ & $\{[7,10)\}$\\
-%             $\noderef{f}$ & 1 & ~ & $\noderef{d}$ & $\noderef{h}$ & $\{[0,10)\}$\\
-%             $\noderef{g}$ & 1 & ~ & $\noderef{e}$ & $\noderef{i}$ & $\{[0,10)\}$\\
-%             $\noderef{i}$ & 1 & ~ & $\noderef{f}$ & $\noderef{k}$ & $\{[0,10)\}$\\
-%             $\noderef{k}$ & 2 & ~ & $\noderef{g}$ & $\noderef{i}$ & $\{[0,10)\}$\\
-%             $\noderef{n}$ & 3 & ~ & $\noderef{h}$ & $\noderef{k}$ & $\{[0,10)\}$\\
-%             \multicolumn{2}{c}{~}& ~ & $\noderef{i}$ & $\noderef{n}$ & $\{[0,10)\}$\\
-%             \multicolumn{2}{c}{~}& ~ & $\noderef{k}$ & $\noderef{n}$ & $\{[0,10)\}$
-%         \end{tabular}
-%     \end{center}
-%     % NOTE: I know  the "correct" placement for table captions is on top, but I
-%     % always find this quite weird (since many of our figs are quite "table
-%     % like" anyway. If anyone feels strongly about putting the caption above
-%     % the table please just move it back.
-%     \caption{\label{tab-gARG-data}
-%     Data encoding of the genome ARG shown in Fig.~\ref{fig-arg-in-pedigree} with the
-% 	node table on the left (having node and time columns) and the edge table on the
-% 	right (having child, parent, and intervals columns).
-%     }
-% \end{table}
-
-\begin{figure}
+\begin{figure}[ht]
 	\begin{center}

### 2023-10-25 paper.tex

@@
 
-% Other representations of ARGs, beyond straightforward implementations
-% of a Griffiths-style
-% eARG~\citep[e.g.][]{ignatieva2021kwarg,rasmussen2014genome}
-% and \texttt{tskit}'s gARG format, are of course possible.
-% The ARGON simulator~\citep{palamara2016argon}, for example,
-% stores ARGs
-% ARG-Needle~\citep{zhan2023towards}, for example, uses a format based
-% on hap
-
 % \cite{shipilina2023origin} discuss the idea of a ``haplotype block''
@@
 feasible in practice, leading to a surge of interest
-in development of ARG-based inference methods, their evaluation
-and application.
+in inference methods, their evaluation, and their application to biological questions.
 The prospect of ARGs being used routinely within population
@@
 The flexibility of the gARG encoding contrasts with the classical
-``event'' ARG (eARG), which is more limited in what can be described,
-because of its origins in an idealised mathematical model,
-and the assumption of having complete information about
-recombination events.
+``event'' ARG (eARG) encoding, which is more limited in what can be described.
+Importantly, gARGs do not require fully precise estimates of
+ancestral recombination events,
+and allow us to directly express important forms of temporal uncertainty.
 
 Fully decoupling the general concept of an ARG from the coalescent
-with recombination stochastic process is an important step.
-While this has proven to be a useful and
+with recombination stochastic process (henceforth, ``coalescent'')
+is an important step.
+While the coalescent has proven to be a useful and
 robust
@@
 violate its assumptions.
-For instance, several human datasets now consist of hundreds of thousands of
+One key assumption is that
+sample size $n$ is much less than the effective population size, $N_e$.
+Several human datasets now consist of hundreds of thousands of
 genomes~\citep{turnbull2018hundred, bycroft2018genome,
@@
 halldorsson2022sequences},
-and so sample size is substantially \emph{larger} than the
+and so sample size is an order of magnitude \emph{larger} than the
 usually assumed $N_e$ values.
-Agricultural datasets are an even more extreme departure from coalescent
-assumptions, with hundreds of thousands of samples embedded in
+Agricultural datasets are an even more extreme departure from this
+assumption, with hundreds of thousands of samples embedded in
 multi-generational pedigrees~\citep{hayes20191000,Ros-Freixedes2020}
@@
 less~\citep{MacLeod2013,Makanjuola2020,Hall2016,Porcnic2016}.
-Even if sufficiently complex demographic models~\citep{gower2022demes}
-encompassing hundreds of populations, explosive growth rates and
-myriad interconnections of migration, were somehow estimated and
-provided as input, ARGs sampled from
-the coalescent cannot capture the complexities of realistic family structures, which often involve overlapping generations and extreme variation in lifetime reproductive success.
-Flexible ARG encoding should accommodate such cases.
-
-Given that there is no single model that can capture the complexities
-of modern datasets across species (and taxa; see discussion of SARS-CoV-2
-below), we cannot rely on sampling from a distribution to capture
-the uncertainty in our ARG estimates. Indeed, even if the coalescent
-were a suitable model for datasets such as UKB and we had a means of
-sampling from its distribution of ARGs at the scale of hundreds of thousands
-of samples, we could only explore the tiniest corner of such an incomprehensibly
-large space of possibilities.
+A model assuming a single $N_e$ would be a
+drastic over-simplification of course, but
+even if sufficiently complex demographic models~\citep{gower2022demes}
+encompassing hundreds of populations, explosive growth rates and myriad
+interconnections of migration, were somehow estimated and provided as input,
+ARGs sampled from the coalescent cannot capture the complexities
+of family structure in these
+datasets~\citep[e.g.][]{turnbull2018hundred,Ros-Freixedes2020}.
+Another core assumption of the coalescent model is that the genome
+(or at least the region under study) is short enough that the number of extant
+lineages remains much smaller than $N_e$ at all times.
+High-quality whole genome assemblies are now available
+for many species % Can we say something more concrete here?
+and projects are under way to obtain
+them for tens of thousands more~\citep{darwin2022sequence,lewin2022earth},
+and so we can expect inferred ARGs to routinely span whole chromosomes.
+
+Recent large-scale methods have simplified the inference problem by
+making a single, deterministic best-guess at ARG
+inference~\citep{kelleher2019inferring,speidel2019method,zhan2023towards}.
+Even under strict parsimony conditions and for small sample sizes, the
+% Citation? This is something people have figured out, right?
+number of plausible ARGs compatible with a given dataset is vast.
+Thus, although it is clearly an oversimplification to arbitrarily
+choose \emph{one} best guess, it is not clear that sampling more
+when sample sizes are large will achieve much.
+At the scale of millions of samples,
+we could only ever explore the tiniest corner of the incomprehensibly
+large space of likely ARGs.
 Therefore, it is important to systematically
-describe and utilise uncertainty about ARG inference.
+describe and utilise uncertainty about ARG inference, and to
+incorporates uncertainty encountered during inference into the returned ARG.
 One approach, enabled by the gARG encoding described here, is to allow
@@
 uncertainty about node ages and recombination breakpoint positions, is an important
-aspect of future work.
+aspect of future work. Fully utilising such uncertainty in downstream
+applications is also vital, of course.
 
 The timing, positions, and even the number of recombination events is generally
-not possible to infer precisely from genome sequencing data. For ARGs under
+not possible to infer precisely from genome sequencing data. Under
 coalescent-based models, the proportion of recombination events that change the
@@
 position can only be inferred approximately, depending on how much information
-can be elucidated from mutations in the surrounding genomic region. For general
-ARGs, for instance those representing pedigrees, or genealogies of organisms with
-low mutation rates and complex patterns of recombination, such inference can be
-even more challenging. The fact that the eARG encoding \emph{requires}
-precise information about recombination is therefore a fundamental limitation, which
-is a central message of this paper.
-% This might be where we mention Fisher's junctions and IBD?
-% https://github.com/tskit-dev/what-is-an-arg-paper/issues/121
-
+can be elucidated from mutations in the surrounding genomic region.
+% For general
+% ARGs, for instance those representing pedigrees, or genealogies of organisms with
+% low mutation rates and complex patterns of recombination, such inference can be
+% even more challenging.
+The fact that the eARG encoding \emph{requires}
+precise information about recombination is therefore a fundamental limitation.
 % % Do downstream applications actually make use of such full ARGs?
@@
 for downstream applications.
-% To date, few make use such detailed recombination information.
-Most applications work by examining local trees independently.
+Many applications work by examining local trees independently.
 For example, the \relate\ selection test~\citep{speidel2019method}
@@
 type of machine learning algorithm
-that takes ``temporal'' correlations into account. Thus,
-while SIA does use some of the information about local tree correlation,
-clearly much of the detail about recombination
-events in an ARG is lost.
+that takes ``temporal'' correlations into account.
+Thus, while SIA takes advantage of information about local tree correlation,
+it is in quite an indirect way, and
+clearly much of the detail about recombination events in an ARG is lost.
 The \texttt{tsdate} algorithm (and related naive estimator
@@
 
-% Note: haven't said anything about the coalescent here. Do we need to?
 The advantages of a model-agnostic representation that naturally
-incorporates uncertainty about the ordering of events in an ARG are well-illustrated by~\cite{zhan2023towards}, who
+incorporates uncertainty about the ordering of events in an ARG
+are well-illustrated by~\cite{zhan2023towards}, who
 inferred ARGs using millions of SARS-CoV-2 sequences from the
-GISAID database~\citep{shu2017gisaid}. In contrast to typical human sequencing datasets, the SARS-CoV-2 data is
+GISAID database~\citep{shu2017gisaid}.
+In contrast to typical human sequencing datasets, the SARS-CoV-2 data is
 sampled continuously through time, sometimes with tens of thousands of sequences collected per day, with relatively little genetic diversity to
@@
 also present in the dataset) and unary nodes (samples that have
-one other sampled descendant). Recombination is known
-to be an important factor in the evolution of SARS-CoV-2~\citep{vaninsberghe2021recombinant,jackson2021generation,ignatieva2022ongoing},
-and the ARGs also contain recombination nodes.
+one direct descendant).
+Recombination is an important factor in the evolution of
+SARS-CoV-2~\citep{vaninsberghe2021recombinant,jackson2021generation,ignatieva2022ongoing},
+% Tried to jazz this up a bit, was quite dry
+and the inferred ARGs contain an unprecedented level of detail about
+the combined processes of viral mutation and recombination.
 Because parental sequences
@@
 of each part of the recombinant sequence without arbitrarily assigning
-times or orderings for these events. This demonstrates the flexibility of the gARG
-encoding in capturing recombinant genealogies regardless of the underlying
-(possibly unknowable) generative evolutionary process.
+times or orderings for these events.
 
@@
 In comparing tree-by-tree along the genome, the effects of recombination
-are only incorporated in an indirect manner through the correlations
+are incorporated in quite an indirect manner through the correlations
 between the local trees, rather than through directly taking into account
@@
 % https://bmcgenomics.biomedcentral.com/articles/10.1186/s12864-020-07011-0
-Tree distance metrics usually have at least $O(n^2)$ time complexity
+Tree distance metrics often have at least $O(n^2)$ time complexity
 and therefore cannot be
@@
 A recent trend has been to move away from such
-tree distance-based approaches and to examine more properties of the inferred ARGs.
-\citet{brandt2021evaluation} focussed on the
-distributions of pairwise MRCA times;
-\citet{deng2021distribution} derived an approximation to the distribution of the
-waiting distances between local trees;
-\citet{ignatieva2023distribution}
-derived the distribution of the genomic span of an edge and that of a
-clade of samples. In each case, simulation studies demonstrated
+tree distance-based approaches and to examine more
+properties of the inferred ARGs, such as
+distributions of pairwise MRCA times~\citep{brandt2021evaluation},
+waiting distances between local trees~\citep{deng2021distribution},
+and the genomic span of an edge or clade of
+samples~\citep{ignatieva2023distribution}.
+In each case, simulation studies demonstrated
 substantial differences between these quantities in simulated and reconstructed ARGs,
 that were not captured using tree-by-tree comparisons.
-% Not sure this is the right place for this sentence
-These evaluations have been based on
-ground truth data from highly idealised simulations, with sample sizes limited to at most a few thousand (typically much fewer).
-The effects of the richness of real data
-on modern biobank-scale or multi-generational pedigree datasets
-are almost entirely unknown (beyond studies using very simplistic error models).
+Evaluations to-date have almost all been based
+on ground truth data from highly idealised simulations,
+with sample sizes limited to at most a few thousand (typically much fewer).
+Beyond the effects of very simplistic error models~\cite[e.g.][]{kelleher2019inferring},
+the effects of the richness of real data
+at biobank-scale on ARG inference are almost entirely unknown.
 The development of ARG evaluation metrics that take into account more of the
@@
 % get good local trees
-specific inference goals, sharing a common ARG interchange format.
+targetting particular inference goals, sharing a common ARG interchange format.
 The output of these diverse inference methods could then be
@@
 of classical concepts, capable of describing even the bewildering complexity
-of contemporary datasets and describing important forms of temporal uncertainty.
+of contemporary datasets and capturing important forms of temporal uncertainty.
 The \texttt{tskit} library is a high-quality open-source gARG implementation,

### 2023-10-25 paper.tex
Pass through intro

@@
 is usually referred to as an ancestral recombination graph (ARG),
+% FIXME this sentance needs to change. We want to capture more of the
+% idea of the formalisms falling behind actual practise and the
+% confusion this is causing
 although there is some confusion about the interpretation and little
@@
 schaefer2021ancestral,wohns2022unified,zhang2023biobank,zhan2023towards}
-% and data representation~\citep{kelleher2016efficient,kelleher2018efficient}
 have raised the realistic prospect of ARG-based analysis becoming a standard part
@@
 % Para 2: What are they, though?
-Although it is widely accepted that ARGs are important, there is significant
+Although it is widely accepted that ARGs are important, there is some
 confusion about what, precisely, an ARG \emph{is}.
-Originally, ARGs were defined by Griffiths and colleagues as an alternative
+Originally, Griffiths and colleagues defined the ARG as an alternative
 formulation of the coalescent with recombination~\citep{hudson1983properties},
@@
 among ancestral lineages is formalised as a
-% Note: please don't add "mathematical graph". If a reader doesn't know what a
-% graph is they're not going to understand this paper.
 graph~\citep{griffiths1991two,ethier1990two,griffiths1996ancestral,griffiths1997ancestral}.
-Subsequently, an ARG has come to be thought of as a data
+Subsequently, ARGs have come to be thought of as a data
 structure~\citep{minichiello2006mapping}, i.e.\ describing
-the \emph{outcome} of such a random process or an
-% Note on earlier version: introducing a different term for ARG here is just confusing
+a \emph{realisation} of such a random process or an
 inference from a sample of genome sequences.
@@
 kelleher2019inferring, speidel2019method, baumdicker2021efficient, zhang2023biobank}.
-This broad usage seems a useful generalisation,
-but is not compatible with the usual concrete definition of an ARG.
-This mismatch has led to a range of problems that, if unchecked,
+This broad usage is a useful generalisation,
+but is not, however, compatible with the available formal definitions
+derived from the original Griffiths approach.
+This mismatch between usage and definition has led to significant
+problems, which if not addressed,
 may impede progress in applying ARGs to understand the biological world.
 
-% Para 3: what is the problem?
-% FIXME current version isn't great, but it's better than previous effort.
-[FIXME this paragrph needs a substantial update, need to reread the
-eARG and gARG sections again to be clear what we say we're doing.
-We should focus on the concrete interchange aspects rather than
-coalescenty type things.]
-The mismatch between the classical Griffiths definition of an ARG and the
-modern interpretation is the source of substantial confusion,
-and, we argue, a significant limiting factor for the field.
-% actual outputs of these modern methods is a central issue.
-As this definition is essentially a reformulation of the
-coalescent with recombination (henceforth ``the coalescent''; we
-will assume the incorporation of recombination for the purposes of
-this article) in graph-theoretic form,
-the distinction between the simplifying assumptions of the model
-and the necessary requirements for a data structure are not clear.
-The consequences of conflating model with data structure
-are subtle, varied, and deep.
-% TODO rephrase, not great currently
-We argue that the prevailing definition of an ARG
-imposes unnecessary limitations, and that with a broader definition
-we can reassess what we should be attempting to estimate,
-how we measure success, and the expected limitations on the precision
-and accuracy of our inferences.
-
-% What do we do?
-In this paper, we attempt to address these problems by
-providing a precise definition of an ARG that is general enough to encompass
-the output of popular simulation and inference methods.
-We propose a definition of genetic genealogy in terms of specific genomes
-and their intervals of genetic inheritance, and refer to this as
-a ``genome ARG'', or ``gARG''.
-This contrasts with the classical Griffiths approach,
-based on recording evolutionary events, which we call
-an ``event ARG'', or ``eARG''.
-We show that the gARG encoding has a range of advantages over the
-eARG approach, being flexible, efficient and, in particular,
-allowing us to systematically encode uncertainty about
-the precise details of recombination events.
-
-% TODO: Provide a map of the paper's sections and how they tie together in support of gARG as a way forward.
-% Here is a rough draft - it would be good to state the most important message (in one sentence or so) from each of the sections to help with frontloading key messages
-% Genome ARGs &
-% Event ARGs
-In the following, we first define the genome ARG encoding
-and contrast it with the event ARG approach.
-We then discuss the important concept of ancestral material in
-Section~\ref{sec-ancestral-material}
-and [DO SOME STUFF].
-% Converting and eARG to gARG &
-% ARGs and local trees &
-% Locally unary nodes &
-% Levels of simplification
-This simplification is closely related to operations needed to convert eARGs to
-various forms of gARGs,
-which opens a rich set of details about ARGs and local trees, local unary nodes, and
-levels of simplification.
-% Precision of recombination information
-We then highlight the important difference between the required level of precision
-about recombination information for eARGs compared to gARGs.
-% Computational efficiency and data interchange
-The way this recombination information is stored in eARGs and gARGs has important
-implications for computational efficiency, which is critical if we want to use ARGs
-for modern genomic datasets.
-% Discussion
-We wrap with a discussion highlighting that WHAT-IS-THE-KEY-SINGLE-MESSAGE-NOT-COVERED-ABOVE?
-Finally, the literature on ARGs and associated aspects is vast, particularly
-due to different definitions and implementations. We cover some important aspects
+% % Para 3: what is the problem?
+The classical formal definition of an ARG is expressed in terms of
+genetic inheritance events: the coalescence and recombination events
+that occur in the history of a sample. We show that this approach
+to representing ARGs (which we refer to as the ``Event ARG'', or eARG
+encoding) is inherently limiting, insufficient to represent
+the complexities of modern datasets, and incompatible with
+the outputs of modern large-scale simulation and inference
+methods~\citep{kelleher2016efficient,speidel2019method,
+kelleher2019inferring,zhang2023biobank}.
+We address these problems by providing a more flexible formulation
+of an ARG data structure, which we refer to as the ``Genome ARG''
+or gARG encoding. We show that it is fully compatible with the
+classical eARG approach, while providing much more flexibility
+in the types of genetic inheritance structures that can be encompassed,
+including the outputs of modern simulation and inference methods.
+In particular, we show that in contrast to the eARG encoding which
+requires fully precise estimates of each recombination event,
+the gARG encoding allows us to systematically describe uncertainty
+the temporal ordering of multiple recombinations.
+
+% This is nice to have, but probably won't work in the final version
+% without section numbers. Let's deal with that when we get there I guess.
+We begin in Section~\ref{sec-gARG} by providing a precise formal definition
+of a gARG, illustrated by an example ARG embedded in pedigree.
+We then provide a similar definition of the classical eARG
+approach in Section~\ref{sec-eARG}, and consider its limitations
+in the context of contemporary datasets.
+We then discuss the import concept of ``ancestral material''
+in Section~\ref{sec-ancestral-material}, and how this relates
+to the process of converting an eARG to a gARG.
+Using this new definition for the remainder of the paper,
+we continue
+in Section\ref{sec-ARG-and-local-trees} by considering the relationship
+between an ARG and its local trees and show (contrary to the
+% I want to put a (suitably encoded) here somewhere, but the sentence is
+% very long already.
+prevailing view) that the sequence of local trees contains
+precisely as much information as the corresponding ARG.
+The gARG view also opens a rich new set of details about ARGs,
+including the ideas of locally unary nodes
+(Section~\ref{sec-locally-unary-edges}),
+the levels of detail that can be represented in an ARG
+(Section~\ref{sec-ARG-simplification}),
+and (relatedly) the degrees of precision about recombination
+that can be stored (Section~\ref{sec-precision}).
+These ideas have important practical considerations,
+which we illustrate by examining the qualitative properties of the
+ARGs inferred by four recent methods for a classical benchmark
+dataset in Section~\ref{sec-example-inferred-args}.
+We then discuss how the gARG framework can be
+efficiently implementated in Section~\ref{sec-efficiency},
+and finish with an assessment of the key challenges facing
+the field in the Discussion.
+Finally, the literature on ARGs and associated aspects is large and
+confusing, and we cover some important aspects
 in appendices, including
 a brief history of ancestral graphs (Appendix~\ref{sec-arg-history}),
-description of the Big and Little ARGs (Appendix~\ref{sec-big-and-little-arg}),
-survey of ARG inference methods (Appendix~\ref{sec-survey-arg-infer}), and
-ARGs at an individual vs cell lineage level (Appendix~\ref{sec-cell-lineages-and-args}).
+a description of the Big and Little ARG stochastic processes
+ (Appendix~\ref{sec-big-and-little-arg}),
+a survey of ARG inference methods (Appendix~\ref{sec-survey-arg-infer}),
+and a discussion of ARGs at an individual vs cell
+lineage level (Appendix~\ref{sec-cell-lineages-and-args}).
 
@@
 for each $1 \leq u \leq n$. The ``ARG arity'' of a node $u$ is the
-number of children it has in the graph topology, i.e.,
+number of children it has in the graph topology, i.e.\
 $a_u = |\{v: (v, u, I) \in E\}|$. Thus, the local arity is less than

### 2023-10-26 paper.tex
edits from JEre
tweaks

@@
 local trees.
-The ARG inference methods \arbores~\citep{heine2018bridging} and
-\espalier~\citep{rasmussen2022espalier} are illustrative in this context.
-Both begin by splitting an input sequence alignment into
+The ARG inference method
+\espalier~\citep{rasmussen2022espalier} is illustrative in this context.
+It begins by splitting an input sequence alignment into
 segments that are assumed to be non-recombining. Within
@@
 the internal nodes, and decide which nodes persist
-across multiple local trees. \arbores~bridges the gaps
-between the inferred local trees via a coalescent-based
-sampling scheme.
+across multiple local trees. 
+% TODO add a sentence here saying how it does this. 
+% that's what we're interested in illustrating here.
 \espalier\ is specifically interested
@@
 induced by recombination between these partially reconciled local trees.
+A similar problem is tackled by \arbores~\citep{heine2018bridging}, which explores a space of ARGs by fixing a starting ARG, and then repeatedly erasing and resimulating candidate SPR moves connecting two fixed, leaf-labelled local trees.
+Exploration is done by varying the positions of these boundary trees along the sequence.
 Inferring the SPRs between leaf-labelled trees is

### 2023-10-26 paper.tex
Minor edits

@@
 confusion about what, precisely, an ARG \emph{is}.
-Originally, Griffiths and colleagues defined the ARG as an alternative
+% The grammar is a nightmare here - the stochastic process is singular, right?
+In its original form, the ARG is an alternative
 formulation of the coalescent with recombination~\citep{hudson1983properties},
@@
 This mismatch between usage and definition has led to significant
-problems, which if not addressed,
+problems, which if not addressed
 may impede progress in applying ARGs to understand the biological world.
@@
 % % Para 3: what is the problem?
-The classical formal definition of an ARG is expressed in terms of
-genetic inheritance events: the coalescence and recombination events
+The classical formal definition of an ARG is derived from the
+stochastic process, and expressed in terms of events:
+the coalescence and recombination events
 that occur in the history of a sample. We show that this approach
@@
 in the types of genetic inheritance structures that can be encompassed,
-including the outputs of modern simulation and inference methods.
-In particular, we show that in contrast to the eARG encoding which
-requires fully precise estimates of each recombination event,
+including the outputs of modern methods.
+In particular we show that, in contrast with the eARG encoding which
+requires precise estimates of each recombination event,
 the gARG encoding allows us to systematically describe uncertainty
@@
 approach in Section~\ref{sec-eARG}, and consider its limitations
-in the context of contemporary datasets.
+in the context of current datasets and research questions.
 We then discuss the import concept of ``ancestral material''
@@
 we continue
-in Section\ref{sec-ARG-and-local-trees} by considering the relationship
-between an ARG and its local trees and show (contrary to the
-% I want to put a (suitably encoded) here somewhere, but the sentence is
-% very long already.
-prevailing view) that the sequence of local trees contains
+in Section~\ref{sec-ARG-and-local-trees} by considering the relationship
+between an ARG and its local trees. Contrary to the
+prevailing view, we show that a (suitably encoded) sequence of local trees contains
 precisely as much information as the corresponding ARG.
-The gARG view also opens a rich new set of details about ARGs,
+The gARG encoding opens a rich new set of details about ARGs,
 including the ideas of locally unary nodes
@@
 These ideas have important practical considerations,
-which we illustrate by examining the qualitative properties of the
+which we illustrate by examining the qualitative properties of
 ARGs inferred by four recent methods for a classical benchmark
@@
 the field in the Discussion.
-Finally, the literature on ARGs and associated aspects is large and
-confusing, and we cover some important aspects
+Finally, the literature on ARGs is large and
+confusing, and we attempty to clarify some important aspects
 in appendices, including
@@
 We will also use the term ``genome'' in its
-more common sense of ``the genome'' of a species, denoting the full set
-of chromosomes, and hope that the distinction
-will be clear from the context.
+more common sense of ``the genome'' of a species,
+and hope that the distinction will be clear from the context.
 We are not concerned here with mutational processes or observed sequences,
@@
 coordinates over which inheritance occured.
+This is sufficient to capture the effects of any form of
+genetic inheritance (excluding structural variation).
+
 We can define a gARG formally as follows.
@@
 inheritance intervals $I$.
-Here, the terms parent and child are used in graph sense;
+Here, the terms parent and child are used in the graph sense;
 these nodes respectively represent ancestor and descendant genomes,
@@
 node $u\in N$ is associated with a time $\tau_u$.
-The topological relationships
+The topological ordering
 defined by the directed graph structure, however,
- is sufficient for many applications,
+ is sufficient for many applications
 and we will assume that node dates are not known here.
@@
 with which arbitrary additional information can be associated.
-% We can delete then next two sentances if we want, but nodes are
-% integers and this is a vital element of the model.
-Concretely defining nodes as integers in this way has some
-significant advantages. In particular, it provides a direct link
-between an ARG and its local trees in terms of a well-defined
-combinatorial object (Section~\ref{sec-ARG-and-local-trees}),
-as well as enabling straightforward connection between
-definitions and practical implementation (Section~\ref{sec-efficiency}).
+% Note: I don't see why we have to apologise for defining nodes as
+% integers, that's just the definition. I don't think anyone really
+% has a problem with this.
+% Concretely defining nodes as integers in this way has some
+% significant advantages. In particular, it provides a direct link
+% between an ARG and its local trees in terms of a well-defined
+% combinatorial object (Section~\ref{sec-ARG-and-local-trees}),
+% as well as enabling straightforward connection between
+% definitions and practical implementation (Section~\ref{sec-efficiency}).
 
@@
 of genetic inheritance and constituent genomes.
-To avoid confusion with genomic intervals, nodes are
-labelled with lowercase alphabetical letters
-rather than integer identifiers.
+Here, and  throughout,
+nodes labelled with lowercase alphabetical letters
+rather than integer identifiers to avoid confusion with genomic intervals.
 Thus, individual $D_1$ is composed
@@
 
-% These definitions follow naturally from the outcome of events
-% associated with the inheritance of DNA between generations.
-% % FIXME this sentance doesn't follow well from the first, needs a
-% % different bridge before listing important properties.
-% % Maybe the "listicle" format here isn't right anyway.
-% As such,
-% there are some important consequences to these definitions.
-% Firstly, there is
-% no limitation on a mating system, ploidy, or age structure in a gARG.
-% By making the
-% basic unit an individual's genomes,
-% any pattern of inheritance among mixed ploidy (e.g.\ haplodiploid
-% species) can be accommodated.
-% Secondly, because we record the intervals of inheritance
-% between genomes,
-% arbitrary patterns of genetic inheritance can be directly expressed, such as
-% gene conversions,
-% multiple simultaneous recombinations,
-% and so on.
-% % Wilder: give forward ref to section on efficiency here, mention tree sequence
-% % TODO, should probably drop this point.
-% Thirdly, local trees that we generate for a particular position on the
-% genome can have nodes of any arity (number of child nodes), including ``unary''
-% nodes with a single child (see Section~\ref{sec-locally-unary-edges}).
-% % TODO should rephrase this from parent/child to ancestor/descendant?
-% Fourth, recombination is modelled in terms of
-% its inputs (parental genomes) and output (child genome) with associated inheritance intervals.
-% Thus we will have a child node (the recombinant; e.g.\ node \noderef{a}
-% Fig.~\ref{fig-arg-in-pedigree}) and two (or more; see Section~\ref{sec-ARG-simplification})
-% % TODO: check above if we want to refer to sec-ARG-simplification or some other!
-% parental nodes from which this child has inherited genome segments.
-% % TODO rephrase the above, it'll confuse people. Just need to reiterate the
-% % points about events, and maybe lead into the next section
-% Finally, gARGs are not defined in terms of the
-% coalescent with recombination (or any other) stochastic process.
-% Thus, the gARG encoding is focused on faithfully representing the
-% \emph{outcome} of the processes of genetic inheritance
-% with recombination,
-% and is agnostic to the details of those processes.
-
-% Firstly, a gARG may contain nodes that are both samples \emph{and}
-% internal nodes in the local trees.
-% This ability is useful when we have pedigree information
-% along with genetic data
-% \cite[e.g.][]{hayes20191000,RosFreixedes2022,anderson2022genes},
-% where we may have many generations of internal sample nodes
-% whose genomes have been sequenced.
-% Another situation in which it is important that
-% we do not assume samples are all contemporary ``leaf'' nodes
-% (or that all internal nodes are either coalescences or recombinations)
-% is when we are incorporating ancient genomes into ARG
-% inference~\citep{speidel2021inferring,wohns2022unified}.
-% Similarly, a gARG may contain nodes that
-% do not correspond to any particular event in the ancestry of the samples.
-% For example, node \noderef{h} is the direct ancestor of \noderef{d} in
-% Fig~\ref{fig-arg-in-pedigree}, and is not the product of either a
-% coalescence or a recombination. Such nodes would usually be removed
-% from the graph so that \noderef{d} descends directly from \noderef{k} (see
-% the ``\nameref{ARG_simplification}'' section) but there is no necessity for this from
-% a representational perspective.
-% Secondly, because we do not classify nodes by event type
-% (common ancestor or recombination) there is no limitation
-% on the complexity that can be modelled---multiple recombinations
-% and coalescences can happen simultaneously in the same genome,
-% a common occurance in large deeply sequenced
-% pedigrees~\citep{hayes20191000,RosFreixedes2022}.
-% Finally, there is a subtle and important point about how recombination
-% is represented: which genome in the organismal lifecycle
-% do nodes represent?
-% In principle, we can interpret nodes as representing genomes
-% at any point in the life cycle [supplementary Figure ?]. Hudson
-% views nodes as representing gametes~\citep{hudson1983properties}.
-% However, we argue that the most intuitive
-% approach is to let nodes represent the genomes present in diploid
-% individuals in each generation, as in
-% Fig~\ref{fig-arg-in-pedigree}A. In this case recombination
-% is not represented by a single ``recombination node'' (as in an eARG), but with
-% two parent genomes between which recombination takes place
-% (e.g.\ nodes \noderef{e} and \noderef{f}), and a
-% single resulting ``recombinant'' genome (e.g.\ \noderef{c}). The
-% edges that link this recombinant with its two parents carry the information
-% concerning breakpoint location that would be associated with a
-% recombination node in an eARG. [TODO finish up here with more
-% clarity on what this means in terms of the required specifity
-% of recombination events, and put in a forward ref to the simplification
-% section where we talk about stacking up recombination events
-% and their identifiability.]
-
-\section{Event ARGs}
-\label{sec-eARG}
 % % I feel like sophisticated readers would at this point be saying,
@@
 % % what's the point of this paper?
-The genome ARG framework defined in the previous section is
-in many ways a clarification of existing treatments
+The genome ARG framework defined here is
+in many ways simply a clarification of existing treatments
 \cite[e.g.][]{mathieson2020ancestry,shipilina2023origin},
@@
 % with representation details
-Note that here, and throughout, we are not questioning the form of the
+It is important to note that here, and throughout,
+we are not questioning the form of the
 actual ancestral processes that occur in nature, but rather how we
-\emph{represent} the outcomes of such processes in a practical sense.
+\emph{represent} the outcomes of such processes in a practical manner.
 % Again, need to remind the reader that there's an actual point
@@
 These practical details,
-as demonstrated through the remainder of the article, have important
+as demonstrated in later sections, have important
 consequences not only for how methods exchange information about
@@
 
-% section is in many ways simply a clarification of existing definitions,
-% it does provide substantial advantages (as shown in later sections).
-% To allow us to concretely discuss and characterise the prevailing
-% definition of an ARG, we begin by surveying some
-% recent definitions from the literature.
-% \citet{brandt2021evaluation} follow the original Griffiths definition of
-% an ARG by coupling it directly to a generative process, stating
-% that the ``full ancestral recombination graph (ARG) is a structure that encodes all
-% coalescence and recombination events resulting from the stochastic process of
-% the coalescent with recombination.''
-% \citet{shipilina2023origin} emphasise the true existence of the
-% corresponding events (rather than the generative process) by saying that
-% ``[t]he ARG describes the complete
-% ancestry of a sample of genomes through a series of real coalescence and
-% recombination events.''
-% \cite{mathieson2020ancestry}
-% define an ARG as ``a subset of the pedigree...
-% contain[ing] only those edges along which inherited segments of DNA have been
-% transmitted, and only those nodes corresponding to ancestors in which there was
-% a recombination or coalescence event.''
-
-% The prevailing view, captured in these representative definitions, is that an
-% ARG is composed of \emph{events}, and so we will loosely label this approach as
-% ``event ARGs'', or ``eARGs''. Whether we view nodes in an ARG as events or
-% genomes is essentially a philosophical question and in most senses the two are
-% equivalent (see Appendix~\ref{sec-cell-lineages-and-args}, however, for a discussion
-% of some subtle issues around \emph{which} genome corresponds to events
-% of various types).
-% How we represent and reason about the differential inheritance of
-% genetic matierial \emph{between} nodes in an ARG, however, is of fundamental
-% importance, and the central theme of this article.
-% % Need to make this point strongly for the Nicks. Not sure this is the
-% % right place, though. Maybe it's good to remind them again here though...
-% Note that here, and throughout, we are not questioning the form of the
-% actual ancestral processes that occur in nature, but rather how we
-% \emph{represent} such realised histories in a practical sense.
-
-\begin{figure}
-\centering
-\tikzmath{\x1 =0; \x2=8;\xx=12.5; \x3=14; \xt=18;}
-\begin{tikzpicture}[x=5mm, y=5mm, node distance=2mm and 20mm]
-\tikzset{greynode/.style={circle,fill,inner sep=1},
-nodelabel/.style={font=\footnotesize}}
-
-\node [anchor=north west] at (\x1,6) {A};
-\node [anchor=north west] at (\x2,6) {B};
-% \node [anchor=north west] at (\xt,6) {C};
-
-%%% (A) ARG
-
-\node (s0) [greynode] at (\x1 + 0, 0) {};
-\node (s1) [greynode] at (\x1 + 3, 0) {};
-\node (s2) [greynode] at (\x1 + 6, 0) {};
-\node (s3) [greynode] at (\x1 + 3, 1) {};
-\node (s4) [greynode] at (\x1 + 1, 2) {};
-\node (s5) [greynode] at (\x1 + 5, 3) {};
-\node (s6) [greynode] at (\x1 + 3, 4) {};
-
-\draw (s1) -- (s3);
-\draw (s0) |- (s4);
-\draw (s4) -- (\x1 + 2,2) |- (s3);
-\draw (s4) |- (s6);
-\draw (s3) -- (\x1 + 4,1) |- (s5);
-\draw (s2) |- (s5);
-\draw (s5) |- (s6);
-
-%%% (B) Trees
-\node (l0) [greynode] at (\x2 + 0, 0) {};
-\node (l1) [greynode] at (\x2 + 2, 0) {};
-\node (l2) [greynode] at (\x2 + 3, 0) {};
-\node (l3) [greynode] at (\x2 + 1, 2) {};
-\node (l4) [greynode] at (\x2 + 2, 4) {};
-
-\draw (l0) |- (l3);
-\draw (l1) |- (l3);
-\draw (l2) |- (l4);
-\draw (l3) |- (l4);
-
-\node (r0) [greynode] at (\x3 + 0, 0) {};
-\node (r1) [greynode] at (\x3 + 1, 0) {};
-\node (r2) [greynode] at (\x3 + 3, 0) {};
-\node (r3) [greynode] at (\x3 + 2, 3) {};
-\node (r4) [greynode] at (\x3 + 1, 4) {};
-
-\draw (r0) |- (r4);
-\draw (r1) |- (r3);
-\draw (r2) |- (r3);
-\draw (r3) |- (r4);
-
-\foreach \u/\lab in {
-        s0/$\noderef{a}$, s1/$\noderef{b}$, s2/$\noderef{c}$,
-        r0/$\noderef{a}$, r1/$\noderef{b}$, r2/$\noderef{c}$,
-        l0/$\noderef{a}$, l1/$\noderef{b}$, l2/$\noderef{c}$}
-    \node[nodelabel,anchor=south] at ([yshift=-12pt]\u) {\lab};
-
-\foreach \u/\lab in {
-        s6/$\noderef{g}$,
-        l4/$\noderef{g}$,
-        r4/$\noderef{g}$}
-    \node[nodelabel,anchor=south] at (\u) {\lab};
-
-\node [nodelabel,anchor=north west] at ($(s3) + (\x1 + 0,0)$) {$x$};
-\foreach \u/\lab in {
-        s4/$\noderef{e}$,
-        l3/$\noderef{e}$}
-    \node[nodelabel,anchor=south west] at (\u) {\lab};
-\foreach \u/\lab in {
-        s5/$\noderef{f}$,
-        r3/$\noderef{f}$}
-    \node[nodelabel,anchor=south east] at (\u) {\lab};
-\foreach \u/\lab in {s3/$\noderef{d}$, s6/$\noderef{g}$}
-    \node[nodelabel,anchor=south] at (\u) {\lab};
-
-\draw[dashed] (\xx,0) -- (\xx, 4);
-\node[nodelabel,anchor=north] at (\xx,0) {$x$};
-
-% %%% (C) Encoding
-% \node [nodelabel,anchor=north west] at ($(\xt,5)$) {
-% \begin{tabular}{c|c|l}
-% % \multicolumn{2}{c}{Breakpoints}\\
-% Node & Breakpoint & Parents\\
-% \hline
-% $\noderef{a}$ & $\varnothing$ & [\noderef{e}]\\
-% $\noderef{b}$ & $\varnothing$ & [\noderef{d}] \\
-% $\noderef{c}$ & $\varnothing$ & [\noderef{f}]\\
-% $\noderef{d}$ & $x$ & [\noderef{e}, \noderef{f}] \\
-% $\noderef{e}$ & $\varnothing$ & [\noderef{g}]\\
-% $\noderef{f}$ & $\varnothing$ & [\noderef{g}]\\
-% $\noderef{g}$ & $\varnothing$ & []\\
-% \end{tabular}};
-
-\end{tikzpicture}
-\caption{\label{fig-event-arg}
-A classical event ARG. (A) Standard graph depiction with
-breakpoint $x$ associated with the recombination node \noderef{d}.
-Nodes \noderef{e}, \noderef{f} and \noderef{g} are common ancestor events.
-(B) Corresponding local trees to the left and right of breakpoint $x$
-(note these are shown in the conventional form in which only coalescences
-within the local tree are included; see Section~\ref{sec-ARG-and-local-trees}
-% TODO: check above if we want to refer to sec-ARG-and-local-trees or some other!
-for a discussion of this important point).
-}
-\end{figure}
-
+\section{Event ARGs}
+\label{sec-eARG}
 % The prevailing view is that ARGs are composed of events, and so we
@@
 left of $x$ from \noderef{e} and to the right of $x$ from \noderef{f}.
+\begin{figure}
+\centering
+\tikzmath{\x1 =0; \x2=8;\xx=12.5; \x3=14; \xt=18;}
+\begin{tikzpicture}[x=5mm, y=5mm, node distance=2mm and 20mm]
+\tikzset{greynode/.style={circle,fill,inner sep=1},
+nodelabel/.style={font=\footnotesize}}
+
+\node [anchor=north west] at (\x1,6) {A};
+\node [anchor=north west] at (\x2,6) {B};
+% \node [anchor=north west] at (\xt,6) {C};
+
+%%% (A) ARG
+
+\node (s0) [greynode] at (\x1 + 0, 0) {};
+\node (s1) [greynode] at (\x1 + 3, 0) {};
+\node (s2) [greynode] at (\x1 + 6, 0) {};
+\node (s3) [greynode] at (\x1 + 3, 1) {};
+\node (s4) [greynode] at (\x1 + 1, 2) {};
+\node (s5) [greynode] at (\x1 + 5, 3) {};
+\node (s6) [greynode] at (\x1 + 3, 4) {};
+
+\draw (s1) -- (s3);
+\draw (s0) |- (s4);
+\draw (s4) -- (\x1 + 2,2) |- (s3);
+\draw (s4) |- (s6);
+\draw (s3) -- (\x1 + 4,1) |- (s5);
+\draw (s2) |- (s5);
+\draw (s5) |- (s6);
+
+%%% (B) Trees
+\node (l0) [greynode] at (\x2 + 0, 0) {};
+\node (l1) [greynode] at (\x2 + 2, 0) {};
+\node (l2) [greynode] at (\x2 + 3, 0) {};
+\node (l3) [greynode] at (\x2 + 1, 2) {};
+\node (l4) [greynode] at (\x2 + 2, 4) {};
+
+\draw (l0) |- (l3);
+\draw (l1) |- (l3);
+\draw (l2) |- (l4);
+\draw (l3) |- (l4);
+
+\node (r0) [greynode] at (\x3 + 0, 0) {};
+\node (r1) [greynode] at (\x3 + 1, 0) {};
+\node (r2) [greynode] at (\x3 + 3, 0) {};
+\node (r3) [greynode] at (\x3 + 2, 3) {};
+\node (r4) [greynode] at (\x3 + 1, 4) {};
+
+\draw (r0) |- (r4);
+\draw (r1) |- (r3);
+\draw (r2) |- (r3);
+\draw (r3) |- (r4);
+
+\foreach \u/\lab in {
+        s0/$\noderef{a}$, s1/$\noderef{b}$, s2/$\noderef{c}$,
+        r0/$\noderef{a}$, r1/$\noderef{b}$, r2/$\noderef{c}$,
+        l0/$\noderef{a}$, l1/$\noderef{b}$, l2/$\noderef{c}$}
+    \node[nodelabel,anchor=south] at ([yshift=-12pt]\u) {\lab};
+
+\foreach \u/\lab in {
+        s6/$\noderef{g}$,
+        l4/$\noderef{g}$,
+        r4/$\noderef{g}$}
+    \node[nodelabel,anchor=south] at (\u) {\lab};
+
+\node [nodelabel,anchor=north west] at ($(s3) + (\x1 + 0,0)$) {$x$};
+\foreach \u/\lab in {
+        s4/$\noderef{e}$,
+        l3/$\noderef{e}$}
+    \node[nodelabel,anchor=south west] at (\u) {\lab};
+\foreach \u/\lab in {
+        s5/$\noderef{f}$,
+        r3/$\noderef{f}$}
+    \node[nodelabel,anchor=south east] at (\u) {\lab};
+\foreach \u/\lab in {s3/$\noderef{d}$, s6/$\noderef{g}$}
+    \node[nodelabel,anchor=south] at (\u) {\lab};
+
+\draw[dashed] (\xx,0) -- (\xx, 4);
+\node[nodelabel,anchor=north] at (\xx,0) {$x$};
+
+% %%% (C) Encoding
+% \node [nodelabel,anchor=north west] at ($(\xt,5)$) {
+% \begin{tabular}{c|c|l}
+% % \multicolumn{2}{c}{Breakpoints}\\
+% Node & Breakpoint & Parents\\
+% \hline
+% $\noderef{a}$ & $\varnothing$ & [\noderef{e}]\\
+% $\noderef{b}$ & $\varnothing$ & [\noderef{d}] \\
+% $\noderef{c}$ & $\varnothing$ & [\noderef{f}]\\
+% $\noderef{d}$ & $x$ & [\noderef{e}, \noderef{f}] \\
+% $\noderef{e}$ & $\varnothing$ & [\noderef{g}]\\
+% $\noderef{f}$ & $\varnothing$ & [\noderef{g}]\\
+% $\noderef{g}$ & $\varnothing$ & []\\
+% \end{tabular}};
+
+\end{tikzpicture}
+\caption{\label{fig-event-arg}
+A classical event ARG. (A) Standard graph depiction with
+breakpoint $x$ associated with the recombination node \noderef{d}.
+Nodes \noderef{e}, \noderef{f} and \noderef{g} are common ancestor events.
+(B) Corresponding local trees to the left and right of breakpoint $x$
+(note these are shown in the conventional form in which only coalescences
+within the local tree are included; see Section~\ref{sec-ARG-and-local-trees}
+% TODO: check above if we want to refer to sec-ARG-and-local-trees or some other!
+for a discussion of this important point).
+}
+\end{figure}
 
@@
 because many ancestral nodes will not be reachable from the
-samples at a given position.
+samples at a given position (so their corresponding entries in $\pi$ will be zero).
 
@@
 Firstly, the internal nodes of evolutionary trees are usually
-considered to be \emph{unlabelled}.
-Evolutionary trees usually regard internal nodes as unlabelled
-because they most often correspond
-to anonymous hypothetical ancestors.
-Internal ARG nodes may likewise correspond to anonymous hypothetical
-ancestors, but if we do not label the tree nodes we lose the fact that
+considered to be \emph{unlabelled}, or equivalently, labelled by the leaves which they subtend.
+The same canonical labelling cannot be used for internal ARG nodes because the leaves they subtend will typically vary by genomic position.
+If we do not label the tree nodes in a way which is persistent across the sequence of local trees in the ARG, we lose the fact that
 the \emph{same} ancestors sometimes persist across multiple trees.
@@
 but we still record the fact that genome \noderef{c} inherits from \noderef{f}
-in the local tree. Thus, \noderef{c} has a single child in this
-interval and it is unary in this local tree.
-% TODO note this is CA but not coalescent, and refer to where we discuss this.
+in the local tree. Thus, \noderef{f} has a single child in this
+interval and it is unary in this local tree: in other words is is a \emph{locally unary} node.
+% TODO note that node e is CA but not coalescent, and refer to where we discuss this.
 In another example, \noderef{e} is binary in the graph, but is
-unary in each of the local trees in which it is present.
+locally unary in all local trees in which it is present.
 This is because \noderef{a} and \noderef{c} are both recombinant
@@
 and \noderef{c} inherits from \noderef{e} over $[7, 10)$.
-The intervals over which
-we know that a node inherits from another, but where there
-is no local coalescence, are marked by unary nodes in the
-local trees.
 
@@
 is rare (e.g.~SARS-CoV-2; see Discussion).
-\emph{Locally} unary nodes, those which have one child
-in one or more local trees, are a common and important feature
+In contrast, locally unary nodes, are a common and important feature
 of many different types of ARG.

### 2023-10-26 paper.tex
Edits from Yan and JEre

--- 2023-10-26 paper.tex
@@
 we continue
-in Section\ref{sec-ARG-and-local-trees} by considering the relationship
-between an ARG and its local trees and show (contrary to the
-% I want to put a (suitably encoded) here somewhere, but the sentence is
-% very long already.
-prevailing view) that the sequence of local trees contains
+in Section~\ref{sec-ARG-and-local-trees} by considering the relationship
+between an ARG and its local trees. Contrary to the
+prevailing view, we show that a (suitably encoded) sequence of local trees contains
 precisely as much information as the corresponding ARG.

### 2023-10-26 paper.tex

--- 2023-10-26 paper.tex
@@
 \label{sec-eARG}
-% The prevailing view is that ARGs are composed of events, and so we
-% characterise this view
-% Besides viewing ARGs as being composed of events, another
-% feature of the definitions given above is that they do not
-% explicitly describe
-% % TODO check this! Do they talk about how trees are embedded?
-% the details of how such events are represented.
-Where the concrete details of an ARG data structure
+% Trying to be totally, boringly, abundantly, blatantly clear here what we're
+% talking about.
+In this section we define the classical view of an ARG data structure,
+and illustrate its limitations. We are
+interested in the details of how ARGs are described mathematically,
+and as a consequence, how they are represented in a practical sense
+as the output of inference programs.
+Where such concrete details of an ARG data structure
 (the encoding) are
 % TODO more?
-specified~\citep[e.g.][]{
+mathematically defined~\citep[e.g.][]{
 rasmussen2014genome,gusfield2014recombinatorics,ignatieva2021kwarg}
-they follow the approach described by Griffiths and colleagues.
-Because it is focused on recording the details of events,
-we will refer to this Griffiths encoding as an ``Event ARG'', or eARG.
-In an eARG, we have two different types of node in the graph
+they follow the approach described by Griffiths and colleagues
+(but see~\citet{parida2011minimal} and \citet{zhang2023biobank}
+for notable exceptions), and the majority of
+% Note the "an" here to avoid cases like ARGweaver
+ARG inference methods use it as an output format
+\citep[(TODO exhaustive list)][]{rasmussen2014genome,ignatieva2021kwarg}.
+In this Griffiths encoding, we have two different types of node in the graph
 representing the common ancestor and recombination events
@@
 This is done by tracing rootwards through the graph from the samples,
-taking decisions about which outbound edge to follow through
+making decisions about which outbound edge to follow through
 recombination nodes based on the breakpoint
 position~\citep{griffiths1996ancestral}.
-Fig.~\ref{fig-event-arg} shows an example of a classical
+Because it is focused on recording events and their properties,
+we will refer to this Griffiths encoding as the ``Event ARG'', or eARG
+encoding. Fig.~\ref{fig-event-arg} shows an example of a classical
 eARG with three sample genomes (\noderef{a}, \noderef{b}, and \noderef{c})
@@
 left of $x$ from \noderef{e} and to the right of $x$ from \noderef{f}.
+
 \begin{figure}
@@
 
-While the Griffiths approach of annotating recombination nodes
-with a breakpoint in an eARG is a concise and elegant way of describing
-realisations of the coalescent, it has some significant
-issues when applied to the complexities of present-day datasets.
-Because we are explicitly modelling the \emph{properties} of events,
-the encoding must be extended to incorporate
-different types of event that are not simple crossover recombination
-or common ancestry.
-Processes such as gene conversion, for example, can in principle
-be included, but either require additional node types
-or, for instance, multiple recombination nodes joined by a zero-length edge.
-Modelling multiple recombinant offspring of the same parents
-(while keeping track of their identity as siblings)
-as occurs
-in Wright-Fisher models (e.g.\ Fig.~\ref{fig-arg-in-pedigree}),
-or densely sampled multi-generational
-data~\citep[e.g.][]{turnbull2018hundred,hayes20191000,RosFreixedes2022}
-% circumlocutions isn't technically correct, but I like the roundabout
-% overly verbose feel it gives.
-would require more involved circumlocutions.
-These and other problems can be worked around
-(although, we would ideally like everyone to use the
-\emph{same} workarounds if methods are to interoperate; see the Discussion),
-but it is much simpler to use the gARG encoding described
-in the previous section.
-By recording the \emph{outcome} of events in terms of the
-intervals of genetic inheritance on edges instead of recording the
-properties of the events themselves, we remove the need to explicitly
-model different types of event in the data structure.
-More fundamentally, because gARGs
-allow us to directly describe
-genetic inheritance in multiple non-contiguous genome intervals
-between an ancestor and a descendant,
-we can encode the outcome of
-\emph{multiple} genetic inheritance events, and to
-systematically describe the uncertainty around their
-ordering (see Section~\ref{sec-precision}).
-
-There is also a certain clarity gained by explicitly modeling nodes
-in the inheritance graph as an individual's genomes.
+% eARGs are representationally for limited interchange
+While the Griffiths approach of annotating recombination nodes with a
+breakpoint in an eARG is a concise and elegant way of describing realisations
+of the coalescent, it has some significant issues when applied to the
+complexities of present-day datasets. The eARG encoding explicitly models only
+two different types of event and thus anything that is not a single crossover
+recombination or common ancestor event, must be incorporated
+either in a roundabout way using these
+events, or by adding new types of event to the encoding. For example, gene
+conversion could be accommodated either by stipulating a third type of event
+(annotated by two breakpoints and corresponding traversal conventions for
+recovering the local trees) or by two recombination nodes joined by a
+zero-length edge. From the perspective of practical interchange of data between
+inference methods and downstream applications, both
+workarounds are problematic, and the gARG encoding described in the previous
+section offers a much simpler solution.
+Modern datasets contain many more complexities which the eARG encoding
+is ill-suited to represent (see Discussion).
+
+There is also a certain clarity gained by explicitly modelling nodes
+in the inheritance graph as genomes.
 Outside of the context of a
@@
 For example, \emph{which} genome along a chain of cell divisions should be
-regarded as the one where an event occured
+regarded as the one where an event occurred
 (Appendix~\ref{sec-cell-lineages-and-args}),
@@
 within a single individual should be regarded as one or multiple events are
-debateable.
+debatable.
 % If we want to have an ARG software ecosystem then such fireside
@@
 are unproductive.
-
-% By regarding nodes as an individual's genomes and
-% explicitly annotating edges with the intervals of genetic inheritance
-% between those genomes, we can capture the \emph{outcome} of a
-% range of different types of event without needing to explicitly model
-% the details of these events in the ARG data structure.
-% More fundamentally, because gARGs
-% allow us to directly describe
-% genetic inheritance in multiple non-contiguous genome intervals
-% between an ancestor and a descendant,
-% we can encode the outcome of
-% \emph{multiple} genetic inheritance events, and to
-% systematically describe the uncertainty around their
-% ordering (see Section~\ref{sec-precision}).
-
-% While the additional generality provided by the gARG representation
-% is useful, it is in reality a small change that does not
-% lead to significantly different time or space complexity in computational
-% tasks.
-% The vitally important and fundamental distinction between
-% a gARG and an eARG is that gARGs
-% allow us to directly describe
-% genetic inheritance in multiple non-contiguous genome intervals
-% between an ancestor and a descendant, which cannot be done
-% by specifying recombination events and breakpoints.
-% % FIXME this sentence is problematic. We should recast to talk about
-% % the outcomes of events, not talk about simulateousness (which will
-% % confuse people
-% Thus, we have the means to encode the \emph{outcome} of
-% multiple genetic inheritance events, and to
-% systematically describe the uncertainty around their
-% ordering (see Section~\ref{sec-precision}).
-% As we describe in the following sections, freeing ourselves from the requirement
-% of recording specific events and
-% the additional generality of the gARG encoding
-% has significant advantages. It also raises questions about what our goals
-% for inferring ARGs should be, and how we evaluate the success
-% of such inferences.
-
-% A common feature of these definitions is that they are expressed
-% in terms of evolutionary \emph{events}, and so for the purposes of
-% discussion we can label
-% this prevailing interpretation as an ``event ARG'', or eARG.
-% (Many definitions also emphasise
-% the necessity of exhaustively capturing \emph{all} events, which we
-% examine in later sections.)
-% Another common feature of these definitions is that they do not explicitly
-% discuss the mechanism by which local trees vary along the genome.
-% The topology of an eARG defines the common ancestor and
-% recombination events that occurred,
-% but the graph structure alone is not sufficient to define the local trees,
-% or equivalently, how ``ancestral material'' flows along
-% the edges of the ARG (see Section~\ref{sec-eARG-to-gARG}).
-
-% TODO reference forward to the cell ARG appendix. Ultimately
-% every event corresponds to a particular genome in a particular
-% individual or cell.
-% Whether we regard nodes as events or genomes is something of a
-% philosophical question, and in most senses the two are equivalent
-% (but see the discussion on ``unary nodes'' in Section~\ref{sec-locally-unary-edges}).
-% From a concrete data-structure encoding perspective, the only real
-% difference between an eARG and a gARG is that in an eARG we store recombination
-% \emph{breakpoints} at nodes of out-degree two (recombination nodes)
-% whereas in a gARG we store \emph{inheritance intervals} on edges.
-% A key advantage of associating inheritance intervals with
-% edges in a gARG is that it allows us to describe genetic inheritance events other
-% than simple crossovers, such as multiple recombinations and
-% gene conversions. By explicitly stating the exact regions of genome
-% that are transmitted along an edge, we generalise the representation
-% to encompass any process that leads to the differential inheritance of
-% genetic material along the genome.
-
-% While the additional generality provided by the gARG representation
-% is useful, it is in reality a small change that does not
-% lead to significantly different time or space complexity in computational
-% tasks.
-% The vitally important and fundamental distinction between
-% a gARG and an eARG is that gARGs
-% allow us to directly describe
-% genetic inheritance in multiple non-contiguous genome intervals
-% between an ancestor and a descendant, which cannot be done
-% by specifying recombination events and breakpoints.
-% % FIXME this sentence is problematic. We should recast to talk about
-% % the outcomes of events, not talk about simulateousness (which will
-% % confuse people
-% Thus, we have the means to encode the \emph{outcome} of
-% multiple genetic inheritance events, and to
-% systematically describe the uncertainty around their
-% ordering (see Section~\ref{sec-precision}).
-% As we describe in the following sections, freeing ourselves from the requirement
-% of recording specific events and
-% the additional generality of the gARG encoding
-% has significant advantages. It also raises questions about what our goals
-% for inferring ARGs should be, and how we evaluate the success
-% of such inferences.
 

### 2023-10-26 paper.tex
minor edits s5 and s6

@@
 interval, node \noderef{e} is not ancestral to any
-of the samples: it carries \emph{non ancestral} material.
+of the samples: it carries non ancestral material.
 This process of local coalescence continues through the
@@
 time in a ``prospective'' manner. Genetic inheritance relationships
-are recorded exhaustively, generation-by-generation
+are recorded exhaustively, generation-by-generation,
 leading to a rapid build-up of information, much of which
@@
 duplicate its graph topology, and then add inheritance annotations
-to each of the gARG's edges (Fig.~\ref{fig-ancestry-resolution}B).
+to each of the gARG's edges (Fig.~\ref{fig-ancestry-resolution}B) as follows.
 If a given node is a common ancestor event, we annotate the single
@@
 Thus, the final step is to use the ``simplify'' algorithm to perform this
-resolution (Fig.~\ref{fig-ancestry-resolution}C).
+resolution and produce the sample-resolved
+gARG (Fig.~\ref{fig-ancestry-resolution}C).
 
@@
 allows for ``local'' inspection
-of an ARG in a way that is not possible by storing
-common ancestor and recombination events in an eARG. Because
-the ancestral material is stored with each edge, the
+of an ARG in a way that is not possible in an eARG.
+Because the ancestral material is stored with each edge, the
 cumulative effects of events over time can be reasoned
@@
 This combinatorial approach provides at least
-one novel insight into the relationship between ARGs and local trees.
+one novel insight, clarifying the fundamental relationship between
+ARGs and local trees.
 If we are given the sequence of local trees for a gARG
@@
 by the oriented tree $\pi^x_1\dots\pi^x_n$. This path does not directly
-contain information about branching points, and so it is helpful
-to define the ``arity''.
+contain information about branching points, and defining a
+node's arity (number of child nodes) is therefore useful.
 The ``local arity'' of a node is the number of children it has
@@
 This distinction between ARG and local tree arity is mainly
-of interest when we consider \emph{unary} nodes, those that have
+of interest when we consider unary nodes, those that have
 a single child.
@@
 
+[FIXME this isn't true, add comment about eARG conversion]
 Outside of the context of simulation, ARG-unary nodes are
@@
 cannot fully reason about how a local tree is embedded in an ARG
-without them.
-
-Nodes with one child are not a standard feature of evolutionary trees
-or the ARG literature
-(although \citet{mathieson2020ancestry} show examples in which the ARGs
-contain unary nodes, and state that paths are
-``implicitly passing through points representing specific individuals.'').
-We are usually interested in
-nodes that have at least two children because these
-are in principle detectable from mutational differences
-arising in the respective subtrees.
-A tree branch may correspond to many generations
-of individuals (or indeed cell divisions; see Appendix \ref{sec-cell-lineages-and-args})
-and there is usually
-little information about these intermediates, and hence limited utility
-in modelling them.
-As we see in the next section, however, there is a range of
-different ways in which ARG nodes can be locally unary. Furthermore,
-several inference methods infer ARGs containing locally unary nodes
-(see Section~\ref{sec-precision}).
+without them. As we see in the next section, both ARG and locally
+unary nodes occur in a range of different scenarios.
+
+% NOTE: this para doesn't serve much purpose now I think.
+
+% Nodes with one child are not a standard feature of evolutionary trees
+% or the ARG literature
+% (although \citet{mathieson2020ancestry} show examples in which the ARGs
+% contain unary nodes, and state that paths are
+% ``implicitly passing through points representing specific individuals.'').
+% We are usually interested in
+% nodes that have at least two children because these
+% are in principle detectable from mutational differences
+% arising in the respective subtrees.
+% A tree branch may correspond to many generations
+% of individuals (or indeed cell divisions; see Appendix \ref{sec-cell-lineages-and-args})
+% and there is usually
+% little information about these intermediates, and hence limited utility
+% in modelling them.
+% As we see in the next section, however, there is a range of
+% different ways in which ARG nodes can be locally unary. Furthermore,
+% several inference methods infer ARGs containing locally unary nodes
+% (see Section~\ref{sec-precision}).
 
@@
 \label{sec-ARG-simplification}
-ARG simplification is a powerful tool, and can do far more
-than removing unreachable nodes and edges in a prospective ARG
-and resolving ancestral material in an eARG.
+ARG simplification is a powerful tool.
 In general, we can think of
-ARG simplification as the process
+simplification as the process
 of removing nodes and re-writing edges (and their inheritance annotations)

### 2023-10-26 paper.tex
edits for S7. Remove strong assertions about identifiability

@@
 The graph visualisations in Fig.~\ref{fig-simplification} have
-three novel features.
+three novel features which require some explanation.
 Firstly, edge weights (the thickness of the lines joining
@@
 whereas nodes in which ancestral material never coalesces have a
-coalescent span of 0\% (such nodes will be extremely difficult to
-identify from the distribution of shared mutations in the samples).
+coalescent span of 0\%.
+% JK: Let's not get into discussions of identifiability or inferability
+% here, we're just trying to show the things that exist in the true history.
+% (such nodes will be extremely difficult to
+% identify from the distribution of shared mutations in the samples).
 % This is useful why? I can't come up with a quick way of saying
@@
 % JK: Hmm, not sure we want to jump into inferability here?
-All these features will impact what kind of an ARG we can
-hope to infer from sampled genomes.
+% All these features will impact what kind of an ARG we can
+% hope to infer from sampled genomes.
 
@@
 In Fig.~\ref{fig-simplification}A the local trees (right)
-contain many unary nodes, and as we successively simplify,
-the local trees contain fewer
-unary nodes (Fig.~\ref{fig-simplification}B,C) until
-we reach Fig.~\ref{fig-simplification}D, where the local trees
-have no unary nodes.
-A node is unary in a local tree covering the genome interval $I$
-if genetic inheritance passes through that ARG node,
-but no coalescence occurs in the interval $I$.
+contain many unary nodes, fewer as  we successively
+simplify (Fig.~\ref{fig-simplification}B,C),
+until we reach Fig.~\ref{fig-simplification}D, where are none.
+
+The first level of simplification that we can perform is based only
+on the graph topology.
+An example of graph topology that we may consider redundant
+(or non-identifiable) is a
+``diamond''~\citep{rasmussen2014genome}
+in which the two parent nodes of a recombination immediately
+join again into a common ancestor (e.g.~\noderef{j}, \noderef{l}, \noderef{m}
+and \noderef{n} in Fig.~\ref{fig-simplification}A).
+Unless we are specifically
+interested in the recombination event or these ancestral genomes,
+the diamond can be replaced by a single edge without loss of
+information.
+More generally, any subgraph that is singly-connected in both the leafward and
+rootward direction (a ``super-diamond'') can be replaced by one edge.
+This definition includes the case
+of a node that has one inbound and one outbound edge, such as
+nodes \noderef{f} and \noderef{h}.
+Fig.~\ref{fig-simplification}B shows the result of this type of
+graph topology simplification.
+
+Simplifying away diamonds will remove many unary nodes from the
+local trees, but there can still be nodes that are unary in all
+of the local trees. In particular, a node can represent a recombinant
+with multiple parents in the graph but only a single child (e.g.\ node \noderef{n}
+in Fig.~\ref{fig-simplification}B), or can represent a common ancestor with
+multiple children in the graph but in which no coalescence takes place
+in the local trees
+(node \noderef{r} in Fig.~\ref{fig-simplification}B).
 The distinction between the ``common ancestry'' of two or more genomes
@@
 unary in these three trees.
-
-The first level of simplification that we can perform is to remove
-parts of the graph topology that are invisible to the samples.
-An example of such topology is a
-``diamond''~\citep{rasmussen2014genome}
-in which the two parent nodes of a recombination immediately
-join again into a common ancestor (e.g.~\noderef{j}, \noderef{l}, \noderef{m}
-and \noderef{n} in Fig.~\ref{fig-simplification}A).
-Unless we are specifically
-interested in the recombination event or these ancestral genomes,
-there is no information in this topology and the diamond can be
-replaced by a single edge. More generally, any
-subgraph that is singly-connected in both the leafward and
-rootward direction (a ``super-diamond'') is non-identifiable and can be
-replaced by one edge. This definition includes the case
-of a node that has one inbound and one outbound edge, such as
-nodes \noderef{f} and \noderef{h}.
-Fig.~\ref{fig-simplification}B shows the result of this type of
-graph topology simplification.
-
-Simplifying away diamonds will remove many unary nodes from the
-local trees, but there can still be nodes that are unary in all
-of the local trees. In particular, a node can represent a recombinant
-with multiple parents in the graph but only a single child (e.g.\ node \noderef{n}
-in Fig.~\ref{fig-simplification}B), or can represent a common ancestor with
-multiple children in the graph but in which no coalescence takes place
-in the local trees
-(node \noderef{r} in Fig.~\ref{fig-simplification}B).
-% point out that this is why Hudson refers to CA nodes, not coalescence nodes
 Such nodes are not singly connected in the graph, but are nevertheless unary in
@@
 (e.g. \noderef{n}) are still present, and represent precise
-information about the time, genomic location and tree
-branches involved
-in the recombination event.
+information about the time, genomic location and
+lineages involved in the recombination event.
 
@@
 at position 87 has been lost entirely.
-
-% The point of these examples is to emphasise that
-% information about recombination in an ARG is not ``all
-% or nothing'': there is a gradation of detail
-% about recombination, in terms of genomic location, timing,
-% and the ancestral lineages involved.
-% While there has been a recent tendency to emphasise the
-% importance of ``full ARGs''
-% \citep[e.g.][]{deng2021distribution,brandt2021evaluation,
-% rasmussen2022espalier}
-% which contain complete information about all recombination
-% events, it is reasonable to question the value of
-% such precise information and whether it is inferrable.
-% [SOME MORE STUFF]
-
-% Ancestral recombination graphs are often characterised
-% as consisting of a complete record of the genetic history
-% of a sample. For example,
-% \cite{rasmussen2014genome} state that an ``ARG provides a record of all
-% coalescence and recombination events since the divergence of the sequences
-% under study'' and
-% according to \cite{deng2021distribution} an ARG
-% ``provides all the information about the genealogical history of a sample,
-% including the locations of recombination events.'' It is worth
-% questioning, however, whether such comprehensive information
-% will always be available, and in particular whether our
-% representation of genealogical history should \emph{require}
-% such precision. Simulations will usually generate complete
-% information about the history of a sample, of course, but
-% we may not have sufficient information to infer such detail from sampled genomes.
-
-% % The full ancestral recombination graph (ARG) is a structure that encodes all
-% % coalescence and recombination events resulting from the stochastic process of
-% % the coalescent with recombination.
-% \citet{brandt2021evaluation} define a ``full'' ARG as
-% ``a structure that encodes all
-% coalescence and recombination events resulting from the stochastic process of
-% the coalescent with recombination'' (thus going against the general
-% historial trend discussed in the XXX section).
-% \citet{rasmussen2022espalier} have a less restrictive definition, in
-% which a ``full'' ARG is one that contains complete information about
-% recombination events, without necessarily being tied to a particular
-% stochastic process.
-% While [something positive], it is important to be clear on
-% what the overall goal of such precise inferences are,
-% what the limitations on when they can be made are,
-% and their downstream utility in applications.
-
-% % What is the goal of having fully precise ARGs?
-% % 1) studying the actual recombination events;
-% % 2) evaluating the likelihoods
-% The main reasons for wanting precise estimates of all recombination
-% events are to either compute a likelihood of the inferred ARG
-% under a statistical model, or to study the recombinants in
-% detail~\citep{rasmussen2022espalier}.
-% % Explain that we're really interested in the details of recombination
-% % itself with a small number of samples.
-% In the latter case, we might
-% have a small number of samples~\citep{guo2022recombination}.
-
-% Beyond cases in which visual inspection of the trees and the
-% effects of recombination is feasible, the main application
-% of having completely precise estimates of recombination events is
-% to evaluate the likelihood of the estimate under some
-% statistical model. The only statistical model available
-% is the coalescent with recombination
-% and its approximations~\citep{mcvean2005approximating,marjoram2006fast}.
-% In this case, the utility of a ``full`` ARG for some dataset
-% is limited by how well the dataset fits the assumptions
-% of the model.
-
-% % JK: This is very rough, just jotting down the basic ideas,
-% % will need substantial revision.
-% It is not clear that such complete information will always
-% (or indeed ever) be available from inferences~\cite{shipilina2023origin}.
-%  We may be
-% able to precisely identify the location and timing
-% of recent recombinations (particularly if detailed pedigree
-% information is available) but beyond this, the information
-% required simply isn't present. When sampling from a
-% process such as the Sequentially Markov
-% Coalescent~\citep{mcvean2005approximating,marjoram2006fast}
-% like ARGweaver~\citep{rasmussen2014genome}, it is true that
-% we do obtain a fully resolved ARG with precise
-% information about recombination, but it is reasonable
-% to question how much information from the data is used
-% to support older recombinations, in particular.
-
-% % This idea of a continuum of information about recombination
-% % being inferrable and representable is consistent with the ideas
-% % discussed by~\cite{shipilina2023origin}. There a haplotype block
-% % is synonymous with coalescent nodes in the ARG, which we can generalise
-% % as nodes in a gARG with at least two children. In principle, these
-% % nodes are identifiable by the mutations that occur on the outbound
-% % edges of the node.
-
-% It is helpful to consider the difference between fully
-% resolved binary trees and those that contain polytomies.
-% The situation is in fact directly analogous: an ARG that
-% contains fully resolved recombination information must have
-% at two most parents per node.
-
-% In general, discussions tend to contrast an ARG
-% on one hand as holding \emph{complete} information about recombination
-% (and therefore between-tree correlation structure)
-% and a sequence of local trees on the other (with, by implication,
-% no direct information about between-tree correlations).
-% This is a false dichotomy:
-% a \emph{continuum} of information about recombination and between-tree
-% correlation structure can be represented and inferred, and all
-% such levels of information are useful, even if only for computational
-% efficiency.
 

### 2023-10-26 paper.tex
Updates from Overleaf

--- 2023-10-26 paper.tex
@@
 % pedantically, ARG is not really a "word"
-The term ``ARG'' is now most often used in quite a general
+The term ``ARG'' is now frequently used in a general
 sense~\citep[e.g.][]{mathieson2020ancestry,hejase2020summary,
@@
 that occur in the history of a sample. We show that this approach
-to representing ARGs (which we refer to as the ``Event ARG'', or eARG
+to representing ARGs (which we refer to as the ``event ARG'', or eARG
 encoding) is inherently limiting, insufficient to represent
@@
 We address these problems by providing a more flexible formulation
-of an ARG data structure, which we refer to as the ``Genome ARG''
+of an ARG data structure, which we refer to as the ``genome ARG''
 or gARG encoding. We show that it is fully compatible with the
@@
 We define a genome as the complete set of genetic material that a child
-inherits from one parent. A diploid individual  (we assume diploids here
-for clarity, but the definitions apply to organisms of arbitrary ploidy)
+inherits from one parent. A diploid individual  
 % Q: will someone nitpick that we're assuming diploid obligate sexual species?
-therefore carries two genomes, one inherited from each parent.
+therefore carries two genomes, one inherited from each parent (we assume diploids here
+for clarity, but the definitions apply to organisms of arbitrary ploidy).
 We will also use the term ``genome'' in its
@@
  we ``annotate'' the edges with the genome
-coordinates over which inheritance occured.
+coordinates over which inheritance occurred.
 This is sufficient to capture the effects of any form of
@@
 node $u\in N$ is associated with a time $\tau_u$.
-The topological ordering
-defined by the directed graph structure, however,
+However, the topological ordering
+defined by the directed graph structure
  is sufficient for many applications
-and we will assume that node dates are not known here.
+thus we do not require node dates to be known.
 % This should be obvious, but seems like it needs to be said
@@
 Here, and throughout,
-nodes labelled with lowercase alphabetical letters
-rather than integer identifiers to avoid confusion with genomic intervals.
-Thus, individual $D_1$ is composed
+nodes are labelled with lowercase alphabetical letters
+rather than integer identifiers, to avoid confusion with genomic intervals.
+Thus individual $D_1$ is composed
 of genomes \noderef{a} and \noderef{b}, which are inherited from its
-two parents, $D_3$ and $D_4$. Each inherited genome may be the recombined product
-of the genomes
-% FIXME "contained within" is a poor phrase here, use "constituent" or something
-contained within a parent.
+two parents $D_3$ and $D_4$. Each inherited genome may be the recombined product
+of the two genomes belonging to an individual parent.
 In this example,
@@
 The local trees span the three genome regions delineated
-by the two recombination breakpoints that gave rise to these genomes.
-See Section~\ref{sec-ARG-and-local-trees} for further details on
-how the local trees are embedded in an ARG.
+by the two recombination breakpoints that gave rise to these genomes;
+further discussion on
+how local trees are embedded in an ARG is provided in Section~\ref{sec-ARG-and-local-trees}
 % Following the above definitions, corresponding data encoding for
@@
 \citep[(TODO exhaustive list)][]{rasmussen2014genome,ignatieva2021kwarg}.
-In this Griffiths encoding, we have two different types of node in the graph
+In this Griffiths encoding, we have two different types of node in the graph,
 representing the common ancestor and recombination events
@@
 Because it is focused on recording events and their properties,
-we will refer to this Griffiths encoding as the ``Event ARG'', or eARG
+we will refer to this Griffiths encoding as the ``event ARG'', or eARG
 encoding. Fig.~\ref{fig-event-arg} shows an example of a classical
@@
 Indeed, there is considerable variation in the number of recombinations inferred by
-different methods, with e.g.~\argweaver\ (Fig.~\ref{fig-inferred-args} B) suggesting
-far more than are required under minimal parsimony assumptions (Fig.~\ref{fig-inferred-args} A).
+different methods, with e.g.~\argweaver\ (Fig.~\ref{fig-inferred-args}B) suggesting
+far more than are required under minimal parsimony assumptions (Fig.~\ref{fig-inferred-args}A).
 
@@
 (hence Fig.~\ref{fig-inferred-args}A,B disagrees with Fig.~\ref{fig-inferred-args}D
-in this order)
-More generally, the number of nodes in Fig.~\ref{fig-inferred-args} C
+in this order).
+More generally, the number of nodes in Fig.~\ref{fig-inferred-args}C
 that have more than two children hints at considerable uncertainty in
@@
 of a gARG provided here by stipulating a
-simple tabular representation and nodes and edges,
+simple tabular representation of nodes and edges,
 and also defining a concise representation of
@@
 describe and utilise uncertainty about ARG inference, and to
-incorporates uncertainty encountered during inference into the returned ARG.
+incorporate uncertainty encountered during inference into the returned ARG.
 One approach, enabled by the gARG encoding described here, is to allow
@@
 events only a small proportion are actually detectable from the data, assuming
-human-like mutation rates \citep{myers2002detection,hayman2023recoverability}.
+human-like mutation and recombination rates \citep{myers2002detection,hayman2023recoverability}.
 Even when a recombination event \emph{is} detectable, its timing and breakpoint
@@
 In comparing tree-by-tree along the genome, the effects of recombination
-are incorporated in quite an indirect manner through the correlations
-between the local trees, rather than through directly taking into account
+are incorporated in a rather indirect manner through the correlations
+between the local trees, instead of directly taking into account
 the persistence of nodes and edges across multiple trees.
@@
 These methods operate on representations
-of the ``Little ARG'', and are computationally expensive, being
+of the ``Little ARG'' (see Appendix \ref{sec-big-and-little-arg}), and are computationally expensive, being
 applicable to at most hundreds of samples consisting of tens or hundreds of

### 2023-10-26 paper.tex
delete dead stuff
Add data availability section

@@
 % (such nodes will be extremely difficult to
-% identify from the distribution of shared mutations in the samples).
-% This is useful why? I can't come up with a quick way of saying
-% this. In some senses it's a measure of identifiability?
-% JK: Hmm, not sure we want to jump into inferability here?
-% All these features will impact what kind of an ARG we can
-% hope to infer from sampled genomes.
 
@@
 To focus on the ARG topology rather than estimates of node times, the Y position of each node
-is arbitrary, taken from the ranked order of nodes output by each software method.
-The version numbers of inference software, the specific inference parameters
-needed to reproduce the ARGs, and (where necessary) functions to convert outputs to the
-\texttt{tskit} gARG format, are available from our GitHub repository.
+is arbitrary, taken from the ranked order of nodes output by each software method (see the ``Data Availability''
+section for details of inference software and conversion methods).
 
@@
 
+\section*{Data Availability}
+
+The public GitHub repository at \url{https://github.com/tskit-dev/what-is-an-arg-paper} can be
+used to reproduce all figures and tables in this paper. In particular this includes
+the ARG used in Fig.~\ref{fig-ancestry-resolution}; the simulation code and functions
+used to generate Fig.~\ref{fig-simplification}; and for Fig.~\ref{fig-inferred-args},
+the software versions, parameter settings, and (where necessary) functions to convert
+software outputs to the \texttt{tskit} gARG format.
+
 \section*{Acknowledgements}

### 2023-10-26 paper.tex
Move Wiuf & Hein caveat
I think it distracts the reader to put the caveat about "not a standard simulated ARG" at the end of the section. This is my attempt to shorten the text and put something up front instead. I think it makes the section end on a stronger note.

--- 2023-10-26 paper.tex
@@
 Any eARG can be converted to a sample-resolved gARG using the same machinery,
-and the two-step process is illustrated in Fig.~\ref{fig-ancestry-resolution}.
-The first step is to take the input eARG (Fig.~\ref{fig-ancestry-resolution}A),
+and the two-step process is shown in Fig.~\ref{fig-ancestry-resolution}.
+In order to show a graph with non-ancestral material, rather than a simulated
+Little ARG (Appendix~\ref{sec-big-and-little-arg}), we use a
+classic example from the literature
+\citet{wiuf1999recombination}, hand-crafted to
+highlight several important ARG properties. The first conversion
+step is to take the input eARG 
+Fig.~\ref{fig-ancestry-resolution}A),
 duplicate its graph topology, and then add inheritance annotations
@@
 each time.
-
-Note that the \citet{wiuf1999recombination} eARG
-in Fig.~\ref{fig-ancestry-resolution} is not particularly
-representative, because inference or simulation methods will usually
-only generate ARGs containing nodes and edges ancestral to the sample
-(see the discussion of the ``Big ARG'' stochastic process in
-Appendix~\ref{sec-big-and-little-arg}, however).
-Nonetheless, it is an instructive example from the literature which highlights several
-important properties of ARGs, and the general point about
-the need to resolve ancestral material ``on the fly'' for eARG traversals
-holds.
 

### 2023-10-26 paper.tex
Edits from Yan, Jere

--- 2023-10-26 paper.tex
@@
 \citep[(TODO exhaustive list)][]{rasmussen2014genome,ignatieva2021kwarg}.
-In this Griffiths encoding, we have two different types of node in the graph,
+In this Griffiths encoding, as well as sample nodes,
+we have two additional types of node in the graph,
 representing the common ancestor and recombination events
@@
 Because it is focused on recording events and their properties,
-we will refer to this Griffiths encoding as the ``event ARG'', or eARG
+we will refer to this Griffiths encoding as the ``event ARG'' or eARG
 encoding. Fig.~\ref{fig-event-arg} shows an example of a classical
@@
 \caption{\label{fig-event-arg}
-A classical event ARG. (A) Standard graph depiction with
+A classical event ARG (eARG). (A) Standard graph depiction with
 breakpoint $x$ associated with the recombination node \noderef{d}.
@@
 breakpoint in an eARG is a concise and elegant way of describing realisations
-of the coalescent, it has some significant issues when applied to the
+of the coalescent with recombination, it has some significant issues when applied to the
 complexities of present-day datasets. The eARG encoding explicitly models only
@@
 on the edges of an ARG.
-In Fig.~\ref{fig-arg-in-pedigree}, for example, we have
+For example, in Fig.~\ref{fig-arg-in-pedigree} we have
 four sample genomes, \noderef{a}--\noderef{d}. As we
 trace their genetic ancestry into the previous generation
-(\noderef{e}--\noderef{h}), we can think of the ancestral
-material of the samples propagating through the graph
-backwards in time. In the interval $[2, 7)$, there is a
+(\noderef{e}--\noderef{h}), we can think of their ancestral
+material propagating through the graph
+backwards in time. In the region $[2, 7)$, there is a
 local coalescence where nodes \noderef{a} and \noderef{c}
-find a common ancestor in \noderef{f}. Thus, in this interval,
+find a common ancestor in \noderef{f}. Thus, in this region,
 the total number of genome segments that are ancestral to the
-sample is reduced from four to three. Note that in the same
-interval, node \noderef{e} is not ancestral to any
-of the samples: it carries non ancestral material.
+sample is reduced from four to three. Fig.~\ref{fig-arg-in-pedigree}A
+illustrates this by (shaded) ancestral material being present
+in only three nodes (\noderef{f}, \noderef{g}, and \noderef{h}) in this region,
+while node \noderef{e} is blank
+as it carries \emph{non ancestral} material.
 This process of local coalescence continues through the
@@
 (see Appendix~\ref{sec-big-and-little-arg}).
-The ability to \emph{store} the resolved ancestral material
-for a sample
-in an ARG is also a key distinction between the eARG and gARG
+The ability to \emph{store} resolved ancestral material
+is also a key distinction between the eARG and gARG
 encodings. Because an eARG stores only the graph topology and
 recombination breakpoints, there is no way to locally
-ascertain the ancestral material without traversing the graph
-rootwards from the samples,
+ascertain ancestral material without traversing the graph
+rootwards from the sample nodes,
 resolving the effects of recombination and common ancestor events.
@@
 a sample through a pre-existing graph is a well-studied problem,
-and central to recent advances in individual-based forwards-time
+and central to recent advances in individual-based forward-time
 simulations~\citep{kelleher2018efficient,haller2018tree}.
@@
 Section~\ref{sec-ARG-simplification}), and is also the key enabling factor when
-incorporating propective ARGs into
+incorporating prospective ARGs into
 forward-time simulation.
@@
 This distinction between ARG and local tree arity is mainly
-of interest when we consider unary nodes, those that have
-a single child.
+of interest when we consider those that have
+a single child: that are \emph{unary}.
 Returning to the example in Fig~\ref{fig-arg-in-pedigree}, nodes
@@
 in the local tree. Thus, \noderef{f} has a single child in this
-interval and it is unary in this local tree: in other words is is a \emph{locally unary} node.
+interval and it is unary in this local tree: in other words it is a \emph{locally unary} node.
 % TODO note that node e is CA but not coalescent, and refer to where we discuss this.
 In another example, \noderef{e} is binary in the graph, but is
-locally unary in all local trees in which it is present.
+locally unary in all trees in which it is present.
 This is because \noderef{a} and \noderef{c} are both recombinant
@@
 reveals that many non-recombinant (green) nodes are
-locally unary (see section \ref{sec-locally-unary-edges}). In other words,
+locally unary (see Section \ref{sec-locally-unary-edges}). In other words,
 in some regions of the genome these nodes are present in the ancestry without
@@
 The Little ARG process avoids non-genetic ancestors at the cost of more complex
-dynamics and state space. We also demonstrate that evaluating the sampling probability
-of either process---a key quantity in many statistical approaches---requires that the
+dynamics and state space.
+We also demonstrate that applications relying on the grouping of inheritance pathways into ancestral lineages, such as likelihood-based inference under the coalescent, requires that the
 gARG (or eARG) data structure be interpreted in a model-specific way.
@@
 (B) A corresponding realisation of a Big ARG, which augments Hudson's algorithm
-by tracking nonancestral lineages. The result is a simpler state space and
+by tracking non-ancestral lineages. The result is a simpler state space and
 dynamics, at the cost of extra nodes and edges, highlighted in red, which do
@@
 interval of time of the realisations.
-In order to evaluate these rates, and hence the sampling probability
-\cite[see e.g.][Equation (3)]{mahmoudi2022bayesian},
+In order to evaluate these rates, e.g.\ for likelihood-based inference
+\citep{baumdicker2021efficient, mahmoudi2022bayesian},
 it is necessary to know the number of lineages and number of extant links
@@
 Some representations may not provide this information.
-For example, in the gARG encoding depicted in Figure \ref{fig-ancestry-resolution}C
-it is clear that a recombination takes place between nodes \noderef{b} and
-\noderef{d} as well as \noderef{e}, but the exact time of the event is ambiguous,
-and thus so is the number of lineages during the time interval. In fact, this
-information cannot be recovered from the gARG encoding used in Figure \ref{fig-ancestry-resolution}C,
-and alternative conventions must be introduced to resolve such ambiguities.
-For the likelihood-based inference algorithms for the coalescent with recombination,
-it is sufficient to create \emph{two} gARG nodes at the time of the recombination event:
-see the appendix of \cite{baumdicker2021efficient} for details of evaluating
-coalescent with recombination likelihoods using this convention.
-This is also the interpretation depicted in
-Figure \ref{hudson_vs_bigARG}, but it means that the two edges above node \noderef{b}
-in Figure \ref{fig-ancestry-resolution} should correspond to only one lineage,
-along which all $m-1$ links are available for recombination.
-The lineage then splits into two at the time of nodes \noderef{d} and \noderef{e}.
-Nodes \noderef{k}, \noderef{l}, and \noderef{m} in Figure \ref{fig-ancestry-resolution}
-demonstrate that the same issue can affect the number of links available for recombination:
-without an external convention, the exact time at which the trapped ancestral material on
-node \noderef{k} ceases to be available for an effective recombination in the Little ARG process.
+For example, in the gARG encoding depicted in Figure \ref{fig-ancestry-resolution}C,
+it is clear that a recombination takes place between nodes \noderef{i}, \noderef{k} and \noderef{j}.
+But the exact time of the recombination event is ambiguous: it could take place at any time between node \noderef{i} and its parents and produce the same gARG.
+Because a recombination increases the number of extant lineages by one (in the rootward direction of time), the number of lineages during the same time interval is ambiguous as well.
+In fact, this information cannot be recovered from the gARG encoding used in Figure \ref{fig-ancestry-resolution}C without an extrinsic convention.
+For the coalescent,
+it is sufficient to create \emph{two} gARG nodes at the time of the recombination event, with the interpretation that the two rootward edges from node \noderef{i} in Figure \ref{fig-ancestry-resolution}C belong to the same lineage until the time of nodes \noderef{k} and \noderef{j}, and split into two separate lineages at that time point.
+Similarly, the trapped, non-ancestral links along that lineage remain available for effective recombination (i.e.\ one which splits up ancestral material) for the same time interval.
+This interpretation is highlighted in
+Figure \ref{hudson_vs_bigARG} by drawing only one vertical edge between a recombinant child and its two parents.
 

### 2023-10-26 paper.tex

@@
 \label{sec-example-inferred-args}
+The scalability gains made by recent ARG inference methods
+such as \relate~\citep{speidel2019method} and
+\tsinfer~\citep{kelleher2019inferring} have
+have been, in part, due to inferring lower levels of precision about
+recombination than classical methods.
+Neither method infers
+explicit recombination events, and therefore their outputs cannot
+be described using the classical eARG formalisms (Section~\ref{sec-eARG}).
+% Could do, something like
+% this mismatch has led some to characterise their output as a
+% series of unrelated local trees [cite], which lacks nuance of the reality.
+% Or maybe we don't bother: what's to be gained from pointing out people's
+% mistakes here? Hopefully the rich detail of reality will come through
+% and replace old ideas.
+Nonetheless, both methods produce estimates in which nodes and edges
+persist across multiple trees, naturally captured by the gARG approach.
+To illustrate the varying levels of information captured
+by current methods, and some qualitative differences between them,
+Fig.~\ref{fig-inferred-args} shows graphical depictions of
+example ARGs produced by four tools using substantially
+different inference strategies.
+
+% Indeed, a
+% common statement in the literature is that modern tools such as
+% \tsinfer~\citep{kelleher2019inferring} and \relate~\citep{speidel2019method}
+% construct not an ARG but ``only'' a sequence of local trees
+% \citep[e.g.][]{hejase2020summary, brandt2021evaluation, ignatieva2021kwarg}.
+% While this statement is correct inasmuch as neither method attempts to infer
+% the explicit recombination events that cause adjacent trees to differ, it is
+% certainly not the case that they infer independent local trees. In fact, both
+% methods take advantage of genome-wide mutation information to construct
+% genealogies, and as a result nodes and edges persist across multiple local
+% trees. Their results can therefore be thought of as an inheritance graph, and
+% represented in the gARG format.
 
@@
 region of the ADH locus~\citep{kreitman1983nucleotide}.
-Results for four different methods based on
-% FIXME: this is the wrong place to be summarising. Name the methods
-% here and summarise in the text
-(A) parsimony, (B) sampling the SMC$^\prime$, (C) copying from
-inferred ancestral haplotypes, and (D) correlated local tree estimation.
+Results for four different methods:
+(A) \kwarg; (B) \argweaver; (C) \tsinfer; and (D) \relate. See the
+text for details of these methods.
 Edge colours are based on the time of the edge's child node
 (lighter: older; darker: younger).
-X and Y positions of graph nodes are arbitrary;
-line width and node colour are as described in Fig.~\ref{fig-simplification}.
-Graphics on the bottom row show the genome positions, relative to the start of the ADH gene,
-for each graph edge from the corresponding ARG. Edge intervals are drawn as horizontal lines,
-stacked in time order (edges with youngest children at the bottom);
-vertical dashed lines indicate breakpoints between local trees.
+Vertical and horizontal positions of graph nodes are arbitrary.
+Line width and node colour are as described in Fig.~\ref{fig-simplification}.
+Graphics on the bottom row show the genome positions, relative to the start of
+the ADH gene, for each graph edge from the corresponding ARG. Edge intervals
+are drawn as horizontal lines, stacked in time order (edges with youngest
+children at the bottom); vertical dashed lines indicate breakpoints between
+local trees.
 }
@@
 
-The scalability gains made by recent ARG inference methods have been, in part, due to inferring
-lower levels of precision about recombination. Indeed, a common statement in the literature is
-that modern tools such as \tsinfer~\citep{kelleher2019inferring} and \relate~\citep{speidel2019method}
-construct not an ARG but ``only'' a sequence of local
-trees \citep[e.g.][]{hejase2020summary, brandt2021evaluation, ignatieva2021kwarg}. While this statement
-is correct inasmuch as neither method attempts to infer the explicit recombination events that cause
-adjacent trees to differ, it is certainly not the case that they infer independent local trees.
-In fact, both methods take advantage of genome-wide mutation information to construct genealogies,
-and as a result nodes and edges persist across multiple local trees.
-Their results can therefore be thought of as an inheritance graph, and represented in the gARG format.
-
-To illustrate the varying levels of information that can be captured in a gARG,
-Fig.~\ref{fig-inferred-args} presents graphical representations of genealogies output by two tools that
-\emph{do} explicitly infer recombination events, namely \kwarg~\citep{ignatieva2021kwarg} and
-\argweaver~\citep{rasmussen2014genome},
-and two which do not (\tsinfer\ and \relate).
-All inferences are based on the
-same underlying dataset from \citet{kreitman1983nucleotide}, a standard benchmark in the classical
-ARG literature, for which the minimum number of recombination breakpoints in the absence of recurrent
-mutation is 7~\citep{song2003parsimonious}. This dataset comprises 43 biallelic SNPs spanning 2.4Kb of the
-\emph{D.~melanogaster} ADH locus on chromosome 2L. Where required for inference purposes we
-assume a mutation rate of
-$5.49\times10^{-9}$ per site per generation
-and a recombination rate of
-$8.4\times10^{-9}$ per site per generation
-\citep{adrion2020community,lauterbur2022expanding},
-together with
-a constant effective population size of 1,000,000.
-To focus on the ARG topology rather than estimates of node times, the Y position of each node
-is arbitrary, taken from the ranked order of nodes output by each software method.
-The version numbers of inference software, the specific inference parameters
-needed to reproduce the ARGs, and (where necessary) functions to convert outputs to the
-\texttt{tskit} gARG format, are available from our GitHub repository.
-
-Note that both \kwarg\ and \argweaver\ produce many ARGs
-(those shown in Fig.~\ref{fig-inferred-args}A,B are arbitrarily selected examples).
-This stems from the fact that \kwarg\ searches the space of plausible ARGs,
-outputting minimal ones using heuristics, and \argweaver\ samples
-from the posterior distribution over
-ARGs given the data and a fixed model (a discretised version of the SMC$^\prime$ \citep{marjoram2006fast},
-an approximation of the full coalescent with recombination).
-In contrast, \tsinfer\ and \relate\ (Fig.~\ref{fig-inferred-args}C,D) are
-deterministic methods, aiming to find a single ``best guess'' ARG.
-The ability of a gARG to encode the effects of recombination, rather
-than having to precisely specify each event, is an important contributor to this aim,
-as most recombination events cannot be uniquely identified from the data.
-Indeed, there is considerable variation in the number of recombinations inferred by
-different methods, with e.g.~\argweaver\ (Fig.~\ref{fig-inferred-args}B) suggesting
-far more than are required under minimal parsimony assumptions (Fig.~\ref{fig-inferred-args}A).
+The first two methods explicitly infer recombination events.
+\kwarg~\citep{ignatieva2021kwarg} is a parsimony based approach
+which searches the space of plausible ARGs,
+outputting minimal ones using heuristics.
+\argweaver~\citep{rasmussen2014genome} on the other hand is
+model-based, sampling from a discretised version of the
+% could spell this out, but in reality everyone knows what the SMC
+% is, and if they don't spelling it out won't help them.
+SMC~\citep{mcvean2005approximating,marjoram2006fast}.
+Note that both \kwarg\ and \argweaver\ produce many ARGs,
+and those shown in Fig.~\ref{fig-inferred-args} are arbitrarily selected
+examples.
+While the second two methods
+both produce a single best-guess estimate and
+do not explicitly infer recombination events, they are based on
+quite different principles.
+\texttt{Tsinfer} works in a two-step process, first generating ancestral
+haplotypes via heuristics and then inferring inheritance relationships
+between them using the Li and Stephens model~\citep{li2003modeling}.
+\texttt{Relate} works by [TODO: Ana can you fill in please?]
+See Appendix~\ref{sec-survey-arg-infer} for more details on these
+and other inference methods.
+
+Inferred ARGs are based on the \citet{kreitman1983nucleotide} dataset,
+a standard benchmark in the classical ARG
+literature.
+It consists of 43 biallelic SNPs spanning 2.4Kb of the \emph{D.~melanogaster} ADH
+locus on chromosome 2L. Where required for inference purposes we assume a
+mutation rate of $5.49\times10^{-9}$
+and recombination rate of $8.4\times10^{-9}$ per site per generation
+% NOTE: we should be citing the original sources here too.
+\citep{adrion2020community,lauterbur2023expanding},
+% Note: would have been simpler to use the stdposim estimate
+% of 1.7M from Li et al 2006. (not that it matters)
+and constant effective population size of $10^6$.
+Full details of the software used and code to generate these
+figures are available on GitHub (see Data Availability).
 
 % Here we describe recombination features
-A rough feel for the amount of recombination in each ARG is given by the distribution
-of number of parents per node, as highlighted by the node colour scheme.
-In Fig.~\ref{fig-inferred-args}A,B, each recombination event is tagged by
-a node with exactly 2 parents and one child. Assuming that the breakpoint for each event
-% TODO "infinite breakpoints" is goign to confuse people. What are we
-% clarifying here? Citation necessary if we keep.
-occurs at a different location (the ``infinite breakpoints'' model), then the
-number of breakpoints should exactly equal the number of 2-parent (brown) nodes in the plot.
-In contrast, Figure Fig.~\ref{fig-inferred-args}C,D have ARG nodes with multiple
-parents and multiple children. These indicate deliberate imprecision, where one
-or more recombinations have been ``pushed down'' onto a more recent, identifiable node.
-Several of these nodes have more than two parents (in red), representing
-multiple recombinations inferred to have occurred in an unresolvable order.
-Nevertheless, the number of breakpoints in these two ARGs is smaller than the
-number of multiple-parent ARG nodes, showing that several multiple-parent nodes
-must share breakpoint positions. The sharing of breakpoint locations
-between these nodes indicates that in \tsinfer\ and \relate\ ARGs it may be
-difficult to condense each transition between trees into a set of simple SPR moves.
-Overall, Fig.~\ref{fig-inferred-args} shows that \tsinfer\ and \relate\ ARGs contain a level
-of detail about recombination that lies somewhere between a sequence of unrelated local
-trees on one extreme and an ARG with precisely specified recombination events on the other.
+Considering~Fig~\ref{fig-inferred-args}, we can see that
+there is considerable variation in the number of recombination breakpoints
+inferred by different methods, with e.g.~\argweaver\ suggesting
+far more than the 7 required for this dataset under minimal parsimony
+assumptions~\citep{song2003parsimonious}.
+A sense of the amount of recombination in each ARG is provided by
+the node colouring scheme, which shows the number of parents for each node.
+In Fig.~\ref{fig-inferred-args}A,B, each recombination event
+corresponds to a node with exactly 2 parents and one child.
+As these methods explicitly infer a recombination event for each
+breakpoint, the number of breakpoints equals the number of 2-parent
+(brown) nodes.
+In contrast, Fig.~\ref{fig-inferred-args}C,D do not have this
+straightforward
+relationship between the number of nodes with multiple parents
+and breakpoints along the genome.
+In both ARGs the number of breakpoints is smaller
+than the number of multiple-parent ARG nodes, showing
+that several multiple-parent nodes must share breakpoint positions.
+There are also ARG nodes with multiple parents and multiple children,
+where one or more recombinations have been pushed down onto a
+more recent node.
+% Several of these nodes have more than two
+% % parents (in red), representing multiple recombinations inferred to have
+% occurred in an unresolvable order.
+% Nevertheless, the number of breakpoints in
+% these two ARGs is smaller than the number of multiple-parent ARG nodes, showing
+% that several multiple-parent nodes must share breakpoint positions.
+As a consequence, it may be difficult to condense each transition between trees
+in these ARGs into a set of SPR operations.
+% The sharing
+% of breakpoint locations between these nodes indicates that in \tsinfer\ and
+% \relate\ ARGs it may be difficult to condense each transition between trees
+% into a set of simple SPR moves.
+% Overall, Fig.~\ref{fig-inferred-args} shows
+% that \tsinfer\ and \relate\ ARGs contain a level of detail about recombination
+% that lies somewhere between a sequence of unrelated local trees on one extreme
+% and an ARG with precisely specified recombination events on the other.
 
 % Here we move on to coalescence
-As for inferred coalescences, the shading (``coalescent span'')
-reveals that many non-recombinant (green) nodes are
-locally unary (see Section \ref{sec-locally-unary-edges}). In other words,
-in some regions of the genome these nodes are present in the ancestry without
-corresponding to coalescence of ancestral material. As argued previously, this can provide useful
-information about shared ancestry. Note however, that such unary segments are not inferred
-in the \relate\ ARG: other than the samples it has entirely solid nodes.
-Another clear difference involves coalescent nodes with more than two children (polytomies);
-in Appendix \ref{sec-cell-lineages-and-args} we argue that this reflects
-uncertainty in the order of coalescent events, in an equivalent
-manner to the imprecision associated with recombination nodes having more
-than two parents. Only \tsinfer\ creates such polytomies, the clearest of which
-involves the nodes \noderef{Fr-F}, \noderef{Wa-F}, and \noderef{Af-F}. These
-happen to have identical sequences, but because \kwarg, \argweaver, and \relate\ require
-bifurcating trees, they are forced to pick an arbitrary order of branching
-(hence Fig.~\ref{fig-inferred-args}A,B disagrees with Fig.~\ref{fig-inferred-args}D
-in this order).
-More generally, the number of nodes in Fig.~\ref{fig-inferred-args}C
-that have more than two children hints at considerable uncertainty in
-the order of branching elsewhere in the ARG.
-
-%% possible TODO - mention something about multiple local roots in
-% tsinfer and Relate versus the other two.
-
-%The example of successive simplification of a simulated ARG
-%in Fig.~\ref{fig-simplification} illustrates the point
-%that is is possible to \emph{store} arbitrary levels of precision
-%about recombination in a gARG.
-%It is also possible to \emph{infer} various levels of precision about
-%recombination.
-%To illustrate this, and to demonstrate the structural
-%heterogeneity of the ARGs inferred  by current methods,
-%Fig.~\ref{fig-inferred-args} shows the results of ARG inference
-% To illustrate these points and to emphasise the heterogeneity
-% of the ARGs inferred by current methods, we inferred ARGs for the
-%for the classical \citet{kreitman1983nucleotide} benchmark dataset
-%using  \kwarg~\citep{ignatieva2021kwarg},
-%\argweaver~\citep{rasmussen2014genome,hubisz2020inference},
-%\tsinfer~\citep{kelleher2019inferring},
-%and \relate~\citep{speidel2019method}.
-% While there is some consistency among the inferred ARGs (such as the samples
-% FR-F, WA-F and Af-F being grouped into the same clade), there
-% are also considerable differences.
-%The minimum number
-%of recombination breakpoints required in the absence of recurrent mutation
-%is 7 for this dataset~\citep{song2003parsimonious},
-%and the number of inferred breakpoints
-%under the input mutation and recombination rates
-%varies from 2 (\relate) to 17 (\argweaver).
-% TODO this needs a bit of rewriting with the rest of the paper in mind
-% kwarg and argweaver are inferring eARGs, but we need to be careful
-% not to confuse people.
-%The ARGs inferred by the different methods also have major differences
-%in how recombination information is represented.
-%Essentially,
-%\kwarg\ and \argweaver\ infer eARGs that contain
-%specific recombination \emph{events} for each breakpoint.
-%On the other hand, \tsinfer\ and
-%\relate\ do not infer events but rather
-%infer gARGs with by differing levels of precision about recombination.
-
-% TODO add additional explanation of the bottom row, if enough isn't written below
-% https://github.com/tskit-dev/what-is-an-arg-paper/issues/386
+Shading within nodes in Fig~\ref{fig-inferred-args} indicates the
+fraction of the node's span over which it is coalescent
+(Section~\ref{sec-locally-unary-edges}). For example, brown nodes
+in Fig~\ref{fig-inferred-args}A,B are clear because there is
+no local coalescence at these recombination nodes
+(they are ``graph unary'', and so local coalescence is impossible).
+The significant number of partially shaded nodes in
+Fig~\ref{fig-inferred-args}A,B and C demonstates that
+\kwarg, \argweaver\ and \tsinfer\ all contain locally unary nodes.
+Another difference between methods highlighted in this figure
+is the presence of polytomies, which only
+\tsinfer\ creates. The most obvious example involves
+involves nodes \noderef{Fr-F}, \noderef{Wa-F}, and \noderef{Af-F},
+which happen happen to have identical sequences.
+Because \kwarg, \argweaver, and \relate\ require
+bifurcating trees by design, each picks an arbitrary order of branching
+(hence Fig.~\ref{fig-inferred-args}A,B disagrees
+with Fig.~\ref{fig-inferred-args}D in this order).
+% JK: I don't quite get this, so readers probably won't either. Let's
+% stick a few clear messages.
+% More generally, the number of nodes in Fig.~\ref{fig-inferred-args}C
+% that have more than two children hints at considerable uncertainty in
+% the order of branching elsewhere in the ARG.
+
 The bottom row of Fig~\ref{fig-inferred-args} shows the extent along
 the genome to which graph edges are shared between multiple trees.
-This illustrates that both \tsinfer\ and (to a lesser extent) \relate\ have edges
+All four methods infer nodes and edges that are shared between
+multiple tree, to varying degrees.
+For example, all of the methods infer
+infer that \noderef{Af-f}, \noderef{Fr-f}, and \noderef{Wa-f}
+form a clade along the entire sequence.
+In particular, we can
+see both \tsinfer\ and (to a lesser extent) \relate\ have edges
 that span multiple tree boundaries, indicating that they are not inferring
 a series of \emph{unrelated} local trees.
-%(i.e., with either unlabelled internal
-%nodes or differently labelled nodes in each tree).
-For example, both infer that \noderef{Af-f}, \noderef{Fr-f}, and \noderef{Wa-f}
-share a common ancestor along
-the entire sequence. However, in comparison to \kwarg\ and \argweaver\ neither
+However, in comparison to \kwarg\ and \argweaver\ neither
 method results in extensive node sharing in the
-oldest time periods. This primarily reflects the fact that they do not encode topology
-above the local MRCAs, also indicated by multiple roots in the ARG
-(nodes with zero parents, in blue), compared to the
-single ARG root in the \kwarg\ and \argweaver\ genealogies. In terms of inference,
-\tsinfer\ and \relate\ show a pragmatic representation: we do not expect
-genetic variation to be very informative about topological relationships
-between nodes that that are older than the local MRCAs.
+oldest time periods.
+Overall, Fig.~\ref{fig-inferred-args} shows
+that \tsinfer\ and \relate\ ARGs contain a level of detail
+that lies somewhere between a sequence of unrelated local trees on one extreme
+and an ARG with precisely specified recombination events on the other
+(Fig~\ref{fig-inferred-args}A,B).
+% JK: Let's not get into this.
+% This primarily reflects the fact that they do not encode topology
+% above the local MRCAs, also indicated by multiple roots in the ARG
+% (nodes with zero parents, in blue), compared to the
+% single ARG root in the \kwarg\ and \argweaver\ genealogies. In terms of inference,
+% \tsinfer\ and \relate\ show a pragmatic representation: we do not expect
+% genetic variation to be very informative about topological relationships
+% between nodes that that are older than the local MRCAs.
 

### 2023-10-27 paper.tex
Updates from Yan, Gregor, Jerome

@@
 where the stochastic process of coalescence and recombination
-among ancestral lineages is formalised as a
+among ancestral lineages was formalised by Griffiths and colleagues as a
 graph~\citep{griffiths1991two,ethier1990two,griffiths1996ancestral,griffiths1997ancestral}.
@@
 structure~\citep{minichiello2006mapping}, i.e.\ describing
-a \emph{realisation} of such a random process or an
+a \emph{realisation} of such a random process or of an
 inference from a sample of genome sequences.
@@
 stochastic process, and expressed in terms of events:
-the coalescence and recombination events
+the common ancestor and recombination events
 that occur in the history of a sample. We show that this approach
-to representing ARGs (which we refer to as the ``event ARG'', or eARG
+to representing ARGs (which we call the ``event ARG'', or eARG
 encoding) is inherently limiting, insufficient to represent
@@
 We address these problems by providing a more flexible formulation
-of an ARG data structure, which we refer to as the ``genome ARG''
+of an ARG data structure, which we call the ``genome ARG''
 or gARG encoding. We show that it is fully compatible with the
 classical eARG approach, while providing much more flexibility
-in the types of genetic inheritance structures that can be encompassed,
-including the outputs of modern methods.
-In particular we show that, in contrast with the eARG encoding which
+in how genetic inheritance can be represented,
+and encompasses the outputs of modern methods.
+Moreover we show that, in contrast with the eARG encoding which
 requires precise estimates of each recombination event,
 the gARG encoding allows us to systematically describe uncertainty
-the temporal ordering of multiple recombinations.
+about the temporal ordering of multiple recombinations.
 
@@
 in the context of current datasets and research questions.
-We then discuss the import concept of ``ancestral material''
+We then discuss the important concept of ``ancestral material''
 in Section~\ref{sec-ancestral-material}, and how this relates
@@
 In order to capture the effects of recombination
-(and other processes that lead to partial genome inheritance)
  we ``annotate'' the edges with the genome
 coordinates over which inheritance occurred.
-This is sufficient to capture the effects of any form of
-genetic inheritance (excluding structural variation).
+This is sufficient to describe the effects of inheritance under
+any form of homologous recombination (such as multiple crossovers,
+gene conversion events, and many forms of bacterial and viral recombination).
 
@@
 the gARG for a given set of individuals is embedded in their pedigree.
-The figure shows the pedigree of eight diploid individuals, along with paths
-of genetic inheritance and constituent genomes.
+The figure shows the pedigree of eight diploid individuals
+and their sixteen constituent genomes (each consisting of a single chromosome),
+along with paths of genetic inheritance.
 Here, and throughout,
@@
 (but see~\citet{parida2011minimal} and \citet{zhang2023biobank}
-for notable exceptions), and the majority of
+for notable exceptions), and a large number of
 % Note the "an" here to avoid cases like ARGweaver
 ARG inference methods use it as an output format
-\citep[(TODO exhaustive list)][]{rasmussen2014genome,ignatieva2021kwarg}.
-In this Griffiths encoding, as well as sample nodes,
-we have two additional types of node in the graph,
+\citep{song2004minimum,song2005efficient,rasmussen2014genome,heine2018bridging,ignatieva2021kwarg}.
+In this Griffiths encoding, as well as nodes marking sampling events,
+we have two additional node types in the graph,
 representing the common ancestor and recombination events
@@
 encoding. Fig.~\ref{fig-event-arg} shows an example of a classical
-eARG with three sample genomes (\noderef{a}, \noderef{b}, and \noderef{c})
+eARG with three sample genomes (\noderef{a}, \noderef{b}, and \noderef{c}),
+three common ancestor events (\noderef{e}, \noderef{f}, and \noderef{g})
 and a single recombination event (node \noderef{d}) with a breakpoint
@@
 while node \noderef{e} is blank
-as it carries \emph{non ancestral} material.
+as it carries \emph{non-ancestral} material.
 This process of local coalescence continues through the
@@
 forward-time simulation. 
+% Define "sample-resolved" here for later use
+We refer to a gARG that has been simplified with respect to a set of 
+samples, such that the inheritance annotations on its edges contain
+no non-ancestral material,  as sample-resolved.
 
@@
 Converting the \citet[][Fig.~1]{wiuf1999recombination} example
-to a sample-resolved gARG. (A) The original eARG; square nodes represent events, with
-each recombination (red) containing a breakpoint.
+to a sample-resolved gARG. (A) The original eARG; square nodes 
+represent sampling (black), common ancestor (blue), and recombination (red) events;
+the latter contain breakpoint positions.
 (B) The corresponding gARG with breakpoints directly converted to
@@
 Coalescence with respect to the sample is indicated by shaded bars, as
-in Fig~\ref{fig-arg-in-pedigree}A; nodes \noderef{n}, \noderef{o}, \noderef{p}, \noderef{q} have incomplete
+in Fig.~\ref{fig-arg-in-pedigree}A; nodes \noderef{n}, \noderef{o}, \noderef{p}, \noderef{q} have truncated
 bars showing that local ancestry of entirely coalesced regions is omitted.
@@
 
-Any eARG can be converted to a sample-resolved gARG using the same machinery,
-and the two-step process is illustrated in Fig.~\ref{fig-ancestry-resolution}.
+Any eARG can be converted to a sample-resolved gARG
+via a two-step process illustrated in Fig.~\ref{fig-ancestry-resolution}.
 The first step is to take the input eARG (Fig.~\ref{fig-ancestry-resolution}A),
@@
 inheritance intervals we get on the edges in a prospective gARG
-produced by a forwards-time simulation, which are concerned with recording
-the local genetic relationship between a parent and child genome and not
+produced by a forward-time simulation, which are concerned with recording
+the direct genetic relationship between a parent and child genome and are not
 necessarily minimal in terms of the resolved ancestral material of a sample.
-Thus, the final step is to use the ``simplify'' algorithm to perform this
-resolution and produce the sample-resolved
-gARG (Fig.~\ref{fig-ancestry-resolution}C).
+Thus, the final step is to use the ``simplify'' algorithm to perform the required
+sample resolution (Fig.~\ref{fig-ancestry-resolution}C).
 
@@
 sample-resolved gARG because all segments of the genome have
-fully coalesced before it is reached. Likewise, the edge
+fully coalesced in \noderef{k} and \noderef{p} before \noderef{q} is reached.
+Likewise, the edge
 between \noderef{g} and \noderef{j} is omitted because the recombination
@@
 of an ARG in a way that is not possible in an eARG.
-Because the ancestral material is stored with each edge, the
+Because the ancestral material is stored with each edge of a gARG, the
 cumulative effects of events over time can be reasoned
@@
 in Fig.~\ref{fig-ancestry-resolution} is not particularly
-representative, because inference or simulation methods will usually
+representative, because inference or simulation methods usually
 only generate ARGs containing nodes and edges ancestral to the sample
-(see the discussion of the ``Big ARG'' stochastic process in
-Appendix~\ref{sec-big-and-little-arg}, however).
+(but see the discussion of the ``Big ARG'' stochastic process in
+Appendix~\ref{sec-big-and-little-arg}).
 Nonetheless, it is an instructive example from the literature which highlights several
@@
 (Appendix~\ref{sec-cell-lineages-and-args}).
-These chains of single-parent inheritance over time give rise,
-by definition, to a tree structure.
+By definition, these chains of single-parent inheritance give rise,
+over time, to a tree structure.
 As a result of recombination, adjacent nucleotides can have
-different chains of inheritance, and
+different paths of inheritance, and
 an ARG encodes the entire ensemble of local trees along the
@@
 % different ways in the same sentence then.
-genome for a given sample.
+genome for a given set of sample nodes.
 % This is for Yan - to me this is a given, but maybe some people feel
@@
 Precisely defining the process by which local trees are extracted
-from an ARG is essential to our understanding of how they
+from an ARG is essential to our understanding of how ARGs and local trees
 are related, and we require a concrete mathematical structure
@@
 If we are given the sequence of local trees for a gARG
-encoded as an oriented tree along with the genome interval
+encoded as an oriented tree, along with the genome interval
 covered by each tree, then we can recover precisely the same ARG.
@@
 then simply combine the ``local ARGs'' for each position $x$
+% YW: The above "combining" operation isn't obviously lossless to me: 
+% we could potentially lose information in the different
+% combinations of intervals (e.g. if we use 2 non-combined intervals
+% to represent a diamond in a 1-RE-node ARG). It seems like we might
+% need an additional gARG spec to define what combinations are equivalent.
+% But no-one else has commented on this, so I guess we just let it lie.
 to recover precisely the same set of edges as the original ARG.
@@
 tree and ARG nodes clear.
-
 The second property of how evolutionary trees
 are conventionally represented that is unhelpful in the context of ARGs is their
-focus on branching points (coalescences), i.e.\
-nodes that have two or more children.
+focus on branching points (coalescences), i.e.\ nodes that have two or more children.
 As the introductory paragraph of this section emphasised,
@@
 The local tree at a given position records the \emph{path} through
-the ARG, and if we discard sections of that path that are not
-branching points in the local tree, we lose information about the ARG.
-We return to this important point in the following two sections.
+the ARG; if this path omits nodes that are not
+branching points in the local tree (such as \noderef{e} in Fig.~\ref{fig-arg-in-pedigree}),
+we lose information about the ARG.
+We return to this point in the following two sections,
+when we discuss ``locally unary'' nodes and the simplification process.
 
@@
 trees that we can derive from a known ARG (as just discussed),
-and the ARGs that we can derive from a sequence of \emph{estimated}
+and an ARG that we can derive from a sequence of \emph{estimated}
 local trees.
@@
 A similar problem is tackled by \arbores~\citep{heine2018bridging}, which explores a space of ARGs by fixing a starting ARG, and then repeatedly erasing and resimulating candidate SPR moves connecting two fixed, leaf-labelled local trees.
-Exploration is done by varying the positions of these boundary trees along the sequence.
 Inferring the SPRs between leaf-labelled trees is
@@
 
-This distinction between ARG and local tree arity is mainly
+This distinction between ARG and local arity is mainly
 of interest when we consider those that have
 a single child: that are \emph{unary}.
-Returning to the example in Fig~\ref{fig-arg-in-pedigree}, nodes
+Returning to the example in Fig.~\ref{fig-arg-in-pedigree}, nodes
 \noderef{g} and \noderef{h}
-are ARG-unary (Fig~\ref{fig-arg-in-pedigree}B), and are consequently
-also unary in the local trees (Fig~\ref{fig-arg-in-pedigree}C).
+are ARG-unary (Fig.~\ref{fig-arg-in-pedigree}B), and are consequently
+also unary in the local trees (Fig.~\ref{fig-arg-in-pedigree}C).
 On the other hand, node \noderef{f} has two children
@@
 in the local tree. Thus, \noderef{f} has a single child in this
-interval and it is unary in this local tree: in other words it is a \emph{locally unary} node.
-% TODO note that node e is CA but not coalescent, and refer to where we discuss this.
-In another example, \noderef{e} is binary in the graph, but is
+interval: in other words it is a \emph{locally unary} node.
+In another example, \noderef{e} is binary in the graph, being a common
+ancestor of \noderef{a} and \noderef{c}, but is
 locally unary in all trees in which it is present.
-This is because \noderef{a} and \noderef{c} are both recombinant
-offspring of \noderef{e} and \noderef{f}, and,
-in this example,
-\noderef{a} inherits from
-\noderef{e} over the interval $[0, 2)$,
-and \noderef{c} inherits from \noderef{e} over $[7, 10)$.
+This is because no ancestral material coalesces there:
+\noderef{a} inherits genetic material from the far left hand end of
+\noderef{e},
+while \noderef{c} only inherits the (disjoint) right hand end.
 
@@
 
-% NOTE: this para doesn't serve much purpose now I think.
-
-% Nodes with one child are not a standard feature of evolutionary trees
-% or the ARG literature
-% (although \citet{mathieson2020ancestry} show examples in which the ARGs
-% contain unary nodes, and state that paths are
-% ``implicitly passing through points representing specific individuals.'').
-% We are usually interested in
-% nodes that have at least two children because these
-% are in principle detectable from mutational differences
-% arising in the respective subtrees.
-% A tree branch may correspond to many generations
-% of individuals (or indeed cell divisions; see Appendix \ref{sec-cell-lineages-and-args})
-% and there is usually
-% little information about these intermediates, and hence limited utility
-% in modelling them.
-% As we see in the next section, however, there is a range of
-% different ways in which ARG nodes can be locally unary. Furthermore,
-% several inference methods infer ARGs containing locally unary nodes
-% (see Section~\ref{sec-precision}).
-
 \section{Levels of simplification}
@@
 of removing nodes and re-writing edges (and their inheritance annotations)
-to remove various types of redundancy.
-The redundancy that we are interested in
+to remove various types of redundancy, much of which
 % We are interested in both graph and local unary, so qualifying this
@@
 in Fig.~\ref{fig-simplification};
-the graph is shown on the left and the corresponding sequence
+the gARG representation is shown on the left and the corresponding sequence
 of local trees on the right.
@@
 Levels of ARG simplification.
-(A) An example ARG simulated from a diploid Wright-Fisher model.
+(A) An example gARG simulated from a diploid Wright-Fisher model.
 (B) Remove all
@@
 using a single node to represent
-a recombination event (e.g.~Fig~\ref{fig-ancestry-resolution}),
+a recombination event (e.g.~Fig.~\ref{fig-ancestry-resolution}),
 or two to represent the parent genomes is usually not important
 (either is possible in the gARG encoding)
-and the most convenient approach will vary by application.
-Note also that polytomies are a natural feature of such
-a Wright-Fisher model
-(e.g.~node \noderef{k} in Fig.~\ref{fig-simplification}).
+and the most convenient approach will vary by application 
+(discussed e.g.\ in Appendix~\ref{sec-big-and-little-arg}).
+Note also that polytomies (e.g.~Fig.~\ref{fig-simplification} node \noderef{k})
+are a natural feature of such a Wright-Fisher model 
+(but see Appendix~\ref{sec-cell-lineages-and-args}).
 
@@
 coalescent span of 0\%.
-% JK: Let's not get into discussions of identifiability or inferability
-% here, we're just trying to show the things that exist in the true history.
-% (such nodes will be extremely difficult to
 
 Returning to the main topic of this section,
-Fig.~\ref{fig-simplification}A is the direct output of a Wright-Fisher
-simulation in which we retain all nodes involved in recombination
+Fig.~\ref{fig-simplification}A is original simulation output in 
+which we retain all nodes involved in recombination
 or common ancestry events. This is the true history, and contains
@@
 until we reach Fig.~\ref{fig-simplification}D, where are none.
+A more detailed version of the displayed ARGs, showing the edge
+intervals, is provided as Supplementary Fig.~\ref{fig-simplification-with-edges}.
 
@@
 nodes \noderef{f} and \noderef{h}.
+% Note that f and h are only removable because we are using the 2-RE-node
+% representation. That means when we remove these nodes, we lose information about
+% the time of recombination, which we wouldn't do if we were using the
+% 1-RE-node representation (because the node would be multiply connected
+% in the rootward direction). We definitely don't want to get into this
+% level of detail here though, so we'll just pass over it without comment.
 Fig.~\ref{fig-simplification}B shows the result of this type of
@@
 common ancestor but non-coalescent nodes can produce graph nodes with
-more than two children (e.g.~node \noderef{s}). These represent the
+more than two children (e.g.~node \noderef{s}). Both cases represent the
 merged
 \emph{effects} of multiple evolutionary events in a single node (genome), and the
-ARG no longer contains the intermediate genomes representing those events.
+ARG no longer contains the intermediate genomes corresponding to
+those events (see also Appendix~\ref{sec-cell-lineages-and-args}).
 
@@
 between them using the Li and Stephens model~\citep{li2003modeling}.
-\texttt{Relate} works by [TODO: Ana can you fill in please?]
+\texttt{Relate} first reconstructs local tree topologies across the genome, using a variant of the Li and Stephens model to estimate the ordering of coalescence events in each tree, and then estimates branch lengths using MCMC with a coalescent-based prior.
 See Appendix~\ref{sec-survey-arg-infer} for more details on these
@@
 % Here we describe recombination features
-Considering~Fig~\ref{fig-inferred-args}, we can see that
+Considering~Fig.~\ref{fig-inferred-args}, we can see that
 there is considerable variation in the number of recombination breakpoints
@@
 % Here we move on to coalescence
-Shading within nodes in Fig~\ref{fig-inferred-args} indicates the
+Shading within nodes in Fig.~\ref{fig-inferred-args} indicates the
 fraction of the node's span over which it is coalescent
 (Section~\ref{sec-locally-unary-edges}). For example, brown nodes
-in Fig~\ref{fig-inferred-args}A,B are clear because there is
+in Fig.~\ref{fig-inferred-args}A,B are clear because there is
 no local coalescence at these recombination nodes
-(they are ``graph unary'', and so local coalescence is impossible).
+(they are ``ARG unary'', and so local coalescence is impossible).
 The significant number of partially shaded nodes in
-Fig~\ref{fig-inferred-args}A,B and C demonstates that
+Fig.~\ref{fig-inferred-args}A,B and C demonstrates that
 \kwarg, \argweaver\ and \tsinfer\ all contain locally unary nodes.
@@
 All four methods infer nodes and edges that are shared between
-multiple tree, to varying degrees.
+multiple trees, to varying degrees.
 For example, all of the methods infer
@@
 \label{sec-efficiency}
-The gARG encoding discussed defined here
+The gARG encoding defined here
 leads to highly efficient storage and processing of ARG data,
@@
 few minor details in which the formal details of gARGs
-provided here differ from their practical implementation in
+provided in Section~\ref{sec-gARG} differ from their practical implementation in
 \texttt{tskit}.
@@
 This denormalisation of the gARG data model is for efficiency purposes.
-Secondly, $-1$ is used to denote the presence of
-roots in an oriented tree, rather than $0$ as used here for notational
-simplicity.
+Secondly, zero- rather than one-based indexing is used for 
+nodes in ARGs and oriented trees; consequently  $-1$ is used to denote the presence of
+roots (rather than $0$ as used here for notational simplicity).
 
@@
 processes and instead use the term as a more general representation of any
-recombinant genetic ancestry seems useful, and we have
+recombinant genetic ancestry seems useful; we have
 tried to clarify and systematise it here. Thus
@@
 Fully decoupling the general concept of an ARG from the coalescent
-with recombination stochastic process (henceforth, ``coalescent'')
+with recombination (henceforth, ``coalescent'')
 is an important step.
@@
 Thus, although it is clearly an oversimplification to arbitrarily
-choose \emph{one} best guess, it is not clear that sampling more
+choose \emph{one} best guess, 
+it is not clear that generating many guesses
 when sample sizes are large will achieve much.
@@
 sampled continuously through time, sometimes with tens of thousands of sequences collected per day, with relatively little genetic diversity to
-distinguish them. The reconstructed ARGs thus contain polytomies (representing
-both the uncertainty around coalescence times and the presence of true
-multiple mergers), non-leaf sample nodes (sequences with descendants
-also present in the dataset) and unary nodes (samples that have
-one direct descendant).
+distinguish them. The reconstructed ARGs thus contain polytomies
+and non-leaf sample nodes (sequences with descendants also present
+in the dataset) many of which only have a single descendant (i.e. are ARG-unary).
 Recombination is an important factor in the evolution of
@@
 
+\section*{Data Availability}
+The public GitHub repository at 
+\url{https://github.com/tskit-dev/what-is-an-arg-paper} can be
+used to reproduce all figures and tables in this paper. 
+In particular this includes
+the ARG used in Fig.~\ref{fig-ancestry-resolution}; 
+the simulation code and functions
+used to generate Fig.~\ref{fig-simplification}; 
+and for Fig.~\ref{fig-inferred-args},
+the software versions, parameter settings, and 
+(where necessary) functions to convert
+software outputs to the \texttt{tskit} gARG format.
+
 \bibliographystyle{plainnat}
@@
 In fact, this information cannot be recovered from the gARG encoding used in Figure \ref{fig-ancestry-resolution}C without an extrinsic convention.
-For the coalescent,
+For the basic coalescent with recombination,
 it is sufficient to create \emph{two} gARG nodes at the time of the recombination event, with the interpretation that the two rootward edges from node \noderef{i} in Figure \ref{fig-ancestry-resolution}C belong to the same lineage until the time of nodes \noderef{k} and \noderef{j}, and split into two separate lineages at that time point.
@@
 \label{sec-cell-lineages-and-args}
-As we have argued, an important benefit of the gARG representation is that it
-can concatenate multiple events by encoding the outcome of their effects
-on a single node. In the case of recombination events, this manifests itself as
-a single node with three or more parents (e.g.\ Fig.~\ref{fig-simplification}C, node \noderef{e}).
-Here we make the argument that the same principle underlies ARGs containing
-coalescence points that involve three or more children. These ``polytomies'' are generated not only
-by Wright-Fisher dynamics such as those in Fig.~\ref{fig-simplification}C (see node \noderef{k}), but also
-by life histories that involve ``burst'' reproduction events, which are often modelled by a lambda
-coalescent process [AND OTHERS]\citep{wakely2008coalescent}.
+In eukaryotic organisms, ARGs are a result of the cellular processes of mitosis
+(which leads to common ancestry and coalescence)
+and meiosis (which leads to recombination, including both crossovers and gene conversion).
 
@@
 The ARG paths for the lower arm of the sampled chromosomes (green) is highlighted as a
-think solid line and involves a single recombination event and four coalescence events
+thick solid line and involves a single recombination event and four coalescence events
 (highlighted as deep red and blue squares within individuals $D_5$, $D_{10}$, and $D_{13}$).
@@
 
-At the base of this concept is a subtle and important point about which genomes are represented by gARG nodes.
-In principle, we can interpret nodes as representing genomes
-at any point in the life cycle, for instance, \citet{hudson1983properties} views nodes as representing gametes,
-whereas in Fig.~\ref{fig-arg-in-pedigree}A, we present them as genomes within a diploid individual.
-In reality, however, recombination and replication of DNA occur at a cellular level,
-and respectively correspond to the processes of meiosis and mitosis. Therefore, a comprehensive gARG would not
-trace genetic inheritance between diploid individuals, but between cells during the
-life cycle of these individuals. This is illustrated in Fig.~\ref{fig-cell-lines}, where we show a simplified
-schematic of cellular inheritance in a diploid population over four partially overlapping generations.
-In this figure, thick black lines trace an ARG that describes the ancestry of the lower arm of the
-four sampled chromosomes. The ``events'' underlying the ARG are shown as red and blue squares,
-and the potential ARG nodes are shown as green chromosomal regions. Note that a number of these nodes are
-``pass-through'' nodes with one parent and one child, which would normally be omitted.
-
-The unusual feature of this particular example is that three ARG lineages coalesce in one historical individual,
-$D_{10}$, whose three children all happen to be ancestors of the sampled chromosomes.
-We can see that the process that generates these three child lineages actually consists of two successive bifurcations:
-necessitated by the fact that the only known method of reproducing DNA is by (semi-conservative) duplication.
-However, if we take each node in the ARG to represents the generalised genome of an individual (rather than the specific
-genome of a cell within an individual) we end up representing this as a instantaneous trifucation or polytomy.
-
-There is biological significance to this cellular-level inheritance process. Were a mutation to occur
-during the first cell division of $D_{10}$, the two gametes produced by the cells
-in the left half of $D_{10}$ could share a mutation not present in the right hand gamete. In other words,
-the relationship between three (or more) gametes that originate, without recombination, from the same parental
-chromosomes will nevertheless be strictly bifurcating; two of the gametes have a closer relationship with the
-third more distant. Of course, unless mutations occur between successive rounds of cell division, the between-cell
-relationships will be impossible to infer from the data.
-
-% As with concatenating multiple recombination events, the gARG representation allows us to concatenate within-individual
-% coalescence events such that we are not obliged to represent precisely the underlying cellular process. In humans and
-% other multicellular diploids, it seems most intuitive to take the nodes in the gARG to represent two haploid genomes within
-% a newly fertilised diploid individual. If this individual has 3 or more children, and these children feature in the
-% eventual ancestry of the samples, the gARG approach outlined in this paper means that they can be legitimately represented
-% by a polytomy in the local trees.
+In Fig.~\ref{fig-simplification} we provide a schematic of the events (coloured red and blue squares) and
+the genomes (chromosome icons) that occur in the cellular germline of a simplified, diploid multicellular
+hermaphrodite eukaryote with partially overlapping generations. Here, events are not represented by a single
+genome: rather, a genome can be associated with (or ``tag'') an event above or below it.
+The gARG representation allows any such genome to be arbitrarily distant in time or number of cell
+divisions from the event itself (see Section~\ref{sec-precision}).
+
+Thick black lines trace an ARG that describes the ancestry of the lower arm of the
+four sampled chromosomes, and these the lines pass through gARG nodes (green chromosomal regions).
+A number of these nodes have only one parent and one child in the ARG; these would often be omitted
+(taken as implicit) in a minimally simplified gARG.
+
+In particular, this schematic illustrates a case where three ARG lineages coalesce within
+a single individual ($D_{10}$), corresponding to what is normally represented by a polytomy
+(e.g. Fig.~\ref{fig-simplification} node \noderef{k}). The process that generates
+these three lineages actually consists of two successive bifurcations, reflecting the fact that
+the only known method of reproducing DNA is by (semi-conservative) duplication. Under this schematic,
+polytomies therefore represent a gARG-enabled merging of multiple coalescent events into an older
+node, just as ARG nodes with more than 2 parents represent a merging of multiple recombination
+events into a younger node.
+
+It is worth noting that ``true'' polytomies in ARGs are a feature of model assumptions
+rather than biological reality. Indeed, given a large enough mutation rate
+(or single cell sequencing at a sufficiently detailed level) it should be possible to
+distinguish sequential rounds of mitosis within a single multicellular organism.
+For example, a mutation in the first cell division of $D_{10}$ could be shared between
+the two gametes produced by the cells in the left half of $D_{10}$ but not shared by
+the right hand gamete. For an ARG that has been accurately inferred from real data,
+a polytomy is therefore making a (deliberately) imprecise statement about the order
+of coalescence, rather than reflecting of a single instantaneous event; this
+interpretation fits naturally into the gARG framework.
 

### 2023-10-27 paper.tex
Clarify espalier

@@
 further discussion on
-how local trees are embedded in an ARG is provided in Section~\ref{sec-ARG-and-local-trees}
+how local trees are embedded in an ARG is provided in Section~\ref{sec-ARG-and-local-trees}.
 % Following the above definitions, corresponding data encoding for
@@
 (Appendix~\ref{sec-cell-lineages-and-args}).
-By definition, these chains of single-parent inheritance give rise,
-over time, to a tree structure.
+These paths of single-parent inheritance give rise,
+by definition, to a tree structure.
 As a result of recombination, adjacent nucleotides can have
@@
 % significant! ;)
-These statements that an ARG
-contains more information than its local trees are only true
-if we use a representation of the local trees that discards
-information about those local trees that can be derived from the ARG.
+These
+% and similar, \citep[e.g.][]{}
+statements that an ARG
+contains more information than its local trees are true
+if we represent local trees in their conventional form,
+but this form discards important information that is available from the ARG.
 
@@
 represented that lead to this
-% We can rephrase... suggestions?
 disagreement about the relationship between local trees and an ARG.
-Firstly, the internal nodes of evolutionary trees are usually
-considered to be \emph{unlabelled}, or equivalently, labelled by the leaves which they subtend.
-The same canonical labelling cannot be used for internal ARG nodes because the leaves they subtend will typically vary by genomic position.
-If we do not label the tree nodes in a way which is persistent across the sequence of local trees in the ARG, we lose the fact that
-the \emph{same} ancestors sometimes persist across multiple trees.
-Defining ARG nodes as integers and using the oriented tree encoding
-explicitly labels internal nodes, and makes the relationship between
-tree and ARG nodes clear.
+Firstly, the internal nodes of evolutionary trees are usually considered to be
+\emph{unlabelled}, or equivalently, labelled by the leaves which they subtend.
+The same canonical labelling cannot be used for internal ARG nodes because the
+leaves they subtend will typically vary by genomic position. If we do not label
+the tree nodes in a way which is persistent across the sequence of local trees
+in the ARG, we lose the fact that the \emph{same} ancestors sometimes persist
+across multiple trees.
+Defining ARG nodes as integers and using the oriented
+tree encoding explicitly labels internal nodes, and makes the relationship
+between tree and ARG nodes clear and precise.
+
+% Note: good to keep to paras here to logically separate the two important
+% points.
 The second property of how evolutionary trees
@@
 the ARG; if this path omits nodes that are not
-branching points in the local tree (such as \noderef{e} in Fig.~\ref{fig-arg-in-pedigree}),
+branching points (such as \noderef{e} in Fig.~\ref{fig-arg-in-pedigree}),
 we lose information about the ARG.
@@
 across multiple local trees.
-% TODO add a sentence here saying how it does this.
-% that's what we're interested in illustrating here.
-\espalier\ is specifically interested
-in the details of recombination events,
-and also attempts to infer
+\espalier\ approaches this task
+by identifying maximimal subtrees
+that do not change between pairs of adjacent
+local trees and then
+heuristically exploring the space of possible
+rearrangements of these subtrees.
+To derive precise details about recombination events,
+\espalier\ then attempts to infer
 the precise subtree prune-and-regraft (SPR)
@@
 induced by recombination between these partially reconciled local trees.
-A similar problem is tackled by \arbores~\citep{heine2018bridging}, which explores a space of ARGs by fixing a starting ARG, and then repeatedly erasing and resimulating candidate SPR moves connecting two fixed, leaf-labelled local trees.
+% A similar problem is tackled by \arbores~\citep{heine2018bridging}, which
+% explores a space of ARGs by fixing a starting ARG, and then repeatedly erasing
+% and resimulating candidate SPR moves connecting two fixed, leaf-labelled local
+% trees.
 Inferring the SPRs between leaf-labelled trees is

### 2023-10-27 paper.tex
Clarify

@@
 sequences along the genome''.
-% Could cite some more? E.g. KwARG paper says
-% "We note that ARGs contain more information than local trees"
-% If even Ana and Jotun are wrong about this it then it must be
-% significant! ;)
 These
+% % Could cite some more? E.g. KwARG paper says
+% % "We note that ARGs contain more information than local trees"
+% % Would need a few more examples though, which would require some digging.
+%
 % and similar, \citep[e.g.][]{}
@@
 contains more information than its local trees are true
-if we represent local trees in their conventional form,
-but this form discards important information that is available from the ARG.
+if we represent local trees in their conventional forms,
+but these forms discard important information that is
+available in an ARG.
 

### 2023-10-27 paper.tex
Clarify arg-unary

@@
 induced by recombination between these partially reconciled local trees.
+% NOTE: dropping this because it doesn't clarify the actual point.
 % A similar problem is tackled by \arbores~\citep{heine2018bridging}, which
@@
 This distinction between ARG and local arity is mainly
-of interest when we consider those that have
-a single child: that are \emph{unary}.
+of interest when we consider nodes that have
+a single child: those that are \emph{unary}.
 Returning to the example in Fig.~\ref{fig-arg-in-pedigree}, nodes
@@
 but we still record the fact that genome \noderef{c} inherits from \noderef{f}
-in the local tree. Thus, \noderef{f} has a single child in this
-interval: in other words it is a \emph{locally unary} node.
+in the local tree. Thus, node \noderef{f} has a single child in this
+interval: it is \emph{locally unary}.
 In another example, \noderef{e} is binary in the graph, being a common
@@
 locally unary in all trees in which it is present.
-This is because no ancestral material coalesces there:
+This is because no ancestral material coalesces in \noderef{e}:
 \noderef{a} inherits genetic material from the far left hand end of
@@
 
-[FIXME this isn't true, add comment about eARG conversion]
-Outside of the context of simulation, ARG-unary nodes are
-only likely to occur in longitudinal datasets where genetic
+ARG-unary nodes occur when an ARG node has one child and one or
+more parents.
+Nodes with one child and one parent (``pass-through'' nodes)
+occur where we wish the record the passage of ancestral material
+through a known node. For example, in simulations
+it is sometimes useful to record the passage of
+ancestral material through known pedigree individuals
+regardless of whether common ancestry occurs.
+ARG-unary nodes are also occur in inferences
+from longitudinal datasets where genetic
 data is sampled at many timepoints and recombination
 is rare (e.g.~SARS-CoV-2; see Discussion).
-In contrast, locally unary nodes, are a common and important feature
+Nodes with one child and two parents arise when
+we model a recombination event using a single node
+in the classical manner
+(e.g.\ Fig.~\ref{fig-ancestry-resolution}).
+
+Locally unary nodes are a common and important feature
 of many different types of ARG.
@@
 cannot fully reason about how a local tree is embedded in an ARG
-without them. As we see in the next section, both ARG and locally
-unary nodes occur in a range of different scenarios.
+without them. As we see in the next two sections, both ARG and locally
+unary nodes occur in various scenarios, and are
+produced by a range of current inference methods.
 

### 2023-10-27 paper.tex
Clear up a few TODOs

@@
 \caption{\label{fig-arg-in-pedigree}
-% TODO caption needs another pass. Once the full section has been written
-% it'll be clearer what we need to say in the caption vs the text.
 An example genome ARG (gARG) embedded in a pedigree.
@@
 ARG inference methods use it as an output format
-\citep{song2004minimum,song2005efficient,rasmussen2014genome,heine2018bridging,ignatieva2021kwarg}.
+\citep[e.g.][]{song2004minimum,song2005efficient,rasmussen2014genome,
+heine2018bridging,ignatieva2021kwarg}.
 In this Griffiths encoding, as well as nodes marking sampling events,
@@
 within the local tree are included; see Section~\ref{sec-ARG-and-local-trees}
-% TODO: check above if we want to refer to sec-ARG-and-local-trees or some other!
 for a discussion of this important point).
@@
 roots (rather than $0$ as used here for notational simplicity).
-
-% \cite{shipilina2023origin} discuss the idea of a ``haplotype block''
-% (equivalent to the ``bricked tree sequence'' of
-% \cite{nowbandegani2023extremely},
-% % TODO look this up. I expect it's basically the same??
-% and the block-based ARG data structure of
-% \citep{palamara2016argon}),
-% here we consider the unique sets of
-% samples that coalesce at a node in the ARG over a particular
-% genome interval, and the fundamental limitations they place
-% on ARG inference.
-% [FIXME find somewhere natural for this. The last clause is too vague - what
-% are the implications?]
-% Alternative gARG encodings split transmitted intervals and associate each with
-% a new node, resulting in a single genome being represented by a group of
-% multiple nodes (or ``blocks'' \citep{palamara2016argon}), with implications for
-
-%%%% Some old text about the coalescent we might still want to plunder
-% % If we want to use the coalescent, then really the assumptions
-% % need to make sense.
-% % n << Ne is a silly assumption in today's datasets
-% A core assumption of the coalescent is that the sample size $n$
-% is much less than the effective population size, $N_e$.
-% Several human datasets now consist of hundreds of thousands of
-% genomes~\citep{bycroft2018genome,karczewski2020mutational,tanjo2021practical},
-% and so sample size is substantially \emph{larger} than $N_e$
-% (often assumed to be $10^4$ in humans).
-% Agricultural datasets are even more extreme.
-% For example,
-% the US dairy cattle database alone currently comprises more than 6 million
-% animals with SNP array
-% genotypes\footnote{\url{https://queries.uscdcb.com/Genotype/counts.html}},
-% while the effective population size in modern dairy cattle breeds is
-% less than 100 and decreasing~\citep{MacLeod2013,Makanjuola2020}.
-% An extreme sign of these breeding practices is that there are only two ancestral
-% Y-chromosome lineages present in today's US Holstein dairy breed~\citep{Yue2015}.
-% Simlarly, the 1000 bull genomes project~\citep{hayes20191000}
-% comprises close to 7000 genomes, which are part of multi-generation pedigrees
-% with millions of animals and extensive SNP array genotype and phenotype
-% data \citep[e.g.][]{Cesarani2022}.
-% In another example, genome and SNP array data for 440,610 individuals within
-% 7 multi-generation pedigrees~\citep{whalen2018,Johnsson2021,Ros-Freixedes2020}
-% were combined to infer recombination in pigs~\citep{RosFreixedes2022}.
-% Effective population size in modern pig breeds is also less than 100 due to
-% intense selection and directed reproduction \citep{Hall2016,Porcnic2016}.
-% Another core assumption of the coalescent model is that the genome (or
-% at least the region under study) is short enough that the number of extant
-% lineages remains much smaller than $N_e$ at all times. Whole
-% genome sequences have been available for model organisms
-% for over a decade now, % True? Citation?
-% and indeed complete chromosome-level assemblies are possible
-% in humans~\citep{miga2020telomere}.
-% Projects are under way to obtain high-quality assemblies
-% for all eukaryotic species in Britain and Ireland~\citep{darwin2022sequence}
-% and ultimately worldwide~\citep{lewin2022earth}.
 
@@
 
-% TODO revisit number here
 \section*{Appendix}

### 2023-10-27 paper.tex
Tweaks to s1 and s2

@@
 where the stochastic process of coalescence and recombination
-among ancestral lineages was formalised by Griffiths and colleagues as a
-graph~\citep{griffiths1991two,ethier1990two,griffiths1996ancestral,griffiths1997ancestral}.
+among ancestral lineages was formalised as a
+graph by Griffiths and colleagues~\citep{griffiths1991two,
+ethier1990two,griffiths1996ancestral,griffiths1997ancestral}.
 Subsequently, ARGs have come to be thought of as a data
 structure~\citep{minichiello2006mapping}, i.e.\ describing
-a \emph{realisation} of such a random process or of an
-inference from a sample of genome sequences.
+a \emph{realisation} of such a random process,
+or an inference from a set of sampled genome sequences.
 The distinction between stochastic process
@@
 and (relatedly) the degrees of precision about recombination
-that can be stored (Section~\ref{sec-precision}).
+that can be stored and we may seek to infer (Section~\ref{sec-precision}).
 These ideas have important practical considerations,
@@
 node $u\in N$ is associated with a time $\tau_u$.
-However, the topological ordering
-defined by the directed graph structure
- is sufficient for many applications
-thus we do not require node dates to be known.
+The topological ordering defined by the directed graph structure
+is sufficient for many applications, however,
+and node dates are not required for our purposes here.
 % This should be obvious, but seems like it needs to be said
@@
 nodes are labelled with lowercase alphabetical letters
-rather than integer identifiers, to avoid confusion with genomic intervals.
+rather than integer identifiers to avoid confusion with genomic intervals.
 Thus individual $D_1$ is composed
@@
 genome (C).
-% Jerome: this feels unecessary and "genealogy" is often interpreted as
-% "pedigree" so needs to be used very carefully.
-% The local trees represent the genealogy of genomes \noderef{a}--\noderef{d},
-% which changes between local trees due to recombination.
 The local trees span the three genome regions delineated
 by the two recombination breakpoints that gave rise to these genomes;
-further discussion on
-how local trees are embedded in an ARG is provided in Section~\ref{sec-ARG-and-local-trees}.
-% Following the above definitions, corresponding data encoding for
-% this gARG is shown in Tab.~\ref{tab-gARG-data}.
+see Section~\ref{sec-ARG-and-local-trees} for details
+on how local trees are embedded in an ARG.
 

### 2023-10-27 paper.tex
Tweak

@@
 (the encoding) are
-% TODO more?
+% WIP - it's surprisingly hard to find papers where an ARG is defined
+% concretely
 mathematically defined~\citep[e.g.][]{
-rasmussen2014genome,gusfield2014recombinatorics,ignatieva2021kwarg}
+gusfield2014recombinatorics,ignatieva2021kwarg}
 they follow the approach described by Griffiths and colleagues

### 2023-10-27 paper.tex
Yans updates

--- 2023-10-27 paper.tex
@@
 \caption{\label{fig-arg-in-pedigree}
+% TODO caption needs another pass. Once the full section has been written
+% it'll be clearer what we need to say in the caption vs the text.
 An example genome ARG (gARG) embedded in a pedigree.
@@
 ARG inference methods use it as an output format
-\citep[e.g.][]{song2004minimum,song2005efficient,rasmussen2014genome,
-heine2018bridging,ignatieva2021kwarg}.
+\citep{song2004minimum,song2005efficient,rasmussen2014genome,heine2018bridging,ignatieva2021kwarg}.
 In this Griffiths encoding, as well as nodes marking sampling events,
@@
 within the local tree are included; see Section~\ref{sec-ARG-and-local-trees}
+% TODO: check above if we want to refer to sec-ARG-and-local-trees or some other!
 for a discussion of this important point).
@@
 (Appendix~\ref{sec-cell-lineages-and-args}).
-These paths of single-parent inheritance give rise,
-by definition, to a tree structure.
+By definition, these chains of single-parent inheritance give rise,
+over time, to a tree structure.
 As a result of recombination, adjacent nucleotides can have
@@
 sequences along the genome''.
-These
-% % Could cite some more? E.g. KwARG paper says
-% % "We note that ARGs contain more information than local trees"
-% % Would need a few more examples though, which would require some digging.
-%
-% and similar, \citep[e.g.][]{}
-statements that an ARG
-contains more information than its local trees are true
-if we represent local trees in their conventional forms,
-but these forms discard important information that is
-available in an ARG.
+% Could cite some more? E.g. KwARG paper says
+% "We note that ARGs contain more information than local trees"
+% If even Ana and Jotun are wrong about this it then it must be
+% significant! ;)
+These statements that an ARG
+contains more information than its local trees are only true
+if we use a representation of the local trees that discards
+information about those local trees that can be derived from the ARG.
 
@@
 represented that lead to this
+% We can rephrase... suggestions?
 disagreement about the relationship between local trees and an ARG.
-Firstly, the internal nodes of evolutionary trees are usually considered to be
-\emph{unlabelled}, or equivalently, labelled by the leaves which they subtend.
-The same canonical labelling cannot be used for internal ARG nodes because the
-leaves they subtend will typically vary by genomic position. If we do not label
-the tree nodes in a way which is persistent across the sequence of local trees
-in the ARG, we lose the fact that the \emph{same} ancestors sometimes persist
-across multiple trees.
-Defining ARG nodes as integers and using the oriented
-tree encoding explicitly labels internal nodes, and makes the relationship
-between tree and ARG nodes clear and precise.
-
-% Note: good to keep to paras here to logically separate the two important
-% points.
+Firstly, the internal nodes of evolutionary trees are usually
+considered to be \emph{unlabelled}, or equivalently, labelled by the leaves which they subtend.
+The same canonical labelling cannot be used for internal ARG nodes because the leaves they subtend will typically vary by genomic position.
+If we do not label the tree nodes in a way which is persistent across the sequence of local trees in the ARG, we lose the fact that
+the \emph{same} ancestors sometimes persist across multiple trees.
+Defining ARG nodes as integers and using the oriented tree encoding
+explicitly labels internal nodes, and makes the relationship between
+tree and ARG nodes clear.
 The second property of how evolutionary trees
@@
 the ARG; if this path omits nodes that are not
-branching points (such as \noderef{e} in Fig.~\ref{fig-arg-in-pedigree}),
+branching points in the local tree (such as \noderef{e} in Fig.~\ref{fig-arg-in-pedigree}),
 we lose information about the ARG.
@@
 across multiple local trees.
-\espalier\ approaches this task
-by identifying maximimal subtrees
-that do not change between pairs of adjacent
-local trees and then
-heuristically exploring the space of possible
-rearrangements of these subtrees.
-To derive precise details about recombination events,
-\espalier\ then attempts to infer
+% TODO add a sentence here saying how it does this.
+% that's what we're interested in illustrating here.
+\espalier\ is specifically interested
+in the details of recombination events,
+and also attempts to infer
 the precise subtree prune-and-regraft (SPR)
@@
 induced by recombination between these partially reconciled local trees.
-% NOTE: dropping this because it doesn't clarify the actual point.
-% A similar problem is tackled by \arbores~\citep{heine2018bridging}, which
-% explores a space of ARGs by fixing a starting ARG, and then repeatedly erasing
-% and resimulating candidate SPR moves connecting two fixed, leaf-labelled local
-% trees.
+A similar problem is tackled by \arbores~\citep{heine2018bridging}, which explores a space of ARGs by fixing a starting ARG, and then repeatedly erasing and resimulating candidate SPR moves connecting two fixed, leaf-labelled local trees.
 Inferring the SPRs between leaf-labelled trees is
@@
 This distinction between ARG and local arity is mainly
-of interest when we consider nodes that have
-a single child: those that are \emph{unary}.
+of interest when we consider those that have
+a single child: that are \emph{unary}.
 Returning to the example in Fig.~\ref{fig-arg-in-pedigree}, nodes
@@
 but we still record the fact that genome \noderef{c} inherits from \noderef{f}
-in the local tree. Thus, node \noderef{f} has a single child in this
-interval: it is \emph{locally unary}.
+in the local tree. Thus, \noderef{f} has a single child in this
+interval: in other words it is a \emph{locally unary} node.
 In another example, \noderef{e} is binary in the graph, being a common
@@
 locally unary in all trees in which it is present.
-This is because no ancestral material coalesces in \noderef{e}:
+This is because no ancestral material coalesces there:
 \noderef{a} inherits genetic material from the far left hand end of
@@
 
-ARG-unary nodes occur when an ARG node has one child and one or
-more parents.
-Nodes with one child and one parent (``pass-through'' nodes)
-occur where we wish the record the passage of ancestral material
-through a known node. For example, in simulations
-it is sometimes useful to record the passage of
-ancestral material through known pedigree individuals
-regardless of whether common ancestry occurs.
-ARG-unary nodes are also occur in inferences
-from longitudinal datasets where genetic
+[FIXME this isn't true, add comment about eARG conversion]
+Outside of the context of simulation, ARG-unary nodes are
+only likely to occur in longitudinal datasets where genetic
 data is sampled at many timepoints and recombination
 is rare (e.g.~SARS-CoV-2; see Discussion).
-Nodes with one child and two parents arise when
-we model a recombination event using a single node
-in the classical manner
-(e.g.\ Fig.~\ref{fig-ancestry-resolution}).
-
-Locally unary nodes are a common and important feature
+In contrast, locally unary nodes, are a common and important feature
 of many different types of ARG.
@@
 cannot fully reason about how a local tree is embedded in an ARG
-without them. As we see in the next two sections, both ARG and locally
-unary nodes occur in various scenarios, and are
-produced by a range of current inference methods.
+without them. As we see in the next section, both ARG and locally
+unary nodes occur in a range of different scenarios.
 
@@
 roots (rather than $0$ as used here for notational simplicity).
+
+% \cite{shipilina2023origin} discuss the idea of a ``haplotype block''
+% (equivalent to the ``bricked tree sequence'' of
+% \cite{nowbandegani2023extremely},
+% % TODO look this up. I expect it's basically the same??
+% and the block-based ARG data structure of
+% \citep{palamara2016argon}),
+% here we consider the unique sets of
+% samples that coalesce at a node in the ARG over a particular
+% genome interval, and the fundamental limitations they place
+% on ARG inference.
+% [FIXME find somewhere natural for this. The last clause is too vague - what
+% are the implications?]
+% Alternative gARG encodings split transmitted intervals and associate each with
+% a new node, resulting in a single genome being represented by a group of
+% multiple nodes (or ``blocks'' \citep{palamara2016argon}), with implications for
+
+%%%% Some old text about the coalescent we might still want to plunder
+% % If we want to use the coalescent, then really the assumptions
+% % need to make sense.
+% % n << Ne is a silly assumption in today's datasets
+% A core assumption of the coalescent is that the sample size $n$
+% is much less than the effective population size, $N_e$.
+% Several human datasets now consist of hundreds of thousands of
+% genomes~\citep{bycroft2018genome,karczewski2020mutational,tanjo2021practical},
+% and so sample size is substantially \emph{larger} than $N_e$
+% (often assumed to be $10^4$ in humans).
+% Agricultural datasets are even more extreme.
+% For example,
+% the US dairy cattle database alone currently comprises more than 6 million
+% animals with SNP array
+% genotypes\footnote{\url{https://queries.uscdcb.com/Genotype/counts.html}},
+% while the effective population size in modern dairy cattle breeds is
+% less than 100 and decreasing~\citep{MacLeod2013,Makanjuola2020}.
+% An extreme sign of these breeding practices is that there are only two ancestral
+% Y-chromosome lineages present in today's US Holstein dairy breed~\citep{Yue2015}.
+% Simlarly, the 1000 bull genomes project~\citep{hayes20191000}
+% comprises close to 7000 genomes, which are part of multi-generation pedigrees
+% with millions of animals and extensive SNP array genotype and phenotype
+% data \citep[e.g.][]{Cesarani2022}.
+% In another example, genome and SNP array data for 440,610 individuals within
+% 7 multi-generation pedigrees~\citep{whalen2018,Johnsson2021,Ros-Freixedes2020}
+% were combined to infer recombination in pigs~\citep{RosFreixedes2022}.
+% Effective population size in modern pig breeds is also less than 100 due to
+% intense selection and directed reproduction \citep{Hall2016,Porcnic2016}.
+% Another core assumption of the coalescent model is that the genome (or
+% at least the region under study) is short enough that the number of extant
+% lineages remains much smaller than $N_e$ at all times. Whole
+% genome sequences have been available for model organisms
+% for over a decade now, % True? Citation?
+% and indeed complete chromosome-level assemblies are possible
+% in humans~\citep{miga2020telomere}.
+% Projects are under way to obtain high-quality assemblies
+% for all eukaryotic species in Britain and Ireland~\citep{darwin2022sequence}
+% and ultimately worldwide~\citep{lewin2022earth}.
 
@@
 
+% TODO revisit number here
 \section*{Appendix}

### 2023-10-27 paper.tex
Various updates

--- 2023-10-27 paper.tex
@@
 This broad usage is a useful generalisation,
-but is not, however, compatible with the available formal definitions
+but is not compatible with the available formal definitions
 derived from the original Griffiths approach.
@@
 The topological ordering defined by the directed graph structure
+and an arrow of time (telling us which direction is pastwards)
 is sufficient for many applications, however,
-and node dates are not required for our purposes here.
+and we assume node dates are not known here.
 % This should be obvious, but seems like it needs to be said
@@
 (the encoding) are
-% TODO more?
 mathematically defined~\citep[e.g.][]{
-rasmussen2014genome,gusfield2014recombinatorics,ignatieva2021kwarg}
+gusfield2014recombinatorics,ignatieva2021kwarg}
 they follow the approach described by Griffiths and colleagues
@@
 heine2018bridging,ignatieva2021kwarg}.
-In this Griffiths encoding, as well as nodes marking sampling events,
-we have two additional node types in the graph,
+% Griffiths never mentions "sampling events", so let's not get
+% into that. It's not an important detail.
+In this Griffiths encoding we have two types of internal node in the graph,
 representing the common ancestor and recombination events
@@
 are used to construct the local trees.
-This is done by tracing rootwards through the graph from the samples,
+This is done by tracing pastwards through the graph from the samples,
 making decisions about which outbound edge to follow through
@@
 ascertain ancestral material without traversing the graph
-rootwards from the sample nodes,
+pastwards from the sample nodes,
 resolving the effects of recombination and common ancestor events.
@@
 
-% \section{Retrospective vs prospective ARGs}
-% \label{sec-retro-pro}
-% The ARG concept arose from a \emph{retrospective} view of inheritance.
-% For example the definition of an eARG, as discussed above, is invariably
-% couched in terms of the events that occured in the history of some set of sampled
-% genomes. Similarly, the genome ARG in Fig.~\ref{fig-arg-in-pedigree}
-% is retrospective because it only includes information
-% about genetic inheritance relative to the sample individuals
-% $D_1$ and $D_2$ (genomes \noderef{a}--\noderef{d}).
-% That is, only the inheritance
-% history of the sample genomes \noderef{a}--\noderef{d} is recorded in the ARG.
-% % Gregor & Jerome think that "inheritance" is clearer here than "transmission"
-% Genetic inheritance must have occurred between (e.g.)
-% $D_8$ and her daughter $D_6$, but this is not recorded because
-% it is not relevant to the genetic ancestry of the sample genomes.
-
-% There is no particular reason, however, that ARGs must be
-% retrospective. Recent advances in individual-based forward-time simulation
-% ~\citep{kelleher2018efficient,haller2018tree} depend upon the ability to
-% record gARGs in a \emph{prospective} manner.
-% In a prospective gARG, the inheritance intervals $I$
-% on an edge joining ancestor genome $v$ and descendant $u$
-% define \emph{all} regions of genome that $u$ inherited from $v$.
-% Thus, these inheritance intervals are
-% concerned only with the genetic relationship
-% between the nodes in question, and are not defined with respect
-% to any given set of samples. They simply record the immediate
-% genetic inheritance between two genomes
-% (in the simplest case, between a parent and a child).
-% \citet{kelleher2018efficient} refer to the structure
-% recording all such forwards-in-time genetic relationships
-% as an ``embellished pedigree`` (and suggest the term ``nedigree'' as a shorthand).
-% Describing this instead as a ``prospective ARG'', emphasising
-% the use of the same data structures and the direct connection with
-% the usual retrospective interpretation of an ARG, may be
-% more useful.
-
-% The exhaustive recording of full inheritance intervals from generation
-% to generation leads to a rapid build-up of information, much of which
-% may not be relevant to the genetic ancestry of the current population.
-% This redundancy can be periodically removed using the ``simplify'' algorithm,
-% described in detail by \citet{kelleher2018efficient}. Simplification
-% allows efficient subsetting of ARGs (removing both samples and certain
-% levels of detail about recombination, as discussed in Section~\ref{sec-ARG-simplification}),
-% and is also the key enabling factor when incorporating propective ARGs into
-% forward-time simulation.
-% The core algorithm involves the concept of
-% ``ancestral material''~\citep{wiuf1999ancestry,wiuf1999recombination}:
-% the genomic intervals ancestral to a set of samples,
-% scattered across lineages (the edges of the graph)
-% at some time in the past. In the next section, we outline how
-% tracing ancestral material allows the identification and removal of nodes
-% and inheritance intervals, such that
-% only genetic ancestry relevant to the current population is retained.
-
 \section{ARGs and local trees}
@@
 We begin by setting $\pi_u = 0$ for each $1 \leq u \leq n$.
-Then, for each sample node in $S$ we trace its path rootwards through the
+Then, for each sample node in $S$ we trace its path pastwards through the
 ARG for position $x$, and record this path in $\pi$.
@@
 model recombination \emph{events} via an ARG node, but simply record
-the \emph{outcome} of a recombination via outbound edges to
+the \emph{outcome} of a recombination via edges to
 the parent's two genomes.
-Thus, a recombinant node may also correspond to a coalescence
-(e.g.~node \noderef{g} in Fig.~\ref{fig-simplification}).
+Thus, a recombinant node such as node \noderef{g} in Fig.~\ref{fig-simplification}
+may also correspond to a coalescence.
 The distinction of
 using a single node to represent
-a recombination event (e.g.~Fig.~\ref{fig-ancestry-resolution}),
-or two to represent the parent genomes is usually not important
-(either is possible in the gARG encoding)
-and the most convenient approach will vary by application
+a recombination event, as is done in ~Fig.~\ref{fig-ancestry-resolution},
+or two to represent the parent genomes, as in Fig.~\ref{fig-simplification},  is usually not important.
+Either is possible in the gARG encoding, and the most convenient approach will vary by application
 (discussed e.g.\ in Appendix~\ref{sec-big-and-little-arg}).
-Note also that polytomies (e.g.~Fig.~\ref{fig-simplification} node \noderef{k})
-are a natural feature of such a Wright-Fisher model
+Note also that node \noderef{k} in Fig.~\ref{fig-simplification} has three children.
+Polytomies like this are a natural feature of such a Wright-Fisher model
 (but see Appendix~\ref{sec-cell-lineages-and-args}).
@@
 is reachable from the samples in the local trees)
-over which it is coalescent (has more than one child). Nodes which
+over which it has more than one child. Nodes which
 are never locally unary therefore have a coalescent span of 100\%,
@@
 Returning to the main topic of this section,
-Fig.~\ref{fig-simplification}A is original simulation output in
+Fig.~\ref{fig-simplification}A is the original simulation output, in
 which we retain all nodes involved in recombination
@@
 Nonetheless, both methods produce estimates in which nodes and edges
-persist across multiple trees, naturally captured by the gARG approach.
+persist across multiple trees, creating
+inheritance graphs which fit naturally into the gARG formulation.
 To illustrate the varying levels of information captured
@@
 clearly much of the detail about recombination events in an ARG is lost.
-The \texttt{tsdate} algorithm (and related naive estimator
-of ancestral location) uses much more between-tree
-information~\citep{wohns2022unified}, explicitly using the gARG
-encoding to reason about ancestral nodes.
+The \texttt{tsdate} algorithm and related naive estimator
+of ancestral location explicitly use the gARG encoding
+to reason about ancestral nodes~\citep{wohns2022unified}.
 The main application for fully precise ARGs thus far has been
@@
 % get good local trees
-targetting particular inference goals, sharing a common ARG interchange format.
+targeting particular inference goals, sharing a common ARG interchange format.
 The output of these diverse inference methods could then be
@@
 for helpful discussions and comments on the manuscript.
-
 % Ordered alphabetically by surname
@@
 % TODO Jerome: add EPSRC and NIH HG011395-0 grants.
-Jerome Kelleher acknowledges support from the Robertson Foundation.
+Jerome Kelleher acknowledges support from the Robertson Foundation,
+NIH (research grants HG011395 and HG012473) and 
+EPSRC (research grant EP/X024881/1).
 Jere Koskela acknowledges support from EPSRC research grant EP/V049208/1.
@@
 
-In Fig.~\ref{fig-simplification} we provide a schematic of the events (coloured red and blue squares) and
+Fig.~\ref{fig-simplification} shows a schematic of the events (coloured red and blue squares) and
 the genomes (chromosome icons) that occur in the cellular germline of a simplified, diploid multicellular

### 2023-10-27 paper.tex

@@
 as the output of inference programs.
-Where such concrete details of an ARG data structure
-(the encoding) are
-mathematically defined~\citep[e.g.][]{
-gusfield2014recombinatorics,ignatieva2021kwarg}
+Where details of an ARG data structure (the encoding) are
+% Three papers 10 years apart will do here, right?
+provided~\citep[e.g.][]{
+wiuf1999recombination,gusfield2014recombinatorics,hayman2023recoverability}
 they follow the approach described by Griffiths and colleagues
@@
 % Griffiths never mentions "sampling events", so let's not get
-% into that. It's not an important detail.
+% into that. It's not an important detail, and we don't want to be
+% overly precise about something that people haven't actually written
+% down.
 In this Griffiths encoding we have two types of internal node in the graph,
@@
 within a single individual should be regarded as one or multiple events are
-debatable.
+debatable points.
 % If we want to have an ARG software ecosystem then such fireside

### 2023-10-27 paper.tex

@@
 The structure representing such recombinant genetic ancestry
-is usually referred to as an ancestral recombination graph (ARG),
-% FIXME this sentance needs to change. We want to capture more of the
-% idea of the formalisms falling behind actual practise and the
-% confusion this is causing
-although there is some confusion about the interpretation and little
-agreement on specific details.
-We propose a concrete definition of ARGs
-in terms of genomes and their intervals of genetic inheritance (gARG)
-and contrast this with the classical event-based definition (eARG).
-We show that eARGs are limited in the patterns
-of genetic inheritance that can be represented,
-and require a precise description of the details of all recombination events.
-We show, in contrast, that the gARG definition
-can fully capture the richness of modern large-scale datasets,
-enables fine-grained
+is usually referred to as an ancestral recombination graph (ARG).
+% Be specific about what we're talking about.
+ARGs are classically defined by the graph of
+% We can play a bit loose here, please don't change to "common ancestor"
+coalescence and recombination events in the history of sample,
+with the differential inheritance structure encoded by the
+crossover breakpoints associated with recombination events.
+We show that this ``event ARG'' (eARG) encoding
+is limited in the patterns of genetic inheritance that can be represented,
+being insufficient to describe the outputs of modern ARG simulation
+and inference methods,
+and ill-suited to capturing the complex patterns of
+relatedness that exist in contemporary datasets.
+We propose an alternative ``genome ARG'' (gARG) encoding
+expressed in terms of
+% we haven't define a "genome" yet, so need to give a bit of context
+specific genomes and their intervals of genetic inheritance.
+We show it can fully capture the richness of modern large-scale datasets
+and is compatible with modern simulation and inference methods.
+Moreover, we show that, unlike the eARG encoding which requires
+a precise estimate of all recombination events,
+the gARG encoding enables fine-grained
 levels of precision about recombination to be represented,
-and forms the basis of an efficient computational framework.
+opening up a rich set of additional details.
 \end{abstract}

### 2023-10-27 paper.tex
Updates from Overleaf

--- 2023-10-27 paper.tex
@@
 
-ARG-unary nodes occur when an ARG node has one child and one or
-more parents.
-Nodes with one child and one parent (``pass-through'' nodes)
-occur where we wish the record the passage of ancestral material
+An ARG node with one child and one or
+more parents is ARG-unary.
+If it has one child and only one parent, it represents a ``pass-through'' node:
+these occur where we wish the record the passage of ancestral material
 through a known node. For example, in simulations
@@
 regardless of whether common ancestry occurs.
-ARG-unary nodes are also occur in inferences
-from longitudinal datasets where genetic
-data is sampled at many timepoints and recombination
-is rare (e.g.~SARS-CoV-2; see Discussion).
 Nodes with one child and two parents arise when
@@
 (e.g.\ Fig.~\ref{fig-ancestry-resolution}).
+It is also possible for sample nodes to be ARG-unary, for example
+in inferences from longitudinal datasets where genetic
+data is sampled at many timepoints and recombination
+is rare (e.g.~SARS-CoV-2; see Discussion).
 

### 2023-10-27 paper.tex
Yans updates

@@
 specific genomes and their intervals of genetic inheritance.
-We show it can fully capture the richness of modern large-scale datasets
+We show this can fully capture the richness of modern large-scale datasets
 and is compatible with modern simulation and inference methods.
@@
 
-An ARG node with one child and one or
-more parents is ARG-unary.
-If it has one child and only one parent, it represents a ``pass-through'' node:
+% We get discussion of ARG-unary nodes out of the way first: these are
+% relatively easy to understand (but it's worth emphasising that
+% eARG-like recombination nodes are ARG-unary, as that might not be clear
+By definition, ARG-unary nodes have one child but can have one or
+more parents.
+A node with one child and only one parent represents a ``pass-through'' node:
 these occur where we wish the record the passage of ancestral material
@@
 
-Locally unary nodes are a common and important feature
+% Moving on from ARG-unary to the more general case
+More generally, locally unary nodes,
+which can have one or more children in the graph,
+are a common and important feature
 of many different types of ARG.
@@
 clearly much of the detail about recombination events in an ARG is lost.
+% You don't even need to define event nodes when using a graph-based
+% algorithm like tsdate
 The \texttt{tsdate} algorithm and related naive estimator
@@
 \label{sec-cell-lineages-and-args}
-In eukaryotic organisms, ARGs are a result of the cellular processes of mitosis
-(which leads to common ancestry and coalescence)
-and meiosis (which leads to recombination, including both crossovers and gene conversion).
-
 \begin{figure}
@@
 
+In eukaryotic organisms, ARGs are a result of the cellular processes of mitosis
+(which leads to common ancestry and coalescence)
+and meiosis (which leads to recombination, including both crossovers and gene conversion).
 Fig.~\ref{fig-simplification} shows a schematic of the events (coloured red and blue squares) and
 the genomes (chromosome icons) that occur in the cellular germline of a simplified, diploid multicellular
-hermaphrodite eukaryote with partially overlapping generations. Here, events are not represented by a single
+hermaphrodite eukaryote with partially overlapping generations. 
+% It may be worth putting in something here about events being a slippery concept
+% (we have already mentioned this at the end of Section 3
+Here, an event is not represented by a specific
 genome: rather, a genome can be associated with (or ``tag'') an event above or below it.

### 2023-10-30 paper.tex
Edits from Yan

@@
 % We can play a bit loose here, please don't change to "common ancestor"
-coalescence and recombination events in the history of sample,
+coalescence and recombination events in the history of a sample,
 with the differential inheritance structure encoded by the
@@
 specific genomes and their intervals of genetic inheritance.
-We show this can fully capture the richness of modern large-scale datasets
+We show that the gARG encoding can fully capture the richness of modern large-scale datasets
 and is compatible with modern simulation and inference methods.
 Moreover, we show that, unlike the eARG encoding which requires
-a precise estimate of all recombination events,
+% TODO rephrase "to be detailed" is ambiguous
+all recombination events to be detailed,
 the gARG encoding enables fine-grained
@@
 text for details of these methods.
-Edge colours are based on the time of the edge's child node
+Edge colours indicate time of the edge's child node
 (lighter: older; darker: younger).
@@
 Line width and node colour are as described in Fig.~\ref{fig-simplification}.
-Graphics on the bottom row show the genome positions, relative to the start of
+Bottom row graphics show the genome positions, relative to the start of
 the ADH gene, for each graph edge from the corresponding ARG. Edge intervals
 are drawn as horizontal lines, stacked in time order (edges with youngest
-children at the bottom); vertical dashed lines indicate breakpoints between
+children at the bottom); vertical dashed lines denote breakpoints between
 local trees.
@@
 between them using the Li and Stephens model~\citep{li2003modeling}.
-\texttt{Relate} first reconstructs local tree topologies across the genome, using a variant of the Li and Stephens model to estimate the ordering of coalescence events in each tree, and then estimates branch lengths using MCMC with a coalescent-based prior.
+\texttt{Relate} first reconstructs local tree topologies across the genome,
+using a variant of the Li and Stephens model to estimate the
+ordering of coalescence events in each tree, and then estimates
+branch lengths using MCMC with a coalescent-based prior.
 See Appendix~\ref{sec-survey-arg-infer} for more details on these
@@
 It consists of 43 biallelic SNPs spanning 2.4Kb of the \emph{D.~melanogaster} ADH
-locus on chromosome 2L. Where required for inference purposes we assume a
-mutation rate of $5.49\times10^{-9}$
-and recombination rate of $8.4\times10^{-9}$ per site per generation
+locus on chromosome 2L. Where required for inference purposes we assume
+mutation and recombination rates of $5.49\times10^{-9}$
+and $2.40463\times10^{-9}$ per site per generation
 % NOTE: we should be citing the original sources here too.
-\citep{adrion2020community,lauterbur2023expanding},
-% Note: would have been simpler to use the stdposim estimate
-% of 1.7M from Li et al 2006. (not that it matters)
-and constant effective population size of $10^6$.
+\citep{schrider2013rates,comeron2012many}
+and a constant effective population size of $1720600$ \citep{li2006inferring}, all
+values being taken from the \texttt{stdpopsim} library
+\citep{adrion2020community,lauterbur2023expanding}.
 Full details of the software used and code to generate these
@@
 % Here we describe recombination features
-Considering~Fig.~\ref{fig-inferred-args}, we can see that
-there is considerable variation in the number of recombination breakpoints
+Considering Fig.~\ref{fig-inferred-args}, we can see that
+there is substantial variation in the number of recombination breakpoints
 inferred by different methods, with e.g.~\argweaver\ suggesting
@@
 relationship between the number of nodes with multiple parents
-and breakpoints along the genome.
+and number of breakpoints along the genome.
 In both ARGs the number of breakpoints is smaller
@@
 is the presence of polytomies, which only
-\tsinfer\ creates. The most obvious example involves
+\tsinfer\ creates. The most obvious example
 involves nodes \noderef{Fr-F}, \noderef{Wa-F}, and \noderef{Af-F},
-which happen happen to have identical sequences.
+which happen to have identical sequences.
 Because \kwarg, \argweaver, and \relate\ require
 bifurcating trees by design, each picks an arbitrary order of branching
-(hence Fig.~\ref{fig-inferred-args}A,B disagrees
-with Fig.~\ref{fig-inferred-args}D in this order).
-% JK: I don't quite get this, so readers probably won't either. Let's
-% stick a few clear messages.
+(hence Fig.~\ref{fig-inferred-args}A and B disagree
+in this order, and Fig.~\ref{fig-inferred-args}D even shows different
+orders in different trees).
+% JK: I don't quite get the commented out sentence below, so readers probably
+% won't either. Let's stick to a few clear messages.
 % More generally, the number of nodes in Fig.~\ref{fig-inferred-args}C
@@
 uncertainty about node ages and recombination breakpoint positions, is an important
-aspect of future work. Fully utilising such uncertainty in downstream
-applications is also vital, of course.
+aspect of future work. 
+% Obvious, but nobody is doing it, so needs to be said. 
+% TODO rephrase to say this is an open question.
+Fully utilising such uncertainty in downstream
+applications is also vital.
 
@@
 
-In eukaryotic organisms, ARGs are a result of the cellular processes of mitosis
+In eukaryotes, ARGs are a result of the cellular processes of mitosis
 (which leads to common ancestry and coalescence)
@@
 Here, an event is not represented by a specific
-genome: rather, a genome can be associated with (or ``tag'') an event above or below it.
-The gARG representation allows any such genome to be arbitrarily distant in time or number of cell
-divisions from the event itself (see Section~\ref{sec-precision}).
+genome. Rather, a genome can be associated with (or ``tag'') an event above or below it (and as argued in
+Section~\ref{sec-precision}, the gARG representation allows these associated genomes to be arbitrarily distant
+in time or number of cell
+divisions from the event itself).
 
 Thick black lines trace an ARG that describes the ancestry of the lower arm of the
-four sampled chromosomes, and these the lines pass through gARG nodes (green chromosomal regions).
+four sampled chromosomes, and these lines pass through gARG nodes (green chromosomal regions).
 A number of these nodes have only one parent and one child in the ARG; these would often be omitted
@@
 
-It is worth noting that ``true'' polytomies in ARGs are a feature of model assumptions
-rather than biological reality. Indeed, given a large enough mutation rate
-(or single cell sequencing at a sufficiently detailed level) it should be possible to
-distinguish sequential rounds of mitosis within a single multicellular organism.
-For example, a mutation in the first cell division of $D_{10}$ could be shared between
+Note that mutations can occur along any lineage in Fig.~\ref{fig-cell-lines}. This means, for example, 
+that a mutation in the first cell division of $D_{10}$ could be shared between
 the two gametes produced by the cells in the left half of $D_{10}$ but not shared by
-the right hand gamete. For an ARG that has been accurately inferred from real data,
-a polytomy is therefore making a (deliberately) imprecise statement about the order
-of coalescence, rather than reflecting of a single instantaneous event; this
-interpretation fits naturally into the gARG framework.
+the right hand gamete. With a high enough germline mutation rate (or equivalently, a long enough genome), each
+round of mitotic genome duplication within a single multicellular organism could
+potentially be distinguished.
+This implies that in ARGs that have been inferred from real data, polytomies represent a lack
+of mutational information (rather than marking a single ``multifurcation event'').
+The ability to represent polytomies, and conceptualise them as a (deliberately) imprecise statement about the order
+of coalescence within and between individuals,
+fits naturally into the gARG framework presented in this paper.
 

### 2023-10-30 paper.tex
tweak

@@
 Finally, the literature on ARGs is large and
-confusing, and we attempty to clarify some important aspects
+confusing, and we attempt to clarify some important aspects
 in appendices, including

### 2023-10-30 paper.tex
First-pass at high-falutin interpretation

@@
 \newcommand{\argweaver}[0]{\texttt{ARGweaver}}
+\newcommand{\argweaverD}[0]{\texttt{ARGweaver-D}}
 \newcommand{\relate}[0]{\texttt{Relate}}
@@
 % We can play a bit loose here, please don't change to "common ancestor"
-coalescence and recombination events in the history of a sample,
+coalescence and recombination events in the history of a sample of genomes,
 with the differential inheritance structure encoded by the
@@
 about the temporal ordering of multiple recombinations.
+
+Aside from the stochastic process interpretation, there
+is another concept of an ARG that we must be careful to distinguish,
+that of the \emph{actual} ARG underlying a sample of genomes
+and ultimately an entire species. This ARG is the true history,
+perfectly resolved into binary splits and mergers by
+the cellular processes of meiosis and mitosis, regardless
+of sampling density or population processes
+(Appendix~\ref{sec-cell-lineages-and-args}).
+Just as this true, ideal ARG unquestionably exists it
+can also never be entirely known
+% NOTE: preempting some discussion here:
+% Mutation rate is (a) not high enough in existing species to
+% resolve, (b) probably couldn't be if organisms are to survive,
+% and (c) such a high mutation rate would probably even obscure
+% the more ancient past.
+because of a fundamental lack of mutational information.
+This ideal ARG is also not a sample from the coalescent
+or any other population genetics model: by necessity, and design,
+such mathematical constructs are caricatures of biological reality,
+allowing us to draw general conclusions about population processes
+without the burden of unnecessary detail.
+The scale of present-day datasets breaks fundamental assumptions
+made by such models, and we must therefore be careful to articulate
+our goals in inference.
+Mathematical models are insufficient to describe the complexities
+of biological reality when sampling is at population scale,
+and the true ARG can never be known in its entirity.
+A reasonable goal to set, then, is to estimate \emph{aspects}
+of the true ARG and to capture uncertainty where it inevitably exists.
+This is the approach taken by recent inference
+methods~\citep{kelleher2019inferring,speidel2019method,zhang2023biobank},
+by neither fully imposing the constraints of a mathematical
+model or trying estimate the true ARG is its compete detail, they
+have made substantial progress. However, our vocabularly
+and mathematical formalisms, rooted in both the ideal and
+mathematical concepts of an ARG, are insufficient to capture the
+richness of these new structures.
 
@@
 Genomes \noderef{a} and \noderef{c} are the product of two independent
-recombination events (red) between
+meioses (recombination events in red) between
 the paternal genomes \noderef{e}
@@
 differential inheritance of genetic material between genomes.
-% This is definitely needed for the Nicks, who are always thinking
-% about the deeper processes and are quite happy being loosey-goosey
-% with representation details
+% % This is definitely needed for the Nicks, who are always thinking
+% % about the deeper processes and are quite happy being loosey-goosey
+% % with representation details
 It is important to note that here, and throughout,
@@
 and $2.40463\times10^{-9}$ per site per generation
-% NOTE: we should be citing the original sources here too.
 \citep{schrider2013rates,comeron2012many}
@@
 \citep{adrion2020community,lauterbur2023expanding}.
-Full details of the software used and code to generate these
-figures are available on GitHub (see Data Availability).
+Software versions were \kwarg\ v1.0, \argweaverD\ (2019),
+\tsinfer v0.3.1, and \relate v1.1.9: full details and replicable code for generating
+these figures are available on GitHub (see Data Availability).
 
@@
 We are grateful to Nick Barton, Gideon Bradburd, Alex Lewanski and Andrew Vaughan
-for helpful discussions and comments on the manuscript.
+for helpful discussions and comments on the manuscript, and 
+to Iain Mathieson and Aylwyn Scally for interesting discussions that 
+originally motivated this work.
 % Ordered alphabetically by surname
@@
 Ancestral recombination events and coalescence events are shown as red and blue squares respectively.
-The ARG paths for the lower arm of the sampled chromosomes (green) is highlighted as a
+A realised ARG path for the lower arm of the sampled chromosomes (green) is highlighted as a
 thick solid line and involves a single recombination event and four coalescence events
@@
 Here, an event is not represented by a specific
-genome. Rather, a genome can be associated with (or ``tag'') an event above or below it (and as argued in
+genome. Rather, genomes can be associated with (or ``tag'') events above or below them (and as argued in
 Section~\ref{sec-precision}, the gARG representation allows these associated genomes to be arbitrarily distant
 in time or number of cell
-divisions from the event itself).
+divisions from the event itself). For example, tagging the two genomes above a recombination event leads to the
+2 node representation seen in Figs.~\ref{fig-simplification} and \ref{hudson_vs_bigARG}, whereas tagging the genome
+below a recombination event leads to the more conventional graphs in Figs.~\ref{fig-ancestry-resolution} and \ref{fig-inferred-args}A,B.
 

### 2023-10-30 paper.tex
WIP

@@
 and subfields use the term
-differently (see Appendix~\ref{sec-arg-history}).
+differently (Appendix~\ref{sec-arg-history}).
+Another concept of an ARG that we must be careful to distinguish,
+is that of the \emph{actual} ARG underlying a sample of genomes
+and ultimately an entire species. This ARG is the true history,
+perfectly resolved into binary splits and mergers by
+the cellular processes of meiosis and mitosis, regardless
+of sampling density or population processes
+(Appendix~\ref{sec-cell-lineages-and-args}).
+Just as this true, ideal ARG unquestionably exists,
+it can also never be entirely known,
+% NOTE: preempting some discussion here:
+% Mutation rate is (a) not high enough in existing species to
+% resolve, (b) probably couldn't be if organisms are to survive,
+% and (c) such a high mutation rate would probably even obscure
+% the more ancient past.
+because of a fundamental lack of mutational information.
+
+The scale of modern datasets makes the
+relationship between the true ARG and
+population genetic models important to articulate.
+The coalescent with recombination (henceforth, ``coalescent'')
+is a mathematical model that defines a distribution of ARGs
+for a given set of parameters.
+It has proven to be a useful and robust
+model~\citep{wakeley2012gene,bhaskar2014distortion,nelson2020accounting},
+but many modern datasets have properties that grossly
+violate its assumptions.
+One key assumption is that
+sample size $n$ is much less than the effective population size, $N_e$.
+Several human datasets now consist of hundreds of thousands of
+genomes~\citep{turnbull2018hundred, bycroft2018genome,
+karczewski2020mutational,tanjo2021practical,
+halldorsson2022sequences},
+and so sample size is an order of magnitude \emph{larger} than the
+usually assumed $N_e$ values.
+Agricultural datasets are an even more extreme departure from this
+assumption, with hundreds of thousands of samples embedded in
+multi-generational pedigrees~\citep{hayes20191000,Ros-Freixedes2020}
+and effective population sizes of 100 and even
+less~\citep{MacLeod2013,Makanjuola2020,Hall2016,Porcnic2016}.
+A model assuming a single $N_e$ would be a
+drastic over-simplification in these cases of course, but
+even if sufficiently complex demographic models~\citep{gower2022demes}
+encompassing hundreds of populations, explosive growth rates and myriad
+interconnections of migration, were somehow estimated and provided as input,
+ARGs sampled from the coalescent cannot capture the complexities
+of family structure in these
+datasets~\citep[e.g.][]{turnbull2018hundred,Ros-Freixedes2020}.
+Another core assumption of the coalescent model is that the genome
+(or at least the region under study) is short enough that the number of extant
+% TODO do we need to explain this a bit more?
+lineages remains much smaller than $N_e$ at all times in the past.
+High-quality whole genome assemblies are now available
+for many species % Can we say something more specific here?
+and projects are under way to obtain
+them for tens of thousands more~\citep{darwin2022sequence,lewin2022earth}.
+Vast quantities of whole genome sequence data across a broad range
+of species exists today.
+
+The true ARG underlying a modern population-scale dataset
+is therefore not a sample from the coalescent, or any other
+population genetics model.
+
+% Mathematical models are insufficient to describe the complexities
+% of biological reality when sampling is at population scale,
+% and the true ARG can never be known in its entirity.
+
+A reasonable goal to set, then, is to estimate \emph{aspects}
+of the true ARG and to capture uncertainty where it inevitably exists.
+This is the approach taken by recent inference
+methods~\citep{kelleher2019inferring,speidel2019method,zhang2023biobank},
+by neither fully imposing the constraints of a mathematical
+model or trying estimate the true ARG is its compete detail, they
+have made substantial progress. However, our vocabularly
+and mathematical formalisms, rooted in both the ideal and
+mathematical concepts of an ARG, are insufficient to capture the
+richness of these new structures.
+
+% This ideal ARG is also not a sample from the coalescent
+% or any other population genetics model: by necessity, and design,
+% such mathematical constructs are caricatures of biological reality,
+% allowing us to draw general conclusions about population processes
+% without the burden of unnecessary detail.
+% The scale of present-day datasets breaks fundamental assumptions
+% made by such models, and we must therefore be careful to articulate
+% our goals in inference.
+
 % pedantically, ARG is not really a "word"
@@
 This broad usage is a useful generalisation,
-but is not compatible with the available formal definitions
+but is not compatible with the available formal
+data structure definitions
 derived from the original Griffiths approach.
@@
 
-Aside from the stochastic process interpretation, there
-is another concept of an ARG that we must be careful to distinguish,
-that of the \emph{actual} ARG underlying a sample of genomes
-and ultimately an entire species. This ARG is the true history,
-perfectly resolved into binary splits and mergers by
-the cellular processes of meiosis and mitosis, regardless
-of sampling density or population processes
-(Appendix~\ref{sec-cell-lineages-and-args}).
-Just as this true, ideal ARG unquestionably exists it
-can also never be entirely known
-% NOTE: preempting some discussion here:
-% Mutation rate is (a) not high enough in existing species to
-% resolve, (b) probably couldn't be if organisms are to survive,
-% and (c) such a high mutation rate would probably even obscure
-% the more ancient past.
-because of a fundamental lack of mutational information.
-This ideal ARG is also not a sample from the coalescent
-or any other population genetics model: by necessity, and design,
-such mathematical constructs are caricatures of biological reality,
-allowing us to draw general conclusions about population processes
-without the burden of unnecessary detail.
-The scale of present-day datasets breaks fundamental assumptions
-made by such models, and we must therefore be careful to articulate
-our goals in inference.
-Mathematical models are insufficient to describe the complexities
-of biological reality when sampling is at population scale,
-and the true ARG can never be known in its entirity.
-A reasonable goal to set, then, is to estimate \emph{aspects}
-of the true ARG and to capture uncertainty where it inevitably exists.
-This is the approach taken by recent inference
-methods~\citep{kelleher2019inferring,speidel2019method,zhang2023biobank},
-by neither fully imposing the constraints of a mathematical
-model or trying estimate the true ARG is its compete detail, they
-have made substantial progress. However, our vocabularly
-and mathematical formalisms, rooted in both the ideal and
-mathematical concepts of an ARG, are insufficient to capture the
-richness of these new structures.
+% A data structure is a means of encoding information, and
+% a precise description of the structure allows us to mathematically
+% reason about its properties.
+
+% We must also distinguish between a specific realisation
+% the \emph{actual} ARG underlying a sample of genomes
 
@@
 
-Fully decoupling the general concept of an ARG from the coalescent
-with recombination (henceforth, ``coalescent'')
-is an important step.
-While the coalescent has proven to be a useful and
-robust
-model~\citep{wakeley2012gene,bhaskar2014distortion,nelson2020accounting},
-many modern datasets have properties that grossly
-violate its assumptions.
-One key assumption is that
-sample size $n$ is much less than the effective population size, $N_e$.
-Several human datasets now consist of hundreds of thousands of
-genomes~\citep{turnbull2018hundred, bycroft2018genome,
-karczewski2020mutational,tanjo2021practical,
-halldorsson2022sequences},
-and so sample size is an order of magnitude \emph{larger} than the
-usually assumed $N_e$ values.
-Agricultural datasets are an even more extreme departure from this
-assumption, with hundreds of thousands of samples embedded in
-multi-generational pedigrees~\citep{hayes20191000,Ros-Freixedes2020}
-and effective population sizes of 100 and even
-less~\citep{MacLeod2013,Makanjuola2020,Hall2016,Porcnic2016}.
-A model assuming a single $N_e$ would be a
-drastic over-simplification of course, but
-even if sufficiently complex demographic models~\citep{gower2022demes}
-encompassing hundreds of populations, explosive growth rates and myriad
-interconnections of migration, were somehow estimated and provided as input,
-ARGs sampled from the coalescent cannot capture the complexities
-of family structure in these
-datasets~\citep[e.g.][]{turnbull2018hundred,Ros-Freixedes2020}.
-Another core assumption of the coalescent model is that the genome
-(or at least the region under study) is short enough that the number of extant
-lineages remains much smaller than $N_e$ at all times.
-High-quality whole genome assemblies are now available
-for many species % Can we say something more concrete here?
-and projects are under way to obtain
-them for tens of thousands more~\citep{darwin2022sequence,lewin2022earth},
-and so we can expect inferred ARGs to routinely span whole chromosomes.
+% Fully decoupling the general concept of an ARG from the coalescent
+% with recombination (henceforth, ``coalescent'')
+% is an important step.
+% While the coalescent has proven to be a useful and
+% robust
+% model~\citep{wakeley2012gene,bhaskar2014distortion,nelson2020accounting},
+% many modern datasets have properties that grossly
+% violate its assumptions.
+% One key assumption is that
+% sample size $n$ is much less than the effective population size, $N_e$.
+% Several human datasets now consist of hundreds of thousands of
+% genomes~\citep{turnbull2018hundred, bycroft2018genome,
+% karczewski2020mutational,tanjo2021practical,
+% halldorsson2022sequences},
+% and so sample size is an order of magnitude \emph{larger} than the
+% usually assumed $N_e$ values.
+% Agricultural datasets are an even more extreme departure from this
+% assumption, with hundreds of thousands of samples embedded in
+% multi-generational pedigrees~\citep{hayes20191000,Ros-Freixedes2020}
+% and effective population sizes of 100 and even
+% less~\citep{MacLeod2013,Makanjuola2020,Hall2016,Porcnic2016}.
+% A model assuming a single $N_e$ would be a
+% drastic over-simplification of course, but
+% even if sufficiently complex demographic models~\citep{gower2022demes}
+% encompassing hundreds of populations, explosive growth rates and myriad
+% interconnections of migration, were somehow estimated and provided as input,
+% ARGs sampled from the coalescent cannot capture the complexities
+% of family structure in these
+% datasets~\citep[e.g.][]{turnbull2018hundred,Ros-Freixedes2020}.
+% Another core assumption of the coalescent model is that the genome
+% (or at least the region under study) is short enough that the number of extant
+% lineages remains much smaller than $N_e$ at all times.
+% High-quality whole genome assemblies are now available
+% for many species % Can we say something more concrete here?
+% and projects are under way to obtain
+% them for tens of thousands more~\citep{darwin2022sequence,lewin2022earth},
+% and so we can expect inferred ARGs to routinely span whole chromosomes.
 

### 2023-10-31 paper.tex
para on the true arg and relation to models

@@
 differently (Appendix~\ref{sec-arg-history}).
-Another concept of an ARG that we must be careful to distinguish,
+Another concept of an ARG that we must be careful to distinguish
 is that of the \emph{actual} ARG underlying a sample of genomes
@@
 (Appendix~\ref{sec-cell-lineages-and-args}).
-Just as this true, ideal ARG unquestionably exists,
-it can also never be entirely known,
-% NOTE: preempting some discussion here:
-% Mutation rate is (a) not high enough in existing species to
-% resolve, (b) probably couldn't be if organisms are to survive,
-% and (c) such a high mutation rate would probably even obscure
-% the more ancient past.
-because of a fundamental lack of mutational information.
-
-The scale of modern datasets makes the
-relationship between the true ARG and
-population genetic models important to articulate.
+Just as this true, species ARG unquestionably exists,
+it can also never be entirely known.
+% TODO: I'm not happy with the phrasing here, anyone want to jump in?
+Even if a species
+had a sufficiently high mutation rate to identify every
+recent branching point uniquely (and somehow managed to survive)
+this high rate of mutations would inevitably obscure
+more ancient structure.
+% Must be a citation for this?
+Similarly, it seems unlikely that the details
+of ancient recombinations can ever be known with much precision.
+
+% Popgen models and the species ARG
+Species-level ARGs may seem fanciful, but the scale of
+modern datasets makes it necessary for us to grapple with the idea.
+The UK Biobank (UKB), for example, has
+genotype data for around 500,000 humans~\citep{bycroft2018genome},
+and exome~\citep{backman2021exome} and
+whole genome sequence~\citep{halldorsson2022sequences} data
+for large subsets of this cohort.
+UKB is just one of many such projects~\citep[e.g.][]{turnbull2018hundred,
+karczewski2020mutational,tanjo2021practical}, and
+agricultural datasets are even more
+detailed~\citep[e.g.][]{hayes20191000,Ros-Freixedes2020}.
+While we are not sequencing the entire species yet,
+we are certainly sampling at a scale where the assumptions of
+population genetics models are substantially broken.
+By necessity and design, such mathematical models are caricatures
+of biological reality,
+allowing us to draw general conclusions about population processes
+without the burden of unnecessary detail.
 The coalescent with recombination (henceforth, ``coalescent'')
-is a mathematical model that defines a distribution of ARGs
+is a powerful model that defines a distribution of ARGs
 for a given set of parameters.
-It has proven to be a useful and robust
-model~\citep{wakeley2012gene,bhaskar2014distortion,nelson2020accounting},
-but many modern datasets have properties that grossly
-violate its assumptions.
-One key assumption is that
-sample size $n$ is much less than the effective population size, $N_e$.
-Several human datasets now consist of hundreds of thousands of
-genomes~\citep{turnbull2018hundred, bycroft2018genome,
-karczewski2020mutational,tanjo2021practical,
-halldorsson2022sequences},
-and so sample size is an order of magnitude \emph{larger} than the
+A key assumption of the coalescent is that
+sample size is much less than the effective population size ($N_e$)
+and although the model has proved surprisingly robust to violations
+of this assumption~\citep{wakeley2012gene,bhaskar2014distortion,
+nelson2020accounting},
+sample size can now be an order of magnitude or more \emph{larger} than the
 usually assumed $N_e$ values.
-Agricultural datasets are an even more extreme departure from this
-assumption, with hundreds of thousands of samples embedded in
-multi-generational pedigrees~\citep{hayes20191000,Ros-Freixedes2020}
-and effective population sizes of 100 and even
-less~\citep{MacLeod2013,Makanjuola2020,Hall2016,Porcnic2016}.
 A model assuming a single $N_e$ would be a
@@
 ARGs sampled from the coalescent cannot capture the complexities
-of family structure in these
-datasets~\citep[e.g.][]{turnbull2018hundred,Ros-Freixedes2020}.
-Another core assumption of the coalescent model is that the genome
-(or at least the region under study) is short enough that the number of extant
-% TODO do we need to explain this a bit more?
-lineages remains much smaller than $N_e$ at all times in the past.
-High-quality whole genome assemblies are now available
-for many species % Can we say something more specific here?
-and projects are under way to obtain
-them for tens of thousands more~\citep{darwin2022sequence,lewin2022earth}.
-Vast quantities of whole genome sequence data across a broad range
-of species exists today.
-
+of family structure or the vagaries of history inherent in these
+datasets.
+
+% What is the problem?
 The true ARG underlying a modern population-scale dataset
-is therefore not a sample from the coalescent, or any other
-population genetics model.
-
-% Mathematical models are insufficient to describe the complexities
-% of biological reality when sampling is at population scale,
-% and the true ARG can never be known in its entirity.
-
-A reasonable goal to set, then, is to estimate \emph{aspects}
+therefore cannot be known in its entirety, and the value
+of attempting to describe the intricate details of the recent past
+through mathematical models is questionable.
+A pragmatic goal that we may set, then, is to infer \emph{aspects}
 of the true ARG and to capture uncertainty where it inevitably exists.
 This is the approach taken by recent inference
+% TODO should we mention SARGE? I don't think it's in the same class, but
+% maybe I'm wrong?
 methods~\citep{kelleher2019inferring,speidel2019method,zhang2023biobank},
-by neither fully imposing the constraints of a mathematical
-model or trying estimate the true ARG is its compete detail, they
-have made substantial progress. However, our vocabularly
-and mathematical formalisms, rooted in both the ideal and
-mathematical concepts of an ARG, are insufficient to capture the
-richness of these new structures.
-
-% This ideal ARG is also not a sample from the coalescent
-% or any other population genetics model: by necessity, and design,
-% such mathematical constructs are caricatures of biological reality,
-% allowing us to draw general conclusions about population processes
-% without the burden of unnecessary detail.
-% The scale of present-day datasets breaks fundamental assumptions
-% made by such models, and we must therefore be careful to articulate
-% our goals in inference.
-
-% pedantically, ARG is not really a "word"
-The term ``ARG'' is now frequently used in a general
+which estimate an approximate version of the true
+ARG, informed, but not strictly constrained, by population genetic
+models and parsimony.
+While these methods have made substantial progress there
+are, however, some significant problems in this emerging field.
+Our terminology and mathematical formalisms for ARGs,
+ultimately derived from the coalescent,
+are ill-suited to describing these approximate structures.
+% TODO add some citations of examples?
+Discussions are forced to contrast a ``full'' ARG, containing
+a precise description of every recombination or coalescence event,
+with a sequence of disconnected local trees.
+This view is depauperate, obscuring many shades of important
+and interesting detail between these two extremes.
+Although the term ``ARG'' is now often used in a general
 sense~\citep[e.g.][]{mathieson2020ancestry,hejase2020summary,
 schaefer2021ancestral,harris2023using,zhang2023biobank,fan2023likelihood},
-as a data structure
+as informally
 encompassing the varied outputs from modern simulation and
 inference methods~\citep{rasmussen2014genome, palamara2016argon, haller2018tree,
-kelleher2019inferring, speidel2019method, baumdicker2021efficient, zhang2023biobank}.
-This broad usage is a useful generalisation,
-but is not compatible with the available formal
-data structure definitions
-derived from the original Griffiths approach.
-This mismatch between usage and definition has led to significant
-problems, which if not addressed
-may impede progress in applying ARGs to understand the biological world.
-
-% % Para 3: what is the problem?
-The classical formal definition of an ARG is derived from the
-stochastic process, and expressed in terms of events:
-the common ancestor and recombination events
-that occur in the history of a sample. We show that this approach
-to representing ARGs (which we call the ``event ARG'', or eARG
-encoding) is inherently limiting, insufficient to represent
-the complexities of modern datasets, and incompatible with
-the outputs of modern large-scale simulation and inference
-methods~\citep{kelleher2016efficient,speidel2019method,
-kelleher2019inferring,zhang2023biobank}.
-We address these problems by providing a more flexible formulation
-of an ARG data structure, which we call the ``genome ARG''
-or gARG encoding. We show that it is fully compatible with the
-classical eARG approach, while providing much more flexibility
-in how genetic inheritance can be represented,
-and encompasses the outputs of modern methods.
-Moreover we show that, in contrast with the eARG encoding which
-requires precise estimates of each recombination event,
-the gARG encoding allows us to systematically describe uncertainty
-about the temporal ordering of multiple recombinations.
+kelleher2019inferring, speidel2019method, baumdicker2021efficient,
+zhang2023biobank},
+there is no corresponding mathematical description of an ARG data structure.
+Classical formalisms, based on the Griffiths approach, are insufficient to
+describe these rich details.
+Given that ARGs have already been estimated for the 500,000
+humans in UKB~\citep{kelleher2019inferring,zhang2023biobank} and
+more than a million SARS-CoV-2 genomes~\citep{zhan2023towards},
+it is vital that our formal methods and terminology catch up
+with this reality.
+
+% % pedantically, ARG is not really a "word"
+% The term ``ARG'' is now frequently used in a general
+% sense~\citep[e.g.][]{mathieson2020ancestry,hejase2020summary,
+% schaefer2021ancestral,harris2023using,zhang2023biobank,fan2023likelihood},
+% as a data structure
+% encompassing the varied outputs from modern simulation and
+% inference methods~\citep{rasmussen2014genome, palamara2016argon, haller2018tree,
+% kelleher2019inferring, speidel2019method, baumdicker2021efficient, zhang2023biobank}.
+% This broad usage is a useful generalisation,
+% but is not compatible with the available formal
+% data structure definitions
+% derived from the original Griffiths approach.
+% This mismatch between usage and definition has led to significant
+% problems, which if not addressed
+% may impede progress in applying ARGs to understand the biological world.
+
+% % % Para 3: what is the problem?
+% The classical formal definition of an ARG is derived from the
+% stochastic process, and expressed in terms of events:
+% the common ancestor and recombination events
+% that occur in the history of a sample. We show that this approach
+% to representing ARGs (which we call the ``event ARG'', or eARG
+% encoding) is inherently limiting, insufficient to represent
+% the complexities of modern datasets, and incompatible with
+% the outputs of modern large-scale simulation and inference
+% methods~\citep{kelleher2016efficient,speidel2019method,
+% kelleher2019inferring,zhang2023biobank}.
+% We address these problems by providing a more flexible formulation
+% of an ARG data structure, which we call the ``genome ARG''
+% or gARG encoding. We show that it is fully compatible with the
+% classical eARG approach, while providing much more flexibility
+% in how genetic inheritance can be represented,
+% and encompasses the outputs of modern methods.
+% Moreover we show that, in contrast with the eARG encoding which
+% requires precise estimates of each recombination event,
+% the gARG encoding allows us to systematically describe uncertainty
+% about the temporal ordering of multiple recombinations.
 

### 2023-10-31 paper.tex
Problem para

@@
 schaefer2021ancestral,harris2023using,zhang2023biobank,fan2023likelihood},
-as informally
+informally
 encompassing the varied outputs from modern simulation and
@@
 
-% % pedantically, ARG is not really a "word"
-% The term ``ARG'' is now frequently used in a general
-% sense~\citep[e.g.][]{mathieson2020ancestry,hejase2020summary,
-% schaefer2021ancestral,harris2023using,zhang2023biobank,fan2023likelihood},
-% as a data structure
-% encompassing the varied outputs from modern simulation and
-% inference methods~\citep{rasmussen2014genome, palamara2016argon, haller2018tree,
-% kelleher2019inferring, speidel2019method, baumdicker2021efficient, zhang2023biobank}.
-% This broad usage is a useful generalisation,
-% but is not compatible with the available formal
-% data structure definitions
-% derived from the original Griffiths approach.
-% This mismatch between usage and definition has led to significant
-% problems, which if not addressed
-% may impede progress in applying ARGs to understand the biological world.
-
-% % % Para 3: what is the problem?
-% The classical formal definition of an ARG is derived from the
-% stochastic process, and expressed in terms of events:
-% the common ancestor and recombination events
-% that occur in the history of a sample. We show that this approach
-% to representing ARGs (which we call the ``event ARG'', or eARG
-% encoding) is inherently limiting, insufficient to represent
-% the complexities of modern datasets, and incompatible with
-% the outputs of modern large-scale simulation and inference
-% methods~\citep{kelleher2016efficient,speidel2019method,
-% kelleher2019inferring,zhang2023biobank}.
-% We address these problems by providing a more flexible formulation
-% of an ARG data structure, which we call the ``genome ARG''
-% or gARG encoding. We show that it is fully compatible with the
-% classical eARG approach, while providing much more flexibility
-% in how genetic inheritance can be represented,
-% and encompasses the outputs of modern methods.
-% Moreover we show that, in contrast with the eARG encoding which
-% requires precise estimates of each recombination event,
-% the gARG encoding allows us to systematically describe uncertainty
-% about the temporal ordering of multiple recombinations.
+We address these problems here by providing a simple formal description of an
+ARG data structure, based on recording specific genomes and their intervals of
+genetic inheritance. We call this the ``genome ARG'', or gARG encoding. We
+contrast this with the classical formal definition of an ARG, based on
+recording common ancestor and recombination events, which we refer to as the
+``event ARG'' or eARG encoding. We show that the new gARG encoding it is a
+substantial generalisation of the classical eARG approach,  providing much more
+flexibility in how genetic inheritance can be represented, and encompassing the
+outputs of modern methods. We show that the gARG approach can represent many
+different types of approximation, in particular allowing us to systematically
+describe uncertainty about the temporal ordering of multiple recombinations. It
+is important to note that throughout we are interested in the details of these
+competing mathematical formulations and their practical consequences.
+% Do we need to say this? Probably not?
+We are not questioning the form of the true ARG, or its existance.
 
@@
 % reason about its properties.
-
-% We must also distinguish between a specific realisation
-% the \emph{actual} ARG underlying a sample of genomes
 

### 2023-10-31 paper.tex
Tweaks from Yan

--- 2023-10-30 paper.tex
@@
 and subfields use the term
-differently (Appendix~\ref{sec-arg-history}).
-Another concept of an ARG that we must be careful to distinguish,
-is that of the \emph{actual} ARG underlying a sample of genomes
-and ultimately an entire species. This ARG is the true history,
-perfectly resolved into binary splits and mergers by
-the cellular processes of meiosis and mitosis, regardless
-of sampling density or population processes
-(Appendix~\ref{sec-cell-lineages-and-args}).
-Just as this true, ideal ARG unquestionably exists,
-it can also never be entirely known,
-% NOTE: preempting some discussion here:
-% Mutation rate is (a) not high enough in existing species to
-% resolve, (b) probably couldn't be if organisms are to survive,
-% and (c) such a high mutation rate would probably even obscure
-% the more ancient past.
-because of a fundamental lack of mutational information.
-
-The scale of modern datasets makes the
-relationship between the true ARG and
-population genetic models important to articulate.
-The coalescent with recombination (henceforth, ``coalescent'')
-is a mathematical model that defines a distribution of ARGs
-for a given set of parameters.
-It has proven to be a useful and robust
-model~\citep{wakeley2012gene,bhaskar2014distortion,nelson2020accounting},
-but many modern datasets have properties that grossly
-violate its assumptions.
-One key assumption is that
-sample size $n$ is much less than the effective population size, $N_e$.
-Several human datasets now consist of hundreds of thousands of
-genomes~\citep{turnbull2018hundred, bycroft2018genome,
-karczewski2020mutational,tanjo2021practical,
-halldorsson2022sequences},
-and so sample size is an order of magnitude \emph{larger} than the
-usually assumed $N_e$ values.
-Agricultural datasets are an even more extreme departure from this
-assumption, with hundreds of thousands of samples embedded in
-multi-generational pedigrees~\citep{hayes20191000,Ros-Freixedes2020}
-and effective population sizes of 100 and even
-less~\citep{MacLeod2013,Makanjuola2020,Hall2016,Porcnic2016}.
-A model assuming a single $N_e$ would be a
-drastic over-simplification in these cases of course, but
-even if sufficiently complex demographic models~\citep{gower2022demes}
-encompassing hundreds of populations, explosive growth rates and myriad
-interconnections of migration, were somehow estimated and provided as input,
-ARGs sampled from the coalescent cannot capture the complexities
-of family structure in these
-datasets~\citep[e.g.][]{turnbull2018hundred,Ros-Freixedes2020}.
-Another core assumption of the coalescent model is that the genome
-(or at least the region under study) is short enough that the number of extant
-% TODO do we need to explain this a bit more?
-lineages remains much smaller than $N_e$ at all times in the past.
-High-quality whole genome assemblies are now available
-for many species % Can we say something more specific here?
-and projects are under way to obtain
-them for tens of thousands more~\citep{darwin2022sequence,lewin2022earth}.
-Vast quantities of whole genome sequence data across a broad range
-of species exists today.
-
-The true ARG underlying a modern population-scale dataset
-is therefore not a sample from the coalescent, or any other
-population genetics model.
-
-% Mathematical models are insufficient to describe the complexities
-% of biological reality when sampling is at population scale,
-% and the true ARG can never be known in its entirity.
-
-A reasonable goal to set, then, is to estimate \emph{aspects}
-of the true ARG and to capture uncertainty where it inevitably exists.
-This is the approach taken by recent inference
-methods~\citep{kelleher2019inferring,speidel2019method,zhang2023biobank},
-by neither fully imposing the constraints of a mathematical
-model or trying estimate the true ARG is its compete detail, they
-have made substantial progress. However, our vocabularly
-and mathematical formalisms, rooted in both the ideal and
-mathematical concepts of an ARG, are insufficient to capture the
-richness of these new structures.
-
-% This ideal ARG is also not a sample from the coalescent
-% or any other population genetics model: by necessity, and design,
-% such mathematical constructs are caricatures of biological reality,
-% allowing us to draw general conclusions about population processes
-% without the burden of unnecessary detail.
-% The scale of present-day datasets breaks fundamental assumptions
-% made by such models, and we must therefore be careful to articulate
-% our goals in inference.
-
+differently (see Appendix~\ref{sec-arg-history}).
 % pedantically, ARG is not really a "word"
@@
 This broad usage is a useful generalisation,
-but is not compatible with the available formal
-data structure definitions
+but is not compatible with the available formal definitions
 derived from the original Griffiths approach.
@@
 about the temporal ordering of multiple recombinations.
-
-% A data structure is a means of encoding information, and
-% a precise description of the structure allows us to mathematically
-% reason about its properties.
-
-% We must also distinguish between a specific realisation
-% the \emph{actual} ARG underlying a sample of genomes
 
@@
 differential inheritance of genetic material between genomes.
-% % This is definitely needed for the Nicks, who are always thinking
-% % about the deeper processes and are quite happy being loosey-goosey
-% % with representation details
+% This is definitely needed for the Nicks, who are always thinking
+% about the deeper processes and are quite happy being loosey-goosey
+% with representation details
 It is important to note that here, and throughout,
@@
 
-% Fully decoupling the general concept of an ARG from the coalescent
-% with recombination (henceforth, ``coalescent'')
-% is an important step.
-% While the coalescent has proven to be a useful and
-% robust
-% model~\citep{wakeley2012gene,bhaskar2014distortion,nelson2020accounting},
-% many modern datasets have properties that grossly
-% violate its assumptions.
-% One key assumption is that
-% sample size $n$ is much less than the effective population size, $N_e$.
-% Several human datasets now consist of hundreds of thousands of
-% genomes~\citep{turnbull2018hundred, bycroft2018genome,
-% karczewski2020mutational,tanjo2021practical,
-% halldorsson2022sequences},
-% and so sample size is an order of magnitude \emph{larger} than the
-% usually assumed $N_e$ values.
-% Agricultural datasets are an even more extreme departure from this
-% assumption, with hundreds of thousands of samples embedded in
-% multi-generational pedigrees~\citep{hayes20191000,Ros-Freixedes2020}
-% and effective population sizes of 100 and even
-% less~\citep{MacLeod2013,Makanjuola2020,Hall2016,Porcnic2016}.
-% A model assuming a single $N_e$ would be a
-% drastic over-simplification of course, but
-% even if sufficiently complex demographic models~\citep{gower2022demes}
-% encompassing hundreds of populations, explosive growth rates and myriad
-% interconnections of migration, were somehow estimated and provided as input,
-% ARGs sampled from the coalescent cannot capture the complexities
-% of family structure in these
-% datasets~\citep[e.g.][]{turnbull2018hundred,Ros-Freixedes2020}.
-% Another core assumption of the coalescent model is that the genome
-% (or at least the region under study) is short enough that the number of extant
-% lineages remains much smaller than $N_e$ at all times.
-% High-quality whole genome assemblies are now available
-% for many species % Can we say something more concrete here?
-% and projects are under way to obtain
-% them for tens of thousands more~\citep{darwin2022sequence,lewin2022earth},
-% and so we can expect inferred ARGs to routinely span whole chromosomes.
+Fully decoupling the general concept of an ARG from the coalescent
+with recombination (henceforth, ``coalescent'')
+is an important step.
+While the coalescent has proven to be a useful and
+robust
+model~\citep{wakeley2012gene,bhaskar2014distortion,nelson2020accounting},
+many modern datasets have properties that grossly
+violate its assumptions.
+One key assumption is that
+sample size $n$ is much less than the effective population size, $N_e$.
+Several human datasets now consist of hundreds of thousands of
+genomes~\citep{turnbull2018hundred, bycroft2018genome,
+karczewski2020mutational,tanjo2021practical,
+halldorsson2022sequences},
+and so sample size is an order of magnitude \emph{larger} than the
+usually assumed $N_e$ values.
+Agricultural datasets are an even more extreme departure from this
+assumption, with hundreds of thousands of samples embedded in
+multi-generational pedigrees~\citep{hayes20191000,Ros-Freixedes2020}
+and effective population sizes of 100 and even
+less~\citep{MacLeod2013,Makanjuola2020,Hall2016,Porcnic2016}.
+A model assuming a single $N_e$ would be a
+drastic over-simplification of course, but
+even if sufficiently complex demographic models~\citep{gower2022demes}
+encompassing hundreds of populations, explosive growth rates and myriad
+interconnections of migration, were somehow estimated and provided as input,
+ARGs sampled from the coalescent cannot capture the complexities
+of family structure in these
+datasets~\citep[e.g.][]{turnbull2018hundred,Ros-Freixedes2020}.
+Another core assumption of the coalescent model is that the genome
+(or at least the region under study) is short enough that the number of extant
+lineages remains much smaller than $N_e$ at all times.
+High-quality whole genome assemblies are now available
+for many species % Can we say something more concrete here?
+and projects are under way to obtain
+them for tens of thousands more~\citep{darwin2022sequence,lewin2022earth},
+and so we can expect inferred ARGs to routinely span whole chromosomes.
 

### 2023-10-31 paper.tex

--- 2023-10-31 paper.tex
@@
 We address these problems here by providing a simple formal description of an
-ARG data structure, based on recording specific genomes and their intervals of
-genetic inheritance. We call this the ``genome ARG'', or gARG encoding. We
+ARG data structure, based on recording the intervals of
+genetic inheritance between specific genomes.
+We call this the ``genome ARG'', or gARG encoding. We
 contrast this with the classical formal definition of an ARG, based on
@@
 breakpoint in an eARG is a concise and elegant way of describing realisations
-of the coalescent with recombination, it has some significant issues when applied to the
-complexities of present-day datasets. The eARG encoding explicitly models only
+of the coalescent, it is inherently limited when
+implemented literally. The eARG encoding explicitly models only
 two different types of event and thus anything that is not a single crossover
@@
 section offers a much simpler solution.
-Modern datasets contain many more complexities which the eARG encoding
-is ill-suited to represent (see Discussion).
+
+Aside from these obvious practical challenges arising from a
+literal implementation of the Griffiths approach, there is a deeper
+issue with the implicit strategy of basing an ARG data structure on
+recording events and their properties (e.g.\ the crossover breakpoint
+for a recombination event). The fundamental problem with this approach
+is that it assumes that all events are \emph{knowable}, and does not
+provide any obvious mechanism for either aggregating multiple events
+or expressing uncertainty about them. While this is not a
+problem when describing the results of simulations (where all details
+are perfectly known), it is a major issue when we wish to
+formally describe the output of inference methods, particularly
+as datasets approach the population scale.
+As discussed in the introduction,
+the precise details of all events in these vast ARGs can never be
+known, and a data structure that \emph{enforces} complete
+precision is therefore an impediment to progress.
 

### 2023-11-01 paper.tex
Various updates to intro

@@
 % Be specific about what we're talking about.
-ARGs are classically defined by the graph of
+An ARG is classically defined as a graph of
 % We can play a bit loose here, please don't change to "common ancestor"
 coalescence and recombination events in the history of a sample of genomes,
-with the differential inheritance structure encoded by the
-crossover breakpoints associated with recombination events.
+encoding an inheritance structure where
+crossover breakpoints are associated with recombination events.
 We show that this ``event ARG'' (eARG) encoding
 is limited in the patterns of genetic inheritance that can be represented,
-being insufficient to describe the outputs of modern ARG simulation
+being insufficient to describe the outputs of recent ARG simulation
 and inference methods,
@@
 specific genomes and their intervals of genetic inheritance.
-We show that the gARG encoding can fully capture the richness of modern large-scale datasets
+We show that the gARG encoding can fully capture the richness of large-scale datasets
 and is compatible with modern simulation and inference methods.
-Moreover, we show that, unlike the eARG encoding which requires
+We also show that, unlike the eARG encoding which requires
 % TODO rephrase "to be detailed" is ambiguous
-all recombination events to be detailed,
+all recombination events to be encoded,
 the gARG encoding enables fine-grained
@@
 % Para 1: ARGs are cool!
-Estimating the genetic genealogy of a set of genome sequences
+% NOTE: slight inaccuracy of "DNA" here offset by making the problem concrete.
+Estimating the genetic genealogy of a set of DNA sequences
 under the influence of recombination,
@@
 caused by recombination, and encodes a sequence of correlated genealogical
-trees along the genome.
+trees along the genome
+(see \citet{lewanski2023era} for a biologically oriented introduction).
 Recent breakthroughs
@@
 follow~\citep{harris2019database,harris2023using}.
-\citet{lewanski2023era} provide a biologically oriented introduction to ARGs
-and further discuss their potential impacts.
 
@@
 a \emph{realisation} of such a random process,
-or an inference from a set of sampled genome sequences.
+or an inferred ancestry of a sample of genomes.
 The distinction between stochastic process
@@
 is that of the \emph{actual} ARG underlying a sample of genomes
-and ultimately an entire species. This ARG is the true history,
+and ultimately an entire species or higher grouping. 
+This ARG is the true history,
 perfectly resolved into binary splits and mergers by
@@
 (Appendix~\ref{sec-cell-lineages-and-args}).
-Just as this true, species ARG unquestionably exists,
+Just as this true ARG unquestionably exists,
 it can also never be entirely known.
 % TODO: I'm not happy with the phrasing here, anyone want to jump in?
-Even if a species
-had a sufficiently high mutation rate to identify every
-recent branching point uniquely (and somehow managed to survive)
-this high rate of mutations would inevitably obscure
-more ancient structure.
+Even if the mutation rate were high enough to identify every
+recent branching point uniquely (while remaining biologically survivable),
+such a high rate would saturate the genome with mutations, inevitably obscuring
+past history.
 % Must be a citation for this?
-Similarly, it seems unlikely that the details
+Similarly, it is unlikely that the details
 of ancient recombinations can ever be known with much precision.
@@
 karczewski2020mutational,tanjo2021practical}, and
-agricultural datasets are even more
-detailed~\citep[e.g.][]{hayes20191000,Ros-Freixedes2020}.
-While we are not sequencing the entire species yet,
+agricultural datasets are even
+richer in terms of generations~\citep[e.g.][]{hayes20191000,Ros-Freixedes2020}.
+While we are not sequencing entire species,
 we are certainly sampling at a scale where the assumptions of
@@
 without the burden of unnecessary detail.
+ % TODO we're singling out the coalescent here, but it's just an 
+ % exmaple. The DTWF is no better an approximation of the last 5
+ % generations of UKB, e.g.
 The coalescent with recombination (henceforth, ``coalescent'')
-is a powerful model that defines a distribution of ARGs
+is a powerful model that defines a distribution over ARGs
 for a given set of parameters.
@@
 interconnections of migration, were somehow estimated and provided as input,
-ARGs sampled from the coalescent cannot capture the complexities
+ARGs sampled from the coalescent cannot capture the full complexities
 of family structure or the vagaries of history inherent in these
@@
 therefore cannot be known in its entirety, and the value
+% TODO rephrase heavily!
 of attempting to describe the intricate details of the recent past
@@
 recording common ancestor and recombination events, which we refer to as the
-``event ARG'' or eARG encoding. We show that the new gARG encoding it is a
+``event ARG'' or eARG encoding. We show that the new gARG encoding is a
 substantial generalisation of the classical eARG approach,  providing much more
@@
 % Do we need to say this? Probably not?
-We are not questioning the form of the true ARG, or its existance.
-
-% A data structure is a means of encoding information, and
-% a precise description of the structure allows us to mathematically
-% reason about its properties.
+We are not questioning the form of the true ARG, or its existence.
 
@@
 node $u\in N$ is associated with a time $\tau_u$. 
+A node can represent any genome along a 
+chain of cell divisions~\ref{sec-cell-lineages-and-args},
+or can be interpreted as representing one of ``the'' genomes of an
+individual. How we assign dates will therefore vary according to application.
 The topological ordering defined by the directed graph structure
@@
 For example, \emph{which} genome along a chain of cell divisions should be
-regarded as the one where an event occurred
-(Appendix~\ref{sec-cell-lineages-and-args}),
-or whether multiple recombinations
+regarded as the one where an event occurred,
+or whether multiple coalescences 
 within a single individual should be regarded as one or multiple events are
-debatable points.
+debatable points (Appendix~\ref{sec-cell-lineages-and-args}).
 % If we want to have an ARG software ecosystem then such fireside
@@
 time in a ``prospective'' manner. Genetic inheritance relationships
-are recorded exhaustively, generation-by-generation,
+and mutations are recorded exhaustively, generation-by-generation,
 leading to a rapid build-up of information, much of which
@@
 are related, and we require a concrete mathematical structure
-to describe the local trees.
+to describe the local trees. It is important to note that 
+although the following discussion is phrased in terms of the 
+gARG encoding,  the arguments apply equally to eARGs
+because any eARG can be converted to a 
+% Note, being sample-resolve is not necessary for following argument so omitting
+gARG without loss of information (Section~\ref{sec-ancestral-material}).
 
@@
 \citep{adrion2020community,lauterbur2023expanding}.
-Software versions were \kwarg\ v1.0, \argweaverD\ (2019),
-\tsinfer v0.3.1, and \relate v1.1.9: full details and replicable code for generating
+Software versions were \kwarg\~v1.0, \argweaverD\~(2019),
+\tsinfer\~v0.3.1, and \relate\~v1.1.9: full details and replicable code for generating
 these figures are available on GitHub (see Data Availability).
@@
 a rich supporting software ecosystem.
-Ideally, this would comprise a wide range of different
+Ideally, this would comprise a wide range of 
 inference methods specialised to different organisms, types

### 2023-11-01 paper.tex

@@
 % Be specific about what we're talking about.
+% TODO rephrase, this is wooly now
 An ARG is classically defined as a graph of
-% We can play a bit loose here, please don't change to "common ancestor"
 coalescence and recombination events in the history of a sample of genomes,
+% TODO this is vague
 encoding an inheritance structure where
@@
 and is compatible with modern simulation and inference methods.
+% TODO rewrite
 We also show that, unlike the eARG encoding which requires
-% TODO rephrase "to be detailed" is ambiguous
 all recombination events to be encoded,
@@
 % The grammar is a nightmare here - the stochastic process is singular, right?
-In its original form, the ARG is an alternative
+In its original form,
+developed by Griffiths and colleagues,
+the ARG is an alternative
 formulation of the coalescent with recombination~\citep{hudson1983properties},
 where the stochastic process of coalescence and recombination
-among ancestral lineages was formalised as a
-graph by Griffiths and colleagues~\citep{griffiths1991two,
+among ancestral lineages is formalised
+as a graph~\citep{griffiths1991two,
 ethier1990two,griffiths1996ancestral,griffiths1997ancestral}.
-Subsequently, ARGs have come to be thought of as a data
+Subsequently, an ARG has come to be thought of as a data
 structure~\citep{minichiello2006mapping}, i.e.\ describing
@@
 The distinction between stochastic process
-and data structure is not clear cut, however,
-and subfields use the term
+and data structure is not clear cut, however, and subfields use the term
 differently (Appendix~\ref{sec-arg-history}).
-Another concept of an ARG that we must be careful to distinguish
-is that of the \emph{actual} ARG underlying a sample of genomes
-and ultimately an entire species or higher grouping. 
-This ARG is the true history,
+Other subtly different concepts that we must be careful to distinguish
+are the ``true'' ARG, % I don't see a better word for "true" here
+describing the actual history of a sample of genomes,
+and a ``population'' ARG which is
+the true ARG of every individual in a population.
+% Note: I started off talking about the "ultimate" ARG of everone
+% who ever lived, but didn't find it that helpful.
+True ARGs are the actual history,
 perfectly resolved into binary splits and mergers by
@@
 (Appendix~\ref{sec-cell-lineages-and-args}).
-Just as this true ARG unquestionably exists,
-it can also never be entirely known.
-% TODO: I'm not happy with the phrasing here, anyone want to jump in?
-Even if the mutation rate were high enough to identify every
-recent branching point uniquely (while remaining biologically survivable),
-such a high rate would saturate the genome with mutations, inevitably obscuring
-past history.
-% Must be a citation for this?
-Similarly, it is unlikely that the details
-of ancient recombinations can ever be known with much precision.
-
-% Popgen models and the species ARG
-Species-level ARGs may seem fanciful, but the scale of
+Although population ARGs unquestionably exist,
+they can also never be entirely known because of a fundamental
+lack of mutational information.
+Even if mutation rate were high enough to uniquely identify every
+recent branching point,
+such a high rate would saturate the genome with mutations
+and obscure deeper history.
+
+% Population ARGs are here, now.
+Population ARGs may seem fanciful, but the scale of
 modern datasets makes it necessary for us to grapple with the idea.
@@
 for large subsets of this cohort.
-UKB is just one of many such projects~\citep[e.g.][]{turnbull2018hundred,
-karczewski2020mutational,tanjo2021practical}, and
-agricultural datasets are even
-richer in terms of generations~\citep[e.g.][]{hayes20191000,Ros-Freixedes2020}.
-While we are not sequencing entire species,
-we are certainly sampling at a scale where the assumptions of
-population genetics models are substantially broken.
-By necessity and design, such mathematical models are caricatures
-of biological reality,
-allowing us to draw general conclusions about population processes
-without the burden of unnecessary detail.
- % TODO we're singling out the coalescent here, but it's just an 
- % exmaple. The DTWF is no better an approximation of the last 5
- % generations of UKB, e.g.
-The coalescent with recombination (henceforth, ``coalescent'')
-is a powerful model that defines a distribution over ARGs
-for a given set of parameters.
-A key assumption of the coalescent is that
-sample size is much less than the effective population size ($N_e$)
-and although the model has proved surprisingly robust to violations
-of this assumption~\citep{wakeley2012gene,bhaskar2014distortion,
-nelson2020accounting},
-sample size can now be an order of magnitude or more \emph{larger} than the
-usually assumed $N_e$ values.
-A model assuming a single $N_e$ would be a
-drastic over-simplification in these cases of course, but
-even if sufficiently complex demographic models~\citep{gower2022demes}
-encompassing hundreds of populations, explosive growth rates and myriad
-interconnections of migration, were somehow estimated and provided as input,
-ARGs sampled from the coalescent cannot capture the full complexities
-of family structure or the vagaries of history inherent in these
-datasets.
-
-% What is the problem?
-The true ARG underlying a modern population-scale dataset
-therefore cannot be known in its entirety, and the value
-% TODO rephrase heavily!
-of attempting to describe the intricate details of the recent past
-through mathematical models is questionable.
-A pragmatic goal that we may set, then, is to infer \emph{aspects}
-of the true ARG and to capture uncertainty where it inevitably exists.
-This is the approach taken by recent inference
-% TODO should we mention SARGE? I don't think it's in the same class, but
-% maybe I'm wrong?
-methods~\citep{kelleher2019inferring,speidel2019method,zhang2023biobank},
-which estimate an approximate version of the true
-ARG, informed, but not strictly constrained, by population genetic
-models and parsimony.
-While these methods have made substantial progress there
-are, however, some significant problems in this emerging field.
-Our terminology and mathematical formalisms for ARGs,
-ultimately derived from the coalescent,
-are ill-suited to describing these approximate structures.
-% TODO add some citations of examples?
-Discussions are forced to contrast a ``full'' ARG, containing
-a precise description of every recombination or coalescence event,
-with a sequence of disconnected local trees.
-This view is depauperate, obscuring many shades of important
-and interesting detail between these two extremes.
+UKB is just one of many such population-scale sequencing
+projects~\citep[e.g.][]{turnbull2018hundred,
+karczewski2020mutational,tanjo2021practical}.
+Agricultural datasets are on a similar scale,
+and also include dense multi-generational sampling
+and near-perfect pedigree
+information~\citep[e.g.][]{hayes20191000,Ros-Freixedes2020}.
+Recent advances have made it possible to actually
+\emph{estimate} ARGs at population scale:
+ARGs have already been inferred for the 500,000
+humans in UKB~\citep{kelleher2019inferring,zhang2023biobank}
+and over a million SARS-CoV-2 genomes~\citep{zhan2023towards}.
+While this new population-scale reality presents many
+exciting opportunities, it also poses substantial challenges
+to existing methodologies.
+
+% What's the problem?
+A major problem currently facing the field is that
+classical mathematical formalisms and terminology
+cannot adequately describe these vast inferred ARGs.
+Fundamentally, these formalisms assume that an
+entire ARG is known with complete precision,
+and are not suited to describing partial or approximate knowledge.
+As we are actively inferring ARGs at the population scale,
+and such ARGs can never be known in complete detail,
+there is currently a substantial gap between our theoretical frameworks
+and practical application.
+The breakthroughs in scale achieved by recent
+% Tedious to cite again, but I don't want to leave Relate out of the
+% motivatory discussion
+methods~\citep[e.g.][]{kelleher2019inferring,speidel2019method,zhang2023biobank}
+are all based (in different ways) on inferring
+approximate \emph{structures},
+instead of a complete, fully detailed history.
+% Not sure the point is coming across that well here, but it's a
+% really important point and a significant
+It is important to distinguish here between
+structures and models: whether an inference method is based
+on heuristics or a rigorous mathematical model is orthogonal
+to the level of detail provided in its estimate.
+One could heuristically estimate a fully precise ARG,
+% this would be a cool thing to do!
+or statistically sample a partial,
+approximate ARG under a model
+such as the coalescent. % with recomb is surely unnecessary here
 Although the term ``ARG'' is now often used in a general
@@
 informally
-encompassing the varied outputs from modern simulation and
+encompassing the varied approximate structures output by modern simulation and
 inference methods~\citep{rasmussen2014genome, palamara2016argon, haller2018tree,
@@
 zhang2023biobank},
-there is no corresponding mathematical description of an ARG data structure.
-Classical formalisms, based on the Griffiths approach, are insufficient to
-describe these rich details.
-Given that ARGs have already been estimated for the 500,000
-humans in UKB~\citep{kelleher2019inferring,zhang2023biobank} and
-more than a million SARS-CoV-2 genomes~\citep{zhan2023towards},
-it is vital that our formal methods and terminology catch up
-with this reality.
+there is no corresponding mathematical description
+that is sufficiently general.
 
@@
 competing mathematical formulations and their practical consequences.
-% Do we need to say this? Probably not?
-We are not questioning the form of the true ARG, or its existence.
+% True ARGs, being perfectly resolved, can ultimately be described in
+% either formalism
 

### 2023-11-01 paper.tex
JK updates, addressing most outstanding comments

@@
 in the context of current datasets and research questions.
-We then discuss the important concept of ``ancestral material''
-in Section~\ref{sec-ancestral-material}, and how this relates
+Following this, we discuss the important concept of ``ancestral material''
+in Section~\ref{sec-ancestral-material}, and how it relates
 to the process of converting an eARG to a gARG.
@@
 with which arbitrary additional information can be associated.
-% Note: I don't see why we have to apologise for defining nodes as
-% integers, that's just the definition. I don't think anyone really
-% has a problem with this.
-% Concretely defining nodes as integers in this way has some
-% significant advantages. In particular, it provides a direct link
-% between an ARG and its local trees in terms of a well-defined
-% combinatorial object (Section~\ref{sec-ARG-and-local-trees}),
-% as well as enabling straightforward connection between
-% definitions and practical implementation (Section~\ref{sec-efficiency}).
 
@@
 resolves ancestral material.
-Simplification allows efficient subsetting of ARGs (removing both samples and
-certain levels of detail about recombination, as discussed in
-Section~\ref{sec-ARG-simplification}), and is also the key enabling factor when
-incorporating prospective ARGs into
-forward-time simulation.
+Efficient simplification is the key enabling factor for 
+this prospective-ARG based approach to forward-time simulation,
+which can be orders of magnitude faster than standard
+sequence-based methods
+(see Section~\ref{sec-ARG-simplification} for 
+other applications).
 % Define "sample-resolved" here for later use
@@
 ARGs and local trees.
-If we are given the sequence of local trees for a gARG
+Suppose we are given a gARG defined by a set of nodes and edges.
+There is no requirement on the structure of this ARG beyond
+the basic definitions: it could correspond to an ARG in which
+every recombination event is exactly specified 
+(e.g.\ Fig.~\ref{fig-ancestry-resolution})
+or one in which local trees are entirely disjoint
+(i.e.\ only the sample nodes are shared between them).
+If we  are given the sequence of local trees for this gARG
 encoded as an oriented tree, along with the genome interval
-covered by each tree, then we can recover precisely the same ARG.
+covered by each tree, we can recover the original gARG exactly.
 % We do this site-by-site for simplicity, but hopefully it's obvious
@@
 ``does not contain all of the information in the underlying ARG: it lacks the
-number of recombinations occuring at each site, the times at which
+number of recombinations occurring at each site, the times at which
 recombinations occurred, and the specific sequences involved as recombination
@@
 for downstream applications.
-Many applications work by examining local trees independently.
+Many applications work by examining local trees independently,
+making detailed information about recombination events superfluous.
 For example, the \relate\ selection test~\citep{speidel2019method}
@@
 clearly much of the detail about recombination events in an ARG is lost.
-% You don't even need to define event nodes when using a graph-based
-% algorithm like tsdate
-The \texttt{tsdate} algorithm and related naive estimator
-of ancestral location explicitly use the gARG encoding
-to reason about ancestral nodes~\citep{wohns2022unified}.
+% Mentioning tsdate here doesn't advance the 
+% narrative I think.
+% The \texttt{tsdate} algorithm and related naive estimator
+% of ancestral location explicitly use the gARG encoding
+% to reason about ancestral nodes~\citep{wohns2022unified}.
 The main application for fully precise ARGs thus far has been
@@
 Recombination is an important factor in the evolution of
-SARS-CoV-2~\citep{vaninsberghe2021recombinant,jackson2021generation,ignatieva2022ongoing},
+SARS-CoV-2 \citep{vaninsberghe2021recombinant,jackson2021generation,ignatieva2022ongoing},
 % Tried to jazz this up a bit, was quite dry
@@
 valuable and timely addition to the field.
+% Someone should do this!
+Using ARGs simulated from observed pedigree data~\citep{anderson2022genes} as 
+ground-truth would also add a valuable dimension to 
+our understanding of how well methods perform when faced with 
+realistic population and family structure.
 

### 2023-11-01 paper.tex
New last para

@@
 them for tens of thousands more~\citep{darwin2022sequence,lewin2022earth},
-and so we can expect inferred ARGs to routinely span whole chromosomes.
+and so we can expect inferred ARGs to routinely span
+large fractions of a chromosome.
 
@@
 we could only ever explore the tiniest corner of the incomprehensibly
-large space of likely ARGs.
+large space of plausible ARGs.
 Therefore, it is important to systematically
@@
 One approach, enabled by the gARG encoding described here, is to allow
-nodes to have (rootward) in-degree greater than two (polytomies
+nodes to have more than two children (polytomies
 representing uncertainty over the ordering of coalescence events) or
-out-degree greater than two (representing uncertainty over the ordering
+more than two parents (representing uncertainty over the ordering
 of multiple recombination events). Development of other methods to capture, for example,
@@
 events only a small proportion are actually detectable from the data, assuming
-human-like mutation and recombination rates \citep{myers2002detection,hayman2023recoverability}.
+human-like mutation and recombination
+rates \citep{myers2002detection,hayman2023recoverability}.
 Even when a recombination event \emph{is} detectable, its timing and breakpoint
@@
 precise information about recombination is therefore a fundamental limitation.
+
 % % Do downstream applications actually make use of such full ARGs?
@@
 precise ARGs from data,
-we should also consider the value that such precise estimates provide
+we should also consider the value that such exact estimates provide
 for downstream applications.
@@
 and non-leaf sample nodes (sequences with descendants also present
-in the dataset) many of which only have a single descendant (i.e. are ARG-unary).
+in the dataset) many of which only have a single child (i.e.\ are ARG-unary).
 Recombination is an important factor in the evolution of
@@
 In each case, simulation studies demonstrated
-substantial differences between these quantities in simulated and reconstructed ARGs,
+substantial differences between these quantities in simulated and
+reconstructed ARGs
 that were not captured using tree-by-tree comparisons.
@@
 Ideally, this would comprise a wide range of
-inference methods specialised to different organisms, types
-and scales of data and
-% I'm thinking about things like Espalier specifically wanting
-% to characterise the recombs, vs Relate mainly wanting to
-% get good local trees
-targeting particular inference goals, sharing a common ARG interchange format.
-The output of these diverse inference methods could then be
-processed by numerous downstream applications,
-enabling users to freely compare the outputs of different
-inference methods on their data, and the result of different
-analyses on these inferences.
+inference methods specialised to different organisms,
+inference goals, and types and scales of data.
+If these diverse inference methods share a common,
+well-defined data format, their outputs could then be
+processed by many different downstream applications
+without the productivity-sapping problems
+of converting between (partially incompatible)
+program output formats~\citep{excoffier2006computer}.
 Earlier efforts to standardise ARG interchange shared this vision,
 but did not succeed~\citep{cardona2008extended,mcgill2013graphml}.
-Without a well-defined and efficient shared format
-for ARG data interchange,
-users will suffer from problems with converting data from one program's
-output to another, an unfortunate hallmark of population genetics
-software for many years~\citep{excoffier2006computer}.
-Current methods tend to tightly couple downstream analysis
-of the inferred ARG with the inference
-itself within the same software package,
+Current methods tend to tightly couple both ARG inference and downstream
+analysis within the same software package,
 which is ultimately not compatible with the widespread use
@@
 of classical concepts, capable of describing even the bewildering complexity
-of contemporary datasets and capturing important forms of temporal uncertainty.
-The \texttt{tskit} library is a high-quality open-source gARG implementation,
+of contemporary datasets and encompassing a wide range of approximate
+% Relate, tsinfer, ARGneedle all have different approx structures
+ARG structures, and would be a reasonable basis for such a community
+interchange format.
+
+% This is all a bit boring, maybe, but I think the people we want to
+% influence into using tskit will need to have this spelled out for them.
+Rigorously defining interchange formats~\cite[e.g.][]{kelleher2019htsget}
+is difficult and time-consuming,
+and no matter how precise the specification, in practise
+it is the \emph{implementations} that determines how well methods
+interoperate. The SAM/BAM file format~\citep{li2009sequence}
+is an instructive example. Originally developed as part of the
+1000 Genomes project~\citep{10002015global} to address the fragmented
+software ecosystem that existed at the time~\citep{danecek2021twelve},
+SAM/BAM has since become ubiquitous in bioinformatics pipelines.
+The excellent interoperability between methods
+exchanging alignment data is largely attributable to the
+success of \texttt{htslib}~\citep{bonfield2021htslib}, the
+software library that \emph{implements} SAM/BAM and several other
+foundational bioinformatics file formats.
+Today, there are dozens  % hundreds? Could check on github I guess?
+of tools that use \texttt{htslib} to
+support SAM/BAM and it this shared use of community software
+infrastructure that guarantees the smooth flow of data between
+applications.
+The emerging ARG software ecosystem could similarly benefit
+from the adoption of such shared community infrastructure
+to handle the mundane and time-consuming details of data interchange.
+The \texttt{tskit} library (Section~\ref{sec-efficiency}
+is a high-quality open-source gARG implementation,
 with proven efficiency and
@@
 that is already in widespread use.
-Adopting it as a community standard,
-and utilising its rich suite of operations,
-may accelerate progress and help to
-establish ARG methods within mainstream genetics.
+% TODO needs rephrasing
+Adopting it as a community standard may ease software implementation burden
+on researchers, freeing their time to address the many fascinating open
+questions and challenges that exist.
 

### 2023-11-02 paper.tex
Minor tweaks

@@
 to handle the mundane and time-consuming details of data interchange.
-The \texttt{tskit} library (Section~\ref{sec-efficiency}
+The \texttt{tskit} library (Section~\ref{sec-efficiency})
 is a high-quality open-source gARG implementation,

### 2023-11-02 paper.tex
Peters comments
closes #424

@@
 ethier1990two,griffiths1996ancestral,griffiths1997ancestral}.
+% Move into plural, an ARG is an instance of the data structure in this view
 Subsequently, an ARG has come to be thought of as a data
@@
 the true ARG of every individual in a population.
-% Note: I started off talking about the "ultimate" ARG of everone
-% who ever lived, but didn't find it that helpful.
 True ARGs are the actual history,
@@
 genotype data for around 500,000 humans~\citep{bycroft2018genome},
-and exome~\citep{backman2021exome} and
+along with exome~\citep{backman2021exome} and
 whole genome sequence~\citep{halldorsson2022sequences} data
-for large subsets of this cohort.
+for large subsets of the cohort.
 UKB is just one of many such population-scale sequencing
@@
 is a key concept in understanding the overall inheritance structure
-of an ARG. It denotes the genomic intervals ancestral to a set of samples
+of an ARG (here, and throughout, we use the general term ``ARG'' when
+the details of the specific encoding are not important).
+It denotes the genomic intervals ancestral to a set of samples
 on the edges of an ARG.
@@
 Then, the edges of the ``local ARG'' for this tree is given by
-$E^x = \{(u, \pi_u^x, \{x\} \mid \pi_u^x \neq 0\}$. Since the ARG
+$E^x = \{(u, \pi_u^x, \{x\}) \mid \pi_u^x \neq 0\}$. Because the ARG
 edges are defined by $(c, p, I)$ tuples, where the set $I$ defines
@@
 
-% Indeed, a
-% common statement in the literature is that modern tools such as
-% \tsinfer~\citep{kelleher2019inferring} and \relate~\citep{speidel2019method}
-% construct not an ARG but ``only'' a sequence of local trees
-% \citep[e.g.][]{hejase2020summary, brandt2021evaluation, ignatieva2021kwarg}.
-% While this statement is correct inasmuch as neither method attempts to infer
-% the explicit recombination events that cause adjacent trees to differ, it is
-% certainly not the case that they infer independent local trees. In fact, both
-% methods take advantage of genome-wide mutation information to construct
-% genealogies, and as a result nodes and edges persist across multiple local
-% trees. Their results can therefore be thought of as an inheritance graph, and
-% represented in the gARG format.
-
 \begin{figure} \begin{center}
@@
 \citep{adrion2020community,lauterbur2023expanding}.
-Software versions were \kwarg\~v1.0, \argweaverD\~(2019),
-\tsinfer\~v0.3.1, and \relate\~v1.1.9: full details and replicable code for generating
+Software versions were \kwarg~v1.0, \argweaverD~(2019),
+\tsinfer~v0.3.1, and \relate~v1.1.9.
+Full details and code for generating
 these figures are available on GitHub (see Data Availability).
@@
 \section*{Acknowledgements}
-We are grateful to Nick Barton, Gideon Bradburd, Alex Lewanski and Andrew Vaughan
-for helpful discussions and comments on the manuscript, and
+We are grateful to Nick Barton, Gideon Bradburd, Alex Lewanski, Andrew Vaughan
+and Peter Ralph for helpful discussions and comments on the manuscript, and
 to Iain Mathieson and Aylwyn Scally for interesting discussions that
@@
 % Anastasia Ignatieva acknowledges support from TODO
-% TODO Jerome: add EPSRC and NIH HG011395-0 grants.
 Jerome Kelleher acknowledges support from the Robertson Foundation,
@@
 
-%It is important to note that parsimonious ARGs
-%have quite different properties from realisations of the
-%ARG \emph{process} as modelled by Hudson, Griffiths, and colleagues.
-%Firstly, branch lengths are not typically estimated,
-%% why do we need the following two lines
-%and node times are required in order to compute a likelihood under the
-%coalescent with recombination (see section XXX).
-%Secondly, even if branch lengths were estimated,
-%the realisations would have a very small likelihood since
-%the coalescent with recombination is often highly \emph{un}parsimonious
-%in terms of recombination events.
-
 The goal of this historical overview is to illustrate that the meaning of the term ``ARG" now strongly

### 2023-11-02 paper.tex

@@
 \begin{abstract}
-New developments have made it possible to infer genetic genealogies in
-the presence of recombination at scale, enabling many
-downstream applications in population and statistical genetics.
-The structure representing such recombinant genetic ancestry
-is usually referred to as an ancestral recombination graph (ARG).
-% Be specific about what we're talking about.
-% TODO rephrase, this is wooly now
-An ARG is classically defined as a graph of
-coalescence and recombination events in the history of a sample of genomes,
-% TODO this is vague
-encoding an inheritance structure where
-crossover breakpoints are associated with recombination events.
-We show that this ``event ARG'' (eARG) encoding
-is limited in the patterns of genetic inheritance that can be represented,
-being insufficient to describe the outputs of recent ARG simulation
-and inference methods,
-and ill-suited to capturing the complex patterns of
-relatedness that exist in contemporary datasets.
-We propose an alternative ``genome ARG'' (gARG) encoding
-expressed in terms of
+As a result of recombination, adjacent nucleotides can have
+different paths of genetic inheritance and therefore
+the genealogical trees for
+% surely saying nucleotides and "sample of sequences" is concrete enough
+% here? This paper isn't for a general audience, let's just say what
+% we're talking about in direct language understandable by population
+% geneticists.
+a sample of sequences vary along the genome.
+The structure representing this ensemble of correlated local
+trees is usually referred to as an ancestral recombination graph (ARG).
+New developments have made it possible to infer ARGs at scale,
+enabling many new applications in population and statistical genetics.
+This rapid progress, however, has led to a
+% Controversial, perhaps, but true?
+substantial gap opening between theory and practice.
+Standard mathematical formalisms,
+% "detailing" here is intended to convey the we keep details about these
+% events like recomb breakpoint.
+based on exhaustively detailing the ``events'' that occur
+in the history of a sample,
+are insufficient to describe the outputs of current methods.
+We argue, moreover, that the precise estimates of
+\emph{all} events required
+by such formalisms are fundamentally unsuited to the realities
+of modern, population-scale datasets.
+We propose an alternative mathematical formulation that
+encompasses the outputs of recent methods and
+can capture the full richness of modern large-scale datasets.
+By defining this ARG encoding in terms of
 % we haven't define a "genome" yet, so need to give a bit of context
-specific genomes and their intervals of genetic inheritance.
-We show that the gARG encoding can fully capture the richness of large-scale datasets
-and is compatible with modern simulation and inference methods.
-% TODO rewrite
-We also show that, unlike the eARG encoding which requires
-all recombination events to be encoded,
-the gARG encoding enables fine-grained
-levels of precision about recombination to be represented,
-opening up a rich set of additional details.
+specific genomes and their intervals of genetic inheritance,
+we avoid the necessity of exhaustively listing (and estimating)
+\emph{all} events. The effects of multiple events can be
+aggregated in different ways, providing a natural way to express many
+forms of approximate and partial knowledge about the
+recombinant ancestry of a sample.
 \end{abstract}

### 2023-11-02 paper.tex
Minor edits

@@
 a sample of sequences vary along the genome.
-The structure representing this ensemble of correlated local
-trees is usually referred to as an ancestral recombination graph (ARG).
+The structure capturing the details of these intricately interwoven paths of 
+inheritance is referred to as an ancestral recombination graph (ARG).
 New developments have made it possible to infer ARGs at scale,
@@
 Although population ARGs unquestionably exist,
-they can also never be entirely known because of a fundamental
+they can also never be entirely known, in part because of a fundamental
 lack of mutational information.
@@
 Fundamentally, these formalisms assume that an
-entire ARG is known with complete precision,
+ARG is known in complete detail
 and are not suited to describing partial or approximate knowledge.
@@
 The breakthroughs in scale achieved by recent
-% Tedious to cite again, but I don't want to leave Relate out of the
-% motivatory discussion
 methods~\citep[e.g.][]{kelleher2019inferring,speidel2019method,zhang2023biobank}
-are all based (in different ways) on inferring
-approximate \emph{structures},
-instead of a complete, fully detailed history.
+are all based, in different ways, on inferring
+approximate \emph{structures}
+instead of a complete and fully detailed history.
 % Not sure the point is coming across that well here, but it's a
@@
 zhang2023biobank},
-there is no corresponding mathematical description
+there is no corresponding mathematical definition
 that is sufficiently general.
 
-We address these problems here by providing a simple formal description of an
+We address this problem by providing a simple formal definition of an
 ARG data structure, based on recording the intervals of
@@
 to the process of converting an eARG to a gARG.
-Using this new definition for the remainder of the paper,
-we continue
+We continue
 in Section~\ref{sec-ARG-and-local-trees} by considering the relationship
 between an ARG and its local trees. Contrary to the
-prevailing view, we show that a (suitably encoded) sequence of local trees contains
+prevailing view, we show that a suitably encoded sequence of local trees contains
 precisely as much information as the corresponding ARG.
@@
 (Section~\ref{sec-ARG-simplification}),
-and (relatedly) the degrees of precision about recombination
+and the degrees of precision about recombination
 that can be stored and we may seek to infer (Section~\ref{sec-precision}).
@@
 A node can represent any genome along a
+% A gamete *is* one of the genomes along a chain of cell divisions? 
+% I don't see what adding a gamete clarifies, and we cite Hudson plenty
 chain of cell divisions~\ref{sec-cell-lineages-and-args},
-or can be interpreted as representing one of ``the'' genomes of an
-individual. How we assign dates will therefore vary according to application.
+or can be interpreted as representing one of the genomes of an
+individual; how we assign dates  to nodes
+will therefore vary according to application.
 The topological ordering defined by the directed graph structure
@@
 Genomes \noderef{a} and \noderef{c} are the product of two independent
-meioses (recombination events in red) between
+meioses (recombination events, in red) between
 the paternal genomes \noderef{e}
@@
 recording events and their properties (e.g.\ the crossover breakpoint
-for a recombination event). The fundamental problem with this approach
-is that it assumes that all events are \emph{knowable}, and does not
+for a recombination event). The fundamental problem
+is that this approach assumes all events are \emph{knowable}, and does not
 provide any obvious mechanism for either aggregating multiple events
@@
 the two gametes produced by the cells in the left half of $D_{10}$ but not shared by
-the right hand gamete. With a high enough germline mutation rate (or equivalently, a long enough genome), each
-round of mitotic genome duplication within a single multicellular organism could
+the right hand gamete. With a high enough mutation rate (or equivalently, a long enough genome), each
+round of mitotic germline genome duplication within a single multicellular organism could
 potentially be distinguished.

### 2023-11-02 cover-letter/cover-letter.tex
Initial outline of cover-letter

\documentclass{letter}

\signature{Jerome Kelleher}

\address{Big Data Institute\\University of Oxford\\UK}
\begin{document}

\begin{letter}{GENETICS}

\opening{Dear Editors,}

I am writing on behalf of my coauthors to submit our manuscript entitled
\emph{A general and efficient representation of ancestral recombination graphs}
for consideration for publication in GENETICS
as a ``Theoretical Population and Evolutionary Genetics''
Investigations article.

\closing{Sincerely,}

\end{letter}
\end{document}

### 2023-11-02 cover-letter/cover-letter.tex

@@
 
-\address{Big Data Institute\\University of Oxford\\UK}
+\address{Big Data Institute\\University of Oxford}
 \begin{document}
@@
 \emph{A general and efficient representation of ancestral recombination graphs}
-for consideration for publication in GENETICS
-as a ``Theoretical Population and Evolutionary Genetics''
-Investigations article.
+for consideration for publication as a
+full-length article in the
+``Theoretical Population and Evolutionary Genetics'' section.
+
+Ancestral recombination graphs (ARGs) are currently the subject of
+intense research interest, with new inference methods, evaluations,
+and applications regularly appearing.
+In their recent preprint\footnote{The era of the ARG:
+an empiricist’s guide to ancestral recombination graphs, \emph{arXiv}, 2023}
+ (a biologically oriented primer
+on ARGs), Lewanski et al.\ have hailed the ``Era of the ARG'', and
+argue eloquently for their widespread application.
+
+Despite this general optimism, however, we believe there is a
+a fundamental problem in the area, which will substantially hamper
+progress if not addressed.
+The crux of the matter is that it is no longer clear
+what an ARG \emph{is}, precisely.
+While the classical Griffiths definition that
+an ARG is the set of coalescence and recombination events
+in the history of a sample is straightforward,
+it is not sufficient to describe the output of methods
+such as \texttt{msprime}, \texttt{tsinfer}, \texttt{Relate}
+or \texttt{ARGneedle} (all of which are now routinely
+referred to as ``ARG'' methods). This mismatch
+between formal definition, informal usage and the
+actual output of current methods is the source of deep
+and persistent confusion.
+
+We show that the classical event-based approach is
+fundamentally limited, and propose a simple new
+formal framework which we call the ``gARG''
+encoding. We demonstrate that it generalises
+the classical definitions, is fully compatible
+with the output of the methods listed above,
+% clarifies inference goals,
+and opens up many avenues for future research.
+
+As this this manuscript addresses a question
+that is central to current research trends,
+we believe that it will be of substantial interest
+to the readers of GENETICS. We hope that you agree.
 

### 2023-11-02 cover-letter/cover-letter.tex

@@
 
-As this this manuscript addresses a question
+As this manuscript addresses a question
 that is central to current research trends,

### 2023-11-02 paper.tex
tweak

@@
 left of $x$ from \noderef{e} and to the right of $x$ from \noderef{f}.
+% Some eARG implementations incorporate additional information to
+% distinguish left and right parents into the
+% encoding~\citep[e.g.][]{gusfield2014recombinatorics,ignatieva2021kwarg}.
 
@@
 
-In eukaryotes, ARGs are a result of the cellular processes of mitosis
-(which leads to common ancestry and coalescence)
-and meiosis (which leads to recombination, including both crossovers and gene conversion).
-Fig.~\ref{fig-simplification} shows a schematic of the events (coloured red and blue squares) and
-the genomes (chromosome icons) that occur in the cellular germline of a simplified, diploid multicellular
+In eukaryotes, ARGs are a result of the cellular processes of mitosis and meiosis.
+Mitosis leads to common ancestor events,
+and meiosis leads to recombination events (both crossover and gene conversion).
+Fig.~\ref{fig-simplification} shows a schematic of the events  and
+the genomes (chromosome icons) that occur in the cellular
+germline of a simplified, diploid multicellular
 hermaphrodite eukaryote with partially overlapping generations.
-% It may be worth putting in something here about events being a slippery concept
-% (we have already mentioned this at the end of Section 3
 Here, an event is not represented by a specific
-genome. Rather, genomes can be associated with (or ``tag'') events above or below them (and as argued in
-Section~\ref{sec-precision}, the gARG representation allows these associated genomes to be arbitrarily distant
-in time or number of cell
-divisions from the event itself). For example, tagging the two genomes above a recombination event leads to the
-2 node representation seen in Figs.~\ref{fig-simplification} and \ref{hudson_vs_bigARG}, whereas tagging the genome
-below a recombination event leads to the more conventional graphs in Figs.~\ref{fig-ancestry-resolution} and \ref{fig-inferred-args}A,B.
-
-Thick black lines trace an ARG that describes the ancestry of the lower arm of the
-four sampled chromosomes, and these lines pass through gARG nodes (green chromosomal regions).
-A number of these nodes have only one parent and one child in the ARG; these would often be omitted
-(taken as implicit) in a minimally simplified gARG.
-
-In particular, this schematic illustrates a case where three ARG lineages coalesce within
-a single individual ($D_{10}$), corresponding to what is normally represented by a polytomy
-(e.g. Fig.~\ref{fig-simplification} node \noderef{k}). The process that generates
-these three lineages actually consists of two successive bifurcations, reflecting the fact that
-the only known method of reproducing DNA is by (semi-conservative) duplication. Under this schematic,
-polytomies therefore represent a gARG-enabled merging of multiple coalescent events into an older
-node, just as ARG nodes with more than 2 parents represent a merging of multiple recombination
-events into a younger node.
-
-Note that mutations can occur along any lineage in Fig.~\ref{fig-cell-lines}. This means, for example,
-that a mutation in the first cell division of $D_{10}$ could be shared between
-the two gametes produced by the cells in the left half of $D_{10}$ but not shared by
-the right hand gamete. With a high enough mutation rate (or equivalently, a long enough genome), each
+genome. Rather, genomes can be associated with, or ``tag'', events above
+(ancestral to) or below (descended from) them.
+For example, tagging the two genomes above a recombination event leads to the
+two node representation seen in
+Figs.~\ref{fig-simplification} and \ref{hudson_vs_bigARG},
+whereas tagging the genome
+below a recombination event leads to the more
+conventional graphs in Figs.~\ref{fig-ancestry-resolution} and \ref{fig-inferred-args}A,B.
+
+%% JK:  What does this add that is not in caption?
+% Thick black lines trace an ARG that describes the ancestry of the lower arm of
+% the four sampled chromosomes, and these lines pass through gARG nodes (green
+% chromosomal regions). A number of these nodes have only one parent and one
+% child in the ARG; these would often be omitted (taken as implicit) in a
+% minimally simplified gARG.
+
+The schematic illustrates an important point about the biological reality of polytomies.
+Three linages coalesce in the left-hand genome of individual $D_{10}$, but do
+so as the result of two successive bifurcations. This is \emph{necessarily} so,
+because the only known method of reproducing DNA is by (semi-conservative) duplication.
+Whether this polytomy is resolvable depends on the available mutational data.
+Mutations can occur along any cell lineages. For example,
+a mutation in the first cell division of $D_{10}$ could be shared between
+the two gametes produced by the cells in the left half of $D_{10}$
+but not shared by the right hand gamete.
+With enough mutations, each
 round of mitotic germline genome duplication within a single multicellular organism could
-potentially be distinguished.
-This implies that in ARGs that have been inferred from real data, polytomies represent a lack
-of mutational information (rather than marking a single ``multifurcation event'').
-The ability to represent polytomies, and conceptualise them as a (deliberately) imprecise statement about the order
-of coalescence within and between individuals,
-fits naturally into the gARG framework presented in this paper.
+in principle be distinguished.
+
+% Note that mutations can occur along any lineage in Fig.~\ref{fig-cell-lines}. This means, for example,
+% that a mutation in the first cell division of $D_{10}$ could be shared between
+% the two gametes produced by the cells in the left half of $D_{10}$ but not shared by
+% the right hand gamete. With a high enough mutation rate (or equivalently, a long enough genome), each
+% round of mitotic germline genome duplication within a single multicellular organism could
+% potentially be distinguished.
+% This implies that in ARGs that have been inferred from real data, polytomies represent a lack
+% of mutational information (rather than marking a single ``multifurcation event'').
+% The ability to represent polytomies, and conceptualise them as a (deliberately) imprecise statement about the order
+% of coalescence within and between individuals,
+% fits naturally into the gARG framework presented in this paper.
 

### 2023-11-02 paper.tex
Updates from Overleaf

--- 2023-11-02 paper.tex
@@
 
-\bibliographystyle{plainnat}
+\bibliographystyle{refstyle}
 \bibliography{paper}
@@
 Ancestral recombination events and coalescence events are shown as red and blue squares respectively.
-A realised ARG path for the lower arm of the sampled chromosomes (green) is highlighted as a
-thick solid line and involves a single recombination event and four coalescence events
+A realised ARG path for the lower arm of the sampled chromosomes is highlighted as a
+thick solid line, passing through a set of potential gARG nodes (green). This ARG
+involves a single recombination event and four coalescence events
 (highlighted as deep red and blue squares within individuals $D_5$, $D_{10}$, and $D_{13}$).

### 2023-11-02 paper.tex
Tweaks

@@
 \usepackage[symbol]{footmisc}
-% JK: turning this off for the moment as I keep clicking through on links
-% to the bibliography while reading the text and it's intensely annoying.
-% Can reinstate when we're ready to preprint
-% \usepackage[hidelinks]{hyperref}
+\usepackage[hidelinks]{hyperref}
 
@@
 it is the \emph{implementations} that determines how well methods
-interoperate. The SAM/BAM file format~\citep{li2009sequence}
+interoperate. The BAM file format~\citep{li2009sequence}
 is an instructive example. Originally developed as part of the
@@
 software ecosystem that existed at the time~\citep{danecek2021twelve},
-SAM/BAM has since become ubiquitous in bioinformatics pipelines.
+BAM has since become ubiquitous in bioinformatics pipelines.
 The excellent interoperability between methods
@@
 success of \texttt{htslib}~\citep{bonfield2021htslib}, the
-software library that \emph{implements} SAM/BAM and several other
+software library that \emph{implements} BAM and several other
 foundational bioinformatics file formats.
-Today, there are dozens  % hundreds? Could check on github I guess?
-of tools that use \texttt{htslib} to
-support SAM/BAM and it this shared use of community software
+Today, there are thousands of software projects
+using \texttt{htslib}~\citep{bonfield2021htslib},
+and it this shared use of community software
 infrastructure that guarantees the smooth flow of data between

### 2023-11-03 cover-letter/cover-letter.tex
Typos and tweaks.

@@
 Despite this general optimism, however, we believe there is a
-a fundamental problem in the area, which will substantially hamper
+fundamental problem in the area, which will substantially hamper
 progress if not addressed.

### 2023-11-03 paper.tex
Typos and tweaks.

@@
 % geneticists.
-a sample of sequences vary along the genome.
+a sample of DNA sequences vary along the genome.
 The structure capturing the details of these intricately interwoven paths of
@@
 are insufficient to describe the outputs of current methods.
-We argue, moreover, that the precise estimates of
+We argue that the precise estimates of
 \emph{all} events required
@@
 specific genomes and their intervals of genetic inheritance,
-we avoid the necessity of exhaustively listing (and estimating)
+we avoid the need to exhaustively list (and estimate)
 \emph{all} events. The effects of multiple events can be
@@
 Broadly speaking, an ARG describes the different paths of genetic inheritance
-caused by recombination, and encodes a sequence of correlated genealogical
-trees along the genome
+caused by recombination, encapsulating the resulting complex web of genetic ancestry
 (see \citet{lewanski2023era} for a biologically oriented introduction).
@@
 the true ARG of every individual in a population.
-True ARGs are the actual history,
+True ARGs are the underlying, real history,
 perfectly resolved into binary splits and mergers by
@@
 (Appendix~\ref{sec-cell-lineages-and-args}).
-Although population ARGs unquestionably exist,
+Although true population ARGs unquestionably exist,
 they can also never be entirely known, in part because of a fundamental
@@
 chain of cell divisions~\ref{sec-cell-lineages-and-args},
-or can be interpreted as representing one of the genomes of an
-individual; how we assign dates  to nodes
+or can be interpreted as representing one of the genomes of a
+potentially long-lived individual; how we assign precise times to nodes
 will therefore vary according to application.
@@
 As we proceed backwards in time, generation by generation, the
-extant lineages chose parents randomly.
+extant lineages choose parents randomly.
 With a certain probability recombination occurs, and the ancestral
@@
 simplify (Fig.~\ref{fig-simplification}B,C),
-until we reach Fig.~\ref{fig-simplification}D, where are none.
+until we reach Fig.~\ref{fig-simplification}D, where there are none.
 A more detailed version of the displayed ARGs, showing the edge
@@
 such as \relate~\citep{speidel2019method} and
-\tsinfer~\citep{kelleher2019inferring} have
+\tsinfer~\citep{kelleher2019inferring} 
 have been, in part, due to inferring lower levels of precision about
@@
 (BBS/E/D/30002275, BBS/E/RL/230001A, and BBS/E/RL/230001C).
-% Anastasia Ignatieva acknowledges support from TODO
 Jerome Kelleher acknowledges support from the Robertson Foundation,
@@
 Jere Koskela acknowledges support from EPSRC research grant EP/V049208/1.
-% Anthony Wilder Wohns acknowledges support from TODO
 
@@
 to reason about and infer recombinant ancestry as a single object,
-in a way that not possible within Hudson's framework, which emphasised
+in a way that is not possible within Hudson's framework, which emphasised
 instead the collection of local trees along the genome

### 2023-11-03 paper.tex
Update intro and abstract

@@
 are insufficient to describe the outputs of current methods.
-We argue that the precise estimates of
-\emph{all} events required
-by such formalisms are fundamentally unsuited to the realities
+Moreover, we argue that the underlying assumption
+that all events can be known and precisely estimated
+is fundamentally unsuited to the realities
 of modern, population-scale datasets.
@@
 Broadly speaking, an ARG describes the different paths of genetic inheritance
-caused by recombination, encapsulating the resulting complex web of genetic ancestry
+caused by recombination, encapsulating the resulting complex web of genetic
+ancestry
 (see \citet{lewanski2023era} for a biologically oriented introduction).
@@
 the true ARG of every individual in a population.
-True ARGs are the underlying, real history,
+True ARGs are the underlying real history of a sample,
 perfectly resolved into binary splits and mergers by
@@
 (Appendix~\ref{sec-cell-lineages-and-args}).
-Although true population ARGs unquestionably exist,
+Although population-scale true ARGs unquestionably exist,
 they can also never be entirely known, in part because of a fundamental
@@
 instead of a complete and fully detailed history.
-% Not sure the point is coming across that well here, but it's a
-% really important point and a significant
-It is important to distinguish here between
+% This is a digression from the maing point of the paragraph, so
+% making it a parenthetical aside.
+(Note that it is important to distinguish here between
 structures and models: whether an inference method is based
@@
 approximate ARG under a model
-such as the coalescent. % with recomb is surely unnecessary here
+such as the coalescent.)
 Although the term ``ARG'' is now often used in a general
@@
 substantial generalisation of the classical eARG approach,  providing much more
-flexibility in how genetic inheritance can be represented, and encompassing the
+flexibility in how genetic inheritance can be represented, and encompasses the
 outputs of modern methods. We show that the gARG approach can represent many
@@
 in the context of current datasets and research questions.
-Following this, we discuss the important concept of ``ancestral material''
+Following this, we discuss the important concept of ancestral material
 in Section~\ref{sec-ancestral-material}, and how it relates

### 2023-11-03 paper.tex
Clarify what nodes are

@@
 
+How nodes are interpreted, exactly, is application dependent.
+Following \citet{hudson1983properties}, we can view nodes
+as representing gametes, or we can imagine them representing
+the genomes present in cells immediately before or after
+some instantaneous event (Appendix~\ref{sec-cell-lineages-and-args}).
+A node can represent any genome along a chain of cell divisions
+or can be interpreted as representing one of the genomes of a
+potentially long-lived individual. All these interpretations
+are potentially useful, and equally valid under the assumptions
+of the gARG encoding.
 In many settings, nodes are dated, i.e.\ each
-node $u\in N$ is associated with a time $\tau_u$.
-A node can represent any genome along a
-% A gamete *is* one of the genomes along a chain of cell divisions?
-% I don't see what adding a gamete clarifies, and we cite Hudson plenty
-chain of cell divisions~\ref{sec-cell-lineages-and-args},
-or can be interpreted as representing one of the genomes of a
-potentially long-lived individual; how we assign precise times to nodes
-will therefore vary according to application.
+node $u\in N$ is associated with a time $\tau_u$,
+and how we assign precise times will vary by application.
 The topological ordering defined by the directed graph structure

### 2023-11-03 paper.tex
Clarify eARG amibiguity

@@
 ancestral lineages. Recombination nodes are annotated with
-the corresponding crossover breakpoints. These breakpoints
+the corresponding crossover breakpoints, and these breakpoints
 are used to construct the local trees.
@@
 at position $x$.
-We assume here that \noderef{d} inherits genetic material to the
+Assigning a breakpoint to a recombination node is
+not sufficient to uniquely define the local trees, and either
+some additional ordering rules~\citep[e.g.][]{griffiths1996ancestral} or
+explicit information~\citep[e.g.][]{gusfield2014recombinatorics,ignatieva2021kwarg}
+is required to distinguish the left and right parents.
+We assume in Fig.~\ref{fig-event-arg} that \noderef{d} inherits genetic material to the
 left of $x$ from \noderef{e} and to the right of $x$ from \noderef{f}.
-% Some eARG implementations incorporate additional information to
-% distinguish left and right parents into the
-% encoding~\citep[e.g.][]{gusfield2014recombinatorics,ignatieva2021kwarg}.
 

### 2023-11-03 paper.tex
Updates from Overleaf

--- 2023-11-03 paper.tex
@@
 We then discuss how the gARG framework can be
-efficiently implementated in Section~\ref{sec-efficiency},
+efficiently implemented in Section~\ref{sec-efficiency},
 and finish with an assessment of the key challenges facing
@@
 % YW: The above "combining" operation isn't obviously lossless to me:
-% we could potentially lose information in the different
-% combinations of intervals (e.g. if we use 2 non-combined intervals
-% to represent a diamond in a 1-RE-node ARG). It seems like we might
-% need an additional gARG spec to define what combinations are equivalent.
+% we could potentially lose information in the different combinations
+% of intervals (e.g. if we use 2 non-combined intervals to represent a 
+% diamond in a 1-RE-node ARG). It seems like we might need an additional
+% gARG spec to define what combinations of intervals are equivalent.
 % But no-one else has commented on this, so I guess we just let it lie.
@@
 % into a set of simple SPR moves.
-% Overall, Fig.~\ref{fig-inferred-args} shows
-% that \tsinfer\ and \relate\ ARGs contain a level of detail about recombination
-% that lies somewhere between a sequence of unrelated local trees on one extreme
-% and an ARG with precisely specified recombination events on the other.
 

### 2023-11-03 paper.tex
Minor fixes througout

@@
 (see Section~\ref{sec-ARG-simplification} for
-other applications).
+other applications of ARG simplification).
 % Define "sample-resolved" here for later use
@@
 \espalier\ approaches this task
-by identifying maximimal subtrees
+by identifying maximal subtrees
 that do not change between pairs of adjacent
@@
 rearrangements of these subtrees.
-To derive precise details about recombination events,
+To derive details about recombination events,
 \espalier\ then attempts to infer
@@
 through a series of simplification steps
-in Fig.~\ref{fig-simplification};
-the gARG representation is shown on the left and the corresponding sequence
-of local trees on the right.
+in Fig.~\ref{fig-simplification}.
 
@@
 the parent's two genomes.
-Thus, a recombinant node such as node \noderef{g} in Fig.~\ref{fig-simplification}
+Thus, a recombinant node such as \noderef{g} in Fig.~\ref{fig-simplification}
 may also correspond to a coalescence.
@@
 until we reach Fig.~\ref{fig-simplification}D, where there are none.
-A more detailed version of the displayed ARGs, showing the edge
-intervals, is provided as Supplementary Fig.~\ref{fig-simplification-with-edges}.
 
@@
 \citep{schrider2013rates,comeron2012many}
-and a constant effective population size of $1720600$ \citep{li2006inferring}, all
-values being taken from the \texttt{stdpopsim} library
-\citep{adrion2020community,lauterbur2023expanding}.
+and a constant effective population size of 1,720,600 \citep{li2006inferring},
+as provided by the \texttt{stdpopsim}
+catalog~\citep{adrion2020community,lauterbur2023expanding}.
 Software versions were \kwarg~v1.0, \argweaverD~(2019),
@@
 In Fig.~\ref{fig-inferred-args}A,B, each recombination event
-corresponds to a node with exactly 2 parents and one child.
+corresponds to a node with exactly two parents and one child.
 As these methods explicitly infer a recombination event for each
-breakpoint, the number of breakpoints equals the number of 2-parent
+breakpoint, the number of breakpoints equals the number of two-parent
 (brown) nodes.
@@
 more recent node.
-% Several of these nodes have more than two
-% % parents (in red), representing multiple recombinations inferred to have
-% occurred in an unresolvable order.
-% Nevertheless, the number of breakpoints in
-% these two ARGs is smaller than the number of multiple-parent ARG nodes, showing
-% that several multiple-parent nodes must share breakpoint positions.
 As a consequence, it may be difficult to condense each transition between trees
 in these ARGs into a set of SPR operations.
-% The sharing
-% of breakpoint locations between these nodes indicates that in \tsinfer\ and
-% \relate\ ARGs it may be difficult to condense each transition between trees
-% into a set of simple SPR moves.
 
@@
 Fig.~\ref{fig-inferred-args}A,B and C demonstrates that
-\kwarg, \argweaver\ and \tsinfer\ all contain locally unary nodes.
+the \kwarg, \argweaver\ and \tsinfer\ ARGs all contain locally unary nodes.
 Another difference between methods highlighted in this figure
@@
 simulations~\citep{kelleher2018efficient,haller2018tree},
-inference from data~\citep{kelleher2019inferring,wohns2022unified},
+inference from
+data~\citep{kelleher2019inferring,wohns2022unified,zhan2023towards},
 and calculation of population genetics statistics~\citep{ralph2020efficiently}.
@@
 Tskit is mature software, widely used in population genetics, and
-has been incorporated into several downstream
+has been incorporated into numerous downstream
 applications~\citep[e.g.,][]{haller2019slim,speidel2019method,
@@
 \texttt{tskit}.
-Firstly, ``edges'' in tree sequence terminology would be perhaps better be
+Firstly, ``edges'' in tree sequence terminology would perhaps be better
 described as ``edge-intervals'', as each describes a single contiguous
@@
 making a single, deterministic best-guess at ARG
-inference~\citep{kelleher2019inferring,speidel2019method,zhan2023towards}.
+inference~\citep{kelleher2019inferring,speidel2019method,zhang2023biobank,zhan2023towards}.
 Even under strict parsimony conditions and for small sample sizes, the
@@
 nodes to have more than two children (polytomies
-representing uncertainty over the ordering of coalescence events) or
+representing uncertainty over the ordering of coalescence events, Appendix~\ref{sec-cell-lineages-and-args}) or
 more than two parents (representing uncertainty over the ordering
-of multiple recombination events). Development of other methods to capture, for example,
+of multiple recombination events, Section~\ref{sec-ARG-simplification}). Development of other methods to capture, for example,
 uncertainty about node ages and recombination breakpoint positions, is an important
@@
 and comparing this to the inferred version via pairwise comparison of
-local trees along the genome, for example
-using tree distance
+local trees along the genome using tree distance
 metrics~\citep[e.g.][]{robinson1981comparison,kendall2016mapping},
@@
 The performance of tree distance metrics varies by
-application~\citep{kuhner2015assessing}, and the correct approach
+application~\citep{kuhner2015practical}, and the correct approach
 to handling subtleties such as polytomies
@@
 % https://bmcgenomics.biomedcentral.com/articles/10.1186/s12864-020-07011-0
-Tree distance metrics often have at least $O(n^2)$ time complexity
+Tree distance metrics often have $O(n^2)$ time complexity or worse
 and therefore cannot be
@@
 without the productivity-sapping problems
-of converting between (partially incompatible)
-program output formats~\citep{excoffier2006computer}.
+of converting between partially incompatible
+formats~\citep{excoffier2006computer}.
 Earlier efforts to standardise ARG interchange shared this vision,
@@
 and no matter how precise the specification, in practise
-it is the \emph{implementations} that determines how well methods
-interoperate. The BAM file format~\citep{li2009sequence}
+it is the \emph{implementations} that determine how well methods
+interoperate. The BAM read alignment format~\citep{li2009sequence}
 is an instructive example. Originally developed as part of the
@@
 \section*{Acknowledgements}
-We are grateful to Nick Barton, Gideon Bradburd, Alex Lewanski, Andrew Vaughan
-and Peter Ralph for helpful discussions and comments on the manuscript, and
+We are grateful to Nick Barton, Gideon Bradburd, Alex Lewanski,
+Peter Ralph and Andrew Vaughan
+for helpful discussions and comments on the manuscript, and
 to Iain Mathieson and Aylwyn Scally for interesting discussions that

### 2023-11-03 paper.tex
Minor
Yans changes

--- 2023-11-03 paper.tex
@@
 more recent node.
-% Several of these nodes have more than two
-% % parents (in red), representing multiple recombinations inferred to have
-% occurred in an unresolvable order.
-% Nevertheless, the number of breakpoints in
-% these two ARGs is smaller than the number of multiple-parent ARG nodes, showing
-% that several multiple-parent nodes must share breakpoint positions.
 As a consequence, it may be difficult to condense each transition between trees

### 2023-11-03 paper.tex
Fix

--- 2023-11-03 paper.tex
@@
 in Fig.~\ref{fig-simplification}A).
-% FIXME this needs some work
 Secondly, node colours denote the number of parents that they
@@
 aspect of future work.
-% Obvious, but nobody is doing it, so needs to be said.
-% TODO rephrase to say this is an open question.
-Fully utilising such uncertainty in downstream
-applications is also vital.
+How this uncertainty can be utilised in downstream applications
+is an open question.
 
@@
 that is already in widespread use.
-% TODO needs rephrasing
 Adopting it as a community standard may ease software implementation burden
@@
 We are grateful to Nick Barton, Gideon Bradburd, Alex Lewanski,
-Peter Ralph and Andrew Vaughan
+Peter Ralph and Andrew Vaughn
 for helpful discussions and comments on the manuscript, and

### 2023-11-03 paper.tex
Tweak

@@
 Following \citet{hudson1983properties}, we can view nodes
-as representing gametes, or we can imagine them representing
-the genomes present in cells immediately before or after
+as representing gametes, or we can imagine them representing,
+for example, the genomes present in cells immediately before or after
 some instantaneous event (Appendix~\ref{sec-cell-lineages-and-args}).

### 2023-11-03 paper.tex

@@
 forms of approximate and partial knowledge about the
-recombinant ancestry of a sample.
+recombinant ancestry of a sample of genomes.
 \end{abstract}
@@
 cannot adequately describe these vast inferred ARGs.
-Fundamentally, these formalisms assume that an
-ARG is known in complete detail
-and are not suited to describing partial or approximate knowledge.
+Fundamentally these formalisms assume that an
+ARG is known in complete detail: they
+are not suited to describing partial or approximate knowledge.
 As we are actively inferring ARGs at the population scale,
-and such ARGs can never be known in complete detail,
-there is currently a substantial gap between our theoretical frameworks
+and such ARGs can never be fully known,
+there is currently a substantial gap between classical representations
 and practical application.
@@
 instead of a complete and fully detailed history.
-% This is a digression from the maing point of the paragraph, so
+% This is a digression from the main point of the paragraph, so
 % making it a parenthetical aside.
@@
 that occurred at node \noderef{d}. Given these two local trees
-in isolation
+in isolation,
 % (without further information about the corresponding ARG)
@@
 inferred by different methods, with e.g.~\argweaver\ suggesting
-far more than the 7 required for this dataset under minimal parsimony
+far more than the seven required for this dataset under minimal parsimony
 assumptions~\citep{song2003parsimonious}.
@@
 within the context of stochastic
-processes and instead use the term as a more general representation of any
+processes, and instead use the term as a more general representation of any
 recombinant genetic ancestry seems useful; we have
@@
 for many species % Can we say something more concrete here?
-and projects are under way to obtain
+with projects under way to obtain
 them for tens of thousands more~\citep{darwin2022sequence,lewin2022earth},
@@
 Thus, although it is clearly an oversimplification to arbitrarily
-choose \emph{one} best guess,
-it is not clear that generating many guesses
-when sample sizes are large will achieve much.
+choose \emph{one} best guess, if sample sizes are large
+it is not clear that generating many guesses will achieve much.
 At the scale of millions of samples,
@@
 Therefore, it is important to systematically
-describe and utilise uncertainty about ARG inference, and to
+describe and use uncertainty about ARG inference, and to
 incorporate uncertainty encountered during inference into the returned ARG.
@@
 aspect of future work.
-How this uncertainty can be utilised in downstream applications
+How this uncertainty can be used in downstream applications
 is an open question.
@@
 The schematic illustrates an important point about the biological reality of polytomies.
-Three linages coalesce in the left-hand genome of individual $D_{10}$, but do
+Three lineages coalesce in the left-hand genome of individual $D_{10}$, but do
 so as the result of two successive bifurcations. This is \emph{necessarily} so,

### 2024-03-20 paper.tex
Edits to Section 3 (eARGs)

--- 2023-11-03 paper.tex
@@
 \label{sec-eARG}
-% Trying to be totally, boringly, abundantly, blatantly clear here what we're
-% talking about.
-In this section we define the classical view of an ARG data structure,
-and illustrate its limitations. We are
-interested in the details of how ARGs are described mathematically,
-and as a consequence, how they are represented in a practical sense
-as the output of inference programs.
-Where details of an ARG data structure (the encoding) are
-% Three papers 10 years apart will do here, right?
-provided~\citep[e.g.][]{
+In this section we define a common (and classical) view of an ARG data structure,
+described explicitly in several publications~\citep[e.g.][]{
 wiuf1999recombination,gusfield2014recombinatorics,hayman2023recoverability}
-they follow the approach described by Griffiths and colleagues
 (but see~\citet{parida2011minimal} and \citet{zhang2023biobank}
-for notable exceptions), and a large number of
-% Note the "an" here to avoid cases like ARGweaver
-ARG inference methods use it as an output format
+for notable exceptions), and used as an output format by multiple ARG inferences tools
 \citep[e.g.][]{song2004minimum,song2005efficient,rasmussen2014genome,
@@
 % down.
-In this Griffiths encoding we have two types of internal node in the graph,
+In this encoding we have two types of internal node in the graph,
 representing the common ancestor and recombination events
@@
 Because it is focused on recording events and their properties,
-we will refer to this Griffiths encoding as the ``event ARG'' or eARG
-encoding. Fig.~\ref{fig-event-arg} shows an example of a classical
+we will refer to this encoding as the ``event ARG'' or eARG
+encoding. Fig.~\ref{fig-event-arg} shows an example of an
 eARG with three sample genomes (\noderef{a}, \noderef{b}, and \noderef{c}),
@@
 % eARGs are representationally for limited interchange
-While the Griffiths approach of annotating recombination nodes with a
+While this approach of annotating recombination nodes with a
 breakpoint in an eARG is a concise and elegant way of describing realisations
@@
 recovering the local trees) or by two recombination nodes joined by a
-zero-length edge. From the perspective of practical interchange of data between
-inference methods and downstream applications, both
-workarounds are problematic, and the gARG encoding described in the previous
-section offers a much simpler solution.
-
-Aside from these obvious practical challenges arising from a
-literal implementation of the Griffiths approach, there is a deeper
+zero-length edge. The gARG encoding described in the previous
+section offers a simpler and more direct solution.
+
+Aside from these practical challenges, there is also a deeper
 issue with the implicit strategy of basing an ARG data structure on
 recording events and their properties (e.g.\ the crossover breakpoint
-for a recombination event). The fundamental problem
-is that this approach assumes all events are \emph{knowable}, and does not
-provide any obvious mechanism for either aggregating multiple events
+for a recombination event). This approach assumes all events are 
+\emph{knowable}, and does not
+provide an obvious mechanism for either aggregating multiple events
 or expressing uncertainty about them. While this is not a
 problem when describing the results of simulations (where all details
-are perfectly known), it is a major issue when we wish to
+are perfectly known), it is an issue when we wish to
 formally describe the output of inference methods, particularly
@@
 the precise details of all events in these vast ARGs can never be
-known, and a data structure that \emph{enforces} complete
-precision is therefore an impediment to progress.
+known, so it is disadvantageous to use a data structure that enforces complete
+precision.
 
@@
 % discussions won't help.
-From the perspective of a concrete data structure,
-ideally forming the basis of an ecosystem of interoperable
-inference and analysis methods, such debates
-are unproductive.
+The gARG format avoids the need for such considerations,
+so is more naturally compatible with forming an ecosystem of interoperable
+inference and analysis methods.
 

### 2024-03-21 paper.tex
Structural updates addressing comments

--- 2023-11-03 paper.tex
@@
 enabling many new applications in population and statistical genetics.
-This rapid progress, however, has led to a
-% Controversial, perhaps, but true?
-substantial gap opening between theory and practice.
-Standard mathematical formalisms,
+This rapid progress has led to a diversity of ARG definitions and representations.
+
+We show that a synthesis of these views can be achieved by thinking
+in terms of genomes and transmission
+intervals.
+% Standard mathematical formalisms,
 % "detailing" here is intended to convey the we keep details about these
 % events like recomb breakpoint.
-based on exhaustively detailing the ``events'' that occur
-in the history of a sample,
-are insufficient to describe the outputs of current methods.
+% based on exhaustively detailing the ``events'' that occur
+% in the history of a sample,
+%are insufficient to describe the outputs of current methods.
 Moreover, we argue that the underlying assumption
@@
 differently (Appendix~\ref{sec-arg-history}).
-Other subtly different concepts that we must be careful to distinguish
-are the ``true'' ARG, % I don't see a better word for "true" here
-describing the actual history of a sample of genomes,
-and a ``population'' ARG which is
-the true ARG of every individual in a population.
-True ARGs are the underlying real history of a sample,
-perfectly resolved into binary splits and mergers by
-the cellular processes of meiosis and mitosis, regardless
-of sampling density or population processes
-(Appendix~\ref{sec-cell-lineages-and-args}).
-Although population-scale true ARGs unquestionably exist,
-they can also never be entirely known, in part because of a fundamental
-lack of mutational information.
-Even if mutation rate were high enough to uniquely identify every
-recent branching point,
-such a high rate would saturate the genome with mutations
-and obscure deeper history.
+
+%Other subtly different concepts that we must be careful to distinguish
+%are the ``true'' ARG, % I don't see a better word for "true" here
+%describing the actual history of a sample of genomes,
+%and a ``population'' ARG which is
+%the true ARG of every individual in a population.
+%True ARGs are the underlying real history of a sample,
+%perfectly resolved into binary splits and mergers by
+%%the cellular processes of meiosis and mitosis, regardless
+%of sampling density or population processes
+%(Appendix~\ref{sec-cell-lineages-and-args}).
+%Although population-scale true ARGs unquestionably exist,
+%they can also never be entirely known, in part because of a fundamental
+%lack of mutational information.
+%Even if mutation rate were high enough to uniquely identify every
+%recent branching point,
+%such a high rate would saturate the genome with mutations
+%and obscure deeper history.
 
 % Population ARGs are here, now.
-Population ARGs may seem fanciful, but the scale of
-modern datasets makes it necessary for us to grapple with the idea.
-The UK Biobank (UKB), for example, has
-genotype data for around 500,000 humans~\citep{bycroft2018genome},
-along with exome~\citep{backman2021exome} and
-whole genome sequence~\citep{halldorsson2022sequences} data
-for large subsets of the cohort.
-UKB is just one of many such population-scale sequencing
-projects~\citep[e.g.][]{turnbull2018hundred,
-karczewski2020mutational,tanjo2021practical}.
-Agricultural datasets are on a similar scale,
-and also include dense multi-generational sampling
-and near-perfect pedigree
-information~\citep[e.g.][]{hayes20191000,Ros-Freixedes2020}.
-Recent advances have made it possible to actually
-\emph{estimate} ARGs at population scale:
-ARGs have already been inferred for the 500,000
-humans in UKB~\citep{kelleher2019inferring,zhang2023biobank}
-and over a million SARS-CoV-2 genomes~\citep{zhan2023towards}.
-While this new population-scale reality presents many
-exciting opportunities, it also poses substantial challenges
-to existing methodologies.
+%Population ARGs may seem fanciful, but the scale of
+%modern datasets makes it necessary for us to grapple with the idea.
+%The UK Biobank (UKB), for example, has
+%genotype data for around 500,000 humans~\citep{bycroft2018genome},
+%along with exome~\citep{backman2021exome} and
+%whole genome sequence~\citep{halldorsson2022sequences} data
+%for large subsets of the cohort.
+%UKB is just one of many such population-scale sequencing
+%projects~\citep[e.g.][]{turnbull2018hundred,
+%karczewski2020mutational,tanjo2021practical}.
+%Agricultural datasets are on a similar scale,
+%and also include dense multi-generational sampling
+%and near-perfect pedigree
+%information~\citep[e.g.][]{hayes20191000,Ros-Freixedes2020}.
+%Recent advances have made it possible to actually
+%\emph{estimate} ARGs at population scale:
+%ARGs have already been inferred for the 500,000
+%humans in UKB~\citep{kelleher2019inferring,zhang2023biobank}
+%and over a million SARS-CoV-2 genomes~\citep{zhan2023towards}.
+%While this new population-scale reality presents many
+%exciting opportunities, it also poses substantial challenges
+%to existing methodologies.
 
 % What's the problem?
-A major problem currently facing the field is that
-classical mathematical formalisms and terminology
-cannot adequately describe these vast inferred ARGs.
-Fundamentally, these formalisms assume that an
-ARG is known in complete detail
-and are not suited to describing partial or approximate knowledge.
-As we are actively inferring ARGs at the population scale,
-and such ARGs can never be known in complete detail,
-there is currently a substantial gap between our theoretical frameworks
-and practical application.
+%A major problem currently facing the field is that
+%classical mathematical formalisms and terminology
+%cannot adequately describe these vast inferred ARGs.
+%Fundamentally, these formalisms assume that an
+%ARG is known in complete detail
+%and are not suited to describing partial or approximate knowledge.
+%As we are actively inferring ARGs at the population scale,
+%and such ARGs can never be known in complete detail,
+%there is currently a substantial gap between our theoretical frameworks
+%and practical application.
+
 The breakthroughs in scale achieved by recent
@@
 % without section numbers. Let's deal with that when we get there I guess.
-We begin in Section~\ref{sec-gARG} by providing a precise formal definition
-of a gARG, illustrated by an example ARG embedded in pedigree.
-We then provide a similar definition of the classical eARG
-approach in Section~\ref{sec-eARG}, and consider its limitations
-in the context of current datasets and research questions.
-Following this, we discuss the important concept of ancestral material
-in Section~\ref{sec-ancestral-material}, and how it relates
-to the process of converting an eARG to a gARG.
-We continue
-in Section~\ref{sec-ARG-and-local-trees} by considering the relationship
-between an ARG and its local trees. Contrary to the
-prevailing view, we show that a suitably encoded sequence of local trees contains
-precisely as much information as the corresponding ARG.
-The gARG encoding opens a rich new set of details about ARGs,
-including the ideas of locally unary nodes
-(Section~\ref{sec-locally-unary-edges}),
-the levels of detail that can be represented in an ARG
-(Section~\ref{sec-ARG-simplification}),
-and the degrees of precision about recombination
-that can be stored and we may seek to infer (Section~\ref{sec-precision}).
-These ideas have important practical considerations,
-which we illustrate by examining the qualitative properties of
-ARGs inferred by four recent methods for a classical benchmark
-dataset in Section~\ref{sec-example-inferred-args}.
-We then discuss how the gARG framework can be
-efficiently implemented in Section~\ref{sec-efficiency},
-and finish with an assessment of the key challenges facing
-the field in the Discussion.
-Finally, the literature on ARGs is large and
-confusing, and we attempt to clarify some important aspects
-in appendices, including
-a brief history of ancestral graphs (Appendix~\ref{sec-arg-history}),
-a description of the Big and Little ARG stochastic processes
- (Appendix~\ref{sec-big-and-little-arg}),
-a survey of ARG inference methods (Appendix~\ref{sec-survey-arg-infer}),
-and a discussion of ARGs at an individual vs cell
-lineage level (Appendix~\ref{sec-cell-lineages-and-args}).
+% We begin in Section~\ref{sec-gARG} by providing a precise formal definition
+% of a gARG, illustrated by an example ARG embedded in pedigree.
+% We then provide a similar definition of the classical eARG
+% approach in Section~\ref{sec-eARG}, and consider its limitations
+% in the context of current datasets and research questions.
+% Following this, we discuss the important concept of ancestral material
+% in Section~\ref{sec-ancestral-material}, and how it relates
+% to the process of converting an eARG to a gARG.
+% We continue
+% in Section~\ref{sec-ARG-and-local-trees} by considering the relationship
+% between an ARG and its local trees. Contrary to the
+% prevailing view, we show that a suitably encoded sequence of local trees contains
+% precisely as much information as the corresponding ARG.
+% The gARG encoding opens a rich new set of details about ARGs,
+% including the ideas of locally unary nodes
+% (Section~\ref{sec-locally-unary-edges}),
+% the levels of detail that can be represented in an ARG
+% (Section~\ref{sec-ARG-simplification}),
+% and the degrees of precision about recombination
+% that can be stored and we may seek to infer (Section~\ref{sec-precision}).
+% These ideas have important practical considerations,
+% which we illustrate by examining the qualitative properties of
+% ARGs inferred by four recent methods for a classical benchmark
+% dataset in Section~\ref{sec-example-inferred-args}.
+% We then discuss how the gARG framework can be
+% efficiently implemented in Section~\ref{sec-efficiency},
+% and finish with an assessment of the key challenges facing
+% the field in the Discussion.
+% Finally, the literature on ARGs is large and
+% confusing, and we attempt to clarify some important aspects
+% in appendices, including
+% a brief history of ancestral graphs (Appendix~\ref{sec-arg-history}),
+% a description of the Big and Little ARG stochastic processes
+%  (Appendix~\ref{sec-big-and-little-arg}),
+% a survey of ARG inference methods (Appendix~\ref{sec-survey-arg-infer}),
+% and a discussion of ARGs at an individual vs cell
+% lineage level (Appendix~\ref{sec-cell-lineages-and-args}).
 
@@
 
-\section{ARGs and local trees}
-\label{sec-ARG-and-local-trees}
-% (We use ``local'' rather than
-% ``marginal''
-% % could list loads here, but not much point.
-% trees~\citep[e.g.][]{griffiths1996ancestral,minichiello2006mapping,kelleher2016efficient}
-% to avoid the potential confusion with the
-% statistical concept of marginal distributions.)
-The relationship between an ARG and its corresponding
-local trees is subtle and important.
-A fundamental property of genetics is that a
-given DNA nucleotide is inherited from exactly one parent genome,
-both at an organismal and cell-by-cell level
-(Appendix~\ref{sec-cell-lineages-and-args}).
-These paths of single-parent inheritance give rise,
-by definition, to a tree structure.
-As a result of recombination, adjacent nucleotides can have
-different paths of inheritance, and
-an ARG encodes the entire ensemble of local trees along the
-% Said "sample of genomes" originally, but using term "genome" in two
-% different ways in the same sentence then.
-genome for a given set of sample nodes.
-% This is for Yan - to me this is a given, but maybe some people feel
-% the following is overly formal?
-Precisely defining the process by which local trees are extracted
-from an ARG is essential to our understanding of how ARGs and local trees
-are related, and we require a concrete mathematical structure
-to describe the local trees. It is important to note that
-although the following discussion is phrased in terms of the
-gARG encoding,  the arguments apply equally to eARGs
-because any eARG can be converted to a
-% Note, being sample-resolve is not necessary for following argument so omitting
-gARG without loss of information (Section~\ref{sec-ancestral-material}).
-
-Oriented trees provide a
-convenient formalism to capture these parent-child relationships
-in a well-defined combinatorial object.
-% There's no need for a comma
-Let $\pi_1\dots\pi_n$ be a sequence of integers, such that $\pi_u$
-denotes the parent of node $u$, and $\pi_u = 0$ if $u$ is a
-root~\cite[p.\ 461]{knuth11combinatorial}.
-This encoding is particularly useful to describe
-evolutionary trees because parent-child relationships are
-important but the ordering of children at a node is
-not~\citep{kelleher2013coalescent,kelleher2014coalescent,
-kelleher2016efficient}.
-Thus, for a given gARG with nodes $\{1, \dots, n\}$ and
-edges $E$ (Section~\ref{sec-gARG}), we recover the local tree
-at position $x$ as follows.
-We begin by setting $\pi_u = 0$ for each $1 \leq u \leq n$.
-Then, for each sample node in $S$ we trace its path pastwards through the
-ARG for position $x$, and record this path in $\pi$.
-Specifically, at a given node $u$,
-we find an edge $(c, p, I) \in E$ such that $u = c$ and $x \in I$, and set
-$\pi_c \leftarrow p$. We then set $u \leftarrow p$, and repeat
-until either $\pi_u \neq 0$ (indicating we have traversed this section
-of the ARG already on the path from another sample) or there
-is no matching outbound edge (indicating we are at a root).
-Note that the local trees for an ARG are ``sparse''~\citep{kelleher2016efficient},
-because many ancestral nodes will not be reachable from the
-samples at a given position (so their corresponding entries in $\pi$ will be zero).
-
-This combinatorial approach provides at least
-one novel insight, clarifying the fundamental relationship between
-ARGs and local trees.
-Suppose we are given a gARG defined by a set of nodes and edges.
-There is no requirement on the structure of this ARG beyond
-the basic definitions: it could correspond to an ARG in which
-every recombination event is exactly specified
-(e.g.\ Fig.~\ref{fig-ancestry-resolution})
-or one in which local trees are entirely disjoint
-(i.e.\ only the sample nodes are shared between them).
-If we  are given the sequence of local trees for this gARG
-encoded as an oriented tree, along with the genome interval
-covered by each tree, we can recover the original gARG exactly.
-% We do this site-by-site for simplicity, but hopefully it's obvious
-% that this isn't necessary?
-More formally, suppose we are given the local tree $\pi^x_1\dots\pi^x_n$
-for each nucleotide position $1 \leq x \leq L$ on a genome of length $L$.
-Then, the edges of the ``local ARG'' for this tree is given by
-$E^x = \{(u, \pi_u^x, \{x\}) \mid \pi_u^x \neq 0\}$. Because the ARG
-edges are defined by $(c, p, I)$ tuples, where the set $I$ defines
-the positions over which node $c$ inherits from parent $p$, we can
-then simply combine the ``local ARGs'' for each position $x$
-% YW: The above "combining" operation isn't obviously lossless to me:
-% we could potentially lose information in the different combinations
-% of intervals (e.g. if we use 2 non-combined intervals to represent a
-% diamond in a 1-RE-node ARG). It seems like we might need an additional
-% gARG spec to define what combinations of intervals are equivalent.
-% But no-one else has commented on this, so I guess we just let it lie.
-to recover precisely the same set of edges as the original ARG.
-Thus, under this definition, there
-is a one-to-one correspondence between an ARG and
-the sequence of local trees that it encodes.
-
-This is not the prevailing view, however.
-\cite{kuhner2017consensus} argue that the
-``interval-tree'' representation of
-an ARG (the local trees and the genome intervals they cover)
-``does not contain all of the information in the underlying ARG: it lacks the
-number of recombinations occurring at each site, the times at which
-recombinations occurred, and the specific sequences involved as recombination
-partners.''
-\cite{shipilina2023origin} discuss the same ideas,
-and note that the
-``full ARG...~contains more information than the series of tree
-sequences along the genome''.
-These
-% % Could cite some more? E.g. KwARG paper says
-% % "We note that ARGs contain more information than local trees"
-% % Would need a few more examples though, which would require some digging.
-%
-% and similar, \citep[e.g.][]{}
-statements that an ARG
-contains more information than its local trees are true
-if we represent local trees in their conventional forms,
-but these forms discard important information that is
-available in an ARG.
-
-There are two properties of how evolutionary trees are conventionally
-represented that lead to this
-disagreement about the relationship between local trees and an ARG.
-Firstly, the internal nodes of evolutionary trees are usually considered to be
-\emph{unlabelled}, or equivalently, labelled by the leaves which they subtend.
-The same canonical labelling cannot be used for internal ARG nodes because the
-leaves they subtend will typically vary by genomic position. If we do not label
-the tree nodes in a way which is persistent across the sequence of local trees
-in the ARG, we lose the fact that the \emph{same} ancestors sometimes persist
-across multiple trees.
-Defining ARG nodes as integers and using the oriented
-tree encoding explicitly labels internal nodes, and makes the relationship
-between tree and ARG nodes clear and precise.
-
-% Note: good to keep to paras here to logically separate the two important
-% points.
-The second property of how evolutionary trees
-are conventionally represented that is unhelpful in the context of ARGs is their
-focus on branching points (coalescences), i.e.\ nodes that have two or more children.
-As the introductory paragraph of this section emphasised,
-parent-child relationships are what fundamentally define a tree,
-and branching points can be seen as incidental. This is reflected
-by the oriented tree encoding which simply stores the local
-parent-child relationships, and does not, for example,
-directly tell us how many children a particular node has.
-The local tree at a given position records the \emph{path} through
-the ARG; if this path omits nodes that are not
-branching points (such as \noderef{e} in Fig.~\ref{fig-arg-in-pedigree}),
-we lose information about the ARG.
-We return to this point in the following two sections,
-when we discuss ``locally unary'' nodes and the simplification process.
-
-It is important that we make the distinction here between the local
-trees that we can derive from a known ARG (as just discussed),
-and an ARG that we can derive from a sequence of \emph{estimated}
+\section{A diversity of structures}
+
+3 PARAS. - enough outline to let readewr know what's in the appendices.
+
+There are lots of new inference techniques: they all produce slightly different
+structures, all of which can be though of a types of ARGs...
+
+\begin{figure} \begin{center}
+\includegraphics[width=\textwidth]{illustrations/inference.pdf} \end{center}
+\caption{\label{fig-inferred-args} Inference of sample-resolved ARGs for 11
+\textit{Drosophila melanogaster} DNA sequences over a 2.4kb
+region of the ADH locus~\citep{kreitman1983nucleotide}.
+Results for four different methods:
+(A) \kwarg; (B) \argweaver; (C) \tsinfer; and (D) \relate. See the
+text for details of these methods.
+Edge colours indicate time of the edge's child node
+(lighter: older; darker: younger).
+Vertical and horizontal positions of graph nodes are arbitrary.
+Line width and node colour are as described in Fig.~\ref{fig-simplification}.
+Bottom row graphics show the genome positions, relative to the start of
+the ADH gene, for each graph edge from the corresponding ARG. Edge intervals
+are drawn as horizontal lines, stacked in time order (edges with youngest
+children at the bottom); vertical dashed lines denote breakpoints between
 local trees.
-The ARG inference method
-\espalier~\citep{rasmussen2022espalier} is illustrative in this context.
-It begins by splitting an input sequence alignment into
-segments that are assumed to be non-recombining. Within
-each segment, an initial local tree is estimated using
-standard phylogenetic methods. By necessity, these local
-trees will contain internal nodes that are unlabelled and
-consist only of branching points: there is no information
-shared between the independent tree estimation steps
-across segments. Part of the task of stitching
-these trees together
-into an ARG is then, essentially, to generate labels for
-the internal nodes, and decide which nodes persist
-across multiple local trees.
-\espalier\ approaches this task
-by identifying maximal subtrees
-that do not change between pairs of adjacent
-local trees and then
-heuristically exploring the space of possible
-rearrangements of these subtrees.
-To derive details about recombination events,
-\espalier\ then attempts to infer
-the precise subtree prune-and-regraft (SPR)
-operations~\citep{hein1990reconstructing,song2003on,song2006properties}
-induced by recombination between these partially reconciled local trees.
-% NOTE: dropping this because it doesn't clarify the actual point.
-% A similar problem is tackled by \arbores~\citep{heine2018bridging}, which
-% explores a space of ARGs by fixing a starting ARG, and then repeatedly erasing
-% and resimulating candidate SPR moves connecting two fixed, leaf-labelled local
-% trees.
-Inferring the SPRs between leaf-labelled trees is
-NP-hard~\citep{hein1996complexity,allen2001subtree,bordewich2005computational},
-but it is unclear what the complexity is when there
-is a degree of internal node sharing between trees.
-% I don't really know if this is true, it's just a way of wrapping the
-% section up by saying that other things we thought were settled might
-% be worth thinking about again.
-The combinatorial formulation of ARGs and local trees provided here
-may help clarify these fundamental questions.
-
-\section{Locally unary nodes}
-\label{sec-locally-unary-edges}
-As discussed in the previous section, the local tree at a given position
-$x$ is best seen as the path through the ARG at that position, defined
-by the oriented tree $\pi^x_1\dots\pi^x_n$. This path does not directly
-contain information about branching points, and defining a
-node's arity (number of child nodes) is therefore useful.
-The ``local arity'' of a node is the number of children it has
-in the local tree at position $x$, i.e., $a^x_u = |\{v : \pi^x_v = u\}|$
-for each $1 \leq u \leq n$. The ``ARG arity'' of a node $u$ is the
-number of children it has in the graph topology, i.e.\
-$a_u = |\{v: (v, u, I) \in E\}|$. Thus, the local arity is less than
-or equal to the ARG arity (more precisely, $0 \leq a^x_u \leq a_u$),
-and the local arity of a node may change as we move along the genome.
-
-This distinction between ARG and local arity is mainly
-of interest when we consider nodes that have
-a single child: those that are \emph{unary}.
-Returning to the example in Fig.~\ref{fig-arg-in-pedigree}, nodes
-\noderef{g} and \noderef{h}
-are ARG-unary (Fig.~\ref{fig-arg-in-pedigree}B), and are consequently
-also unary in the local trees (Fig.~\ref{fig-arg-in-pedigree}C).
-On the other hand, node \noderef{f} has two children
-in the graph, but is binary only
-in the local tree covering the interval $[2, 7)$,
-representing the coalescence of samples \noderef{a} and \noderef{c}
-in this genome region. Over the interval $[0, 2)$ no coalescence occurs,
-but we still record the fact that genome \noderef{c} inherits from \noderef{f}
-in the local tree. Thus, node \noderef{f} has a single child in this
-interval: it is \emph{locally unary}.
-In another example, \noderef{e} is binary in the graph, being a common
-ancestor of \noderef{a} and \noderef{c}, but is
-locally unary in all trees in which it is present.
-This is because no ancestral material coalesces in \noderef{e}:
-\noderef{a} inherits genetic material from the far left hand end of
-\noderef{e},
-while \noderef{c} only inherits the (disjoint) right hand end.
-
-% We get discussion of ARG-unary nodes out of the way first: these are
-% relatively easy to understand (but it's worth emphasising that
-% eARG-like recombination nodes are ARG-unary, as that might not be clear
-By definition, ARG-unary nodes have one child but can have one or
-more parents.
-A node with one child and only one parent represents a ``pass-through'' node:
-these occur where we wish the record the passage of ancestral material
-through a known node. For example, in simulations
-it is sometimes useful to record the passage of
-ancestral material through known pedigree individuals
-regardless of whether common ancestry occurs.
-Nodes with one child and two parents arise when
-we model a recombination event using a single node
-in the classical manner
-(e.g.\ Fig.~\ref{fig-ancestry-resolution}).
-It is also possible for sample nodes to be ARG-unary, for example
-in inferences from longitudinal datasets where genetic
-data is sampled at many timepoints and recombination
-is rare (e.g.~SARS-CoV-2; see Discussion).
-
-% Moving on from ARG-unary to the more general case
-More generally, locally unary nodes,
-which can have one or more children in the graph,
-are a common and important feature
-of many different types of ARG.
-As discussed in the previous section,
-without these nodes marking the passage
-of ancestral material through specific ancestors, the local trees
-lack information about events other than local coalescence.
-% Not sure if switching example will confuse people here, but would
-% like to contrast with example where we don't use unary nodes to
-% emphasise the point that we lose info
-For example, the local trees for the classical event ARG
-depicted in Fig.~\ref{fig-event-arg}B follow the usual conventions
-and do not include any information about the recombination
-that occurred at node \noderef{d}. Given these two local trees
-in isolation
-% (without further information about the corresponding ARG)
-we lack specific information about the recombination.
-Explicitly recording that node \noderef{d} lies on the
-branch joining \noderef{b} to \noderef{e} in the left hand
-tree, and \noderef{b} to \noderef{f} in the right hand tree
-resolves all ambiguity, and makes the collection of local
-trees exactly equivalent to the corresponding ARG (see previous section).
-Unary nodes are a vital link between ARGs and local trees, and we
-cannot fully reason about how a local tree is embedded in an ARG
-without them. As we see in the next two sections, both ARG and locally
-unary nodes occur in various scenarios, and are
-produced by a range of current inference methods.
-
-\section{Levels of simplification}
-\label{sec-ARG-simplification}
-ARG simplification is a powerful tool.
-In general, we can think of
-simplification as the process
-of removing nodes and re-writing edges (and their inheritance annotations)
-to remove various types of redundancy, much of which
-% We are interested in both graph and local unary, so qualifying this
-% would just make it less readable, IMO
-revolves around the presence of unary nodes (see previous section).
-We illustrate this successive removal of redundancy
-through a series of simplification steps
-in Fig.~\ref{fig-simplification}.
+}
+\end{figure}
+
+These give different levels of precision about recombination (see new Appendix X)
+
+You can see this as different levels of detail on the same underlying truth. Here's a figure:
 
@@
 
-The ARG in Fig.~\ref{fig-simplification}A is the output of a
-backwards-time Wright-Fisher simulation for a sample of two diploid
-individuals (population size $N=10$), and follows a similar process
-to the methods described by~\cite{nelson2020accounting}.
-As we proceed backwards in time, generation by generation, the
-extant lineages choose parents randomly.
-With a certain probability recombination occurs, and the ancestral
-material of a lineage is split between the two parental
-genomes. Local coalescence occurs
-when lineages with overlapping ancestral material choose the same
-parent genome.
-Note that in this simulation we do not explicitly
-model recombination \emph{events} via an ARG node, but simply record
-the \emph{outcome} of a recombination via edges to
-the parent's two genomes.
-Thus, a recombinant node such as \noderef{g} in Fig.~\ref{fig-simplification}
-may also correspond to a coalescence.
-The distinction of
-using a single node to represent
-a recombination event, as is done in ~Fig.~\ref{fig-ancestry-resolution},
-or two to represent the parent genomes, as in Fig.~\ref{fig-simplification},  is usually not important.
-Either is possible in the gARG encoding, and the most convenient approach will vary by application
-(discussed e.g.\ in Appendix~\ref{sec-big-and-little-arg}).
-Note also that node \noderef{k} in Fig.~\ref{fig-simplification} has three children.
-Polytomies like this are a natural feature of such a Wright-Fisher model
-(but see Appendix~\ref{sec-cell-lineages-and-args}).
-
-The graph visualisations in Fig.~\ref{fig-simplification} have
-three novel features which require some explanation.
-Firstly, edge weights (the thickness of the lines joining
-nodes) correspond to the length of the inheritance intervals
-they are annotated with. This allows us to distinguish
-edges that persist across many local trees from those that are
-less influential (contrast the edge
-$(\noderef{g}, \noderef{h})$
-with $(\noderef{g}, \noderef{i})$
-in Fig.~\ref{fig-simplification}A).
-Secondly, node colours denote the number of parents that they
-have in the graph, allowing us to easily see roots (those
-with zero parents), recombinants (those with two parents)
-and more complex situations arising from simplification (see below).
-Thirdly, the shading intensity of a node denotes the ``coalescent span'', the
-fraction of the node's span (the length of genome in which it
-is reachable from the samples in the local trees)
-over which it has more than one child. Nodes which
-are never locally unary therefore have a coalescent span of 100\%,
-whereas nodes in which ancestral material never coalesces have a
-coalescent span of 0\%.
-
-Returning to the main topic of this section,
-Fig.~\ref{fig-simplification}A is the original simulation output, in
-which we retain all nodes involved in recombination
-or common ancestry events. This is the true history, and contains
-a very high level of detail, some of which may be considered
-redundant (or, from another perspective, unobservable).
-In Fig.~\ref{fig-simplification}A the local trees (right)
-contain many unary nodes, fewer as  we successively
-simplify (Fig.~\ref{fig-simplification}B,C),
-until we reach Fig.~\ref{fig-simplification}D, where there are none.
-
-The first level of simplification that we can perform is based only
-on the graph topology.
-An example of graph topology that we may consider redundant
-(or non-identifiable) is a
-``diamond''~\citep{rasmussen2014genome}
-in which the two parent nodes of a recombination immediately
-join again into a common ancestor (e.g.~\noderef{j}, \noderef{l}, \noderef{m}
-and \noderef{n} in Fig.~\ref{fig-simplification}A).
-Unless we are specifically
-interested in the recombination event or these ancestral genomes,
-the diamond can be replaced by a single edge without loss of
-information.
-More generally, any subgraph that is singly-connected in both the leafward and
-rootward direction (a ``super-diamond'') can be replaced by one edge.
-This definition includes the case
-of a node that has one inbound and one outbound edge, such as
-nodes \noderef{f} and \noderef{h}.
-% Note that f and h are only removable because we are using the 2-RE-node
-% representation. That means when we remove these nodes, we lose information about
-% the time of recombination, which we wouldn't do if we were using the
-% 1-RE-node representation (because the node would be multiply connected
-% in the rootward direction). We definitely don't want to get into this
-% level of detail here though, so we'll just pass over it without comment.
-Fig.~\ref{fig-simplification}B shows the result of this type of
-graph topology simplification.
-
-Simplifying away diamonds will remove many unary nodes from the
-local trees, but there can still be nodes that are unary in all
-of the local trees. In particular, a node can represent a recombinant
-with multiple parents in the graph but only a single child (e.g.\ node \noderef{n}
-in Fig.~\ref{fig-simplification}B), or can represent a common ancestor with
-multiple children in the graph but in which no coalescence takes place
-in the local trees
-(node \noderef{r} in Fig.~\ref{fig-simplification}B).
-The distinction between the ``common ancestry'' of two or more genomes
-in an ancestral genome and the ``coalescence'' which may or may
-not occur in the local trees is
-important~\citep{hudson1983testing,kelleher2016efficient}.
-Consider \noderef{e} in Fig.~\ref{fig-simplification}A,
-for example. We can see from the graph that it is a common
-ancestor of samples \noderef{a} and \noderef{b}, but
-it does not correspond to any coalescence in the
-local trees to the left of position $44$, and is therefore
-unary in these three trees.
-Such nodes are not singly connected in the graph, but are nevertheless unary in
-all of the local trees.
-The operation to remove them
-therefore requires knowledge not just of the graph topology but also of the
-ancestral material associated with the edges.
-As we see in Fig.~\ref{fig-simplification}C,
-removal of recombinant nodes can produce graph nodes with
-more than two parents (e.g.~node \noderef{e}); and likewise, removal of
-common ancestor but non-coalescent nodes can produce graph nodes with
-more than two children (e.g.~node \noderef{s}). Both cases represent the
-merged
-\emph{effects} of multiple evolutionary events in a single node (genome), and the
-ARG no longer contains the intermediate genomes corresponding to
-those events (see also Appendix~\ref{sec-cell-lineages-and-args}).
-
-The remaining nodes are MRCAs of some subset of the samples
-at \emph{some} positions along the genome. We still have
-some unary nodes in the local trees, but these nodes will
-correspond to a coalescence in at least one other
-local tree. For example, node  \noderef{k} is unary in the fourth tree
-of Fig.~\ref{fig-simplification}C, but is either binary
-or ternary in all other local trees (recall this is a Wright-Fisher
-simulation). The final level of simplification is to alter the edge annotations
-such that, although no nodes are removed from the graph, all
-unary nodes disappear from the local trees (Fig.~\ref{fig-simplification}D).
-Note that although this last stage produces simpler local trees, by
-removing information about the exact paths taken by lineages through
-the graph, we lose potentially useful information about shared edges
-between trees.
-The \texttt{msprime} simulator, and the version of Hudson's algorithm described
-by~\citet{kelleher2016efficient}, produces ARGs
-that are fully simplified (i.e., contain no locally unary nodes).
-It is not difficult, however, to update
-these methods to record information about the passage of ancestral
-material through genomes under a range of conditions.
-
-An important consequence of simplifying ARGs to remove
-unary nodes in local trees is that we lose some information
-about recombination
-events. This is related to the amount of \emph{precision} about
-recombination events that we store and
-can hope to infer from sampled genomes, which is the topic of the next
-section.
-
-\section{Precision of recombination information}
-\label{sec-precision}
-% The ARGs in Fig.~\ref{fig-simplification} represent different
-% levels of detail about the same ancestral history. They represent
-% the same set of recombination and common ancestor events,
-As illustrated in Fig.~\ref{fig-simplification}, successive levels
-of ARG simplification reduce the amount of information about the
-history of the sample that is stored. Some of the information lost,
-e.g.\ ``diamond'' removal (Fig.~\ref{fig-simplification}A),
-seems like a reasonable tradeoff for a simpler structure.
-The consequences of other simplifications, however, are
-more subtle and relate directly to what can be known about
-recombination events and the levels of precision that
-we should seek to infer about them.
-
-The ARGs in Fig.~\ref{fig-simplification} contain different
-numbers of local trees ($6$, $5$, $5$ and $4$ respectively for A through
-D). When we move from A to B the local trees
-for the intervals $[44,61)$ and $[61,87)$ are merged because
-the only differences between them are their paths through
-nodes \noderef{l} and \noderef{m}. These nodes that participated
-in the diamond are removed from the ARG, and we have lost
-all information about the corresponding recombination at
-position 61. Other nodes (e.g.\ \noderef{o} and \noderef{p})
-have also been removed but these represent the \emph{parents}
-of recombinants. The recombinant nodes themselves
-(e.g. \noderef{n}) are still present, and represent precise
-information about the time, genomic location and
-lineages involved in the recombination event.
-
-Fig.~\ref{fig-simplification}C has the same number of local trees
-as Fig.~\ref{fig-simplification}B, but has less precise information
-about recombination. Continuing the previous example, node
-\noderef{n} has been removed from the graph because it was unary
-in all of the local trees; its outbound edges to \noderef{s}
-and \noderef{q} have effectively been ``pushed down''
-to \noderef{e} (which is retained because it is the coalescent
-parent of \noderef{a} and \noderef{b} over the interval
-$[44, 100)$). We
-have therefore lost precision about
-the \emph{timing} of this recombination event, and know only
-that it must have occurred between the times of node \noderef{e}
-and \noderef{q}.
-
-Fig.~\ref{fig-simplification}D removes all unary nodes from the
-local trees, and further reduces the precision of
-recombination information. Node \noderef{e} has not been
-removed from the graph because it is coalescent in the
-final tree, but we no longer know that the recombination
-event at position 30 was ancestral to it, or have
-any indication of its timings. Furthermore,
-trees for $[44, 87)$ and $[87, 100)$ were only distinguishable
-by the passage of the former tree through nodes \noderef{e}
-and \noderef{q}, and so the recombination on node \noderef{g}
-at position 87 has been lost entirely.
-
-\section{Example inferred ARGs}
-\label{sec-example-inferred-args}
-The scalability gains made by recent ARG inference methods
-such as \relate~\citep{speidel2019method} and
-\tsinfer~\citep{kelleher2019inferring}
-have been, in part, due to inferring lower levels of precision about
-recombination than classical methods.
-Neither method infers
-explicit recombination events, and therefore their outputs cannot
-be described using the classical eARG formalisms (Section~\ref{sec-eARG}).
-% Could do, something like
-% this mismatch has led some to characterise their output as a
-% series of unrelated local trees [cite], which lacks nuance of the reality.
-% Or maybe we don't bother: what's to be gained from pointing out people's
-% mistakes here? Hopefully the rich detail of reality will come through
-% and replace old ideas.
-Nonetheless, both methods produce estimates in which nodes and edges
-persist across multiple trees, creating
-inheritance graphs which fit naturally into the gARG formulation.
-To illustrate the varying levels of information captured
-by current methods, and some qualitative differences between them,
-Fig.~\ref{fig-inferred-args} shows graphical depictions of
-example ARGs produced by four tools using substantially
-different inference strategies.
-
-\begin{figure} \begin{center}
-\includegraphics[width=\textwidth]{illustrations/inference.pdf} \end{center}
-\caption{\label{fig-inferred-args} Inference of sample-resolved ARGs for 11
-\textit{Drosophila melanogaster} DNA sequences over a 2.4kb
-region of the ADH locus~\citep{kreitman1983nucleotide}.
-Results for four different methods:
-(A) \kwarg; (B) \argweaver; (C) \tsinfer; and (D) \relate. See the
-text for details of these methods.
-Edge colours indicate time of the edge's child node
-(lighter: older; darker: younger).
-Vertical and horizontal positions of graph nodes are arbitrary.
-Line width and node colour are as described in Fig.~\ref{fig-simplification}.
-Bottom row graphics show the genome positions, relative to the start of
-the ADH gene, for each graph edge from the corresponding ARG. Edge intervals
-are drawn as horizontal lines, stacked in time order (edges with youngest
-children at the bottom); vertical dashed lines denote breakpoints between
-local trees.
-}
-\end{figure}
-
-The first two methods explicitly infer recombination events.
-\kwarg~\citep{ignatieva2021kwarg} is a parsimony based approach
-which searches the space of plausible ARGs,
-outputting minimal ones using heuristics.
-\argweaver~\citep{rasmussen2014genome} on the other hand is
-model-based, sampling from a discretised version of the
-% could spell this out, but in reality everyone knows what the SMC
-% is, and if they don't spelling it out won't help them.
-SMC~\citep{mcvean2005approximating,marjoram2006fast}.
-Note that both \kwarg\ and \argweaver\ produce many ARGs,
-and those shown in Fig.~\ref{fig-inferred-args} are arbitrarily selected
-examples.
-While the second two methods
-both produce a single best-guess estimate and
-do not explicitly infer recombination events, they are based on
-quite different principles.
-\texttt{Tsinfer} works in a two-step process, first generating ancestral
-haplotypes via heuristics and then inferring inheritance relationships
-between them using the Li and Stephens model~\citep{li2003modeling}.
-\texttt{Relate} first reconstructs local tree topologies across the genome,
-using a variant of the Li and Stephens model to estimate the
-ordering of coalescence events in each tree, and then estimates
-branch lengths using MCMC with a coalescent-based prior.
-See Appendix~\ref{sec-survey-arg-infer} for more details on these
-and other inference methods.
-
-Inferred ARGs are based on the \citet{kreitman1983nucleotide} dataset,
-a standard benchmark in the classical ARG
-literature.
-It consists of 43 biallelic SNPs spanning 2.4Kb of the \emph{D.~melanogaster} ADH
-locus on chromosome 2L. Where required for inference purposes we assume
-mutation and recombination rates of $5.49\times10^{-9}$
-and $2.40463\times10^{-9}$ per site per generation
-\citep{schrider2013rates,comeron2012many}
-and a constant effective population size of 1,720,600 \citep{li2006inferring},
-as provided by the \texttt{stdpopsim}
-catalog~\citep{adrion2020community,lauterbur2023expanding}.
-Software versions were \kwarg~v1.0, \argweaverD~(2019),
-\tsinfer~v0.3.1, and \relate~v1.1.9.
-Full details and code for generating
-these figures are available on GitHub (see Data Availability).
-
-% Here we describe recombination features
-Considering Fig.~\ref{fig-inferred-args}, we can see that
-there is substantial variation in the number of recombination breakpoints
-inferred by different methods, with e.g.~\argweaver\ suggesting
-far more than the 7 required for this dataset under minimal parsimony
-assumptions~\citep{song2003parsimonious}.
-A sense of the amount of recombination in each ARG is provided by
-the node colouring scheme, which shows the number of parents for each node.
-In Fig.~\ref{fig-inferred-args}A,B, each recombination event
-corresponds to a node with exactly two parents and one child.
-As these methods explicitly infer a recombination event for each
-breakpoint, the number of breakpoints equals the number of two-parent
-(brown) nodes.
-In contrast, Fig.~\ref{fig-inferred-args}C,D do not have this
-straightforward
-relationship between the number of nodes with multiple parents
-and number of breakpoints along the genome.
-In both ARGs the number of breakpoints is smaller
-than the number of multiple-parent ARG nodes, showing
-that several multiple-parent nodes must share breakpoint positions.
-There are also ARG nodes with multiple parents and multiple children,
-where one or more recombinations have been pushed down onto a
-more recent node.
-As a consequence, it may be difficult to condense each transition between trees
-in these ARGs into a set of SPR operations.
-
-% Here we move on to coalescence
-Shading within nodes in Fig.~\ref{fig-inferred-args} indicates the
-fraction of the node's span over which it is coalescent
-(Section~\ref{sec-locally-unary-edges}). For example, brown nodes
-in Fig.~\ref{fig-inferred-args}A,B are clear because there is
-no local coalescence at these recombination nodes
-(they are ``ARG unary'', and so local coalescence is impossible).
-The significant number of partially shaded nodes in
-Fig.~\ref{fig-inferred-args}A,B and C demonstrates that
-the \kwarg, \argweaver\ and \tsinfer\ ARGs all contain locally unary nodes.
-Another difference between methods highlighted in this figure
-is the presence of polytomies, which only
-\tsinfer\ creates. The most obvious example
-involves nodes \noderef{Fr-F}, \noderef{Wa-F}, and \noderef{Af-F},
-which happen to have identical sequences.
-Because \kwarg, \argweaver, and \relate\ require
-bifurcating trees by design, each picks an arbitrary order of branching
-(hence Fig.~\ref{fig-inferred-args}A and B disagree
-in this order, and Fig.~\ref{fig-inferred-args}D even shows different
-orders in different trees).
-% JK: I don't quite get the commented out sentence below, so readers probably
-% won't either. Let's stick to a few clear messages.
-% More generally, the number of nodes in Fig.~\ref{fig-inferred-args}C
-% that have more than two children hints at considerable uncertainty in
-% the order of branching elsewhere in the ARG.
-
-The bottom row of Fig~\ref{fig-inferred-args} shows the extent along
-the genome to which graph edges are shared between multiple trees.
-All four methods infer nodes and edges that are shared between
-multiple trees, to varying degrees.
-For example, all of the methods infer
-infer that \noderef{Af-f}, \noderef{Fr-f}, and \noderef{Wa-f}
-form a clade along the entire sequence.
-In particular, we can
-see both \tsinfer\ and (to a lesser extent) \relate\ have edges
-that span multiple tree boundaries, indicating that they are not inferring
-a series of \emph{unrelated} local trees.
-However, in comparison to \kwarg\ and \argweaver\ neither
-method results in extensive node sharing in the
-oldest time periods.
-Overall, Fig.~\ref{fig-inferred-args} shows
-that \tsinfer\ and \relate\ ARGs contain a level of detail
-that lies somewhere between a sequence of unrelated local trees on one extreme
-and an ARG with precisely specified recombination events on the other
-(Fig~\ref{fig-inferred-args}A,B).
-% JK: Let's not get into this.
-% This primarily reflects the fact that they do not encode topology
-% above the local MRCAs, also indicated by multiple roots in the ARG
-% (nodes with zero parents, in blue), compared to the
-% single ARG root in the \kwarg\ and \argweaver\ genealogies. In terms of inference,
-% \tsinfer\ and \relate\ show a pragmatic representation: we do not expect
-% genetic variation to be very informative about topological relationships
-% between nodes that that are older than the local MRCAs.
-
-\section{Implementation and efficiency}
-\label{sec-efficiency}
-The gARG encoding defined here
-leads to highly efficient storage and processing of ARG data,
-and has already been in use for several years.
-The succinct tree sequence data structure
-(usually known as a ``tree sequence'' for brevity)
-is a practical gARG implementation focused on efficiency.
-It was originally developed as part of the \texttt{msprime}
-simulator~\citep{kelleher2016efficient} and has subsequently been
-extended and applied to forward-time
-simulations~\citep{kelleher2018efficient,haller2018tree},
-inference from
-data~\citep{kelleher2019inferring,wohns2022unified,zhan2023towards},
-and calculation of population genetics statistics~\citep{ralph2020efficiently}.
-The succinct tree sequence encoding extends the basic definition
-of a gARG provided here by stipulating a
-simple tabular representation of nodes and edges,
-and also defining a concise representation of
-sequence variation using the ``site'' and  ``mutation'' tables.
-The key property of the succinct tree sequence encoding
-that makes it an efficient substrate for defining analysis
-algorithms is that it allows us to sequentially
-recover the local trees along the genome very efficiently,
-and in a way that allows us to reason about the \emph{differences}
-between those trees~\citep{kelleher2016efficient,ralph2020efficiently}.
-
-The \texttt{tskit} library is a liberally
-licensed open source toolkit that provides a comprehensive suite
-of tools for working with gARGs (encoded as a succinct tree sequence).
-Based on core functionality written
-in C, it provides interfaces in C, Python and Rust.
-Tskit is mature software, widely used in population genetics, and
-has been incorporated into numerous downstream
-applications~\citep[e.g.,][]{haller2019slim,speidel2019method,
-adrion2020community,
-terasaki2021geonomics,
-baumdicker2021efficient,
-fan2022genealogical,
-guo2022recombination,
-korfmann2022weak,
-mahmoudi2022bayesian,
-petr2022slendr,
-rasmussen2022espalier,
-zhang2023biobank,
-nowbandegani2023extremely,
-ignatieva2023distribution,
-fan2023likelihood}.
-The technical details of \texttt{tskit}, and how it provides an
-efficient and portable platform for ARG-based analysis, are beyond
-the scope of this manuscript.
-In the interest of avoiding confusion, however, we list a
-few minor details in which the formal details of gARGs
-provided in Section~\ref{sec-gARG} differ from their practical implementation in
-\texttt{tskit}.
-Firstly, ``edges'' in tree sequence terminology would perhaps be better
-described as ``edge-intervals'', as each describes a single contiguous
-interval of genome inheritance between a pair of nodes.
-This denormalisation of the gARG data model is for efficiency purposes.
-Secondly, zero- rather than one-based indexing is used for
-nodes in ARGs and oriented trees; consequently  $-1$ is used to denote the presence of
-roots (rather than $0$ as used here for notational simplicity).
+%%% Pasted from previous para
+
+The timing, positions, and even the number of recombination events is generally
+not possible to infer precisely from genome sequencing data. Under
+coalescent-based models, the proportion of recombination events that change the
+ARG topology grows very slowly with sample size \citep{hein2004gene}, and of those
+events only a small proportion are actually detectable from the data, assuming
+human-like mutation and recombination
+rates \citep{myers2002detection,hayman2023recoverability}.
+Even when a recombination event \emph{is} detectable, its timing and breakpoint
+position can only be inferred approximately, depending on how much information
+can be elucidated from mutations in the surrounding genomic region.
+% For general
+% ARGs, for instance those representing pedigrees, or genealogies of organisms with
+% low mutation rates and complex patterns of recombination, such inference can be
+% even more challenging.
+The fact that the eARG encoding \emph{requires}
+precise information about recombination is therefore a fundamental limitation.
 
@@
 
-The timing, positions, and even the number of recombination events is generally
-not possible to infer precisely from genome sequencing data. Under
-coalescent-based models, the proportion of recombination events that change the
-ARG topology grows very slowly with sample size \citep{hein2004gene}, and of those
-events only a small proportion are actually detectable from the data, assuming
-human-like mutation and recombination
-rates \citep{myers2002detection,hayman2023recoverability}.
-Even when a recombination event \emph{is} detectable, its timing and breakpoint
-position can only be inferred approximately, depending on how much information
-can be elucidated from mutations in the surrounding genomic region.
-% For general
-% ARGs, for instance those representing pedigrees, or genealogies of organisms with
-% low mutation rates and complex patterns of recombination, such inference can be
-% even more challenging.
-The fact that the eARG encoding \emph{requires}
-precise information about recombination is therefore a fundamental limitation.
-
 % % Do downstream applications actually make use of such full ARGs?
-Besides the inherent limitations that exist on inferring fully
-precise ARGs from data,
-we should also consider the value that such exact estimates provide
-for downstream applications.
-Many applications work by examining local trees independently,
-making detailed information about recombination events superfluous.
-For example, the \relate\ selection test~\citep{speidel2019method}
-obtains $p$-values by computing clade size probabilities conditional
-on the timing of coalescence events in a given local tree.
-In their method
-for estimating dispersal rates and the locations of genetic
-ancestors,
-\cite{osmond2021estimating} downsample trees along the genome
-so that they can be regarded as approximately independent.
-Similarly, \cite{fan2023likelihood} compute the likelihood
-of an ARG under a particular demographic model as the product
-over a sample of widely-separated local trees, assumed to be independent.
-The SIA method for detecting selection~\citep{hejase2022deep}
-encodes local trees as a set of lineage counts at discrete
-time intervals, and uses these as feature for a
-type of machine learning algorithm
-that takes ``temporal'' correlations into account.
-Thus, while SIA takes advantage of information about local tree correlation,
-it is in quite an indirect way, and
-clearly much of the detail about recombination events in an ARG is lost.
+%Besides the inherent limitations that exist on inferring fully
+%precise ARGs from data,
+%we should also consider the value that such exact estimates provide
+%for downstream applications.
+%Many applications work by examining local trees independently,
+%making detailed information about recombination events superfluous.
+%For example, the \relate\ selection test~\citep{speidel2019method}
+%obtains $p$-values by computing clade size probabilities conditional
+%on the timing of coalescence events in a given local tree.
+%In their method
+%for estimating dispersal rates and the locations of genetic
+%ancestors,
+%\cite{osmond2021estimating} downsample trees along the genome
+%so that they can be regarded as approximately independent.
+%Similarly, \cite{fan2023likelihood} compute the likelihood
+%of an ARG under a particular demographic model as the product
+%over a sample of widely-separated local trees, assumed to be independent.
+%The SIA method for detecting selection~\citep{hejase2022deep}
+%encodes local trees as a set of lineage counts at discrete
+%time intervals, and uses these as feature for a
+%type of machine learning algorithm
+%that takes ``temporal'' correlations into account.
+%Thus, while SIA takes advantage of information about local tree correlation,
+%it is in quite an indirect way, and
+%clearly much of the detail about recombination events in an ARG is lost.
+
 % Mentioning tsdate here doesn't advance the
@@
 % to reason about ancestral nodes~\citep{wohns2022unified}.
-The main application for fully precise ARGs thus far has been
-to compute a likelihood under the
-coalescent~\citep[e.g.][]{kuhner2000maximum,mahmoudi2022bayesian,
-guo2022recombination},
-which currently requires the details of all recombination
-events to be known.
-
-The advantages of a model-agnostic representation that naturally
-incorporates uncertainty about the ordering of events in an ARG
-are well-illustrated by~\cite{zhan2023towards}, who
-inferred ARGs using millions of SARS-CoV-2 sequences from the
-GISAID database~\citep{shu2017gisaid}.
-In contrast to typical human sequencing datasets, the SARS-CoV-2 data is
-sampled continuously through time, sometimes with tens of thousands of sequences collected per day, with relatively little genetic diversity to
-distinguish them. The reconstructed ARGs thus contain polytomies
-and non-leaf sample nodes (sequences with descendants also present
-in the dataset) many of which only have a single child (i.e.\ are ARG-unary).
-Recombination is an important factor in the evolution of
-SARS-CoV-2 \citep{vaninsberghe2021recombinant,jackson2021generation,ignatieva2022ongoing},
-% Tried to jazz this up a bit, was quite dry
-and the inferred ARGs contain an unprecedented level of detail about
-the combined processes of viral mutation and recombination.
-Because parental sequences
-are generally never sampled themselves, and often
-a recombinant strain is the product of multiple recombination events,
-uncertainty around this is captured by recording the ancestry
-of each part of the recombinant sequence without arbitrarily assigning
-times or orderings for these events.
+% The main application for fully precise ARGs thus far has been
+% to compute a likelihood under the
+% coalescent~\citep[e.g.][]{kuhner2000maximum,mahmoudi2022bayesian,
+% guo2022recombination},
+% which currently requires the details of all recombination
+% events to be known.
+
+% The advantages of a model-agnostic representation that naturally
+% incorporates uncertainty about the ordering of events in an ARG
+% are well-illustrated by~\cite{zhan2023towards}, who
+% inferred ARGs using millions of SARS-CoV-2 sequences from the
+% GISAID database~\citep{shu2017gisaid}.
+% In contrast to typical human sequencing datasets, the SARS-CoV-2 data is
+% sampled continuously through time, sometimes with tens of thousands of sequences collected per day, with relatively little genetic diversity to
+% distinguish them. The reconstructed ARGs thus contain polytomies
+% and non-leaf sample nodes (sequences with descendants also present
+% in the dataset) many of which only have a single child (i.e.\ are ARG-unary).
+% Recombination is an important factor in the evolution of
+% SARS-CoV-2 \citep{vaninsberghe2021recombinant,jackson2021generation,ignatieva2022ongoing},
+% % Tried to jazz this up a bit, was quite dry
+% and the inferred ARGs contain an unprecedented level of detail about
+% the combined processes of viral mutation and recombination.
+% Because parental sequences
+% are generally never sampled themselves, and often
+% a recombinant strain is the product of multiple recombination events,
+% uncertainty around this is captured by recording the ancestry
+% of each part of the recombinant sequence without arbitrarily assigning
+% times or orderings for these events.
 
@@
 
+\section{ARGs and local trees}
+\label{sec-ARG-and-local-trees}
+% (We use ``local'' rather than
+% ``marginal''
+% % could list loads here, but not much point.
+% trees~\citep[e.g.][]{griffiths1996ancestral,minichiello2006mapping,kelleher2016efficient}
+% to avoid the potential confusion with the
+% statistical concept of marginal distributions.)
+The relationship between an ARG and its corresponding
+local trees is subtle and important.
+A fundamental property of genetics is that a
+given DNA nucleotide is inherited from exactly one parent genome,
+both at an organismal and cell-by-cell level
+(Appendix~\ref{sec-cell-lineages-and-args}).
+These paths of single-parent inheritance give rise,
+by definition, to a tree structure.
+As a result of recombination, adjacent nucleotides can have
+different paths of inheritance, and
+an ARG encodes the entire ensemble of local trees along the
+% Said "sample of genomes" originally, but using term "genome" in two
+% different ways in the same sentence then.
+genome for a given set of sample nodes.
+% This is for Yan - to me this is a given, but maybe some people feel
+% the following is overly formal?
+Precisely defining the process by which local trees are extracted
+from an ARG is essential to our understanding of how ARGs and local trees
+are related, and we require a concrete mathematical structure
+to describe the local trees. It is important to note that
+although the following discussion is phrased in terms of the
+gARG encoding,  the arguments apply equally to eARGs
+because any eARG can be converted to a
+% Note, being sample-resolve is not necessary for following argument so omitting
+gARG without loss of information (Section~\ref{sec-ancestral-material}).
+
+Oriented trees provide a
+convenient formalism to capture these parent-child relationships
+in a well-defined combinatorial object.
+% There's no need for a comma
+Let $\pi_1\dots\pi_n$ be a sequence of integers, such that $\pi_u$
+denotes the parent of node $u$, and $\pi_u = 0$ if $u$ is a
+root~\cite[p.\ 461]{knuth11combinatorial}.
+This encoding is particularly useful to describe
+evolutionary trees because parent-child relationships are
+important but the ordering of children at a node is
+not~\citep{kelleher2013coalescent,kelleher2014coalescent,
+kelleher2016efficient}.
+Thus, for a given gARG with nodes $\{1, \dots, n\}$ and
+edges $E$ (Section~\ref{sec-gARG}), we recover the local tree
+at position $x$ as follows.
+We begin by setting $\pi_u = 0$ for each $1 \leq u \leq n$.
+Then, for each sample node in $S$ we trace its path pastwards through the
+ARG for position $x$, and record this path in $\pi$.
+Specifically, at a given node $u$,
+we find an edge $(c, p, I) \in E$ such that $u = c$ and $x \in I$, and set
+$\pi_c \leftarrow p$. We then set $u \leftarrow p$, and repeat
+until either $\pi_u \neq 0$ (indicating we have traversed this section
+of the ARG already on the path from another sample) or there
+is no matching outbound edge (indicating we are at a root).
+Note that the local trees for an ARG are ``sparse''~\citep{kelleher2016efficient},
+because many ancestral nodes will not be reachable from the
+samples at a given position (so their corresponding entries in $\pi$ will be zero).
+
+This combinatorial approach provides at least
+one novel insight, clarifying the fundamental relationship between
+ARGs and local trees.
+Suppose we are given a gARG defined by a set of nodes and edges.
+There is no requirement on the structure of this ARG beyond
+the basic definitions: it could correspond to an ARG in which
+every recombination event is exactly specified
+(e.g.\ Fig.~\ref{fig-ancestry-resolution})
+or one in which local trees are entirely disjoint
+(i.e.\ only the sample nodes are shared between them).
+If we  are given the sequence of local trees for this gARG
+encoded as an oriented tree, along with the genome interval
+covered by each tree, we can recover the original gARG exactly.
+% We do this site-by-site for simplicity, but hopefully it's obvious
+% that this isn't necessary?
+More formally, suppose we are given the local tree $\pi^x_1\dots\pi^x_n$
+for each nucleotide position $1 \leq x \leq L$ on a genome of length $L$.
+Then, the edges of the ``local ARG'' for this tree is given by
+$E^x = \{(u, \pi_u^x, \{x\}) \mid \pi_u^x \neq 0\}$. Because the ARG
+edges are defined by $(c, p, I)$ tuples, where the set $I$ defines
+the positions over which node $c$ inherits from parent $p$, we can
+then simply combine the ``local ARGs'' for each position $x$
+% YW: The above "combining" operation isn't obviously lossless to me:
+% we could potentially lose information in the different combinations
+% of intervals (e.g. if we use 2 non-combined intervals to represent a
+% diamond in a 1-RE-node ARG). It seems like we might need an additional
+% gARG spec to define what combinations of intervals are equivalent.
+% But no-one else has commented on this, so I guess we just let it lie.
+to recover precisely the same set of edges as the original ARG.
+Thus, under this definition, there
+is a one-to-one correspondence between an ARG and
+the sequence of local trees that it encodes.
+
+This is not the prevailing view, however.
+\cite{kuhner2017consensus} argue that the
+``interval-tree'' representation of
+an ARG (the local trees and the genome intervals they cover)
+``does not contain all of the information in the underlying ARG: it lacks the
+number of recombinations occurring at each site, the times at which
+recombinations occurred, and the specific sequences involved as recombination
+partners.''
+\cite{shipilina2023origin} discuss the same ideas,
+and note that the
+``full ARG...~contains more information than the series of tree
+sequences along the genome''.
+These
+% % Could cite some more? E.g. KwARG paper says
+% % "We note that ARGs contain more information than local trees"
+% % Would need a few more examples though, which would require some digging.
+%
+% and similar, \citep[e.g.][]{}
+statements that an ARG
+contains more information than its local trees are true
+if we represent local trees in their conventional forms,
+but these forms discard important information that is
+available in an ARG.
+
+There are two properties of how evolutionary trees are conventionally
+represented that lead to this
+disagreement about the relationship between local trees and an ARG.
+Firstly, the internal nodes of evolutionary trees are usually considered to be
+\emph{unlabelled}, or equivalently, labelled by the leaves which they subtend.
+The same canonical labelling cannot be used for internal ARG nodes because the
+leaves they subtend will typically vary by genomic position. If we do not label
+the tree nodes in a way which is persistent across the sequence of local trees
+in the ARG, we lose the fact that the \emph{same} ancestors sometimes persist
+across multiple trees.
+Defining ARG nodes as integers and using the oriented
+tree encoding explicitly labels internal nodes, and makes the relationship
+between tree and ARG nodes clear and precise.
+
+% Note: good to keep to paras here to logically separate the two important
+% points.
+The second property of how evolutionary trees
+are conventionally represented that is unhelpful in the context of ARGs is their
+focus on branching points (coalescences), i.e.\ nodes that have two or more children.
+As the introductory paragraph of this section emphasised,
+parent-child relationships are what fundamentally define a tree,
+and branching points can be seen as incidental. This is reflected
+by the oriented tree encoding which simply stores the local
+parent-child relationships, and does not, for example,
+directly tell us how many children a particular node has.
+The local tree at a given position records the \emph{path} through
+the ARG; if this path omits nodes that are not
+branching points (such as \noderef{e} in Fig.~\ref{fig-arg-in-pedigree}),
+we lose information about the ARG.
+We return to this point in the following two sections,
+when we discuss ``locally unary'' nodes and the simplification process.
+
+It is important that we make the distinction here between the local
+trees that we can derive from a known ARG (as just discussed),
+and an ARG that we can derive from a sequence of \emph{estimated}
+local trees.
+The ARG inference method
+\espalier~\citep{rasmussen2022espalier} is illustrative in this context.
+It begins by splitting an input sequence alignment into
+segments that are assumed to be non-recombining. Within
+each segment, an initial local tree is estimated using
+standard phylogenetic methods. By necessity, these local
+trees will contain internal nodes that are unlabelled and
+consist only of branching points: there is no information
+shared between the independent tree estimation steps
+across segments. Part of the task of stitching
+these trees together
+into an ARG is then, essentially, to generate labels for
+the internal nodes, and decide which nodes persist
+across multiple local trees.
+\espalier\ approaches this task
+by identifying maximal subtrees
+that do not change between pairs of adjacent
+local trees and then
+heuristically exploring the space of possible
+rearrangements of these subtrees.
+To derive details about recombination events,
+\espalier\ then attempts to infer
+the precise subtree prune-and-regraft (SPR)
+operations~\citep{hein1990reconstructing,song2003on,song2006properties}
+induced by recombination between these partially reconciled local trees.
+% NOTE: dropping this because it doesn't clarify the actual point.
+% A similar problem is tackled by \arbores~\citep{heine2018bridging}, which
+% explores a space of ARGs by fixing a starting ARG, and then repeatedly erasing
+% and resimulating candidate SPR moves connecting two fixed, leaf-labelled local
+% trees.
+Inferring the SPRs between leaf-labelled trees is
+NP-hard~\citep{hein1996complexity,allen2001subtree,bordewich2005computational},
+but it is unclear what the complexity is when there
+is a degree of internal node sharing between trees.
+% I don't really know if this is true, it's just a way of wrapping the
+% section up by saying that other things we thought were settled might
+% be worth thinking about again.
+The combinatorial formulation of ARGs and local trees provided here
+may help clarify these fundamental questions.
+
+\section{Locally unary nodes}
+\label{sec-locally-unary-edges}
+As discussed in the previous section, the local tree at a given position
+$x$ is best seen as the path through the ARG at that position, defined
+by the oriented tree $\pi^x_1\dots\pi^x_n$. This path does not directly
+contain information about branching points, and defining a
+node's arity (number of child nodes) is therefore useful.
+The ``local arity'' of a node is the number of children it has
+in the local tree at position $x$, i.e., $a^x_u = |\{v : \pi^x_v = u\}|$
+for each $1 \leq u \leq n$. The ``ARG arity'' of a node $u$ is the
+number of children it has in the graph topology, i.e.\
+$a_u = |\{v: (v, u, I) \in E\}|$. Thus, the local arity is less than
+or equal to the ARG arity (more precisely, $0 \leq a^x_u \leq a_u$),
+and the local arity of a node may change as we move along the genome.
+
+This distinction between ARG and local arity is mainly
+of interest when we consider nodes that have
+a single child: those that are \emph{unary}.
+Returning to the example in Fig.~\ref{fig-arg-in-pedigree}, nodes
+\noderef{g} and \noderef{h}
+are ARG-unary (Fig.~\ref{fig-arg-in-pedigree}B), and are consequently
+also unary in the local trees (Fig.~\ref{fig-arg-in-pedigree}C).
+On the other hand, node \noderef{f} has two children
+in the graph, but is binary only
+in the local tree covering the interval $[2, 7)$,
+representing the coalescence of samples \noderef{a} and \noderef{c}
+in this genome region. Over the interval $[0, 2)$ no coalescence occurs,
+but we still record the fact that genome \noderef{c} inherits from \noderef{f}
+in the local tree. Thus, node \noderef{f} has a single child in this
+interval: it is \emph{locally unary}.
+In another example, \noderef{e} is binary in the graph, being a common
+ancestor of \noderef{a} and \noderef{c}, but is
+locally unary in all trees in which it is present.
+This is because no ancestral material coalesces in \noderef{e}:
+\noderef{a} inherits genetic material from the far left hand end of
+\noderef{e},
+while \noderef{c} only inherits the (disjoint) right hand end.
+
+% We get discussion of ARG-unary nodes out of the way first: these are
+% relatively easy to understand (but it's worth emphasising that
+% eARG-like recombination nodes are ARG-unary, as that might not be clear
+By definition, ARG-unary nodes have one child but can have one or
+more parents.
+A node with one child and only one parent represents a ``pass-through'' node:
+these occur where we wish the record the passage of ancestral material
+through a known node. For example, in simulations
+it is sometimes useful to record the passage of
+ancestral material through known pedigree individuals
+regardless of whether common ancestry occurs.
+Nodes with one child and two parents arise when
+we model a recombination event using a single node
+in the classical manner
+(e.g.\ Fig.~\ref{fig-ancestry-resolution}).
+It is also possible for sample nodes to be ARG-unary, for example
+in inferences from longitudinal datasets where genetic
+data is sampled at many timepoints and recombination
+is rare (e.g.~SARS-CoV-2; see Discussion).
+
+% Moving on from ARG-unary to the more general case
+More generally, locally unary nodes,
+which can have one or more children in the graph,
+are a common and important feature
+of many different types of ARG.
+As discussed in the previous section,
+without these nodes marking the passage
+of ancestral material through specific ancestors, the local trees
+lack information about events other than local coalescence.
+% Not sure if switching example will confuse people here, but would
+% like to contrast with example where we don't use unary nodes to
+% emphasise the point that we lose info
+For example, the local trees for the classical event ARG
+depicted in Fig.~\ref{fig-event-arg}B follow the usual conventions
+and do not include any information about the recombination
+that occurred at node \noderef{d}. Given these two local trees
+in isolation
+% (without further information about the corresponding ARG)
+we lack specific information about the recombination.
+Explicitly recording that node \noderef{d} lies on the
+branch joining \noderef{b} to \noderef{e} in the left hand
+tree, and \noderef{b} to \noderef{f} in the right hand tree
+resolves all ambiguity, and makes the collection of local
+trees exactly equivalent to the corresponding ARG (see previous section).
+Unary nodes are a vital link between ARGs and local trees, and we
+cannot fully reason about how a local tree is embedded in an ARG
+without them. As we see in the next two sections, both ARG and locally
+unary nodes occur in various scenarios, and are
+produced by a range of current inference methods.
+
+\section{Levels of simplification}
+\label{sec-ARG-simplification}
+ARG simplification is a powerful tool.
+In general, we can think of
+simplification as the process
+of removing nodes and re-writing edges (and their inheritance annotations)
+to remove various types of redundancy, much of which
+% We are interested in both graph and local unary, so qualifying this
+% would just make it less readable, IMO
+revolves around the presence of unary nodes (see previous section).
+We illustrate this successive removal of redundancy
+through a series of simplification steps
+in Fig.~\ref{fig-simplification}.
+
+The ARG in Fig.~\ref{fig-simplification}A is the output of a
+backwards-time Wright-Fisher simulation for a sample of two diploid
+individuals (population size $N=10$), and follows a similar process
+to the methods described by~\cite{nelson2020accounting}.
+As we proceed backwards in time, generation by generation, the
+extant lineages choose parents randomly.
+With a certain probability recombination occurs, and the ancestral
+material of a lineage is split between the two parental
+genomes. Local coalescence occurs
+when lineages with overlapping ancestral material choose the same
+parent genome.
+Note that in this simulation we do not explicitly
+model recombination \emph{events} via an ARG node, but simply record
+the \emph{outcome} of a recombination via edges to
+the parent's two genomes.
+Thus, a recombinant node such as \noderef{g} in Fig.~\ref{fig-simplification}
+may also correspond to a coalescence.
+The distinction of
+using a single node to represent
+a recombination event, as is done in ~Fig.~\ref{fig-ancestry-resolution},
+or two to represent the parent genomes, as in Fig.~\ref{fig-simplification},  is usually not important.
+Either is possible in the gARG encoding, and the most convenient approach will vary by application
+(discussed e.g.\ in Appendix~\ref{sec-big-and-little-arg}).
+Note also that node \noderef{k} in Fig.~\ref{fig-simplification} has three children.
+Polytomies like this are a natural feature of such a Wright-Fisher model
+(but see Appendix~\ref{sec-cell-lineages-and-args}).
+
+The graph visualisations in Fig.~\ref{fig-simplification} have
+three novel features which require some explanation.
+Firstly, edge weights (the thickness of the lines joining
+nodes) correspond to the length of the inheritance intervals
+they are annotated with. This allows us to distinguish
+edges that persist across many local trees from those that are
+less influential (contrast the edge
+$(\noderef{g}, \noderef{h})$
+with $(\noderef{g}, \noderef{i})$
+in Fig.~\ref{fig-simplification}A).
+Secondly, node colours denote the number of parents that they
+have in the graph, allowing us to easily see roots (those
+with zero parents), recombinants (those with two parents)
+and more complex situations arising from simplification (see below).
+Thirdly, the shading intensity of a node denotes the ``coalescent span'', the
+fraction of the node's span (the length of genome in which it
+is reachable from the samples in the local trees)
+over which it has more than one child. Nodes which
+are never locally unary therefore have a coalescent span of 100\%,
+whereas nodes in which ancestral material never coalesces have a
+coalescent span of 0\%.
+
+Returning to the main topic of this section,
+Fig.~\ref{fig-simplification}A is the original simulation output, in
+which we retain all nodes involved in recombination
+or common ancestry events. This is the true history, and contains
+a very high level of detail, some of which may be considered
+redundant (or, from another perspective, unobservable).
+In Fig.~\ref{fig-simplification}A the local trees (right)
+contain many unary nodes, fewer as  we successively
+simplify (Fig.~\ref{fig-simplification}B,C),
+until we reach Fig.~\ref{fig-simplification}D, where there are none.
+
+The first level of simplification that we can perform is based only
+on the graph topology.
+An example of graph topology that we may consider redundant
+(or non-identifiable) is a
+``diamond''~\citep{rasmussen2014genome}
+in which the two parent nodes of a recombination immediately
+join again into a common ancestor (e.g.~\noderef{j}, \noderef{l}, \noderef{m}
+and \noderef{n} in Fig.~\ref{fig-simplification}A).
+Unless we are specifically
+interested in the recombination event or these ancestral genomes,
+the diamond can be replaced by a single edge without loss of
+information.
+More generally, any subgraph that is singly-connected in both the leafward and
+rootward direction (a ``super-diamond'') can be replaced by one edge.
+This definition includes the case
+of a node that has one inbound and one outbound edge, such as
+nodes \noderef{f} and \noderef{h}.
+% Note that f and h are only removable because we are using the 2-RE-node
+% representation. That means when we remove these nodes, we lose information about
+% the time of recombination, which we wouldn't do if we were using the
+% 1-RE-node representation (because the node would be multiply connected
+% in the rootward direction). We definitely don't want to get into this
+% level of detail here though, so we'll just pass over it without comment.
+Fig.~\ref{fig-simplification}B shows the result of this type of
+graph topology simplification.
+
+Simplifying away diamonds will remove many unary nodes from the
+local trees, but there can still be nodes that are unary in all
+of the local trees. In particular, a node can represent a recombinant
+with multiple parents in the graph but only a single child (e.g.\ node \noderef{n}
+in Fig.~\ref{fig-simplification}B), or can represent a common ancestor with
+multiple children in the graph but in which no coalescence takes place
+in the local trees
+(node \noderef{r} in Fig.~\ref{fig-simplification}B).
+The distinction between the ``common ancestry'' of two or more genomes
+in an ancestral genome and the ``coalescence'' which may or may
+not occur in the local trees is
+important~\citep{hudson1983testing,kelleher2016efficient}.
+Consider \noderef{e} in Fig.~\ref{fig-simplification}A,
+for example. We can see from the graph that it is a common
+ancestor of samples \noderef{a} and \noderef{b}, but
+it does not correspond to any coalescence in the
+local trees to the left of position $44$, and is therefore
+unary in these three trees.
+Such nodes are not singly connected in the graph, but are nevertheless unary in
+all of the local trees.
+The operation to remove them
+therefore requires knowledge not just of the graph topology but also of the
+ancestral material associated with the edges.
+As we see in Fig.~\ref{fig-simplification}C,
+removal of recombinant nodes can produce graph nodes with
+more than two parents (e.g.~node \noderef{e}); and likewise, removal of
+common ancestor but non-coalescent nodes can produce graph nodes with
+more than two children (e.g.~node \noderef{s}). Both cases represent the
+merged
+\emph{effects} of multiple evolutionary events in a single node (genome), and the
+ARG no longer contains the intermediate genomes corresponding to
+those events (see also Appendix~\ref{sec-cell-lineages-and-args}).
+
+The remaining nodes are MRCAs of some subset of the samples
+at \emph{some} positions along the genome. We still have
+some unary nodes in the local trees, but these nodes will
+correspond to a coalescence in at least one other
+local tree. For example, node  \noderef{k} is unary in the fourth tree
+of Fig.~\ref{fig-simplification}C, but is either binary
+or ternary in all other local trees (recall this is a Wright-Fisher
+simulation). The final level of simplification is to alter the edge annotations
+such that, although no nodes are removed from the graph, all
+unary nodes disappear from the local trees (Fig.~\ref{fig-simplification}D).
+Note that although this last stage produces simpler local trees, by
+removing information about the exact paths taken by lineages through
+the graph, we lose potentially useful information about shared edges
+between trees.
+The \texttt{msprime} simulator, and the version of Hudson's algorithm described
+by~\citet{kelleher2016efficient}, produces ARGs
+that are fully simplified (i.e., contain no locally unary nodes).
+It is not difficult, however, to update
+these methods to record information about the passage of ancestral
+material through genomes under a range of conditions.
+
+An important consequence of simplifying ARGs to remove
+unary nodes in local trees is that we lose some information
+about recombination
+events. This is related to the amount of \emph{precision} about
+recombination events that we store and
+can hope to infer from sampled genomes, which is the topic of the next
+section.
+
+\section{Precision of recombination information}
+\label{sec-precision}
+% The ARGs in Fig.~\ref{fig-simplification} represent different
+% levels of detail about the same ancestral history. They represent
+% the same set of recombination and common ancestor events,
+As illustrated in Fig.~\ref{fig-simplification}, successive levels
+of ARG simplification reduce the amount of information about the
+history of the sample that is stored. Some of the information lost,
+e.g.\ ``diamond'' removal (Fig.~\ref{fig-simplification}A),
+seems like a reasonable tradeoff for a simpler structure.
+The consequences of other simplifications, however, are
+more subtle and relate directly to what can be known about
+recombination events and the levels of precision that
+we should seek to infer about them.
+
+The ARGs in Fig.~\ref{fig-simplification} contain different
+numbers of local trees ($6$, $5$, $5$ and $4$ respectively for A through
+D). When we move from A to B the local trees
+for the intervals $[44,61)$ and $[61,87)$ are merged because
+the only differences between them are their paths through
+nodes \noderef{l} and \noderef{m}. These nodes that participated
+in the diamond are removed from the ARG, and we have lost
+all information about the corresponding recombination at
+position 61. Other nodes (e.g.\ \noderef{o} and \noderef{p})
+have also been removed but these represent the \emph{parents}
+of recombinants. The recombinant nodes themselves
+(e.g. \noderef{n}) are still present, and represent precise
+information about the time, genomic location and
+lineages involved in the recombination event.
+
+Fig.~\ref{fig-simplification}C has the same number of local trees
+as Fig.~\ref{fig-simplification}B, but has less precise information
+about recombination. Continuing the previous example, node
+\noderef{n} has been removed from the graph because it was unary
+in all of the local trees; its outbound edges to \noderef{s}
+and \noderef{q} have effectively been ``pushed down''
+to \noderef{e} (which is retained because it is the coalescent
+parent of \noderef{a} and \noderef{b} over the interval
+$[44, 100)$). We
+have therefore lost precision about
+the \emph{timing} of this recombination event, and know only
+that it must have occurred between the times of node \noderef{e}
+and \noderef{q}.
+
+Fig.~\ref{fig-simplification}D removes all unary nodes from the
+local trees, and further reduces the precision of
+recombination information. Node \noderef{e} has not been
+removed from the graph because it is coalescent in the
+final tree, but we no longer know that the recombination
+event at position 30 was ancestral to it, or have
+any indication of its timings. Furthermore,
+trees for $[44, 87)$ and $[87, 100)$ were only distinguishable
+by the passage of the former tree through nodes \noderef{e}
+and \noderef{q}, and so the recombination on node \noderef{g}
+at position 87 has been lost entirely.
+
+\section{Example inferred ARGs}
+\label{sec-example-inferred-args}
+The scalability gains made by recent ARG inference methods
+such as \relate~\citep{speidel2019method} and
+\tsinfer~\citep{kelleher2019inferring}
+have been, in part, due to inferring lower levels of precision about
+recombination than classical methods.
+Neither method infers
+explicit recombination events, and therefore their outputs cannot
+be described using the classical eARG formalisms (Section~\ref{sec-eARG}).
+% Could do, something like
+% this mismatch has led some to characterise their output as a
+% series of unrelated local trees [cite], which lacks nuance of the reality.
+% Or maybe we don't bother: what's to be gained from pointing out people's
+% mistakes here? Hopefully the rich detail of reality will come through
+% and replace old ideas.
+Nonetheless, both methods produce estimates in which nodes and edges
+persist across multiple trees, creating
+inheritance graphs which fit naturally into the gARG formulation.
+To illustrate the varying levels of information captured
+by current methods, and some qualitative differences between them,
+Fig.~\ref{fig-inferred-args} shows graphical depictions of
+example ARGs produced by four tools using substantially
+different inference strategies.
+
+THE FIGURE OF INFERRED ARGS WAS HERE
+
+The first two methods explicitly infer recombination events.
+\kwarg~\citep{ignatieva2021kwarg} is a parsimony based approach
+which searches the space of plausible ARGs,
+outputting minimal ones using heuristics.
+\argweaver~\citep{rasmussen2014genome} on the other hand is
+model-based, sampling from a discretised version of the
+% could spell this out, but in reality everyone knows what the SMC
+% is, and if they don't spelling it out won't help them.
+SMC~\citep{mcvean2005approximating,marjoram2006fast}.
+Note that both \kwarg\ and \argweaver\ produce many ARGs,
+and those shown in Fig.~\ref{fig-inferred-args} are arbitrarily selected
+examples.
+While the second two methods
+both produce a single best-guess estimate and
+do not explicitly infer recombination events, they are based on
+quite different principles.
+\texttt{Tsinfer} works in a two-step process, first generating ancestral
+haplotypes via heuristics and then inferring inheritance relationships
+between them using the Li and Stephens model~\citep{li2003modeling}.
+\texttt{Relate} first reconstructs local tree topologies across the genome,
+using a variant of the Li and Stephens model to estimate the
+ordering of coalescence events in each tree, and then estimates
+branch lengths using MCMC with a coalescent-based prior.
+See Appendix~\ref{sec-survey-arg-infer} for more details on these
+and other inference methods.
+
+Inferred ARGs are based on the \citet{kreitman1983nucleotide} dataset,
+a standard benchmark in the classical ARG
+literature.
+It consists of 43 biallelic SNPs spanning 2.4Kb of the \emph{D.~melanogaster} ADH
+locus on chromosome 2L. Where required for inference purposes we assume
+mutation and recombination rates of $5.49\times10^{-9}$
+and $2.40463\times10^{-9}$ per site per generation
+\citep{schrider2013rates,comeron2012many}
+and a constant effective population size of 1,720,600 \citep{li2006inferring},
+as provided by the \texttt{stdpopsim}
+catalog~\citep{adrion2020community,lauterbur2023expanding}.
+Software versions were \kwarg~v1.0, \argweaverD~(2019),
+\tsinfer~v0.3.1, and \relate~v1.1.9.
+Full details and code for generating
+these figures are available on GitHub (see Data Availability).
+
+% Here we describe recombination features
+Considering Fig.~\ref{fig-inferred-args}, we can see that
+there is substantial variation in the number of recombination breakpoints
+inferred by different methods, with e.g.~\argweaver\ suggesting
+far more than the 7 required for this dataset under minimal parsimony
+assumptions~\citep{song2003parsimonious}.
+A sense of the amount of recombination in each ARG is provided by
+the node colouring scheme, which shows the number of parents for each node.
+In Fig.~\ref{fig-inferred-args}A,B, each recombination event
+corresponds to a node with exactly two parents and one child.
+As these methods explicitly infer a recombination event for each
+breakpoint, the number of breakpoints equals the number of two-parent
+(brown) nodes.
+In contrast, Fig.~\ref{fig-inferred-args}C,D do not have this
+straightforward
+relationship between the number of nodes with multiple parents
+and number of breakpoints along the genome.
+In both ARGs the number of breakpoints is smaller
+than the number of multiple-parent ARG nodes, showing
+that several multiple-parent nodes must share breakpoint positions.
+There are also ARG nodes with multiple parents and multiple children,
+where one or more recombinations have been pushed down onto a
+more recent node.
+As a consequence, it may be difficult to condense each transition between trees
+in these ARGs into a set of SPR operations.
+
+% Here we move on to coalescence
+Shading within nodes in Fig.~\ref{fig-inferred-args} indicates the
+fraction of the node's span over which it is coalescent
+(Section~\ref{sec-locally-unary-edges}). For example, brown nodes
+in Fig.~\ref{fig-inferred-args}A,B are clear because there is
+no local coalescence at these recombination nodes
+(they are ``ARG unary'', and so local coalescence is impossible).
+The significant number of partially shaded nodes in
+Fig.~\ref{fig-inferred-args}A,B and C demonstrates that
+the \kwarg, \argweaver\ and \tsinfer\ ARGs all contain locally unary nodes.
+Another difference between methods highlighted in this figure
+is the presence of polytomies, which only
+\tsinfer\ creates. The most obvious example
+involves nodes \noderef{Fr-F}, \noderef{Wa-F}, and \noderef{Af-F},
+which happen to have identical sequences.
+Because \kwarg, \argweaver, and \relate\ require
+bifurcating trees by design, each picks an arbitrary order of branching
+(hence Fig.~\ref{fig-inferred-args}A and B disagree
+in this order, and Fig.~\ref{fig-inferred-args}D even shows different
+orders in different trees).
+% JK: I don't quite get the commented out sentence below, so readers probably
+% won't either. Let's stick to a few clear messages.
+% More generally, the number of nodes in Fig.~\ref{fig-inferred-args}C
+% that have more than two children hints at considerable uncertainty in
+% the order of branching elsewhere in the ARG.
+
+The bottom row of Fig~\ref{fig-inferred-args} shows the extent along
+the genome to which graph edges are shared between multiple trees.
+All four methods infer nodes and edges that are shared between
+multiple trees, to varying degrees.
+For example, all of the methods infer
+infer that \noderef{Af-f}, \noderef{Fr-f}, and \noderef{Wa-f}
+form a clade along the entire sequence.
+In particular, we can
+see both \tsinfer\ and (to a lesser extent) \relate\ have edges
+that span multiple tree boundaries, indicating that they are not inferring
+a series of \emph{unrelated} local trees.
+However, in comparison to \kwarg\ and \argweaver\ neither
+method results in extensive node sharing in the
+oldest time periods.
+Overall, Fig.~\ref{fig-inferred-args} shows
+that \tsinfer\ and \relate\ ARGs contain a level of detail
+that lies somewhere between a sequence of unrelated local trees on one extreme
+and an ARG with precisely specified recombination events on the other
+(Fig~\ref{fig-inferred-args}A,B).
+% JK: Let's not get into this.
+% This primarily reflects the fact that they do not encode topology
+% above the local MRCAs, also indicated by multiple roots in the ARG
+% (nodes with zero parents, in blue), compared to the
+% single ARG root in the \kwarg\ and \argweaver\ genealogies. In terms of inference,
+% \tsinfer\ and \relate\ show a pragmatic representation: we do not expect
+% genetic variation to be very informative about topological relationships
+% between nodes that that are older than the local MRCAs.
+
+\section{Implementation and efficiency}
+\label{sec-efficiency}
+The gARG encoding defined here
+leads to highly efficient storage and processing of ARG data,
+and has already been in use for several years.
+The succinct tree sequence data structure
+(usually known as a ``tree sequence'' for brevity)
+is a practical gARG implementation focused on efficiency.
+It was originally developed as part of the \texttt{msprime}
+simulator~\citep{kelleher2016efficient} and has subsequently been
+extended and applied to forward-time
+simulations~\citep{kelleher2018efficient,haller2018tree},
+inference from
+data~\citep{kelleher2019inferring,wohns2022unified,zhan2023towards},
+and calculation of population genetics statistics~\citep{ralph2020efficiently}.
+The succinct tree sequence encoding extends the basic definition
+of a gARG provided here by stipulating a
+simple tabular representation of nodes and edges,
+and also defining a concise representation of
+sequence variation using the ``site'' and  ``mutation'' tables.
+The key property of the succinct tree sequence encoding
+that makes it an efficient substrate for defining analysis
+algorithms is that it allows us to sequentially
+recover the local trees along the genome very efficiently,
+and in a way that allows us to reason about the \emph{differences}
+between those trees~\citep{kelleher2016efficient,ralph2020efficiently}.
+
+The \texttt{tskit} library is a liberally
+licensed open source toolkit that provides a comprehensive suite
+of tools for working with gARGs (encoded as a succinct tree sequence).
+Based on core functionality written
+in C, it provides interfaces in C, Python and Rust.
+Tskit is mature software, widely used in population genetics, and
+has been incorporated into numerous downstream
+applications~\citep[e.g.,][]{haller2019slim,speidel2019method,
+adrion2020community,
+terasaki2021geonomics,
+baumdicker2021efficient,
+fan2022genealogical,
+guo2022recombination,
+korfmann2022weak,
+mahmoudi2022bayesian,
+petr2022slendr,
+rasmussen2022espalier,
+zhang2023biobank,
+nowbandegani2023extremely,
+ignatieva2023distribution,
+fan2023likelihood}.
+The technical details of \texttt{tskit}, and how it provides an
+efficient and portable platform for ARG-based analysis, are beyond
+the scope of this manuscript.
+In the interest of avoiding confusion, however, we list a
+few minor details in which the formal details of gARGs
+provided in Section~\ref{sec-gARG} differ from their practical implementation in
+\texttt{tskit}.
+Firstly, ``edges'' in tree sequence terminology would perhaps be better
+described as ``edge-intervals'', as each describes a single contiguous
+interval of genome inheritance between a pair of nodes.
+This denormalisation of the gARG data model is for efficiency purposes.
+Secondly, zero- rather than one-based indexing is used for
+nodes in ARGs and oriented trees; consequently  $-1$ is used to denote the presence of
+roots (rather than $0$ as used here for notational simplicity).
+
 \section{Cell lineages and ARGs}

### 2024-04-04 paper.tex
Edits from Yan and Gregor

@@
 %are insufficient to describe the outputs of current methods.
-Moreover, we argue that the underlying assumption
-that all events can be known and precisely estimated
-is fundamentally unsuited to the realities
+We argue that classical ARG formulations, which assume
+that all events can be known and precisely estimated,
+are fundamentally unsuited to the realities
 of modern, population-scale datasets.
-We propose an alternative mathematical formulation that
+We describe an alternative formulation that
 encompasses the outputs of recent methods and
@@
 % making it a parenthetical aside.
-(Note that it is important to distinguish here between
+It is important to distinguish here between
 structures and models: whether an inference method is based
@@
 approximate ARG under a model
-such as the coalescent.)
+such as the coalescent.
 Although the term ``ARG'' is now often used in a general
@@
 
-We address this problem by providing a simple formal definition of an
+We highlight this problem and show how to address it by a simple formal definition of an
 ARG data structure, based on recording the intervals of
@@
 different types of approximation, in particular allowing us to systematically
-describe uncertainty about the temporal ordering of multiple recombinations. It
-is important to note that throughout we are interested in the details of these
-competing mathematical formulations and their practical consequences.
-% True ARGs, being perfectly resolved, can ultimately be described in
-% either formalism
+describe uncertainty about the temporal ordering of multiple recombinations.
 
@@
 % % with representation details
-It is important to note that here, and throughout,
-we are not questioning the form of the
-actual ancestral processes that occur in nature, but rather how we
-\emph{represent} the outcomes of such processes in a practical manner.
+In the next section we contrast this gARG representation with a classical event-based (eARG) representation of the same underlying process.
+The choice between using a gARG or an eARG to represent the general process of ancestral inheritance has
 % Again, need to remind the reader that there's an actual point
 % to all this getting stuck into the details.
-These practical details,
-as demonstrated in later sections, have important
-consequences not only for how methods exchange information about
-simulated and inferred ARGs, but more fundamentally in
-how we set our goals for inference and evaluate the success of results.
+important
+consequences not just when addressing the practical need to store and compare ARGs, but more fundamentally when establishing goals for inferring ARGs and evaluating the success of such inferences.
 
@@
 \label{sec-eARG}
-In this section we define a common (and classical) view of an ARG data structure,
+A common, classical view of an ARG data structure,
 described explicitly in several publications~\citep[e.g.][]{
-wiuf1999recombination,gusfield2014recombinatorics,hayman2023recoverability}
+wiuf1999recombination,gusfield2014recombinatorics,hayman2023recoverability},
+interprets nodes not as genomes but as historical \emph{events}
 (but see~\citet{parida2011minimal} and \citet{zhang2023biobank}
-for notable exceptions), and used as an output format by multiple ARG inferences tools
+for notable exceptions).
+This eARG encoding is the basis of the output formats created by multiple ARG inference tools
 \citep[e.g.][]{song2004minimum,song2005efficient,rasmussen2014genome,
@@
 % down.
-In this encoding we have two types of internal node in the graph,
-representing the common ancestor and recombination events
+In this case, two types of internal node exist in the graph. These
+represent the common ancestor and recombination events
 in the history of a sample.
@@
 recombination nodes based on the breakpoint
-position~\citep{griffiths1996ancestral}.
-Because it is focused on recording events and their properties,
-we will refer to this encoding as the ``event ARG'' or eARG
-encoding. Fig.~\ref{fig-event-arg} shows an example of an
+position~\citep{griffiths1996ancestral}. Fig.~\ref{fig-event-arg} shows an example of an
 eARG with three sample genomes (\noderef{a}, \noderef{b}, and \noderef{c}),
@@
 implemented literally. The eARG encoding explicitly models only
-two different types of event and thus anything that is not a single crossover
-recombination or common ancestor event, must be incorporated
+two different types of event; thus anything that is not a single crossover
+recombination or common ancestor event must be incorporated
 either in a roundabout way using these
@@
 find a common ancestor in \noderef{f}. Thus, in this region,
-the total number of genome segments that are ancestral to the
-sample is reduced from four to three. Fig.~\ref{fig-arg-in-pedigree}A
+we have three genome segments that are ancestral to the
+four samples. Fig.~\ref{fig-arg-in-pedigree}A
 illustrates this by (shaded) ancestral material being present
@@
 leading to a rapid build-up of information, much of which
-will not be relevant to the genetic ancestry of the current population.
+will not be relevant to the genetic ancestry of a future population.
 This redundancy is periodically removed using the ``simplify''
@@
 the direct genetic relationship between a parent and child genome and are not
-necessarily minimal in terms of the resolved ancestral material of a sample.
-Thus, the final step is to use the ``simplify'' algorithm to perform the required
-sample resolution (Fig.~\ref{fig-ancestry-resolution}C).
+necessarily minimal in terms of the ancestral material of a sample.
+Thus, the final step is to use the ``simplify'' algorithm to resolve the
+ancestral material of the samples (Fig.~\ref{fig-ancestry-resolution}C).
 
@@
 
-3 PARAS. - enough outline to let readewr know what's in the appendices.
-
-There are lots of new inference techniques: they all produce slightly different
-structures, all of which can be though of a types of ARGs...
+There are a number of tools for inferring genetic genealogies,
+and the graph structures that they produce differ in important aspects.
+These varying representations have all been termed ``ARGs'' by different authors,
+and each describes more than a simple sequence of uncorrelated trees.
+By way of illustration, Fig.~\ref{fig-inferred-args} depicts the output of four
+different ``ARG inference'' tools run on a common dataset,
+with the resulting ARGs converted into a standard gARG format (cite tskit).
 
@@
 Results for four different methods:
-(A) \kwarg; (B) \argweaver; (C) \tsinfer; and (D) \relate. See the
-text for details of these methods.
+(A) \kwarg; (B) \argweaver; (C) \tsinfer; and (D) \relate, converted to the standard \emph{tskit} gARG encoding. See
+Appendix \ref{sec-example-inferred-args} for details of these methods.
 Edge colours indicate time of the edge's child node
@@
 
-These give different levels of precision about recombination (see new Appendix X)
-
-You can see this as different levels of detail on the same underlying truth. Here's a figure:
+As is well-known (cite, perhaps Gusfield?), many different ancestral graph topologies are compatible with the same resulting
+sequence data. Hence it is unsurprising that the four graphs differ,
+for example in the numbers of nodes and their relationship to each other.
+However, more detailed inspection of Fig.~\ref{fig-inferred-args} reveals a
+diversity of representation that is only possible via the gARG encoding.
+For instance, although the most parsimonious ARG in Fig.~\ref{fig-inferred-args}A
+has far fewer ancestral events than an ARG instance sampled
+from from the coalescent with recombination in Fig.~\ref{fig-inferred-args}B,
+both consist only of nodes with either two parents or two children (but not both).
+These nodes correspond to recombination and common ancestor events in the equivalent eARG.
+In contrast, more scalable heuristic inference methods (Fig.~\ref{fig-inferred-args}C and D)
+create ARGs in which a node can simultaneously have more that than two children and more than two parents.
+Note also that the first three methods infer nodes that have a
+``coalescence span'' of less than 100\%, indicating that in the local trees, ancestral lineages exist 
+that pass through these nodes without the nodes marking a coalescent point. Such ``locally-unary''
+nodes are an notable feature of the local trees defined by an ARG
+(see Appendix \ref{sec-locally-unary-edges}).
+
+These differences, in particular the number of parents and children, and the coalescence span, reflect the
+degree of precision represented in the inferred ARGs. Indeed, it is important to note that the gARG encoding
+allows a single underlying genetic genealogy to be represented at
+differing levels of detail. The concept is illustrated in Fig.~\ref{fig-simplification},
+which shows different levels of ``simplification'' (Appendix~\ref{sec-ARG-simplification}) of the
+same underlying simulated ARG. 
+The full ARG, with all coalescent and recombination events represented by
+separate genomes, is shown in Fig.~\ref{fig-simplification}A. Simpler representations can be formed by removing ``unknowable'' nodes (Fig.~\ref{fig-simplification}B) and 
+collapsing multiple recombinations into a single child or multiple coalescences into
+a single parent (Fig.~\ref{fig-simplification}C, also see Appendix~\ref{sec-cell-lineages-and-args}).
+Finally, if we chose to represent only local coalescence points (Fig.~\ref{fig-simplification}D),
+we end up with a ``fully simplified'' ARG
+of the sort inferred in Fig.~\ref{fig-inferred-args}D. Note that in all cases, the equivalent
+trees only differ in the locally-unary nodes they contain; topological relationships
+between the samples remain identical.
 
@@
 
-The timing, positions, and even the number of recombination events is generally
-not possible to infer precisely from genome sequencing data. Under
+The ability to represent an ARG to differing degrees of precision is a powerful
+feature of gARGs. In particular, when inferring ARGs from genome
+sequencing data, the timing, positions, and even the number of recombination
+events is generally not possible to infer precisely. For example, under
 coalescent-based models, the proportion of recombination events that change the
@@
 position can only be inferred approximately, depending on how much information
-can be elucidated from mutations in the surrounding genomic region.
+can be elucidated from mutations in the surrounding genomic region. Because a gARG
+can encode a diversity of ARG structures, it
+allows this fundamental uncertainty in inference to be appropriately represented
+(Appendix~\ref{sec-precision}.
 % For general
@@
 % even more challenging.
-The fact that the eARG encoding \emph{requires}
-precise information about recombination is therefore a fundamental limitation.
+%The fact that the eARG encoding \emph{requires}
+%precise information about recombination is therefore a fundamental limitation.
 
@@
 A necessary first step is a degree of terminological clarity.
-As discussed in Appendix~\ref{sec-arg-history}, the term
+As reviewed in Appendix~\ref{sec-arg-history}, the term
 ``ancestral recombination graph" has several
@@
 processes and instead use the term as a more general representation of any
-recombinant genetic ancestry seems useful; we have
-tried to clarify and systematise it here. Thus
+recombinant genetic ancestry seems useful.
+Our intent here is to clarify and systematise this decoupling. Thus
 we can think of an ARG as any structure that encodes the
@@
 the influence of recombination. The ``genome'' ARG (gARG) encoding
-made explicit here is one way we can concretely
+is one way we can concretely
 define such recombinant ancestry, which we have shown is both
@@
 using \texttt{htslib}~\citep{bonfield2021htslib},
-and it this shared use of community software
+and it is this shared use of community software
 infrastructure that guarantees the smooth flow of data between
@@
 In particular this includes
-the ARG used in Fig.~\ref{fig-ancestry-resolution};
+the ARG used in Fig.~\ref{fig-ancestry-resolution} as well as
 the simulation code and functions
-used to generate Fig.~\ref{fig-simplification};
-and for Fig.~\ref{fig-inferred-args},
-the software versions, parameter settings, and
+used to generate Fig.~\ref{fig-inferred-args} and Fig.~\ref{fig-simplification}.
+This includes the software versions, parameter settings, and
 (where necessary) functions to convert
@@
 branching points (such as \noderef{e} in Fig.~\ref{fig-arg-in-pedigree}),
-we lose information about the ARG.
-We return to this point in the following two sections,
-when we discuss ``locally unary'' nodes and the simplification process.
-
-It is important that we make the distinction here between the local
-trees that we can derive from a known ARG (as just discussed),
-and an ARG that we can derive from a sequence of \emph{estimated}
+information about the ARG is lost.
+We expand on this point in Appendices~\ref{sec-locally-unary-edges} and
+\ref{sec-ARG-simplification}
+where we which discuss ``locally unary'' nodes and the simplification process.
+
+It is important to make the distinction here between the local
+trees that can be derived from a known ARG (as just discussed),
+and an ARG that can be derived from a sequence of \emph{estimated}
 local trees.
@@
 \label{sec-locally-unary-edges}
-As discussed in the previous section, the local tree at a given position
+As discussed in Appendix~\ref{sec-ARG-and-local-trees}, the local tree at a given position
 $x$ is best seen as the path through the ARG at that position, defined
@@
 a single child: those that are \emph{unary}.
-Returning to the example in Fig.~\ref{fig-arg-in-pedigree}, nodes
+For the example in Fig.~\ref{fig-arg-in-pedigree}, nodes
 \noderef{g} and \noderef{h}
@@
 interval: it is \emph{locally unary}.
-In another example, \noderef{e} is binary in the graph, being a common
+In the same figure, \noderef{e} is binary in the graph, being a common
 ancestor of \noderef{a} and \noderef{c}, but is
@@
 of many different types of ARG.
-As discussed in the previous section,
+As discussed in Appendix~\ref{sec-ARG-and-local-trees},
 without these nodes marking the passage
@@
 % would just make it less readable, IMO
-revolves around the presence of unary nodes (see previous section).
-We illustrate this successive removal of redundancy
-through a series of simplification steps
+revolves around the presence of unary nodes (Appendix~\ref{sec-locally-unary-edges}).
+This successive removal of redundancy
+through a series of simplification steps is shown
 in Fig.~\ref{fig-simplification}.
@@
 recombination events that we store and
-can hope to infer from sampled genomes, which is the topic of the next
-section.
+can hope to infer from sampled genomes, which is the topic of
+Appendix~\ref{sec-precision}.
 
@@
 different inference strategies.
-
-THE FIGURE OF INFERRED ARGS WAS HERE
 
@@
 and meiosis leads to recombination events (both crossover and gene conversion).
-Fig.~\ref{fig-simplification} shows a schematic of the events  and
+Fig.~\ref{fig-cell-lines} shows a schematic of the events  and
 the genomes (chromosome icons) that occur in the cellular
@@
 two node representation seen in
-Figs.~\ref{fig-simplification} and \ref{hudson_vs_bigARG},
+Fig.~\ref{fig-simplification}A and Fig.~\ref{hudson_vs_bigARG},
 whereas tagging the genome

### 2024-04-04 paper.tex
Updated references

@@
 This rapid progress has led to a diversity of ARG definitions and representations.
-
-We show that a synthesis of these views can be achieved by thinking
-in terms of genomes and transmission
-intervals.
+% We show that a synthesis of these views can be achieved by thinking
+% in terms of genomes and transmission
+% intervals.
 % Standard mathematical formalisms,
@@
 ancestry
-(see \citet{lewanski2023era} for a biologically oriented introduction).
+(see \citet{lewanski2024era,brandt2024promise} for biologically oriented
+introductions).
 Recent breakthroughs
@@
 methods~\citep{rasmussen2014genome,kelleher2019inferring,speidel2019method,
-schaefer2021ancestral,wohns2022unified,zhang2023biobank,zhan2023towards}
+schaefer2021ancestral,wohns2022unified,zhang2023biobank,zhan2023towards,deng2024robust}
 have raised the realistic prospect of ARG-based analysis becoming a standard part
@@
 ignatieva2023distribution,
-fan2023likelihood}
+fan2023likelihood,
+grundler2024geographic,
+huang2024estimating}
 and many more are sure to
@@
 mahmoudi2022bayesian,
-petr2022slendr,
+petr2023slendr,
 rasmussen2022espalier,
@@
 ignatieva2023distribution,
-fan2023likelihood}.
+fan2023likelihood,
+tsambos2023link,
+tagami2024tstrait}.
 The technical details of \texttt{tskit}, and how it provides an

### 2024-04-04 paper.tex
Move implementation and efficiency back into main text
Partial recordering of appendixes

@@
 The breakthroughs in scale achieved by recent
-methods~\citep[e.g.][]{kelleher2019inferring,speidel2019method,zhang2023biobank}
-are all based, in different ways, on inferring
+methods are all based, in different ways, on inferring
 approximate \emph{structures}
-instead of a complete and fully detailed history.
-% This is a digression from the maing point of the paragraph, so
-% making it a parenthetical aside.
+instead of a complete and fully detailed history
+(see Appendix~\ref{sec-survey-arg-infer} for a brief survey).
 It is important to distinguish here between
@@
 by the two recombination breakpoints that gave rise to these genomes;
-see Section~\ref{sec-ARG-and-local-trees} for details
+see Appendix~\ref{sec-ARG-and-local-trees} for details
 on how local trees are embedded in an ARG.
@@
 % down.
-In this case, two types of internal node exist in the graph. These
+In this encoding there are two types of internal node in the graph. These
 represent the common ancestor and recombination events
@@
 % discussions won't help.
-The gARG format avoids the need for such considerations,
-so is more naturally compatible with forming an ecosystem of interoperable
-inference and analysis methods.
+% The gARG formulation avoids the need for such considerations,
+% so is more naturally compatible with forming an ecosystem of interoperable
+% inference and analysis methods.
 
@@
 %precise information about recombination is therefore a fundamental limitation.
+
+\section{Implementation and efficiency}
+\label{sec-efficiency}
+The gARG encoding defined here
+leads to highly efficient storage and processing of ARG data,
+and has already been in use for several years.
+The succinct tree sequence data structure
+(usually known as a ``tree sequence'' for brevity)
+is a practical gARG implementation focused on efficiency.
+It was originally developed as part of the \texttt{msprime}
+simulator~\citep{kelleher2016efficient} and has subsequently been
+extended and applied to forward-time
+simulations~\citep{kelleher2018efficient,haller2018tree},
+inference from
+data~\citep{kelleher2019inferring,wohns2022unified,zhan2023towards},
+and calculation of population genetics statistics~\citep{ralph2020efficiently}.
+The succinct tree sequence encoding extends the basic definition
+of a gARG provided here by stipulating a
+simple tabular representation of nodes and edges,
+and also defining a concise representation of
+sequence variation using the ``site'' and  ``mutation'' tables.
+The key property of the succinct tree sequence encoding
+that makes it an efficient substrate for defining analysis
+algorithms is that it allows us to sequentially
+recover the local trees along the genome very efficiently,
+and in a way that allows us to reason about the \emph{differences}
+between those trees~\citep{kelleher2016efficient,ralph2020efficiently}.
+
+The \texttt{tskit} library is a liberally
+licensed open source toolkit that provides a comprehensive suite
+of tools for working with gARGs (encoded as a succinct tree sequence).
+Based on core functionality written
+in C, it provides interfaces in C, Python and Rust.
+Tskit is mature software, widely used in population genetics, and
+has been incorporated into numerous downstream
+applications~\citep[e.g.,][]{haller2019slim,speidel2019method,
+adrion2020community,
+terasaki2021geonomics,
+baumdicker2021efficient,
+fan2022genealogical,
+guo2022recombination,
+korfmann2022weak,
+mahmoudi2022bayesian,
+petr2023slendr,
+rasmussen2022espalier,
+zhang2023biobank,
+nowbandegani2023extremely,
+ignatieva2023distribution,
+fan2023likelihood,
+tsambos2023link,
+tagami2024tstrait}.
+The technical details of \texttt{tskit}, and how it provides an
+efficient and portable platform for ARG-based analysis, are beyond
+the scope of this manuscript.
+In the interest of avoiding confusion, however, we list a
+few minor details in which the formal details of gARGs
+provided in Section~\ref{sec-gARG} differ from their practical implementation in
+\texttt{tskit}.
+Firstly, ``edges'' in tree sequence terminology would perhaps be better
+described as ``edge-intervals'', as each describes a single contiguous
+interval of genome inheritance between a pair of nodes.
+This denormalisation of the gARG data model is for efficiency purposes.
+Secondly, zero- rather than one-based indexing is used for
+nodes in ARGs and oriented trees; consequently  $-1$ is used to denote the presence of
+roots (rather than $0$ as used here for notational simplicity).
 
@@
 
-\section{The Big and Little ARG}
-\label{sec-big-and-little-arg}
-Here we review two important stochastic processes that construct ARGs:
-the ``Big" ARG process of \cite{griffiths1997ancestral}, and the ``Little" ARG process of
- \cite{hudson1983properties}. The Big ARG process is mathematically simpler
- but is computationally intractable due to generating a vast number of ancestors
- which contribute no genetic material to the initial sample.
-The Little ARG process avoids non-genetic ancestors at the cost of more complex
-dynamics and state space.
-We also demonstrate that applications relying on the grouping of inheritance pathways into ancestral lineages, such as likelihood-based inference under the coalescent, requires that the
-gARG (or eARG) data structure be interpreted in a model-specific way.
-
-A generic state of the Little ARG process consists of a finite collection of lineages $L$,
-each of which is a list of disjoint ancestry segments $(\ell, r, a)$, where
-$[\ell, r)$ is a half-closed genomic interval and $a$ is an integer
-tracking the number of samples to which the lineage is ancestral over that interval.
-We also usually track the node associated with each segment, but
-that is not important for our purposes here so we omit it to lighten notation.
-The initial condition for a sample of $n$ genomes of length $m$ consists of $n$ lineages
-of the form $\{(0, m, 1)\}$. The process traverses a series of common ancestor and
-recombination events backwards in time.
-Recombination events happen at rate $\rho \nu / (m - 1)$,
-where $\rho \geq 0$ is a per-genome recombination rate and
- \[
- \nu = \sum_{x \in L}\left( \max_{(\ell, r, a) \in x}r
-     - \min_{(\ell, r, a) \in x}\ell - 1 \right)
- \]
- is the number of available ``links" surrounded by ancestral material.
- At a recombination event we choose one of these links uniformly and break it,
- replacing the original lineage in $L$ with two new lineages containing the ancestral material
- to the left and right of the break point, respectively.
-
-Common ancestor events occur at rate $\binom{|L|}{2}$.
-In a common ancestor event, two uniformly sampled lineages have their segments
-merged into a single ancestor lineage, which is added to $L$.
-If the lineages have overlapping intervals of ancestry,
-say, $(\ell, r, a_1)$ and $(\ell, r, a_2)$, a
-\emph{coalescence} occurs. The result is a segment
-$(\ell, r, a_1 + a_2)$, and if $a_1 + a_2 < n$ it is included in the
-ancestor lineage. Otherwise, if $a_1 + a_2 = n$, we have found
-the most recent common ancestor of all samples in the interval $[\ell, r)$
-and do not need to simulate its history any further.
-Non-overlapping intervals from the two lineages are included
- in the ancestor lineage without changes. Eventually,
-we find resultant lineages in which all segments have fully coalesced,
-and so the number of extant lineages gradually falls to zero.
-
-In the Griffiths formulation (the Big ARG process), each edge in the graph corresponds to an extant
-lineage and nodes are events in the process. The $n$ initial leaf nodes are
-sampling events. Common ancestor events occur at rate $\binom{|L|}{2}$.
-When a common ancestor event happens, two uniformly chosen lineages
-merge into a common ancestor lineage.
-Recombination events happen at rate $|L| \rho$. Here, we choose a lineage (i.e.\ edge) uniformly,
-and a breakpoint $0 < x < m$ uniformly on its genome. We terminate the edge at a
-node, record the breakpoint, and start two new edges from this node. The process
-then continues until there is only one lineage left (the Grand Most Recent
-Common Ancestor, GMRCA), which is guaranteed to
-happen in finite time because of the quadratic rate of coalescing vs.\ linear rate of branching.
-
-The state-space of the Big ARG process is much simpler than that of the Little ARG process,
-which greatly facilitates mathematical reasoning. This simplicity comes at a
-substantial cost, however, if we wish to use it as a practical means of
-simulating recombinant ancestries.
-The number of events in the Big ARG all the way back to the GMRCA
-is $O(e^\rho)$~\citep{griffiths1997ancestral}, whereas the number
-of events required to simulate the Little ARG is
-$O(\rho^2)$~\citep{hein2004gene,baumdicker2021efficient}.
-This disparity arises because the majority of the events in the Big ARG are
-recombination events which occur outside of ancestral material,
-and these do not have any bearing on the ancestry of the initial sample.
-Because we don't keep track of the distribution of ancestral material during the process,
-we generate a vastly larger graph.
-
-\begin{figure}
-\centering
-% FIXME this is a quick nasty hack to make the figure a bit smaller and
-% prevent flushing all the other figures to the end of the document.
-\scalebox{0.54}{
-\begin{tabular}{cc}
-\begin{tikzpicture}
-	\node [anchor=north west] at (-3.5,9.3) {\LARGE \textbf{A}};
-	\draw (-0.4, -0.2) -- (0.4, -0.2) -- (0.4, -0.3) -- (-0.4, -0.3) -- (-0.4, -0.2);
-	\draw (-0.1, -0.1) -- (-0.1, -0.4);
-	\node [label=below:{\small $v$}] at (-0.1, -0.2) {};
-	\draw (3.6, -0.2) -- (4.4, -0.2) -- (4.4, -0.3) -- (3.6, -0.3) -- (3.6, -0.2);
-	\draw (5.6, -0.2) -- (6.4, -0.2) -- (6.4, -0.3) -- (5.6, -0.3) -- (5.6, -0.2);
-
-	\draw (4,0) -- (4, 1.2) -- (6, 1.2) -- (6,0);
-	\draw (4.6, 1) -- (5.4, 1) -- (5.4, 0.9) -- (4.6, 0.9) -- (4.6, 1);
-	\draw (5.1, 1.1) -- (5.1, 0.8);
-	\node [label=below:{\small $w$}] at (5.1, 1) {};
-
-	\draw (0, 0) -- (0, 1.7) -- (-1,1.7) -- (1,1.7);
-	\draw (-1.4, 1.5) -- (-1.1, 1.5) -- (-1.1, 1.4) -- (-1.4, 1.4) -- (-1.4, 1.5);
-	\draw (-1.1, 1.45) -- (-0.6, 1.45);
-	\draw (0.6, 1.45) -- (0.9, 1.45);
-	\draw (0.9, 1.5) -- (1.4, 1.5) -- (1.4, 1.4) -- (0.9, 1.4) -- (0.9, 1.5);
-	\draw (1.1, 1.6) -- (1.1, 1.3);
-	\node [label=below:{\small $w$}] at (1.1, 1.5) {};
-
-	\draw (5, 1.2) -- (5, 2.4) -- (4,2.4) -- (6,2.4);
-	\draw (3.6, 2.2) -- (4.1, 2.2) -- (4.1, 2.1) -- (3.6, 2.1) -- (3.6, 2.2);
-	\draw (4.1, 2.15) -- (4.4, 2.15);
-	\draw (5.6, 2.15) -- (6.1, 2.15);
-	\draw (6.1, 2.2) -- (6.4, 2.2) -- (6.4, 2.1) -- (6.1, 2.1) -- (6.1, 2.2);
-
-	\draw (1, 1.7) -- (1, 3.1) -- (0,3.1) -- (2,3.1);
-	\draw (-0.4, 2.85) -- (0.1, 2.85);
-	\draw (0.1, 2.9) -- (0.4, 2.9) -- (0.4, 2.8) -- (0.1, 2.8) -- (0.1, 2.9);
-	\draw (1.6, 2.85) -- (1.9, 2.85);
-	\draw (1.9, 2.9) -- (2.1, 2.9) -- (2.1, 2.8) -- (1.9, 2.8) -- (1.9, 2.9);
-	\draw (2.1, 2.85) -- (2.4, 2.85);
-
-	\draw (-1, 1.7) -- (-1, 4.5) -- (0,4.5) -- (0,3.1);
-	\node [scale=0.2,label=above left:{\small \noderef{a}}] at (-0.5,4.5) {a};
-	\draw (-0.9, 4.3) -- (-0.6, 4.3) -- (-0.6, 4.2) -- (-0.9, 4.2) -- (-0.9, 4.3);
-	\draw (-0.6, 4.25) -- (-0.4, 4.25);
-	\draw (-0.4, 4.3) -- (-0.1, 4.3) -- (-0.1, 4.2) -- (-0.4, 4.2) -- (-0.4, 4.3);
-	\draw (-0.5, 4.4) -- (-0.5, 4.1);
-	\node [label=below:{\small $x$}] at (-0.5, 4.3) {};
-
-	\draw (4,2.4) -- (4,5.4) -- (2,5.4) -- (2,3.1);
-	\node [scale=0.2,label=above left:{\small \noderef{b}}] at (3,5.4) {b};
-	\draw (2.6, 5.2) -- (2.9, 5.2) -- (2.9, 5.1) -- (2.6, 5.1) -- (2.6, 5.2);
-	\draw (2.9, 5.15) -- (3.4, 5.15);
-
-	\draw (-0.5, 4.5) -- (-0.5, 6) -- (-1.5,6) -- (0.5,6);
-	\draw (-1.9, 5.75) -- (-1.4, 5.75);
-	\draw (-1.4, 5.8) -- (-1.1, 5.8) -- (-1.1, 5.7) -- (-1.4, 5.7) -- (-1.4, 5.8);
-	\draw (0.1, 5.8) -- (0.4, 5.8) -- (0.4, 5.7) -- (0.1, 5.7) -- (0.1, 5.8);
-	\draw (0.4, 5.75) -- (0.9, 5.75);
-
-	\draw (0.5, 6) -- (0.5, 7.2) -- (3,7.2) -- (3,5.4);
-	\node [scale=0.2,label=above:{\small \noderef{c}}] at (1.75,7.2) {c};
-	\draw (1.35, 6.95) -- (2.15, 6.95);
-
-	\draw (-1.5, 6) -- (-1.5, 8);
-	\draw (6, 2.4) -- (6, 8);
-
-	% Edge annotations above each event
-	\node [label=left:{\small$\{(0,m,1)\}$}] at (0.2,0.6) {};
-	\node [label=left:{\small$\{(0,m,1)\}$}] at (4.2,0.4) {};
-	\node [label=right:{\small$\{(0,m,1)\}$}] at (5.8,0.4) {};
-	\node [label=left:{\small$\{(0,m,2)\}$}] at (5.2,1.6) {};
-	\node [label=left:{\small$\{(0,v,1)\}$}] at (-0.8,2.3) {};
-	\node [label=right:{\small$\{(v,m,1)\}$}] at (0.8,2.1) {};
-	\node [label=right:{\small$\{(0,w,2)\}$}] at (3.8,3.9) {};
-	\node [label=left:{\small$\{(w,m,2)\}$}] at (6.2,5) {};
-	\node [label=right:{\small$\{(w,m,1)\}$}] at (-0.2,3.8) {};
-	\node [label=right:{\small$\{(v,w,1)\}$}] at (1.8,4.1) {};
-	\node [label=left:{\small$\{(0,v,1), (w,m,1)\}$}] at (-0.3,5.2) {};
-	\node [label=right:{\small$\{(0,v,2)\}$}] at (2.8,6.4) {};
-	\node [label=left:{\small$\{(w,m,1)\}$}] at (-1.3,7) {};
-	\node [label=right:{\small$\{(0,v,1)\}$}] at (0.3,6.4) {};
-
-	% Dashed lines for start and end times
-	\draw[dashed] (-2, 0) -- (11, 0);
-	\node [label=left:{$t = 0$}] at (-2,0) {};
-	\draw[dashed] (-2, 8) -- (11, 8);
-	\node [label=left:{$t = T$}] at (-2,8) {};
-
-	% Numbers of extant ancestors and links, from top to bottom
-	\node[label=right:{CA \; RE}] at (7.5, 8.2) {};
-	\node[label=right:{$\binom{2}{2}$ \; $\frac{2 m - 2 w - 2}{m - 1} \rho$}] at (7.5, 7.6) {};
-	\node[label=right:{$\binom{4}{2}$ \; $\frac{2 m + 2 v - 2 w - 4}{m - 1} \rho$}] at (7.5, 6.6) {};
-	\node[label=right:{$\binom{3}{2}$ \; $\frac{2 m + v - w - 3}{m - 1} \rho$}] at (7.5, 5.7) {};
-	\node[label=right:{$\binom{4}{2}$ \; $\frac{2 m + w - v - 4}{m - 1} \rho$}] at (7.5, 4.95) {};
-	\node[label=right:{$\binom{5}{2}$ \; $\frac{2 m - 5}{m - 1} \rho$}] at (7.5, 3.8) {};
-	\node[label=right:{$\binom{4}{2}$ \; $\frac{2 m - 4}{m -1} \rho$}] at (7.5, 2.75) {};
-	\node[label=right:{$\binom{3}{2}$ \; $\frac{2 m - 3}{m - 1} \rho$}] at (7.5, 2.05) {};
-	\node[label=right:{$\binom{2}{2}$ \; $2 \rho$}] at (7.5, 1.45) {};
-	\node[label=right:{$\binom{3}{2}$ \; $3 \rho$}] at (7.5, 0.6) {};
-
-	% Gray dashed lines to visually separate holding times
-	\draw[color=gray, dashed] (7.5, 1.2) -- (11, 1.2);
-	\draw[color=gray, dashed] (7.5, 1.7) -- (11, 1.7);
-	\draw[color=gray, dashed] (7.5, 2.4) -- (11, 2.4);
-	\draw[color=gray, dashed] (7.5, 3.1) -- (11, 3.1);
-	\draw[color=gray, dashed] (7.5, 4.5) -- (11, 4.5);
-	\draw[color=gray, dashed] (7.5, 5.4) -- (11, 5.4);
-	\draw[color=gray, dashed] (7.5, 6) -- (11, 6);
-	\draw[color=gray, dashed] (7.5, 7.2) -- (11, 7.2);
-\end{tikzpicture}
-&
-\begin{tikzpicture}
-	\node [anchor=north west] at (-3.5,9.3) {\LARGE \textbf{B}};
-	\draw (-0.4, -0.2) -- (0.4, -0.2) -- (0.4, -0.3) -- (-0.4, -0.3) -- (-0.4, -0.2);
-	\draw (-0.1, -0.1) -- (-0.1, -0.4);
-	\node [label=below:{\small $v$}] at (-0.1, -0.2) {};
-	\draw (3.6, -0.2) -- (4.4, -0.2) -- (4.4, -0.3) -- (3.6, -0.3) -- (3.6, -0.2);
-	\draw (5.6, -0.2) -- (6.4, -0.2) -- (6.4, -0.3) -- (5.6, -0.3) -- (5.6, -0.2);
-
-	\draw (4,0) -- (4, 1.2) -- (6, 1.2) -- (6,0);
-	\draw (4.6, 1) -- (5.4, 1) -- (5.4, 0.9) -- (4.6, 0.9) -- (4.6, 1);
-	\draw (5.1, 1.1) -- (5.1, 0.8);
-	\node [label=below:{\small $w$}] at (5.1, 1) {};
-
-	\draw (0, 0) -- (0, 1.7) -- (-1,1.7) -- (1,1.7);
-	\draw (-1.4, 1.5) -- (-1.1, 1.5) -- (-1.1, 1.4) -- (-1.4, 1.4) -- (-1.4, 1.5);
-	\draw (-1.1, 1.45) -- (-0.6, 1.45);
-	\draw (0.6, 1.45) -- (0.9, 1.45);
-	\draw (0.9, 1.5) -- (1.4, 1.5) -- (1.4, 1.4) -- (0.9, 1.4) -- (0.9, 1.5);
-	\draw (1.1, 1.6) -- (1.1, 1.3);
-	\node [label=below:{\small $w$}] at (1.1, 1.5) {};
-
-	\draw (5, 1.2) -- (5, 2.4) -- (4,2.4) -- (6,2.4);
-	\draw (3.6, 2.2) -- (4.1, 2.2) -- (4.1, 2.1) -- (3.6, 2.1) -- (3.6, 2.2);
-	\draw (4.1, 2.15) -- (4.4, 2.15);
-	\draw (5.6, 2.15) -- (6.1, 2.15);
-	\draw (6.1, 2.2) -- (6.4, 2.2) -- (6.4, 2.1) -- (6.1, 2.1) -- (6.1, 2.2);
-	\draw [color=red](5.8, 2.3) -- (5.8, 2.0);
-	\node [label={[red]below:{\small $y$}}] at (5.8, 2.1) {};
-
-	\draw (1, 1.7) -- (1, 3.1) -- (0,3.1) -- (2,3.1);
-	\draw (-0.4, 2.85) -- (0.1, 2.85);
-	\draw (0.1, 2.9) -- (0.4, 2.9) -- (0.4, 2.8) -- (0.1, 2.8) -- (0.1, 2.9);
-	\draw (1.6, 2.85) -- (1.9, 2.85);
-	\draw (1.9, 2.9) -- (2.1, 2.9) -- (2.1, 2.8) -- (1.9, 2.8) -- (1.9, 2.9);
-	\draw (2.1, 2.85) -- (2.4, 2.85);
-
-	\draw (-1, 1.7) -- (-1, 4.5) -- (0,4.5) -- (0,3.1);
-	\draw (-0.9, 4.3) -- (-0.6, 4.3) -- (-0.6, 4.2) -- (-0.9, 4.2) -- (-0.9, 4.3);
-	\draw (-0.6, 4.25) -- (-0.4, 4.25);
-	\draw (-0.4, 4.3) -- (-0.1, 4.3) -- (-0.1, 4.2) -- (-0.4, 4.2) -- (-0.4, 4.3);
-	\draw (-0.5, 4.4) -- (-0.5, 4.1);
-	\node [label=below:{\small $x$}] at (-0.5, 4.3) {};
-
-	\draw (6, 2.4) -- (6, 3.8) -- (7, 3.8);
-	\draw [color=red](6, 3.8) -- (5, 3.8);
-	\draw (4.6, 3.55) -- (5.4, 3.55);
-	\draw (6.6, 3.55) -- (7.1, 3.55);
-	\draw (7.1, 3.6) -- (7.4, 3.6) -- (7.4, 3.5) -- (7.1, 3.5) -- (7.1, 3.6);
-
-	\draw (4,2.4) -- (4,5.4) -- (2,5.4) -- (2,3.1);
-	\draw (2.6, 5.2) -- (2.9, 5.2) -- (2.9, 5.1) -- (2.6, 5.1) -- (2.6, 5.2);
-	\draw (2.9, 5.15) -- (3.4, 5.15);
-
-	\draw (-0.5, 4.5) -- (-0.5, 6) -- (-1.5,6) -- (0.5,6);
-	\draw (-1.9, 5.75) -- (-1.4, 5.75);
-	\draw (-1.4, 5.8) -- (-1.1, 5.8) -- (-1.1, 5.7) -- (-1.4, 5.7) -- (-1.4, 5.8);
-	\draw (0.1, 5.8) -- (0.4, 5.8) -- (0.4, 5.7) -- (0.1, 5.7) -- (0.1, 5.8);
-	\draw (0.4, 5.75) -- (0.9, 5.75);
-
-	\draw [color=red](5, 3.8) -- (5, 6.6) -- (4, 6.6);
-	\draw (4,6.6) -- (3,6.6) -- (3, 5.4);
-	\draw (3.6, 6.4) -- (3.9, 6.4) -- (3.9, 6.3) -- (3.6, 6.3) -- (3.6, 6.4);
-	\draw (3.9, 6.35) -- (4.4, 6.35);
-
-	\draw (0.5, 6) -- (0.5, 7.2) -- (4,7.2) -- (4,6.6);
-	\draw (1.85, 6.95) -- (2.65, 6.95);
-
-	\draw (-1.5, 6) -- (-1.5, 8);
-	\draw [color=red](2.25, 7.2) -- (2.25, 8);
-	\draw (7, 3.8) -- (7, 8);
-
-	% Dashed lines for start and end times
-	\draw[dashed] (-2, 0) -- (9.1, 0);
-	\node [label=left:{$t = 0$}] at (-2,0) {};
-	\draw[dashed] (-2, 8) -- (9.1, 8);
-	\node [label=left:{$t = T$}] at (-2,8) {};
-
-	% Numbers of extant ancestors and links, from top to bottom
-	\node[label=right:{CA \; RE}] at (7.5, 8.2) {};
-	\node[label=right:{$\binom{3}{2}$ \; $3 \rho$}] at (7.5, 7.6) {};
-	\node[label=right:{$\binom{4}{2}$ \; $4 \rho$}] at (7.5, 6.9) {};
-	\node[label=right:{$\binom{5}{2}$ \; $5 \rho$}] at (7.5, 6.3) {};
-	\node[label=right:{$\binom{4}{2}$ \; $4 \rho$}] at (7.5, 5.7) {};
-	\node[label=right:{$\binom{5}{2}$ \; $5 \rho$}] at (7.5, 4.95) {};
-	\node[label=right:{$\binom{6}{2}$ \; $6 \rho$}] at (7.5, 4.15) {};
-	\node[label=right:{$\binom{5}{2}$ \; $5 \rho$}] at (7.5, 3.45) {};
-	\node[label=right:{$\binom{4}{2}$ \; $4 \rho$}] at (7.5, 2.75) {};
-	\node[label=right:{$\binom{3}{2}$ \; $3 \rho$}] at (7.5, 2.05) {};
-	\node[label=right:{$\binom{2}{2}$ \; $2 \rho$}] at (7.5, 1.45) {};
-	\node[label=right:{$\binom{3}{2}$ \; $3 \rho$}] at (7.5, 0.6) {};
-
-	% Gray dashed lines to visually separate holding times
-	\draw[color=gray, dashed] (7.5, 1.2) -- (9.1, 1.2);
-	\draw[color=gray, dashed] (7.5, 1.7) -- (9.1, 1.7);
-	\draw[color=gray, dashed] (7.5, 2.4) -- (9.1, 2.4);
-	\draw[color=gray, dashed] (7.5, 3.1) -- (9.1, 3.1);
-	\draw[color=gray, dashed] (7.5, 3.8) -- (9.1, 3.8);
-	\draw[color=gray, dashed] (7.5, 4.5) -- (9.1, 4.5);
-	\draw[color=gray, dashed] (7.5, 5.4) -- (9.1, 5.4);
-	\draw[color=gray, dashed] (7.5, 6) -- (9.1, 6);
-	\draw[color=gray, dashed] (7.5, 6.6) -- (9.1, 6.6);
-	\draw[color=gray, dashed] (7.5, 7.2) -- (9.1, 7.2);
-\end{tikzpicture}
-\end{tabular}
-}
-\caption{(A)
-A realisation of the graph traversed by Hudson's algorithm started from a
-sample of three chromosomes with $m$ discrete sites each at time $t = 0$, and
-propagated until time $T$. The MRCA on the genetic interval $[v, w)$ is reached
-at event \noderef{b}, while that on $[0, v)$ is reached at event \noderef{c}.
-The non-ancestral segment $[v, w)$ above
-\noderef{a} contributes to the rate of effective recombinations because it
-is trapped between ancestral segments. The two columns titled CA and RE
-are the respective rates of mergers and recombinations when
-the recombination rate is $\rho$.
-(B) A corresponding realisation of a Big ARG, which augments Hudson's algorithm
-by tracking non-ancestral lineages. The result is a simpler state space and
-dynamics, at the cost of extra nodes and edges, highlighted in red, which do
-not affect the local tree at any site.
-Recombination positions are labelled alphabetically in time, and their ordering
-along the genome is $y < v < x < w$, of which the first only appears in panel B.
-There are two separate recombination events at link $w$.
-}
-\label{hudson_vs_bigARG}
-\end{figure}
-
-Figure \ref{hudson_vs_bigARG} illustrates the more complex state space
-of the Little ARG process, as well as the extra events which occur in the Big ARG process.
-Moreover, it depicts the rates of common ancestors and recombination events in each
-interval of time of the realisations.
-In order to evaluate these rates, e.g.\ for likelihood-based inference
-\citep{baumdicker2021efficient, mahmoudi2022bayesian},
-it is necessary to know the number of lineages and number of extant links
-available for recombination in each time interval.
-Some representations may not provide this information.
-For example, in the gARG encoding depicted in Figure \ref{fig-ancestry-resolution}C,
-it is clear that a recombination takes place between nodes \noderef{i}, \noderef{k} and \noderef{j}.
-But the exact time of the recombination event is ambiguous: it could take place at any time between node \noderef{i} and its parents and produce the same gARG.
-Because a recombination increases the number of extant lineages by one (in the rootward direction of time), the number of lineages during the same time interval is ambiguous as well.
-In fact, this information cannot be recovered from the gARG encoding used in Figure \ref{fig-ancestry-resolution}C without an extrinsic convention.
-For the basic coalescent with recombination,
-it is sufficient to create \emph{two} gARG nodes at the time of the recombination event, with the interpretation that the two rootward edges from node \noderef{i} in Figure \ref{fig-ancestry-resolution}C belong to the same lineage until the time of nodes \noderef{k} and \noderef{j}, and split into two separate lineages at that time point.
-Similarly, the trapped, non-ancestral links along that lineage remain available for effective recombination (i.e.\ one which splits up ancestral material) for the same time interval.
-This interpretation is highlighted in
-Figure \ref{hudson_vs_bigARG} by drawing only one vertical edge between a recombinant child and its two parents.
-
-% Full quote:
-% This process simplifies mathematics on the account that the notion of an
-% ancestor will have a less restrictive meaning than usual:
-% An ``ancestral'' sequence in the birth and death process
-% need not have any genetic material in common with a
-% sequence descended from it.
-
 \section{Survey of ARG inference methods}
 \label{sec-survey-arg-infer}
-
 The problem of reconstructing ARGs for samples of recombining sequences has
@@
 % events and times explicitly, e.g.\ using nodes.
+
+\section{Cell lineages and ARGs}
+\label{sec-cell-lineages-and-args}
+\begin{figure}
+\begin{center}
+    \includegraphics[width=0.8\textwidth]{illustrations/cell-lines}
+\end{center}
+\caption{\label{fig-cell-lines}
+Cellular inheritance of a single chromosome in a diploid population.
+Individuals (blue) contain diploid cells (white circles enclosing a homologous pair of chromosomes).
+For clarity, only two rounds of mitotic germ-line cell division are shown per individual, and
+meiosis is not illustrated in detail.
+Lines show prospective inheritance paths for all chromosomes. Solid lines show all possible
+retrospective ancestry paths for four chosen chromosomes (indicated by square black ``sampling events'')
+sampled from 3 diploid individuals ($D_1$, $D_2$, $D_3$) in the current generation.
+Ancestral recombination events and coalescence events are shown as red and blue squares respectively.
+A realised ARG path for the lower arm of the sampled chromosomes is highlighted as a
+thick solid line, passing through a set of potential gARG nodes (green). This ARG
+involves a single recombination event and four coalescence events
+(highlighted as deep red and blue squares within individuals $D_5$, $D_{10}$, and $D_{13}$).
+ARG lineages also show gametic genomes, contained within shaded circles.
+As in Fig.~\ref{fig-arg-in-pedigree}A, inherited regions within the sampled chromosome arm are
+shaded by the number of descendant samples.
+}
+\end{figure}
+
+In eukaryotes, ARGs are a result of the cellular processes of mitosis and meiosis.
+Mitosis leads to common ancestor events,
+and meiosis leads to recombination events (both crossover and gene conversion).
+Fig.~\ref{fig-cell-lines} shows a schematic of the events  and
+the genomes (chromosome icons) that occur in the cellular
+germline of a simplified, diploid multicellular
+hermaphrodite eukaryote with partially overlapping generations.
+Here, an event is not represented by a specific
+genome. Rather, genomes can be associated with, or ``tag'', events above
+(ancestral to) or below (descended from) them.
+For example, tagging the two genomes above a recombination event leads to the
+two node representation seen in
+Fig.~\ref{fig-simplification}A and Fig.~\ref{hudson_vs_bigARG},
+whereas tagging the genome
+below a recombination event leads to the more
+conventional graphs in Figs.~\ref{fig-ancestry-resolution} and \ref{fig-inferred-args}A,B.
+
+The schematic illustrates an important point about the biological reality of polytomies.
+Three lineages coalesce in the left-hand genome of individual $D_{10}$, but do
+so as the result of two successive bifurcations. This is \emph{necessarily} so,
+because the only known method of reproducing DNA is by (semi-conservative) duplication.
+Whether this polytomy is resolvable depends on the available mutational data.
+Mutations can occur along any cell lineages. For example,
+a mutation in the first cell division of $D_{10}$ could be shared between
+the two gametes produced by the cells in the left half of $D_{10}$
+but not shared by the right hand gamete.
+With enough mutations, each
+round of mitotic germline genome duplication within a single multicellular organism could
+in principle be distinguished.
 
@@
 
+\section{The Big and Little ARG}
+\label{sec-big-and-little-arg}
+Here we review two important stochastic processes that construct ARGs:
+the ``Big" ARG process of \cite{griffiths1997ancestral}, and the ``Little" ARG process of
+ \cite{hudson1983properties}. The Big ARG process is mathematically simpler
+ but is computationally intractable due to generating a vast number of ancestors
+ which contribute no genetic material to the initial sample.
+The Little ARG process avoids non-genetic ancestors at the cost of more complex
+dynamics and state space.
+We also demonstrate that applications relying on the grouping of inheritance pathways into ancestral lineages, such as likelihood-based inference under the coalescent, requires that the
+gARG (or eARG) data structure be interpreted in a model-specific way.
+
+A generic state of the Little ARG process consists of a finite collection of lineages $L$,
+each of which is a list of disjoint ancestry segments $(\ell, r, a)$, where
+$[\ell, r)$ is a half-closed genomic interval and $a$ is an integer
+tracking the number of samples to which the lineage is ancestral over that interval.
+We also usually track the node associated with each segment, but
+that is not important for our purposes here so we omit it to lighten notation.
+The initial condition for a sample of $n$ genomes of length $m$ consists of $n$ lineages
+of the form $\{(0, m, 1)\}$. The process traverses a series of common ancestor and
+recombination events backwards in time.
+Recombination events happen at rate $\rho \nu / (m - 1)$,
+where $\rho \geq 0$ is a per-genome recombination rate and
+ \[
+ \nu = \sum_{x \in L}\left( \max_{(\ell, r, a) \in x}r
+     - \min_{(\ell, r, a) \in x}\ell - 1 \right)
+ \]
+ is the number of available ``links" surrounded by ancestral material.
+ At a recombination event we choose one of these links uniformly and break it,
+ replacing the original lineage in $L$ with two new lineages containing the ancestral material
+ to the left and right of the break point, respectively.
+
+Common ancestor events occur at rate $\binom{|L|}{2}$.
+In a common ancestor event, two uniformly sampled lineages have their segments
+merged into a single ancestor lineage, which is added to $L$.
+If the lineages have overlapping intervals of ancestry,
+say, $(\ell, r, a_1)$ and $(\ell, r, a_2)$, a
+\emph{coalescence} occurs. The result is a segment
+$(\ell, r, a_1 + a_2)$, and if $a_1 + a_2 < n$ it is included in the
+ancestor lineage. Otherwise, if $a_1 + a_2 = n$, we have found
+the most recent common ancestor of all samples in the interval $[\ell, r)$
+and do not need to simulate its history any further.
+Non-overlapping intervals from the two lineages are included
+ in the ancestor lineage without changes. Eventually,
+we find resultant lineages in which all segments have fully coalesced,
+and so the number of extant lineages gradually falls to zero.
+
+In the Griffiths formulation (the Big ARG process), each edge in the graph corresponds to an extant
+lineage and nodes are events in the process. The $n$ initial leaf nodes are
+sampling events. Common ancestor events occur at rate $\binom{|L|}{2}$.
+When a common ancestor event happens, two uniformly chosen lineages
+merge into a common ancestor lineage.
+Recombination events happen at rate $|L| \rho$. Here, we choose a lineage (i.e.\ edge) uniformly,
+and a breakpoint $0 < x < m$ uniformly on its genome. We terminate the edge at a
+node, record the breakpoint, and start two new edges from this node. The process
+then continues until there is only one lineage left (the Grand Most Recent
+Common Ancestor, GMRCA), which is guaranteed to
+happen in finite time because of the quadratic rate of coalescing vs.\ linear rate of branching.
+
+The state-space of the Big ARG process is much simpler than that of the Little ARG process,
+which greatly facilitates mathematical reasoning. This simplicity comes at a
+substantial cost, however, if we wish to use it as a practical means of
+simulating recombinant ancestries.
+The number of events in the Big ARG all the way back to the GMRCA
+is $O(e^\rho)$~\citep{griffiths1997ancestral}, whereas the number
+of events required to simulate the Little ARG is
+$O(\rho^2)$~\citep{hein2004gene,baumdicker2021efficient}.
+This disparity arises because the majority of the events in the Big ARG are
+recombination events which occur outside of ancestral material,
+and these do not have any bearing on the ancestry of the initial sample.
+Because we don't keep track of the distribution of ancestral material during the process,
+we generate a vastly larger graph.
+
+\begin{figure}
+\centering
+% FIXME this is a quick nasty hack to make the figure a bit smaller and
+% prevent flushing all the other figures to the end of the document.
+\scalebox{0.54}{
+\begin{tabular}{cc}
+\begin{tikzpicture}
+	\node [anchor=north west] at (-3.5,9.3) {\LARGE \textbf{A}};
+	\draw (-0.4, -0.2) -- (0.4, -0.2) -- (0.4, -0.3) -- (-0.4, -0.3) -- (-0.4, -0.2);
+	\draw (-0.1, -0.1) -- (-0.1, -0.4);
+	\node [label=below:{\small $v$}] at (-0.1, -0.2) {};
+	\draw (3.6, -0.2) -- (4.4, -0.2) -- (4.4, -0.3) -- (3.6, -0.3) -- (3.6, -0.2);
+	\draw (5.6, -0.2) -- (6.4, -0.2) -- (6.4, -0.3) -- (5.6, -0.3) -- (5.6, -0.2);
+
+	\draw (4,0) -- (4, 1.2) -- (6, 1.2) -- (6,0);
+	\draw (4.6, 1) -- (5.4, 1) -- (5.4, 0.9) -- (4.6, 0.9) -- (4.6, 1);
+	\draw (5.1, 1.1) -- (5.1, 0.8);
+	\node [label=below:{\small $w$}] at (5.1, 1) {};
+
+	\draw (0, 0) -- (0, 1.7) -- (-1,1.7) -- (1,1.7);
+	\draw (-1.4, 1.5) -- (-1.1, 1.5) -- (-1.1, 1.4) -- (-1.4, 1.4) -- (-1.4, 1.5);
+	\draw (-1.1, 1.45) -- (-0.6, 1.45);
+	\draw (0.6, 1.45) -- (0.9, 1.45);
+	\draw (0.9, 1.5) -- (1.4, 1.5) -- (1.4, 1.4) -- (0.9, 1.4) -- (0.9, 1.5);
+	\draw (1.1, 1.6) -- (1.1, 1.3);
+	\node [label=below:{\small $w$}] at (1.1, 1.5) {};
+
+	\draw (5, 1.2) -- (5, 2.4) -- (4,2.4) -- (6,2.4);
+	\draw (3.6, 2.2) -- (4.1, 2.2) -- (4.1, 2.1) -- (3.6, 2.1) -- (3.6, 2.2);
+	\draw (4.1, 2.15) -- (4.4, 2.15);
+	\draw (5.6, 2.15) -- (6.1, 2.15);
+	\draw (6.1, 2.2) -- (6.4, 2.2) -- (6.4, 2.1) -- (6.1, 2.1) -- (6.1, 2.2);
+
+	\draw (1, 1.7) -- (1, 3.1) -- (0,3.1) -- (2,3.1);
+	\draw (-0.4, 2.85) -- (0.1, 2.85);
+	\draw (0.1, 2.9) -- (0.4, 2.9) -- (0.4, 2.8) -- (0.1, 2.8) -- (0.1, 2.9);
+	\draw (1.6, 2.85) -- (1.9, 2.85);
+	\draw (1.9, 2.9) -- (2.1, 2.9) -- (2.1, 2.8) -- (1.9, 2.8) -- (1.9, 2.9);
+	\draw (2.1, 2.85) -- (2.4, 2.85);
+
+	\draw (-1, 1.7) -- (-1, 4.5) -- (0,4.5) -- (0,3.1);
+	\node [scale=0.2,label=above left:{\small \noderef{a}}] at (-0.5,4.5) {a};
+	\draw (-0.9, 4.3) -- (-0.6, 4.3) -- (-0.6, 4.2) -- (-0.9, 4.2) -- (-0.9, 4.3);
+	\draw (-0.6, 4.25) -- (-0.4, 4.25);
+	\draw (-0.4, 4.3) -- (-0.1, 4.3) -- (-0.1, 4.2) -- (-0.4, 4.2) -- (-0.4, 4.3);
+	\draw (-0.5, 4.4) -- (-0.5, 4.1);
+	\node [label=below:{\small $x$}] at (-0.5, 4.3) {};
+
+	\draw (4,2.4) -- (4,5.4) -- (2,5.4) -- (2,3.1);
+	\node [scale=0.2,label=above left:{\small \noderef{b}}] at (3,5.4) {b};
+	\draw (2.6, 5.2) -- (2.9, 5.2) -- (2.9, 5.1) -- (2.6, 5.1) -- (2.6, 5.2);
+	\draw (2.9, 5.15) -- (3.4, 5.15);
+
+	\draw (-0.5, 4.5) -- (-0.5, 6) -- (-1.5,6) -- (0.5,6);
+	\draw (-1.9, 5.75) -- (-1.4, 5.75);
+	\draw (-1.4, 5.8) -- (-1.1, 5.8) -- (-1.1, 5.7) -- (-1.4, 5.7) -- (-1.4, 5.8);
+	\draw (0.1, 5.8) -- (0.4, 5.8) -- (0.4, 5.7) -- (0.1, 5.7) -- (0.1, 5.8);
+	\draw (0.4, 5.75) -- (0.9, 5.75);
+
+	\draw (0.5, 6) -- (0.5, 7.2) -- (3,7.2) -- (3,5.4);
+	\node [scale=0.2,label=above:{\small \noderef{c}}] at (1.75,7.2) {c};
+	\draw (1.35, 6.95) -- (2.15, 6.95);
+
+	\draw (-1.5, 6) -- (-1.5, 8);
+	\draw (6, 2.4) -- (6, 8);
+
+	% Edge annotations above each event
+	\node [label=left:{\small$\{(0,m,1)\}$}] at (0.2,0.6) {};
+	\node [label=left:{\small$\{(0,m,1)\}$}] at (4.2,0.4) {};
+	\node [label=right:{\small$\{(0,m,1)\}$}] at (5.8,0.4) {};
+	\node [label=left:{\small$\{(0,m,2)\}$}] at (5.2,1.6) {};
+	\node [label=left:{\small$\{(0,v,1)\}$}] at (-0.8,2.3) {};
+	\node [label=right:{\small$\{(v,m,1)\}$}] at (0.8,2.1) {};
+	\node [label=right:{\small$\{(0,w,2)\}$}] at (3.8,3.9) {};
+	\node [label=left:{\small$\{(w,m,2)\}$}] at (6.2,5) {};
+	\node [label=right:{\small$\{(w,m,1)\}$}] at (-0.2,3.8) {};
+	\node [label=right:{\small$\{(v,w,1)\}$}] at (1.8,4.1) {};
+	\node [label=left:{\small$\{(0,v,1), (w,m,1)\}$}] at (-0.3,5.2) {};
+	\node [label=right:{\small$\{(0,v,2)\}$}] at (2.8,6.4) {};
+	\node [label=left:{\small$\{(w,m,1)\}$}] at (-1.3,7) {};
+	\node [label=right:{\small$\{(0,v,1)\}$}] at (0.3,6.4) {};
+
+	% Dashed lines for start and end times
+	\draw[dashed] (-2, 0) -- (11, 0);
+	\node [label=left:{$t = 0$}] at (-2,0) {};
+	\draw[dashed] (-2, 8) -- (11, 8);
+	\node [label=left:{$t = T$}] at (-2,8) {};
+
+	% Numbers of extant ancestors and links, from top to bottom
+	\node[label=right:{CA \; RE}] at (7.5, 8.2) {};
+	\node[label=right:{$\binom{2}{2}$ \; $\frac{2 m - 2 w - 2}{m - 1} \rho$}] at (7.5, 7.6) {};
+	\node[label=right:{$\binom{4}{2}$ \; $\frac{2 m + 2 v - 2 w - 4}{m - 1} \rho$}] at (7.5, 6.6) {};
+	\node[label=right:{$\binom{3}{2}$ \; $\frac{2 m + v - w - 3}{m - 1} \rho$}] at (7.5, 5.7) {};
+	\node[label=right:{$\binom{4}{2}$ \; $\frac{2 m + w - v - 4}{m - 1} \rho$}] at (7.5, 4.95) {};
+	\node[label=right:{$\binom{5}{2}$ \; $\frac{2 m - 5}{m - 1} \rho$}] at (7.5, 3.8) {};
+	\node[label=right:{$\binom{4}{2}$ \; $\frac{2 m - 4}{m -1} \rho$}] at (7.5, 2.75) {};
+	\node[label=right:{$\binom{3}{2}$ \; $\frac{2 m - 3}{m - 1} \rho$}] at (7.5, 2.05) {};
+	\node[label=right:{$\binom{2}{2}$ \; $2 \rho$}] at (7.5, 1.45) {};
+	\node[label=right:{$\binom{3}{2}$ \; $3 \rho$}] at (7.5, 0.6) {};
+
+	% Gray dashed lines to visually separate holding times
+	\draw[color=gray, dashed] (7.5, 1.2) -- (11, 1.2);
+	\draw[color=gray, dashed] (7.5, 1.7) -- (11, 1.7);
+	\draw[color=gray, dashed] (7.5, 2.4) -- (11, 2.4);
+	\draw[color=gray, dashed] (7.5, 3.1) -- (11, 3.1);
+	\draw[color=gray, dashed] (7.5, 4.5) -- (11, 4.5);
+	\draw[color=gray, dashed] (7.5, 5.4) -- (11, 5.4);
+	\draw[color=gray, dashed] (7.5, 6) -- (11, 6);
+	\draw[color=gray, dashed] (7.5, 7.2) -- (11, 7.2);
+\end{tikzpicture}
+&
+\begin{tikzpicture}
+	\node [anchor=north west] at (-3.5,9.3) {\LARGE \textbf{B}};
+	\draw (-0.4, -0.2) -- (0.4, -0.2) -- (0.4, -0.3) -- (-0.4, -0.3) -- (-0.4, -0.2);
+	\draw (-0.1, -0.1) -- (-0.1, -0.4);
+	\node [label=below:{\small $v$}] at (-0.1, -0.2) {};
+	\draw (3.6, -0.2) -- (4.4, -0.2) -- (4.4, -0.3) -- (3.6, -0.3) -- (3.6, -0.2);
+	\draw (5.6, -0.2) -- (6.4, -0.2) -- (6.4, -0.3) -- (5.6, -0.3) -- (5.6, -0.2);
+
+	\draw (4,0) -- (4, 1.2) -- (6, 1.2) -- (6,0);
+	\draw (4.6, 1) -- (5.4, 1) -- (5.4, 0.9) -- (4.6, 0.9) -- (4.6, 1);
+	\draw (5.1, 1.1) -- (5.1, 0.8);
+	\node [label=below:{\small $w$}] at (5.1, 1) {};
+
+	\draw (0, 0) -- (0, 1.7) -- (-1,1.7) -- (1,1.7);
+	\draw (-1.4, 1.5) -- (-1.1, 1.5) -- (-1.1, 1.4) -- (-1.4, 1.4) -- (-1.4, 1.5);
+	\draw (-1.1, 1.45) -- (-0.6, 1.45);
+	\draw (0.6, 1.45) -- (0.9, 1.45);
+	\draw (0.9, 1.5) -- (1.4, 1.5) -- (1.4, 1.4) -- (0.9, 1.4) -- (0.9, 1.5);
+	\draw (1.1, 1.6) -- (1.1, 1.3);
+	\node [label=below:{\small $w$}] at (1.1, 1.5) {};
+
+	\draw (5, 1.2) -- (5, 2.4) -- (4,2.4) -- (6,2.4);
+	\draw (3.6, 2.2) -- (4.1, 2.2) -- (4.1, 2.1) -- (3.6, 2.1) -- (3.6, 2.2);
+	\draw (4.1, 2.15) -- (4.4, 2.15);
+	\draw (5.6, 2.15) -- (6.1, 2.15);
+	\draw (6.1, 2.2) -- (6.4, 2.2) -- (6.4, 2.1) -- (6.1, 2.1) -- (6.1, 2.2);
+	\draw [color=red](5.8, 2.3) -- (5.8, 2.0);
+	\node [label={[red]below:{\small $y$}}] at (5.8, 2.1) {};
+
+	\draw (1, 1.7) -- (1, 3.1) -- (0,3.1) -- (2,3.1);
+	\draw (-0.4, 2.85) -- (0.1, 2.85);
+	\draw (0.1, 2.9) -- (0.4, 2.9) -- (0.4, 2.8) -- (0.1, 2.8) -- (0.1, 2.9);
+	\draw (1.6, 2.85) -- (1.9, 2.85);
+	\draw (1.9, 2.9) -- (2.1, 2.9) -- (2.1, 2.8) -- (1.9, 2.8) -- (1.9, 2.9);
+	\draw (2.1, 2.85) -- (2.4, 2.85);
+
+	\draw (-1, 1.7) -- (-1, 4.5) -- (0,4.5) -- (0,3.1);
+	\draw (-0.9, 4.3) -- (-0.6, 4.3) -- (-0.6, 4.2) -- (-0.9, 4.2) -- (-0.9, 4.3);
+	\draw (-0.6, 4.25) -- (-0.4, 4.25);
+	\draw (-0.4, 4.3) -- (-0.1, 4.3) -- (-0.1, 4.2) -- (-0.4, 4.2) -- (-0.4, 4.3);
+	\draw (-0.5, 4.4) -- (-0.5, 4.1);
+	\node [label=below:{\small $x$}] at (-0.5, 4.3) {};
+
+	\draw (6, 2.4) -- (6, 3.8) -- (7, 3.8);
+	\draw [color=red](6, 3.8) -- (5, 3.8);
+	\draw (4.6, 3.55) -- (5.4, 3.55);
+	\draw (6.6, 3.55) -- (7.1, 3.55);
+	\draw (7.1, 3.6) -- (7.4, 3.6) -- (7.4, 3.5) -- (7.1, 3.5) -- (7.1, 3.6);
+
+	\draw (4,2.4) -- (4,5.4) -- (2,5.4) -- (2,3.1);
+	\draw (2.6, 5.2) -- (2.9, 5.2) -- (2.9, 5.1) -- (2.6, 5.1) -- (2.6, 5.2);
+	\draw (2.9, 5.15) -- (3.4, 5.15);
+
+	\draw (-0.5, 4.5) -- (-0.5, 6) -- (-1.5,6) -- (0.5,6);
+	\draw (-1.9, 5.75) -- (-1.4, 5.75);
+	\draw (-1.4, 5.8) -- (-1.1, 5.8) -- (-1.1, 5.7) -- (-1.4, 5.7) -- (-1.4, 5.8);
+	\draw (0.1, 5.8) -- (0.4, 5.8) -- (0.4, 5.7) -- (0.1, 5.7) -- (0.1, 5.8);
+	\draw (0.4, 5.75) -- (0.9, 5.75);
+
+	\draw [color=red](5, 3.8) -- (5, 6.6) -- (4, 6.6);
+	\draw (4,6.6) -- (3,6.6) -- (3, 5.4);
+	\draw (3.6, 6.4) -- (3.9, 6.4) -- (3.9, 6.3) -- (3.6, 6.3) -- (3.6, 6.4);
+	\draw (3.9, 6.35) -- (4.4, 6.35);
+
+	\draw (0.5, 6) -- (0.5, 7.2) -- (4,7.2) -- (4,6.6);
+	\draw (1.85, 6.95) -- (2.65, 6.95);
+
+	\draw (-1.5, 6) -- (-1.5, 8);
+	\draw [color=red](2.25, 7.2) -- (2.25, 8);
+	\draw (7, 3.8) -- (7, 8);
+
+	% Dashed lines for start and end times
+	\draw[dashed] (-2, 0) -- (9.1, 0);
+	\node [label=left:{$t = 0$}] at (-2,0) {};
+	\draw[dashed] (-2, 8) -- (9.1, 8);
+	\node [label=left:{$t = T$}] at (-2,8) {};
+
+	% Numbers of extant ancestors and links, from top to bottom
+	\node[label=right:{CA \; RE}] at (7.5, 8.2) {};
+	\node[label=right:{$\binom{3}{2}$ \; $3 \rho$}] at (7.5, 7.6) {};
+	\node[label=right:{$\binom{4}{2}$ \; $4 \rho$}] at (7.5, 6.9) {};
+	\node[label=right:{$\binom{5}{2}$ \; $5 \rho$}] at (7.5, 6.3) {};
+	\node[label=right:{$\binom{4}{2}$ \; $4 \rho$}] at (7.5, 5.7) {};
+	\node[label=right:{$\binom{5}{2}$ \; $5 \rho$}] at (7.5, 4.95) {};
+	\node[label=right:{$\binom{6}{2}$ \; $6 \rho$}] at (7.5, 4.15) {};
+	\node[label=right:{$\binom{5}{2}$ \; $5 \rho$}] at (7.5, 3.45) {};
+	\node[label=right:{$\binom{4}{2}$ \; $4 \rho$}] at (7.5, 2.75) {};
+	\node[label=right:{$\binom{3}{2}$ \; $3 \rho$}] at (7.5, 2.05) {};
+	\node[label=right:{$\binom{2}{2}$ \; $2 \rho$}] at (7.5, 1.45) {};
+	\node[label=right:{$\binom{3}{2}$ \; $3 \rho$}] at (7.5, 0.6) {};
+
+	% Gray dashed lines to visually separate holding times
+	\draw[color=gray, dashed] (7.5, 1.2) -- (9.1, 1.2);
+	\draw[color=gray, dashed] (7.5, 1.7) -- (9.1, 1.7);
+	\draw[color=gray, dashed] (7.5, 2.4) -- (9.1, 2.4);
+	\draw[color=gray, dashed] (7.5, 3.1) -- (9.1, 3.1);
+	\draw[color=gray, dashed] (7.5, 3.8) -- (9.1, 3.8);
+	\draw[color=gray, dashed] (7.5, 4.5) -- (9.1, 4.5);
+	\draw[color=gray, dashed] (7.5, 5.4) -- (9.1, 5.4);
+	\draw[color=gray, dashed] (7.5, 6) -- (9.1, 6);
+	\draw[color=gray, dashed] (7.5, 6.6) -- (9.1, 6.6);
+	\draw[color=gray, dashed] (7.5, 7.2) -- (9.1, 7.2);
+\end{tikzpicture}
+\end{tabular}
+}
+\caption{(A)
+A realisation of the graph traversed by Hudson's algorithm started from a
+sample of three chromosomes with $m$ discrete sites each at time $t = 0$, and
+propagated until time $T$. The MRCA on the genetic interval $[v, w)$ is reached
+at event \noderef{b}, while that on $[0, v)$ is reached at event \noderef{c}.
+The non-ancestral segment $[v, w)$ above
+\noderef{a} contributes to the rate of effective recombinations because it
+is trapped between ancestral segments. The two columns titled CA and RE
+are the respective rates of mergers and recombinations when
+the recombination rate is $\rho$.
+(B) A corresponding realisation of a Big ARG, which augments Hudson's algorithm
+by tracking non-ancestral lineages. The result is a simpler state space and
+dynamics, at the cost of extra nodes and edges, highlighted in red, which do
+not affect the local tree at any site.
+Recombination positions are labelled alphabetically in time, and their ordering
+along the genome is $y < v < x < w$, of which the first only appears in panel B.
+There are two separate recombination events at link $w$.
+}
+\label{hudson_vs_bigARG}
+\end{figure}
+
+Figure \ref{hudson_vs_bigARG} illustrates the more complex state space
+of the Little ARG process, as well as the extra events which occur in the Big ARG process.
+Moreover, it depicts the rates of common ancestors and recombination events in each
+interval of time of the realisations.
+In order to evaluate these rates, e.g.\ for likelihood-based inference
+\citep{baumdicker2021efficient, mahmoudi2022bayesian},
+it is necessary to know the number of lineages and number of extant links
+available for recombination in each time interval.
+Some representations may not provide this information.
+For example, in the gARG encoding depicted in Figure \ref{fig-ancestry-resolution}C,
+it is clear that a recombination takes place between nodes \noderef{i}, \noderef{k} and \noderef{j}.
+But the exact time of the recombination event is ambiguous: it could take place at any time between node \noderef{i} and its parents and produce the same gARG.
+Because a recombination increases the number of extant lineages by one (in the rootward direction of time), the number of lineages during the same time interval is ambiguous as well.
+In fact, this information cannot be recovered from the gARG encoding used in Figure \ref{fig-ancestry-resolution}C without an extrinsic convention.
+For the basic coalescent with recombination,
+it is sufficient to create \emph{two} gARG nodes at the time of the recombination event, with the interpretation that the two rootward edges from node \noderef{i} in Figure \ref{fig-ancestry-resolution}C belong to the same lineage until the time of nodes \noderef{k} and \noderef{j}, and split into two separate lineages at that time point.
+Similarly, the trapped, non-ancestral links along that lineage remain available for effective recombination (i.e.\ one which splits up ancestral material) for the same time interval.
+This interpretation is highlighted in
+Figure \ref{hudson_vs_bigARG} by drawing only one vertical edge between a recombinant child and its two parents.
+
+% Full quote:
+% This process simplifies mathematics on the account that the notion of an
+% ancestor will have a less restrictive meaning than usual:
+% An ``ancestral'' sequence in the birth and death process
+% need not have any genetic material in common with a
+% sequence descended from it.
+
 \section{Locally unary nodes}
@@
 
-\section{Implementation and efficiency}
-\label{sec-efficiency}
-The gARG encoding defined here
-leads to highly efficient storage and processing of ARG data,
-and has already been in use for several years.
-The succinct tree sequence data structure
-(usually known as a ``tree sequence'' for brevity)
-is a practical gARG implementation focused on efficiency.
-It was originally developed as part of the \texttt{msprime}
-simulator~\citep{kelleher2016efficient} and has subsequently been
-extended and applied to forward-time
-simulations~\citep{kelleher2018efficient,haller2018tree},
-inference from
-data~\citep{kelleher2019inferring,wohns2022unified,zhan2023towards},
-and calculation of population genetics statistics~\citep{ralph2020efficiently}.
-The succinct tree sequence encoding extends the basic definition
-of a gARG provided here by stipulating a
-simple tabular representation of nodes and edges,
-and also defining a concise representation of
-sequence variation using the ``site'' and  ``mutation'' tables.
-The key property of the succinct tree sequence encoding
-that makes it an efficient substrate for defining analysis
-algorithms is that it allows us to sequentially
-recover the local trees along the genome very efficiently,
-and in a way that allows us to reason about the \emph{differences}
-between those trees~\citep{kelleher2016efficient,ralph2020efficiently}.
-
-The \texttt{tskit} library is a liberally
-licensed open source toolkit that provides a comprehensive suite
-of tools for working with gARGs (encoded as a succinct tree sequence).
-Based on core functionality written
-in C, it provides interfaces in C, Python and Rust.
-Tskit is mature software, widely used in population genetics, and
-has been incorporated into numerous downstream
-applications~\citep[e.g.,][]{haller2019slim,speidel2019method,
-adrion2020community,
-terasaki2021geonomics,
-baumdicker2021efficient,
-fan2022genealogical,
-guo2022recombination,
-korfmann2022weak,
-mahmoudi2022bayesian,
-petr2023slendr,
-rasmussen2022espalier,
-zhang2023biobank,
-nowbandegani2023extremely,
-ignatieva2023distribution,
-fan2023likelihood,
-tsambos2023link,
-tagami2024tstrait}.
-The technical details of \texttt{tskit}, and how it provides an
-efficient and portable platform for ARG-based analysis, are beyond
-the scope of this manuscript.
-In the interest of avoiding confusion, however, we list a
-few minor details in which the formal details of gARGs
-provided in Section~\ref{sec-gARG} differ from their practical implementation in
-\texttt{tskit}.
-Firstly, ``edges'' in tree sequence terminology would perhaps be better
-described as ``edge-intervals'', as each describes a single contiguous
-interval of genome inheritance between a pair of nodes.
-This denormalisation of the gARG data model is for efficiency purposes.
-Secondly, zero- rather than one-based indexing is used for
-nodes in ARGs and oriented trees; consequently  $-1$ is used to denote the presence of
-roots (rather than $0$ as used here for notational simplicity).
-
-\section{Cell lineages and ARGs}
-\label{sec-cell-lineages-and-args}
-\begin{figure}
-\begin{center}
-    \includegraphics[width=0.8\textwidth]{illustrations/cell-lines}
-\end{center}
-\caption{\label{fig-cell-lines}
-Cellular inheritance of a single chromosome in a diploid population.
-Individuals (blue) contain diploid cells (white circles enclosing a homologous pair of chromosomes).
-For clarity, only two rounds of mitotic germ-line cell division are shown per individual, and
-meiosis is not illustrated in detail.
-Lines show prospective inheritance paths for all chromosomes. Solid lines show all possible
-retrospective ancestry paths for four chosen chromosomes (indicated by square black ``sampling events'')
-sampled from 3 diploid individuals ($D_1$, $D_2$, $D_3$) in the current generation.
-Ancestral recombination events and coalescence events are shown as red and blue squares respectively.
-A realised ARG path for the lower arm of the sampled chromosomes is highlighted as a
-thick solid line, passing through a set of potential gARG nodes (green). This ARG
-involves a single recombination event and four coalescence events
-(highlighted as deep red and blue squares within individuals $D_5$, $D_{10}$, and $D_{13}$).
-ARG lineages also show gametic genomes, contained within shaded circles.
-As in Fig.~\ref{fig-arg-in-pedigree}A, inherited regions within the sampled chromosome arm are
-shaded by the number of descendant samples.
-}
-\end{figure}
-
-In eukaryotes, ARGs are a result of the cellular processes of mitosis and meiosis.
-Mitosis leads to common ancestor events,
-and meiosis leads to recombination events (both crossover and gene conversion).
-Fig.~\ref{fig-cell-lines} shows a schematic of the events  and
-the genomes (chromosome icons) that occur in the cellular
-germline of a simplified, diploid multicellular
-hermaphrodite eukaryote with partially overlapping generations.
-Here, an event is not represented by a specific
-genome. Rather, genomes can be associated with, or ``tag'', events above
-(ancestral to) or below (descended from) them.
-For example, tagging the two genomes above a recombination event leads to the
-two node representation seen in
-Fig.~\ref{fig-simplification}A and Fig.~\ref{hudson_vs_bigARG},
-whereas tagging the genome
-below a recombination event leads to the more
-conventional graphs in Figs.~\ref{fig-ancestry-resolution} and \ref{fig-inferred-args}A,B.
-
-%% JK:  What does this add that is not in caption?
-% Thick black lines trace an ARG that describes the ancestry of the lower arm of
-% the four sampled chromosomes, and these lines pass through gARG nodes (green
-% chromosomal regions). A number of these nodes have only one parent and one
-% child in the ARG; these would often be omitted (taken as implicit) in a
-% minimally simplified gARG.
-
-The schematic illustrates an important point about the biological reality of polytomies.
-Three lineages coalesce in the left-hand genome of individual $D_{10}$, but do
-so as the result of two successive bifurcations. This is \emph{necessarily} so,
-because the only known method of reproducing DNA is by (semi-conservative) duplication.
-Whether this polytomy is resolvable depends on the available mutational data.
-Mutations can occur along any cell lineages. For example,
-a mutation in the first cell division of $D_{10}$ could be shared between
-the two gametes produced by the cells in the left half of $D_{10}$
-but not shared by the right hand gamete.
-With enough mutations, each
-round of mitotic germline genome duplication within a single multicellular organism could
-in principle be distinguished.
-
-% Note that mutations can occur along any lineage in Fig.~\ref{fig-cell-lines}. This means, for example,
-% that a mutation in the first cell division of $D_{10}$ could be shared between
-% the two gametes produced by the cells in the left half of $D_{10}$ but not shared by
-% the right hand gamete. With a high enough mutation rate (or equivalently, a long enough genome), each
-% round of mitotic germline genome duplication within a single multicellular organism could
-% potentially be distinguished.
-% This implies that in ARGs that have been inferred from real data, polytomies represent a lack
-% of mutational information (rather than marking a single ``multifurcation event'').
-% The ability to represent polytomies, and conceptualise them as a (deliberately) imprecise statement about the order
-% of coalescence within and between individuals,
-% fits naturally into the gARG framework presented in this paper.
-
 \clearpage

### 2024-04-04 paper.tex
Finalise first three appendixes

@@
 and data structure is not clear cut, however, and subfields use the term
-differently (Appendix~\ref{sec-arg-history}).
+differently (Appendix~\ref{sec-arg-history},\ref{sec-big-and-little-arg}).
 
@@
 instead of a complete and fully detailed history
-(see Appendix~\ref{sec-survey-arg-infer} for a brief survey).
+(Appendix~\ref{sec-survey-arg-infer}).
 It is important to distinguish here between
@@
 
+\section{The Big and Little ARG}
+\label{sec-big-and-little-arg}
+Here we review two important stochastic processes that construct ARGs:
+the ``Big" ARG process of \cite{griffiths1997ancestral}, and the ``Little" ARG process of
+ \cite{hudson1983properties}. The Big ARG process is mathematically simpler
+ but is computationally intractable due to generating a vast number of ancestors
+ which contribute no genetic material to the initial sample.
+The Little ARG process avoids non-genetic ancestors at the cost of more complex
+dynamics and state space.
+We also demonstrate that applications relying on the grouping of inheritance pathways into ancestral lineages, such as likelihood-based inference under the coalescent, requires that the
+gARG (or eARG) data structure be interpreted in a model-specific way.
+
+A generic state of the Little ARG process consists of a finite collection of lineages $L$,
+each of which is a list of disjoint ancestry segments $(\ell, r, a)$, where
+$[\ell, r)$ is a half-closed genomic interval and $a$ is an integer
+tracking the number of samples to which the lineage is ancestral over that interval.
+We also usually track the node associated with each segment, but
+that is not important for our purposes here so we omit it to lighten notation.
+The initial condition for a sample of $n$ genomes of length $m$ consists of $n$ lineages
+of the form $\{(0, m, 1)\}$. The process traverses a series of common ancestor and
+recombination events backwards in time.
+Recombination events happen at rate $\rho \nu / (m - 1)$,
+where $\rho \geq 0$ is a per-genome recombination rate and
+ \[
+ \nu = \sum_{x \in L}\left( \max_{(\ell, r, a) \in x}r
+     - \min_{(\ell, r, a) \in x}\ell - 1 \right)
+ \]
+ is the number of available ``links" surrounded by ancestral material.
+ At a recombination event we choose one of these links uniformly and break it,
+ replacing the original lineage in $L$ with two new lineages containing the ancestral material
+ to the left and right of the break point, respectively.
+
+Common ancestor events occur at rate $\binom{|L|}{2}$.
+In a common ancestor event, two uniformly sampled lineages have their segments
+merged into a single ancestor lineage, which is added to $L$.
+If the lineages have overlapping intervals of ancestry,
+say, $(\ell, r, a_1)$ and $(\ell, r, a_2)$, a
+\emph{coalescence} occurs. The result is a segment
+$(\ell, r, a_1 + a_2)$, and if $a_1 + a_2 < n$ it is included in the
+ancestor lineage. Otherwise, if $a_1 + a_2 = n$, we have found
+the most recent common ancestor of all samples in the interval $[\ell, r)$
+and do not need to simulate its history any further.
+Non-overlapping intervals from the two lineages are included
+ in the ancestor lineage without changes. Eventually,
+we find resultant lineages in which all segments have fully coalesced,
+and so the number of extant lineages gradually falls to zero.
+
+In the Griffiths formulation (the Big ARG process), each edge in the graph corresponds to an extant
+lineage and nodes are events in the process. The $n$ initial leaf nodes are
+sampling events. Common ancestor events occur at rate $\binom{|L|}{2}$.
+When a common ancestor event happens, two uniformly chosen lineages
+merge into a common ancestor lineage.
+Recombination events happen at rate $|L| \rho$. Here, we choose a lineage (i.e.\ edge) uniformly,
+and a breakpoint $0 < x < m$ uniformly on its genome. We terminate the edge at a
+node, record the breakpoint, and start two new edges from this node. The process
+then continues until there is only one lineage left (the Grand Most Recent
+Common Ancestor, GMRCA), which is guaranteed to
+happen in finite time because of the quadratic rate of coalescing vs.\ linear rate of branching.
+
+The state-space of the Big ARG process is much simpler than that of the Little ARG process,
+which greatly facilitates mathematical reasoning. This simplicity comes at a
+substantial cost, however, if we wish to use it as a practical means of
+simulating recombinant ancestries.
+The number of events in the Big ARG all the way back to the GMRCA
+is $O(e^\rho)$~\citep{griffiths1997ancestral}, whereas the number
+of events required to simulate the Little ARG is
+$O(\rho^2)$~\citep{hein2004gene,baumdicker2021efficient}.
+This disparity arises because the majority of the events in the Big ARG are
+recombination events which occur outside of ancestral material,
+and these do not have any bearing on the ancestry of the initial sample.
+Because we don't keep track of the distribution of ancestral material during the process,
+we generate a vastly larger graph.
+
+\begin{figure}
+\centering
+% FIXME this is a quick nasty hack to make the figure a bit smaller and
+% prevent flushing all the other figures to the end of the document.
+\scalebox{0.54}{
+\begin{tabular}{cc}
+\begin{tikzpicture}
+	\node [anchor=north west] at (-3.5,9.3) {\LARGE \textbf{A}};
+	\draw (-0.4, -0.2) -- (0.4, -0.2) -- (0.4, -0.3) -- (-0.4, -0.3) -- (-0.4, -0.2);
+	\draw (-0.1, -0.1) -- (-0.1, -0.4);
+	\node [label=below:{\small $v$}] at (-0.1, -0.2) {};
+	\draw (3.6, -0.2) -- (4.4, -0.2) -- (4.4, -0.3) -- (3.6, -0.3) -- (3.6, -0.2);
+	\draw (5.6, -0.2) -- (6.4, -0.2) -- (6.4, -0.3) -- (5.6, -0.3) -- (5.6, -0.2);
+
+	\draw (4,0) -- (4, 1.2) -- (6, 1.2) -- (6,0);
+	\draw (4.6, 1) -- (5.4, 1) -- (5.4, 0.9) -- (4.6, 0.9) -- (4.6, 1);
+	\draw (5.1, 1.1) -- (5.1, 0.8);
+	\node [label=below:{\small $w$}] at (5.1, 1) {};
+
+	\draw (0, 0) -- (0, 1.7) -- (-1,1.7) -- (1,1.7);
+	\draw (-1.4, 1.5) -- (-1.1, 1.5) -- (-1.1, 1.4) -- (-1.4, 1.4) -- (-1.4, 1.5);
+	\draw (-1.1, 1.45) -- (-0.6, 1.45);
+	\draw (0.6, 1.45) -- (0.9, 1.45);
+	\draw (0.9, 1.5) -- (1.4, 1.5) -- (1.4, 1.4) -- (0.9, 1.4) -- (0.9, 1.5);
+	\draw (1.1, 1.6) -- (1.1, 1.3);
+	\node [label=below:{\small $w$}] at (1.1, 1.5) {};
+
+	\draw (5, 1.2) -- (5, 2.4) -- (4,2.4) -- (6,2.4);
+	\draw (3.6, 2.2) -- (4.1, 2.2) -- (4.1, 2.1) -- (3.6, 2.1) -- (3.6, 2.2);
+	\draw (4.1, 2.15) -- (4.4, 2.15);
+	\draw (5.6, 2.15) -- (6.1, 2.15);
+	\draw (6.1, 2.2) -- (6.4, 2.2) -- (6.4, 2.1) -- (6.1, 2.1) -- (6.1, 2.2);
+
+	\draw (1, 1.7) -- (1, 3.1) -- (0,3.1) -- (2,3.1);
+	\draw (-0.4, 2.85) -- (0.1, 2.85);
+	\draw (0.1, 2.9) -- (0.4, 2.9) -- (0.4, 2.8) -- (0.1, 2.8) -- (0.1, 2.9);
+	\draw (1.6, 2.85) -- (1.9, 2.85);
+	\draw (1.9, 2.9) -- (2.1, 2.9) -- (2.1, 2.8) -- (1.9, 2.8) -- (1.9, 2.9);
+	\draw (2.1, 2.85) -- (2.4, 2.85);
+
+	\draw (-1, 1.7) -- (-1, 4.5) -- (0,4.5) -- (0,3.1);
+	\node [scale=0.2,label=above left:{\small \noderef{a}}] at (-0.5,4.5) {a};
+	\draw (-0.9, 4.3) -- (-0.6, 4.3) -- (-0.6, 4.2) -- (-0.9, 4.2) -- (-0.9, 4.3);
+	\draw (-0.6, 4.25) -- (-0.4, 4.25);
+	\draw (-0.4, 4.3) -- (-0.1, 4.3) -- (-0.1, 4.2) -- (-0.4, 4.2) -- (-0.4, 4.3);
+	\draw (-0.5, 4.4) -- (-0.5, 4.1);
+	\node [label=below:{\small $x$}] at (-0.5, 4.3) {};
+
+	\draw (4,2.4) -- (4,5.4) -- (2,5.4) -- (2,3.1);
+	\node [scale=0.2,label=above left:{\small \noderef{b}}] at (3,5.4) {b};
+	\draw (2.6, 5.2) -- (2.9, 5.2) -- (2.9, 5.1) -- (2.6, 5.1) -- (2.6, 5.2);
+	\draw (2.9, 5.15) -- (3.4, 5.15);
+
+	\draw (-0.5, 4.5) -- (-0.5, 6) -- (-1.5,6) -- (0.5,6);
+	\draw (-1.9, 5.75) -- (-1.4, 5.75);
+	\draw (-1.4, 5.8) -- (-1.1, 5.8) -- (-1.1, 5.7) -- (-1.4, 5.7) -- (-1.4, 5.8);
+	\draw (0.1, 5.8) -- (0.4, 5.8) -- (0.4, 5.7) -- (0.1, 5.7) -- (0.1, 5.8);
+	\draw (0.4, 5.75) -- (0.9, 5.75);
+
+	\draw (0.5, 6) -- (0.5, 7.2) -- (3,7.2) -- (3,5.4);
+	\node [scale=0.2,label=above:{\small \noderef{c}}] at (1.75,7.2) {c};
+	\draw (1.35, 6.95) -- (2.15, 6.95);
+
+	\draw (-1.5, 6) -- (-1.5, 8);
+	\draw (6, 2.4) -- (6, 8);
+
+	% Edge annotations above each event
+	\node [label=left:{\small$\{(0,m,1)\}$}] at (0.2,0.6) {};
+	\node [label=left:{\small$\{(0,m,1)\}$}] at (4.2,0.4) {};
+	\node [label=right:{\small$\{(0,m,1)\}$}] at (5.8,0.4) {};
+	\node [label=left:{\small$\{(0,m,2)\}$}] at (5.2,1.6) {};
+	\node [label=left:{\small$\{(0,v,1)\}$}] at (-0.8,2.3) {};
+	\node [label=right:{\small$\{(v,m,1)\}$}] at (0.8,2.1) {};
+	\node [label=right:{\small$\{(0,w,2)\}$}] at (3.8,3.9) {};
+	\node [label=left:{\small$\{(w,m,2)\}$}] at (6.2,5) {};
+	\node [label=right:{\small$\{(w,m,1)\}$}] at (-0.2,3.8) {};
+	\node [label=right:{\small$\{(v,w,1)\}$}] at (1.8,4.1) {};
+	\node [label=left:{\small$\{(0,v,1), (w,m,1)\}$}] at (-0.3,5.2) {};
+	\node [label=right:{\small$\{(0,v,2)\}$}] at (2.8,6.4) {};
+	\node [label=left:{\small$\{(w,m,1)\}$}] at (-1.3,7) {};
+	\node [label=right:{\small$\{(0,v,1)\}$}] at (0.3,6.4) {};
+
+	% Dashed lines for start and end times
+	\draw[dashed] (-2, 0) -- (11, 0);
+	\node [label=left:{$t = 0$}] at (-2,0) {};
+	\draw[dashed] (-2, 8) -- (11, 8);
+	\node [label=left:{$t = T$}] at (-2,8) {};
+
+	% Numbers of extant ancestors and links, from top to bottom
+	\node[label=right:{CA \; RE}] at (7.5, 8.2) {};
+	\node[label=right:{$\binom{2}{2}$ \; $\frac{2 m - 2 w - 2}{m - 1} \rho$}] at (7.5, 7.6) {};
+	\node[label=right:{$\binom{4}{2}$ \; $\frac{2 m + 2 v - 2 w - 4}{m - 1} \rho$}] at (7.5, 6.6) {};
+	\node[label=right:{$\binom{3}{2}$ \; $\frac{2 m + v - w - 3}{m - 1} \rho$}] at (7.5, 5.7) {};
+	\node[label=right:{$\binom{4}{2}$ \; $\frac{2 m + w - v - 4}{m - 1} \rho$}] at (7.5, 4.95) {};
+	\node[label=right:{$\binom{5}{2}$ \; $\frac{2 m - 5}{m - 1} \rho$}] at (7.5, 3.8) {};
+	\node[label=right:{$\binom{4}{2}$ \; $\frac{2 m - 4}{m -1} \rho$}] at (7.5, 2.75) {};
+	\node[label=right:{$\binom{3}{2}$ \; $\frac{2 m - 3}{m - 1} \rho$}] at (7.5, 2.05) {};
+	\node[label=right:{$\binom{2}{2}$ \; $2 \rho$}] at (7.5, 1.45) {};
+	\node[label=right:{$\binom{3}{2}$ \; $3 \rho$}] at (7.5, 0.6) {};
+
+	% Gray dashed lines to visually separate holding times
+	\draw[color=gray, dashed] (7.5, 1.2) -- (11, 1.2);
+	\draw[color=gray, dashed] (7.5, 1.7) -- (11, 1.7);
+	\draw[color=gray, dashed] (7.5, 2.4) -- (11, 2.4);
+	\draw[color=gray, dashed] (7.5, 3.1) -- (11, 3.1);
+	\draw[color=gray, dashed] (7.5, 4.5) -- (11, 4.5);
+	\draw[color=gray, dashed] (7.5, 5.4) -- (11, 5.4);
+	\draw[color=gray, dashed] (7.5, 6) -- (11, 6);
+	\draw[color=gray, dashed] (7.5, 7.2) -- (11, 7.2);
+\end{tikzpicture}
+&
+\begin{tikzpicture}
+	\node [anchor=north west] at (-3.5,9.3) {\LARGE \textbf{B}};
+	\draw (-0.4, -0.2) -- (0.4, -0.2) -- (0.4, -0.3) -- (-0.4, -0.3) -- (-0.4, -0.2);
+	\draw (-0.1, -0.1) -- (-0.1, -0.4);
+	\node [label=below:{\small $v$}] at (-0.1, -0.2) {};
+	\draw (3.6, -0.2) -- (4.4, -0.2) -- (4.4, -0.3) -- (3.6, -0.3) -- (3.6, -0.2);
+	\draw (5.6, -0.2) -- (6.4, -0.2) -- (6.4, -0.3) -- (5.6, -0.3) -- (5.6, -0.2);
+
+	\draw (4,0) -- (4, 1.2) -- (6, 1.2) -- (6,0);
+	\draw (4.6, 1) -- (5.4, 1) -- (5.4, 0.9) -- (4.6, 0.9) -- (4.6, 1);
+	\draw (5.1, 1.1) -- (5.1, 0.8);
+	\node [label=below:{\small $w$}] at (5.1, 1) {};
+
+	\draw (0, 0) -- (0, 1.7) -- (-1,1.7) -- (1,1.7);
+	\draw (-1.4, 1.5) -- (-1.1, 1.5) -- (-1.1, 1.4) -- (-1.4, 1.4) -- (-1.4, 1.5);
+	\draw (-1.1, 1.45) -- (-0.6, 1.45);
+	\draw (0.6, 1.45) -- (0.9, 1.45);
+	\draw (0.9, 1.5) -- (1.4, 1.5) -- (1.4, 1.4) -- (0.9, 1.4) -- (0.9, 1.5);
+	\draw (1.1, 1.6) -- (1.1, 1.3);
+	\node [label=below:{\small $w$}] at (1.1, 1.5) {};
+
+	\draw (5, 1.2) -- (5, 2.4) -- (4,2.4) -- (6,2.4);
+	\draw (3.6, 2.2) -- (4.1, 2.2) -- (4.1, 2.1) -- (3.6, 2.1) -- (3.6, 2.2);
+	\draw (4.1, 2.15) -- (4.4, 2.15);
+	\draw (5.6, 2.15) -- (6.1, 2.15);
+	\draw (6.1, 2.2) -- (6.4, 2.2) -- (6.4, 2.1) -- (6.1, 2.1) -- (6.1, 2.2);
+	\draw [color=red](5.8, 2.3) -- (5.8, 2.0);
+	\node [label={[red]below:{\small $y$}}] at (5.8, 2.1) {};
+
+	\draw (1, 1.7) -- (1, 3.1) -- (0,3.1) -- (2,3.1);
+	\draw (-0.4, 2.85) -- (0.1, 2.85);
+	\draw (0.1, 2.9) -- (0.4, 2.9) -- (0.4, 2.8) -- (0.1, 2.8) -- (0.1, 2.9);
+	\draw (1.6, 2.85) -- (1.9, 2.85);
+	\draw (1.9, 2.9) -- (2.1, 2.9) -- (2.1, 2.8) -- (1.9, 2.8) -- (1.9, 2.9);
+	\draw (2.1, 2.85) -- (2.4, 2.85);
+
+	\draw (-1, 1.7) -- (-1, 4.5) -- (0,4.5) -- (0,3.1);
+	\draw (-0.9, 4.3) -- (-0.6, 4.3) -- (-0.6, 4.2) -- (-0.9, 4.2) -- (-0.9, 4.3);
+	\draw (-0.6, 4.25) -- (-0.4, 4.25);
+	\draw (-0.4, 4.3) -- (-0.1, 4.3) -- (-0.1, 4.2) -- (-0.4, 4.2) -- (-0.4, 4.3);
+	\draw (-0.5, 4.4) -- (-0.5, 4.1);
+	\node [label=below:{\small $x$}] at (-0.5, 4.3) {};
+
+	\draw (6, 2.4) -- (6, 3.8) -- (7, 3.8);
+	\draw [color=red](6, 3.8) -- (5, 3.8);
+	\draw (4.6, 3.55) -- (5.4, 3.55);
+	\draw (6.6, 3.55) -- (7.1, 3.55);
+	\draw (7.1, 3.6) -- (7.4, 3.6) -- (7.4, 3.5) -- (7.1, 3.5) -- (7.1, 3.6);
+
+	\draw (4,2.4) -- (4,5.4) -- (2,5.4) -- (2,3.1);
+	\draw (2.6, 5.2) -- (2.9, 5.2) -- (2.9, 5.1) -- (2.6, 5.1) -- (2.6, 5.2);
+	\draw (2.9, 5.15) -- (3.4, 5.15);
+
+	\draw (-0.5, 4.5) -- (-0.5, 6) -- (-1.5,6) -- (0.5,6);
+	\draw (-1.9, 5.75) -- (-1.4, 5.75);
+	\draw (-1.4, 5.8) -- (-1.1, 5.8) -- (-1.1, 5.7) -- (-1.4, 5.7) -- (-1.4, 5.8);
+	\draw (0.1, 5.8) -- (0.4, 5.8) -- (0.4, 5.7) -- (0.1, 5.7) -- (0.1, 5.8);
+	\draw (0.4, 5.75) -- (0.9, 5.75);
+
+	\draw [color=red](5, 3.8) -- (5, 6.6) -- (4, 6.6);
+	\draw (4,6.6) -- (3,6.6) -- (3, 5.4);
+	\draw (3.6, 6.4) -- (3.9, 6.4) -- (3.9, 6.3) -- (3.6, 6.3) -- (3.6, 6.4);
+	\draw (3.9, 6.35) -- (4.4, 6.35);
+
+	\draw (0.5, 6) -- (0.5, 7.2) -- (4,7.2) -- (4,6.6);
+	\draw (1.85, 6.95) -- (2.65, 6.95);
+
+	\draw (-1.5, 6) -- (-1.5, 8);
+	\draw [color=red](2.25, 7.2) -- (2.25, 8);
+	\draw (7, 3.8) -- (7, 8);
+
+	% Dashed lines for start and end times
+	\draw[dashed] (-2, 0) -- (9.1, 0);
+	\node [label=left:{$t = 0$}] at (-2,0) {};
+	\draw[dashed] (-2, 8) -- (9.1, 8);
+	\node [label=left:{$t = T$}] at (-2,8) {};
+
+	% Numbers of extant ancestors and links, from top to bottom
+	\node[label=right:{CA \; RE}] at (7.5, 8.2) {};
+	\node[label=right:{$\binom{3}{2}$ \; $3 \rho$}] at (7.5, 7.6) {};
+	\node[label=right:{$\binom{4}{2}$ \; $4 \rho$}] at (7.5, 6.9) {};
+	\node[label=right:{$\binom{5}{2}$ \; $5 \rho$}] at (7.5, 6.3) {};
+	\node[label=right:{$\binom{4}{2}$ \; $4 \rho$}] at (7.5, 5.7) {};
+	\node[label=right:{$\binom{5}{2}$ \; $5 \rho$}] at (7.5, 4.95) {};
+	\node[label=right:{$\binom{6}{2}$ \; $6 \rho$}] at (7.5, 4.15) {};
+	\node[label=right:{$\binom{5}{2}$ \; $5 \rho$}] at (7.5, 3.45) {};
+	\node[label=right:{$\binom{4}{2}$ \; $4 \rho$}] at (7.5, 2.75) {};
+	\node[label=right:{$\binom{3}{2}$ \; $3 \rho$}] at (7.5, 2.05) {};
+	\node[label=right:{$\binom{2}{2}$ \; $2 \rho$}] at (7.5, 1.45) {};
+	\node[label=right:{$\binom{3}{2}$ \; $3 \rho$}] at (7.5, 0.6) {};
+
+	% Gray dashed lines to visually separate holding times
+	\draw[color=gray, dashed] (7.5, 1.2) -- (9.1, 1.2);
+	\draw[color=gray, dashed] (7.5, 1.7) -- (9.1, 1.7);
+	\draw[color=gray, dashed] (7.5, 2.4) -- (9.1, 2.4);
+	\draw[color=gray, dashed] (7.5, 3.1) -- (9.1, 3.1);
+	\draw[color=gray, dashed] (7.5, 3.8) -- (9.1, 3.8);
+	\draw[color=gray, dashed] (7.5, 4.5) -- (9.1, 4.5);
+	\draw[color=gray, dashed] (7.5, 5.4) -- (9.1, 5.4);
+	\draw[color=gray, dashed] (7.5, 6) -- (9.1, 6);
+	\draw[color=gray, dashed] (7.5, 6.6) -- (9.1, 6.6);
+	\draw[color=gray, dashed] (7.5, 7.2) -- (9.1, 7.2);
+\end{tikzpicture}
+\end{tabular}
+}
+\caption{(A)
+A realisation of the graph traversed by Hudson's algorithm started from a
+sample of three chromosomes with $m$ discrete sites each at time $t = 0$, and
+propagated until time $T$. The MRCA on the genetic interval $[v, w)$ is reached
+at event \noderef{b}, while that on $[0, v)$ is reached at event \noderef{c}.
+The non-ancestral segment $[v, w)$ above
+\noderef{a} contributes to the rate of effective recombinations because it
+is trapped between ancestral segments. The two columns titled CA and RE
+are the respective rates of mergers and recombinations when
+the recombination rate is $\rho$.
+(B) A corresponding realisation of a Big ARG, which augments Hudson's algorithm
+by tracking non-ancestral lineages. The result is a simpler state space and
+dynamics, at the cost of extra nodes and edges, highlighted in red, which do
+not affect the local tree at any site.
+Recombination positions are labelled alphabetically in time, and their ordering
+along the genome is $y < v < x < w$, of which the first only appears in panel B.
+There are two separate recombination events at link $w$.
+}
+\label{hudson_vs_bigARG}
+\end{figure}
+
+Figure \ref{hudson_vs_bigARG} illustrates the more complex state space
+of the Little ARG process, as well as the extra events which occur in the Big ARG process.
+Moreover, it depicts the rates of common ancestors and recombination events in each
+interval of time of the realisations.
+In order to evaluate these rates, e.g.\ for likelihood-based inference
+\citep{baumdicker2021efficient, mahmoudi2022bayesian},
+it is necessary to know the number of lineages and number of extant links
+available for recombination in each time interval.
+Some representations may not provide this information.
+For example, in the gARG encoding depicted in Figure \ref{fig-ancestry-resolution}C,
+it is clear that a recombination takes place between nodes \noderef{i}, \noderef{k} and \noderef{j}.
+But the exact time of the recombination event is ambiguous: it could take place at any time between node \noderef{i} and its parents and produce the same gARG.
+Because a recombination increases the number of extant lineages by one (in the rootward direction of time), the number of lineages during the same time interval is ambiguous as well.
+In fact, this information cannot be recovered from the gARG encoding used in Figure \ref{fig-ancestry-resolution}C without an extrinsic convention.
+For the basic coalescent with recombination,
+it is sufficient to create \emph{two} gARG nodes at the time of the recombination event, with the interpretation that the two rootward edges from node \noderef{i} in Figure \ref{fig-ancestry-resolution}C belong to the same lineage until the time of nodes \noderef{k} and \noderef{j}, and split into two separate lineages at that time point.
+Similarly, the trapped, non-ancestral links along that lineage remain available for effective recombination (i.e.\ one which splits up ancestral material) for the same time interval.
+This interpretation is highlighted in
+Figure \ref{hudson_vs_bigARG} by drawing only one vertical edge between a recombinant child and its two parents.
+
+% Full quote:
+% This process simplifies mathematics on the account that the notion of an
+% ancestor will have a less restrictive meaning than usual:
+% An ``ancestral'' sequence in the birth and death process
+% need not have any genetic material in common with a
+% sequence descended from it.
+
 \section{Survey of ARG inference methods}
@@
 
-\section{The Big and Little ARG}
-\label{sec-big-and-little-arg}
-Here we review two important stochastic processes that construct ARGs:
-the ``Big" ARG process of \cite{griffiths1997ancestral}, and the ``Little" ARG process of
- \cite{hudson1983properties}. The Big ARG process is mathematically simpler
- but is computationally intractable due to generating a vast number of ancestors
- which contribute no genetic material to the initial sample.
-The Little ARG process avoids non-genetic ancestors at the cost of more complex
-dynamics and state space.
-We also demonstrate that applications relying on the grouping of inheritance pathways into ancestral lineages, such as likelihood-based inference under the coalescent, requires that the
-gARG (or eARG) data structure be interpreted in a model-specific way.
-
-A generic state of the Little ARG process consists of a finite collection of lineages $L$,
-each of which is a list of disjoint ancestry segments $(\ell, r, a)$, where
-$[\ell, r)$ is a half-closed genomic interval and $a$ is an integer
-tracking the number of samples to which the lineage is ancestral over that interval.
-We also usually track the node associated with each segment, but
-that is not important for our purposes here so we omit it to lighten notation.
-The initial condition for a sample of $n$ genomes of length $m$ consists of $n$ lineages
-of the form $\{(0, m, 1)\}$. The process traverses a series of common ancestor and
-recombination events backwards in time.
-Recombination events happen at rate $\rho \nu / (m - 1)$,
-where $\rho \geq 0$ is a per-genome recombination rate and
- \[
- \nu = \sum_{x \in L}\left( \max_{(\ell, r, a) \in x}r
-     - \min_{(\ell, r, a) \in x}\ell - 1 \right)
- \]
- is the number of available ``links" surrounded by ancestral material.
- At a recombination event we choose one of these links uniformly and break it,
- replacing the original lineage in $L$ with two new lineages containing the ancestral material
- to the left and right of the break point, respectively.
-
-Common ancestor events occur at rate $\binom{|L|}{2}$.
-In a common ancestor event, two uniformly sampled lineages have their segments
-merged into a single ancestor lineage, which is added to $L$.
-If the lineages have overlapping intervals of ancestry,
-say, $(\ell, r, a_1)$ and $(\ell, r, a_2)$, a
-\emph{coalescence} occurs. The result is a segment
-$(\ell, r, a_1 + a_2)$, and if $a_1 + a_2 < n$ it is included in the
-ancestor lineage. Otherwise, if $a_1 + a_2 = n$, we have found
-the most recent common ancestor of all samples in the interval $[\ell, r)$
-and do not need to simulate its history any further.
-Non-overlapping intervals from the two lineages are included
- in the ancestor lineage without changes. Eventually,
-we find resultant lineages in which all segments have fully coalesced,
-and so the number of extant lineages gradually falls to zero.
-
-In the Griffiths formulation (the Big ARG process), each edge in the graph corresponds to an extant
-lineage and nodes are events in the process. The $n$ initial leaf nodes are
-sampling events. Common ancestor events occur at rate $\binom{|L|}{2}$.
-When a common ancestor event happens, two uniformly chosen lineages
-merge into a common ancestor lineage.
-Recombination events happen at rate $|L| \rho$. Here, we choose a lineage (i.e.\ edge) uniformly,
-and a breakpoint $0 < x < m$ uniformly on its genome. We terminate the edge at a
-node, record the breakpoint, and start two new edges from this node. The process
-then continues until there is only one lineage left (the Grand Most Recent
-Common Ancestor, GMRCA), which is guaranteed to
-happen in finite time because of the quadratic rate of coalescing vs.\ linear rate of branching.
-
-The state-space of the Big ARG process is much simpler than that of the Little ARG process,
-which greatly facilitates mathematical reasoning. This simplicity comes at a
-substantial cost, however, if we wish to use it as a practical means of
-simulating recombinant ancestries.
-The number of events in the Big ARG all the way back to the GMRCA
-is $O(e^\rho)$~\citep{griffiths1997ancestral}, whereas the number
-of events required to simulate the Little ARG is
-$O(\rho^2)$~\citep{hein2004gene,baumdicker2021efficient}.
-This disparity arises because the majority of the events in the Big ARG are
-recombination events which occur outside of ancestral material,
-and these do not have any bearing on the ancestry of the initial sample.
-Because we don't keep track of the distribution of ancestral material during the process,
-we generate a vastly larger graph.
-
-\begin{figure}
-\centering
-% FIXME this is a quick nasty hack to make the figure a bit smaller and
-% prevent flushing all the other figures to the end of the document.
-\scalebox{0.54}{
-\begin{tabular}{cc}
-\begin{tikzpicture}
-	\node [anchor=north west] at (-3.5,9.3) {\LARGE \textbf{A}};
-	\draw (-0.4, -0.2) -- (0.4, -0.2) -- (0.4, -0.3) -- (-0.4, -0.3) -- (-0.4, -0.2);
-	\draw (-0.1, -0.1) -- (-0.1, -0.4);
-	\node [label=below:{\small $v$}] at (-0.1, -0.2) {};
-	\draw (3.6, -0.2) -- (4.4, -0.2) -- (4.4, -0.3) -- (3.6, -0.3) -- (3.6, -0.2);
-	\draw (5.6, -0.2) -- (6.4, -0.2) -- (6.4, -0.3) -- (5.6, -0.3) -- (5.6, -0.2);
-
-	\draw (4,0) -- (4, 1.2) -- (6, 1.2) -- (6,0);
-	\draw (4.6, 1) -- (5.4, 1) -- (5.4, 0.9) -- (4.6, 0.9) -- (4.6, 1);
-	\draw (5.1, 1.1) -- (5.1, 0.8);
-	\node [label=below:{\small $w$}] at (5.1, 1) {};
-
-	\draw (0, 0) -- (0, 1.7) -- (-1,1.7) -- (1,1.7);
-	\draw (-1.4, 1.5) -- (-1.1, 1.5) -- (-1.1, 1.4) -- (-1.4, 1.4) -- (-1.4, 1.5);
-	\draw (-1.1, 1.45) -- (-0.6, 1.45);
-	\draw (0.6, 1.45) -- (0.9, 1.45);
-	\draw (0.9, 1.5) -- (1.4, 1.5) -- (1.4, 1.4) -- (0.9, 1.4) -- (0.9, 1.5);
-	\draw (1.1, 1.6) -- (1.1, 1.3);
-	\node [label=below:{\small $w$}] at (1.1, 1.5) {};
-
-	\draw (5, 1.2) -- (5, 2.4) -- (4,2.4) -- (6,2.4);
-	\draw (3.6, 2.2) -- (4.1, 2.2) -- (4.1, 2.1) -- (3.6, 2.1) -- (3.6, 2.2);
-	\draw (4.1, 2.15) -- (4.4, 2.15);
-	\draw (5.6, 2.15) -- (6.1, 2.15);
-	\draw (6.1, 2.2) -- (6.4, 2.2) -- (6.4, 2.1) -- (6.1, 2.1) -- (6.1, 2.2);
-
-	\draw (1, 1.7) -- (1, 3.1) -- (0,3.1) -- (2,3.1);
-	\draw (-0.4, 2.85) -- (0.1, 2.85);
-	\draw (0.1, 2.9) -- (0.4, 2.9) -- (0.4, 2.8) -- (0.1, 2.8) -- (0.1, 2.9);
-	\draw (1.6, 2.85) -- (1.9, 2.85);
-	\draw (1.9, 2.9) -- (2.1, 2.9) -- (2.1, 2.8) -- (1.9, 2.8) -- (1.9, 2.9);
-	\draw (2.1, 2.85) -- (2.4, 2.85);
-
-	\draw (-1, 1.7) -- (-1, 4.5) -- (0,4.5) -- (0,3.1);
-	\node [scale=0.2,label=above left:{\small \noderef{a}}] at (-0.5,4.5) {a};
-	\draw (-0.9, 4.3) -- (-0.6, 4.3) -- (-0.6, 4.2) -- (-0.9, 4.2) -- (-0.9, 4.3);
-	\draw (-0.6, 4.25) -- (-0.4, 4.25);
-	\draw (-0.4, 4.3) -- (-0.1, 4.3) -- (-0.1, 4.2) -- (-0.4, 4.2) -- (-0.4, 4.3);
-	\draw (-0.5, 4.4) -- (-0.5, 4.1);
-	\node [label=below:{\small $x$}] at (-0.5, 4.3) {};
-
-	\draw (4,2.4) -- (4,5.4) -- (2,5.4) -- (2,3.1);
-	\node [scale=0.2,label=above left:{\small \noderef{b}}] at (3,5.4) {b};
-	\draw (2.6, 5.2) -- (2.9, 5.2) -- (2.9, 5.1) -- (2.6, 5.1) -- (2.6, 5.2);
-	\draw (2.9, 5.15) -- (3.4, 5.15);
-
-	\draw (-0.5, 4.5) -- (-0.5, 6) -- (-1.5,6) -- (0.5,6);
-	\draw (-1.9, 5.75) -- (-1.4, 5.75);
-	\draw (-1.4, 5.8) -- (-1.1, 5.8) -- (-1.1, 5.7) -- (-1.4, 5.7) -- (-1.4, 5.8);
-	\draw (0.1, 5.8) -- (0.4, 5.8) -- (0.4, 5.7) -- (0.1, 5.7) -- (0.1, 5.8);
-	\draw (0.4, 5.75) -- (0.9, 5.75);
-
-	\draw (0.5, 6) -- (0.5, 7.2) -- (3,7.2) -- (3,5.4);
-	\node [scale=0.2,label=above:{\small \noderef{c}}] at (1.75,7.2) {c};
-	\draw (1.35, 6.95) -- (2.15, 6.95);
-
-	\draw (-1.5, 6) -- (-1.5, 8);
-	\draw (6, 2.4) -- (6, 8);
-
-	% Edge annotations above each event
-	\node [label=left:{\small$\{(0,m,1)\}$}] at (0.2,0.6) {};
-	\node [label=left:{\small$\{(0,m,1)\}$}] at (4.2,0.4) {};
-	\node [label=right:{\small$\{(0,m,1)\}$}] at (5.8,0.4) {};
-	\node [label=left:{\small$\{(0,m,2)\}$}] at (5.2,1.6) {};
-	\node [label=left:{\small$\{(0,v,1)\}$}] at (-0.8,2.3) {};
-	\node [label=right:{\small$\{(v,m,1)\}$}] at (0.8,2.1) {};
-	\node [label=right:{\small$\{(0,w,2)\}$}] at (3.8,3.9) {};
-	\node [label=left:{\small$\{(w,m,2)\}$}] at (6.2,5) {};
-	\node [label=right:{\small$\{(w,m,1)\}$}] at (-0.2,3.8) {};
-	\node [label=right:{\small$\{(v,w,1)\}$}] at (1.8,4.1) {};
-	\node [label=left:{\small$\{(0,v,1), (w,m,1)\}$}] at (-0.3,5.2) {};
-	\node [label=right:{\small$\{(0,v,2)\}$}] at (2.8,6.4) {};
-	\node [label=left:{\small$\{(w,m,1)\}$}] at (-1.3,7) {};
-	\node [label=right:{\small$\{(0,v,1)\}$}] at (0.3,6.4) {};
-
-	% Dashed lines for start and end times
-	\draw[dashed] (-2, 0) -- (11, 0);
-	\node [label=left:{$t = 0$}] at (-2,0) {};
-	\draw[dashed] (-2, 8) -- (11, 8);
-	\node [label=left:{$t = T$}] at (-2,8) {};
-
-	% Numbers of extant ancestors and links, from top to bottom
-	\node[label=right:{CA \; RE}] at (7.5, 8.2) {};
-	\node[label=right:{$\binom{2}{2}$ \; $\frac{2 m - 2 w - 2}{m - 1} \rho$}] at (7.5, 7.6) {};
-	\node[label=right:{$\binom{4}{2}$ \; $\frac{2 m + 2 v - 2 w - 4}{m - 1} \rho$}] at (7.5, 6.6) {};
-	\node[label=right:{$\binom{3}{2}$ \; $\frac{2 m + v - w - 3}{m - 1} \rho$}] at (7.5, 5.7) {};
-	\node[label=right:{$\binom{4}{2}$ \; $\frac{2 m + w - v - 4}{m - 1} \rho$}] at (7.5, 4.95) {};
-	\node[label=right:{$\binom{5}{2}$ \; $\frac{2 m - 5}{m - 1} \rho$}] at (7.5, 3.8) {};
-	\node[label=right:{$\binom{4}{2}$ \; $\frac{2 m - 4}{m -1} \rho$}] at (7.5, 2.75) {};
-	\node[label=right:{$\binom{3}{2}$ \; $\frac{2 m - 3}{m - 1} \rho$}] at (7.5, 2.05) {};
-	\node[label=right:{$\binom{2}{2}$ \; $2 \rho$}] at (7.5, 1.45) {};
-	\node[label=right:{$\binom{3}{2}$ \; $3 \rho$}] at (7.5, 0.6) {};
-
-	% Gray dashed lines to visually separate holding times
-	\draw[color=gray, dashed] (7.5, 1.2) -- (11, 1.2);
-	\draw[color=gray, dashed] (7.5, 1.7) -- (11, 1.7);
-	\draw[color=gray, dashed] (7.5, 2.4) -- (11, 2.4);
-	\draw[color=gray, dashed] (7.5, 3.1) -- (11, 3.1);
-	\draw[color=gray, dashed] (7.5, 4.5) -- (11, 4.5);
-	\draw[color=gray, dashed] (7.5, 5.4) -- (11, 5.4);
-	\draw[color=gray, dashed] (7.5, 6) -- (11, 6);
-	\draw[color=gray, dashed] (7.5, 7.2) -- (11, 7.2);
-\end{tikzpicture}
-&
-\begin{tikzpicture}
-	\node [anchor=north west] at (-3.5,9.3) {\LARGE \textbf{B}};
-	\draw (-0.4, -0.2) -- (0.4, -0.2) -- (0.4, -0.3) -- (-0.4, -0.3) -- (-0.4, -0.2);
-	\draw (-0.1, -0.1) -- (-0.1, -0.4);
-	\node [label=below:{\small $v$}] at (-0.1, -0.2) {};
-	\draw (3.6, -0.2) -- (4.4, -0.2) -- (4.4, -0.3) -- (3.6, -0.3) -- (3.6, -0.2);
-	\draw (5.6, -0.2) -- (6.4, -0.2) -- (6.4, -0.3) -- (5.6, -0.3) -- (5.6, -0.2);
-
-	\draw (4,0) -- (4, 1.2) -- (6, 1.2) -- (6,0);
-	\draw (4.6, 1) -- (5.4, 1) -- (5.4, 0.9) -- (4.6, 0.9) -- (4.6, 1);
-	\draw (5.1, 1.1) -- (5.1, 0.8);
-	\node [label=below:{\small $w$}] at (5.1, 1) {};
-
-	\draw (0, 0) -- (0, 1.7) -- (-1,1.7) -- (1,1.7);
-	\draw (-1.4, 1.5) -- (-1.1, 1.5) -- (-1.1, 1.4) -- (-1.4, 1.4) -- (-1.4, 1.5);
-	\draw (-1.1, 1.45) -- (-0.6, 1.45);
-	\draw (0.6, 1.45) -- (0.9, 1.45);
-	\draw (0.9, 1.5) -- (1.4, 1.5) -- (1.4, 1.4) -- (0.9, 1.4) -- (0.9, 1.5);
-	\draw (1.1, 1.6) -- (1.1, 1.3);
-	\node [label=below:{\small $w$}] at (1.1, 1.5) {};
-
-	\draw (5, 1.2) -- (5, 2.4) -- (4,2.4) -- (6,2.4);
-	\draw (3.6, 2.2) -- (4.1, 2.2) -- (4.1, 2.1) -- (3.6, 2.1) -- (3.6, 2.2);
-	\draw (4.1, 2.15) -- (4.4, 2.15);
-	\draw (5.6, 2.15) -- (6.1, 2.15);
-	\draw (6.1, 2.2) -- (6.4, 2.2) -- (6.4, 2.1) -- (6.1, 2.1) -- (6.1, 2.2);
-	\draw [color=red](5.8, 2.3) -- (5.8, 2.0);
-	\node [label={[red]below:{\small $y$}}] at (5.8, 2.1) {};
-
-	\draw (1, 1.7) -- (1, 3.1) -- (0,3.1) -- (2,3.1);
-	\draw (-0.4, 2.85) -- (0.1, 2.85);
-	\draw (0.1, 2.9) -- (0.4, 2.9) -- (0.4, 2.8) -- (0.1, 2.8) -- (0.1, 2.9);
-	\draw (1.6, 2.85) -- (1.9, 2.85);
-	\draw (1.9, 2.9) -- (2.1, 2.9) -- (2.1, 2.8) -- (1.9, 2.8) -- (1.9, 2.9);
-	\draw (2.1, 2.85) -- (2.4, 2.85);
-
-	\draw (-1, 1.7) -- (-1, 4.5) -- (0,4.5) -- (0,3.1);
-	\draw (-0.9, 4.3) -- (-0.6, 4.3) -- (-0.6, 4.2) -- (-0.9, 4.2) -- (-0.9, 4.3);
-	\draw (-0.6, 4.25) -- (-0.4, 4.25);
-	\draw (-0.4, 4.3) -- (-0.1, 4.3) -- (-0.1, 4.2) -- (-0.4, 4.2) -- (-0.4, 4.3);
-	\draw (-0.5, 4.4) -- (-0.5, 4.1);
-	\node [label=below:{\small $x$}] at (-0.5, 4.3) {};
-
-	\draw (6, 2.4) -- (6, 3.8) -- (7, 3.8);
-	\draw [color=red](6, 3.8) -- (5, 3.8);
-	\draw (4.6, 3.55) -- (5.4, 3.55);
-	\draw (6.6, 3.55) -- (7.1, 3.55);
-	\draw (7.1, 3.6) -- (7.4, 3.6) -- (7.4, 3.5) -- (7.1, 3.5) -- (7.1, 3.6);
-
-	\draw (4,2.4) -- (4,5.4) -- (2,5.4) -- (2,3.1);
-	\draw (2.6, 5.2) -- (2.9, 5.2) -- (2.9, 5.1) -- (2.6, 5.1) -- (2.6, 5.2);
-	\draw (2.9, 5.15) -- (3.4, 5.15);
-
-	\draw (-0.5, 4.5) -- (-0.5, 6) -- (-1.5,6) -- (0.5,6);
-	\draw (-1.9, 5.75) -- (-1.4, 5.75);
-	\draw (-1.4, 5.8) -- (-1.1, 5.8) -- (-1.1, 5.7) -- (-1.4, 5.7) -- (-1.4, 5.8);
-	\draw (0.1, 5.8) -- (0.4, 5.8) -- (0.4, 5.7) -- (0.1, 5.7) -- (0.1, 5.8);
-	\draw (0.4, 5.75) -- (0.9, 5.75);
-
-	\draw [color=red](5, 3.8) -- (5, 6.6) -- (4, 6.6);
-	\draw (4,6.6) -- (3,6.6) -- (3, 5.4);
-	\draw (3.6, 6.4) -- (3.9, 6.4) -- (3.9, 6.3) -- (3.6, 6.3) -- (3.6, 6.4);
-	\draw (3.9, 6.35) -- (4.4, 6.35);
-
-	\draw (0.5, 6) -- (0.5, 7.2) -- (4,7.2) -- (4,6.6);
-	\draw (1.85, 6.95) -- (2.65, 6.95);
-
-	\draw (-1.5, 6) -- (-1.5, 8);
-	\draw [color=red](2.25, 7.2) -- (2.25, 8);
-	\draw (7, 3.8) -- (7, 8);
-
-	% Dashed lines for start and end times
-	\draw[dashed] (-2, 0) -- (9.1, 0);
-	\node [label=left:{$t = 0$}] at (-2,0) {};
-	\draw[dashed] (-2, 8) -- (9.1, 8);
-	\node [label=left:{$t = T$}] at (-2,8) {};
-
-	% Numbers of extant ancestors and links, from top to bottom
-	\node[label=right:{CA \; RE}] at (7.5, 8.2) {};
-	\node[label=right:{$\binom{3}{2}$ \; $3 \rho$}] at (7.5, 7.6) {};
-	\node[label=right:{$\binom{4}{2}$ \; $4 \rho$}] at (7.5, 6.9) {};
-	\node[label=right:{$\binom{5}{2}$ \; $5 \rho$}] at (7.5, 6.3) {};
-	\node[label=right:{$\binom{4}{2}$ \; $4 \rho$}] at (7.5, 5.7) {};
-	\node[label=right:{$\binom{5}{2}$ \; $5 \rho$}] at (7.5, 4.95) {};
-	\node[label=right:{$\binom{6}{2}$ \; $6 \rho$}] at (7.5, 4.15) {};
-	\node[label=right:{$\binom{5}{2}$ \; $5 \rho$}] at (7.5, 3.45) {};
-	\node[label=right:{$\binom{4}{2}$ \; $4 \rho$}] at (7.5, 2.75) {};
-	\node[label=right:{$\binom{3}{2}$ \; $3 \rho$}] at (7.5, 2.05) {};
-	\node[label=right:{$\binom{2}{2}$ \; $2 \rho$}] at (7.5, 1.45) {};
-	\node[label=right:{$\binom{3}{2}$ \; $3 \rho$}] at (7.5, 0.6) {};
-
-	% Gray dashed lines to visually separate holding times
-	\draw[color=gray, dashed] (7.5, 1.2) -- (9.1, 1.2);
-	\draw[color=gray, dashed] (7.5, 1.7) -- (9.1, 1.7);
-	\draw[color=gray, dashed] (7.5, 2.4) -- (9.1, 2.4);
-	\draw[color=gray, dashed] (7.5, 3.1) -- (9.1, 3.1);
-	\draw[color=gray, dashed] (7.5, 3.8) -- (9.1, 3.8);
-	\draw[color=gray, dashed] (7.5, 4.5) -- (9.1, 4.5);
-	\draw[color=gray, dashed] (7.5, 5.4) -- (9.1, 5.4);
-	\draw[color=gray, dashed] (7.5, 6) -- (9.1, 6);
-	\draw[color=gray, dashed] (7.5, 6.6) -- (9.1, 6.6);
-	\draw[color=gray, dashed] (7.5, 7.2) -- (9.1, 7.2);
-\end{tikzpicture}
-\end{tabular}
-}
-\caption{(A)
-A realisation of the graph traversed by Hudson's algorithm started from a
-sample of three chromosomes with $m$ discrete sites each at time $t = 0$, and
-propagated until time $T$. The MRCA on the genetic interval $[v, w)$ is reached
-at event \noderef{b}, while that on $[0, v)$ is reached at event \noderef{c}.
-The non-ancestral segment $[v, w)$ above
-\noderef{a} contributes to the rate of effective recombinations because it
-is trapped between ancestral segments. The two columns titled CA and RE
-are the respective rates of mergers and recombinations when
-the recombination rate is $\rho$.
-(B) A corresponding realisation of a Big ARG, which augments Hudson's algorithm
-by tracking non-ancestral lineages. The result is a simpler state space and
-dynamics, at the cost of extra nodes and edges, highlighted in red, which do
-not affect the local tree at any site.
-Recombination positions are labelled alphabetically in time, and their ordering
-along the genome is $y < v < x < w$, of which the first only appears in panel B.
-There are two separate recombination events at link $w$.
-}
-\label{hudson_vs_bigARG}
-\end{figure}
-
-Figure \ref{hudson_vs_bigARG} illustrates the more complex state space
-of the Little ARG process, as well as the extra events which occur in the Big ARG process.
-Moreover, it depicts the rates of common ancestors and recombination events in each
-interval of time of the realisations.
-In order to evaluate these rates, e.g.\ for likelihood-based inference
-\citep{baumdicker2021efficient, mahmoudi2022bayesian},
-it is necessary to know the number of lineages and number of extant links
-available for recombination in each time interval.
-Some representations may not provide this information.
-For example, in the gARG encoding depicted in Figure \ref{fig-ancestry-resolution}C,
-it is clear that a recombination takes place between nodes \noderef{i}, \noderef{k} and \noderef{j}.
-But the exact time of the recombination event is ambiguous: it could take place at any time between node \noderef{i} and its parents and produce the same gARG.
-Because a recombination increases the number of extant lineages by one (in the rootward direction of time), the number of lineages during the same time interval is ambiguous as well.
-In fact, this information cannot be recovered from the gARG encoding used in Figure \ref{fig-ancestry-resolution}C without an extrinsic convention.
-For the basic coalescent with recombination,
-it is sufficient to create \emph{two} gARG nodes at the time of the recombination event, with the interpretation that the two rootward edges from node \noderef{i} in Figure \ref{fig-ancestry-resolution}C belong to the same lineage until the time of nodes \noderef{k} and \noderef{j}, and split into two separate lineages at that time point.
-Similarly, the trapped, non-ancestral links along that lineage remain available for effective recombination (i.e.\ one which splits up ancestral material) for the same time interval.
-This interpretation is highlighted in
-Figure \ref{hudson_vs_bigARG} by drawing only one vertical edge between a recombinant child and its two parents.
-
-% Full quote:
-% This process simplifies mathematics on the account that the notion of an
-% ancestor will have a less restrictive meaning than usual:
-% An ``ancestral'' sequence in the birth and death process
-% need not have any genetic material in common with a
-% sequence descended from it.
-
 \section{Locally unary nodes}

### 2024-04-04 paper.tex
Finish reordering

@@
 
-% This is nice to have, but probably won't work in the final version
-% without section numbers. Let's deal with that when we get there I guess.
-% We begin in Section~\ref{sec-gARG} by providing a precise formal definition
-% of a gARG, illustrated by an example ARG embedded in pedigree.
-% We then provide a similar definition of the classical eARG
-% approach in Section~\ref{sec-eARG}, and consider its limitations
-% in the context of current datasets and research questions.
-% Following this, we discuss the important concept of ancestral material
-% in Section~\ref{sec-ancestral-material}, and how it relates
-% to the process of converting an eARG to a gARG.
-% We continue
-% in Section~\ref{sec-ARG-and-local-trees} by considering the relationship
-% between an ARG and its local trees. Contrary to the
-% prevailing view, we show that a suitably encoded sequence of local trees contains
-% precisely as much information as the corresponding ARG.
-% The gARG encoding opens a rich new set of details about ARGs,
-% including the ideas of locally unary nodes
-% (Section~\ref{sec-locally-unary-edges}),
-% the levels of detail that can be represented in an ARG
-% (Section~\ref{sec-ARG-simplification}),
-% and the degrees of precision about recombination
-% that can be stored and we may seek to infer (Section~\ref{sec-precision}).
-% These ideas have important practical considerations,
-% which we illustrate by examining the qualitative properties of
-% ARGs inferred by four recent methods for a classical benchmark
-% dataset in Section~\ref{sec-example-inferred-args}.
-% We then discuss how the gARG framework can be
-% efficiently implemented in Section~\ref{sec-efficiency},
-% and finish with an assessment of the key challenges facing
-% the field in the Discussion.
-% Finally, the literature on ARGs is large and
-% confusing, and we attempt to clarify some important aspects
-% in appendices, including
-% a brief history of ancestral graphs (Appendix~\ref{sec-arg-history}),
-% a description of the Big and Little ARG stochastic processes
-%  (Appendix~\ref{sec-big-and-little-arg}),
-% a survey of ARG inference methods (Appendix~\ref{sec-survey-arg-infer}),
-% and a discussion of ARGs at an individual vs cell
-% lineage level (Appendix~\ref{sec-cell-lineages-and-args}).
-
 \section{Genome ARGs}
@@
 (note these are shown in the conventional form in which only coalescences
-within the local tree are included; see Section~\ref{sec-ARG-and-local-trees}
+within the local tree are included; see Appendix~\ref{sec-ARG-and-local-trees}
 for a discussion of this important point).
@@
 sequence-based methods
-(see Section~\ref{sec-ARG-simplification} for
+(see Appendix~\ref{sec-ARG-simplification} for
 other applications of ARG simplification).
@@
 the scope of this manuscript.
-In the interest of avoiding confusion, however, we list a
-few minor details in which the formal details of gARGs
-provided in Section~\ref{sec-gARG} differ from their practical implementation in
-\texttt{tskit}.
-Firstly, ``edges'' in tree sequence terminology would perhaps be better
-described as ``edge-intervals'', as each describes a single contiguous
-interval of genome inheritance between a pair of nodes.
-This denormalisation of the gARG data model is for efficiency purposes.
-Secondly, zero- rather than one-based indexing is used for
-nodes in ARGs and oriented trees; consequently  $-1$ is used to denote the presence of
-roots (rather than $0$ as used here for notational simplicity).
+% In the interest of avoiding confusion, however, we list a
+% few minor details in which the formal details of gARGs
+% provided in Section~\ref{sec-gARG} differ from their practical implementation in
+% \texttt{tskit}.
+% Firstly, ``edges'' in tree sequence terminology would perhaps be better
+% described as ``edge-intervals'', as each describes a single contiguous
+% interval of genome inheritance between a pair of nodes.
+% This denormalisation of the gARG data model is for efficiency purposes.
+% Secondly, zero- rather than one-based indexing is used for
+% nodes in ARGs and oriented trees; consequently  $-1$ is used to denote the presence of
+% roots (rather than $0$ as used here for notational simplicity).
 
@@
 nodes to have more than two children (polytomies
-representing uncertainty over the ordering of coalescence events, Appendix~\ref{sec-cell-lineages-and-args}) or
+representing uncertainty over the ordering of coalescence events, 
+Appendix~\ref{sec-cell-lineages-and-args}) or
 more than two parents (representing uncertainty over the ordering
-of multiple recombination events, Section~\ref{sec-ARG-simplification}). Development of other methods to capture, for example,
+of multiple recombination events, Appendix~\ref{sec-ARG-simplification}). 
+Development of other methods to capture, for example,
 uncertainty about node ages and recombination breakpoint positions, is an important
@@
 data is sampled at many timepoints and recombination
-is rare (e.g.~SARS-CoV-2; see Discussion).
+is rare, e.g.~SARS-CoV-2~\citep{zhan2023towards}.
 
@@
 fraction of the node's span over which it is coalescent
-(Section~\ref{sec-locally-unary-edges}). For example, brown nodes
+(Appendix~\ref{sec-locally-unary-edges}). For example, brown nodes
 in Fig.~\ref{fig-inferred-args}A,B are clear because there is

### 2024-04-04 paper.tex
WIP

@@
 
-Note that the \citet{wiuf1999recombination} eARG
-in Fig.~\ref{fig-ancestry-resolution} is not particularly
-representative, because inference or simulation methods usually
-only generate ARGs containing nodes and edges ancestral to the sample
-(but see the discussion of the ``Big ARG'' stochastic process in
-Appendix~\ref{sec-big-and-little-arg}).
-Nonetheless, it is an instructive example from the literature which highlights several
-important properties of ARGs, and the general point about
-the need to resolve ancestral material ``on the fly'' for eARG traversals
-holds.
+% Note that the \citet{wiuf1999recombination} eARG
+% in Fig.~\ref{fig-ancestry-resolution} is not particularly
+% representative, because inference or simulation methods usually
+% only generate ARGs containing nodes and edges ancestral to the sample
+% (but see the discussion of the ``Big ARG'' stochastic process in
+% Appendix~\ref{sec-big-and-little-arg}).
+% Nonetheless, it is an instructive example from the literature which highlights several
+% important properties of ARGs, and the general point about
+% the need to resolve ancestral material ``on the fly'' for eARG traversals
+% holds.
 
 \section{A diversity of structures}
-
-There are a number of tools for inferring genetic genealogies,
-and the graph structures that they produce differ in important aspects.
-These varying representations have all been termed ``ARGs'' by different authors,
-and each describes more than a simple sequence of uncorrelated trees.
-By way of illustration, Fig.~\ref{fig-inferred-args} depicts the output of four
-different ``ARG inference'' tools run on a common dataset,
-with the resulting ARGs converted into a standard gARG format (cite tskit).
+A key goal of this perspective is to highlight the heterogeneity of
+the graph structures inferred by modern ARG inference methods.
+To illustrate this point, Fig.~\ref{fig-inferred-args} shows the 
+output of four methods run on the 
+\cite{kreitman1983nucleotide} dataset, often used as an
+ARG inference benchmark. One source of differences between 
+the inferences is the underlying model. For instance, 
+the ARG in Fig.~\ref{fig-inferred-args}A is inferred by 
+\kwarg\ under parsimony, and therefore has many fewer 
+ancestral nodes then Fig.~\ref{fig-inferred-args}B, which 
+is sampled from the coalescent with recombination with an 
+appropriate recombination rate parameter. These differences
+are well understood.
 
@@
 Results for four different methods:
-(A) \kwarg; (B) \argweaver; (C) \tsinfer; and (D) \relate, converted to the standard \emph{tskit} gARG encoding. See
+(A) \kwarg; (B) \argweaver; (C) \tsinfer; and (D) \relate, converted 
+to the standard \emph{tskit} gARG encoding. See
 Appendix \ref{sec-example-inferred-args} for details of these methods.
@@
 
-As is well-known (cite, perhaps Gusfield?), many different ancestral graph topologies are compatible with the same resulting
-sequence data. Hence it is unsurprising that the four graphs differ,
-for example in the numbers of nodes and their relationship to each other.
-However, more detailed inspection of Fig.~\ref{fig-inferred-args} reveals a
-diversity of representation that is only possible via the gARG encoding.
-For instance, although the most parsimonious ARG in Fig.~\ref{fig-inferred-args}A
-has far fewer ancestral events than an ARG instance sampled
-from from the coalescent with recombination in Fig.~\ref{fig-inferred-args}B,
-both consist only of nodes with either two parents or two children (but not both).
-These nodes correspond to recombination and common ancestor events in the equivalent eARG.
-In contrast, more scalable heuristic inference methods (Fig.~\ref{fig-inferred-args}C and D)
-create ARGs in which a node can simultaneously have more that than two children and more than two parents.
-Note also that the first three methods infer nodes that have a
-``coalescence span'' of less than 100\%, indicating that in the local trees, ancestral lineages exist 
-that pass through these nodes without the nodes marking a coalescent point. Such ``locally-unary''
-nodes are an notable feature of the local trees defined by an ARG
-(see Appendix \ref{sec-locally-unary-edges}).
-
-These differences, in particular the number of parents and children, and the coalescence span, reflect the
-degree of precision represented in the inferred ARGs. Indeed, it is important to note that the gARG encoding
-allows a single underlying genetic genealogy to be represented at
-differing levels of detail. The concept is illustrated in Fig.~\ref{fig-simplification},
-which shows different levels of ``simplification'' (Appendix~\ref{sec-ARG-simplification}) of the
-same underlying simulated ARG. 
-The full ARG, with all coalescent and recombination events represented by
-separate genomes, is shown in Fig.~\ref{fig-simplification}A. Simpler representations can be formed by removing ``unknowable'' nodes (Fig.~\ref{fig-simplification}B) and 
-collapsing multiple recombinations into a single child or multiple coalescences into
-a single parent (Fig.~\ref{fig-simplification}C, also see Appendix~\ref{sec-cell-lineages-and-args}).
-Finally, if we chose to represent only local coalescence points (Fig.~\ref{fig-simplification}D),
-we end up with a ``fully simplified'' ARG
-of the sort inferred in Fig.~\ref{fig-inferred-args}D. Note that in all cases, the equivalent
-trees only differ in the locally-unary nodes they contain; topological relationships
+The differences between ARGs that we wish to draw attention to here 
+are more subtle
+
+As is well-known (cite, perhaps Gusfield?), many different ancestral graph
+topologies are compatible with the same resulting sequence data. Hence it is
+unsurprising that the four graphs differ, for example in the numbers of nodes
+and their relationship to each other. However, more detailed inspection of
+Fig.~\ref{fig-inferred-args} reveals a diversity of representation that is only
+possible via the gARG encoding. For instance, although the most parsimonious
+ARG in Fig.~\ref{fig-inferred-args}A has far fewer ancestral events than an ARG
+instance sampled from from the coalescent with recombination in
+Fig.~\ref{fig-inferred-args}B, both consist only of nodes with either two
+parents or two children (but not both). These nodes correspond to recombination
+and common ancestor events in the equivalent eARG. In contrast, more scalable
+heuristic inference methods (Fig.~\ref{fig-inferred-args}C and D) create ARGs
+in which a node can simultaneously have more that than two children and more
+than two parents. Note also that the first three methods infer nodes that have
+a ``coalescence span'' of less than 100\%, indicating that in the local trees,
+ancestral lineages exist that pass through these nodes without the nodes
+marking a coalescent point. Such ``locally-unary'' nodes are an notable feature
+of the local trees defined by an ARG (see Appendix
+\ref{sec-locally-unary-edges}).
+
+These differences, in particular the number of parents and children, and the
+coalescence span, reflect the degree of precision represented in the inferred
+ARGs. Indeed, it is important to note that the gARG encoding allows a single
+underlying genetic genealogy to be represented at differing levels of detail.
+The concept is illustrated in Fig.~\ref{fig-simplification}, which shows
+different levels of ``simplification'' (Appendix~\ref{sec-ARG-simplification})
+of the same underlying simulated ARG. The full ARG, with all coalescent and
+recombination events represented by separate genomes, is shown in
+Fig.~\ref{fig-simplification}A. Simpler representations can be formed by
+removing ``unknowable'' nodes (Fig.~\ref{fig-simplification}B) and collapsing
+multiple recombinations into a single child or multiple coalescences into a
+single parent (Fig.~\ref{fig-simplification}C, also see
+Appendix~\ref{sec-cell-lineages-and-args}). Finally, if we chose to represent
+only local coalescence points (Fig.~\ref{fig-simplification}D), we end up with
+a ``fully simplified'' ARG of the sort inferred in
+Fig.~\ref{fig-inferred-args}D. Note that in all cases, the equivalent trees
+only differ in the locally-unary nodes they contain; topological relationships
 between the samples remain identical.

### 2024-04-05 paper.tex
Add link et al paper
Closes #446

@@
 caused by recombination, encapsulating the resulting complex web of genetic
-ancestry
-(see \citet{lewanski2024era,brandt2024promise} for biologically oriented
-introductions).
+ancestry.
 Recent breakthroughs
@@
 fan2023likelihood,
+link2023tree,
 grundler2024geographic,
@@
 differently (Appendix~\ref{sec-arg-history},\ref{sec-big-and-little-arg}).
-
-%Other subtly different concepts that we must be careful to distinguish
-%are the ``true'' ARG, % I don't see a better word for "true" here
-%describing the actual history of a sample of genomes,
-%and a ``population'' ARG which is
-%the true ARG of every individual in a population.
-%True ARGs are the underlying real history of a sample,
-%perfectly resolved into binary splits and mergers by
-%%the cellular processes of meiosis and mitosis, regardless
-%of sampling density or population processes
-%(Appendix~\ref{sec-cell-lineages-and-args}).
-%Although population-scale true ARGs unquestionably exist,
-%they can also never be entirely known, in part because of a fundamental
-%lack of mutational information.
-%Even if mutation rate were high enough to uniquely identify every
-%recent branching point,
-%such a high rate would saturate the genome with mutations
-%and obscure deeper history.
-
-% Population ARGs are here, now.
-%Population ARGs may seem fanciful, but the scale of
-%modern datasets makes it necessary for us to grapple with the idea.
-%The UK Biobank (UKB), for example, has
-%genotype data for around 500,000 humans~\citep{bycroft2018genome},
-%along with exome~\citep{backman2021exome} and
-%whole genome sequence~\citep{halldorsson2022sequences} data
-%for large subsets of the cohort.
-%UKB is just one of many such population-scale sequencing
-%projects~\citep[e.g.][]{turnbull2018hundred,
-%karczewski2020mutational,tanjo2021practical}.
-%Agricultural datasets are on a similar scale,
-%and also include dense multi-generational sampling
-%and near-perfect pedigree
-%information~\citep[e.g.][]{hayes20191000,Ros-Freixedes2020}.
-%Recent advances have made it possible to actually
-%\emph{estimate} ARGs at population scale:
-%ARGs have already been inferred for the 500,000
-%humans in UKB~\citep{kelleher2019inferring,zhang2023biobank}
-%and over a million SARS-CoV-2 genomes~\citep{zhan2023towards}.
-%While this new population-scale reality presents many
-%exciting opportunities, it also poses substantial challenges
-%to existing methodologies.
-
-% What's the problem?
-%A major problem currently facing the field is that
-%classical mathematical formalisms and terminology
-%cannot adequately describe these vast inferred ARGs.
-%Fundamentally, these formalisms assume that an
-%ARG is known in complete detail
-%and are not suited to describing partial or approximate knowledge.
-%As we are actively inferring ARGs at the population scale,
-%and such ARGs can never be known in complete detail,
-%there is currently a substantial gap between our theoretical frameworks
-%and practical application.
-
-The breakthroughs in scale achieved by recent
 methods are all based, in different ways, on inferring
@@
 that is sufficiently general.
+
+(see \citet{lewanski2024era,brandt2024promise} for biologically oriented
+introductions)
 
@@
 that we wish to perform on an ARG will require resolving
-the ancestral material with respect to a sample.
+the ancestral material with respect to a set of samples.
 % this is weak, but would be good to emphasise this point.

### 2024-04-05 paper.tex
Update introduction

@@
 differently (Appendix~\ref{sec-arg-history},\ref{sec-big-and-little-arg}).
-methods are all based, in different ways, on inferring
-approximate \emph{structures}
+% TODO rework these two sentences. Point about models vs structures 
+% is an important one, we need to make that.
+Recent large-scale methods infer approximate \emph{structures}
 instead of a complete and fully detailed history
-(Appendix~\ref{sec-survey-arg-infer}).
-It is important to distinguish here between
-structures and models: whether an inference method is based
+(Appendix~\ref{sec-survey-arg-infer}), and it is
+therefore important to distinguish between
+approximate structures and approximate models.
+Whether an inference method is based
 on heuristics or a rigorous mathematical model is orthogonal
 to the level of detail provided in its estimate.
-One could heuristically estimate a fully precise ARG,
-% this would be a cool thing to do!
-or statistically sample a partial,
-approximate ARG under a model
-such as the coalescent.
-Although the term ``ARG'' is now often used in a general
-sense~\citep[e.g.][]{mathieson2020ancestry,hejase2020summary,
+The term ``ARG'' therefore has many different meanings, varying 
+over time and depending on context. 
+There is an emerging consensus to use the term
+in quite a general sense~\citep[e.g.][]{mathieson2020ancestry,hejase2020summary,
 schaefer2021ancestral,harris2023using,zhang2023biobank,fan2023likelihood},
 informally
-encompassing the varied approximate structures output by modern simulation and
+encompassing the varied structures output by modern simulation and
 inference methods~\citep{rasmussen2014genome, palamara2016argon, haller2018tree,
 kelleher2019inferring, speidel2019method, baumdicker2021efficient,
-zhang2023biobank},
-there is no corresponding mathematical definition
-that is sufficiently general.
-
-(see \citet{lewanski2024era,brandt2024promise} for biologically oriented
-introductions)
-
-We highlight this problem and show how to address it by a simple formal definition of an
-ARG data structure, based on recording the intervals of
-genetic inheritance between specific genomes.
-We call this the ``genome ARG'', or gARG encoding. We
-contrast this with the classical formal definition of an ARG, based on
-recording common ancestor and recombination events, which we refer to as the
-``event ARG'' or eARG encoding. We show that the new gARG encoding is a
-substantial generalisation of the classical eARG approach,  providing much more
-flexibility in how genetic inheritance can be represented, and encompasses the
-outputs of modern methods. We show that the gARG approach can represent many
-different types of approximation, in particular allowing us to systematically
-describe uncertainty about the temporal ordering of multiple recombinations.
+zhang2023biobank}. 
+However, there is currently no formal definition or systematic 
+discussion that unifies these different structures, 
+ultimately stifling progress in this vibrant research area.
+
+In this perspective we provide a simple formal definition
+of an ARG data structure, which generalises classical 
+definitions and encompasses the output of modern simulation and inference 
+methods. We show that different levels of approximation
+are possible using this structure, illustrated via examples.
+% This is a bit crap, but do want to make this point about tskit for the 
+% "only read the introduction" crowd.
+The proposed ARG definition is the basis of the widely-used \texttt{tskit}
+library which provides a powerful software platform
+for ARG-based analysis and, we argue, would be a useful community standard.
+This perspective is intended for ``ARG practitioners'', who we hope
+will find the detailed examples and technical appendices useful.
+Readers seeking an introduction to ARGs and their applications are 
+directed to \citet{lewanski2024era}
+and \citet{brandt2024promise}.
 

### 2024-04-05 paper.tex
Cut some more stuff

@@
 or can be interpreted as representing one of the genomes of a
-potentially long-lived individual. All these interpretations
-are potentially useful, and equally valid under the assumptions
-of the gARG encoding.
+potentially long-lived individual.
 In many settings, nodes are dated, i.e.\ each
@@
 
-% % I feel like sophisticated readers would at this point be saying,
-% % ah, yeah, that's basically what an ARG is just in different words;
-% % what's the point of this paper?
-The genome ARG framework defined here is
-in many ways simply a clarification of existing treatments
-\cite[e.g.][]{mathieson2020ancestry,shipilina2023origin},
-adding concrete details to describe the
-differential inheritance of genetic material between genomes.
-% % This is definitely needed for the Nicks, who are always thinking
-% % about the deeper processes and are quite happy being loosey-goosey
-% % with representation details
-In the next section we contrast this gARG representation with a classical event-based (eARG) representation of the same underlying process.
+% % % I feel like sophisticated readers would at this point be saying,
+% % % ah, yeah, that's basically what an ARG is just in different words;
+% % % what's the point of this paper?
+% The genome ARG framework defined here is
+% in many ways simply a clarification of existing treatments
+% \cite[e.g.][]{mathieson2020ancestry,shipilina2023origin},
+% adding concrete details to describe the
+% differential inheritance of genetic material between genomes.
+% % % This is definitely needed for the Nicks, who are always thinking
+% % % about the deeper processes and are quite happy being loosey-goosey
+% % % with representation details
+% In the next section we contrast this gARG representation with a classical event-based (eARG) representation of the same underlying process.
 The choice between using a gARG or an eARG to represent the general process of ancestral inheritance has
-% Again, need to remind the reader that there's an actual point
-% to all this getting stuck into the details.
-important
-consequences not just when addressing the practical need to store and compare ARGs, but more fundamentally when establishing goals for inferring ARGs and evaluating the success of such inferences.
+% % Again, need to remind the reader that there's an actual point
+% % to all this getting stuck into the details.
+% important
+% consequences not just when addressing the practical need to store and compare ARGs, but more fundamentally when establishing goals for inferring ARGs and evaluating the success of such inferences.
 
@@
 for notable exceptions).
-This eARG encoding is the basis of the output formats created by multiple ARG inference tools
+This Event ARG (eARG) encoding is the basis of the output formats 
+created by multiple ARG inference tools
 \citep[e.g.][]{song2004minimum,song2005efficient,rasmussen2014genome,
@@
 % down.
-In this encoding there are two types of internal node in the graph. These
-represent the common ancestor and recombination events
+In this encoding there are two types of internal node in the graph,
+representing the common ancestor and recombination events
 in the history of a sample.
@@
 breakpoint in an eARG is a concise and elegant way of describing realisations
-of the coalescent, it is inherently limited when
-implemented literally. The eARG encoding explicitly models only
+of the coalescent, it has limitations.
+The eARG encoding explicitly models only
 two different types of event; thus anything that is not a single crossover
@@
 events, or by adding new types of event to the encoding. For example, gene
-conversion could be accommodated either by stipulating a third type of event
+% Graham asked for coalescences with GC papers, but this is the only one
+% I know of?
+conversion~\citep{wiuf2000coalescent} could 
+be accommodated either by stipulating a third type of event
 (annotated by two breakpoints and corresponding traversal conventions for
@@
 as datasets approach the population scale.
-As discussed in the introduction,
-the precise details of all events in these vast ARGs can never be
-known, so it is disadvantageous to use a data structure that enforces complete
-precision.
-
-There is also a certain clarity gained by explicitly modelling nodes
-in the inheritance graph as genomes.
-Outside of the context of a
-% The point being, events are perfectly well defined in the models
-% but not in reality
-mathematical model, an ``event'' is a slippery concept.
-For example, \emph{which} genome along a chain of cell divisions should be
-regarded as the one where an event occurred,
-or whether multiple coalescences
-within a single individual should be regarded as one or multiple events are
-debatable points (Appendix~\ref{sec-cell-lineages-and-args}).
-% If we want to have an ARG software ecosystem then such fireside
-% discussions won't help.
-% The gARG formulation avoids the need for such considerations,
-% so is more naturally compatible with forming an ecosystem of interoperable
-% inference and analysis methods.
+% As discussed in the introduction,
+% the precise details of all events in these vast ARGs can never be
+% known, so it is disadvantageous to use a data structure that enforces complete
+% precision.
+
+% There is also a certain clarity gained by explicitly modelling nodes
+% in the inheritance graph as genomes.
+% Outside of the context of a
+% % The point being, events are perfectly well defined in the models
+% % but not in reality
+% mathematical model, an ``event'' is a slippery concept.
+% For example, \emph{which} genome along a chain of cell divisions should be
+% regarded as the one where an event occurred,
+% or whether multiple coalescences
+% within a single individual should be regarded as one or multiple events are
+% debatable points (Appendix~\ref{sec-cell-lineages-and-args}).
+% % If we want to have an ARG software ecosystem then such fireside
+% % discussions won't help.
+% % The gARG formulation avoids the need for such considerations,
+% % so is more naturally compatible with forming an ecosystem of interoperable
+% % inference and analysis methods.
 
@@
 Hudson's seminal simulation
-algorithm~\citep{hudson1983testing,kelleher2016efficient},
-and the key distinguishing feature between the
-``Big'' and ``Little'' ARG stochastic processes
-(see Appendix~\ref{sec-big-and-little-arg}).
-The ability to \emph{store} resolved ancestral material
+algorithm~\citep{hudson1983testing,kelleher2016efficient}.
+The ability to store resolved ancestral material
 is also a key distinction between the eARG and gARG

### 2024-04-05 paper.tex
Minor updates

@@
 
+% NEEDS REWRITE
 \begin{abstract}
@@
 To illustrate this point, Fig.~\ref{fig-inferred-args} shows the 
-output of four methods run on the 
-\cite{kreitman1983nucleotide} dataset, often used as an
-ARG inference benchmark. One source of differences between 
+output of 
+\kwarg~\citep{ignatieva2021kwarg},
+\argweaver~\citep{rasmussen2014genome},
+\tsinfer~\citep{kelleher2019inferring}
+and \relate~\citep{speidel2019method}
+on the classical \cite{kreitman1983nucleotide} dataset. 
+One source of differences between 
 the inferences is the underlying model. For instance, 
-the ARG in Fig.~\ref{fig-inferred-args}A is inferred by 
-\kwarg\ under parsimony, and therefore has many fewer 
+the ARG in Fig.~\ref{fig-inferred-args}A is inferred 
+under parsimony, and therefore has many fewer 
 ancestral nodes then Fig.~\ref{fig-inferred-args}B, which 
 is sampled from the coalescent with recombination with an 
-appropriate recombination rate parameter. These differences
+appropriate recombination rate parameter. Such differences
 are well understood.
@@
 
-The differences between ARGs that we wish to draw attention to here 
+[WIP] The differences between ARGs that we wish to draw attention to here 
 are more subtle

### 2024-04-17 paper.tex
Add Korfman ref
Closes #454

@@
 tsambos2023link,
-tagami2024tstrait}.
+tagami2024tstrait,
+korfmann2024simultaneous}.
 The technical details of \texttt{tskit}, and how it provides an

### 2024-04-17 paper.tex
Pass through Discussion

@@
 grundler2024geographic,
-huang2024estimating}
+huang2024estimating,
+korfmann2024simultaneous,
+deraje2024inferring}
 and many more are sure to
@@
 In this perspective we provide a simple formal definition
-of an ARG data structure, which generalises classical 
+of an ARG data structure which generalises classical 
 definitions and encompasses the output of modern simulation and inference 
@@
 formally describe the output of inference methods, particularly
-as datasets approach the population scale.
-% As discussed in the introduction,
-% the precise details of all events in these vast ARGs can never be
-% known, so it is disadvantageous to use a data structure that enforces complete
-% precision.
+as datasets approach the population scale~\citep[e.g.][]{
+turnbull2018hundred, bycroft2018genome,hayes20191000,
+Ros-Freixedes2020,karczewski2020mutational,tanjo2021practical,
+halldorsson2022sequences}.
 
@@
 is a key concept in understanding the overall inheritance structure
-of an ARG (here, and throughout, we use the general term ``ARG'' when
-the details of the specific encoding are not important).
+of an ARG.
 It denotes the genomic intervals ancestral to a set of samples
@@
 \label{sec-efficiency}
-The gARG encoding defined here
-leads to highly efficient storage and processing of ARG data,
-and has already been in use for several years.
+The gARG encoding leads to highly efficient storage and processing of ARG data,
+and has been in use for several years.
 The succinct tree sequence data structure
@@
 \label{sec-discussion}
-Recent breakthroughs have finally made large-scale ARG inference
-feasible in practice, leading to a surge of interest
-in inference methods, their evaluation, and their application to biological questions.
-The prospect of ARGs being used routinely within population
-and statistical genetics is tantalising,
-but in reality there is substantial work to be done to
-enable this.
-A necessary first step is a degree of terminological clarity.
-As reviewed in Appendix~\ref{sec-arg-history}, the term
-``ancestral recombination graph" has several
-subtly different interpretations, depending on context.
-The trend to decouple ARGs from their original definition
-within the context of stochastic
-processes and instead use the term as a more general representation of any
-recombinant genetic ancestry seems useful.
-Our intent here is to clarify and systematise this decoupling. Thus
-we can think of an ARG as any structure that encodes the
-reticulate genetic ancestry of a sample of colinear sequences under
-the influence of recombination. The ``genome'' ARG (gARG) encoding
-is one way we can concretely
-define such recombinant ancestry, which we have shown is both
-flexible and efficient.
-The flexibility of the gARG encoding contrasts with the classical
-``event'' ARG (eARG) encoding, which is more limited in what can be described.
-Importantly, gARGs do not require fully precise estimates of
-ancestral recombination events,
-and allow us to directly express important forms of temporal uncertainty.
-
-Fully decoupling the general concept of an ARG from the coalescent
-with recombination (henceforth, ``coalescent'')
-is an important step.
-While the coalescent has proven to be a useful and
-robust
-model~\citep{wakeley2012gene,bhaskar2014distortion,nelson2020accounting},
-many modern datasets have properties that grossly
-violate its assumptions.
-One key assumption is that
-sample size $n$ is much less than the effective population size, $N_e$.
-Several human datasets now consist of hundreds of thousands of
-genomes~\citep{turnbull2018hundred, bycroft2018genome,
-karczewski2020mutational,tanjo2021practical,
-halldorsson2022sequences},
-and so sample size is an order of magnitude \emph{larger} than the
-usually assumed $N_e$ values.
-Agricultural datasets are an even more extreme departure from this
-assumption, with hundreds of thousands of samples embedded in
-multi-generational pedigrees~\citep{hayes20191000,Ros-Freixedes2020}
-and effective population sizes of 100 and even
-less~\citep{MacLeod2013,Makanjuola2020,Hall2016,Porcnic2016}.
-A model assuming a single $N_e$ would be a
-drastic over-simplification of course, but
-even if sufficiently complex demographic models~\citep{gower2022demes}
-encompassing hundreds of populations, explosive growth rates and myriad
-interconnections of migration, were somehow estimated and provided as input,
-ARGs sampled from the coalescent cannot capture the complexities
-of family structure in these
-datasets~\citep[e.g.][]{turnbull2018hundred,Ros-Freixedes2020}.
-Another core assumption of the coalescent model is that the genome
-(or at least the region under study) is short enough that the number of extant
-lineages remains much smaller than $N_e$ at all times.
-High-quality whole genome assemblies are now available
-for many species % Can we say something more concrete here?
-and projects are under way to obtain
-them for tens of thousands more~\citep{darwin2022sequence,lewin2022earth},
-and so we can expect inferred ARGs to routinely span
-large fractions of a chromosome.
-
-Recent large-scale methods have simplified the inference problem by
-making a single, deterministic best-guess at ARG
-inference~\citep{kelleher2019inferring,speidel2019method,zhang2023biobank,zhan2023towards}.
-Even under strict parsimony conditions and for small sample sizes, the
+% Recent breakthroughs have finally made large-scale ARG inference
+% feasible in practice, leading to a surge of interest
+% in inference methods, their evaluation, and their application to biological questions.
+% The prospect of ARGs being used routinely within population
+% and statistical genetics is tantalising,
+% but in reality there is substantial work to be done to
+% enable this.
+% A necessary first step is a degree of terminological clarity.
+% As reviewed in Appendix~\ref{sec-arg-history}, the term
+% ``ancestral recombination graph" has several
+% subtly different interpretations, depending on context.
+% The trend to decouple ARGs from their original definition
+% within the context of stochastic
+% processes and instead use the term as a more general representation of any
+% recombinant genetic ancestry seems useful.
+% Our intent here is to clarify and systematise this decoupling. Thus
+% we can think of an ARG as any structure that encodes the
+% reticulate genetic ancestry of a sample of colinear sequences under
+% the influence of recombination. The ``genome'' ARG (gARG) encoding
+% is one way we can concretely
+% define such recombinant ancestry, which we have shown is both
+% flexible and efficient.
+% The flexibility of the gARG encoding contrasts with the classical
+% ``event'' ARG (eARG) encoding, which is more limited in what can be described.
+% Importantly, gARGs do not require fully precise estimates of
+% ancestral recombination events,
+% and allow us to directly express important forms of temporal uncertainty.
+
+% Fully decoupling the general concept of an ARG from the coalescent
+% with recombination (henceforth, ``coalescent'')
+% is an important step.
+% While the coalescent has proven to be a useful and
+% robust
+% model~\citep{wakeley2012gene,bhaskar2014distortion,nelson2020accounting},
+% many modern datasets have properties that grossly
+% violate its assumptions.
+% One key assumption is that
+% sample size $n$ is much less than the effective population size, $N_e$.
+% Several human datasets now consist of hundreds of thousands of
+% genomes~\citep{turnbull2018hundred, bycroft2018genome,
+% karczewski2020mutational,tanjo2021practical,
+% halldorsson2022sequences},
+% and so sample size is an order of magnitude \emph{larger} than the
+% usually assumed $N_e$ values.
+% Agricultural datasets are an even more extreme departure from this
+% assumption, with hundreds of thousands of samples embedded in
+% multi-generational pedigrees~\citep{hayes20191000,Ros-Freixedes2020}
+% and effective population sizes of 100 and even
+% less~\citep{MacLeod2013,Makanjuola2020,Hall2016,Porcnic2016}.
+% A model assuming a single $N_e$ would be a
+% drastic over-simplification of course, but
+% even if sufficiently complex demographic models~\citep{gower2022demes}
+% encompassing hundreds of populations, explosive growth rates and myriad
+% interconnections of migration, were somehow estimated and provided as input,
+% ARGs sampled from the coalescent cannot capture the complexities
+% of family structure in these
+% datasets~\citep[e.g.][]{turnbull2018hundred,Ros-Freixedes2020}.
+% Another core assumption of the coalescent model is that the genome
+% (or at least the region under study) is short enough that the number of extant
+% lineages remains much smaller than $N_e$ at all times.
+% High-quality whole genome assemblies are now available
+% for many species % Can we say something more concrete here?
+% and projects are under way to obtain
+% them for tens of thousands more~\citep{darwin2022sequence,lewin2022earth},
+% and so we can expect inferred ARGs to routinely span
+% large fractions of a chromosome.
+
+Tremendous progress has been made in recent years on the long-standing
+problem of ARG inference, there is now a range of practically 
+applicable methods available.
+Methods targetting large-scale datasets tend to simplify
+the inference problem by
+making a single, deterministic 
+best-guess~\citep{kelleher2019inferring,speidel2019method,zhang2023biobank,zhan2023towards}.
+Even under strict parsimony conditions and for small sample sizes the
 % Citation? This is something people have figured out, right?
-number of plausible ARGs compatible with a given dataset is vast.
-Thus, although it is clearly an oversimplification to arbitrarily
-choose \emph{one} best guess,
-it is not clear that generating many guesses
-when sample sizes are large will achieve much.
-At the scale of millions of samples,
-we could only ever explore the tiniest corner of the incomprehensibly
-large space of plausible ARGs.
-Therefore, it is important to systematically
-describe and utilise uncertainty about ARG inference, and to
-incorporate uncertainty encountered during inference into the returned ARG.
-One approach, enabled by the gARG encoding described here, is to allow
-nodes to have more than two children (polytomies
-representing uncertainty over the ordering of coalescence events, 
-Appendix~\ref{sec-cell-lineages-and-args}) or
-more than two parents (representing uncertainty over the ordering
-of multiple recombination events, Appendix~\ref{sec-ARG-simplification}). 
+number of plausible ARGs compatible with a given dataset is vast,
+and it is therefore not clear that generating many guesses
+when sample sizes are large will achieve much in terms of capturing uncertainty.
+An alternative approach to 
+is to incorporate uncertainty encountered during inference into the returned ARG.
+The gARG encoding described here enables particular kinds of uncertainty
+to be incorporated directly into the topology:
+nodes that have more than two children (polytomies)
+represent uncertainty over the ordering of coalescence events 
+(Appendix~\ref{sec-cell-lineages-and-args}), 
+and those have have more than two parents 
+represent uncertainty over the ordering of multiple recombination events
+(Appendix~\ref{sec-ARG-simplification}). 
 Development of other methods to capture, for example,
@@
 aspect of future work.
-How this uncertainty can be utilised in downstream applications
-is an open question.
+% How this uncertainty can be utilised in downstream applications
+% is an open question.
 
@@
 
-This view of ARGs,
-decoupled from generative models and
-without the hard requirement
-of complete precision on all historical events, may clarify inference goals and improve
-methods for evaluation.
+% This view of ARGs,
+% decoupled from generative models and
+% without the hard requirement
+% of complete precision on all historical events, may clarify inference goals and improve
+% methods for evaluation.
+Another important avenue for future work is to develop improved methods 
+to evaluate and benchmark inference quality.
 In most cases,
@@
 In comparing tree-by-tree along the genome, the effects of recombination
-are incorporated in a rather indirect manner through the correlations
+are incorporated in an indirect manner through the correlations
 between the local trees, instead of directly taking into account

### 2024-04-17 paper.tex
Thanks Graham
Closes #451
Pass through "diversity" section

@@
 \section{A diversity of structures}
-A key goal of this perspective is to highlight the heterogeneity of
-the graph structures inferred by modern ARG inference methods.
-To illustrate this point, Fig.~\ref{fig-inferred-args} shows the 
-output of 
-\kwarg~\citep{ignatieva2021kwarg},
-\argweaver~\citep{rasmussen2014genome},
-\tsinfer~\citep{kelleher2019inferring}
-and \relate~\citep{speidel2019method}
-on the classical \cite{kreitman1983nucleotide} dataset. 
-One source of differences between 
-the inferences is the underlying model. For instance, 
-the ARG in Fig.~\ref{fig-inferred-args}A is inferred 
-under parsimony, and therefore has many fewer 
-ancestral nodes then Fig.~\ref{fig-inferred-args}B, which 
-is sampled from the coalescent with recombination with an 
-appropriate recombination rate parameter. Such differences
-are well understood.
-
 \begin{figure} \begin{center}
@@
 \end{figure}
-
-[WIP] The differences between ARGs that we wish to draw attention to here 
-are more subtle
-
-As is well-known (cite, perhaps Gusfield?), many different ancestral graph
-topologies are compatible with the same resulting sequence data. Hence it is
-unsurprising that the four graphs differ, for example in the numbers of nodes
-and their relationship to each other. However, more detailed inspection of
-Fig.~\ref{fig-inferred-args} reveals a diversity of representation that is only
-possible via the gARG encoding. For instance, although the most parsimonious
-ARG in Fig.~\ref{fig-inferred-args}A has far fewer ancestral events than an ARG
-instance sampled from from the coalescent with recombination in
-Fig.~\ref{fig-inferred-args}B, both consist only of nodes with either two
-parents or two children (but not both). These nodes correspond to recombination
-and common ancestor events in the equivalent eARG. In contrast, more scalable
-heuristic inference methods (Fig.~\ref{fig-inferred-args}C and D) create ARGs
-in which a node can simultaneously have more that than two children and more
-than two parents. Note also that the first three methods infer nodes that have
-a ``coalescence span'' of less than 100\%, indicating that in the local trees,
-ancestral lineages exist that pass through these nodes without the nodes
-marking a coalescent point. Such ``locally-unary'' nodes are an notable feature
-of the local trees defined by an ARG (see Appendix
-\ref{sec-locally-unary-edges}).
-
-These differences, in particular the number of parents and children, and the
-coalescence span, reflect the degree of precision represented in the inferred
-ARGs. Indeed, it is important to note that the gARG encoding allows a single
-underlying genetic genealogy to be represented at differing levels of detail.
-The concept is illustrated in Fig.~\ref{fig-simplification}, which shows
-different levels of ``simplification'' (Appendix~\ref{sec-ARG-simplification})
-of the same underlying simulated ARG. The full ARG, with all coalescent and
-recombination events represented by separate genomes, is shown in
-Fig.~\ref{fig-simplification}A. Simpler representations can be formed by
-removing ``unknowable'' nodes (Fig.~\ref{fig-simplification}B) and collapsing
-multiple recombinations into a single child or multiple coalescences into a
-single parent (Fig.~\ref{fig-simplification}C, also see
-Appendix~\ref{sec-cell-lineages-and-args}). Finally, if we chose to represent
-only local coalescence points (Fig.~\ref{fig-simplification}D), we end up with
-a ``fully simplified'' ARG of the sort inferred in
-Fig.~\ref{fig-inferred-args}D. Note that in all cases, the equivalent trees
-only differ in the locally-unary nodes they contain; topological relationships
-between the samples remain identical.
+A key goal of this perspective is to highlight the heterogeneity of
+the graph structures inferred by modern ARG inference methods.
+To illustrate this point, Fig.~\ref{fig-inferred-args} shows the 
+output of 
+\kwarg~\citep{ignatieva2021kwarg},
+\argweaver~\citep{rasmussen2014genome},
+\tsinfer~\citep{kelleher2019inferring}
+and \relate~\citep{speidel2019method}
+on the classical \cite{kreitman1983nucleotide} dataset. 
+The ARGs in Fig.~\ref{fig-inferred-args}A and B are precise 
+estimates (Appendix~\ref{sec-precision}),
+with each node corresponding to a common ancestor 
+or recombination event, or  equivalently, either having two children
+or two parents.
+In contrast the ARGs in Fig.~\ref{fig-inferred-args}C and D do not 
+have this clear-cut interpretation, and the nodes
+can simultaneously have more that than two children and more
+than two parents. Another dimension of variability among the ARGs
+is that the first three methods infer nodes that have
+a ``coalescence span''  greater than 0 and less than 100\%, indicating that there 
+are nodes that are ``locally unary'' (Appendix~\ref{sec-locally-unary-edges}),
+but mark a coalescence between lineages elsewhere along the sequence.
 
@@
 
-%%% Pasted from previous para
-
-The ability to represent an ARG to differing degrees of precision is a powerful
-feature of gARGs. In particular, when inferring ARGs from genome
+A key feature of the gARG encoding is that it enables these varying levels 
+of precision to be represented, and brings these nuanced features to light.
+These ideas are illustrated in Fig.~\ref{fig-simplification}, which shows
+different levels of ``simplification'' (Appendix~\ref{sec-ARG-simplification})
+of the same underlying simulated ARG. The full ARG, with all coalescent and
+recombination events represented by separate genomes, is shown in
+Fig.~\ref{fig-simplification}A. Simpler representations can be formed by
+removing ``unknowable'' nodes (Fig.~\ref{fig-simplification}B) and collapsing
+multiple recombinations into a single child or multiple coalescences into a
+single parent (Fig.~\ref{fig-simplification}C). 
+Finally, Fig.~\ref{fig-simplification}D is a ``fully simplified''
+ARG, in which only coalescences in local trees are retained.
+Note that while ARGs of this type (produced by default by the \texttt{msprime}
+simulator, for example) lack a significant level of detail, they still
+retain the key feature of shared node identity across local trees.
+
+This ability to represent an ARG to differing degrees of precision is a powerful
+feature. In particular, when inferring ARGs from genome
 sequencing data, the timing, positions, and even the number of recombination
@@
 allows this fundamental uncertainty in inference to be appropriately represented
-(Appendix~\ref{sec-precision}.
+(Appendix~\ref{sec-precision}).
+
 % For general
@@
 aspect of future work.
-% How this uncertainty can be utilised in downstream applications
-% is an open question.
+How this uncertainty can be utilised in downstream applications
+is an open question.
 
@@
 originally motivated this work.
+We are particularly grateful to Graham Coop for detailed feedback
+on and earlier version of the manuscript.
 % Ordered alphabetically by surname

### 2024-04-17 paper.tex
Updated abstract

@@
 inheritance is referred to as an ancestral recombination graph (ARG).
-New developments have made it possible to infer ARGs at scale,
-enabling many new applications in population and statistical genetics.
-This rapid progress has led to a diversity of ARG definitions and representations.
-% We show that a synthesis of these views can be achieved by thinking
-% in terms of genomes and transmission
-% intervals.
-% Standard mathematical formalisms,
-% "detailing" here is intended to convey the we keep details about these
-% events like recomb breakpoint.
-% based on exhaustively detailing the ``events'' that occur
-% in the history of a sample,
-%are insufficient to describe the outputs of current methods.
-We argue that classical ARG formulations, which assume
-that all events can be known and precisely estimated,
-are fundamentally unsuited to the realities
-of modern, population-scale datasets.
-We describe an alternative formulation that
-encompasses the outputs of recent methods and
-can capture the full richness of modern large-scale datasets.
-By defining this ARG encoding in terms of
-% we haven't define a "genome" yet, so need to give a bit of context
-specific genomes and their intervals of genetic inheritance,
-we avoid the need to exhaustively list (and estimate)
-\emph{all} events. The effects of multiple events can be
-aggregated in different ways, providing a natural way to express many
-forms of approximate and partial knowledge about the
-recombinant ancestry of a sample.
+% New developments have made it possible to infer ARGs at scale,
+% enabling many new applications in population and statistical genetics.
+% This rapid progress has led to a diversity of ARG definitions and representations.
+Classical formalisms have focused on mapping 
+coalescence and recombination events to the nodes in an ARG.
+We show that this approach is out of step with modern developments,
+which do not represent genetic inheritance in terms of these events 
+or explicitly infer them.
+We present a simple formalism that defines an ARG in terms 
+of specific genomes and their intervals of genetic inheritance,
+and show how it generalises these classical treatments and 
+encompasses the outputs of recent methods. We discuss 
+nuances arising from this more general structure, and argue that it
+forms an appropriate basis for a software standard in this rapidly 
+growing field.
 \end{abstract}

### 2024-04-17 paper.tex
Update

@@
 
-% NEEDS REWRITE
 \begin{abstract}
@@
 the genealogical trees for
-% surely saying nucleotides and "sample of sequences" is concrete enough
-% here? This paper isn't for a general audience, let's just say what
-% we're talking about in direct language understandable by population
-% geneticists.
 a sample of DNA sequences vary along the genome.

### 2024-04-17 paper.tex
Light editing up to App C

@@
 widespread use~\citep{hudson2002generating,kelleher2016efficient,
-baumdicker2021efficient}.
+kelleher2020coalescent,baumdicker2021efficient}.
 In the 1990s, Griffiths and colleagues revisited the
@@
 etheridge2009coalescent,
+baumdicker2014AGTG,
 gonzalezcasanova2018duality,
 koskela2019robust}.
-% Could mention the Ancestral Gene Transfer graph here too, within
-% something like "Similarly the Ancestral Gene Transfer Graph [cite]
-% and [other ancestral graphs, if they exist?] model genealogical
-% processes in bacteria as a graph.
 
@@
 has similarly been achieved through resorting to principled
-approximations~\citep{rasmussen2014genome,mahmoudi2022bayesian},
+approximations~\citep{rasmussen2014genome,mahmoudi2022bayesian,deng2024robust},
 or moving away from the coalescent with recombination altogether and seeking
@@
 wilton2015smc,griffiths2016coalescent},
-as well as, more commonly, the concrete realisation of ancestry from a
+as well as the concrete realisation of ancestry from a
 process~\citep[e.g.][]{gusfield2014recombinatorics,mathieson2020ancestry,brandt2021evaluation}.
@@
 We also demonstrate that applications relying on the grouping of inheritance pathways into ancestral lineages, such as likelihood-based inference under the coalescent, requires that the
-gARG (or eARG) data structure be interpreted in a model-specific way.
+ARG data structure be interpreted in a model-specific way.
 
@@
 
-In the Griffiths formulation (the Big ARG process), each edge in the graph corresponds to an extant
+In the Big ARG process each edge in the graph corresponds to an extant
 lineage and nodes are events in the process. The $n$ initial leaf nodes are

### 2024-04-18 paper.tex
Tweaks from Gregor

@@
 coalescence and recombination events to the nodes in an ARG.
-We show that this approach is out of step with modern developments,
+This approach is out of step with modern developments,
 which do not represent genetic inheritance in terms of these events 
@@
 \label{sec-eARG}
-A common, classical view of an ARG data structure,
+A classical view of an ARG data structure,
 described explicitly in several publications~\citep[e.g.][]{
@@
 material propagating through the graph
-backwards in time. In the region $[2, 7)$, there is a
+pastwards. In the region $[2, 7)$, there is a
 local coalescence where nodes \noderef{a} and \noderef{c}
@@
 \argweaver~\citep{rasmussen2014genome},
-\tsinfer~\citep{kelleher2019inferring}
+\tsinfer~\citep{kelleher2019inferring},
 and \relate~\citep{speidel2019method}
@@
 We are particularly grateful to Graham Coop for detailed feedback
-on and earlier version of the manuscript.
+on an earlier version of the manuscript.
 % Ordered alphabetically by surname

### 2024-04-18 paper.tex
Add SINGER to review
Closes #448

@@
 been of interest since the ARG was first defined. Early methods focused on
-finding \emph{parsimonious} ARGs, i.e.\ those with a minimal number of
+finding parsimonious ARGs, i.e.\ those with a minimal number of
 recombination events~\citep{hein1990reconstructing}. Two main approaches
@@
 approximate models \citep{didelot2010inference, heine2018bridging,
-hubisz2020mapping,hubisz2020inference, medina2020speeding}. The most scalable
-method, \texttt{ARGWeaver} \citep{rasmussen2014genome}, can be applied to
-dozens of mammal-like genomes \citep{hubisz2020inference}.
+hubisz2020mapping,hubisz2020inference, medina2020speeding,deng2024robust}. 
+The most scalable
+method, \texttt{SINGER}, can be applied to
+hundreds of human genomes~\citep{deng2024robust}.
 

### 2024-04-18 paper.tex
Small tweaks to appendices

@@
 Three lineages coalesce in the left-hand genome of individual $D_{10}$, but do
-so as the result of two successive bifurcations. This is \emph{necessarily} so,
+so as the result of two successive bifurcations. This is necessarily so,
 because the only known method of reproducing DNA is by (semi-conservative) duplication.
@@
 resolves all ambiguity, and makes the collection of local
-trees exactly equivalent to the corresponding ARG (see previous section).
+trees exactly equivalent to the corresponding ARG.
 Unary nodes are a vital link between ARGs and local trees, and we
 cannot fully reason about how a local tree is embedded in an ARG
-without them. As we see in the next two sections, both ARG and locally
-unary nodes occur in various scenarios, and are
-produced by a range of current inference methods.
+without them.
 
@@
 a recombination event, as is done in ~Fig.~\ref{fig-ancestry-resolution},
-or two to represent the parent genomes, as in Fig.~\ref{fig-simplification},  is usually not important.
-Either is possible in the gARG encoding, and the most convenient approach will vary by application
-(discussed e.g.\ in Appendix~\ref{sec-big-and-little-arg}).
+or two to represent the parent genomes, as in Fig.~\ref{fig-simplification},  
+is often not important.
+Either is possible in the gARG encoding, and 
+the most convenient approach will vary by application.
 Note also that node \noderef{k} in Fig.~\ref{fig-simplification} has three children.
-Polytomies like this are a natural feature of such a Wright-Fisher model
-(but see Appendix~\ref{sec-cell-lineages-and-args}).
+Polytomies like this are a natural feature of such a Wright-Fisher model.
+See Appendix~\ref{sec-cell-lineages-and-args} for a discussion of the 
+ultimate biological interpretation of these topological considerations.
 
@@
 
-An important consequence of simplifying ARGs to remove
-unary nodes in local trees is that we lose some information
-about recombination
-events. This is related to the amount of \emph{precision} about
-recombination events that we store and
-can hope to infer from sampled genomes, which is the topic of
-Appendix~\ref{sec-precision}.
-
 \section{Precision of recombination information}

### 2024-04-18 paper.tex
Fix figure ref
Closes #443

@@
 two node representation seen in
-Fig.~\ref{fig-simplification}A and Fig.~\ref{hudson_vs_bigARG},
+Fig.~\ref{fig-arg-in-pedigree}A and Fig.~\ref{hudson_vs_bigARG},
 whereas tagging the genome

### 2024-04-18 paper.tex

@@
 \section*{Acknowledgements}
-We are grateful to Nick Barton, Gideon Bradburd, Alex Lewanski,
+We are grateful to Nick Barton, Gideon Bradburd, 
+Castedo Ellerman, Alex Lewanski,
 Peter Ralph and Andrew Vaughn

### 2024-04-18 paper.tex

@@
 \section*{Data Availability}
-The public GitHub repository at
-\url{https://github.com/tskit-dev/what-is-an-arg-paper} can be
-used to reproduce all figures and tables in this paper.
-In particular this includes
-the ARG used in Fig.~\ref{fig-ancestry-resolution} as well as
-the simulation code and functions
-used to generate Fig.~\ref{fig-inferred-args} and Fig.~\ref{fig-simplification}.
-This includes the software versions, parameter settings, and
-(where necessary) functions to convert
-software outputs to the \texttt{tskit} gARG format.
+All code used to generate figures and run analyses is available on GitHub at
+\url{https://github.com/tskit-dev/what-is-an-arg-paper}.
 

### 2024-04-18 paper.tex

@@
 differently (Appendix~\ref{sec-arg-history},\ref{sec-big-and-little-arg}).
-% TODO rework these two sentences. Point about models vs structures 
-% is an important one, we need to make that.
-Recent large-scale methods infer approximate \emph{structures}
-instead of a complete and fully detailed history
-(Appendix~\ref{sec-survey-arg-infer}), and it is
-therefore important to distinguish between
-approximate structures and approximate models.
-Whether an inference method is based
-on heuristics or a rigorous mathematical model is orthogonal
-to the level of detail provided in its estimate.
+% % TODO rework these two sentences. Point about models vs structures 
+% % is an important one, we need to make that.
+% Recent large-scale methods infer approximate \emph{structures}
+% instead of a complete and fully detailed history
+% (Appendix~\ref{sec-survey-arg-infer}), and it is
+% therefore important to distinguish between
+% approximate structures and approximate models.
+% Whether an inference method is based
+% on heuristics or a rigorous mathematical model is orthogonal
+% to the level of detail provided in its estimate.
 The term ``ARG'' therefore has many different meanings, varying 
 over time and depending on context. 
-There is an emerging consensus to use the term
+There is, however, an emerging consensus to use the term
 in quite a general sense~\citep[e.g.][]{mathieson2020ancestry,hejase2020summary,
@@
 This perspective is intended for ``ARG practitioners'', who we hope
-will find the detailed examples and technical appendices useful.
+will find the detailed examples, technical appendices, and 
+comprehensive bibliography useful.
 Readers seeking an introduction to ARGs and their applications are 

### 2024-04-18 paper.tex

@@
 applicable methods available.
-Methods targetting large-scale datasets tend to simplify
+Methods targeting large-scale datasets tend to simplify
 the inference problem by

### 2024-04-18 cover-letter/cover-letter-resubmit.tex
Initial version of resubmission letter

\documentclass{letter}

\signature{Jerome Kelleher}

\address{Big Data Institute\\University of Oxford}
\begin{document}

\begin{letter}{GENETICS}

\opening{Dear Graham,}

I am writing on behalf of my coauthors to resubmit our 
manuscript entitled
\emph{A general and efficient representation of ancestral recombination graphs}
for consideration for publication as a Perspective.

We are deeply grateful for your detailed feedback, which has allowed us 
to identify our target audience and clarified the purpose of this article.
We are absolutely in agreement that it is more suited to the Perspective
format, and are very happy for it to be considered potentially appropriate
for publication in Genetics.

We have heavily revised the manuscript, substantially shortening it
and moving some more technical and descriptive section to appendices.
We have attached a detailed point-by-point response in the 
\texttt{response-to-reviewers.pdf} file, along with a 
latex-diff of the differences between the current and previous submissions.

Given your influence on the manuscript, we have included you in the
Acknowledgements, and thank you again for taking the time to give
such thoughtful and detailed feedback.

\closing{Sincerely,}

\end{letter}
\end{document}

### 2024-04-18 cover-letter/cover-letter-resubmit.tex
Update cover letter

@@
 We have heavily revised the manuscript, substantially shortening it
-and moving some more technical and descriptive section to appendices.
+and moving some more technical and descriptive sections to appendices.
 We have attached a detailed point-by-point response in the 
@@
 
-Given your influence on the manuscript, we have included you in the
+Given your major influence on the manuscript, we have included you in the
 Acknowledgements, and thank you again for taking the time to give
-such thoughtful and detailed feedback.
+such helpful, thoughtful and detailed feedback.
 

### 2024-04-18 response-to-reviewers.tex
Add response to reviewers

% LaTeX rebuttal letter example.
%
% Copyright 2019 Friedemann Zenke, fzenke.net
%
% Based on examples by Dirk Eddelbuettel, Fran and others from
% https://tex.stackexchange.com/questions/2317/latex-style-or-macro-for-detailed-response-to-referee-report
%
% Licensed under cc by-sa 3.0 with attribution required.

\documentclass[11pt]{article}
\usepackage[utf8]{inputenc}
\usepackage{lipsum} % to generate some filler text
\usepackage{fullpage}

% import Eq and Section references from the main manuscript where needed
% \usepackage{xr}
% \externaldocument{manuscript}

% package needed for optional arguments
\usepackage{xifthen}
% define counters for reviewers and their points
\newcounter{reviewer}
\setcounter{reviewer}{0}
\newcounter{point}[reviewer]
\setcounter{point}{0}

% This refines the format of how the reviewer/point reference will appear.
\renewcommand{\thepoint}{\thereviewer.\arabic{point}}

% command declarations for reviewer points and our responses
\newcommand{\reviewersection}{\stepcounter{reviewer} \bigskip \hrule
                  \section*{Reviewer \thereviewer}}

\newenvironment{point}
   {\refstepcounter{point} \bigskip \noindent {\textbf{Reviewer~Point~\thepoint} } ---\ }
   {\par }

\newcommand{\shortpoint}[1]{\refstepcounter{point}  \bigskip \noindent
    {\textbf{Reviewer~Point~\thepoint} } ---~#1\par }

\newenvironment{reply}
   {\medskip \noindent \begin{sf}\textbf{Reply}:\  }
   {\medskip \end{sf}}

\newcommand{\shortreply}[2][]{\medskip \noindent \begin{sf}\textbf{Reply}:\  #2
    \ifthenelse{\equal{#1}{}}{}{ \hfill \footnotesize (#1)}%
    \medskip \end{sf}}

\begin{document}

\section*{Response to the editor}
% General intro text goes here
Thank you for considering this manuscript for publication.

\subsection*{Specific comments}
% We address your specific comments below:
% % \reviewersection

\section*{Response to the reviewers}
% General intro text goes here
We thank the reviewers for their close reading of our manuscript and
insightful comments. In the following we address the points raised
in turn.

% Let's start point-by-point with Reviewer 1
\reviewersection

% Point one description
\begin{point}
STUFF
\end{point}

\begin{reply}
Responds
\end{reply}

\end{document}

### 2024-04-18 response-to-reviewers.tex

@@
 
+% Associate Editor's overview & review
+\reviewersection
+
+\begin{point}
+I gained a number of insights from reading the paper, however, I wonder who the
+intended audience of the paper is. It is not a research paper, indeed many of
+the technical details are given in other papers. Nor is it written at a
+particularly accessible level for people trying to understand ARGs and tree
+sequences.The paper is advocating for the adoption of "genome ARG (gARG)" and
+its simplified versions, but the authors need to be much clearer who they want
+to adopt this and in what level. Separating our structure for storing large
+ARGs estimated from data, from the ``true" ARG seems like a very useful argument
+to make but this gets lost a lot in the details.
+\end{point}
+\begin{reply}
+We have clarified the purpose of the article by explicitly identifying 
+``ARG practitioners'' as the intended audience. We hope that it may
+serve to clarify some points of confusion among this community,
+and provide a useful starting point into a large and confusing literature
+for those new to the field.
+\end{reply}
+
+\begin{point}
+The ``event ARG'' is hard for people to grapple with on first brush, however, in
+my view, the authors' presentation here risks making ARGs even less intuitive.
+\end{point}
+\begin{reply}
+We have attempted to simplify this by cutting out extraneous detail and 
+focusing on the most important points that distinguish an Event ARG
+from a Genome ARG.
+\end{reply}
+
+\begin{point}
+My view is that the authors should switch the article to being a "Perspective"
+or "Review" and shorten it by moving a lot of the technical detail to the
+appendices. I suspect this switch will help people understand that this is the
+author's take on a topic, and also help it be read through by a far broader
+audience. People interested in the details can then work through the appendices
+without getting lost in them on the first read through. I would be interested
+to see such an article published in Genetics, as I think it would be an
+important contribution and help clarify a complex field. I'm happy for the
+authors to argue for keeping different sections in the main paper, but in doing
+so they need to be more clearly justified to the reader in the paper.
+\end{point}
+\begin{reply}
+We have followed this approach, and kept only what we regard as sections
+essential to the main messages of the paper in the main text. Briefly,
+we clearly need the Genome ARGs and Event ARGs sections, and the
+Ancestral material and sample resolution section is critical to understanding 
+how they relate to each other. We have added a new ``A Diversity of
+Structures'' section, which highlights some important nuances possible
+when using gARGs as illustrated by Figures 4 and 5. We regard Section 6 as 
+vital, as it explicitly draws the connection between gARGs, eARGs and the 
+tskit library.
+\end{reply}
+
+\begin{point}
+The authors make a strong case that the data does not contain information about
+certain aspects of the ARG, that does not mean that probabilistic statements
+about these events cannot be made under particular models (or that these models
+cannot be learnt from data). Coalescent inference often involves integrating
+out events that we cannot observe in the data and dealing with great
+uncertainty in the timing of events. I think the authors should perhaps be more
+open to the view that for some applications we may want to build back in more
+events even if they are not "needed" in some minimum construction. At the
+moment it reads a little dogmatically, when I suspect that is not the authors
+true position.
+\end{point}
+\begin{reply}
+We have cut this text.
+\end{reply}
+
+\begin{point}
+It seems in general that many in the field, at least methods-savvy folks, are
+on board with tree sequences representation of data. If the "succinct tree
+sequence data structure (usually known as a "tree sequence" for brevity) is a
+practical gARG implementation", why not just leave "tree sequence" as the term
+we use. It is far less clear to me that we need to label yet another object an
+ARG, without muddying the definition of an ARG even further. More justification
+is needed to help the reader understand what the authors are arguing here.
+\end{point}
+\begin{reply}
+This was our initial position, but it has become clear that ``tree sequence''
+is regularly confused with ``a sequence of disconnected trees'' and there 
+are numerous inaccurate statements along these lines in the literature. 
+There was a deep-seated
+confusion about the relationship between what is classically considered an 
+ARG, and the structures that Relate, tsinfer and ARG-Needle produce. We feel
+strongly that the only way to resolve this is to show that ``tree sequences''
+\emph{are} ARGs, and that there needs to be \emph{some} terminology to distinguish
+the classical ``full'' ARG that is driven by events, and the more 
+general structure used by tskit. Thus, in order to tackle this 
+``tree sequences aren't proper ARGs'' sentiment, the terminology must contain
+``ARG''. Our hope is that the term Genome ARG will not be necessary
+in the future, and we will simply say ``ARG''.
+
+We hope that the shorter main text will make the requirement for a 
+simple general definition of an ARG data structure clear, without 
+needing to explicitly mention these concerns in the text.
+\end{reply}
+
+\begin{point}
+ The authors could also do with having a clearer separation of the "gARG" from the
+"fully simplified gARG" as for a lot of the paper it feels like they informally
+move back and forth of arguing the strengths of both without being fully clear
+on which they are talking about.
+\end{point}
+\begin{reply}
+We have clarified this point in several places by stating when we assume a 
+fully simplified ARG and not.
+\end{reply}
+
+\begin{point}
+Introduction
+
+The authors state that concepts : "that we must be careful to distinguish are
+the "true" ARG, describing the actual history of a sample of genomes, and a
+"population" ARG which is the
+true ARG of every individual in a population.". But given that they're both
+"true" ARGs, one for a sample and one for the whole population sample it's not
+clear what is that different here. Also from the description I think the
+authors mean the population ARG to not include every individual who has ever
+lived, but it's not clear.
+
+"population-scale true ARGs unquestionably exist, they can also never be
+entirely known," that's true of all sample ARGs
+
+"perfectly resolved into binary splits and mergers" - not clear which direction
+time is running in this statement.
+\end{point}
+\begin{reply}
+We have cut these passages.
+\end{reply}
+
+\begin{point}
+"For example, gene conversion could be accommodated either by stipulating a
+third type of event (annotated by two breakpoints and corresponding traversal
+conventions for recovering the local trees)" -Cite the coalescent with gene
+conversion papers
+\end{point}
+\begin{reply}
+Thanks for pointing out this omission.
+We have cited Wiuf and Hein, 2000, but are not aware of other coalescent 
+with gene conversion papers. We would be happy to cite others if you 
+provide references.
+\end{reply}
+
+\begin{point}
+"we will refer to this Griffiths encoding as the "event ARG" - It feels a
+little strange to attach a single person's name to an approach that the paper
+takes a fairly negative take on. I understand there's no ill feeling here, but
+it's not clear how well this comes across.
+\end{point}
+\begin{reply}
+Thanks for pointing this out --- you are absolutely right. We only mention
+Griffiths in a few places now, and avoid attaching his name to the structure.
+\end{reply}
+
+\begin{point}
+Briefly define "Big ARG" and "Little ARG" in the main text as they are referred
+to multiple times.
+\end{point}
+\begin{reply}
+They are no longer mentioned in the main text.
+\end{reply}
+
+\begin{point}
+Section 4 and Figure 3. - There's a lot of detail in this section to then get
+to this sentence at the end: "Note that the Wiuf and Hein (1999b) eARG in Fig.
+3 is not particularly representative, because inference or simulation methods
+usually only generate ARGs containing nodes and edges ancestral to the sample".
+So this feels like the authors have attached a lot of baggage to the eARG that
+it didn't actually have.
+\end{point}
+\begin{reply}
+We have cut this paragraph. 
+\end{reply}
+
+\begin{point}
+Sections 5 \& 6 similarly are fairly technical and it's not fully clear who they
+are aimed at, Could these sections be moved to the appendix?
+\end{point}
+\begin{reply}
+This has been done.
+\end{reply}
+
+\begin{point}
+Section 7 and 8: Again there is a lot of detail here, much of which amounts to
+a long description of Figure 4. 
+\end{point}
+\begin{reply}
+We have moved these to appendices.
+\end{reply}
+
+\begin{point}
+It is not fully clear if the authors want the
+bottom graph in Figure 4 to be the end product (a "fully simplified gARG"), if
+that is the case they need to state this up front and label it that way in the
+graph.
+\end{point}
+\begin{reply}
+We have added a clarification to the caption to say that this is `often
+described as ``fully simplified"'.
+\end{reply}
+
+\begin{point}
+ As I mention above, it's not really clear to me that labeling 4D an ARG
+is really useful, given how much has been stripped away.
+\end{point}
+\begin{reply}
+On this point we disagree. As discussed above and argued in the text, 
+we feel that it is imperative
+that there is a simple formal definition of an ARG that corresponds to
+informal usage. Having some arbitrary threshold of ``sufficient structure''
+for something to be considered a `proper' ARG can surely only lead to more
+confusion. The only reasonable path forward is for the term ARG to become
+general, and for the community to invent adjectives to classify
+the different types of ARG.
+\end{reply}
+
+\begin{point}
+Section 9: I'm not sure what the goal of this section is.
+\end{point}
+\begin{reply}
+This section has been moved to an appendix. It is intended to clarify
+the common features of existing ARG inference methods, as well as 
+highlight their differences.
+\end{reply}
+
+\begin{point}
+Discussion:
+"Fully decoupling the general concept of an ARG from the coalescent with
+recombination (hence-forth, "coalescent") is an important step." Not sure it
+matters whether the coalescent
+is a good or bad model for data. What matters is the separation of the idea of
+genetic genealogy from stochastic models that generate such structures. That
+separation occurred long before this paper, and I'm not sure that discussing
+the limitations of the stochastic model adds much to what is already a long
+paper. More generally, please work through the discussion with the aim of
+significantly shortening it.
+\end{point}
+\begin{reply}
+We have cut this passage and several others, substantially shortening 
+the discussion.
+\end{reply}
+
+\begin{point}
+\end{point}
+\begin{reply}
+\end{reply}
+
+\begin{point}
+\end{point}
+\begin{reply}
+\end{reply}
+
+\begin{point}
+\end{point}
+\begin{reply}
+\end{reply}
+
 \subsection*{Specific comments}

### 2024-04-18 cover-letter/cover-letter-resubmit.tex

@@
 \texttt{response-to-reviewers.pdf} file, along with a 
-latex-diff of the differences between the current and previous submissions.
+latex-diff of the differences between the current and previous submissions
+(which does a reasonable but not fully accurate job of capturing the large-scale
+movements of text).
 

### 2024-04-19 paper.tex
Address Graham's comments
Squash a few typos

@@
 i.e.\ are unary everywhere (e.g.\ \noderef{n} and \noderef{r}).
-(D) Rewrite edges to bypass nodes in local trees in which they are unary.
+(D) Rewrite edges to bypass nodes in local trees in which they are unary
+(often described as ``fully simplified'').
 In each case, the graph is shown on the left
@@
 (Appendix~\ref{sec-cell-lineages-and-args}), 
-and those have have more than two parents 
+and those that have more than two parents 
 represent uncertainty over the ordering of multiple recombination events
@@
 The \texttt{msprime} simulator, and the version of Hudson's algorithm described
-by~\citet{kelleher2016efficient}, produces ARGs
+by~\citet{kelleher2016efficient}, produce ARGs
 that are fully simplified (i.e., contain no locally unary nodes).

### 2024-04-19 paper.tex

@@
 zhang2023biobank}. 
-However, there is currently no formal definition or systematic 
+There is currently no formal definition or systematic 
 discussion that unifies these different structures,
-ultimately stifling progress in this vibrant research area.
+however, stifling progress in this vibrant research area.
 

### 2024-04-22 response-to-reviewers.tex
Tidy up response

@@
 % command declarations for reviewer points and our responses
-\newcommand{\reviewersection}{\stepcounter{reviewer} \bigskip \hrule
-                  \section*{Reviewer \thereviewer}}
+\newcommand{\reviewersection}{\stepcounter{reviewer} \bigskip \hrule}
+                  % \section*{Reviewer \thereviewer}}
 
 \newenvironment{point}
-   {\refstepcounter{point} \bigskip \noindent {\textbf{Reviewer~Point~\thepoint} } ---\ }
+   {\refstepcounter{point} \bigskip \noindent {\textbf{Point~\thepoint} } ---\ }
    {\par }
@@
 % General intro text goes here
-Thank you for considering this manuscript for publication.
+Thank you for considering this manuscript for publication.  We have addressed
+the points you and Reviewer 1 raised below in turn.
 
@@
 \reviewersection
+\section*{Associate Editor's comments}
 
@@
 particularly accessible level for people trying to understand ARGs and tree
-sequences.The paper is advocating for the adoption of "genome ARG (gARG)" and
+sequences.The paper is advocating for the adoption of ``genome ARG (gARG)" and
 its simplified versions, but the authors need to be much clearer who they want
@@
 \begin{point}
-My view is that the authors should switch the article to being a "Perspective"
-or "Review" and shorten it by moving a lot of the technical detail to the
+My view is that the authors should switch the article to being a ``Perspective"
+or ``Review" and shorten it by moving a lot of the technical detail to the
 appendices. I suspect this switch will help people understand that this is the
@@
 \begin{reply}
-We have followed this approach, and kept only what we regard as sections
+We agree that the article is much more suited to the ``Perspective'' format,
+and have heavily revised it following your suggestions.
+We have kept only what we regard as sections
 essential to the main messages of the paper in the main text. Briefly,
 we clearly need the Genome ARGs and Event ARGs sections, and the
-Ancestral material and sample resolution section is critical to understanding 
+``Ancestral material and sample resolution'' section is critical to understanding 
 how they relate to each other. We have added a new ``A Diversity of
@@
 open to the view that for some applications we may want to build back in more
-events even if they are not "needed" in some minimum construction. At the
+events even if they are not ``needed" in some minimum construction. At the
 moment it reads a little dogmatically, when I suspect that is not the authors
@@
 It seems in general that many in the field, at least methods-savvy folks, are
-on board with tree sequences representation of data. If the "succinct tree
-sequence data structure (usually known as a "tree sequence" for brevity) is a
-practical gARG implementation", why not just leave "tree sequence" as the term
+on board with tree sequences representation of data. If the ``succinct tree
+sequence data structure (usually known as a ``tree sequence" for brevity) is a
+practical gARG implementation", why not just leave ``tree sequence" as the term
 we use. It is far less clear to me that we need to label yet another object an
@@
 \begin{point}
- The authors could also do with having a clearer separation of the "gARG" from the
-"fully simplified gARG" as for a lot of the paper it feels like they informally
+ The authors could also do with having a clearer separation of the ``gARG" from the
+``fully simplified gARG" as for a lot of the paper it feels like they informally
 move back and forth of arguing the strengths of both without being fully clear
@@
 \begin{point}
-Introduction
-
-The authors state that concepts : "that we must be careful to distinguish are
-the "true" ARG, describing the actual history of a sample of genomes, and a
-"population" ARG which is the
+(Introduction)
+The authors state that concepts : ``that we must be careful to distinguish are
+the ``true" ARG, describing the actual history of a sample of genomes, and a
+``population" ARG which is the
 true ARG of every individual in a population.". But given that they're both
-"true" ARGs, one for a sample and one for the whole population sample it's not
+``true" ARGs, one for a sample and one for the whole population sample it's not
 clear what is that different here. Also from the description I think the
@@
 
-"population-scale true ARGs unquestionably exist, they can also never be
+``population-scale true ARGs unquestionably exist, they can also never be
 entirely known," that's true of all sample ARGs
 
-"perfectly resolved into binary splits and mergers" - not clear which direction
+``perfectly resolved into binary splits and mergers" - not clear which direction
 time is running in this statement.
@@
 \begin{point}
-"For example, gene conversion could be accommodated either by stipulating a
+``For example, gene conversion could be accommodated either by stipulating a
 third type of event (annotated by two breakpoints and corresponding traversal
@@
 \begin{point}
-"we will refer to this Griffiths encoding as the "event ARG" - It feels a
+``we will refer to this Griffiths encoding as the ``event ARG" - It feels a
 little strange to attach a single person's name to an approach that the paper
@@
 \begin{point}
-Briefly define "Big ARG" and "Little ARG" in the main text as they are referred
+Briefly define ``Big ARG" and ``Little ARG" in the main text as they are referred
 to multiple times.
@@
 Section 4 and Figure 3. - There's a lot of detail in this section to then get
-to this sentence at the end: "Note that the Wiuf and Hein (1999b) eARG in Fig.
+to this sentence at the end: ``Note that the Wiuf and Hein (1999b) eARG in Fig.
 3 is not particularly representative, because inference or simulation methods
@@
 It is not fully clear if the authors want the
-bottom graph in Figure 4 to be the end product (a "fully simplified gARG"), if
+bottom graph in Figure 4 to be the end product (a ``fully simplified gARG"), if
 that is the case they need to state this up front and label it that way in the
@@
 general, and for the community to invent adjectives to classify
-the different types of ARG.
+the different types of ARG. As we point out in the introduction, 
+stuctures like this (the default output of msprime) are
+informally defined as ARGs by emerging consensus, and so the key point 
+of this manuscript is to provide a corresponding formal definition.
 \end{reply}
@@
 Discussion:
-"Fully decoupling the general concept of an ARG from the coalescent with
-recombination (hence-forth, "coalescent") is an important step." Not sure it
+``Fully decoupling the general concept of an ARG from the coalescent with
+recombination (hence-forth, ``coalescent") is an important step." Not sure it
 matters whether the coalescent
@@
 
+% Let's start point-by-point with Reviewer 1
+\reviewersection
+\section*{Reviewer 1}
+
 \begin{point}
+In this article, Wong et al. describe a data structure called the ``genome
+ARG''
+(gARG), designed to effectively represent information in ancestral
+recombination graphs. The authors have done an excellent job of clearly
+explaining the concept and offering a detailed comparison of different ARG
+definitions. The practicality of the gARG encoding is already well recognized
+in the community through various applications, such as msprime, forward-time
+simulation, and the efficient computation of population genetic statistics.
 \end{point}
 \begin{reply}
+Thank you for the kind words, we are glad you find the exposition clear.
 \end{reply}
@@
 \begin{point}
+However, I have reservations about categorizing this work as a standalone
+``Investigation'' article. It seems to align more closely with the nature of a
+review article. While the authors claim to be ``proposing an alternative
+mathematical formulation," this characterization might be somewhat of an
+overstatement. The contribution, while valuable, appears to be more in the
+realm of summarizing and clarifying existing concepts rather than introducing a
+fundamentally new mathematical approach.
 \end{point}
 \begin{reply}
+We are absolutely in agreement, and have restructured the article as a
+``Perspective''.
 \end{reply}
@@
 \begin{point}
+The figures provided are generally helpful; however, the font size in some of
+them is quite small, making them difficult to read. For instance, the node
+labels in Figure 3C are impossible to read without zooming in maximally. I
+suggest increasing the font size in most figures to ensure clear legibility.
 \end{point}
 \begin{reply}
-\end{reply}
-
-\subsection*{Specific comments}
-% We address your specific comments below:
-% % \reviewersection
-
-\section*{Response to the reviewers}
-% General intro text goes here
-We thank the reviewers for their close reading of our manuscript and
-insightful comments. In the following we address the points raised
-in turn.
-
-% Let's start point-by-point with Reviewer 1
-\reviewersection
-
-% Point one description
-\begin{point}
-STUFF
-\end{point}
-
-\begin{reply}
-Responds
+We have modified Figure 3 to make it more legible, and will endeavour to 
+adapt the figures to be legible in the final published typesetting 
+(should it be accepted!).
 \end{reply}

### 2024-04-22 response-to-reviewers.tex

--- 2024-04-18 response-to-reviewers.tex
@@
 
-\begin{point}
-\end{point}
-\begin{reply}
-\end{reply}
-
-\begin{point}
-\end{point}
-\begin{reply}
-\end{reply}
-
-\begin{point}
-\end{point}
-\begin{reply}
-\end{reply}
-
 \subsection*{Specific comments}
@@
 
-% Point one description
 \begin{point}
-STUFF
+In this article, Wong et al. describe a data structure called the ``genome
+ARG''
+(gARG), designed to effectively represent information in ancestral
+recombination graphs. The authors have done an excellent job of clearly
+explaining the concept and offering a detailed comparison of different ARG
+definitions. The practicality of the gARG encoding is already well recognized
+in the community through various applications, such as msprime, forward-time
+simulation, and the efficient computation of population genetic statistics.
 \end{point}
+\begin{reply}
+Thank you for the kind words, we are glad you find the exposition clear.
+\end{reply}
 
+\begin{point}
+However, I have reservations about categorizing this work as a standalone
+``Investigation'' article. It seems to align more closely with the nature of a
+review article. While the authors claim to be "proposing an alternative
+mathematical formulation," this characterization might be somewhat of an
+overstatement. The contribution, while valuable, appears to be more in the
+realm of summarizing and clarifying existing concepts rather than introducing a
+fundamentally new mathematical approach.
+\end{point}
 \begin{reply}
-Responds
+We are absolutely in agreement, and have restructured the article as a
+``Perspective''.
+\end{reply}
+
+\begin{point}
+The figures provided are generally helpful; however, the font size in some of
+them is quite small, making them difficult to read. For instance, the node
+labels in Figure 3C are impossible to read without zooming in maximally. I
+suggest increasing the font size in most figures to ensure clear legibility.
+\end{point}
+\begin{reply}
+We have modified Figure 3 to make it more legible, and will endeavour to 
+adapt the figures to be legible in the final published typesetting 
+(should it be accepted!).
 \end{reply}

### 2024-04-22 paper.tex
Rough initial conversion to GENETICS

@@
-\documentclass{article}
-\usepackage[letterpaper,top=2cm,bottom=2.5cm,left=3cm,right=3cm,marginparwidth=1.75cm]{geometry}
-\usepackage[round]{natbib}
+\documentclass[9pt,twocolumn,twoside]{gsajnl}
+\articletype{inv} % article type
+
+% \usepackage[round]{natbib}
 \usepackage{amsmath,amssymb,amsfonts}%
-\usepackage{geometry}%
-\usepackage{color}
 \usepackage{graphicx}
 \usepackage{authblk}
-\usepackage{nameref}
-\usepackage[right]{lineno}
-\usepackage{subcaption}
 \usepackage{tikz}
@@
 \usepackage{url}
-\usepackage[symbol]{footmisc}
-\usepackage[hidelinks]{hyperref}
+% \usepackage[symbol]{footmisc}
+% \usepackage[hidelinks]{hyperref}
+
+% Packages from msprime 1.0 paper
+% \usepackage{amsmath,amssymb}%
+% \usepackage{graphicx}
+% \usepackage{authblk}
+% \usepackage{dsfont}
+% \usepackage{environ}%
+% \usepackage{caption}
+% \usepackage{subcaption}
+% \usepackage{tikz}
+% \usetikzlibrary{calc,positioning}
+% \usepackage{booktabs}
 
@@
 
-\begin{document}
-
-\title{\vspace{-1.5em} \bf
-A general and efficient representation of ancestral recombination graphs}
+\title{A general and efficient representation of ancestral recombination graphs}
 
@@
 \author[1$\dagger$]{Jerome~Kelleher}
-\affil[ ]{\mbox{}\vspace{-2.5em}}
-
-\maketitle
-
-\setlength{\skip\footins}{1em}
-\setlength{\footnotemargin}{0.5em}
-\renewcommand{\thefootnote}{\arabic{footnote}}
-\footnotetext[1]{Big Data Institute, Li Ka Shing Centre for Health
-Information and Discovery, University of Oxford, UK}
-\footnotetext[2]{School of Mathematics and Statistics, University of Glasgow, UK}
-\footnotetext[3]{Department of Statistics, University of Oxford, UK}
-\footnotetext[4]{School of Mathematics, Statistics and Physics, Newcastle University, UK}
-\footnotetext[5]{Department of Statistics, University of Warwick, UK}
-\footnotetext[6]{The Roslin Institute and Royal (Dick) School of Veterinary Studies, University of Edinburgh, UK}
-\footnotetext[7]{Broad Institute of MIT and Harvard, Cambridge, USA}
-\footnotetext[8]{Department of Genetics, Stanford University School of Medicine, Stanford, USA}
-\renewcommand{\thefootnote}{\fnsymbol{footnote}}
-\footnotetext[1]{Joint second author, listed alphabetically}
-\footnotetext[2]{Correspondence: jerome.kelleher@bdi.ox.ac.uk}
-
-\vspace{-1em}
+% \affil[ ]{\mbox{}\vspace{-2.5em}}
+
+\affil[1]{Big Data Institute, Li Ka Shing Centre for Health Information and Discovery, University of Oxford, OX3 7LF, UK}
+\affil[2]{School of Mathematics and Statistics, University of Glasgow, UK}
+\affil[3]{Department of Statistics, University of Oxford, UK}
+\affil[4]{School of Mathematics, Statistics and Physics, Newcastle University, UK}
+\affil[5]{Department of Statistics, University of Warwick, UK}
+\affil[6]{The Roslin Institute and Royal (Dick) School of Veterinary Studies, University of Edinburgh, UK}
+\affil[7]{Broad Institute of MIT and Harvard, Cambridge, USA}
+\affil[8]{Department of Genetics, Stanford University School of Medicine, Stanford, USA}
+
+\keywords{Ancestral Recombination Graphs}
+
+\runningtitle{Ancestral Recombination Graphs}
+\runningauthor{Wong \textit{et al.}}
 
@@
 
-\textbf{Keywords:} Ancestral recombination graphs
-
-\linenumbers
+\begin{document}
+\maketitle
+
+\thispagestyle{firststyle}
+\marginmark
+\firstpagefootnote
+
+% Use the \equalcontrib command to mark authors with equal
+% contributions, using the relevant superscript numbers
+% \equalcontrib{1}
+% \equalcontrib{2}
+\makeatletter
+\newcommand{\equalcontribfirst}[1]{\@authfootnote{#1}{Joint second author,
+listed alphabetically.}}
+\equalcontribfirst{$\star$} % {Denotes shared first authorship, listed alphabetically}
+% \equalcontriblast{$\dagger$} % {Denotes shared senior authorship, listed alphabetically}
+% % \equalcontrib{$\ddagger$}{Denotes corresponding author}
+
+% \footnotetext[1]{}
+\correspondingauthoraffiliation{$\dagger$}{\url{jerome.kelleher@bdi.ox.ac.uk}}
+\vspace{-33pt}% Only used for adjusting extra space in the left column of the first page
+
+% \textbf{Keywords:} Ancestral recombination graphs
+
+% \linenumbers
 \section{Introduction}
@@
 
-\begin{figure}
+\begin{figure*}
 \begin{center}
@@
 }
-\end{figure}
+\end{figure*}
 
@@
 
-\begin{figure}
+\begin{figure*}
 \centering
@@
 \section{A diversity of structures}
-\begin{figure} \begin{center}
+\begin{figure*} \begin{center}
 \includegraphics[width=\textwidth]{illustrations/inference.pdf} \end{center}
@@
 }
-\end{figure}
+\end{figure*}
 A key goal of this perspective is to highlight the heterogeneity of
@@
 
-\bibliographystyle{refstyle}
+% \bibliographystyle{refstyle}
 \bibliography{paper}
@@
 \label{hudson_vs_bigARG}
-\end{figure}
+\end{figure*}
 
@@
 \label{sec-cell-lineages-and-args}
-\begin{figure}
+\begin{figure*}
 \begin{center}
@@
 
-\begin{figure}[ht]
+\begin{figure*}[ht]
 	\begin{center}
@@
 	}
-\end{figure}
+\end{figure*}
 

### 2024-06-03 response-to-reviewers-2.tex

% LaTeX rebuttal letter example.
%
% Copyright 2019 Friedemann Zenke, fzenke.net
%
% Based on examples by Dirk Eddelbuettel, Fran and others from
% https://tex.stackexchange.com/questions/2317/latex-style-or-macro-for-detailed-response-to-referee-report
%
% Licensed under cc by-sa 3.0 with attribution required.

\documentclass[11pt]{article}
\usepackage[utf8]{inputenc}
\usepackage{lipsum} % to generate some filler text
\usepackage{fullpage}

% import Eq and Section references from the main manuscript where needed
% \usepackage{xr}
% \externaldocument{manuscript}

% package needed for optional arguments
\usepackage{xifthen}
% define counters for reviewers and their points
\newcounter{reviewer}
\setcounter{reviewer}{0}
\newcounter{point}[reviewer]
\setcounter{point}{0}

% This refines the format of how the reviewer/point reference will appear.
\renewcommand{\thepoint}{\thereviewer.\arabic{point}}

% command declarations for reviewer points and our responses
\newcommand{\reviewersection}{\stepcounter{reviewer} \bigskip \hrule}
                  % \section*{Reviewer \thereviewer}}

\newenvironment{point}
   {\refstepcounter{point} \bigskip \noindent {\textbf{Point~\thepoint} } ---\ }
   {\par }

\newcommand{\shortpoint}[1]{\refstepcounter{point}  \bigskip \noindent
    {\textbf{Reviewer~Point~\thepoint} } ---~#1\par }

\newenvironment{reply}
   {\medskip \noindent \begin{sf}\textbf{Reply}:\  }
   {\medskip \end{sf}}

\newcommand{\shortreply}[2][]{\medskip \noindent \begin{sf}\textbf{Reply}:\  #2
    \ifthenelse{\equal{#1}{}}{}{ \hfill \footnotesize (#1)}%
    \medskip \end{sf}}

\begin{document}

\section*{Response to the editor}
% General intro text goes here
We thank you again for a careful reading of this manuscript, and the further suggestions for improving its clarity and flow. We have addressed the points you raised below in turn.

% Associate Editor's overview & review
\reviewersection
\section*{Associate Editor's comments}

\begin{point}
My remaining broad concern is that the paper is still in places somewhat narrow about the goals of future ARG development. I certainly see the practical utility of dropping inference down to some minimum "knowable" structure that can be reconstructed using deterministic algorithms for very large datasets. However, probabilistic reconstructions of some form of ARG with more explicit events is also a reasonable goal moving forwards (e.g. for some applications we may want a subset of the recombination events explicitly included). There are a few places where the paper still comes across as overly dogmatic about the minimum "knowable" ARG being the only goal (although the discussion casts a broader view).
\end{point}
\begin{reply}
We have gone through the article and, in addition to the suggestions made below, have rephrased 
parts to make it clear that a gARG can be used to encode a \emph{variety} of ARG structures, whether events are or are not explicitly inferred by the reconstruction method. We specifically state at the end of \emph{A diversity of structures} that
\begin{quote}
A gARG can encode a diversity of ARG structures, including 
those where events \emph{are} recorded explicitly, and those where
they are treated as fundamentally uncertain and thus not explicitly inferred (Appendix H).
\end{quote}
\end{reply}

\begin{point}
Abstract: "This approach is out of step with modern developments, which do not represent genetic inheritance in terms of these events or explicitly infer them." So this is on the places where I feel like the authors state things too strongly. The authors, and some others, approaches have taken this path, but folks can agree that the gARG is a good idea and yet think that explicitly inferring details of recombination events is a `modern' goal.
\end{point}
\begin{reply}
We have changed this to "This approach is out of step with many modern developments, however,..."
\end{reply}

\begin{point}
"Broadly speaking, an ARG describes the different paths of genetic inheritance caused by recombination, encapsulating the resulting complex web of genetic ancestry " - add "of a set of samples". Also I'd say "genetic ancestors", as ancestry is tied up with genetic ancestry groups in peoples' minds.
\end{point}
\begin{reply}
Amended as suggested.
\end{reply}

\begin{point}
"We define a genome as the complete set of genetic material that a child inherits from one parent. A diploid individual therefore carries two genomes, one inherited from each parent (we assume diploids here for clarity, but the definitions apply to organisms of arbitrary ploidy). " -Excludes Y, mtDNA, and X as written, please revise, e.g. talk about autosomal genome.
\end{point}
\begin{reply}
Amended as suggested.
\end{reply}

\begin{point}
"The topology of a gARG specifies that genetic inheritance occurred between particular ancestors and descendants, " -struggle slightly with word "particular" here as the identity of the ancestors is not known. Deleting "particular" is likely sufficient.
\end{point}
\begin{reply}
Amended as suggested.
\end{reply}

\begin{point}
"This is sufficient to describe the effects of inheritance under any form of homologous recombination (such as multiple crossovers,..." -do you mean multiple crossovers during a single round of meiosis.
\end{point}
\begin{reply}
Yes - amended to clarify this.
\end{reply}

\begin{point}
"In this encoding there are two types of internal node in the graph, representing the common ancestor and recombination events in the history of a sample. " stipulate that these are most recent common ancestor events.
\end{point}
\begin{reply}
Amended as suggested.
\end{reply}

\begin{point}
"This approach assumes all events are knowable, and does not provide an obvious mechanism for either aggregating multiple events or expressing uncertainty about them. While this is not a problem when describing the results of simulations". -Maybe one way to flip this around would be to say that because it arose from tracking a particular stochastic process it has these properties. Also I don't think it assumes that all events are knowable, eg we could construct some parsimonious ARG or probabilistic ARG. If we wish to express uncertainty about events we usually give draws from the posterior etc. I agree that might be computational prohibitive with large samples etc, but it seems like place to take a broad view. This seems like a place to acknowledge that for some applications we might want to explicitly reconstruct the events.
\end{point}
\begin{reply}
We have rephrased this part to read
\begin{quote}
This approach necessitates that all events are recorded explicitly, and does not 
provide an obvious mechanism for either aggregating multiple events
or expressing uncertainty about them. While this is not a
problem when describing the results of simulations, for instance (where all details
are perfectly known), it is an issue when we wish to
formally describe the output of inference methods which do not
necessarily attempt to infer events that are not \emph{knowable} from the data, 
particularly as datasets approach the population scale...
\end{quote}
\end{reply}

\begin{point}
"A key feature of the gARG encoding is that it enables these varying levels of precision to be represented, and brings these nuanced features to light." -the word nuanced feels strange here.
\end{point}
\begin{reply}
We have deleted the second part of this sentence.
\end{reply}

\begin{point}
"Simpler representations can be formed by removing "unknowable" nodes (Fig. 5B)" -unknowable is vague here, do you mean bubbles along a single lineage?
\end{point}
\begin{reply}
We've added a clarification that this refers to nodes such as those in singly-connected graph components.
\end{reply}

\begin{point}
"The gARG encoding leads to highly efficient storage and processing of ARG data, "-As gARG has various levels of precision, perhaps this needs to state that the "gARG encoding can lead to..." or be more precise that this is a reduced precision level.
\end{point}
\begin{reply}
Amended as suggested to add "can lead to".
\end{reply}

\begin{point}
"The succinct tree sequence data structure (usually known as a "tree sequence" for brevity) is a practical gARG implementation focused on efficiency." - If the tree sequence is focused at a particular level of gARG simplification be precise about this.
\end{point}
\begin{reply}
We have left this sentence as is, since the tree sequence structure can record gARGs at various levels of simplification.
\end{reply}

\begin{point}
"Methods targeting large-scale datasets tend to simplify the inference problem by making a single, deterministic best-guess " --I think this is the best guess of the topology, and the uncertainty in times given the ARG is downstream of this. If so please clarify. Also I'd perhaps explicitly acknowledge Deng et al (SINGER), e.g. "deterministic best-guess of the topology (see Deng et al for parallel developments addressing uncertainty with somewhat small sample sizes)" or something like that. While these deterministic approaches are a strong way forward for human biobank scale data, it's good to be highlight parallel developments that might be key to other applications.
\end{point}
\begin{reply}
We have mentioned this as suggested.
\end{reply}

\end{document}

### 2024-06-03 paper.tex
Ana's changes

@@
 \usepackage{url}
-% \usepackage[symbol]{footmisc}
-% \usepackage[hidelinks]{hyperref}
 
@@
 coalescence and recombination events to the nodes in an ARG.
-This approach is out of step with modern developments,
+This approach is out of step with many modern developments, however,
 which do not represent genetic inheritance in terms of these events 
@@
 caused by recombination, encapsulating the resulting complex web of genetic
-ancestry.
+ancestors of a set of samples.
 Recent breakthroughs
@@
 % Q: will someone nitpick that we're assuming diploid obligate sexual species?
-therefore carries two genomes, one inherited from each parent (we assume diploids here
+therefore carries two genomes, one inherited from each parent (we assume diploids and consider nuclear autosomal DNA here
 for clarity, but the definitions apply to organisms of arbitrary ploidy).
@@
 The topology of a gARG specifies that genetic inheritance
-occurred between particular
+occurred between
 ancestors and descendants, but the graph connectivity
@@
 This is sufficient to describe the effects of inheritance under
-any form of homologous recombination (such as multiple crossovers,
+any form of homologous recombination (such as multiple crossovers during a single round of meiosis,
 gene conversion events, and many forms of bacterial and viral recombination).
@@
 In this encoding there are two types of internal node in the graph,
-representing the common ancestor and recombination events
+representing the most recent common ancestor and recombination events
 in the history of a sample.
@@
 
-\begin{figure*}
+\begin{figure}
 \centering
@@
 }
-\end{figure*}
+\end{figure}
 
@@
 recording events and their properties (e.g.\ the crossover breakpoint
-for a recombination event). This approach assumes all events are 
-\emph{knowable}, and does not
+for a recombination event). This approach 
+% assumes all events are \emph{knowable}, and does not
+necessitates that all events are recorded explicitly, and does not 
 provide an obvious mechanism for either aggregating multiple events
 or expressing uncertainty about them. While this is not a
-problem when describing the results of simulations (where all details
+problem when describing the results of simulations, for instance (where all details
 are perfectly known), it is an issue when we wish to
-formally describe the output of inference methods, particularly
+formally describe the output of inference methods which do not
+necessarily attempt to infer events that are not \emph{knowable} from the data, particularly
 as datasets approach the population scale~\citep[e.g.][]{
@@
 A key feature of the gARG encoding is that it enables these varying levels 
-of precision to be represented, and brings these nuanced features to light.
+of precision to be represented.
 These ideas are illustrated in Fig.~\ref{fig-simplification}, which shows
@@
 Fig.~\ref{fig-simplification}A. Simpler representations can be formed by
-removing ``unknowable'' nodes (Fig.~\ref{fig-simplification}B) and collapsing
+removing ``unknowable'' nodes such as those in singly-connected graph components (Fig.~\ref{fig-simplification}B) and collapsing
 multiple recombinations into a single child or multiple coalescences into a
@@
 position can only be inferred approximately, depending on how much information
-can be elucidated from mutations in the surrounding genomic region. Because a gARG
-can encode a diversity of ARG structures, it
-allows this fundamental uncertainty in inference to be appropriately represented
-(Appendix~\ref{sec-precision}).
+can be elucidated from mutations in the surrounding genomic region. 
+A gARG can encode a diversity of ARG structures, including 
+those where events \emph{are} recorded explicitly, and those where
+they are treated as fundamentally uncertain and thus not explicitly inferred (Appendix~\ref{sec-precision}).
+% Because a gARG
+% can encode a diversity of ARG structures, it
+% allows this fundamental uncertainty in inference to be appropriately represented
+% (Appendix~\ref{sec-precision})
 
@@
 \label{sec-efficiency}
-The gARG encoding leads to highly efficient storage and processing of ARG data,
+The gARG encoding can lead to highly efficient storage and processing of ARG data,
 and has been in use for several years.
@@
 making a single, deterministic 
-best-guess~\citep{kelleher2019inferring,speidel2019method,zhang2023biobank,zhan2023towards}.
+best-guess~\citep{kelleher2019inferring,speidel2019method,zhang2023biobank,zhan2023towards} 
+(but see \citep{deng2024robust} for recent
+developments in capturing uncertainty using a Bayesian framework, for relatively small
+sample sizes).
 Even under strict parsimony conditions and for small sample sizes the

### 2024-06-03 paper.tex

@@
 for a recombination event). This approach 
-% assumes all events are \emph{knowable}, and does not
-necessitates that all events are recorded explicitly, and does not 
+requires all events to be recorded explicitly, and does not 
 provide an obvious mechanism for either aggregating multiple events
-or expressing uncertainty about them. While this is not a
-problem when describing the results of simulations, for instance (where all details
-are perfectly known), it is an issue when we wish to
-formally describe the output of inference methods which do not
-necessarily attempt to infer events that are not \emph{knowable} from the data, particularly
-as datasets approach the population scale~\citep[e.g.][]{
+or expressing uncertainty about them. This is not a
+problem when describing the results of simulations, where all details
+are perfectly known. However, it can be an issue when we wish to
+formally describe the output of various inference methods, particularly
+those that avoid inferring events that are not \emph{knowable} from the data:
+a useful approach as datasets approach the population scale~\citep[e.g.][]{
 turnbull2018hundred, bycroft2018genome,hayes20191000,
 Ros-Freixedes2020,karczewski2020mutational,tanjo2021practical,
-halldorsson2022sequences}.
+halldorsson2022sequences}).
 

### 2024-06-03 paper.tex
Update paper.tex

@@
 Ros-Freixedes2020,karczewski2020mutational,tanjo2021practical,
-halldorsson2022sequences}).
+halldorsson2022sequences}.
 

### 2024-06-04 paper.tex
Update "probabilistic sampling" bit

@@
 requires all events to be recorded explicitly, and does not 
-provide an obvious mechanism for either aggregating multiple events
-or expressing uncertainty about them. This is not a
-problem when describing the results of simulations, where all details
-are perfectly known. However, it can be an issue when we wish to
-formally describe the output of various inference methods, particularly
-those that avoid inferring events that are not \emph{knowable} from the data:
-a useful approach as datasets approach the population scale~\citep[e.g.][]{
+provide an obvious mechanism for aggregating multiple, potentially
+unresolvable, events.
+As datasets approach the population scale~\citep[e.g.][]{
 turnbull2018hundred, bycroft2018genome,hayes20191000,
 Ros-Freixedes2020,karczewski2020mutational,tanjo2021practical,
-halldorsson2022sequences}.
+halldorsson2022sequences} representing such uncertainty 
+directly through the data structure is a useful alternative to 
+classical methods based on probabilisitic sampling.
 
@@
 to handle the mundane and time-consuming details of data interchange.
-The \texttt{tskit} library (Section~\ref{sec-efficiency})
+The \texttt{tskit} library 
 is a high-quality open-source gARG implementation,

### 2024-06-04 paper.tex
Fix typo

@@
 directly through the data structure is a useful alternative to 
-classical methods based on probabilisitic sampling.
+classical methods based on probabilistic sampling.
 

### 2024-06-04 response-to-reviewers-2.tex
Minor tidy-ups and blandifications

@@
 \begin{point}
-My remaining broad concern is that the paper is still in places somewhat narrow about the goals of future ARG development.
-I certainly see the practical utility of dropping inference down to some minimum "knowable" structure that can be reconstructed using deterministic algorithms for very large datasets.
-However, probabilistic reconstructions of some form of ARG with more explicit events is also a reasonable goal moving forwards (e.g.
-for some applications we may want a subset of the recombination events explicitly included).
-There are a few places where the paper still comes across as overly dogmatic about the minimum "knowable" ARG being the only goal (although the discussion casts a broader view).
+My remaining broad concern is that the paper is still in places somewhat narrow
+about the goals of future ARG development. I certainly see the practical
+utility of dropping inference down to some minimum ``knowable'' structure that
+can be reconstructed using deterministic algorithms for very large datasets.
+However, probabilistic reconstructions of some form of ARG with more explicit
+events is also a reasonable goal moving forwards (e.g. for some applications we
+may want a subset of the recombination events explicitly included). There are a
+few places where the paper still comes across as overly dogmatic about the
+minimum ``knowable'' ARG being the only goal (although the discussion casts a
+broader view).
 \end{point}
 \begin{reply}
-We have gone through the article and, in addition to the suggestions made below, have rephrased 
-parts to make it clear that a gARG can be used to encode a \emph{variety} of ARG structures, whether events are or are not explicitly inferred by the reconstruction method.
-We specifically state at the end of \emph{A diversity of structures} that
+We have gone through the article and, in addition to the suggestions made
+below, have rephrased parts to make it clear that a gARG can be used to encode
+a \emph{variety} of ARG structures, whether events are or are not explicitly
+inferred by the reconstruction method. We specifically state at the end of
+\emph{A diversity of structures} that
 \begin{quote}
@@
 \begin{point}
-Abstract: "This approach is out of step with modern developments, which do not represent genetic inheritance in terms of these events or explicitly infer them." So this is on the places where I feel like the authors state things too strongly.
-The authors, and some others, approaches have taken this path, but folks can agree that the gARG is a good idea and yet think that explicitly inferring details of recombination events is a `modern' goal.
+Abstract: ``This approach is out of step with modern developments, which do not
+represent genetic inheritance in terms of these events or explicitly infer
+them.'' So this is on the places where I feel like the authors state things too
+strongly. The authors, and some others, approaches have taken this path, but
+folks can agree that the gARG is a good idea and yet think that explicitly
+inferring details of recombination events is a `modern' goal.
 \end{point}
 \begin{reply}
-We have changed this to "This approach is out of step with many modern developments, however,..."
+We have changed this to ``This approach is out of step with some modern developments, 
+however,...''
 \end{reply}
@@
 \begin{point}
-"Broadly speaking, an ARG describes the different paths of genetic inheritance caused by recombination, encapsulating the resulting complex web of genetic ancestry " - add "of a set of samples".
-Also I'd say "genetic ancestors", as ancestry is tied up with genetic ancestry groups in peoples' minds.
+``Broadly speaking, an ARG describes the different paths of genetic inheritance
+caused by recombination, encapsulating the resulting complex web of genetic
+ancestry'' - add ``of a set of samples''. Also I'd say ``genetic ancestors'', as
+ancestry is tied up with genetic ancestry groups in peoples' minds.
 \end{point}
@@
 \begin{point}
-"We define a genome as the complete set of genetic material that a child inherits from one parent.
-A diploid individual therefore carries two genomes, one inherited from each parent (we assume diploids here for clarity, but the definitions apply to organisms of arbitrary ploidy).
-" -Excludes Y, mtDNA, and X as written, please revise, e.g.
-talk about autosomal genome.
+``We define a genome as the complete set of genetic material that a child
+inherits from one parent. A diploid individual therefore carries two genomes,
+one inherited from each parent (we assume diploids here for clarity, but the
+definitions apply to organisms of arbitrary ploidy). '' -Excludes Y, mtDNA, and
+X as written, please revise, e.g. talk about autosomal genome.
 \end{point}
@@
 \begin{point}
-"The topology of a gARG specifies that genetic inheritance occurred between particular ancestors and descendants, " -struggle slightly with word "particular" here as the identity of the ancestors is not known.
-Deleting "particular" is likely sufficient.
+``The topology of a gARG specifies that genetic inheritance occurred between
+particular ancestors and descendants, '' -struggle slightly with word
+``particular" here as the identity of the ancestors is not known. Deleting
+``particular" is likely sufficient.
 \end{point}
@@
 \begin{point}
-"This is sufficient to describe the effects of inheritance under any form of homologous recombination (such as multiple crossovers,..." -do you mean multiple crossovers during a single round of meiosis.
+``This is sufficient to describe the effects of inheritance under any form of
+homologous recombination (such as multiple crossovers,..." -do you mean
+multiple crossovers during a single round of meiosis.
 \end{point}
@@
 \begin{point}
-"In this encoding there are two types of internal node in the graph, representing the common ancestor and recombination events in the history of a sample.
-" stipulate that these are most recent common ancestor events.
+``In this encoding there are two types of internal node in the graph,
+representing the common ancestor and recombination events in the history of a
+sample. " stipulate that these are most recent common ancestor events.
 \end{point}
@@
 \begin{point}
-"This approach assumes all events are knowable, and does not provide an obvious mechanism for either aggregating multiple events or expressing uncertainty about them.
-While this is not a problem when describing the results of simulations".
--Maybe one way to flip this around would be to say that because it arose from tracking a particular stochastic process it has these properties.
-Also I don't think it assumes that all events are knowable, eg we could construct some parsimonious ARG or probabilistic ARG.
-If we wish to express uncertainty about events we usually give draws from the posterior etc.
-I agree that might be computational prohibitive with large samples etc, but it seems like place to take a broad view.
-This seems like a place to acknowledge that for some applications we might want to explicitly reconstruct the events.
+``This approach assumes all events are knowable, and does not provide an obvious
+mechanism for either aggregating multiple events or expressing uncertainty
+about them. While this is not a problem when describing the results of
+simulations''. -Maybe one way to flip this around would be to say that because
+it arose from tracking a particular stochastic process it has these properties.
+Also I don't think it assumes that all events are knowable, eg we could
+construct some parsimonious ARG or probabilistic ARG. If we wish to express
+uncertainty about events we usually give draws from the posterior etc. I agree
+that might be computational prohibitive with large samples etc, but it seems
+like place to take a broad view. This seems like a place to acknowledge that
+for some applications we might want to explicitly reconstruct the events.
 \end{point}
@@
 \begin{quote}
-This approach necessitates that all events are recorded explicitly, and does not 
-provide an obvious mechanism for either aggregating multiple events
-or expressing uncertainty about them. While this is not a
-problem when describing the results of simulations, for instance (where all details
-are perfectly known), it is an issue when we wish to
-formally describe the output of inference methods which do not
-necessarily attempt to infer events that are not \emph{knowable} from the data, 
-particularly as datasets approach the population scale...
 \end{quote}
+This approach 
+requires all events to be recorded explicitly, and does not 
+provide an obvious mechanism for aggregating multiple, potentially
+unresolvable, events.
+As datasets approach the population scale [citations]
+representing such uncertainty 
+directly through the data structure is a useful alternative to 
+classical methods based on probabilistic sampling.
 \end{reply}
@@
 \begin{point}
-"A key feature of the gARG encoding is that it enables these varying levels of precision to be represented, and brings these nuanced features to light." -the word nuanced feels strange here.
+``A key feature of the gARG encoding is that it enables these varying levels of
+precision to be represented, and brings these nuanced features to light." -the
+word nuanced feels strange here.
 \end{point}
@@
 \begin{point}
-"Simpler representations can be formed by removing "unknowable" nodes (Fig. 5B)" -unknowable is vague here, do you mean bubbles along a single lineage?
+``Simpler representations can be formed by removing ``unknowable" nodes (Fig.
+5B)" -unknowable is vague here, do you mean bubbles along a single lineage?
 \end{point}
 \begin{reply}
-We've added a clarification that this refers to nodes such as those in singly-connected graph components.
+We've added a clarification that this refers to nodes such as those 
+in singly-connected graph components.
 \end{reply}
@@
 \begin{point}
-"The gARG encoding leads to highly efficient storage and processing of ARG data, "-As gARG has various levels of precision, perhaps this needs to state that the "gARG encoding can lead to..." or be more precise that this is a reduced precision level.
+``The gARG encoding leads to highly efficient storage and processing of ARG
+data, "-As gARG has various levels of precision, perhaps this needs to state
+that the "gARG encoding can lead to..." or be more precise that this is a
+reduced precision level.
 \end{point}
 \begin{reply}
-Amended as suggested to add "can lead to".
+Amended as suggested to add ``can lead to".
 \end{reply}
@@
 \begin{point}
-"The succinct tree sequence data structure (usually known as a "tree sequence" for brevity) is a practical gARG implementation focused on efficiency." - If the tree sequence is focused at a particular level of gARG simplification be precise about this.
+``The succinct tree sequence data structure (usually known as a ``tree sequence"
+for brevity) is a practical gARG implementation focused on efficiency." - If
+the tree sequence is focused at a particular level of gARG simplification be
+precise about this.
 \end{point}
 \begin{reply}
-We have left this sentence as is, since the tree sequence structure can record gARGs at various levels of simplification.
+We have left this sentence as is, since the tree sequence structure 
+can record gARGs at various levels of simplification.
 \end{reply}
@@
 \begin{point}
-"Methods targeting large-scale datasets tend to simplify the inference problem by making a single, deterministic best-guess " --I think this is the best guess of the topology, and the uncertainty in times given the ARG is downstream of this.
-If so please clarify.
-Also I'd perhaps explicitly acknowledge Deng et al (SINGER), e.g.
-"deterministic best-guess of the topology (see Deng et al for parallel developments addressing uncertainty with somewhat small sample sizes)" or something like that.
-While these deterministic approaches are a strong way forward for human biobank scale data, it's good to be highlight parallel developments that might be key to other applications.
+``Methods targeting large-scale datasets tend to simplify the inference problem
+by making a single, deterministic best-guess " --I think this is the best guess
+of the topology, and the uncertainty in times given the ARG is downstream of
+this. If so please clarify. Also I'd perhaps explicitly acknowledge Deng et al
+(SINGER), e.g. ``deterministic best-guess of the topology (see Deng et al for
+parallel developments addressing uncertainty with somewhat small sample sizes)"
+or something like that. While these deterministic approaches are a strong way
+forward for human biobank scale data, it's good to be highlight parallel
+developments that might be key to other applications.
 \end{point}

### 2024-06-04 cover-letter/cover-letter-resubmit-2.tex
Add cover letter

--- 2024-04-18 cover-letter/cover-letter-resubmit.tex
@@
 manuscript entitled
-\emph{A general and efficient representation of ancestral recombination graphs}
-for consideration for publication as a Perspective.
+\emph{A general and efficient representation of ancestral recombination
+graphs}.
 
-We are deeply grateful for your detailed feedback, which has allowed us 
-to identify our target audience and clarified the purpose of this article.
-We are absolutely in agreement that it is more suited to the Perspective
-format, and are very happy for it to be considered potentially appropriate
-for publication in Genetics.
+We are delighed that it is potentially suitable for publication in GENETICS,
+and have endeavored to address the points you have raised.
 
-We have heavily revised the manuscript, substantially shortening it
-and moving some more technical and descriptive section to appendices.
 We have attached a detailed point-by-point response in the 
-\texttt{response-to-reviewers.pdf} file, along with a 
+\texttt{response-to-reviewers-2.pdf} file, along with a 
 latex-diff of the differences between the current and previous submissions.
 
-Given your influence on the manuscript, we have included you in the
-Acknowledgements, and thank you again for taking the time to give
-such thoughtful and detailed feedback.
+Thank you again for your careful and helpful input throughout this process.
 

### 2024-06-04 paper.tex

@@
 coalescence and recombination events to the nodes in an ARG.
-This approach is out of step with many modern developments, however,
+This approach is out of step with some modern developments, however,
 which do not represent genetic inheritance in terms of these events 
@@
 recording events and their properties (e.g.\ the crossover breakpoint
-for a recombination event). This approach 
+for a recombination event). 
+This approach 
 requires all events to be recorded explicitly, and does not 

### 2024-06-07 paper.tex
Move data availability
Add funding and Conflict of interest section
Remove supplementary figure

@@
 
+\section*{Data Availability}
+All code used to generate figures and run analyses is available on GitHub at
+\url{https://github.com/tskit-dev/what-is-an-arg-paper}.
+
 \section*{Acknowledgements}
@@
 on an earlier version of the manuscript.
+
+\section*{Funding}
 % Ordered alphabetically by surname
@@
 
-\section*{Data Availability}
-All code used to generate figures and run analyses is available on GitHub at
-\url{https://github.com/tskit-dev/what-is-an-arg-paper}.
+\section*{Conflicts of Interest}
+The authors declare that there is no conflict of interest.
 
@@
 
-\clearpage
-\renewcommand\thefigure{S\arabic{figure}}
-\setcounter{figure}{0}
-\renewcommand\thetable{S\arabic{table}}
-\setcounter{table}{0}
-
-\section*{Supplementary Material}
-
-\begin{figure*}[ht]
-	\begin{center}
-		\includegraphics[width=0.8\textwidth]{illustrations/simplification-with-edges.pdf}
-	\end{center}
-	\caption{\label{fig-simplification-with-edges}
-	Example ARGs Fig.~\ref{fig-simplification}A--D,
-    with edges annotated with inheritance intervals.
-	}
-\end{figure*}
+% \clearpage
+% \renewcommand\thefigure{S\arabic{figure}}
+% \setcounter{figure}{0}
+% \renewcommand\thetable{S\arabic{table}}
+% \setcounter{table}{0}
+
+% \section*{Supplementary Material}
+
+% \begin{figure*}[ht]
+% 	\begin{center}
+% 		\includegraphics[width=0.8\textwidth]{illustrations/simplification-with-edges.pdf}
+% 	\end{center}
+% 	\caption{\label{fig-simplification-with-edges}
+% 	Example ARGs Fig.~\ref{fig-simplification}A--D,
+%     with edges annotated with inheritance intervals.
+% 	}
+% \end{figure*}
 

### 2024-06-07 paper.tex
Hardcode S1 fig reference

@@
 In the interest of visual clarity, inheritance intervals are not shown
-on the graph edges; Supplementary Fig.~\ref{fig-simplification-with-edges}
+on the graph edges; Supplementary Fig.~S1
 shows the graphs with these inheritance intervals included.

### 2024-06-10 paper.tex
Add postcodes

@@
 \affil[1]{Big Data Institute, Li Ka Shing Centre for Health Information and Discovery, University of Oxford, OX3 7LF, UK}
-\affil[2]{School of Mathematics and Statistics, University of Glasgow, UK}
-\affil[3]{Department of Statistics, University of Oxford, UK}
-\affil[4]{School of Mathematics, Statistics and Physics, Newcastle University, UK}
-\affil[5]{Department of Statistics, University of Warwick, UK}
-\affil[6]{The Roslin Institute and Royal (Dick) School of Veterinary Studies, University of Edinburgh, UK}
-\affil[7]{Broad Institute of MIT and Harvard, Cambridge, USA}
-\affil[8]{Department of Genetics, Stanford University School of Medicine, Stanford, USA}
+\affil[2]{School of Mathematics and Statistics, University of Glasgow, G12 8TA, UK}
+\affil[3]{Department of Statistics, University of Oxford, OX1 3LB, UK}
+\affil[4]{School of Mathematics, Statistics and Physics, Newcastle University, NE1 7RU, UK}
+\affil[5]{Department of Statistics, University of Warwick, CV4 7AL, UK}
+\affil[6]{The Roslin Institute and Royal (Dick) School of Veterinary Studies, University of Edinburgh, EH25 9RG, UK}
+\affil[7]{Broad Institute of MIT and Harvard, Cambridge, MA 02142, USA}
+\affil[8]{Department of Genetics, Stanford University School of Medicine, Stanford, CA 94305-5101, USA}
 

### 2024-06-26 paper.tex
Update node labels
And change some captions to avoid colour mentions

@@
 \usepackage{url}
+
+% Uncommment to extract tikz pictures
+% \usetikzlibrary{external}
+% \tikzexternalize
 
@@
 
-\newcommand{\noderef}[1]{\textsf{#1}}
+\newcommand{\noderef}[1]{\textit{#1}}
 \newcommand{\tsinfer}[0]{\texttt{tsinfer}}
@@
 caused by recombination, encapsulating the resulting complex web of genetic
-ancestors of a set of samples.
+ancestors of a set of sampled genomes.
 Recent breakthroughs
@@
 An example genome ARG (gARG) embedded in a pedigree.
-(A) Diploid individuals (blue), visualised in a highly inbred pedigree and
-labelled $D_1$ to $D_8$,
+\textbf{a.} Diploid individuals (shaded background squares / circles), visualised in a highly inbred pedigree and
+labelled $d_1$ to $d_8$,
 contain both paternal and maternal  genomes
-labelled \noderef{a} to \noderef{p}. Black lines show inheritance paths connecting
-genomes in the current generation (\noderef{a} to \noderef{d}) with their ancestors.
-Genomes \noderef{a} and \noderef{c} are the product of two independent
-meioses (recombination events, in red) between
-the paternal genomes \noderef{e}
-and \noderef{f}, and regions of genome inherited are shown with shaded colour.
-Genomes are shaded such that where, backwards in time,
-they merge into a common ancestor, the merged region is darker.
-(B) The corresponding gARG along with inheritance annotations on all edges
-(partial inheritance in bold).
-(C) The corresponding local trees.
+labelled \noderef{A} to \noderef{P}. Lines show inheritance paths connecting
+genomes in the current generation (\noderef{A} to \noderef{D}) with their ancestors.
+Genomes \noderef{A} and \noderef{C} are the product of two independent
+meioses (recombination events, with italicised breakpoint positions) between
+the paternal genomes \noderef{E}
+and \noderef{F}; the regions of inherited genome are shaded.
+Further shading highlights genomic regions that merge into a common ancestor:
+merged regions gradually darken until fully coalesced.
+\textbf{b.} The corresponding gARG along with inheritance annotations on all edges
+(partial genomic inheritance in bold).
+\textbf{c.} The corresponding local trees.
 }
@@
 Here, and throughout,
-nodes are labelled with lowercase alphabetical letters
+nodes are labelled with uppercase alphabetical letters
 rather than integer identifiers to avoid confusion with genomic intervals.
-Thus individual $D_1$ is composed
-of genomes \noderef{a} and \noderef{b}, which are inherited from its
-two parents $D_3$ and $D_4$. Each inherited genome may be the recombined product
+Thus individual $d_1$ is composed
+of genomes \noderef{A} and \noderef{B}, which are inherited from its
+two parents $d_3$ and $d_4$. Each inherited genome may be the recombined product
 of the two genomes belonging to an individual parent.
 In this example,
-genome \noderef{b} was inherited directly from $D_4$'s genome \noderef{g} without
+genome \noderef{B} was inherited directly from $d_4$'s genome \noderef{G} without
 recombination, whereas
-genome \noderef{a} is the recombinant product of
-$D_2$'s genomes \noderef{e} and \noderef{f} crossing over at position 2.
-Specifically, genome \noderef{a} inherited the (half-closed)
-interval $[0, 2)$ from genome \noderef{e} and $[2, 10)$ from genome \noderef{f}.
+genome \noderef{A} is the recombinant product of
+$d_2$'s genomes \noderef{E} and \noderef{F} crossing over at position 2.
+Specifically, genome \noderef{A} inherited the (half-closed)
+interval $[0, 2)$ from genome \noderef{E} and $[2, 10)$ from genome \noderef{F}.
 These intervals are shown attached to the corresponding graph edges.
 The figure shows the annotated pedigree with realised inheritance of genomes
-between generations (A), the corresponding gARG (B), and finally the corresponding
+between generations (a), the corresponding gARG (b), and finally the corresponding
 sequence of local trees along the
-genome (C).
+genome (c).
 The local trees span the three genome regions delineated
@@
 position~\citep{griffiths1996ancestral}. Fig.~\ref{fig-event-arg} shows an example of an
-eARG with three sample genomes (\noderef{a}, \noderef{b}, and \noderef{c}),
-three common ancestor events (\noderef{e}, \noderef{f}, and \noderef{g})
-and a single recombination event (node \noderef{d}) with a breakpoint
+eARG with three sample genomes (\noderef{A}, \noderef{B}, and \noderef{C}),
+three common ancestor events (\noderef{E}, \noderef{F}, and \noderef{G})
+and a single recombination event (node \noderef{D}) with a breakpoint
 at position $x$.
@@
 is required to distinguish the left and right parents.
-We assume in Fig.~\ref{fig-event-arg} that \noderef{d} inherits genetic material to the
-left of $x$ from \noderef{e} and to the right of $x$ from \noderef{f}.
+We assume in Fig.~\ref{fig-event-arg} that \noderef{D} inherits genetic material to the
+left of $x$ from \noderef{E} and to the right of $x$ from \noderef{F}.
 
@@
 \centering
-\tikzmath{\x1 =0; \x2=8;\xx=12.5; \x3=14; \xt=18;}
+\tikzmath{\x1 =0; \x2=8.5;\xx=12.5; \x3=13.5; \xt=18;}
 \begin{tikzpicture}[x=5mm, y=5mm, node distance=2mm and 20mm]
@@
 
-\node [anchor=north west] at (\x1,6) {A};
-\node [anchor=north west] at (\x2,6) {B};
+\node [anchor=north west] at (\x1-.5,5.5) {\textsf{\textbf{\LARGE a}}};
+\node [anchor=north west] at (\x2-.5,5.5) {\textsf{\textbf{\LARGE b}}};
 % \node [anchor=north west] at (\xt,6) {C};
 
-%%% (A) ARG
+%%% (a) ARG
 
@@
 
-%%% (B) Trees
+%%% (b) Trees
 \node (l0) [greynode] at (\x2 + 0, 0) {};
@@
 \foreach \u/\lab in {
-        s0/$\noderef{a}$, s1/$\noderef{b}$, s2/$\noderef{c}$,
-        r0/$\noderef{a}$, r1/$\noderef{b}$, r2/$\noderef{c}$,
-        l0/$\noderef{a}$, l1/$\noderef{b}$, l2/$\noderef{c}$}
+        s0/$\noderef{A}$, s1/$\noderef{B}$, s2/$\noderef{B}$,
+        r0/$\noderef{A}$, r1/$\noderef{B}$, r2/$\noderef{B}$,
+        l0/$\noderef{A}$, l1/$\noderef{B}$, l2/$\noderef{B}$}
     \node[nodelabel,anchor=south] at ([yshift=-12pt]\u) {\lab};
@@
 \foreach \u/\lab in {
-        s6/$\noderef{g}$,
-        l4/$\noderef{g}$,
-        r4/$\noderef{g}$}
+        s6/$\noderef{G}$,
+        l4/$\noderef{G}$,
+        r4/$\noderef{G}$}
     \node[nodelabel,anchor=south] at (\u) {\lab};
@@
 \foreach \u/\lab in {
-        s4/$\noderef{e}$,
-        l3/$\noderef{e}$}
+        s4/$\noderef{E}$,
+        l3/$\noderef{E}$}
     \node[nodelabel,anchor=south west] at (\u) {\lab};
 \foreach \u/\lab in {
-        s5/$\noderef{f}$,
-        r3/$\noderef{f}$}
+        s5/$\noderef{F}$,
+        r3/$\noderef{F}$}
     \node[nodelabel,anchor=south east] at (\u) {\lab};
-\foreach \u/\lab in {s3/$\noderef{d}$, s6/$\noderef{g}$}
+\foreach \u/\lab in {s3/$\noderef{D}$, s6/$\noderef{G}$}
     \node[nodelabel,anchor=south] at (\u) {\lab};
@@
 
-% %%% (C) Encoding
+% %%% (c) Encoding
 % \node [nodelabel,anchor=north west] at ($(\xt,5)$) {
@@
 % \hline
-% $\noderef{a}$ & $\varnothing$ & [\noderef{e}]\\
-% $\noderef{b}$ & $\varnothing$ & [\noderef{d}] \\
-% $\noderef{c}$ & $\varnothing$ & [\noderef{f}]\\
-% $\noderef{d}$ & $x$ & [\noderef{e}, \noderef{f}] \\
-% $\noderef{e}$ & $\varnothing$ & [\noderef{g}]\\
-% $\noderef{f}$ & $\varnothing$ & [\noderef{g}]\\
-% $\noderef{g}$ & $\varnothing$ & []\\
+% $\noderef{A}$ & $\varnothing$ & [\noderef{E}]\\
+% $\noderef{B}$ & $\varnothing$ & [\noderef{D}] \\
+% $\noderef{C}$ & $\varnothing$ & [\noderef{F}]\\
+% $\noderef{D}$ & $x$ & [\noderef{E}, \noderef{F}] \\
+% $\noderef{E}$ & $\varnothing$ & [\noderef{G}]\\
+% $\noderef{F}$ & $\varnothing$ & [\noderef{G}]\\
+% $\noderef{G}$ & $\varnothing$ & []\\
 % \end{tabular}};
@@
 \caption{\label{fig-event-arg}
-A classical event ARG (eARG). (A) Standard graph depiction with
-breakpoint $x$ associated with the recombination node \noderef{d}.
-Nodes \noderef{e}, \noderef{f} and \noderef{g} are common ancestor events.
-(B) Corresponding local trees to the left and right of breakpoint $x$
+A classical event ARG (eARG). \textbf{a.} Standard graph depiction with
+breakpoint $x$ associated with the recombination node \noderef{D}.
+Nodes \noderef{E}, \noderef{F} and \noderef{G} are common ancestor events.
+\textbf{b.} Corresponding local trees to the left and right of breakpoint $x$
 (note these are shown in the conventional form in which only coalescences
-within the local tree are included; see Appendix~\ref{sec-ARG-and-local-trees}
-for a discussion of this important point).
+within the local tree are included, hence \noderef{D} is omitted; see
+Appendix~\ref{sec-ARG-and-local-trees} for a discussion of this important point).
 }
@@
 For example, in Fig.~\ref{fig-arg-in-pedigree} we have
-four sample genomes, \noderef{a}--\noderef{d}. As we
+four sample genomes, \noderef{A}--\noderef{D}. As we
 trace their genetic ancestry into the previous generation
-(\noderef{e}--\noderef{h}), we can think of their ancestral
+(\noderef{E}--\noderef{H}), we can think of their ancestral
 material propagating through the graph
 pastwards. In the region $[2, 7)$, there is a
-local coalescence where nodes \noderef{a} and \noderef{c}
-find a common ancestor in \noderef{f}. Thus, in this region,
+local coalescence where nodes \noderef{A} and \noderef{C}
+find a common ancestor in \noderef{F}. Thus, in this region,
 we have three genome segments that are ancestral to the
-four samples. Fig.~\ref{fig-arg-in-pedigree}A
+four samples. Fig.~\ref{fig-arg-in-pedigree}a
 illustrates this by (shaded) ancestral material being present
-in only three nodes (\noderef{f}, \noderef{g}, and \noderef{h}) in this region,
-while node \noderef{e} is blank
+in only three nodes (\noderef{F}, \noderef{G}, and \noderef{H}) in this region,
+while node \noderef{E} is blank
 as it carries \emph{non-ancestral} material.
@@
 graph, until all samples reach their most recent common
-ancestor in node \noderef{n}.
+ancestor in node \noderef{N}.
 
@@
 Converting the \citet[][Fig.~1]{wiuf1999recombination} example
-to a sample-resolved gARG. (A) The original eARG; square nodes
-represent sampling (black), common ancestor (blue), and recombination (red) events;
-the latter contain breakpoint positions.
-(B) The corresponding gARG with breakpoints directly converted to
+to a sample-resolved gARG. \textbf{a.} The original eARG; nodes
+represent sampling, common ancestor, and recombination events (small shaded, small blue, and large red rectangles respectively); the latter contain breakpoint positions.
+\textbf{b.} The corresponding gARG with breakpoints directly converted to
 edges annotated with inheritance intervals.
-(C) The sample-resolved gARG resulting from simplifying with respect
-to the sample genomes, \noderef{a}, \noderef{b}, and \noderef{c}.
+\textbf{c.} The sample-resolved gARG resulting from simplifying with respect
+to the sample genomes, \noderef{A}, \noderef{B}, and \noderef{C}.
 Dashed lines show edges that are
-no longer present (in practice, nodes \noderef{g}, \noderef{j}, and \noderef{q} would also be removed).
+no longer present (in practice, nodes \noderef{G}, \noderef{J}, and \noderef{Q} would also be removed).
 Coalescence with respect to the sample is indicated by shaded bars, as
-in Fig.~\ref{fig-arg-in-pedigree}A; nodes \noderef{n}, \noderef{o}, \noderef{p}, \noderef{q} have truncated
+in Fig.~\ref{fig-arg-in-pedigree}a; nodes \noderef{N}, \noderef{O}, \noderef{P}, \noderef{Q} have truncated
 bars showing that local ancestry of entirely coalesced regions is omitted.
@@
 via a two-step process illustrated in Fig.~\ref{fig-ancestry-resolution}.
-The first step is to take the input eARG (Fig.~\ref{fig-ancestry-resolution}A),
+The first step is to take the input eARG (Fig.~\ref{fig-ancestry-resolution}a),
 duplicate its graph topology, and then add inheritance annotations
-to each of the gARG's edges (Fig.~\ref{fig-ancestry-resolution}B) as follows.
+to each of the gARG's edges (Fig.~\ref{fig-ancestry-resolution}b) as follows.
 If a given node is a common ancestor event, we annotate the single
@@
 Thus, the final step is to use the ``simplify'' algorithm to resolve the
-ancestral material of the samples (Fig.~\ref{fig-ancestry-resolution}C).
-
-The sample-resolved gARG of Fig.~\ref{fig-ancestry-resolution}C
+ancestral material of the samples (Fig.~\ref{fig-ancestry-resolution}c).
+
+The sample-resolved gARG of Fig.~\ref{fig-ancestry-resolution}c
 differs in some important ways to the
-original eARG (Fig.~\ref{fig-ancestry-resolution}A).
+original eARG (Fig.~\ref{fig-ancestry-resolution}a).
 Firstly, we can see that some nodes and edges have been removed entirely
 from the graph.
-The ``grand MRCA'' \noderef{q} is omitted from the
+The ``grand MRCA'' \noderef{Q} is omitted from the
 sample-resolved gARG because all segments of the genome have
-fully coalesced in \noderef{k} and \noderef{p} before \noderef{q} is reached.
+fully coalesced in \noderef{K} and \noderef{P} before \noderef{Q} is reached.
 Likewise, the edge
-between \noderef{g} and \noderef{j} is omitted because the recombination
-event at position $5$ (represented by node \noderef{g})
+between \noderef{G} and \noderef{J} is omitted because the recombination
+event at position $5$ (represented by node \noderef{G})
 fell in non-ancestral material.
 More generally, we can see that the sample resolved
-gARG of Fig.~\ref{fig-ancestry-resolution}C
+gARG of Fig.~\ref{fig-ancestry-resolution}c
 allows for ``local'' inspection
@@
 Results for four different methods:
-(A) \kwarg; (B) \argweaver; (C) \tsinfer; and (D) \relate, converted 
+\textbf{a.} \kwarg; \textbf{b.} \argweaver; \textbf{c.} \tsinfer; and \textbf{d.} \relate, converted 
 to the standard \emph{tskit} gARG encoding. See
@@
 on the classical \cite{kreitman1983nucleotide} dataset. 
-The ARGs in Fig.~\ref{fig-inferred-args}A and B are precise 
+The ARGs in Fig.~\ref{fig-inferred-args}a, b are precise 
 estimates (Appendix~\ref{sec-precision}),
@@
 or two parents.
-In contrast the ARGs in Fig.~\ref{fig-inferred-args}C and D do not 
+In contrast the ARGs in Fig.~\ref{fig-inferred-args}c, d do not 
 have this clear-cut interpretation, and the nodes
@@
 Levels of ARG simplification.
-(A) An example gARG simulated from a diploid Wright-Fisher model.
-(B) Remove all
-singly-connected graph components (e.g., diamonds such as \noderef{jlnm}).
-(C) Remove nodes that never represent coalescences,
-i.e.\ are unary everywhere (e.g.\ \noderef{n} and \noderef{r}).
-(D) Rewrite edges to bypass nodes in local trees in which they are unary
+\textbf{a.} An example gARG simulated from a diploid Wright-Fisher model.
+\textbf{b.} Remove all
+singly-connected graph components (e.g., diamonds such as \noderef{JLMN}).
+\textbf{c.} Remove nodes that never represent coalescences,
+i.e.\ are unary everywhere (e.g.\ \noderef{N} and \noderef{R}).
+\textbf{d.} Rewrite edges to bypass nodes in local trees in which they are unary
 (often described as ``fully simplified'').
@@
 recombination events represented by separate genomes, is shown in
-Fig.~\ref{fig-simplification}A. Simpler representations can be formed by
-removing ``unknowable'' nodes such as those in singly-connected graph components (Fig.~\ref{fig-simplification}B) and collapsing
+Fig.~\ref{fig-simplification}a. Simpler representations can be formed by
+removing ``unknowable'' nodes such as those in singly-connected graph components (Fig.~\ref{fig-simplification}b) and collapsing
 multiple recombinations into a single child or multiple coalescences into a
-single parent (Fig.~\ref{fig-simplification}C). 
-Finally, Fig.~\ref{fig-simplification}D is a ``fully simplified''
+single parent (Fig.~\ref{fig-simplification}c). 
+Finally, Fig.~\ref{fig-simplification}d is a ``fully simplified''
 ARG, in which only coalescences in local trees are retained.
@@
 \begin{tikzpicture}
-	\node [anchor=north west] at (-3.5,9.3) {\LARGE \textbf{A}};
+	\node [anchor=north west] at (-3.1,9.3)  {\textsf{\textbf{\huge a}}};
 	\draw (-0.4, -0.2) -- (0.4, -0.2) -- (0.4, -0.3) -- (-0.4, -0.3) -- (-0.4, -0.2);
@@
 	\draw (-1, 1.7) -- (-1, 4.5) -- (0,4.5) -- (0,3.1);
-	\node [scale=0.2,label=above left:{\small \noderef{a}}] at (-0.5,4.5) {a};
+	\node [scale=0.2,label=above left:{\small \noderef{A}}] at (-0.5,4.5) {a};
 	\draw (-0.9, 4.3) -- (-0.6, 4.3) -- (-0.6, 4.2) -- (-0.9, 4.2) -- (-0.9, 4.3);
@@
 	\draw (4,2.4) -- (4,5.4) -- (2,5.4) -- (2,3.1);
-	\node [scale=0.2,label=above left:{\small \noderef{b}}] at (3,5.4) {b};
+	\node [scale=0.2,label=above left:{\small \noderef{B}}] at (3,5.4) {b};
 	\draw (2.6, 5.2) -- (2.9, 5.2) -- (2.9, 5.1) -- (2.6, 5.1) -- (2.6, 5.2);
@@
 	\draw (0.5, 6) -- (0.5, 7.2) -- (3,7.2) -- (3,5.4);
-	\node [scale=0.2,label=above:{\small \noderef{c}}] at (1.75,7.2) {c};
+	\node [scale=0.2,label=above:{\small \noderef{C}}] at (1.75,7.2) {c};
 	\draw (1.35, 6.95) -- (2.15, 6.95);
@@
 	\draw[color=gray, dashed] (7.5, 7.2) -- (11, 7.2);
-\end{tikzpicture}
-&
-\begin{tikzpicture}
-	\node [anchor=north west] at (-3.5,9.3) {\LARGE \textbf{B}};
+
+        \begin{scope}[shift={(15,0)}]
+
+	\node [anchor=north west] at (-3.1,9.3) {\textsf{\textbf{\huge b}}};
 	\draw (-0.4, -0.2) -- (0.4, -0.2) -- (0.4, -0.3) -- (-0.4, -0.3) -- (-0.4, -0.2);
@@
 	\draw[color=gray, dashed] (7.5, 7.2) -- (9.1, 7.2);
+	\end{scope}
 \end{tikzpicture}
@@
 }
-\caption{(A)
+\caption{\textbf{a.}
 A realisation of the graph traversed by Hudson's algorithm started from a
@@
 propagated until time $T$. The MRCA on the genetic interval $[v, w)$ is reached
-at event \noderef{b}, while that on $[0, v)$ is reached at event \noderef{c}.
+at event \noderef{B}, while that on $[0, v)$ is reached at event \noderef{C}.
 The non-ancestral segment $[v, w)$ above
-\noderef{a} contributes to the rate of effective recombinations because it
+\noderef{A} contributes to the rate of effective recombinations because it
 is trapped between ancestral segments. The two columns titled CA and RE
@@
 the recombination rate is $\rho$.
-(B) A corresponding realisation of a Big ARG, which augments Hudson's algorithm
+\textbf{b.} A corresponding realisation of a Big ARG, which augments Hudson's algorithm
 by tracking non-ancestral lineages. The result is a simpler state space and
@@
 Some representations may not provide this information.
-For example, in the gARG encoding depicted in Figure \ref{fig-ancestry-resolution}C,
-it is clear that a recombination takes place between nodes \noderef{i}, \noderef{k} and \noderef{j}.
-But the exact time of the recombination event is ambiguous: it could take place at any time between node \noderef{i} and its parents and produce the same gARG.
+For example, in the gARG encoding depicted in Figure \ref{fig-ancestry-resolution}c,
+it is clear that a recombination takes place between nodes \noderef{I}, \noderef{K} and \noderef{J}.
+But the exact time of the recombination event is ambiguous: it could take place at any time between node \noderef{I} and its parents and produce the same gARG.
 Because a recombination increases the number of extant lineages by one (in the rootward direction of time), the number of lineages during the same time interval is ambiguous as well.
-In fact, this information cannot be recovered from the gARG encoding used in Figure \ref{fig-ancestry-resolution}C without an extrinsic convention.
+In fact, this information cannot be recovered from the gARG encoding used in Figure \ref{fig-ancestry-resolution}c without an extrinsic convention.
 For the basic coalescent with recombination,
-it is sufficient to create \emph{two} gARG nodes at the time of the recombination event, with the interpretation that the two rootward edges from node \noderef{i} in Figure \ref{fig-ancestry-resolution}C belong to the same lineage until the time of nodes \noderef{k} and \noderef{j}, and split into two separate lineages at that time point.
+it is sufficient to create \emph{two} gARG nodes at the time of the recombination event, with the interpretation that the two rootward edges from node \noderef{I} in Figure \ref{fig-ancestry-resolution}c belong to the same lineage until the time of nodes \noderef{K} and \noderef{J}, and split into two separate lineages at that time point.
 Similarly, the trapped, non-ancestral links along that lineage remain available for effective recombination (i.e.\ one which splits up ancestral material) for the same time interval.
@@
 retrospective ancestry paths for four chosen chromosomes (indicated by square black ``sampling events'')
-sampled from 3 diploid individuals ($D_1$, $D_2$, $D_3$) in the current generation.
+sampled from 3 diploid individuals ($d_1$, $d_2$, $d_3$) in the current generation.
 Ancestral recombination events and coalescence events are shown as red and blue squares respectively.
@@
 involves a single recombination event and four coalescence events
-(highlighted as deep red and blue squares within individuals $D_5$, $D_{10}$, and $D_{13}$).
+(highlighted as deep red and blue squares within individuals $d_5$, $d_{10}$, and $d_{13}$).
 ARG lineages also show gametic genomes, contained within shaded circles.
-As in Fig.~\ref{fig-arg-in-pedigree}A, inherited regions within the sampled chromosome arm are
+As in Fig.~\ref{fig-arg-in-pedigree}a, inherited regions within the sampled chromosome arm are
 shaded by the number of descendant samples.
@@
 two node representation seen in
-Fig.~\ref{fig-arg-in-pedigree}A and Fig.~\ref{hudson_vs_bigARG},
+Fig.~\ref{fig-arg-in-pedigree}a and Fig.~\ref{hudson_vs_bigARG},
 whereas tagging the genome
 below a recombination event leads to the more
-conventional graphs in Figs.~\ref{fig-ancestry-resolution} and \ref{fig-inferred-args}A,B.
+conventional graphs in Figs.~\ref{fig-ancestry-resolution} and \ref{fig-inferred-args}a,b.
 
 The schematic illustrates an important point about the biological reality of polytomies.
-Three lineages coalesce in the left-hand genome of individual $D_{10}$, but do
+Three lineages coalesce in the left-hand genome of individual $d_{10}$, but do
 so as the result of two successive bifurcations. This is necessarily so,
@@
 Mutations can occur along any cell lineages. For example,
-a mutation in the first cell division of $D_{10}$ could be shared between
-the two gametes produced by the cells in the left half of $D_{10}$
+a mutation in the first cell division of $d_{10}$ could be shared between
+the two gametes produced by the cells in the left half of $d_{10}$
 but not shared by the right hand gamete.
@@
 the ARG; if this path omits nodes that are not
-branching points (such as \noderef{e} in Fig.~\ref{fig-arg-in-pedigree}),
+branching points (such as \noderef{E} in Fig.~\ref{fig-arg-in-pedigree}),
 information about the ARG is lost.
@@
 For the example in Fig.~\ref{fig-arg-in-pedigree}, nodes
-\noderef{g} and \noderef{h}
-are ARG-unary (Fig.~\ref{fig-arg-in-pedigree}B), and are consequently
-also unary in the local trees (Fig.~\ref{fig-arg-in-pedigree}C).
-On the other hand, node \noderef{f} has two children
+\noderef{G} and \noderef{H}
+are ARG-unary (Fig.~\ref{fig-arg-in-pedigree}b), and are consequently
+also unary in the local trees (Fig.~\ref{fig-arg-in-pedigree}c).
+On the other hand, node \noderef{F} has two children
 in the graph, but is binary only
 in the local tree covering the interval $[2, 7)$,
-representing the coalescence of samples \noderef{a} and \noderef{c}
+representing the coalescence of samples \noderef{A} and \noderef{C}
 in this genome region. Over the interval $[0, 2)$ no coalescence occurs,
-but we still record the fact that genome \noderef{c} inherits from \noderef{f}
-in the local tree. Thus, node \noderef{f} has a single child in this
+but we still record the fact that genome \noderef{C} inherits from \noderef{F}
+in the local tree. Thus, node \noderef{F} has a single child in this
 interval: it is \emph{locally unary}.
-In the same figure, \noderef{e} is binary in the graph, being a common
-ancestor of \noderef{a} and \noderef{c}, but is
+In the same figure, \noderef{E} is binary in the graph, being a common
+ancestor of \noderef{A} and \noderef{C}, but is
 locally unary in all trees in which it is present.
-This is because no ancestral material coalesces in \noderef{e}:
-\noderef{a} inherits genetic material from the far left hand end of
-\noderef{e},
-while \noderef{c} only inherits the (disjoint) right hand end.
+This is because no ancestral material coalesces in \noderef{E}:
+\noderef{A} inherits genetic material from the far left hand end of
+\noderef{E},
+while \noderef{C} only inherits the (disjoint) right hand end.
 
@@
 For example, the local trees for the classical event ARG
-depicted in Fig.~\ref{fig-event-arg}B follow the usual conventions
+depicted in Fig.~\ref{fig-event-arg}b follow the usual conventions
 and do not include any information about the recombination
-that occurred at node \noderef{d}. Given these two local trees
+that occurred at node \noderef{D}. Given these two local trees
 in isolation
@@
 we lack specific information about the recombination.
-Explicitly recording that node \noderef{d} lies on the
-branch joining \noderef{b} to \noderef{e} in the left hand
-tree, and \noderef{b} to \noderef{f} in the right hand tree
+Explicitly recording that node \noderef{D} lies on the
+branch joining \noderef{B} to \noderef{E} in the left hand
+tree, and \noderef{B} to \noderef{F} in the right hand tree
 resolves all ambiguity, and makes the collection of local
@@
 
-The ARG in Fig.~\ref{fig-simplification}A is the output of a
+The ARG in Fig.~\ref{fig-simplification}a is the output of a
 backwards-time Wright-Fisher simulation for a sample of two diploid
@@
 the parent's two genomes.
-Thus, a recombinant node such as \noderef{g} in Fig.~\ref{fig-simplification}
+Thus, a recombinant node such as \noderef{G} in Fig.~\ref{fig-simplification}
 may also correspond to a coalescence.
@@
 the most convenient approach will vary by application.
-Note also that node \noderef{k} in Fig.~\ref{fig-simplification} has three children.
+Note also that node \noderef{K} in Fig.~\ref{fig-simplification} has three children.
 Polytomies like this are a natural feature of such a Wright-Fisher model.
@@
 less influential (contrast the edge
-$(\noderef{g}, \noderef{h})$
-with $(\noderef{g}, \noderef{i})$
-in Fig.~\ref{fig-simplification}A).
+$(\noderef{G}, \noderef{H})$
+with $(\noderef{G}, \noderef{I})$
+in Fig.~\ref{fig-simplification}a).
 Secondly, node colours denote the number of parents that they
@@
 Returning to the main topic of this section,
-Fig.~\ref{fig-simplification}A is the original simulation output, in
+Fig.~\ref{fig-simplification}a is the original simulation output, in
 which we retain all nodes involved in recombination
@@
 redundant (or, from another perspective, unobservable).
-In Fig.~\ref{fig-simplification}A the local trees (right)
+In Fig.~\ref{fig-simplification}a the local trees (right)
 contain many unary nodes, fewer as  we successively
-simplify (Fig.~\ref{fig-simplification}B,C),
-until we reach Fig.~\ref{fig-simplification}D, where there are none.
+simplify (Fig.~\ref{fig-simplification}b,c),
+until we reach Fig.~\ref{fig-simplification}d, where there are none.
 
@@
 in which the two parent nodes of a recombination immediately
-join again into a common ancestor (e.g.~\noderef{j}, \noderef{l}, \noderef{m}
-and \noderef{n} in Fig.~\ref{fig-simplification}A).
+join again into a common ancestor (e.g.~\noderef{J}, \noderef{L}, \noderef{M}
+and \noderef{N} in Fig.~\ref{fig-simplification}a).
 Unless we are specifically
@@
 of a node that has one inbound and one outbound edge, such as
-nodes \noderef{f} and \noderef{h}.
+nodes \noderef{F} and \noderef{H}.
 % Note that f and h are only removable because we are using the 2-RE-node
@@
 % level of detail here though, so we'll just pass over it without comment.
-Fig.~\ref{fig-simplification}B shows the result of this type of
+Fig.~\ref{fig-simplification}b shows the result of this type of
 graph topology simplification.
@@
 of the local trees. In particular, a node can represent a recombinant
-with multiple parents in the graph but only a single child (e.g.\ node \noderef{n}
-in Fig.~\ref{fig-simplification}B), or can represent a common ancestor with
+with multiple parents in the graph but only a single child (e.g.\ node \noderef{N}
+in Fig.~\ref{fig-simplification}b), or can represent a common ancestor with
 multiple children in the graph but in which no coalescence takes place
 in the local trees
-(node \noderef{r} in Fig.~\ref{fig-simplification}B).
+(node \noderef{R} in Fig.~\ref{fig-simplification}b).
 The distinction between the ``common ancestry'' of two or more genomes
@@
 important~\citep{hudson1983testing,kelleher2016efficient}.
-Consider \noderef{e} in Fig.~\ref{fig-simplification}A,
+Consider \noderef{E} in Fig.~\ref{fig-simplification}a,
 for example. We can see from the graph that it is a common
-ancestor of samples \noderef{a} and \noderef{b}, but
+ancestor of samples \noderef{A} and \noderef{B}, but
 it does not correspond to any coalescence in the
@@
 ancestral material associated with the edges.
-As we see in Fig.~\ref{fig-simplification}C,
+As we see in Fig.~\ref{fig-simplification}c,
 removal of recombinant nodes can produce graph nodes with
-more than two parents (e.g.~node \noderef{e}); and likewise, removal of
+more than two parents (e.g.~node \noderef{E}); and likewise, removal of
 common ancestor but non-coalescent nodes can produce graph nodes with
-more than two children (e.g.~node \noderef{s}). Both cases represent the
+more than two children (e.g.~node \noderef{S}). Both cases represent the
 merged
@@
 correspond to a coalescence in at least one other
-local tree. For example, node  \noderef{k} is unary in the fourth tree
-of Fig.~\ref{fig-simplification}C, but is either binary
+local tree. For example, node  \noderef{K} is unary in the fourth tree
+of Fig.~\ref{fig-simplification}c, but is either binary
 or ternary in all other local trees (recall this is a Wright-Fisher
@@
 such that, although no nodes are removed from the graph, all
-unary nodes disappear from the local trees (Fig.~\ref{fig-simplification}D).
+unary nodes disappear from the local trees (Fig.~\ref{fig-simplification}d).
 Note that although this last stage produces simpler local trees, by
@@
 history of the sample that is stored. Some of the information lost,
-e.g.\ ``diamond'' removal (Fig.~\ref{fig-simplification}A),
+e.g.\ ``diamond'' removal (Fig.~\ref{fig-simplification}a),
 seems like a reasonable tradeoff for a simpler structure.
@@
 the only differences between them are their paths through
-nodes \noderef{l} and \noderef{m}. These nodes that participated
+nodes \noderef{L} and \noderef{M}. These nodes that participated
 in the diamond are removed from the ARG, and we have lost
 all information about the corresponding recombination at
-position 61. Other nodes (e.g.\ \noderef{o} and \noderef{p})
+position 61. Other nodes (e.g.\ \noderef{O} and \noderef{P})
 have also been removed but these represent the \emph{parents}
 of recombinants. The recombinant nodes themselves
-(e.g. \noderef{n}) are still present, and represent precise
+(e.g. \noderef{N}) are still present, and represent precise
 information about the time, genomic location and
@@
 
-Fig.~\ref{fig-simplification}C has the same number of local trees
-as Fig.~\ref{fig-simplification}B, but has less precise information
+Fig.~\ref{fig-simplification}c has the same number of local trees
+as Fig.~\ref{fig-simplification}b, but has less precise information
 about recombination. Continuing the previous example, node
-\noderef{n} has been removed from the graph because it was unary
-in all of the local trees; its outbound edges to \noderef{s}
-and \noderef{q} have effectively been ``pushed down''
-to \noderef{e} (which is retained because it is the coalescent
-parent of \noderef{a} and \noderef{b} over the interval
+\noderef{N} has been removed from the graph because it was unary
+in all of the local trees; its outbound edges to \noderef{S}
+and \noderef{Q} have effectively been ``pushed down''
+to \noderef{E} (which is retained because it is the coalescent
+parent of \noderef{A} and \noderef{B} over the interval
 $[44, 100)$). We
@@
 the \emph{timing} of this recombination event, and know only
-that it must have occurred between the times of node \noderef{e}
-and \noderef{q}.
-
-Fig.~\ref{fig-simplification}D removes all unary nodes from the
+that it must have occurred between the times of node \noderef{E}
+and \noderef{Q}.
+
+Fig.~\ref{fig-simplification}d removes all unary nodes from the
 local trees, and further reduces the precision of
-recombination information. Node \noderef{e} has not been
+recombination information. Node \noderef{E} has not been
 removed from the graph because it is coalescent in the
@@
 trees for $[44, 87)$ and $[87, 100)$ were only distinguishable
-by the passage of the former tree through nodes \noderef{e}
-and \noderef{q}, and so the recombination on node \noderef{g}
+by the passage of the former tree through nodes \noderef{E}
+and \noderef{Q}, and so the recombination on node \noderef{G}
 at position 87 has been lost entirely.
@@
 the node colouring scheme, which shows the number of parents for each node.
-In Fig.~\ref{fig-inferred-args}A,B, each recombination event
+In Fig.~\ref{fig-inferred-args}a,b, each recombination event
 corresponds to a node with exactly two parents and one child.
@@
 (brown) nodes.
-In contrast, Fig.~\ref{fig-inferred-args}C,D do not have this
+In contrast, Fig.~\ref{fig-inferred-args}c,d do not have this
 straightforward
@@
 (Appendix~\ref{sec-locally-unary-edges}). For example, brown nodes
-in Fig.~\ref{fig-inferred-args}A,B are clear because there is
+in Fig.~\ref{fig-inferred-args}a,b are clear because there is
 no local coalescence at these recombination nodes
@@
 The significant number of partially shaded nodes in
-Fig.~\ref{fig-inferred-args}A,B and C demonstrates that
+Fig.~\ref{fig-inferred-args}a,b and c demonstrates that
 the \kwarg, \argweaver\ and \tsinfer\ ARGs all contain locally unary nodes.
@@
 bifurcating trees by design, each picks an arbitrary order of branching
-(hence Fig.~\ref{fig-inferred-args}A and B disagree
-in this order, and Fig.~\ref{fig-inferred-args}D even shows different
+(hence Fig.~\ref{fig-inferred-args}a and b disagree
+in this order, and Fig.~\ref{fig-inferred-args}d even shows different
 orders in different trees).
@@
 % won't either. Let's stick to a few clear messages.
-% More generally, the number of nodes in Fig.~\ref{fig-inferred-args}C
+% More generally, the number of nodes in Fig.~\ref{fig-inferred-args}c
 % that have more than two children hints at considerable uncertainty in
@@
 and an ARG with precisely specified recombination events on the other
-(Fig~\ref{fig-inferred-args}A,B).
+(Fig~\ref{fig-inferred-args}a,b).
 % JK: Let's not get into this.
@@
 
-% \clearpage
-% \renewcommand\thefigure{S\arabic{figure}}
-% \setcounter{figure}{0}
-% \renewcommand\thetable{S\arabic{table}}
-% \setcounter{table}{0}
-
-% \section*{Supplementary Material}
-
-% \begin{figure*}[ht]
+%\clearpage
+%\pagenumbering{gobble} 
+%\renewcommand\thefigure{S\arabic{figure}}
+%\setcounter{figure}{0}
+%\renewcommand\thetable{S\arabic{table}}
+%\setcounter{table}{0}
+%\onecolumn
+%\section*{Supplementary Material {\normalfont for ``A general and efficient representation of ancestral recombination graphs'', Wong \textit{et al.} 2024}}
+%\pagestyle{plain}
+%\begin{figure}[ht]
 %	\begin{center}
-% 		\includegraphics[width=0.8\textwidth]{illustrations/simplification-with-edges.pdf}
+%		\includegraphics[width=0.975\textwidth]{illustrations/simplification-with-edges.pdf}
 %	\end{center}
-% 	\caption{\label{fig-simplification-with-edges}
-% 	Example ARGs Fig.~\ref{fig-simplification}A--D,
-%     with edges annotated with inheritance intervals.
+%	\caption{
+%		\label{fig-simplification-with-edges}
+%		Example ARGs from Fig.~\ref{fig-simplification}a--d, with edges annotated with inheritance intervals.
 %	}
-% \end{figure*}
+%\end{figure}
 

### 2024-06-26 argutils/viz.py

@@ -227,24 +227,27 @@ def draw(
             nodelist=np.where(use)[0],
         )
 
-    for font_color, labels in labels_by_colour.items():
-        text = nx.draw_networkx_labels(
-            G,
-            pos,
-            font_color=font_color,
-            font_size=font_size,
-            labels=labels,
-            ax=ax,
-        )
-        if rotated_sample_labels:
-            y_below_extra = 0.0145 * np.diff(ax.get_ylim())  # 2% of the y range
-            for nd, t in text.items():
-                if is_sample[nd]:
-                    x, y = t.get_position()
-                    t.set_rotation(-90)
-                    t.set_position((x, y - y_below_extra))
-                    t.set_va('top')
-    # Assign edge colors using the ordering of edges in the networkx representation
+    with mpl.rc_context({'font.style': 'italic'}):
+        for font_color, labels in labels_by_colour.items():
+            text = nx.draw_networkx_labels(
+                G,
+                pos,
+                font_color=font_color,
+                font_size=font_size,
+                font_family="Georgia",
+                labels=labels,
+                ax=ax,
+                verticalalignment="center_baseline",
+            )
+            if rotated_sample_labels:
+                y_below_extra = 0.0145 * np.diff(ax.get_ylim())  # 2% of the y range
+                for nd, t in text.items():
+                    if is_sample[nd]:
+                        x, y = t.get_position()
+                        t.set_rotation(-90)
+                        t.set_position((x, y - y_below_extra))
+                        t.set_va('top')
+        # Assign edge colors using the ordering of edges in the networkx representation
     if edge_colors is not None:
         ordered_edge_color = []
         for edge_id, edge in enumerate(G.edges()):
@@ -282,7 +285,7 @@ def label_nodes(ts, labels=None):
     will be labelled with numbers rather than ascii uppercase letters.
     """
     if labels is None:
-        labels = {i: lab for i, lab in enumerate(string.ascii_lowercase)}
+        labels = {i: lab for i, lab in enumerate(string.ascii_uppercase)}
     tables = ts.dump_tables()
 
     tables.nodes.clear()
